### 2022-08-29 EvolutionaryRescue1.tex
Add files via upload

\documentclass[11pt,a4paper]{article}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{tikz,pgfplots}
\usepackage{gensymb}
\usepackage{graphicx}

\usepackage{epstopdf}
%\usepackage{listings}
\usepackage{tikz,fp,fp-random,todonotes}
\usepackage[procnames]{listings}
\usepackage{afterpage}
%\usepackage[margin=1in]{geometry}
\usepackage{titlesec}
\usepackage{float}
\usepackage{amsthm}
\usepackage{wrapfig}
\usepackage{booktabs,caption}
\usepackage[flushleft]{threeparttable}
\usepackage{scrextend}
\usepackage{stfloats}
\usepackage{mathtools}
\usepackage{bm}
\usepackage{stackengine}
\usepackage[hidelinks]{hyperref}
\usepackage{siunitx}
\usepackage{subfig}
\usepackage{authblk}
\usepackage[margin=0.7in]{geometry}

\addtokomafont{labelinglabel}{\sffamily}

\usetikzlibrary{calc}
\usetikzlibrary{backgrounds}
\usetikzlibrary{calc}
\usetikzlibrary{decorations.markings}
\usetikzlibrary{arrows}
\usetikzlibrary{positioning}
\usetikzlibrary{decorations.text}

\usetikzlibrary{shadings}
\usetikzlibrary{backgrounds}
\usetikzlibrary{calc}
\usetikzlibrary{decorations.markings}
\usetikzlibrary{arrows}
\usetikzlibrary{positioning}
\usetikzlibrary{decorations.text}

\allowdisplaybreaks
\renewcommand{\floatpagefraction}{.85}

\newcommand{\x}{{\bf x}}
\renewcommand{\d}{{\rm d}}
\newcommand{\e}{{\rm e}}
\newcommand{\ii}{{\rm i}}

\newcommand{\tmi}{\tau_0\wedge\tau}
\newcommand{\tma}{\tau_0\vee\tau}
\newcommand{\taua}{\tau_{\rm A}}

\begin{document}
\title{Evolutionary rescue}
\author[1]{Remus Stana}
\author[1]{Yoav Ram}
\author[2]{‪Daniel Weissman}
\affil[1]{School of Zoology, Tel Aviv University, Tel Aviv-Yafo, Israel}
\affil[2]{Department of Physics, Emory University, Atlanta, United States of America}
\maketitle
\begin{abstract}
\end{abstract}

\section{Introduction}
The question of what role does chromosomal instability (CIN) play in the emergence of cancer has been studied extensively in the past decades \cite{michor2005can,christine2018understanding,nowak2002role,pavelka2010dr,komarova2003mutation,zhu2018cellular}. One hypothesis is that CIN facilitates tumorgenesis by accelerating the removal of tumor suppression genes (TSG) and subsequent appearance of cancer. The deletion of tumor suppression genes can happen in two ways (see Figure \ref{figureCIN}): two point mutations deleting both alleles of the TSG or one point mutation and one chromosomal loss event. The case of two chromosomal loss events is taken into consideration because the fitness cost would be very large and the cell would die.

Populations adapted to a certain environment are vulnerable to change in the environment which might cause extinction of the population. Adaptation is a race against time as the population size decreases in the new environment. Evolutionary rescue is the process where the population acquires mutations which increases the fitness of the population in the new environment such that the extinction in averted. We are interested in how likely is a population to adapt to the changing environment through evolution. Examples of such changes include climate change, invasive species or the onset of drug therapies. There exists three possible ways for a population to survive environmental degradation: migration to a new habitat similar to the one before the onset of environmental changes \cite{cobbold2020should}, adaptation by phenotypic plasticity which involves no changes in the genotype\cite{carja2019evolutionary,carja2017evolutionary} and adaptation through genetic mutations\cite{uecker2014evolutionary,uecker2016role,uecker2011fixation}.

Majority of studies assume that the fitness of the wildtype and mutant are time homogeneous. An exception is \cite{marrec2020adapt} fitness of wildtype and mutant are taken to be functions of time.

\begin{figure}
\centering
\begin{tikzpicture}[very thick,decoration={
    markings,
    mark=at position 0.5 with {\arrow{>}}}
    ] 
\edef\x{6.5} 
\edef\y{-0.5} 
\edef\r{1.25}

\draw [postaction={decorate}, thick](-0.25,-2) -- (-0.25+5,0) node [midway, above] {$2u$};
\draw [postaction={decorate}, thick](-0.25,-2) -- (-0.25+5,-4) node [midway, above] {$2p$};
\draw [postaction={decorate}, thick](-0.25+5+\r,-0) -- (-0.25+10-\r,-0) node [midway, above] {$u$};
\draw [postaction={decorate}, thick](-0.25+5+\r,-4) -- (-0.25+10-\r,-4) node [midway, above] {$u$};
\draw [postaction={decorate}, thick](-0.25+5,-0) -- (-0.25+10,-4) node [midway, above] {$p$};

\filldraw[color=red!60, fill=red!5, very thick](-0.25,-2) circle (\r);
\filldraw[color=red!60, fill=red!5, very thick](-0.25+5,0) circle (\r);
\filldraw[color=red!60, fill=red!5, very thick](-0.25+5,-4) circle (\r);
\filldraw[color=red!60, fill=red!5, very thick](-0.25+10,0) circle (\r);
\filldraw[color=red!60, fill=red!5, very thick](-0.25+10,-4) circle (\r);

\draw [thick](0.05,-1.25) -- (0.05,-2.75);
\draw [thick](-0.55,-1.25) -- (-0.55,-2.75);

\draw [thick](0.05+5,-1.25+2) -- (0.05+5,-2.75+2);
\draw [thick](-0.55+5,-1.25+2) -- (-0.55+5,-2.75+2);

\draw [thick](0.05+10,-1.25+2) -- (0.05+10,-2.75+2);
\draw [thick](-0.55+10,-1.25+2) -- (-0.55+10,-2.75+2);

\draw [thick](-0.55+5,-1.25-2) -- (-0.55+5,-2.75-2);

\draw [thick](-0.55+10,-1.25-2) -- (-0.55+10,-2.75-2);
\end{tikzpicture}
\caption{
}
\label{figureCIN}
\end{figure}

\section{Model}
We model evolutionary rescue of cancer cells through aneuploidy with the help of multitype branching processes: we have three populations of cancer cells, wildtype, aneuploidy and mutant cells which proliferate and die with rates $\lambda_w,  \lambda_a,  \lambda_m$ and $\mu_w, \mu_a, \mu_m$, respectively. Wildtype cells can acquire aneuploidy at rate $u$ while aneuploidy and wildtype cells can mutate with rate $v$ (se Figure \ref{figureAneuploidy}). The probability that a single mutant cell will survive is given by\cite{allen2010introduction}:
\begin{equation}
p_m=\left\{
  \begin{array}{@{}ll@{}}
  \frac{\lambda_m-\mu_m}{\lambda_m},\quad\text{if }\lambda_m>\mu_m \\
   0,\quad\text{else}.
  \end{array}\right.
  \end{equation}
 If the population originally consists of a single aneuploidy cell, then the probability that the population will survive is given by the following quadratic equation:
\begin{align}
1-p_a=\frac{\mu_a}{\lambda_a+\mu_a+v}+\frac{v}{\lambda_a+\mu_a+v}\left(1-p_m\right)+\frac{\lambda_a}{\lambda_a+\mu_a+v}\left(1-p_a\right)^2,
\end{align}
whose smallest solution is the survival probability:
\begin{align}\label{survproba}
p_a=\frac{\lambda_a-\mu_a-v+\sqrt{\left(\lambda_a-\mu_a-v\right)^2+4\lambda_avp_m}}{2\lambda_a}.
\end{align}
Analogously, the survival probability of a population consisting initially of one wildtype cell satisfies the quadratic equation:
\begin{align}\nonumber
1-p_w&=\frac{\mu_w}{\lambda_w+\mu_w+u+v}+\frac{u}{\lambda_w+\mu_w+u+v}\left(1-p_a\right)+\frac{\lambda_w}{\lambda_w+\mu_w+u+v}\left(1-p_w\right)^2\\
&+\frac{v}{\lambda_w+\mu_w+u+v}\left(1-p_m\right),
\end{align}
where the smallest solution gives the survival probability:
\begin{align}\label{survprobw}
p_w=\frac{\lambda_w-\mu_w-u-v+\sqrt{\left(\lambda_w-\mu_w-u-v\right)^2+4\lambda_w\left(up_a+vp_m\right)}}{2\lambda_w}.
\end{align}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
\centering
\begin{tikzpicture}[very thick,decoration={
    markings,
    mark=at position 0.5 with {\arrow{>}}}, middlearrow/.style 2 args={
        decoration={             
            markings, 
            mark=at position 0.5 with {\arrow[xshift=3.333pt]{triangle 45}, \node[#1] {#2};}
        },
        postaction={decorate}
    }]
\edef\x{6.5} 
\edef\y{-0.5} 
\edef\r{1.25}

\edef\R{1} 
\edef\Y{0.25}

\draw[black,thick,middlearrow={above}{$\lambda_w$},text=black] (-0.25,\Y) arc (270:-90:\R);
\draw[black,thick,middlearrow={above}{$\lambda_a$},text=black] (-0.25+5,\Y) arc (270:-90:\R);
\draw[black,thick,middlearrow={above}{$\lambda_m$},text=black] (-0.25+10,\Y) arc (270:-90:\R);

\draw [postaction={decorate}, thick](-0.25,0) -- (-0.25+5,0) node [midway, above] {$u$};
\draw [postaction={decorate}, thick](-0.25+5+\r,-0) -- (-0.25+10-\r,-0) node [midway, above] {$v$};

\draw[thick, postaction={decorate}] (-0.25,0) arc (180:360:5) node [midway, above] {$v$};

\filldraw[color=green!60, fill=green!5, very thick](-0.25,0) circle (\r);
\filldraw[color=green!60, fill=green!5, very thick](-0.25+5,0) circle (\r);
\filldraw[color=green!60, fill=green!5, very thick](-0.25+10,0) circle (\r);

\draw [thick, postaction={decorate}](-0.25,-1.25) -- (-0.25,-2.75) node [midway, left] {$\mu_w$};

\draw [thick, postaction={decorate}](-0.25+5,-1.25) -- (-0.25+5,-2.75) node [midway, right] {$\mu_a$};

\draw [thick, postaction={decorate}](-0.25+10,-1.25) -- (-0.25+10,-2.75) node [midway, right] {$\mu_m$};

\edef\y{-0.25}

\node at (\y,0) {Wildtype};
\node at (\y+5,0) {Aneuploidy};
\node at (\y+10,0) {Mutant};
\end{tikzpicture}
\caption{Diagram of the evolutionary rescue model where the population of cancer cells is subdivided into three populations of cancer cells, wildtype, aneuploidy and mutant cells which proliferate and die with rates $\lambda_w,  \lambda_a,  \lambda_m$ and $\mu_w, \mu_a, \mu_m$, respectively. Wildtype cells can acquire aneuploidy at rate $u$ while aneuploidy and wildtype cells can mutate with rate $v$.}
\label{figureAneuploidy}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{First case: $4\lambda_avp_m<\left(\lambda_a-\mu_a-v\right)^2$}
We assume that $\mu_w>\lambda_w$ and $\mu_a<\lambda_a$ and, as a result, we rewrite \eqref{survproba} and \eqref{survprobw} as:
\begin{align*}
p_a&=\frac{\lambda_a-\mu_a-v}{2\lambda_a}\left(1+\sqrt{1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}}\right),\\
p_w&=\frac{\lambda_w-\mu_w-u-v}{2\lambda_w}\left(1-\sqrt{1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}}\right).
\end{align*}
Making use of the Taylor series expansions:
\begin{align*}
\left(1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}\right)^{\frac{1}{2}}&=1+\frac{2\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}+\cdots\\
\left(1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}\right)^{\frac{1}{2}}&=1+\frac{2\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_a-\mu_a-u-v\right)^2}+\cdots
\end{align*}
we obtain the following approximation for the survival probability of a population consisting of a single individual wildtype cell:
\begin{align}\label{survprobwinitial}
p_w&\approx\frac{vp_m+up_a}{\lambda_a-\mu_a-u-v}\\
\nonumber
&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{v\left(\lambda_a-\mu_a-u\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-u\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\ \label{survprobw}
&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{v\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right],
\end{align}
where in the last line we have used the fact that $u,v\ll1$. Using the notational convention
\begin{equation}\label{notationalconv}
\Delta_i=\lambda_i-\mu_i,
\end{equation}
we write \eqref{survprobw} as
\begin{equation}\label{survprobwapprox}
p_w=-\frac{1}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right).
\end{equation}
Given an initial population consisting of $N$ wildtype cancer cells, the probability that the population will survive is given by: 
\begin{align}
p_{est}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right],
\end{align}
which we plot in Figure \ref{SurvPlotNData} as a function of $N$ and in Figure \ref{SurvPlot} as a function of $\lambda_w$.

We want to improve the accuracy of our approximation by taking into consideration the second term of the Taylor series expansion:
\begin{align*}
\left(1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}\right)^{\frac{1}{2}}=1+\frac{2\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}-\frac{\left(\lambda_avp_m\right)^2}{4\left(\lambda_a-\mu_a-v\right)^4}+\cdots,
\end{align*}
which gives us the following approximation for $p_a$:
\begin{align}
p_a=\frac{\lambda_a-\mu_a-v}{\lambda_a}+\frac{vp_m}{\lambda_a-\mu_a-v}-\frac{\lambda_a\left(vp_m\right)^2}{8\left(\lambda_a-\mu_a-v\right)^3}
\end{align}
From which we deduce that:
\begin{align}\nonumber
p_w&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{v\left(\lambda_a-\mu_a-u\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-u\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}-\frac{uv^2\lambda_a\left(\lambda_m-\mu_m\right)^2}{8\lambda_m^2\left(\lambda_a-\mu_a-v\right)^3}\right]\\ \label{survprobw}
&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{v\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}-\frac{uv^2\lambda_a\left(\lambda_m-\mu_m\right)^2}{8\lambda_m^2\left(\lambda_a-\mu_a\right)^3}\right].
\end{align}
Using the notations described in \eqref{notationalconv} we write the above equation as:
\begin{equation}\label{survprobwapproxcorrected}
p_w=-\frac{1}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}-\frac{uv^2\lambda_a\Delta_m^2}{8\lambda_m^2\Delta_a^3}\right).
\end{equation}
Given an initial population consisting of $N$ wildtype cancer cells, the probability that the population will survive is given by: 
\begin{align}
p_{est}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right].
\end{align}
\section{Second case: $4\lambda_avp_m>\left(\lambda_a-\mu_a-v\right)^2$}
If we assume that  $4\lambda_avp_m>\left(\lambda_a-\mu_a-v\right)^2$ then we write:
\begin{equation}
p_a=\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}\left(1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\lambda_avp_m}\right)^{\frac12}}{2\lambda_a}
\end{equation}
and using the following Taylor series expansion:
\begin{align*}
\left(1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\lambda_avp_m}\right)^{\frac{1}{2}}=1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{8\lambda_avp_m}+\cdots
\end{align*}
we obtain:
\begin{align*}
p_a&\approx\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}\left[1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{8\lambda_avp_m}\right]}{2\lambda_a}\\
&=\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\sqrt{\lambda_avp_m}}}{2\lambda_a}\\
&\approx\sqrt{\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_a\lambda_m}}
\end{align*}
In the last line of the above we have used the following inequalities:
\begin{align*}
\lambda_a-\mu_a,\,v,\,\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\sqrt{\lambda_avp_m}}\ll1.
\end{align*}
As a result, we have from \eqref{survprobwinitial} the probability of rescue of a population starting from one wildtype individual:
\begin{align*}
p_w&\approx\frac{1}{\lambda_a-\mu_a-u-v}\left[v\frac{\lambda_m-\mu_m}{\lambda_m}+u\sqrt{\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_a\lambda_m}}\right]\\
&=\sqrt{\frac{\lambda_m-\mu_m}{\lambda_m}}\frac{\sqrt{v}}{\lambda_a-\mu_a-u-v}\left[\frac{u}{\sqrt{\lambda_a}}+\sqrt{\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}}\right]\\
&=\sqrt{\frac{\Delta_m}{\lambda_m}}\frac{\sqrt{v}}{\Delta_a-u-v}\left[\frac{u}{\sqrt{\lambda_a}}+\sqrt{\frac{v\Delta_m}{\lambda_m}}\right],
\end{align*}
where in the last line we have used the notations defined in \eqref{notationalconv}.

Given an initial population consisting of $N$ wildtype cancer cells, the probability that the population will survive is given by: 
\begin{align*}
p_{est}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[-N\sqrt{\frac{\Delta_m}{\lambda_m}}\frac{\sqrt{v}}{\Delta_a-u-v}\left(\frac{u}{\sqrt{\lambda_a}}+\sqrt{\frac{v\Delta_m}{\lambda_m}}\right)\right].
\end{align*}

\section{Standing genetic variation}
Until now we have assumed that the initial population of cells consisted entirely of wildtype cells. We modify this assumtion such that the initial population includes a fraction $f$ of cells with aneuploidy. The probability of evolutionary rescue by the cells with aneuploidy from the initial population is:
\begin{equation*}
p_{old}=1-\left(1-p_a\right)^{fN}\approx 1-\e^{-fNp_a}.
\end{equation*}
The total probability of evolutionary rescue is given by:
\begin{align}\nonumber
p_{total}&=p_{new}+\left(1-p_{new}\right)p_{old}\\
&=1-\e^{-\left[\left(1-f\right)p_w+fp_a\right]N}.
\end{align}
The fraction of the cases in which the population is rescued by the standing genetic variation is given by:
\begin{align*}
F\left(f\right)=\frac{ 1-\e^{-fNp_a}}{1-\e^{-\left[\left(1-f\right)p_w+fp_a\right]N}}.
\end{align*}
We let $F=\frac{1}{2}$ and we obtain:
\begin{equation*}
f^*\approx\frac{p_w}{p_w+p_a},
\end{equation*}
wher we have used the expansion $\e^x\approx 1+x$.
We plot $F$ and $f^*$ in Figure \ref{FractionPlot}.

\section{Discussion}

\begin{figure}[!t]
 \vspace*{1\baselineskip}
\includegraphics[width=1\textwidth]{SurvPlot.pdf}
\caption{Plot of the probability of survival of a population as a function of the proliferation rate of the wildtype cells. The blue line represents the exact solution \eqref{survprobw} and the orange line line represents the approximation \eqref{survprobwapprox}. Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $N=75, \lambda_a=1+10^{-2},\lambda_m=1+10^{-3},\mu_w=1,\mu_a=1,\mu_m=1.$}
\label{SurvPlot}
\end{figure}

\begin{figure}[!t]
 \vspace*{1\baselineskip}
\includegraphics[width=1\textwidth]{SurvPlotNData.pdf}
\caption{Plot of the probability of survival of a population as a function of the initial population size of wildtype cells. The blue line represents the exact solution \eqref{survprobw}, the orange line line represents the approximation \eqref{survprobwapprox}, the green line represents the first order correction \eqref{survprobwapproxcorrected} and the red dots represents stochastic simulations. For the simulations we have chosen the following parameters: $\lambda_w=1-10^{-2}, \lambda_a=1+10^{-2},\lambda_m=1+10^{-3},\mu_w=1,\mu_a=1,\mu_m=1.$}
\label{SurvPlotNData}
\end{figure}

\begin{figure}[!t]
 \vspace*{1\baselineskip}
\includegraphics[width=1\textwidth]{FractionPlot.pdf}
\caption{}
\label{FractionPlot}
\end{figure}

\bibliographystyle{unsrt}
\bibliography{green2021}
\end{document}

### 2022-10-02 EvolutionaryRescue1.tex
Add files via upload

@@
 \usetikzlibrary{decorations.text}
+\usetikzlibrary{shapes.multipart}
 
@@
 \author[1]{Remus Stana}
+\author[2]{‪Daniel Weissman}
 \author[1]{Yoav Ram}
-\author[2]{‪Daniel Weissman}
 \affil[1]{School of Zoology, Tel Aviv University, Tel Aviv-Yafo, Israel}
@@
 \begin{abstract}
+Evolutionary rescue is the process by which a population is able to survive a sudden environmental change which initially causes the population to decline towards extinction.
+A prime example of evolutionary rescue is the ability of cancer to survive being exposed to various treatments.
+We are interested in the mechanisms through which a population of cancer cells are able to adapt to chemotherapy, and in particular, the role played by chromosomal instability (aneuploidy).
+Cancer cells which have aneuploidy are hypothesized to have a higher fitness in an environment altered by anti-cancer drugs as they have incomplete pathways which drugs activate in order to kill the cells.
+Aneuploidy is highly prevalent in tumors and certain drugs which attempt to combat cancers through increasing chromosomal instability.
+As a result, the question we wish to answer is how aneuploidy impacts the fate of the population of cancer cells.
+We propose to model evolutionary rescue with the help of multi-type branching processes to obtain the probability that cancer will survive.
+Additionally, we will utilize large genomic datasets to asses the effects of aneuploidy on the probability of evolutionary rescue.
 \end{abstract}
@@
 \section{Model}
-We model evolutionary rescue of cancer cells through aneuploidy with the help of multitype branching processes: we have three populations of cancer cells, wildtype, aneuploidy and mutant cells which proliferate and die with rates $\lambda_w,  \lambda_a,  \lambda_m$ and $\mu_w, \mu_a, \mu_m$, respectively.
+We model evolutionary rescue of cancer cells through aneuploidy with the help of multitype branching processes: we have three populations of cancer cells, wildtype euploid $w_t$, wildtype aneuploid $a_t$ and mutant euploid cells $m_t$ which proliferate and die with rates $\lambda_w,  \lambda_a,  \lambda_m$ and $\mu_w, \mu_a, \mu_m$, respectively.
 Wildtype cells can acquire aneuploidy at rate $u$ while aneuploidy and wildtype cells can mutate with rate $v$ (se Figure \ref{figureAneuploidy}).
-The probability that a single mutant cell will survive is given by\cite{allen2010introduction}:
-\begin{equation}
-p_m=\left\{
-  \begin{array}{@{}ll@{}}
-  \frac{\lambda_m-\mu_m}{\lambda_m},\quad\text{if }\lambda_m>\mu_m \\
-   0,\quad\text{else}.
-  \end{array}\right.
-  \end{equation}
+Changes in the number of wildtype euploid, wildtype aneuploid and mutant euploid cells occur at rates:
+\begin{align*}
+w_t&\rightarrow w_t+1:\quad \lambda_ww_t,\\
+w_t&\rightarrow w_t-1:\quad \left(\mu_w+u\right)w_t,\\
+a_t&\rightarrow a_t+1:\quad \lambda_aa_t+uw_t,\\
+a_t&\rightarrow a_t-1:\quad \left(\mu_a-w\right)a_t,\\
+m_t&\rightarrow m_t+1:\quad \lambda_am_t+wa_t,\\
+m_t&\rightarrow m_t-1:\quad \mu_am_t.
+\end{align*}
+The survival probability of a population consisting initially of one wildtype cell satisfies the quadratic equation:
+\begin{align}\nonumber
+1-p_w&=\frac{\mu_w}{\lambda_w+\mu_w+u+v}+\frac{u}{\lambda_w+\mu_w+u+v}\left(1-p_a\right)+\frac{\lambda_w}{\lambda_w+\mu_w+u+v}\left(1-p_w\right)^2\\ \label{quadraticeqev1}
+&+\frac{v}{\lambda_w+\mu_w+u+v}\left(1-p_m\right),
+\end{align}
+where the $p_a$ and $p_w$ are the survival probabilities of a population consisting initially of one aneuploidy and wildtype cell, respectively.
+
+ The smallest solution of the quadratic equation \eqref{quadraticeqev1} gives the survival probability:
+\begin{align}\label{survprobw}
+p_w=\frac{\lambda_w-\mu_w-u-v+\sqrt{\left(\lambda_w-\mu_w-u-v\right)^2+4\lambda_w\left(up_a+vp_m\right)}}{2\lambda_w}.
+\end{align}
 If the population originally consists of a single aneuploidy cell, then the probability that the population will survive is given by the following quadratic equation:
@@
 \end{align}
-Analogously, the survival probability of a population consisting initially of one wildtype cell satisfies the quadratic equation:
-\begin{align}\nonumber
-1-p_w&=\frac{\mu_w}{\lambda_w+\mu_w+u+v}+\frac{u}{\lambda_w+\mu_w+u+v}\left(1-p_a\right)+\frac{\lambda_w}{\lambda_w+\mu_w+u+v}\left(1-p_w\right)^2\\
-&+\frac{v}{\lambda_w+\mu_w+u+v}\left(1-p_m\right),
-\end{align}
-where the smallest solution gives the survival probability:
-\begin{align}\label{survprobw}
-p_w=\frac{\lambda_w-\mu_w-u-v+\sqrt{\left(\lambda_w-\mu_w-u-v\right)^2+4\lambda_w\left(up_a+vp_m\right)}}{2\lambda_w}.
-\end{align}
-
+The probability that a single mutant cell will survive is given by \cite{allen2010introduction}:
+\begin{equation}
+p_m=\left\{
+  \begin{array}{@{}ll@{}}
+  \frac{\lambda_m-\mu_m}{\lambda_m},\quad\text{if }\lambda_m>\mu_m \\
+   0,\quad\text{else}.
+  \end{array}\right.
+  \end{equation}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+\edef\x{0.25} 
 \edef\y{-0.25}
 
-\node at (\y,0) {Wildtype};
-\node at (\y+5,0) {Aneuploidy};
-\node at (\y+10,0) {Mutant};
+\node[text width=2cm,align=justify] at (\y,\x)  
+{\begin{center}
+Wildtype  euploid
+\end{center}};
+
+\node[text width=2cm,align=justify] at (\y+5,\x)  
+{\begin{center}
+Wildtype  aneuploid
+\end{center}};
+
+\node[text width=2cm,align=justify] at (\y+10,\x)  
+{\begin{center}
+Mutant  euploid
+\end{center}};
 \end{tikzpicture}
-\caption{Diagram of the evolutionary rescue model where the population of cancer cells is subdivided into three populations of cancer cells, wildtype, aneuploidy and mutant cells which proliferate and die with rates $\lambda_w,  \lambda_a,  \lambda_m$ and $\mu_w, \mu_a, \mu_m$, respectively.
-Wildtype cells can acquire aneuploidy at rate $u$ while aneuploidy and wildtype cells can mutate with rate $v$.}
+\caption{Diagram of the evolutionary rescue model where the population of cancer cells is subdivided to wildtype euploid, wildtype aneuploid and mutant euploid cells that divide and die with rates $\lambda_w,  \lambda_a,  \lambda_m$ and $\mu_w, \mu_a, \mu_m$, respectively.
+Wildtype cells can acquire aneuploidy at rate $u$.
+Aneuploid and wildtype cells can mutate with rate $v$.}
 \label{figureAneuploidy}

### 2022-10-12 EvolutionaryRescue1.tex
Add files via upload

@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{figure}
+\centering
+\begin{tikzpicture}[very thick,decoration={
+    markings,
+    mark=at position 0.5 with {\arrow{>}}}
+    ] 
+\edef\x{6.5} 
+\edef\y{-0.5} 
+\edef\r{1.25}
+
+\draw [postaction={decorate}, thick](-0.25,-4) -- (-0.25+5,0) node [midway, above] {$u$};
+%\draw [postaction={decorate}, thick](-0.25,-2) -- (-0.25+5,-4) node [midway, above] {$2p$};
+\draw [postaction={decorate}, thick](-0.25+5,-0) -- (-0.25+10,-4) node [midway, above] {$v$};
+%\draw [postaction={decorate}, thick](-0.25+5+\r,-4) -- (-0.25+10-\r,-4) node [midway, above] {$u$};
+%\draw [postaction={decorate}, thick](-0.25+5,-0) -- (-0.25+10,-4) node [midway, above] {$p$};
+\draw [postaction={decorate}, thick]  (-0.25+10,-4) -- (-0.25+5,-8) node [midway, above] {$u$};
+\draw [postaction={decorate}, thick](-0.25,-4)  -- (-0.25+5,-8)  node [midway, above] {$v$};
+
+\filldraw[color=red!60, fill=red!5, very thick](-0.25,-4) circle (\r);
+\filldraw[color=red!60, fill=red!5, very thick](-0.25+5,0) circle (\r);
+%\filldraw[color=red!60, fill=red!5, very thick](-0.25+5,-4) circle (\r);
+\filldraw[color=red!60, fill=red!5, very thick](-0.25+10,-4) circle (\r);
+%\filldraw[color=red!60, fill=red!5, very thick](-0.25+10,-4) circle (\r);
+\filldraw[color=red!60, fill=red!5, very thick](-0.25+5,-8) circle (\r);
+
+\draw [thick](0.05,-1.25+2-4) -- (0.05,-2.75+2-4);
+\draw [thick](-0.55,-1.25+2-4) -- (-0.55,-2.75+2-4);
+
+\draw [thick](0.15+5,-1.25+2) -- (0.15+5,-2.75+2);
+\draw [thick](-0.65+5,-1.25+2) -- (-0.65+5,-2.75+2);
+\draw [thick](-0.25+5,-1.25+2) -- (-0.25+5,-2.75+2);
+
+\draw [thick](0.15+10,-1.25+2-4) -- (0.15+10,-2.75+2-4);
+\draw [thick](-0.25+10,-1.25+2-4) -- (-0.25+10,-2.75+2-4);
+\draw [,color=blue,thick](-0.7+10,-1.25+2-4) -- (-0.7+10,-2.75+2-4);
+
+\draw [thick](0.05+5,-1.25+2-8) -- (0.05+5,-2.75+2-8);
+\draw [,color=blue,thick](-0.55+5,-1.25+2-8) -- (-0.55+5,-2.75+2-8);
+%\draw [thick](-0.55+5,-1.25-2) -- (-0.55+5,-2.75-2);
+
+%\draw [thick](-0.55+10,-1.25-2) -- (-0.55+10,-2.75-2);
+\end{tikzpicture}
+\caption{
+}
+\label{figuretrisomy}
+\end{figure}
+
 \section{Model}
@@
 Changes in the number of wildtype euploid, wildtype aneuploid and mutant euploid cells occur at rates:
-\begin{align*}
+\begin{subequations}
+\begin{flalign}
 w_t&\rightarrow w_t+1:\quad \lambda_ww_t,\\
-w_t&\rightarrow w_t-1:\quad \left(\mu_w+u\right)w_t,\\
+w_t&\rightarrow w_t-1:\quad \left(\mu_w+u+v\right)w_t,\\
 a_t&\rightarrow a_t+1:\quad \lambda_aa_t+uw_t,\\
-a_t&\rightarrow a_t-1:\quad \left(\mu_a-w\right)a_t,\\
-m_t&\rightarrow m_t+1:\quad \lambda_am_t+wa_t,\\
+a_t&\rightarrow a_t-1:\quad \left(\mu_a+v\right)a_t,\\
+m_t&\rightarrow m_t+1:\quad \lambda_am_t+va_t+vm_t,\\
 m_t&\rightarrow m_t-1:\quad \mu_am_t.
-\end{align*}
+\end{flalign}
+\end{subequations}
+Given that we have used branching processes, the lineages produced by cells from the initial population will be independent of each other.
+This is not a biologically realistic model as cells compete for resources in tumors.
+We will study a more biologically realistic model in Section.
 The survival probability of a population consisting initially of one wildtype cell satisfies the quadratic equation:
@@
 \end{align}
-The probability that a single mutant cell will survive is given by \cite{allen2010introduction}:
+The probability that a single mutant cell will survive is given by (see Appendix \ref{AppendixSurvLin}):
 \begin{equation}
@@
 The blue line represents the exact solution \eqref{survprobw}, the orange line line represents the approximation \eqref{survprobwapprox}, the green line represents the first order correction \eqref{survprobwapproxcorrected} and the red dots represents stochastic simulations.
-For the simulations we have chosen the following parameters: $\lambda_w=1-10^{-2}, \lambda_a=1+10^{-2},\lambda_m=1+10^{-3},\mu_w=1,\mu_a=1,\mu_m=1.$}
+For the simulations we have chosen the following parameters: $\lambda_w=1-10^{-2}, \lambda_a=1+10^{-2},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1.$}
 \label{SurvPlotNData}
@@
 \bibliography{green2021}
+
+\appendix
+
+\section{Gillespie algorithm}
+\begin{subequations}
+\begin{flalign}
+(+1,0,0)&:\quad \lambda_ww_t,\\
+(-1,0,0)&:\quad \mu_ww_t,\\
+(-1,+1,0)&:\quad uw_t,\\
+(-1,0,+1)&:\quad vw_t,\\
+(0,+1,0)&:\quad \lambda_aa_t,\\
+(0,-1,0)&:\quad \mu_aa_t,\\
+(0,-1,+1)&:\quad va_t,\\
+(0,0,+1)&:\quad \lambda_am_t,\\
+(0,0,-1)&:\quad \mu_am_t.
+\end{flalign}
+\end{subequations}
+
+\section{Survival probability of a mutant lineage}\label{AppendixSurvLin}
+\begin{align*}
+p_{i+j,i}\left(\Delta t\right)=\left\{
+  \begin{array}{@{}ll@{}}
+  \mu i\Delta t+o\left(\Delta t\right),\quad j=-1 \\
+  \lambda i\Delta t+o\left(\Delta t\right),\quad j=1 \\
+    1-\left(\lambda+\mu\right) i\Delta t+o\left(\Delta t\right),\quad j=0 \\
+   o\left(\Delta t\right),\quad j\neq-1,0,1.
+  \end{array}\right.
+\end{align*}
+
+\begin{align*}
+\frac{dp_i\left(t\right)}{dt}&=\lambda\left(i-1\right)p_{i-1}\left(t\right)+\mu\left(i+1\right)p_{i+1}\left(t\right)-\left(\lambda+\mu\right)ip_i\left(t\right)\\
+\frac{dp_0\left(t\right)}{dt}&=\mu p_1\left(t\right)
+\end{align*}
+
+\begin{align*}
+\mathcal{P}\left(z,t\right)=\left\{
+  \begin{array}{@{}ll@{}}
+  \left(\frac{\e^{t\left(\mu-\lambda\right)}\left(\lambda z-\mu\right)-\mu\left(z-1\right)}{\e^{t\left(\mu-\lambda\right)}\left(\lambda z-\mu\right)-\lambda\left(z-1\right)}\right)^N,\quad \text{if }\lambda\neq\mu \\
+  \left(\frac{1-\left(\lambda t-1\right)\left(z-1\right)}{1-\lambda t\left(z-1\right)}\right)^N,\quad \text{if }\lambda=\mu
+  \end{array}\right.
+\end{align*}
+
+\begin{equation*}
+\mathcal{P}\left(z,t\right)=\sum_{i=0}^\infty p_i\left(t\right)z^i \quad\text{and}\quad p_i\left(t\right)=\left.\frac{1}{i!}\frac{\partial^i\mathcal{P}}{\partial z^i}\right|_{z=0}
+\end{equation*}
+
+\begin{align*}
+p_0\left(t\right)=\left\{
+  \begin{array}{@{}ll@{}}
+  \left(\frac{\mu-\mu\e^{\left(\mu-\lambda\right)t}}{\lambda-\mu\e^{\left(\mu-\lambda\right)t}}\right)^N,\quad \text{if }\lambda\neq\mu \\
+  \left(\frac{\lambda t}{1+\lambda t}\right)^N,\quad \text{if }\lambda=\mu
+  \end{array}\right.
+\end{align*}
+
+\begin{align*}
+p_0\left(\infty\right)=\lim_{t\rightarrow\infty}p_0\left(t\right)=\left\{
+  \begin{array}{@{}ll@{}}
+  1,\quad \text{if }\lambda\leq\mu \\
+  \left(\frac{\mu}{\lambda}\right)^N,\quad \text{if }\lambda>\mu
+  \end{array}\right.
+\end{align*}
 \end{document}

### 2022-11-07 Python Code/EvolutionaryRescueLogistic2.py
Add files via upload

@@ -1,0 +1,157 @@
+#import matplotlib.inline
+import numpy as np
+import matplotlib.pyplot as plt
+import numba
+import scipy.stats as stats
+from scipy.integrate import solve_ivp
+import seaborn as sns
+sns.set_context('talk')
+
+from functools import partial
+
+def plot_mp(t, n, a, m, label='', ax=None):
+    if ax is None:
+        fig, ax = plt.subplots()
+    ax.plot(t, n, lw=3, label='wildtype ' + label)
+    ax.plot(t, a, lw=3, label='aneuploidy ' + label)
+    ax.plot(t, m, lw=3, label='mutant ' + label)
+    ax.set_xlabel('Time')
+    ax.set_ylabel('Count')
+    ax.legend()
+    return ax
+
+@numba.jit # 2-fold faster
+def get_rates(n, a, m, λn, λa, λm, μn, μa, μm, v, w):
+    return np.array([
+        (λn + μn)*n,    
+        λn * n,     
+        v * n,
+        w * n,
+        (λa + μa*(1 - (n+a+m)/K))*a, 
+        λa * a,
+        w * a,
+        (λm + μm*(1 - (n+a+m)/K))*m,    
+        λm * m,
+    ])
+    
+@numba.jit # 2-fold faster
+def draw_time(rates):
+ #   print(rates,'r')
+    total_rate = rates.sum()
+    
+#    print(rates.sum(),'hello')
+    if total_rate > 0:
+        return np.random.exponential(1/total_rate)
+    if total_rate == 0:
+        return 1
+
+# @numba.jit # jit causes errors with multinomial
+def draw_reaction(rates):
+    if rates.sum() > 0:
+        rates /= rates.sum()
+        return np.random.multinomial(1, rates).argmax()
+    if rates.sum() == 0:
+        return 9
+
+#rates = get_rates(3, 100, βm, βp, γp)
+#rates /= rates.sum()
+    
+updates = np.array([
+    [1, 0, 0],  # growth wild type
+    [-1, 0, 0], # death wild type
+    [-1, 1, 0], #mutation w->a
+    [-1, 0, 1], #mutation w->m
+    [0, 1, 0],  # growth aneuploidy type
+    [0, -1, 0],  # death aneuploidy type
+    [0, -1, 1], # #mutation a->m
+    [0, 0, 1],  # growth mutant type
+    [0, 0, -1],   # death mutant type
[…]

### 2022-11-07 Python Code/EvolutionaryRescueLogistic2.py
Delete Python Code directory

@@ -1,157 +1,0 @@
-#import matplotlib.inline
-import numpy as np
-import matplotlib.pyplot as plt
-import numba
-import scipy.stats as stats
-from scipy.integrate import solve_ivp
-import seaborn as sns
-sns.set_context('talk')
-
-from functools import partial
-
-def plot_mp(t, n, a, m, label='', ax=None):
-    if ax is None:
-        fig, ax = plt.subplots()
-    ax.plot(t, n, lw=3, label='wildtype ' + label)
-    ax.plot(t, a, lw=3, label='aneuploidy ' + label)
-    ax.plot(t, m, lw=3, label='mutant ' + label)
-    ax.set_xlabel('Time')
-    ax.set_ylabel('Count')
-    ax.legend()
-    return ax
-
-@numba.jit # 2-fold faster
-def get_rates(n, a, m, λn, λa, λm, μn, μa, μm, v, w):
-    return np.array([
-        (λn + μn)*n,    
-        λn * n,     
-        v * n,
-        w * n,
-        (λa + μa*(1 - (n+a+m)/K))*a, 
-        λa * a,
-        w * a,
-        (λm + μm*(1 - (n+a+m)/K))*m,    
-        λm * m,
-    ])
-    
-@numba.jit # 2-fold faster
-def draw_time(rates):
- #   print(rates,'r')
-    total_rate = rates.sum()
-    
-#    print(rates.sum(),'hello')
-    if total_rate > 0:
-        return np.random.exponential(1/total_rate)
-    if total_rate == 0:
-        return 1
-
-# @numba.jit # jit causes errors with multinomial
-def draw_reaction(rates):
-    if rates.sum() > 0:
-        rates /= rates.sum()
-        return np.random.multinomial(1, rates).argmax()
-    if rates.sum() == 0:
-        return 9
-
-#rates = get_rates(3, 100, βm, βp, γp)
-#rates /= rates.sum()
-    
-updates = np.array([
-    [1, 0, 0],  # growth wild type
-    [-1, 0, 0], # death wild type
-    [-1, 1, 0], #mutation w->a
-    [-1, 0, 1], #mutation w->m
-    [0, 1, 0],  # growth aneuploidy type
-    [0, -1, 0],  # death aneuploidy type
-    [0, -1, 1], # #mutation a->m
-    [0, 0, 1],  # growth mutant type
-    [0, 0, -1],   # death mutant type
[…]

### 2022-11-07 Python Code/EvolutionaryRescueGompertz.py
Add files via upload

@@ -1,0 +1,170 @@
+#import matplotlib.inline
+import numpy as np
+import matplotlib.pyplot as plt
+import numba
+import scipy.stats as stats
+from scipy.integrate import solve_ivp
+import seaborn as sns
+sns.set_context('talk')
+
+from functools import partial
+
+def plot_mp(t, n, a, m, label='', ax=None):
+    if ax is None:
+        fig, ax = plt.subplots()
+    ax.plot(t, n, lw=3, label='wildtype ' + label)
+    ax.plot(t, a, lw=3, label='aneuploidy ' + label)
+    ax.plot(t, m, lw=3, label='mutant ' + label)
+    ax.set_xlabel('Time')
+    ax.set_ylabel('Count')
+    ax.legend()
+    return ax
+
+def funsum(n,a,m):
+    
+    if n+a+m==0:
+        
+        return 1
+    
+    if n+a+m>0:
+        
+        return n+a+m
+        
+        
+
+@numba.jit # 2-fold faster
+def get_rates(n, a, m, λn, λa, λm, μn, μa, μm, v, w):
+    return np.array([
+        λn * n,    
+        μn * n,     
+        v * n,
+        w * n,
+        (λa - μa*np.log(funsum(n,a,m)/K))*a, 
+        λa * a,
+        w * a,
+        (λm - μm* np.log(funsum(n,a,m)/K))*m,    
+        λm * m,
+    ])
+    
+@numba.jit # 2-fold faster
+def draw_time(rates):
+ #   print(rates,'r')
+    total_rate = rates.sum()
+    
+#    print(rates.sum(),'hello')
+    if total_rate > 0:
+        return np.random.exponential(1/total_rate)
+    if total_rate == 0:
+        return 1
+
+# @numba.jit # jit causes errors with multinomial
+def draw_reaction(rates):
+    if rates.sum() > 0:
+        rates /= rates.sum()
+        return np.random.multinomial(1, rates).argmax()
+    if rates.sum() == 0:
+        return 9
+
+#rates = get_rates(3, 100, βm, βp, γp)
+#rates /= rates.sum()
+    
+updates = np.array([
+    [1, 0, 0],  # growth wild type
+    [-1, 0, 0], # death wild type
+    [-1, 1, 0], #mutation w->a
+    [-1, 0, 1], #mutation w->m
+    [0, 1, 0],  # growth aneuploidy type
[…]

### 2022-11-07 EvolutionaryRescue1.tex
Add files via upload

@@
 \begin{document}
-\title{Evolutionary rescue}
+\title{Modeling the effect of aneuploidy on cancer evolution}
 \author[1]{Remus Stana}
@@
 \section{First case: $4\lambda_avp_m<\left(\lambda_a-\mu_a-v\right)^2$}
-We assume that $\mu_w>\lambda_w$ and $\mu_a<\lambda_a$ and, as a result, we rewrite \eqref{survproba} and \eqref{survprobw} as:
+\subsection{$\lambda_a>\mu_a$ and $\lambda_w<\mu_w$}
+We assume that $\lambda_a>\mu_a$ and $\lambda_w<\mu_w$ and, as a result, we rewrite \eqref{survprobw} and \eqref{survproba} as:
 \begin{align*}
@@
 \left(1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}\right)^{\frac{1}{2}}&=1+\frac{2\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}+\cdots\\
-\left(1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}\right)^{\frac{1}{2}}&=1+\frac{2\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_a-\mu_a-u-v\right)^2}+\cdots
+\left(1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}\right)^{\frac{1}{2}}&=1+\frac{2\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}+\cdots
 \end{align*}
@@
 \begin{align}\label{survprobwinitial}
-p_w&\approx\frac{vp_m+up_a}{\lambda_a-\mu_a-u-v}\\
+p_w&\approx\frac{vp_m+up_a}{\lambda_w-\mu_w-u-v}\\
 \nonumber
-&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{v\left(\lambda_a-\mu_a-u\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-u\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\ \label{survprobw}
-&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{v\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right],
+&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{v\left(\lambda_a-\mu_a-u\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-u\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\ \label{survprobw2}
+&\approx-\frac{1}{\lambda_w-\mu_w}\left[\frac{v\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right],
 \end{align}
@@
 \begin{align}\nonumber
-p_w&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{v\left(\lambda_a-\mu_a-u\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-u\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}-\frac{uv^2\lambda_a\left(\lambda_m-\mu_m\right)^2}{8\lambda_m^2\left(\lambda_a-\mu_a-v\right)^3}\right]\\ \label{survprobw}
-&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{v\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}-\frac{uv^2\lambda_a\left(\lambda_m-\mu_m\right)^2}{8\lambda_m^2\left(\lambda_a-\mu_a\right)^3}\right].
+p_w&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{v\left(\lambda_a-\mu_a-u\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-u\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}-\frac{uv^2\lambda_a\left(\lambda_m-\mu_m\right)^2}{8\lambda_m^2\left(\lambda_a-\mu_a-v\right)^3}\right]\\ \label{survprobw3}
+&\approx-\frac{1}{\lambda_w-\mu_w}\left[\frac{v\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}-\frac{uv^2\lambda_a\left(\lambda_m-\mu_m\right)^2}{8\lambda_m^2\left(\lambda_a-\mu_a\right)^3}\right].
 \end{align}
@@
 \end{align}
+\subsection{$\lambda_a<\mu_a$ and $\lambda_w<\mu_w$}
+We assume that $\lambda_a<\mu_a$ and $\lambda_w<\mu_w$ and, as a result, we rewrite \eqref{survprobw} and \eqref{survproba} as:
+\begin{align*}
+p_a&=\frac{\lambda_a-\mu_a-v}{2\lambda_a}\left(1-\sqrt{1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}}\right),\\
+p_w&=\frac{\lambda_w-\mu_w-u-v}{2\lambda_w}\left(1-\sqrt{1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}}\right).
+\end{align*}
+\begin{align}\label{survprobwinitial}
+p_w&\approx\frac{vp_m+up_a}{\lambda_w-\mu_w-u-v}\\
+\nonumber
+&\approx\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-v\right)}-\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\ \label{survprobw2}
+&\approx\frac{v}{\lambda_m\left(\lambda_w-\mu_w\right)}\left[\frac{u\left(\lambda_m-\mu_m\right)}{\left(\lambda_a-\mu_a\right)}-\left(\lambda_m-\mu_m\right)\right],
+\end{align}
+where in the last line we have used the fact that $u,v\ll1$. Using the notational convention
+\begin{equation}\label{notationalconv}
+\Delta_i=\lambda_i-\mu_i,
+\end{equation}
+we write \eqref{survprobw} as
+\begin{equation}\label{survprobwapprox}
+p_w=\frac{v}{\lambda_m\Delta_w}\left(\frac{u\Delta_m}{\Delta_a}-\Delta_m\right).
+\end{equation}
+Given an initial population consisting of $N$ wildtype cancer cells, the probability that the population will survive is given by: 
+\begin{align}
+p_{est}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{vN}{\lambda_m\Delta_w}\left(\Delta_m-\frac{u\Delta_m}{\Delta_a}\right)\right],
+\end{align}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Second case: $4\lambda_avp_m>\left(\lambda_a-\mu_a-v\right)^2$}
@@
 \bibliographystyle{unsrt}
-\bibliography{green2021}
+\bibliography{evo2022}
 
@@
 \section{Survival probability of a mutant lineage}\label{AppendixSurvLin}
+The infinitesimal transition probabilities for the simple birth and death process:
 \begin{align*}
@@
 \end{align*}
+The forward Kolmogorov differential equations are:
+\begin{subequations}
+\label{KolmeqSurv}
+\begin{flalign}
+\frac{dp_i\left(t\right)}{dt}&=\lambda\left(i-1\right)p_{i-1}\left(t\right)+\mu\left(i+1\right)p_{i+1}\left(t\right)-\left(\lambda+\mu\right)ip_i\left(t\right),\\
+\frac{dp_0\left(t\right)}{dt}&=\mu p_1\left(t\right),
+\end{flalign}
+\end{subequations}
+for $i=1,2,\dots$ with initial conditions $p_i\left(0\right)=\delta_{iN}$.
 
-\begin{align*}
-\frac{dp_i\left(t\right)}{dt}&=\lambda\left(i-1\right)p_{i-1}\left(t\right)+\mu\left(i+1\right)p_{i+1}\left(t\right)-\left(\lambda+\mu\right)ip_i\left(t\right)\\
-\frac{dp_0\left(t\right)}{dt}&=\mu p_1\left(t\right)
-\end{align*}
-
+The probability generating function is defined as:
+\begin{equation*}
+\mathcal{P}\left(z,t\right)=\sum_{i=0}^\infty p_i\left(t\right)z^i
+\end{equation*}
+which we obtain by multiplying \eqref{KolmeqSurv} by $z^i$ and summing over $i$:
 \begin{align*}
@@
   \begin{array}{@{}ll@{}}
-  \left(\frac{\e^{t\left(\mu-\lambda\right)}\left(\lambda z-\mu\right)-\mu\left(z-1\right)}{\e^{t\left(\mu-\lambda\right)}\left(\lambda z-\mu\right)-\lambda\left(z-1\right)}\right)^N,\quad \text{if }\lambda\neq\mu \\
-  \left(\frac{1-\left(\lambda t-1\right)\left(z-1\right)}{1-\lambda t\left(z-1\right)}\right)^N,\quad \text{if }\lambda=\mu
+  \left(\frac{\e^{t\left(\mu-\lambda\right)}\left(\lambda z-\mu\right)-\mu\left(z-1\right)}{\e^{t\left(\mu-\lambda\right)}\left(\lambda z-\mu\right)-\lambda\left(z-1\right)}\right)^N,\quad \text{if }\lambda\neq\mu, \\
+  \left(\frac{1-\left(\lambda t-1\right)\left(z-1\right)}{1-\lambda t\left(z-1\right)}\right)^N,\quad \text{if }\lambda=\mu.
   \end{array}\right.
 \end{align*}
-
+The probability $p_i$ can be obtained from the probability generating function as:
 \begin{equation*}
-\mathcal{P}\left(z,t\right)=\sum_{i=0}^\infty p_i\left(t\right)z^i \quad\text{and}\quad p_i\left(t\right)=\left.\frac{1}{i!}\frac{\partial^i\mathcal{P}}{\partial z^i}\right|_{z=0}
+p_i\left(t\right)=\left.\frac{1}{i!}\frac{\partial^i\mathcal{P}}{\partial z^i}\right|_{z=0},
 \end{equation*}
-
+and the probability of extinction is given by:
 \begin{align*}
@@
   \begin{array}{@{}ll@{}}
-  \left(\frac{\mu-\mu\e^{\left(\mu-\lambda\right)t}}{\lambda-\mu\e^{\left(\mu-\lambda\right)t}}\right)^N,\quad \text{if }\lambda\neq\mu \\
-  \left(\frac{\lambda t}{1+\lambda t}\right)^N,\quad \text{if }\lambda=\mu
+  \left(\frac{\mu-\mu\e^{\left(\mu-\lambda\right)t}}{\lambda-\mu\e^{\left(\mu-\lambda\right)t}}\right)^N,\quad \text{if }\lambda\neq\mu, \\
+  \left(\frac{\lambda t}{1+\lambda t}\right)^N,\quad \text{if }\lambda=\mu.
   \end{array}\right.
 \end{align*}
-
+When $t\rightarrow\infty$ the extinction probability has the following expression:
 \begin{align*}
@@
   \begin{array}{@{}ll@{}}
-  1,\quad \text{if }\lambda\leq\mu \\
-  \left(\frac{\mu}{\lambda}\right)^N,\quad \text{if }\lambda>\mu
+  1,\quad \text{if }\lambda\leq\mu, \\
+  \left(\frac{\mu}{\lambda}\right)^N,\quad \text{if }\lambda>\mu.
   \end{array}\right.

### 2022-12-04 EvolutionaryRescue1.tex
Add files via upload

@@
 \section{Introduction}
-The question of what role does chromosomal instability (CIN) play in the emergence of cancer has been studied extensively in the past decades \cite{michor2005can,christine2018understanding,nowak2002role,pavelka2010dr,komarova2003mutation,zhu2018cellular}.
+%The question of what role does chromosomal instability (CIN) play in the emergence of cancer has been studied extensively in the past decades \cite{michor2005can,christine2018understanding,nowak2002role,pavelka2010dr,komarova2003mutation,zhu2018cellular}.
 One hypothesis is that CIN facilitates tumorgenesis by accelerating the removal of tumor suppression genes (TSG) and subsequent appearance of cancer.
@@
 The case of two chromosomal loss events is taken into consideration because the fitness cost would be very large and the cell would die.
-
-Populations adapted to a certain environment are vulnerable to change in the environment which might cause extinction of the population.
+%
+%Populations adapted to a certain environment are vulnerable to change in the environment which might cause extinction of the population.
 Adaptation is a race against time as the population size decreases in the new environment.
@@
 There exists three possible ways for a population to survive environmental degradation: migration to a new habitat similar to the one before the onset of environmental changes \cite{cobbold2020should}, adaptation by phenotypic plasticity which involves no changes in the genotype\cite{carja2019evolutionary,carja2017evolutionary} and adaptation through genetic mutations\cite{uecker2014evolutionary,uecker2016role,uecker2011fixation}.
+%
+%Majority of studies assume that the fitness of the wildtype and mutant are time homogeneous. An exception is \cite{marrec2020adapt} fitness of wildtype and mutant are taken to be functions of time.
 
-Majority of studies assume that the fitness of the wildtype and mutant are time homogeneous. An exception is \cite{marrec2020adapt} fitness of wildtype and mutant are taken to be functions of time.
+%%%%%%
+
+Populations adapted to a certain environment are vulnerable to environmental changes, which might cause extinction of the population.
+Examples of such environmental changes include climate change, invasive species or the onset of drug therapies.
+Adaptation is a race against time as the population size decreases in the new environment\cite{tanaka2022surviving}.
+{\em Evolutionary rescue} is the process where the population acquires mutations that increase the fitness of the population in the new environment such that extinction is averted.
+It is mathematically equivalent to the problem of crossing of fitness valley \cite{weissman2009rate,weissman2010rate}.
+We are interested in how likely is a population to adapt to the changing environment through evolutionary adaptation.
+There are three potential ways for a population to survive environmental degradation: migration to a new habitat similar to the one before the onset of environmental change \cite{cobbold2020should}; adaptation by phenotypic plasticity without genetic change \cite{carja2019evolutionary,carja2017evolutionary,levien2021non}; and adaptation through genetic mutations \cite{uecker2014evolutionary,uecker2016role,uecker2011fixation}.
+
+Aberrations in chromosome copy number are exploited by cancer cells in order to survive under conditions of selective pressure such as chemotherapy. % REF
+Cancer cells are likely to be aneuploid, which also correlates with poor patient outcomes \cite{ben2020context}.
+One mechanism by which aneuploidy has been hypothesized to rescue cancer is by providing phenotypic variation and increasing heterogeneity in a tumor.
+
+The role of chromosomal instability (CIN) in the emergence of cancer has been studied extensively in the past decades \cite{michor2005can,christine2018understanding,nowak2002role,pavelka2010dr,komarova2003mutation,zhu2018cellular}.
+One hypothesis is that CIN facilitates tumor genesis by accelerating the removal of tumor suppression genes (TSG) and subsequent appearance of cancer.
+The deletion of tumor suppression genes can happen in two ways: two point mutations deleting both alleles of the TSG (assuming a diploid genotype), or one point mutation and one chromosomal loss event.
+Initial studies have shown that aneuploidy can have a significant role in the deletion of the the tumor suppressing genes when compared to two consecutive point mutations\cite{nowak2002role,komarova2003mutation,michor2005can,komarova2008selective}.
+% REF ok
+However, when taking into account that the appearance of aneuploidy requires a mutation to trigger chromosomal instability, the probability that CIN precedes tumor genesis is highly unlikely.
+
+Majority of studies assume that the fitness of the wildtype and mutant are independent of time.
+An exception is \cite{marrec2020adapt}, where the fitness of the wildtype and mutant are time dependent.
+Additionally, \cite{uecker2011fixation} investigated the probability of fixation of a beneficial mutation in a variable environment with arbitrary time-dependent selection coefficient and population size.
 
@@
 \draw [thick](-0.25+10,-1.25+2-4) -- (-0.25+10,-2.75+2-4);
-\draw [,color=blue,thick](-0.7+10,-1.25+2-4) -- (-0.7+10,-2.75+2-4);
+\draw [,color=blue,very thick](-0.7+10,-1.25+2-4) -- (-0.7+10,-2.75+2-4);
 
 \draw [thick](0.05+5,-1.25+2-8) -- (0.05+5,-2.75+2-8);
-\draw [,color=blue,thick](-0.55+5,-1.25+2-8) -- (-0.55+5,-2.75+2-8);
+\draw [,color=blue,very thick](-0.55+5,-1.25+2-8) -- (-0.55+5,-2.75+2-8);
 %\draw [thick](-0.55+5,-1.25-2) -- (-0.55+5,-2.75-2);
@@
 \end{tikzpicture}
-\caption{
+\caption{Diagram of the evolutionary rescue process which involves aneuploidy.
+The wildtype cell can either acquire trisomy at rate $u$ or aquire a fitness increasing mutation at rate $v$.
+Aneuploidy cells can  aquire a fitness increasing mutation at rate $v$.
+Finally, the aneuploidy cells can lose trisomy at rate $u$.
 }
@@
 \begin{align}\label{survprobwinitial}
-p_w&\approx\frac{vp_m+up_a}{\lambda_w-\mu_w-u-v}\\
+p_w&\approx-\frac{vp_m+up_a}{\lambda_w-\mu_w-u-v}\\
 \nonumber
@@
 \begin{align}
-p_{est}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right].
+p_{est}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}-\frac{uv^2\lambda_a\Delta_m^2}{8\lambda_m^2\Delta_a^3}\right)\right].
 \end{align}
@@
 \begin{align}\label{survprobwinitial}
-p_w&\approx\frac{vp_m+up_a}{\lambda_w-\mu_w-u-v}\\
+p_w&\approx-\frac{vp_m+up_a}{\lambda_w-\mu_w-u-v}\\
 \nonumber
 &\approx\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-v\right)}-\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\ \label{survprobw2}
-&\approx\frac{v}{\lambda_m\left(\lambda_w-\mu_w\right)}\left[\frac{u\left(\lambda_m-\mu_m\right)}{\left(\lambda_a-\mu_a\right)}-\left(\lambda_m-\mu_m\right)\right],
+&\approx\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_w-\mu_w\right)}\left[\frac{u}{\left(\lambda_a-\mu_a\right)}-1\right],
 \end{align}
@@
 \begin{equation}\label{survprobwapprox}
-p_w=\frac{v}{\lambda_m\Delta_w}\left(\frac{u\Delta_m}{\Delta_a}-\Delta_m\right).
+p_w=\frac{v\Delta_m}{\lambda_m\Delta_w}\left(\frac{u}{\Delta_a}-1\right).
 \end{equation}
@@
 \begin{align}
-p_{est}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{vN}{\lambda_m\Delta_w}\left(\Delta_m-\frac{u\Delta_m}{\Delta_a}\right)\right],
+p_{est}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right],
 \end{align}
+
+\begin{figure}[!t]
+ \vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{P_est.pdf}
+\caption{}
+\label{P_est}
+\end{figure}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{figure}[!t]
+ \vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{DeleteriousPlot.pdf}
+\caption{}
+\label{DeleteriousPlot}
+\end{figure}
 \section{Second case: $4\lambda_avp_m>\left(\lambda_a-\mu_a-v\right)^2$}
@@
 \begin{align*}
-p_w&\approx\frac{1}{\lambda_a-\mu_a-u-v}\left[v\frac{\lambda_m-\mu_m}{\lambda_m}+u\sqrt{\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_a\lambda_m}}\right]\\
-&=\sqrt{\frac{\lambda_m-\mu_m}{\lambda_m}}\frac{\sqrt{v}}{\lambda_a-\mu_a-u-v}\left[\frac{u}{\sqrt{\lambda_a}}+\sqrt{\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}}\right]\\
-&=\sqrt{\frac{\Delta_m}{\lambda_m}}\frac{\sqrt{v}}{\Delta_a-u-v}\left[\frac{u}{\sqrt{\lambda_a}}+\sqrt{\frac{v\Delta_m}{\lambda_m}}\right],
+p_w&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[v\frac{\lambda_m-\mu_m}{\lambda_m}+u\sqrt{\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_a\lambda_m}}\right]\\
+&=-\sqrt{\frac{\lambda_m-\mu_m}{\lambda_m}}\frac{\sqrt{v}}{\lambda_w-\mu_w-u-v}\left[\frac{u}{\sqrt{\lambda_a}}+\sqrt{\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}}\right]\\
+&=-\sqrt{\frac{\Delta_m}{\lambda_m}}\frac{\sqrt{v}}{\Delta_w-u-v}\left[\frac{u}{\sqrt{\lambda_a}}+\sqrt{\frac{v\Delta_m}{\lambda_m}}\right],
 \end{align*}
@@
 \begin{align*}
-p_{est}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[-N\sqrt{\frac{\Delta_m}{\lambda_m}}\frac{\sqrt{v}}{\Delta_a-u-v}\left(\frac{u}{\sqrt{\lambda_a}}+\sqrt{\frac{v\Delta_m}{\lambda_m}}\right)\right].
+p_{est}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[N\sqrt{\frac{\Delta_m}{\lambda_m}}\frac{\sqrt{v}}{\Delta_w-u-v}\left(\frac{u}{\sqrt{\lambda_a}}+\sqrt{\frac{v\Delta_m}{\lambda_m}}\right)\right].
 \end{align*}
 
+\begin{equation}
+p_{est}\sim\left\{
+  \begin{array}{@{}ll@{}}
+  1-\exp\left[N\sqrt{\frac{\Delta_m}{\lambda_m}}\frac{\sqrt{v}}{\Delta_w-u-v}\left(\frac{u}{\sqrt{\lambda_a}}+\sqrt{\frac{v\Delta_m}{\lambda_m}}\right)\right],\quad\text{if }4\lambda_avp_m>\left(\Delta_a-v\right)^2,\\
+   1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right],\quad\text{if }\Delta_a<0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2,\\
+   1-\exp\left[\frac{N}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right],\quad\text{if }\Delta_a>0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2.
+  \end{array}\right.
+\end{equation}
+\section{Logistic}
+We want to have birth and death rates which depends on the population size of wildtype $w$, aneuploidy $a$ and mutant $m$ cells:
+\begin{align*}
+&\lambda_w'=\lambda_w,\quad\mu_w'=\mu_w\\
+&\lambda_a'=C_1+\lambda_a\left(1-\frac{w+a+m}{K}\right),\quad \mu_a'=C_1\\
+&\lambda_m'=C_2+\lambda_m\left(1-\frac{w+a+m}{K}\right),\quad \mu_m'=C_2
+\end{align*}
+where $C_1, C_2>0$ are constants.
 \section{Standing genetic variation}
@@
 
+\section{Mean time}
+
 \section{Discussion}
@@
 \label{SurvPlot}
+\end{figure}
+
+\begin{figure}[!t]
+ \vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{HeatMap_p_est.pdf}
+\caption{}
+\label{SurvHeatMapPlot}
 \end{figure}

### 2022-12-04 Python Code/EvolutionaryRescueExponentialDeleteriousAneuploidy.py
Add files via upload

@@ -1,0 +1,150 @@
+#import matplotlib.inline
+import numpy as np
+import matplotlib.pyplot as plt
+import numba
+import scipy.stats as stats
+from scipy.integrate import solve_ivp
+import seaborn as sns
+sns.set_context('talk')
+
+from functools import partial
+
+def plot_mp(t, n, a, m, label='', ax=None):
+    if ax is None:
+        fig, ax = plt.subplots()
+    ax.plot(t, n, lw=3, label='wildtype ' + label)
+    ax.plot(t, a, lw=3, label='aneuploidy ' + label)
+    ax.plot(t, m, lw=3, label='mutant ' + label)
+    ax.set_xlabel('Time')
+    ax.set_ylabel('Count')
+    ax.legend()
+    return ax
+
+@numba.jit # 2-fold faster
+def get_rates(n, a, m, λn, λa, λm, μn, μa, μm, v, w):
+    return np.array([
+        λn * n,    
+        μn * n,     
+        v * n,
+        w * n,
+        λa * a, 
+        μa * a,
+        w * a,
+        λm * m,    
+        μm * m,
+    ])
+    
+@numba.jit # 2-fold faster
+def draw_time(rates):
+ #   print(rates,'r')
+    total_rate = rates.sum()
+    
+#    print(rates.sum(),'hello')
+    if total_rate > 0:
+        return np.random.exponential(1/total_rate)
+    if total_rate == 0:
+        return 1
+
+# @numba.jit # jit causes errors with multinomial
+def draw_reaction(rates):
+    if rates.sum() > 0:
+        rates /= rates.sum()
+        return np.random.multinomial(1, rates).argmax()
+    if rates.sum() == 0:
+        return 9
+
+#rates = get_rates(3, 100, βm, βp, γp)
+#rates /= rates.sum()
+    
+updates = np.array([
+    [1, 0, 0],  # growth wild type
+    [-1, 0, 0], # death wild type
+    [-1, 1, 0], #mutation w->a
+    [-1, 0, 1], #mutation w->m
+    [0, 1, 0],  # growth aneuploidy type
+    [0, -1, 0],  # death aneuploidy type
+    [0, -1, 1], # #mutation a->m
+    [0, 0, 1],  # growth mutant type
+    [0, 0, -1],   # death mutant type
+    [0, 0, 0]
+])
[…]

### 2023-01-01 EvolutionaryRescue1.tex
Add files via upload

@@
 There are three potential ways for a population to survive environmental degradation: migration to a new habitat similar to the one before the onset of environmental change \cite{cobbold2020should}; adaptation by phenotypic plasticity without genetic change \cite{carja2019evolutionary,carja2017evolutionary,levien2021non}; and adaptation through genetic mutations \cite{uecker2014evolutionary,uecker2016role,uecker2011fixation}.
+
+Chromosomal instability (CIN) is the process during mitosis in which cells suffer from chromosome segration defects, which leads to a state called aneuploidy where cells are characterized by structural changes of the chromosomes and copy number alterations \cite{schukken2018cin}.
 
@@
 Given an initial population consisting of $N$ wildtype cancer cells, the probability that the population will survive is given by: 
-\begin{align}
+\begin{align}\label{AneuploidyDeleteriousApprox}
 p_{est}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right],
@@
 \includegraphics[width=1\textwidth]{P_est.pdf}
-\caption{}
+\caption{Plot of the survival probability of an initial population consisting of $10^{4}$ wildtype cells as a function of $\Delta_a=\lambda_a-\mu_a$ for various values of $\Delta_w=\lambda_a-\mu_a$.
+The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximations \eqref{AneuploidyDeleteriousApprox}.}
 \label{P_est}
@@
  \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{DeleteriousPlot.pdf}
+\includegraphics[width=1\textwidth]{P_est_divergence.pdf}
+\caption{Plot of the survival probability of an initial population consisting of $10^{4}$ wildtype cells as a function of $\Delta_w=\lambda_w-\mu_w$ for various values of $\Delta_a=\lambda_a-\mu_a$.
+The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximations \eqref{AneuploidyDeleteriousApprox}.}
+\label{P_est}
+\end{figure}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{figure}[!t]
+ \vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{DeleteriousTauLeapPlot.pdf}
 \caption{}
@@
 \end{figure}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Second case: $4\lambda_avp_m>\left(\lambda_a-\mu_a-v\right)^2$}
@@
 &=\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\sqrt{\lambda_avp_m}}}{2\lambda_a}\\
-&\approx\sqrt{\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_a\lambda_m}}
+&=\frac{\left(\lambda_a-\mu_a-v+2\sqrt{\lambda_avp_m}\right)^2+4\lambda_avp_m}{8\lambda_a\sqrt{\lambda_avp_m}}\\
+&=\frac{4\lambda_avp_m+4\lambda_avp_m\left(1+\frac{\lambda_a-\mu_a-v}{2\sqrt{\lambda_avp_m}}\right)^2}{8\lambda_a\sqrt{\lambda_avp_m}}\\
+&=\frac{1}{2\lambda_a}\left(\lambda_a-\mu_a-v+2\sqrt{\lambda_avp_m}\right)
 \end{align*}
-In the last line of the above we have used the following inequalities:
-\begin{align*}
-\lambda_a-\mu_a,\,v,\,\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\sqrt{\lambda_avp_m}}\ll1.
-\end{align*}
+%In the last line of the above we have used the following inequalities:
+%\begin{align*}
+%\lambda_a-\mu_a,\,v,\,\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\sqrt{\lambda_avp_m}}\ll1.
+%\end{align*}
 As a result, we have from \eqref{survprobwinitial} the probability of rescue of a population starting from one wildtype individual:
 \begin{align*}
-p_w&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[v\frac{\lambda_m-\mu_m}{\lambda_m}+u\sqrt{\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_a\lambda_m}}\right]\\
-&=-\sqrt{\frac{\lambda_m-\mu_m}{\lambda_m}}\frac{\sqrt{v}}{\lambda_w-\mu_w-u-v}\left[\frac{u}{\sqrt{\lambda_a}}+\sqrt{\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}}\right]\\
-&=-\sqrt{\frac{\Delta_m}{\lambda_m}}\frac{\sqrt{v}}{\Delta_w-u-v}\left[\frac{u}{\sqrt{\lambda_a}}+\sqrt{\frac{v\Delta_m}{\lambda_m}}\right],
+p_w&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[v\frac{\lambda_m-\mu_m}{\lambda_m}+\frac{u}{2\lambda_a}\left(\lambda_a-\mu_a-v+2\sqrt{\lambda_avp_m}\right)\right]\\
+&=-\frac{1}{\lambda_w-\mu_w-u-v}\left[v\frac{\lambda_m-\mu_m}{\lambda_m}+\frac{u}{2\lambda_a}\left(\lambda_a-\mu_a-v\right)+u\sqrt{\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_a\lambda_m}}\right]\\
+&=-\frac{1}{\Delta_w-u-v}\left[v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right],
 \end{align*}
@@
 \begin{align*}
-p_{est}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[N\sqrt{\frac{\Delta_m}{\lambda_m}}\frac{\sqrt{v}}{\Delta_w-u-v}\left(\frac{u}{\sqrt{\lambda_a}}+\sqrt{\frac{v\Delta_m}{\lambda_m}}\right)\right].
+p_{est}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w-u-v}\left(v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right)\right].
 \end{align*}
@@
   \begin{array}{@{}ll@{}}
-  1-\exp\left[N\sqrt{\frac{\Delta_m}{\lambda_m}}\frac{\sqrt{v}}{\Delta_w-u-v}\left(\frac{u}{\sqrt{\lambda_a}}+\sqrt{\frac{v\Delta_m}{\lambda_m}}\right)\right],\quad\text{if }4\lambda_avp_m>\left(\Delta_a-v\right)^2,\\
+  1-\exp\left[\frac{N}{\Delta_w-u-v}\left(v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right)\right],\quad\text{if }4\lambda_avp_m>\left(\Delta_a-v\right)^2,\\
    1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right],\quad\text{if }\Delta_a<0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2,\\

### 2023-01-04 EvolutionaryRescue1.tex
Add files via upload

@@
 \caption{Plot of the survival probability of an initial population consisting of $10^{4}$ wildtype cells as a function of $\Delta_a=\lambda_a-\mu_a$ for various values of $\Delta_w=\lambda_a-\mu_a$.
-The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximations \eqref{AneuploidyDeleteriousApprox}.}
+The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximation \eqref{AneuploidyDeleteriousApprox}.}
 \label{P_est}
@@
 \includegraphics[width=1\textwidth]{DeleteriousTauLeapPlot.pdf}
-\caption{}
+\caption{Plot of the probability of survival of a population as a function of the initial population size of wildtype cells.
+The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximation \eqref{pestthirdcase}.}
 \label{DeleteriousPlot}
@@
 Given an initial population consisting of $N$ wildtype cancer cells, the probability that the population will survive is given by: 
-\begin{align*}
-p_{est}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w-u-v}\left(v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right)\right].
-\end{align*}
+\begin{align}\label{pestthirdcase}
+p_{est}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w-u-v}\left(v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right)\right],
+\end{align}
+which we plot in Figure \ref{DeleteriousPlot} where we compare with numerical simulations and the exact result \eqref{survprobw}.
 
+The probability of evolutionary rescue is given by:
 \begin{equation}

### 2023-01-10 EvolutionaryRescue1.tex
Add files via upload

@@
 \author[1]{Remus Stana}
-\author[2]{‪Daniel Weissman}
+\author[2]{Daniel Weissman}
 \author[1]{Yoav Ram}
@@
 \affil[2]{Department of Physics, Emory University, Atlanta, United States of America}
+
 \maketitle
+
 \begin{abstract}
@@
 \draw [thick](0.05+5,-1.25+2) -- (0.05+5,-2.75+2);
-\draw [thick](-0.55+5,-1.25+2) -- (-0.55+5,-2.75+2);
+\draw [very thick, blue](-0.55+5,-1.25+2) -- (-0.55+5,-2.75+2);
 
-\draw [thick](0.05+10,-1.25+2) -- (0.05+10,-2.75+2);
-\draw [thick](-0.55+10,-1.25+2) -- (-0.55+10,-2.75+2);
+\draw [very thick, blue](0.05+10,-1.25+2) -- (0.05+10,-2.75+2);
+\draw [very thick, blue](-0.55+10,-1.25+2) -- (-0.55+10,-2.75+2);
 
@@
 
-\draw [thick](-0.55+10,-1.25-2) -- (-0.55+10,-2.75-2);
+\draw [very thick, blue](-0.55+10,-1.25-2) -- (-0.55+10,-2.75-2);
 \end{tikzpicture}
-\caption{
+\caption{Diagram of tumorgenesis by which a cell loses both of its tumor suppressing cells.
+The first tumor suppressing gene can be eliminated by a mutation at rate $2u$ or through chromosomal instability with rate $2p$.
+The last TSG can either be eliminated through mutation at rate $u$ or through chromosomal instability with rate $p$.
+Grey vertical lines represent active TSGs while the blue vertical lines represent inactivated TSGs.
 }
@@
 \end{equation}
-\section{Logistic}
-We want to have birth and death rates which depends on the population size of wildtype $w$, aneuploidy $a$ and mutant $m$ cells:
-\begin{align*}
-&\lambda_w'=\lambda_w,\quad\mu_w'=\mu_w\\
-&\lambda_a'=C_1+\lambda_a\left(1-\frac{w+a+m}{K}\right),\quad \mu_a'=C_1\\
-&\lambda_m'=C_2+\lambda_m\left(1-\frac{w+a+m}{K}\right),\quad \mu_m'=C_2
-\end{align*}
-where $C_1, C_2>0$ are constants.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\section{Logistic}
+%We want to have birth and death rates which depends on the population size of wildtype $w$, aneuploidy $a$ and mutant $m$ cells:
+%\begin{align*}
+%&\lambda_w'=\lambda_w,\quad\mu_w'=\mu_w\\
+%&\lambda_a'=C_1+\lambda_a\left(1-\frac{w+a+m}{K}\right),\quad \mu_a'=C_1\\
+%&\lambda_m'=C_2+\lambda_m\left(1-\frac{w+a+m}{K}\right),\quad \mu_m'=C_2
+%\end{align*}
+%where $C_1, C_2>0$ are constants.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Standing genetic variation}
@@
 \begin{align*}
-F\left(f\right)=\frac{ 1-\e^{-fNp_a}}{1-\e^{-\left[\left(1-f\right)p_w+fp_a\right]N}}.
+F\left(f\right)=\frac{p_{old}}{p_{total}}=\frac{ 1-\e^{-fNp_a}}{1-\e^{-\left[\left(1-f\right)p_w+fp_a\right]N}}.
 \end{align*}
 We let $F=\frac{1}{2}$ and we obtain:
-\begin{equation*}
+\begin{equation}\label{halfeqstandvar}
 f^*\approx\frac{p_w}{p_w+p_a},
-\end{equation*}
+\end{equation}
 wher we have used the expansion $\e^x\approx 1+x$.
@@
 \includegraphics[width=1\textwidth]{HeatMap_p_est.pdf}
-\caption{}
+\caption{Contours of the survival probability of an initial population consisting of $10^{3}$ wildtype cells as a function of the
+   parameters $\Delta_w=\lambda_w-\mu_w$ and $\Delta_a=\lambda_a-\mu_a$.}
 \label{SurvHeatMapPlot}
@@
 \includegraphics[width=1\textwidth]{FractionPlot.pdf}
-\caption{}
+\caption{Plot of the fraction of the cases the population is rescue by standing genetic variation as a function of the fraction of initial cells which are aneuploid.
+The red vertical line highlights the the value of $f$ for which half the times the population is rescues by aneuploid cells while the pink line is our approximation \eqref{halfeqstandvar}.
+For this plot we have chosen the following parameters: $N=10^2, \lambda_w=1-10^{-2}, \lambda_a=1+10^{-2},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1.$}
 \label{FractionPlot}

### 2023-01-31 EvolutionaryRescue1.tex
Add files via upload

@@
 \draw [thick](-0.25+10,-1.25+2-4) -- (-0.25+10,-2.75+2-4);
-\draw [,color=blue,very thick](-0.7+10,-1.25+2-4) -- (-0.7+10,-2.75+2-4);
+\draw [color=blue,very thick](-0.7+10,-1.25+2-4) -- (-0.7+10,-2.75+2-4);
 
 \draw [thick](0.05+5,-1.25+2-8) -- (0.05+5,-2.75+2-8);
-\draw [,color=blue,very thick](-0.55+5,-1.25+2-8) -- (-0.55+5,-2.75+2-8);
+\draw [color=blue,very thick](-0.55+5,-1.25+2-8) -- (-0.55+5,-2.75+2-8);
 %\draw [thick](-0.55+5,-1.25-2) -- (-0.55+5,-2.75-2);
@@
 
+\begin{align}
+\Delta_a^*=2vp_m+v+2\sqrt{vp_m\left(vp_m+\mu_a+v\right)}
+\end{align}
+
 The probability of evolutionary rescue is given by:
@@
 \section{Mean time}
+%%%
+\begin{figure}[!t]
+ \vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{MeanTimeGrowthAneuploidyPlot.pdf}
+\caption{}
+\label{MeanTimeGrowthAneuploidyPlot}
+\end{figure}
+%%%
 
+\begin{align}
+\tau=\tau_1+\tau_2
+\end{align}
+inhomogeneus Poisson process
+\begin{align}
+\tau_1\sim \text{Poisson}\left(\frac{1}{up_aW\left(t\right)}\right)
+\end{align}
+where 
+\begin{equation}
+W\left(t\right)=W_0\e^{\Delta_wt}
+\end{equation}
+\begin{align}
+f\left(\tau_1\right)=R\left(\tau_1\right)\e^{-\int_0^{\tau_1} R(t)\d t}
+\end{align}
+\begin{align*}
+P\left(\tau_1<t\right)&=P\left(\tau_1<t|M\left(t\rightarrow\infty\right)\neq0\right)P\left(M\left(t\rightarrow\infty\right)\neq0\right)\\
+&+P\left(\tau_1<t|M\left(t\rightarrow\infty\right)=0\right)P\left(M\left(t\rightarrow\infty\right)=0\right)\\
+P\left(\tau_1<t\right)&=P\left(\tau_1<t|M\left(t\rightarrow\infty\right)\neq0\right)P\left(M\left(t\rightarrow\infty\right)\neq0\right)
+\end{align*}
+\begin{align}
+P\left(\tau_1<t|M\left(t\rightarrow\infty\right)\neq0\right)=\frac{P\left(\tau_1<t\right)}{1-\left(1-p_w\right)^N}
+\end{align}
+where we have used
+\begin{equation}
+P\left(M\left(t\rightarrow\infty\right)\neq0\right)=1-\left(1-p_w\right)^N
+\end{equation}
+\begin{align}
+f\left(\tau_1|M\left(t\rightarrow\infty\right)\neq0\right)=\frac{R\left(\tau_1\right)\e^{-\int_0^{\tau_1} R(t)\d t}}{1-\left(1-p_w\right)^N}
+\end{align}
+\begin{align}
+\mathbb{E}\left[\tau_1\right]&=\frac{\int_0^\infty\e^{-\int_0^\tau R(t)\d t}\d\tau}{1-\left(1-p_w\right)^N}\\
+&=\frac{\int_0^\infty\e^{-uNp_a\frac{\e^{\Delta_w\tau}-1}{\Delta_w}}\,\d\tau}{1-\left(1-p_w\right)^N}
+\end{align}
+\begin{align}
+P_2\left(\tau_2\right)&=\int_0^\infty \d w P\left(W_l=w\right)\left[1-\e^{-vp_mW_l\left(\tau_2\right)}\right]\\
+&=1-\mathbb{E}\left[\e^{-vp_mW_l\left(\tau_2\right)}\right]
+\end{align}
+\begin{align}
+a_\pm=\frac{2-\delta-y\pm\sqrt{\left(\delta+y\right)^2-4y}}{2\left(1-\delta\right)}
+\end{align}
+\begin{align}
+P_2\left(\tau_2\right)=\frac{\left(a_+-1\right)\left(1-a_-\right)\left(1-\e^{\left[-\left(1-\delta\right)\left(a_+-a_-\right)t\right]}\right)}{a_+-1+\left(1-a_-\right)\e^{\left[-\left(1-\delta\right)\left(a_+-a_-\right)t\right]}}
+\end{align}
+with $y=vp_m$ in $a_\pm$.
+\begin{align}
+\mathbb{E}\left[\tau_2\right]=\int_0^\infty\left(1-\frac{P_2\left(t\right)}{p_m}\right)\d t
+\end{align}
 \section{Discussion}
@@
 The red vertical line highlights the the value of $f$ for which half the times the population is rescues by aneuploid cells while the pink line is our approximation \eqref{halfeqstandvar}.
-For this plot we have chosen the following parameters: $N=10^2, \lambda_w=1-10^{-2}, \lambda_a=1+10^{-2},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1.$}
+For this plot we have chosen the following parameters: $N=10^3, \lambda_w=1-10^{-2}, \lambda_a=1-10^{-4},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1.$}
 \label{FractionPlot}

### 2023-02-06 EvolutionaryRescue1.tex
Add files via upload

@@
 Additionally, \cite{uecker2011fixation} investigated the probability of fixation of a beneficial mutation in a variable environment with arbitrary time-dependent selection coefficient and population size.
+
+Most studies on evolutionary rescue are interested in the probability that at least one mutation rescues the population but the question of wheter multiple mutations contribute to the survival of the population is less explored.
+One exception is \cite{wilson2017soft} which show that evolutionary rescue is significantly inhanced by soft selective sweeps when multiple mutations contribute.
+
+Evolutionary rescue which requires two succesive mutations has been investigated by \cite{martin2013probability} with the help of diffusion approximation.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\section{Logistic}
-%We want to have birth and death rates which depends on the population size of wildtype $w$, aneuploidy $a$ and mutant $m$ cells:
-%\begin{align*}
-%&\lambda_w'=\lambda_w,\quad\mu_w'=\mu_w\\
-%&\lambda_a'=C_1+\lambda_a\left(1-\frac{w+a+m}{K}\right),\quad \mu_a'=C_1\\
-%&\lambda_m'=C_2+\lambda_m\left(1-\frac{w+a+m}{K}\right),\quad \mu_m'=C_2
-%\end{align*}
-%where $C_1, C_2>0$ are constants.
+\section{Logistic}
+We want to have birth and death rates which depends on the population size of wildtype $w$, aneuploidy $a$ and mutant $m$ cells:
+\begin{align*}
+&\lambda_w'=\lambda_w,\quad\mu_w'=\mu_w\\
+&\lambda_a'=C_1+\lambda_a\left(1-\frac{w+a+m}{K}\right),\quad \mu_a'=C_1\\
+&\lambda_m'=C_2+\lambda_m\left(1-\frac{w+a+m}{K}\right),\quad \mu_m'=C_2
+\end{align*}
+where $C_1, C_2>0$ are constants.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 P\left(\tau_1<t\right)&=P\left(\tau_1<t|M\left(t\rightarrow\infty\right)\neq0\right)P\left(M\left(t\rightarrow\infty\right)\neq0\right)
+\end{align*}
+where we used
+\begin{align*}
+P\left(\tau_1<t|M\left(t\rightarrow\infty\right)=0\right)=0
 \end{align*}

### 2023-02-07 EvolutionaryRescue1.tex
started revising and adding comments

@@
-\documentclass[11pt,a4paper]{article}
+\documentclass[12pt]{extarticle}
+\usepackage{geometry}
+\geometry{
+a4paper,
+total={170mm,257mm},
+left=20mm,
+top=20mm,
+headheight=12pt
+}
+
+\usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent
 \usepackage{amsmath}
@@
 \usepackage{afterpage}
-%\usepackage[margin=1in]{geometry}
-\usepackage{titlesec}
 \usepackage{float}
@@
 \usepackage{scrextend}
+\usepackage[all]{nowidow}
 \usepackage{stfloats}
@@
 \usepackage{authblk}
-\usepackage[margin=0.7in]{geometry}
+\usepackage{cleveref}
+
+% less space before sections 
+% \titlespacing*{<command>}{<left>}{<before-sep>}{<after-sep>}
+\usepackage{titlesec}
+\titlespacing*{\section}
+{0pt}{2ex plus 1ex minus .2ex}{2ex plus .2ex}
+\titlespacing*{\subsection}
+{0pt}{1ex plus 1ex minus .2ex}{1ex plus .2ex}
+\titlespacing*{\paragraph}
+{0pt}{1ex plus 1ex minus .2ex}{1ex plus .2ex}
 
@@
 
+% appendix
+\usepackage[title,page]{appendix}
+\usepackage{chngcntr}
+
+% Supplementary
+% https://support.authorea.com/en-us/article/how-to-create-an-appendix-section-or-supplementary-information-1g25i5a/
+\newcommand{\beginsupplement}{%
+      	\setcounter{table}{0}
+        \renewcommand{\thetable}{S\arabic{table}}%
+        \setcounter{figure}{0}
+        \renewcommand{\thefigure}{S\arabic{figure}}%
+		\setcounter{equation}{0}
+        \renewcommand{\theequation}{A\arabic{equation}}%
+}
+
+% line numbers
+\usepackage[displaymath, mathlines]{lineno}
+\renewcommand\linenumberfont{\normalfont\small\sffamily}
+\linenumbers
+\modulolinenumbers[2]
+
+% Title page
+\title{Modeling the effect of aneuploidy on cancer evolution}
+% Authors
+\renewcommand\Affilfont{\small}
+
+\author[1]{Remus Stana}
+\author[2]{Daniel B. Weissman}
+\author[1,*]{Yoav Ram}
+\affil[1]{School of Zoology, Tel Aviv University}
+\affil[2]{Department of Physics, Emory University}
+\affil[*]{Corresponding author: yoav@yoavram.com}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{document}
-\title{Modeling the effect of aneuploidy on cancer evolution}
-\author[1]{Remus Stana}
-\author[2]{Daniel Weissman}
-\author[1]{Yoav Ram}
-\affil[1]{School of Zoology, Tel Aviv University, Tel Aviv-Yafo, Israel}
-\affil[2]{Department of Physics, Emory University, Atlanta, United States of America}
-
 \maketitle
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{abstract}
@@
 
-\section{Introduction}
+\newpage
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section*{Introduction}
+
+% TODO remove the commented out paragraphs if you dont need them.
+
 %The question of what role does chromosomal instability (CIN) play in the emergence of cancer has been studied extensively in the past decades \cite{michor2005can,christine2018understanding,nowak2002role,pavelka2010dr,komarova2003mutation,zhu2018cellular}.
-One hypothesis is that CIN facilitates tumorgenesis by accelerating the removal of tumor suppression genes (TSG) and subsequent appearance of cancer.
+One hypothesis is that CIN facilitates tumorigenesis by accelerating the removal of tumor suppression genes (TSG) and subsequent appearance of cancer.
 The deletion of tumor suppression genes can happen in two ways (see Figure \ref{figureCIN}): two point mutations deleting both alleles of the TSG or one point mutation and one chromosomal loss event.
@@
 
-Populations adapted to a certain environment are vulnerable to environmental changes, which might cause extinction of the population.
-Examples of such environmental changes include climate change, invasive species or the onset of drug therapies.
-Adaptation is a race against time as the population size decreases in the new environment\cite{tanaka2022surviving}.
-{\em Evolutionary rescue} is the process where the population acquires mutations that increase the fitness of the population in the new environment such that extinction is averted.
-It is mathematically equivalent to the problem of crossing of fitness valley \cite{weissman2009rate,weissman2010rate}.
-We are interested in how likely is a population to adapt to the changing environment through evolutionary adaptation.
-There are three potential ways for a population to survive environmental degradation: migration to a new habitat similar to the one before the onset of environmental change \cite{cobbold2020should}; adaptation by phenotypic plasticity without genetic change \cite{carja2019evolutionary,carja2017evolutionary,levien2021non}; and adaptation through genetic mutations \cite{uecker2014evolutionary,uecker2016role,uecker2011fixation}.
+% OVERVIEW OF INTRODUCTION:
+% - background on CIN in cancer
+% - what hasn't been done? (aneuploidy + drug resistance)
+% - why is that important?
+% - how we tackle this background
+% - summary of our analysis
+% history of evolutionary rescue literature (that is not specific to cancer) should move to literature, and we should have a clear statement on what we add to that literature.
 
-Chromosomal instability (CIN) is the process during mitosis in which cells suffer from chromosome segration defects, which leads to a state called aneuploidy where cells are characterized by structural changes of the chromosomes and copy number alterations \cite{schukken2018cin}.
+\paragraph{Aneuploidy in cancer.} Chromosomal instability (CIN) is the mitotic process in which cells suffer from chromosome mis-segregation that leads to aneuploidy, where cells are characterized by structural changes of the chromosomes and copy number alterations \cite{schukken2018cin}.
+Interestingly, aberrations in chromosome copy number are have been shown to allow cancer cells to survive under stressful conditions such as drug therapy. % REF
+Indeed, cancer cells are often likely to be aneuploid, and aneuploidy is associated with poor patient outcomes \cite{ben2020context}.
 
-Aberrations in chromosome copy number are exploited by cancer cells in order to survive under conditions of selective pressure such as chemotherapy. % REF
-Cancer cells are likely to be aneuploid, which also correlates with poor patient outcomes \cite{ben2020context}.
-One mechanism by which aneuploidy has been hypothesized to rescue cancer is by providing phenotypic variation and increasing heterogeneity in a tumor.
+% this belong in the next paragraph, or remove it altogether
+% One mechanism by which aneuploidy has been hypothesized to affect cancer is by providing phenotypic variation and increasing heterogeneity in a tumor.
 
@@
 Initial studies have shown that aneuploidy can have a significant role in the deletion of the the tumor suppressing genes when compared to two consecutive point mutations\cite{nowak2002role,komarova2003mutation,michor2005can,komarova2008selective}.
-% REF ok
-However, when taking into account that the appearance of aneuploidy requires a mutation to trigger chromosomal instability, the probability that CIN precedes tumor genesis is highly unlikely.
+% TODO mention if studies are theoretical , experimental, or clinical
+However, when taking into account that the appearance of aneuploidy requires a mutation to trigger CIN, the probability that CIN precedes tumor genesis is highly unlikely.
 
-Majority of studies assume that the fitness of the wildtype and mutant are independent of time.
+\paragraph{Evolutionary rescue.} Populations adapted to a certain environment are vulnerable to environmental changes, which might cause extinction of the population.
+Examples of such environmental changes include climate change, invasive species or the onset of drug therapies.
+Adaptation is a race against time as the population size decreases in the new environment~\cite{tanaka2022surviving}.
+{\em Evolutionary rescue} is the process where the population acquires a trait that increases fitness in the new environment such that extinction is averted.
+It is mathematically equivalent to the problem of crossing of fitness valley \cite{weissman2009rate,weissman2010rate}.
+There are three potential ways for a population to survive environmental change: migration to a new habitat similar to the one before the onset of environmental change \cite{cobbold2020should}; adaptation by phenotypic plasticity without genetic modification \cite{carja2019evolutionary,carja2017evolutionary,levien2021non}; and adaptation through genetic modifications, e.g., mutation \cite{uecker2014evolutionary,uecker2016role,uecker2011fixation}.
+
+Models of evolutionary rescue usually assume that the fitness of the wildtype and mutant are independent of time.
 An exception is \cite{marrec2020adapt}, where the fitness of the wildtype and mutant are time dependent.
-Additionally, \cite{uecker2011fixation} investigated the probability of fixation of a beneficial mutation in a variable environment with arbitrary time-dependent selection coefficient and population size.
+Additionally, \cite{uecker2011fixation} investigated the probability of fixation of a beneficial mutation in a variable environment with arbitrary time-dependent selection coefficent and population size.
+Most models focus on the probability that at least one mutation rescues the population.
+How multiple mutations contribute to the survival of the population is less explored.
+One exception is \cite{wilson2017soft} which showed that evolutionary rescue is significantly enhanced by soft selective sweeps when multiple mutations contribute.
+Evolutionary rescue that requires two successive mutations has been investigated by \cite{martin2013probability} with the help of diffusion approximation.
 
-Most studies on evolutionary rescue are interested in the probability that at least one mutation rescues the population but the question of wheter multiple mutations contribute to the survival of the population is less explored.
-One exception is \cite{wilson2017soft} which show that evolutionary rescue is significantly inhanced by soft selective sweeps when multiple mutations contribute.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section*{Model}
 
-Evolutionary rescue which requires two succesive mutations has been investigated by \cite{martin2013probability} with the help of diffusion approximation.
+We follow the number of three types of cancer cells at time $t$: wildtype euploid, $w_t$; wildtype aneuploid, $a_t$; and mutant euploid, $m_t$. 
+These cells grow and die with rates $\lambda_k$ and $\mu_k$ for $k=w, a, m$.
+Wildtype cells become aneuploid at rate $u$. Aneuploid and wildtype cells mutate with rate $v$ (Figure \ref{figureAneuploidy}). 
+Thus, the changes in the number of each cell type is described by
+\begin{subequations}
+\begin{flalign}
+w_t&\rightarrow w_t+1:\quad \lambda_ww_t,\\
+w_t&\rightarrow w_t-1:\quad \left(\mu_w+u+v\right)w_t,\\
+a_t&\rightarrow a_t+1:\quad \lambda_aa_t+uw_t,\\
+a_t&\rightarrow a_t-1:\quad \left(\mu_a+v\right)a_t,\\
+m_t&\rightarrow m_t+1:\quad \lambda_am_t+va_t+vm_t,\\
+m_t&\rightarrow m_t-1:\quad \mu_am_t.
+\end{flalign}
+\end{subequations}
 
+The difference between growth and death is $\Delta_k = \lambda_k-\mu_k$.
+We assume that $\Delta_w<0$ due to drug therapy (or some other stress), and that $\Delta_m$ due to resistance conferred by the mutant.
+We analyze two cases: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, % REF: https://www.nature.com/articles/nrmicro.2016.34
+$0>\Delta_a>\Delta_w$.
+
+% TODO explain briefly what evolutionary rescue means in this model.
+To analyze evolutionary rescue in this model, we use the framework of multitype branching processes. % REF Harris? https://doi.org/10.1016/j.jtbi.2021.110849
+This allows us to find explicit expressions for the survival probability of the population--the probability that it does not become extinct due to stressful conditions, e.g., drug therapy.
+Note that branching processes imply that lineages produced by cells from the initial population grow and die independently of each other.
+This is a necessary simplification that is not biologically realistic, as cells usually compete for resources.
+A more biologically realistic model...
+% add the description of the more realistic model here
+
+The survival probability of a population consisting initially of one wildtype cell satisfies the quadratic equation:
+\begin{equation} \label{quadraticeqev1}
+\begin{aligned}
+1-p_w = & \frac{\mu_w}{\lambda_w+\mu_w+u+v} + 
+		  \frac{u}{\lambda_w+\mu_w+u+v}\left(1-p_a\right) + \\
+		  & \frac{\lambda_w}{\lambda_w+\mu_w+u+v}\left(1-p_w\right)^2 +
+		  \frac{v}{\lambda_w+\mu_w+u+v}\left(1-p_m\right) ,
+\end{aligned}
+\end{equation}
+where the $p_w$, $p_a$, and $p_m$ are the survival probabilities of a population consisting initially of one wildtype cell, one aneuploid cell, and one mutant cell, respectively.
+
+The smallest solution of the quadratic equation \cref{quadraticeqev1} gives the survival probability, which still depends on both $p_a$ and $p_m$,
+\begin{align}\label{survprobw}
+p_w=\frac{\lambda_w-\mu_w-u-v+\sqrt{\left(\lambda_w-\mu_w-u-v\right)^2+4\lambda_w\left(up_a+vp_m\right)}}{2\lambda_w} .
+\end{align}
+
+If the population originally consists of a single aneuploidy cell, then the probability that the population will survive is given by the quadratic equation
+\begin{align}
+1-p_a=\frac{\mu_a}{\lambda_a+\mu_a+v}+\frac{v}{\lambda_a+\mu_a+v}\left(1-p_m\right)+\frac{\lambda_a}{\lambda_a+\mu_a+v}\left(1-p_a\right)^2,
+\end{align}
+whose smallest solution is the survival probability:
+\begin{align}\label{survproba}
+p_a=\frac{\lambda_a-\mu_a-v+\sqrt{\left(\lambda_a-\mu_a-v\right)^2+4\lambda_avp_m}}{2\lambda_a}.
+\end{align}
+
+The probability that a single mutant cell survives is given by (Appendix \ref{AppendixSurvLin}):
+\begin{equation}
+p_m = \left\{
+  \begin{array}{@{}ll@{}}
+  \frac{\lambda_m-\mu_m}{\lambda_m},\quad\text{if }\lambda_m>\mu_m \\
+   0,\quad\text{else}.
+  \end{array}\right.
+\end{equation}
+
+% TODO add subsection on the Gillespie simulations
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% TODO combine all diagram figures to a single figure with multiple panels. and make it fit in a single page.
 \begin{figure}
@@
 \end{tikzpicture}
-\caption{Diagram of tumorgenesis by which a cell loses both of its tumor suppressing cells.
+\caption{Diagram of tumorigenesis by which a cell loses both of its tumor suppressing cells.
 The first tumor suppressing gene can be eliminated by a mutation at rate $2u$ or through chromosomal instability with rate $2p$.
@@
 \caption{Diagram of the evolutionary rescue process which involves aneuploidy.
-The wildtype cell can either acquire trisomy at rate $u$ or aquire a fitness increasing mutation at rate $v$.
-Aneuploidy cells can  aquire a fitness increasing mutation at rate $v$.
+The wildtype cell can either acquire trisomy at rate $u$ or acquire a fitness increasing mutation at rate $v$.
+Aneuploidy cells can  acquire a fitness increasing mutation at rate $v$.
 Finally, the aneuploidy cells can lose trisomy at rate $u$.
@@
 
-\section{Model}
-We model evolutionary rescue of cancer cells through aneuploidy with the help of multitype branching processes: we have three populations of cancer cells, wildtype euploid $w_t$, wildtype aneuploid $a_t$ and mutant euploid cells $m_t$ which proliferate and die with rates $\lambda_w,  \lambda_a,  \lambda_m$ and $\mu_w, \mu_a, \mu_m$, respectively.
-Wildtype cells can acquire aneuploidy at rate $u$ while aneuploidy and wildtype cells can mutate with rate $v$ (se Figure \ref{figureAneuploidy}).
-Changes in the number of wildtype euploid, wildtype aneuploid and mutant euploid cells occur at rates:
-\begin{subequations}
-\begin{flalign}
-w_t&\rightarrow w_t+1:\quad \lambda_ww_t,\\
-w_t&\rightarrow w_t-1:\quad \left(\mu_w+u+v\right)w_t,\\
-a_t&\rightarrow a_t+1:\quad \lambda_aa_t+uw_t,\\
-a_t&\rightarrow a_t-1:\quad \left(\mu_a+v\right)a_t,\\
-m_t&\rightarrow m_t+1:\quad \lambda_am_t+va_t+vm_t,\\
-m_t&\rightarrow m_t-1:\quad \mu_am_t.
-\end{flalign}
-\end{subequations}
-Given that we have used branching processes, the lineages produced by cells from the initial population will be independent of each other.
-This is not a biologically realistic model as cells compete for resources in tumors.
-We will study a more biologically realistic model in Section.
-The survival probability of a population consisting initially of one wildtype cell satisfies the quadratic equation:
-\begin{align}\nonumber
-1-p_w&=\frac{\mu_w}{\lambda_w+\mu_w+u+v}+\frac{u}{\lambda_w+\mu_w+u+v}\left(1-p_a\right)+\frac{\lambda_w}{\lambda_w+\mu_w+u+v}\left(1-p_w\right)^2\\ \label{quadraticeqev1}
-&+\frac{v}{\lambda_w+\mu_w+u+v}\left(1-p_m\right),
-\end{align}
-where the $p_a$ and $p_w$ are the survival probabilities of a population consisting initially of one aneuploidy and wildtype cell, respectively.
-
- The smallest solution of the quadratic equation \eqref{quadraticeqev1} gives the survival probability:
-\begin{align}\label{survprobw}
-p_w=\frac{\lambda_w-\mu_w-u-v+\sqrt{\left(\lambda_w-\mu_w-u-v\right)^2+4\lambda_w\left(up_a+vp_m\right)}}{2\lambda_w}.
-\end{align}
-If the population originally consists of a single aneuploidy cell, then the probability that the population will survive is given by the following quadratic equation:
-\begin{align}
-1-p_a=\frac{\mu_a}{\lambda_a+\mu_a+v}+\frac{v}{\lambda_a+\mu_a+v}\left(1-p_m\right)+\frac{\lambda_a}{\lambda_a+\mu_a+v}\left(1-p_a\right)^2,
-\end{align}
-whose smallest solution is the survival probability:
-\begin{align}\label{survproba}
-p_a=\frac{\lambda_a-\mu_a-v+\sqrt{\left(\lambda_a-\mu_a-v\right)^2+4\lambda_avp_m}}{2\lambda_a}.
-\end{align}
-The probability that a single mutant cell will survive is given by (see Appendix \ref{AppendixSurvLin}):
-\begin{equation}
-p_m=\left\{
-  \begin{array}{@{}ll@{}}
-  \frac{\lambda_m-\mu_m}{\lambda_m},\quad\text{if }\lambda_m>\mu_m \\
-   0,\quad\text{else}.
-  \end{array}\right.
-  \end{equation}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\section{First case: $4\lambda_avp_m<\left(\lambda_a-\mu_a-v\right)^2$}
-\subsection{$\lambda_a>\mu_a$ and $\lambda_w<\mu_w$}
-We assume that $\lambda_a>\mu_a$ and $\lambda_w<\mu_w$ and, as a result, we rewrite \eqref{survprobw} and \eqref{survproba} as:
+\newpage
+\section*{Results}
+
+% OVERVIEW OF RESULTS
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% TODO give a descriptive name instead of a mathematical formula
+\subsection*{First case: $4\lambda_avp_m<\left(\lambda_a-\mu_a-v\right)^2$}
+
+\paragraph{Aneuploid cells are drug resistant.} 
+We first assume wildtype cells are susceptible to the drug, $\lambda_w<\mu_w$, whereas aneuploid cells are resistant, $\lambda_a>\mu_a$.
+Thus, we rewrite \cref{survprobw,survproba} as
 \begin{align*}
@@
 \end{align*}
-Making use of the Taylor series expansions:
-\begin{align*}
-\left(1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}\right)^{\frac{1}{2}}&=1+\frac{2\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}+\cdots\\
-\left(1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}\right)^{\frac{1}{2}}&=1+\frac{2\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}+\cdots
-\end{align*}
-we obtain the following approximation for the survival probability of a population consisting of a single individual wildtype cell:
+Using a quadratic Taylor expansion, $\sqrt{1+x}=1+x/2+O(x^2)$, 
+%\begin{align*}
+%\left(1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}\right)^{\frac{1}{2}}&=1+\frac{2\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}+\cdots\\
+%\left(1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}\right)^{\frac{1}{2}}&=1+\frac{2\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}+\cdots
+%\end{align*}
+we obtain the following approximation for the survival probability of a population consisting of a single individual wildtype cell (assuming $u,v \ll 1$),
 \begin{align}\label{survprobwinitial}
-p_w&\approx-\frac{vp_m+up_a}{\lambda_w-\mu_w-u-v}\\
+p_w 
+&\approx -\frac{vp_m+up_a}{\lambda_w-\mu_w-u-v}\\
 \nonumber
-&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{v\left(\lambda_a-\mu_a-u\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-u\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\ \label{survprobw2}
-&\approx-\frac{1}{\lambda_w-\mu_w}\left[\frac{v\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right],
+%&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{v\left(\lambda_a-\mu_a-u\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-u\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\ \label{survprobw2}
+&\approx-\frac{1}{\lambda_w-\mu_w}\left[\frac{v\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right].
 \end{align}
-where in the last line we have used the fact that $u,v\ll1$. Using the notational convention
-\begin{equation}\label{notationalconv}
-\Delta_i=\lambda_i-\mu_i,
-\end{equation}
-we write \eqref{survprobw} as
+
+We write \cref{survprobw} as
 \begin{equation}\label{survprobwapprox}
@@
 \end{equation}
-Given an initial population consisting of $N$ wildtype cancer cells, the probability that the population will survive is given by: 
+So, given an initial population of $N$ wildtype cells, the probability that the population will survive is given by
 \begin{align}
-p_{est}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right],
+p_{est}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right] .
 \end{align}
-which we plot in Figure \ref{SurvPlotNData} as a function of $N$ and in Figure \ref{SurvPlot} as a function of $\lambda_w$.
+\Cref{SurvPlot} shows that the survival probability, $p_{est}$, quickly grows as the wildtype growth rate, $\lambda_w$, increases.
+% Todo I dont think this figure should show lambda_w>1, as in that case the wildtype is resistant and there is no rescue scenario.
+\Cref{SurvPlotNData} show $p_{est}$ as a function of $N$, including comparison of our approximation % ref to equation
+and simulation results.
 
@@
 \end{align}
-\subsection{$\lambda_a<\mu_a$ and $\lambda_w<\mu_w$}
+
+\paragraph{$\lambda_a<\mu_a$ and $\lambda_w<\mu_w$.}
 We assume that $\lambda_a<\mu_a$ and $\lambda_w<\mu_w$ and, as a result, we rewrite \eqref{survprobw} and \eqref{survproba} as:
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{figure}[!t]
  \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{P_est.pdf}
+\includegraphics[width=1\textwidth]{Figures/P_est.pdf}
 \caption{Plot of the survival probability of an initial population consisting of $10^{4}$ wildtype cells as a function of $\Delta_a=\lambda_a-\mu_a$ for various values of $\Delta_w=\lambda_a-\mu_a$.
@@
 \end{figure}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
  \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{P_est_divergence.pdf}
+\includegraphics[width=1\textwidth]{Figures/P_est_divergence.pdf}
 \caption{Plot of the survival probability of an initial population consisting of $10^{4}$ wildtype cells as a function of $\Delta_w=\lambda_w-\mu_w$ for various values of $\Delta_a=\lambda_a-\mu_a$.
@@
  \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{DeleteriousTauLeapPlot.pdf}
+\includegraphics[width=1\textwidth]{Figures/DeleteriousTauLeapPlot.pdf}
 \caption{Plot of the probability of survival of a population as a function of the initial population size of wildtype cells.
@@
 \end{figure}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Second case: $4\lambda_avp_m>\left(\lambda_a-\mu_a-v\right)^2$}
+\subsection*{Second case: $4\lambda_avp_m>\left(\lambda_a-\mu_a-v\right)^2$}
 If we assume that  $4\lambda_avp_m>\left(\lambda_a-\mu_a-v\right)^2$ then we write:
@@
 \end{equation}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Logistic}
+\subsection*{Logistic growth}
 We want to have birth and death rates which depends on the population size of wildtype $w$, aneuploidy $a$ and mutant $m$ cells:
@@
 where $C_1, C_2>0$ are constants.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Standing genetic variation}
+\subsection*{Standing genetic variation}
 Until now we have assumed that the initial population of cells consisted entirely of wildtype cells.
@@
 
-\section{Mean time}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection*{Rescue time}
 %%%
@@
  \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{MeanTimeGrowthAneuploidyPlot.pdf}
+\includegraphics[width=1\textwidth]{Figures/MeanTimeGrowthAneuploidyPlot.pdf}
 \caption{}
@@
 \end{align}
-\section{Discussion}
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{figure}[!t]
  \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{SurvPlot.pdf}
+\includegraphics[width=1\textwidth]{Figures/SurvPlot.pdf}
 \caption{Plot of the probability of survival of a population as a function of the proliferation rate of the wildtype cells.
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{figure}[!t]
  \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{HeatMap_p_est.pdf}
+\includegraphics[width=1\textwidth]{Figures/HeatMap_p_est.pdf}
 \caption{Contours of the survival probability of an initial population consisting of $10^{3}$ wildtype cells as a function of the
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{figure}[!t]
  \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{SurvPlotNData.pdf}
+\includegraphics[width=1\textwidth]{Figures/SurvPlotNData.pdf}
 \caption{Plot of the probability of survival of a population as a function of the initial population size of wildtype cells.
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{figure}[!t]
  \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{FractionPlot.pdf}
+\includegraphics[width=1\textwidth]{Figures/FractionPlot.pdf}
 \caption{Plot of the fraction of the cases the population is rescue by standing genetic variation as a function of the fraction of initial cells which are aneuploid.
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section*{Discussion}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%\section*{References}
+\nolinenumbers
+%\bibliographystyle{agsm}
+%\bibliographystyle{apalike}
+%\bibliographystyle{unsrtnat}
 \bibliographystyle{unsrt}
 \bibliography{evo2022}
+\pagebreak
 
-\appendix
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\appendix
+\begin{appendices}
+\renewcommand{\theequation}{\thesection\arabic{equation}}
+\counterwithin*{equation}{section}
 
-\section{Gillespie algorithm}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section*{Gillespie algorithm}
 \begin{subequations}
@@
 
-\section{Survival probability of a mutant lineage}\label{AppendixSurvLin}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section*{Survival probability of a mutant lineage}\label{AppendixSurvLin}
 The infinitesimal transition probabilities for the simple birth and death process:
@@
 \end{align*}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\end{appendices}
+
+\newpage
+\section*{Supplementary Figures}
+\beginsupplement % https://support.authorea.com/en-us/article/how-to-create-an-appendix-section-or-supplementary-information-1g25i5a/
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \end{document}

### 2023-02-19 EvolutionaryRescue1.tex
remus update feb 17

@@
 \usepackage{graphicx}
+\usepackage{subcaption}
 
@@
 \usepackage{siunitx}
-\usepackage{subfig}
+%\usepackage{subfig}
 \usepackage{authblk}
@@
 \renewcommand\linenumberfont{\normalfont\small\sffamily}
-\linenumbers
+%\linenumbers
 \modulolinenumbers[2]
@@
 
-% TODO remove the commented out paragraphs if you dont need them.
-
-%The question of what role does chromosomal instability (CIN) play in the emergence of cancer has been studied extensively in the past decades \cite{michor2005can,christine2018understanding,nowak2002role,pavelka2010dr,komarova2003mutation,zhu2018cellular}.
-One hypothesis is that CIN facilitates tumorigenesis by accelerating the removal of tumor suppression genes (TSG) and subsequent appearance of cancer.
-The deletion of tumor suppression genes can happen in two ways (see Figure \ref{figureCIN}): two point mutations deleting both alleles of the TSG or one point mutation and one chromosomal loss event.
-The case of two chromosomal loss events is taken into consideration because the fitness cost would be very large and the cell would die.
-%
-%Populations adapted to a certain environment are vulnerable to change in the environment which might cause extinction of the population.
-Adaptation is a race against time as the population size decreases in the new environment.
-Evolutionary rescue is the process where the population acquires mutations which increases the fitness of the population in the new environment such that the extinction in averted.
-We are interested in how likely is a population to adapt to the changing environment through evolution.
-Examples of such changes include climate change, invasive species or the onset of drug therapies.
-There exists three possible ways for a population to survive environmental degradation: migration to a new habitat similar to the one before the onset of environmental changes \cite{cobbold2020should}, adaptation by phenotypic plasticity which involves no changes in the genotype\cite{carja2019evolutionary,carja2017evolutionary} and adaptation through genetic mutations\cite{uecker2014evolutionary,uecker2016role,uecker2011fixation}.
-%
-%Majority of studies assume that the fitness of the wildtype and mutant are time homogeneous. An exception is \cite{marrec2020adapt} fitness of wildtype and mutant are taken to be functions of time.
+% TODO remove the commented out paragraphs if you dont need them. OK
 
@@
 \paragraph{Aneuploidy in cancer.} Chromosomal instability (CIN) is the mitotic process in which cells suffer from chromosome mis-segregation that leads to aneuploidy, where cells are characterized by structural changes of the chromosomes and copy number alterations \cite{schukken2018cin}.
-Interestingly, aberrations in chromosome copy number are have been shown to allow cancer cells to survive under stressful conditions such as drug therapy. % REF
+Interestingly, aberrations in chromosome copy number have been shown to allow cancer cells to survive under stressful conditions such as drug therapy. % REF
 Indeed, cancer cells are often likely to be aneuploid, and aneuploidy is associated with poor patient outcomes \cite{ben2020context}.
@@
 The deletion of tumor suppression genes can happen in two ways: two point mutations deleting both alleles of the TSG (assuming a diploid genotype), or one point mutation and one chromosomal loss event.
-Initial studies have shown that aneuploidy can have a significant role in the deletion of the the tumor suppressing genes when compared to two consecutive point mutations\cite{nowak2002role,komarova2003mutation,michor2005can,komarova2008selective}.
-% TODO mention if studies are theoretical , experimental, or clinical
+Initial theoretical studies have shown that aneuploidy can have a significant role in the deletion of the the tumor suppressing genes when compared to two consecutive point mutations\cite{nowak2002role,komarova2003mutation,michor2005can,komarova2008selective}.
+% TODO mention if studies are theoretical , experimental, or clinical OK
 However, when taking into account that the appearance of aneuploidy requires a mutation to trigger CIN, the probability that CIN precedes tumor genesis is highly unlikely.
@@
 We assume that $\Delta_w<0$ due to drug therapy (or some other stress), and that $\Delta_m$ due to resistance conferred by the mutant.
-We analyze two cases: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, % REF: https://www.nature.com/articles/nrmicro.2016.34
-$0>\Delta_a>\Delta_w$.
+We analyze two cases: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, % REF: https://www.nature.com/articles/nrmicro.2016.34 OK
+$0>\Delta_a>\Delta_w$  \cite{brauner2016distinguishing}.
 
-% TODO explain briefly what evolutionary rescue means in this model.
-To analyze evolutionary rescue in this model, we use the framework of multitype branching processes. % REF Harris? https://doi.org/10.1016/j.jtbi.2021.110849
+% TODO explain briefly what evolutionary rescue means in this model. OK
+In our model rescues means that the aneuploid cells are able to either reverse ($\Delta a>0$ case) or delay ($0>\Delta_a>\Delta_w$ case) the extinction of the cancer cell population until mutant cells are generated and rescue the population.
+To analyze evolutionary rescue in this model, we use the framework of multitype branching processes \cite{rybnikov2021fitness,harris1963theory}.
+% REF Harris?
+https://doi.org/10.1016/j.jtbi.2021.110849 OK
 This allows us to find explicit expressions for the survival probability of the population--the probability that it does not become extinct due to stressful conditions, e.g., drug therapy.
@@
 This is a necessary simplification that is not biologically realistic, as cells usually compete for resources.
-A more biologically realistic model...
-% add the description of the more realistic model here
+A more biologically realistic model would include limited resources with competition between the lineages and spatial structure which plays an important role in the development of cancer \cite{martens2011spatial}.
+% add the description of the more realistic model here OK
 
@@
 % TODO add subsection on the Gillespie simulations 
+\subsection*{Gillespie simulations.} 
+Simulations were performed using a Gillespie algorithm\cite{gillespie1976general,gillespie1977exact} programmed in Python.
+The wildtype population consisting of $N$ cells is assumed to be declining exponentially at the beginning of the simulations.
+The algorithm we utilized proceeds at time $t$ as follows:
+\begin{enumerate}
+\item First we draw the time step $\Delta t$ from an exponential distribution whose rate parameter is equal to the sum of the rates of all possible events.
+\item Secondly we select a specific event $i$ according to its relative probability:
+\begin{equation*}
+p_i=\frac{\nu_i}{\sum \nu_j},
+\end{equation*}
+where $\nu_i$ is the rate of event $i$.
+\item Finally, we update the population size and all the event rates for time $t+\Delta t$.
+\end{enumerate}
+We repeat steps 1 to 3 until either the population of cancer cells dies out or the mutant population reaches a level at which there is a $99\%$ probability that 
+it will survive in perpetuity.
+\paragraph{Data availability.} All code used to generate the figures in this article is available online at: {\color{blue}https://github.com/yoavram-lab/EvolutionaryRescue}.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% TODO combine all diagram figures to a single figure with multiple panels. and make it fit in a single page.
+% TODO combine all diagram figures to a single figure with multiple panels. and make it fit in a single page. OK
+
+%%%%%%
+
 \begin{figure}
+\begin{subfigure}{\textwidth}
 \centering
-\begin{tikzpicture}[very thick,decoration={
-    markings,
-    mark=at position 0.5 with {\arrow{>}}}
-    ] 
-\edef\x{6.5} 
-\edef\y{-0.5} 
-\edef\r{1.25}
+\includegraphics[width=0.55\textwidth]{Figures/figureAneuploidy1.pdf}
+\caption{}
+\end{subfigure}
 
-\draw [postaction={decorate}, thick](-0.25,-2) -- (-0.25+5,0) node [midway, above] {$2u$};
-\draw [postaction={decorate}, thick](-0.25,-2) -- (-0.25+5,-4) node [midway, above] {$2p$};
-\draw [postaction={decorate}, thick](-0.25+5+\r,-0) -- (-0.25+10-\r,-0) node [midway, above] {$u$};
-\draw [postaction={decorate}, thick](-0.25+5+\r,-4) -- (-0.25+10-\r,-4) node [midway, above] {$u$};
-\draw [postaction={decorate}, thick](-0.25+5,-0) -- (-0.25+10,-4) node [midway, above] {$p$};
+\bigskip
 
-\filldraw[color=red!60, fill=red!5, very thick](-0.25,-2) circle (\r);
-\filldraw[color=red!60, fill=red!5, very thick](-0.25+5,0) circle (\r);
-\filldraw[color=red!60, fill=red!5, very thick](-0.25+5,-4) circle (\r);
-\filldraw[color=red!60, fill=red!5, very thick](-0.25+10,0) circle (\r);
-\filldraw[color=red!60, fill=red!5, very thick](-0.25+10,-4) circle (\r);
+\begin{subfigure}{\textwidth}
+\centering
+\includegraphics[width=0.55\textwidth]{Figures/figureAneuploidy2.pdf}
+\caption{}
+\end{subfigure}
 
-\draw [thick](0.05,-1.25) -- (0.05,-2.75);
-\draw [thick](-0.55,-1.25) -- (-0.55,-2.75);
+\bigskip
 
-\draw [thick](0.05+5,-1.25+2) -- (0.05+5,-2.75+2);
-\draw [very thick, blue](-0.55+5,-1.25+2) -- (-0.55+5,-2.75+2);
+\begin{subfigure}{\textwidth}
+\centering
+\includegraphics[width=0.55\textwidth]{Figures/figureAneuploidy3.pdf}
+\caption{}
+\label{figureAneuploidy}
+\end{subfigure}
 
-\draw [very thick, blue](0.05+10,-1.25+2) -- (0.05+10,-2.75+2);
-\draw [very thick, blue](-0.55+10,-1.25+2) -- (-0.55+10,-2.75+2);
-
-\draw [thick](-0.55+5,-1.25-2) -- (-0.55+5,-2.75-2);
-
-\draw [very thick, blue](-0.55+10,-1.25-2) -- (-0.55+10,-2.75-2);
-\end{tikzpicture}
-\caption{Diagram of tumorigenesis by which a cell loses both of its tumor suppressing cells.
+\caption{$\left(\text{a}\right)$ Diagram of tumorigenesis by which a cell loses both of its tumor suppressing cells.
 The first tumor suppressing gene can be eliminated by a mutation at rate $2u$ or through chromosomal instability with rate $2p$.
@@
 Grey vertical lines represent active TSGs while the blue vertical lines represent inactivated TSGs.
-}
-\label{figureCIN}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{figure}
-\centering
-\begin{tikzpicture}[very thick,decoration={
-    markings,
-    mark=at position 0.5 with {\arrow{>}}}
-    ] 
-\edef\x{6.5} 
-\edef\y{-0.5} 
-\edef\r{1.25}
-
-\draw [postaction={decorate}, thick](-0.25,-4) -- (-0.25+5,0) node [midway, above] {$u$};
-%\draw [postaction={decorate}, thick](-0.25,-2) -- (-0.25+5,-4) node [midway, above] {$2p$};
-\draw [postaction={decorate}, thick](-0.25+5,-0) -- (-0.25+10,-4) node [midway, above] {$v$};
-%\draw [postaction={decorate}, thick](-0.25+5+\r,-4) -- (-0.25+10-\r,-4) node [midway, above] {$u$};
-%\draw [postaction={decorate}, thick](-0.25+5,-0) -- (-0.25+10,-4) node [midway, above] {$p$};
-\draw [postaction={decorate}, thick]  (-0.25+10,-4) -- (-0.25+5,-8) node [midway, above] {$u$};
-\draw [postaction={decorate}, thick](-0.25,-4)  -- (-0.25+5,-8)  node [midway, above] {$v$};
-
-\filldraw[color=red!60, fill=red!5, very thick](-0.25,-4) circle (\r);
-\filldraw[color=red!60, fill=red!5, very thick](-0.25+5,0) circle (\r);
-%\filldraw[color=red!60, fill=red!5, very thick](-0.25+5,-4) circle (\r);
-\filldraw[color=red!60, fill=red!5, very thick](-0.25+10,-4) circle (\r);
-%\filldraw[color=red!60, fill=red!5, very thick](-0.25+10,-4) circle (\r);
-\filldraw[color=red!60, fill=red!5, very thick](-0.25+5,-8) circle (\r);
-
-\draw [thick](0.05,-1.25+2-4) -- (0.05,-2.75+2-4);
-\draw [thick](-0.55,-1.25+2-4) -- (-0.55,-2.75+2-4);
-
-\draw [thick](0.15+5,-1.25+2) -- (0.15+5,-2.75+2);
-\draw [thick](-0.65+5,-1.25+2) -- (-0.65+5,-2.75+2);
-\draw [thick](-0.25+5,-1.25+2) -- (-0.25+5,-2.75+2);
-
-\draw [thick](0.15+10,-1.25+2-4) -- (0.15+10,-2.75+2-4);
-\draw [thick](-0.25+10,-1.25+2-4) -- (-0.25+10,-2.75+2-4);
-\draw [color=blue,very thick](-0.7+10,-1.25+2-4) -- (-0.7+10,-2.75+2-4);
-
-\draw [thick](0.05+5,-1.25+2-8) -- (0.05+5,-2.75+2-8);
-\draw [color=blue,very thick](-0.55+5,-1.25+2-8) -- (-0.55+5,-2.75+2-8);
-%\draw [thick](-0.55+5,-1.25-2) -- (-0.55+5,-2.75-2);
-
-%\draw [thick](-0.55+10,-1.25-2) -- (-0.55+10,-2.75-2);
-\end{tikzpicture}
-\caption{Diagram of the evolutionary rescue process which involves aneuploidy.
+$\left(\text{b}\right)$ Diagram of the evolutionary rescue process which involves aneuploidy.
 The wildtype cell can either acquire trisomy at rate $u$ or acquire a fitness increasing mutation at rate $v$.
@@
 Finally, the aneuploidy cells can lose trisomy at rate $u$.
-}
-\label{figuretrisomy}
+$\left(\text{c}\right)$ Diagram of the evolutionary rescue model where the population of cancer cells is subdivided to wildtype euploid, wildtype aneuploid and mutant euploid cells that divide and die with rates $\lambda_w,  \lambda_a,  \lambda_m$ and $\mu_w, \mu_a, \mu_m$, respectively.
+Wildtype cells can acquire aneuploidy at rate $u$.
+Aneuploid and wildtype cells can mutate with rate $v$.}
 \end{figure}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}
-\centering
-\begin{tikzpicture}[very thick,decoration={
-    markings,
-    mark=at position 0.5 with {\arrow{>}}}, middlearrow/.style 2 args={
-        decoration={             
-            markings, 
-            mark=at position 0.5 with {\arrow[xshift=3.333pt]{triangle 45}, \node[#1] {#2};}
-        },
-        postaction={decorate}
-    }]
-\edef\x{6.5} 
-\edef\y{-0.5} 
-\edef\r{1.25}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\edef\R{1} 
-\edef\Y{0.25}
-
-\draw[black,thick,middlearrow={above}{$\lambda_w$},text=black] (-0.25,\Y) arc (270:-90:\R);
-\draw[black,thick,middlearrow={above}{$\lambda_a$},text=black] (-0.25+5,\Y) arc (270:-90:\R);
-\draw[black,thick,middlearrow={above}{$\lambda_m$},text=black] (-0.25+10,\Y) arc (270:-90:\R);
-
-\draw [postaction={decorate}, thick](-0.25,0) -- (-0.25+5,0) node [midway, above] {$u$};
-\draw [postaction={decorate}, thick](-0.25+5+\r,-0) -- (-0.25+10-\r,-0) node [midway, above] {$v$};
-
-\draw[thick, postaction={decorate}] (-0.25,0) arc (180:360:5) node [midway, above] {$v$};
-
-\filldraw[color=green!60, fill=green!5, very thick](-0.25,0) circle (\r);
-\filldraw[color=green!60, fill=green!5, very thick](-0.25+5,0) circle (\r);
-\filldraw[color=green!60, fill=green!5, very thick](-0.25+10,0) circle (\r);
-
-\draw [thick, postaction={decorate}](-0.25,-1.25) -- (-0.25,-2.75) node [midway, left] {$\mu_w$};
-
-\draw [thick, postaction={decorate}](-0.25+5,-1.25) -- (-0.25+5,-2.75) node [midway, right] {$\mu_a$};
-
-\draw [thick, postaction={decorate}](-0.25+10,-1.25) -- (-0.25+10,-2.75) node [midway, right] {$\mu_m$};
-
-\edef\x{0.25} 
-\edef\y{-0.25}
-
-\node[text width=2cm,align=justify] at (\y,\x)  
-{\begin{center}
-Wildtype  euploid
-\end{center}};
-
-\node[text width=2cm,align=justify] at (\y+5,\x)  
-{\begin{center}
-Wildtype  aneuploid
-\end{center}};
-
-\node[text width=2cm,align=justify] at (\y+10,\x)  
-{\begin{center}
-Mutant  euploid
-\end{center}};
-\end{tikzpicture}
-\caption{Diagram of the evolutionary rescue model where the population of cancer cells is subdivided to wildtype euploid, wildtype aneuploid and mutant euploid cells that divide and die with rates $\lambda_w,  \lambda_a,  \lambda_m$ and $\mu_w, \mu_a, \mu_m$, respectively.
-Wildtype cells can acquire aneuploidy at rate $u$.
-Aneuploid and wildtype cells can mutate with rate $v$.}
-\label{figureAneuploidy}
-\end{figure}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\newpage
+%\newpage
 \section*{Results}
@@
 We write \cref{survprobw} as
-\begin{equation}\label{survprobwapprox}
+\begin{equation}\label{survprobwapprox1}
 p_w=-\frac{1}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right).
@@
 So, given an initial population of $N$ wildtype cells, the probability that the population will survive is given by
-\begin{align}
+\begin{align}\label{aneuploidyresistentfirstapprox}
 p_{est}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right] .
@@
 % Todo I dont think this figure should show lambda_w>1, as in that case the wildtype is resistant and there is no rescue scenario.
-\Cref{SurvPlotNData} show $p_{est}$ as a function of $N$, including comparison of our approximation % ref to equation
+OK
+\Cref{SurvPlotNData} show $p_{est}$ as a function of $N$, including comparison of our approximation \eqref{aneuploidyresistentfirstapprox} % ref to equation OK
 and simulation results.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{figure}[!t]
+ \vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/P_est_growth.pdf}
+\caption{Plot of the probability of survival of a population as a function of the proliferation rate of the wildtype cells.
+The blue line represents the exact solution \eqref{survprobw} and the orange line line represents the approximation \eqref{survprobwapprox1}.
+Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $N=10^4, \lambda_a=1+10^{-2},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1.$}
+\label{SurvPlot}
+\end{figure}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 we write \eqref{survprobw} as
-\begin{equation}\label{survprobwapprox}
+\begin{equation}\label{survprobwapprox2}
 p_w=\frac{v\Delta_m}{\lambda_m\Delta_w}\left(\frac{u}{\Delta_a}-1\right).
@@
 which we plot in Figure \ref{DeleteriousPlot} where we compare with numerical simulations and the exact result \eqref{survprobw}.
-
+The transition between the regimes defined by \eqref{AneuploidyDeleteriousApprox} and \eqref{pestthirdcase} respectively occurs at:
 \begin{align}
-\Delta_a^*=2vp_m+v+2\sqrt{vp_m\left(vp_m+\mu_a+v\right)}
+\Delta_a^*=2vp_m+v+2\sqrt{vp_m\left(vp_m+\mu_a+v\right)}.
 \end{align}
@@
 %%%
-
+We wish to calculate the mean time for the appearence of the first mutant which rescues the cancer cell population. Let $\tau$ be the time the first mutant cell appears which rescues the population:
 \begin{align}
-\tau=\tau_1+\tau_2
+\tau=\tau_1+\tau_2,
 \end{align}
+where $\tau_1$ is the time at which the first aneuploid cell appears whose lineage will rescue the population and $\tau_2$ the time which this aneuploidy lineage drifts until it produces the first mutant which rescues the population.
 inhomogeneus Poisson process
 \begin{align}
-\tau_1\sim \text{Poisson}\left(\frac{1}{up_aW\left(t\right)}\right)
+\tau_1\sim \text{Poisson}\left(\frac{1}{up_aW\left(t\right)}\right),
 \end{align}
@@
 \end{align*}
-where we used
+where we used the fact that evolutionary rescue is impossible when the mutant population is destined to be zero:
 \begin{align*}
-P\left(\tau_1<t|M\left(t\rightarrow\infty\right)=0\right)=0
+P\left(\tau_1<t|M\left(t\rightarrow\infty\right)=0\right)=0.
 \end{align*}
+As a result, we obtain 
 \begin{align}
@@
  \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/SurvPlot.pdf}
-\caption{Plot of the probability of survival of a population as a function of the proliferation rate of the wildtype cells.
-The blue line represents the exact solution \eqref{survprobw} and the orange line line represents the approximation \eqref{survprobwapprox}.
-Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $N=75, \lambda_a=1+10^{-2},\lambda_m=1+10^{-3},\mu_w=1,\mu_a=1,\mu_m=1.$}
-\label{SurvPlot}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[!t]
- \vspace*{1\baselineskip}
 \includegraphics[width=1\textwidth]{Figures/HeatMap_p_est.pdf}
@@
 \caption{Plot of the probability of survival of a population as a function of the initial population size of wildtype cells.
-The blue line represents the exact solution \eqref{survprobw}, the orange line line represents the approximation \eqref{survprobwapprox}, the green line represents the first order correction \eqref{survprobwapproxcorrected} and the red dots represents stochastic simulations.
+The blue line represents the exact solution \eqref{survprobw}, the orange line line represents the approximation \eqref{aneuploidyresistentfirstapprox}, the green line represents the first order correction \eqref{survprobwapproxcorrected} and the red dots represents stochastic simulations.
 For the simulations we have chosen the following parameters: $\lambda_w=1-10^{-2}, \lambda_a=1+10^{-2},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1.$}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section*{Diffusion approximation}\label{AppendixDiffApprox}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \end{appendices}

### 2023-02-19 EvolutionaryRescue1.tex
finished editing

@@
 \paragraph{Aneuploidy in cancer.} Chromosomal instability (CIN) is the mitotic process in which cells suffer from chromosome mis-segregation that leads to aneuploidy, where cells are characterized by structural changes of the chromosomes and copy number alterations \cite{schukken2018cin}.
-Interestingly, aberrations in chromosome copy number have been shown to allow cancer cells to survive under stressful conditions such as drug therapy. % REF
+Interestingly, aberrations in chromosome copy number have been shown to allow cancer cells to survive under stressful conditions such as drug therapy.
 Indeed, cancer cells are often likely to be aneuploid, and aneuploidy is associated with poor patient outcomes \cite{ben2020context}.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Model}
+\section*{Methods}
+\subsection*{Evolutionary model}
 
-We follow the number of three types of cancer cells at time $t$: wildtype euploid, $w_t$; wildtype aneuploid, $a_t$; and mutant euploid, $m_t$. 
+We follow the number of cancer cells that belong to three different types at time $t$: wildtype euploid, $w_t$; wildtype aneuploid, $a_t$; and mutant euploid, $m_t$. 
 These cells grow and die with rates $\lambda_k$ and $\mu_k$ for $k=w, a, m$.
@@
   \begin{array}{@{}ll@{}}
-  \frac{\lambda_m-\mu_m}{\lambda_m},\quad\text{if }\lambda_m>\mu_m \\
-   0,\quad\text{else}.
+  \frac{\lambda_m-\mu_m}{\lambda_m},\quad \text{if }\lambda_m>\mu_m \\
+   0,\quad \text{else} .
   \end{array}\right.
@@
 % TODO add subsection on the Gillespie simulations 
-\subsection*{Gillespie simulations.} 
-Simulations were performed using a Gillespie algorithm\cite{gillespie1976general,gillespie1977exact} programmed in Python.
-The wildtype population consisting of $N$ cells is assumed to be declining exponentially at the beginning of the simulations.
-The algorithm we utilized proceeds at time $t$ as follows:
-\begin{enumerate}
-\item First we draw the time step $\Delta t$ from an exponential distribution whose rate parameter is equal to the sum of the rates of all possible events.
-\item Secondly we select a specific event $i$ according to its relative probability:
-\begin{equation*}
-p_i=\frac{\nu_i}{\sum \nu_j},
-\end{equation*}
-where $\nu_i$ is the rate of event $i$.
-\item Finally, we update the population size and all the event rates for time $t+\Delta t$.
-\end{enumerate}
-We repeat steps 1 to 3 until either the population of cancer cells dies out or the mutant population reaches a level at which there is a $99\%$ probability that 
-it will survive in perpetuity.
-\paragraph{Data availability.} All code used to generate the figures in this article is available online at: {\color{blue}https://github.com/yoavram-lab/EvolutionaryRescue}.
+\subsection*{Gillespie simulations} 
+Simulations were performed using a Gillespie algorithm \cite{gillespie1976general,gillespie1977exact} implemented in Python \cite{pyhon}.
+The simulation monitors the number of cells of each type: wildtype, aneuploid, and mutant. 
+The wildtype population initially consists of $N$ cells, whereas the other cell types are initially absent.
+
+At each iteration of the simulation loop we compute the rate of each even, $\nu_i$, for event $i$. % TODO explains what events can occur and how you compute their rates.
+We then draw the time until the next even, $\Delta t$, from an exponential distribution $\textit{Exp}(\sum_j \nu_j)$ whose rate parameter is the sum of the rates of all events.
+Then, we randomly determine which event occurred, where the probability for even $i$ is $p_i=\frac{\nu_i}{\sum \nu_j}$.
+Finally, we update the population size according to the event that occurred and update the time from $t$ to $t+\Delta t$.
+We repeat these iterations until either the population dies out (the number of cells of all types is zero) or the number of mutant cells is high enough so that it's extinction probability is $<1\%$.
+% TODO how do you determine the extinction prob?
+ref to equaiton
+
+% TODO do you use tau leap? if so, update the text accordingly.
+
+\paragraph{Data availability.} All code used to generate the figures in this article is available online at \href{https://github.com/yoavram-lab/EvolutionaryRescue}{https://github.com/yoavram-lab/EvolutionaryRescue}.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% TODO : should this move to methods? we only use this for simulations, right? should be in the Gillespie section?
 \subsection*{Logistic growth}
@@
 \subsection*{Standing genetic variation}
-Until now we have assumed that the initial population of cells consisted entirely of wildtype cells.
-We modify this assumtion such that the initial population includes a fraction $f$ of cells with aneuploidy.
-The probability of evolutionary rescue by the cells with aneuploidy from the initial population is:
+So far we have assumed that the initial population of cells consisted entirely of wildtype cells.
+We now modify this assumption so that the initial population includes a fraction $f$ of cells with aneuploidy.
+The probability of evolutionary rescue by cells with aneuploidy from the initial population is
 \begin{equation*}
-p_{old}=1-\left(1-p_a\right)^{fN}\approx 1-\e^{-fNp_a}.
+p_{old} = 1-\left(1-p_a\right)^{fN}\approx 1-\e^{-fNp_a}.
 \end{equation*}
-The total probability of evolutionary rescue is given by:
+The total probability of evolutionary rescue is given by
 \begin{align}\nonumber
-p_{total}&=p_{new}+\left(1-p_{new}\right)p_{old}\\
-&=1-\e^{-\left[\left(1-f\right)p_w+fp_a\right]N}.
+p_{total} 	&= p_{new}+\left(1-p_{new}\right)p_{old}\\
+			&= 1-\exp\left(-\left[\left(1-f\right)p_w + fp_a\right]N\right) .
 \end{align}
-The fraction of the cases in which the population is rescued by the standing genetic variation is given by:
-\begin{align*}
-F\left(f\right)=\frac{p_{old}}{p_{total}}=\frac{ 1-\e^{-fNp_a}}{1-\e^{-\left[\left(1-f\right)p_w+fp_a\right]N}}.
-\end{align*}
-We let $F=\frac{1}{2}$ and we obtain:
+
+The fraction of cases in which the population is rescued by the standing genetic variation is given by $F\left(f\right)=\frac{p_{old}}{p_{total}}$.
+%=\frac{ 1-\e^{-fNp_a}}{1-\e^{-\left[\left(1-f\right)p_w+fp_a\right]N}}.
+Setting $F=\frac{1}{2}$, we use the expansion $\e^x \approx 1+x$ to obtain
 \begin{equation}\label{halfeqstandvar}
@@
 \end{equation}
-wher we have used the expansion $\e^x\approx 1+x$.
-We plot $F$ and $f^*$ in Figure \ref{FractionPlot}.
+See \Cref{FractionPlot} for a demonstration of $F$ and $f*$.
 
@@
 %%%
-We wish to calculate the mean time for the appearence of the first mutant which rescues the cancer cell population. Let $\tau$ be the time the first mutant cell appears which rescues the population:
-\begin{align}
-\tau=\tau_1+\tau_2,
-\end{align}
-where $\tau_1$ is the time at which the first aneuploid cell appears whose lineage will rescue the population and $\tau_2$ the time which this aneuploidy lineage drifts until it produces the first mutant which rescues the population.
+We calculate the mean time for the appearance of the first mutant that rescues the cancer cell population.
+Let $\tau=\tau_1+\tau_2$ be the time the first mutant cell that appears and rescues the population, where $\tau_1$ is the time at which the first aneuploid cell appears whose lineage will rescue the population and $\tau_2$ the time for this aneuploidy lineage to produce the first mutant that rescues the population.
+% something is missing in the next line...?
+% the entire section is missing text...
 inhomogeneus Poisson process

### 2023-02-26 EvolutionaryRescue1.tex
revised methods section and fig 1

@@
 \author[1]{Remus Stana}
-\author[2]{Daniel B. Weissman}
+\author[2]{Uri Ben-David}
+\author[3]{Daniel B. Weissman}
 \author[1,*]{Yoav Ram}
-\affil[1]{School of Zoology, Tel Aviv University}
-\affil[2]{Department of Physics, Emory University}
+\affil[1]{School of Zoology, Faculty of Life Sciences, Tel Aviv University}
+\affil[2]{Department of Human Molecular Genetics and Biochemistry, Faculty of Medicine, Tel Aviv University}
+\affil[3]{Department of Physics, Emory University}
 \affil[*]{Corresponding author: yoav@yoavram.com}
@@
 We assume that $\Delta_w<0$ due to drug therapy (or some other stress), and that $\Delta_m$ due to resistance conferred by the mutant.
-We analyze two cases: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, % REF: https://www.nature.com/articles/nrmicro.2016.34 OK
-$0>\Delta_a>\Delta_w$  \cite{brauner2016distinguishing}.
+We analyze two cases: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_w$  \cite{brauner2016distinguishing}.
 
-% TODO explain briefly what evolutionary rescue means in this model. OK
-In our model rescues means that the aneuploid cells are able to either reverse ($\Delta a>0$ case) or delay ($0>\Delta_a>\Delta_w$ case) the extinction of the cancer cell population until mutant cells are generated and rescue the population.
-To analyze evolutionary rescue in this model, we use the framework of multitype branching processes \cite{rybnikov2021fitness,harris1963theory}.
-% REF Harris?
-https://doi.org/10.1016/j.jtbi.2021.110849 OK
+Here, \emph{evolutionary rescue} occurs when the aneuploid cells are able to either prevent (when $\Delta a>0$) or delay (when $0>\Delta_a>\Delta_w$) the extinction of the cancer-cell population before the adaptive mutant cells appear and fixate in the populaiton.
+To analyze evolutionary rescue in this model, we use the framework of \emph{multitype branching processes} \cite{rybnikov2021fitness,harris1963theory}. 
 This allows us to find explicit expressions for the survival probability of the population--the probability that it does not become extinct due to stressful conditions, e.g., drug therapy.
 Note that branching processes imply that lineages produced by cells from the initial population grow and die independently of each other.
-This is a necessary simplification that is not biologically realistic, as cells usually compete for resources.
-A more biologically realistic model would include limited resources with competition between the lineages and spatial structure which plays an important role in the development of cancer \cite{martens2011spatial}.
-% add the description of the more realistic model here OK
+This is a necessary simplification, but it is unrealistic, as cells usually compete for resources.
+A more realistic model may include competition on limited resources between lineages, as well as spatial structure, which may play an important role in the development of cancer \cite{martens2011spatial}.
 
-The survival probability of a population consisting initially of one wildtype cell satisfies the quadratic equation:
+The survival probability of a population consisting initially of one wildtype cell satisfies the quadratic equation,
 \begin{equation} \label{quadraticeqev1}
@@
 
-If the population originally consists of a single aneuploidy cell, then the probability that the population will survive is given by the quadratic equation
+If the population originally consists of a single aneuploid cell, then the probability that the population will survive is given by the quadratic equation
 \begin{align}
@@
    0,\quad \text{else} .
-  \end{array}\right.
+  \end{array}\right .
 \end{equation}
@@
 % TODO add subsection on the Gillespie simulations 
-\subsection*{Gillespie simulations} 
-Simulations were performed using a Gillespie algorithm \cite{gillespie1976general,gillespie1977exact} implemented in Python \cite{pyhon}.
+\subsection*{Evolutionary simulation} 
+Simulations are performed using a \emph{Gillespie algorithm} \cite{gillespie1976general,gillespie1977exact} implemented in Python \cite{python}.
 The simulation monitors the number of cells of each type: wildtype, aneuploid, and mutant. 
@@
 
-At each iteration of the simulation loop we compute the rate of each even, $\nu_i$, for event $i$. % TODO explains what events can occur and how you compute their rates.
-We then draw the time until the next even, $\Delta t$, from an exponential distribution $\textit{Exp}(\sum_j \nu_j)$ whose rate parameter is the sum of the rates of all events.
-Then, we randomly determine which event occurred, where the probability for even $i$ is $p_i=\frac{\nu_i}{\sum \nu_j}$.
-Finally, we update the population size according to the event that occurred and update the time from $t$ to $t+\Delta t$.
-We repeat these iterations until either the population dies out (the number of cells of all types is zero) or the number of mutant cells is high enough so that it's extinction probability is $<1\%$.
+At each iteration of the simulation loop we compute the rate $\nu_j$ of each event $j$. % TODO explains what events can occur and how you compute their rates, refer to illustration figure
+We then draw the time until the next event, $\Delta t$, from an exponential distribution whose rate parameter is the sum of the rates of all events, such that $\Delta t \sim \textit{Exp}(\sum_j \nu_j)$.
+Then, we randomly determine which event occurred, where the probability for event $j$ is $p_j=\nu_j/\sum_i \nu_i$.
+Finally, we update the number of cells of each type according to the event that occurred and update the time from $t$ to $t+\Delta t$.
+% TODO more details - you need to specify the events and how they change the number of cells.
+We repeat these iterations until either the population becomes extinct (the number of cells of all types is zero) or the number of mutant cells is high enough so that its extinction probability is $<1\%$.
 % TODO how do you determine the extinction prob?
-ref to equaiton
+ref to equation
 
@@
 
-\paragraph{Data availability.} All code used to generate the figures in this article is available online at \href{https://github.com/yoavram-lab/EvolutionaryRescue}{https://github.com/yoavram-lab/EvolutionaryRescue}.
+\paragraph{Data availability.} All source code is available online at \href{https://github.com/yoavram-lab/EvolutionaryRescue}{https://github.com/yoavram-lab/EvolutionaryRescue}.
 
@@
 \begin{figure}
-\begin{subfigure}{\textwidth}
-\centering
-\includegraphics[width=0.55\textwidth]{Figures/figureAneuploidy1.pdf}
-\caption{}
-\end{subfigure}
+%\begin{subfigure}{\textwidth}
+%\centering
+%\includegraphics[width=0.55\textwidth]{Figures/figureAneuploidy1.pdf}
+%\caption{}
+%\end{subfigure}
+%
+%\bigskip
+%
+%\begin{subfigure}{\textwidth}
+%\centering
+%\includegraphics[width=0.55\textwidth]{Figures/figureAneuploidy2.pdf}
+%\caption{}
+%\end{subfigure}
+%
+%\bigskip
 
-\bigskip
-
-\begin{subfigure}{\textwidth}
-\centering
-\includegraphics[width=0.55\textwidth]{Figures/figureAneuploidy2.pdf}
-\caption{}
-\end{subfigure}
-
-\bigskip
-
-\begin{subfigure}{\textwidth}
+%\begin{subfigure}{\textwidth}
 \centering
 \includegraphics[width=0.55\textwidth]{Figures/figureAneuploidy3.pdf}
-\caption{}
+%\caption{}
 \label{figureAneuploidy}
-\end{subfigure}
+%\end{subfigure}
 
-\caption{$\left(\text{a}\right)$ Diagram of tumorigenesis by which a cell loses both of its tumor suppressing cells.
+\caption{\textbf{Model illustration.}
+% TODO why do we need (a) and (b)? does it relate to our model?
+%\textbf{(a)} Diagram of tumorigenesis by which a cell loses both of its tumor suppressing cells.
 The first tumor suppressing gene can be eliminated by a mutation at rate $2u$ or through chromosomal instability with rate $2p$.
@@
 Grey vertical lines represent active TSGs while the blue vertical lines represent inactivated TSGs.
-$\left(\text{b}\right)$ Diagram of the evolutionary rescue process which involves aneuploidy.
-The wildtype cell can either acquire trisomy at rate $u$ or acquire a fitness increasing mutation at rate $v$.
-Aneuploidy cells can  acquire a fitness increasing mutation at rate $v$.
-Finally, the aneuploidy cells can lose trisomy at rate $u$.
-$\left(\text{c}\right)$ Diagram of the evolutionary rescue model where the population of cancer cells is subdivided to wildtype euploid, wildtype aneuploid and mutant euploid cells that divide and die with rates $\lambda_w,  \lambda_a,  \lambda_m$ and $\mu_w, \mu_a, \mu_m$, respectively.
-Wildtype cells can acquire aneuploidy at rate $u$.
-Aneuploid and wildtype cells can mutate with rate $v$.}
+%\textbf{(b)} Evolutionary model with aneuploidy and mutation. Wildtype and aneuploid cells acquire a beneficial mutation at rate $v$. Diploid cells can become trisomic at rate $u$, and vice versa.
+%\textbf{(c)} 
+%Diagram of the evolutionary model. 
+A population of cancer cells is subdivided to wildtype euploid, wildtype aneuploid and mutant euploid cells, which divide with rates $\lambda_w$, $\lambda_a$, and $\lambda_m$, respectively, and die at rates $\mu_w$, $\mu_a$, and $\mu_m$, respectively.
+Wildtype cells can become aneuploid at rate $u$. Both aneuploid and wildtype cells can acquire a beneficial mutation with rate $v$.}
 \end{figure}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Survival probability of a mutant lineage}\label{AppendixSurvLin}
+\section{Survival probability of a mutant lineage}\label{AppendixSurvLin}
 The infinitesimal transition probabilities for the simple birth and death process:

### 2023-03-12 EvolutionaryRescue1.tex
revised

@@
 \usepackage{subcaption}
+\usepackage{xurl}
 
@@
 The deletion of tumor suppression genes can happen in two ways: two point mutations deleting both alleles of the TSG (assuming a diploid genotype), or one point mutation and one chromosomal loss event.
-Initial theoretical studies have shown that aneuploidy can have a significant role in the deletion of the the tumor suppressing genes when compared to two consecutive point mutations\cite{nowak2002role,komarova2003mutation,michor2005can,komarova2008selective}.
+Initial theoretical studies have shown that aneuploidy can have a significant role in the deletion of the the tumor suppressing genes when compared to two consecutive point mutations \cite{nowak2002role,komarova2003mutation,michor2005can,komarova2008selective}.
 % TODO mention if studies are theoretical , experimental, or clinical OK
@@
 
-Here, \emph{evolutionary rescue} occurs when the aneuploid cells are able to either prevent (when $\Delta a>0$) or delay (when $0>\Delta_a>\Delta_w$) the extinction of the cancer-cell population before the adaptive mutant cells appear and fixate in the populaiton.
+Here, \emph{evolutionary rescue} occurs when the aneuploid cells are able to either prevent (when $\Delta_a>0$) or delay (when $0>\Delta_a>\Delta_w$) the extinction of the cancer-cell population before the adaptive mutant cells appear and fixate in the populaiton.
 To analyze evolutionary rescue in this model, we use the framework of \emph{multitype branching processes} \cite{rybnikov2021fitness,harris1963theory}. 
@@
 
-% TODO add subsection on the Gillespie simulations 
+% TODO add subsection on the Gillespie simulations OK
 \subsection*{Evolutionary simulation} 
@@
 The simulation monitors the number of cells of each type: wildtype, aneuploid, and mutant. 
-The wildtype population initially consists of $N$ cells, whereas the other cell types are initially absent.
+The wildtype population initially consists of $w_0$ cells, whereas the other cell types are initially absent.
 
-At each iteration of the simulation loop we compute the rate $\nu_j$ of each event $j$. % TODO explains what events can occur and how you compute their rates, refer to illustration figure
+At each iteration of the simulation loop we compute the rate $\nu_j$ of each event $j$. % TODO explains what events can occur and how you compute their rates, refer to illustration figure  OK
+The sate of the stochastic system at time $t$ is represented by the triplet $\left(w_t,a_t,m_t\right)$ which can change in the time interval $\Delta_t$ according to the following events with appropriate rates (see Figure \ref{figureAneuploidy}):
+\begin{subequations}
+\begin{flalign*}
+(+1,0,0)&:\quad \lambda_ww_t\quad\left(\text{birth euploid cell}\right),\\
+(-1,0,0)&:\quad \mu_ww_t\quad\left(\text{death euploid cell}\right),\\
+(-1,+1,0)&:\quad uw_t\quad\left(\text{euploid cell aquires aneuploidy}\right),\\
+(-1,0,+1)&:\quad vw_t\quad\left(\text{euploid cell aquires mutation}\right),\\
+(0,+1,0)&:\quad \lambda_aa_t\quad\left(\text{birth aneuploid cell}\right),\\
+(0,-1,0)&:\quad \mu_aa_t\quad\left(\text{death aneuploid cell}\right),\\
+(0,-1,+1)&:\quad va_t\quad\left(\text{aneuploid cell aquires mutation}\right),\\
+(0,0,+1)&:\quad \lambda_am_t\quad\left(\text{birth mutant cell}\right),\\
+(0,0,-1)&:\quad \mu_am_t\quad\left(\text{death mutant cell}\right).
+\end{flalign*}
+\end{subequations}
 We then draw the time until the next event, $\Delta t$, from an exponential distribution whose rate parameter is the sum of the rates of all events, such that $\Delta t \sim \textit{Exp}(\sum_j \nu_j)$.
@@
 % TODO more details - you need to specify the events and how they change the number of cells.
+OK
 We repeat these iterations until either the population becomes extinct (the number of cells of all types is zero) or the number of mutant cells is high enough so that its extinction probability is $<1\%$.
 % TODO how do you determine the extinction prob?
-ref to equation
-
-% TODO do you use tau leap? if so, update the text accordingly.
-
-\paragraph{Data availability.} All source code is available online at \href{https://github.com/yoavram-lab/EvolutionaryRescue}{https://github.com/yoavram-lab/EvolutionaryRescue}.
+ref to equation OK
+We continue running the simulation until:
+\begin{equation*}
+m_t>\left\lfloor-\frac{3\log10}{\log\left(\frac{\mu_m}{\lambda_m}\right)}\right\rfloor+1,
+\end{equation*}
+which is the population size at which the extinction probability of the population of mutant cell is less then $1\%$. 
+% TODO do you use tau leap? if so, update the text accordingly. OK
+When the size of the initial population is large we utilize $\tau$-leaping where the increment of population change in a fixed time interval $\Delta t$ is Poisson distributed with mean $\nu_i\Delta t$ for each population $i$ \cite{gillespie2001approximate}.
+If the increment is negative and larger then the subpopulation size then updated
+\paragraph{Data availability.} All source code is available online at \url{https://github.com/yoavram-lab/EvolutionaryRescue}.
 
@@
 \begin{figure}
-%\begin{subfigure}{\textwidth}
-%\centering
-%\includegraphics[width=0.55\textwidth]{Figures/figureAneuploidy1.pdf}
-%\caption{}
-%\end{subfigure}
-%
-%\bigskip
-%
-%\begin{subfigure}{\textwidth}
-%\centering
-%\includegraphics[width=0.55\textwidth]{Figures/figureAneuploidy2.pdf}
-%\caption{}
-%\end{subfigure}
-%
-%\bigskip
-
-%\begin{subfigure}{\textwidth}
 \centering
@@
 %\caption{}
-\label{figureAneuploidy}
+
 %\end{subfigure}
@@
 \caption{\textbf{Model illustration.}
-% TODO why do we need (a) and (b)? does it relate to our model?
-%\textbf{(a)} Diagram of tumorigenesis by which a cell loses both of its tumor suppressing cells.
-The first tumor suppressing gene can be eliminated by a mutation at rate $2u$ or through chromosomal instability with rate $2p$.
-The last TSG can either be eliminated through mutation at rate $u$ or through chromosomal instability with rate $p$.
-Grey vertical lines represent active TSGs while the blue vertical lines represent inactivated TSGs.
-%\textbf{(b)} Evolutionary model with aneuploidy and mutation. Wildtype and aneuploid cells acquire a beneficial mutation at rate $v$. Diploid cells can become trisomic at rate $u$, and vice versa.
-%\textbf{(c)} 
-%Diagram of the evolutionary model. 
 A population of cancer cells is subdivided to wildtype euploid, wildtype aneuploid and mutant euploid cells, which divide with rates $\lambda_w$, $\lambda_a$, and $\lambda_m$, respectively, and die at rates $\mu_w$, $\mu_a$, and $\mu_m$, respectively.
 Wildtype cells can become aneuploid at rate $u$. Both aneuploid and wildtype cells can acquire a beneficial mutation with rate $v$.}
+\label{figureAneuploidy}
 \end{figure}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% TODO give a descriptive name instead of a mathematical formula
-\subsection*{First case: $4\lambda_avp_m<\left(\lambda_a-\mu_a-v\right)^2$}
+% TODO give a descriptive name instead of a mathematical formula OK
+%\subsection*{First case: $4\lambda_avp_m<\left(\lambda_a-\mu_a-v\right)^2$}
+\subsection*{First case: Highly deleterious or highly beneficial aneuploidy}
 
@@
 Using a quadratic Taylor expansion, $\sqrt{1+x}=1+x/2+O(x^2)$, 
-%\begin{align*}
-%\left(1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}\right)^{\frac{1}{2}}&=1+\frac{2\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}+\cdots\\
-%\left(1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}\right)^{\frac{1}{2}}&=1+\frac{2\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}+\cdots
-%\end{align*}
 we obtain the following approximation for the survival probability of a population consisting of a single individual wildtype cell (assuming $u,v \ll 1$),
@@
 \caption{Plot of the probability of survival of a population as a function of the proliferation rate of the wildtype cells.
-The blue line represents the exact solution \eqref{survprobw} and the orange line line represents the approximation \eqref{survprobwapprox1}.
-Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $N=10^4, \lambda_a=1+10^{-2},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1.$}
+The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximation \eqref{survprobwapprox1}.
+Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $N=10^4, \lambda_a=1+10^{-2},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
+The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
+The error bars are present but are not visible given the fact that we have used $10^5$ simulations for each combination of parameters.}
 \label{SurvPlot}
@@
 \caption{Plot of the survival probability of an initial population consisting of $10^{4}$ wildtype cells as a function of $\Delta_a=\lambda_a-\mu_a$ for various values of $\Delta_w=\lambda_a-\mu_a$.
-The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximation \eqref{AneuploidyDeleteriousApprox}.}
+The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximation \eqref{AneuploidyDeleteriousApprox}.
+The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.}
 \label{P_est}
@@
 \caption{Plot of the survival probability of an initial population consisting of $10^{4}$ wildtype cells as a function of $\Delta_w=\lambda_w-\mu_w$ for various values of $\Delta_a=\lambda_a-\mu_a$.
-The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximations \eqref{AneuploidyDeleteriousApprox}.}
+The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximations \eqref{AneuploidyDeleteriousApprox}.
+The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
+The value highlighted in grey is the threshold $\Delta_a^*$ from \eqref{thresholdvalueaneuploid} which marks the transition between the regime dictated by \eqref{AneuploidyDeleteriousApprox} to the one dictated by \eqref{pestthirdcase}.
+}
 \label{P_est}
@@
 \caption{Plot of the probability of survival of a population as a function of the initial population size of wildtype cells.
-The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximation \eqref{pestthirdcase}.}
+The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximation \eqref{pestthirdcase}.
+The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
+The error bars are present but are not visible given the fact that we have used $n=10^5$ simulations for each combination of parameters.}
 \label{DeleteriousPlot}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Second case: $4\lambda_avp_m>\left(\lambda_a-\mu_a-v\right)^2$}
+%\subsection*{Second case: $4\lambda_avp_m>\left(\lambda_a-\mu_a-v\right)^2$}
+\subsection*{Second case: Neutral aneuploidy}
 If we assume that  $4\lambda_avp_m>\left(\lambda_a-\mu_a-v\right)^2$ then we write:
@@
 \end{align*}
-%In the last line of the above we have used the following inequalities:
-%\begin{align*}
-%\lambda_a-\mu_a,\,v,\,\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\sqrt{\lambda_avp_m}}\ll1.
-%\end{align*}
 As a result, we have from \eqref{survprobwinitial} the probability of rescue of a population starting from one wildtype individual:
@@
 The transition between the regimes defined by \eqref{AneuploidyDeleteriousApprox} and \eqref{pestthirdcase} respectively occurs at:
-\begin{align}
+\begin{align}\label{thresholdvalueaneuploid}
 \Delta_a^*=2vp_m+v+2\sqrt{vp_m\left(vp_m+\mu_a+v\right)}.
@@
 &\lambda_w'=\lambda_w,\quad\mu_w'=\mu_w\\
-&\lambda_a'=C_1+\lambda_a\left(1-\frac{w+a+m}{K}\right),\quad \mu_a'=C_1\\
-&\lambda_m'=C_2+\lambda_m\left(1-\frac{w+a+m}{K}\right),\quad \mu_m'=C_2
+&\lambda_a'=C_1+\left(\lambda_a-\mu_a\right)\left(1-\frac{w+a+m}{K}\right),\quad \mu_a'=C_1\\
+&\lambda_m'=C_2+\left(\lambda_m-\mu_m\right)\left(1-\frac{w+a+m}{K}\right),\quad \mu_m'=C_2
 \end{align*}
 where $C_1, C_2>0$ are constants.
+We perform stochastic simulations for different values of the carrying capacity $K$ and we plot the results in Figure \ref{SurvPlotNDataLogisticKComplete}.
+We observe that as $K$ increases the simulations converge to the analytic result which is because the carrying capacity is much larger then the population size of aneuploid cells for which the probability that the population is rescued is certain.
+
+\begin{figure}[!t]
+ \vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/SurvPlotNDataLogisticKComplete.pdf}
+\caption{Plot of the survival probability of an initial population consisting of wildtype cells as a function of the initial population size.
+Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_w=1-10^{-1},\lambda_w=1+10^{-4},\lambda_m=1+10^{-1},\mu_w=1,\mu_m=1,u=10^{-2},v=10^{-7}, C_1=C_2=1$.}
+\label{SurvPlotNDataLogisticKComplete}
+\end{figure}
 
@@
 \includegraphics[width=1\textwidth]{Figures/MeanTimeGrowthAneuploidyPlot.pdf}
-\caption{}
+\caption{Plot of the mean time until the appearence of a resistence mutation which rescues the population in the case when evolutionary rescue is possible only through mutation but not aneuploidy and mutation.
+Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_w=1-10^{-1},\lambda_m=1+10^{-1},\mu_w=1,\mu_m=1,u=10^{-2},v=10^{-7}$.}
 \label{MeanTimeGrowthAneuploidyPlot}
 \end{figure}
-%%%
 We calculate the mean time for the appearance of the first mutant that rescues the cancer cell population.
@@
 % the entire section is missing text...
-inhomogeneus Poisson process
+
+The time $\tau_1$ can be aproximated by a inhomogeneus Poisson process:
 \begin{align}
-\tau_1\sim \text{Poisson}\left(\frac{1}{up_aW\left(t\right)}\right),
+\tau_1\sim \text{Poisson}\left(\frac{1}{up_aw_t}\right),
 \end{align}
-where 
+where $w_t$ is the size of the wildtype population at time $t$: 
 \begin{equation}
-W\left(t\right)=W_0\e^{\Delta_wt}
+w_t=W_0\e^{\Delta_wt}.
 \end{equation}
+As a result, the probability density function of $\tau_1$ is given by:
 \begin{align}
@@
 \end{align}
+We are interested in the time to appearence of the first succesful aneuploid cell conditional on population surving we have the following:
 \begin{align*}
@@
 \end{equation}
+The time $\tau_1$ conditional on evolutionary rescue is given by:
 \begin{align}
@@
 \end{align}
+The expectation of $\tau_1$ is:
+\begin{align}\nonumber
+\mathbb{E}\left[\tau_1\right]&=\frac{\int_0^\infty\e^{-\int_0^\tau R(t)\d t}\d\tau}{1-\left(1-p_w\right)^N}\\
+&=\frac{\int_0^\infty\e^{-uNp_a\frac{\e^{\Delta_w\tau}-1}{\Delta_w}}\,\d\tau}{1-\left(1-p_w\right)^N}.
+\end{align}
+For an aneuploid cell, let $a_\tau$ be the number of its aneuploid descendents at time $\tau$.
+Each of these descentends have a probability $v\d \tau$ of producing a mutant which has a probability $p_m$ to survive.
+Consequently, the probability that the lineage descendent from a single aneuploid cell produces succesful mutant in the time interval $\left[\tau,\tau+\d \tau\right]$ is $vp_ma_t\d \tau$.
+Since the number of succesful mutants is a Poisson process, the probability of evolutionary rescue, given $a_\tau$, is given by:
 \begin{align}
-\mathbb{E}\left[\tau_1\right]&=\frac{\int_0^\infty\e^{-\int_0^\tau R(t)\d t}\d\tau}{1-\left(1-p_w\right)^N}\\
-&=\frac{\int_0^\infty\e^{-uNp_a\frac{\e^{\Delta_w\tau}-1}{\Delta_w}}\,\d\tau}{1-\left(1-p_w\right)^N}
+1-\exp\left(-\int_0^t vp_ma_\tau\d \tau\right)=1-\exp\left(-vp_mA\left(t\right)\right),
 \end{align}
+where $A\left(t\right)$ is the weight of the lineage, defined as:
+\begin{equation}
+A\left(t\right)=\int_0^t a_\tau\d \tau.
+\end{equation}
+As a result, to obtain $p_2$ we average over all possible values of $A\left(t\right)$:
 \begin{align}
-P_2\left(\tau_2\right)&=\int_0^\infty \d w P\left(W_l=w\right)\left[1-\e^{-vp_mW_l\left(\tau_2\right)}\right]\\
-&=1-\mathbb{E}\left[\e^{-vp_mW_l\left(\tau_2\right)}\right]
+p_2\left(t\right)&=\int_0^\infty \d w P\left(W_l=w\right)\left[1-\e^{-vp_mA\left(t\right)}\right]\\
+&=1-\mathbb{E}\left[\e^{-vp_mA\left(t\right)}\right]
 \end{align}
+Following the steps outlines in \cite{weissman2009rate} we obtain:
 \begin{align}
-a_\pm=\frac{2-\delta-y\pm\sqrt{\left(\delta+y\right)^2-4y}}{2\left(1-\delta\right)}
+p_2\left(t\right)=\frac{\left(a_+-1\right)\left(1-a_-\right)\left(1-\e^{-\left(1-\delta\right)\left(a_+-a_-\right)t}\right)}{a_+-1+\left(1-a_-\right)\e^{-\left(1-\delta\right)\left(a_+-a_-\right)t}},
 \end{align}
+where
 \begin{align}
-P_2\left(\tau_2\right)=\frac{\left(a_+-1\right)\left(1-a_-\right)\left(1-\e^{\left[-\left(1-\delta\right)\left(a_+-a_-\right)t\right]}\right)}{a_+-1+\left(1-a_-\right)\e^{\left[-\left(1-\delta\right)\left(a_+-a_-\right)t\right]}}
+a_\pm=\frac{2-\delta-y\pm\sqrt{\left(\delta+y\right)^2-4y}}{2\left(1-\delta\right)},
 \end{align}
 with $y=vp_m$ in $a_\pm$.
+
+The cumulative distribution of $T_2$ is given by $\frac{p_2\left(t\right)}{p_m}$ from which we obtain the mean time:
 \begin{align}
-\mathbb{E}\left[\tau_2\right]=\int_0^\infty\left(1-\frac{P_2\left(t\right)}{p_m}\right)\d t
+\mathbb{E}\left[\tau_2\right]=\int_0^\infty\left(1-\frac{P_2\left(t\right)}{p_m}\right)\d t.
 \end{align}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 The blue line represents the exact solution \eqref{survprobw}, the orange line line represents the approximation \eqref{aneuploidyresistentfirstapprox}, the green line represents the first order correction \eqref{survprobwapproxcorrected} and the red dots represents stochastic simulations.
-For the simulations we have chosen the following parameters: $\lambda_w=1-10^{-2}, \lambda_a=1+10^{-2},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1.$}
+For the simulations we have chosen the following parameters: $\lambda_w=1-10^{-2}, \lambda_a=1+10^{-2},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
+The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.}
 \label{SurvPlotNData}
@@
 The red vertical line highlights the the value of $f$ for which half the times the population is rescues by aneuploid cells while the pink line is our approximation \eqref{halfeqstandvar}.
-For this plot we have chosen the following parameters: $N=10^3, \lambda_w=1-10^{-2}, \lambda_a=1-10^{-4},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1.$}
+For this plot we have chosen the following parameters: $N=10^3, \lambda_w=1-10^{-2}, \lambda_a=1-10^{-4},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
+The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
+The value of $0.332$ highlighted in red is the value of the initial fraction of the population which in aneuploid for which half of the cases of evolutionary rescue is due to initial aneuploid cell population.}
 \label{FractionPlot}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Gillespie algorithm}
-\begin{subequations}
-\begin{flalign}
-(+1,0,0)&:\quad \lambda_ww_t,\\
-(-1,0,0)&:\quad \mu_ww_t,\\
-(-1,+1,0)&:\quad uw_t,\\
-(-1,0,+1)&:\quad vw_t,\\
-(0,+1,0)&:\quad \lambda_aa_t,\\
-(0,-1,0)&:\quad \mu_aa_t,\\
-(0,-1,+1)&:\quad va_t,\\
-(0,0,+1)&:\quad \lambda_am_t,\\
-(0,0,-1)&:\quad \mu_am_t.
-\end{flalign}
-\end{subequations}
+%\section*{Gillespie algorithm}
+%\begin{subequations}
+%\begin{flalign}
+%(+1,0,0)&:\quad \lambda_ww_t,\\
+%(-1,0,0)&:\quad \mu_ww_t,\\
+%(-1,+1,0)&:\quad uw_t,\\
+%(-1,0,+1)&:\quad vw_t,\\
+%(0,+1,0)&:\quad \lambda_aa_t,\\
+%(0,-1,0)&:\quad \mu_aa_t,\\
+%(0,-1,+1)&:\quad va_t,\\
+%(0,0,+1)&:\quad \lambda_am_t,\\
+%(0,0,-1)&:\quad \mu_am_t.
+%\end{flalign}
+%\end{subequations}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section*{Mean time}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+We define the following events
+\begin{align*}
+\text{EvR}&=\text{``evolutionary rescue occurs''},\\
+\text{EvRA}&=\text{``evolutionary rescue occurs through aneuploidy''},\\
+\text{EvRM}&=\text{``evolutionary rescue occurs through mutation''}.
+\end{align*}
+As a result, the mean time to appearence of the first mutant cell which rescues the population given evolutionary rescue can be expanded as:
+\begin{align*}
+\mathbb{E}\left[\tau|\text{EvR}\right]&=\mathbb{E}\left[\tau|\text{EvR},\text{EvRA}\cap\text{EvRM}^{C}\right]p\left(\text{EvRA}\cap\text{EvRM}^{C}|\text{EvR}\right)\\
+&+\mathbb{E}\left[\tau|\text{EvR},\text{EvRM}\cap\text{EvRA}^{C}\right]p\left(\text{EvRM}\cap\text{EvRA}^{C}|\text{EvR}\right)\\
+&+\mathbb{E}\left[\tau|\text{EvR},\text{EvRM}\cap\text{EvRA}\right]p\left(\text{EvRM}\cap\text{EvRA}|\text{EvR}\right).
+\end{align*}
+The joint probabilities in the previous lines can be written as:
+\begin{align*}
+p\left(\text{EvRA}\cap\text{EvRM}^{C}|\text{EvR}\right)&=\frac{p\left(\text{EvRA}\cap\text{EvRM}^{C}\right)}{p\left(\text{EvR}\right)}
+\approx\frac{p\left(\text{EvRA}\right)\left[1-p\left(\text{EvRM}\right)\right]}{p\left(\text{EvR}\right)},\\
+p\left(\text{EvRM}\cap\text{EvRA}^{C}|\text{EvR}\right)&=\frac{p\left(\text{EvRM}\cap\text{EvRA}^{C}\right)}{p\left(\text{EvR}\right)}
+\approx\frac{p\left(\text{EvRM}\right)\left[1-p\left(\text{EvRA}\right)\right]}{p\left(\text{EvR}\right)},\\
+p\left(\text{EvRM}\cap\text{EvRA}|\text{EvR}\right)&=\frac{p\left(\text{EvRM}\cap\text{EvRA}\right)}{p\left(\text{EvR}\right)}\approx\frac{p\left(\text{EvRM}\right)p\left(\text{EvRA}\right)}{p\left(\text{EvR}\right)}\ll1.
+\end{align*}
+As a result, we have:
+\begin{align*}
+\mathbb{E}\left[\tau|\text{EvR}\right]&\approx\mathbb{E}\left[\tau|\text{EvR},\text{EvRA}\cap\text{EvRM}^{C}\right]p\left(\text{EvRA}\cap\text{EvRM}^{C}|\text{EvR}\right)\\
+&+\mathbb{E}\left[\tau|\text{EvR},\text{EvRM}\cap\text{EvRA}^{C}\right]p\left(\text{EvRM}\cap\text{EvRA}^{C}|\text{EvR}\right).
+\end{align*}
+%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section*{Diffusion approximation}\label{AppendixDiffApprox}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{equation}
+r=\lim_{\Delta t\rightarrow0}\frac{\mathbb{E}\left(\Delta n_t|n_t\right)}{\Delta_t n_t}
+\end{equation}
+\begin{equation}
+\sigma=\lim_{\Delta t\rightarrow0}\frac{\mathbb{V}ar\left(\Delta n_t|n_t\right)}{\Delta_t n_t}
+\end{equation}
+\begin{align}
+\lambda_1=v\pi_f\frac{w_0}{|r_w|}
+\end{align}
+where
+\begin{align}
+\bar{\pi}_f=\int_{0}^\infty\int_{0}^\infty \left(1-\e^{\frac{-2r}{\sigma}}\right) f_r\left(r,\sigma\right)\d r\d\sigma
+\end{align}
+Let $f\left(r,\sigma\right)=\delta\left(r-r_m\right)\delta\left(\sigma-\sigma_m\right)$ then:
+\begin{align}
+\bar{\pi}_f=1-\e^{\frac{-2r_m}{\sigma_m}}
+\end{align}
+\begin{align}
+\lambda_1=v\left(1-\e^{\frac{-2r_m}{\sigma_m}}\right)\frac{N_0}{|\Delta_w|}
+\end{align}
+\begin{align}
+\lambda_2&=\frac{uw_0}{|r_w|}\int_{0}^\infty\int_{0}^\infty \left(1-\pi_f\left(r,q\right)\right) f_r\left(r,\sigma\right) p_1^*\left(r,\sigma\right)\d r\d\sigma\\
+&=\frac{uw_0}{|\Delta_w|}p_1^*\left(r_a,\sigma_a\right)\\
+&=\frac{uw_0}{|\Delta_w|}\left[1-\exp\left(-\frac{|\Delta_a|}{|\lambda_a+\mu_a|}\right)\left(\sqrt{1+\frac{2\sigma_a^2}{r_a^2}u^*}-1\right)\right]
+\end{align}
+where
+\begin{equation}
+u^*=v\left(1-\e^{\frac{-2r_m}{\sigma_m}}\right)
+\end{equation}
+The number of rescue mutations has a Poisson distribution with rate $\lambda_1+\lambda_2$. As a result, the probability of evolutionary rescue is given by:
+\begin{equation}
+p_{est}=1-\e^{-\left(\lambda_1+\lambda_2\right)}
+\end{equation}
+
 \end{appendices}

### 2023-05-11 EvolutionaryRescue1.tex
Remus first update

@@
 We assume that $\Delta_w<0$ due to drug therapy (or some other stress), and that $\Delta_m$ due to resistance conferred by the mutant.
-We analyze two cases: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_w$  \cite{brauner2016distinguishing}.
+We analyze two cases: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are susceptible, $0>\Delta_a>\Delta_w$  \cite{brauner2016distinguishing}.
 
@@
 The wildtype population initially consists of $w_0$ cells, whereas the other cell types are initially absent.
+\begin{figure}
+\begin{center}
+  \begin{tabular}{| l |c| c | c | c |}
+    \hline
+    Symbol & Name & Value & Units & References \\ \hline
+    $N$ & Tumor size at beginning of treatment& $10^7-10^9$ & cells  & \cite{del2009does} \\ \hline
+    $\lambda_w$ & Wildtype birth rate&  &   & 1 \\ \hline
+    $\mu_w$ & Wildtype death rate&  &   & 1 \\ \hline
+    $\lambda_a$ & Aneuploid birth rate&  &   & 1 \\ \hline
+    $\mu_a$ & Aneuploid death rate&  &   & 1 \\ \hline
+    $\lambda_m$ & Mutant birth rate&  &   & 1 \\ \hline
+    $\mu_m$ & Mutant death rate&  &   & 1 \\ \hline
+    $u$ & Chromosomal instability rate& $10^{-3}-10^{-2}$ & 1$\slash$cell division  & \cite{nowak2004evolutionary,bakker2023predicting} \\ \hline
+    $v$ & Mutation rate& $10^{-7}-10^{-9}$ &  1$\slash$gene$\slash$cell division  & \cite{nowak2004evolutionary} \\ \hline
+  \end{tabular}
+\caption{.}
+  \label{t1}
+\end{center}
+\end{figure}
 
@@
 OK
-We repeat these iterations until either the population becomes extinct (the number of cells of all types is zero) or the number of mutant cells is high enough so that its extinction probability is $<1\%$.
+We repeat these iterations until either the population becomes extinct (the number of cells of all types is zero) or the number of mutant cells is high enough so that its extinction probability is $<0.1\%$.
 % TODO how do you determine the extinction prob?
@@
 \end{equation*}
-which is the population size at which the extinction probability of the population of mutant cell is less then $1\%$. 
+which is the population size at which the extinction probability of the population of mutant cell is less then $0.1\%$. 
 % TODO do you use tau leap? if so, update the text accordingly. OK
 When the size of the initial population is large we utilize $\tau$-leaping where the increment of population change in a fixed time interval $\Delta t$ is Poisson distributed with mean $\nu_i\Delta t$ for each population $i$ \cite{gillespie2001approximate}.
-If the increment is negative and larger then the subpopulation size then updated
+If the increment is negative and larger then the subpopulation size then updated to be zero.
 \paragraph{Data availability.} All source code is available online at \url{https://github.com/yoavram-lab/EvolutionaryRescue}.
@@
 \centering
-\includegraphics[width=0.55\textwidth]{Figures/figureAneuploidy3.pdf}
+\includegraphics[width=0.75\textwidth]{Figures/figureAneuploidy.pdf}
 %\caption{}
@@
 \begin{align}\label{aneuploidyresistentfirstapprox}
-p_{est}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right] .
+p_{resc}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right] .
 \end{align}
-\Cref{SurvPlot} shows that the survival probability, $p_{est}$, quickly grows as the wildtype growth rate, $\lambda_w$, increases.
+\Cref{SurvPlot} for $N=10^4$ and \Cref{SurvPlotLargeN} for $N=10^8$  show that the survival probability, $p_{resc}$, quickly grows as the wildtype growth rate, $\lambda_w$, increases.
 % Todo I dont think this figure should show lambda_w>1, as in that case the wildtype is resistant and there is no rescue scenario.
 OK
-\Cref{SurvPlotNData} show $p_{est}$ as a function of $N$, including comparison of our approximation \eqref{aneuploidyresistentfirstapprox} % ref to equation OK
+\Cref{SurvPlotNData} show $p_{rescue}$ as a function of $N$, including comparison of our approximation \eqref{aneuploidyresistentfirstapprox} % ref to equation OK
 and simulation results.
@@
  \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/P_est_growth.pdf}
+\includegraphics[width=1\textwidth]{Figures/CombinedSubplot.pdf}
 \caption{Plot of the probability of survival of a population as a function of the proliferation rate of the wildtype cells.
+The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximation \eqref{aneuploidyresistentfirstapprox} for the upper plot and \eqref{AneuploidyDeleteriousApprox} for the lower plot.
+Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $N=10^4, \lambda_a=1+10^{-2},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
+Plot of the survival probability of an initial population consisting of $10^{4}$ wildtype cells as a function of $\Delta_a=\lambda_a-\mu_a$ for various values of $\Delta_w=\lambda_a-\mu_a$.
+The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.}
+\label{SurvPlot}
+\end{figure}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{figure}[!t]
+ \vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/CombinedSubplotLargePop.pdf}
+\caption{Plot of the probability of survival of a population as a function of the proliferation rate of the wildtype cells.
+The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximation \eqref{aneuploidyresistentfirstapprox} for the upper plot and \eqref{AneuploidyDeleteriousApprox} for the lower plot.
+Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $N=10^8, \lambda_a=1+10^{-2},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
+Plot of the survival probability of an initial population consisting of $10^{4}$ wildtype cells as a function of $\Delta_a=\lambda_a-\mu_a$ for various values of $\Delta_w=\lambda_a-\mu_a$.
+The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.}
+\label{SurvPlotLargeN}
+\end{figure}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\begin{figure}[!t]
+% \vspace*{1\baselineskip}
+%\includegraphics[width=1\textwidth]{Figures/P_est_growth.pdf}
+%\caption{Plot of the probability of survival of a population as a function of the proliferation rate of the wildtype cells.
 The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximation \eqref{survprobwapprox1}.
@@
 The error bars are present but are not visible given the fact that we have used $10^5$ simulations for each combination of parameters.}
-\label{SurvPlot}
-\end{figure}
+%\label{SurvPlot}
+%\end{figure}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \begin{align}
-p_a=\frac{\lambda_a-\mu_a-v}{\lambda_a}+\frac{vp_m}{\lambda_a-\mu_a-v}-\frac{\lambda_a\left(vp_m\right)^2}{8\left(\lambda_a-\mu_a-v\right)^3}
+p_a=\frac{\lambda_a-\mu_a-v}{\lambda_a}+\frac{vp_m}{\lambda_a-\mu_a-v}-\frac{\lambda_a\left(vp_m\right)^2}{8\left(\lambda_a-\mu_a-v\right)^3}.
 \end{align}
@@
 Given an initial population consisting of $N$ wildtype cancer cells, the probability that the population will survive is given by: 
-\begin{align}
-p_{est}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}-\frac{uv^2\lambda_a\Delta_m^2}{8\lambda_m^2\Delta_a^3}\right)\right].
+\begin{align}\label{survprobresistantaneuploidy}
+p_{resc}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}-\frac{uv^2\lambda_a\Delta_m^2}{8\lambda_m^2\Delta_a^3}\right)\right].
 \end{align}
@@
 \end{align*}
+As a result, we can approximate:
 \begin{align}\label{survprobwinitial}
@@
 \begin{align}\label{AneuploidyDeleteriousApprox}
-p_{est}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right],
+p_{resc}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right],
 \end{align}
-
+which we plot in Figure \ref{P_est} for initial population size $N=10^4$ and Figure \ref{P_est_large_N} for initial population size $N=10^8$.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[!t]
- \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/P_est.pdf}
-\caption{Plot of the survival probability of an initial population consisting of $10^{4}$ wildtype cells as a function of $\Delta_a=\lambda_a-\mu_a$ for various values of $\Delta_w=\lambda_a-\mu_a$.
+%\begin{figure}[!t]
+% \vspace*{1\baselineskip}
+%\includegraphics[width=1\textwidth]{Figures/P_est.pdf}
+%\caption{Plot of the survival probability of an initial population consisting of $10^{4}$ wildtype cells as a function of $\Delta_a=\lambda_a-\mu_a$ for various values of $\Delta_w=\lambda_a-\mu_a$.
 The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximation \eqref{AneuploidyDeleteriousApprox}.
 The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.}
-\label{P_est}
-\end{figure}
-
+%\label{P_est}
+%\end{figure}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \caption{Plot of the survival probability of an initial population consisting of $10^{4}$ wildtype cells as a function of $\Delta_w=\lambda_w-\mu_w$ for various values of $\Delta_a=\lambda_a-\mu_a$.
-The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximations \eqref{AneuploidyDeleteriousApprox}.
+The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximations \eqref{AneuploidyDeleteriousApprox} and \eqref{pestthirdcase}.
 The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
@@
 \label{P_est}
+\end{figure}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{figure}[!t]
+ \vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/P_est_divergenceLargePopulation.pdf}
+\caption{Plot of the survival probability of an initial population consisting of $10^{8}$ wildtype cells as a function of $\Delta_a=\lambda_a-\mu_a$ for various values of $\Delta_w=\lambda_w-\mu_w$.
+The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximations \eqref{AneuploidyDeleteriousApprox}.
+The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
+}
+\label{P_est_large_N}
 \end{figure}
@@
 \begin{equation}
-p_a=\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}\left(1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\lambda_avp_m}\right)^{\frac12}}{2\lambda_a}
+p_a=\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}\left(1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\lambda_avp_m}\right)^{\frac12}}{2\lambda_a},
 \end{equation}
@@
 \begin{align*}
-\left(1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\lambda_avp_m}\right)^{\frac{1}{2}}=1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{8\lambda_avp_m}+\cdots
+\left(1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\lambda_avp_m}\right)^{\frac{1}{2}}=1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{8\lambda_avp_m}+\cdots,
 \end{align*}
@@
 &=\frac{4\lambda_avp_m+4\lambda_avp_m\left(1+\frac{\lambda_a-\mu_a-v}{2\sqrt{\lambda_avp_m}}\right)^2}{8\lambda_a\sqrt{\lambda_avp_m}}\\
-&=\frac{1}{2\lambda_a}\left(\lambda_a-\mu_a-v+2\sqrt{\lambda_avp_m}\right)
+&=\frac{1}{2\lambda_a}\left(\lambda_a-\mu_a-v+2\sqrt{\lambda_avp_m}\right).
 \end{align*}
@@
 \begin{align}\label{pestthirdcase}
-p_{est}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w-u-v}\left(v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right)\right],
+p_{resc}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w-u-v}\left(v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right)\right],
 \end{align}
@@
 The probability of evolutionary rescue is given by:
-\begin{equation}
-p_{est}\sim\left\{
+\begin{equation}\label{probresc}
+p_{resc}\sim\left\{
   \begin{array}{@{}ll@{}}
@@
 \begin{align*}
-&\lambda_w'=\lambda_w,\quad\mu_w'=\mu_w\\
-&\lambda_a'=C_1+\left(\lambda_a-\mu_a\right)\left(1-\frac{w+a+m}{K}\right),\quad \mu_a'=C_1\\
-&\lambda_m'=C_2+\left(\lambda_m-\mu_m\right)\left(1-\frac{w+a+m}{K}\right),\quad \mu_m'=C_2
+&\lambda_w'=\lambda_w,\quad\mu_w'=\mu_w,\\
+&\lambda_a'=C_1+\left(\lambda_a-\mu_a\right)\left(1-\frac{w+a+m}{K}\right),\quad \mu_a'=C_1,\\
+&\lambda_m'=C_2+\left(\lambda_m-\mu_m\right)\left(1-\frac{w+a+m}{K}\right),\quad \mu_m'=C_2,
 \end{align*}
@@
 \caption{Plot of the survival probability of an initial population consisting of wildtype cells as a function of the initial population size.
-Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_w=1-10^{-1},\lambda_w=1+10^{-4},\lambda_m=1+10^{-1},\mu_w=1,\mu_m=1,u=10^{-2},v=10^{-7}, C_1=C_2=1$.}
+Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_w=1-10^{-1},\lambda_a=1+10^{-4},\lambda_m=1+10^{-1},\mu_w=1,\mu_m=1,u=10^{-2},v=10^{-7}, C_1=C_2=1$.}
 \label{SurvPlotNDataLogisticKComplete}
@@
 \subsection*{Standing genetic variation}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{figure}[!t]
+ \vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/FractionPlot.pdf}
+\caption{Plot of the fraction of the cases the population is rescue by standing genetic variation as a function of the fraction of initial cells which are aneuploid.
+The red vertical line highlights the the value of $f$ for which half the times the population is rescues by aneuploid cells while the pink line is our approximation \eqref{halfeqstandvar}.
+For this plot we have chosen the following parameters: $N=10^3, \lambda_w=1-10^{-2}, \lambda_a=1-10^{-4},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
+The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
+The value of $0.332$ highlighted in red is the value of the initial fraction of the population which in aneuploid for which half of the cases of evolutionary rescue is due to initial aneuploid cell population.}
+\label{FractionPlot}
+\end{figure}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 So far we have assumed that the initial population of cells consisted entirely of wildtype cells.
@@
 \begin{equation}\label{halfeqstandvar}
-f^*\approx\frac{p_w}{p_w+p_a},
+f^*\approx\frac{p_w}{p_w+p_a}.
 \end{equation}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection*{Contribution of aneuploidy to the evolutionary rescue of cancer}
+We wish to understand the contribution of aneuploidy to evolutionary rescue of the cancer cell population.
+For this purpose we define the ratio of the probability of evolutionary rescue when aneuploidy can play a role in rescue ($u>0$) to the probability where acquisition of aneuploidy is not possible ($u=0$):
+\begin{equation}\label{ratiorescueexact}
+H=\frac{\left.p_{resc}\right|_{u>0}}{\left.p_{resc}\right|_{u=0}}
+\end{equation}
+As a result, we obtain from \eqref{probresc} the approximation for the ratio:
+\begin{equation}\label{ratiorescue}
+H\sim\left\{
+  \begin{array}{@{}ll@{}}
+  \frac{1-\exp\left[\frac{N}{\Delta_w-u-v}\left(v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right)\right]}{1-\exp\left[\frac{vN\Delta_m}{\left(\Delta_w-v\right)\lambda_m}\right]},\quad\text{if }4\lambda_avp_m>\left(\Delta_a-v\right)^2,\\
+   \frac{1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right]}{1-\exp\left(\frac{v\Delta_mN}{\lambda_m\Delta_w}\right)},\quad\text{if }\Delta_a<0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2,\\
+   \frac{1-\exp\left[\frac{N}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right]}{1-\exp\left[\frac{N}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{v\Delta_m}{\lambda_m}\right)\right]},\quad\text{if }\Delta_a>0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2.
+  \end{array}\right.
+\end{equation}
+We plot \eqref{ratiorescue} in Figure \ref{RatioEvolRescue} for both resistent and susceptible aneuploidy as a function of the proliferation rate of the wildtype cells and in Figure \ref{RatioEvolRescuePopulationSize} as a function of the initial population size of wildtype cells.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{figure}[!t]
+ \vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/RatioEvolRescue.pdf}
+\caption{Plot of the ratio of the probability of evolutionary rescue when aneuploidy can play a role in rescue ($u>0$) to the probability where acquisition of aneuploidy is not possible ($u=0$) as a function of the proliferation rate of the wildtype cells.
+The continuous lines represent the exact result \eqref{ratiorescueexact} while the dashed lines represent the approximation \eqref{ratiorescue}.
+Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_w=1-10^{-1},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
+The upper plot show the case when aneuploidy is resistant while the lower plot shows the case when it is susceptible.}
+\label{RatioEvolRescue}
+\end{figure}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{figure}[!t]
+ \vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/RatioEvolRescuePopulationSize.pdf}
+\caption{Plot of the ratio of the probability of evolutionary rescue when aneuploidy can play a role in rescue ($u>0$) to the probability where acquisition of aneuploidy is not possible ($u=0$) as a function of the initial population size of wildtype cells.
+The continuous lines represent the exact result \eqref{ratiorescueexact} while the dashed lines represent the approximation \eqref{ratiorescue}.
+Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
+The upper plot show the case when aneuploidy is resistant while the lower plot shows the case when it is susceptible.
+The vertical red line in the lower plot separates the regime $Nv<1$ from the the $Nv>1$.}
+\label{RatioEvolRescuePopulationSize}
+\end{figure}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
  \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/MeanTimeGrowthAneuploidyPlot.pdf}
-\caption{Plot of the mean time until the appearence of a resistence mutation which rescues the population in the case when evolutionary rescue is possible only through mutation but not aneuploidy and mutation.
+\includegraphics[width=1\textwidth]{Figures/MeanTimeGrowthMutantDirectPlot.pdf}
+\caption{Plot of the mean time until the appearance of a resistance mutation which rescues the population in the case when evolutionary rescue is possible only through mutation but not aneuploidy and mutation.
 Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_w=1-10^{-1},\lambda_m=1+10^{-1},\mu_w=1,\mu_m=1,u=10^{-2},v=10^{-7}$.}
@@
 We calculate the mean time for the appearance of the first mutant that rescues the cancer cell population.
-Let $\tau=\tau_1+\tau_2$ be the time the first mutant cell that appears and rescues the population, where $\tau_1$ is the time at which the first aneuploid cell appears whose lineage will rescue the population and $\tau_2$ the time for this aneuploidy lineage to produce the first mutant that rescues the population.
+This can occur either through the pathway $wildtype\rightarrow aneuploid \rightarrow mutant$ or through the pathway $wildtype \rightarrow mutant$.
+We start with the second pathway: let $T_1$ be the time at which the first mutant cell appears which rescues the population when evolutionary rescue is only possible through mutation.
+We are interested in the mean time $\tau_1=\mathbb{E}\left[T_1\right]$.
+
 % something is missing in the next line...?
@@
 
-The time $\tau_1$ can be aproximated by a inhomogeneus Poisson process:
-\begin{align}
-\tau_1\sim \text{Poisson}\left(\frac{1}{up_aw_t}\right),
-\end{align}
+The number of successful mutants generated until time $t$ can be approximated by a inhomogeneous Poisson process with rate $R\left(t\right)=up_aw_t$
+%\begin{align}
+%\tau_1\sim \text{Poisson}\left(\frac{1}{up_aw_t}\right),
+%\end{align}
 where $w_t$ is the size of the wildtype population at time $t$: 
 \begin{equation}
-w_t=W_0\e^{\Delta_wt}.
+w_t=N\e^{\Delta_wt}.
 \end{equation}
-As a result, the probability density function of $\tau_1$ is given by:
-\begin{align}
-f\left(\tau_1\right)=R\left(\tau_1\right)\e^{-\int_0^{\tau_1} R(t)\d t}
-\end{align}
-We are interested in the time to appearence of the first succesful aneuploid cell conditional on population surving we have the following:
+We are interested in the time to appearance of the first successful mutant cell conditional on population surviving:
 \begin{align*}
-P\left(\tau_1<t\right)&=P\left(\tau_1<t|M\left(t\rightarrow\infty\right)\neq0\right)P\left(M\left(t\rightarrow\infty\right)\neq0\right)\\
-&+P\left(\tau_1<t|M\left(t\rightarrow\infty\right)=0\right)P\left(M\left(t\rightarrow\infty\right)=0\right)\\
-P\left(\tau_1<t\right)&=P\left(\tau_1<t|M\left(t\rightarrow\infty\right)\neq0\right)P\left(M\left(t\rightarrow\infty\right)\neq0\right)
+P\left(T_1<t\right)&=P\left(T_1<t|m_{t\rightarrow\infty}\neq0\right)P\left(m_{t\rightarrow\infty}\neq0\right)\\
+&+P\left(T_1<t|m_{t\rightarrow\infty}=0\right)P\left(m_{t\rightarrow\infty}=0\right).
+\end{align*}
+As a result, the cumulative distribution function can be written as:
+\begin{align*}
+P\left(T_1<t\right)&=P\left(T_1<t|m_{t\rightarrow\infty}\neq0\right)P\left(m_{t\rightarrow\infty}\neq0\right),
 \end{align*}
@@
 \begin{align*}
-P\left(\tau_1<t|M\left(t\rightarrow\infty\right)=0\right)=0.
+P\left(T_1<t|m_{t\rightarrow\infty}=0\right)=0.
 \end{align*}
@@
 \begin{align}
-P\left(\tau_1<t|M\left(t\rightarrow\infty\right)\neq0\right)=\frac{P\left(\tau_1<t\right)}{1-\left(1-p_w\right)^N}
+P\left(T_1<t|m_{t\rightarrow\infty}\neq0\right)=\frac{P\left(\tau_1<t\right)}{1-\left(1-p_w\right)^N},
 \end{align}
@@
 \begin{equation}
-P\left(M\left(t\rightarrow\infty\right)\neq0\right)=1-\left(1-p_w\right)^N
+P\left(m_{t\rightarrow\infty}\neq0\right)=1-\left(1-p_w\right)^N.
 \end{equation}
-The time $\tau_1$ conditional on evolutionary rescue is given by:
+The probability density function of $T_1$ is given by:
 \begin{align}
-f\left(\tau_1|M\left(t\rightarrow\infty\right)\neq0\right)=\frac{R\left(\tau_1\right)\e^{-\int_0^{\tau_1} R(t)\d t}}{1-\left(1-p_w\right)^N}
+f_{T_1}\left(t_1\right)=R\left(t_1\right)\e^{-\int_0^{t_1} R(t)\d t}.
 \end{align}
-The expectation of $\tau_1$ is:
-\begin{align}\nonumber
-\mathbb{E}\left[\tau_1\right]&=\frac{\int_0^\infty\e^{-\int_0^\tau R(t)\d t}\d\tau}{1-\left(1-p_w\right)^N}\\
-&=\frac{\int_0^\infty\e^{-uNp_a\frac{\e^{\Delta_w\tau}-1}{\Delta_w}}\,\d\tau}{1-\left(1-p_w\right)^N}.
+As a result, the time $T_1$ conditional on evolutionary rescue is given by:
+\begin{align}
+f_{T_1}\left(t_1|m_{t\rightarrow\infty}\neq0\right)=\frac{R\left(t_1\right)\e^{-\int_0^{t_1} R(t)\d t}}{1-\left(1-p_w\right)^N}.
 \end{align}
-For an aneuploid cell, let $a_\tau$ be the number of its aneuploid descendents at time $\tau$.
+The expectation of $T_1$ is:
+\begin{align}\label{meantime1}
+\tau_1=\mathbb{E}\left[T_1\right]&=\frac{\int_0^\infty\e^{-\int_0^\tau R(t)\d t}\d\tau}{1-\left(1-p_w\right)^N}=\frac{\int_0^\infty\e^{-uNp_a\frac{\e^{\Delta_w\tau}-1}{\Delta_w}}\,\d\tau}{1-\left(1-p_w\right)^N}.
+\end{align}
+
+%Let $T=T_1+T_2$ be the time the first mutant cell that appears and rescues the population, where $T_1$ is the time at which the first aneuploid cell appears whose lineage will rescue the population and $T_2$ the time for this aneuploidy lineage to produce the first mutant that rescues the population.
+The expected mean time $\tau_1=\mathbb{E}\left[T_1\right]$ is identical to \eqref{meantime1} with adequate parameters.
+%
+%For an aneuploid cell, let $a_\tau$ be the number of its aneuploid descendents at time $\tau$.
 Each of these descentends have a probability $v\d \tau$ of producing a mutant which has a probability $p_m$ to survive.
 Consequently, the probability that the lineage descendent from a single aneuploid cell produces succesful mutant in the time interval $\left[\tau,\tau+\d \tau\right]$ is $vp_ma_t\d \tau$.
-Since the number of succesful mutants is a Poisson process, the probability of evolutionary rescue, given $a_\tau$, is given by:
+Since the number of succesful mutants is a Poisson process, the probability of evolutionary rescue occurring until time $t$, given $a_\tau$, is given by:
+%\begin{align}
+%1-\exp\left(-\int_0^t vp_ma_\tau\d \tau\right)=1-\exp\left(-vp_mA_t\right),
+%\end{align}
+%where $A_t$ is the weight of the lineage until time $t$, defined as:
+%\begin{equation}
+%A_t=\int_0^t a_\tau\d \tau.
+%\end{equation}
+%As a result, to obtain $p_a\left(t\right)$ we average over all possible values of $A_t$:
+%\begin{align}
+%p_a\left(t\right)&=\int_0^\infty \d a P\left(A_T=a\right)\left[1-\e^{-vp_mA_t}\right]\\
+%&=1-\mathbb{E}\left[\e^{-vp_mA_t}\right]
+%\end{align}
+%Following the steps outlines in \cite{weissman2009rate} we obtain:
+%\begin{align}
+%p_a\left(t\right)=\frac{\left(a_+-1\right)\left(1-a_-\right)\left(1-\e^{-\left(1-\delta\right)\left(a_+-a_-\right)t}\right)}{a_+-1+\left(1-a_-\right)\e^{-\left(1-\delta\right)\left(a_+-a_-\right)t}},
+%\end{align}
+%where
+%\begin{align}
+%a_\pm=\frac{2-\delta-vp_m\pm\sqrt{\left(\delta+vp_m\right)^2-4vp_m}}{2\left(1-\delta\right)}.
+%\end{align}
+%and
+%\begin{align}
+%\delta=\frac{\Delta_a}{\Delta_w}
+%\end{align}
+%
+%The cumulative distribution of $T_2$ is given by $\frac{p_a\left(t\right)}{p_m}$ from which we obtain the mean time:
+%\begin{align}
+%\tau_2=\mathbb{E}\left[T_2\right]=\int_0^\infty\left(1-\frac{p_a\left(t\right)}{p_a}\right)\d t.
+%\end{align}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+When $Nu\gg1$ the aneuploid population can be assumed to be deterministic and approximated by the solution to the system of ODEs:
+\begin{equation}
+a_t=\frac{Nu\e^{\Delta_wt}}{\Delta_w-\Delta_a}\left[1-\e^{\left(\Delta_w-\Delta_a\right)t}\right].
+\end{equation}
+As a result, when $N\gg1$ the number of successful mutants created by direct mutation or though aneuploidy are an inhomogeneous Poisson processes with the rates:
+
+%\begin{align}
+%f_{T_2}\left(t_2|m_{t\rightarrow\infty}\neq0\right)=\frac{R_2\left(t_2\right)\e^{-\int_0^{t_2} R_2(t)\d t}}{1-\left(1-p_w\right)^N}
+%\end{align}
+
+%\begin{align}\label{meantime2}
+%\tau_2=\mathbb{E}\left[T_2\right]&=\frac{\int_0^\infty\e^{-\int_0^\tau R_2(t)\d t}\d\tau}{1-\left(1-p_w\right)^N}=\frac{\int_0^\infty\e^{-uNp_a\frac{\e^{\Delta_w\tau}-1}{\Delta_w}}\,\d\tau}{1-\left(1-p_w\right)^N}.
+%\end{align}
+%
+%\begin{align}\label{meantime3}
+%\tau_3=\frac{\int_0^\infty\e^{-vNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}}\,\d\tau}{1-\left(1-p_w\right)^N}.
+%\end{align}
+
+\begin{align*}
+r_1\left(t\right)&=vp_m\int_0^ta_{\tau}\,\d\tau=\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_wt}-1}{\Delta_w}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right),\\
+r_2\left(t\right)&=vp_m\int_0^tw_{\tau}\,\d\tau=uNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}.
+\end{align*}
+For large initial population sizes we can assume that both rescue mutations produced through direct mutation and aneuploidy are independent and, as a result, they can be merged into a single Poisson process with rate $\left(r_1+r_2\right)\left(t\right)$.
+Consequently, the mean time to the appearance of the first rescue mutant is:
 \begin{align}
-1-\exp\left(-\int_0^t vp_ma_\tau\d \tau\right)=1-\exp\left(-vp_mA\left(t\right)\right),
+\tau_2=\frac{\int_0^\infty\e^{-\left(r_1+r_2\right)}\,\d\tau}{1-\left(1-p_w\right)^N}=\frac{\int_0^\infty\exp\left[-\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_wt}-1}{\Delta_w}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right)-uNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}\right]\,\d\tau}{1-\left(1-p_w\right)^N},
 \end{align}
-where $A\left(t\right)$ is the weight of the lineage, defined as:
-\begin{equation}
-A\left(t\right)=\int_0^t a_\tau\d \tau.
-\end{equation}
-As a result, to obtain $p_2$ we average over all possible values of $A\left(t\right)$:
-\begin{align}
-p_2\left(t\right)&=\int_0^\infty \d w P\left(W_l=w\right)\left[1-\e^{-vp_mA\left(t\right)}\right]\\
-&=1-\mathbb{E}\left[\e^{-vp_mA\left(t\right)}\right]
-\end{align}
-Following the steps outlines in \cite{weissman2009rate} we obtain:
-\begin{align}
-p_2\left(t\right)=\frac{\left(a_+-1\right)\left(1-a_-\right)\left(1-\e^{-\left(1-\delta\right)\left(a_+-a_-\right)t}\right)}{a_+-1+\left(1-a_-\right)\e^{-\left(1-\delta\right)\left(a_+-a_-\right)t}},
-\end{align}
-where
-\begin{align}
-a_\pm=\frac{2-\delta-y\pm\sqrt{\left(\delta+y\right)^2-4y}}{2\left(1-\delta\right)},
-\end{align}
-with $y=vp_m$ in $a_\pm$.
+which we plot in Figure \ref{MeanTimeDeleteriousAneuploidyPlot} as a function of the initial population size.
 
-The cumulative distribution of $T_2$ is given by $\frac{p_2\left(t\right)}{p_m}$ from which we obtain the mean time:
-\begin{align}
-\mathbb{E}\left[\tau_2\right]=\int_0^\infty\left(1-\frac{P_2\left(t\right)}{p_m}\right)\d t.
-\end{align}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{figure}[!t]
  \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/HeatMap_p_est.pdf}
-\caption{Contours of the survival probability of an initial population consisting of $10^{3}$ wildtype cells as a function of the
-   parameters $\Delta_w=\lambda_w-\mu_w$ and $\Delta_a=\lambda_a-\mu_a$.}
-\label{SurvHeatMapPlot}
+\includegraphics[width=1\textwidth]{Figures/MeanTimeDeleteriousPlot.pdf}
+\caption{Plot of the mean time until the appearance of a resistance mutation which rescues the population in the case when evolutionary rescue is possible 
+ through mutation and aneuploidy.
+Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_w=1-10^{-1},\lambda_a=1-10^{-2},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1,u=10^{-2},v=10^{-7}$.}
+\label{MeanTimeDeleteriousAneuploidyPlot}
+\end{figure}
+
+\begin{align}\nonumber
+I&=\frac{\tau_2}{\tau_1}=\frac{\int_0^\infty\exp\left[-\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_wt}-1}{\Delta_w}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right)-uNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}\right]\,\d\tau}{\int_0^\infty\e^{-uNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}}\,\d\tau}\times\frac{1-\left(1-\left.p_w\right|_{u=0}\right)^N}{1-\left(1-\left.p_w\right|_{u>0}\right)^N}\\
+&=\frac{\int_0^\infty\exp\left[-\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_wt}-1}{\Delta_w}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right)-uNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}\right]\,\d\tau}{\int_0^\infty\e^{-uNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}}\,\d\tau}\frac{1}{H}
+\end{align}
+where $H$, is the ratio of the probability of evolutionary rescue with and without aneuploidy, defined in \eqref{ratiorescueexact}.
+
+\begin{figure}[!t]
+ \vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/MeanTimeRatioInitialPopulationSize.pdf}
+\caption{Plot of the probability of survival of a population as a function of the initial population size of wildtype cells.
+The blue line represents the exact solution \eqref{survprobw}, the orange line line represents the approximation \eqref{aneuploidyresistentfirstapprox}, the green line represents the first order correction \eqref{survprobwapproxcorrected} and the red dots represents stochastic simulations.
+For the simulations we have chosen the following parameters: $\lambda_w=1-10^{-2}, \lambda_a=1+10^{-2},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
+The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.}
+\label{MeanTimeRatioInitialPopulationSize}
 \end{figure}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[!t]
- \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/FractionPlot.pdf}
-\caption{Plot of the fraction of the cases the population is rescue by standing genetic variation as a function of the fraction of initial cells which are aneuploid.
-The red vertical line highlights the the value of $f$ for which half the times the population is rescues by aneuploid cells while the pink line is our approximation \eqref{halfeqstandvar}.
-For this plot we have chosen the following parameters: $N=10^3, \lambda_w=1-10^{-2}, \lambda_a=1-10^{-4},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
-The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
-The value of $0.332$ highlighted in red is the value of the initial fraction of the population which in aneuploid for which half of the cases of evolutionary rescue is due to initial aneuploid cell population.}
-\label{FractionPlot}
-\end{figure}
 
+\newpage
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section*{Discussion}
+
+In this paper, we have modelled a population of cancer cells which are exposed to chemotherapeutic drugs and decline towards extinction.
+Evolutionary rescue  is the process where the population acquires a trait that increases fitness in the new environment such that extinction is averted.
+We have derived the probability of evolutionary rescue of the population of cancer cells under various demographic scenarious.
+The cancer cell population can escape extinction either through direct mutation or through mutation from aneuploidy.
+We have used multitype branching processes to study our model (Figure \ref{figureAneuploidy}) which allows us to obtain exact solutions for the probability of evolutionary rescue.
+
+The case when the aneuploid cells are resistant can be approximated by the one step evolutionary rescue process where the aneuploidy rescues the population (Figure \ref{SurvPlot}).
+However, when the growth rate of the aneuploid cells is negative then they cannot rescue the population and they can only act as a stepping stone (Figure \ref{P_est}) through which the mutant can be obtain in a more expedient fashion, given that the aneuploid population declines slower then the wildtype population, compared to the case of direct mutation from the wildtype.
+
+We observe from Figure \ref{RatioEvolRescue} that aneuploidy has a significant contribution towards evlutionary rescue.
+When aneuploidy is slightly increasing ($\Delta_a=10^{-3}$) the probability of evolutionary rescue is three orders of magnitude larger when aneuploidy is present compared to the case when aneuploidy is not present.
+
+For our model we have assumed that cancer cell lineages are independent of each other.
+However this is not always true as cancer cells compete for resources which has an effect on the probability of evolutionary rescue (Figure \ref{SurvPlotNDataLogisticKComplete}).
+Future work should include using density dependent branching process in order to better model the conditions under cancer cells proliferate.
+
+The presence of aneuploid cancer cells at the onset of chemotherapy can facillitate evolutionary rescue by acting as a stepping stone for the appearance of resistant mutant cells.
+From Figure \ref{FractionPlot} we observe that, for even a relative small fraction of the initial population being composed of aneuploid cells, evolutionary rescue is more likely to occour through the initial aneuploidy.
+
+We propose experiments similar to the ones highlighted in \cite{martin2013probability} in order to test the predictions of our model.
+For example, in order to study the effects of initial population size on the probability of evolutionary rescue we propose to derive a large culture mass from a single cancer cell in permissive conditions and then dilute to a wide range of stating population sizes ($10^7-10^9$).
+Afterwards, we expose the population to anti-cancer drug which induces aneuploidy or to saline solution for control.
+Final density, in both cases, would be measured by optical density and the results compared to predictions from out model.
+
+We observe from equations \eqref{survprobw} and \eqref{probresc} that the probability of evolutionary rescue increases when the initial population size increases, the wildtype population does not decline too quickly, the mutation and aneuploidy rates are high and the probabilities $p_a$ and $p_m$ are elevated.
 
@@
 \begin{equation*}
-\mathcal{P}\left(z,t\right)=\sum_{i=0}^\infty p_i\left(t\right)z^i
+\mathcal{P}\left(z,t\right)=\sum_{i=0}^\infty p_i\left(t\right)z^i,
 \end{equation*}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Mean time}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-We define the following events
-\begin{align*}
-\text{EvR}&=\text{``evolutionary rescue occurs''},\\
-\text{EvRA}&=\text{``evolutionary rescue occurs through aneuploidy''},\\
-\text{EvRM}&=\text{``evolutionary rescue occurs through mutation''}.
-\end{align*}
-As a result, the mean time to appearence of the first mutant cell which rescues the population given evolutionary rescue can be expanded as:
-\begin{align*}
-\mathbb{E}\left[\tau|\text{EvR}\right]&=\mathbb{E}\left[\tau|\text{EvR},\text{EvRA}\cap\text{EvRM}^{C}\right]p\left(\text{EvRA}\cap\text{EvRM}^{C}|\text{EvR}\right)\\
-&+\mathbb{E}\left[\tau|\text{EvR},\text{EvRM}\cap\text{EvRA}^{C}\right]p\left(\text{EvRM}\cap\text{EvRA}^{C}|\text{EvR}\right)\\
-&+\mathbb{E}\left[\tau|\text{EvR},\text{EvRM}\cap\text{EvRA}\right]p\left(\text{EvRM}\cap\text{EvRA}|\text{EvR}\right).
-\end{align*}
-The joint probabilities in the previous lines can be written as:
-\begin{align*}
-p\left(\text{EvRA}\cap\text{EvRM}^{C}|\text{EvR}\right)&=\frac{p\left(\text{EvRA}\cap\text{EvRM}^{C}\right)}{p\left(\text{EvR}\right)}
-\approx\frac{p\left(\text{EvRA}\right)\left[1-p\left(\text{EvRM}\right)\right]}{p\left(\text{EvR}\right)},\\
-p\left(\text{EvRM}\cap\text{EvRA}^{C}|\text{EvR}\right)&=\frac{p\left(\text{EvRM}\cap\text{EvRA}^{C}\right)}{p\left(\text{EvR}\right)}
-\approx\frac{p\left(\text{EvRM}\right)\left[1-p\left(\text{EvRA}\right)\right]}{p\left(\text{EvR}\right)},\\
-p\left(\text{EvRM}\cap\text{EvRA}|\text{EvR}\right)&=\frac{p\left(\text{EvRM}\cap\text{EvRA}\right)}{p\left(\text{EvR}\right)}\approx\frac{p\left(\text{EvRM}\right)p\left(\text{EvRA}\right)}{p\left(\text{EvR}\right)}\ll1.
-\end{align*}
-As a result, we have:
-\begin{align*}
-\mathbb{E}\left[\tau|\text{EvR}\right]&\approx\mathbb{E}\left[\tau|\text{EvR},\text{EvRA}\cap\text{EvRM}^{C}\right]p\left(\text{EvRA}\cap\text{EvRM}^{C}|\text{EvR}\right)\\
-&+\mathbb{E}\left[\tau|\text{EvR},\text{EvRM}\cap\text{EvRA}^{C}\right]p\left(\text{EvRM}\cap\text{EvRA}^{C}|\text{EvR}\right).
-\end{align*}
+%\section*{Mean time}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%We define the following events
+%\begin{align*}
+%\text{EvR}&=\text{``evolutionary rescue occurs''},\\
+%\text{EvRA}&=\text{``evolutionary rescue occurs through aneuploidy''},\\
+%\text{EvRM}&=\text{``evolutionary rescue occurs through mutation''}.
+%\end{align*}
+%As a result, the mean time to appearence of the first mutant cell which rescues the population given evolutionary rescue can be expanded as:
+%\begin{align*}
+%\mathbb{E}\left[\tau|\text{EvR}\right]&=\mathbb{E}\left[\tau|\text{EvR},\text{EvRA}\cap\text{EvRM}^{C}\right]p\left(\text{EvRA}\cap\text{EvRM}^{C}|\text{EvR}\right)\\
+%&+\mathbb{E}\left[\tau|\text{EvR},\text{EvRM}\cap\text{EvRA}^{C}\right]p\left(\text{EvRM}\cap\text{EvRA}^{C}|\text{EvR}\right)\\
+%&+\mathbb{E}\left[\tau|\text{EvR},\text{EvRM}\cap\text{EvRA}\right]p\left(\text{EvRM}\cap\text{EvRA}|\text{EvR}\right).
+%\end{align*}
+%The joint probabilities in the previous lines can be written as:
+%\begin{align*}
+%p\left(\text{EvRA}\cap\text{EvRM}^{C}|\text{EvR}\right)&=\frac{p\left(\text{EvRA}\cap\text{EvRM}^{C}\right)}{p\left(\text{EvR}\right)}
+%\approx\frac{p\left(\text{EvRA}\right)\left[1-p\left(\text{EvRM}\right)\right]}{p\left(\text{EvR}\right)},\\
+%p\left(\text{EvRM}\cap\text{EvRA}^{C}|\text{EvR}\right)&=\frac{p\left(\text{EvRM}\cap\text{EvRA}^{C}\right)}{p\left(\text{EvR}\right)}
+%\approx\frac{p\left(\text{EvRM}\right)\left[1-p\left(\text{EvRA}\right)\right]}{p\left(\text{EvR}\right)},\\
+%p\left(\text{EvRM}\cap\text{EvRA}|\text{EvR}\right)&=\frac{p\left(\text{EvRM}\cap\text{EvRA}\right)}{p\left(\text{EvR}\right)}\approx\frac{p\left(\text{EvRM}\right)p\left(\text{EvRA}\right)}{p\left(\text{EvR}\right)}\ll1.
+%\end{align*}
+%As a result, we have:
+%\begin{align*}
+%\mathbb{E}\left[\tau|\text{EvR}\right]&\approx\mathbb{E}\left[\tau|\text{EvR},\text{EvRA}\cap\text{EvRM}^{C}\right]p\left(\text{EvRA}\cap\text{EvRM}^{C}|\text{EvR}\right)\\
+%&+\mathbb{E}\left[\tau|\text{EvR},\text{EvRM}\cap\text{EvRA}^{C}\right]p\left(\text{EvRM}\cap\text{EvRA}^{C}|\text{EvR}\right).
+%\end{align*}
 %%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{equation}
-r=\lim_{\Delta t\rightarrow0}\frac{\mathbb{E}\left(\Delta n_t|n_t\right)}{\Delta_t n_t}
-\end{equation}
-\begin{equation}
-\sigma=\lim_{\Delta t\rightarrow0}\frac{\mathbb{V}ar\left(\Delta n_t|n_t\right)}{\Delta_t n_t}
-\end{equation}
+An alternative method to obtain the probability of evolutionary rescue is to utilize a Feller diffusion approximation which is governed by two parameters: the growth rate $r$ and the reproductive variance $\sigma$.
+The two parameters are obtained from the underlying demographic process as the infinitesimal relative change in mean and variance of $n_t$ over and an infinitesimally small time interval $\Delta t$:
+\begin{align*}
+r&=\lim_{\Delta t\rightarrow0}\frac{\mathbb{E}\left(\Delta n_t|n_t\right)}{\Delta_t n_t},\\
+\sigma&=\lim_{\Delta t\rightarrow0}\frac{\mathbb{V}ar\left(\Delta n_t|n_t\right)}{\Delta_t n_t}.
+\end{align*}
+The rate at which mutants are generated directly from the wildtype is: 
 \begin{align}
-\lambda_1=v\pi_f\frac{w_0}{|r_w|}
+\theta_1=v\bar{\pi}_f\frac{N}{|r_w|},
 \end{align}
@@
 \begin{align}
-\bar{\pi}_f=\int_{0}^\infty\int_{0}^\infty \left(1-\e^{\frac{-2r}{\sigma}}\right) f_r\left(r,\sigma\right)\d r\d\sigma
+\bar{\pi}_f=\int_{0}^\infty\int_{0}^\infty \left(1-\e^{-\frac{2r}{\sigma}}\right) f_r\left(r,\sigma\right)\d r\d\sigma.
 \end{align}
-Let $f\left(r,\sigma\right)=\delta\left(r-r_m\right)\delta\left(\sigma-\sigma_m\right)$ then:
+Letting $f_r\left(r,\sigma\right)=\delta\left(r-r_m\right)\delta\left(\sigma-\sigma_m\right)$ then:
 \begin{align}
-\bar{\pi}_f=1-\e^{\frac{-2r_m}{\sigma_m}}
+\bar{\pi}_f=1-\e^{-\frac{2r_m}{\sigma_m}},
 \end{align}
+and, as a result, we have:
 \begin{align}
-\lambda_1=v\left(1-\e^{\frac{-2r_m}{\sigma_m}}\right)\frac{N_0}{|\Delta_w|}
+\theta_1=v\left(1-\e^{-\frac{2r_m}{\sigma_m}}\right)\frac{N}{|\Delta_w|}.
 \end{align}
-\begin{align}
-\lambda_2&=\frac{uw_0}{|r_w|}\int_{0}^\infty\int_{0}^\infty \left(1-\pi_f\left(r,q\right)\right) f_r\left(r,\sigma\right) p_1^*\left(r,\sigma\right)\d r\d\sigma\\
-&=\frac{uw_0}{|\Delta_w|}p_1^*\left(r_a,\sigma_a\right)\\
-&=\frac{uw_0}{|\Delta_w|}\left[1-\exp\left(-\frac{|\Delta_a|}{|\lambda_a+\mu_a|}\right)\left(\sqrt{1+\frac{2\sigma_a^2}{r_a^2}u^*}-1\right)\right]
-\end{align}
+The rate at which mutants are generated indirectly from the wildtype through aneuploidy is: 
+\begin{align*}
+\theta_2&=\frac{uN}{|r_w|}\int_{0}^\infty\int_{0}^\infty \left(1-\pi_f\left(r,q\right)\right) f_r\left(r,\sigma\right) p_1^*\left(r,\sigma\right)\d r\d\sigma\\
+&=\frac{uN}{|\Delta_w|}p_1^*\left(r_a,\sigma_a\right)\\
+&=\frac{uN}{|\Delta_w|}\left[1-\exp\left(-\frac{|r_a|}{\sigma_a^2}\left(\sqrt{1+\frac{2\sigma_a^2}{r_a^2}u^*}-1\right)\right)\right]\\
+&=\frac{uN}{|\Delta_w|}\left[1-\exp\left(-\frac{|\Delta_a|}{(\lambda_a+\mu_a)^2}\left(\sqrt{1+\frac{2(\lambda_a+\mu_a)^2}{\Delta_a^2}u^*}-1\right)\right)\right],
+\end{align*}
 where
 \begin{equation}
-u^*=v\left(1-\e^{\frac{-2r_m}{\sigma_m}}\right)
+u^*=v\left(1-\e^{-\frac{2r_m}{\sigma_m}}\right).
 \end{equation}
-The number of rescue mutations has a Poisson distribution with rate $\lambda_1+\lambda_2$. As a result, the probability of evolutionary rescue is given by:
-\begin{equation}
-p_{est}=1-\e^{-\left(\lambda_1+\lambda_2\right)}
+The number of rescue mutations has a Poisson distribution with rate $\theta_1+\theta_2$. As a result, the probability of evolutionary rescue is given by:
+\begin{equation}\label{FellerApprox}
+p_{rescue}=1-\e^{-\left(\theta_1+\theta_2\right)},
 \end{equation}
-
+which we plot in Figure \ref{P_est_diffusion} as a function of $\Delta_w$.
 \end{appendices}
@@
 \beginsupplement % https://support.authorea.com/en-us/article/how-to-create-an-appendix-section-or-supplementary-information-1g25i5a/
-
+\begin{figure}[!t]
+ \vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/P_est_diffusion.pdf}
+\caption{Plot of the survival probability of an initial population consisting of $w_0=10^{4}$ wildtype cells as a function of $\Delta_a=\lambda_a-\mu_a$ for various values of $\Delta_w=\lambda_a-\mu_a$.
+The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the Feller diffusion approximation \eqref{FellerApprox}.
+The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/w_0}$ where $p$ is the mean probability of evolutionary rescue.}
+\label{P_est_diffusion}
+\end{figure}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2023-05-29 EvolutionaryRescue1.tex
ain

@@
 \newcommand{\e}{{\rm e}}
+\newcommand{\erfc}{{\rm erfc}}
 \newcommand{\ii}{{\rm i}}
@@
 
-The smallest solution of the quadratic equation \cref{quadraticeqev1} gives the survival probability, which still depends on both $p_a$ and $p_m$,
+The smallest solution of the quadratic equation \cref{quadraticeqev1} gives the survival probability \cite{uecker2015adaptive}, which still depends on both $p_a$ and $p_m$,
 \begin{align}\label{survprobw}
@@
 
+An alternative method to obtain $p_m$ would be to solve the following quadratic equation:
+\begin{align}
+1-p_m=\frac{\mu_m}{\lambda_m+\mu_m}+\frac{\lambda_m}{\lambda_m+\mu_m}\left(1-p_m\right)^2.
+\end{align}
 % TODO add subsection on the Gillespie simulations OK
@@
 The wildtype population initially consists of $w_0$ cells, whereas the other cell types are initially absent.
-\begin{figure}
+\begin{table}
 \begin{center}
@@
     $N$ & Tumor size at beginning of treatment& $10^7-10^9$ & cells  & \cite{del2009does} \\ \hline
-    $\lambda_w$ & Wildtype birth rate&  &   & 1 \\ \hline
-    $\mu_w$ & Wildtype death rate&  &   & 1 \\ \hline
-    $\lambda_a$ & Aneuploid birth rate&  &   & 1 \\ \hline
-    $\mu_a$ & Aneuploid death rate&  &   & 1 \\ \hline
-    $\lambda_m$ & Mutant birth rate&  &   & 1 \\ \hline
-    $\mu_m$ & Mutant death rate&  &   & 1 \\ \hline
+    $\lambda_w$ & Wildtype birth rate& 0.14 & 1/days  & \cite{bozic2013evolutionary} \\ \hline
+    $\mu_w$ & Wildtype death rate& 0.17 & 1/days  & \cite{bozic2013evolutionary} \\ \hline
+    $\lambda_a$ & Aneuploid birth rate& 0.14 & 1/days  & * \\ \hline
+    $\mu_a$ & Aneuploid death rate& $0.13-0.17$ & 1/days  & * \\ \hline
+    $\lambda_m$ & Mutant birth rate& 0.14 & 1/days  & \cite{bozic2013evolutionary} \\ \hline
+    $\mu_m$ & Mutant death rate& 0.13 & 1/days  & \cite{bozic2013evolutionary} \\ \hline
     $u$ & Chromosomal instability rate& $10^{-3}-10^{-2}$ & 1$\slash$cell division  & \cite{nowak2004evolutionary,bakker2023predicting} \\ \hline
@@
   \end{tabular}
-\caption{.}
-  \label{t1}
+\caption{\label{demo-table}\textbf{Model parameters.} Table of parameters used in our model (see Figure \ref{figureAneuploidy}).
+The * symbol in the reference column means that for those parameters the values have not been selected from a paper.
+For the aneuploid birth rate we have chosen the same value for the wildtype and mutant birth rates.
+For the anueploid death rate we have chosen an intermediate value between the wildtype and mutant death rates.}
+  \label{table1}
 \end{center}
-\end{figure}
+\end{table}
 
@@
 
-\caption{\textbf{Model illustration.}
+\caption{
 A population of cancer cells is subdivided to wildtype euploid, wildtype aneuploid and mutant euploid cells, which divide with rates $\lambda_w$, $\lambda_a$, and $\lambda_m$, respectively, and die at rates $\mu_w$, $\mu_a$, and $\mu_m$, respectively.
@@
 %&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{v\left(\lambda_a-\mu_a-u\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-u\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\ \label{survprobw2}
-&\approx-\frac{1}{\lambda_w-\mu_w}\left[\frac{v\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right].
+&\approx-\frac{1}{\lambda_w-\mu_w}\left[\frac{u\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right].
 \end{align}
@@
 \begin{equation}\label{survprobwapprox1}
-p_w=-\frac{1}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right).
+p_w=-\frac{1}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right).
 \end{equation}
@@
 \begin{align}\label{aneuploidyresistentfirstapprox}
-p_{resc}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right] .
+p_{resc}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right] .
 \end{align}
@@
 OK
-\Cref{SurvPlotNData} show $p_{rescue}$ as a function of $N$, including comparison of our approximation \eqref{aneuploidyresistentfirstapprox} % ref to equation OK
-and simulation results.
 
@@
 \includegraphics[width=1\textwidth]{Figures/CombinedSubplot.pdf}
-\caption{Plot of the probability of survival of a population as a function of the proliferation rate of the wildtype cells.
+\caption{Plot of the probability of evolutionary rescue of a population, consisting initially of $N$ wildtype cells, as a function of the proliferation rate of the wildtype cells for various values of the proliferation rate of the aneuploid cells.
 The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximation \eqref{aneuploidyresistentfirstapprox} for the upper plot and \eqref{AneuploidyDeleteriousApprox} for the lower plot.
+The dots represent numerical simulations where the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
 Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $N=10^4, \lambda_a=1+10^{-2},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
-Plot of the survival probability of an initial population consisting of $10^{4}$ wildtype cells as a function of $\Delta_a=\lambda_a-\mu_a$ for various values of $\Delta_w=\lambda_a-\mu_a$.
-The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.}
+}
 \label{SurvPlot}
@@
 \includegraphics[width=1\textwidth]{Figures/CombinedSubplotLargePop.pdf}
-\caption{Plot of the probability of survival of a population as a function of the proliferation rate of the wildtype cells.
+\caption{Plot of the probability of evolutionary rescue of a population, consisting initially of $N$ wildtype cells, as a function of the proliferation rate of the wildtype cells for various values of the proliferation rate of the aneuploid cells.
 The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximation \eqref{aneuploidyresistentfirstapprox} for the upper plot and \eqref{AneuploidyDeleteriousApprox} for the lower plot.
-Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $N=10^8, \lambda_a=1+10^{-2},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
-Plot of the survival probability of an initial population consisting of $10^{4}$ wildtype cells as a function of $\Delta_a=\lambda_a-\mu_a$ for various values of $\Delta_w=\lambda_a-\mu_a$.
-The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.}
+The dots represent numerical simulations where the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
+Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $N=10^8, \lambda_w=0.14, \lambda_a=0.14,\lambda_m=0.14,\mu_m=0.13, u=10^{-2}, v=10^{-7}$.
+}
 \label{SurvPlotLargeN}
@@
 \begin{align}\nonumber
-p_w&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{v\left(\lambda_a-\mu_a-u\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-u\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}-\frac{uv^2\lambda_a\left(\lambda_m-\mu_m\right)^2}{8\lambda_m^2\left(\lambda_a-\mu_a-v\right)^3}\right]\\ \label{survprobw3}
-&\approx-\frac{1}{\lambda_w-\mu_w}\left[\frac{v\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}-\frac{uv^2\lambda_a\left(\lambda_m-\mu_m\right)^2}{8\lambda_m^2\left(\lambda_a-\mu_a\right)^3}\right].
+p_w&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{u\left(\lambda_a-\mu_a-v\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-v\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}-\frac{uv^2\lambda_a\left(\lambda_m-\mu_m\right)^2}{8\lambda_m^2\left(\lambda_a-\mu_a-v\right)^3}\right]\\ \label{survprobw3}
+&\approx-\frac{1}{\lambda_w-\mu_w}\left[\frac{u\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}-\frac{uv^2\lambda_a\left(\lambda_m-\mu_m\right)^2}{8\lambda_m^2\left(\lambda_a-\mu_a\right)^3}\right].
 \end{align}
@@
 \begin{equation}\label{survprobwapproxcorrected}
-p_w=-\frac{1}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}-\frac{uv^2\lambda_a\Delta_m^2}{8\lambda_m^2\Delta_a^3}\right).
+p_w=-\frac{1}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}-\frac{uv^2\lambda_a\Delta_m^2}{8\lambda_m^2\Delta_a^3}\right).
 \end{equation}
@@
 \begin{align}\label{survprobresistantaneuploidy}
-p_{resc}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}-\frac{uv^2\lambda_a\Delta_m^2}{8\lambda_m^2\Delta_a^3}\right)\right].
-\end{align}
+p_{resc}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}-\frac{uv^2\lambda_a\Delta_m^2}{8\lambda_m^2\Delta_a^3}\right)\right].
+\end{align}
+\Cref{SurvPlotNData} show $p_{rescue}$ as a function of $N$, including comparison of our first approximation \eqref{aneuploidyresistentfirstapprox} % ref to equation OK
+and simulation results.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{figure}[!t]
+ \vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/SurvPlotNData.pdf}
+\caption{Plot of the probability of survival of a population as a function of the initial population size of wildtype cells.
+The blue line represents the exact solution \eqref{survprobw}, the orange line line represents the approximation \eqref{aneuploidyresistentfirstapprox}, the green line represents the first order correction \eqref{survprobwapproxcorrected} and the red dots represents stochastic simulations.
+For the simulations we have chosen the following parameters: $\lambda_w=0.14, \lambda_a=0.14,\lambda_m=0.14,\mu_w=0.17,\mu_a=0.135,\mu_m=0.13$.
+The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n=100$ is the number of simulations.}
+\label{SurvPlotNData}
+\end{figure}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \includegraphics[width=1\textwidth]{Figures/P_est_divergence.pdf}
-\caption{Plot of the survival probability of an initial population consisting of $10^{4}$ wildtype cells as a function of $\Delta_w=\lambda_w-\mu_w$ for various values of $\Delta_a=\lambda_a-\mu_a$.
+\caption{Plot of the probability of evolutionary rescue, of an initial population, consisting of $N$ wildtype cells, as a function of proliferation rate of the wildtype cells $\Delta_w=\lambda_w-\mu_w$ for various values of the proliferation rate of the aneuploid cells $\Delta_a=\lambda_a-\mu_a$.
 The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximations \eqref{AneuploidyDeleteriousApprox} and \eqref{pestthirdcase}.
-The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
+The dots represent numerical simulations where the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
 The value highlighted in grey is the threshold $\Delta_a^*$ from \eqref{thresholdvalueaneuploid} which marks the transition between the regime dictated by \eqref{AneuploidyDeleteriousApprox} to the one dictated by \eqref{pestthirdcase}.
+Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $N=10^4,\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
 }
@@
 \includegraphics[width=1\textwidth]{Figures/P_est_divergenceLargePopulation.pdf}
-\caption{Plot of the survival probability of an initial population consisting of $10^{8}$ wildtype cells as a function of $\Delta_a=\lambda_a-\mu_a$ for various values of $\Delta_w=\lambda_w-\mu_w$.
+\caption{Plot of the probability of evolutionary rescue, of an initial population, consisting of $N$ wildtype cells, as a function of proliferation rate of the wildtype cells $\Delta_w=\lambda_w-\mu_w$ for various values of the proliferation rate of the aneuploid cells $\Delta_a=\lambda_a-\mu_a$.
 The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximations \eqref{AneuploidyDeleteriousApprox}.
-The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
+The dots represent numerical simulations where the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
+Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $N=10^8,\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
 }
@@
 \includegraphics[width=1\textwidth]{Figures/DeleteriousTauLeapPlot.pdf}
-\caption{Plot of the probability of survival of a population as a function of the initial population size of wildtype cells.
-The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximation \eqref{pestthirdcase}.
-The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
-The error bars are present but are not visible given the fact that we have used $n=10^5$ simulations for each combination of parameters.}
+\caption{Plot of the probability of evolutionary rescue of a population, consisting of $N$ wildtype cells, as a function of the initial population size of wildtype cells.
+The blue line represent the exact result \eqref{survprobw} while the orange lines represent the approximation \eqref{pestthirdcase}.
+The green dots represent numerical simulations where the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
+The error bars are present but are not visible given the fact that we have used $n=10^5$ simulations for each combination of parameters.
+Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.}
 \label{DeleteriousPlot}
@@
    1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right],\quad\text{if }\Delta_a<0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2,\\
-   1-\exp\left[\frac{N}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right],\quad\text{if }\Delta_a>0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2.
+   1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right],\quad\text{if }\Delta_a>0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2.
   \end{array}\right.
@@
  \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/SurvPlotNDataLogisticKComplete.pdf}
-\caption{Plot of the survival probability of an initial population consisting of wildtype cells as a function of the initial population size.
-Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_w=1-10^{-1},\lambda_a=1+10^{-4},\lambda_m=1+10^{-1},\mu_w=1,\mu_m=1,u=10^{-2},v=10^{-7}, C_1=C_2=1$.}
+\includegraphics[width=1\textwidth]{Figures/SurvPlotNDataLogisticK.pdf}
+\caption{Plot of the probability of evolutionary rescue of a population, consisting of $N$ wildtype cells, as a function of the initial population size of wildtype cells where maximum population size is constrained by the carrying capacity $K$.
+The blue line represent the exact result \eqref{survprobw} while the orange dots represent numerical simulations where the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
+Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_w=1-10^{-1},\lambda_a=1+10^{-4},\lambda_m=1+10^{-1},\mu_w=1,\mu_m=1,u=10^{-2},v=10^{-7}, C_1=C_2=1, K=10^9$.}
 \label{SurvPlotNDataLogisticKComplete}
@@
 The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
-The value of $0.332$ highlighted in red is the value of the initial fraction of the population which in aneuploid for which half of the cases of evolutionary rescue is due to initial aneuploid cell population.}
+The value of $0.332$ highlighted in red is the value of the initial fraction of the population which in aneuploid for which half of the cases of evolutionary rescue are due to the initial aneuploid cell population.}
 \label{FractionPlot}
@@
 \begin{equation}\label{ratiorescueexact}
-H=\frac{\left.p_{resc}\right|_{u>0}}{\left.p_{resc}\right|_{u=0}}
+H=\frac{\left.p_{resc}\right|_{u>0}}{\left.p_{resc}\right|_{u=0}}.
 \end{equation}
@@
    \frac{1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right]}{1-\exp\left(\frac{v\Delta_mN}{\lambda_m\Delta_w}\right)},\quad\text{if }\Delta_a<0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2,\\
-   \frac{1-\exp\left[\frac{N}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right]}{1-\exp\left[\frac{N}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{v\Delta_m}{\lambda_m}\right)\right]},\quad\text{if }\Delta_a>0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2.
+   \frac{1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right]}{1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\right]},\quad\text{if }\Delta_a>0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2.
   \end{array}\right.
 \end{equation}
+
+%\begin{equation}
+%H\sim\left\{
+%  \begin{array}{@{}ll@{}}
+%  \frac{1-\exp\left[\frac{N}{\Delta_w-u-v}\left(v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right)\right]}{1-\exp\left[\frac{vN\Delta_m}{\left(\Delta_w-v\right)\lambda_m}\right]},\quad\text{if }4\lambda_avp_m>\left(\Delta_a-v\right)^2,\\
+%   \frac{1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right]}{1-\exp\left(\frac{v\Delta_mN}{\lambda_m\Delta_w}\right)},\quad\text{if }\Delta_a<0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2,\\
+%   \frac{1-\exp\left(\frac{u\Delta_aN}{\lambda_a\Delta_w}\right)}{1-\exp\left(\frac{v\Delta_mN}{\lambda_m\Delta_w}\right)},\quad\text{if }\Delta_a>0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2.
+%  \end{array}\right.
+%\end{equation}
 We plot \eqref{ratiorescue} in Figure \ref{RatioEvolRescue} for both resistent and susceptible aneuploidy as a function of the proliferation rate of the wildtype cells and in Figure \ref{RatioEvolRescuePopulationSize} as a function of the initial population size of wildtype cells.
@@
 The continuous lines represent the exact result \eqref{ratiorescueexact} while the dashed lines represent the approximation \eqref{ratiorescue}.
+The upper plot show the case when aneuploidy is resistant while the lower plot shows the case when it is susceptible.
 Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_w=1-10^{-1},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
-The upper plot show the case when aneuploidy is resistant while the lower plot shows the case when it is susceptible.}
+}
 \label{RatioEvolRescue}
@@
 The continuous lines represent the exact result \eqref{ratiorescueexact} while the dashed lines represent the approximation \eqref{ratiorescue}.
-Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
 The upper plot show the case when aneuploidy is resistant while the lower plot shows the case when it is susceptible.
-The vertical red line in the lower plot separates the regime $Nv<1$ from the the $Nv>1$.}
+Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_w=0.14, \lambda_a=0.14,\lambda_m=0.14,\mu_w=0.17,\mu_m=0.13, u=10^{-2}, v=10^{-7}$.
+}
 \label{RatioEvolRescuePopulationSize}
@@
 \caption{Plot of the mean time until the appearance of a resistance mutation which rescues the population in the case when evolutionary rescue is possible only through mutation but not aneuploidy and mutation.
+The blue line represents the approximation \eqref{meantime1} and the dashed red line represents the first order approximation \eqref{limitapprox}.
+The orange dots represent the numerical simulations while the error bars represent the interval centered at the mean which containing $95\%$ of the simulated values.
 Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_w=1-10^{-1},\lambda_m=1+10^{-1},\mu_w=1,\mu_m=1,u=10^{-2},v=10^{-7}$.}
@@
 \end{align}
+The fration in exponential of the integrand in \eqref{meantime1} can approximated as:
+\begin{equation*}
+\frac{\e^{\Delta_w\tau}-1}{\Delta_w}=\frac{1+\Delta_w\tau+O(\tau^2)-1}{\Delta_w}=\tau+O(\tau^2).
+\end{equation*}
+As a result, the mean time $\tau_1$ can be simplified:
+\begin{align}\label{limitapprox}
+\tau_1\approx\left(1+\e^{-Np_w}\right)\int_0^\infty\e^{-uNp_a\tau}\,\d\tau=\frac{\left(1+\e^{-Np_w}\right)}{uNp_a},
+\end{align}
+where in the previous line we have used the approximation:
+\begin{equation*}
+\frac{1}{1-\e^{-Np_w}}\approx1+\e^{-Np_w}.
+\end{equation*}
+We plot the expansion \eqref{limitapprox} in Figure \ref{MeanTimeGrowthAneuploidyPlot} and observe that it is a very good fit for intermediate and large values of the initial wildtype population size.
 
@@
 r_1\left(t\right)&=vp_m\int_0^ta_{\tau}\,\d\tau=\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_wt}-1}{\Delta_w}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right),\\
-r_2\left(t\right)&=vp_m\int_0^tw_{\tau}\,\d\tau=uNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}.
+r_2\left(t\right)&=vp_m\int_0^tw_{\tau}\,\d\tau=vNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}.
 \end{align*}
@@
 Consequently, the mean time to the appearance of the first rescue mutant is:
-\begin{align}
-\tau_2=\frac{\int_0^\infty\e^{-\left(r_1+r_2\right)}\,\d\tau}{1-\left(1-p_w\right)^N}=\frac{\int_0^\infty\exp\left[-\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_wt}-1}{\Delta_w}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right)-uNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}\right]\,\d\tau}{1-\left(1-p_w\right)^N},
+\begin{align}\label{meantimet2}
+\tau_2=\frac{\int_0^\infty\e^{-\left(r_1+r_2\right)}\,\d\tau}{1-\left(1-p_w\right)^N}=\frac{\int_0^\infty\exp\left[-\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_w\tau}-1}{\Delta_w}-\frac{\e^{\Delta_a\tau}-1}{\Delta_a}\right)-vNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}\right]\,\d\tau}{1-\left(1-p_w\right)^N},
 \end{align}
@@
 
+We wish to obtain a simpler formula for $\tau_2$ in an analogous way to \eqref{limitapprox}. For this, we make use of the following expansions:
+\begin{align*}
+\frac{\e^{\Delta_w\tau}-1}{\Delta_w}&=\frac{1+\Delta_w\tau+\frac{\Delta_w^2\tau^2}{2}+O(\tau^3)-1}{\Delta_w}=\tau+\frac{\Delta_w}{2}\tau^2+O(\tau^3).\\
+\frac{\e^{\Delta_a\tau}-1}{\Delta_a}&=\frac{1+\Delta_a\tau+\frac{\Delta_a^2\tau^2}{2}+O(\tau^3)-1}{\Delta_a}=\tau+\frac{\Delta_w}{2}\tau^2+O(\tau^3),
+\end{align*}
+which allow us to write:
+\begin{equation*}
+\frac{\e^{\Delta_w\tau}-1}{\Delta_w}-\frac{\e^{\Delta_a\tau}-1}{\Delta_a}\approx\frac{\left(\Delta_w-\Delta_a\right)\tau^2}{2}.
+\end{equation*}
+As a result, the integrand in \eqref{meantimet2} can be written as:
+\begin{align*}
+&\exp\left[-\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_w\tau}-1}{\Delta_w}-\frac{\e^{\Delta_a\tau}-1}{\Delta_a}\right)-vNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}\right]\approx\exp\left(-uvNp_m\tau^2-vNp_m\tau\right)\\
+&=\exp\left(\frac{vNp_m}{2}\right)\exp\left[-\frac{uvNp_m}{2}\left(\tau+\frac{1}{u}\right)\right].
+\end{align*}
+Consequently, the mean time $\tau_2$ is obtained to be:
+\begin{align}\label{limitapprox2}
+\tau_2\approx\left[1+\exp\left(-Np_w\right)\right]\exp\left(\frac{vNp_m}{2u}\right)\frac{\erfc\left(\sqrt{\frac{vNp_m}{2u}}\right)}{\sqrt{\frac{2uvNp_m}{\pi}}},
+\end{align}
+where $\erfc$ is the complementary error function.
+We plot the expansion \eqref{limitapprox2} in Figure \ref{MeanTimeDeleteriousAneuploidyPlot} and observe that it is a very good fit for large values of the initial wildtype population size.
+
+If we select only linear terms in the following expansions:
+\begin{align*}
+\frac{\e^{\Delta_w\tau}-1}{\Delta_w}&=\frac{1+\Delta_w\tau+O(\tau^2)-1}{\Delta_w}=\tau+O(\tau^2).\\
+\frac{\e^{\Delta_a\tau}-1}{\Delta_a}&=\frac{1+\Delta_a\tau+O(\tau^2)-1}{\Delta_a}=\tau+O(\tau^2),
+\end{align*}
+we obtain the first order approximation for $\tau_2$:
+\begin{align}\label{limitapprox3}
+\tau_2\approx\left(1+\e^{-Np_w}\right)\int_0^\infty\e^{-uNp_m\tau}\,\d\tau=\frac{\left(1+\e^{-Np_w}\right)}{uNp_m},
+\end{align}
+which we plot in Figure \ref{MeanTimeDeleteriousAneuploidyPlot} and observe that it offer as a good a fit to \eqref{meantimet2} as \eqref{limitapprox2}.
+Additionally, we observe that for large initial wildtype populations sizes direct mutation drives evolutionary rescue while aneuploidy plays a role for intermediate sized tumors.
+This is consistent with the information obtained from Figure \ref{RatioEvolRescuePopulationSize} where aneuploidy improves the probability of evolutionary rescue only  for small and intermediate values of $N$.
 \begin{figure}[!t]
@@
  through mutation and aneuploidy.
-Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_w=1-10^{-1},\lambda_a=1-10^{-2},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1,u=10^{-2},v=10^{-7}$.}
+Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_w=1-10^{-1},\lambda_a=1-10^{-2},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1,u=10^{-2},v=10^{-7}$.
+The blue line represents the approximation \eqref{meantimet2}, the dashed red line represents the second order approximation \eqref{limitapprox2} and the green line is first order approximation \eqref{limitapprox3}.
+The orange dots represent the numerical simulations while the error bars represent the interval centered at the mean which containing $95\%$ of the simulated values.}
 \label{MeanTimeDeleteriousAneuploidyPlot}
@@
 
+\subsection*{Contribution of aneuploidy to mean evolutionary rescue time}
 \begin{align}\nonumber
-I&=\frac{\tau_2}{\tau_1}=\frac{\int_0^\infty\exp\left[-\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_wt}-1}{\Delta_w}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right)-uNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}\right]\,\d\tau}{\int_0^\infty\e^{-uNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}}\,\d\tau}\times\frac{1-\left(1-\left.p_w\right|_{u=0}\right)^N}{1-\left(1-\left.p_w\right|_{u>0}\right)^N}\\
-&=\frac{\int_0^\infty\exp\left[-\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_wt}-1}{\Delta_w}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right)-uNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}\right]\,\d\tau}{\int_0^\infty\e^{-uNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}}\,\d\tau}\frac{1}{H}
+I&=\frac{\tau_2}{\tau_1}=\frac{\int_0^\infty\exp\left[-\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_wt}-1}{\Delta_w}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right)-vNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}\right]\,\d\tau}{\int_0^\infty\e^{-uNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}}\,\d\tau}\times\frac{1-\left(1-\left.p_w\right|_{u=0}\right)^N}{1-\left(1-\left.p_w\right|_{u>0}\right)^N}\\\label{eqMeanTimeRatioInitialPopulationSize}
+&=\frac{\int_0^\infty\exp\left[-\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_wt}-1}{\Delta_w}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right)-vNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}\right]\,\d\tau}{\int_0^\infty\e^{-vNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}}\,\d\tau}\frac{1}{H},
 \end{align}
 where $H$, is the ratio of the probability of evolutionary rescue with and without aneuploidy, defined in \eqref{ratiorescueexact}.
+We plot \eqref{eqMeanTimeRatioInitialPopulationSize} in Figure \ref{MeanTimeRatioInitialPopulationSize} as a function of the initial wildtype population for varying values of the Malthusian fitness of aneuploid cells $\Delta_a$.
 
@@
 \includegraphics[width=1\textwidth]{Figures/MeanTimeRatioInitialPopulationSize.pdf}
-\caption{Plot of the probability of survival of a population as a function of the initial population size of wildtype cells.
-The blue line represents the exact solution \eqref{survprobw}, the orange line line represents the approximation \eqref{aneuploidyresistentfirstapprox}, the green line represents the first order correction \eqref{survprobwapproxcorrected} and the red dots represents stochastic simulations.
-For the simulations we have chosen the following parameters: $\lambda_w=1-10^{-2}, \lambda_a=1+10^{-2},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
-The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.}
+\caption{Plot of the ratio of the mean time to evolutionary rescue when aneuploidy can play a role in rescue ($u>0$) to the mean time where acquisition of aneuploidy is not possible ($u=0$) as a function of the initial population size of wildtype cells.
+The continuous lines represent the approximation \eqref{eqMeanTimeRatioInitialPopulationSize}.
+Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_w=1-10^{-1},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.}
 \label{MeanTimeRatioInitialPopulationSize}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[!t]
- \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/SurvPlotNData.pdf}
-\caption{Plot of the probability of survival of a population as a function of the initial population size of wildtype cells.
-The blue line represents the exact solution \eqref{survprobw}, the orange line line represents the approximation \eqref{aneuploidyresistentfirstapprox}, the green line represents the first order correction \eqref{survprobwapproxcorrected} and the red dots represents stochastic simulations.
-For the simulations we have chosen the following parameters: $\lambda_w=1-10^{-2}, \lambda_a=1+10^{-2},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
-The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.}
-\label{SurvPlotNData}
 \end{figure}
@@
 
-We observe from Figure \ref{RatioEvolRescue} that aneuploidy has a significant contribution towards evlutionary rescue.
-When aneuploidy is slightly increasing ($\Delta_a=10^{-3}$) the probability of evolutionary rescue is three orders of magnitude larger when aneuploidy is present compared to the case when aneuploidy is not present.
+We observe from Figure \ref{RatioEvolRescue} that aneuploidy has a significant contribution towards evolutionary rescue.
+When aneuploidy is slightly increasing ($\Delta_a=10^{-3}$) the probability of evolutionary rescue is three orders of magnitude larger when aneuploidy is present compared to the case when aneuploidy is not present under the parameters previously described for tumors (see Table \ref{table1}).
 
 For our model we have assumed that cancer cell lineages are independent of each other.
-However this is not always true as cancer cells compete for resources which has an effect on the probability of evolutionary rescue (Figure \ref{SurvPlotNDataLogisticKComplete}).
+However this is not always true as cancer cells compete for resources which can have an effect on the probability of evolutionary rescue.
+We observe that this is not the case when the carrying capacity if sufficiently large the probability of evolutionary rescue is not impacted by the logistic model (see Figure \ref{SurvPlotNDataLogisticKComplete}).
 Future work should include using density dependent branching process in order to better model the conditions under cancer cells proliferate.
@@
 We observe from equations \eqref{survprobw} and \eqref{probresc} that the probability of evolutionary rescue increases when the initial population size increases, the wildtype population does not decline too quickly, the mutation and aneuploidy rates are high and the probabilities $p_a$ and $p_m$ are elevated.
+
+The probability of evolutionary rescue is enhanced by aneuploidy for small and intermediate sized tumors (see Figure \ref{RatioEvolRescuePopulationSize}).
+As a result, aneuploidy is unlikely to to contribute to primary tumors overcoming chemotherapy but it can contribute to the evolutionary rescue of secondary tumors whose size might be below the detection threshold of $\sim10^7$  \cite{bozic2013evolutionary}.
+Given the fact that the mean time for small and intermediate tumors to overcome chemotherapy can be of the order of 100 days (see Figure \ref{MeanTimeDeleteriousAneuploidyPlot})  this could explain the reappearance of cancer even after initial remission.
 
@@
 \end{align*}
+Setting $N=1$ and using the fact that the probability of survival is $p=1-p_0$ we obtain:
+\begin{equation*}
+p =  \left\{\begin{array}{@{}ll@{}}
+  1,\quad \text{if }\lambda\leq\mu, \\
+  \frac{\lambda-\mu}{\lambda},\quad \text{if }\lambda>\mu.
+  \end{array}\right.
+\end{equation*}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2023-05-30 EvolutionaryRescue1.tex
revised intro, results, figures - need to proceed from rescue prob

@@
 \usepackage{subcaption}
-\usepackage{xurl}
+%\usepackage{xurl}
 
@@
 
+% NatBib
+%\usepackage[super,comma]{natbib}
+%\usepackage[numbers,super,compress]{natbib}
+\usepackage[round,colon,authoryear]{natbib}
+
 % Title page
@@
 \author[1,*]{Yoav Ram}
-\affil[1]{School of Zoology, Faculty of Life Sciences, Tel Aviv University}
-\affil[2]{Department of Human Molecular Genetics and Biochemistry, Faculty of Medicine, Tel Aviv University}
-\affil[3]{Department of Physics, Emory University}
+\affil[1]{School of Zoology, Faculty of Life Sciences, Tel Aviv University, Tel Aviv, Israel}
+\affil[2]{Department of Human Molecular Genetics and Biochemistry, Faculty of Medicine, Tel Aviv University, Tel Aviv, Israel}
+\affil[3]{Department of Physics, Emory University, Atlanta, GA}
 \affil[*]{Corresponding author: yoav@yoavram.com}
@@
 
-\paragraph{Aneuploidy in cancer.} Chromosomal instability (CIN) is the mitotic process in which cells suffer from chromosome mis-segregation that leads to aneuploidy, where cells are characterized by structural changes of the chromosomes and copy number alterations \cite{schukken2018cin}.
+\paragraph{Aneuploidy in cancer.} Chromosomal instability (CIN) is the mitotic process in which cells suffer from chromosome mis-segregation that leads to aneuploidy, where cells are characterized by structural changes of the chromosomes and copy number alterations \citep{schukken2018cin}.
 Interestingly, aberrations in chromosome copy number have been shown to allow cancer cells to survive under stressful conditions such as drug therapy.
-Indeed, cancer cells are often likely to be aneuploid, and aneuploidy is associated with poor patient outcomes \cite{ben2020context}.
+Indeed, cancer cells are often likely to be aneuploid, and aneuploidy is associated with poor patient outcomes \citep{ben2020context}.
 
@@
 
-The role of chromosomal instability (CIN) in the emergence of cancer has been studied extensively in the past decades \cite{michor2005can,christine2018understanding,nowak2002role,pavelka2010dr,komarova2003mutation,zhu2018cellular}.
+The role of chromosomal instability (CIN) in the emergence of cancer has been studied extensively in the past decades \citep{michor2005can,christine2018understanding,nowak2002role,pavelka2010dr,komarova2003mutation,zhu2018cellular}.
 One hypothesis is that CIN facilitates tumor genesis by accelerating the removal of tumor suppression genes (TSG) and subsequent appearance of cancer.
 The deletion of tumor suppression genes can happen in two ways: two point mutations deleting both alleles of the TSG (assuming a diploid genotype), or one point mutation and one chromosomal loss event.
-Initial theoretical studies have shown that aneuploidy can have a significant role in the deletion of the the tumor suppressing genes when compared to two consecutive point mutations \cite{nowak2002role,komarova2003mutation,michor2005can,komarova2008selective}.
-% TODO mention if studies are theoretical , experimental, or clinical OK
+Initial theoretical studies have shown that aneuploidy can have a significant role in the deletion of the the tumor suppressing genes when compared to two consecutive point mutations \citep{nowak2002role,komarova2003mutation,michor2005can,komarova2008selective}.
 However, when taking into account that the appearance of aneuploidy requires a mutation to trigger CIN, the probability that CIN precedes tumor genesis is highly unlikely.
@@
 Examples of such environmental changes include climate change, invasive species or the onset of drug therapies.
-Adaptation is a race against time as the population size decreases in the new environment~\cite{tanaka2022surviving}.
-{\em Evolutionary rescue} is the process where the population acquires a trait that increases fitness in the new environment such that extinction is averted.
-It is mathematically equivalent to the problem of crossing of fitness valley \cite{weissman2009rate,weissman2010rate}.
-There are three potential ways for a population to survive environmental change: migration to a new habitat similar to the one before the onset of environmental change \cite{cobbold2020should}; adaptation by phenotypic plasticity without genetic modification \cite{carja2019evolutionary,carja2017evolutionary,levien2021non}; and adaptation through genetic modifications, e.g., mutation \cite{uecker2014evolutionary,uecker2016role,uecker2011fixation}.
-
-Models of evolutionary rescue usually assume that the fitness of the wildtype and mutant are independent of time.
-An exception is \cite{marrec2020adapt}, where the fitness of the wildtype and mutant are time dependent.
-Additionally, \cite{uecker2011fixation} investigated the probability of fixation of a beneficial mutation in a variable environment with arbitrary time-dependent selection coefficent and population size.
+Adaptation is a race against time as the population size decreases in the new environment~\citep{tanaka2022surviving}.
+\emph{Evolutionary rescue} is the process where the population acquires a trait that increases fitness in the new environment such that extinction is averted.
+It is mathematically equivalent to the problem of crossing of fitness valley \citep{weissman2009rate,weissman2010rate}.
+There are three potential ways for a population to survive environmental change: migration to a new habitat similar to the one before the onset of environmental change \citep{cobbold2020should}; adaptation by phenotypic plasticity without genetic modification \citep{carja2019evolutionary,carja2017evolutionary,levien2021non}; and adaptation through genetic modifications, e.g., mutation \citep{uecker2014evolutionary,uecker2016role,uecker2011fixation}.
+
+Models of evolutionary rescue usually assume that the fitness of the wildtype and mutant are homogeneous in time.
+An exception was given by \citet{marrec2020adapt}, who modeled the fitness of the wildtype and mutant as time dependent.
+Additionally, \citet{uecker2011fixation} investigated the probability of fixation of a beneficial mutation in a variable environment with arbitrary time-dependent selection coefficient and population size.
 Most models focus on the probability that at least one mutation rescues the population.
-How multiple mutations contribute to the survival of the population is less explored.
-One exception is \cite{wilson2017soft} which showed that evolutionary rescue is significantly enhanced by soft selective sweeps when multiple mutations contribute.
-Evolutionary rescue that requires two successive mutations has been investigated by \cite{martin2013probability} with the help of diffusion approximation.
+How multiple mutations contribute to the survival of the population is less explored, but \citet{wilson2017soft} have shown that evolutionary rescue is significantly enhanced by soft selective sweeps when multiple mutations contribute.
+Evolutionary rescue that requires two successive mutations has been investigated using diffusion approximation by \citet{martin2013probability}.
 
@@
 
-We follow the number of cancer cells that belong to three different types at time $t$: wildtype euploid, $w_t$; wildtype aneuploid, $a_t$; and mutant euploid, $m_t$. 
-These cells grow and die with rates $\lambda_k$ and $\mu_k$ for $k=w, a, m$.
-Wildtype cells become aneuploid at rate $u$. Aneuploid and wildtype cells mutate with rate $v$ (Figure \ref{figureAneuploidy}). 
-Thus, the changes in the number of each cell type is described by
-\begin{subequations}
-\begin{flalign}
-w_t&\rightarrow w_t+1:\quad \lambda_ww_t,\\
-w_t&\rightarrow w_t-1:\quad \left(\mu_w+u+v\right)w_t,\\
-a_t&\rightarrow a_t+1:\quad \lambda_aa_t+uw_t,\\
-a_t&\rightarrow a_t-1:\quad \left(\mu_a+v\right)a_t,\\
-m_t&\rightarrow m_t+1:\quad \lambda_am_t+va_t+vm_t,\\
-m_t&\rightarrow m_t-1:\quad \mu_am_t.
-\end{flalign}
-\end{subequations}
-
-The difference between growth and death is $\Delta_k = \lambda_k-\mu_k$.
-We assume that $\Delta_w<0$ due to drug therapy (or some other stress), and that $\Delta_m$ due to resistance conferred by the mutant.
-We analyze two cases: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are susceptible, $0>\Delta_a>\Delta_w$  \cite{brauner2016distinguishing}.
-
-Here, \emph{evolutionary rescue} occurs when the aneuploid cells are able to either prevent (when $\Delta_a>0$) or delay (when $0>\Delta_a>\Delta_w$) the extinction of the cancer-cell population before the adaptive mutant cells appear and fixate in the populaiton.
-To analyze evolutionary rescue in this model, we use the framework of \emph{multitype branching processes} \cite{rybnikov2021fitness,harris1963theory}. 
-This allows us to find explicit expressions for the survival probability of the population--the probability that it does not become extinct due to stressful conditions, e.g., drug therapy.
-Note that branching processes imply that lineages produced by cells from the initial population grow and die independently of each other.
-This is a necessary simplification, but it is unrealistic, as cells usually compete for resources.
-A more realistic model may include competition on limited resources between lineages, as well as spatial structure, which may play an important role in the development of cancer \cite{martens2011spatial}.
-
-The survival probability of a population consisting initially of one wildtype cell satisfies the quadratic equation,
-\begin{equation} \label{quadraticeqev1}
-\begin{aligned}
-1-p_w = & \frac{\mu_w}{\lambda_w+\mu_w+u+v} + 
-		  \frac{u}{\lambda_w+\mu_w+u+v}\left(1-p_a\right) + \\
-		  & \frac{\lambda_w}{\lambda_w+\mu_w+u+v}\left(1-p_w\right)^2 +
-		  \frac{v}{\lambda_w+\mu_w+u+v}\left(1-p_m\right) ,
-\end{aligned}
-\end{equation}
-where the $p_w$, $p_a$, and $p_m$ are the survival probabilities of a population consisting initially of one wildtype cell, one aneuploid cell, and one mutant cell, respectively.
-
-The smallest solution of the quadratic equation \cref{quadraticeqev1} gives the survival probability \cite{uecker2015adaptive}, which still depends on both $p_a$ and $p_m$,
-\begin{align}\label{survprobw}
-p_w=\frac{\lambda_w-\mu_w-u-v+\sqrt{\left(\lambda_w-\mu_w-u-v\right)^2+4\lambda_w\left(up_a+vp_m\right)}}{2\lambda_w} .
-\end{align}
-
-If the population originally consists of a single aneuploid cell, then the probability that the population will survive is given by the quadratic equation
-\begin{align}
-1-p_a=\frac{\mu_a}{\lambda_a+\mu_a+v}+\frac{v}{\lambda_a+\mu_a+v}\left(1-p_m\right)+\frac{\lambda_a}{\lambda_a+\mu_a+v}\left(1-p_a\right)^2,
-\end{align}
-whose smallest solution is the survival probability:
-\begin{align}\label{survproba}
-p_a=\frac{\lambda_a-\mu_a-v+\sqrt{\left(\lambda_a-\mu_a-v\right)^2+4\lambda_avp_m}}{2\lambda_a}.
-\end{align}
-
-The probability that a single mutant cell survives is given by (Appendix \ref{AppendixSurvLin}):
-\begin{equation}
-p_m = \left\{
-  \begin{array}{@{}ll@{}}
-  \frac{\lambda_m-\mu_m}{\lambda_m},\quad \text{if }\lambda_m>\mu_m \\
-   0,\quad \text{else} .
-  \end{array}\right .
-\end{equation}
-
-An alternative method to obtain $p_m$ would be to solve the following quadratic equation:
-\begin{align}
-1-p_m=\frac{\mu_m}{\lambda_m+\mu_m}+\frac{\lambda_m}{\lambda_m+\mu_m}\left(1-p_m\right)^2.
-\end{align}
-% TODO add subsection on the Gillespie simulations OK
+We follow the number of cancer cells that have one of three different genotypes at time $t$: wildtype, $w_t$; aneuploid, $a_t$; and mutant, $m_t$. 
+These cells divide and die with rates $\lambda_k$ and $\mu_k$ (for $k=w, a, m$).
+The difference between the division and death rate is $\Delta_k = \lambda_k-\mu_k$.
+We assume the population of cells is under a strong stress, such as drug therapy, to which the wildtype genotype is susceptible and therefore $\Delta_w<0$, whereas the mutant is resistant to the stress, $\Delta_m>0$
+We analyze two cases: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_w$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}.
+Wildtype cells may missegregate to become aneuploids at rate $u$. Both aneuploid and wildtype cells may mutate to become mutants at rate $v$ (\Cref{figureAneuploidy}). 
+%Thus, the changes in the number of each cell type is described by 
+%\begin{equation}
+%\begin{aligned}
+%w_t&\rightarrow w_t+1:\quad \lambda_ww_t,\\
+%w_t&\rightarrow w_t-1:\quad \left(\mu_w+u+v\right)w_t,\\
+%a_t&\rightarrow a_t+1:\quad \lambda_aa_t+uw_t,\\
+%a_t&\rightarrow a_t-1:\quad \left(\mu_a+v\right)a_t,\\
+%m_t&\rightarrow m_t+1:\quad \lambda_am_t+va_t+vm_t,\\
+%m_t&\rightarrow m_t-1:\quad \mu_am_t.
+%\end{aligned}
+%\end{equation}
+
+%%%%%%%%%%%
+\begin{figure}[!t]
+\centering
+\includegraphics[width=0.5\textwidth]{Figures/figureAneuploidy.pdf}
+\caption{
+\textbf{Model illustration.}
+% TODO in the figure change labels to Wildtype, Aneuploid, Mutant (instead of Euploid etc)
+A population of cancer cells is composed of wildtype, aneuploid, and mutant cells, which divide with rates $\lambda_w$, $\lambda_a$, and $\lambda_m$ and die at rates $\mu_w$, $\mu_a$, and $\mu_m$, respectively.
+Wildtype cells can become aneuploid at rate $u$.
+Both aneuploid and wildtype cells can acquire a beneficial mutation with rate $v$.
+Color denotes the relative growth rates of the three genotypes such that $\lambda_w - \mu_w < \lambda_a - \mu_a < \lambda_m - \mu_m$.
+}
+\label{figureAneuploidy}
+\end{figure}
+%%%%%%%%%%%
+
+Evolutionary rescue occurs when mutant cells appear and fixate ($m_t>>1$) in the population before the population becomes extinct ($w_t=a_t=m_t=0$).
+Aneuploidy may contribute to evolutionary rescue by either preventing (when $\Delta_a>0$) or delaying (when $0>\Delta_a>\Delta_w$) the extinction of the population before mutant cells appear and fixate.
+To analyze evolutionary rescue in this model, we use the framework of \emph{multitype branching processes} \citep{rybnikov2021fitness,harris1963theory}. 
+This allows us to find explicit expressions for the \emph{survival probability}: the probability that the population does not become extinct.
+Note that branching processes assume that lineages produced by cells from the initial population divide and die independently of each other, which may be unrealistic, as cells usually compete for resources.
+A more realistic model may include competition for limited resources and spatial structure, which may play an important role in the development of cancer \citep[e.g.,][]{martens2011spatial}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection*{Evolutionary simulation} 
-Simulations are performed using a \emph{Gillespie algorithm} \cite{gillespie1976general,gillespie1977exact} implemented in Python \cite{python}.
+Simulations are performed using a \emph{Gillespie algorithm} \citep{gillespie1976general,gillespie1977exact} implemented in Python \citep{python}.
 The simulation monitors the number of cells of each type: wildtype, aneuploid, and mutant. 
 The wildtype population initially consists of $w_0$ cells, whereas the other cell types are initially absent.
+
 \begin{table}
 \begin{center}
-  \begin{tabular}{| l |c| c | c | c |}
+  \begin{tabular}{| l |p{5cm}| c | c | p{3cm} |}
     \hline
-    Symbol & Name & Value & Units & References \\ \hline
-    $N$ & Tumor size at beginning of treatment& $10^7-10^9$ & cells  & \cite{del2009does} \\ \hline
-    $\lambda_w$ & Wildtype birth rate& 0.14 & 1/days  & \cite{bozic2013evolutionary} \\ \hline
-    $\mu_w$ & Wildtype death rate& 0.17 & 1/days  & \cite{bozic2013evolutionary} \\ \hline
-    $\lambda_a$ & Aneuploid birth rate& 0.14 & 1/days  & * \\ \hline
-    $\mu_a$ & Aneuploid death rate& $0.13-0.17$ & 1/days  & * \\ \hline
-    $\lambda_m$ & Mutant birth rate& 0.14 & 1/days  & \cite{bozic2013evolutionary} \\ \hline
-    $\mu_m$ & Mutant death rate& 0.13 & 1/days  & \cite{bozic2013evolutionary} \\ \hline
-    $u$ & Chromosomal instability rate& $10^{-3}-10^{-2}$ & 1$\slash$cell division  & \cite{nowak2004evolutionary,bakker2023predicting} \\ \hline
-    $v$ & Mutation rate& $10^{-7}-10^{-9}$ &  1$\slash$gene$\slash$cell division  & \cite{nowak2004evolutionary} \\ \hline
+     & Name & Value & Units & References \\ \hline
+    $N$ & Tumor size at beginning of treatment& $10^7-10^9$ & cells  & \citet{del2009does} \\ \hline
+    $\lambda_w$ & Wildtype birth rate& 0.14 & 1/days  & \citep{bozic2013evolutionary} \\ \hline
+    $\mu_w$ & Wildtype death rate& 0.17 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
+    $\lambda_a$  & Aneuploid birth rate$^\ast$ & 0.14 & 1/days  & - \\ \hline
+    $\mu_a$ & Aneuploid death rate$^\ast$ & $0.13-0.17$ & 1/days  & - \\ \hline
+    $\lambda_m$ & Mutant birth rate& 0.14 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
+    $\mu_m$ & Mutant death rate& 0.13 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
+    $u$ & Chromosomal instability rate& $10^{-3}-10^{-2}$ & 1$\slash$cell division  & \citet{nowak2004evolutionary,bakker2023predicting} \\ \hline
+    $v$ & Mutation rate& $10^{-7}-10^{-9}$ &  1$\slash$gene$\slash$cell division  & \citet{nowak2004evolutionary} \\ \hline
   \end{tabular}
-\caption{\label{demo-table}\textbf{Model parameters.} Table of parameters used in our model (see Figure \ref{figureAneuploidy}).
-The * symbol in the reference column means that for those parameters the values have not been selected from a paper.
-For the aneuploid birth rate we have chosen the same value for the wildtype and mutant birth rates.
-For the anueploid death rate we have chosen an intermediate value between the wildtype and mutant death rates.}
+\caption{\textbf{Model parameters.} 
+Aneuploid birth rate $\lambda_a$ is set to the same value as the wildtype and mutant birth rates, $\lambda_w$ and $\lambda_m$.
+Aneuploid death rate $\mu_a$ is set to an intermediate value between the wildtype and mutant death rates, $\mu_w$ and $\mu_m$.}
   \label{table1}
@@
 
-At each iteration of the simulation loop we compute the rate $\nu_j$ of each event $j$. % TODO explains what events can occur and how you compute their rates, refer to illustration figure  OK
-The sate of the stochastic system at time $t$ is represented by the triplet $\left(w_t,a_t,m_t\right)$ which can change in the time interval $\Delta_t$ according to the following events with appropriate rates (see Figure \ref{figureAneuploidy}):
+The state of the stochastic system at time $t$ is represented by the triplet $\left(w_t,a_t,m_t\right)$.
+The following describes the events that may occur (right column), the rates at which they occur (middle column), and the effect these events have on the state (\Cref{figureAneuploidy}):
 \begin{subequations}
 \begin{flalign*}
-(+1,0,0)&:\quad \lambda_ww_t\quad\left(\text{birth euploid cell}\right),\\
-(-1,0,0)&:\quad \mu_ww_t\quad\left(\text{death euploid cell}\right),\\
-(-1,+1,0)&:\quad uw_t\quad\left(\text{euploid cell aquires aneuploidy}\right),\\
-(-1,0,+1)&:\quad vw_t\quad\left(\text{euploid cell aquires mutation}\right),\\
-(0,+1,0)&:\quad \lambda_aa_t\quad\left(\text{birth aneuploid cell}\right),\\
-(0,-1,0)&:\quad \mu_aa_t\quad\left(\text{death aneuploid cell}\right),\\
-(0,-1,+1)&:\quad va_t\quad\left(\text{aneuploid cell aquires mutation}\right),\\
-(0,0,+1)&:\quad \lambda_am_t\quad\left(\text{birth mutant cell}\right),\\
-(0,0,-1)&:\quad \mu_am_t\quad\left(\text{death mutant cell}\right).
+(+1,0,0)&:\quad \lambda_ww_t\quad\left(\text{birth of wildtype cell}\right),\\
+(-1,0,0)&:\quad \mu_ww_t\quad\left(\text{death of wildtype cell}\right),\\
+(-1,+1,0)&:\quad uw_t\quad\left(\text{wildtype cell becomes aneuploid}\right),\\
+(-1,0,+1)&:\quad vw_t\quad\left(\text{wildtype cell becomes mutant}\right),\\
+(0,+1,0)&:\quad \lambda_aa_t\quad\left(\text{birth of aneuploid cell}\right),\\
+(0,-1,0)&:\quad \mu_aa_t\quad\left(\text{death of aneuploid cell}\right),\\
+(0,-1,+1)&:\quad va_t\quad\left(\text{aneuploid cell becomes mutant}\right),\\
+(0,0,+1)&:\quad \lambda_am_t\quad\left(\text{birth of mutant cell}\right),\\
+(0,0,-1)&:\quad \mu_am_t\quad\left(\text{death of mutant cell}\right).
 \end{flalign*}
 \end{subequations}
+Each iteration of the simulation loop starts with computing the rates $\nu_j$ of each event $j$.
 We then draw the time until the next event, $\Delta t$, from an exponential distribution whose rate parameter is the sum of the rates of all events, such that $\Delta t \sim \textit{Exp}(\sum_j \nu_j)$.
@@
 OK
-We repeat these iterations until either the population becomes extinct (the number of cells of all types is zero) or the number of mutant cells is high enough so that its extinction probability is $<0.1\%$.
-% TODO how do you determine the extinction prob?
-ref to equation OK
-We continue running the simulation until:
+We repeat these iterations until either the population becomes extinct (the number of cells of all types is zero) or the number of mutant cells is high enough so that its extinction probability is $<0.1\%$, that is until
 \begin{equation*}
@@
 \end{equation*}
-which is the population size at which the extinction probability of the population of mutant cell is less then $0.1\%$. 
-% TODO do you use tau leap? if so, update the text accordingly. OK
-When the size of the initial population is large we utilize $\tau$-leaping where the increment of population change in a fixed time interval $\Delta t$ is Poisson distributed with mean $\nu_i\Delta t$ for each population $i$ \cite{gillespie2001approximate}.
-If the increment is negative and larger then the subpopulation size then updated to be zero.
-\paragraph{Data availability.} All source code is available online at \url{https://github.com/yoavram-lab/EvolutionaryRescue}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% TODO combine all diagram figures to a single figure with multiple panels. and make it fit in a single page. OK
-
-%%%%%%
-
-\begin{figure}
-\centering
-\includegraphics[width=0.75\textwidth]{Figures/figureAneuploidy.pdf}
-%\caption{}
-
-%\end{subfigure}
-
-\caption{
-A population of cancer cells is subdivided to wildtype euploid, wildtype aneuploid and mutant euploid cells, which divide with rates $\lambda_w$, $\lambda_a$, and $\lambda_m$, respectively, and die at rates $\mu_w$, $\mu_a$, and $\mu_m$, respectively.
-Wildtype cells can become aneuploid at rate $u$. Both aneuploid and wildtype cells can acquire a beneficial mutation with rate $v$.}
-\label{figureAneuploidy}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\paragraph{$\tau$-leaping.}
+% TODO "is large" - how large?
+When the size of the initial population is large we utilize $\tau$-leaping \citep{gillespie2001approximate}, where change in number of cells of genotype $i$ in a fixed time interval $\Delta t$ is Poisson distributed with mean $\nu_i\Delta t$.
+If the increment is negative and larger then the subpopulation size then updated to be zero. % TODO "and larger..."??
+
+\paragraph{Density-dependent growth.}
+To simulate birth and death rates that depend on the number of cells in the population, we transform the rates of division and death to the following:
+\begin{align*}
+\lambda_w' &= \lambda_w, \\
+\mu_w' &= \mu_w,\\
+\lambda_a' &= C_1+\left(\lambda_a-\mu_a\right)\left(1-\frac{w+a+m}{K}\right),\\ 
+\mu_a' &= C_1,\\
+\lambda_m' &= C_2+\left(\lambda_m-\mu_m\right)\left(1-\frac{w+a+m}{K}\right),\\ 
+\mu_m' &= C_2,
+\end{align*}
+where $C_1, C_2>0$ are constants. % TODO what constants?
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection*{Code and data availability.} All source code is available online at \url{https://github.com/yoavram-lab/EvolutionaryRescue}.
 
@@
 
-% OVERVIEW OF RESULTS
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% TODO give a descriptive name instead of a mathematical formula OK
-%\subsection*{First case: $4\lambda_avp_m<\left(\lambda_a-\mu_a-v\right)^2$}
-\subsection*{First case: Highly deleterious or highly beneficial aneuploidy}
-
-\paragraph{Aneuploid cells are drug resistant.} 
-We first assume wildtype cells are susceptible to the drug, $\lambda_w<\mu_w$, whereas aneuploid cells are resistant, $\lambda_a>\mu_a$.
-Thus, we rewrite \cref{survprobw,survproba} as
-\begin{align*}
-p_a&=\frac{\lambda_a-\mu_a-v}{2\lambda_a}\left(1+\sqrt{1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}}\right),\\
-p_w&=\frac{\lambda_w-\mu_w-u-v}{2\lambda_w}\left(1-\sqrt{1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}}\right).
-\end{align*}
-Using a quadratic Taylor expansion, $\sqrt{1+x}=1+x/2+O(x^2)$, 
-we obtain the following approximation for the survival probability of a population consisting of a single individual wildtype cell (assuming $u,v \ll 1$),
-\begin{align}\label{survprobwinitial}
+% TODO: OVERVIEW OF RESULTS
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection*{Survival probability}
+
+Let $p_w$, $p_a$, and $p_m$ be the survival probabilities of a population consisting initially of single wildtype cell, aneuploid cell, or mutant cell, respectively.
+The complements $1-p_w$, $1-p_a$, and $1-p_m$ are the extinction probabilities, which satisfy each its respective equation,
+\begin{equation} \label{extinction_prob}
+\begin{aligned}
+1-p_w = &\frac{\mu_w}{\lambda_w+\mu_w+u+v} + 
+		  \frac{u}{\lambda_w+\mu_w+u+v}\left(1-p_a\right) + \\
+		  & \frac{\lambda_w}{\lambda_w+\mu_w+u+v}\left(1-p_w\right)^2 +
+		  \frac{v}{\lambda_w+\mu_w+u+v}\left(1-p_m\right) ,\\
+1-p_a = &\frac{\mu_a}{\lambda_a+\mu_a+v}+\frac{v}{\lambda_a+\mu_a+v}\left(1-p_m\right)+\frac{\lambda_a}{\lambda_a+\mu_a+v}\left(1-p_a\right)^2 ,\\
+1-p_m = &\frac{\mu_m}{\lambda_m+\mu_m}+\frac{\lambda_m}{\lambda_m+\mu_m}\left(1-p_m\right)^2 .	 
+\end{aligned}
+\end{equation}
+
+The survival probabilities are given by the smallest solution for each quadratic equation \citep{uecker2015adaptive}. Therefore we have
+\begin{equation}\label{survival_prob}
+\begin{aligned}
+p_w &= \frac{\lambda_w-\mu_w-u-v+\sqrt{\left(\lambda_w-\mu_w-u-v\right)^2+4\lambda_w\left(up_a+vp_m\right)}}{2\lambda_w} ,\\
+p_a &= \frac{\lambda_a-\mu_a-v+\sqrt{\left(\lambda_a-\mu_a-v\right)^2+4\lambda_avp_m}}{2\lambda_a}, \\
+p_m &= \left\{
+  \begin{array}{@{}ll@{}}
+  \frac{\lambda_m-\mu_m}{\lambda_m},& \lambda_m \ge \mu_m \\
+   0, & \lambda_m \le \mu_m . %TODO in the appendix this is 1, not 0
+  \end{array}\right .  
+\end{aligned}
+\end{equation}
+See Appendix \ref{AppendixSurvLin} for the derivation of $p_m$. %TODO why do we need Appendix 1 if we can just find p_m as solution to quadratic equation?
+Note that the equation for $p_w$ depends on both $p_a$ and $p_m$, and the equation for $p_a$ depends on $p_m$.
+To proceed, we can plug the solution for $p_m$ and $p_a$ into the solution for $p_w$. We perform this for three different scenarios.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsubsection*{Scenario 1: Aneuploid cells are drug resistant}
+
+We first assume that aneuploidy provides partial resistance to drug therapy, $\lambda_a>\mu_a$, and that this resistance is significant, $\left(\lambda_a-\mu_a-v\right)^2 > 4\lambda_a v p_m$.
+We thus rewrite \cref{survival_prob} as
+\begin{align*}
+p_w&=\frac{\lambda_w-\mu_w-u-v}{2\lambda_w}\left(1-\sqrt{1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}}\right) ,
+\text{and} \\
+p_a&=\frac{\lambda_a-\mu_a-v}{2\lambda_a}\left(1+\sqrt{1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}}\right) . 
+\end{align*}
+Using the quadratic Taylor expansion $\sqrt{1+x}=1+x/2+O(x^2)$ and assuming $u,v \ll 1$,
+we obtain the following approximation for the survival probability of a population initially consisting of a single wildtype cell,
+\begin{align}\label{survprobwapprox1}
 p_w 
@@
 %&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{v\left(\lambda_a-\mu_a-u\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-u\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\ \label{survprobw2}
-&\approx-\frac{1}{\lambda_w-\mu_w}\left[\frac{u\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right].
-\end{align}
-
-We write \cref{survprobw} as
-\begin{equation}\label{survprobwapprox1}
-p_w=-\frac{1}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right).
-\end{equation}
-So, given an initial population of $N$ wildtype cells, the probability that the population will survive is given by
-\begin{align}\label{aneuploidyresistentfirstapprox}
-p_{resc}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right] .
-\end{align}
-\Cref{SurvPlot} for $N=10^4$ and \Cref{SurvPlotLargeN} for $N=10^8$  show that the survival probability, $p_{resc}$, quickly grows as the wildtype growth rate, $\lambda_w$, increases.
-% Todo I dont think this figure should show lambda_w>1, as in that case the wildtype is resistant and there is no rescue scenario.
-OK
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[!t]
- \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/CombinedSubplot.pdf}
-\caption{Plot of the probability of evolutionary rescue of a population, consisting initially of $N$ wildtype cells, as a function of the proliferation rate of the wildtype cells for various values of the proliferation rate of the aneuploid cells.
-The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximation \eqref{aneuploidyresistentfirstapprox} for the upper plot and \eqref{AneuploidyDeleteriousApprox} for the lower plot.
-The dots represent numerical simulations where the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
-Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $N=10^4, \lambda_a=1+10^{-2},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
-}
-\label{SurvPlot}
-\end{figure}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[!t]
- \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/CombinedSubplotLargePop.pdf}
-\caption{Plot of the probability of evolutionary rescue of a population, consisting initially of $N$ wildtype cells, as a function of the proliferation rate of the wildtype cells for various values of the proliferation rate of the aneuploid cells.
-The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximation \eqref{aneuploidyresistentfirstapprox} for the upper plot and \eqref{AneuploidyDeleteriousApprox} for the lower plot.
-The dots represent numerical simulations where the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
-Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $N=10^8, \lambda_w=0.14, \lambda_a=0.14,\lambda_m=0.14,\mu_m=0.13, u=10^{-2}, v=10^{-7}$.
-}
-\label{SurvPlotLargeN}
-\end{figure}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\begin{figure}[!t]
-% \vspace*{1\baselineskip}
-%\includegraphics[width=1\textwidth]{Figures/P_est_growth.pdf}
-%\caption{Plot of the probability of survival of a population as a function of the proliferation rate of the wildtype cells.
-The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximation \eqref{survprobwapprox1}.
-Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $N=10^4, \lambda_a=1+10^{-2},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
-The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
-The error bars are present but are not visible given the fact that we have used $10^5$ simulations for each combination of parameters.}
-%\label{SurvPlot}
-%\end{figure}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-We want to improve the accuracy of our approximation by taking into consideration the second term of the Taylor series expansion:
+&\approx-\frac{1}{\lambda_w-\mu_w}\left[\frac{u\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\
+\end{align}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\paragraph{Second-order approximation.}
+To improve our approximation, we can consider the second term of the Taylor series expansion,
 \begin{align*}
@@
 \end{align*}
-which gives us the following approximation for $p_a$:
+which gives us the following approximation,
 \begin{align}
-p_a=\frac{\lambda_a-\mu_a-v}{\lambda_a}+\frac{vp_m}{\lambda_a-\mu_a-v}-\frac{\lambda_a\left(vp_m\right)^2}{8\left(\lambda_a-\mu_a-v\right)^3}.
-\end{align}
-From which we deduce that:
+p_a \approx \frac{\lambda_a-\mu_a-v}{\lambda_a}+\frac{vp_m}{\lambda_a-\mu_a-v}-\frac{\lambda_a\left(vp_m\right)^2}{8\left(\lambda_a-\mu_a-v\right)^3} .
+\end{align}
+We therefore have
 \begin{align}\nonumber
 p_w&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{u\left(\lambda_a-\mu_a-v\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-v\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}-\frac{uv^2\lambda_a\left(\lambda_m-\mu_m\right)^2}{8\lambda_m^2\left(\lambda_a-\mu_a-v\right)^3}\right]\\ \label{survprobw3}
-&\approx-\frac{1}{\lambda_w-\mu_w}\left[\frac{u\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}-\frac{uv^2\lambda_a\left(\lambda_m-\mu_m\right)^2}{8\lambda_m^2\left(\lambda_a-\mu_a\right)^3}\right].
-\end{align}
-Using the notations described in \eqref{notationalconv} we write the above equation as:
+&\approx-\frac{1}{\lambda_w-\mu_w}\left[\frac{u\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}-\frac{uv^2\lambda_a\left(\lambda_m-\mu_m\right)^2}{8\lambda_m^2\left(\lambda_a-\mu_a\right)^3}\right],
+\end{align}
+and using $\Delta_k=\lambda_k-\mu_k$, we can write the above equation as
 \begin{equation}\label{survprobwapproxcorrected}
-p_w=-\frac{1}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}-\frac{uv^2\lambda_a\Delta_m^2}{8\lambda_m^2\Delta_a^3}\right).
-\end{equation}
-Given an initial population consisting of $N$ wildtype cancer cells, the probability that the population will survive is given by: 
-\begin{align}\label{survprobresistantaneuploidy}
-p_{resc}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}-\frac{uv^2\lambda_a\Delta_m^2}{8\lambda_m^2\Delta_a^3}\right)\right].
-\end{align}
-\Cref{SurvPlotNData} show $p_{rescue}$ as a function of $N$, including comparison of our first approximation \eqref{aneuploidyresistentfirstapprox} % ref to equation OK
-and simulation results.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[!t]
- \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/SurvPlotNData.pdf}
-\caption{Plot of the probability of survival of a population as a function of the initial population size of wildtype cells.
-The blue line represents the exact solution \eqref{survprobw}, the orange line line represents the approximation \eqref{aneuploidyresistentfirstapprox}, the green line represents the first order correction \eqref{survprobwapproxcorrected} and the red dots represents stochastic simulations.
-For the simulations we have chosen the following parameters: $\lambda_w=0.14, \lambda_a=0.14,\lambda_m=0.14,\mu_w=0.17,\mu_a=0.135,\mu_m=0.13$.
-The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n=100$ is the number of simulations.}
-\label{SurvPlotNData}
-\end{figure}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\paragraph{$\lambda_a<\mu_a$ and $\lambda_w<\mu_w$.}
-We assume that $\lambda_a<\mu_a$ and $\lambda_w<\mu_w$ and, as a result, we rewrite \eqref{survprobw} and \eqref{survproba} as:
-\begin{align*}
-p_a&=\frac{\lambda_a-\mu_a-v}{2\lambda_a}\left(1-\sqrt{1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}}\right),\\
-p_w&=\frac{\lambda_w-\mu_w-u-v}{2\lambda_w}\left(1-\sqrt{1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}}\right).
-\end{align*}
-As a result, we can approximate:
-\begin{align}\label{survprobwinitial}
+p_w \approx -\frac{1}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}-\frac{uv^2\lambda_a\Delta_m^2}{8\lambda_m^2\Delta_a^3}\right).
+\end{equation}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsubsection*{Scenario 2: Aneuploid cells are drug tolerant.}
+
+We now assume that aneuploidy provides tolerance to the drug, that is, the number of aneuploid cells significantly declines over time, but at a lower rate than the number of wildtype cells, $\lambda_w - \mu_w < \lambda_a - \mu_a < 0$.
+We also assume that both declines are significant, $\left(\lambda_a-\mu_a-v\right)^2 > 4\lambda_a v p_m$.
+We rewrite \cref{survival_prob} as
+\begin{align*}
+p_w&=\frac{\lambda_w-\mu_w-u-v}{2\lambda_w}\left(1-\sqrt{1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}}\right) ,
+\text{and} \\
+p_a&=\frac{\lambda_a-\mu_a-v}{2\lambda_a}\left(1-\sqrt{1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}}\right) .
+\end{align*}
+Since $u,v\ll1$, this can be approximated by % TODO how do you do the first approx? taylor?
+\begin{equation}\label{survprobwinitial}
+\begin{aligned}
 p_w&\approx-\frac{vp_m+up_a}{\lambda_w-\mu_w-u-v}\\
-\nonumber
-&\approx\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-v\right)}-\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\ \label{survprobw2}
-&\approx\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_w-\mu_w\right)}\left[\frac{u}{\left(\lambda_a-\mu_a\right)}-1\right],
-\end{align}
-where in the last line we have used the fact that $u,v\ll1$. Using the notational convention
-\begin{equation}\label{notationalconv}
-\Delta_i=\lambda_i-\mu_i,
-\end{equation}
-we write \eqref{survprobw} as
-\begin{equation}\label{survprobwapprox2}
-p_w=\frac{v\Delta_m}{\lambda_m\Delta_w}\left(\frac{u}{\Delta_a}-1\right).
-\end{equation}
-Given an initial population consisting of $N$ wildtype cancer cells, the probability that the population will survive is given by: 
-\begin{align}\label{AneuploidyDeleteriousApprox}
-p_{resc}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right],
-\end{align}
-which we plot in Figure \ref{P_est} for initial population size $N=10^4$ and Figure \ref{P_est_large_N} for initial population size $N=10^8$.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\begin{figure}[!t]
-% \vspace*{1\baselineskip}
-%\includegraphics[width=1\textwidth]{Figures/P_est.pdf}
-%\caption{Plot of the survival probability of an initial population consisting of $10^{4}$ wildtype cells as a function of $\Delta_a=\lambda_a-\mu_a$ for various values of $\Delta_w=\lambda_a-\mu_a$.
-The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximation \eqref{AneuploidyDeleteriousApprox}.
-The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.}
-%\label{P_est}
-%\end{figure}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[!t]
- \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/P_est_divergence.pdf}
-\caption{Plot of the probability of evolutionary rescue, of an initial population, consisting of $N$ wildtype cells, as a function of proliferation rate of the wildtype cells $\Delta_w=\lambda_w-\mu_w$ for various values of the proliferation rate of the aneuploid cells $\Delta_a=\lambda_a-\mu_a$.
-The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximations \eqref{AneuploidyDeleteriousApprox} and \eqref{pestthirdcase}.
-The dots represent numerical simulations where the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
-The value highlighted in grey is the threshold $\Delta_a^*$ from \eqref{thresholdvalueaneuploid} which marks the transition between the regime dictated by \eqref{AneuploidyDeleteriousApprox} to the one dictated by \eqref{pestthirdcase}.
-Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $N=10^4,\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
-}
-\label{P_est}
-\end{figure}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{figure}[!t]
- \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/P_est_divergenceLargePopulation.pdf}
-\caption{Plot of the probability of evolutionary rescue, of an initial population, consisting of $N$ wildtype cells, as a function of proliferation rate of the wildtype cells $\Delta_w=\lambda_w-\mu_w$ for various values of the proliferation rate of the aneuploid cells $\Delta_a=\lambda_a-\mu_a$.
-The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximations \eqref{AneuploidyDeleteriousApprox}.
-The dots represent numerical simulations where the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
-Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $N=10^8,\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
-}
-\label{P_est_large_N}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[!t]
- \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/DeleteriousTauLeapPlot.pdf}
-\caption{Plot of the probability of evolutionary rescue of a population, consisting of $N$ wildtype cells, as a function of the initial population size of wildtype cells.
-The blue line represent the exact result \eqref{survprobw} while the orange lines represent the approximation \eqref{pestthirdcase}.
-The green dots represent numerical simulations where the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
-The error bars are present but are not visible given the fact that we have used $n=10^5$ simulations for each combination of parameters.
-Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.}
-\label{DeleteriousPlot}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\subsection*{Second case: $4\lambda_avp_m>\left(\lambda_a-\mu_a-v\right)^2$}
-\subsection*{Second case: Neutral aneuploidy}
-If we assume that  $4\lambda_avp_m>\left(\lambda_a-\mu_a-v\right)^2$ then we write:
+&\approx\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-v\right)}-\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\ %\label{survprobw2}
+&\approx\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_w-\mu_w\right)}\left[\frac{u}{\left(\lambda_a-\mu_a\right)}-1\right]\\
+%&=\frac{v\Delta_m}{\lambda_m\Delta_w}\left(\frac{u}{\Delta_a}-1\right) .
+\end{aligned}
+\end{equation}
+%Using the notational convention
+%\begin{equation}\label{notationalconv}
+%\Delta_i=\lambda_i-\mu_i,
+%\end{equation}
+%we write \eqref{survprobw} as
+%\begin{equation}\label{survprobwapprox2}
+%p_w=\frac{v\Delta_m}{\lambda_m\Delta_w}\left(\frac{u}{\Delta_a}-1\right).
+%\end{equation}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsubsection*{Scenario 3: Aneuploid cells are non-growing.}
+
+We now assume that the growth rate of aneuploid cells is close to zero (either positive or negative), such that  $\left(\lambda_a-\mu_a-v\right)^2 < 4\lambda_avp_m$.
+We rewrite \cref{survival_prob} as
 \begin{equation}
-p_a=\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}\left(1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\lambda_avp_m}\right)^{\frac12}}{2\lambda_a},
-\end{equation}
-and using the following Taylor series expansion:
-\begin{align*}
+p_a=\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}\left(1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\lambda_avp_m}\right)^{\frac12}}{2\lambda_a} .
+\end{equation}
+Using a following Taylor series expansion
+\begin{equation*}
 \left(1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\lambda_avp_m}\right)^{\frac{1}{2}}=1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{8\lambda_avp_m}+\cdots,
-\end{align*}
-we obtain:
+\end{equation*}
+we obtain the approximation
 \begin{align*}
@@
 \end{align*}
-As a result, we have from \eqref{survprobwinitial} the probability of rescue of a population starting from one wildtype individual:
-\begin{align*}
+From \Cref{survprobwinitial}, % TODO I dont see how the referred equation plays a part here
+the survival probability of a population starting from one wildtype individual is
+\begin{align}\label{scenario3}
 p_w&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[v\frac{\lambda_m-\mu_m}{\lambda_m}+\frac{u}{2\lambda_a}\left(\lambda_a-\mu_a-v+2\sqrt{\lambda_avp_m}\right)\right]\\
-&=-\frac{1}{\lambda_w-\mu_w-u-v}\left[v\frac{\lambda_m-\mu_m}{\lambda_m}+\frac{u}{2\lambda_a}\left(\lambda_a-\mu_a-v\right)+u\sqrt{\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_a\lambda_m}}\right]\\
-&=-\frac{1}{\Delta_w-u-v}\left[v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right],
-\end{align*}
-where in the last line we have used the notations defined in \eqref{notationalconv}.
-
-Given an initial population consisting of $N$ wildtype cancer cells, the probability that the population will survive is given by: 
-\begin{align}\label{pestthirdcase}
-p_{resc}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w-u-v}\left(v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right)\right],
-\end{align}
-which we plot in Figure \ref{DeleteriousPlot} where we compare with numerical simulations and the exact result \eqref{survprobw}.
-The transition between the regimes defined by \eqref{AneuploidyDeleteriousApprox} and \eqref{pestthirdcase} respectively occurs at:
-\begin{align}\label{thresholdvalueaneuploid}
-\Delta_a^*=2vp_m+v+2\sqrt{vp_m\left(vp_m+\mu_a+v\right)}.
-\end{align}
-
-The probability of evolutionary rescue is given by:
-\begin{equation}\label{probresc}
-p_{resc}\sim\left\{
+&=-\frac{1}{\lambda_w-\mu_w-u-v}\left[v\frac{\lambda_m-\mu_m}{\lambda_m}+\frac{u}{2\lambda_a}\left(\lambda_a-\mu_a-v\right)+u\sqrt{\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_a\lambda_m}}\right].
+%&=-\frac{1}{\Delta_w-u-v}\left[v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right],
+\end{align}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection*{Evolutionary rescue probability}
+
+To estimate the rescue probability $p_{rescue}$, we assume independence between clonal lineages starting from an initial population of $N$ wildtype cells.
+Thus, the rescue probability is given by 
+\begin{equation}
+p_{rescue}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w} ,
+\end{equation}
+where the approximation $(1-p_w)\approx e^{-p_w}$ assumes that $p_w$ (but not $N p_w$) is small.
+
+Now, we can apply the approximations for the survival probability $p_w$ from \Cref{survprobwapprox1,survprobwinitial,scenario3} and substituting $\Delta_k=\lambda_k-\mu_k$
+
+The rescue probability is given by
+\begin{equation}\label{rescue_prob}
+p_{rescue} = \left\{
   \begin{array}{@{}ll@{}}
@@
 
+%Note that if instead \cref{survprobwapprox1} we apply \cref{survprobwapproxcorrected}, we have
+%\begin{align}\label{survprobresistantaneuploidy}
+%p_{rescue}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}-\frac{uv^2\lambda_a\Delta_m^2}{8\lambda_m^2\Delta_a^3}\right)\right].
+%\end{align}
+
+In \Cref{SurvPlot,SurvPlotLargeN} we explore the effects of the wildtype and aneuploid growth rates on the rescue probability for small and large population sizes ($N=10^4$ and $N=10^8$, respectively).
+
+\Cref{SurvPlotNData} show $p_{rescue}$ as a function of $N$, including comparison of our first approximation \eqref{aneuploidyresistentfirstapprox} and simulation results.
+
+\Cref{P_est,P_est_large_N} show the rescue probability for initial population sizes $N=10^4$ and $N=10^8$, respectively. % TODO check figure refs
+
+In \Cref{DeleteriousPlot}, we compare the exact result (\cref{survival_prob}) with numerical simulations.
+The transition between the regimes defined by \cref{AneuploidyDeleteriousApprox} and \cref{pestthirdcase} respectively occurs at:
+\begin{align}\label{thresholdvalueaneuploid}
+\Delta_a^*=2vp_m+v+2\sqrt{vp_m\left(vp_m+\mu_a+v\right)}.
+\end{align}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% TODO : should this move to methods? we only use this for simulations, right? should be in the Gillespie section?
-\subsection*{Logistic growth}
-We want to have birth and death rates which depends on the population size of wildtype $w$, aneuploidy $a$ and mutant $m$ cells:
-\begin{align*}
-&\lambda_w'=\lambda_w,\quad\mu_w'=\mu_w,\\
-&\lambda_a'=C_1+\left(\lambda_a-\mu_a\right)\left(1-\frac{w+a+m}{K}\right),\quad \mu_a'=C_1,\\
-&\lambda_m'=C_2+\left(\lambda_m-\mu_m\right)\left(1-\frac{w+a+m}{K}\right),\quad \mu_m'=C_2,
-\end{align*}
-where $C_1, C_2>0$ are constants.
+\paragraph*{Density-dependent growth}
+
 We perform stochastic simulations for different values of the carrying capacity $K$ and we plot the results in Figure \ref{SurvPlotNDataLogisticKComplete}.
@@
 
-\begin{figure}[!t]
- \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/SurvPlotNDataLogisticK.pdf}
-\caption{Plot of the probability of evolutionary rescue of a population, consisting of $N$ wildtype cells, as a function of the initial population size of wildtype cells where maximum population size is constrained by the carrying capacity $K$.
-The blue line represent the exact result \eqref{survprobw} while the orange dots represent numerical simulations where the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
-Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_w=1-10^{-1},\lambda_a=1+10^{-4},\lambda_m=1+10^{-1},\mu_w=1,\mu_m=1,u=10^{-2},v=10^{-7}, C_1=C_2=1, K=10^9$.}
-\label{SurvPlotNDataLogisticKComplete}
-\end{figure}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection*{Standing genetic variation}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[!t]
- \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/FractionPlot.pdf}
-\caption{Plot of the fraction of the cases the population is rescue by standing genetic variation as a function of the fraction of initial cells which are aneuploid.
-The red vertical line highlights the the value of $f$ for which half the times the population is rescues by aneuploid cells while the pink line is our approximation \eqref{halfeqstandvar}.
-For this plot we have chosen the following parameters: $N=10^3, \lambda_w=1-10^{-2}, \lambda_a=1-10^{-4},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
-The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
-The value of $0.332$ highlighted in red is the value of the initial fraction of the population which in aneuploid for which half of the cases of evolutionary rescue are due to the initial aneuploid cell population.}
-\label{FractionPlot}
-\end{figure}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 So far we have assumed that the initial population of cells consisted entirely of wildtype cells.
@@
 \begin{equation}\label{ratiorescueexact}
-H=\frac{\left.p_{resc}\right|_{u>0}}{\left.p_{resc}\right|_{u=0}}.
-\end{equation}
-As a result, we obtain from \eqref{probresc} the approximation for the ratio:
+H=\frac{\left.p_{rescue}\right|_{u>0}}{\left.p_{rescue}\right|_{u=0}}.
+\end{equation}
+As a result, we obtain from \eqref{rescue_prob} the approximation for the ratio:
 \begin{equation}\label{ratiorescue}
@@
 We plot \eqref{ratiorescue} in Figure \ref{RatioEvolRescue} for both resistent and susceptible aneuploidy as a function of the proliferation rate of the wildtype cells and in Figure \ref{RatioEvolRescuePopulationSize} as a function of the initial population size of wildtype cells.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[!t]
- \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/RatioEvolRescue.pdf}
-\caption{Plot of the ratio of the probability of evolutionary rescue when aneuploidy can play a role in rescue ($u>0$) to the probability where acquisition of aneuploidy is not possible ($u=0$) as a function of the proliferation rate of the wildtype cells.
-The continuous lines represent the exact result \eqref{ratiorescueexact} while the dashed lines represent the approximation \eqref{ratiorescue}.
-The upper plot show the case when aneuploidy is resistant while the lower plot shows the case when it is susceptible.
-Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_w=1-10^{-1},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
-}
-\label{RatioEvolRescue}
-\end{figure}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[!t]
- \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/RatioEvolRescuePopulationSize.pdf}
-\caption{Plot of the ratio of the probability of evolutionary rescue when aneuploidy can play a role in rescue ($u>0$) to the probability where acquisition of aneuploidy is not possible ($u=0$) as a function of the initial population size of wildtype cells.
-The continuous lines represent the exact result \eqref{ratiorescueexact} while the dashed lines represent the approximation \eqref{ratiorescue}.
-The upper plot show the case when aneuploidy is resistant while the lower plot shows the case when it is susceptible.
-Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_w=0.14, \lambda_a=0.14,\lambda_m=0.14,\mu_w=0.17,\mu_m=0.13, u=10^{-2}, v=10^{-7}$.
-}
-\label{RatioEvolRescuePopulationSize}
-\end{figure}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection*{Rescue time}
-%%%
-\begin{figure}[!t]
- \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/MeanTimeGrowthMutantDirectPlot.pdf}
-\caption{Plot of the mean time until the appearance of a resistance mutation which rescues the population in the case when evolutionary rescue is possible only through mutation but not aneuploidy and mutation.
-The blue line represents the approximation \eqref{meantime1} and the dashed red line represents the first order approximation \eqref{limitapprox}.
-The orange dots represent the numerical simulations while the error bars represent the interval centered at the mean which containing $95\%$ of the simulated values.
-Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_w=1-10^{-1},\lambda_m=1+10^{-1},\mu_w=1,\mu_m=1,u=10^{-2},v=10^{-7}$.}
-\label{MeanTimeGrowthAneuploidyPlot}
-\end{figure}
 We calculate the mean time for the appearance of the first mutant that rescues the cancer cell population.
@@
 %\end{align}
-%Following the steps outlines in \cite{weissman2009rate} we obtain:
+%Following the steps outlines in \citep{weissman2009rate} we obtain:
 %\begin{align}
@@
 This is consistent with the information obtained from Figure \ref{RatioEvolRescuePopulationSize} where aneuploidy improves the probability of evolutionary rescue only  for small and intermediate values of $N$.
-\begin{figure}[!t]
+\begin{figure}[!h]
  \vspace*{1\baselineskip}
@@
 
-\begin{figure}[!t]
+\begin{figure}[!h]
  \vspace*{1\baselineskip}
@@
 
-We propose experiments similar to the ones highlighted in \cite{martin2013probability} in order to test the predictions of our model.
+We propose experiments similar to the ones highlighted in \citep{martin2013probability} in order to test the predictions of our model.
 For example, in order to study the effects of initial population size on the probability of evolutionary rescue we propose to derive a large culture mass from a single cancer cell in permissive conditions and then dilute to a wide range of stating population sizes ($10^7-10^9$).
@@
 
-We observe from equations \eqref{survprobw} and \eqref{probresc} that the probability of evolutionary rescue increases when the initial population size increases, the wildtype population does not decline too quickly, the mutation and aneuploidy rates are high and the probabilities $p_a$ and $p_m$ are elevated.
+We observe from equations \eqref{survprobw} and \eqref{rescue_prob} that the probability of evolutionary rescue increases when the initial population size increases, the wildtype population does not decline too quickly, the mutation and aneuploidy rates are high and the probabilities $p_a$ and $p_m$ are elevated.
 
 The probability of evolutionary rescue is enhanced by aneuploidy for small and intermediate sized tumors (see Figure \ref{RatioEvolRescuePopulationSize}).
-As a result, aneuploidy is unlikely to to contribute to primary tumors overcoming chemotherapy but it can contribute to the evolutionary rescue of secondary tumors whose size might be below the detection threshold of $\sim10^7$  \cite{bozic2013evolutionary}.
+As a result, aneuploidy is unlikely to to contribute to primary tumors overcoming chemotherapy but it can contribute to the evolutionary rescue of secondary tumors whose size might be below the detection threshold of $\sim10^7$  \citep{bozic2013evolutionary}.
 Given the fact that the mean time for small and intermediate tumors to overcome chemotherapy can be of the order of 100 days (see Figure \ref{MeanTimeDeleteriousAneuploidyPlot})  this could explain the reappearance of cancer even after initial remission.
@@
 
+\section*{Figures}
+
+% TODO merge figures 2 and 3 to a single figure here
+%\begin{figure}[h!]
+%\begin{subfigure}{0.5\textwidth}
+%      \centering
+%      \includegraphics[width=\textwidth]{Figures/CombinedSubplot.pdf}      
+%      \label{SurvPlot}
+%\end{subfigure}
+%\begin{subfigure}{0.5\textwidth}
+%      \centering
+%      \includegraphics[width=\textwidth]{Figures/CombinedSubplotLargePop.pdf}      
+%      \label{SurvPlotLargeN}
+%\end{subfigure}
+%\caption{Survival probability for large-effect aneuploidy.}
+%\label{}
+%\end{figure}  
+%%%%%%%%%%%
+\begin{figure}[!h]
+ \vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/CombinedSubplot.pdf}
+\caption{Plot of the probability of evolutionary rescue of a population, consisting initially of $N$ wildtype cells, as a function of the proliferation rate of the wildtype cells for various values of the proliferation rate of the aneuploid cells.
+The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximation \eqref{aneuploidyresistentfirstapprox} for the upper plot and \eqref{AneuploidyDeleteriousApprox} for the lower plot.
+The dots represent numerical simulations where the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
+Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $N=10^4, \lambda_a=1+10^{-2},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
+}
+\label{SurvPlot}
+\end{figure}
+%%%%%%%%%%%
+\begin{figure}[!h]
+ \vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/CombinedSubplotLargePop.pdf}
+\caption{Plot of the probability of evolutionary rescue of a population, consisting initially of $N$ wildtype cells, as a function of the proliferation rate of the wildtype cells for various values of the proliferation rate of the aneuploid cells.
+The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximation \eqref{aneuploidyresistentfirstapprox} for the upper plot and \eqref{AneuploidyDeleteriousApprox} for the lower plot.
+The dots represent numerical simulations where the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
+Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $N=10^8, \lambda_w=0.14, \lambda_a=0.14,\lambda_m=0.14,\mu_m=0.13, u=10^{-2}, v=10^{-7}$.
+}
+\label{SurvPlotLargeN}
+\end{figure}
+%%%%%%%%%%%
+%\begin{figure}[!h]
+% \vspace*{1\baselineskip}
+%\includegraphics[width=1\textwidth]{Figures/P_est_growth.pdf}
+%\caption{Plot of the probability of survival of a population as a function of the proliferation rate of the wildtype cells.
+The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximation \eqref{survprobwapprox1}.
+Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $N=10^4, \lambda_a=1+10^{-2},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
+The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
+The error bars are present but are not visible given the fact that we have used $10^5$ simulations for each combination of parameters.}
+%\label{SurvPlot}
+%\end{figure}
+%%%%%%%%%%%
+\begin{figure}[!h]
+ \vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/SurvPlotNData.pdf}
+\caption{Plot of the probability of survival of a population as a function of the initial population size of wildtype cells.
+The blue line represents the exact solution \eqref{survprobw}, the orange line line represents the approximation \eqref{aneuploidyresistentfirstapprox}, the green line represents the first order correction \eqref{survprobwapproxcorrected} and the red dots represents stochastic simulations.
+For the simulations we have chosen the following parameters: $\lambda_w=0.14, \lambda_a=0.14,\lambda_m=0.14,\mu_w=0.17,\mu_a=0.135,\mu_m=0.13$.
+The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n=100$ is the number of simulations.}
+\label{SurvPlotNData}
+\end{figure}
+%%%%%%%%%%%
+%\begin{figure}[!h]
+% \vspace*{1\baselineskip}
+%\includegraphics[width=1\textwidth]{Figures/P_est.pdf}
+%\caption{Plot of the survival probability of an initial population consisting of $10^{4}$ wildtype cells as a function of $\Delta_a=\lambda_a-\mu_a$ for various values of $\Delta_w=\lambda_a-\mu_a$.
+The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximation \eqref{AneuploidyDeleteriousApprox}.
+The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.}
+%\label{P_est}
+%\end{figure}
+%%%%%%%%%%%
+\begin{figure}[!h]
+ \vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/P_est_divergence.pdf}
+\caption{Plot of the probability of evolutionary rescue, of an initial population, consisting of $N$ wildtype cells, as a function of proliferation rate of the wildtype cells $\Delta_w=\lambda_w-\mu_w$ for various values of the proliferation rate of the aneuploid cells $\Delta_a=\lambda_a-\mu_a$.
+The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximations \eqref{AneuploidyDeleteriousApprox} and \eqref{pestthirdcase}.
+The dots represent numerical simulations where the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
+The value highlighted in grey is the threshold $\Delta_a^*$ from \eqref{thresholdvalueaneuploid} which marks the transition between the regime dictated by \eqref{AneuploidyDeleteriousApprox} to the one dictated by \eqref{pestthirdcase}.
+Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $N=10^4,\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
+}
+\label{P_est}
+\end{figure}
+%%%%%%%%%%%
+\begin{figure}[!h]
+ \vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/P_est_divergenceLargePopulation.pdf}
+\caption{Plot of the probability of evolutionary rescue, of an initial population, consisting of $N$ wildtype cells, as a function of proliferation rate of the wildtype cells $\Delta_w=\lambda_w-\mu_w$ for various values of the proliferation rate of the aneuploid cells $\Delta_a=\lambda_a-\mu_a$.
+The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximations \eqref{AneuploidyDeleteriousApprox}.
+The dots represent numerical simulations where the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
+Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $N=10^8,\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
+}
+\label{P_est_large_N}
+\end{figure}
+%%%%%%%%%%%
+\begin{figure}[!h]
+ \vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/DeleteriousTauLeapPlot.pdf}
+\caption{Plot of the probability of evolutionary rescue of a population, consisting of $N$ wildtype cells, as a function of the initial population size of wildtype cells.
+The blue line represent the exact result \eqref{survprobw} while the orange lines represent the approximation \eqref{pestthirdcase}.
+The green dots represent numerical simulations where the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
+The error bars are present but are not visible given the fact that we have used $n=10^5$ simulations for each combination of parameters.
+Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.}
+\label{DeleteriousPlot}
+\end{figure}
+%%%%%%%%%%%
+\begin{figure}[!h]
+ \vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/SurvPlotNDataLogisticK.pdf}
+\caption{Plot of the probability of evolutionary rescue of a population, consisting of $N$ wildtype cells, as a function of the initial population size of wildtype cells where maximum population size is constrained by the carrying capacity $K$.
+The blue line represent the exact result \eqref{survprobw} while the orange dots represent numerical simulations where the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
+Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_w=1-10^{-1},\lambda_a=1+10^{-4},\lambda_m=1+10^{-1},\mu_w=1,\mu_m=1,u=10^{-2},v=10^{-7}, C_1=C_2=1, K=10^9$.}
+\label{SurvPlotNDataLogisticKComplete}
+\end{figure}
+%%%%%%%%%%%
+\begin{figure}[!h]
+ \vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/RatioEvolRescue.pdf}
+\caption{Plot of the ratio of the probability of evolutionary rescue when aneuploidy can play a role in rescue ($u>0$) to the probability where acquisition of aneuploidy is not possible ($u=0$) as a function of the proliferation rate of the wildtype cells.
+The continuous lines represent the exact result \eqref{ratiorescueexact} while the dashed lines represent the approximation \eqref{ratiorescue}.
+The upper plot show the case when aneuploidy is resistant while the lower plot shows the case when it is susceptible.
+Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_w=1-10^{-1},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
+}
+\label{RatioEvolRescue}
+\end{figure}
+%%%%%%%%%%%
+\begin{figure}[!h]
+ \vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/RatioEvolRescuePopulationSize.pdf}
+\caption{Plot of the ratio of the probability of evolutionary rescue when aneuploidy can play a role in rescue ($u>0$) to the probability where acquisition of aneuploidy is not possible ($u=0$) as a function of the initial population size of wildtype cells.
+The continuous lines represent the exact result \eqref{ratiorescueexact} while the dashed lines represent the approximation \eqref{ratiorescue}.
+The upper plot show the case when aneuploidy is resistant while the lower plot shows the case when it is susceptible.
+Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_w=0.14, \lambda_a=0.14,\lambda_m=0.14,\mu_w=0.17,\mu_m=0.13, u=10^{-2}, v=10^{-7}$.
+}
+\label{RatioEvolRescuePopulationSize}
+\end{figure}
+%%%%%%%%%%%
+\begin{figure}[!h]
+ \vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/MeanTimeGrowthMutantDirectPlot.pdf}
+\caption{Plot of the mean time until the appearance of a resistance mutation which rescues the population in the case when evolutionary rescue is possible only through mutation but not aneuploidy and mutation.
+The blue line represents the approximation \eqref{meantime1} and the dashed red line represents the first order approximation \eqref{limitapprox}.
+The orange dots represent the numerical simulations while the error bars represent the interval centered at the mean which containing $95\%$ of the simulated values.
+Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_w=1-10^{-1},\lambda_m=1+10^{-1},\mu_w=1,\mu_m=1,u=10^{-2},v=10^{-7}$.}
+\label{MeanTimeGrowthAneuploidyPlot}
+\end{figure}
+%%%%%%%%%%%
+\begin{figure}[!h]
+ \vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/FractionPlot.pdf}
+\caption{Plot of the fraction of the cases the population is rescue by standing genetic variation as a function of the fraction of initial cells which are aneuploid.
+The red vertical line highlights the the value of $f$ for which half the times the population is rescues by aneuploid cells while the pink line is our approximation \eqref{halfeqstandvar}.
+For this plot we have chosen the following parameters: $N=10^3, \lambda_w=1-10^{-2}, \lambda_a=1-10^{-4},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
+The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
+The value of $0.332$ highlighted in red is the value of the initial fraction of the population which in aneuploid for which half of the cases of evolutionary rescue are due to the initial aneuploid cell population.}
+\label{FractionPlot}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %\section*{References}
 \nolinenumbers
-%\bibliographystyle{agsm}
-%\bibliographystyle{apalike}
 %\bibliographystyle{unsrtnat}
-\bibliographystyle{unsrt}
+\bibliographystyle{agsm}
 \bibliography{evo2022}
@@
 \beginsupplement % https://support.authorea.com/en-us/article/how-to-create-an-appendix-section-or-supplementary-information-1g25i5a/
-\begin{figure}[!t]
+\begin{figure}[!h]
  \vspace*{1\baselineskip}

### 2023-05-31 EvolutionaryRescue1.tex

@@
 We assume the population of cells is under a strong stress, such as drug therapy, to which the wildtype genotype is susceptible and therefore $\Delta_w<0$, whereas the mutant is resistant to the stress, $\Delta_m>0$
-We analyze two cases: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_w$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}.
+We analyze three scenarios: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_w$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing or "barely growing", that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$.
 Wildtype cells may missegregate to become aneuploids at rate $u$. Both aneuploid and wildtype cells may mutate to become mutants at rate $v$ (\Cref{figureAneuploidy}). 
@@
 
-Evolutionary rescue occurs when mutant cells appear and fixate ($m_t>>1$) in the population before the population becomes extinct ($w_t=a_t=m_t=0$).
-Aneuploidy may contribute to evolutionary rescue by either preventing (when $\Delta_a>0$) or delaying (when $0>\Delta_a>\Delta_w$) the extinction of the population before mutant cells appear and fixate.
-To analyze evolutionary rescue in this model, we use the framework of \emph{multitype branching processes} \citep{rybnikov2021fitness,harris1963theory}. 
-This allows us to find explicit expressions for the \emph{survival probability}: the probability that the population does not become extinct.
-Note that branching processes assume that lineages produced by cells from the initial population divide and die independently of each other, which may be unrealistic, as cells usually compete for resources.
-A more realistic model may include competition for limited resources and spatial structure, which may play an important role in the development of cancer \citep[e.g.,][]{martens2011spatial}.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
      & Name & Value & Units & References \\ \hline
-    $N$ & Tumor size at beginning of treatment& $10^7-10^9$ & cells  & \citet{del2009does} \\ \hline
-    $\lambda_w$ & Wildtype birth rate& 0.14 & 1/days  & \citep{bozic2013evolutionary} \\ \hline
+    $N$ & Initial tumor size & $10^7-10^9$ & cells  & \citet{del2009does} \\ \hline
+    $\lambda_w$ & Wildtype division rate& 0.14 & 1/days  & \citep{bozic2013evolutionary} \\ \hline
     $\mu_w$ & Wildtype death rate& 0.17 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
-    $\lambda_a$  & Aneuploid birth rate$^\ast$ & 0.14 & 1/days  & - \\ \hline
+    $\lambda_a$  & Aneuploid division rate$^\ast$ & 0.14 & 1/days  & - \\ \hline
     $\mu_a$ & Aneuploid death rate$^\ast$ & $0.13-0.17$ & 1/days  & - \\ \hline
-    $\lambda_m$ & Mutant birth rate& 0.14 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
+    $\lambda_m$ & Mutant division rate& 0.14 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
     $\mu_m$ & Mutant death rate& 0.13 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
-    $u$ & Chromosomal instability rate& $10^{-3}-10^{-2}$ & 1$\slash$cell division  & \citet{nowak2004evolutionary,bakker2023predicting} \\ \hline
+    $u$ & Missegregation rate& $10^{-3}-10^{-2}$ & 1$\slash$cell division  & \citet{nowak2004evolutionary,bakker2023predicting} \\ \hline
     $v$ & Mutation rate& $10^{-7}-10^{-9}$ &  1$\slash$gene$\slash$cell division  & \citet{nowak2004evolutionary} \\ \hline
@@
 \paragraph{Density-dependent growth.}
+
+In our analysis we assume that lineages produced by cells from the initial population divide and die independently of each other, which may be unrealistic, as cells usually compete for resources.
+A more realistic model includes competition for limited resources and spatial structure, which may play an important role in the development of cancer \citep[e.g.,][]{martens2011spatial}.
 To simulate birth and death rates that depend on the number of cells in the population, we transform the rates of division and death to the following:
@@
 \subsection*{Survival probability}
+
+To analyze evolutionary rescue in this model, we use the framework of \emph{multitype branching processes} \citep{rybnikov2021fitness,harris1963theory}. 
+This allows us to find explicit expressions for the \emph{survival probability}: the probability that a lineage descended from a single cell does not become extinct.
 
@@
 p_a &= \frac{\lambda_a-\mu_a-v+\sqrt{\left(\lambda_a-\mu_a-v\right)^2+4\lambda_avp_m}}{2\lambda_a}, \\
-p_m &= \left\{
-  \begin{array}{@{}ll@{}}
-  \frac{\lambda_m-\mu_m}{\lambda_m},& \lambda_m \ge \mu_m \\
-   0, & \lambda_m \le \mu_m . %TODO in the appendix this is 1, not 0
-  \end{array}\right .  
+p_m &= \frac{\lambda_m-\mu_m}{\lambda_m} .
 \end{aligned} 
 \end{equation}
-See Appendix \ref{AppendixSurvLin} for the derivation of $p_m$. %TODO why do we need Appendix 1 if we can just find p_m as solution to quadratic equation?
 Note that the equation for $p_w$ depends on both $p_a$ and $p_m$, and the equation for $p_a$ depends on $p_m$.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{Scenario 1: Aneuploid cells are drug resistant}
+\subsubsection*{Scenario 1: Aneuploid cells are partially resistant}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\paragraph{Second-order approximation.}
+\paragraph{Second-order approximation.} % TODO consider moving to Appendix
 To improve our approximation, we can consider the second term of the Taylor series expansion,
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{Scenario 2: Aneuploid cells are drug tolerant.}
-
-We now assume that aneuploidy provides tolerance to the drug, that is, the number of aneuploid cells significantly declines over time, but at a lower rate than the number of wildtype cells, $\lambda_w - \mu_w < \lambda_a - \mu_a < 0$.
-We also assume that both declines are significant, $\left(\lambda_a-\mu_a-v\right)^2 > 4\lambda_a v p_m$.
+\subsubsection*{Scenario 2: Aneuploid cells are tolerant.}
+
+We now assume that aneuploidy provides tolerance to drug therapy, that is, the number of aneuploid cells significantly declines over time, but at a lower rate than the number of wildtype cells, $\lambda_w - \mu_w < \lambda_a - \mu_a < 0$.
+We also assume that the decline are significant, $\left(\lambda_a-\mu_a-v\right)^2 > 4\lambda_a v p_m$.
 We rewrite \cref{survival_prob} as
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{Scenario 3: Aneuploid cells are non-growing.}
+\subsubsection*{Scenario 3: Aneuploid cells are non-growing} % TODO consider this name
 
@@
 we obtain the approximation
-\begin{align*}
+\begin{equation}
+\begin{aligned}
 p_a&\approx\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}\left[1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{8\lambda_avp_m}\right]}{2\lambda_a}\\
@@
 &=\frac{1}{2\lambda_a}\left(\lambda_a-\mu_a-v+2\sqrt{\lambda_avp_m}\right).
-\end{align*}
+\end{aligned}
+\end{equation}
 From \Cref{survprobwinitial}, % TODO I dont see how the referred equation plays a part here
 the survival probability of a population starting from one wildtype individual is
-\begin{align}\label{scenario3}
+\begin{equation}\label{scenario3}
+\begin{aligned}
 p_w&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[v\frac{\lambda_m-\mu_m}{\lambda_m}+\frac{u}{2\lambda_a}\left(\lambda_a-\mu_a-v+2\sqrt{\lambda_avp_m}\right)\right]\\
@@
 %&=-\frac{1}{\Delta_w-u-v}\left[v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right],
-\end{align}
+\end{aligned}
+\end{equation}
 
@@
 \subsection*{Evolutionary rescue probability}
+
+Evolutionary rescue occurs when mutant cells appear and fixate ($m_t>>1$) in the population before the population becomes extinct ($w_t=a_t=m_t=0$).
+Aneuploidy may contribute to evolutionary rescue by either preventing (when $\Delta_a>0$) or delaying (when $0>\Delta_a>\Delta_w$) the extinction of the population before mutant cells appear and fixate.
 
@@
 
-Now, we can apply the approximations for the survival probability $p_w$ from \Cref{survprobwapprox1,survprobwinitial,scenario3} and substituting $\Delta_k=\lambda_k-\mu_k$
-
-The rescue probability is given by
+Now, we apply the approximations for the survival probability $p_w$ from \Cref{survprobwapprox1,survprobwinitial,scenario3}.
+Therefore, substituting $\Delta_k=\lambda_k-\mu_k$ , the rescue probability is given by
 \begin{equation}\label{rescue_prob}
@@
    \frac{1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right]}{1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\right]},\quad\text{if }\Delta_a>0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2.
+% TODO this term can be simplified/approximated
   \end{array}\right.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 When $Nu\gg1$ the aneuploid population can be assumed to be deterministic and approximated by the solution to the system of ODEs:
@@
 This is consistent with the information obtained from Figure \ref{RatioEvolRescuePopulationSize} where aneuploidy improves the probability of evolutionary rescue only  for small and intermediate values of $N$.
-\begin{figure}[!h]
- \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/MeanTimeDeleteriousPlot.pdf}
-\caption{Plot of the mean time until the appearance of a resistance mutation which rescues the population in the case when evolutionary rescue is possible 
- through mutation and aneuploidy.
-Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_w=1-10^{-1},\lambda_a=1-10^{-2},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1,u=10^{-2},v=10^{-7}$.
-The blue line represents the approximation \eqref{meantimet2}, the dashed red line represents the second order approximation \eqref{limitapprox2} and the green line is first order approximation \eqref{limitapprox3}.
-The orange dots represent the numerical simulations while the error bars represent the interval centered at the mean which containing $95\%$ of the simulated values.}
-\label{MeanTimeDeleteriousAneuploidyPlot}
-\end{figure}
-
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection*{Contribution of aneuploidy to mean evolutionary rescue time}
@@
 
-\begin{figure}[!h]
- \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/MeanTimeRatioInitialPopulationSize.pdf}
-\caption{Plot of the ratio of the mean time to evolutionary rescue when aneuploidy can play a role in rescue ($u>0$) to the mean time where acquisition of aneuploidy is not possible ($u=0$) as a function of the initial population size of wildtype cells.
-The continuous lines represent the approximation \eqref{eqMeanTimeRatioInitialPopulationSize}.
-Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_w=1-10^{-1},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.}
-\label{MeanTimeRatioInitialPopulationSize}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\newpage
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section*{Figures}
+%\section*{References}
+\newpage
+\nolinenumbers
+%\bibliographystyle{unsrtnat}
+\bibliographystyle{agsm}
+\bibliography{evo2022}
+\newpage
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 %%%%%%%%%%%
-\begin{figure}[!h]
+\begin{figure}[p]
+% small population size
+% A: partial resistance
+% B: tolerance
  \vspace*{1\baselineskip}
@@
 %%%%%%%%%%%
-\begin{figure}[!h]
+\begin{figure}[p]
+% large population size
+% A: partial resistance
+% B: tolerance
  \vspace*{1\baselineskip}
@@
 %%%%%%%%%%%
-%\begin{figure}[!h]
-% \vspace*{1\baselineskip}
-%\includegraphics[width=1\textwidth]{Figures/P_est_growth.pdf}
-%\caption{Plot of the probability of survival of a population as a function of the proliferation rate of the wildtype cells.
-The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximation \eqref{survprobwapprox1}.
-Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $N=10^4, \lambda_a=1+10^{-2},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
-The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
-The error bars are present but are not visible given the fact that we have used $10^5$ simulations for each combination of parameters.}
-%\label{SurvPlot}
-%\end{figure}
-%%%%%%%%%%%
-\begin{figure}[!h]
+\begin{figure}[p]
+% partial resistance, compare 1st and 2nd order approximation and exact
  \vspace*{1\baselineskip}
@@
 %%%%%%%%%%%
-%\begin{figure}[!h]
-% \vspace*{1\baselineskip}
-%\includegraphics[width=1\textwidth]{Figures/P_est.pdf}
-%\caption{Plot of the survival probability of an initial population consisting of $10^{4}$ wildtype cells as a function of $\Delta_a=\lambda_a-\mu_a$ for various values of $\Delta_w=\lambda_a-\mu_a$.
-The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximation \eqref{AneuploidyDeleteriousApprox}.
-The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.}
-%\label{P_est}
-%\end{figure}
-%%%%%%%%%%%
-\begin{figure}[!h]
+\begin{figure}[p]
+% small population size
+% switch from tolerance to non-growing
  \vspace*{1\baselineskip}
@@
 %%%%%%%%%%%
-\begin{figure}[!h]
+\begin{figure}[p]
+% large population size
+% switch from tolerance to non-growing
  \vspace*{1\baselineskip}
@@
 %%%%%%%%%%%
-\begin{figure}[!h]
+\begin{figure}[p]
+% non-growing slightly declining, compare 1st order approximation to exact and simulation
  \vspace*{1\baselineskip}
@@
 %%%%%%%%%%%
-\begin{figure}[!h]
+\begin{figure}[p]
+% density-dependent simulations, barely growing (or maybe partial resistance), with exact formula
  \vspace*{1\baselineskip}
@@
 %%%%%%%%%%%
-\begin{figure}[!h]
+\begin{figure}[p]
+% A partial resistance
+% B tolerance
+% function of wildtype growth rate
  \vspace*{1\baselineskip}
@@
 %%%%%%%%%%%
-\begin{figure}[!h]
+\begin{figure}[p]
+% A partial resistance
+% B tolerance
+% function of population size
  \vspace*{1\baselineskip}
@@
 %%%%%%%%%%%
-\begin{figure}[!h]
+\begin{figure}[p]
+% partial resistance - aneuploidy rescues the pop
  \vspace*{1\baselineskip}
@@
 %%%%%%%%%%%
-\begin{figure}[!h]
+\begin{figure}[p]
+% slightly decreasing (need to check)
  \vspace*{1\baselineskip}
@@
 \end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\section*{References}
-\nolinenumbers
-%\bibliographystyle{unsrtnat}
-\bibliographystyle{agsm}
-\bibliography{evo2022}
-\pagebreak
-
+%%%%%%%%%%%
+\begin{figure}[p]
+% tolerance
+ \vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/MeanTimeDeleteriousPlot.pdf}
+\caption{Plot of the mean time until the appearance of a resistance mutation which rescues the population in the case when evolutionary rescue is possible 
+ through mutation and aneuploidy.
+Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_w=1-10^{-1},\lambda_a=1-10^{-2},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1,u=10^{-2},v=10^{-7}$.
+The blue line represents the approximation \eqref{meantimet2}, the dashed red line represents the second order approximation \eqref{limitapprox2} and the green line is first order approximation \eqref{limitapprox3}.
+The orange dots represent the numerical simulations while the error bars represent the interval centered at the mean which containing $95\%$ of the simulated values.}
+\label{MeanTimeDeleteriousAneuploidyPlot}
+\end{figure}
+%%%%%%%%%%%
+\begin{figure}[p]
+% tolerance... maybe third is slightly decreasing
+ \vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/MeanTimeRatioInitialPopulationSize.pdf}
+\caption{Plot of the ratio of the mean time to evolutionary rescue when aneuploidy can play a role in rescue ($u>0$) to the mean time where acquisition of aneuploidy is not possible ($u=0$) as a function of the initial population size of wildtype cells.
+The continuous lines represent the approximation \eqref{eqMeanTimeRatioInitialPopulationSize}.
+Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_w=1-10^{-1},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.}
+\label{MeanTimeRatioInitialPopulationSize}
+\end{figure}
+%%%%%%%%%%%
+\begin{figure}[p]
+% diffusion approx instead of branching process
+% tolerance
+ \vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/P_est_diffusion.pdf}
+\caption{Plot of the survival probability of an initial population consisting of $w_0=10^{4}$ wildtype cells as a function of $\Delta_a=\lambda_a-\mu_a$ for various values of $\Delta_w=\lambda_a-\mu_a$.
+The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the Feller diffusion approximation \eqref{FellerApprox}.
+The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/w_0}$ where $p$ is the mean probability of evolutionary rescue.}
+\label{P_est_diffusion}
+\end{figure}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Survival probability of a mutant lineage}\label{AppendixSurvLin}
-The infinitesimal transition probabilities for the simple birth and death process:
-\begin{align*}
-p_{i+j,i}\left(\Delta t\right)=\left\{
-  \begin{array}{@{}ll@{}}
-  \mu i\Delta t+o\left(\Delta t\right),\quad j=-1 \\
-  \lambda i\Delta t+o\left(\Delta t\right),\quad j=1 \\
-    1-\left(\lambda+\mu\right) i\Delta t+o\left(\Delta t\right),\quad j=0 \\
-   o\left(\Delta t\right),\quad j\neq-1,0,1.
-  \end{array}\right.
-\end{align*}
-The forward Kolmogorov differential equations are:
-\begin{subequations}
-\label{KolmeqSurv}
-\begin{flalign}
-\frac{dp_i\left(t\right)}{dt}&=\lambda\left(i-1\right)p_{i-1}\left(t\right)+\mu\left(i+1\right)p_{i+1}\left(t\right)-\left(\lambda+\mu\right)ip_i\left(t\right),\\
-\frac{dp_0\left(t\right)}{dt}&=\mu p_1\left(t\right),
-\end{flalign}
-\end{subequations}
-for $i=1,2,\dots$ with initial conditions $p_i\left(0\right)=\delta_{iN}$.
-
-The probability generating function is defined as:
-\begin{equation*}
-\mathcal{P}\left(z,t\right)=\sum_{i=0}^\infty p_i\left(t\right)z^i,
-\end{equation*}
-which we obtain by multiplying \eqref{KolmeqSurv} by $z^i$ and summing over $i$:
-\begin{align*}
-\mathcal{P}\left(z,t\right)=\left\{
-  \begin{array}{@{}ll@{}}
-  \left(\frac{\e^{t\left(\mu-\lambda\right)}\left(\lambda z-\mu\right)-\mu\left(z-1\right)}{\e^{t\left(\mu-\lambda\right)}\left(\lambda z-\mu\right)-\lambda\left(z-1\right)}\right)^N,\quad \text{if }\lambda\neq\mu, \\
-  \left(\frac{1-\left(\lambda t-1\right)\left(z-1\right)}{1-\lambda t\left(z-1\right)}\right)^N,\quad \text{if }\lambda=\mu.
-  \end{array}\right.
-\end{align*}
-The probability $p_i$ can be obtained from the probability generating function as:
-\begin{equation*}
-p_i\left(t\right)=\left.\frac{1}{i!}\frac{\partial^i\mathcal{P}}{\partial z^i}\right|_{z=0},
-\end{equation*}
-and the probability of extinction is given by:
-\begin{align*}
-p_0\left(t\right)=\left\{
-  \begin{array}{@{}ll@{}}
-  \left(\frac{\mu-\mu\e^{\left(\mu-\lambda\right)t}}{\lambda-\mu\e^{\left(\mu-\lambda\right)t}}\right)^N,\quad \text{if }\lambda\neq\mu, \\
-  \left(\frac{\lambda t}{1+\lambda t}\right)^N,\quad \text{if }\lambda=\mu.
-  \end{array}\right.
-\end{align*}
-When $t\rightarrow\infty$ the extinction probability has the following expression:
-\begin{align*}
-p_0\left(\infty\right)=\lim_{t\rightarrow\infty}p_0\left(t\right)=\left\{
-  \begin{array}{@{}ll@{}}
-  1,\quad \text{if }\lambda\leq\mu, \\
-  \left(\frac{\mu}{\lambda}\right)^N,\quad \text{if }\lambda>\mu.
-  \end{array}\right.
-\end{align*}
-Setting $N=1$ and using the fact that the probability of survival is $p=1-p_0$ we obtain:
-\begin{equation*}
-p =  \left\{\begin{array}{@{}ll@{}}
-  1,\quad \text{if }\lambda\leq\mu, \\
-  \frac{\lambda-\mu}{\lambda},\quad \text{if }\lambda>\mu.
-  \end{array}\right.
-\end{equation*}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\newpage
-\section*{Supplementary Figures}
-\beginsupplement % https://support.authorea.com/en-us/article/how-to-create-an-appendix-section-or-supplementary-information-1g25i5a/
-\begin{figure}[!h]
- \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/P_est_diffusion.pdf}
-\caption{Plot of the survival probability of an initial population consisting of $w_0=10^{4}$ wildtype cells as a function of $\Delta_a=\lambda_a-\mu_a$ for various values of $\Delta_w=\lambda_a-\mu_a$.
-The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the Feller diffusion approximation \eqref{FellerApprox}.
-The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/w_0}$ where $p$ is the mean probability of evolutionary rescue.}
-\label{P_est_diffusion}
-\end{figure}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2023-05-31 EvolutionaryRescue1.tex
reorganize figures and rescue results -> now going to merge figures

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Evolutionary simulation} 
+\subsection*{Stochastic simulations} 
 Simulations are performed using a \emph{Gillespie algorithm} \citep{gillespie1976general,gillespie1977exact} implemented in Python \citep{python}.
@@
 
-Evolutionary rescue occurs when mutant cells appear and fixate ($m_t>>1$) in the population before the population becomes extinct ($w_t=a_t=m_t=0$).
+In our model, \emph{evolutionary rescue} occurs when resistant cells appear and fixate ($m_t \gg 1$) in the population before the population becomes extinct ($w_t=a_t=m_t=0$).
 Aneuploidy may contribute to evolutionary rescue by either preventing (when $\Delta_a>0$) or delaying (when $0>\Delta_a>\Delta_w$) the extinction of the population before mutant cells appear and fixate.
 
-To estimate the rescue probability $p_{rescue}$, we assume independence between clonal lineages starting from an initial population of $N$ wildtype cells.
+To estimate the rescue probability $p_{rescue}$, we assume independence between clonal lineages starting from an initial population of $N$ wildtype cells (we check the effect of density-dependent growth on our results below).
 Thus, the rescue probability is given by 
@@
 
-Now, we apply the approximations for the survival probability $p_w$ from \Cref{survprobwapprox1,survprobwinitial,scenario3}.
-Therefore, substituting $\Delta_k=\lambda_k-\mu_k$ , the rescue probability is given by
+Applying the approximations for the survival probability $p_w$ from \cref{survprobwapprox1,survprobwinitial,scenario3} and substituting $\Delta_k=\lambda_k-\mu_k$, we find that the rescue probability can be approximated by
 \begin{equation}\label{rescue_prob}
-p_{rescue} = \left\{
-  \begin{array}{@{}ll@{}}
-  1-\exp\left[\frac{N}{\Delta_w-u-v}\left(v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right)\right],\quad\text{if }4\lambda_avp_m>\left(\Delta_a-v\right)^2,\\
-   1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right],\quad\text{if }\Delta_a<0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2,\\
-   1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right],\quad\text{if }\Delta_a>0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2.
-  \end{array}\right.
-\end{equation}
-
+\begin{aligned}
+&p_{rescue} \approx \\
+  &\begin{cases}
+  1-\exp\left[\frac{N}{\Delta_w-u-v}\left(v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right)\right] ,&
+  4\lambda_avp_m>\left(\Delta_a-v\right)^2 ,\\
+   1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right] ,&
+   \Delta_a<0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2 ,\\
+   1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right] ,&
+   \Delta_a>0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2 .
+  \end{cases}
+\end{aligned}
+\end{equation}
+
+% TODO do we need this?
 %Note that if instead \cref{survprobwapprox1} we apply \cref{survprobwapproxcorrected}, we have
@@
 %\end{align}
+%However, both approximations seem to work very well (\Cref{compare_1st_2nd_order_approx}).
+
+We validate these approximations by comparing them to results of stochastic evolutionary simulations.
+We find that the approximations work very well (\Cref{rescue_prob_wt_growth_small_N,rescue_prob_wt_growth_large_N,rescue_prob_an_growth_small_N,rescue_prob_an_growth_large_N,rescue_prob_N_nongrow,rescue_prob_N_logistic}).
 
@@
 }
-\label{SurvPlot}
+\label{rescue_prob_wt_growth_small_N}
 \end{figure}
@@
 }
-\label{SurvPlotLargeN}
-\end{figure}
-%%%%%%%%%%%
-\begin{figure}[p]
-% partial resistance, compare 1st and 2nd order approximation and exact
- \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/SurvPlotNData.pdf}
-\caption{Plot of the probability of survival of a population as a function of the initial population size of wildtype cells.
-The blue line represents the exact solution \eqref{survprobw}, the orange line line represents the approximation \eqref{aneuploidyresistentfirstapprox}, the green line represents the first order correction \eqref{survprobwapproxcorrected} and the red dots represents stochastic simulations.
-For the simulations we have chosen the following parameters: $\lambda_w=0.14, \lambda_a=0.14,\lambda_m=0.14,\mu_w=0.17,\mu_a=0.135,\mu_m=0.13$.
-The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n=100$ is the number of simulations.}
-\label{SurvPlotNData}
+\label{rescue_prob_wt_growth_large_N}
 \end{figure}
@@
 }
-\label{P_est}
+\label{rescue_prob_an_growth_small_N}
 \end{figure}
@@
 }
-\label{P_est_large_N}
+\label{rescue_prob_an_growth_large_N}
 \end{figure}
@@
 Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.}
-\label{DeleteriousPlot}
+\label{rescue_prob_N_nongrow}
 \end{figure}
@@
 Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_w=1-10^{-1},\lambda_a=1+10^{-4},\lambda_m=1+10^{-1},\mu_w=1,\mu_m=1,u=10^{-2},v=10^{-7}, C_1=C_2=1, K=10^9$.}
-\label{SurvPlotNDataLogisticKComplete}
+\label{rescue_prob_N_logistic}
 \end{figure}
@@
 %%%%%%%%%%%
-\begin{figure}[p]
-% diffusion approx instead of branching process
-% tolerance
- \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/P_est_diffusion.pdf}
-\caption{Plot of the survival probability of an initial population consisting of $w_0=10^{4}$ wildtype cells as a function of $\Delta_a=\lambda_a-\mu_a$ for various values of $\Delta_w=\lambda_a-\mu_a$.
+%\begin{figure}[p]
+%% diffusion approx instead of branching process
+%% tolerance
+% \vspace*{1\baselineskip}
+%\includegraphics[width=1\textwidth]{Figures/P_est_diffusion.pdf}
+%\caption{Plot of the survival probability of an initial population consisting of $w_0=10^{4}$ wildtype cells as a function of $\Delta_a=\lambda_a-\mu_a$ for various values of $\Delta_w=\lambda_a-\mu_a$.
 The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the Feller diffusion approximation \eqref{FellerApprox}.
 The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/w_0}$ where $p$ is the mean probability of evolutionary rescue.}
-\label{P_est_diffusion}
-\end{figure}
+%\label{P_est_diffusion}
+%\end{figure}
+%%%%%%%%%%%
+%\begin{figure}[p]
+% partial resistance, compare 1st and 2nd order approximation and exact
+%\vspace*{1\baselineskip}
+%\includegraphics[width=1\textwidth]{Figures/SurvPlotNData.pdf}
+%\caption{Plot of the probability of survival of a population as a function of the initial population size of wildtype cells.
+The blue line represents the exact solution \eqref{survprobw}, the orange line line represents the approximation \eqref{aneuploidyresistentfirstapprox}, the green line represents the first order correction \eqref{survprobwapproxcorrected} and the red dots represents stochastic simulations.
+For the simulations we have chosen the following parameters: $\lambda_w=0.14, \lambda_a=0.14,\lambda_m=0.14,\mu_w=0.17,\mu_a=0.135,\mu_m=0.13$.
+The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n=100$ is the number of simulations.}
+%\label{compare_1st_2nd_order_approx}
+%\end{figure}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %\appendix
-\begin{appendices}
-\renewcommand{\theequation}{\thesection\arabic{equation}}
-\counterwithin*{equation}{section}
+%\begin{appendices}
+%\renewcommand{\theequation}{\thesection\arabic{equation}}
+%\counterwithin*{equation}{section}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Diffusion approximation}\label{AppendixDiffApprox}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-An alternative method to obtain the probability of evolutionary rescue is to utilize a Feller diffusion approximation which is governed by two parameters: the growth rate $r$ and the reproductive variance $\sigma$.
+%\section*{Diffusion approximation}\label{AppendixDiffApprox}
+%
+%An alternative method to obtain the probability of evolutionary rescue is to utilize a Feller diffusion approximation which is governed by two parameters: the growth rate $r$ and the reproductive variance $\sigma$.
 The two parameters are obtained from the underlying demographic process as the infinitesimal relative change in mean and variance of $n_t$ over and an infinitesimally small time interval $\Delta t$:
-\begin{align*}
-r&=\lim_{\Delta t\rightarrow0}\frac{\mathbb{E}\left(\Delta n_t|n_t\right)}{\Delta_t n_t},\\
-\sigma&=\lim_{\Delta t\rightarrow0}\frac{\mathbb{V}ar\left(\Delta n_t|n_t\right)}{\Delta_t n_t}.
-\end{align*}
-The rate at which mutants are generated directly from the wildtype is: 
-\begin{align}
-\theta_1=v\bar{\pi}_f\frac{N}{|r_w|},
-\end{align}
-where
-\begin{align}
-\bar{\pi}_f=\int_{0}^\infty\int_{0}^\infty \left(1-\e^{-\frac{2r}{\sigma}}\right) f_r\left(r,\sigma\right)\d r\d\sigma.
-\end{align}
-Letting $f_r\left(r,\sigma\right)=\delta\left(r-r_m\right)\delta\left(\sigma-\sigma_m\right)$ then:
-\begin{align}
-\bar{\pi}_f=1-\e^{-\frac{2r_m}{\sigma_m}},
-\end{align}
-and, as a result, we have:
-\begin{align}
-\theta_1=v\left(1-\e^{-\frac{2r_m}{\sigma_m}}\right)\frac{N}{|\Delta_w|}.
-\end{align}
-The rate at which mutants are generated indirectly from the wildtype through aneuploidy is: 
-\begin{align*}
-\theta_2&=\frac{uN}{|r_w|}\int_{0}^\infty\int_{0}^\infty \left(1-\pi_f\left(r,q\right)\right) f_r\left(r,\sigma\right) p_1^*\left(r,\sigma\right)\d r\d\sigma\\
-&=\frac{uN}{|\Delta_w|}p_1^*\left(r_a,\sigma_a\right)\\
-&=\frac{uN}{|\Delta_w|}\left[1-\exp\left(-\frac{|r_a|}{\sigma_a^2}\left(\sqrt{1+\frac{2\sigma_a^2}{r_a^2}u^*}-1\right)\right)\right]\\
-&=\frac{uN}{|\Delta_w|}\left[1-\exp\left(-\frac{|\Delta_a|}{(\lambda_a+\mu_a)^2}\left(\sqrt{1+\frac{2(\lambda_a+\mu_a)^2}{\Delta_a^2}u^*}-1\right)\right)\right],
-\end{align*}
-where
-\begin{equation}
-u^*=v\left(1-\e^{-\frac{2r_m}{\sigma_m}}\right).
-\end{equation}
-The number of rescue mutations has a Poisson distribution with rate $\theta_1+\theta_2$. As a result, the probability of evolutionary rescue is given by:
-\begin{equation}\label{FellerApprox}
-p_{rescue}=1-\e^{-\left(\theta_1+\theta_2\right)},
-\end{equation}
-which we plot in Figure \ref{P_est_diffusion} as a function of $\Delta_w$.
-\end{appendices}
+%\begin{align*}
+%r&=\lim_{\Delta t\rightarrow0}\frac{\mathbb{E}\left(\Delta n_t|n_t\right)}{\Delta_t n_t},\\
+%\sigma&=\lim_{\Delta t\rightarrow0}\frac{\mathbb{V}ar\left(\Delta n_t|n_t\right)}{\Delta_t n_t}.
+%\end{align*}
+%The rate at which mutants are generated directly from the wildtype is: 
+%\begin{align}
+%\theta_1=v\bar{\pi}_f\frac{N}{|r_w|},
+%\end{align}
+%where
+%\begin{align}
+%\bar{\pi}_f=\int_{0}^\infty\int_{0}^\infty \left(1-\e^{-\frac{2r}{\sigma}}\right) f_r\left(r,\sigma\right)\d r\d\sigma.
+%\end{align}
+%Letting $f_r\left(r,\sigma\right)=\delta\left(r-r_m\right)\delta\left(\sigma-\sigma_m\right)$ then:
+%\begin{align}
+%\bar{\pi}_f=1-\e^{-\frac{2r_m}{\sigma_m}},
+%\end{align}
+%and, as a result, we have:
+%\begin{align}
+%\theta_1=v\left(1-\e^{-\frac{2r_m}{\sigma_m}}\right)\frac{N}{|\Delta_w|}.
+%\end{align}
+%The rate at which mutants are generated indirectly from the wildtype through aneuploidy is: 
+%\begin{align*}
+%\theta_2&=\frac{uN}{|r_w|}\int_{0}^\infty\int_{0}^\infty \left(1-\pi_f\left(r,q\right)\right) f_r\left(r,\sigma\right) p_1^*\left(r,\sigma\right)\d r\d\sigma\\
+%&=\frac{uN}{|\Delta_w|}p_1^*\left(r_a,\sigma_a\right)\\
+%&=\frac{uN}{|\Delta_w|}\left[1-\exp\left(-\frac{|r_a|}{\sigma_a^2}\left(\sqrt{1+\frac{2\sigma_a^2}{r_a^2}u^*}-1\right)\right)\right]\\
+%&=\frac{uN}{|\Delta_w|}\left[1-\exp\left(-\frac{|\Delta_a|}{(\lambda_a+\mu_a)^2}\left(\sqrt{1+\frac{2(\lambda_a+\mu_a)^2}{\Delta_a^2}u^*}-1\right)\right)\right],
+%\end{align*}
+%where
+%\begin{equation}
+%u^*=v\left(1-\e^{-\frac{2r_m}{\sigma_m}}\right).
+%\end{equation}
+%The number of rescue mutations has a Poisson distribution with rate $\theta_1+\theta_2$. As a result, the probability of evolutionary rescue is given by:
+%\begin{equation}\label{FellerApprox}
+%p_{rescue}=1-\e^{-\left(\theta_1+\theta_2\right)},
+%\end{equation}
+%which we plot in Figure \ref{P_est_diffusion} as a function of $\Delta_w$.
+
+%\end{appendices}
 

### 2023-05-31 EvolutionaryRescue1.tex
Update

--- 2023-05-30 EvolutionaryRescue1.tex
@@
 \usepackage{subcaption}
-%\usepackage{xurl}
+\usepackage{xurl}
 
@@
 
-% NatBib
-%\usepackage[super,comma]{natbib}
-%\usepackage[numbers,super,compress]{natbib}
-\usepackage[round,colon,authoryear]{natbib}
-
 % Title page
@@
 \author[1,*]{Yoav Ram}
-\affil[1]{School of Zoology, Faculty of Life Sciences, Tel Aviv University, Tel Aviv, Israel}
-\affil[2]{Department of Human Molecular Genetics and Biochemistry, Faculty of Medicine, Tel Aviv University, Tel Aviv, Israel}
-\affil[3]{Department of Physics, Emory University, Atlanta, GA}
+\affil[1]{School of Zoology, Faculty of Life Sciences, Tel Aviv University}
+\affil[2]{Department of Human Molecular Genetics and Biochemistry, Faculty of Medicine, Tel Aviv University}
+\affil[3]{Department of Physics, Emory University}
 \affil[*]{Corresponding author: yoav@yoavram.com}
@@
 
-\paragraph{Aneuploidy in cancer.} Chromosomal instability (CIN) is the mitotic process in which cells suffer from chromosome mis-segregation that leads to aneuploidy, where cells are characterized by structural changes of the chromosomes and copy number alterations \citep{schukken2018cin}.
+\paragraph{Aneuploidy in cancer.} Chromosomal instability (CIN) is the mitotic process in which cells suffer from chromosome mis-segregation that leads to aneuploidy, where cells are characterized by structural changes of the chromosomes and copy number alterations \cite{schukken2018cin}.
 Interestingly, aberrations in chromosome copy number have been shown to allow cancer cells to survive under stressful conditions such as drug therapy.
-Indeed, cancer cells are often likely to be aneuploid, and aneuploidy is associated with poor patient outcomes \citep{ben2020context}.
+Indeed, cancer cells are often likely to be aneuploid, and aneuploidy is associated with poor patient outcomes \cite{ben2020context}.
 
@@
 
-The role of chromosomal instability (CIN) in the emergence of cancer has been studied extensively in the past decades \citep{michor2005can,christine2018understanding,nowak2002role,pavelka2010dr,komarova2003mutation,zhu2018cellular}.
+The role of chromosomal instability (CIN) in the emergence of cancer has been studied extensively in the past decades \cite{michor2005can,christine2018understanding,nowak2002role,pavelka2010dr,komarova2003mutation,zhu2018cellular}.
 One hypothesis is that CIN facilitates tumor genesis by accelerating the removal of tumor suppression genes (TSG) and subsequent appearance of cancer.
 The deletion of tumor suppression genes can happen in two ways: two point mutations deleting both alleles of the TSG (assuming a diploid genotype), or one point mutation and one chromosomal loss event.
-Initial theoretical studies have shown that aneuploidy can have a significant role in the deletion of the the tumor suppressing genes when compared to two consecutive point mutations \citep{nowak2002role,komarova2003mutation,michor2005can,komarova2008selective}.
+Initial theoretical studies have shown that aneuploidy can have a significant role in the deletion of the the tumor suppressing genes when compared to two consecutive point mutations \cite{nowak2002role,komarova2003mutation,michor2005can,komarova2008selective}.
+% TODO mention if studies are theoretical , experimental, or clinical OK
 However, when taking into account that the appearance of aneuploidy requires a mutation to trigger CIN, the probability that CIN precedes tumor genesis is highly unlikely.
@@
 Examples of such environmental changes include climate change, invasive species or the onset of drug therapies.
-Adaptation is a race against time as the population size decreases in the new environment~\citep{tanaka2022surviving}.
-\emph{Evolutionary rescue} is the process where the population acquires a trait that increases fitness in the new environment such that extinction is averted.
-It is mathematically equivalent to the problem of crossing of fitness valley \citep{weissman2009rate,weissman2010rate}.
-There are three potential ways for a population to survive environmental change: migration to a new habitat similar to the one before the onset of environmental change \citep{cobbold2020should}; adaptation by phenotypic plasticity without genetic modification \citep{carja2019evolutionary,carja2017evolutionary,levien2021non}; and adaptation through genetic modifications, e.g., mutation \citep{uecker2014evolutionary,uecker2016role,uecker2011fixation}.
-
-Models of evolutionary rescue usually assume that the fitness of the wildtype and mutant are homogeneous in time.
-An exception was given by \citet{marrec2020adapt}, who modeled the fitness of the wildtype and mutant as time dependent.
-Additionally, \citet{uecker2011fixation} investigated the probability of fixation of a beneficial mutation in a variable environment with arbitrary time-dependent selection coefficient and population size.
+Adaptation is a race against time as the population size decreases in the new environment~\cite{tanaka2022surviving}.
+{\em Evolutionary rescue} is the process where the population acquires a trait that increases fitness in the new environment such that extinction is averted.
+It is mathematically equivalent to the problem of crossing of fitness valley \cite{weissman2009rate,weissman2010rate}.
+There are three potential ways for a population to survive environmental change: migration to a new habitat similar to the one before the onset of environmental change \cite{cobbold2020should}; adaptation by phenotypic plasticity without genetic modification \cite{carja2019evolutionary,carja2017evolutionary,levien2021non}; and adaptation through genetic modifications, e.g., mutation \cite{uecker2014evolutionary,uecker2016role,uecker2011fixation}.
+
+Models of evolutionary rescue usually assume that the fitness of the wildtype and mutant are independent of time.
+An exception is \cite{marrec2020adapt}, where the fitness of the wildtype and mutant are time dependent.
+Additionally, \cite{uecker2011fixation} investigated the probability of fixation of a beneficial mutation in a variable environment with arbitrary time-dependent selection coefficent and population size.
 Most models focus on the probability that at least one mutation rescues the population.
-How multiple mutations contribute to the survival of the population is less explored, but \citet{wilson2017soft} have shown that evolutionary rescue is significantly enhanced by soft selective sweeps when multiple mutations contribute.
-Evolutionary rescue that requires two successive mutations has been investigated using diffusion approximation by \citet{martin2013probability}.
+How multiple mutations contribute to the survival of the population is less explored.
+One exception is \cite{wilson2017soft} which showed that evolutionary rescue is significantly enhanced by soft selective sweeps when multiple mutations contribute.
+Evolutionary rescue that requires two successive mutations has been investigated by \cite{martin2013probability} with the help of diffusion approximation.
 
@@
 
-We follow the number of cancer cells that have one of three different genotypes at time $t$: wildtype, $w_t$; aneuploid, $a_t$; and mutant, $m_t$. 
-These cells divide and die with rates $\lambda_k$ and $\mu_k$ (for $k=w, a, m$).
-The difference between the division and death rate is $\Delta_k = \lambda_k-\mu_k$.
-We assume the population of cells is under a strong stress, such as drug therapy, to which the wildtype genotype is susceptible and therefore $\Delta_w<0$, whereas the mutant is resistant to the stress, $\Delta_m>0$
-We analyze two cases: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_w$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}.
-Wildtype cells may missegregate to become aneuploids at rate $u$. Both aneuploid and wildtype cells may mutate to become mutants at rate $v$ (\Cref{figureAneuploidy}). 
-%Thus, the changes in the number of each cell type is described by 
-%\begin{equation}
-%\begin{aligned}
-%w_t&\rightarrow w_t+1:\quad \lambda_ww_t,\\
-%w_t&\rightarrow w_t-1:\quad \left(\mu_w+u+v\right)w_t,\\
-%a_t&\rightarrow a_t+1:\quad \lambda_aa_t+uw_t,\\
-%a_t&\rightarrow a_t-1:\quad \left(\mu_a+v\right)a_t,\\
-%m_t&\rightarrow m_t+1:\quad \lambda_am_t+va_t+vm_t,\\
-%m_t&\rightarrow m_t-1:\quad \mu_am_t.
-%\end{aligned}
-%\end{equation}
-
-%%%%%%%%%%%
-\begin{figure}[!t]
-\centering
-\includegraphics[width=0.5\textwidth]{Figures/figureAneuploidy.pdf}
-\caption{
-\textbf{Model illustration.}
-% TODO in the figure change labels to Wildtype, Aneuploid, Mutant (instead of Euploid etc)
-A population of cancer cells is composed of wildtype, aneuploid, and mutant cells, which divide with rates $\lambda_w$, $\lambda_a$, and $\lambda_m$ and die at rates $\mu_w$, $\mu_a$, and $\mu_m$, respectively.
-Wildtype cells can become aneuploid at rate $u$.
-Both aneuploid and wildtype cells can acquire a beneficial mutation with rate $v$.
-Color denotes the relative growth rates of the three genotypes such that $\lambda_w - \mu_w < \lambda_a - \mu_a < \lambda_m - \mu_m$.
-}
-\label{figureAneuploidy}
-\end{figure}
-%%%%%%%%%%%
-
-Evolutionary rescue occurs when mutant cells appear and fixate ($m_t>>1$) in the population before the population becomes extinct ($w_t=a_t=m_t=0$).
-Aneuploidy may contribute to evolutionary rescue by either preventing (when $\Delta_a>0$) or delaying (when $0>\Delta_a>\Delta_w$) the extinction of the population before mutant cells appear and fixate.
-To analyze evolutionary rescue in this model, we use the framework of \emph{multitype branching processes} \citep{rybnikov2021fitness,harris1963theory}. 
-This allows us to find explicit expressions for the \emph{survival probability}: the probability that the population does not become extinct.
-Note that branching processes assume that lineages produced by cells from the initial population divide and die independently of each other, which may be unrealistic, as cells usually compete for resources.
-A more realistic model may include competition for limited resources and spatial structure, which may play an important role in the development of cancer \citep[e.g.,][]{martens2011spatial}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+We follow the number of cancer cells that belong to three different types at time $t$: wildtype euploid, $w_t$; wildtype aneuploid, $a_t$; and mutant euploid, $m_t$. 
+These cells grow and die with rates $\lambda_k$ and $\mu_k$ for $k=w, a, m$.
+Wildtype cells become aneuploid at rate $u$. Aneuploid and wildtype cells mutate with rate $v$ (Figure \ref{figureAneuploidy}). 
+Thus, the changes in the number of each cell type is described by
+\begin{subequations}
+\begin{flalign}
+w_t&\rightarrow w_t+1:\quad \lambda_ww_t,\\
+w_t&\rightarrow w_t-1:\quad \left(\mu_w+u+v\right)w_t,\\
+a_t&\rightarrow a_t+1:\quad \lambda_aa_t+uw_t,\\
+a_t&\rightarrow a_t-1:\quad \left(\mu_a+v\right)a_t,\\
+m_t&\rightarrow m_t+1:\quad \lambda_am_t+va_t+vm_t,\\
+m_t&\rightarrow m_t-1:\quad \mu_am_t.
+\end{flalign}
+\end{subequations}
+
+The difference between growth and death is $\Delta_k = \lambda_k-\mu_k$.
+We assume that $\Delta_w<0$ due to drug therapy (or some other stress), and that $\Delta_m$ due to resistance conferred by the mutant.
+We analyze two cases: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are susceptible, $0>\Delta_a>\Delta_w$  \cite{brauner2016distinguishing}.
+
+Here, \emph{evolutionary rescue} occurs when the aneuploid cells are able to either prevent (when $\Delta_a>0$) or delay (when $0>\Delta_a>\Delta_w$) the extinction of the cancer-cell population before the adaptive mutant cells appear and fixate in the populaiton.
+To analyze evolutionary rescue in this model, we use the framework of \emph{multitype branching processes} \cite{rybnikov2021fitness,harris1963theory}. 
+This allows us to find explicit expressions for the survival probability of the population--the probability that it does not become extinct due to stressful conditions, e.g., drug therapy.
+Note that branching processes imply that lineages produced by cells from the initial population grow and die independently of each other.
+This is a necessary simplification, but it is unrealistic, as cells usually compete for resources.
+A more realistic model may include competition on limited resources between lineages, as well as spatial structure, which may play an important role in the development of cancer \cite{martens2011spatial}.
+
+The survival probability of a population consisting initially of one wildtype cell satisfies the quadratic equation,
+\begin{equation} \label{quadraticeqev1}
+\begin{aligned}
+1-p_w = & \frac{\mu_w}{\lambda_w+\mu_w+u+v} + 
+		  \frac{u}{\lambda_w+\mu_w+u+v}\left(1-p_a\right) + \\
+		  & \frac{\lambda_w}{\lambda_w+\mu_w+u+v}\left(1-p_w\right)^2 +
+		  \frac{v}{\lambda_w+\mu_w+u+v}\left(1-p_m\right) ,
+\end{aligned}
+\end{equation}
+where the $p_w$, $p_a$, and $p_m$ are the survival probabilities of a population consisting initially of one wildtype cell, one aneuploid cell, and one mutant cell, respectively.
+
+The smallest solution of the quadratic equation \cref{quadraticeqev1} gives the survival probability \cite{uecker2015adaptive}, which still depends on both $p_a$ and $p_m$,
+\begin{align}\label{survprobw}
+p_w=\frac{\lambda_w-\mu_w-u-v+\sqrt{\left(\lambda_w-\mu_w-u-v\right)^2+4\lambda_w\left(up_a+vp_m\right)}}{2\lambda_w} .
+\end{align}
+
+If the population originally consists of a single aneuploid cell, then the probability that the population will survive is given by the quadratic equation
+\begin{align}
+1-p_a=\frac{\mu_a}{\lambda_a+\mu_a+v}+\frac{v}{\lambda_a+\mu_a+v}\left(1-p_m\right)+\frac{\lambda_a}{\lambda_a+\mu_a+v}\left(1-p_a\right)^2,
+\end{align}
+whose smallest solution is the survival probability:
+\begin{align}\label{survproba}
+p_a=\frac{\lambda_a-\mu_a-v+\sqrt{\left(\lambda_a-\mu_a-v\right)^2+4\lambda_avp_m}}{2\lambda_a}.
+\end{align}
+
+The probability that a single mutant cell survives is given by (Appendix \ref{AppendixSurvLin}):
+\begin{equation}
+p_m = \left\{
+  \begin{array}{@{}ll@{}}
+  \frac{\lambda_m-\mu_m}{\lambda_m},\quad \text{if }\lambda_m>\mu_m \\
+   0,\quad \text{else} .
+  \end{array}\right .
+\end{equation}
+
+An alternative method to obtain $p_m$ would be to solve the following quadratic equation:
+\begin{align}
+1-p_m=\frac{\mu_m}{\lambda_m+\mu_m}+\frac{\lambda_m}{\lambda_m+\mu_m}\left(1-p_m\right)^2.
+\end{align}
+% TODO add subsection on the Gillespie simulations OK
 \subsection*{Evolutionary simulation} 
-Simulations are performed using a \emph{Gillespie algorithm} \citep{gillespie1976general,gillespie1977exact} implemented in Python \citep{python}.
+Simulations are performed using a \emph{Gillespie algorithm} \cite{gillespie1976general,gillespie1977exact} implemented in Python \cite{python}.
 The simulation monitors the number of cells of each type: wildtype, aneuploid, and mutant. 
 The wildtype population initially consists of $w_0$ cells, whereas the other cell types are initially absent.
-
 \begin{table}
 \begin{center}
-  \begin{tabular}{| l |p{5cm}| c | c | p{3cm} |}
+  \begin{tabular}{| l |c| c | c | c |}
     \hline
-     & Name & Value & Units & References \\ \hline
-    $N$ & Tumor size at beginning of treatment& $10^7-10^9$ & cells  & \citet{del2009does} \\ \hline
-    $\lambda_w$ & Wildtype birth rate& 0.14 & 1/days  & \citep{bozic2013evolutionary} \\ \hline
-    $\mu_w$ & Wildtype death rate& 0.17 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
-    $\lambda_a$  & Aneuploid birth rate$^\ast$ & 0.14 & 1/days  & - \\ \hline
-    $\mu_a$ & Aneuploid death rate$^\ast$ & $0.13-0.17$ & 1/days  & - \\ \hline
-    $\lambda_m$ & Mutant birth rate& 0.14 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
-    $\mu_m$ & Mutant death rate& 0.13 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
-    $u$ & Chromosomal instability rate& $10^{-3}-10^{-2}$ & 1$\slash$cell division  & \citet{nowak2004evolutionary,bakker2023predicting} \\ \hline
-    $v$ & Mutation rate& $10^{-7}-10^{-9}$ &  1$\slash$gene$\slash$cell division  & \citet{nowak2004evolutionary} \\ \hline
+    Symbol & Name & Value & Units & References \\ \hline
+    $N$ & Tumor size at beginning of treatment& $10^7-10^9$ & cells  & \cite{del2009does} \\ \hline
+    $\lambda_w$ & Wildtype birth rate& 0.14 & 1/days  & \cite{bozic2013evolutionary} \\ \hline
+    $\mu_w$ & Wildtype death rate& 0.17 & 1/days  & \cite{bozic2013evolutionary} \\ \hline
+    $\lambda_a$ & Aneuploid birth rate& 0.14 & 1/days  & * \\ \hline
+    $\mu_a$ & Aneuploid death rate& $0.13-0.17$ & 1/days  & * \\ \hline
+    $\lambda_m$ & Mutant birth rate& 0.14 & 1/days  & \cite{bozic2013evolutionary} \\ \hline
+    $\mu_m$ & Mutant death rate& 0.13 & 1/days  & \cite{bozic2013evolutionary} \\ \hline
+    $u$ & Chromosomal instability rate& $10^{-3}-10^{-2}$ & 1$\slash$cell division  & \cite{nowak2004evolutionary,bakker2023predicting} \\ \hline
+    $v$ & Mutation rate& $10^{-7}-10^{-9}$ &  1$\slash$gene$\slash$cell division  & \cite{nowak2004evolutionary} \\ \hline
   \end{tabular}
-\caption{\textbf{Model parameters.} 
-Aneuploid birth rate $\lambda_a$ is set to the same value as the wildtype and mutant birth rates, $\lambda_w$ and $\lambda_m$.
-Aneuploid death rate $\mu_a$ is set to an intermediate value between the wildtype and mutant death rates, $\mu_w$ and $\mu_m$.}
+\caption{\label{demo-table}\textbf{Model parameters.} Table of parameters used in our model (see Figure \ref{figureAneuploidy}).
+The * symbol in the reference column means that for those parameters the values have not been selected from a paper.
+For the aneuploid birth rate we have chosen the same value for the wildtype and mutant birth rates.
+For the anueploid death rate we have chosen an intermediate value between the wildtype and mutant death rates.}
   \label{table1}
@@
 
-The state of the stochastic system at time $t$ is represented by the triplet $\left(w_t,a_t,m_t\right)$.
-The following describes the events that may occur (right column), the rates at which they occur (middle column), and the effect these events have on the state (\Cref{figureAneuploidy}):
+At each iteration of the simulation loop we compute the rate $\nu_j$ of each event $j$. % TODO explains what events can occur and how you compute their rates, refer to illustration figure  OK
+The sate of the stochastic system at time $t$ is represented by the triplet $\left(w_t,a_t,m_t\right)$ which can change in the time interval $\Delta_t$ according to the following events with appropriate rates (see Figure \ref{figureAneuploidy}):
 \begin{subequations}
 \begin{flalign*}
-(+1,0,0)&:\quad \lambda_ww_t\quad\left(\text{birth of wildtype cell}\right),\\
-(-1,0,0)&:\quad \mu_ww_t\quad\left(\text{death of wildtype cell}\right),\\
-(-1,+1,0)&:\quad uw_t\quad\left(\text{wildtype cell becomes aneuploid}\right),\\
-(-1,0,+1)&:\quad vw_t\quad\left(\text{wildtype cell becomes mutant}\right),\\
-(0,+1,0)&:\quad \lambda_aa_t\quad\left(\text{birth of aneuploid cell}\right),\\
-(0,-1,0)&:\quad \mu_aa_t\quad\left(\text{death of aneuploid cell}\right),\\
-(0,-1,+1)&:\quad va_t\quad\left(\text{aneuploid cell becomes mutant}\right),\\
-(0,0,+1)&:\quad \lambda_am_t\quad\left(\text{birth of mutant cell}\right),\\
-(0,0,-1)&:\quad \mu_am_t\quad\left(\text{death of mutant cell}\right).
+(+1,0,0)&:\quad \lambda_ww_t\quad\left(\text{birth euploid cell}\right),\\
+(-1,0,0)&:\quad \mu_ww_t\quad\left(\text{death euploid cell}\right),\\
+(-1,+1,0)&:\quad uw_t\quad\left(\text{euploid cell aquires aneuploidy}\right),\\
+(-1,0,+1)&:\quad vw_t\quad\left(\text{euploid cell aquires mutation}\right),\\
+(0,+1,0)&:\quad \lambda_aa_t\quad\left(\text{birth aneuploid cell}\right),\\
+(0,-1,0)&:\quad \mu_aa_t\quad\left(\text{death aneuploid cell}\right),\\
+(0,-1,+1)&:\quad va_t\quad\left(\text{aneuploid cell aquires mutation}\right),\\
+(0,0,+1)&:\quad \lambda_am_t\quad\left(\text{birth mutant cell}\right),\\
+(0,0,-1)&:\quad \mu_am_t\quad\left(\text{death mutant cell}\right).
 \end{flalign*}
 \end{subequations}
-Each iteration of the simulation loop starts with computing the rates $\nu_j$ of each event $j$.
 We then draw the time until the next event, $\Delta t$, from an exponential distribution whose rate parameter is the sum of the rates of all events, such that $\Delta t \sim \textit{Exp}(\sum_j \nu_j)$.
@@
 OK
-We repeat these iterations until either the population becomes extinct (the number of cells of all types is zero) or the number of mutant cells is high enough so that its extinction probability is $<0.1\%$, that is until
+We repeat these iterations until either the population becomes extinct (the number of cells of all types is zero) or the number of mutant cells is high enough so that its extinction probability is $<0.1\%$.
+% TODO how do you determine the extinction prob?
+ref to equation OK
+We continue running the simulation until:
 \begin{equation*}
@@
 \end{equation*}
-
-\paragraph{$\tau$-leaping.}
-% TODO "is large" - how large?
-When the size of the initial population is large we utilize $\tau$-leaping \citep{gillespie2001approximate}, where change in number of cells of genotype $i$ in a fixed time interval $\Delta t$ is Poisson distributed with mean $\nu_i\Delta t$.
-If the increment is negative and larger then the subpopulation size then updated to be zero. % TODO "and larger..."??
-
-\paragraph{Density-dependent growth.}
-To simulate birth and death rates that depend on the number of cells in the population, we transform the rates of division and death to the following:
-\begin{align*}
-\lambda_w' &= \lambda_w, \\
-\mu_w' &= \mu_w,\\
-\lambda_a' &= C_1+\left(\lambda_a-\mu_a\right)\left(1-\frac{w+a+m}{K}\right),\\ 
-\mu_a' &= C_1,\\
-\lambda_m' &= C_2+\left(\lambda_m-\mu_m\right)\left(1-\frac{w+a+m}{K}\right),\\ 
-\mu_m' &= C_2,
-\end{align*}
-where $C_1, C_2>0$ are constants. % TODO what constants?
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Code and data availability.} All source code is available online at \url{https://github.com/yoavram-lab/EvolutionaryRescue}.
+which is the population size at which the extinction probability of the population of mutant cell is less then $0.1\%$. 
+% TODO do you use tau leap? if so, update the text accordingly. OK
+When the size of the initial population is large we utilize $\tau$-leaping where the increment of population change in a fixed time interval $\Delta t$ is Poisson distributed with mean $\nu_i\Delta t$ for each population $i$ \cite{gillespie2001approximate}.
+If the increment is negative and larger then the subpopulation size then updated to be zero.
+\paragraph{Data availability.} All source code is available online at \url{https://github.com/yoavram-lab/EvolutionaryRescue}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% TODO combine all diagram figures to a single figure with multiple panels. and make it fit in a single page. OK
+
+%%%%%%
+
+\begin{figure}
+\centering
+\includegraphics[width=0.75\textwidth]{Figures/figureAneuploidy.pdf}
+%\caption{}
+
+%\end{subfigure}
+
+\caption{
+A population of cancer cells is subdivided to wildtype euploid, wildtype aneuploid and mutant euploid cells, which divide with rates $\lambda_w$, $\lambda_a$, and $\lambda_m$, respectively, and die at rates $\mu_w$, $\mu_a$, and $\mu_m$, respectively.
+Wildtype cells can become aneuploid at rate $u$. Both aneuploid and wildtype cells can acquire a beneficial mutation with rate $v$.}
+\label{figureAneuploidy}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-% TODO: OVERVIEW OF RESULTS
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Survival probability}
-
-Let $p_w$, $p_a$, and $p_m$ be the survival probabilities of a population consisting initially of single wildtype cell, aneuploid cell, or mutant cell, respectively.
-The complements $1-p_w$, $1-p_a$, and $1-p_m$ are the extinction probabilities, which satisfy each its respective equation,
-\begin{equation} \label{extinction_prob}
-\begin{aligned}
-1-p_w = &\frac{\mu_w}{\lambda_w+\mu_w+u+v} + 
-		  \frac{u}{\lambda_w+\mu_w+u+v}\left(1-p_a\right) + \\
-		  & \frac{\lambda_w}{\lambda_w+\mu_w+u+v}\left(1-p_w\right)^2 +
-		  \frac{v}{\lambda_w+\mu_w+u+v}\left(1-p_m\right) ,\\
-1-p_a = &\frac{\mu_a}{\lambda_a+\mu_a+v}+\frac{v}{\lambda_a+\mu_a+v}\left(1-p_m\right)+\frac{\lambda_a}{\lambda_a+\mu_a+v}\left(1-p_a\right)^2 ,\\
-1-p_m = &\frac{\mu_m}{\lambda_m+\mu_m}+\frac{\lambda_m}{\lambda_m+\mu_m}\left(1-p_m\right)^2 .	 
-\end{aligned}
-\end{equation}
-
-The survival probabilities are given by the smallest solution for each quadratic equation \citep{uecker2015adaptive}. Therefore we have
-\begin{equation}\label{survival_prob}
-\begin{aligned}
-p_w &= \frac{\lambda_w-\mu_w-u-v+\sqrt{\left(\lambda_w-\mu_w-u-v\right)^2+4\lambda_w\left(up_a+vp_m\right)}}{2\lambda_w} ,\\
-p_a &= \frac{\lambda_a-\mu_a-v+\sqrt{\left(\lambda_a-\mu_a-v\right)^2+4\lambda_avp_m}}{2\lambda_a}, \\
-p_m &= \left\{
-  \begin{array}{@{}ll@{}}
-  \frac{\lambda_m-\mu_m}{\lambda_m},& \lambda_m \ge \mu_m \\
-   0, & \lambda_m \le \mu_m . %TODO in the appendix this is 1, not 0
-  \end{array}\right .  
-\end{aligned}
-\end{equation}
-See Appendix \ref{AppendixSurvLin} for the derivation of $p_m$. %TODO why do we need Appendix 1 if we can just find p_m as solution to quadratic equation?
-Note that the equation for $p_w$ depends on both $p_a$ and $p_m$, and the equation for $p_a$ depends on $p_m$.
-To proceed, we can plug the solution for $p_m$ and $p_a$ into the solution for $p_w$. We perform this for three different scenarios.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{Scenario 1: Aneuploid cells are drug resistant}
-
-We first assume that aneuploidy provides partial resistance to drug therapy, $\lambda_a>\mu_a$, and that this resistance is significant, $\left(\lambda_a-\mu_a-v\right)^2 > 4\lambda_a v p_m$.
-We thus rewrite \cref{survival_prob} as
-\begin{align*}
-p_w&=\frac{\lambda_w-\mu_w-u-v}{2\lambda_w}\left(1-\sqrt{1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}}\right) ,
-\text{and} \\
-p_a&=\frac{\lambda_a-\mu_a-v}{2\lambda_a}\left(1+\sqrt{1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}}\right) . 
-\end{align*}
-Using the quadratic Taylor expansion $\sqrt{1+x}=1+x/2+O(x^2)$ and assuming $u,v \ll 1$,
-we obtain the following approximation for the survival probability of a population initially consisting of a single wildtype cell,
-\begin{align}\label{survprobwapprox1}
+% OVERVIEW OF RESULTS
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% TODO give a descriptive name instead of a mathematical formula OK
+%\subsection*{First case: $4\lambda_avp_m<\left(\lambda_a-\mu_a-v\right)^2$}
+\subsection*{First case: Highly deleterious or highly beneficial aneuploidy}
+
+\paragraph{Aneuploid cells are drug resistant.} 
+We first assume wildtype cells are susceptible to the drug, $\lambda_w<\mu_w$, whereas aneuploid cells are resistant, $\lambda_a>\mu_a$.
+Thus, we rewrite \cref{survprobw,survproba} as
+\begin{align*}
+p_a&=\frac{\lambda_a-\mu_a-v}{2\lambda_a}\left(1+\sqrt{1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}}\right),\\
+p_w&=\frac{\lambda_w-\mu_w-u-v}{2\lambda_w}\left(1-\sqrt{1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}}\right).
+\end{align*}
+Using a quadratic Taylor expansion, $\sqrt{1+x}=1+x/2+O(x^2)$, 
+we obtain the following approximation for the survival probability of a population consisting of a single individual wildtype cell (assuming $u,v \ll 1$),
+\begin{align}\label{survprobwinitial}
 p_w 
@@
 %&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{v\left(\lambda_a-\mu_a-u\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-u\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\ \label{survprobw2}
-&\approx-\frac{1}{\lambda_w-\mu_w}\left[\frac{u\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\
-\end{align}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\paragraph{Second-order approximation.}
-To improve our approximation, we can consider the second term of the Taylor series expansion,
+&\approx-\frac{1}{\lambda_w-\mu_w}\left[\frac{u\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right].
+\end{align}
+
+We write \cref{survprobw} as
+\begin{equation}\label{survprobwapprox1}
+p_w=-\frac{1}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right).
+\end{equation}
+So, given an initial population of $N$ wildtype cells, the probability that the population will survive is given by
+\begin{align}\label{aneuploidyresistentfirstapprox}
+p_{resc}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right] .
+\end{align}
+\Cref{SurvPlot} for $N=10^4$ and \Cref{SurvPlotLargeN} for $N=10^8$  show that the survival probability, $p_{resc}$, quickly grows as the wildtype growth rate, $\lambda_w$, increases.
+% Todo I dont think this figure should show lambda_w>1, as in that case the wildtype is resistant and there is no rescue scenario.
+OK
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{figure}[!t]
+ \vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/CombinedSubplot.pdf}
+\caption{Plot of the probability of evolutionary rescue of a population, consisting initially of $N$ wildtype cells, as a function of the proliferation rate of the wildtype cells for various values of the proliferation rate of the aneuploid cells.
+The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximation \eqref{aneuploidyresistentfirstapprox} for the upper plot and \eqref{AneuploidyDeleteriousApprox} for the lower plot.
+The dots represent numerical simulations where the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
+Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $N=10^4, \lambda_a=1+10^{-2},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
+}
+\label{SurvPlot}
+\end{figure}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{figure}[!t]
+ \vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/CombinedSubplotLargePop.pdf}
+\caption{Plot of the probability of evolutionary rescue of a population, consisting initially of $N$ wildtype cells, as a function of the proliferation rate of the wildtype cells for various values of the proliferation rate of the aneuploid cells.
+The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximation \eqref{aneuploidyresistentfirstapprox} for the upper plot and \eqref{AneuploidyDeleteriousApprox} for the lower plot.
+The dots represent numerical simulations where the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
+Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $N=10^8, \lambda_w=0.14, \lambda_a=0.14,\lambda_m=0.14,\mu_m=0.13, u=10^{-2}, v=10^{-7}$.
+}
+\label{SurvPlotLargeN}
+\end{figure}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\begin{figure}[!t]
+% \vspace*{1\baselineskip}
+%\includegraphics[width=1\textwidth]{Figures/P_est_growth.pdf}
+%\caption{Plot of the probability of survival of a population as a function of the proliferation rate of the wildtype cells.
+The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximation \eqref{survprobwapprox1}.
+Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $N=10^4, \lambda_a=1+10^{-2},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
+The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
+The error bars are present but are not visible given the fact that we have used $10^5$ simulations for each combination of parameters.}
+%\label{SurvPlot}
+%\end{figure}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+We want to improve the accuracy of our approximation by taking into consideration the second term of the Taylor series expansion:
 \begin{align*}
@@
 \end{align*}
-which gives us the following approximation,
+which gives us the following approximation for $p_a$:
 \begin{align}
-p_a \approx \frac{\lambda_a-\mu_a-v}{\lambda_a}+\frac{vp_m}{\lambda_a-\mu_a-v}-\frac{\lambda_a\left(vp_m\right)^2}{8\left(\lambda_a-\mu_a-v\right)^3} .
-\end{align}
-We therefore have
+p_a=\frac{\lambda_a-\mu_a-v}{\lambda_a}+\frac{vp_m}{\lambda_a-\mu_a-v}-\frac{\lambda_a\left(vp_m\right)^2}{8\left(\lambda_a-\mu_a-v\right)^3}.
+\end{align}
+From which we deduce that:
 \begin{align}\nonumber
 p_w&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{u\left(\lambda_a-\mu_a-v\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-v\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}-\frac{uv^2\lambda_a\left(\lambda_m-\mu_m\right)^2}{8\lambda_m^2\left(\lambda_a-\mu_a-v\right)^3}\right]\\ \label{survprobw3}
-&\approx-\frac{1}{\lambda_w-\mu_w}\left[\frac{u\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}-\frac{uv^2\lambda_a\left(\lambda_m-\mu_m\right)^2}{8\lambda_m^2\left(\lambda_a-\mu_a\right)^3}\right],
-\end{align}
-and using $\Delta_k=\lambda_k-\mu_k$, we can write the above equation as
+&\approx-\frac{1}{\lambda_w-\mu_w}\left[\frac{u\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}-\frac{uv^2\lambda_a\left(\lambda_m-\mu_m\right)^2}{8\lambda_m^2\left(\lambda_a-\mu_a\right)^3}\right].
+\end{align}
+Using the notations described in \eqref{notationalconv} we write the above equation as:
 \begin{equation}\label{survprobwapproxcorrected}
-p_w \approx -\frac{1}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}-\frac{uv^2\lambda_a\Delta_m^2}{8\lambda_m^2\Delta_a^3}\right).
-\end{equation}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{Scenario 2: Aneuploid cells are drug tolerant.}
-
-We now assume that aneuploidy provides tolerance to the drug, that is, the number of aneuploid cells significantly declines over time, but at a lower rate than the number of wildtype cells, $\lambda_w - \mu_w < \lambda_a - \mu_a < 0$.
-We also assume that both declines are significant, $\left(\lambda_a-\mu_a-v\right)^2 > 4\lambda_a v p_m$.
-We rewrite \cref{survival_prob} as
-\begin{align*}
-p_w&=\frac{\lambda_w-\mu_w-u-v}{2\lambda_w}\left(1-\sqrt{1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}}\right) ,
-\text{and} \\
-p_a&=\frac{\lambda_a-\mu_a-v}{2\lambda_a}\left(1-\sqrt{1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}}\right) .
-\end{align*}
-Since $u,v\ll1$, this can be approximated by % TODO how do you do the first approx? taylor?
-\begin{equation}\label{survprobwinitial}
-\begin{aligned}
+p_w=-\frac{1}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}-\frac{uv^2\lambda_a\Delta_m^2}{8\lambda_m^2\Delta_a^3}\right).
+\end{equation}
+Given an initial population consisting of $N$ wildtype cancer cells, the probability that the population will survive is given by: 
+\begin{align}\label{survprobresistantaneuploidy}
+p_{resc}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}-\frac{uv^2\lambda_a\Delta_m^2}{8\lambda_m^2\Delta_a^3}\right)\right].
+\end{align}
+\Cref{SurvPlotNData} show $p_{rescue}$ as a function of $N$, including comparison of our first approximation \eqref{aneuploidyresistentfirstapprox} % ref to equation OK
+and simulation results.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{figure}[!t]
+ \vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/SurvPlotNData.pdf}
+\caption{Plot of the probability of survival of a population as a function of the initial population size of wildtype cells.
+The blue line represents the exact solution \eqref{survprobw}, the orange line line represents the approximation \eqref{aneuploidyresistentfirstapprox}, the green line represents the first order correction \eqref{survprobwapproxcorrected} and the red dots represents stochastic simulations.
+For the simulations we have chosen the following parameters: $\lambda_w=0.14, \lambda_a=0.14,\lambda_m=0.14,\mu_w=0.17,\mu_a=0.135,\mu_m=0.13$.
+The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n=100$ is the number of simulations.}
+\label{SurvPlotNData}
+\end{figure}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\paragraph{$\lambda_a<\mu_a$ and $\lambda_w<\mu_w$.}
+We assume that $\lambda_a<\mu_a$ and $\lambda_w<\mu_w$ and, as a result, we rewrite \eqref{survprobw} and \eqref{survproba} as:
+\begin{align*}
+p_a&=\frac{\lambda_a-\mu_a-v}{2\lambda_a}\left(1-\sqrt{1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}}\right),\\
+p_w&=\frac{\lambda_w-\mu_w-u-v}{2\lambda_w}\left(1-\sqrt{1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}}\right).
+\end{align*}
+As a result, we can approximate:
+\begin{align}\label{survprobwinitial}
 p_w&\approx-\frac{vp_m+up_a}{\lambda_w-\mu_w-u-v}\\
-&\approx\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-v\right)}-\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\ %\label{survprobw2}
-&\approx\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_w-\mu_w\right)}\left[\frac{u}{\left(\lambda_a-\mu_a\right)}-1\right]\\
-%&=\frac{v\Delta_m}{\lambda_m\Delta_w}\left(\frac{u}{\Delta_a}-1\right) .
-\end{aligned}
-\end{equation}
-%Using the notational convention
-%\begin{equation}\label{notationalconv}
-%\Delta_i=\lambda_i-\mu_i,
-%\end{equation}
-%we write \eqref{survprobw} as
-%\begin{equation}\label{survprobwapprox2}
-%p_w=\frac{v\Delta_m}{\lambda_m\Delta_w}\left(\frac{u}{\Delta_a}-1\right).
-%\end{equation}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{Scenario 3: Aneuploid cells are non-growing.}
-
-We now assume that the growth rate of aneuploid cells is close to zero (either positive or negative), such that  $\left(\lambda_a-\mu_a-v\right)^2 < 4\lambda_avp_m$.
-We rewrite \cref{survival_prob} as
+\nonumber
+&\approx\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-v\right)}-\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\ \label{survprobw2}
+&\approx\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_w-\mu_w\right)}\left[\frac{u}{\left(\lambda_a-\mu_a\right)}-1\right],
+\end{align}
+where in the last line we have used the fact that $u,v\ll1$. Using the notational convention
+\begin{equation}\label{notationalconv}
+\Delta_i=\lambda_i-\mu_i,
+\end{equation}
+we write \eqref{survprobw} as
+\begin{equation}\label{survprobwapprox2}
+p_w=\frac{v\Delta_m}{\lambda_m\Delta_w}\left(\frac{u}{\Delta_a}-1\right).
+\end{equation}
+Given an initial population consisting of $N$ wildtype cancer cells, the probability that the population will survive is given by: 
+\begin{align}\label{AneuploidyDeleteriousApprox}
+p_{resc}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right],
+\end{align}
+which we plot in Figure \ref{P_est} for initial population size $N=10^4$ and Figure \ref{P_est_large_N} for initial population size $N=10^8$.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\begin{figure}[!t]
+% \vspace*{1\baselineskip}
+%\includegraphics[width=1\textwidth]{Figures/P_est.pdf}
+%\caption{Plot of the survival probability of an initial population consisting of $10^{4}$ wildtype cells as a function of $\Delta_a=\lambda_a-\mu_a$ for various values of $\Delta_w=\lambda_a-\mu_a$.
+The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximation \eqref{AneuploidyDeleteriousApprox}.
+The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.}
+%\label{P_est}
+%\end{figure}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{figure}[!t]
+ \vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/P_est_divergence.pdf}
+\caption{Plot of the probability of evolutionary rescue, of an initial population, consisting of $N$ wildtype cells, as a function of proliferation rate of the wildtype cells $\Delta_w=\lambda_w-\mu_w$ for various values of the proliferation rate of the aneuploid cells $\Delta_a=\lambda_a-\mu_a$.
+The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximations \eqref{AneuploidyDeleteriousApprox} and \eqref{pestthirdcase}.
+The dots represent numerical simulations where the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
+The value highlighted in grey is the threshold $\Delta_a^*$ from \eqref{thresholdvalueaneuploid} which marks the transition between the regime dictated by \eqref{AneuploidyDeleteriousApprox} to the one dictated by \eqref{pestthirdcase}.
+Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $N=10^4,\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
+}
+\label{P_est}
+\end{figure}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{figure}[!t]
+ \vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/P_est_divergenceLargePopulation.pdf}
+\caption{Plot of the probability of evolutionary rescue, of an initial population, consisting of $N$ wildtype cells, as a function of proliferation rate of the wildtype cells $\Delta_w=\lambda_w-\mu_w$ for various values of the proliferation rate of the aneuploid cells $\Delta_a=\lambda_a-\mu_a$.
+The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximations \eqref{AneuploidyDeleteriousApprox}.
+The dots represent numerical simulations where the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
+Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $N=10^8,\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
+}
+\label{P_est_large_N}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{figure}[!t]
+ \vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/DeleteriousTauLeapPlot.pdf}
+\caption{Plot of the probability of evolutionary rescue of a population, consisting of $N$ wildtype cells, as a function of the initial population size of wildtype cells.
+The blue line represent the exact result \eqref{survprobw} while the orange lines represent the approximation \eqref{pestthirdcase}.
+The green dots represent numerical simulations where the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
+The error bars are present but are not visible given the fact that we have used $n=10^5$ simulations for each combination of parameters.
+Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.}
+\label{DeleteriousPlot}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\subsection*{Second case: $4\lambda_avp_m>\left(\lambda_a-\mu_a-v\right)^2$}
+\subsection*{Second case: Neutral aneuploidy}
+If we assume that  $4\lambda_avp_m>\left(\lambda_a-\mu_a-v\right)^2$ then we write:
 \begin{equation}
-p_a=\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}\left(1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\lambda_avp_m}\right)^{\frac12}}{2\lambda_a} .
-\end{equation}
-Using a following Taylor series expansion
-\begin{equation*}
+p_a=\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}\left(1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\lambda_avp_m}\right)^{\frac12}}{2\lambda_a},
+\end{equation}
+and using the following Taylor series expansion:
+\begin{align*}
 \left(1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\lambda_avp_m}\right)^{\frac{1}{2}}=1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{8\lambda_avp_m}+\cdots,
-\end{equation*}
-we obtain the approximation
+\end{align*}
+we obtain:
 \begin{align*}
@@
 \end{align*}
-From \Cref{survprobwinitial}, % TODO I dont see how the referred equation plays a part here
-the survival probability of a population starting from one wildtype individual is
-\begin{align}\label{scenario3}
+As a result, we have from \eqref{survprobwinitial} the probability of rescue of a population starting from one wildtype individual:
+\begin{align*}
 p_w&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[v\frac{\lambda_m-\mu_m}{\lambda_m}+\frac{u}{2\lambda_a}\left(\lambda_a-\mu_a-v+2\sqrt{\lambda_avp_m}\right)\right]\\
-&=-\frac{1}{\lambda_w-\mu_w-u-v}\left[v\frac{\lambda_m-\mu_m}{\lambda_m}+\frac{u}{2\lambda_a}\left(\lambda_a-\mu_a-v\right)+u\sqrt{\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_a\lambda_m}}\right].
-%&=-\frac{1}{\Delta_w-u-v}\left[v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right],
-\end{align}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Evolutionary rescue probability}
-
-To estimate the rescue probability $p_{rescue}$, we assume independence between clonal lineages starting from an initial population of $N$ wildtype cells.
-Thus, the rescue probability is given by 
-\begin{equation}
-p_{rescue}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w} ,
-\end{equation}
-where the approximation $(1-p_w)\approx e^{-p_w}$ assumes that $p_w$ (but not $N p_w$) is small.
-
-Now, we can apply the approximations for the survival probability $p_w$ from \Cref{survprobwapprox1,survprobwinitial,scenario3} and substituting $\Delta_k=\lambda_k-\mu_k$
-
-The rescue probability is given by
-\begin{equation}\label{rescue_prob}
-p_{rescue} = \left\{
+&=-\frac{1}{\lambda_w-\mu_w-u-v}\left[v\frac{\lambda_m-\mu_m}{\lambda_m}+\frac{u}{2\lambda_a}\left(\lambda_a-\mu_a-v\right)+u\sqrt{\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_a\lambda_m}}\right]\\
+&=-\frac{1}{\Delta_w-u-v}\left[v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right],
+\end{align*}
+where in the last line we have used the notations defined in \eqref{notationalconv}.
+
+Given an initial population consisting of $N$ wildtype cancer cells, the probability that the population will survive is given by: 
+\begin{align}\label{pestthirdcase}
+p_{resc}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w-u-v}\left(v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right)\right],
+\end{align}
+which we plot in Figure \ref{DeleteriousPlot} where we compare with numerical simulations and the exact result \eqref{survprobw}.
+The transition between the regimes defined by \eqref{AneuploidyDeleteriousApprox} and \eqref{pestthirdcase} respectively occurs at:
+\begin{align}\label{thresholdvalueaneuploid}
+\Delta_a^*=2vp_m+v+2\sqrt{vp_m\left(vp_m+\mu_a+v\right)}.
+\end{align}
+
+The probability of evolutionary rescue is given by:
+\begin{equation}\label{probresc}
+p_{resc}\sim\left\{
   \begin{array}{@{}ll@{}}
@@
 
-%Note that if instead \cref{survprobwapprox1} we apply \cref{survprobwapproxcorrected}, we have
-%\begin{align}\label{survprobresistantaneuploidy}
-%p_{rescue}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}-\frac{uv^2\lambda_a\Delta_m^2}{8\lambda_m^2\Delta_a^3}\right)\right].
-%\end{align}
-
-In \Cref{SurvPlot,SurvPlotLargeN} we explore the effects of the wildtype and aneuploid growth rates on the rescue probability for small and large population sizes ($N=10^4$ and $N=10^8$, respectively).
-
-\Cref{SurvPlotNData} show $p_{rescue}$ as a function of $N$, including comparison of our first approximation \eqref{aneuploidyresistentfirstapprox} and simulation results.
-
-\Cref{P_est,P_est_large_N} show the rescue probability for initial population sizes $N=10^4$ and $N=10^8$, respectively. % TODO check figure refs
-
-In \Cref{DeleteriousPlot}, we compare the exact result (\cref{survival_prob}) with numerical simulations.
-The transition between the regimes defined by \cref{AneuploidyDeleteriousApprox} and \cref{pestthirdcase} respectively occurs at:
-\begin{align}\label{thresholdvalueaneuploid}
-\Delta_a^*=2vp_m+v+2\sqrt{vp_m\left(vp_m+\mu_a+v\right)}.
-\end{align}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\paragraph*{Density-dependent growth}
-
+% TODO : should this move to methods? we only use this for simulations, right? should be in the Gillespie section?
+\subsection*{Logistic growth}
+We want to have birth and death rates which depends on the population size of wildtype $w$, aneuploidy $a$ and mutant $m$ cells:
+\begin{align*}
+&\lambda_w'=\lambda_w,\quad\mu_w'=\mu_w,\\
+&\lambda_a'=C_1+\left(\lambda_a-\mu_a\right)\left(1-\frac{w+a+m}{K}\right),\quad \mu_a'=C_1,\\
+&\lambda_m'=C_2+\left(\lambda_m-\mu_m\right)\left(1-\frac{w+a+m}{K}\right),\quad \mu_m'=C_2,
+\end{align*}
+where $C_1, C_2>0$ are constants.
 We perform stochastic simulations for different values of the carrying capacity $K$ and we plot the results in Figure \ref{SurvPlotNDataLogisticKComplete}.
@@
 
+\begin{figure}[!t]
+ \vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/SurvPlotNDataLogisticK.pdf}
+\caption{Plot of the probability of evolutionary rescue of a population, consisting of $N$ wildtype cells, as a function of the initial population size of wildtype cells where maximum population size is constrained by the carrying capacity $K$.
+The blue line represent the exact result \eqref{survprobw} while the orange dots represent numerical simulations where the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
+Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_w=1-10^{-1},\lambda_a=1+10^{-4},\lambda_m=1+10^{-1},\mu_w=1,\mu_m=1,u=10^{-2},v=10^{-7}, C_1=C_2=1, K=10^9$.}
+\label{SurvPlotNDataLogisticKComplete}
+\end{figure}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection*{Standing genetic variation}
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{figure}[!t]
+ \vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/FractionPlot.pdf}
+\caption{Plot of the fraction of the cases the population is rescue by standing genetic variation as a function of the fraction of initial cells which are aneuploid.
+The red vertical line highlights the the value of $f$ for which half the times the population is rescues by aneuploid cells while the pink line is our approximation \eqref{halfeqstandvar}.
+For this plot we have chosen the following parameters: $N=10^3, \lambda_w=1-10^{-2}, \lambda_a=1-10^{-4},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
+The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
+The value of $0.332$ highlighted in red is the value of the initial fraction of the population which in aneuploid for which half of the cases of evolutionary rescue are due to the initial aneuploid cell population.}
+\label{FractionPlot}
+\end{figure}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 So far we have assumed that the initial population of cells consisted entirely of wildtype cells.
@@
 \begin{equation}\label{ratiorescueexact}
-H=\frac{\left.p_{rescue}\right|_{u>0}}{\left.p_{rescue}\right|_{u=0}}.
-\end{equation}
-As a result, we obtain from \eqref{rescue_prob} the approximation for the ratio:
+H=\frac{\left.p_{resc}\right|_{u>0}}{\left.p_{resc}\right|_{u=0}}.
+\end{equation}
+As a result, we obtain from \eqref{probresc} the approximation for the ratio:
 \begin{equation}\label{ratiorescue}
@@
 We plot \eqref{ratiorescue} in Figure \ref{RatioEvolRescue} for both resistent and susceptible aneuploidy as a function of the proliferation rate of the wildtype cells and in Figure \ref{RatioEvolRescuePopulationSize} as a function of the initial population size of wildtype cells.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{figure}[!t]
+ \vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/RatioEvolRescue.pdf}
+\caption{Plot of the ratio of the probability of evolutionary rescue when aneuploidy can play a role in rescue ($u>0$) to the probability where acquisition of aneuploidy is not possible ($u=0$) as a function of the proliferation rate of the wildtype cells.
+The continuous lines represent the exact result \eqref{ratiorescueexact} while the dashed lines represent the approximation \eqref{ratiorescue}.
+The upper plot show the case when aneuploidy is resistant while the lower plot shows the case when it is susceptible.
+Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_w=1-10^{-1},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
+}
+\label{RatioEvolRescue}
+\end{figure}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{figure}[!t]
+ \vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/RatioEvolRescuePopulationSize.pdf}
+\caption{Plot of the ratio of the probability of evolutionary rescue when aneuploidy can play a role in rescue ($u>0$) to the probability where acquisition of aneuploidy is not possible ($u=0$) as a function of the initial population size of wildtype cells.
+The continuous lines represent the exact result \eqref{ratiorescueexact} while the dashed lines represent the approximation \eqref{ratiorescue}.
+The upper plot show the case when aneuploidy is resistant while the lower plot shows the case when it is susceptible.
+Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_w=0.14, \lambda_a=0.14,\lambda_m=0.14,\mu_w=0.17,\mu_m=0.13, u=10^{-2}, v=10^{-7}$.
+}
+\label{RatioEvolRescuePopulationSize}
+\end{figure}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection*{Rescue time}
+%%%
+\begin{figure}[!t]
+ \vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/MeanTimeGrowthMutantDirectPlot.pdf}
+\caption{Plot of the mean time until the appearance of a resistance mutation which rescues the population in the case when evolutionary rescue is possible only through mutation but not aneuploidy and mutation.
+The blue line represents the approximation \eqref{meantime1} and the dashed red line represents the first order approximation \eqref{limitapprox}.
+The orange dots represent the numerical simulations while the error bars represent the interval centered at the mean which containing $95\%$ of the simulated values.
+Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_w=1-10^{-1},\lambda_m=1+10^{-1},\mu_w=1,\mu_m=1,u=10^{-2},v=10^{-7}$.}
+\label{MeanTimeGrowthAneuploidyPlot}
+\end{figure}
 We calculate the mean time for the appearance of the first mutant that rescues the cancer cell population.
@@
 \end{align}
-The fration in exponential of the integrand in \eqref{meantime1} can approximated as:
+The fration in the exponential of the integrand in \eqref{meantime1} can approximated as:
 \begin{equation*}
@@
 %\end{align}
-%Following the steps outlines in \citep{weissman2009rate} we obtain:
+%Following the steps outlines in \cite{weissman2009rate} we obtain:
 %\begin{align}
@@
 \begin{align*}
-\frac{\e^{\Delta_w\tau}-1}{\Delta_w}&=\frac{1+\Delta_w\tau+\frac{\Delta_w^2\tau^2}{2}+O(\tau^3)-1}{\Delta_w}=\tau+\frac{\Delta_w}{2}\tau^2+O(\tau^3).\\
+\frac{\e^{\Delta_w\tau}-1}{\Delta_w}&=\frac{1+\Delta_w\tau+\frac{\Delta_w^2\tau^2}{2}+O(\tau^3)-1}{\Delta_w}=\tau+\frac{\Delta_w}{2}\tau^2+O(\tau^3),\\
 \frac{\e^{\Delta_a\tau}-1}{\Delta_a}&=\frac{1+\Delta_a\tau+\frac{\Delta_a^2\tau^2}{2}+O(\tau^3)-1}{\Delta_a}=\tau+\frac{\Delta_w}{2}\tau^2+O(\tau^3),
@@
 \begin{align*}
-\frac{\e^{\Delta_w\tau}-1}{\Delta_w}&=\frac{1+\Delta_w\tau+O(\tau^2)-1}{\Delta_w}=\tau+O(\tau^2).\\
+\frac{\e^{\Delta_w\tau}-1}{\Delta_w}&=\frac{1+\Delta_w\tau+O(\tau^2)-1}{\Delta_w}=\tau+O(\tau^2),\\
 \frac{\e^{\Delta_a\tau}-1}{\Delta_a}&=\frac{1+\Delta_a\tau+O(\tau^2)-1}{\Delta_a}=\tau+O(\tau^2),
@@
 This is consistent with the information obtained from Figure \ref{RatioEvolRescuePopulationSize} where aneuploidy improves the probability of evolutionary rescue only  for small and intermediate values of $N$.
-\begin{figure}[!h]
+\begin{figure}[!t]
  \vspace*{1\baselineskip}
@@
 
-\begin{figure}[!h]
+\begin{figure}[!t]
  \vspace*{1\baselineskip}
@@
 
-We propose experiments similar to the ones highlighted in \citep{martin2013probability} in order to test the predictions of our model.
+We propose experiments similar to the ones highlighted in \cite{martin2013probability} in order to test the predictions of our model.
 For example, in order to study the effects of initial population size on the probability of evolutionary rescue we propose to derive a large culture mass from a single cancer cell in permissive conditions and then dilute to a wide range of stating population sizes ($10^7-10^9$).
@@
 
-We observe from equations \eqref{survprobw} and \eqref{rescue_prob} that the probability of evolutionary rescue increases when the initial population size increases, the wildtype population does not decline too quickly, the mutation and aneuploidy rates are high and the probabilities $p_a$ and $p_m$ are elevated.
+We observe from equations \eqref{survprobw} and \eqref{probresc} that the probability of evolutionary rescue increases when the initial population size increases, the wildtype population does not decline too quickly, the mutation and aneuploidy rates are high and the probabilities $p_a$ and $p_m$ are elevated.
 
 The probability of evolutionary rescue is enhanced by aneuploidy for small and intermediate sized tumors (see Figure \ref{RatioEvolRescuePopulationSize}).
-As a result, aneuploidy is unlikely to to contribute to primary tumors overcoming chemotherapy but it can contribute to the evolutionary rescue of secondary tumors whose size might be below the detection threshold of $\sim10^7$  \citep{bozic2013evolutionary}.
+As a result, aneuploidy is unlikely to to contribute to primary tumors overcoming chemotherapy but it can contribute to the evolutionary rescue of secondary tumors whose size might be below the detection threshold of $\sim10^7$  \cite{bozic2013evolutionary}.
 Given the fact that the mean time for small and intermediate tumors to overcome chemotherapy can be of the order of 100 days (see Figure \ref{MeanTimeDeleteriousAneuploidyPlot})  this could explain the reappearance of cancer even after initial remission.
@@
 
-\section*{Figures}
-
-% TODO merge figures 2 and 3 to a single figure here
-%\begin{figure}[h!]
-%\begin{subfigure}{0.5\textwidth}
-%      \centering
-%      \includegraphics[width=\textwidth]{Figures/CombinedSubplot.pdf}      
-%      \label{SurvPlot}
-%\end{subfigure}
-%\begin{subfigure}{0.5\textwidth}
-%      \centering
-%      \includegraphics[width=\textwidth]{Figures/CombinedSubplotLargePop.pdf}      
-%      \label{SurvPlotLargeN}
-%\end{subfigure}
-%\caption{Survival probability for large-effect aneuploidy.}
-%\label{}
-%\end{figure}  
-%%%%%%%%%%%
-\begin{figure}[!h]
- \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/CombinedSubplot.pdf}
-\caption{Plot of the probability of evolutionary rescue of a population, consisting initially of $N$ wildtype cells, as a function of the proliferation rate of the wildtype cells for various values of the proliferation rate of the aneuploid cells.
-The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximation \eqref{aneuploidyresistentfirstapprox} for the upper plot and \eqref{AneuploidyDeleteriousApprox} for the lower plot.
-The dots represent numerical simulations where the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
-Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $N=10^4, \lambda_a=1+10^{-2},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
-}
-\label{SurvPlot}
-\end{figure}
-%%%%%%%%%%%
-\begin{figure}[!h]
- \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/CombinedSubplotLargePop.pdf}
-\caption{Plot of the probability of evolutionary rescue of a population, consisting initially of $N$ wildtype cells, as a function of the proliferation rate of the wildtype cells for various values of the proliferation rate of the aneuploid cells.
-The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximation \eqref{aneuploidyresistentfirstapprox} for the upper plot and \eqref{AneuploidyDeleteriousApprox} for the lower plot.
-The dots represent numerical simulations where the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
-Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $N=10^8, \lambda_w=0.14, \lambda_a=0.14,\lambda_m=0.14,\mu_m=0.13, u=10^{-2}, v=10^{-7}$.
-}
-\label{SurvPlotLargeN}
-\end{figure}
-%%%%%%%%%%%
-%\begin{figure}[!h]
-% \vspace*{1\baselineskip}
-%\includegraphics[width=1\textwidth]{Figures/P_est_growth.pdf}
-%\caption{Plot of the probability of survival of a population as a function of the proliferation rate of the wildtype cells.
-The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximation \eqref{survprobwapprox1}.
-Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $N=10^4, \lambda_a=1+10^{-2},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
-The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
-The error bars are present but are not visible given the fact that we have used $10^5$ simulations for each combination of parameters.}
-%\label{SurvPlot}
-%\end{figure}
-%%%%%%%%%%%
-\begin{figure}[!h]
- \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/SurvPlotNData.pdf}
-\caption{Plot of the probability of survival of a population as a function of the initial population size of wildtype cells.
-The blue line represents the exact solution \eqref{survprobw}, the orange line line represents the approximation \eqref{aneuploidyresistentfirstapprox}, the green line represents the first order correction \eqref{survprobwapproxcorrected} and the red dots represents stochastic simulations.
-For the simulations we have chosen the following parameters: $\lambda_w=0.14, \lambda_a=0.14,\lambda_m=0.14,\mu_w=0.17,\mu_a=0.135,\mu_m=0.13$.
-The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n=100$ is the number of simulations.}
-\label{SurvPlotNData}
-\end{figure}
-%%%%%%%%%%%
-%\begin{figure}[!h]
-% \vspace*{1\baselineskip}
-%\includegraphics[width=1\textwidth]{Figures/P_est.pdf}
-%\caption{Plot of the survival probability of an initial population consisting of $10^{4}$ wildtype cells as a function of $\Delta_a=\lambda_a-\mu_a$ for various values of $\Delta_w=\lambda_a-\mu_a$.
-The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximation \eqref{AneuploidyDeleteriousApprox}.
-The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.}
-%\label{P_est}
-%\end{figure}
-%%%%%%%%%%%
-\begin{figure}[!h]
- \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/P_est_divergence.pdf}
-\caption{Plot of the probability of evolutionary rescue, of an initial population, consisting of $N$ wildtype cells, as a function of proliferation rate of the wildtype cells $\Delta_w=\lambda_w-\mu_w$ for various values of the proliferation rate of the aneuploid cells $\Delta_a=\lambda_a-\mu_a$.
-The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximations \eqref{AneuploidyDeleteriousApprox} and \eqref{pestthirdcase}.
-The dots represent numerical simulations where the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
-The value highlighted in grey is the threshold $\Delta_a^*$ from \eqref{thresholdvalueaneuploid} which marks the transition between the regime dictated by \eqref{AneuploidyDeleteriousApprox} to the one dictated by \eqref{pestthirdcase}.
-Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $N=10^4,\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
-}
-\label{P_est}
-\end{figure}
-%%%%%%%%%%%
-\begin{figure}[!h]
- \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/P_est_divergenceLargePopulation.pdf}
-\caption{Plot of the probability of evolutionary rescue, of an initial population, consisting of $N$ wildtype cells, as a function of proliferation rate of the wildtype cells $\Delta_w=\lambda_w-\mu_w$ for various values of the proliferation rate of the aneuploid cells $\Delta_a=\lambda_a-\mu_a$.
-The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximations \eqref{AneuploidyDeleteriousApprox}.
-The dots represent numerical simulations where the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
-Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $N=10^8,\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
-}
-\label{P_est_large_N}
-\end{figure}
-%%%%%%%%%%%
-\begin{figure}[!h]
- \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/DeleteriousTauLeapPlot.pdf}
-\caption{Plot of the probability of evolutionary rescue of a population, consisting of $N$ wildtype cells, as a function of the initial population size of wildtype cells.
-The blue line represent the exact result \eqref{survprobw} while the orange lines represent the approximation \eqref{pestthirdcase}.
-The green dots represent numerical simulations where the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
-The error bars are present but are not visible given the fact that we have used $n=10^5$ simulations for each combination of parameters.
-Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.}
-\label{DeleteriousPlot}
-\end{figure}
-%%%%%%%%%%%
-\begin{figure}[!h]
- \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/SurvPlotNDataLogisticK.pdf}
-\caption{Plot of the probability of evolutionary rescue of a population, consisting of $N$ wildtype cells, as a function of the initial population size of wildtype cells where maximum population size is constrained by the carrying capacity $K$.
-The blue line represent the exact result \eqref{survprobw} while the orange dots represent numerical simulations where the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
-Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_w=1-10^{-1},\lambda_a=1+10^{-4},\lambda_m=1+10^{-1},\mu_w=1,\mu_m=1,u=10^{-2},v=10^{-7}, C_1=C_2=1, K=10^9$.}
-\label{SurvPlotNDataLogisticKComplete}
-\end{figure}
-%%%%%%%%%%%
-\begin{figure}[!h]
- \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/RatioEvolRescue.pdf}
-\caption{Plot of the ratio of the probability of evolutionary rescue when aneuploidy can play a role in rescue ($u>0$) to the probability where acquisition of aneuploidy is not possible ($u=0$) as a function of the proliferation rate of the wildtype cells.
-The continuous lines represent the exact result \eqref{ratiorescueexact} while the dashed lines represent the approximation \eqref{ratiorescue}.
-The upper plot show the case when aneuploidy is resistant while the lower plot shows the case when it is susceptible.
-Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_w=1-10^{-1},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
-}
-\label{RatioEvolRescue}
-\end{figure}
-%%%%%%%%%%%
-\begin{figure}[!h]
- \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/RatioEvolRescuePopulationSize.pdf}
-\caption{Plot of the ratio of the probability of evolutionary rescue when aneuploidy can play a role in rescue ($u>0$) to the probability where acquisition of aneuploidy is not possible ($u=0$) as a function of the initial population size of wildtype cells.
-The continuous lines represent the exact result \eqref{ratiorescueexact} while the dashed lines represent the approximation \eqref{ratiorescue}.
-The upper plot show the case when aneuploidy is resistant while the lower plot shows the case when it is susceptible.
-Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_w=0.14, \lambda_a=0.14,\lambda_m=0.14,\mu_w=0.17,\mu_m=0.13, u=10^{-2}, v=10^{-7}$.
-}
-\label{RatioEvolRescuePopulationSize}
-\end{figure}
-%%%%%%%%%%%
-\begin{figure}[!h]
- \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/MeanTimeGrowthMutantDirectPlot.pdf}
-\caption{Plot of the mean time until the appearance of a resistance mutation which rescues the population in the case when evolutionary rescue is possible only through mutation but not aneuploidy and mutation.
-The blue line represents the approximation \eqref{meantime1} and the dashed red line represents the first order approximation \eqref{limitapprox}.
-The orange dots represent the numerical simulations while the error bars represent the interval centered at the mean which containing $95\%$ of the simulated values.
-Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_w=1-10^{-1},\lambda_m=1+10^{-1},\mu_w=1,\mu_m=1,u=10^{-2},v=10^{-7}$.}
-\label{MeanTimeGrowthAneuploidyPlot}
-\end{figure}
-%%%%%%%%%%%
-\begin{figure}[!h]
- \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/FractionPlot.pdf}
-\caption{Plot of the fraction of the cases the population is rescue by standing genetic variation as a function of the fraction of initial cells which are aneuploid.
-The red vertical line highlights the the value of $f$ for which half the times the population is rescues by aneuploid cells while the pink line is our approximation \eqref{halfeqstandvar}.
-For this plot we have chosen the following parameters: $N=10^3, \lambda_w=1-10^{-2}, \lambda_a=1-10^{-4},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
-The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
-The value of $0.332$ highlighted in red is the value of the initial fraction of the population which in aneuploid for which half of the cases of evolutionary rescue are due to the initial aneuploid cell population.}
-\label{FractionPlot}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %\section*{References}
 \nolinenumbers
+%\bibliographystyle{agsm}
+%\bibliographystyle{apalike}
 %\bibliographystyle{unsrtnat}
-\bibliographystyle{agsm}
+\bibliographystyle{unsrt}
 \bibliography{evo2022}
@@
 \beginsupplement % https://support.authorea.com/en-us/article/how-to-create-an-appendix-section-or-supplementary-information-1g25i5a/
-\begin{figure}[!h]
+\begin{figure}[!t]
  \vspace*{1\baselineskip}

### 2023-05-31 EvolutionaryRescue1.tex
start merging figures

--- 2023-05-31 EvolutionaryRescue1.tex
@@
 Thus, the rescue probability is given by 
-\begin{equation}
+\begin{equation}\label{rescue_prob}
 p_{rescue}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w} ,
@@
 Applying the approximations for the survival probability $p_w$ from \cref{survprobwapprox1,survprobwinitial,scenario3} and substituting $\Delta_k=\lambda_k-\mu_k$, we find that the rescue probability can be approximated by
-\begin{equation}\label{rescue_prob}
+\begin{equation}\label{rescue_prob_approx}
 \begin{aligned}
@@
 We validate these approximations by comparing them to results of stochastic evolutionary simulations.
-We find that the approximations work very well (\Cref{rescue_prob_wt_growth_small_N,rescue_prob_wt_growth_large_N,rescue_prob_an_growth_small_N,rescue_prob_an_growth_large_N,rescue_prob_N_nongrow,rescue_prob_N_logistic}).
+We find that the approximations work very well (\Cref{rescue_prob_wt_growth_small_large_N,rescue_prob_an_growth_small_N,rescue_prob_an_growth_large_N,rescue_prob_N_nongrow,rescue_prob_N_logistic}).
 
@@
 % B: tolerance
- \vspace*{1\baselineskip}
+\begin{subfigure}{0.5\textwidth}
 \includegraphics[width=1\textwidth]{Figures/CombinedSubplot.pdf}
-\caption{Plot of the probability of evolutionary rescue of a population, consisting initially of $N$ wildtype cells, as a function of the proliferation rate of the wildtype cells for various values of the proliferation rate of the aneuploid cells.
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=1\textwidth]{Figures/CombinedSubplotLargePop.pdf}
+\end{subfigure}
+\caption{
+\textbf{Evolutionary rescue probability increases with wildtype and aneuploid growth rates.}
+Reduce probability is very high when aneuploidy provides partial resistance, in an initially small tumor (\textbf{Aleft}, $N=10^4$) and even more so in an initially large tumor (\textbf{Aright}, $N=10^8$).
+% TODO change panel labels to ABCD
+When aneuploidy provides tolerance (\textbf{Bleft}, $N=10^4$; (\textbf{Bright}, $N=10^8$), the rescue probability is much lower.
+However, in all cases rescue probability increase with both the wildtype growth rate (x-axis) and the aneuploidy growth rate (colors).
+Markers represent simulation results with 95\% CI; solid and dashed lines for the exact formula (\cref{survival_prob} in \cref{rescue_prob}); % TODO do you use 1-(1-pw)^N or 1-e(-Np_w)?
+dashed lines for the approximate formula (\cref{rescue_prob_approx}), demonstrating that they all agree.
+% TODO add parameter valuesEvolutionaryRescue1.tex
+}
+%\caption{Plot of the probability of evolutionary rescue of a population, consisting initially of $N$ wildtype cells, as a function of the proliferation rate of the wildtype cells for various values of the proliferation rate of the aneuploid cells.
 The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximation \eqref{aneuploidyresistentfirstapprox} for the upper plot and \eqref{AneuploidyDeleteriousApprox} for the lower plot.
@@
 }
-\label{rescue_prob_wt_growth_small_N}
-\end{figure}
-%%%%%%%%%%%
-\begin{figure}[p]
-% large population size
-% A: partial resistance
-% B: tolerance
- \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/CombinedSubplotLargePop.pdf}
-\caption{Plot of the probability of evolutionary rescue of a population, consisting initially of $N$ wildtype cells, as a function of the proliferation rate of the wildtype cells for various values of the proliferation rate of the aneuploid cells.
+\label{rescue_prob_wt_growth_small_large_N}
+\end{figure}
+%%%%%%%%%%%
+%\begin{figure}[p]
+%% large population size
+%% A: partial resistance
+%% B: tolerance
+% \vspace*{1\baselineskip}
+%\includegraphics[width=1\textwidth]{Figures/CombinedSubplotLargePop.pdf}
+%\caption{Plot of the probability of evolutionary rescue of a population, consisting initially of $N$ wildtype cells, as a function of the proliferation rate of the wildtype cells for various values of the proliferation rate of the aneuploid cells.
 The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximation \eqref{aneuploidyresistentfirstapprox} for the upper plot and \eqref{AneuploidyDeleteriousApprox} for the lower plot.
@@
 }
-\label{rescue_prob_wt_growth_large_N}
-\end{figure}
+%\label{rescue_prob_wt_growth_large_N}
+%\end{figure}
 %%%%%%%%%%%

### 2023-06-01 EvolutionaryRescue1.tex
continue revising results and figure captions

--- 2023-05-31 EvolutionaryRescue1.tex
@@
 \centering
-\includegraphics[width=0.5\textwidth]{Figures/figureAneuploidy.pdf}
+\includegraphics[width=0.75\textwidth]{Figures/figureAneuploidy.pdf}
 \caption{
@@
 % TODO in the figure change labels to Wildtype, Aneuploid, Mutant (instead of Euploid etc)
-A population of cancer cells is composed of wildtype, aneuploid, and mutant cells, which divide with rates $\lambda_w$, $\lambda_a$, and $\lambda_m$ and die at rates $\mu_w$, $\mu_a$, and $\mu_m$, respectively.
+\textbf{(A)} A population of cancer cells is composed of wildtype, aneuploid, and mutant cells, which divide with rates $\lambda_w$, $\lambda_a$, and $\lambda_m$ and die at rates $\mu_w$, $\mu_a$, and $\mu_m$, respectively.
 Wildtype cells can become aneuploid at rate $u$.
@@
 Color denotes the relative growth rates of the three genotypes such that $\lambda_w - \mu_w < \lambda_a - \mu_a < \lambda_m - \mu_m$.
+\textbf{(B)} The wildtype and the mutant are susceptible and resistant, respectively, to the drug.
+The aneuploid may be tolerant, non-growing, or or partially resistant.
 }
@@
 We validate these approximations by comparing them to results of stochastic evolutionary simulations.
-We find that the approximations work very well (\Cref{rescue_prob_wt_growth_small_N,rescue_prob_wt_growth_large_N,rescue_prob_an_growth_small_N,rescue_prob_an_growth_large_N,rescue_prob_N_nongrow,rescue_prob_N_logistic}).
+We find that the approximations work very well (\Cref{rescue_prob_wt_growth,rescue_prob_an_growth,rescue_prob_N_nongrow,rescue_prob_N_logistic}).
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\paragraph*{Density-dependent growth}
-
-We perform stochastic simulations for different values of the carrying capacity $K$ and we plot the results in Figure \ref{SurvPlotNDataLogisticKComplete}.
-We observe that as $K$ increases the simulations converge to the analytic result which is because the carrying capacity is much larger then the population size of aneuploid cells for which the probability that the population is rescued is certain.
+\paragraph*{Density-dependent growth.}
+
+In our analysis we used branching processes, which assume that growth (division and death) are density-independent.
+However, in some cases growth may be limited by resources (oxygen, nutrients, etc.) and therefore depend on cell density.
+We therefore performed stochastic simulations of a logistic growth model with carrying capacity $K$ (Methods). 
+We find that ... (\Cref{rescue_prob_N_logistic}). % TODO check what we find; the figure just shows agreement of exact and stochastic
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Standing genetic variation}
-
-So far we have assumed that the initial population of cells consisted entirely of wildtype cells.
-We now modify this assumption so that the initial population includes a fraction $f$ of cells with aneuploidy.
-The probability of evolutionary rescue by cells with aneuploidy from the initial population is
+\paragraph*{Standing genetic variation}
+
+In the above we assumed that upon beginning of drug therapy, the initial tumor consisted entirely of wildtype cells.
+However, aneuploid cells are likely generated even before onset of treatment at some rate $\tilde{u} \le u$ (because the treatment itself may promote generation of aneuploid cells REF), which are likely to have a deleterious effect (REF).
+But if the number of cells in the tumor $N$ is large, as expected if drug treatment is applied, there may already be a fraction $f=\tilde{u}/s$ of aneuploid cells in the population, where $s$ is the cost of aneuploidy (REF).
+
+In this scenario, The probability of evolutionary rescue by cells with aneuploidy from the initial population is
 \begin{equation*}
-p_{old} = 1-\left(1-p_a\right)^{fN}\approx 1-\e^{-fNp_a}.
+% redefine: p_total, p_denovo=p_resuce from eq above; p_stdvar is the new thing
+p_{old} = 1-\left(1-p_a\right)^{fN}\approx 1-\e^{-fNp_a}. % TODO should be p_new?
 \end{equation*}
@@
 \begin{align}\nonumber
-p_{total} 	&= p_{new}+\left(1-p_{new}\right)p_{old}\\
+p_{total} 	&= p_{new}+\left(1-p_{new}\right)p_{old}\\ % TODO what is p_new?
 			&= 1-\exp\left(-\left[\left(1-f\right)p_w + fp_a\right]N\right) .
@@
 Setting $F=\frac{1}{2}$, we use the expansion $\e^x \approx 1+x$ to obtain
-\begin{equation}\label{halfeqstandvar}
+\begin{equation}\label{halfeqstandvar} % TODO not sure this is interesting
 f^*\approx\frac{p_w}{p_w+p_a}.
 \end{equation}
+% TODO maybe better to show p_total / p_denovo as function of u or N.
 See \Cref{FractionPlot} for a demonstration of $F$ and $f*$.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Contribution of aneuploidy to the evolutionary rescue of cancer}
-We wish to understand the contribution of aneuploidy to evolutionary rescue of the cancer cell population.
-For this purpose we define the ratio of the probability of evolutionary rescue when aneuploidy can play a role in rescue ($u>0$) to the probability where acquisition of aneuploidy is not possible ($u=0$):
+\subsection*{Effect of aneuploidy on evolutionary rescue}
+To determine the extent to which aneuploidy may affect evolutionary rescue, we define $H$ to be the ratio of the rescue probability with aneuploidy ($u>0$) and the rescue probability without aneuploidy ($u=0$),
 \begin{equation}\label{ratiorescueexact}
-H=\frac{\left.p_{rescue}\right|_{u>0}}{\left.p_{rescue}\right|_{u=0}}.
-\end{equation}
-As a result, we obtain from \eqref{rescue_prob} the approximation for the ratio:
+H = \frac{p_{rescue}(u>0)}{p_{rescue}(u=0)}.
+\end{equation}
+
+Plugging in our approximations from \cref{rescue_prob}, we have
 \begin{equation}\label{ratiorescue}
-H\sim\left\{
-  \begin{array}{@{}ll@{}}
-  \frac{1-\exp\left[\frac{N}{\Delta_w-u-v}\left(v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right)\right]}{1-\exp\left[\frac{vN\Delta_m}{\left(\Delta_w-v\right)\lambda_m}\right]},\quad\text{if }4\lambda_avp_m>\left(\Delta_a-v\right)^2,\\
-   \frac{1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right]}{1-\exp\left(\frac{v\Delta_mN}{\lambda_m\Delta_w}\right)},\quad\text{if }\Delta_a<0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2,\\
-   \frac{1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right]}{1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\right]},\quad\text{if }\Delta_a>0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2.
-% TODO this term can be simplified/approximated
-  \end{array}\right.
-\end{equation}
-
-%\begin{equation}
-%H\sim\left\{
-%  \begin{array}{@{}ll@{}}
-%  \frac{1-\exp\left[\frac{N}{\Delta_w-u-v}\left(v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right)\right]}{1-\exp\left[\frac{vN\Delta_m}{\left(\Delta_w-v\right)\lambda_m}\right]},\quad\text{if }4\lambda_avp_m>\left(\Delta_a-v\right)^2,\\
-%   \frac{1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right]}{1-\exp\left(\frac{v\Delta_mN}{\lambda_m\Delta_w}\right)},\quad\text{if }\Delta_a<0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2,\\
-%   \frac{1-\exp\left(\frac{u\Delta_aN}{\lambda_a\Delta_w}\right)}{1-\exp\left(\frac{v\Delta_mN}{\lambda_m\Delta_w}\right)},\quad\text{if }\Delta_a>0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2.
-%  \end{array}\right.
-%\end{equation}
-We plot \eqref{ratiorescue} in Figure \ref{RatioEvolRescue} for both resistent and susceptible aneuploidy as a function of the proliferation rate of the wildtype cells and in Figure \ref{RatioEvolRescuePopulationSize} as a function of the initial population size of wildtype cells.
+H = \begin{cases}
+\frac{1-\exp\left[\frac{N}{\Delta_w-u-v}\left(v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right)\right]}{1-\exp\left[\frac{vN\Delta_m}{\left(\Delta_w-v\right)\lambda_m}\right]} ,&
+4\lambda_avp_m>\left(\Delta_a-v\right)^2 ,\\
+\frac{1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right]}{1-\exp\left(\frac{v\Delta_mN}{\lambda_m\Delta_w}\right)}, ,&
+\Delta_a<0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2 ,\\
+\frac{1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right]}{1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\right]} ,&
+\Delta_a>0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2. % TODO this term can be simplified?
+  \end{cases}
+\end{equation}
+
+We find that the rescue ratio increase with the aneuploidy growth rate $\Delta_a$, because the better aneuploid cells are in growth, the better they are at rescuing the population (when they provide partial resistance) or delaying the extinction of the population (when they provide tolerance).
+However, the rescue decreases with the wildtype growth rate $\Delta_w$, because the better the wildtype is at growth, the less is depends on aneuploidy for rescue or delay, and the more likely it is to directly produce mutant cells, rather than relying on aneuploid cells for producing mutant cells (\Cref{RatioEvolRescue}).
+The effect of the initial tumor size $N$ is mostly the same as the wildtype growth rate, except when aneuploidy provides tolerance and the population size is very large, in which case the ratio ...
+% TODO does H go to 1 in this case?
+is it lower than 1 for delta_a=-0.001?
+why?
 
@@
 
-% TODO merge figures 2 and 3 to a single figure here
-%\begin{figure}[h!]
-%\begin{subfigure}{0.5\textwidth}
-%      \centering
-%      \includegraphics[width=\textwidth]{Figures/CombinedSubplot.pdf}      
-%      \label{SurvPlot}
-%\end{subfigure}
-%\begin{subfigure}{0.5\textwidth}
-%      \centering
-%      \includegraphics[width=\textwidth]{Figures/CombinedSubplotLargePop.pdf}      
-%      \label{SurvPlotLargeN}
-%\end{subfigure}
-%\caption{Survival probability for large-effect aneuploidy.}
-%\label{}
-%\end{figure}  
-%%%%%%%%%%%
 \begin{figure}[p]
@@
 % B: tolerance
- \vspace*{1\baselineskip}
+\begin{subfigure}{0.5\textwidth}
 \includegraphics[width=1\textwidth]{Figures/CombinedSubplot.pdf}
-\caption{Plot of the probability of evolutionary rescue of a population, consisting initially of $N$ wildtype cells, as a function of the proliferation rate of the wildtype cells for various values of the proliferation rate of the aneuploid cells.
-The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximation \eqref{aneuploidyresistentfirstapprox} for the upper plot and \eqref{AneuploidyDeleteriousApprox} for the lower plot.
-The dots represent numerical simulations where the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
-Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $N=10^4, \lambda_a=1+10^{-2},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=1\textwidth]{Figures/CombinedSubplotLargePop.pdf}
+\end{subfigure}
+\caption{
+\textbf{Evolutionary rescue probability with partially resistant or tolerant aneuploid cells.}
+Rescue probability is very high when aneuploidy provides partial resistance, in an initially small tumor (\textbf{Aleft}, $N=10^4$) and even more so in an initially large tumor (\textbf{Aright}, $N=10^8$).
+% TODO change panel labels to ABCD
+When aneuploidy provides tolerance (\textbf{Bleft}, $N=10^4$; (\textbf{Bright}, $N=10^8$), the rescue probability is much lower. 
+In both scenarios, rescue probability increase with both the wildtype growth rate (x-axis) and the aneuploidy growth rate (colors).
+Markers represent simulation results with 95\% CI; solid and dashed lines for the exact formula (\cref{survival_prob} in \cref{rescue_prob}); % TODO do you use 1-(1-pw)^N or 1-e(-Np_w)?
+dashed lines for the approximate formula (\cref{rescue_prob_approx}), demonstrating that they all agree.
+% TODO add parameter values
 }
-\label{rescue_prob_wt_growth_small_N}
-\end{figure}
-%%%%%%%%%%%
-\begin{figure}[p]
-% large population size
-% A: partial resistance
-% B: tolerance
- \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/CombinedSubplotLargePop.pdf}
-\caption{Plot of the probability of evolutionary rescue of a population, consisting initially of $N$ wildtype cells, as a function of the proliferation rate of the wildtype cells for various values of the proliferation rate of the aneuploid cells.
-The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximation \eqref{aneuploidyresistentfirstapprox} for the upper plot and \eqref{AneuploidyDeleteriousApprox} for the lower plot.
-The dots represent numerical simulations where the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
-Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $N=10^8, \lambda_w=0.14, \lambda_a=0.14,\lambda_m=0.14,\mu_m=0.13, u=10^{-2}, v=10^{-7}$.
-}
-\label{rescue_prob_wt_growth_large_N}
+\label{rescue_prob_wt_growth}
 \end{figure}
@@
 % switch from tolerance to non-growing
- \vspace*{1\baselineskip}
+\begin{subfigure}{0.5\textwidth}
 \includegraphics[width=1\textwidth]{Figures/P_est_divergence.pdf}
-\caption{Plot of the probability of evolutionary rescue, of an initial population, consisting of $N$ wildtype cells, as a function of proliferation rate of the wildtype cells $\Delta_w=\lambda_w-\mu_w$ for various values of the proliferation rate of the aneuploid cells $\Delta_a=\lambda_a-\mu_a$.
-The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximations \eqref{AneuploidyDeleteriousApprox} and \eqref{pestthirdcase}.
-The dots represent numerical simulations where the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
-The value highlighted in grey is the threshold $\Delta_a^*$ from \eqref{thresholdvalueaneuploid} which marks the transition between the regime dictated by \eqref{AneuploidyDeleteriousApprox} to the one dictated by \eqref{pestthirdcase}.
-Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $N=10^4,\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=1\textwidth]{Figures/P_est_divergenceLargePopulation.pdf}
+\end{subfigure}
+\caption{\textbf{Evolutionary rescue probability with tolerant or non-growing aneuploid cells.}
+Rescue probability grows with the aneuploid growth rate $\Delta_a$ (x-axis), and is much higher in an initially large tumor than in a small one (\textbf(A) $N=10^4$; \textbf{(B)} $N=10^8$). 
+Markers represent simulation results with 95\% CI; solid and dashed lines for the exact formula (\cref{survival_prob} in \cref{rescue_prob}); % TODO do you use 1-(1-pw)^N or 1-e(-Np_w)?
+dashed lines for the approximate formula (\cref{rescue_prob_approx}).
+The approximation agrees with the simulation and exact solution when the initial tumor size is large (panel B).
+When the tumor size is small (panel A), we switch between the approximation for tolerant and for non-growing aneuploid cells; the switch occurs at $\Delta_a^*=2vp_m+v+2\sqrt{vp_m\left(vp_m+\mu_a+v\right)}$.
+% TODO why no switch in panel B?
+% TODO add parameter values
+% TODO more simulations for 0 > Delta_a > Delta^*_a
 }
-\label{rescue_prob_an_growth_small_N}
-\end{figure}
-%%%%%%%%%%%
-\begin{figure}[p]
-% large population size
-% switch from tolerance to non-growing
- \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/P_est_divergenceLargePopulation.pdf}
-\caption{Plot of the probability of evolutionary rescue, of an initial population, consisting of $N$ wildtype cells, as a function of proliferation rate of the wildtype cells $\Delta_w=\lambda_w-\mu_w$ for various values of the proliferation rate of the aneuploid cells $\Delta_a=\lambda_a-\mu_a$.
-The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the approximations \eqref{AneuploidyDeleteriousApprox}.
-The dots represent numerical simulations where the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
-Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $N=10^8,\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
-}
-\label{rescue_prob_an_growth_large_N}
+\label{rescue_prob_an_growth}
 \end{figure}
@@
 \begin{figure}[p]
-% A partial resistance
-% B tolerance
-% function of wildtype growth rate
+% slightly decreasing (need to check)
  \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/RatioEvolRescue.pdf}
-\caption{Plot of the ratio of the probability of evolutionary rescue when aneuploidy can play a role in rescue ($u>0$) to the probability where acquisition of aneuploidy is not possible ($u=0$) as a function of the proliferation rate of the wildtype cells.
-The continuous lines represent the exact result \eqref{ratiorescueexact} while the dashed lines represent the approximation \eqref{ratiorescue}.
-The upper plot show the case when aneuploidy is resistant while the lower plot shows the case when it is susceptible.
-Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_w=1-10^{-1},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
+\includegraphics[width=1\textwidth]{Figures/FractionPlot.pdf}
+\caption{\textbf{Effect of standing variation on evolutionary rescue.}
+In aneuploid cells already exist in the population at the onset of drug therapy as standing genetic variation, then evolutionary rescue is more likely... % TODO think about this figure...
+%Plot of the fraction of the cases the population is rescue by standing genetic variation as a function of the fraction of initial cells which are aneuploid.
+The red vertical line highlights the the value of $f$ for which half the times the population is rescues by aneuploid cells while the pink line is our approximation \eqref{halfeqstandvar}.
+For this plot we have chosen the following parameters: $N=10^3, \lambda_w=1-10^{-2}, \lambda_a=1-10^{-4},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
+The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
+The value of $0.332$ highlighted in red is the value of the initial fraction of the population which in aneuploid for which half of the cases of evolutionary rescue are due to the initial aneuploid cell population.
 }
-\label{RatioEvolRescue}
+\label{FractionPlot}
 \end{figure}
@@
 % B tolerance
-% function of population size
- \vspace*{1\baselineskip}
+% left function of wildtype growth rate 
+% right function of population size
+% TODO fix panel labels
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=1\textwidth]{Figures/RatioEvolRescue.pdf}
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
 \includegraphics[width=1\textwidth]{Figures/RatioEvolRescuePopulationSize.pdf}
-\caption{Plot of the ratio of the probability of evolutionary rescue when aneuploidy can play a role in rescue ($u>0$) to the probability where acquisition of aneuploidy is not possible ($u=0$) as a function of the initial population size of wildtype cells.
+\end{subfigure}
+\caption{\textbf{Effect of aneuploidy on evolutionary rescue.}
+The ratio of rescue probability with and without aneuploid ($H$, \cref{ratiorescue}) increases with the aneuploid growth rate (colors) and decreases with the wildtype growth rates and initial tumor size (x-axis), except for large tumors where aneuploid cells are tolerant (see right side of panel B-right).
+\textbf{(A-left,A-right)} Aneuploidy provides partial resistance. \textbf{(B-left, B-right)} Aneuploidy provides tolerance.  
+Solid and dashed lines apply $p_{rescue}$ from the exact formula of  (\cref{survival_prob} in \cref{rescue_prob}); % TODO do you use 1-(1-pw)^N or 1-e(-Np_w)?
+dashed lines apply $p_{rescue}$ from the approximate formula (\cref{rescue_prob_approx}), with good agreement.
+% TODO parameter values
+}
+\label{rescue_ratio}
+\end{figure}
+%%%%%%%%%%%
+%\begin{figure}[p]
+%% A partial resistance
+%% B tolerance
+%% function of population size
+% \vspace*{1\baselineskip}
+%\caption{Plot of the ratio of the probability of evolutionary rescue when aneuploidy can play a role in rescue ($u>0$) to the probability where acquisition of aneuploidy is not possible ($u=0$) as a function of the initial population size of wildtype cells.
 The continuous lines represent the exact result \eqref{ratiorescueexact} while the dashed lines represent the approximation \eqref{ratiorescue}.
@@
 }
-\label{RatioEvolRescuePopulationSize}
-\end{figure}
+%\label{RatioEvolRescuePopulationSize}
+%\end{figure}
 %%%%%%%%%%%
@@
 \label{MeanTimeGrowthAneuploidyPlot}
-\end{figure}
-%%%%%%%%%%%
-\begin{figure}[p]
-% slightly decreasing (need to check)
- \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/FractionPlot.pdf}
-\caption{Plot of the fraction of the cases the population is rescue by standing genetic variation as a function of the fraction of initial cells which are aneuploid.
-The red vertical line highlights the the value of $f$ for which half the times the population is rescues by aneuploid cells while the pink line is our approximation \eqref{halfeqstandvar}.
-For this plot we have chosen the following parameters: $N=10^3, \lambda_w=1-10^{-2}, \lambda_a=1-10^{-4},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
-The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
-The value of $0.332$ highlighted in red is the value of the initial fraction of the population which in aneuploid for which half of the cases of evolutionary rescue are due to the initial aneuploid cell population.}
-\label{FractionPlot}
 \end{figure}

### 2023-06-01 EvolutionaryRescue1.tex
started on mean time text

@@
 We find that the rescue ratio increase with the aneuploidy growth rate $\Delta_a$, because the better aneuploid cells are in growth, the better they are at rescuing the population (when they provide partial resistance) or delaying the extinction of the population (when they provide tolerance).
-However, the rescue decreases with the wildtype growth rate $\Delta_w$, because the better the wildtype is at growth, the less is depends on aneuploidy for rescue or delay, and the more likely it is to directly produce mutant cells, rather than relying on aneuploid cells for producing mutant cells (\Cref{RatioEvolRescue}).
+However, the rescue decreases with the wildtype growth rate $\Delta_w$, because the better the wildtype is at growth, the less is depends on aneuploidy for rescue or delay, and the more likely it is to directly produce mutant cells, rather than relying on aneuploid cells for producing mutant cells (\Cref{rescue_ratio}).
 The effect of the initial tumor size $N$ is mostly the same as the wildtype growth rate, except when aneuploidy provides tolerance and the population size is very large, in which case the ratio ...
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Rescue time}
-We calculate the mean time for the appearance of the first mutant that rescues the cancer cell population.
-This can occur either through the pathway $wildtype\rightarrow aneuploid \rightarrow mutant$ or through the pathway $wildtype \rightarrow mutant$.
-We start with the second pathway: let $T_1$ be the time at which the first mutant cell appears which rescues the population when evolutionary rescue is only possible through mutation.
-We are interested in the mean time $\tau_1=\mathbb{E}\left[T_1\right]$.
-
-% something is missing in the next line...?
-% the entire section is missing text...
-
-The number of successful mutants generated until time $t$ can be approximated by a inhomogeneous Poisson process with rate $R\left(t\right)=up_aw_t$
-%\begin{align}
-%\tau_1\sim \text{Poisson}\left(\frac{1}{up_aw_t}\right),
-%\end{align}
-where $w_t$ is the size of the wildtype population at time $t$: 
+\subsection*{Evolutionary rescue time}
+
+Even evolutionary rescue occurs, it may take a long time; therefore, it is crucial to estimate the mean waiting time for rescue and the effect aneuploidy may have on it. 
+We therefore calculate the mean time for the appearance of the first mutant that rescues the cell population.
+This can occur either through the evolutionary trajectory $wildtype \rightarrow mutant$ or through the trajectory $wildtype \rightarrow aneuploid \rightarrow mutant$.
+We start with the former.
+
+Assuming no aneuploidy ($u=0$), we define $T_1$ to be the time at which the first mutant cell appears that will avoid extinction and will therefore rescue the population.
+Note that if extinction occurs, that is $m_{t\rightarrow\infty}=0$, then it is implied that $T_1=\infty$, and vice versa if $T_1<\infty$ then $m_{t\rightarrow\infty}\neq0$.
+
+The number of successful mutants generated until time $t$ can be approximated by an inhomogeneous Poisson process with rate $R\left(t\right) = u p_a w_t$
+where $w_t$ is the number of wildtype cells at time $t$, $w_t=N\e^{\Delta_w t}$.
+The probability density function of $T_1$ is thus
+$R\left(t\right)\e^{-\int_0^{t}{R(z)\d z}}$.
+Therefore, the probability density function of $(T_1 \mid T_1 < \infty)$ is
+$f_1(t) = \frac{R\left(t\right)\e^{-\int_0^{t} R(z)\d z}}{p_{rescue}}$. % Todo is this correct? I made changes
+
+We are interested in the mean conditional time, $\tau_1=\mathbb{E}\left[T_1 \mid T_1<\infty\right]$, which is given by
+\begin{align}\label{meantime1}
+\tau_1 =
+\int_{0}^{\infty}{t f_1(t) \d t} = 
+\frac{\int_{0}^{\infty}{t u p_a N\e^{\Delta_w t} \e^{-\int_0^{t} u p_a N\e^{\Delta_w z} \d z} \d t}}{p_{rescue}} .
+\end{align}
 \begin{equation}
-w_t=N\e^{\Delta_wt}.
-\end{equation}
-We are interested in the time to appearance of the first successful mutant cell conditional on population surviving:
-\begin{align*}
-P\left(T_1<t\right)&=P\left(T_1<t|m_{t\rightarrow\infty}\neq0\right)P\left(m_{t\rightarrow\infty}\neq0\right)\\
-&+P\left(T_1<t|m_{t\rightarrow\infty}=0\right)P\left(m_{t\rightarrow\infty}=0\right).
-\end{align*}
-As a result, the cumulative distribution function can be written as:
-\begin{align*}
-P\left(T_1<t\right)&=P\left(T_1<t|m_{t\rightarrow\infty}\neq0\right)P\left(m_{t\rightarrow\infty}\neq0\right),
-\end{align*}
-where we used the fact that evolutionary rescue is impossible when the mutant population is destined to be zero:
-\begin{align*}
-P\left(T_1<t|m_{t\rightarrow\infty}=0\right)=0.
-\end{align*}
-As a result, we obtain 
-\begin{align}
-P\left(T_1<t|m_{t\rightarrow\infty}\neq0\right)=\frac{P\left(\tau_1<t\right)}{1-\left(1-p_w\right)^N},
-\end{align}
-where we have used
-\begin{equation}
-P\left(m_{t\rightarrow\infty}\neq0\right)=1-\left(1-p_w\right)^N.
-\end{equation}
-The probability density function of $T_1$ is given by:
-\begin{align}
-f_{T_1}\left(t_1\right)=R\left(t_1\right)\e^{-\int_0^{t_1} R(t)\d t}.
-\end{align}
-As a result, the time $T_1$ conditional on evolutionary rescue is given by:
-\begin{align}
-f_{T_1}\left(t_1|m_{t\rightarrow\infty}\neq0\right)=\frac{R\left(t_1\right)\e^{-\int_0^{t_1} R(t)\d t}}{1-\left(1-p_w\right)^N}.
-\end{align}
-The expectation of $T_1$ is:
-\begin{align}\label{meantime1}
-\tau_1=\mathbb{E}\left[T_1\right]&=\frac{\int_0^\infty\e^{-\int_0^\tau R(t)\d t}\d\tau}{1-\left(1-p_w\right)^N}=\frac{\int_0^\infty\e^{-uNp_a\frac{\e^{\Delta_w\tau}-1}{\Delta_w}}\,\d\tau}{1-\left(1-p_w\right)^N}.
-\end{align}
-The fration in exponential of the integrand in \eqref{meantime1} can approximated as:
+\frac{\int_0^\infty\e^{-\int_0^t R(z)\d z}\d t}{1-\left(1-p_w\right)^N}=\frac{\int_0^\infty\e^{-uNp_a\frac{\e^{\Delta_w t}-1}{\Delta_w}}\,\d t}{1-\left(1-p_w\right)^N}
+\end{equation}
+% YOAV HERE - dont understand how the previous two lines are the same...
+
+The fraction in exponential of the integrand in \eqref{meantime1} can approximated as:
 \begin{equation*}
-\frac{\e^{\Delta_w\tau}-1}{\Delta_w}=\frac{1+\Delta_w\tau+O(\tau^2)-1}{\Delta_w}=\tau+O(\tau^2).
+\frac{\e^{\Delta_w t}-1}{\Delta_w}=\frac{1+\Delta_w t+O(t^2)-1}{\Delta_w}=t+O(t^2).
 \end{equation*}

### 2023-06-04 EvolutionaryRescue1.tex
going over todos with RS and started revising mean time

@@
 \renewcommand\linenumberfont{\normalfont\small\sffamily}
-%\linenumbers
+\linenumbers
 \modulolinenumbers[2]
@@
 \section*{Introduction}
-
-% TODO remove the commented out paragraphs if you dont need them. OK
 
@@
 \textbf{Model illustration.}
-% TODO in the figure change labels to Wildtype, Aneuploid, Mutant (instead of Euploid etc)
 \textbf{(A)} A population of cancer cells is composed of wildtype, aneuploid, and mutant cells, which divide with rates $\lambda_w$, $\lambda_a$, and $\lambda_m$ and die at rates $\mu_w$, $\mu_a$, and $\mu_m$, respectively.
@@
 Finally, we update the number of cells of each type according to the event that occurred and update the time from $t$ to $t+\Delta t$.
-% TODO more details - you need to specify the events and how they change the number of cells.
-OK
 We repeat these iterations until either the population becomes extinct (the number of cells of all types is zero) or the number of mutant cells is high enough so that its extinction probability is $<0.1\%$, that is until
@@
 \paragraph{$\tau$-leaping.}
-% TODO "is large" - how large?
-When the size of the initial population is large we utilize $\tau$-leaping \citep{gillespie2001approximate}, where change in number of cells of genotype $i$ in a fixed time interval $\Delta t$ is Poisson distributed with mean $\nu_i\Delta t$.
-If the increment is negative and larger then the subpopulation size then updated to be zero. % TODO "and larger..."??
+When simulations are slow (e.g.
+due to large population size), we utilize $\tau$-leaping \citep{gillespie2001approximate}, where change in number of cells of genotype $i$ in a fixed time interval $\Delta t$ is Poisson distributed with mean $\nu_i\Delta t$.
+If the change in number of cells is negative and larger then the subpopulation size then the subpopulation size is updated to be zero.
 
@@
 \end{align*}
-where $C_1, C_2>0$ are constants. % TODO what constants?
+where $C_1, C_2>0$ are constants. % TODO what constants? see Remus version.
 
@@
 We rewrite \cref{survival_prob} as
-\begin{align*}
+\begin{equation}
+\begin{aligned}
 p_w&=\frac{\lambda_w-\mu_w-u-v}{2\lambda_w}\left(1-\sqrt{1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}}\right) ,
@@
 p_a&=\frac{\lambda_a-\mu_a-v}{2\lambda_a}\left(1-\sqrt{1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}}\right) .
-\end{align*}
-Since $u,v\ll1$, this can be approximated by % TODO how do you do the first approx? taylor?
+\end{aligned}
+\end{equation}
+Since $u,v\ll1$, the term in the root can be approximated using a 1st-order Taylor expansion. So, substituting the expressions for $p_a$ and $p_m$, we have
 \begin{equation}\label{survprobwinitial}
@@
 \end{equation}
-%Using the notational convention
-%\begin{equation}\label{notationalconv}
-%\Delta_i=\lambda_i-\mu_i,
-%\end{equation}
-%we write \eqref{survprobw} as
-%\begin{equation}\label{survprobwapprox2}
-%p_w=\frac{v\Delta_m}{\lambda_m\Delta_w}\left(\frac{u}{\Delta_a}-1\right).
-%\end{equation}
 
@@
 \end{equation}
-From \Cref{survprobwinitial}, % TODO I dont see how the referred equation plays a part here
-the survival probability of a population starting from one wildtype individual is
+Plugging this in \cref{survprobwinitial}, the survival probability of a population starting from one wildtype individual is
 \begin{equation}\label{scenario3}
@@
 Thus, the rescue probability is given by 
-\begin{equation}\label{rescue_prob}
-p_{rescue}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w} ,
-\end{equation}
+\begin{subequations}
+\begin{align}
+p_{rescue} &= 1-\left(1-p_w\right)^N \label{rescue_prob} \\
+& \approx 1-\e^{-Np_w} , \label{rescue_prob2}
+\end{align}
+\end{subequations}
 where the approximation $(1-p_w)\approx e^{-p_w}$ assumes that $p_w$ (but not $N p_w$) is small.
 
-Applying the approximations for the survival probability $p_w$ from \cref{survprobwapprox1,survprobwinitial,scenario3} and substituting $\Delta_k=\lambda_k-\mu_k$, we find that the rescue probability can be approximated by
+Applying the approximations for the survival probability $p_w$ from \cref{survprobwapprox1,survprobwinitial,scenario3} in \cref{rescue_prob2} and substituting $\Delta_k=\lambda_k-\mu_k$, we find that the rescue probability can be approximated by
 \begin{equation}\label{rescue_prob_approx}
@@
 
-In \Cref{SurvPlot,SurvPlotLargeN} we explore the effects of the wildtype and aneuploid growth rates on the rescue probability for small and large population sizes ($N=10^4$ and $N=10^8$, respectively).
-
-\Cref{SurvPlotNData} show $p_{rescue}$ as a function of $N$, including comparison of our first approximation \eqref{aneuploidyresistentfirstapprox} and simulation results.
-
-\Cref{P_est,P_est_large_N} show the rescue probability for initial population sizes $N=10^4$ and $N=10^8$, respectively. % TODO check figure refs
-
-In \Cref{DeleteriousPlot}, we compare the exact result (\cref{survival_prob}) with numerical simulations.
-The transition between the regimes defined by \cref{AneuploidyDeleteriousApprox} and \cref{pestthirdcase} respectively occurs at:
-\begin{align}\label{thresholdvalueaneuploid}
-\Delta_a^*=2vp_m+v+2\sqrt{vp_m\left(vp_m+\mu_a+v\right)}.
-\end{align}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 In our analysis we used branching processes, which assume that growth (division and death) are density-independent.
-However, in some cases growth may be limited by resources (oxygen, nutrients, etc.) and therefore depend on cell density.
+However, growth may be limited by resources (oxygen, nutrients, etc.) and therefore depend on cell density.
 We therefore performed stochastic simulations of a logistic growth model with carrying capacity $K$ (Methods). 
-We find that ... (\Cref{rescue_prob_N_logistic}). % TODO check what we find; the figure just shows agreement of exact and stochastic
+We find that our approximations agree with results of simulations with density-dependent growth for biologically relevant parameter values (\Cref{rescue_prob_N_logistic}).
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\paragraph*{Standing genetic variation}
+\paragraph*{Standing vs. de-novo genetic variation}
 
@@
 However, aneuploid cells are likely generated even before onset of treatment at some rate $\tilde{u} \le u$ (because the treatment itself may promote generation of aneuploid cells REF), which are likely to have a deleterious effect (REF).
+% TODO refs
 But if the number of cells in the tumor $N$ is large, as expected if drug treatment is applied, there may already be a fraction $f=\tilde{u}/s$ of aneuploid cells in the population, where $s$ is the cost of aneuploidy (REF).
 
-In this scenario, The probability of evolutionary rescue by cells with aneuploidy from the initial population is
+In this scenario, the probability of evolutionary rescue by cells with aneuploidy from the initial population is
 \begin{equation*}
 % redefine: p_total, p_denovo=p_resuce from eq above; p_stdvar is the new thing
-p_{old} = 1-\left(1-p_a\right)^{fN}\approx 1-\e^{-fNp_a}. % TODO should be p_new?
+p_{sgv} = 1-\left(1-p_a\right)^{fN}\approx 1-\e^{-fNp_a}. 
 \end{equation*}
@@
 \begin{align}\nonumber
-p_{total} 	&= p_{new}+\left(1-p_{new}\right)p_{old}\\ % TODO what is p_new?
+p_{rescue} 	&= p_{sgv}+\left(1-p_{sgv}\right)p_{de-novo} \\
 			&= 1-\exp\left(-\left[\left(1-f\right)p_w + fp_a\right]N\right) .
@@
 
-The fraction of cases in which the population is rescued by the standing genetic variation is given by $F\left(f\right)=\frac{p_{old}}{p_{total}}$.
-%=\frac{ 1-\e^{-fNp_a}}{1-\e^{-\left[\left(1-f\right)p_w+fp_a\right]N}}.
-Setting $F=\frac{1}{2}$, we use the expansion $\e^x \approx 1+x$ to obtain
-\begin{equation}\label{halfeqstandvar} % TODO not sure this is interesting
-f^*\approx\frac{p_w}{p_w+p_a}.
-\end{equation}
-% TODO maybe better to show p_total / p_denovo as function of u or N.
-See \Cref{FractionPlot} for a demonstration of $F$ and $f*$.
+The fraction of cases in which the population is rescued by pre-existing aneuploid cells (i.e., standing genetic variation) is given by $F\left(f\right)=\frac{p_{sgv}}{p_{total}}$ (\Cref{FractionPlot}).
 
@@
 However, the rescue decreases with the wildtype growth rate $\Delta_w$, because the better the wildtype is at growth, the less is depends on aneuploidy for rescue or delay, and the more likely it is to directly produce mutant cells, rather than relying on aneuploid cells for producing mutant cells (\Cref{rescue_ratio}).
-The effect of the initial tumor size $N$ is mostly the same as the wildtype growth rate, except when aneuploidy provides tolerance and the population size is very large, in which case the ratio ...
-% TODO does H go to 1 in this case?
-is it lower than 1 for delta_a=-0.001?
-why?
+The effect of the initial tumor size $N$ is the similar to that of the wildtype growth rate. 
+Importantly, in large tumors, the ratio converges to unity, that is, aneuploidy does not affect the probability for evolutionary rescue. % TODO can we find this analytical?
 
@@
 Assuming no aneuploidy ($u=0$), we define $T_1$ to be the time at which the first mutant cell appears that will avoid extinction and will therefore rescue the population.
-Note that if extinction occurs, that is $m_{t\rightarrow\infty}=0$, then it is implied that $T_1=\infty$, and vice versa if $T_1<\infty$ then $m_{t\rightarrow\infty}\neq0$.
-
-The number of successful mutants generated until time $t$ can be approximated by an inhomogeneous Poisson process with rate $R\left(t\right) = u p_a w_t$
-where $w_t$ is the number of wildtype cells at time $t$, $w_t=N\e^{\Delta_w t}$.
+Note that if extinction occurs, that is $m_{\infty}=0$, then it is implied that $T_1=\infty$, and vice versa if $T_1<\infty$ then $m_{\infty}>0$.
+
+The number of successful mutants generated until time $t$ can be approximated by an inhomogeneous Poisson process with rate $R\left(t\right) = u p_a w_t$,
+where $w_t=N\e^{\Delta_w t}$ is the number of wildtype cells at time $t$.
+Note that 
+\begin{equation}\label{integralR}
+\int_0^{t}{R(z)\d z} = 
+u p_a N \frac{\exp[{\Delta_w t}]-1}{\Delta_w} \approx 
+u p_a N t,
+\end{equation}
+by integrating the exponential and because $\frac{\exp[\Delta_w t]-1}{\Delta_w}=\frac{1+\Delta_w t+O(t^2)-1}{\Delta_w}=t+O(t^2)$.
 The probability density function of $T_1$ is thus
-$R\left(t\right)\e^{-\int_0^{t}{R(z)\d z}}$.
+$R\left(t\right)\exp\left(-\int_0^{t}{R(z)\d z}\right)$. % TODO ref
 Therefore, the probability density function of $(T_1 \mid T_1 < \infty)$ is
-$f_1(t) = \frac{R\left(t\right)\e^{-\int_0^{t} R(z)\d z}}{p_{rescue}}$. % Todo is this correct? I made changes
+$f_1(t) = \frac{R\left(t\right)\exp\left(-\int_0^{t}{R(z)\d z}\right)}{p_{rescue}}$. 
+\\
 
 We are interested in the mean conditional time, $\tau_1=\mathbb{E}\left[T_1 \mid T_1<\infty\right]$, which is given by
-\begin{align}\label{meantime1}
+\begin{equation}\label{meantime1}
+\begin{aligned}
 \tau_1 =
 \int_{0}^{\infty}{t f_1(t) \d t} = 
-\frac{\int_{0}^{\infty}{t u p_a N\e^{\Delta_w t} \e^{-\int_0^{t} u p_a N\e^{\Delta_w z} \d z} \d t}}{p_{rescue}} .
+\frac{\int_{0}^{\infty}{tR(t)\exp\left(-\int_0^{t}{R(z)\d z}\right) \d t}}{p_{rescue}} = 
+\frac{\int_{0}^{\infty}{\exp\left(-\int_0^{t}{R(z)\d z}\right) \d t}}{p_{rescue}}
+\end{aligned}
+\end{equation}
+after applying integration by parts.
+Therefore, plugging \cref{integralR,rescue_prob2} in \cref{meantime1}, 
+\begin{align}\label{limitapprox}
+\tau_1 =
+\frac{\int_{0}^{\infty}{\exp\left(-u p_a N t\right) \d t}}{1-\e^{-Np_w}} \approx
+\left(1+\e^{-Np_w}\right)\int_0^\infty\e^{-u p_a N t} \d t =
+\frac{1+\e^{-Np_w}}{u p_a N} ,
 \end{align}
-\begin{equation}
-\frac{\int_0^\infty\e^{-\int_0^t R(z)\d z}\d t}{1-\left(1-p_w\right)^N}=\frac{\int_0^\infty\e^{-uNp_a\frac{\e^{\Delta_w t}-1}{\Delta_w}}\,\d t}{1-\left(1-p_w\right)^N}
-\end{equation}
-% YOAV HERE - dont understand how the previous two lines are the same...
-
-The fraction in exponential of the integrand in \eqref{meantime1} can approximated as:
-\begin{equation*}
-\frac{\e^{\Delta_w t}-1}{\Delta_w}=\frac{1+\Delta_w t+O(t^2)-1}{\Delta_w}=t+O(t^2).
-\end{equation*}
-As a result, the mean time $\tau_1$ can be simplified:
-\begin{align}\label{limitapprox}
-\tau_1\approx\left(1+\e^{-Np_w}\right)\int_0^\infty\e^{-uNp_a\tau}\,\d\tau=\frac{\left(1+\e^{-Np_w}\right)}{uNp_a},
-\end{align}
-where in the previous line we have used the approximation:
-\begin{equation*}
-\frac{1}{1-\e^{-Np_w}}\approx1+\e^{-Np_w}.
-\end{equation*}
+where we use the approximation $\left(1-\e^{-Np_w}\right)^{-1}\approx1+\e^{-Np_w}$ and performed the integration of the exponent.
+% TODO continue from here
 We plot the expansion \eqref{limitapprox} in Figure \ref{MeanTimeGrowthAneuploidyPlot} and observe that it is a very good fit for intermediate and large values of the initial wildtype population size.
@@
 %\begin{align}
-%f_{T_2}\left(t_2|m_{t\rightarrow\infty}\neq0\right)=\frac{R_2\left(t_2\right)\e^{-\int_0^{t_2} R_2(t)\d t}}{1-\left(1-p_w\right)^N}
+%f_{T_2}\left(t_2|m_{\infty}\neq0\right)=\frac{R_2\left(t_2\right)\e^{-\int_0^{t_2} R_2(t)\d t}}{1-\left(1-p_w\right)^N}
 %\end{align}
@@
 
-In this paper, we have modelled a population of cancer cells which are exposed to chemotherapeutic drugs and decline towards extinction.
 Evolutionary rescue  is the process where the population acquires a trait that increases fitness in the new environment such that extinction is averted.
-We have derived the probability of evolutionary rescue of the population of cancer cells under various demographic scenarious.
-The cancer cell population can escape extinction either through direct mutation or through mutation from aneuploidy.
-We have used multitype branching processes to study our model (Figure \ref{figureAneuploidy}) which allows us to obtain exact solutions for the probability of evolutionary rescue.
-
-The case when the aneuploid cells are resistant can be approximated by the one step evolutionary rescue process where the aneuploidy rescues the population (Figure \ref{SurvPlot}).
-However, when the growth rate of the aneuploid cells is negative then they cannot rescue the population and they can only act as a stepping stone (Figure \ref{P_est}) through which the mutant can be obtain in a more expedient fashion, given that the aneuploid population declines slower then the wildtype population, compared to the case of direct mutation from the wildtype.
-
-We observe from Figure \ref{RatioEvolRescue} that aneuploidy has a significant contribution towards evolutionary rescue.
-When aneuploidy is slightly increasing ($\Delta_a=10^{-3}$) the probability of evolutionary rescue is three orders of magnitude larger when aneuploidy is present compared to the case when aneuploidy is not present under the parameters previously described for tumors (see Table \ref{table1}).
-
-For our model we have assumed that cancer cell lineages are independent of each other.
-However this is not always true as cancer cells compete for resources which can have an effect on the probability of evolutionary rescue.
-We observe that this is not the case when the carrying capacity if sufficiently large the probability of evolutionary rescue is not impacted by the logistic model (see Figure \ref{SurvPlotNDataLogisticKComplete}).
-Future work should include using density dependent branching process in order to better model the conditions under cancer cells proliferate.
-
-The presence of aneuploid cancer cells at the onset of chemotherapy can facillitate evolutionary rescue by acting as a stepping stone for the appearance of resistant mutant cells.
-From Figure \ref{FractionPlot} we observe that, for even a relative small fraction of the initial population being composed of aneuploid cells, evolutionary rescue is more likely to occour through the initial aneuploidy.
-
-We propose experiments similar to the ones highlighted in \citep{martin2013probability} in order to test the predictions of our model.
-For example, in order to study the effects of initial population size on the probability of evolutionary rescue we propose to derive a large culture mass from a single cancer cell in permissive conditions and then dilute to a wide range of stating population sizes ($10^7-10^9$).
-Afterwards, we expose the population to anti-cancer drug which induces aneuploidy or to saline solution for control.
-Final density, in both cases, would be measured by optical density and the results compared to predictions from out model.
-
-We observe from equations \eqref{survprobw} and \eqref{rescue_prob} that the probability of evolutionary rescue increases when the initial population size increases, the wildtype population does not decline too quickly, the mutation and aneuploidy rates are high and the probabilities $p_a$ and $p_m$ are elevated.
-
-The probability of evolutionary rescue is enhanced by aneuploidy for small and intermediate sized tumors (see Figure \ref{RatioEvolRescuePopulationSize}).
-As a result, aneuploidy is unlikely to to contribute to primary tumors overcoming chemotherapy but it can contribute to the evolutionary rescue of secondary tumors whose size might be below the detection threshold of $\sim10^7$  \citep{bozic2013evolutionary}.
-Given the fact that the mean time for small and intermediate tumors to overcome chemotherapy can be of the order of 100 days (see Figure \ref{MeanTimeDeleteriousAneuploidyPlot})  this could explain the reappearance of cancer even after initial remission.
-
+Here, we have modeled a tumor--a population of cancer cells--exposed to drug therapy that causes the cell population to decline towards extinction.
+The cancer cell population can escape extinction either by a mutation that confers resistance, or by first generating aneuploid cells in which the effect of the drug is diminished, and then producing a mutation that confers full resistance (\Cref{figureAneuploidy}).
+
+Using multitype branching processes, we derived the probability of evolutionary rescue of the population of cancer cells under various scenarios for the effect of aneuploidy, including both tolerance and partial resistance to the drug.
+We obtained both exact and approximate expressions for the probability of evolutionary rescue.
+As expected, our analytic results in \cref{rescue_prob} show that the probability of evolutionary rescue increases with the initial tumor size $N$, the wildtype growth rate $\Delta_w=\lambda_w-\mu_w$, and the mutation $v$ and aneuploidy $u$ rates.
+
+When aneuploid cells are partially resistant to the drug ($\Delta_w\ll0\ll\Delta_a\ll\Delta_m$), evolutionary rescue can be approximated by a one-step evolutionary rescue process where aneuploidy itself rescues the population (\Cref{rescue_prob_wt_growth}).
+When aneuploidy only provides tolerance to the drug ($\Delta_w\ll\Delta_a\ll0\ll\Delta_m$), it cannot rescue the population.
+Instead, aneuploidy acts as a \emph{stepping stone} through which the resistant mutant can appear in a more expedient fashion, given that the aneuploid cell population declines slower then the wildtype cell population.
+In this case, aneuploidy provides two benefits.
+First, it delays the extinction of the population--providing more time for appearance of the resistance mutations.
+Second, it increases the population size relative to a wildtype population--providing more cells for generating mutations, i.e., it increases the mutation supply.
+
+We find that aneuploidy can have a significant effect on evolutionary rescue (\Cref{rescue_ratio}).
+For example, when aneuploidy cells are "barely-resistant" (they grow at a very low rate, $\Delta_a=10^{-3}$) the probability of evolutionary rescue is 1000-fold higher with aneuploidy than without it (for parameters previously described in cancer \Cref{table1}).
+Interestingly, aneuploidy is unlikely to contribute to evolutionary rescue in primary tumors, as the number of cells in such tumors ($N>10^7$) % TODO is this correct? 
+is large enough for the appearance of resistant mutation directly before the extinction of wildtype cells (\Cref{rescue_ratio}).
+However, aneuploidy may play a crucial role in evolutionary rescue of secondary tumors, whose size may be below the detection threshold of $\sim10^7$  \citep{bozic2013evolutionary}.
+Given the fact that the mean time for such secondary tumors to overcome chemotherapy can be of the order of 100 days (\Cref{MeanTimeDeleteriousAneuploidyPlot}), this can explain the reappearance of cancer even after initial remission.
+
+We hypothesized that presence of "standing variation"-- % TODO ref
+a subpopulation of aneuploid cancer cells--at the onset of chemotherapy may facilitate evolutionary rescue by reducing the waiting time for the appearance of aneuploid cells.
+Indeed, we observe that even when a small fraction of the initial tumor is aneuploid, evolutionary rescue is more likely to occur through this existing standing variation, rather then through "de novo" aneuploid cells (\Cref{FractionPlot}).
+
+We have assumed that cancer cell lineages are independent of each other.
+However, this may not be the case, as cancer cells compete for resources (e.g., blood supply).
+Nevertheless, we find that when the carrying capacity if large % TODO HOW LARGE?
+our approximation for the probability of evolutionary rescue agrees with results of stochastic simulations with density-dependent growth  \Cref{rescue_prob_N_logistic}).
+Future work may focus on scenarios with small carrying capacity by analysing density-dependent branching processes. % TODO ref?
+
+Our model predictions may be tested by experiments \citep{martin2013probability}.
+For example, to study the effects of initial tumor size on the probability of evolutionary rescue, a large culture mass can be propagated from a single cancer cell in permissive conditions and then diluted to a  range of starting tumor sizes.
+Afterwards, these tumors may be exposed to anti-cancer drugs that induces aneuploidy % TODO ref
+or to saline solution for control. Cell density can then be measured and compared to the predictions of our model. % TODO not clear what is to be compared... what does our model predict...?
+
+% TODO discussion of mean time
+
+% TODO summary
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+{\small
+\section*{Acknowledgements}
+We thank XXX for discussions and comments. % TODO who?
+This work was supported in part by
+the Israel Science Foundation (ISF 552/19, YR),
+the US–Israel Binational Science Foundation (BSF 2021276, YR), 
+Minerva Stiftung Center for Lab Evolution (YR), 
+and the Ela Kodesz Institute for Research on Cancer Development and Prevention (RS).
+}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %\section*{References}
-\newpage
+%\newpage
 \nolinenumbers
@@
 In both scenarios, rescue probability increase with both the wildtype growth rate (x-axis) and the aneuploidy growth rate (colors).
-Markers represent simulation results with 95\% CI; solid and dashed lines for the exact formula (\cref{survival_prob} in \cref{rescue_prob}); % TODO do you use 1-(1-pw)^N or 1-e(-Np_w)?
-dashed lines for the approximate formula (\cref{rescue_prob_approx}), demonstrating that they all agree.
+Markers represent simulation results with 95\% CI; solid and dashed lines for the exact formula (\cref{survival_prob} in \cref{rescue_prob}); dashed lines for the approximate formula (\cref{rescue_prob_approx}), demonstrating that they all agree.
 % TODO add parameter values
@@
 \includegraphics[width=1\textwidth]{Figures/SurvPlotNDataLogisticK.pdf}
+% TODO revise caption
 \caption{Plot of the probability of evolutionary rescue of a population, consisting of $N$ wildtype cells, as a function of the initial population size of wildtype cells where maximum population size is constrained by the carrying capacity $K$.
@@
 \caption{\textbf{Effect of standing variation on evolutionary rescue.}
-In aneuploid cells already exist in the population at the onset of drug therapy as standing genetic variation, then evolutionary rescue is more likely... % TODO think about this figure...
+In aneuploid cells already exist in the population at the onset of drug therapy as standing genetic variation, then evolutionary rescue is more likely... 
+% TODO think about this figure...
+% TODO maybe better to show p_total / p_denovo as function of u or N.
 %Plot of the fraction of the cases the population is rescue by standing genetic variation as a function of the fraction of initial cells which are aneuploid.
@@
 \textbf{(A-left,A-right)} Aneuploidy provides partial resistance. \textbf{(B-left, B-right)} Aneuploidy provides tolerance.  
-Solid and dashed lines apply $p_{rescue}$ from the exact formula of  (\cref{survival_prob} in \cref{rescue_prob}); % TODO do you use 1-(1-pw)^N or 1-e(-Np_w)?
-dashed lines apply $p_{rescue}$ from the approximate formula (\cref{rescue_prob_approx}), with good agreement.
+Solid and dashed lines apply $p_{rescue}$ from the exact formula of  (\cref{survival_prob} in \cref{rescue_prob}); dashed lines apply $p_{rescue}$ from the approximate formula (\cref{rescue_prob_approx}), with good agreement.
 % TODO parameter values

### 2023-06-04 EvolutionaryRescue1.tex

@@
 \begin{abstract}
-Evolutionary rescue is the process by which a population is able to survive a sudden environmental change which initially causes the population to decline towards extinction.
+%Evolutionary rescue is the process by which a population is able to survive a sudden environmental change which initially causes the population to decline towards extinction.
 A prime example of evolutionary rescue is the ability of cancer to survive being exposed to various treatments.
@@
 We validate these approximations by comparing them to results of stochastic evolutionary simulations.
-We find that the approximations work very well (\Cref{rescue_prob_wt_growth,rescue_prob_an_growth,rescue_prob_N_nongrow,rescue_prob_N_logistic}).
+We find that the approximations work very well (\Cref{rescue_prob_wt_growth,rescue_prob_an_growth,rescue_prob_N,rescue_prob_N}).
 
@@
 We therefore performed stochastic simulations of a logistic growth model with carrying capacity $K$ (Methods). 
-We find that our approximations agree with results of simulations with density-dependent growth for biologically relevant parameter values (\Cref{rescue_prob_N_logistic}).
+We find that our approximations agree with results of simulations with density-dependent growth for biologically relevant parameter values (\Cref{rescue_prob_N}).
 
@@
 \end{align}
-where we use the approximation $\left(1-\e^{-Np_w}\right)^{-1}\approx1+\e^{-Np_w}$ and performed the integration of the exponent.
-% TODO continue from here
-We plot the expansion \eqref{limitapprox} in Figure \ref{MeanTimeGrowthAneuploidyPlot} and observe that it is a very good fit for intermediate and large values of the initial wildtype population size.
+where we use the approximation $\left(1-\e^{-Np_w}\right)^{-1}\approx1+\e^{-Np_w}$ and integrate the exponent.
+\Cref{MeanTimeGrowthAneuploidyPlot} show the agreement between this approximating and simulation results for intermediate and large tumor sizes.
+
+\textbf{THE REST OF THIS SECTION REQUIRES EDITING BY YOAV}
 
@@
 \end{align}
-which we plot in Figure \ref{MeanTimeDeleteriousAneuploidyPlot} as a function of the initial population size.
+which we plot in Figure \ref{MeanTimeGrowthAneuploidyPlot} as a function of the initial population size.
 
@@
 where $\erfc$ is the complementary error function.
-We plot the expansion \eqref{limitapprox2} in Figure \ref{MeanTimeDeleteriousAneuploidyPlot} and observe that it is a very good fit for large values of the initial wildtype population size.
+We plot the expansion \eqref{limitapprox2} in Figure \ref{MeanTimeGrowthAneuploidyPlot} and observe that it is a very good fit for large values of the initial wildtype population size.
 
@@
 \end{align}
-which we plot in Figure \ref{MeanTimeDeleteriousAneuploidyPlot} and observe that it offer as a good a fit to \eqref{meantimet2} as \eqref{limitapprox2}.
+which we plot in Figure \ref{MeanTimeGrowthAneuploidyPlot} and observe that it offer as a good a fit to \eqref{meantimet2} as \eqref{limitapprox2}.
 Additionally, we observe that for large initial wildtype populations sizes direct mutation drives evolutionary rescue while aneuploidy plays a role for intermediate sized tumors.
-This is consistent with the information obtained from Figure \ref{RatioEvolRescuePopulationSize} where aneuploidy improves the probability of evolutionary rescue only  for small and intermediate values of $N$.
+This is consistent with the information obtained from \Cref{rescue_ratio} where aneuploidy improves the probability of evolutionary rescue only  for small and intermediate values of $N$.
 
@@
 However, aneuploidy may play a crucial role in evolutionary rescue of secondary tumors, whose size may be below the detection threshold of $\sim10^7$  \citep{bozic2013evolutionary}.
-Given the fact that the mean time for such secondary tumors to overcome chemotherapy can be of the order of 100 days (\Cref{MeanTimeDeleteriousAneuploidyPlot}), this can explain the reappearance of cancer even after initial remission.
+Given the fact that the mean time for such secondary tumors to overcome chemotherapy can be of the order of 100 days (\Cref{MeanTimeGrowthAneuploidyPlot}), this can explain the reappearance of cancer even after initial remission.
 
@@
 Nevertheless, we find that when the carrying capacity if large % TODO HOW LARGE?
-our approximation for the probability of evolutionary rescue agrees with results of stochastic simulations with density-dependent growth  \Cref{rescue_prob_N_logistic}).
+our approximation for the probability of evolutionary rescue agrees with results of stochastic simulations with density-dependent growth  \Cref{rescue_prob_N}).
 Future work may focus on scenarios with small carrying capacity by analysing density-dependent branching processes. % TODO ref?
@@
 \section*{Acknowledgements}
-We thank XXX for discussions and comments. % TODO who?
+%We thank XXX for discussions and comments. % TODO who?
 This work was supported in part by
@@
 and the Ela Kodesz Institute for Research on Cancer Development and Prevention (RS).
+% TODO add funding for Uri and Daniel
 }
@@
 \textbf{Evolutionary rescue probability with partially resistant or tolerant aneuploid cells.}
-Rescue probability is very high when aneuploidy provides partial resistance, in an initially small tumor (\textbf{Aleft}, $N=10^4$) and even more so in an initially large tumor (\textbf{Aright}, $N=10^8$).
+Rescue probability is very high when aneuploidy provides partial resistance ($\lambda_a=0.01$), in an initially small tumor (\textbf{Aleft}, $N=10^4$) and even more so in an initially large tumor (\textbf{Aright}, $N=10^8$).
 % TODO change panel labels to ABCD
@@
 Markers represent simulation results with 95\% CI; solid and dashed lines for the exact formula (\cref{survival_prob} in \cref{rescue_prob}); dashed lines for the approximate formula (\cref{rescue_prob_approx}), demonstrating that they all agree.
-% TODO add parameter values
+Here, aneuploidy rate is $u=10^{-2}$ and mutation rate is $v=10^{-7}$.
+% TODO add missing parameters
 }
@@
 Rescue probability grows with the aneuploid growth rate $\Delta_a$ (x-axis), and is much higher in an initially large tumor than in a small one (\textbf(A) $N=10^4$; \textbf{(B)} $N=10^8$). 
-Markers represent simulation results with 95\% CI; solid and dashed lines for the exact formula (\cref{survival_prob} in \cref{rescue_prob}); % TODO do you use 1-(1-pw)^N or 1-e(-Np_w)?
+Markers for simulation results with 95\% CI; 
+solid lines for the exact formula (\cref{survival_prob} in \cref{rescue_prob});
 dashed lines for the approximate formula (\cref{rescue_prob_approx}).
@@
 \begin{figure}[p]
+\begin{subfigure}{0.5\textwidth}
 % non-growing slightly declining, compare 1st order approximation to exact and simulation
- \vspace*{1\baselineskip}
 \includegraphics[width=1\textwidth]{Figures/DeleteriousTauLeapPlot.pdf}
-\caption{Plot of the probability of evolutionary rescue of a population, consisting of $N$ wildtype cells, as a function of the initial population size of wildtype cells.
-The blue line represent the exact result \eqref{survprobw} while the orange lines represent the approximation \eqref{pestthirdcase}.
-The green dots represent numerical simulations where the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
-The error bars are present but are not visible given the fact that we have used $n=10^5$ simulations for each combination of parameters.
-Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.}
-\label{rescue_prob_N_nongrow}
-\end{figure}
-%%%%%%%%%%%
-\begin{figure}[p]
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
 % density-dependent simulations, barely growing (or maybe partial resistance), with exact formula
- \vspace*{1\baselineskip}
 \includegraphics[width=1\textwidth]{Figures/SurvPlotNDataLogisticK.pdf}
-% TODO revise caption
-\caption{Plot of the probability of evolutionary rescue of a population, consisting of $N$ wildtype cells, as a function of the initial population size of wildtype cells where maximum population size is constrained by the carrying capacity $K$.
-The blue line represent the exact result \eqref{survprobw} while the orange dots represent numerical simulations where the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
-Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_w=1-10^{-1},\lambda_a=1+10^{-4},\lambda_m=1+10^{-1},\mu_w=1,\mu_m=1,u=10^{-2},v=10^{-7}, C_1=C_2=1, K=10^9$.}
-\label{rescue_prob_N_logistic}
+\end{subfigure}
+\caption{\textbf{Evolutionary rescue probability for variable initial tumor size.}
+\textbf{(A)} Comparison of simulation results (markers with 95\% CI, too small to appear with $10^5$ simulations per marker), the exact formula (blue line, \cref{survival_prob} in \cref{rescue_prob}) and the approximate formula (orange line, \cref{rescue_prob_approx}).
+\textbf{(B)} Comparison of results of simulations  with density-dependent growth (markers with with 95\% CI) and the exact formula (blue line, \cref{survival_prob} in \cref{rescue_prob}).
+Here, $K=10^9$.
+% TODO add parameter values
+}
+\label{rescue_prob_N}
 \end{figure}
@@
 % TODO maybe better to show p_total / p_denovo as function of u or N.
-%Plot of the fraction of the cases the population is rescue by standing genetic variation as a function of the fraction of initial cells which are aneuploid.
-The red vertical line highlights the the value of $f$ for which half the times the population is rescues by aneuploid cells while the pink line is our approximation \eqref{halfeqstandvar}.
-For this plot we have chosen the following parameters: $N=10^3, \lambda_w=1-10^{-2}, \lambda_a=1-10^{-4},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.
-The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n$ is the number of simulations.
-The value of $0.332$ highlighted in red is the value of the initial fraction of the population which in aneuploid for which half of the cases of evolutionary rescue are due to the initial aneuploid cell population.
+% TODO add parameter values
 }
@@
 %%%%%%%%%%%
-%\begin{figure}[p]
-%% A partial resistance
-%% B tolerance
-%% function of population size
-% \vspace*{1\baselineskip}
-%\caption{Plot of the ratio of the probability of evolutionary rescue when aneuploidy can play a role in rescue ($u>0$) to the probability where acquisition of aneuploidy is not possible ($u=0$) as a function of the initial population size of wildtype cells.
-The continuous lines represent the exact result \eqref{ratiorescueexact} while the dashed lines represent the approximation \eqref{ratiorescue}.
-The upper plot show the case when aneuploidy is resistant while the lower plot shows the case when it is susceptible.
-Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_w=0.14, \lambda_a=0.14,\lambda_m=0.14,\mu_w=0.17,\mu_m=0.13, u=10^{-2}, v=10^{-7}$.
+\begin{figure}[p]
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=1\textwidth]{Figures/MeanTimeDeleteriousPlot.pdf}
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=1\textwidth]{Figures/MeanTimeGrowthMutantDirectPlot.pdf}
+\end{subfigure}
+\caption{\textbf{Evolutionary rescue time.}
+Shown is the mean time for appearance of a resistance mutation the leads to evolutionary rescue \textbf{(left)} with ($u>0$) and \textbf{(right)} without ($u=0$ aneuploidy.
+Our inhomogeneous Poisson-process approximations (solid blue lines, right: \cref{meantime1}, left: \cref{meantimet2}) is in agreement with simulation results (orange markers with 95\% CI).
+Our 1st-order (dashed red lines, right: \cref{limitapprox}, left: \cref{limitapprox2}) and 2nd-order (green line, left: \cref{limitapprox3}) approximations work well when the initial tumor size is large (here $>10^8$ cells).
+% TODO parameter values
 }
-%\label{RatioEvolRescuePopulationSize}
-%\end{figure}
-%%%%%%%%%%%
-\begin{figure}[p]
-% partial resistance - aneuploidy rescues the pop
- \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/MeanTimeGrowthMutantDirectPlot.pdf}
-\caption{Plot of the mean time until the appearance of a resistance mutation which rescues the population in the case when evolutionary rescue is possible only through mutation but not aneuploidy and mutation.
-The blue line represents the approximation \eqref{meantime1} and the dashed red line represents the first order approximation \eqref{limitapprox}.
-The orange dots represent the numerical simulations while the error bars represent the interval centered at the mean which containing $95\%$ of the simulated values.
-Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_w=1-10^{-1},\lambda_m=1+10^{-1},\mu_w=1,\mu_m=1,u=10^{-2},v=10^{-7}$.}
 \label{MeanTimeGrowthAneuploidyPlot} 
-\end{figure}
-%%%%%%%%%%%
-\begin{figure}[p]
-% tolerance
- \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/MeanTimeDeleteriousPlot.pdf}
-\caption{Plot of the mean time until the appearance of a resistance mutation which rescues the population in the case when evolutionary rescue is possible 
- through mutation and aneuploidy.
-Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_w=1-10^{-1},\lambda_a=1-10^{-2},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1,u=10^{-2},v=10^{-7}$.
-The blue line represents the approximation \eqref{meantimet2}, the dashed red line represents the second order approximation \eqref{limitapprox2} and the green line is first order approximation \eqref{limitapprox3}.
-The orange dots represent the numerical simulations while the error bars represent the interval centered at the mean which containing $95\%$ of the simulated values.}
-\label{MeanTimeDeleteriousAneuploidyPlot}
 \end{figure}
@@
 \includegraphics[width=1\textwidth]{Figures/MeanTimeRatioInitialPopulationSize.pdf}
-\caption{Plot of the ratio of the mean time to evolutionary rescue when aneuploidy can play a role in rescue ($u>0$) to the mean time where acquisition of aneuploidy is not possible ($u=0$) as a function of the initial population size of wildtype cells.
-The continuous lines represent the approximation \eqref{eqMeanTimeRatioInitialPopulationSize}.
-Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $\lambda_w=1-10^{-1},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1$.}
+\caption{\textbf{Ratio of evolutionary rescue time with and without aneuploidy.}
+The ratio of the mean time to appearance of a resistance mutation that leads to evolutionary rescue with ($u>0$) and without ($u=0$) aneuploidy for variable initial tumor sizes (\cref{eqMeanTimeRatioInitialPopulationSize}) when aneuploidy provides tolerance to the drug ($\Delta_a\ll0$).
+When the initial tumor size is not large ($<10^8$), aneuploidy can decrease the rescue time by 10-100-fold.
+\emph{I THINK THERE IS A MISTAKE IN THE BLUE LINE} % TODO
+% TODO parameter values 
+}
 \label{MeanTimeRatioInitialPopulationSize}

### 2023-06-04 EvolutionaryRescue1.tex
change title
update figure

@@
 % Title page
-\title{Modeling the effect of aneuploidy on cancer evolution}
+\title{The role of aneuploidy in the evolution of cancer drug resistance}
 % Authors
@@
 % right function of population size
-% TODO fix panel labels
 \begin{subfigure}{0.5\textwidth}
@@
 \caption{\textbf{Effect of aneuploidy on evolutionary rescue.}
-The ratio of rescue probability with and without aneuploid ($H$, \cref{ratiorescue}) increases with the aneuploid growth rate (colors) and decreases with the wildtype growth rates and initial tumor size (x-axis), except for large tumors where aneuploid cells are tolerant (see right side of panel B-right).
-\textbf{(A-left,A-right)} Aneuploidy provides partial resistance. \textbf{(B-left, B-right)} Aneuploidy provides tolerance.  
+The ratio of rescue probability with and without aneuploid ($H$, \cref{ratiorescue}) increases with the aneuploid growth rate (colors) and decreases with the wildtype growth rates and initial tumor size (x-axes), except for large tumors where where the ratio converges to unity.
+\textbf{(A-left,A-right)} Aneuploidy provides partial resistance.
+\textbf{(B-left, B-right)} Aneuploidy provides tolerance.  
 Solid and dashed lines apply $p_{rescue}$ from the exact formula of  (\cref{survival_prob} in \cref{rescue_prob}); dashed lines apply $p_{rescue}$ from the approximate formula (\cref{rescue_prob_approx}), with good agreement.
 % TODO parameter values
+% TODO start y at 1
+% TODO fix panel labels
 }

### 2023-06-14 EvolutionaryRescue1.tex
add todo

@@
 % history of evolutionary rescue literature (that is not specific to cancer) should move to literature, and we should have a clear statement on what we add to that literature.
+
+% TODO two papers that show aneuploidy provides resistance
+% Lukow, Devon A., Erin L.
+Sausville, Pavit Suri, Narendra Kumar Chunduri, Angela Wieland, Justin Leu, Joan C.
+Smith, et al.
+2021.
+“Chromosomal Instability Accelerates the Evolution of Resistance to Anti-Cancer Therapies.” Developmental Cell 56 (17): 2427-2439.e4.
+https://doi.org/10.1016/j.devcel.2021.07.009.
+% Rutledge, Samuel D., Temple A.
+Douglas, Joshua M.
+Nicholson, Maria Vila-Casadesús, Courtney L.
+Kantzler, Darawalee Wangsa, Monika Barroso-Vilares, Shiv D.
+Kale, Elsa Logarinho, and Daniela Cimini.
+2016.
+“Selective Advantage of Trisomic Human Cells Cultured in Non-Standard Conditions.” Scientific Reports 6 (June 2015): 1–12.
+https://doi.org/10.1038/srep22828.
 
@@
 after applying integration by parts.
-Therefore, plugging \cref{integralR,rescue_prob2} in \cref{meantime1}, 
+Therefore, plugging \cref{integralR,rescue_prob2} in \cref{meantime1}, the probability density function of $T_1$ is given by:
+\begin{align}
+f_{T_1}\left(t_1\right)=R\left(t_1\right)\e^{-\int_0^{t_1} R(t)\d t}.
+\end{align}
+As a result, the time $T_1$ conditional on evolutionary rescue is given by:
+\begin{align}
+f_{T_1}\left(t_1|m_{t\rightarrow\infty}\neq0\right)=\frac{R\left(t_1\right)\e^{-\int_0^{t_1} R(t)\d t}}{1-\left(1-p_w\right)^N}.
+\end{align}
+The expectation of $T_1$ is:
+\begin{align}\label{meantime1}
+\tau_1=\mathbb{E}\left[T_1\right]&=\frac{\int_0^\infty\e^{-\int_0^\tau R(t)\d t}\d\tau}{1-\left(1-p_w\right)^N}=\frac{\int_0^\infty\e^{-uNp_a\frac{\e^{\Delta_w\tau}-1}{\Delta_w}}\,\d\tau}{1-\left(1-p_w\right)^N}.
+\end{align}
+The fration in the exponential of the integrand in \eqref{meantime1} can approximated as:
+\begin{equation*}
+\frac{\e^{\Delta_w\tau}-1}{\Delta_w}=\frac{1+\Delta_w\tau+O(\tau^2)-1}{\Delta_w}=\tau+O(\tau^2).
+\end{equation*}
+As a result, the mean time $\tau_1$ can be simplified:
 \begin{align}\label{limitapprox}

### 2023-06-19 EvolutionaryRescue1.tex
revised preamble, moved figs and table, fixed bug

@@
 
-\usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent
+%\usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent
+\usepackage{graphicx} % Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
+% TeX will automatically convert eps --> pdf in pdflatex
+\usepackage[labelfont=bf]{caption}
+\usepackage{float}
+
+\usepackage{amssymb,amsmath,amsthm}
+\usepackage{commath}
+\usepackage[hyphens]{url}
+\usepackage[dvipsnames]{xcolor}
+\usepackage[unicode=true,colorlinks=true,urlcolor=CadetBlue,citecolor=black,linkcolor=black]{hyperref}
+\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
+\usepackage{authblk}
+\usepackage{longtable}
+\usepackage{multirow}
+\usepackage{booktabs}
+\usepackage{lipsum}  
+\usepackage[title,page]{appendix}
+\usepackage{chngcntr}
+%\usepackage{end float}
+\usepackage{cleveref}
+ \usepackage{subcaption}
+
+%SetFonts
+% newtxtext+newtxmath
+\usepackage{newtxtext} %loads helv for ss, txtt for tt
 \usepackage{amsmath}
-\usepackage{amsfonts}
-\usepackage{tikz,pgfplots}
-\usepackage{gensymb}
-\usepackage{graphicx}
-\usepackage{subcaption}
-%\usepackage{xurl}
-
-\usepackage{epstopdf}
-%\usepackage{listings}
-\usepackage{tikz,fp,fp-random,todonotes}
-\usepackage[procnames]{listings}
-\usepackage{afterpage}
-\usepackage{float}
-\usepackage{amsthm}
-\usepackage{wrapfig}
-\usepackage{booktabs,caption}
-\usepackage[flushleft]{threeparttable}
-\usepackage{scrextend}
-\usepackage[all]{nowidow}
-\usepackage{stfloats}
-\usepackage{mathtools}
-\usepackage{bm}
-\usepackage{stackengine}
-\usepackage[hidelinks]{hyperref}
-\usepackage{siunitx}
-%\usepackage{subfig}
-\usepackage{authblk}
-\usepackage{cleveref}
+\usepackage[bigdelims]{newtxmath}
+\usepackage[T1]{fontenc}
+\usepackage{textcomp}
+%SetFonts
 
 % less space before sections 
-% \titlespacing*{<command>}{<left>}{<before-sep>}{<after-sep>}
-\usepackage{titlesec}
-\titlespacing*{\section}
-{0pt}{2ex plus 1ex minus .2ex}{2ex plus .2ex}
-\titlespacing*{\subsection}
-{0pt}{1ex plus 1ex minus .2ex}{1ex plus .2ex}
-\titlespacing*{\paragraph}
-{0pt}{1ex plus 1ex minus .2ex}{1ex plus .2ex}
-
-\addtokomafont{labelinglabel}{\sffamily}
-
-\usetikzlibrary{calc}
-\usetikzlibrary{backgrounds}
-\usetikzlibrary{calc}
-\usetikzlibrary{decorations.markings}
-\usetikzlibrary{arrows}
-\usetikzlibrary{positioning}
-\usetikzlibrary{decorations.text}
-
-\usetikzlibrary{shadings}
-\usetikzlibrary{backgrounds}
-\usetikzlibrary{calc}
-\usetikzlibrary{decorations.markings}
-\usetikzlibrary{arrows}
-\usetikzlibrary{positioning}
-\usetikzlibrary{decorations.text}
-\usetikzlibrary{shapes.multipart}
-
-\allowdisplaybreaks
-\renewcommand{\floatpagefraction}{.85}
-
+% https://tex.stackexchange.com/a/101126
+%\usepackage{titlesec}
+%\titlespacing*{\section}{0pt}{0.5\baselineskip}{0\baselineskip}
+%\titlespacing*{\subsection}{0pt}{0.5\baselineskip}{0\baselineskip}
+
+% Species names
+%% Meta-Command for defining new species macros
+\usepackage{xspace}
+
+\newcommand{\species}[3]{%
+  \newcommand{#1}{\gdef#1{\textit{#3}\xspace}\textit{#2}\xspace}}
+  \species{\yeast}{Saccharomyces cerevisiae}{S.~cerevisiae}
+
+% line numbers
+\usepackage[displaymath, mathlines]{lineno}
+\renewcommand\linenumberfont{\normalfont\small\sffamily}
+\linenumbers
+\modulolinenumbers[2]
+
+% Yoav & Lee commands
+\newcommand*{\tr}{^\intercal}
+\let\vec\mathbf
+\newcommand{\matrx}[1]{{\Big[ \stackrel{}{#1}\Big]}}
+\newcommand{\diag}[1]{\mbox{diag}\matrx{#1}}
+\newcommand{\goesto}{\rightarrow}
+\newcommand{\dspfrac}[2]{\frac{\displaystyle #1}{\displaystyle #2} }
+\newtheorem{theorem}{Theorem}
+\newtheorem{corollary}{Corollary}
+\newtheorem{lemma}{Lemma}
+\newtheorem{remark}{Remark}
+\newtheorem{result}{Result}
+\renewcommand\qedsymbol{} % no square at end of proof
+\newcommand{\cl}{\mathbf{L}}
+\newcommand{\cj}{\mathbf{J}}
+\newcommand{\ci}{\mathbf{I}}
+\newcommand{\E}{\mathbf{E}}
+\DeclareMathOperator{\sign}{sign}
+\renewcommand{\d}[1]{\ensuremath{\operatorname{d}\!{#1}}}
+
+% Remus commands
 \newcommand{\x}{{\bf x}}
@@
 \newcommand{\taua}{\tau_{\rm A}}
-
-% appendix
-\usepackage[title,page]{appendix}
-\usepackage{chngcntr}
 
@@
 
-% line numbers
-\usepackage[displaymath, mathlines]{lineno}
-\renewcommand\linenumberfont{\normalfont\small\sffamily}
-\linenumbers
-\modulolinenumbers[2]
+% autoref
+\def\equationautorefname{Eq.}
 
 % NatBib
-%\usepackage[super,comma]{natbib}
-%\usepackage[numbers,super,compress]{natbib}
-\usepackage[round,colon,authoryear]{natbib}
+\usepackage[comma,sort]{natbib}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-% this belong in the next paragraph, or remove it altogether
-% One mechanism by which aneuploidy has been hypothesized to affect cancer is by providing phenotypic variation and increasing heterogeneity in a tumor.
-
 The role of chromosomal instability (CIN) in the emergence of cancer has been studied extensively in the past decades \citep{michor2005can,christine2018understanding,nowak2002role,pavelka2010dr,komarova2003mutation,zhu2018cellular}.
@@
 Wildtype cells may missegregate to become aneuploids at rate $u$. Both aneuploid and wildtype cells may mutate to become mutants at rate $v$ (\Cref{figureAneuploidy}).
+
 %Thus, the changes in the number of each cell type is described by 
@@
 
-%%%%%%%%%%%
-\begin{figure}[!t]
-\centering
-\includegraphics[width=0.75\textwidth]{Figures/figureAneuploidy.pdf}
-\caption{
-\textbf{Model illustration.}
-\textbf{(A)} A population of cancer cells is composed of wildtype, aneuploid, and mutant cells, which divide with rates $\lambda_w$, $\lambda_a$, and $\lambda_m$ and die at rates $\mu_w$, $\mu_a$, and $\mu_m$, respectively.
-Wildtype cells can become aneuploid at rate $u$.
-Both aneuploid and wildtype cells can acquire a beneficial mutation with rate $v$.
-Color denotes the relative growth rates of the three genotypes such that $\lambda_w - \mu_w < \lambda_a - \mu_a < \lambda_m - \mu_m$.
-\textbf{(B)} The wildtype and the mutant are susceptible and resistant, respectively, to the drug.
-The aneuploid may be tolerant, non-growing, or or partially resistant.
-}
-\label{figureAneuploidy}
-\end{figure}
-%%%%%%%%%%%
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 The wildtype population initially consists of $w_0$ cells, whereas the other cell types are initially absent.
-
-\begin{table}
-\begin{center}
-  \begin{tabular}{| l |p{5cm}| c | c | p{3cm} |}
-    \hline
-     & Name & Value & Units & References \\ \hline
-    $N$ & Initial tumor size & $10^7-10^9$ & cells  & \citet{del2009does} \\ \hline
-    $\lambda_w$ & Wildtype division rate& 0.14 & 1/days  & \citep{bozic2013evolutionary} \\ \hline
-    $\mu_w$ & Wildtype death rate& 0.17 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
-    $\lambda_a$  & Aneuploid division rate$^\ast$ & 0.14 & 1/days  & - \\ \hline
-    $\mu_a$ & Aneuploid death rate$^\ast$ & $0.13-0.17$ & 1/days  & - \\ \hline
-    $\lambda_m$ & Mutant division rate& 0.14 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
-    $\mu_m$ & Mutant death rate& 0.13 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
-    $u$ & Missegregation rate& $10^{-3}-10^{-2}$ & 1$\slash$cell division  & \citet{nowak2004evolutionary,bakker2023predicting} \\ \hline
-    $v$ & Mutation rate& $10^{-7}-10^{-9}$ &  1$\slash$gene$\slash$cell division  & \citet{nowak2004evolutionary} \\ \hline
-  \end{tabular}
-\caption{\textbf{Model parameters.} 
-Aneuploid birth rate $\lambda_a$ is set to the same value as the wildtype and mutant birth rates, $\lambda_w$ and $\lambda_m$.
-Aneuploid death rate $\mu_a$ is set to an intermediate value between the wildtype and mutant death rates, $\mu_w$ and $\mu_m$.}
-  \label{table1}
-\end{center}
-\end{table}
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \paragraph{$\tau$-leaping.}
@@
 If the change in number of cells is negative and larger then the subpopulation size then the subpopulation size is updated to be zero.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection*{Code and data availability.} All source code is available online at \url{https://github.com/yoavram-lab/EvolutionaryRescue}.
 
-%\newpage
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \section*{Results}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection*{Effect of aneuploidy on evolutionary rescue}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection*{Evolutionary rescue time}
@@
 
-\begin{align*}
+\begin{align}
 r_1\left(t\right)&=vp_m\int_0^ta_{\tau}\,\d\tau=\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_wt}-1}{\Delta_w}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right),\\
 r_2\left(t\right)&=vp_m\int_0^tw_{\tau}\,\d\tau=vNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}.
-\end{align*}
+\end{align}
 For large initial population sizes we can assume that both rescue mutations produced through direct mutation and aneuploidy are independent and, as a result, they can be merged into a single Poisson process with rate $\left(r_1+r_2\right)\left(t\right)$.
@@
 \end{align}
-which we plot in Figure \ref{MeanTimeGrowthAneuploidyPlot} as a function of the initial population size.
+which we plot in \Cref{MeanTimeGrowthAneuploidyPlot} as a function of the initial population size.
 
@@
 where $\erfc$ is the complementary error function.
-We plot the expansion \eqref{limitapprox2} in Figure \ref{MeanTimeGrowthAneuploidyPlot} and observe that it is a very good fit for large values of the initial wildtype population size.
+We plot the expansion \eqref{limitapprox2} in \Cref{MeanTimeGrowthAneuploidyPlot} and observe that it is a very good fit for large values of the initial wildtype population size.
 
@@
 \begin{align}\label{limitapprox3}
-\tau_2\approx\left(1+\e^{-Np_w}\right)\int_0^\infty\e^{-uNp_m\tau}\,\d\tau=\frac{\left(1+\e^{-Np_w}\right)}{uNp_m},
-\end{align}
-which we plot in Figure \ref{MeanTimeGrowthAneuploidyPlot} and observe that it offer as a good a fit to \eqref{meantimet2} as \eqref{limitapprox2}.
+\tau_2\approx\left(1+\e^{-Np_w}\right)\int_0^\infty\e^{-uNp_m\tau}\,\d\tau=\frac{\left(1+\e^{-Np_w}\right)}{uNp_m} ,
+\end{align}
+which we plot in \Cref{MeanTimeGrowthAneuploidyPlot} and observe that it offer as a good a fit to \cref{meantimet2} as \cref{limitapprox2}.
 Additionally, we observe that for large initial wildtype populations sizes direct mutation drives evolutionary rescue while aneuploidy plays a role for intermediate sized tumors.
-This is consistent with the information obtained from \Cref{rescue_ratio} where aneuploidy improves the probability of evolutionary rescue only  for small and intermediate values of $N$.
+This is consistent with the information obtained from \cref{rescue_ratio} where aneuploidy improves the probability of evolutionary rescue only  for small and intermediate values of $N$.
 
@@
 \subsection*{Contribution of aneuploidy to mean evolutionary rescue time}
-\begin{align}\nonumber
+\begin{align}
 I&=\frac{\tau_2}{\tau_1}=\frac{\int_0^\infty\exp\left[-\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_wt}-1}{\Delta_w}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right)-vNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}\right]\,\d\tau}{\int_0^\infty\e^{-uNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}}\,\d\tau}\times\frac{1-\left(1-\left.p_w\right|_{u=0}\right)^N}{1-\left(1-\left.p_w\right|_{u>0}\right)^N}\\\label{eqMeanTimeRatioInitialPopulationSize}
@@
 \end{align}
-where $H$, is the ratio of the probability of evolutionary rescue with and without aneuploidy, defined in \eqref{ratiorescueexact}.
-We plot \eqref{eqMeanTimeRatioInitialPopulationSize} in Figure \ref{MeanTimeRatioInitialPopulationSize} as a function of the initial wildtype population for varying values of the Malthusian fitness of aneuploid cells $\Delta_a$.
+where $H$, is the ratio of the probability of evolutionary rescue with and without aneuploidy, defined in \cref{ratiorescueexact}.
+We plot \cref{eqMeanTimeRatioInitialPopulationSize} in \Cref{MeanTimeRatioInitialPopulationSize} as a function of the initial wildtype population for varying values of the Malthusian fitness of aneuploid cells $\Delta_a$.
 
@@
 %\section*{References}
-%\newpage
+
 \nolinenumbers
@@
 \bibliography{evo2022}
-\newpage
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{figure}[p]
-% small population size
-% A: partial resistance
-% B: tolerance
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{table}
+\begin{center}
+  \begin{tabular}{| l |p{5cm}| c | c | p{3cm} |}
+    \hline
+     & Name & Value & Units & References \\ \hline
+    $N$ & Initial tumor size & $10^7-10^9$ & cells  & \citet{del2009does} \\ \hline
+    $\lambda_w$ & Wildtype division rate& 0.14 & 1/days  & \citep{bozic2013evolutionary} \\ \hline
+    $\mu_w$ & Wildtype death rate& 0.17 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
+    $\lambda_a$  & Aneuploid division rate$^\ast$ & 0.14 & 1/days  & - \\ \hline
+    $\mu_a$ & Aneuploid death rate$^\ast$ & $0.13-0.17$ & 1/days  & - \\ \hline
+    $\lambda_m$ & Mutant division rate& 0.14 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
+    $\mu_m$ & Mutant death rate& 0.13 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
+    $u$ & Missegregation rate& $10^{-3}-10^{-2}$ & 1$\slash$cell division  & \citet{nowak2004evolutionary,bakker2023predicting} \\ \hline
+    $v$ & Mutation rate& $10^{-7}-10^{-9}$ &  1$\slash$gene$\slash$cell division  & \citet{nowak2004evolutionary} \\ \hline
+  \end{tabular}
+\caption{\textbf{Model parameters.} 
+Aneuploid birth rate $\lambda_a$ is set to the same value as the wildtype and mutant birth rates, $\lambda_w$ and $\lambda_m$.
+Aneuploid death rate $\mu_a$ is set to an intermediate value between the wildtype and mutant death rates, $\mu_w$ and $\mu_m$.}
+  \label{table1}
+\end{center}
+\end{table}
+
+%%%%%%%%%%
+
+\begin{figure}
+\centering
+\includegraphics[width=0.75\textwidth]{Figures/figureAneuploidy.pdf}
+\caption{
+\textbf{Model illustration.}
+\textbf{(A)} A population of cancer cells is composed of wildtype, aneuploid, and mutant cells, which divide with rates $\lambda_w$, $\lambda_a$, and $\lambda_m$ and die at rates $\mu_w$, $\mu_a$, and $\mu_m$, respectively.
+Wildtype cells can become aneuploid at rate $u$.
+Both aneuploid and wildtype cells can acquire a beneficial mutation with rate $v$.
+Color denotes the relative growth rates of the three genotypes such that $\lambda_w - \mu_w < \lambda_a - \mu_a < \lambda_m - \mu_m$.
+\textbf{(B)} The wildtype and the mutant are susceptible and resistant, respectively, to the drug.
+The aneuploid may be tolerant, non-growing, or or partially resistant.
+}
+\label{figureAneuploidy}
+\end{figure}
+
+%%%%%%%%%%%
+
+\begin{figure}
+% TODO change panel labels to ABCD
 \begin{subfigure}{0.5\textwidth}
@@
 Rescue probability is very high when aneuploidy provides partial resistance ($\lambda_a=0.01$), in an initially small tumor (\textbf{Aleft}, $N=10^4$) and even more so in an initially large tumor (\textbf{Aright}, $N=10^8$).
-% TODO change panel labels to ABCD
-When aneuploidy provides tolerance (\textbf{Bleft}, $N=10^4$; (\textbf{Bright}, $N=10^8$), the rescue probability is much lower. 
+When aneuploidy provides tolerance (\textbf{Bleft}, $N=10^4$; \textbf{Bright}, $N=10^8$), the rescue probability is much lower. 
 In both scenarios, rescue probability increase with both the wildtype growth rate (x-axis) and the aneuploidy growth rate (colors).
 Markers represent simulation results with 95\% CI; solid and dashed lines for the exact formula (\cref{survival_prob} in \cref{rescue_prob}); dashed lines for the approximate formula (\cref{rescue_prob_approx}), demonstrating that they all agree.
-Here, aneuploidy rate is $u=10^{-2}$ and mutation rate is $v=10^{-7}$.
-% TODO add missing parameters
+Here, division rate $\lambda_w=\lambda_a=\lambda_m=0.14$, (so that growth rate changes due to variable death rate), mutant death rate $\mu_m=0.13$ (so that mutant growth rate $\Delta_m=0.01$), aneuploidy rate $u=10^{-2}$, mutation rate $v=10^{-7}$.
 }
@@
 %%%%%%%%%%%
-\begin{figure}[p]
+
+\begin{figure}
 % small population size
@@
 \end{figure}
+
 %%%%%%%%%%%
-\begin{figure}[p]
+
+\begin{figure}
 \begin{subfigure}{0.5\textwidth}
@@
 \end{figure}
+
 %%%%%%%%%%%
-\begin{figure}[p]
+
+\begin{figure}
 % slightly decreasing (need to check)
@@
 \end{figure}
+
 %%%%%%%%%%%
-\begin{figure}[p]
+
+\begin{figure}
 % A partial resistance
@@
 \end{figure}
+
 %%%%%%%%%%%
-\begin{figure}[p]
+
+\begin{figure}
 % tolerance... maybe third is slightly decreasing
@@
 \end{figure}
-%%%%%%%%%%%
-%\begin{figure}[p]
-%% diffusion approx instead of branching process
-%% tolerance
-% \vspace*{1\baselineskip}
-%\includegraphics[width=1\textwidth]{Figures/P_est_diffusion.pdf}
-%\caption{Plot of the survival probability of an initial population consisting of $w_0=10^{4}$ wildtype cells as a function of $\Delta_a=\lambda_a-\mu_a$ for various values of $\Delta_w=\lambda_a-\mu_a$.
+
+%%%%%%%%%%
+
+\begin{figure}
+% diffusion approx instead of branching process
+% tolerance
+ \vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/P_est_diffusion.pdf}
+\caption{Plot of the survival probability of an initial population consisting of $w_0=10^{4}$ wildtype cells as a function of $\Delta_a=\lambda_a-\mu_a$ for various values of $\Delta_w=\lambda_a-\mu_a$.
 The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the Feller diffusion approximation \eqref{FellerApprox}.
 The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/w_0}$ where $p$ is the mean probability of evolutionary rescue.}
-%\label{P_est_diffusion}
-%\end{figure}
-%%%%%%%%%%%
-%\begin{figure}[p]
+\label{P_est_diffusion}
+\end{figure}
+
+%%%%%%%%%%
+
+\begin{figure}
 % partial resistance, compare 1st and 2nd order approximation and exact
-%\vspace*{1\baselineskip}
-%\includegraphics[width=1\textwidth]{Figures/SurvPlotNData.pdf}
-%\caption{Plot of the probability of survival of a population as a function of the initial population size of wildtype cells.
+\vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/SurvPlotNData.pdf}
+\caption{Plot of the probability of survival of a population as a function of the initial population size of wildtype cells.
 The blue line represents the exact solution \eqref{survprobw}, the orange line line represents the approximation \eqref{aneuploidyresistentfirstapprox}, the green line represents the first order correction \eqref{survprobwapproxcorrected} and the red dots represents stochastic simulations.
@@
 The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n=100$ is the number of simulations.}
-%\label{compare_1st_2nd_order_approx}
-%\end{figure}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\label{compare_1st_2nd_order_approx}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 %\appendix

### 2023-06-19 EvolutionaryRescue1.tex
added parameter values to figures

@@
 \end{align*}
-where $C_1, C_2>0$ are constants. % TODO what constants? see Remus version.
+where $C_1, C_2>0$ are constants and $K$ is the maximum carrying capcity. % TODO what constants? see Remus version.
 
@@
     $u$ & Missegregation rate& $10^{-3}-10^{-2}$ & 1$\slash$cell division  & \citet{nowak2004evolutionary,bakker2023predicting} \\ \hline
-    $v$ & Mutation rate& $10^{-7}-10^{-9}$ &  1$\slash$gene$\slash$cell division  & \citet{nowak2004evolutionary} \\ \hline
+    $v$ & Mutation rate& $10^{-7}-10^{-9}$ &  1$\slash$gene$\slash$cell division  & \citet{nowak2004evolutionary} \\ 
+    % TODO citing Nowak 2004 for the rates is strange because it is a theory paper - cite the original empirical papers
+    \hline
   \end{tabular}
@@
 Markers represent simulation results with 95\% CI; solid and dashed lines for the exact formula (\cref{survival_prob} in \cref{rescue_prob}); dashed lines for the approximate formula (\cref{rescue_prob_approx}), demonstrating that they all agree.
-Here, division rate $\lambda_w=\lambda_a=\lambda_m=0.14$, (so that growth rate changes due to variable death rate), mutant death rate $\mu_m=0.13$ (so that mutant growth rate $\Delta_m=0.01$), aneuploidy rate $u=10^{-2}$, mutation rate $v=10^{-7}$.
+Parameters: division rate $\lambda_w=\lambda_a=\lambda_m=0.14$ (so that growth rate changes due to variable death rate); mutant death rate $\mu_m=0.13$ (so that mutant growth rate $\Delta_m=0.01$); aneuploidy rate $u=10^{-2}$; mutation rate $v=10^{-7}$.
 }
@@
 % switch from tolerance to non-growing
+% TODO why no switch in panel B?
+% TODO more simulations for 0 > Delta_a > Delta^*_a
 \begin{subfigure}{0.5\textwidth}
@@
 When the tumor size is small (panel A), we switch between the approximation for tolerant and for non-growing aneuploid cells; the switch occurs at $\Delta_a^*=2vp_m+v+2\sqrt{vp_m\left(vp_m+\mu_a+v\right)}$.
-% TODO why no switch in panel B?
-% TODO add parameter values
-% TODO more simulations for 0 > Delta_a > Delta^*_a
+Parameters: $\lambda_w=\lambda_a=\lambda_m=0.14$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
 }
@@
 \textbf{(A)} Comparison of simulation results (markers with 95\% CI, too small to appear with $10^5$ simulations per marker), the exact formula (blue line, \cref{survival_prob} in \cref{rescue_prob}) and the approximate formula (orange line, \cref{rescue_prob_approx}).
-\textbf{(B)} Comparison of results of simulations  with density-dependent growth (markers with with 95\% CI) and the exact formula (blue line, \cref{survival_prob} in \cref{rescue_prob}).
-Here, $K=10^9$.
-% TODO add parameter values
+\textbf{(B)} Comparison of results of simulations  with density-dependent growth (markers with with 95\% CI) and the exact formula (blue line, \cref{survival_prob} in \cref{rescue_prob}) with maximum carrying capacity $K=10^9$.
+Parameters: $\lambda_w=\lambda_a=\lambda_m=0.14$; $\mu_w=0.17$; (A) $\mu_a=0.15$, (B) $\mu_a=0.135$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
 }
@@
 In aneuploid cells already exist in the population at the onset of drug therapy as standing genetic variation, then evolutionary rescue is more likely... 
+Parameters: $\lambda_w=\lambda_a=\lambda_m=0.14$; $\mu_w=0.17$; $\mu_a=0.145$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
 % TODO think about this figure...
 % TODO maybe better to show p_total / p_denovo as function of u or N.
-% TODO add parameter values
 }
@@
 Solid and dashed lines apply $p_{rescue}$ from the exact formula of  (\cref{survival_prob} in \cref{rescue_prob}); dashed lines apply $p_{rescue}$ from the approximate formula (\cref{rescue_prob_approx}), with good agreement.
-% TODO parameter values
+Parameters: $N=10^4$; $\lambda_w=\lambda_a=\lambda_m=0.14$; (B) $\mu_w=0.17$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
 % TODO start y at 1
@@
 Our 1st-order (dashed red lines, right: \cref{limitapprox}, left: \cref{limitapprox2}) and 2nd-order (green line, left: \cref{limitapprox3}) approximations work well when the initial tumor size is large (here $>10^8$ cells).
-% TODO parameter values
+Parameters: $\lambda_w=\lambda_a=\lambda_m=0.14$; $\mu_w=0.17$; (A) $\mu_a=0.145$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
 }
@@
 %%%%%%%%%%
-
-\begin{figure}
-% diffusion approx instead of branching process
-% tolerance
- \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/P_est_diffusion.pdf}
-\caption{Plot of the survival probability of an initial population consisting of $w_0=10^{4}$ wildtype cells as a function of $\Delta_a=\lambda_a-\mu_a$ for various values of $\Delta_w=\lambda_a-\mu_a$.
+%
+%\begin{figure}
+%% diffusion approx instead of branching process
+%% tolerance
+% \vspace*{1\baselineskip}
+%\includegraphics[width=1\textwidth]{Figures/P_est_diffusion.pdf}
+%\caption{Plot of the survival probability of an initial population consisting of $w_0=10^{4}$ wildtype cells as a function of $\Delta_a=\lambda_a-\mu_a$ for various values of $\Delta_w=\lambda_a-\mu_a$.
 The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the Feller diffusion approximation \eqref{FellerApprox}.
 The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/w_0}$ where $p$ is the mean probability of evolutionary rescue.}
-\label{P_est_diffusion}
-\end{figure}
-
-%%%%%%%%%%
-
-\begin{figure}
-% partial resistance, compare 1st and 2nd order approximation and exact
-\vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/SurvPlotNData.pdf}
-\caption{Plot of the probability of survival of a population as a function of the initial population size of wildtype cells.
+%\label{P_est_diffusion}
+%\end{figure}
+%
+%%%%%%%%%%%
+%
+%\begin{figure}
+%% partial resistance, compare 1st and 2nd order approximation and exact
+%\vspace*{1\baselineskip}
+%\includegraphics[width=1\textwidth]{Figures/SurvPlotNData.pdf}
+%\caption{Plot of the probability of survival of a population as a function of the initial population size of wildtype cells.
 The blue line represents the exact solution \eqref{survprobw}, the orange line line represents the approximation \eqref{aneuploidyresistentfirstapprox}, the green line represents the first order correction \eqref{survprobwapproxcorrected} and the red dots represents stochastic simulations.
@@
 The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n=100$ is the number of simulations.}
-\label{compare_1st_2nd_order_approx}
-\end{figure}
+%\label{compare_1st_2nd_order_approx}
+%\end{figure}
 

### 2023-06-20 EvolutionaryRescue1.tex
fixed typos

@@
 The difference between the division and death rate is $\Delta_k = \lambda_k-\mu_k$.
-We assume the population of cells is under a strong stress, such as drug therapy, to which the wildtype genotype is susceptible and therefore $\Delta_w<0$, whereas the mutant is resistant to the stress, $\Delta_m>0$
-We analyze three scenarios: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_w$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing or "barely growing", that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$.
+We assume the population of cells is under a strong stress, such as drug therapy, to which the wildtype genotype is susceptible and therefore $\Delta_w<0$, whereas the mutant is resistant to the stress, $\Delta_m>0$.
+We analyze three scenarios: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_w$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing or "barely growing", that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$, in a sense that we will make more precise.
 Wildtype cells may missegregate to become aneuploids at rate $u$. Both aneuploid and wildtype cells may mutate to become mutants at rate $v$ (\Cref{figureAneuploidy}).

### 2023-06-20 EvolutionaryRescue1.tex
typos

@@
 Each iteration of the simulation loop starts with computing the rates $\nu_j$ of each event $j$.
-We then draw the time until the next event, $\Delta t$, from an exponential distribution whose rate parameter is the sum of the rates of all events, such that $\Delta t \sim \textit{Exp}(\sum_j \nu_j)$.
+We then draw the time until the next event, $\Delta t$, from an exponential distribution whose rate parameter is the sum of the rates of all events, such that $\Delta t \sim \text{Exp}(\sum_j \nu_j)$.
 Then, we randomly determine which event occurred, where the probability for event $j$ is $p_j=\nu_j/\sum_i \nu_i$.
@@
 \begin{equation*}
-m_t>\left\lfloor-\frac{3\log10}{\log\left(\frac{\mu_m}{\lambda_m}\right)}\right\rfloor+1,
+m_t>\left\lfloor\frac{3\log10}{\log\left(\frac{\lambda_m}{\mu_m}\right)}\right\rfloor+1.
 \end{equation*}
@@
 
-To analyze evolutionary rescue in this model, we use the framework of \emph{multitype branching processes} \citep{rybnikov2021fitness,harris1963theory}. 
+To analyze evolutionary rescue in this model, we use the framework of \emph{multitype branching processes} \citep{harris1963theory}. 
 This allows us to find explicit expressions for the \emph{survival probability}: the probability that a lineage descended from a single cell does not become extinct.
@@
 
-The survival probabilities are given by the smallest solution for each quadratic equation \citep{uecker2015adaptive}. Therefore we have
+The survival probabilities are given by the smallest solution for each quadratic equation \citep{weissman2009rate,uecker2015adaptive}. Therefore we have
 \begin{equation}\label{survival_prob}

### 2023-06-21 EvolutionaryRescue1.tex
Update EvolutionaryRescue1.tex
added commenting in blue

@@
  \usepackage{subcaption}
+\usepackage{xcolor}
 
@@
 \usepackage{newtxtext} %loads helv for ss, txtt for tt
-\usepackage{amsmath}
 \usepackage[bigdelims]{newtxmath}
@@
 \newcommand{\taua}{\tau_{\rm A}}
+
+% Daniel commands
+
+\newcommand{\daniel}[1]{\textcolor{blue}{#1}}
 
@@
 
+\daniel{I think this section has too much of the mathematical derivations, which should instead be in something like Methods. It should focus on just the results.}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\paragraph*{Density-dependent growth.}
+\paragraph*{Density-dependent growth}
 
@@
 
-When aneuploid cells are partially resistant to the drug ($\Delta_w\ll0\ll\Delta_a\ll\Delta_m$), evolutionary rescue can be approximated by a one-step evolutionary rescue process where aneuploidy itself rescues the population (\Cref{rescue_prob_wt_growth}).
-When aneuploidy only provides tolerance to the drug ($\Delta_w\ll\Delta_a\ll0\ll\Delta_m$), it cannot rescue the population.
+When aneuploid cells are partially resistant to the drug ($\Delta_w\ll$(SOMETHING $\ne0$)$\ll\Delta_a\ll\Delta_m$), evolutionary rescue can be approximated by a one-step evolutionary rescue process where aneuploidy itself rescues the population (\Cref{rescue_prob_wt_growth}).
+When aneuploidy only provides tolerance to the drug ($\Delta_w\ll\Delta_a\ll$(SOMETHING $\ne0$)$\ll\Delta_m$), it cannot rescue the population.
 Instead, aneuploidy acts as a \emph{stepping stone} through which the resistant mutant can appear in a more expedient fashion, given that the aneuploid cell population declines slower then the wildtype cell population.

### 2023-06-21 EvolutionaryRescue1.tex
moved survival probability calculations to Methods

@@
 \newcommand{\daniel}[1]{\textcolor{blue}{#1}}
+\newcommand{\presc}{p_\text{rescue}}
 
@@
 
-\daniel{I think this section has too much of the mathematical derivations, which should instead be in something like Methods. It should focus on just the results.}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Survival probability}
-
-To analyze evolutionary rescue in this model, we use the framework of \emph{multitype branching processes} \citep{harris1963theory}. 
-This allows us to find explicit expressions for the \emph{survival probability}: the probability that a lineage descended from a single cell does not become extinct.
-
-Let $p_w$, $p_a$, and $p_m$ be the survival probabilities of a population consisting initially of single wildtype cell, aneuploid cell, or mutant cell, respectively.
-The complements $1-p_w$, $1-p_a$, and $1-p_m$ are the extinction probabilities, which satisfy each its respective equation,
-\begin{equation} \label{extinction_prob}
-\begin{aligned}
-1-p_w = &\frac{\mu_w}{\lambda_w+\mu_w+u+v} + 
-		  \frac{u}{\lambda_w+\mu_w+u+v}\left(1-p_a\right) + \\
-		  & \frac{\lambda_w}{\lambda_w+\mu_w+u+v}\left(1-p_w\right)^2 +
-		  \frac{v}{\lambda_w+\mu_w+u+v}\left(1-p_m\right) ,\\
-1-p_a = &\frac{\mu_a}{\lambda_a+\mu_a+v}+\frac{v}{\lambda_a+\mu_a+v}\left(1-p_m\right)+\frac{\lambda_a}{\lambda_a+\mu_a+v}\left(1-p_a\right)^2 ,\\
-1-p_m = &\frac{\mu_m}{\lambda_m+\mu_m}+\frac{\lambda_m}{\lambda_m+\mu_m}\left(1-p_m\right)^2 .	 
-\end{aligned}
-\end{equation}
-
-The survival probabilities are given by the smallest solution for each quadratic equation \citep{weissman2009rate,uecker2015adaptive}. Therefore we have
-\begin{equation}\label{survival_prob}
-\begin{aligned}
-p_w &= \frac{\lambda_w-\mu_w-u-v+\sqrt{\left(\lambda_w-\mu_w-u-v\right)^2+4\lambda_w\left(up_a+vp_m\right)}}{2\lambda_w} ,\\
-p_a &= \frac{\lambda_a-\mu_a-v+\sqrt{\left(\lambda_a-\mu_a-v\right)^2+4\lambda_avp_m}}{2\lambda_a}, \\
-p_m &= \frac{\lambda_m-\mu_m}{\lambda_m} .
-\end{aligned} 
-\end{equation}
-Note that the equation for $p_w$ depends on both $p_a$ and $p_m$, and the equation for $p_a$ depends on $p_m$.
-To proceed, we can plug the solution for $p_m$ and $p_a$ into the solution for $p_w$. We perform this for three different scenarios.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{Scenario 1: Aneuploid cells are partially resistant}
-
-We first assume that aneuploidy provides partial resistance to drug therapy, $\lambda_a>\mu_a$, and that this resistance is significant, $\left(\lambda_a-\mu_a-v\right)^2 > 4\lambda_a v p_m$.
-We thus rewrite \cref{survival_prob} as
-\begin{align*}
-p_w&=\frac{\lambda_w-\mu_w-u-v}{2\lambda_w}\left(1-\sqrt{1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}}\right) ,
-\text{and} \\
-p_a&=\frac{\lambda_a-\mu_a-v}{2\lambda_a}\left(1+\sqrt{1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}}\right) . 
-\end{align*}
-Using the quadratic Taylor expansion $\sqrt{1+x}=1+x/2+O(x^2)$ and assuming $u,v \ll 1$,
-we obtain the following approximation for the survival probability of a population initially consisting of a single wildtype cell,
-\begin{align}\label{survprobwapprox1}
-p_w 
-&\approx -\frac{vp_m+up_a}{\lambda_w-\mu_w-u-v}\\
-\nonumber
-%&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{v\left(\lambda_a-\mu_a-u\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-u\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\ \label{survprobw2}
-&\approx-\frac{1}{\lambda_w-\mu_w}\left[\frac{u\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\
-\end{align}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\paragraph{Second-order approximation.} % TODO consider moving to Appendix
-To improve our approximation, we can consider the second term of the Taylor series expansion,
-\begin{align*}
-\left(1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}\right)^{\frac{1}{2}}=1+\frac{2\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}-\frac{\left(\lambda_avp_m\right)^2}{4\left(\lambda_a-\mu_a-v\right)^4}+\cdots,
-\end{align*}
-which gives us the following approximation,
-\begin{align}
-p_a \approx \frac{\lambda_a-\mu_a-v}{\lambda_a}+\frac{vp_m}{\lambda_a-\mu_a-v}-\frac{\lambda_a\left(vp_m\right)^2}{8\left(\lambda_a-\mu_a-v\right)^3} .
-\end{align}
-We therefore have
-\begin{align}\nonumber
-p_w&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{u\left(\lambda_a-\mu_a-v\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-v\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}-\frac{uv^2\lambda_a\left(\lambda_m-\mu_m\right)^2}{8\lambda_m^2\left(\lambda_a-\mu_a-v\right)^3}\right]\\ \label{survprobw3}
-&\approx-\frac{1}{\lambda_w-\mu_w}\left[\frac{u\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}-\frac{uv^2\lambda_a\left(\lambda_m-\mu_m\right)^2}{8\lambda_m^2\left(\lambda_a-\mu_a\right)^3}\right],
-\end{align}
-and using $\Delta_k=\lambda_k-\mu_k$, we can write the above equation as
-\begin{equation}\label{survprobwapproxcorrected}
-p_w \approx -\frac{1}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}-\frac{uv^2\lambda_a\Delta_m^2}{8\lambda_m^2\Delta_a^3}\right).
-\end{equation}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{Scenario 2: Aneuploid cells are tolerant.}
-
-We now assume that aneuploidy provides tolerance to drug therapy, that is, the number of aneuploid cells significantly declines over time, but at a lower rate than the number of wildtype cells, $\lambda_w - \mu_w < \lambda_a - \mu_a < 0$.
-We also assume that the decline are significant, $\left(\lambda_a-\mu_a-v\right)^2 > 4\lambda_a v p_m$.
-We rewrite \cref{survival_prob} as
-\begin{equation}
-\begin{aligned}
-p_w&=\frac{\lambda_w-\mu_w-u-v}{2\lambda_w}\left(1-\sqrt{1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}}\right) ,
-\text{and} \\
-p_a&=\frac{\lambda_a-\mu_a-v}{2\lambda_a}\left(1-\sqrt{1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}}\right) .
-\end{aligned}
-\end{equation}
-Since $u,v\ll1$, the term in the root can be approximated using a 1st-order Taylor expansion. So, substituting the expressions for $p_a$ and $p_m$, we have
-\begin{equation}\label{survprobwinitial}
-\begin{aligned}
-p_w&\approx-\frac{vp_m+up_a}{\lambda_w-\mu_w-u-v}\\
-&\approx\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-v\right)}-\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\ %\label{survprobw2}
-&\approx\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_w-\mu_w\right)}\left[\frac{u}{\left(\lambda_a-\mu_a\right)}-1\right]\\
-%&=\frac{v\Delta_m}{\lambda_m\Delta_w}\left(\frac{u}{\Delta_a}-1\right) .
-\end{aligned}
-\end{equation}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{Scenario 3: Aneuploid cells are non-growing} % TODO consider this name
-
-We now assume that the growth rate of aneuploid cells is close to zero (either positive or negative), such that  $\left(\lambda_a-\mu_a-v\right)^2 < 4\lambda_avp_m$.
-We rewrite \cref{survival_prob} as
-\begin{equation}
-p_a=\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}\left(1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\lambda_avp_m}\right)^{\frac12}}{2\lambda_a} .
-\end{equation}
-Using a following Taylor series expansion
-\begin{equation*}
-\left(1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\lambda_avp_m}\right)^{\frac{1}{2}}=1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{8\lambda_avp_m}+\cdots,
-\end{equation*}
-we obtain the approximation
-\begin{equation}
-\begin{aligned}
-p_a&\approx\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}\left[1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{8\lambda_avp_m}\right]}{2\lambda_a}\\
-&=\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\sqrt{\lambda_avp_m}}}{2\lambda_a}\\
-&=\frac{\left(\lambda_a-\mu_a-v+2\sqrt{\lambda_avp_m}\right)^2+4\lambda_avp_m}{8\lambda_a\sqrt{\lambda_avp_m}}\\
-&=\frac{4\lambda_avp_m+4\lambda_avp_m\left(1+\frac{\lambda_a-\mu_a-v}{2\sqrt{\lambda_avp_m}}\right)^2}{8\lambda_a\sqrt{\lambda_avp_m}}\\
-&=\frac{1}{2\lambda_a}\left(\lambda_a-\mu_a-v+2\sqrt{\lambda_avp_m}\right).
-\end{aligned}
-\end{equation}
-Plugging this in \cref{survprobwinitial}, the survival probability of a population starting from one wildtype individual is
-\begin{equation}\label{scenario3}
-\begin{aligned}
-p_w&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[v\frac{\lambda_m-\mu_m}{\lambda_m}+\frac{u}{2\lambda_a}\left(\lambda_a-\mu_a-v+2\sqrt{\lambda_avp_m}\right)\right]\\
-&=-\frac{1}{\lambda_w-\mu_w-u-v}\left[v\frac{\lambda_m-\mu_m}{\lambda_m}+\frac{u}{2\lambda_a}\left(\lambda_a-\mu_a-v\right)+u\sqrt{\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_a\lambda_m}}\right].
-%&=-\frac{1}{\Delta_w-u-v}\left[v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right],
-\end{aligned}
-\end{equation}
 
@@
 % TODO summary
+
+\section{Methods}
+
+\subsection*{Calculation of survival probabilities of single lineages}
+
+Here we calculate the \emph{survival probability}, the probability that a lineage descended from a single cell does not become extinct.
+Let $p_w$, $p_a$, and $p_m$ be the survival probabilities of a lineage started by a single wildtype cell, aneuploid cell, or mutant cell, respectively.
+We will neglect density dependence, allowing us to treat the lineages as \emph{multitype branching processes} \citep{harris1963theory}.
+The extinction probabilities $1-p_w$, $1-p_a$, and $1-p_m$ then satisfy the equations:
+\begin{equation} \label{extinction_prob}
+\begin{aligned}
+1-p_w = &\frac{\mu_w}{\lambda_w+\mu_w+u+v} + 
+		  \frac{u}{\lambda_w+\mu_w+u+v}\left(1-p_a\right) + \\
+		  & \frac{\lambda_w}{\lambda_w+\mu_w+u+v}\left(1-p_w\right)^2 +
+		  \frac{v}{\lambda_w+\mu_w+u+v}\left(1-p_m\right) ,\\
+1-p_a = &\frac{\mu_a}{\lambda_a+\mu_a+v}+\frac{v}{\lambda_a+\mu_a+v}\left(1-p_m\right)+\frac{\lambda_a}{\lambda_a+\mu_a+v}\left(1-p_a\right)^2 ,\\
+1-p_m = &\frac{\mu_m}{\lambda_m+\mu_m}+\frac{\lambda_m}{\lambda_m+\mu_m}\left(1-p_m\right)^2 .	 
+\end{aligned}
+\end{equation}
+
+The survival probabilities are given by the smallest solution for each quadratic equation \citep{weissman2009rate,uecker2015adaptive}. Therefore we have
+\begin{equation}\label{survival_prob}
+\begin{aligned}
+p_w &= \frac{\lambda_w-\mu_w-u-v+\sqrt{\left(\lambda_w-\mu_w-u-v\right)^2+4\lambda_w\left(up_a+vp_m\right)}}{2\lambda_w} ,\\
+p_a &= \frac{\lambda_a-\mu_a-v+\sqrt{\left(\lambda_a-\mu_a-v\right)^2+4\lambda_avp_m}}{2\lambda_a}, \\
+p_m &= \frac{\lambda_m-\mu_m}{\lambda_m} .
+\end{aligned} 
+\end{equation}
+Note that the equation for $p_w$ depends on both $p_a$ and $p_m$, and the equation for $p_a$ depends on $p_m$.
+To proceed, we can plug the solution for $p_m$ and $p_a$ into the solution for $p_w$. We perform this for three different scenarios.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsubsection*{Scenario 1: Aneuploid cells are partially resistant}
+
+We first assume that aneuploidy provides partial resistance to drug therapy, $\lambda_a>\mu_a$, and that this resistance is significant, $\left(\lambda_a-\mu_a-v\right)^2 > 4\lambda_a v p_m$.
+We thus rewrite \cref{survival_prob} as
+\begin{align*}
+p_w&=\frac{\lambda_w-\mu_w-u-v}{2\lambda_w}\left(1-\sqrt{1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}}\right) ,
+\text{and} \\
+p_a&=\frac{\lambda_a-\mu_a-v}{2\lambda_a}\left(1+\sqrt{1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}}\right) . 
+\end{align*}
+Using the quadratic Taylor expansion $\sqrt{1+x}=1+x/2+O(x^2)$ and assuming $u,v \ll 1$,
+we obtain the following approximation for the survival probability of a population initially consisting of a single wildtype cell,
+\begin{align}\label{survprobwapprox1}
+p_w 
+&\approx -\frac{vp_m+up_a}{\lambda_w-\mu_w-u-v}\\
+\nonumber
+%&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{v\left(\lambda_a-\mu_a-u\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-u\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\ \label{survprobw2}
+&\approx-\frac{1}{\lambda_w-\mu_w}\left[\frac{u\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\
+\end{align}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\paragraph{Second-order approximation.} % TODO consider moving to Appendix
+To improve our approximation, we can consider the second term of the Taylor series expansion,
+\begin{align*}
+\left(1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}\right)^{\frac{1}{2}}=1+\frac{2\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}-\frac{\left(\lambda_avp_m\right)^2}{4\left(\lambda_a-\mu_a-v\right)^4}+\cdots,
+\end{align*}
+which gives us the following approximation,
+\begin{align}
+p_a \approx \frac{\lambda_a-\mu_a-v}{\lambda_a}+\frac{vp_m}{\lambda_a-\mu_a-v}-\frac{\lambda_a\left(vp_m\right)^2}{8\left(\lambda_a-\mu_a-v\right)^3} .
+\end{align}
+We therefore have
+\begin{align}\nonumber
+p_w&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{u\left(\lambda_a-\mu_a-v\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-v\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}-\frac{uv^2\lambda_a\left(\lambda_m-\mu_m\right)^2}{8\lambda_m^2\left(\lambda_a-\mu_a-v\right)^3}\right]\\ \label{survprobw3}
+&\approx-\frac{1}{\lambda_w-\mu_w}\left[\frac{u\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}-\frac{uv^2\lambda_a\left(\lambda_m-\mu_m\right)^2}{8\lambda_m^2\left(\lambda_a-\mu_a\right)^3}\right],
+\end{align}
+and using $\Delta_k=\lambda_k-\mu_k$, we can write the above equation as
+\begin{equation}\label{survprobwapproxcorrected}
+p_w \approx -\frac{1}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}-\frac{uv^2\lambda_a\Delta_m^2}{8\lambda_m^2\Delta_a^3}\right).
+\end{equation}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsubsection*{Scenario 2: Aneuploid cells are tolerant.}
+
+We now assume that aneuploidy provides tolerance to drug therapy, that is, the number of aneuploid cells significantly declines over time, but at a lower rate than the number of wildtype cells, $\lambda_w - \mu_w < \lambda_a - \mu_a < 0$.
+We also assume that the decline are significant, $\left(\lambda_a-\mu_a-v\right)^2 > 4\lambda_a v p_m$.
+We rewrite \cref{survival_prob} as
+\begin{equation}
+\begin{aligned}
+p_w&=\frac{\lambda_w-\mu_w-u-v}{2\lambda_w}\left(1-\sqrt{1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}}\right) ,
+\text{and} \\
+p_a&=\frac{\lambda_a-\mu_a-v}{2\lambda_a}\left(1-\sqrt{1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}}\right) .
+\end{aligned}
+\end{equation}
+Since $u,v\ll1$, the term in the root can be approximated using a 1st-order Taylor expansion. So, substituting the expressions for $p_a$ and $p_m$, we have
+\begin{equation}\label{survprobwinitial}
+\begin{aligned}
+p_w&\approx-\frac{vp_m+up_a}{\lambda_w-\mu_w-u-v}\\
+&\approx\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-v\right)}-\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\ %\label{survprobw2}
+&\approx\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_w-\mu_w\right)}\left[\frac{u}{\left(\lambda_a-\mu_a\right)}-1\right]\\
+%&=\frac{v\Delta_m}{\lambda_m\Delta_w}\left(\frac{u}{\Delta_a}-1\right) .
+\end{aligned}
+\end{equation}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsubsection*{Scenario 3: Aneuploid cells are non-growing} % TODO consider this name
+
+We now assume that the growth rate of aneuploid cells is close to zero (either positive or negative), such that  $\left(\lambda_a-\mu_a-v\right)^2 < 4\lambda_avp_m$.
+We rewrite \cref{survival_prob} as
+\begin{equation}
+p_a=\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}\left(1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\lambda_avp_m}\right)^{\frac12}}{2\lambda_a} .
+\end{equation}
+Using a following Taylor series expansion
+\begin{equation*}
+\left(1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\lambda_avp_m}\right)^{\frac{1}{2}}=1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{8\lambda_avp_m}+\cdots,
+\end{equation*}
+we obtain the approximation
+\begin{equation}
+\begin{aligned}
+p_a&\approx\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}\left[1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{8\lambda_avp_m}\right]}{2\lambda_a}\\
+&=\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\sqrt{\lambda_avp_m}}}{2\lambda_a}\\
+&=\frac{\left(\lambda_a-\mu_a-v+2\sqrt{\lambda_avp_m}\right)^2+4\lambda_avp_m}{8\lambda_a\sqrt{\lambda_avp_m}}\\
+&=\frac{4\lambda_avp_m+4\lambda_avp_m\left(1+\frac{\lambda_a-\mu_a-v}{2\sqrt{\lambda_avp_m}}\right)^2}{8\lambda_a\sqrt{\lambda_avp_m}}\\
+&=\frac{1}{2\lambda_a}\left(\lambda_a-\mu_a-v+2\sqrt{\lambda_avp_m}\right).
+\end{aligned}
+\end{equation}
+Plugging this in \cref{survprobwinitial}, the survival probability of a population starting from one wildtype individual is
+\begin{equation}\label{scenario3}
+\begin{aligned}
+p_w&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[v\frac{\lambda_m-\mu_m}{\lambda_m}+\frac{u}{2\lambda_a}\left(\lambda_a-\mu_a-v+2\sqrt{\lambda_avp_m}\right)\right]\\
+&=-\frac{1}{\lambda_w-\mu_w-u-v}\left[v\frac{\lambda_m-\mu_m}{\lambda_m}+\frac{u}{2\lambda_a}\left(\lambda_a-\mu_a-v\right)+u\sqrt{\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_a\lambda_m}}\right].
+%&=-\frac{1}{\Delta_w-u-v}\left[v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right],
+\end{aligned}
+\end{equation}
 

### 2023-06-21 EvolutionaryRescue1.tex
comments

@@
 
-To estimate the rescue probability $p_{rescue}$, we assume independence between clonal lineages starting from an initial population of $N$ wildtype cells (we check the effect of density-dependent growth on our results below).
+To estimate the rescue probability $\presc$, we assume independence between clonal lineages starting from an initial population of $N$ wildtype cells (we check the effect of density-dependent growth on our results below).
 Thus, the rescue probability is given by 
@@
 \begin{align}
-p_{rescue} &= 1-\left(1-p_w\right)^N \label{rescue_prob} \\
+\presc &= 1-\left(1-p_w\right)^N \label{rescue_prob} \\
 & \approx 1-\e^{-Np_w} , \label{rescue_prob2}
@@
 \end{subequations}
-where the approximation $(1-p_w)\approx e^{-p_w}$ assumes that $p_w$ (but not $N p_w$) is small.
-
-Applying the approximations for the survival probability $p_w$ from \cref{survprobwapprox1,survprobwinitial,scenario3} in \cref{rescue_prob2} and substituting $\Delta_k=\lambda_k-\mu_k$, we find that the rescue probability can be approximated by
+where $p_w$ is the probability that the lineage of a single wild-type cell avoids extinction, and the approximation $(1-p_w)\approx e^{-p_w}$ assumes that $p_w$ (but not $N p_w$) is small.
+
+In the Appendix below, we use the theory of multi-type branching processes to find approximate expressions \cref{survprobwapprox1,survprobwinitial,scenario3} for $p_w$ in different regimes.
+Substituting these  into \cref{rescue_prob2}, we find that the rescue probability can be approximated by
 \begin{equation}\label{rescue_prob_approx}
 \begin{aligned}
-&p_{rescue} \approx \\
+&\presc \approx \\
   &\begin{cases}
@@
 
+\daniel{Need to explain intuitively what all this means.}
+
 % TODO do we need this?
@@
 %\begin{align}\label{survprobresistantaneuploidy}
-%p_{rescue}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}-\frac{uv^2\lambda_a\Delta_m^2}{8\lambda_m^2\Delta_a^3}\right)\right].
+%\presc=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}-\frac{uv^2\lambda_a\Delta_m^2}{8\lambda_m^2\Delta_a^3}\right)\right].
 %\end{align}
@@
 \begin{align}\nonumber
-p_{rescue} 	&= p_{sgv}+\left(1-p_{sgv}\right)p_{de-novo} \\
+\presc 	&= p_{sgv}+\left(1-p_{sgv}\right)p_{de-novo} \\
 			&= 1-\exp\left(-\left[\left(1-f\right)p_w + fp_a\right]N\right) .
@@
 \begin{equation}\label{ratiorescueexact}
-H = \frac{p_{rescue}(u>0)}{p_{rescue}(u=0)}.
+H = \frac{\presc(u>0)}{\presc(u=0)}.
 \end{equation}
@@
 Therefore, the probability density function of $(T_1 \mid T_1 < \infty)$ is
-$f_1(t) = \frac{R\left(t\right)\exp\left(-\int_0^{t}{R(z)\d z}\right)}{p_{rescue}}$. 
+$f_1(t) = \frac{R\left(t\right)\exp\left(-\int_0^{t}{R(z)\d z}\right)}{\presc}$. 
 \\
@@
 \int_{0}^{\infty}{t f_1(t) \d t} = 
-\frac{\int_{0}^{\infty}{tR(t)\exp\left(-\int_0^{t}{R(z)\d z}\right) \d t}}{p_{rescue}} = 
-\frac{\int_{0}^{\infty}{\exp\left(-\int_0^{t}{R(z)\d z}\right) \d t}}{p_{rescue}}
+\frac{\int_{0}^{\infty}{tR(t)\exp\left(-\int_0^{t}{R(z)\d z}\right) \d t}}{\presc} = 
+\frac{\int_{0}^{\infty}{\exp\left(-\int_0^{t}{R(z)\d z}\right) \d t}}{\presc}
 \end{aligned}
@@
 
-\section{Methods}
+\section{Appendices}
 
@@
 \end{equation}
+
+\daniel{Rewrite the below in terms of $\Delta k$.}
 
@@
     $\mu_m$ & Mutant death rate& 0.13 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
+    $\Delta_k \equiv \lambda_k - \mu_k$ & 1/days & Growth rate of type $k$ cells, for $k=w,a,m$\\ \hline
     $u$ & Missegregation rate& $10^{-3}-10^{-2}$ & 1$\slash$cell division  & \citet{nowak2004evolutionary,bakker2023predicting} \\ \hline
@@
 \textbf{(B-left, B-right)} Aneuploidy provides tolerance.  
-Solid and dashed lines apply $p_{rescue}$ from the exact formula of  (\cref{survival_prob} in \cref{rescue_prob}); dashed lines apply $p_{rescue}$ from the approximate formula (\cref{rescue_prob_approx}), with good agreement.
+Solid and dashed lines apply $\presc$ from the exact formula of  (\cref{survival_prob} in \cref{rescue_prob}); dashed lines apply $\presc$ from the approximate formula (\cref{rescue_prob_approx}), with good agreement.
 Parameters: $N=10^4$; $\lambda_w=\lambda_a=\lambda_m=0.14$; (B) $\mu_w=0.17$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
@@
 %\begin{equation}\label{FellerApprox}
-%p_{rescue}=1-\e^{-\left(\theta_1+\theta_2\right)},
+%\presc=1-\e^{-\left(\theta_1+\theta_2\right)},
 %\end{equation}

### 2023-06-21 EvolutionaryRescue1.tex
text clean-ups

@@
 \newcommand{\presc}{p_\text{rescue}}
+\newcommand{\psgv}{p_\text{SGV}}
 
@@
 
-We validate these approximations by comparing them to results of stochastic evolutionary simulations.
-We find that the approximations work very well (\Cref{rescue_prob_wt_growth,rescue_prob_an_growth,rescue_prob_N,rescue_prob_N}).
+These approximations accurately match full stochastic simulations (\Cref{rescue_prob_wt_growth,rescue_prob_an_growth,rescue_prob_N,rescue_prob_N}).
 
@@
 
+\daniel{Need to explain why.}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-In this scenario, the probability of evolutionary rescue by cells with aneuploidy from the initial population is
+In this scenario, the probability $\psgv$ of evolutionary rescue via standing genetic variation, that is, by cells with aneuploidy from the initial population is:
 \begin{equation*}
 % redefine: p_total, p_denovo=p_resuce from eq above; p_stdvar is the new thing
-p_{sgv} = 1-\left(1-p_a\right)^{fN}\approx 1-\e^{-fNp_a}. 
+\psgv = 1-\left(1-p_a\right)^{fN}\approx 1-\e^{-fNp_a}. 
 \end{equation*}
+Here $p_a$ is the probability that a lineage started by an aneuploid cell avoids extinction; see (EQUATION) in the Appendix for its value.
 The total probability of evolutionary rescue is given by
 \begin{align}\nonumber
-\presc 	&= p_{sgv}+\left(1-p_{sgv}\right)p_{de-novo} \\
+\presc 	&= \psgv+\left(1-\psgv\right)p_\text{de novo} \\
 			&= 1-\exp\left(-\left[\left(1-f\right)p_w + fp_a\right]N\right) .
@@
 
-The fraction of cases in which the population is rescued by pre-existing aneuploid cells (i.e., standing genetic variation) is given by $F\left(f\right)=\frac{p_{sgv}}{p_{total}}$ (\Cref{FractionPlot}).
+The fraction of cases in which the population is rescued by pre-existing aneuploid cells (i.e., standing genetic variation) is given by $F\left(f\right)=\frac{\psgv}{p_\text{total}}$ (\Cref{FractionPlot}).
 
@@
 \subsection*{Effect of aneuploidy on evolutionary rescue}
-To determine the extent to which aneuploidy may affect evolutionary rescue, we define $H$ to be the ratio of the rescue probability with aneuploidy ($u>0$) and the rescue probability without aneuploidy ($u=0$),
+To determine the extent to which aneuploidy may affect evolutionary rescue, we define $H$ to be the ratio of the rescue probability with aneuploidy ($u>0$) and the rescue probability without aneuploidy ($u=0$):
 \begin{equation}\label{ratiorescueexact}
-H = \frac{\presc(u>0)}{\presc(u=0)}.
+H \equiv \frac{\presc(u>0)}{\presc(u=0)}.
 \end{equation}
@@
 \begin{equation}\label{ratiorescue}
-H = \begin{cases}
+H \approx \begin{cases}
 \frac{1-\exp\left[\frac{N}{\Delta_w-u-v}\left(v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right)\right]}{1-\exp\left[\frac{vN\Delta_m}{\left(\Delta_w-v\right)\lambda_m}\right]} ,&
-4\lambda_avp_m>\left(\Delta_a-v\right)^2 ,\\
-\frac{1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right]}{1-\exp\left(\frac{v\Delta_mN}{\lambda_m\Delta_w}\right)}, ,&
-\Delta_a<0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2 ,\\
+4\lambda_avp_m \gg \left(\Delta_a-v\right)^2 ,\\
+\frac{1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right]}{1-\exp\left(\frac{v\Delta_mN}{\lambda_m\Delta_w}\right)}, &
+\Delta_a < 0 \quad \text{and} \quad 4\lambda_avp_m \ll \left(\Delta_a-v\right)^2 ,\\
 \frac{1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right]}{1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\right]} ,&
-\Delta_a>0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2. % TODO this term can be simplified?
+\Delta_a>0 \quad \text{and} \quad 4\lambda_avp_m \ll \left(\Delta_a-v\right)^2. % TODO this term can be simplified?
   \end{cases}
 \end{equation}
-
-We find that the rescue ratio increase with the aneuploidy growth rate $\Delta_a$, because the better aneuploid cells are in growth, the better they are at rescuing the population (when they provide partial resistance) or delaying the extinction of the population (when they provide tolerance).
+\daniel{These expressions are impenetrable. Need to be simplified and interpreted.}
+
+Aneuploidy always increases the probability of rescue ($H > 1$). 
+We find that the rescue ratio increase with the aneuploidy growth rate $\Delta_a$, because the better aneuploid cells are in growth, the better they are at directly rescuing the population (when they provide partial resistance) or avoiding extinction long enough to acquire full resistance (when they provide tolerance).
 However, the rescue decreases with the wildtype growth rate $\Delta_w$, because the better the wildtype is at growth, the less is depends on aneuploidy for rescue or delay, and the more likely it is to directly produce mutant cells, rather than relying on aneuploid cells for producing mutant cells (\Cref{rescue_ratio}).
 The effect of the initial tumor size $N$ is the similar to that of the wildtype growth rate. 
-Importantly, in large tumors, the ratio converges to unity, that is, aneuploidy does not affect the probability for evolutionary rescue. % TODO can we find this analytical?
+If the tumor is sufficiently large, then even in the absence of aneuploidy rescue is almost certain to occur, and therefore aneuploidy has only a negligible additional effect.
+% TODO can we find this analytical?
 

### 2023-06-21 EvolutionaryRescue1.tex
started on time section

@@
 
-Even evolutionary rescue occurs, it may take a long time; therefore, it is crucial to estimate the mean waiting time for rescue and the effect aneuploidy may have on it. 
+Even if evolutionary rescue occurs, it may take a long time; therefore, it is crucial to estimate the mean waiting time for rescue and the effect aneuploidy may have on it. 
+\daniel{I'm not sure about this.}
 We therefore calculate the mean time for the appearance of the first mutant that rescues the cell population.
+\daniel{Need to be totally clear here that we're conditioning on rescue.}
 This can occur either through the evolutionary trajectory $wildtype \rightarrow mutant$ or through the trajectory $wildtype \rightarrow aneuploid \rightarrow mutant$.

### 2023-07-16 EvolutionaryRescue1.tex
add fig

--- 2023-06-21 EvolutionaryRescue1.tex
@@
  \usepackage{subcaption}
-\usepackage{xcolor}
 
@@
 \usepackage{newtxtext} %loads helv for ss, txtt for tt
+\usepackage{amsmath}
 \usepackage[bigdelims]{newtxmath}
@@
 
-% Daniel commands
-
-\newcommand{\daniel}[1]{\textcolor{blue}{#1}}
-\newcommand{\presc}{p_\text{rescue}}
-
 % Supplementary
@@
 The difference between the division and death rate is $\Delta_k = \lambda_k-\mu_k$.
-We assume the population of cells is under a strong stress, such as drug therapy, to which the wildtype genotype is susceptible and therefore $\Delta_w<0$, whereas the mutant is resistant to the stress, $\Delta_m>0$.
-We analyze three scenarios: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_w$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing or "barely growing", that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$, in a sense that we will make more precise.
+We assume the population of cells is under a strong stress, such as drug therapy, to which the wildtype genotype is susceptible and therefore $\Delta_w<0$, whereas the mutant is resistant to the stress, $\Delta_m>0$
+We analyze three scenarios: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_w$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing or "barely growing", that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$.
 Wildtype cells may missegregate to become aneuploids at rate $u$. Both aneuploid and wildtype cells may mutate to become mutants at rate $v$ (\Cref{figureAneuploidy}).
@@
 Each iteration of the simulation loop starts with computing the rates $\nu_j$ of each event $j$.
-We then draw the time until the next event, $\Delta t$, from an exponential distribution whose rate parameter is the sum of the rates of all events, such that $\Delta t \sim \text{Exp}(\sum_j \nu_j)$.
+We then draw the time until the next event, $\Delta t$, from an exponential distribution whose rate parameter is the sum of the rates of all events, such that $\Delta t \sim \textit{Exp}(\sum_j \nu_j)$.
 Then, we randomly determine which event occurred, where the probability for event $j$ is $p_j=\nu_j/\sum_i \nu_i$.
@@
 \begin{equation*}
-m_t>\left\lfloor\frac{3\log10}{\log\left(\frac{\lambda_m}{\mu_m}\right)}\right\rfloor+1.
+m_t>\left\lfloor-\frac{3\log10}{\log\left(\frac{\mu_m}{\lambda_m}\right)}\right\rfloor+1,
 \end{equation*}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection*{Survival probability}
+
+To analyze evolutionary rescue in this model, we use the framework of \emph{multitype branching processes} \citep{rybnikov2021fitness,harris1963theory}. 
+This allows us to find explicit expressions for the \emph{survival probability}: the probability that a lineage descended from a single cell does not become extinct.
+
+Let $p_w$, $p_a$, and $p_m$ be the survival probabilities of a population consisting initially of single wildtype cell, aneuploid cell, or mutant cell, respectively.
+The complements $1-p_w$, $1-p_a$, and $1-p_m$ are the extinction probabilities, which satisfy each its respective equation,
+\begin{equation} \label{extinction_prob}
+\begin{aligned}
+1-p_w = &\frac{\mu_w}{\lambda_w+\mu_w+u+v} + 
+		  \frac{u}{\lambda_w+\mu_w+u+v}\left(1-p_a\right) + \\
+		  & \frac{\lambda_w}{\lambda_w+\mu_w+u+v}\left(1-p_w\right)^2 +
+		  \frac{v}{\lambda_w+\mu_w+u+v}\left(1-p_m\right) ,\\
+1-p_a = &\frac{\mu_a}{\lambda_a+\mu_a+v}+\frac{v}{\lambda_a+\mu_a+v}\left(1-p_m\right)+\frac{\lambda_a}{\lambda_a+\mu_a+v}\left(1-p_a\right)^2 ,\\
+1-p_m = &\frac{\mu_m}{\lambda_m+\mu_m}+\frac{\lambda_m}{\lambda_m+\mu_m}\left(1-p_m\right)^2 .	 
+\end{aligned}
+\end{equation}
+
+The survival probabilities are given by the smallest solution for each quadratic equation \citep{uecker2015adaptive}. Therefore we have
+\begin{equation}\label{survival_prob}
+\begin{aligned}
+p_w &= \frac{\lambda_w-\mu_w-u-v+\sqrt{\left(\lambda_w-\mu_w-u-v\right)^2+4\lambda_w\left(up_a+vp_m\right)}}{2\lambda_w} ,\\
+p_a &= \frac{\lambda_a-\mu_a-v+\sqrt{\left(\lambda_a-\mu_a-v\right)^2+4\lambda_avp_m}}{2\lambda_a}, \\
+p_m &= \frac{\lambda_m-\mu_m}{\lambda_m} .
+\end{aligned} 
+\end{equation}
+Note that the equation for $p_w$ depends on both $p_a$ and $p_m$, and the equation for $p_a$ depends on $p_m$.
+To proceed, we can plug the solution for $p_m$ and $p_a$ into the solution for $p_w$. We perform this for three different scenarios.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsubsection*{Scenario 1: Aneuploid cells are partially resistant}
+
+We first assume that aneuploidy provides partial resistance to drug therapy, $\lambda_a>\mu_a$, and that this resistance is significant, $\left(\lambda_a-\mu_a-v\right)^2 > 4\lambda_a v p_m$.
+We thus rewrite \cref{survival_prob} as
+\begin{align*}
+p_w&=\frac{\lambda_w-\mu_w-u-v}{2\lambda_w}\left(1-\sqrt{1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}}\right) ,
+\text{and} \\
+p_a&=\frac{\lambda_a-\mu_a-v}{2\lambda_a}\left(1+\sqrt{1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}}\right) . 
+\end{align*}
+Using the quadratic Taylor expansion $\sqrt{1+x}=1+x/2+O(x^2)$ and assuming $u,v \ll 1$,
+we obtain the following approximation for the survival probability of a population initially consisting of a single wildtype cell,
+\begin{align}\label{survprobwapprox1}
+p_w 
+&\approx -\frac{vp_m+up_a}{\lambda_w-\mu_w-u-v}\\
+\nonumber
+%&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{v\left(\lambda_a-\mu_a-u\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-u\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\ \label{survprobw2}
+&\approx-\frac{1}{\lambda_w-\mu_w}\left[\frac{u\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\
+\end{align}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\paragraph{Second-order approximation.} % TODO consider moving to Appendix
+To improve our approximation, we can consider the second term of the Taylor series expansion,
+\begin{align*}
+\left(1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}\right)^{\frac{1}{2}}=1+\frac{2\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}-\frac{\left(\lambda_avp_m\right)^2}{4\left(\lambda_a-\mu_a-v\right)^4}+\cdots,
+\end{align*}
+which gives us the following approximation,
+\begin{align}
+p_a \approx \frac{\lambda_a-\mu_a-v}{\lambda_a}+\frac{vp_m}{\lambda_a-\mu_a-v}-\frac{\lambda_a\left(vp_m\right)^2}{8\left(\lambda_a-\mu_a-v\right)^3} .
+\end{align}
+We therefore have
+\begin{align}\nonumber
+p_w&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{u\left(\lambda_a-\mu_a-v\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-v\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}-\frac{uv^2\lambda_a\left(\lambda_m-\mu_m\right)^2}{8\lambda_m^2\left(\lambda_a-\mu_a-v\right)^3}\right]\\ \label{survprobw3}
+&\approx-\frac{1}{\lambda_w-\mu_w}\left[\frac{u\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}-\frac{uv^2\lambda_a\left(\lambda_m-\mu_m\right)^2}{8\lambda_m^2\left(\lambda_a-\mu_a\right)^3}\right],
+\end{align}
+and using $\Delta_k=\lambda_k-\mu_k$, we can write the above equation as
+\begin{equation}\label{survprobwapproxcorrected}
+p_w \approx -\frac{1}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}-\frac{uv^2\lambda_a\Delta_m^2}{8\lambda_m^2\Delta_a^3}\right).
+\end{equation}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsubsection*{Scenario 2: Aneuploid cells are tolerant.}
+
+We now assume that aneuploidy provides tolerance to drug therapy, that is, the number of aneuploid cells significantly declines over time, but at a lower rate than the number of wildtype cells, $\lambda_w - \mu_w < \lambda_a - \mu_a < 0$.
+We also assume that the decline are significant, $\left(\lambda_a-\mu_a-v\right)^2 > 4\lambda_a v p_m$.
+We rewrite \cref{survival_prob} as
+\begin{equation}
+\begin{aligned}
+p_w&=\frac{\lambda_w-\mu_w-u-v}{2\lambda_w}\left(1-\sqrt{1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}}\right) ,
+\text{and} \\
+p_a&=\frac{\lambda_a-\mu_a-v}{2\lambda_a}\left(1-\sqrt{1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}}\right) .
+\end{aligned}
+\end{equation}
+Since $u,v\ll1$, the term in the root can be approximated using a 1st-order Taylor expansion. So, substituting the expressions for $p_a$ and $p_m$, we have
+\begin{equation}\label{survprobwinitial}
+\begin{aligned}
+p_w&\approx-\frac{vp_m+up_a}{\lambda_w-\mu_w-u-v}\\
+&\approx\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-v\right)}-\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\ %\label{survprobw2}
+&\approx\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_w-\mu_w\right)}\left[\frac{u}{\left(\lambda_a-\mu_a\right)}-1\right]\\
+%&=\frac{v\Delta_m}{\lambda_m\Delta_w}\left(\frac{u}{\Delta_a}-1\right) .
+\end{aligned}
+\end{equation}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsubsection*{Scenario 3: Aneuploid cells are non-growing} % TODO consider this name
+
+We now assume that the growth rate of aneuploid cells is close to zero (either positive or negative), such that  $\left(\lambda_a-\mu_a-v\right)^2 < 4\lambda_avp_m$.
+We rewrite \cref{survival_prob} as
+\begin{equation}
+p_a=\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}\left(1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\lambda_avp_m}\right)^{\frac12}}{2\lambda_a} .
+\end{equation}
+Using a following Taylor series expansion
+\begin{equation*}
+\left(1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\lambda_avp_m}\right)^{\frac{1}{2}}=1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{8\lambda_avp_m}+\cdots,
+\end{equation*}
+we obtain the approximation
+\begin{equation}
+\begin{aligned}
+p_a&\approx\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}\left[1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{8\lambda_avp_m}\right]}{2\lambda_a}\\
+&=\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\sqrt{\lambda_avp_m}}}{2\lambda_a}\\
+&=\frac{\left(\lambda_a-\mu_a-v+2\sqrt{\lambda_avp_m}\right)^2+4\lambda_avp_m}{8\lambda_a\sqrt{\lambda_avp_m}}\\
+&=\frac{4\lambda_avp_m+4\lambda_avp_m\left(1+\frac{\lambda_a-\mu_a-v}{2\sqrt{\lambda_avp_m}}\right)^2}{8\lambda_a\sqrt{\lambda_avp_m}}\\
+&=\frac{1}{2\lambda_a}\left(\lambda_a-\mu_a-v+2\sqrt{\lambda_avp_m}\right).
+\end{aligned}
+\end{equation}
+Plugging this in \cref{survprobwinitial}, the survival probability of a population starting from one wildtype individual is
+\begin{equation}\label{scenario3}
+\begin{aligned}
+p_w&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[v\frac{\lambda_m-\mu_m}{\lambda_m}+\frac{u}{2\lambda_a}\left(\lambda_a-\mu_a-v+2\sqrt{\lambda_avp_m}\right)\right]\\
+&=-\frac{1}{\lambda_w-\mu_w-u-v}\left[v\frac{\lambda_m-\mu_m}{\lambda_m}+\frac{u}{2\lambda_a}\left(\lambda_a-\mu_a-v\right)+u\sqrt{\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_a\lambda_m}}\right].
+%&=-\frac{1}{\Delta_w-u-v}\left[v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right],
+\end{aligned}
+\end{equation}
 
@@
 
-To estimate the rescue probability $\presc$, we assume independence between clonal lineages starting from an initial population of $N$ wildtype cells (we check the effect of density-dependent growth on our results below).
+To estimate the rescue probability $p_{rescue}$, we assume independence between clonal lineages starting from an initial population of $N$ wildtype cells (we check the effect of density-dependent growth on our results below).
 Thus, the rescue probability is given by 
@@
 \begin{align}
-\presc &= 1-\left(1-p_w\right)^N \label{rescue_prob} \\
+p_{rescue} &= 1-\left(1-p_w\right)^N \label{rescue_prob} \\
 & \approx 1-\e^{-Np_w} , \label{rescue_prob2}
@@
 \end{subequations}
-where $p_w$ is the probability that the lineage of a single wild-type cell avoids extinction, and the approximation $(1-p_w)\approx e^{-p_w}$ assumes that $p_w$ (but not $N p_w$) is small.
-
-In the Appendix below, we use the theory of multi-type branching processes to find approximate expressions \cref{survprobwapprox1,survprobwinitial,scenario3} for $p_w$ in different regimes.
-Substituting these  into \cref{rescue_prob2}, we find that the rescue probability can be approximated by
+where the approximation $(1-p_w)\approx e^{-p_w}$ assumes that $p_w$ (but not $N p_w$) is small.
+
+Applying the approximations for the survival probability $p_w$ from \cref{survprobwapprox1,survprobwinitial,scenario3} in \cref{rescue_prob2} and substituting $\Delta_k=\lambda_k-\mu_k$, we find that the rescue probability can be approximated by
 \begin{equation}\label{rescue_prob_approx}
 \begin{aligned}
-&\presc \approx \\
+&p_{rescue} \approx \\
   &\begin{cases}
@@
 
-\daniel{Need to explain intuitively what all this means.}
-
 % TODO do we need this?
@@
 %\begin{align}\label{survprobresistantaneuploidy}
-%\presc=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}-\frac{uv^2\lambda_a\Delta_m^2}{8\lambda_m^2\Delta_a^3}\right)\right].
+%p_{rescue}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}-\frac{uv^2\lambda_a\Delta_m^2}{8\lambda_m^2\Delta_a^3}\right)\right].
 %\end{align}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\paragraph*{Density-dependent growth}
+\paragraph*{Density-dependent growth.}
 
@@
 \begin{align}\nonumber
-\presc 	&= p_{sgv}+\left(1-p_{sgv}\right)p_{de-novo} \\
+p_{rescue} 	&= p_{sgv}+\left(1-p_{sgv}\right)p_{de-novo} \\
 			&= 1-\exp\left(-\left[\left(1-f\right)p_w + fp_a\right]N\right) .
@@
 \begin{equation}\label{ratiorescueexact}
-H = \frac{\presc(u>0)}{\presc(u=0)}.
+H = \frac{p_{rescue}(u>0)}{p_{rescue}(u=0)}.
 \end{equation}
@@
 Therefore, the probability density function of $(T_1 \mid T_1 < \infty)$ is
-$f_1(t) = \frac{R\left(t\right)\exp\left(-\int_0^{t}{R(z)\d z}\right)}{\presc}$. 
+$f_1(t) = \frac{R\left(t\right)\exp\left(-\int_0^{t}{R(z)\d z}\right)}{p_{rescue}}$. 
 \\
@@
 \int_{0}^{\infty}{t f_1(t) \d t} = 
-\frac{\int_{0}^{\infty}{tR(t)\exp\left(-\int_0^{t}{R(z)\d z}\right) \d t}}{\presc} = 
-\frac{\int_{0}^{\infty}{\exp\left(-\int_0^{t}{R(z)\d z}\right) \d t}}{\presc}
+\frac{\int_{0}^{\infty}{tR(t)\exp\left(-\int_0^{t}{R(z)\d z}\right) \d t}}{p_{rescue}} = 
+\frac{\int_{0}^{\infty}{\exp\left(-\int_0^{t}{R(z)\d z}\right) \d t}}{p_{rescue}}
 \end{aligned}
@@
 
-When aneuploid cells are partially resistant to the drug ($\Delta_w\ll$(SOMETHING $\ne0$)$\ll\Delta_a\ll\Delta_m$), evolutionary rescue can be approximated by a one-step evolutionary rescue process where aneuploidy itself rescues the population (\Cref{rescue_prob_wt_growth}).
-When aneuploidy only provides tolerance to the drug ($\Delta_w\ll\Delta_a\ll$(SOMETHING $\ne0$)$\ll\Delta_m$), it cannot rescue the population.
+When aneuploid cells are partially resistant to the drug ($\Delta_w\ll0\ll\Delta_a\ll\Delta_m$), evolutionary rescue can be approximated by a one-step evolutionary rescue process where aneuploidy itself rescues the population (\Cref{rescue_prob_wt_growth}).
+When aneuploidy only provides tolerance to the drug ($\Delta_w\ll\Delta_a\ll0\ll\Delta_m$), it cannot rescue the population.
 Instead, aneuploidy acts as a \emph{stepping stone} through which the resistant mutant can appear in a more expedient fashion, given that the aneuploid cell population declines slower then the wildtype cell population.
@@
 % TODO summary
-
-\section{Appendices}
-
-\subsection*{Calculation of survival probabilities of single lineages}
-
-Here we calculate the \emph{survival probability}, the probability that a lineage descended from a single cell does not become extinct.
-Let $p_w$, $p_a$, and $p_m$ be the survival probabilities of a lineage started by a single wildtype cell, aneuploid cell, or mutant cell, respectively.
-We will neglect density dependence, allowing us to treat the lineages as \emph{multitype branching processes} \citep{harris1963theory}.
-The extinction probabilities $1-p_w$, $1-p_a$, and $1-p_m$ then satisfy the equations:
-\begin{equation} \label{extinction_prob}
-\begin{aligned}
-1-p_w = &\frac{\mu_w}{\lambda_w+\mu_w+u+v} + 
-		  \frac{u}{\lambda_w+\mu_w+u+v}\left(1-p_a\right) + \\
-		  & \frac{\lambda_w}{\lambda_w+\mu_w+u+v}\left(1-p_w\right)^2 +
-		  \frac{v}{\lambda_w+\mu_w+u+v}\left(1-p_m\right) ,\\
-1-p_a = &\frac{\mu_a}{\lambda_a+\mu_a+v}+\frac{v}{\lambda_a+\mu_a+v}\left(1-p_m\right)+\frac{\lambda_a}{\lambda_a+\mu_a+v}\left(1-p_a\right)^2 ,\\
-1-p_m = &\frac{\mu_m}{\lambda_m+\mu_m}+\frac{\lambda_m}{\lambda_m+\mu_m}\left(1-p_m\right)^2 .	 
-\end{aligned}
-\end{equation}
-
-\daniel{Rewrite the below in terms of $\Delta k$.}
-
-The survival probabilities are given by the smallest solution for each quadratic equation \citep{weissman2009rate,uecker2015adaptive}. Therefore we have
-\begin{equation}\label{survival_prob}
-\begin{aligned}
-p_w &= \frac{\lambda_w-\mu_w-u-v+\sqrt{\left(\lambda_w-\mu_w-u-v\right)^2+4\lambda_w\left(up_a+vp_m\right)}}{2\lambda_w} ,\\
-p_a &= \frac{\lambda_a-\mu_a-v+\sqrt{\left(\lambda_a-\mu_a-v\right)^2+4\lambda_avp_m}}{2\lambda_a}, \\
-p_m &= \frac{\lambda_m-\mu_m}{\lambda_m} .
-\end{aligned} 
-\end{equation}
-Note that the equation for $p_w$ depends on both $p_a$ and $p_m$, and the equation for $p_a$ depends on $p_m$.
-To proceed, we can plug the solution for $p_m$ and $p_a$ into the solution for $p_w$. We perform this for three different scenarios.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{Scenario 1: Aneuploid cells are partially resistant}
-
-We first assume that aneuploidy provides partial resistance to drug therapy, $\lambda_a>\mu_a$, and that this resistance is significant, $\left(\lambda_a-\mu_a-v\right)^2 > 4\lambda_a v p_m$.
-We thus rewrite \cref{survival_prob} as
-\begin{align*}
-p_w&=\frac{\lambda_w-\mu_w-u-v}{2\lambda_w}\left(1-\sqrt{1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}}\right) ,
-\text{and} \\
-p_a&=\frac{\lambda_a-\mu_a-v}{2\lambda_a}\left(1+\sqrt{1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}}\right) . 
-\end{align*}
-Using the quadratic Taylor expansion $\sqrt{1+x}=1+x/2+O(x^2)$ and assuming $u,v \ll 1$,
-we obtain the following approximation for the survival probability of a population initially consisting of a single wildtype cell,
-\begin{align}\label{survprobwapprox1}
-p_w 
-&\approx -\frac{vp_m+up_a}{\lambda_w-\mu_w-u-v}\\
-\nonumber
-%&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{v\left(\lambda_a-\mu_a-u\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-u\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\ \label{survprobw2}
-&\approx-\frac{1}{\lambda_w-\mu_w}\left[\frac{u\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\
-\end{align}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\paragraph{Second-order approximation.} % TODO consider moving to Appendix
-To improve our approximation, we can consider the second term of the Taylor series expansion,
-\begin{align*}
-\left(1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}\right)^{\frac{1}{2}}=1+\frac{2\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}-\frac{\left(\lambda_avp_m\right)^2}{4\left(\lambda_a-\mu_a-v\right)^4}+\cdots,
-\end{align*}
-which gives us the following approximation,
-\begin{align}
-p_a \approx \frac{\lambda_a-\mu_a-v}{\lambda_a}+\frac{vp_m}{\lambda_a-\mu_a-v}-\frac{\lambda_a\left(vp_m\right)^2}{8\left(\lambda_a-\mu_a-v\right)^3} .
-\end{align}
-We therefore have
-\begin{align}\nonumber
-p_w&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{u\left(\lambda_a-\mu_a-v\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-v\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}-\frac{uv^2\lambda_a\left(\lambda_m-\mu_m\right)^2}{8\lambda_m^2\left(\lambda_a-\mu_a-v\right)^3}\right]\\ \label{survprobw3}
-&\approx-\frac{1}{\lambda_w-\mu_w}\left[\frac{u\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}-\frac{uv^2\lambda_a\left(\lambda_m-\mu_m\right)^2}{8\lambda_m^2\left(\lambda_a-\mu_a\right)^3}\right],
-\end{align}
-and using $\Delta_k=\lambda_k-\mu_k$, we can write the above equation as
-\begin{equation}\label{survprobwapproxcorrected}
-p_w \approx -\frac{1}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}-\frac{uv^2\lambda_a\Delta_m^2}{8\lambda_m^2\Delta_a^3}\right).
-\end{equation}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{Scenario 2: Aneuploid cells are tolerant.}
-
-We now assume that aneuploidy provides tolerance to drug therapy, that is, the number of aneuploid cells significantly declines over time, but at a lower rate than the number of wildtype cells, $\lambda_w - \mu_w < \lambda_a - \mu_a < 0$.
-We also assume that the decline are significant, $\left(\lambda_a-\mu_a-v\right)^2 > 4\lambda_a v p_m$.
-We rewrite \cref{survival_prob} as
-\begin{equation}
-\begin{aligned}
-p_w&=\frac{\lambda_w-\mu_w-u-v}{2\lambda_w}\left(1-\sqrt{1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}}\right) ,
-\text{and} \\
-p_a&=\frac{\lambda_a-\mu_a-v}{2\lambda_a}\left(1-\sqrt{1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}}\right) .
-\end{aligned}
-\end{equation}
-Since $u,v\ll1$, the term in the root can be approximated using a 1st-order Taylor expansion. So, substituting the expressions for $p_a$ and $p_m$, we have
-\begin{equation}\label{survprobwinitial}
-\begin{aligned}
-p_w&\approx-\frac{vp_m+up_a}{\lambda_w-\mu_w-u-v}\\
-&\approx\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-v\right)}-\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\ %\label{survprobw2}
-&\approx\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_w-\mu_w\right)}\left[\frac{u}{\left(\lambda_a-\mu_a\right)}-1\right]\\
-%&=\frac{v\Delta_m}{\lambda_m\Delta_w}\left(\frac{u}{\Delta_a}-1\right) .
-\end{aligned}
-\end{equation}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{Scenario 3: Aneuploid cells are non-growing} % TODO consider this name
-
-We now assume that the growth rate of aneuploid cells is close to zero (either positive or negative), such that  $\left(\lambda_a-\mu_a-v\right)^2 < 4\lambda_avp_m$.
-We rewrite \cref{survival_prob} as
-\begin{equation}
-p_a=\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}\left(1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\lambda_avp_m}\right)^{\frac12}}{2\lambda_a} .
-\end{equation}
-Using a following Taylor series expansion
-\begin{equation*}
-\left(1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\lambda_avp_m}\right)^{\frac{1}{2}}=1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{8\lambda_avp_m}+\cdots,
-\end{equation*}
-we obtain the approximation
-\begin{equation}
-\begin{aligned}
-p_a&\approx\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}\left[1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{8\lambda_avp_m}\right]}{2\lambda_a}\\
-&=\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\sqrt{\lambda_avp_m}}}{2\lambda_a}\\
-&=\frac{\left(\lambda_a-\mu_a-v+2\sqrt{\lambda_avp_m}\right)^2+4\lambda_avp_m}{8\lambda_a\sqrt{\lambda_avp_m}}\\
-&=\frac{4\lambda_avp_m+4\lambda_avp_m\left(1+\frac{\lambda_a-\mu_a-v}{2\sqrt{\lambda_avp_m}}\right)^2}{8\lambda_a\sqrt{\lambda_avp_m}}\\
-&=\frac{1}{2\lambda_a}\left(\lambda_a-\mu_a-v+2\sqrt{\lambda_avp_m}\right).
-\end{aligned}
-\end{equation}
-Plugging this in \cref{survprobwinitial}, the survival probability of a population starting from one wildtype individual is
-\begin{equation}\label{scenario3}
-\begin{aligned}
-p_w&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[v\frac{\lambda_m-\mu_m}{\lambda_m}+\frac{u}{2\lambda_a}\left(\lambda_a-\mu_a-v+2\sqrt{\lambda_avp_m}\right)\right]\\
-&=-\frac{1}{\lambda_w-\mu_w-u-v}\left[v\frac{\lambda_m-\mu_m}{\lambda_m}+\frac{u}{2\lambda_a}\left(\lambda_a-\mu_a-v\right)+u\sqrt{\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_a\lambda_m}}\right].
-%&=-\frac{1}{\Delta_w-u-v}\left[v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right],
-\end{aligned}
-\end{equation}
 
@@
     $\mu_m$ & Mutant death rate& 0.13 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
-    $\Delta_k \equiv \lambda_k - \mu_k$ & 1/days & Growth rate of type $k$ cells, for $k=w,a,m$\\ \hline
     $u$ & Missegregation rate& $10^{-3}-10^{-2}$ & 1$\slash$cell division  & \citet{nowak2004evolutionary,bakker2023predicting} \\ \hline
@@
 \textbf{(B-left, B-right)} Aneuploidy provides tolerance.  
-Solid and dashed lines apply $\presc$ from the exact formula of  (\cref{survival_prob} in \cref{rescue_prob}); dashed lines apply $\presc$ from the approximate formula (\cref{rescue_prob_approx}), with good agreement.
+Solid and dashed lines apply $p_{rescue}$ from the exact formula of  (\cref{survival_prob} in \cref{rescue_prob}); dashed lines apply $p_{rescue}$ from the approximate formula (\cref{rescue_prob_approx}), with good agreement.
 Parameters: $N=10^4$; $\lambda_w=\lambda_a=\lambda_m=0.14$; (B) $\mu_w=0.17$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
@@
 % TODO parameter values 
+}
+\label{MeanTimeRatioInitialPopulationSize}
+\end{figure}
+
+%%%%%%%%%%%
+
+\begin{figure}
+ \vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/ThresholdPopulationSize.pdf}
+\caption{\textbf{TODO.}
+TODO
 }
@@
 %\begin{equation}\label{FellerApprox}
-%\presc=1-\e^{-\left(\theta_1+\theta_2\right)},
+%p_{rescue}=1-\e^{-\left(\theta_1+\theta_2\right)},
 %\end{equation}

### 2023-07-16 EvolutionaryRescue1.tex
yoav revised; updaed figs

@@
 
-Even evolutionary rescue occurs, it may take a long time; therefore, it is crucial to estimate the mean waiting time for rescue and the effect aneuploidy may have on it. 
-We therefore calculate the mean time for the appearance of the first mutant that rescues the cell population.
+Even when evolutionary rescue occurs, it may take a long time.
+We therefore wish to estimate the mean waiting time for rescue and the effect aneuploidy may have on it. 
+We calculate the mean time for the appearance of the first mutant that rescues the cell population.
 This can occur either through the evolutionary trajectory $wildtype \rightarrow mutant$ or through the trajectory $wildtype \rightarrow aneuploid \rightarrow mutant$.
@@
 Assuming no aneuploidy ($u=0$), we define $T_1$ to be the time at which the first mutant cell appears that will avoid extinction and will therefore rescue the population.
-Note that if extinction occurs, that is $m_{\infty}=0$, then it is implied that $T_1=\infty$, and vice versa if $T_1<\infty$ then $m_{\infty}>0$.
+Note that if extinction occurs, that is the frequency of mutants after a very long time is zero, $m_{\infty}=0$, then it is implied that $T_1=\infty$, and vice versa if $T_1<\infty$ then $m_{\infty}>0$.
 
@@
 $R\left(t\right)\exp\left(-\int_0^{t}{R(z)\d z}\right)$. % TODO ref
-Therefore, the probability density function of $(T_1 \mid T_1 < \infty)$ is
+Therefore, the probability density function of the conditional random variable $(T_1 \mid T_1 < \infty)$ is
 $f_1(t) = \frac{R\left(t\right)\exp\left(-\int_0^{t}{R(z)\d z}\right)}{p_{rescue}}$. 
@@
 after applying integration by parts.
-Therefore, plugging \cref{integralR,rescue_prob2} in \cref{meantime1}, the probability density function of $T_1$ is given by:
-\begin{align}
-f_{T_1}\left(t_1\right)=R\left(t_1\right)\e^{-\int_0^{t_1} R(t)\d t}.
-\end{align}
-As a result, the time $T_1$ conditional on evolutionary rescue is given by:
-\begin{align}
-f_{T_1}\left(t_1|m_{t\rightarrow\infty}\neq0\right)=\frac{R\left(t_1\right)\e^{-\int_0^{t_1} R(t)\d t}}{1-\left(1-p_w\right)^N}.
-\end{align}
-The expectation of $T_1$ is:
-\begin{align}\label{meantime1}
-\tau_1=\mathbb{E}\left[T_1\right]&=\frac{\int_0^\infty\e^{-\int_0^\tau R(t)\d t}\d\tau}{1-\left(1-p_w\right)^N}=\frac{\int_0^\infty\e^{-uNp_a\frac{\e^{\Delta_w\tau}-1}{\Delta_w}}\,\d\tau}{1-\left(1-p_w\right)^N}.
-\end{align}
-The fration in the exponential of the integrand in \eqref{meantime1} can approximated as:
-\begin{equation*}
-\frac{\e^{\Delta_w\tau}-1}{\Delta_w}=\frac{1+\Delta_w\tau+O(\tau^2)-1}{\Delta_w}=\tau+O(\tau^2).
-\end{equation*}
-As a result, the mean time $\tau_1$ can be simplified:
+Therefore, plugging \cref{integralR,rescue_prob2} in \cref{meantime1}, 
 \begin{align}\label{limitapprox}
 \tau_1 = 
-\frac{\int_{0}^{\infty}{\exp\left(-u p_a N t\right) \d t}}{1-\e^{-Np_w}} \approx
+%\frac{\int_0^\infty\exp\left(-\int_0^\tau R(z)\d z\right)\d t}{1-\left(1-p_w\right)^N} = 
+\frac{\int_0^\infty\e^{-uNp_a\frac{\e^{\Delta_w\tau}-1}{\Delta_w}}\,\d\tau}{1-\left(1-p_w\right)^N} \approx
+\frac{\int_{0}^{\infty}{\exp\left(-u p_a N t\right) \d t}}{1-\e^{-Np_w}} \approx \\
 \left(1+\e^{-Np_w}\right)\int_0^\infty\e^{-u p_a N t} \d t =
@@
 \end{align}
-where we use the approximation $\left(1-\e^{-Np_w}\right)^{-1}\approx1+\e^{-Np_w}$ and integrate the exponent.
-\Cref{MeanTimeGrowthAneuploidyPlot} show the agreement between this approximating and simulation results for intermediate and large tumor sizes.
-
-\textbf{THE REST OF THIS SECTION REQUIRES EDITING BY YOAV}
-
-%Let $T=T_1+T_2$ be the time the first mutant cell that appears and rescues the population, where $T_1$ is the time at which the first aneuploid cell appears whose lineage will rescue the population and $T_2$ the time for this aneuploidy lineage to produce the first mutant that rescues the population.
-The expected mean time $\tau_1=\mathbb{E}\left[T_1\right]$ is identical to \eqref{meantime1} with adequate parameters.
-%
-%For an aneuploid cell, let $a_\tau$ be the number of its aneuploid descendents at time $\tau$.
-Each of these descentends have a probability $v\d \tau$ of producing a mutant which has a probability $p_m$ to survive.
-Consequently, the probability that the lineage descendent from a single aneuploid cell produces succesful mutant in the time interval $\left[\tau,\tau+\d \tau\right]$ is $vp_ma_t\d \tau$.
-Since the number of succesful mutants is a Poisson process, the probability of evolutionary rescue occurring until time $t$, given $a_\tau$, is given by:
-%\begin{align}
-%1-\exp\left(-\int_0^t vp_ma_\tau\d \tau\right)=1-\exp\left(-vp_mA_t\right),
-%\end{align}
-%where $A_t$ is the weight of the lineage until time $t$, defined as:
-%\begin{equation}
-%A_t=\int_0^t a_\tau\d \tau.
-%\end{equation}
-%As a result, to obtain $p_a\left(t\right)$ we average over all possible values of $A_t$:
-%\begin{align}
-%p_a\left(t\right)&=\int_0^\infty \d a P\left(A_T=a\right)\left[1-\e^{-vp_mA_t}\right]\\
-%&=1-\mathbb{E}\left[\e^{-vp_mA_t}\right]
-%\end{align}
-%Following the steps outlines in \citep{weissman2009rate} we obtain:
-%\begin{align}
-%p_a\left(t\right)=\frac{\left(a_+-1\right)\left(1-a_-\right)\left(1-\e^{-\left(1-\delta\right)\left(a_+-a_-\right)t}\right)}{a_+-1+\left(1-a_-\right)\e^{-\left(1-\delta\right)\left(a_+-a_-\right)t}},
-%\end{align}
-%where
-%\begin{align}
-%a_\pm=\frac{2-\delta-vp_m\pm\sqrt{\left(\delta+vp_m\right)^2-4vp_m}}{2\left(1-\delta\right)}.
-%\end{align}
-%and
-%\begin{align}
-%\delta=\frac{\Delta_a}{\Delta_w}
-%\end{align}
-%
-%The cumulative distribution of $T_2$ is given by $\frac{p_a\left(t\right)}{p_m}$ from which we obtain the mean time:
-%\begin{align}
-%\tau_2=\mathbb{E}\left[T_2\right]=\int_0^\infty\left(1-\frac{p_a\left(t\right)}{p_a}\right)\d t.
-%\end{align}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-When $Nu\gg1$ the aneuploid population can be assumed to be deterministic and approximated by the solution to the system of ODEs:
+where we use the approximations 
+$\frac{\e^{\Delta_w\tau}-1}{\Delta_w}=\frac{1+\Delta_w\tau+O(\tau^2)-1}{\Delta_w}=\tau+O(\tau^2)$ and $(1-\e^{-Np_w})^{-1} \approx 1+\e^{-Np_w}$ and integrate the exponent.
+\Cref{MeanTimeGrowthAneuploidyPlot}B show the agreement between this approximating and simulation results for intermediate and large tumor sizes.
+\\
+
+When $Nu\gg1$ the aneuploid frequency dynamics is roughly deterministic and therefore can be approximated by 
 \begin{equation}
-a_t=\frac{Nu\e^{\Delta_wt}}{\Delta_w-\Delta_a}\left[1-\e^{\left(\Delta_w-\Delta_a\right)t}\right].
-\end{equation}
-As a result, when $N\gg1$ the number of successful mutants created by direct mutation or though aneuploidy are an inhomogeneous Poisson processes with the rates:
-
-%\begin{align}
-%f_{T_2}\left(t_2|m_{\infty}\neq0\right)=\frac{R_2\left(t_2\right)\e^{-\int_0^{t_2} R_2(t)\d t}}{1-\left(1-p_w\right)^N}
-%\end{align}
-
-%\begin{align}\label{meantime2}
-%\tau_2=\mathbb{E}\left[T_2\right]&=\frac{\int_0^\infty\e^{-\int_0^\tau R_2(t)\d t}\d\tau}{1-\left(1-p_w\right)^N}=\frac{\int_0^\infty\e^{-uNp_a\frac{\e^{\Delta_w\tau}-1}{\Delta_w}}\,\d\tau}{1-\left(1-p_w\right)^N}.
-%\end{align}
-%
-%\begin{align}\label{meantime3}
-%\tau_3=\frac{\int_0^\infty\e^{-vNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}}\,\d\tau}{1-\left(1-p_w\right)^N}.
-%\end{align}
-
+a_t \approx \frac{Nu\e^{\Delta_wt}}{\Delta_w-\Delta_a}\left[1-\e^{\left(\Delta_w-\Delta_a\right)t}\right].
+\end{equation}
+As a result, when $N\gg1$ % TODO N>>1 ? or Nu >> 1?
+the number of successful mutants created by direct mutation and via aneuploidy can be approximated by inhomogeneous Poisson processes with the rates
 \begin{align}
@@
 \end{align}
-For large initial population sizes we can assume that both rescue mutations produced through direct mutation and aneuploidy are independent and, as a result, they can be merged into a single Poisson process with rate $\left(r_1+r_2\right)\left(t\right)$.
-Consequently, the mean time to the appearance of the first rescue mutant is:
+For large initial population sizes we assume that the two processes are independent and as a result, they can be merged into a single Poisson process with rate $\left(r_1+r_2\right)\left(t\right)$.
+Consequently, the mean time to the appearance of the first rescue mutant is
 \begin{align}\label{meantimet2}
-\tau_2=\frac{\int_0^\infty\e^{-\left(r_1+r_2\right)}\,\d\tau}{1-\left(1-p_w\right)^N}=\frac{\int_0^\infty\exp\left[-\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_w\tau}-1}{\Delta_w}-\frac{\e^{\Delta_a\tau}-1}{\Delta_a}\right)-vNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}\right]\,\d\tau}{1-\left(1-p_w\right)^N},
+\tau_2 = 
+\frac{\int_0^\infty\e^{-\left(r_1+r_2\right)} \d t}{1-\left(1-p_w\right)^N} = 
+\frac{\int_0^\infty\exp\left[-\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_w\tau}-1}{\Delta_w}-\frac{\e^{\Delta_a\tau}-1}{\Delta_a}\right)-vNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}\right] \d t}{1-\left(1-p_w\right)^N},
 \end{align}
-which we plot in \Cref{MeanTimeGrowthAneuploidyPlot} as a function of the initial population size.
-
-We wish to obtain a simpler formula for $\tau_2$ in an analogous way to \eqref{limitapprox}. For this, we make use of the following expansions:
-\begin{align*}
-\frac{\e^{\Delta_w\tau}-1}{\Delta_w}&=\frac{1+\Delta_w\tau+\frac{\Delta_w^2\tau^2}{2}+O(\tau^3)-1}{\Delta_w}=\tau+\frac{\Delta_w}{2}\tau^2+O(\tau^3),\\
-\frac{\e^{\Delta_a\tau}-1}{\Delta_a}&=\frac{1+\Delta_a\tau+\frac{\Delta_a^2\tau^2}{2}+O(\tau^3)-1}{\Delta_a}=\tau+\frac{\Delta_w}{2}\tau^2+O(\tau^3),
-\end{align*}
-which allow us to write:
-\begin{equation*}
-\frac{\e^{\Delta_w\tau}-1}{\Delta_w}-\frac{\e^{\Delta_a\tau}-1}{\Delta_a}\approx\frac{\left(\Delta_w-\Delta_a\right)\tau^2}{2}.
-\end{equation*}
-As a result, the integrand in \eqref{meantimet2} can be written as:
-\begin{align*}
-&\exp\left[-\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_w\tau}-1}{\Delta_w}-\frac{\e^{\Delta_a\tau}-1}{\Delta_a}\right)-vNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}\right]\approx\exp\left(-uvNp_m\tau^2-vNp_m\tau\right)\\
-&=\exp\left(\frac{vNp_m}{2}\right)\exp\left[-\frac{uvNp_m}{2}\left(\tau+\frac{1}{u}\right)\right].
-\end{align*}
-Consequently, the mean time $\tau_2$ is obtained to be:
-\begin{align}\label{limitapprox2}
-\tau_2\approx\left[1+\exp\left(-Np_w\right)\right]\exp\left(\frac{vNp_m}{2u}\right)\frac{\erfc\left(\sqrt{\frac{vNp_m}{2u}}\right)}{\sqrt{\frac{2uvNp_m}{\pi}}},
-\end{align}
-where $\erfc$ is the complementary error function.
-We plot the expansion \eqref{limitapprox2} in \Cref{MeanTimeGrowthAneuploidyPlot} and observe that it is a very good fit for large values of the initial wildtype population size.
-
-If we select only linear terms in the following expansions:
+which we plot in \Cref{MeanTimeGrowthAneuploidyPlot}A as a function of the initial population size, $N$.
+
+We wish to obtain a simpler formula for $\tau_2$, similar to \cref{limitapprox}.
+We thus have the following expansions,
+%\begin{align*}
+%\frac{\e^{\Delta_w\tau}-1}{\Delta_w}&=\frac{1+\Delta_w\tau+\frac{\Delta_w^2\tau^2}{2}+O(\tau^3)-1}{\Delta_w}=\tau+\frac{\Delta_w}{2}\tau^2+O(\tau^3) ,\\
+%\frac{\e^{\Delta_a\tau}-1}{\Delta_a}&=\frac{1+\Delta_a\tau+\frac{\Delta_a^2\tau^2}{2}+O(\tau^3)-1}{\Delta_a}=\tau+\frac{\Delta_w}{2}\tau^2+O(\tau^3) ,
+%\end{align*}
+%which allow us to write
+%\begin{equation*}
+%\frac{\e^{\Delta_w\tau}-1}{\Delta_w}-\frac{\e^{\Delta_a\tau}-1}{\Delta_a} \approx \frac{\left(\Delta_w-\Delta_a\right)\tau^2}{2}.
+%\end{equation*}
+%As a result, the integrand in \cref{meantimet2} can be written as:
+%\begin{align*}
+%&\exp\left[-\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_w\tau}-1}{\Delta_w}-\frac{\e^{\Delta_a\tau}-1}{\Delta_a}\right)-vNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}\right]\approx \\
+%& \exp\left(-uvNp_m\tau^2-vNp_m\tau\right) =
+%\exp\left(\frac{vNp_m}{2}\right)\exp\left[-\frac{uvNp_m}{2}\left(\tau+\frac{1}{u}\right)\right].
+%\end{align*}
+%Consequently, the mean conditional time $\tau_2$ is approximated by
+%\begin{align}\label{limitapprox2}
+%\tau_2 \approx
+%\frac{\left(1+\exp\left(-Np_w\right)\right) \exp\left(\frac{vNp_m}{2u}\right) \erfc\left(\sqrt{\frac{vNp_m}{2u}}\right)}{\sqrt{\frac{2uvNp_m}{\pi}}},
+%\end{align}
+%where $\erfc(z)$ is the complementary error function. 
+%
+%If we apply linear approximations,
 \begin{align*}
@@
 \end{align*}
-we obtain the first order approximation for $\tau_2$:
+which we use to derive a first-order approximation for $\tau_2$,
 \begin{align}\label{limitapprox3}
-\tau_2\approx\left(1+\e^{-Np_w}\right)\int_0^\infty\e^{-uNp_m\tau}\,\d\tau=\frac{\left(1+\e^{-Np_w}\right)}{uNp_m} ,
+\tau_2 \approx
+\left(1+\e^{-Np_w}\right)\int_0^\infty\e^{-uNp_m\tau}\,\d\tau=\frac{\left(1+\e^{-Np_w}\right)}{uNp_m} ,
 \end{align}
-which we plot in \Cref{MeanTimeGrowthAneuploidyPlot} and observe that it offer as a good a fit to \cref{meantimet2} as \cref{limitapprox2}.
+
+\Cref{MeanTimeGrowthAneuploidyPlot}A shows that the approximation \cref{limitapprox3} has a good fit with simulation results for large initial wildtype population size ($N > 2\cdot10^7$).
+An approximation that uses second-order terms $\frac{\Delta_w^2\tau^2}{2}$ and $\frac{\Delta_a^2\tau^2}{2}$ does not perform better (\Cref{MeanTimeGrowthAneuploidyPlot}A).
+
+% TODO YOAV HERE
+
 Additionally, we observe that for large initial wildtype populations sizes direct mutation drives evolutionary rescue while aneuploidy plays a role for intermediate sized tumors.
@@
 \caption{\textbf{Evolutionary rescue probability with tolerant or non-growing aneuploid cells.}
-Rescue probability grows with the aneuploid growth rate $\Delta_a$ (x-axis), and is much higher in an initially large tumor than in a small one (\textbf(A) $N=10^4$; \textbf{(B)} $N=10^8$). 
+Rescue probability grows with the aneuploid growth rate $\Delta_a$ (x-axis), and is much higher in an initially large tumor than in a small one (\textbf{A}: $N=10^4$; \textbf{B}: $N=10^8$). 
 Markers for simulation results with 95\% CI; 

### 2023-07-17 EvolutionaryRescue1.tex
finished revision

@@
 
-\subsection*{Effect of aneuploidy on evolutionary rescue}
-To determine the extent to which aneuploidy may affect evolutionary rescue, we define $H$ to be the ratio of the rescue probability with aneuploidy ($u>0$) and the rescue probability without aneuploidy ($u=0$),
+\subsection*{Effect of aneuploidy on the probability of evolutionary rescue}
+To determine the extent to which aneuploidy may affect evolutionary rescue, we define $H$ to be the ratio of the rescue probability with and without aneuploidy ($u>0$ and $u=0$, respectively),
 \begin{equation}\label{ratiorescueexact}
@@
 %\frac{\int_0^\infty\exp\left(-\int_0^\tau R(z)\d z\right)\d t}{1-\left(1-p_w\right)^N} = 
-\frac{\int_0^\infty\e^{-uNp_a\frac{\e^{\Delta_w\tau}-1}{\Delta_w}}\,\d\tau}{1-\left(1-p_w\right)^N} \approx
+\frac{\int_0^\infty\e^{-uNp_a\frac{\e^{\Delta_w t}-1}{\Delta_w}} \d t}{1-\left(1-p_w\right)^N} \approx
 \frac{\int_{0}^{\infty}{\exp\left(-u p_a N t\right) \d t}}{1-\e^{-Np_w}} \approx \\
@@
 \begin{align}
-r_1\left(t\right)&=vp_m\int_0^ta_{\tau}\,\d\tau=\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_wt}-1}{\Delta_w}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right),\\
-r_2\left(t\right)&=vp_m\int_0^tw_{\tau}\,\d\tau=vNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}.
+r_1\left(t\right)&=vp_m\int_0^ta_{z} \d z = \frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_wt}-1}{\Delta_w}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right),\\
+r_2\left(t\right)&=vp_m\int_0^tw_{z} \d z = vNp_m\frac{\e^{\Delta_w t}-1}{\Delta_w}.
 \end{align}
@@
 \tau_2 = 
-\frac{\int_0^\infty\e^{-\left(r_1+r_2\right)} \d t}{1-\left(1-p_w\right)^N} = 
-\frac{\int_0^\infty\exp\left[-\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_w\tau}-1}{\Delta_w}-\frac{\e^{\Delta_a\tau}-1}{\Delta_a}\right)-vNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}\right] \d t}{1-\left(1-p_w\right)^N},
+\frac{\int_0^\infty\e^{-\left(r_1(t)+r_2(t)\right)} \d t}{1-\left(1-p_w\right)^N} = 
+\frac{\int_0^\infty\exp\left[-\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_w t}-1}{\Delta_w}-\frac{\e^{\Delta_a t}-1}{\Delta_a}\right)-vNp_m\frac{\e^{\Delta_w t}-1}{\Delta_w}\right] \d t}{1-\left(1-p_w\right)^N},
 \end{align}
@@
 \begin{align*}
-\frac{\e^{\Delta_w\tau}-1}{\Delta_w}&=\frac{1+\Delta_w\tau+O(\tau^2)-1}{\Delta_w}=\tau+O(\tau^2),\\
-\frac{\e^{\Delta_a\tau}-1}{\Delta_a}&=\frac{1+\Delta_a\tau+O(\tau^2)-1}{\Delta_a}=\tau+O(\tau^2),
+\frac{\e^{\Delta_w t}-1}{\Delta_w}&=\frac{1+\Delta_w t+O(t^2)-1}{\Delta_w} = t+O(t^2),\\
+\frac{\e^{\Delta_a t}-1}{\Delta_a}&=\frac{1+\Delta_a t+O(t^2)-1}{\Delta_a} = t+O(t^2),
 \end{align*}
@@
 \tau_2 \approx
-\left(1+\e^{-Np_w}\right)\int_0^\infty\e^{-uNp_m\tau}\,\d\tau=\frac{\left(1+\e^{-Np_w}\right)}{uNp_m} ,
+\left(1+\e^{-Np_w}\right)\int_0^\infty\e^{-u N p_m t} \d t = 
+\frac{\left(1+\e^{-Np_w}\right)}{uNp_m} ,
 \end{align}
@@
 \Cref{MeanTimeGrowthAneuploidyPlot}A shows that the approximation \cref{limitapprox3} has a good fit with simulation results for large initial wildtype population size ($N > 2\cdot10^7$).
-An approximation that uses second-order terms $\frac{\Delta_w^2\tau^2}{2}$ and $\frac{\Delta_a^2\tau^2}{2}$ does not perform better (\Cref{MeanTimeGrowthAneuploidyPlot}A).
-
-% TODO YOAV HERE
-
-Additionally, we observe that for large initial wildtype populations sizes direct mutation drives evolutionary rescue while aneuploidy plays a role for intermediate sized tumors.
-This is consistent with the information obtained from \cref{rescue_ratio} where aneuploidy improves the probability of evolutionary rescue only  for small and intermediate values of $N$.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Contribution of aneuploidy to mean evolutionary rescue time}
-\begin{align}
-I&=\frac{\tau_2}{\tau_1}=\frac{\int_0^\infty\exp\left[-\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_wt}-1}{\Delta_w}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right)-vNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}\right]\,\d\tau}{\int_0^\infty\e^{-uNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}}\,\d\tau}\times\frac{1-\left(1-\left.p_w\right|_{u=0}\right)^N}{1-\left(1-\left.p_w\right|_{u>0}\right)^N}\\\label{eqMeanTimeRatioInitialPopulationSize}
-&=\frac{\int_0^\infty\exp\left[-\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_wt}-1}{\Delta_w}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right)-vNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}\right]\,\d\tau}{\int_0^\infty\e^{-vNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}}\,\d\tau}\frac{1}{H},
+An approximation that uses the second-order terms, $\frac{\Delta_w^2 t^2}{2}$ and $\frac{\Delta_a^2 t^2}{2}$, does not perform better (\Cref{MeanTimeGrowthAneuploidyPlot}A).
+\\
+
+Importantly, in \Cref{MeanTimeGrowthAneuploidyPlot} we observe that in initially large tumors, direct mutation drives evolutionary rescue while aneuploidy plays a role for intermediate sized tumors.
+% TODO where do you see this..?
+This is consistent with \Cref{rescue_ratio}B,D, where aneuploidy only increases the probability of evolutionary rescue in small and intermediate tumors ($N<10^7$).
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection*{Effect of aneuploidy on mean evolutionary rescue time}
+
+Next, we wish to find the ratio of the mean rescue time with and without aneuploidy ($u>0$ and $u=0$, respectively),
+\begin{align}\label{eqMeanTimeRatioInitialPopulationSize}
+\frac{\tau_2}{\tau_1}  
+%\frac{\int_0^\infty\exp\left[-\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_w t}-1}{\Delta_w}-\frac{\e^{\Delta_a t}-1}{\Delta_a}\right)-vNp_m\frac{\e^{\Delta_w t}-1}{\Delta_w}\right] \d t}{\int_0^\infty\e^{-uNp_m\frac{\e^{\Delta_w t}-1}{\Delta_w}} \d t} \cdot \frac{1-\left(1-p_w(u=0)\right)^N}{1-\left(1-p_w(u>0)\right)^N}\\
+= H^{-1} \cdot \frac{\int_0^\infty\exp\left[-\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_wt}-1}{\Delta_w}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right)-vNp_m\frac{\e^{\Delta_w t}-1}{\Delta_w}\right] \d t}{\int_0^\infty\e^{-vNp_m\frac{\e^{\Delta_w t}-1}{\Delta_w}} \d t},
 \end{align}
-where $H$, is the ratio of the probability of evolutionary rescue with and without aneuploidy, defined in \cref{ratiorescueexact}.
-We plot \cref{eqMeanTimeRatioInitialPopulationSize} in \Cref{MeanTimeRatioInitialPopulationSize} as a function of the initial wildtype population for varying values of the Malthusian fitness of aneuploid cells $\Delta_a$.
+where $H$ is the ratio of the rescue probability with and without aneuploidy, see \cref{ratiorescueexact}.
+\Cref{MeanTimeRatioInitialPopulationSize} shows how this ratio, $\tau_2/\tau_1$, changes as a function of the initial tumor size, $N$, for different aneuploid growth rates, $\Delta_a$.
 
@@
 
-Evolutionary rescue  is the process where the population acquires a trait that increases fitness in the new environment such that extinction is averted.
-Here, we have modeled a tumor--a population of cancer cells--exposed to drug therapy that causes the cell population to decline towards extinction.
-The cancer cell population can escape extinction either by a mutation that confers resistance, or by first generating aneuploid cells in which the effect of the drug is diminished, and then producing a mutation that confers full resistance (\Cref{figureAneuploidy}).
-
-Using multitype branching processes, we derived the probability of evolutionary rescue of the population of cancer cells under various scenarios for the effect of aneuploidy, including both tolerance and partial resistance to the drug.
-We obtained both exact and approximate expressions for the probability of evolutionary rescue.
-As expected, our analytic results in \cref{rescue_prob} show that the probability of evolutionary rescue increases with the initial tumor size $N$, the wildtype growth rate $\Delta_w=\lambda_w-\mu_w$, and the mutation $v$ and aneuploidy $u$ rates.
-
-When aneuploid cells are partially resistant to the drug ($\Delta_w\ll0\ll\Delta_a\ll\Delta_m$), evolutionary rescue can be approximated by a one-step evolutionary rescue process where aneuploidy itself rescues the population (\Cref{rescue_prob_wt_growth}).
+We have modeled a tumor--a population of cancer cells--exposed to drug therapy that causes the population to decline in size towards potential extinction.
+The cancer cell population can be "evolutionary rescued", or escape extinction, by two paths.
+In the direct path, a sensitive cell acquires a mutation that confers resistance that allows it to rapidly grow.
+In the indirect path, a sensitive cell first becomes aneuploid, which diminishes the effect of the drug, and then an aneuploid cell acquires a mutation that confers resistance (\Cref{figureAneuploidy}).
+
+Using multitype branching processes, we derived the probability of evolutionary rescue of the population of cancer cells under different scenarios for the effect of aneuploidy, ranging from tolerance to partial resistance.
+We obtained exact and approximate expressions for the probability of evolutionary rescue (\cref{rescue_prob}).
+Our results show that the probability of evolutionary rescue increases with the initial tumor size $N$, the sensitive growth rate $\Delta_w$, the mutation rate $v$, and the aneuploidy rate $u$.
+
+When aneuploid cells are partially resistant to the drug ($\Delta_w\ll0\ll\Delta_a\ll\Delta_m$), evolutionary rescue can be approximated by a one-step process in which aneuploidy itself rescues the population (\Cref{rescue_prob_wt_growth}).
 When aneuploidy only provides tolerance to the drug ($\Delta_w\ll\Delta_a\ll0\ll\Delta_m$), it cannot rescue the population.
-Instead, aneuploidy acts as a \emph{stepping stone} through which the resistant mutant can appear in a more expedient fashion, given that the aneuploid cell population declines slower then the wildtype cell population.
+Instead, it acts as a \emph{stepping stone} through which the resistant mutant can appear more rapidly, given that the aneuploid cell population declines slower then the sensitive cell population.
 In this case, aneuploidy provides two benefits.
-First, it delays the extinction of the population--providing more time for appearance of the resistance mutations.
-Second, it increases the population size relative to a wildtype population--providing more cells for generating mutations, i.e., it increases the mutation supply.
-
-We find that aneuploidy can have a significant effect on evolutionary rescue (\Cref{rescue_ratio}).
-For example, when aneuploidy cells are "barely-resistant" (they grow at a very low rate, $\Delta_a=10^{-3}$) the probability of evolutionary rescue is 1000-fold higher with aneuploidy than without it (for parameters previously described in cancer \Cref{table1}).
-Interestingly, aneuploidy is unlikely to contribute to evolutionary rescue in primary tumors, as the number of cells in such tumors ($N>10^7$) % TODO is this correct? 
-is large enough for the appearance of resistant mutation directly before the extinction of wildtype cells (\Cref{rescue_ratio}).
-However, aneuploidy may play a crucial role in evolutionary rescue of secondary tumors, whose size may be below the detection threshold of $\sim10^7$  \citep{bozic2013evolutionary}.
-Given the fact that the mean time for such secondary tumors to overcome chemotherapy can be of the order of 100 days (\Cref{MeanTimeGrowthAneuploidyPlot}), this can explain the reappearance of cancer even after initial remission.
-
-We hypothesized that presence of "standing variation"-- % TODO ref
-a subpopulation of aneuploid cancer cells--at the onset of chemotherapy may facilitate evolutionary rescue by reducing the waiting time for the appearance of aneuploid cells.
-Indeed, we observe that even when a small fraction of the initial tumor is aneuploid, evolutionary rescue is more likely to occur through this existing standing variation, rather then through "de novo" aneuploid cells (\Cref{FractionPlot}).
+First, it delays the extinction of the population--providing more time for appearance of the resistance mutation.
+Second, it increases the population size relative to a sensitive population--providing more cells in which mutations can occur, i.e., it increases the mutation supply, $Nv$.
+
+We find that aneuploidy can have a significant effect on evolutionary rescue (\Cref{rescue_ratio,MeanTimeRatioInitialPopulationSize}).
+For example, when aneuploidy cells are "barely-resistant" (they grow at a very low rate, $\Delta_a=10^{-3}$) the probability of evolutionary rescue is 1,000-fold higher with aneuploidy than without it (for parameters previously described in cancer, see \Cref{table1}).
+Interestingly, aneuploidy is unlikely to contribute to evolutionary rescue in primary tumors in which the number of cells is large enough ($N>10^7$) for the appearance of resistant mutation directly in sensitive cells before these cells become extinct (\Cref{rescue_ratio}).
+However, aneuploidy can have a crucial role in evolutionary rescue of secondary tumors, in which the number of sensitive cells may be below the detection threshold of $\sim10^7$  \citep{bozic2013evolutionary}.
+Given the fact that the mean time for such secondary tumors to overcome chemotherapy can be of the order of 100 days (\Cref{MeanTimeGrowthAneuploidyPlot}), % TODO do we see tau=100 here...?
+this can explain the reappearance of cancer even after initial remission.
+Indeed, we find that the tumor size can decrease by orders of magnitude before it is rescued (\Cref{MinTumorSize}).
+
+We hypothesized that presence of \emph{standing variation} % TODO ref
+--the existence of a subpopulation of aneuploid cancer cells before therapy begins--can facilitate evolutionary rescue by reducing the waiting time for the appearance of aneuploid cells.
+Indeed, we observe that even when a small fraction of the initial tumor is aneuploid, evolutionary rescue is more likely to occur through this existing standing variation, rather then through \emph{de novo} aneuploid cells (\Cref{FractionPlot}).
 
@@
 our approximation for the probability of evolutionary rescue agrees with results of stochastic simulations with density-dependent growth  \Cref{rescue_prob_N}).
-Future work may focus on scenarios with small carrying capacity by analysing density-dependent branching processes. % TODO ref?
+Future work may focus on scenarios with small carrying capacity by analyzing density-dependent branching processes. % TODO ref?
 
@@
 Afterwards, these tumors may be exposed to anti-cancer drugs that induces aneuploidy % TODO ref
-or to saline solution for control. Cell density can then be measured and compared to the predictions of our model. % TODO not clear what is to be compared... what does our model predict...?
-
-% TODO discussion of mean time
+or to saline solution for control. 
+Cell density can then be measured and compared to the predictions of our model. % TODO not clear what is to be compared... what does our model predict...?
+
+% TODO discussion of mean time?
 
@@
 \section*{Acknowledgements}
-%We thank XXX for discussions and comments. % TODO who?
+%We thank Hildegard Uecker for discussions and comments. % TODO who?
 This work was supported in part by
@@
 \begin{figure}
-% TODO change panel labels to ABCD
 \begin{subfigure}{0.5\textwidth}
@@
 \textbf{Evolutionary rescue probability with partially resistant or tolerant aneuploid cells.}
-Rescue probability is very high when aneuploidy provides partial resistance ($\lambda_a=0.01$), in an initially small tumor (\textbf{Aleft}, $N=10^4$) and even more so in an initially large tumor (\textbf{Aright}, $N=10^8$).
-When aneuploidy provides tolerance (\textbf{Bleft}, $N=10^4$; \textbf{Bright}, $N=10^8$), the rescue probability is much lower. 
+Rescue probability is very high when aneuploidy provides partial resistance ($\lambda_a=0.01$), in an initially small tumor (\textbf{A}, $N=10^4$) and even more so in an initially large tumor (\textbf{B}, $N=10^8$).
+When aneuploidy provides tolerance (\textbf{C}, $N=10^4$; \textbf{D}, $N=10^8$), the rescue probability is much lower. 
 In both scenarios, rescue probability increase with both the wildtype growth rate (x-axis) and the aneuploidy growth rate (colors).
@@
 % TODO more simulations for 0 > Delta_a > Delta^*_a
+% TODO simulations in A?
 \begin{subfigure}{0.5\textwidth}
@@
 \begin{figure}
-% A partial resistance
-% B tolerance
-% left function of wildtype growth rate 
-% right function of population size
 \begin{subfigure}{0.5\textwidth}
@@
 The ratio of rescue probability with and without aneuploid ($H$, \cref{ratiorescue}) increases with the aneuploid growth rate (colors) and decreases with the wildtype growth rates and initial tumor size (x-axes), except for large tumors where where the ratio converges to unity.
-\textbf{(A-left,A-right)} Aneuploidy provides partial resistance.
-\textbf{(B-left, B-right)} Aneuploidy provides tolerance.  
+\textbf{(A, B)} Aneuploidy provides partial resistance.
+\textbf{(C, D)} Aneuploidy provides tolerance.  
 Solid and dashed lines apply $p_{rescue}$ from the exact formula of  (\cref{survival_prob} in \cref{rescue_prob}); dashed lines apply $p_{rescue}$ from the approximate formula (\cref{rescue_prob_approx}), with good agreement.
@@
 % TODO start y at 1
-% TODO fix panel labels
 }
@@
 \begin{figure}
-% tolerance... maybe third is slightly decreasing
  \vspace*{1\baselineskip}
@@
 \caption{\textbf{Ratio of evolutionary rescue time with and without aneuploidy.}
-The ratio of the mean time to appearance of a resistance mutation that leads to evolutionary rescue with ($u>0$) and without ($u=0$) aneuploidy for variable initial tumor sizes (\cref{eqMeanTimeRatioInitialPopulationSize}) when aneuploidy provides tolerance to the drug ($\Delta_a\ll0$).
+The ratio of the mean time to appearance of a resistance mutation that leads to evolutionary rescue with ($u>0$) and without ($u=0$) aneuploidy for variable initial tumor sizes (\cref{eqMeanTimeRatioInitialPopulationSize}) when aneuploidy provides tolerance to the drug ($\Delta_a < 0$).
 When the initial tumor size is not large ($<10^8$), aneuploidy can decrease the rescue time by 10-100-fold.
-\emph{I THINK THERE IS A MISTAKE IN THE BLUE LINE} % TODO
 % TODO parameter values 
@@
 }
-\label{MeanTimeRatioInitialPopulationSize}
+\label{MinTumorSize}
 \end{figure}

### 2023-07-17 EvolutionaryRescue1.tex
revise with RS and DBW

@@
 \newcommand{\taua}{\tau_{\rm A}}
+
+% Daniel commands
+
+\newcommand{\daniel}[1]{\textcolor{blue}{#1}}
+\newcommand{\presc}{p_\text{rescue}}
+\newcommand{\psgv}{p_\text{SGV}}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Survival probability}
-
-To analyze evolutionary rescue in this model, we use the framework of \emph{multitype branching processes} \citep{rybnikov2021fitness,harris1963theory}. 
-This allows us to find explicit expressions for the \emph{survival probability}: the probability that a lineage descended from a single cell does not become extinct.
-
-Let $p_w$, $p_a$, and $p_m$ be the survival probabilities of a population consisting initially of single wildtype cell, aneuploid cell, or mutant cell, respectively.
-The complements $1-p_w$, $1-p_a$, and $1-p_m$ are the extinction probabilities, which satisfy each its respective equation,
-\begin{equation} \label{extinction_prob}
-\begin{aligned}
-1-p_w = &\frac{\mu_w}{\lambda_w+\mu_w+u+v} + 
-		  \frac{u}{\lambda_w+\mu_w+u+v}\left(1-p_a\right) + \\
-		  & \frac{\lambda_w}{\lambda_w+\mu_w+u+v}\left(1-p_w\right)^2 +
-		  \frac{v}{\lambda_w+\mu_w+u+v}\left(1-p_m\right) ,\\
-1-p_a = &\frac{\mu_a}{\lambda_a+\mu_a+v}+\frac{v}{\lambda_a+\mu_a+v}\left(1-p_m\right)+\frac{\lambda_a}{\lambda_a+\mu_a+v}\left(1-p_a\right)^2 ,\\
-1-p_m = &\frac{\mu_m}{\lambda_m+\mu_m}+\frac{\lambda_m}{\lambda_m+\mu_m}\left(1-p_m\right)^2 .	 
-\end{aligned}
-\end{equation}
-
-The survival probabilities are given by the smallest solution for each quadratic equation \citep{uecker2015adaptive}. Therefore we have
-\begin{equation}\label{survival_prob}
-\begin{aligned}
-p_w &= \frac{\lambda_w-\mu_w-u-v+\sqrt{\left(\lambda_w-\mu_w-u-v\right)^2+4\lambda_w\left(up_a+vp_m\right)}}{2\lambda_w} ,\\
-p_a &= \frac{\lambda_a-\mu_a-v+\sqrt{\left(\lambda_a-\mu_a-v\right)^2+4\lambda_avp_m}}{2\lambda_a}, \\
-p_m &= \frac{\lambda_m-\mu_m}{\lambda_m} .
-\end{aligned} 
-\end{equation}
-Note that the equation for $p_w$ depends on both $p_a$ and $p_m$, and the equation for $p_a$ depends on $p_m$.
-To proceed, we can plug the solution for $p_m$ and $p_a$ into the solution for $p_w$. We perform this for three different scenarios.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{Scenario 1: Aneuploid cells are partially resistant}
-
-We first assume that aneuploidy provides partial resistance to drug therapy, $\lambda_a>\mu_a$, and that this resistance is significant, $\left(\lambda_a-\mu_a-v\right)^2 > 4\lambda_a v p_m$.
-We thus rewrite \cref{survival_prob} as
-\begin{align*}
-p_w&=\frac{\lambda_w-\mu_w-u-v}{2\lambda_w}\left(1-\sqrt{1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}}\right) ,
-\text{and} \\
-p_a&=\frac{\lambda_a-\mu_a-v}{2\lambda_a}\left(1+\sqrt{1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}}\right) . 
-\end{align*}
-Using the quadratic Taylor expansion $\sqrt{1+x}=1+x/2+O(x^2)$ and assuming $u,v \ll 1$,
-we obtain the following approximation for the survival probability of a population initially consisting of a single wildtype cell,
-\begin{align}\label{survprobwapprox1}
-p_w 
-&\approx -\frac{vp_m+up_a}{\lambda_w-\mu_w-u-v}\\
-\nonumber
-%&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{v\left(\lambda_a-\mu_a-u\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-u\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\ \label{survprobw2}
-&\approx-\frac{1}{\lambda_w-\mu_w}\left[\frac{u\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\
-\end{align}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\paragraph{Second-order approximation.} % TODO consider moving to Appendix
-To improve our approximation, we can consider the second term of the Taylor series expansion,
-\begin{align*}
-\left(1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}\right)^{\frac{1}{2}}=1+\frac{2\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}-\frac{\left(\lambda_avp_m\right)^2}{4\left(\lambda_a-\mu_a-v\right)^4}+\cdots,
-\end{align*}
-which gives us the following approximation,
-\begin{align}
-p_a \approx \frac{\lambda_a-\mu_a-v}{\lambda_a}+\frac{vp_m}{\lambda_a-\mu_a-v}-\frac{\lambda_a\left(vp_m\right)^2}{8\left(\lambda_a-\mu_a-v\right)^3} .
-\end{align}
-We therefore have
-\begin{align}\nonumber
-p_w&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{u\left(\lambda_a-\mu_a-v\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-v\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}-\frac{uv^2\lambda_a\left(\lambda_m-\mu_m\right)^2}{8\lambda_m^2\left(\lambda_a-\mu_a-v\right)^3}\right]\\ \label{survprobw3}
-&\approx-\frac{1}{\lambda_w-\mu_w}\left[\frac{u\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}-\frac{uv^2\lambda_a\left(\lambda_m-\mu_m\right)^2}{8\lambda_m^2\left(\lambda_a-\mu_a\right)^3}\right],
-\end{align}
-and using $\Delta_k=\lambda_k-\mu_k$, we can write the above equation as
-\begin{equation}\label{survprobwapproxcorrected}
-p_w \approx -\frac{1}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}-\frac{uv^2\lambda_a\Delta_m^2}{8\lambda_m^2\Delta_a^3}\right).
-\end{equation}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{Scenario 2: Aneuploid cells are tolerant.}
-
-We now assume that aneuploidy provides tolerance to drug therapy, that is, the number of aneuploid cells significantly declines over time, but at a lower rate than the number of wildtype cells, $\lambda_w - \mu_w < \lambda_a - \mu_a < 0$.
-We also assume that the decline are significant, $\left(\lambda_a-\mu_a-v\right)^2 > 4\lambda_a v p_m$.
-We rewrite \cref{survival_prob} as
-\begin{equation}
-\begin{aligned}
-p_w&=\frac{\lambda_w-\mu_w-u-v}{2\lambda_w}\left(1-\sqrt{1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}}\right) ,
-\text{and} \\
-p_a&=\frac{\lambda_a-\mu_a-v}{2\lambda_a}\left(1-\sqrt{1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}}\right) .
-\end{aligned}
-\end{equation}
-Since $u,v\ll1$, the term in the root can be approximated using a 1st-order Taylor expansion. So, substituting the expressions for $p_a$ and $p_m$, we have
-\begin{equation}\label{survprobwinitial}
-\begin{aligned}
-p_w&\approx-\frac{vp_m+up_a}{\lambda_w-\mu_w-u-v}\\
-&\approx\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-v\right)}-\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\ %\label{survprobw2}
-&\approx\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_w-\mu_w\right)}\left[\frac{u}{\left(\lambda_a-\mu_a\right)}-1\right]\\
-%&=\frac{v\Delta_m}{\lambda_m\Delta_w}\left(\frac{u}{\Delta_a}-1\right) .
-\end{aligned}
-\end{equation}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{Scenario 3: Aneuploid cells are non-growing} % TODO consider this name
-
-We now assume that the growth rate of aneuploid cells is close to zero (either positive or negative), such that  $\left(\lambda_a-\mu_a-v\right)^2 < 4\lambda_avp_m$.
-We rewrite \cref{survival_prob} as
-\begin{equation}
-p_a=\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}\left(1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\lambda_avp_m}\right)^{\frac12}}{2\lambda_a} .
-\end{equation}
-Using a following Taylor series expansion
-\begin{equation*}
-\left(1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\lambda_avp_m}\right)^{\frac{1}{2}}=1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{8\lambda_avp_m}+\cdots,
-\end{equation*}
-we obtain the approximation
-\begin{equation}
-\begin{aligned}
-p_a&\approx\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}\left[1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{8\lambda_avp_m}\right]}{2\lambda_a}\\
-&=\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\sqrt{\lambda_avp_m}}}{2\lambda_a}\\
-&=\frac{\left(\lambda_a-\mu_a-v+2\sqrt{\lambda_avp_m}\right)^2+4\lambda_avp_m}{8\lambda_a\sqrt{\lambda_avp_m}}\\
-&=\frac{4\lambda_avp_m+4\lambda_avp_m\left(1+\frac{\lambda_a-\mu_a-v}{2\sqrt{\lambda_avp_m}}\right)^2}{8\lambda_a\sqrt{\lambda_avp_m}}\\
-&=\frac{1}{2\lambda_a}\left(\lambda_a-\mu_a-v+2\sqrt{\lambda_avp_m}\right).
-\end{aligned}
-\end{equation}
-Plugging this in \cref{survprobwinitial}, the survival probability of a population starting from one wildtype individual is
-\begin{equation}\label{scenario3}
-\begin{aligned}
-p_w&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[v\frac{\lambda_m-\mu_m}{\lambda_m}+\frac{u}{2\lambda_a}\left(\lambda_a-\mu_a-v+2\sqrt{\lambda_avp_m}\right)\right]\\
-&=-\frac{1}{\lambda_w-\mu_w-u-v}\left[v\frac{\lambda_m-\mu_m}{\lambda_m}+\frac{u}{2\lambda_a}\left(\lambda_a-\mu_a-v\right)+u\sqrt{\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_a\lambda_m}}\right].
-%&=-\frac{1}{\Delta_w-u-v}\left[v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right],
-\end{aligned}
-\end{equation}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection*{Evolutionary rescue probability}
@@
 \end{subequations}
-where the approximation $(1-p_w)\approx e^{-p_w}$ assumes that $p_w$ (but not $N p_w$) is small.
+where $p_w$ is the probability that the lineage of a single wild-type cell avoids extinction, and the approximation $(1-p_w)\approx e^{-p_w}$ assumes that $p_w$ (but not necessarily $N p_w$) is small.
 
-Applying the approximations for the survival probability $p_w$ from \cref{survprobwapprox1,survprobwinitial,scenario3} in \cref{rescue_prob2} and substituting $\Delta_k=\lambda_k-\mu_k$, we find that the rescue probability can be approximated by
+The probability for evolutionary rescue $\presc$  (\cref{rescue_prob2}) is $\presc \approx N p_w$ when $N<1/p_w$ and $\presc \approx 1$ when $N > 1/p_w$.
+Therefore, $N^*=1/p_w$ is the threshold tumor size for rescue above which rescue is very likely (\Cref{rescue_prob_N}).
+% TODO add 1/p_w to figure as vertical line
+Therefore, we can write
+\begin{equation}
+\presc = 1-e^{-N/N^*} .
+\end{equation}
+In the Appendix below, we use the theory of multi-type branching processes to find approximate expressions \cref{survprobwapprox1,survprobwinitial,scenario3} for $p_w$ in different regimes. 
+Substituting these  into $N^*=1/p_w$, we find approximations for the threshold tumor size, $N^*$. 
+\\
+
+In the following, an important quantity is $T^* = (4 v \lambda_a \Delta_m/\lambda_m)^{-1/2}$, which is the critical time an aneuploid lineage needs to survive to produce a resistant mutant that avoids random extinction.
+First, if aneuploidy is very rare ($u T^*< 1$), or if aneuploidy is rare ($u < -\Delta_a$) and very sensitive to the drug ($\Delta_a T^* < -1$), then rescue will likely occur by a direct resistance mutation in a sensitive cell, such that
+\begin{equation}
+N_m^* \approx \frac{\abs{\Delta_w}}{v} \cdot \frac{\lambda_m}{\Delta_m} .
+\end{equation}
+Otherwise, aneuploidy is frequent enough ($u > \max{(-\Delta_a, 1/T^*)}$) to affect the evolution of the tumor. 
+The threshold tumor size, $N^*$, can then be approximated by one of the following cases, depending on $\Delta_a T^*$, the change in the log of the aneuploid population size during the critical time,
 \begin{equation}\label{rescue_prob_approx}
 \begin{aligned}
-&p_{rescue} \approx \\
-  &\begin{cases}
-  1-\exp\left[\frac{N}{\Delta_w-u-v}\left(v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right)\right] ,&
-  4\lambda_avp_m>\left(\Delta_a-v\right)^2 ,\\
-   1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right] ,&
-   \Delta_a<0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2 ,\\
-   1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right] ,&
-   \Delta_a>0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2 .
+N_a^* \approx 
+% TODO replace XXX with approximations
+  \frac{\abs{\Delta_w}}{u} \cdot \begin{cases}
+    \frac{\abs{\Delta_a}}{v} \cdot \frac{\lambda_m}{\Delta_m} ,&
+  \Delta_a T^* \ll -1 ,\\ 
+  %\left(\frac{\lambda_a}{v} \cdot \frac{\lambda_m}{\Delta_m}\right)^{1/2} ,&
+  2\lambda_a T^* ,&
+  -1 \ll \Delta_a T^* \ll 1  ,\\ 
+  \frac{\lambda_a}{\Delta_a} ,&
+   \Delta_a T^* \gg 1 .
   \end{cases}
@@
 \end{equation}
+%The second case has a finer approximation by $N^*_a \approx \frac{\abs{\Delta_w}}{u} \cdot \frac{2\lambda_a}{\Delta_a + 1/T^*}$.
+Comparing these approximations to results of stochastic evolutionary simulations, we find that the approximations perform very well (\Cref{rescue_prob_wt_growth,rescue_prob_an_growth,rescue_prob_N,rescue_prob_N}).
 
-% TODO do we need this?
-%Note that if instead \cref{survprobwapprox1} we apply \cref{survprobwapproxcorrected}, we have
-%\begin{align}\label{survprobresistantaneuploidy}
-%p_{rescue}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}-\frac{uv^2\lambda_a\Delta_m^2}{8\lambda_m^2\Delta_a^3}\right)\right].
-%\end{align}
-%However, both approximations seem to work very well (\Cref{compare_1st_2nd_order_approx}).
-
-We validate these approximations by comparing them to results of stochastic evolutionary simulations.
-We find that the approximations work very well (\Cref{rescue_prob_wt_growth,rescue_prob_an_growth,rescue_prob_N,rescue_prob_N}).
+% TODO move this to appendix
+%Substituting these  into \cref{rescue_prob2}, we find that the rescue probability can be approximated by
+%\begin{equation}\label{rescue_prob_approx}
+%\begin{aligned}
+%&\presc \approx \\
+%  &\begin{cases}
+%  1-\exp\left[\frac{N}{\Delta_w-u-v}\left(v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right)\right] ,&
+%  4\lambda_avp_m>\left(\Delta_a-v\right)^2 ,\\
+%   1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right] ,&
+%   \Delta_a<0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2 ,\\
+%   1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right] ,&
+%   \Delta_a>0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2 .
+%  \end{cases}
+%\end{aligned}
+%\end{equation}
 
@@
 The approximation agrees with the simulation and exact solution when the initial tumor size is large (panel B).
-When the tumor size is small (panel A), we switch between the approximation for tolerant and for non-growing aneuploid cells; the switch occurs at $\Delta_a^*=2vp_m+v+2\sqrt{vp_m\left(vp_m+\mu_a+v\right)}$.
+When the tumor size is small (panel A), we switch between the approximation for tolerant and for non-growing aneuploid cells; the switch occurs at $1/T^*$. % TODO ref to eq
 Parameters: $\lambda_w=\lambda_a=\lambda_m=0.14$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
@@
 
-%\appendix
-%\begin{appendices}
-%\renewcommand{\theequation}{\thesection\arabic{equation}}
-%\counterwithin*{equation}{section}
+\appendix
+\begin{appendices}
+\renewcommand{\theequation}{\thesection\arabic{equation}}
+\counterwithin*{equation}{section}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\section*{Gillespie algorithm}
-%\begin{subequations}
-%\begin{flalign}
-%(+1,0,0)&:\quad \lambda_ww_t,\\
-%(-1,0,0)&:\quad \mu_ww_t,\\
-%(-1,+1,0)&:\quad uw_t,\\
-%(-1,0,+1)&:\quad vw_t,\\
-%(0,+1,0)&:\quad \lambda_aa_t,\\
-%(0,-1,0)&:\quad \mu_aa_t,\\
-%(0,-1,+1)&:\quad va_t,\\
-%(0,0,+1)&:\quad \lambda_am_t,\\
-%(0,0,-1)&:\quad \mu_am_t.
-%\end{flalign}
-%\end{subequations}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section*{Survival probability of a single lineage}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\section*{Mean time}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%We define the following events
-%\begin{align*}
-%\text{EvR}&=\text{``evolutionary rescue occurs''},\\
-%\text{EvRA}&=\text{``evolutionary rescue occurs through aneuploidy''},\\
-%\text{EvRM}&=\text{``evolutionary rescue occurs through mutation''}.
-%\end{align*}
-%As a result, the mean time to appearence of the first mutant cell which rescues the population given evolutionary rescue can be expanded as:
-%\begin{align*}
-%\mathbb{E}\left[\tau|\text{EvR}\right]&=\mathbb{E}\left[\tau|\text{EvR},\text{EvRA}\cap\text{EvRM}^{C}\right]p\left(\text{EvRA}\cap\text{EvRM}^{C}|\text{EvR}\right)\\
-%&+\mathbb{E}\left[\tau|\text{EvR},\text{EvRM}\cap\text{EvRA}^{C}\right]p\left(\text{EvRM}\cap\text{EvRA}^{C}|\text{EvR}\right)\\
-%&+\mathbb{E}\left[\tau|\text{EvR},\text{EvRM}\cap\text{EvRA}\right]p\left(\text{EvRM}\cap\text{EvRA}|\text{EvR}\right).
-%\end{align*}
-%The joint probabilities in the previous lines can be written as:
-%\begin{align*}
-%p\left(\text{EvRA}\cap\text{EvRM}^{C}|\text{EvR}\right)&=\frac{p\left(\text{EvRA}\cap\text{EvRM}^{C}\right)}{p\left(\text{EvR}\right)}
-%\approx\frac{p\left(\text{EvRA}\right)\left[1-p\left(\text{EvRM}\right)\right]}{p\left(\text{EvR}\right)},\\
-%p\left(\text{EvRM}\cap\text{EvRA}^{C}|\text{EvR}\right)&=\frac{p\left(\text{EvRM}\cap\text{EvRA}^{C}\right)}{p\left(\text{EvR}\right)}
-%\approx\frac{p\left(\text{EvRM}\right)\left[1-p\left(\text{EvRA}\right)\right]}{p\left(\text{EvR}\right)},\\
-%p\left(\text{EvRM}\cap\text{EvRA}|\text{EvR}\right)&=\frac{p\left(\text{EvRM}\cap\text{EvRA}\right)}{p\left(\text{EvR}\right)}\approx\frac{p\left(\text{EvRM}\right)p\left(\text{EvRA}\right)}{p\left(\text{EvR}\right)}\ll1.
-%\end{align*}
-%As a result, we have:
-%\begin{align*}
-%\mathbb{E}\left[\tau|\text{EvR}\right]&\approx\mathbb{E}\left[\tau|\text{EvR},\text{EvRA}\cap\text{EvRM}^{C}\right]p\left(\text{EvRA}\cap\text{EvRM}^{C}|\text{EvR}\right)\\
-%&+\mathbb{E}\left[\tau|\text{EvR},\text{EvRM}\cap\text{EvRA}^{C}\right]p\left(\text{EvRM}\cap\text{EvRA}^{C}|\text{EvR}\right).
-%\end{align*}
-%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\section*{Diffusion approximation}\label{AppendixDiffApprox}
-%
-%An alternative method to obtain the probability of evolutionary rescue is to utilize a Feller diffusion approximation which is governed by two parameters: the growth rate $r$ and the reproductive variance $\sigma$.
-The two parameters are obtained from the underlying demographic process as the infinitesimal relative change in mean and variance of $n_t$ over and an infinitesimally small time interval $\Delta t$:
-%\begin{align*}
-%r&=\lim_{\Delta t\rightarrow0}\frac{\mathbb{E}\left(\Delta n_t|n_t\right)}{\Delta_t n_t},\\
-%\sigma&=\lim_{\Delta t\rightarrow0}\frac{\mathbb{V}ar\left(\Delta n_t|n_t\right)}{\Delta_t n_t}.
-%\end{align*}
-%The rate at which mutants are generated directly from the wildtype is: 
-%\begin{align}
-%\theta_1=v\bar{\pi}_f\frac{N}{|r_w|},
-%\end{align}
-%where
-%\begin{align}
-%\bar{\pi}_f=\int_{0}^\infty\int_{0}^\infty \left(1-\e^{-\frac{2r}{\sigma}}\right) f_r\left(r,\sigma\right)\d r\d\sigma.
-%\end{align}
-%Letting $f_r\left(r,\sigma\right)=\delta\left(r-r_m\right)\delta\left(\sigma-\sigma_m\right)$ then:
-%\begin{align}
-%\bar{\pi}_f=1-\e^{-\frac{2r_m}{\sigma_m}},
-%\end{align}
-%and, as a result, we have:
-%\begin{align}
-%\theta_1=v\left(1-\e^{-\frac{2r_m}{\sigma_m}}\right)\frac{N}{|\Delta_w|}.
-%\end{align}
-%The rate at which mutants are generated indirectly from the wildtype through aneuploidy is: 
-%\begin{align*}
-%\theta_2&=\frac{uN}{|r_w|}\int_{0}^\infty\int_{0}^\infty \left(1-\pi_f\left(r,q\right)\right) f_r\left(r,\sigma\right) p_1^*\left(r,\sigma\right)\d r\d\sigma\\
-%&=\frac{uN}{|\Delta_w|}p_1^*\left(r_a,\sigma_a\right)\\
-%&=\frac{uN}{|\Delta_w|}\left[1-\exp\left(-\frac{|r_a|}{\sigma_a^2}\left(\sqrt{1+\frac{2\sigma_a^2}{r_a^2}u^*}-1\right)\right)\right]\\
-%&=\frac{uN}{|\Delta_w|}\left[1-\exp\left(-\frac{|\Delta_a|}{(\lambda_a+\mu_a)^2}\left(\sqrt{1+\frac{2(\lambda_a+\mu_a)^2}{\Delta_a^2}u^*}-1\right)\right)\right],
-%\end{align*}
-%where
-%\begin{equation}
-%u^*=v\left(1-\e^{-\frac{2r_m}{\sigma_m}}\right).
-%\end{equation}
-%The number of rescue mutations has a Poisson distribution with rate $\theta_1+\theta_2$. As a result, the probability of evolutionary rescue is given by:
-%\begin{equation}\label{FellerApprox}
-%p_{rescue}=1-\e^{-\left(\theta_1+\theta_2\right)},
-%\end{equation}
-%which we plot in Figure \ref{P_est_diffusion} as a function of $\Delta_w$.
+To analyze evolutionary rescue in this model, we use the framework of \emph{multitype branching processes} \citep{rybnikov2021fitness,harris1963theory}. 
+This allows us to find explicit expressions for the \emph{survival probability}: the probability that a lineage descended from a single cell does not become extinct.
 
-%\end{appendices}
+Let $p_w$, $p_a$, and $p_m$ be the survival probabilities of a population consisting initially of single wildtype cell, aneuploid cell, or mutant cell, respectively.
+The complements $1-p_w$, $1-p_a$, and $1-p_m$ are the extinction probabilities, which satisfy each its respective equation,
+\begin{equation} \label{extinction_prob}
+\begin{aligned}
+1-p_w = &\frac{\mu_w}{\lambda_w+\mu_w+u+v} + 
+		  \frac{u}{\lambda_w+\mu_w+u+v}\left(1-p_a\right) + \\
+		  & \frac{\lambda_w}{\lambda_w+\mu_w+u+v}\left(1-p_w\right)^2 +
+		  \frac{v}{\lambda_w+\mu_w+u+v}\left(1-p_m\right) ,\\
+1-p_a = &\frac{\mu_a}{\lambda_a+\mu_a+v}+\frac{v}{\lambda_a+\mu_a+v}\left(1-p_m\right)+\frac{\lambda_a}{\lambda_a+\mu_a+v}\left(1-p_a\right)^2 ,\\
+1-p_m = &\frac{\mu_m}{\lambda_m+\mu_m}+\frac{\lambda_m}{\lambda_m+\mu_m}\left(1-p_m\right)^2 .	 
+\end{aligned}
+\end{equation}
+
+The survival probabilities are given by the smallest solution for each quadratic equation \citep{uecker2015adaptive}. Therefore we have
+\begin{equation}\label{survival_prob}
+\begin{aligned}
+p_w &= \frac{\lambda_w-\mu_w-u-v+\sqrt{\left(\lambda_w-\mu_w-u-v\right)^2+4\lambda_w\left(up_a+vp_m\right)}}{2\lambda_w} ,\\
+p_a &= \frac{\lambda_a-\mu_a-v+\sqrt{\left(\lambda_a-\mu_a-v\right)^2+4\lambda_avp_m}}{2\lambda_a}, \\
+p_m &= \frac{\lambda_m-\mu_m}{\lambda_m} .
+\end{aligned} 
+\end{equation}
+Note that the equation for $p_w$ depends on both $p_a$ and $p_m$, and the equation for $p_a$ depends on $p_m$.
+To proceed, we can plug the solution for $p_m$ and $p_a$ into the solution for $p_w$. We perform this for three different scenarios.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsubsection*{Scenario 1: Aneuploid cells are partially resistant}
+
+We first assume that aneuploidy provides partial resistance to drug therapy, $\lambda_a>\mu_a$, and that this resistance is significant, $\left(\lambda_a-\mu_a-v\right)^2 > 4\lambda_a v p_m$.
+We thus rewrite \cref{survival_prob} as
+\begin{align*}
+p_w&=\frac{\lambda_w-\mu_w-u-v}{2\lambda_w}\left(1-\sqrt{1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}}\right) ,
+\text{and} \\
+p_a&=\frac{\lambda_a-\mu_a-v}{2\lambda_a}\left(1+\sqrt{1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}}\right) . 
+\end{align*}
+Using the quadratic Taylor expansion $\sqrt{1+x}=1+x/2+O(x^2)$ and assuming $u,v \ll 1$,
+we obtain the following approximation for the survival probability of a population initially consisting of a single wildtype cell,
+\begin{align}\label{survprobwapprox1}
+p_w 
+&\approx -\frac{vp_m+up_a}{\lambda_w-\mu_w-u-v}\\
+\nonumber
+%&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{v\left(\lambda_a-\mu_a-u\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-u\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\ \label{survprobw2}
+&\approx-\frac{1}{\lambda_w-\mu_w}\left[\frac{u\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\
+\end{align}
+Now $u v$ is very small, and if we assume $v \ll u$, we have
+\begin{equation}
+p_w \approx \frac{u}{\abs{\Delta_w}} \cdot \frac{\Delta_a}{\lambda_a} .
+\end{equation}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\paragraph{Second-order approximation.} % TODO consider moving to Appendix
+To improve our approximation, we can consider the second term of the Taylor series expansion,
+\begin{align*}
+\left(1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}\right)^{\frac{1}{2}}=1+\frac{2\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}-\frac{\left(\lambda_avp_m\right)^2}{4\left(\lambda_a-\mu_a-v\right)^4}+\cdots,
+\end{align*}
+which gives us the following approximation,
+\begin{align}
+p_a \approx \frac{\lambda_a-\mu_a-v}{\lambda_a}+\frac{vp_m}{\lambda_a-\mu_a-v}-\frac{\lambda_a\left(vp_m\right)^2}{8\left(\lambda_a-\mu_a-v\right)^3} .
+\end{align}
+We therefore have
+\begin{align}\nonumber
+p_w&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{u\left(\lambda_a-\mu_a-v\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-v\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}-\frac{uv^2\lambda_a\left(\lambda_m-\mu_m\right)^2}{8\lambda_m^2\left(\lambda_a-\mu_a-v\right)^3}\right]\\ \label{survprobw3}
+&\approx-\frac{1}{\lambda_w-\mu_w}\left[\frac{u\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}-\frac{uv^2\lambda_a\left(\lambda_m-\mu_m\right)^2}{8\lambda_m^2\left(\lambda_a-\mu_a\right)^3}\right],
+\end{align}
+and using $\Delta_k=\lambda_k-\mu_k$, we can write the above equation as
+\begin{equation}\label{survprobwapproxcorrected}
+p_w \approx -\frac{1}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}-\frac{uv^2\lambda_a\Delta_m^2}{8\lambda_m^2\Delta_a^3}\right).
+\end{equation}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsubsection*{Scenario 2: Aneuploid cells are tolerant.}
+
+We now assume that aneuploidy provides tolerance to drug therapy, that is, the number of aneuploid cells significantly declines over time, but at a lower rate than the number of wildtype cells, $\lambda_w - \mu_w < \lambda_a - \mu_a < 0$.
+We also assume that the decline are significant, $\left(\lambda_a-\mu_a-v\right)^2 > 4\lambda_a v p_m$.
+We rewrite \cref{survival_prob} as
+\begin{equation}
+\begin{aligned}
+p_w&=\frac{\lambda_w-\mu_w-u-v}{2\lambda_w}\left(1-\sqrt{1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}}\right) ,
+\text{and} \\
+p_a&=\frac{\lambda_a-\mu_a-v}{2\lambda_a}\left(1-\sqrt{1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}}\right) .
+\end{aligned}
+\end{equation}
+Since $u,v\ll1$, the term in the root can be approximated using a 1st-order Taylor expansion. So, substituting the expressions for $p_a$ and $p_m$, we have
+\begin{equation}\label{survprobwinitial}
+\begin{aligned}
+p_w&\approx-\frac{vp_m+up_a}{\lambda_w-\mu_w-u-v}\\
+&\approx\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-v\right)}-\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\ %\label{survprobw2}
+&\approx\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_w-\mu_w\right)}\left[\frac{u}{\left(\lambda_a-\mu_a\right)}-1\right] \\
+&=\frac{v\Delta_m}{\lambda_m \abs{\Delta_w}}\left(\frac{u}{\abs{\Delta_a}}+1\right) .
+\end{aligned}
+\end{equation}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsubsection*{Scenario 3: Aneuploid cells are non-growing} % TODO consider this name
+
+We now assume that the growth rate of aneuploid cells is close to zero (either positive or negative), such that  $\left(\lambda_a-\mu_a-v\right)^2 < 4\lambda_avp_m$.
+We rewrite \cref{survival_prob} as
+\begin{equation}
+p_a=\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}\left(1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\lambda_avp_m}\right)^{\frac12}}{2\lambda_a} .
+\end{equation}
+Using a following Taylor series expansion for small $\left(\lambda_a-\mu_a-v\right)^2 / 4\lambda_avp_m$,
+\begin{equation*}
+\left(1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\lambda_avp_m}\right)^{\frac{1}{2}}=1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{8\lambda_avp_m}+\cdots,
+\end{equation*}
+we obtain the approximation
+\begin{equation}
+\begin{aligned}
+p_a&\approx\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}\left[1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{8\lambda_avp_m}\right]}{2\lambda_a}\\
+&=\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\sqrt{\lambda_avp_m}}}{2\lambda_a}\\
+&=\frac{\left(\lambda_a-\mu_a-v+2\sqrt{\lambda_avp_m}\right)^2+4\lambda_avp_m}{8\lambda_a\sqrt{\lambda_avp_m}}\\
+&=\frac{4\lambda_avp_m+4\lambda_avp_m\left(1+\frac{\lambda_a-\mu_a-v}{2\sqrt{\lambda_avp_m}}\right)^2}{8\lambda_a\sqrt{\lambda_avp_m}}\\
+&=\frac{1}{2\lambda_a}\left(\lambda_a-\mu_a-v+2\sqrt{\lambda_avp_m}\right).
+\end{aligned}
+\end{equation}
+Plugging this in \cref{survprobwinitial}, the survival probability of a population starting from one wildtype individual is
+\begin{equation}\label{scenario3}
+\begin{aligned}
+p_w&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[v\frac{\lambda_m-\mu_m}{\lambda_m}+\frac{u}{2\lambda_a}\left(\lambda_a-\mu_a-v+2\sqrt{\lambda_avp_m}\right)\right]\\
+&=-\frac{1}{\lambda_w-\mu_w-u-v}\left[v\frac{\lambda_m-\mu_m}{\lambda_m}+\frac{u}{2\lambda_a}\left(\lambda_a-\mu_a-v\right)+u\sqrt{\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_a\lambda_m}}\right].
+%&=-\frac{1}{\Delta_w-u-v}\left[v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right],
+\end{aligned}
+\end{equation}
+
+\end{appendices}
 

### 2023-07-17 EvolutionaryRescue1.tex
add condition on v

@@
 We analyze three scenarios: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_w$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing or "barely growing", that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$.
-Wildtype cells may missegregate to become aneuploids at rate $u$. Both aneuploid and wildtype cells may mutate to become mutants at rate $v$ (\Cref{figureAneuploidy}).
+Wildtype cells may missegregate to become aneuploids at rate $u$.
+Both aneuploid and wildtype cells may mutate to become mutants at rate $v$, which we assume is lower than the division rates, $v<\min{(\lambda_w, \lambda_a, \lambda_m)}$.
+See \Cref{figureAneuploidy} for an illustration of the model.
 
@@
 However, growth may be limited by resources (oxygen, nutrients, etc.) and therefore depend on cell density.
-We therefore performed stochastic simulations of a logistic growth model with carrying capacity $K$ (Methods). 
+We therefore performed stochastic simulations of a logistic growth model with carrying capacity $K$ (see Methods). 
 We find that our approximations agree with results of simulations with density-dependent growth for biologically relevant parameter values (\Cref{rescue_prob_N}).

### 2023-07-19 EvolutionaryRescue1.tex
revise again with RS and DBW

@@
 N_a^* \approx 
-% TODO replace XXX with approximations
   \frac{\abs{\Delta_w}}{u} \cdot \begin{cases}
@@
 Comparing these approximations to results of stochastic evolutionary simulations, we find that the approximations perform very well (\Cref{rescue_prob_wt_growth,rescue_prob_an_growth,rescue_prob_N,rescue_prob_N}).
+\\
+
+The ratio of threshold tumor size for rescue by aneuploidy ($u$ is high) or by direct mutation ($u$ is low) is
+\begin{equation}
+\frac{N^*_a}{N^*_m} \approx \begin{cases}
+    \frac{\abs{\Delta_a}}{u} ,&
+  \Delta_a T^* \ll -1 ,\\ 
+  \frac{1}{u}\left(\frac{\lambda_a}{v} \cdot \frac{\lambda_m}{\Delta_m}\right)^{1/2} ,&
+%  \sqrt{\frac{\Delta_a}{u} \cdot v \frac{\Delta_m}{\lambda_m} \cdot \left(u\frac{\Delta_a}{\lambda_a}\right)^{-1}} ,&
+  -1 \ll \Delta_a T^* \ll 1  ,\\ 
+  v \frac{\Delta_m}{\lambda_m} \cdot \left(u \frac{\Delta_a}{\lambda_a}\right)^{-1}  ,&
+   \Delta_a T^* \gg 1 .
+  \end{cases}
+\end{equation}
+In all cases, the threshold size ratio rapidly decreases with the aneuploidy rate $u$ and the aneuploid growth rate $\Delta_a$ increase, although when $\Delta_a$ is close to zero, it has XXX
+
+In the first case, $\abs{\Delta_a}/u$ is  the ratio of the expected time for an aneuploid lineage appears, $1/u$ and the expected time until that lineage disappears, $1/\Delta_a$.
+In the third case, $v \frac{\Delta_m}{\lambda_m} \cdot \left(u \frac{\Delta_a}{\lambda_a}\right)^{-1}$ is the ratio of the rate of formation of resistant mutants that avoid extinction and partially resistant aneuploid that avoid extinction.
+In the second case, $\frac{1}{u}\left(\frac{\lambda_a}{v} \cdot \frac{\lambda_m}{\Delta_m}\right)^{1/2}=\sqrt{\frac{\Delta_a}{u} \cdot v \frac{\Delta_m}{\lambda_m} \cdot \left(u\frac{\Delta_a}{\lambda_a}\right)^{-1}}$, which is the geometric mean of the first and third cases.
+
+Interestingly, if we increase $\lambda_a$ and $\mu_a$ such that $\Delta_a$ remains constant, $T^*$ decreases, and we are in the middle case.
+In this case, increasing $\lambda_a$ should also increase the mutation rate in aneuploid cells, $v$, so overall $d \lambda_a T^*/d\lambda_a = 0$ and $N_a^*$ is unaffected by increasing $\lambda_a$.
+% TODO PLUG IN REAL VALUES
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\paragraph*{Standing vs. de-novo genetic variation}
+\paragraph*{Standing vs. de-novo genetic variation.}
 
 In the above we assumed that upon beginning of drug therapy, the initial tumor consisted entirely of wildtype cells.
-However, aneuploid cells are likely generated even before onset of treatment at some rate $\tilde{u} \le u$ (because the treatment itself may promote generation of aneuploid cells REF), which are likely to have a deleterious effect (REF).
+However, aneuploid cells are likely generated even before onset of treatment at some rate $\tilde{u} \le u$ (because the treatment itself may promote generation of aneuploid cells REF), which are likely to have a deleterious effect, $s$ (REF).
 % TODO refs
-But if the number of cells in the tumor $N$ is large, as expected if drug treatment is applied, there may already be a fraction $f=\tilde{u}/s$ of aneuploid cells in the population, where $s$ is the cost of aneuploidy (REF).
+But if the number of cells in the tumor $N$ is large (as expected if the tumor is to be treated with a drug), there may already be a fraction $f=\tilde{u}/s$ of aneuploid cells in the population (REF).
 
-In this scenario, the probability of evolutionary rescue by cells with aneuploidy from the initial population is
-\begin{equation*}
-% redefine: p_total, p_denovo=p_resuce from eq above; p_stdvar is the new thing
-p_{sgv} = 1-\left(1-p_a\right)^{fN}\approx 1-\e^{-fNp_a}. 
-\end{equation*}
-The total probability of evolutionary rescue is given by
-\begin{align}\nonumber
-p_{rescue} 	&= p_{sgv}+\left(1-p_{sgv}\right)p_{de-novo} \\
-			&= 1-\exp\left(-\left[\left(1-f\right)p_w + fp_a\right]N\right) .
-\end{align}
+Therefore, the threshold tumor size with standing generation variation, $\tilde{N}^*_{a}$, is similar to the ratio with de-novo aneuploidy, except that the sensitive growth rate $\abs{\Delta_w}$ is replaced with the aneuploidy cost, $s$, such that
+\begin{equation}
+\frac{\tilde{N}^*_{a}}{N^*_{a}} = \frac{u}{\tilde{u}} \; \frac{s}{\abs{\Delta_w}} .
+\end{equation}
 
-The fraction of cases in which the population is rescued by pre-existing aneuploid cells (i.e., standing genetic variation) is given by $F\left(f\right)=\frac{p_{sgv}}{p_{total}}$ (\Cref{FractionPlot}).
+Therefore, standing genetic variation will drive adaptation to the drug if $\Delta_w$ is very negative due to a stronger effect of the drug on sensitive cells, or if $s$ is very small due to a low cost of aneuploidy in the pre-drug conditions.
+In contrast, de-novo aneuploids will have a stronger effect on adaptation if the cost $s$ is large, the drug is weak ($\Delta_w$ is small) or if the drug induces the appearance of aneuploid cells ($u > \tilde u$).
+
+% TODO PLUG IN REAL VALUES
+
+%In this scenario, the probability of evolutionary rescue by cells with aneuploidy from the initial population is
+%\begin{equation*}
+%% redefine: p_total, p_denovo=p_resuce from eq above; p_stdvar is the new thing
+%p_{sgv} = 1-\left(1-p_a\right)^{fN}\approx 1-\e^{-fNp_a}. 
+%\end{equation*}
+%The total probability of evolutionary rescue is given by
+%\begin{align}\nonumber
+%p_{rescue} 	&= p_{sgv}+\left(1-p_{sgv}\right)p_{de\;novo} \\
+%			&= 1-\exp\left(-\left[\left(1-f\right)p_w + fp_a\right]N\right) .
+%\end{align}
+%
+%The fraction of cases in which the population is rescued by pre-existing aneuploid cells (i.e., standing genetic variation) is given by $F\left(f\right)=\frac{p_{sgv}}{p_{total}}$ (\Cref{FractionPlot}).
 
@@
 
-\subsection*{Effect of aneuploidy on the probability of evolutionary rescue}
-To determine the extent to which aneuploidy may affect evolutionary rescue, we define $H$ to be the ratio of the rescue probability with and without aneuploidy ($u>0$ and $u=0$, respectively),
-\begin{equation}\label{ratiorescueexact}
-H = \frac{p_{rescue}(u>0)}{p_{rescue}(u=0)}.
-\end{equation}
+% TODO Move to appendix
+%\subsection*{Effect of aneuploidy on the probability of evolutionary rescue}
+%To determine the extent to which aneuploidy may affect evolutionary rescue, we define $H$ to be the ratio of the rescue probability with and without aneuploidy ($u>0$ and $u=0$, respectively),
+%\begin{equation}\label{ratiorescueexact}
+%H = \frac{p_{rescue}(u>0)}{p_{rescue}(u=0)}.
+%\end{equation}
 
-Plugging in our approximations from \cref{rescue_prob}, we have
-\begin{equation}\label{ratiorescue}
-H = \begin{cases}
-\frac{1-\exp\left[\frac{N}{\Delta_w-u-v}\left(v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right)\right]}{1-\exp\left[\frac{vN\Delta_m}{\left(\Delta_w-v\right)\lambda_m}\right]} ,&
-4\lambda_avp_m>\left(\Delta_a-v\right)^2 ,\\
-\frac{1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right]}{1-\exp\left(\frac{v\Delta_mN}{\lambda_m\Delta_w}\right)}, ,&
-\Delta_a<0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2 ,\\
-\frac{1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right]}{1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\right]} ,&
-\Delta_a>0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2. % TODO this term can be simplified?
-  \end{cases}
-\end{equation}
-
-We find that the rescue ratio increase with the aneuploidy growth rate $\Delta_a$, because the better aneuploid cells are in growth, the better they are at rescuing the population (when they provide partial resistance) or delaying the extinction of the population (when they provide tolerance).
-However, the rescue decreases with the wildtype growth rate $\Delta_w$, because the better the wildtype is at growth, the less is depends on aneuploidy for rescue or delay, and the more likely it is to directly produce mutant cells, rather than relying on aneuploid cells for producing mutant cells (\Cref{rescue_ratio}).
-The effect of the initial tumor size $N$ is the similar to that of the wildtype growth rate. 
-Importantly, in large tumors, the ratio converges to unity, that is, aneuploidy does not affect the probability for evolutionary rescue. % TODO can we find this analytical?
+%Plugging in our approximations from \cref{rescue_prob}, we have
+%\begin{equation}\label{ratiorescue}
+%H = \begin{cases}
+%\frac{1-\exp\left[\frac{N}{\Delta_w-u-v}\left(v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right)\right]}{1-\exp\left[\frac{vN\Delta_m}{\left(\Delta_w-v\right)\lambda_m}\right]} ,&
+%4\lambda_avp_m>\left(\Delta_a-v\right)^2 ,\\
+%\frac{1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right]}{1-\exp\left(\frac{v\Delta_mN}{\lambda_m\Delta_w}\right)}, ,&
+%\Delta_a<0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2 ,\\
+%\frac{1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right]}{1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\right]} ,&
+%\Delta_a>0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2. % TODO this term can be simplified?
+%  \end{cases}
+%\end{equation}
+%
+%We find that the rescue ratio increase with the aneuploidy growth rate $\Delta_a$, because the better aneuploid cells are in growth, the better they are at rescuing the population (when they provide partial resistance) or delaying the extinction of the population (when they provide tolerance).
+%However, the rescue decreases with the wildtype growth rate $\Delta_w$, because the better the wildtype is at growth, the less is depends on aneuploidy for rescue or delay, and the more likely it is to directly produce mutant cells, rather than relying on aneuploid cells for producing mutant cells (\Cref{rescue_ratio}).
+%The effect of the initial tumor size $N$ is the similar to that of the wildtype growth rate. 
+%Importantly, in large tumors, the ratio converges to unity, that is, aneuploidy does not affect the probability for evolutionary rescue.
 
@@
 
-\subsection*{Evolutionary rescue time}
+\subsection*{Recurrence time due to evolutionary rescue}
 
@@
     $N$ & Initial tumor size & $10^7-10^9$ & cells  & \citet{del2009does} \\ \hline
-    $\lambda_w$ & Wildtype division rate& 0.14 & 1/days  & \citep{bozic2013evolutionary} \\ \hline
+    $\lambda_w$ & Wildtype division rate& 0.14 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
     $\mu_w$ & Wildtype death rate& 0.17 & 1/days  & \citet{bozic2013evolutionary} \\ \hline

### 2023-07-20 EvolutionaryRescue1.tex
revised evol time section

@@
 Aneuploidy may contribute to evolutionary rescue by either preventing (when $\Delta_a>0$) or delaying (when $0>\Delta_a>\Delta_w$) the extinction of the population before mutant cells appear and fixate.
-
-To estimate the rescue probability $\presc$, we assume independence between clonal lineages starting from an initial population of $N$ wildtype cells (we check the effect of density-dependent growth on our results below).
-Thus, the rescue probability is given by 
-\begin{subequations}
-\begin{align}
-\presc &= 1-\left(1-p_w\right)^N \label{rescue_prob} \\
-& \approx 1-\e^{-Np_w} , \label{rescue_prob2}
-\end{align}
-\end{subequations}
-where $p_w$ is the probability that the lineage of a single wild-type cell avoids extinction, and the approximation $(1-p_w)\approx e^{-p_w}$ assumes that $p_w$ (but not necessarily $N p_w$) is small.
-
-The probability for evolutionary rescue $\presc$  (\cref{rescue_prob2}) is $\presc \approx N p_w$ when $N<1/p_w$ and $\presc \approx 1$ when $N > 1/p_w$.
-Therefore, $N^*=1/p_w$ is the threshold tumor size for rescue above which rescue is very likely (\Cref{rescue_prob_N}).
-% TODO add 1/p_w to figure as vertical line
-Therefore, we can write
-\begin{equation}
-\presc = 1-e^{-N/N^*} .
+We assume independence between clonal lineages starting from an initial population of $N$ wildtype cells (we check the effect of density-dependent growth on our results below).
+We therefore define $p_w$ as the probability that a lineage starting from a single wildtype cell avoids extinction by acquiring drug resistance.
+Thus, $N^*=1/p_w$ is the threshold tumor size above which evolutionary rescue is very likely, and the rescue probability is given by 
+\begin{equation} \label{eq:rescue_prob} 
+\presc = 
+1-\left(1-p_w\right)^N \approx
+1-\e^{-Np_w} = 
+1-e^{-N/N^*} .
 \end{equation}
-In the Appendix below, we use the theory of multi-type branching processes to find approximate expressions \cref{survprobwapprox1,survprobwinitial,scenario3} for $p_w$ in different regimes. 
-Substituting these  into $N^*=1/p_w$, we find approximations for the threshold tumor size, $N^*$. 
+where the approximation $(1-p_w)\approx e^{-p_w}$ assumes that $p_w$ (but not necessarily $N p_w$) is small.
+Indeed, when $N<1/p_w$, then the probability for evolutionary rescue is $\presc \approx N p_w$  and when $N > 1/p_w$, it is $\presc \approx 1$, justifying the definition of $N^*$ as the threshold tumor size.
 \\
 
-In the following, an important quantity is $T^* = (4 v \lambda_a \Delta_m/\lambda_m)^{-1/2}$, which is the critical time an aneuploid lineage needs to survive to produce a resistant mutant that avoids random extinction.
+In the Appendix, we use the theory of multi-type branching processes to find approximate expressions \cref{eq:survprobwapprox1,eq:survprobwinitial,eq:scenario3} for $p_w$ in different regimes. 
+Substituting these  into $N^*=1/p_w$, we find approximations for the threshold tumor size, $N^*$. 
+For these approximations, an important quantity is $T^* = (4 v \lambda_a \Delta_m/\lambda_m)^{-1/2}$, which is the critical time an aneuploid lineage needs to survive to produce a resistant mutant that avoids random extinction.
 First, if aneuploidy is very rare ($u T^*< 1$), or if aneuploidy is rare ($u < -\Delta_a$) and very sensitive to the drug ($\Delta_a T^* < -1$), then rescue will likely occur by a direct resistance mutation in a sensitive cell, such that
-\begin{equation}
+\begin{equation} \label{eq:N_m}
 N_m^* \approx \frac{\abs{\Delta_w}}{v} \cdot \frac{\lambda_m}{\Delta_m} .
 \end{equation}
-Otherwise, aneuploidy is frequent enough ($u > \max{(-\Delta_a, 1/T^*)}$) to affect the evolution of the tumor. 
+Here, $\abs{\Delta_w}/v$ is the ratio of XXX.
+
+Otherwise, aneuploidy is frequent enough ($u > \max{(-\Delta_a, 1/T^*)}$) to affect the evolution of drug resistance. 
 The threshold tumor size, $N^*$, can then be approximated by one of the following cases, depending on $\Delta_a T^*$, the change in the log of the aneuploid population size during the critical time,
-\begin{equation}\label{rescue_prob_approx}
+\begin{equation}  \label{eq:N_a}
 \begin{aligned}
@@
 Comparing these approximations to results of stochastic evolutionary simulations, we find that the approximations perform very well (\Cref{rescue_prob_wt_growth,rescue_prob_an_growth,rescue_prob_N,rescue_prob_N}).
-\\
 
-The ratio of threshold tumor size for rescue by aneuploidy ($u$ is high) or by direct mutation ($u$ is low) is
-\begin{equation}
+Using \cref{eq:N_a,eq:N_m}, we can find the ratio of threshold tumor size for rescue via aneuploidy ($u$ is high) or via direct mutation ($u$ is low),
+\begin{equation} \label{eq:N_ratio}
 \frac{N^*_a}{N^*_m} \approx \begin{cases}
@@
 \end{equation}
-In all cases, the threshold size ratio rapidly decreases with the aneuploidy rate $u$ and the aneuploid growth rate $\Delta_a$ increase, although when $\Delta_a$ is close to zero, it has XXX
+In all cases, the effect of aneuploidy increases (i.e., the threshold size ratio decreases) when the aneuploidy rate $u$ increases.
+Increasing the aneuploid growth rate $\Delta_a$ also leads to an increased role of aneuploidy, but only in the first and third case.
 
-In the first case, $\abs{\Delta_a}/u$ is  the ratio of the expected time for an aneuploid lineage appears, $1/u$ and the expected time until that lineage disappears, $1/\Delta_a$.
-In the third case, $v \frac{\Delta_m}{\lambda_m} \cdot \left(u \frac{\Delta_a}{\lambda_a}\right)^{-1}$ is the ratio of the rate of formation of resistant mutants that avoid extinction and partially resistant aneuploid that avoid extinction.
+In the first case, $\abs{\Delta_a}/u$ is  the ratio of the expected time for an aneuploid lineage to appear, $1/u$, and the expected time until that lineage disappears, $1/\Delta_a$.
+In the third case, $\left(v \frac{\Delta_m}{\lambda_m}\right) / \left(u \frac{\Delta_a}{\lambda_a}\right)$ is the ratio of the rates of formation of resistant mutants that avoid extinction and partially resistant aneuploids that avoid extinction.
 In the second case, $\frac{1}{u}\left(\frac{\lambda_a}{v} \cdot \frac{\lambda_m}{\Delta_m}\right)^{1/2}=\sqrt{\frac{\Delta_a}{u} \cdot v \frac{\Delta_m}{\lambda_m} \cdot \left(u\frac{\Delta_a}{\lambda_a}\right)^{-1}}$, which is the geometric mean of the first and third cases.
 
-Interestingly, if we increase $\lambda_a$ and $\mu_a$ such that $\Delta_a$ remains constant, $T^*$ decreases, and we are in the middle case.
-In this case, increasing $\lambda_a$ should also increase the mutation rate in aneuploid cells, $v$, so overall $d \lambda_a T^*/d\lambda_a = 0$ and $N_a^*$ is unaffected by increasing $\lambda_a$.
+Interestingly, increasing both the aneuploid division rate, $\lambda_a$, and the aneuploid death rate, $\mu_a$, such that the growth rate $\Delta_a$ remains constant, leads to decreases in $T^*$, and therefore to the second case.
+In this case, increasing the division rate $\lambda_a$ should also increase the mutation rate $v$ in aneuploid cells, as mutations mostly occur during division , so overall the threshold tumor size $N_a^*$ is unaffected by the division rate $\lambda_a$ (i.e., $d \lambda_a T^*/d\lambda_a = 0$).
+Thus, if aneuploids rapidly die due to the drug but compensate by rapidly dividing, further increasing  the division rate will \emph{not} facilitate adaptation.
 % TODO PLUG IN REAL VALUES
-
-% TODO move this to appendix
-%Substituting these  into \cref{rescue_prob2}, we find that the rescue probability can be approximated by
-%\begin{equation}\label{rescue_prob_approx}
-%\begin{aligned}
-%&\presc \approx \\
-%  &\begin{cases}
-%  1-\exp\left[\frac{N}{\Delta_w-u-v}\left(v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right)\right] ,&
-%  4\lambda_avp_m>\left(\Delta_a-v\right)^2 ,\\
-%   1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right] ,&
-%   \Delta_a<0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2 ,\\
-%   1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right] ,&
-%   \Delta_a>0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2 .
-%  \end{cases}
-%\end{aligned}
-%\end{equation}
 
@@
 
-In our analysis we used branching processes, which assume that growth (division and death) are density-independent.
+In our analysis we used branching processes, which assume that growth (division and death) is density-independent.
 However, growth may be limited by resources (oxygen, nutrients, etc.) and therefore depend on cell density.
@@
 
-In the above we assumed that upon beginning of drug therapy, the initial tumor consisted entirely of wildtype cells.
-However, aneuploid cells are likely generated even before onset of treatment at some rate $\tilde{u} \le u$ (because the treatment itself may promote generation of aneuploid cells REF), which are likely to have a deleterious effect, $s$ (REF).
+In the above we assumed that at the onset of drug treatment, the initial tumor consisted entirely of wildtype cells that are drug sensitive.
+However, aneuploid cells are likely generated even before onset of treatment at some rate $\tilde{u} \le u$ (because the treatment itself may promote generation of aneuploid cells REF), which are likely to have a deleterious effect in the absence of the drug, $s$ (REF).
 % TODO refs
-But if the number of cells in the tumor $N$ is large (as expected if the tumor is to be treated with a drug), there may already be a fraction $f=\tilde{u}/s$ of aneuploid cells in the population (REF).
+But if the number of cells in the tumor $N$ is large (as expected if the tumor is to be treated with a drug), there may already be a fraction $f \approx \tilde{u}/s$ of aneuploid cells in the population.
 
-Therefore, the threshold tumor size with standing generation variation, $\tilde{N}^*_{a}$, is similar to the ratio with de-novo aneuploidy, except that the sensitive growth rate $\abs{\Delta_w}$ is replaced with the aneuploidy cost, $s$, such that
+Therefore, the threshold tumor size with standing generation variation, $\tilde{N}^*_{a}$, is similar to the ratio with de-novo variation, $N^*_a$, except that the sensitive growth rate $\abs{\Delta_w}$ is replaced with the aneuploidy cost, $s$, such that
 \begin{equation}
@@
 Therefore, standing genetic variation will drive adaptation to the drug if $\Delta_w$ is very negative due to a stronger effect of the drug on sensitive cells, or if $s$ is very small due to a low cost of aneuploidy in the pre-drug conditions.
-In contrast, de-novo aneuploids will have a stronger effect on adaptation if the cost $s$ is large, the drug is weak ($\Delta_w$ is small) or if the drug induces the appearance of aneuploid cells ($u > \tilde u$).
+In contrast, de-novo aneuploids will have a stronger effect on adaptation if the aneuploidy cost $s$ is large, the effect of the drug is weak ($\Delta_w$ is small), or if the drug induces the appearance of aneuploid cells ($u > \tilde u$).
 
@@
 
-%Plugging in our approximations from \cref{rescue_prob}, we have
+%Plugging in our approximations from \cref{eq:rescue_prob}, we have
 %\begin{equation}\label{ratiorescue}
@@
 
-Even when evolutionary rescue occurs, it may take a long time.
-We therefore wish to estimate the mean waiting time for rescue and the effect aneuploidy may have on it. 
-We calculate the mean time for the appearance of the first mutant that rescues the cell population.
-This can occur either through the evolutionary trajectory $wildtype \rightarrow mutant$ or through the trajectory $wildtype \rightarrow aneuploid \rightarrow mutant$.
-We start with the former.
+Even when evolutionary rescue occurs and leads to recurrence of the tumor, it may take a long time.
+The overall expected recurrence time can be estimated by adding two terms: the mean waiting time for evolutionary rescue--the appearance of a resistant lineage that avoid extinction--and the expected time for proliferation of that lineage back to the original tumor size, $N$.
 
-Assuming no aneuploidy ($u=0$), we define $T_1$ to be the time at which the first mutant cell appears that will avoid extinction and will therefore rescue the population.
-Note that if extinction occurs, that is the frequency of mutants after a very long time is zero, $m_{\infty}=0$, then it is implied that $T_1=\infty$, and vice versa if $T_1<\infty$ then $m_{\infty}>0$.
+\paragraph{Evolutionary rescue time.}
+In Appendix C we derive an approximation for $\tau_1$, the expected rescue time without aneuploidy ($u=0$), and $\tau_2$, the expected rescue time with aneuploidy ($u>0$),
+\begin{align}
+\tau_1 &\approx \frac{1+\e^{-Np_w}}{u N p_a} , \label{eq:limitapprox} ,\\
+\tau_2 &\approx \frac{1+\e^{-Np_w}}{u N p_m} , \label{eq:limitapprox3} .
+\end{align}
+\Cref{MeanTimeGrowthAneuploidyPlot} shows the agreement between this approximations and simulation results for intermediate and large tumor sizes.
 
-The number of successful mutants generated until time $t$ can be approximated by an inhomogeneous Poisson process with rate $R\left(t\right) = u p_a w_t$,
-where $w_t=N\e^{\Delta_w t}$ is the number of wildtype cells at time $t$.
-Note that 
-\begin{equation}\label{integralR}
-\int_0^{t}{R(z)\d z} = 
-u p_a N \frac{\exp[{\Delta_w t}]-1}{\Delta_w} \approx 
-u p_a N t,
-\end{equation}
-by integrating the exponential and because $\frac{\exp[\Delta_w t]-1}{\Delta_w}=\frac{1+\Delta_w t+O(t^2)-1}{\Delta_w}=t+O(t^2)$.
-The probability density function of $T_1$ is thus
-$R\left(t\right)\exp\left(-\int_0^{t}{R(z)\d z}\right)$. % TODO ref
-Therefore, the probability density function of the conditional random variable $(T_1 \mid T_1 < \infty)$ is
-$f_1(t) = \frac{R\left(t\right)\exp\left(-\int_0^{t}{R(z)\d z}\right)}{p_{rescue}}$. 
-\\
-
-We are interested in the mean conditional time, $\tau_1=\mathbb{E}\left[T_1 \mid T_1<\infty\right]$, which is given by
-\begin{equation}\label{meantime1}
-\begin{aligned}
-\tau_1 =
-\int_{0}^{\infty}{t f_1(t) \d t} = 
-\frac{\int_{0}^{\infty}{tR(t)\exp\left(-\int_0^{t}{R(z)\d z}\right) \d t}}{p_{rescue}} = 
-\frac{\int_{0}^{\infty}{\exp\left(-\int_0^{t}{R(z)\d z}\right) \d t}}{p_{rescue}}
-\end{aligned}
-\end{equation}
-after applying integration by parts.
-Therefore, plugging \cref{integralR,rescue_prob2} in \cref{meantime1}, 
-\begin{align}\label{limitapprox}
-\tau_1 = 
-%\frac{\int_0^\infty\exp\left(-\int_0^\tau R(z)\d z\right)\d t}{1-\left(1-p_w\right)^N} = 
-\frac{\int_0^\infty\e^{-uNp_a\frac{\e^{\Delta_w t}-1}{\Delta_w}} \d t}{1-\left(1-p_w\right)^N} \approx
-\frac{\int_{0}^{\infty}{\exp\left(-u p_a N t\right) \d t}}{1-\e^{-Np_w}} \approx \\
-\left(1+\e^{-Np_w}\right)\int_0^\infty\e^{-u p_a N t} \d t =
-\frac{1+\e^{-Np_w}}{u p_a N} ,
-\end{align}
-where we use the approximations 
-$\frac{\e^{\Delta_w\tau}-1}{\Delta_w}=\frac{1+\Delta_w\tau+O(\tau^2)-1}{\Delta_w}=\tau+O(\tau^2)$ and $(1-\e^{-Np_w})^{-1} \approx 1+\e^{-Np_w}$ and integrate the exponent.
-\Cref{MeanTimeGrowthAneuploidyPlot}B show the agreement between this approximating and simulation results for intermediate and large tumor sizes.
-\\
-
-When $Nu\gg1$ the aneuploid frequency dynamics is roughly deterministic and therefore can be approximated by 
-\begin{equation}
-a_t \approx \frac{Nu\e^{\Delta_wt}}{\Delta_w-\Delta_a}\left[1-\e^{\left(\Delta_w-\Delta_a\right)t}\right].
-\end{equation}
-As a result, when $N\gg1$ % TODO N>>1 ? or Nu >> 1?
-the number of successful mutants created by direct mutation and via aneuploidy can be approximated by inhomogeneous Poisson processes with the rates
-\begin{align}
-r_1\left(t\right)&=vp_m\int_0^ta_{z} \d z = \frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_wt}-1}{\Delta_w}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right),\\
-r_2\left(t\right)&=vp_m\int_0^tw_{z} \d z = vNp_m\frac{\e^{\Delta_w t}-1}{\Delta_w}.
-\end{align}
-For large initial population sizes we assume that the two processes are independent and as a result, they can be merged into a single Poisson process with rate $\left(r_1+r_2\right)\left(t\right)$.
-Consequently, the mean time to the appearance of the first rescue mutant is
-\begin{align}\label{meantimet2}
-\tau_2 = 
-\frac{\int_0^\infty\e^{-\left(r_1(t)+r_2(t)\right)} \d t}{1-\left(1-p_w\right)^N} = 
-\frac{\int_0^\infty\exp\left[-\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_w t}-1}{\Delta_w}-\frac{\e^{\Delta_a t}-1}{\Delta_a}\right)-vNp_m\frac{\e^{\Delta_w t}-1}{\Delta_w}\right] \d t}{1-\left(1-p_w\right)^N},
-\end{align}
-which we plot in \Cref{MeanTimeGrowthAneuploidyPlot}A as a function of the initial population size, $N$.
-
-We wish to obtain a simpler formula for $\tau_2$, similar to \cref{limitapprox}.
-We thus have the following expansions,
-%\begin{align*}
-%\frac{\e^{\Delta_w\tau}-1}{\Delta_w}&=\frac{1+\Delta_w\tau+\frac{\Delta_w^2\tau^2}{2}+O(\tau^3)-1}{\Delta_w}=\tau+\frac{\Delta_w}{2}\tau^2+O(\tau^3) ,\\
-%\frac{\e^{\Delta_a\tau}-1}{\Delta_a}&=\frac{1+\Delta_a\tau+\frac{\Delta_a^2\tau^2}{2}+O(\tau^3)-1}{\Delta_a}=\tau+\frac{\Delta_w}{2}\tau^2+O(\tau^3) ,
-%\end{align*}
-%which allow us to write
-%\begin{equation*}
-%\frac{\e^{\Delta_w\tau}-1}{\Delta_w}-\frac{\e^{\Delta_a\tau}-1}{\Delta_a} \approx \frac{\left(\Delta_w-\Delta_a\right)\tau^2}{2}.
-%\end{equation*}
-%As a result, the integrand in \cref{meantimet2} can be written as:
-%\begin{align*}
-%&\exp\left[-\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_w\tau}-1}{\Delta_w}-\frac{\e^{\Delta_a\tau}-1}{\Delta_a}\right)-vNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}\right]\approx \\
-%& \exp\left(-uvNp_m\tau^2-vNp_m\tau\right) =
-%\exp\left(\frac{vNp_m}{2}\right)\exp\left[-\frac{uvNp_m}{2}\left(\tau+\frac{1}{u}\right)\right].
-%\end{align*}
-%Consequently, the mean conditional time $\tau_2$ is approximated by
-%\begin{align}\label{limitapprox2}
-%\tau_2 \approx
-%\frac{\left(1+\exp\left(-Np_w\right)\right) \exp\left(\frac{vNp_m}{2u}\right) \erfc\left(\sqrt{\frac{vNp_m}{2u}}\right)}{\sqrt{\frac{2uvNp_m}{\pi}}},
-%\end{align}
-%where $\erfc(z)$ is the complementary error function. 
-%
-%If we apply linear approximations,
-\begin{align*}
-\frac{\e^{\Delta_w t}-1}{\Delta_w}&=\frac{1+\Delta_w t+O(t^2)-1}{\Delta_w} = t+O(t^2),\\
-\frac{\e^{\Delta_a t}-1}{\Delta_a}&=\frac{1+\Delta_a t+O(t^2)-1}{\Delta_a} = t+O(t^2),
-\end{align*}
-which we use to derive a first-order approximation for $\tau_2$,
-\begin{align}\label{limitapprox3}
-\tau_2 \approx
-\left(1+\e^{-Np_w}\right)\int_0^\infty\e^{-u N p_m t} \d t = 
-\frac{\left(1+\e^{-Np_w}\right)}{uNp_m} ,
+Thus, the ratio of the expected rescue time with and without aneuploidy ($u>0$ and $u=0$, respectively),
+\begin{align}\label{eq:eqMeanTimeRatioInitialPopulationSize}
+\frac{\tau_1}{\tau_2}  \approx \frac{p_a}{p_m} = 
+%&\frac{\lambda_m}{\lambda_a} \; \frac{\Delta_a-v+\sqrt{(\Delta_a - v)^2 + (1/T^*)^2}}{2 \Delta_m} \approx \\
+%&\frac{1}{2} \; \frac{\lambda_m}{\Delta_m} \left[ \left(\frac{\lambda_a}{\Delta_a}\right)^{-1} + \sqrt{\frac{\left(\Delta_a\right)^2 + (1/T^*)^2}{\lambda_a^2}}\right] = \\
+\frac{1}{2} \; \frac{\lambda_m}{\Delta_m} \left[ \left(\frac{\lambda_a}{\Delta_a}\right)^{-1} + \sqrt{\left(\frac{\lambda_a}{\Delta_a}\right)^{-2} + 4v \left(\lambda_a\frac{\lambda_m}{\Delta_m}\right)^{-1}}\right] .
 \end{align}
 
-\Cref{MeanTimeGrowthAneuploidyPlot}A shows that the approximation \cref{limitapprox3} has a good fit with simulation results for large initial wildtype population size ($N > 2\cdot10^7$).
-An approximation that uses the second-order terms, $\frac{\Delta_w^2 t^2}{2}$ and $\frac{\Delta_a^2 t^2}{2}$, does not perform better (\Cref{MeanTimeGrowthAneuploidyPlot}A).
-\\
+\Cref{MeanTimeRatioInitialPopulationSize} shows how this ratio, $\tau_1/\tau_2$, changes as a function of the initial tumor size, $N$, for different aneuploid growth rates, $\Delta_a$.
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Effect of aneuploidy on mean evolutionary rescue time}
+\paragraph{Proliferation time.}
 
-Next, we wish to find the ratio of the mean rescue time with and without aneuploidy ($u>0$ and $u=0$, respectively),
-\begin{align}\label{eqMeanTimeRatioInitialPopulationSize}
-\frac{\tau_2}{\tau_1}  
-%\frac{\int_0^\infty\exp\left[-\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_w t}-1}{\Delta_w}-\frac{\e^{\Delta_a t}-1}{\Delta_a}\right)-vNp_m\frac{\e^{\Delta_w t}-1}{\Delta_w}\right] \d t}{\int_0^\infty\e^{-uNp_m\frac{\e^{\Delta_w t}-1}{\Delta_w}} \d t} \cdot \frac{1-\left(1-p_w(u=0)\right)^N}{1-\left(1-p_w(u>0)\right)^N}\\
-= H^{-1} \cdot \frac{\int_0^\infty\exp\left[-\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_wt}-1}{\Delta_w}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right)-vNp_m\frac{\e^{\Delta_w t}-1}{\Delta_w}\right] \d t}{\int_0^\infty\e^{-vNp_m\frac{\e^{\Delta_w t}-1}{\Delta_w}} \d t},
-\end{align}
-where $H$ is the ratio of the rescue probability with and without aneuploidy, see \cref{ratiorescueexact}.
-\Cref{MeanTimeRatioInitialPopulationSize} shows how this ratio, $\tau_2/\tau_1$, changes as a function of the initial tumor size, $N$, for different aneuploid growth rates, $\Delta_a$.
+TODO
 
@@
 
-We have modeled a tumor--a population of cancer cells--exposed to drug therapy that causes the population to decline in size towards potential extinction.
-The cancer cell population can be "evolutionary rescued", or escape extinction, by two paths.
+We have modeled a tumor--a population of cancer cells--exposed to drug treatment that causes the population to decline in size towards potential extinction.
+In this scenario, the tumor can be "evolutionary rescued", or escape extinction, via two paths.
 In the direct path, a sensitive cell acquires a mutation that confers resistance that allows it to rapidly grow.
@@
 Using multitype branching processes, we derived the probability of evolutionary rescue of the population of cancer cells under different scenarios for the effect of aneuploidy, ranging from tolerance to partial resistance.
-We obtained exact and approximate expressions for the probability of evolutionary rescue (\cref{rescue_prob}).
+We obtained exact and approximate expressions for the probability of evolutionary rescue (\cref{eq:rescue_prob}).
 Our results show that the probability of evolutionary rescue increases with the initial tumor size $N$, the sensitive growth rate $\Delta_w$, the mutation rate $v$, and the aneuploidy rate $u$.
@@
 In both scenarios, rescue probability increase with both the wildtype growth rate (x-axis) and the aneuploidy growth rate (colors).
-Markers represent simulation results with 95\% CI; solid and dashed lines for the exact formula (\cref{survival_prob} in \cref{rescue_prob}); dashed lines for the approximate formula (\cref{rescue_prob_approx}), demonstrating that they all agree.
+Markers represent simulation results with 95\% CI; solid and dashed lines for the exact formula (\cref{eq:survival_prob} in \cref{eq:rescue_prob}); dashed lines for the approximate formula (\cref{rescue_prob_approx}), demonstrating that they all agree.
 Parameters: division rate $\lambda_w=\lambda_a=\lambda_m=0.14$ (so that growth rate changes due to variable death rate); mutant death rate $\mu_m=0.13$ (so that mutant growth rate $\Delta_m=0.01$); aneuploidy rate $u=10^{-2}$; mutation rate $v=10^{-7}$.
@@
 Markers for simulation results with 95\% CI; 
-solid lines for the exact formula (\cref{survival_prob} in \cref{rescue_prob});
+solid lines for the exact formula (\cref{eq:survival_prob} in \cref{eq:rescue_prob});
 dashed lines for the approximate formula (\cref{rescue_prob_approx}).
@@
 \caption{\textbf{Evolutionary rescue probability for variable initial tumor size.}
-\textbf{(A)} Comparison of simulation results (markers with 95\% CI, too small to appear with $10^5$ simulations per marker), the exact formula (blue line, \cref{survival_prob} in \cref{rescue_prob}) and the approximate formula (orange line, \cref{rescue_prob_approx}).
-\textbf{(B)} Comparison of results of simulations  with density-dependent growth (markers with with 95\% CI) and the exact formula (blue line, \cref{survival_prob} in \cref{rescue_prob}) with maximum carrying capacity $K=10^9$.
+\textbf{(A)} Comparison of simulation results (markers with 95\% CI, too small to appear with $10^5$ simulations per marker), the exact formula (blue line, \cref{eq:survival_prob} in \cref{eq:rescue_prob}) and the approximate formula (orange line, \cref{rescue_prob_approx}).
+\textbf{(B)} Comparison of results of simulations  with density-dependent growth (markers with with 95\% CI) and the exact formula (blue line, \cref{eq:survival_prob} in \cref{eq:rescue_prob}) with maximum carrying capacity $K=10^9$.
 Parameters: $\lambda_w=\lambda_a=\lambda_m=0.14$; $\mu_w=0.17$; (A) $\mu_a=0.15$, (B) $\mu_a=0.135$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
@@
 \textbf{(C, D)} Aneuploidy provides tolerance.  
-Solid and dashed lines apply $p_{rescue}$ from the exact formula of  (\cref{survival_prob} in \cref{rescue_prob}); dashed lines apply $p_{rescue}$ from the approximate formula (\cref{rescue_prob_approx}), with good agreement.
+Solid and dashed lines apply $p_{rescue}$ from the exact formula of  (\cref{eq:survival_prob} in \cref{eq:rescue_prob}); dashed lines apply $p_{rescue}$ from the approximate formula (\cref{rescue_prob_approx}), with good agreement.
 Parameters: $N=10^4$; $\lambda_w=\lambda_a=\lambda_m=0.14$; (B) $\mu_w=0.17$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
@@
 Shown is the mean time for appearance of a resistance mutation the leads to evolutionary rescue \textbf{(left)} with ($u>0$) and \textbf{(right)} without ($u=0$ aneuploidy.
-Our inhomogeneous Poisson-process approximations (solid blue lines, right: \cref{meantime1}, left: \cref{meantimet2}) is in agreement with simulation results (orange markers with 95\% CI).
-Our 1st-order (dashed red lines, right: \cref{limitapprox}, left: \cref{limitapprox2}) and 2nd-order (green line, left: \cref{limitapprox3}) approximations work well when the initial tumor size is large (here $>10^8$ cells).
+Our inhomogeneous Poisson-process approximations (solid blue lines, right: \cref{eq:meantime1}, left: \cref{meantimet2}) is in agreement with simulation results (orange markers with 95\% CI).
+Our 1st-order (dashed red lines, right: \cref{eq:limitapprox}, left: \cref{limitapprox2}) and 2nd-order (green line, left: \cref{limitapprox3}) approximations work well when the initial tumor size is large (here $>10^8$ cells).
 Parameters: $\lambda_w=\lambda_a=\lambda_m=0.14$; $\mu_w=0.17$; (A) $\mu_a=0.145$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
@@
 \caption{\textbf{Ratio of evolutionary rescue time with and without aneuploidy.}
-The ratio of the mean time to appearance of a resistance mutation that leads to evolutionary rescue with ($u>0$) and without ($u=0$) aneuploidy for variable initial tumor sizes (\cref{eqMeanTimeRatioInitialPopulationSize}) when aneuploidy provides tolerance to the drug ($\Delta_a < 0$).
+The ratio of the mean time to appearance of a resistance mutation that leads to evolutionary rescue with ($u>0$) and without ($u=0$) aneuploidy for variable initial tumor sizes (\cref{eq:eqMeanTimeRatioInitialPopulationSize}) when aneuploidy provides tolerance to the drug ($\Delta_a < 0$).
 When the initial tumor size is not large ($<10^8$), aneuploidy can decrease the rescue time by 10-100-fold.
@@
 
-\appendix
+\newpage 
+\clearpage
+
 \begin{appendices}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Survival probability of a single lineage}
+\section*{Appendix A: Survival probability of a single lineage}\label{sec:appendix-surv-prob}
 
@@
 The complements $1-p_w$, $1-p_a$, and $1-p_m$ are the extinction probabilities, which satisfy each its respective equation,
-\begin{equation} \label{extinction_prob}
+\begin{equation} \label{eq:extinction_prob}
 \begin{aligned}
@@
 The survival probabilities are given by the smallest solution for each quadratic equation \citep{uecker2015adaptive}. Therefore we have
-\begin{equation}\label{survival_prob}
+\begin{equation}\label{eq:survival_prob}
 \begin{aligned}
@@
 We first assume that aneuploidy provides partial resistance to drug therapy, $\lambda_a>\mu_a$, and that this resistance is significant, $\left(\lambda_a-\mu_a-v\right)^2 > 4\lambda_a v p_m$.
-We thus rewrite \cref{survival_prob} as
+We thus rewrite \cref{eq:survival_prob} as
 \begin{align*}
@@
 we obtain the following approximation for the survival probability of a population initially consisting of a single wildtype cell,
-\begin{align}\label{survprobwapprox1}
+\begin{align} \label{eq:survprobwapprox1}
 p_w 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\paragraph{Second-order approximation.} % TODO consider moving to Appendix
+\paragraph{Second-order approximation.} 
 To improve our approximation, we can consider the second term of the Taylor series expansion,
@@
 We also assume that the decline are significant, $\left(\lambda_a-\mu_a-v\right)^2 > 4\lambda_a v p_m$.
-We rewrite \cref{survival_prob} as
+We rewrite \cref{eq:survival_prob} as
 \begin{equation}
@@
 Since $u,v\ll1$, the term in the root can be approximated using a 1st-order Taylor expansion. So, substituting the expressions for $p_a$ and $p_m$, we have
-\begin{equation}\label{survprobwinitial}
+\begin{equation} \label{eq:survprobwinitial}
 \begin{aligned}
@@
 We now assume that the growth rate of aneuploid cells is close to zero (either positive or negative), such that  $\left(\lambda_a-\mu_a-v\right)^2 < 4\lambda_avp_m$.
-We rewrite \cref{survival_prob} as
+We rewrite \cref{eq:survival_prob} as
 \begin{equation}
-p_a=\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}\left(1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\lambda_avp_m}\right)^{\frac12}}{2\lambda_a} .
+p_a = \frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}\left(1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\lambda_avp_m}\right)^{\frac12}}{2\lambda_a} .
 \end{equation}
@@
 \end{equation}
-Plugging this in \cref{survprobwinitial}, the survival probability of a population starting from one wildtype individual is
-\begin{equation}\label{scenario3}
+Plugging this in \cref{eq:survprobwinitial}, the survival probability of a population starting from one wildtype individual is
+\begin{equation}\label{eq:scenario3}
 \begin{aligned}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section*{Appendix B: Evolutionary rescue probability}\label{sec:appendix-rescue-prob}
+
+Substituting \cref{eq:survprobwapprox1,eq:survprobwinitial,eq:scenario3} into \cref{eq:rescue_prob}, the evolutionary rescue probability can be approximated by
+\begin{equation}\label{rescue_prob_approx}
+\begin{aligned}
+&\presc \approx \\
+  &\begin{cases}
+  1-\exp\left[\frac{N}{\Delta_w-u-v}\left(v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right)\right] ,&
+  4\lambda_avp_m>\left(\Delta_a-v\right)^2 ,\\
+   1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right] ,&
+   \Delta_a<0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2 ,\\
+   1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right] ,&
+   \Delta_a>0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2 .
+  \end{cases}
+\end{aligned}
+\end{equation}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section*{Appendix C: Evolutionary rescue time}
+
+We first calculate the expected time for the appearance of the first mutant that rescues the cell population.
+This can occur either through the evolutionary trajectory $wildtype \rightarrow mutant$ or through the trajectory $wildtype \rightarrow aneuploid \rightarrow mutant$.
+We start with the former.
+
+Assuming no aneuploidy ($u=0$), we define $T_1$ to be the time at which the first mutant cell appears that will avoid extinction and will therefore rescue the population.
+Note that if extinction occurs, that is the frequency of mutants after a very long time is zero, $m_{\infty}=0$, then it is implied that $T_1=\infty$, and vice versa if $T_1<\infty$ then $m_{\infty}>0$.
+
+The number of successful mutants generated until time $t$ can be approximated by an inhomogeneous Poisson process with rate $R\left(t\right) = u p_a w_t$,
+where $w_t=N\e^{\Delta_w t}$ is the number of wildtype cells at time $t$.
+Note that 
+\begin{equation}\label{eq:integralR}
+\int_0^{t}{R(z)\d z} = 
+u p_a N \frac{\exp[{\Delta_w t}]-1}{\Delta_w} \approx 
+u p_a N t,
+\end{equation}
+by integrating the exponential and because $\frac{\exp[\Delta_w t]-1}{\Delta_w}=\frac{1+\Delta_w t+O(t^2)-1}{\Delta_w}=t+O(t^2)$.
+The probability density function of $T_1$ is thus
+$R\left(t\right)\exp\left(-\int_0^{t}{R(z)\d z}\right)$. % TODO ref
+Therefore, the probability density function of the conditional random variable $(T_1 \mid T_1 < \infty)$ is
+$f_1(t) = \frac{R\left(t\right)\exp\left(-\int_0^{t}{R(z)\d z}\right)}{p_{rescue}}$. 
+\\
+
+We are interested in the mean conditional time, $\tau_1=\mathbb{E}\left[T_1 \mid T_1<\infty\right]$, which is given by
+\begin{equation}\label{eq:meantime1}
+\begin{aligned}
+\tau_1 =
+\int_{0}^{\infty}{t f_1(t) \d t} = 
+\frac{\int_{0}^{\infty}{tR(t)\exp\left(-\int_0^{t}{R(z)\d z}\right) \d t}}{p_{rescue}} = 
+\frac{\int_{0}^{\infty}{\exp\left(-\int_0^{t}{R(z)\d z}\right) \d t}}{p_{rescue}}
+\end{aligned}
+\end{equation}
+after applying integration by parts.
+Therefore, plugging \cref{eq:integralR,eq:rescue_prob} in \cref{eq:meantime1}, 
+\begin{align}\label{eq:limitapprox_appendix}
+\tau_1 = 
+%\frac{\int_0^\infty\exp\left(-\int_0^\tau R(z)\d z\right)\d t}{1-\left(1-p_w\right)^N} = 
+\frac{\int_0^\infty\e^{-uNp_a\frac{\e^{\Delta_w t}-1}{\Delta_w}} \d t}{1-\left(1-p_w\right)^N} \approx
+\frac{\int_{0}^{\infty}{\exp\left(-u p_a N t\right) \d t}}{1-\e^{-Np_w}} \approx \\
+\left(1+\e^{-Np_w}\right)\int_0^\infty\e^{-u p_a N t} \d t =
+\frac{1+\e^{-Np_w}}{u p_a N} ,
+\end{align}
+where we use the approximations 
+$\frac{\e^{\Delta_w\tau}-1}{\Delta_w}=\frac{1+\Delta_w\tau+O(\tau^2)-1}{\Delta_w}=\tau+O(\tau^2)$ and $(1-\e^{-Np_w})^{-1} \approx 1+\e^{-Np_w}$ and integrate the exponent.
+\Cref{MeanTimeGrowthAneuploidyPlot}B show the agreement between this approximating and simulation results for intermediate and large tumor sizes.
+\\
+
+When $Nu\gg1$ the aneuploid frequency dynamics is roughly deterministic and therefore can be approximated by 
+\begin{equation}
+a_t \approx \frac{Nu\e^{\Delta_wt}}{\Delta_w-\Delta_a}\left[1-\e^{\left(\Delta_w-\Delta_a\right)t}\right].
+\end{equation}
+As a result, when $N\gg1$ % TODO N>>1 ? or Nu >> 1?
+the number of successful mutants created by direct mutation and via aneuploidy can be approximated by inhomogeneous Poisson processes with the rates
+\begin{align}
+r_1\left(t\right)&=vp_m\int_0^ta_{z} \d z = \frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_wt}-1}{\Delta_w}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right),\\
+r_2\left(t\right)&=vp_m\int_0^tw_{z} \d z = vNp_m\frac{\e^{\Delta_w t}-1}{\Delta_w}.
+\end{align}
+For large initial population sizes we assume that the two processes are independent and as a result, they can be merged into a single Poisson process with rate $\left(r_1+r_2\right)\left(t\right)$.
+Consequently, the mean time to the appearance of the first rescue mutant is
+\begin{align}\label{meantimet2}
+\tau_2 = 
+\frac{\int_0^\infty\e^{-\left(r_1(t)+r_2(t)\right)} \d t}{1-\left(1-p_w\right)^N} = 
+\frac{\int_0^\infty\exp\left[-\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_w t}-1}{\Delta_w}-\frac{\e^{\Delta_a t}-1}{\Delta_a}\right)-vNp_m\frac{\e^{\Delta_w t}-1}{\Delta_w}\right] \d t}{1-\left(1-p_w\right)^N},
+\end{align}
+which we plot in \Cref{MeanTimeGrowthAneuploidyPlot}A as a function of the initial population size, $N$.
+
+We wish to obtain a simpler formula for $\tau_2$, similar to \cref{eq:limitapprox}.
+We thus have the following expansions,
+%\begin{align*}
+%\frac{\e^{\Delta_w\tau}-1}{\Delta_w}&=\frac{1+\Delta_w\tau+\frac{\Delta_w^2\tau^2}{2}+O(\tau^3)-1}{\Delta_w}=\tau+\frac{\Delta_w}{2}\tau^2+O(\tau^3) ,\\
+%\frac{\e^{\Delta_a\tau}-1}{\Delta_a}&=\frac{1+\Delta_a\tau+\frac{\Delta_a^2\tau^2}{2}+O(\tau^3)-1}{\Delta_a}=\tau+\frac{\Delta_w}{2}\tau^2+O(\tau^3) ,
+%\end{align*}
+%which allow us to write
+%\begin{equation*}
+%\frac{\e^{\Delta_w\tau}-1}{\Delta_w}-\frac{\e^{\Delta_a\tau}-1}{\Delta_a} \approx \frac{\left(\Delta_w-\Delta_a\right)\tau^2}{2}.
+%\end{equation*}
+%As a result, the integrand in \cref{meantimet2} can be written as:
+%\begin{align*}
+%&\exp\left[-\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_w\tau}-1}{\Delta_w}-\frac{\e^{\Delta_a\tau}-1}{\Delta_a}\right)-vNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}\right]\approx \\
+%& \exp\left(-uvNp_m\tau^2-vNp_m\tau\right) =
+%\exp\left(\frac{vNp_m}{2}\right)\exp\left[-\frac{uvNp_m}{2}\left(\tau+\frac{1}{u}\right)\right].
+%\end{align*}
+%Consequently, the mean conditional time $\tau_2$ is approximated by
+%\begin{align}\label{limitapprox2}
+%\tau_2 \approx
+%\frac{\left(1+\exp\left(-Np_w\right)\right) \exp\left(\frac{vNp_m}{2u}\right) \erfc\left(\sqrt{\frac{vNp_m}{2u}}\right)}{\sqrt{\frac{2uvNp_m}{\pi}}},
+%\end{align}
+%where $\erfc(z)$ is the complementary error function. 
+%
+%If we apply linear approximations,
+\begin{align*}
+\frac{\e^{\Delta_w t}-1}{\Delta_w}&=\frac{1+\Delta_w t+O(t^2)-1}{\Delta_w} = t+O(t^2),\\
+\frac{\e^{\Delta_a t}-1}{\Delta_a}&=\frac{1+\Delta_a t+O(t^2)-1}{\Delta_a} = t+O(t^2),
+\end{align*}
+which we use to derive a first-order approximation for $\tau_2$,
+\begin{align}\label{limitapprox3_appendix}
+\tau_2 \approx
+\left(1+\e^{-Np_w}\right)\int_0^\infty\e^{-u N p_m t} \d t = 
+\frac{\left(1+\e^{-Np_w}\right)}{uNp_m} ,
+\end{align}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \end{appendices}

### 2023-07-23 EvolutionaryRescue1.tex
minor edits
added qualitative description of the different regimes

@@
 \begin{equation*}
-m_t>\left\lfloor-\frac{3\log10}{\log\left(\frac{\mu_m}{\lambda_m}\right)}\right\rfloor+1,
+m_t > \left\lfloor\frac{3\log10}{\log\left(\lambda_m / \mu_m\right)}\right\rfloor + 1.
 \end{equation*}
@@
 \end{equation}
-Here, $\abs{\Delta_w}/v$ is the ratio of XXX.
+Here, $\abs{\Delta_w}/v$ is the ratio of the rate at which wild-type cells are decreasing in number and the rate at which they are mutating.
 
@@
     \frac{\abs{\Delta_a}}{v} \cdot \frac{\lambda_m}{\Delta_m} ,&
-  \Delta_a T^* \ll -1 ,\\ 
+  \Delta_a T^* \ll -1 \text{ (partially sensitive aneuploids)},\\ 
   %\left(\frac{\lambda_a}{v} \cdot \frac{\lambda_m}{\Delta_m}\right)^{1/2} ,&
   2\lambda_a T^* ,&
-  -1 \ll \Delta_a T^* \ll 1  ,\\ 
+  -1 \ll \Delta_a T^* \ll 1  \text{ (stationary aneuploids)},\\ 
   \frac{\lambda_a}{\Delta_a} ,&
-   \Delta_a T^* \gg 1 .
+   \Delta_a T^* \gg 1 \text{ (resistant aneuploids)}.
   \end{cases}
@@
 %The second case has a finer approximation by $N^*_a \approx \frac{\abs{\Delta_w}}{u} \cdot \frac{2\lambda_a}{\Delta_a + 1/T^*}$.
+The first line describes the case in which aneuploids are still effectively killed by the treatment, but not as quickly as the wild type. 
+In the second case, the aneuploids are sufficiently resistant that the size of each aneuploid lineage is expected to remain roughly constant. 
+In both of these first two cases, aneuploidy increases the probability of rescue by slowing or halting the decrease of the cancer population, allowing more opportunities for producing resistant mutants.
+In the third case, aneuploidy provides sufficient resistance for the aneuploid population to re-grow the tumor even without additional resistance mutations.
+Note that in this case there is no dependence on the parameters characterizing mutants or their production ($v$, $\lambda_m$, and $\Delta_m$).
 Comparing these approximations to results of stochastic evolutionary simulations, we find that the approximations perform very well (\Cref{rescue_prob_wt_growth,rescue_prob_an_growth,rescue_prob_N,rescue_prob_N}).
@@
 In all cases, the effect of aneuploidy increases (i.e., the threshold size ratio decreases) when the aneuploidy rate $u$ increases.
-Increasing the aneuploid growth rate $\Delta_a$ also leads to an increased role of aneuploidy, but only in the first and third case.
+Increasing the aneuploid growth rate $\Delta_a$ also leads to an increased role of aneuploidy, although the effect is minor when $|\Delta_a|$ is small compared to $T^*$.
 

### 2023-07-23 EvolutionaryRescue1.tex
edited references

@@
   \end{tabular}
-\caption{\textbf{Model parameters.} 
+\caption{\textbf{Model parameters.} NEED DIFFERENT REFS---THESE ARE MOSTLY THEORY PAPERS. CHECK BIONUMBERS?
 Aneuploid birth rate $\lambda_a$ is set to the same value as the wildtype and mutant birth rates, $\lambda_w$ and $\lambda_m$.
@@
 
-To analyze evolutionary rescue in this model, we use the framework of \emph{multitype branching processes} \citep{rybnikov2021fitness,harris1963theory}. 
+To analyze evolutionary rescue in this model, we use the framework of \emph{multitype branching processes} \citep{harris1963theory, weissman2009rate}. 
 This allows us to find explicit expressions for the \emph{survival probability}: the probability that a lineage descended from a single cell does not become extinct.

### 2023-07-24 EvolutionaryRescue1.tex
fixed tau1 errors; removed linear approx for tau1 and tau2; added main messages and figures outline in comments

@@
 \maketitle
+
+% MESSAGES
+% 1. aneuploidy increases the prob of rescue -- reduces the tumor threshold size for rescue
+% 2. aneuploidy changes the survival curve in a distinct way
+
+% FIGURES
+% Fig 1A: model graph chart
+% Fig 1B: Delta_w, Delta_a, Delta_m illustration
+
+% Fig 2: example simulations - w_t, a_t, m_t over time 
+% 	A u=0; B Delta_a << 0; C Delta_a ~ 0; D Delta_a >>0
+
+% Fig 3A: prob rescue vs N for various Delta_a; markup N*
+% Fig 3B: N* vs Delta_a
+% Fig 3C: N* vs u/v
+
+% Fig 4: survival curves
+% 	A u=0; B Delta_a << 0; C Delta_a ~ 0; D Delta_a >>0
 
@@
 \paragraph{Evolutionary rescue time.}
-In Appendix C we derive an approximation for $\tau_1$, the expected rescue time without aneuploidy ($u=0$), and $\tau_2$, the expected rescue time with aneuploidy ($u>0$),
-\begin{align}
-\tau_1 &\approx \frac{1+\e^{-Np_w}}{u N p_a} , \label{eq:limitapprox} ,\\
-\tau_2 &\approx \frac{1+\e^{-Np_w}}{u N p_m} , \label{eq:limitapprox3} .
-\end{align}
-\Cref{MeanTimeGrowthAneuploidyPlot} shows the agreement between this approximations and simulation results for intermediate and large tumor sizes.
-
-Thus, the ratio of the expected rescue time with and without aneuploidy ($u>0$ and $u=0$, respectively),
-\begin{align}\label{eq:eqMeanTimeRatioInitialPopulationSize}
-\frac{\tau_1}{\tau_2}  \approx \frac{p_a}{p_m} = 
-%&\frac{\lambda_m}{\lambda_a} \; \frac{\Delta_a-v+\sqrt{(\Delta_a - v)^2 + (1/T^*)^2}}{2 \Delta_m} \approx \\
-%&\frac{1}{2} \; \frac{\lambda_m}{\Delta_m} \left[ \left(\frac{\lambda_a}{\Delta_a}\right)^{-1} + \sqrt{\frac{\left(\Delta_a\right)^2 + (1/T^*)^2}{\lambda_a^2}}\right] = \\
-\frac{1}{2} \; \frac{\lambda_m}{\Delta_m} \left[ \left(\frac{\lambda_a}{\Delta_a}\right)^{-1} + \sqrt{\left(\frac{\lambda_a}{\Delta_a}\right)^{-2} + 4v \left(\lambda_a\frac{\lambda_m}{\Delta_m}\right)^{-1}}\right] .
-\end{align}
-
-\Cref{MeanTimeRatioInitialPopulationSize} shows how this ratio, $\tau_1/\tau_2$, changes as a function of the initial tumor size, $N$, for different aneuploid growth rates, $\Delta_a$.
-
-Importantly, in \Cref{MeanTimeGrowthAneuploidyPlot} we observe that in initially large tumors, direct mutation drives evolutionary rescue while aneuploidy plays a role for intermediate sized tumors.
-% TODO where do you see this..?
-This is consistent with \Cref{rescue_ratio}B,D, where aneuploidy only increases the probability of evolutionary rescue in small and intermediate tumors ($N<10^7$).
+In Appendix C we derive an approximation for $\tau_1$, the expected rescue time without aneuploidy ($u=0$), and $\tau_2$, the expected rescue time with aneuploidy ($u>0$).
+\Cref{MeanTimeGrowthAneuploidyPlot} shows the agreement between these approximations and simulation results for intermediate and large tumor sizes.
 
@@
 TODO
+
+% ? Distribution of recurrence time -- with and without aneuploidy
+
+% Kaplan-Meier curves -- x: time; y: P(# cells >= N) -- once for u=0; u>0 with the three scenarios
 
@@
 \section*{Appendix C: Evolutionary rescue time}
+% TODO remove the linear approximation, it is not good.
 
@@
 
-The number of successful mutants generated until time $t$ can be approximated by an inhomogeneous Poisson process with rate $R\left(t\right) = u p_a w_t$,
+The number of successful mutants generated until time $t$ can be approximated by an inhomogeneous Poisson process with rate $R\left(t\right) = v p_m w_t$,
 where $w_t=N\e^{\Delta_w t}$ is the number of wildtype cells at time $t$.
@@
 \int_0^{t}{R(z)\d z} = 
-u p_a N \frac{\exp[{\Delta_w t}]-1}{\Delta_w} \approx 
-u p_a N t,
+v p_m N \frac{\exp[{\Delta_w t}]-1}{\Delta_w} \approx 
+v p_m N t,
 \end{equation}
@@
 Therefore, the probability density function of the conditional random variable $(T_1 \mid T_1 < \infty)$ is
-$f_1(t) = \frac{R\left(t\right)\exp\left(-\int_0^{t}{R(z)\d z}\right)}{p_{rescue}}$. 
+$f_1(t) = \frac{R\left(t\right)\exp\left(-\int_0^{t}{R(z)\d z}\right)}{\presc}$. 
 \\
@@
 %\frac{\int_0^\infty\exp\left(-\int_0^\tau R(z)\d z\right)\d t}{1-\left(1-p_w\right)^N} = 
-\frac{\int_0^\infty\e^{-uNp_a\frac{\e^{\Delta_w t}-1}{\Delta_w}} \d t}{1-\left(1-p_w\right)^N} \approx
-\frac{\int_{0}^{\infty}{\exp\left(-u p_a N t\right) \d t}}{1-\e^{-Np_w}} \approx \\
-\left(1+\e^{-Np_w}\right)\int_0^\infty\e^{-u p_a N t} \d t =
-\frac{1+\e^{-Np_w}}{u p_a N} ,
+\frac{\int_0^\infty\e^{-v N p_m\frac{\e^{\Delta_w t}-1}{\Delta_w}} \d t}{1-\left(1-p_w\right)^N} \approx
+\frac{\int_{0}^{\infty}{\exp\left(-v p_m N t\right) \d t}}{1-\e^{-N p_w}} \approx \\
+\left(1+\e^{-N p_w}\right)\int_0^\infty\e^{-v p_m N t} \d t =
+\frac{1+\e^{-N p_w}}{v p_m N} ,
 \end{align}
 where we use the approximations 
-$\frac{\e^{\Delta_w\tau}-1}{\Delta_w}=\frac{1+\Delta_w\tau+O(\tau^2)-1}{\Delta_w}=\tau+O(\tau^2)$ and $(1-\e^{-Np_w})^{-1} \approx 1+\e^{-Np_w}$ and integrate the exponent.
+$\frac{\e^{\Delta_w\tau}-1}{\Delta_w}=\frac{1+\Delta_w\tau+O(\tau^2)-1}{\Delta_w}=\tau+O(\tau^2)$ and $(1-\e^{-N p_w})^{-1} \approx 1+\e^{-N p_w}$ and integrate the exponent.
 \Cref{MeanTimeGrowthAneuploidyPlot}B show the agreement between this approximating and simulation results for intermediate and large tumor sizes.

### 2023-07-25 EvolutionaryRescue1.tex
new figures

@@
 %%%%%%%%%%%
+% Fig 3A: prob rescue vs N for various Delta_a; markup N*
+% Fig 3B: N* vs Delta_a
+% Fig 3C: N* vs u/v
 
@@
 \begin{subfigure}{0.5\textwidth}
-\includegraphics[width=1\textwidth]{Figures/CombinedSubplot.pdf}
+A\\
+\includegraphics[width=1\textwidth]{Figures/ProbvNPlot.pdf}
+\end{subfigure}
+\\
+\begin{subfigure}{0.5\textwidth}
+B\\
+\includegraphics[width=1\textwidth]{Figures/ThresholdPopulationSizePlot.pdf}
 \end{subfigure}
 \begin{subfigure}{0.5\textwidth}
-\includegraphics[width=1\textwidth]{Figures/CombinedSubplotLargePop.pdf}
+C\\
+\includegraphics[width=1\textwidth]{Figures/ThresholdPopulationSizeVersusRatioPlot.pdf}
 \end{subfigure}
 \caption{
-\textbf{Evolutionary rescue probability with partially resistant or tolerant aneuploid cells.}
-Rescue probability is very high when aneuploidy provides partial resistance ($\lambda_a=0.01$), in an initially small tumor (\textbf{A}, $N=10^4$) and even more so in an initially large tumor (\textbf{B}, $N=10^8$).
-When aneuploidy provides tolerance (\textbf{C}, $N=10^4$; \textbf{D}, $N=10^8$), the rescue probability is much lower. 
-In both scenarios, rescue probability increase with both the wildtype growth rate (x-axis) and the aneuploidy growth rate (colors).
-Markers represent simulation results with 95\% CI; solid and dashed lines for the exact formula (\cref{eq:survival_prob} in \cref{eq:rescue_prob}); dashed lines for the approximate formula (\cref{rescue_prob_approx}), demonstrating that they all agree.
-Parameters: division rate $\lambda_w=\lambda_a=\lambda_m=0.14$ (so that growth rate changes due to variable death rate); mutant death rate $\mu_m=0.13$ (so that mutant growth rate $\Delta_m=0.01$); aneuploidy rate $u=10^{-2}$; mutation rate $v=10^{-7}$.
+\textbf{Aneuploidy facilitates evolutionary rescue of cancer under drug treatment.}
+\textbf{(A)} The probability of evolutionary rescue (i.e.
+the probability that the population does not go to extinction), $\presc$, as a function of the initial tumor size, $N$.
+Dashed line shows the threshold tumor size, $N^*$, above which the probability is very high.
+\textbf{(B)} The threshold tumor size $N_a^*$ as a function of the aneuploid growth rate $\Delta_a$.
+The dashed horizontal line shows $N^*_m$, the threshold tumor size without aneuploidy ($u=0$).
+When aneuploid growth rate is close to or higher than zero, aneuploidy decreases the threshold tumor size, thereby facilitating evolutionary rescue.
+\textbf{(C)} The threshold tumor size $N_a^*$ as a function of the ratio of aneuploidy and mutation rates, $u/v$.
+The dashed horizontal line shows $N^*_m$, the threshold tumor size without aneuploidy ($u=0$).
+When the aneuploidy rate is much higher than the mutation rate, aneuploidy decreases the threshold tumor size, thereby facilitating evolutionary rescue.
 }
-\label{rescue_prob_wt_growth}
+\label{fig:LABEL}
 \end{figure}
@@
 
-\begin{figure}
-% small population size
-% switch from tolerance to non-growing
-% TODO why no switch in panel B?
-% TODO more simulations for 0 > Delta_a > Delta^*_a
-% TODO simulations in A?
-\begin{subfigure}{0.5\textwidth}
-\includegraphics[width=1\textwidth]{Figures/P_est_divergence.pdf}
-\end{subfigure}
-\begin{subfigure}{0.5\textwidth}
-\includegraphics[width=1\textwidth]{Figures/P_est_divergenceLargePopulation.pdf}
-\end{subfigure}
-\caption{\textbf{Evolutionary rescue probability with tolerant or non-growing aneuploid cells.}
-Rescue probability grows with the aneuploid growth rate $\Delta_a$ (x-axis), and is much higher in an initially large tumor than in a small one (\textbf{A}: $N=10^4$; \textbf{B}: $N=10^8$). 
-Markers for simulation results with 95\% CI; 
-solid lines for the exact formula (\cref{eq:survival_prob} in \cref{eq:rescue_prob});
-dashed lines for the approximate formula (\cref{rescue_prob_approx}).
-The approximation agrees with the simulation and exact solution when the initial tumor size is large (panel B).
-When the tumor size is small (panel A), we switch between the approximation for tolerant and for non-growing aneuploid cells; the switch occurs at $1/T^*$. % TODO ref to eq
-Parameters: $\lambda_w=\lambda_a=\lambda_m=0.14$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
-}
-\label{rescue_prob_an_growth}
-\end{figure}
-
-%%%%%%%%%%%
-
-\begin{figure}
-\begin{subfigure}{0.5\textwidth}
-% non-growing slightly declining, compare 1st order approximation to exact and simulation
-\includegraphics[width=1\textwidth]{Figures/DeleteriousTauLeapPlot.pdf}
-\end{subfigure}
-\begin{subfigure}{0.5\textwidth}
-% density-dependent simulations, barely growing (or maybe partial resistance), with exact formula
-\includegraphics[width=1\textwidth]{Figures/SurvPlotNDataLogisticK.pdf}
-\end{subfigure}
-\caption{\textbf{Evolutionary rescue probability for variable initial tumor size.}
-\textbf{(A)} Comparison of simulation results (markers with 95\% CI, too small to appear with $10^5$ simulations per marker), the exact formula (blue line, \cref{eq:survival_prob} in \cref{eq:rescue_prob}) and the approximate formula (orange line, \cref{rescue_prob_approx}).
-\textbf{(B)} Comparison of results of simulations  with density-dependent growth (markers with with 95\% CI) and the exact formula (blue line, \cref{eq:survival_prob} in \cref{eq:rescue_prob}) with maximum carrying capacity $K=10^9$.
-Parameters: $\lambda_w=\lambda_a=\lambda_m=0.14$; $\mu_w=0.17$; (A) $\mu_a=0.15$, (B) $\mu_a=0.135$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
-}
-\label{rescue_prob_N}
-\end{figure}
-
-%%%%%%%%%%%
-
-\begin{figure}
-% slightly decreasing (need to check)
- \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/FractionPlot.pdf}
-\caption{\textbf{Effect of standing variation on evolutionary rescue.}
-In aneuploid cells already exist in the population at the onset of drug therapy as standing genetic variation, then evolutionary rescue is more likely... 
-Parameters: $\lambda_w=\lambda_a=\lambda_m=0.14$; $\mu_w=0.17$; $\mu_a=0.145$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
-% TODO think about this figure...
-% TODO maybe better to show p_total / p_denovo as function of u or N.
-}
-\label{FractionPlot}
-\end{figure}
-
-%%%%%%%%%%%
-
-\begin{figure}
-\begin{subfigure}{0.5\textwidth}
-\includegraphics[width=1\textwidth]{Figures/RatioEvolRescue.pdf}
-\end{subfigure}
-\begin{subfigure}{0.5\textwidth}
-\includegraphics[width=1\textwidth]{Figures/RatioEvolRescuePopulationSize.pdf}
-\end{subfigure}
-\caption{\textbf{Effect of aneuploidy on evolutionary rescue.}
-The ratio of rescue probability with and without aneuploid ($H$, \cref{ratiorescue}) increases with the aneuploid growth rate (colors) and decreases with the wildtype growth rates and initial tumor size (x-axes), except for large tumors where where the ratio converges to unity.
-\textbf{(A, B)} Aneuploidy provides partial resistance.
-\textbf{(C, D)} Aneuploidy provides tolerance.  
-Solid and dashed lines apply $p_{rescue}$ from the exact formula of  (\cref{eq:survival_prob} in \cref{eq:rescue_prob}); dashed lines apply $p_{rescue}$ from the approximate formula (\cref{rescue_prob_approx}), with good agreement.
-Parameters: $N=10^4$; $\lambda_w=\lambda_a=\lambda_m=0.14$; (B) $\mu_w=0.17$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
-% TODO start y at 1
-}
-\label{rescue_ratio}
-\end{figure}
-
-%%%%%%%%%%%
-
-\begin{figure}
-\begin{subfigure}{0.5\textwidth}
-\includegraphics[width=1\textwidth]{Figures/MeanTimeDeleteriousPlot.pdf}
-\end{subfigure}
-\begin{subfigure}{0.5\textwidth}
-\includegraphics[width=1\textwidth]{Figures/MeanTimeGrowthMutantDirectPlot.pdf}
-\end{subfigure}
-\caption{\textbf{Evolutionary rescue time.}
-Shown is the mean time for appearance of a resistance mutation the leads to evolutionary rescue \textbf{(left)} with ($u>0$) and \textbf{(right)} without ($u=0$ aneuploidy.
-Our inhomogeneous Poisson-process approximations (solid blue lines, right: \cref{eq:meantime1}, left: \cref{meantimet2}) is in agreement with simulation results (orange markers with 95\% CI).
-Our 1st-order (dashed red lines, right: \cref{eq:limitapprox}, left: \cref{limitapprox2}) and 2nd-order (green line, left: \cref{limitapprox3}) approximations work well when the initial tumor size is large (here $>10^8$ cells).
-Parameters: $\lambda_w=\lambda_a=\lambda_m=0.14$; $\mu_w=0.17$; (A) $\mu_a=0.145$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
-}
-\label{MeanTimeGrowthAneuploidyPlot} 
-\end{figure}
-
-%%%%%%%%%%%
-
-\begin{figure}
- \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/MeanTimeRatioInitialPopulationSize.pdf}
-\caption{\textbf{Ratio of evolutionary rescue time with and without aneuploidy.}
-The ratio of the mean time to appearance of a resistance mutation that leads to evolutionary rescue with ($u>0$) and without ($u=0$) aneuploidy for variable initial tumor sizes (\cref{eq:eqMeanTimeRatioInitialPopulationSize}) when aneuploidy provides tolerance to the drug ($\Delta_a < 0$).
+%%%%%%%%%%%%
+%
+%\begin{figure}
+%\begin{subfigure}{0.5\textwidth}
+%\includegraphics[width=1\textwidth]{Figures/FILENAME.pdf}
+%\end{subfigure}
+%\begin{subfigure}{0.5\textwidth}
+%\includegraphics[width=1\textwidth]{Figures/FILENAME.pdf}
+%\end{subfigure}
+%\caption{
+%\textbf{TITLE}
+%CAPTION
+%}
+%\label{fig:LABEL}
+%\end{figure}
+%%%%%%%%%%%%
+
+%%%%%%%%%%%%
+%
+%\begin{figure}
+%\begin{subfigure}{0.5\textwidth}
+%\includegraphics[width=1\textwidth]{Figures/CombinedSubplot.pdf}
+%\end{subfigure}
+%\begin{subfigure}{0.5\textwidth}
+%\includegraphics[width=1\textwidth]{Figures/CombinedSubplotLargePop.pdf}
+%\end{subfigure}
+%\caption{
+%\textbf{Evolutionary rescue probability with partially resistant or tolerant aneuploid cells.}
+%Rescue probability is very high when aneuploidy provides partial resistance ($\lambda_a=0.01$), in an initially small tumor (\textbf{A}, $N=10^4$) and even more so in an initially large tumor (\textbf{B}, $N=10^8$).
+%When aneuploidy provides tolerance (\textbf{C}, $N=10^4$; \textbf{D}, $N=10^8$), the rescue probability is much lower. 
+%In both scenarios, rescue probability increase with both the wildtype growth rate (x-axis) and the aneuploidy growth rate (colors).
+%Markers represent simulation results with 95\% CI; solid and dashed lines for the exact formula (\cref{eq:survival_prob} in \cref{eq:rescue_prob}); dashed lines for the approximate formula (\cref{rescue_prob_approx}), demonstrating that they all agree.
+%Parameters: division rate $\lambda_w=\lambda_a=\lambda_m=0.14$ (so that growth rate changes due to variable death rate); mutant death rate $\mu_m=0.13$ (so that mutant growth rate $\Delta_m=0.01$); aneuploidy rate $u=10^{-2}$; mutation rate $v=10^{-7}$.
+%}
+%\label{rescue_prob_wt_growth}
+%\end{figure}
+%%%%%%%%%%%%
+%
+%\begin{figure}
+%% small population size
+%% switch from tolerance to non-growing
+%% TODO why no switch in panel B?
+%% TODO more simulations for 0 > Delta_a > Delta^*_a
+%% TODO simulations in A?
+%\begin{subfigure}{0.5\textwidth}
+%\includegraphics[width=1\textwidth]{Figures/P_est_divergence.pdf}
+%\end{subfigure}
+%\begin{subfigure}{0.5\textwidth}
+%\includegraphics[width=1\textwidth]{Figures/P_est_divergenceLargePopulation.pdf}
+%\end{subfigure}
+%\caption{\textbf{Evolutionary rescue probability with tolerant or non-growing aneuploid cells.}
+%Rescue probability grows with the aneuploid growth rate $\Delta_a$ (x-axis), and is much higher in an initially large tumor than in a small one (\textbf{A}: $N=10^4$; \textbf{B}: $N=10^8$). 
+%Markers for simulation results with 95\% CI; 
+%solid lines for the exact formula (\cref{eq:survival_prob} in \cref{eq:rescue_prob});
+%dashed lines for the approximate formula (\cref{rescue_prob_approx}).
+%The approximation agrees with the simulation and exact solution when the initial tumor size is large (panel B).
+%When the tumor size is small (panel A), we switch between the approximation for tolerant and for non-growing aneuploid cells; the switch occurs at $1/T^*$. % TODO ref to eq
+%Parameters: $\lambda_w=\lambda_a=\lambda_m=0.14$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
+%}
+%\label{rescue_prob_an_growth}
+%\end{figure}
+%
+%%%%%%%%%%%%
+%
+%\begin{figure}
+%\begin{subfigure}{0.5\textwidth}
+%% non-growing slightly declining, compare 1st order approximation to exact and simulation
+%\includegraphics[width=1\textwidth]{Figures/DeleteriousTauLeapPlot.pdf}
+%\end{subfigure}
+%\begin{subfigure}{0.5\textwidth}
+%% density-dependent simulations, barely growing (or maybe partial resistance), with exact formula
+%\includegraphics[width=1\textwidth]{Figures/SurvPlotNDataLogisticK.pdf}
+%\end{subfigure}
+%\caption{\textbf{Evolutionary rescue probability for variable initial tumor size.}
+%\textbf{(A)} Comparison of simulation results (markers with 95\% CI, too small to appear with $10^5$ simulations per marker), the exact formula (blue line, \cref{eq:survival_prob} in \cref{eq:rescue_prob}) and the approximate formula (orange line, \cref{rescue_prob_approx}).
+%\textbf{(B)} Comparison of results of simulations  with density-dependent growth (markers with with 95\% CI) and the exact formula (blue line, \cref{eq:survival_prob} in \cref{eq:rescue_prob}) with maximum carrying capacity $K=10^9$.
+%Parameters: $\lambda_w=\lambda_a=\lambda_m=0.14$; $\mu_w=0.17$; (A) $\mu_a=0.15$, (B) $\mu_a=0.135$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
+%}
+%\label{rescue_prob_N}
+%\end{figure}
+%
+%%%%%%%%%%%%
+%
+%\begin{figure}
+%% slightly decreasing (need to check)
+% \vspace*{1\baselineskip}
+%\includegraphics[width=1\textwidth]{Figures/FractionPlot.pdf}
+%\caption{\textbf{Effect of standing variation on evolutionary rescue.}
+%In aneuploid cells already exist in the population at the onset of drug therapy as standing genetic variation, then evolutionary rescue is more likely... 
+%Parameters: $\lambda_w=\lambda_a=\lambda_m=0.14$; $\mu_w=0.17$; $\mu_a=0.145$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
+%% TODO think about this figure...
+%% TODO maybe better to show p_total / p_denovo as function of u or N.
+%}
+%\label{FractionPlot}
+%\end{figure}
+%
+%%%%%%%%%%%%
+%
+%\begin{figure}
+%\begin{subfigure}{0.5\textwidth}
+%\includegraphics[width=1\textwidth]{Figures/RatioEvolRescue.pdf}
+%\end{subfigure}
+%\begin{subfigure}{0.5\textwidth}
+%\includegraphics[width=1\textwidth]{Figures/RatioEvolRescuePopulationSize.pdf}
+%\end{subfigure}
+%\caption{\textbf{Effect of aneuploidy on evolutionary rescue.}
+%The ratio of rescue probability with and without aneuploid ($H$, \cref{ratiorescue}) increases with the aneuploid growth rate (colors) and decreases with the wildtype growth rates and initial tumor size (x-axes), except for large tumors where where the ratio converges to unity.
+%\textbf{(A, B)} Aneuploidy provides partial resistance.
+%\textbf{(C, D)} Aneuploidy provides tolerance.  
+%Solid and dashed lines apply $p_{rescue}$ from the exact formula of  (\cref{eq:survival_prob} in \cref{eq:rescue_prob}); dashed lines apply $p_{rescue}$ from the approximate formula (\cref{rescue_prob_approx}), with good agreement.
+%Parameters: $N=10^4$; $\lambda_w=\lambda_a=\lambda_m=0.14$; (B) $\mu_w=0.17$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
+%% TODO start y at 1
+%}
+%\label{rescue_ratio}
+%\end{figure}
+%
+%%%%%%%%%%%%
+%
+%\begin{figure}
+%\begin{subfigure}{0.5\textwidth}
+%\includegraphics[width=1\textwidth]{Figures/MeanTimeDeleteriousPlot.pdf}
+%\end{subfigure}
+%\begin{subfigure}{0.5\textwidth}
+%\includegraphics[width=1\textwidth]{Figures/MeanTimeGrowthMutantDirectPlot.pdf}
+%\end{subfigure}
+%\caption{\textbf{Evolutionary rescue time.}
+%Shown is the mean time for appearance of a resistance mutation the leads to evolutionary rescue \textbf{(left)} with ($u>0$) and \textbf{(right)} without ($u=0$ aneuploidy.
+%Our inhomogeneous Poisson-process approximations (solid blue lines, right: \cref{eq:meantime1}, left: \cref{meantimet2}) is in agreement with simulation results (orange markers with 95\% CI).
+%Our 1st-order (dashed red lines, right: \cref{eq:limitapprox}, left: \cref{limitapprox2}) and 2nd-order (green line, left: \cref{limitapprox3}) approximations work well when the initial tumor size is large (here $>10^8$ cells).
+%Parameters: $\lambda_w=\lambda_a=\lambda_m=0.14$; $\mu_w=0.17$; (A) $\mu_a=0.145$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
+%}
+%\label{MeanTimeGrowthAneuploidyPlot} 
+%\end{figure}
+%
+%%%%%%%%%%%%
+%
+%\begin{figure}
+% \vspace*{1\baselineskip}
+%\includegraphics[width=1\textwidth]{Figures/MeanTimeRatioInitialPopulationSize.pdf}
+%\caption{\textbf{Ratio of evolutionary rescue time with and without aneuploidy.}
+%The ratio of the mean time to appearance of a resistance mutation that leads to evolutionary rescue with ($u>0$) and without ($u=0$) aneuploidy for variable initial tumor sizes (\cref{eq:eqMeanTimeRatioInitialPopulationSize}) when aneuploidy provides tolerance to the drug ($\Delta_a < 0$).
 When the initial tumor size is not large ($<10^8$), aneuploidy can decrease the rescue time by 10-100-fold.
-% TODO parameter values 
-}
-\label{MeanTimeRatioInitialPopulationSize}
-\end{figure}
-
-%%%%%%%%%%%
-
-\begin{figure}
- \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/ThresholdPopulationSize.pdf}
-\caption{\textbf{TODO.}
-TODO
-}
-\label{MinTumorSize}
-\end{figure}
+%% TODO parameter values 
+%}
+%\label{MeanTimeRatioInitialPopulationSize}
+%\end{figure}
+%
+%%%%%%%%%%%%
+%
+%\begin{figure}
+% \vspace*{1\baselineskip}
+%\includegraphics[width=1\textwidth]{Figures/ThresholdPopulationSize.pdf}
+%\caption{\textbf{TODO.}
+%TODO
+%}
+%\label{MinTumorSize}
+%\end{figure}
 

### 2023-07-25 EvolutionaryRescue1.tex
beneficial...

@@
     $\mu_m$ & Mutant death rate& 0.13 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
-    $u$ & Missegregation rate& $10^{-3}-10^{-2}$ & 1$\slash$cell division  & \citet{nowak2004evolutionary,bakker2023predicting} \\ \hline
-    $v$ & Mutation rate& $10^{-7}-10^{-9}$ &  1$\slash$gene$\slash$cell division  & \citet{nowak2004evolutionary} \\ 
+    $u$ & Beneficial aneuploidy rate& $10^{-3}-10^{-2}$ & 1 / chromosome / cell division  & \citet{nowak2004evolutionary,bakker2023predicting} \\ \hline
+    $v$ & Beneficial mutation rate& $10^{-7}-10^{-9}$ &  1 / gene / cell division  & \citet{nowak2004evolutionary} \\ 
     % TODO citing Nowak 2004 for the rates is strange because it is a theory paper - cite the original empirical papers
+    % TODO resistance and aneuploidy could increase lambda, not just reduce mu.
     \hline

### 2023-09-05 EvolutionaryRescue1.tex
Remus

--- 2023-07-25 EvolutionaryRescue1.tex
@@
 %\usepackage{end float}
+ \usepackage{subcaption}
  \usepackage{cleveref}
- \usepackage{subcaption}
 
@@
 \begin{abstract}
-%Evolutionary rescue is the process by which a population is able to survive a sudden environmental change which initially causes the population to decline towards extinction.
+Evolutionary rescue is the process by which a population is able to survive a sudden environmental change which initially causes the population to decline towards extinction.
 A prime example of evolutionary rescue is the ability of cancer to survive being exposed to various treatments.
@@
 We propose to model evolutionary rescue with the help of multi-type branching processes to obtain the probability that cancer will survive.
-Additionally, we will utilize large genomic datasets to asses the effects of aneuploidy on the probability of evolutionary rescue.
 \end{abstract}
@@
 
-% TODO two papers that show aneuploidy provides resistance
+% TODO two papers that show aneuploidy provides resistance Ok
 % Lukow, Devon A., Erin L.
@@
 \paragraph{Aneuploidy in cancer.} Chromosomal instability (CIN) is the mitotic process in which cells suffer from chromosome mis-segregation that leads to aneuploidy, where cells are characterized by structural changes of the chromosomes and copy number alterations \citep{schukken2018cin}.
-Interestingly, aberrations in chromosome copy number have been shown to allow cancer cells to survive under stressful conditions such as drug therapy.
+Interestingly, aberrations in chromosome copy number have been shown to allow cancer cells to survive under stressful conditions such as drug therapy \citep{lukow2021chromosomal,rutledge2016selective}.
 Indeed, cancer cells are often likely to be aneuploid, and aneuploidy is associated with poor patient outcomes \citep{ben2020context}.
@@
 \mu_w' &= \mu_w,\\
-\lambda_a' &= C_1+\left(\lambda_a-\mu_a\right)\left(1-\frac{w+a+m}{K}\right),\\ 
-\mu_a' &= C_1,\\
-\lambda_m' &= C_2+\left(\lambda_m-\mu_m\right)\left(1-\frac{w+a+m}{K}\right),\\ 
-\mu_m' &= C_2,
+\lambda_a' &= \lambda_a,\\ 
+\mu_a' &= \mu_a + \lambda_a\frac{w+a+m}{K},\\
+\lambda_m' &= \lambda_m,\\ 
+\mu_m' &= \mu_m + \lambda_m\frac{w+a+m}{K}.
 \end{align*}
-where $C_1, C_2>0$ are constants and $K$ is the maximum carrying capcity. % TODO what constants? see Remus version.
+where $K$ is the maximum carrying capcity. % TODO what constants? see Remus version. Ok
 
@@
 Note that in this case there is no dependence on the parameters characterizing mutants or their production ($v$, $\lambda_m$, and $\Delta_m$).
-Comparing these approximations to results of stochastic evolutionary simulations, we find that the approximations perform very well (\Cref{rescue_prob_wt_growth,rescue_prob_an_growth,rescue_prob_N,rescue_prob_N}).
+Comparing these approximations to results of stochastic evolutionary simulations, we find that the approximations perform very well (\Cref{rescue_prob}).%rescue_prob_wt_growth,rescue_prob_an_growth,rescue_prob_N,rescue_prob_N}).
 
@@
 In the above we assumed that at the onset of drug treatment, the initial tumor consisted entirely of wildtype cells that are drug sensitive.
-However, aneuploid cells are likely generated even before onset of treatment at some rate $\tilde{u} \le u$ (because the treatment itself may promote generation of aneuploid cells REF), which are likely to have a deleterious effect in the absence of the drug, $s$ (REF).
-% TODO refs
+However, aneuploid cells are likely generated even before onset of treatment at some rate $\tilde{u} \le u$ (because the treatment itself may promote generation of aneuploid cells \citep{wang2019molecular,mason2017functional}), which are likely to have a deleterious effect in the absence of the drug, $s$ \citep{replogle2020aneuploidy,giam2015aneuploidy}.
+% TODO refs OK
 But if the number of cells in the tumor $N$ is large (as expected if the tumor is to be treated with a drug), there may already be a fraction $f \approx \tilde{u}/s$ of aneuploid cells in the population.
@@
 \begin{equation}
-\frac{\tilde{N}^*_{a}}{N^*_{a}} = \frac{u}{\tilde{u}} \; \frac{s}{\abs{\Delta_w}} .
+\frac{\tilde{N}^*_{a}}{N^*_{a}} = \frac{u}{\tilde{u}} \; \frac{s}{\abs{\Delta_w}}.
 \end{equation}
@@
 
-TODO
-
+In Appendix D we approximate the mean time for the population of mutant cancer cells to reach the initial population size $N$.
+\Cref{MeanTimeGrowthAneuploidyPlot} shows the agreement between these approximations and simulation results for intermediate and large tumor sizes.
+
+\paragraph{Distribution of recurrence time -- with and without aneuploidy.}
+
+In Appendix E we derive the probability that by time $t$ a succesful mutant has been generated.
+\Cref{ReboundProbability} show the agreement between our formula and simulation results for the case when aneuploidy is present and when it is absent.
 % ? Distribution of recurrence time -- with and without aneuploidy
@@
 
-We find that aneuploidy can have a significant effect on evolutionary rescue (\Cref{rescue_ratio,MeanTimeRatioInitialPopulationSize}).
+We find that aneuploidy can have a significant effect on evolutionary rescue (\Cref{rescue_prob}).
 For example, when aneuploidy cells are "barely-resistant" (they grow at a very low rate, $\Delta_a=10^{-3}$) the probability of evolutionary rescue is 1,000-fold higher with aneuploidy than without it (for parameters previously described in cancer, see \Cref{table1}).
@@
 
-We hypothesized that presence of \emph{standing variation} % TODO ref
+We hypothesized that presence of \emph{standing variation} % TODO ref <- why ref?
 --the existence of a subpopulation of aneuploid cancer cells before therapy begins--can facilitate evolutionary rescue by reducing the waiting time for the appearance of aneuploid cells.
@@
 However, this may not be the case, as cancer cells compete for resources (e.g., blood supply).
-Nevertheless, we find that when the carrying capacity if large % TODO HOW LARGE?
+Nevertheless, we find that when the carrying capacity is large ($K\sim10^7$) % TODO HOW LARGE?
+ok
 our approximation for the probability of evolutionary rescue agrees with results of stochastic simulations with density-dependent growth  \Cref{rescue_prob_N}).
-Future work may focus on scenarios with small carrying capacity by analyzing density-dependent branching processes. % TODO ref?
+Future work may focus on scenarios with small carrying capacity by analyzing density-dependent branching processes~\citep{klebaner1997population}. % TODO ref? OK
 
@@
 For example, to study the effects of initial tumor size on the probability of evolutionary rescue, a large culture mass can be propagated from a single cancer cell in permissive conditions and then diluted to a  range of starting tumor sizes.
-Afterwards, these tumors may be exposed to anti-cancer drugs that induces aneuploidy % TODO ref
-or to saline solution for control. 
+Afterwards, these tumors may be exposed to anti-cancer drugs that induces aneuploidy % TODO ref OK
+or to saline solution for control~\citep{ippolito2021gene}. 
 Cell density can then be measured and compared to the predictions of our model. % TODO not clear what is to be compared... what does our model predict...?
 
-% TODO discussion of mean time?
-
-% TODO summary
+We observe that the presence of aneuploidy has the effect of shortening the time necessary for a mutant to appear which will rescue the tumor (\Cref{MeanTimeGrowthAneuploidyPlot}).
+However, this effect is only true for small tumors ($N<10^5$) as direct mutation is the main mechanism for generation of successful mutants for large and intermediate population sizes.
+% TODO discussion of mean time? ok
+
+Our study quantitatively confirms that aneuploidy plays an important role in tumors overcoming exposure to chemotherapeutic drugs when the tumor size is small or intermediate.
+Very large tumors can escape anti-cancer drugs through direct mutation while smaller ones are able to obtain the beneficial mutaion through an aneuploid intermediary (\Cref{rescue_prob}).
+% TODO summary ok
 
@@
     $N$ & Initial tumor size & $10^7-10^9$ & cells  & \citet{del2009does} \\ \hline
-    $\lambda_w$ & Wildtype division rate& 0.14 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
-    $\mu_w$ & Wildtype death rate& 0.17 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
+    $\lambda_w$ & Wildtype division rate& 0.14 & 1/days  & \citet{bozic2013evolutionary,rew2000cell} \\ \hline
+    $\mu_w$ & Wildtype death rate& $0.15-0.21$ & 1/days  & \citet{bozic2013evolutionary} \\ \hline
     $\lambda_a$  & Aneuploid division rate$^\ast$ & 0.14 & 1/days  & - \\ \hline
-    $\mu_a$ & Aneuploid death rate$^\ast$ & $0.13-0.17$ & 1/days  & - \\ \hline
-    $\lambda_m$ & Mutant division rate& 0.14 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
-    $\mu_m$ & Mutant death rate& 0.13 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
+    $\mu_a$ & Aneuploid death rate$^\ast$ & $0.13-0.21$ & 1/days  & - \\ \hline
+    $\lambda_m$ & Mutant division rate& 0.14 & 1/days  & \citet{bozic2013evolutionary,rew2000cell} \\ \hline
+    $\mu_m$ & Mutant death rate& 0.13 & 1/days  & \citet{bozic2013evolutionary,carlson2003tumor} \\ \hline
     $u$ & Missegregation rate& $10^{-3}-10^{-2}$ & 1$\slash$cell division  & \citet{nowak2004evolutionary,bakker2023predicting} \\ \hline
-    $v$ & Mutation rate& $10^{-7}-10^{-9}$ &  1$\slash$gene$\slash$cell division  & \citet{nowak2004evolutionary} \\ 
+    $v$ & Mutation rate& $10^{-5}-10^{-7}$ &  1$\slash$gene$\slash$cell division  & \citet{nowak2004evolutionary,loeb2001mutator} \\ 
     % TODO citing Nowak 2004 for the rates is strange because it is a theory paper - cite the original empirical papers
@@
   \end{tabular}
-\caption{\textbf{Model parameters.} NEED DIFFERENT REFS---THESE ARE MOSTLY THEORY PAPERS. CHECK BIONUMBERS?
+\caption{\textbf{Model parameters.} %NEED DIFFERENT REFS---THESE ARE MOSTLY THEORY PAPERS. CHECK BIONUMBERS? OK
+The * symbol in the Name column means that for those parameters the values have not been selected from a paper.
 Aneuploid birth rate $\lambda_a$ is set to the same value as the wildtype and mutant birth rates, $\lambda_w$ and $\lambda_m$.
@@
 \centering
-\includegraphics[width=0.75\textwidth]{Figures/figureAneuploidy.pdf}
+\includegraphics[width=\textwidth]{Figures/figureAneuploidy.pdf}
 \caption{
@@
 \textbf{(B)} The wildtype and the mutant are susceptible and resistant, respectively, to the drug.
-The aneuploid may be tolerant, non-growing, or or partially resistant.
+The aneuploid may be tolerant $(4)$, non-growing $(3)$, partially resistant $(2)$ or fully resistant $(1)$.
 }
@@
 the probability that the population does not go to extinction), $\presc$, as a function of the initial tumor size, $N$.
-Dashed line shows the threshold tumor size, $N^*$, above which the probability is very high.
+Dashed vertical line shows the threshold tumor size, $N_a^*$, above which the probability is very high.
+Blue dashed line represents the probability of evolutionary rescue as a function of $N$ without aneuploidy ($u=0$).
+The blue and and red dots represent numerical simulations for the case with aneuploidy and without aneuploidy, respectively.
+The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the  probability of that a succesful mutant has not ben generated and $n=100$ is the number of simulations.
 \textbf{(B)} The threshold tumor size $N_a^*$ as a function of the aneuploid growth rate $\Delta_a$.
 The dashed horizontal line shows $N^*_m$, the threshold tumor size without aneuploidy ($u=0$).
+The red dots represent numerical simulations.
+The inset highlights the case when aneuploidy cancer cells are non-growing.
 When aneuploid growth rate is close to or higher than zero, aneuploidy decreases the threshold tumor size, thereby facilitating evolutionary rescue.
@@
 The dashed horizontal line shows $N^*_m$, the threshold tumor size without aneuploidy ($u=0$).
+The blue line represents the exact formula for threshold tumor size $N_a^*$.
+The red dots represent numerical simulations.
 When the aneuploidy rate is much higher than the mutation rate, aneuploidy decreases the threshold tumor size, thereby facilitating evolutionary rescue.
 }
-\label{fig:LABEL}
+\label{rescue_prob}
+\end{figure}
+%%%%%%%%%%%
+\begin{figure}
+\vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/ReboundProbability.pdf}
+\caption{Plot of the probability that a succesful mutant has not appeared by time $t$.
+The blue line represents the case with aneuploidy ($u>0$), the black line represents the case without aneuploidy ($u=0$).
+The green and and red dots represent numerical simulations for the case with aneuploidy and without aneuploidy, respectively.
+For the simulations we have chosen the following parameters: $\lambda_w=0.14, \lambda_a=0.14,\lambda_m=0.14,\mu_w=0.17,\mu_a=0.1401,\mu_m=0.13, u=0.14\times10^{-2}, v=0.14\times10^{-7}$.
+The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the  probability of that a succesful mutant has not ben generated and $n=100$ is the number of simulations.
+As time increases, aneuploidy plays an important role in helping the cancer cell population escape extinction.}
+\label{ReboundProbability}
 \end{figure}
@@
 %%%%%%%%%%%%
-%
+
 %\begin{figure}
@@
 %\end{figure}
-%
+
 %%%%%%%%%%%%
-%
-%\begin{figure}
-%\begin{subfigure}{0.5\textwidth}
-%% non-growing slightly declining, compare 1st order approximation to exact and simulation
-%\includegraphics[width=1\textwidth]{Figures/DeleteriousTauLeapPlot.pdf}
-%\end{subfigure}
-%\begin{subfigure}{0.5\textwidth}
-%% density-dependent simulations, barely growing (or maybe partial resistance), with exact formula
-%\includegraphics[width=1\textwidth]{Figures/SurvPlotNDataLogisticK.pdf}
-%\end{subfigure}
-%\caption{\textbf{Evolutionary rescue probability for variable initial tumor size.}
-%\textbf{(A)} Comparison of simulation results (markers with 95\% CI, too small to appear with $10^5$ simulations per marker), the exact formula (blue line, \cref{eq:survival_prob} in \cref{eq:rescue_prob}) and the approximate formula (orange line, \cref{rescue_prob_approx}).
-%\textbf{(B)} Comparison of results of simulations  with density-dependent growth (markers with with 95\% CI) and the exact formula (blue line, \cref{eq:survival_prob} in \cref{eq:rescue_prob}) with maximum carrying capacity $K=10^9$.
-%Parameters: $\lambda_w=\lambda_a=\lambda_m=0.14$; $\mu_w=0.17$; (A) $\mu_a=0.15$, (B) $\mu_a=0.135$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
-%}
-%\label{rescue_prob_N}
-%\end{figure}
-%
+
 %%%%%%%%%%%%
@@
 %%%%%%%%%%%%
-%
-%\begin{figure}
-%\begin{subfigure}{0.5\textwidth}
-%\includegraphics[width=1\textwidth]{Figures/MeanTimeDeleteriousPlot.pdf}
-%\end{subfigure}
-%\begin{subfigure}{0.5\textwidth}
-%\includegraphics[width=1\textwidth]{Figures/MeanTimeGrowthMutantDirectPlot.pdf}
-%\end{subfigure}
-%\caption{\textbf{Evolutionary rescue time.}
-%Shown is the mean time for appearance of a resistance mutation the leads to evolutionary rescue \textbf{(left)} with ($u>0$) and \textbf{(right)} without ($u=0$ aneuploidy.
-%Our inhomogeneous Poisson-process approximations (solid blue lines, right: \cref{eq:meantime1}, left: \cref{meantimet2}) is in agreement with simulation results (orange markers with 95\% CI).
-%Our 1st-order (dashed red lines, right: \cref{eq:limitapprox}, left: \cref{limitapprox2}) and 2nd-order (green line, left: \cref{limitapprox3}) approximations work well when the initial tumor size is large (here $>10^8$ cells).
-%Parameters: $\lambda_w=\lambda_a=\lambda_m=0.14$; $\mu_w=0.17$; (A) $\mu_a=0.145$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
-%}
-%\label{MeanTimeGrowthAneuploidyPlot} 
-%\end{figure}
-%
+
 %%%%%%%%%%%%
@@
 %%%%%%%%%%%%
-%
-%\begin{figure}
-% \vspace*{1\baselineskip}
-%\includegraphics[width=1\textwidth]{Figures/ThresholdPopulationSize.pdf}
-%\caption{\textbf{TODO.}
-%TODO
-%}
-%\label{MinTumorSize}
-%\end{figure}
 
@@
 \end{align*}
-Using the quadratic Taylor expansion $\sqrt{1+x}=1+x/2+O(x^2)$ and assuming $u,v \ll 1$,
+Using the quadratic Taylor expansion $\sqrt{1+x}=1+x/2+\mathcal{O}(x^2)$ and assuming $u,v \ll 1$,
 we obtain the following approximation for the survival probability of a population initially consisting of a single wildtype cell,
@@
 \end{equation}
+If we assume that $u>\abs{\Delta_a}$ then we have:
+\begin{equation*}
+p_w\approx\frac{u}{\abs{\Delta_w}}\cdot\frac{v\Delta_m}{\lambda_m \abs{\Delta_a}}.
+\end{equation*}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{Scenario 3: Aneuploid cells are non-growing} % TODO consider this name
+\subsubsection*{Scenario 3: Aneuploid cells are non-growing} % TODO consider this name ok
 
@@
 p_w&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[v\frac{\lambda_m-\mu_m}{\lambda_m}+\frac{u}{2\lambda_a}\left(\lambda_a-\mu_a-v+2\sqrt{\lambda_avp_m}\right)\right]\\
-&=-\frac{1}{\lambda_w-\mu_w-u-v}\left[v\frac{\lambda_m-\mu_m}{\lambda_m}+\frac{u}{2\lambda_a}\left(\lambda_a-\mu_a-v\right)+u\sqrt{\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_a\lambda_m}}\right].
-%&=-\frac{1}{\Delta_w-u-v}\left[v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right],
+&=-\frac{1}{\lambda_w-\mu_w-u-v}\left[v\frac{\lambda_m-\mu_m}{\lambda_m}+\frac{u}{2\lambda_a}\left(\lambda_a-\mu_a-v\right)+u\sqrt{\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_a\lambda_m}}\right]\\
+&\approx-\frac{1}{\Delta_w}\left[v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right].
 \end{aligned}
 \end{equation}
-
+Using the fact that
+\begin{equation*}
+\left(\Delta_a-v\right)^2 < 4\lambda_avp_m\Rightarrow\frac{\Delta_a-v}{2\lambda_a} < \sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}},
+\end{equation*}
+and $v\ll u$ we obtain:
+\begin{equation}
+p_w\approx\frac{u}{\abs{\Delta_w}}\cdot\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}.
+\end{equation}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \section*{Appendix C: Evolutionary rescue time}
-% TODO remove the linear approximation, it is not good.
+% TODO remove the linear approximation, it is not good. OK
 
@@
 \end{equation}
-by integrating the exponential and because $\frac{\exp[\Delta_w t]-1}{\Delta_w}=\frac{1+\Delta_w t+O(t^2)-1}{\Delta_w}=t+O(t^2)$.
+by integrating the exponential and because $\frac{\exp[\Delta_w t]-1}{\Delta_w}=\frac{1+\Delta_w t+\mathcal{O}(t^2)-1}{\Delta_w}=t+O(t^2)$.
 The probability density function of $T_1$ is thus
@@
 
-We wish to obtain a simpler formula for $\tau_2$, similar to \cref{eq:limitapprox}.
-We thus have the following expansions,
+%We wish to obtain a simpler formula for $\tau_2$, similar to \cref{eq:limitapprox}.
+%We thus have the following expansions,
 %\begin{align*}
@@
 %If we apply linear approximations,
+%\begin{align*}
+%\frac{\e^{\Delta_w t}-1}{\Delta_w}&=\frac{1+\Delta_w t+\mathcal{O}(t^2)-1}{\Delta_w} = t+O(t^2),\\
+%\frac{\e^{\Delta_a t}-1}{\Delta_a}&=\frac{1+\Delta_a t+\mathcal{O}(t^2)-1}{\Delta_a} = t+O(t^2),
+%\end{align*}
+%which we use to derive a first-order approximation for $\tau_2$,
+%\begin{align}\label{limitapprox3_appendix}
+%\tau_2 \approx
+%\left(1+\e^{-Np_w}\right)\int_0^\infty\e^{-u N p_m t} \d t = 
+%\frac{\left(1+\e^{-Np_w}\right)}{uNp_m} ,
+%\end{align}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section*{Appendix D: Proliferation time}
+We define the proliferation time to be the time it takes the population of mutant cancer cells to reach the initial tumor size $N$.
+We distinguish between two cases.
+Firstly, when the time it takes for the mutant population to increase by a factor of $e$ is significantly smaller then the time it takes for a mutant cells, which rescues the population, to be generated.
+This is given by the condition:
 \begin{align*}
-\frac{\e^{\Delta_w t}-1}{\Delta_w}&=\frac{1+\Delta_w t+O(t^2)-1}{\Delta_w} = t+O(t^2),\\
-\frac{\e^{\Delta_a t}-1}{\Delta_a}&=\frac{1+\Delta_a t+O(t^2)-1}{\Delta_a} = t+O(t^2),
+\frac{1}{\left(r_1+r_2\right)\left(\tau_2\right)}\gg\frac{1}{\Delta_m}.
 \end{align*}
-which we use to derive a first-order approximation for $\tau_2$,
-\begin{align}\label{limitapprox3_appendix}
-\tau_2 \approx
-\left(1+\e^{-Np_w}\right)\int_0^\infty\e^{-u N p_m t} \d t = 
-\frac{\left(1+\e^{-Np_w}\right)}{uNp_m} ,
+As a result, the prolifereation time is given by~\citep{pompei2023fitness}:
+\begin{align}
+\tau_2'=\tau_2+\frac{\log \Delta_mN}{\Delta_m}.
 \end{align}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+The second case is when the mutant population to increase by a factor of $e$ much faster then new mutants which rescue the population are generated. This is given by the condition:
+\begin{align*}
+\frac{1}{\left(r_1+r_2\right)\left(\tau_2\right)}\ll\frac{1}{\Delta_m}.
+\end{align*}
+As a result, the proliferation time is obtained by solving the following system of ODEs:
+\begin{align*}
+\frac{dw}{dt}&=\Delta_ww,\\
+\frac{da}{dt}&=\Delta_aa+uw,\\
+\frac{dm}{dt}&=\Delta_mm+v\left(w+a\right),
+\end{align*}
+for initial condition $\left(w(0), a(0), m(0)\right)=\left(N,0,0\right)$ and obtaining $\tau_2'$ from $m\left(\tau_2'\right)=N$. Consequently, we obtain the following approximation:
+\begin{equation}
+\tau_2'=\frac{\Delta_w-\Delta_m}{v\Delta_w}\left(1-\frac{u}{\Delta_a-\Delta_m}\right)^{-1},
+\end{equation}
+using the fact that $\e^{x}=1+x+\mathcal{O}\left(x^2\right)$.
+%
+%\begin{align}
+%\tau_2'=
+%   \begin{cases}
+%  \tau_2+\frac{\log \Delta_mN}{\Delta_m} ,&
+% \Delta_m\ll\left(r_1+r_2\right)\left(\tau_2\right) ,\\
+%   \frac{1}{\Delta_a}\log\frac{\left(\Delta_a-\Delta_m\right)\left(\Delta_a\Delta_m-uv\right)}{uv\Delta_w},&
+%    \Delta_m\gg\left(r_1+r_2\right)\left(\tau_2\right)
+%  \end{cases}
+%\end{align}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section*{Appendix E: Distribution of recurrence time}
+The probability that a succesful mutant has been generated by time $t$ is given by:
+\begin{align*}
+P\left(rescue,t\right)&=P\left(T_1<t\right)\\
+&=1-\exp\left\{-\left[r_1\left(t\right)+r_2\left(t\right)\right]\right\}\\
+&=1-\exp\left\{-\left[\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_wt}-1}{\Delta_w}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right)+ vNp_m\frac{\e^{\Delta_w t}-1}{\Delta_w}\right]\right\},
+\end{align*}
+where $T_1$ is the time at which the first mutant cell appears that will avoid extinction and which was defined in Appendix C.
+
+As a result, the probability that a succesful mutant has not been generated by time $t$ is:
+\begin{equation}
+1-P\left(rescue,t\right)=\exp\left\{-\left[\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_wt}-1}{\Delta_w}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right)+ vNp_m\frac{\e^{\Delta_w t}-1}{\Delta_w}\right]\right\}.
+\end{equation}
+\newpage
+\section*{Appendix F: Figures}
+
+\begin{figure}[!htb]
+\begin{subfigure}{0.5\textwidth}
+A\\
+\includegraphics[width=1\textwidth]{Figures/MeanTimeDeleteriousPlot.pdf}
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
+B\\
+\includegraphics[width=1\textwidth]{Figures/MeanTimeGrowthMutantDirectPlot.pdf}
+\end{subfigure}
+\caption{\textbf{Evolutionary rescue time.}
+Shown is the mean time for appearance of a resistance mutation the leads to evolutionary rescue \textbf{(left)} with aneuploidy ($u>0$) and \textbf{(right)} without aneuploidy ($u=0$).
+Our inhomogeneous Poisson-process approximations (solid blue lines, right: \cref{eq:meantime1}, left: \cref{meantimet2}) is in agreement with simulation results (orange markers with 95\% CI). 
+%Our 1st-order (dashed red lines, right: \cref{eq:limitapprox}, left: \cref{limitapprox2}) and 2nd-order (green line, left: \cref{limitapprox3}) approximations work well when the initial tumor size is large (here $>10^8$ cells).
+Parameters: $\lambda_w=\lambda_a=\lambda_m=0.14$; $\mu_w=0.17$; (A) $\mu_a=0.145$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
+}
+\label{MeanTimeGrowthAneuploidyPlot} 
+\end{figure}
+
+\begin{figure}[!htb]
+ \vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/ThresholdPopulationSize.pdf}
+\caption{Minimum size that a tumor, which is exposed to chemotherapeutic drugs, reaches before adaptation sets in and tumor begins growing.
+}
+\label{MinTumorSize}
+\end{figure}
+
+%\begin{figure}[!htb]
+%\begin{subfigure}{0.5\textwidth}
+%% non-growing slightly declining, compare 1st order approximation to exact and simulation
+%\includegraphics[width=1\textwidth]{Figures/DeleteriousTauLeapPlot.pdf}
+%\end{subfigure}
+%\begin{subfigure}{0.5\textwidth}
+%% density-dependent simulations, barely growing (or maybe partial resistance), with exact formula
+%\includegraphics[width=1\textwidth]{Figures/SurvPlotNDataLogisticK.pdf}
+%\end{subfigure}
+%\caption{\textbf{Evolutionary rescue probability for variable initial tumor size.}
+%\textbf{(A)} Comparison of simulation results (markers with 95\% CI, too small to appear with $10^5$ simulations per marker), the exact formula (blue line, \cref{eq:survival_prob} in \cref{eq:rescue_prob}) and the approximate formula (orange line, \cref{rescue_prob_approx}).
+%\textbf{(B)} Comparison of results of simulations  with density-dependent growth (markers with with 95\% CI) and the exact formula (blue line, \cref{eq:survival_prob} in \cref{eq:rescue_prob}) with maximum carrying capacity $K=10^9$.
+%Parameters: $\lambda_w=\lambda_a=\lambda_m=0.14$; $\mu_w=0.17$; (A) $\mu_a=0.15$, (B) $\mu_a=0.135$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
+%}
+%\label{rescue_prob_N}
+%\end{figure}
 

### 2023-11-12 EvolutionaryRescue1.tex
Add files via upload
November Remus

@@
 
-% MESSAGES
+% MESSAGES ok
 % 1. aneuploidy increases the prob of rescue -- reduces the tumor threshold size for rescue
@@
 
-% FIGURES
+% FIGURES ok
 % Fig 1A: model graph chart
@@
 We propose to model evolutionary rescue with the help of multi-type branching processes to obtain the probability that cancer will survive.
+We obtain analytic expressions for the probability of evolutionary rescue as a function of the initial tumor size, the growth rates of the different cancer genotypes and observe that aneuploidy plays an important role in the survival of the cancer cell population when the initial tumor size is not too large and the effect of the anti-cancer drug on the aneuploid cells is not too severe.
 \end{abstract}
@@
 Initial theoretical studies have shown that aneuploidy can have a significant role in the deletion of the the tumor suppressing genes when compared to two consecutive point mutations \citep{nowak2002role,komarova2003mutation,michor2005can,komarova2008selective}.
-However, when taking into account that the appearance of aneuploidy requires a mutation to trigger CIN, the probability that CIN precedes tumor genesis is highly unlikely.
+However, when taking into account that the appearance of aneuploidy requires a mutation to trigger CIN, the probability that CIN precedes tumor genesis is highly unlikely~\citep{komarova2004optimal}.
 
@@
 It is mathematically equivalent to the problem of crossing of fitness valley \citep{weissman2009rate,weissman2010rate}.
-There are three potential ways for a population to survive environmental change: migration to a new habitat similar to the one before the onset of environmental change \citep{cobbold2020should}; adaptation by phenotypic plasticity without genetic modification \citep{carja2019evolutionary,carja2017evolutionary,levien2021non}; and adaptation through genetic modifications, e.g., mutation \citep{uecker2014evolutionary,uecker2016role,uecker2011fixation}.
+There are three potential ways for a population to survive environmental change: migration to a new habitat similar to the one before the onset of environmental change \citep{harsch2014keeping,cobbold2020should,zhou2022range}; adaptation by phenotypic plasticity without genetic modification \citep{carja2019evolutionary,carja2017evolutionary,levien2021non}; and adaptation through genetic modifications, e.g., mutation \citep{uecker2014evolutionary,uecker2016role,uecker2011fixation}.
 
@@
 The difference between the division and death rate is $\Delta_k = \lambda_k-\mu_k$.
-We assume the population of cells is under a strong stress, such as drug therapy, to which the wildtype genotype is susceptible and therefore $\Delta_w<0$, whereas the mutant is resistant to the stress, $\Delta_m>0$
-We analyze three scenarios: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_w$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing or "barely growing", that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$.
-Wildtype cells may missegregate to become aneuploids at rate $u$.
-Both aneuploid and wildtype cells may mutate to become mutants at rate $v$, which we assume is lower than the division rates, $v<\min{(\lambda_w, \lambda_a, \lambda_m)}$.
-See \Cref{figureAneuploidy} for an illustration of the model.
+We assume the population of cells is under a strong stress, such as drug therapy, to which the wildtype genotype is susceptible or sensitive and therefore $\Delta_w<0$, whereas the mutant is resistant to the stress, $\Delta_m>0$.
+We analyze three scenarios: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_w$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing, stationary or "barely growing", that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$.
+We assume that both chromosomal missegregation and mutation occur during the process of mitosis.
+Wildtype cells may divide and then missegregate to become aneuploids at rate $u\lambda_w$.
+Both aneuploid and wildtype cells may divide and mutate to become mutants at rates $v\lambda_{a}$ and $v\lambda_{w}$, respectively, which we assume is lower than the division rates, $v\cdot\max{(\lambda_w, \lambda_a)}<\min{(\lambda_w, \lambda_a, \lambda_m)}$.
+See \Cref{figureAneuploidy} for a schematic representation of the model and \Cref{sampleTrajectories} for sample trajectories of the different genotypes for $u=0 \,\textbf{(A)}, \Delta_a\ll0\, \textbf{(B)},\Delta_a\approx0\, \textbf{(C)},\Delta_a\gg0\, \textbf{(D)}$.
 
@@
 The simulation monitors the number of cells of each type: wildtype, aneuploid, and mutant. 
-The wildtype population initially consists of $w_0$ cells, whereas the other cell types are initially absent.
+The wildtype population initially consists of $w_0=N$ cells, whereas the other cell types are initially absent.
 
 The state of the stochastic system at time $t$ is represented by the triplet $\left(w_t,a_t,m_t\right)$.
-The following describes the events that may occur (right column), the rates at which they occur (middle column), and the effect these events have on the state (\Cref{figureAneuploidy}):
+The following describes the events that may occur (right column), the rates at which they occur (middle column), and the effect these events have on the state (left column, see \Cref{figureAneuploidy}):
 \begin{subequations}
 \begin{flalign*}
-(+1,0,0)&:\quad \lambda_ww_t\quad\left(\text{birth of wildtype cell}\right),\\
+(+1,0,0)&:\quad \lambda_ww_t\left(1-u-v\right)\quad\left(\text{birth of wildtype cell}\right),\\
 (-1,0,0)&:\quad \mu_ww_t\quad\left(\text{death of wildtype cell}\right),\\
-(-1,+1,0)&:\quad uw_t\quad\left(\text{wildtype cell becomes aneuploid}\right),\\
-(-1,0,+1)&:\quad vw_t\quad\left(\text{wildtype cell becomes mutant}\right),\\
-(0,+1,0)&:\quad \lambda_aa_t\quad\left(\text{birth of aneuploid cell}\right),\\
+(0,+1,0)&:\quad u\lambda_ww_t\quad\left(\text{wildtype cell divides and becomes aneuploid}\right),\\
+(0,0,+1)&:\quad v\lambda_ww_t\quad\left(\text{wildtype cell divides and becomes mutant}\right),\\
+(0,+1,0)&:\quad \lambda_aa_t\left(1-v\right)\quad\left(\text{birth of aneuploid cell}\right),\\
 (0,-1,0)&:\quad \mu_aa_t\quad\left(\text{death of aneuploid cell}\right),\\
-(0,-1,+1)&:\quad va_t\quad\left(\text{aneuploid cell becomes mutant}\right),\\
-(0,0,+1)&:\quad \lambda_am_t\quad\left(\text{birth of mutant cell}\right),\\
-(0,0,-1)&:\quad \mu_am_t\quad\left(\text{death of mutant cell}\right).
+(0,0,+1)&:\quad v\lambda_aa_t\quad\left(\text{aneuploid cell divides and becomes mutant}\right),\\
+(0,0,+1)&:\quad \lambda_mm_t\quad\left(\text{birth of mutant cell}\right),\\
+(0,0,-1)&:\quad \mu_mm_t\quad\left(\text{death of mutant cell}\right).
 \end{flalign*}
 \end{subequations}
+For the remaining of this paper we assume that the division rates for wildtype and aneuploid cells can be written as $\lambda_ww_t\left(1-u-v\right)\approx \lambda_ww_t$ and $\lambda_aa_t\left(1-v\right)\approx\lambda_aa_t$ because $u,v\ll1$ (see Table \ref{table1}).
 Each iteration of the simulation loop starts with computing the rates $\nu_j$ of each event $j$.
@@
 \end{align*}
-where $K$ is the maximum carrying capcity. % TODO what constants? see Remus version. Ok
+where $K$ is the maximum carrying capacity.
+The effective carrying capacity of this model is $K_e=K\Delta_a/\lambda_a\approx10^6$ for $K=10^8, \lambda_a=0.0901,\mu_a=0.09$.
+% TODO what constants?
+see Remus version.
+Ok
 
@@
 
-% TODO: OVERVIEW OF RESULTS
+% TODO: OVERVIEW OF RESULTS  ok
 
@@
 \\
-
-In the Appendix, we use the theory of multi-type branching processes to find approximate expressions \cref{eq:survprobwapprox1,eq:survprobwinitial,eq:scenario3} for $p_w$ in different regimes. 
+In the Appendix, we use the theory of multi-type branching processes to find approximate expressions \cref{eq:pw_parttolerant,eq:pw_partrest,eq:pw_tolerant} for $p_w$ in different regimes. 
 Substituting these  into $N^*=1/p_w$, we find approximations for the threshold tumor size, $N^*$. 
-For these approximations, an important quantity is $T^* = (4 v \lambda_a \Delta_m/\lambda_m)^{-1/2}$, which is the critical time an aneuploid lineage needs to survive to produce a resistant mutant that avoids random extinction.
-First, if aneuploidy is very rare ($u T^*< 1$), or if aneuploidy is rare ($u < -\Delta_a$) and very sensitive to the drug ($\Delta_a T^* < -1$), then rescue will likely occur by a direct resistance mutation in a sensitive cell, such that
+For these approximations, an important quantity is $T^* = (4 v \lambda_a^2 \Delta_m/\lambda_m)^{-1/2}$, which is the critical time an aneuploid lineage needs to survive to produce a resistant mutant that avoids random extinction.
+First, if aneuploidy is very rare ($u\lambda_a T^*< 1$), or if aneuploidy is rare ($u\lambda_a < -\Delta_a$) and very sensitive to the drug ($\Delta_a T^* < -1$), then rescue will likely occur by a direct resistance mutation in a sensitive cell, such that
 \begin{equation} \label{eq:N_m}
-N_m^* \approx \frac{\abs{\Delta_w}}{v} \cdot \frac{\lambda_m}{\Delta_m} .
-\end{equation}
-Here, $\abs{\Delta_w}/v$ is the ratio of the rate at which wild-type cells are decreasing in number and the rate at which they are mutating.
-
-Otherwise, aneuploidy is frequent enough ($u > \max{(-\Delta_a, 1/T^*)}$) to affect the evolution of drug resistance. 
+N_m^* \approx \frac{\abs{\Delta_w}}{v\lambda_w} \cdot \frac{\lambda_m}{\Delta_m} .
+\end{equation}
+Here, $\abs{\Delta_w}/\left(v\lambda_w\right)$ is the ratio of the rate at which wild-type cells are decreasing in number and the rate at which they are mutating.
+
+Otherwise, aneuploidy is frequent enough ($u\lambda_a > \max{(-\Delta_a, 1/T^*)}$) to affect the evolution of drug resistance. 
 The threshold tumor size, $N^*$, can then be approximated by one of the following cases, depending on $\Delta_a T^*$, the change in the log of the aneuploid population size during the critical time,
@@
 N_a^* \approx 
-  \frac{\abs{\Delta_w}}{u} \cdot \begin{cases}
-    \frac{\abs{\Delta_a}}{v} \cdot \frac{\lambda_m}{\Delta_m} ,&
-  \Delta_a T^* \ll -1 \text{ (partially sensitive aneuploids)},\\ 
+  \frac{\abs{\Delta_w}}{u\lambda_w} \cdot \begin{cases}
+    \frac{\abs{\Delta_a}}{v\lambda_a} \cdot \frac{\lambda_m}{\Delta_m} ,&
+  \Delta_a T^* \ll -1 \text{ (partially tolerant aneuploids)},\\ 
   %\left(\frac{\lambda_a}{v} \cdot \frac{\lambda_m}{\Delta_m}\right)^{1/2} ,&
@@
 \end{equation}
-%The second case has a finer approximation by $N^*_a \approx \frac{\abs{\Delta_w}}{u} \cdot \frac{2\lambda_a}{\Delta_a + 1/T^*}$.
 The first line describes the case in which aneuploids are still effectively killed by the treatment, but not as quickly as the wild type. 
@@
 Note that in this case there is no dependence on the parameters characterizing mutants or their production ($v$, $\lambda_m$, and $\Delta_m$).
-Comparing these approximations to results of stochastic evolutionary simulations, we find that the approximations perform very well (\Cref{rescue_prob}).%rescue_prob_wt_growth,rescue_prob_an_growth,rescue_prob_N,rescue_prob_N}).
+Comparing these approximations to results of stochastic evolutionary simulations, we find that the approximations perform very well (\Cref{rescue_prob,rescue_threshold}).
+Increasing the the missegregation rate $u$ and aneuploid growth rate $\Delta_a$ reduces the threshold tumor size and facilitates the evolution of drug resistance (\Cref{rescue_threshold}).
 
@@
 \frac{N^*_a}{N^*_m} \approx \begin{cases}
-    \frac{\abs{\Delta_a}}{u} ,&
+    \frac{\abs{\Delta_a}}{u\lambda_a} ,&
   \Delta_a T^* \ll -1 ,\\ 
-  \frac{1}{u}\left(\frac{\lambda_a}{v} \cdot \frac{\lambda_m}{\Delta_m}\right)^{1/2} ,&
-%  \sqrt{\frac{\Delta_a}{u} \cdot v \frac{\Delta_m}{\lambda_m} \cdot \left(u\frac{\Delta_a}{\lambda_a}\right)^{-1}} ,&
+  \frac{1}{u}\left(v \cdot \frac{\Delta_m}{\lambda_m}\right)^{1/2} ,&
   -1 \ll \Delta_a T^* \ll 1  ,\\ 
-  v \frac{\Delta_m}{\lambda_m} \cdot \left(u \frac{\Delta_a}{\lambda_a}\right)^{-1}  ,&
+  v \frac{\Delta_m}{\lambda_m} \cdot \left(u\frac{\Delta_a}{\lambda_a}\right)^{-1}  ,&
    \Delta_a T^* \gg 1 .
@@
 \end{equation}
-In all cases, the effect of aneuploidy increases (i.e., the threshold size ratio decreases) when the aneuploidy rate $u$ increases.
-Increasing the aneuploid growth rate $\Delta_a$ also leads to an increased role of aneuploidy, although the effect is minor when $|\Delta_a|$ is small compared to $T^*$.
-
-In the first case, $\abs{\Delta_a}/u$ is  the ratio of the expected time for an aneuploid lineage to appear, $1/u$, and the expected time until that lineage disappears, $1/\Delta_a$.
+In all cases, the effect of aneuploidy increases (i.e., the threshold size ratio decreases) when the aneuploidy rate $u$ increases (\Cref{rescue_threshold}B).
+Increasing the aneuploid growth rate $\Delta_a$ also leads to an increased role of aneuploidy, although the effect is minor when $|\Delta_a|$ is small compared to $T^*$ (\Cref{rescue_threshold}A).
+Increasing the mutation rate $v$ decreases the effect of aneuploidy and increases the probability of evolutionary rescue.
+
+In the first case, $\abs{\Delta_a}/\left(u\lambda_a\right)$ is  the ratio of the expected time for an aneuploid lineage to appear, $1/\left(u\lambda_a\right)$, and the expected time until that lineage disappears, $1/\abs{\Delta_a}$.
 In the third case, $\left(v \frac{\Delta_m}{\lambda_m}\right) / \left(u \frac{\Delta_a}{\lambda_a}\right)$ is the ratio of the rates of formation of resistant mutants that avoid extinction and partially resistant aneuploids that avoid extinction.
-In the second case, $\frac{1}{u}\left(\frac{\lambda_a}{v} \cdot \frac{\lambda_m}{\Delta_m}\right)^{1/2}=\sqrt{\frac{\Delta_a}{u} \cdot v \frac{\Delta_m}{\lambda_m} \cdot \left(u\frac{\Delta_a}{\lambda_a}\right)^{-1}}$, which is the geometric mean of the first and third cases.
+In the second case, $\frac{1}{u}\left(v \cdot \frac{\Delta_m}{\lambda_m}\right)^{1/2}=\sqrt{\frac{\Delta_a}{u\lambda_a} \cdot v \frac{\Delta_m}{\lambda_m} \cdot \left(u\frac{\Delta_a}{\lambda_a}\right)^{-1}}$, which is the geometric mean of the first and third cases.
 
 Interestingly, increasing both the aneuploid division rate, $\lambda_a$, and the aneuploid death rate, $\mu_a$, such that the growth rate $\Delta_a$ remains constant, leads to decreases in $T^*$, and therefore to the second case.
-In this case, increasing the division rate $\lambda_a$ should also increase the mutation rate $v$ in aneuploid cells, as mutations mostly occur during division , so overall the threshold tumor size $N_a^*$ is unaffected by the division rate $\lambda_a$ (i.e., $d \lambda_a T^*/d\lambda_a = 0$).
+In this case, increasing the division rate $\lambda_a$ should also increase the mutation rate $v\lambda_a$ in aneuploid cells, as mutations mostly occur during division, so overall the threshold tumor size $N_a^*$ is unaffected by the division rate $\lambda_a$ (i.e., $d \lambda_a T^*/d\lambda_a = 0$).
 Thus, if aneuploids rapidly die due to the drug but compensate by rapidly dividing, further increasing  the division rate will \emph{not} facilitate adaptation.
-% TODO PLUG IN REAL VALUES
+For the parameter values of  $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$ we are in the partially resistant aneuploid case and obtain the ratio $N^*_a/N^*_m\approx\abs{\Delta_a}/u\lambda_a=0.11$ which means that the presence of aneuploidy reduces the threshold population by approximately $90\%$.
+We observe that aneuploidy has a significant contribution towards the tumor overcoming chemotherapeutic drugs.
+% TODO PLUG IN REAL VALUES ok
 
@@
 We therefore performed stochastic simulations of a logistic growth model with carrying capacity $K$ (see Methods). 
-We find that our approximations agree with results of simulations with density-dependent growth for biologically relevant parameter values (\Cref{rescue_prob_N}).
+We find that our approximations agree with results of simulations with density-dependent growth for biologically relevant parameter values (\Cref{LogisticPlot}).
 
@@
 % TODO refs OK
-But if the number of cells in the tumor $N$ is large (as expected if the tumor is to be treated with a drug), there may already be a fraction $f \approx \tilde{u}/s$ of aneuploid cells in the population.
+But if the number of cells in the tumor $N$ is large (as expected if the tumor is to be treated with a drug), there may already be a fraction $f \approx \tilde{u}\lambda_w/s$ (we assume that the division rate of the wildtype in a drug-free environment is equal to the division rate of the wildtype in the environment affected by the anti-cancer drug) of aneuploid cells in the population.
 
@@
 In contrast, de-novo aneuploids will have a stronger effect on adaptation if the aneuploidy cost $s$ is large, the effect of the drug is weak ($\Delta_w$ is small), or if the drug induces the appearance of aneuploid cells ($u > \tilde u$).
-
-% TODO PLUG IN REAL VALUES
-
-%In this scenario, the probability of evolutionary rescue by cells with aneuploidy from the initial population is
-%\begin{equation*}
-%% redefine: p_total, p_denovo=p_resuce from eq above; p_stdvar is the new thing
-%p_{sgv} = 1-\left(1-p_a\right)^{fN}\approx 1-\e^{-fNp_a}. 
-%\end{equation*}
-%The total probability of evolutionary rescue is given by
-%\begin{align}\nonumber
-%p_{rescue} 	&= p_{sgv}+\left(1-p_{sgv}\right)p_{de\;novo} \\
-%			&= 1-\exp\left(-\left[\left(1-f\right)p_w + fp_a\right]N\right) .
-%\end{align}
-%
-%The fraction of cases in which the population is rescued by pre-existing aneuploid cells (i.e., standing genetic variation) is given by $F\left(f\right)=\frac{p_{sgv}}{p_{total}}$ (\Cref{FractionPlot}).
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% TODO Move to appendix
-%\subsection*{Effect of aneuploidy on the probability of evolutionary rescue}
-%To determine the extent to which aneuploidy may affect evolutionary rescue, we define $H$ to be the ratio of the rescue probability with and without aneuploidy ($u>0$ and $u=0$, respectively),
-%\begin{equation}\label{ratiorescueexact}
-%H = \frac{p_{rescue}(u>0)}{p_{rescue}(u=0)}.
-%\end{equation}
-
-%Plugging in our approximations from \cref{eq:rescue_prob}, we have
-%\begin{equation}\label{ratiorescue}
-%H = \begin{cases}
-%\frac{1-\exp\left[\frac{N}{\Delta_w-u-v}\left(v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right)\right]}{1-\exp\left[\frac{vN\Delta_m}{\left(\Delta_w-v\right)\lambda_m}\right]} ,&
-%4\lambda_avp_m>\left(\Delta_a-v\right)^2 ,\\
-%\frac{1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right]}{1-\exp\left(\frac{v\Delta_mN}{\lambda_m\Delta_w}\right)}, ,&
-%\Delta_a<0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2 ,\\
-%\frac{1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right]}{1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\right]} ,&
-%\Delta_a>0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2. % TODO this term can be simplified?
-%  \end{cases}
-%\end{equation}
-%
-%We find that the rescue ratio increase with the aneuploidy growth rate $\Delta_a$, because the better aneuploid cells are in growth, the better they are at rescuing the population (when they provide partial resistance) or delaying the extinction of the population (when they provide tolerance).
-%However, the rescue decreases with the wildtype growth rate $\Delta_w$, because the better the wildtype is at growth, the less is depends on aneuploidy for rescue or delay, and the more likely it is to directly produce mutant cells, rather than relying on aneuploid cells for producing mutant cells (\Cref{rescue_ratio}).
-%The effect of the initial tumor size $N$ is the similar to that of the wildtype growth rate. 
-%Importantly, in large tumors, the ratio converges to unity, that is, aneuploidy does not affect the probability for evolutionary rescue.
+Comparing these approximations to results of stochastic evolutionary simulations, we find that the approximations perform very well (\Cref{rescue_denovo}).
+We derive $s$ by using the formula $s=\tilde{u}\lambda_w/f$ and obtaining the fraction of aneuploid cancer cells $f$ from \citep{lukow2021chromosomal} who mixed together wildtype and aneuploid cell at $50:50$ ratio, cultured them in drug-free environment and observed the ratio evolve as a function of time.
+
+For the parameter values of  $\lambda_w=0.1,\mu_w=0.14, u=10^{-2}, \tilde{u}=10^{-3}, s=5\cdot10^{-4}$ the ratio of the threshold population sizes is $\tilde{N}^*_a/N^*_a\approx0.125$ which means that standing genetic variation is the main driver of adaptation even when a relatively small fraction cancer cells are aneuploid in the drug-free environment (in this case $f=20\%$).
+Furthermore, most tumors display significantly higher levels of aneuploidy such that $f\approx 95\%$ even before cancer treatment.
+In this case, the population threshold ratio becomes $\tilde{N}^*_a/N^*_a\approx0.0025$ and standing genetic variation is a very significant driver of adaptation to the chemotherapeutic drug.
+% TODO PLUG IN REAL VALUES ok
 
@@
 Even when evolutionary rescue occurs and leads to recurrence of the tumor, it may take a long time.
-The overall expected recurrence time can be estimated by adding two terms: the mean waiting time for evolutionary rescue--the appearance of a resistant lineage that avoid extinction--and the expected time for proliferation of that lineage back to the original tumor size, $N$.
+When the expected number of resistant lineages that avoid extinction is small the overall expected recurrence time can be estimated by adding two terms: the mean waiting time for evolutionary rescue--the appearance of a resistant lineage that avoid extinction-- (which we denote the mean evolutionary rescue time) and the expected time for proliferation of that lineage back to the original tumor size, $N$ (which we denote the mean proliferation time).
+However, when the expected number of resistant lineages is large the population of mutant cells behaves deterministically and the mean recurrence time cannot be separated into the mean evolutionary rescue time and mean proliferation time.
 
@@
 
-\paragraph{Proliferation time.}
-
-In Appendix D we approximate the mean time for the population of mutant cancer cells to reach the initial population size $N$.
-\Cref{MeanTimeGrowthAneuploidyPlot} shows the agreement between these approximations and simulation results for intermediate and large tumor sizes.
-
-\paragraph{Distribution of recurrence time -- with and without aneuploidy.}
-
-In Appendix E we derive the probability that by time $t$ a succesful mutant has been generated.
+\paragraph{Recurrence time.}
+
+In Appendix D we approximate the mean time for the population of mutant cancer cells to reach the initial population size $N$ which we denote as the recurrence time.
+\Cref{proliferationFigure} shows the agreement between our approximations and simulations.
+For values of $N$ smaller then $10^8$ the mean recurrence time can be approximated as the sum of the mean time for the first rescue muation to appear and the mean time for the lineage generated by that mutation to reach size $N$.
+For $N>10^8$ the behavior of the mutant cancer cell population is deterministic and the mean proliferation time becomes constant.
+
+\paragraph{Distribution of evolutionary rescue time -- with and without aneuploidy.}
+
+In Appendix E we derive the probability that by time $t$ a successful mutant has been generated.
 \Cref{ReboundProbability} show the agreement between our formula and simulation results for the case when aneuploidy is present and when it is absent.
-% ? Distribution of recurrence time -- with and without aneuploidy
-
-% Kaplan-Meier curves -- x: time; y: P(# cells >= N) -- once for u=0; u>0 with the three scenarios
+We observe that aneuploidy start to have an effect on adaptation for timescales greater then 100 days.
+% ? Distribution of recurrence time -- with and without aneuploidy Ok
+% Kaplan-Meier curves -- x: time; y: P(# cells >= N) -- once for u=0; u>0 with the three scenarios Ok
+\paragraph{Distribution of  recurrence time.}
+In Appendix F we derive the probability that mutant cancer cell population reaches $N$ at time $t$.
+\Cref{KaplanMeierfig} shows agreement between our approximations and stochastic simulations for various values of $N$.
 
@@
 
-When aneuploid cells are partially resistant to the drug ($\Delta_w\ll0\ll\Delta_a\ll\Delta_m$), evolutionary rescue can be approximated by a one-step process in which aneuploidy itself rescues the population (\Cref{rescue_prob_wt_growth}).
+\paragraph{Evolutionary rescue}
+When aneuploid cells are partially resistant to the drug ($\Delta_w\ll0\ll\Delta_a\ll\Delta_m$), evolutionary rescue can be approximated by a one-step process in which aneuploidy itself rescues the population (\Cref{rescue_threshold}A).
 When aneuploidy only provides tolerance to the drug ($\Delta_w\ll\Delta_a\ll0\ll\Delta_m$), it cannot rescue the population.
-Instead, it acts as a \emph{stepping stone} through which the resistant mutant can appear more rapidly, given that the aneuploid cell population declines slower then the sensitive cell population.
+Instead, it acts as a \emph{stepping stone} through which the resistant mutant can appear more rapidly, given that the aneuploid cell population declines slower then the sensitive cell population (\Cref{sampleTrajectories}).
 In this case, aneuploidy provides two benefits.
 First, it delays the extinction of the population--providing more time for appearance of the resistance mutation.
-Second, it increases the population size relative to a sensitive population--providing more cells in which mutations can occur, i.e., it increases the mutation supply, $Nv$.
+Second, it increases the population size relative to a sensitive population--providing more cells in which mutations can occur, i.e., it increases the mutation supply, $Nuv\lambda_w\lambda_a/\abs{\Delta_w\Delta_a}$.
 
 We find that aneuploidy can have a significant effect on evolutionary rescue (\Cref{rescue_prob}).
-For example, when aneuploidy cells are "barely-resistant" (they grow at a very low rate, $\Delta_a=10^{-3}$) the probability of evolutionary rescue is 1,000-fold higher with aneuploidy than without it (for parameters previously described in cancer, see \Cref{table1}).
-Interestingly, aneuploidy is unlikely to contribute to evolutionary rescue in primary tumors in which the number of cells is large enough ($N>10^7$) for the appearance of resistant mutation directly in sensitive cells before these cells become extinct (\Cref{rescue_ratio}).
+Interestingly, aneuploidy is unlikely to contribute to evolutionary rescue in primary tumors in which the number of cells is large enough ($N>10^8$) for the appearance of resistant mutation directly in sensitive cells before these cells become extinct (\Cref{rescue_prob}).
 However, aneuploidy can have a crucial role in evolutionary rescue of secondary tumors, in which the number of sensitive cells may be below the detection threshold of $\sim10^7$  \citep{bozic2013evolutionary}.
-Given the fact that the mean time for such secondary tumors to overcome chemotherapy can be of the order of 100 days (\Cref{MeanTimeGrowthAneuploidyPlot}), % TODO do we see tau=100 here...?
+This can have an impact on the recurrence of cancer after the resection of the primary tumor  through secondary tumors which are too small to be detected and for which chemotherapy is employed to prevent  cancer relapse and are estimated to cause the majority of cancer-related deaths~\citep{chaffer2011perspective}.
+
+Given the fact that the mean time for such secondary tumors to overcome chemotherapy can be of the order of 1000 days (\Cref{MeanTimeGrowthAneuploidyPlot}A), % TODO do we see tau=100 here...? OK
 this can explain the reappearance of cancer even after initial remission.
-Indeed, we find that the tumor size can decrease by orders of magnitude before it is rescued (\Cref{MinTumorSize}).
-
-We hypothesized that presence of \emph{standing variation} % TODO ref <- why ref?
+The theoretical predictions for the mean rescue time for tumors smaller then  $10^8$ is greater then 4 years which is consistent with previous estimates of the recurrence time of tumors after resection~\citep{avanzini2019cancer}.
+
+Paradoxically, we observe from \Cref{MeanTimeGrowthAneuploidyPlot} that the mean time of a rescue mutation to appear is significatly shorter for the case when $u=0$ when compared to the case $u>0$, however this can be explained by the fact this mean time is conditioned on evolutionary rescue and, as a result, aneuploidy increase the \emph{window of opportunity} in which a rescue mutation could appear thus increasing the mean time as well (\Cref{sampleTrajectories}).
+
+We hypothesized that presence of \emph{standing variation} 
 --the existence of a subpopulation of aneuploid cancer cells before therapy begins--can facilitate evolutionary rescue by reducing the waiting time for the appearance of aneuploid cells.
-Indeed, we observe that even when a small fraction of the initial tumor is aneuploid, evolutionary rescue is more likely to occur through this existing standing variation, rather then through \emph{de novo} aneuploid cells (\Cref{FractionPlot}).
-
-We have assumed that cancer cell lineages are independent of each other.
-However, this may not be the case, as cancer cells compete for resources (e.g., blood supply).
-Nevertheless, we find that when the carrying capacity is large ($K\sim10^7$) % TODO HOW LARGE?
-ok
-our approximation for the probability of evolutionary rescue agrees with results of stochastic simulations with density-dependent growth  \Cref{rescue_prob_N}).
-Future work may focus on scenarios with small carrying capacity by analyzing density-dependent branching processes~\citep{klebaner1997population}. % TODO ref? OK
-
-Our model predictions may be tested by experiments \citep{martin2013probability}.
+Indeed, we observe that even when a small fraction of the initial tumor is aneuploid, evolutionary rescue is more likely to occur through this existing standing variation, rather then through \emph{de novo} aneuploid cells (\Cref{rescue_denovo}).
+
+We observe from \Cref{KaplanMeierfig} that the probability of the mutant cell population reaching size $N$ is approximately zero before time $\tau_2^r$ which is the recurrence time for the deterministic case.
+This can be explained as follows: in the deterministic case there is a sufficient number of lineages produced such that there exists a lineage where each descendant will only reproduce and not die; the time it takes for this lineage to reach $N$ is the lower bound for the time of all other lineages to reach $N$ and this time cannot be smaller then $\tau_2^r$ by definition.
+Given that for small values of $N$ we expect that at most a single lineage will rescue the tumor, this lineage cannot reach $N$ before $\tau_2^r$ for the deterministic case \cref{eq:t2det}.
+
+\paragraph{Experimental future direction}
+Our model's predictions may be tested by experiments \citep{martin2013probability}.
 For example, to study the effects of initial tumor size on the probability of evolutionary rescue, a large culture mass can be propagated from a single cancer cell in permissive conditions and then diluted to a  range of starting tumor sizes.
@@
 or to saline solution for control~\citep{ippolito2021gene}. 
-Cell density can then be measured and compared to the predictions of our model. % TODO not clear what is to be compared... what does our model predict...?
-
-We observe that the presence of aneuploidy has the effect of shortening the time necessary for a mutant to appear which will rescue the tumor (\Cref{MeanTimeGrowthAneuploidyPlot}).
-However, this effect is only true for small tumors ($N<10^5$) as direct mutation is the main mechanism for generation of successful mutants for large and intermediate population sizes.
+Cell density can then be measured by optical density and a population exposed to the drug is considered extinct if the optical density is lower when compared to the control case.
+We compare to the predictions of our model to see if the tumors with initial size bellow the threshold \eqref{eq:N_a} are more likely to go extinct.
+% TODO not clear what is to be compared...
+what does our model predict...?
+Ok
+
 % TODO discussion of mean time? ok
 
+\paragraph{Directions of future research}
+Even though our model has been designed for cancer it can be extended to understand to understand evolutionary in different biological contexts, for example, how yeast subject to stress can overcome extinction with the help of aneuploidy~\citep{pompei2023fitness}.
+Additionally, the heterogeneity of aneuploidy is a fact the we did not account for, as not all the aneuploidy lineages generated have the same growth rate $\Delta_a$.
+This can be corrected by  fixing the aneuploidy death rates and sampling the aneuploidy growth rates from a distribution $f\left(\Delta_a|\mu_a\right)$~\citet{martin2013probability}.
+Furthermore, tumor heterogeneity is also a factor for pre-existing genetic variation where the fitness of the aneuploid cells can be drawn from a distribution upon which selection acts when the tumor is exposed to a strong stress.
+
+We have assumed that cancer cell lineages are independent of each other.
+However, this may not be the case, as cancer cells compete for resources (e.g., blood supply).
+Nevertheless, we find that when the effective carrying capacity is not too large ($K\sim10^8, K_e\sim10^6$) % TODO HOW LARGE?
+ok
+our approximation for the probability of evolutionary rescue agrees with results of stochastic simulations with density-dependent growth  (\Cref{LogisticPlot}).
+Future work may focus on scenarios with small carrying capacity by analyzing density-dependent branching processes~\citep{klebaner1997population,harris1963theory}. % TODO ref? OK
+
+\paragraph{Conclusions}
+
+Our paper analyzed the role that aneuploidy plays in tumor adaptation to the effects of anti-cancer drugs.
 Our study quantitatively confirms that aneuploidy plays an important role in tumors overcoming exposure to chemotherapeutic drugs when the tumor size is small or intermediate.
-Very large tumors can escape anti-cancer drugs through direct mutation while smaller ones are able to obtain the beneficial mutaion through an aneuploid intermediary (\Cref{rescue_prob}).
+Very large tumors can escape anti-cancer drugs through direct mutation while smaller ones are able to obtain the beneficial mutation through an aneuploid intermediary (\Cref{rescue_prob}).
+As a result, therapies which increase the rate of aneuploidy in tumors in order to combat cancer might have an adverse effect on patient outcomes.
 % TODO summary ok
@@
 \section*{Acknowledgements}
-%We thank Hildegard Uecker for discussions and comments. % TODO who?
+We thank Hildegard Uecker for discussions and comments. % TODO who? ok
 This work was supported in part by
@@
 Minerva Stiftung Center for Lab Evolution (YR), 
-and the Ela Kodesz Institute for Research on Cancer Development and Prevention (RS).
-% TODO add funding for Uri and Daniel
+Ela Kodesz Institute for Research on Cancer Development and Prevention (RS),
+the Simons Foundation (Investigator in Mathematical Modeling of Living Systems $\#508600$, DBW),
+the Sloan Foundation Research Fellowship (FG-2021-16667, DBW),
+the NSF grant ($\#2146260$, DBW),
+{\color{red} Uri Ben-David acknowledgements}
+
+% TODO add funding for Uri and Daniel ok
 }
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
+\newpage
 \begin{table}
@@
     $N$ & Initial tumor size & $10^7-10^9$ & cells  & \citet{del2009does} \\ \hline
-    $\lambda_w$ & Wildtype division rate& 0.14 & 1/days  & \citet{bozic2013evolutionary,rew2000cell} \\ \hline
-    $\mu_w$ & Wildtype death rate& $0.15-0.21$ & 1/days  & \citet{bozic2013evolutionary} \\ \hline
-    $\lambda_a$  & Aneuploid division rate$^\ast$ & 0.14 & 1/days  & - \\ \hline
-    $\mu_a$ & Aneuploid death rate$^\ast$ & $0.13-0.21$ & 1/days  & - \\ \hline
-    $\lambda_m$ & Mutant division rate& 0.14 & 1/days  & \citet{bozic2013evolutionary,rew2000cell} \\ \hline
-    $\mu_m$ & Mutant death rate& 0.13 & 1/days  & \citet{bozic2013evolutionary,carlson2003tumor} \\ \hline
-    $u$ & Missegregation rate& $10^{-3}-10^{-2}$ & 1$\slash$cell division  & \citet{nowak2004evolutionary,bakker2023predicting} \\ \hline
-    $v$ & Mutation rate& $10^{-5}-10^{-7}$ &  1$\slash$gene$\slash$cell division  & \citet{nowak2004evolutionary,loeb2001mutator} \\ 
-    % TODO citing Nowak 2004 for the rates is strange because it is a theory paper - cite the original empirical papers
+    $\lambda_w$ & Wildtype division rate& 0.1 & 1/days  & \citet{bozic2013evolutionary,rew2000cell} \\ \hline
+    $\mu_w$ & Wildtype death rate& $0.11-0.17$ & 1/days  & \citet{bozic2013evolutionary} \\ \hline
+    $\lambda_a$  & Aneuploid division rate$^\ast$ & $0.06-0.1$ & 1/days  & - \\ \hline
+    $\mu_a$ & Aneuploid death rate$^\ast$ & $0.09$ & 1/days  & - \\ \hline
+    $\lambda_m$ & Mutant division rate& 0.1 & 1/days  & \citet{bozic2013evolutionary,rew2000cell} \\ \hline
+    $\mu_m$ & Mutant death rate& 0.09 & 1/days  & \citet{bozic2013evolutionary,carlson2003tumor} \\ \hline
+    $u$ & Missegregation rate& $10^{-3}-10^{-2}$ & 1$\slash$cell division  & \citet{bakker2023predicting} \\ \hline
+    $v$ & Mutation rate& $10^{-7}-10^{-5}$ &  1$\slash$gene$\slash$cell division  & \citet{loeb2001mutator} \\  \hline
+    $\tilde{u}$ & Missegregation rate in the drug free environment$^\ast$& $10^{-3}$ & 1$\slash$cell division  & - \\ \hline
+    $s$ & Selection coefficient of aneuploidy in the drug free environment& $0.3\cdot10^{-3}-10^{-3}$ &  1/days   & \citet{lukow2021chromosomal} \\  
+    %$f$ & Fraction of cancer cells which are aneuploid in the drug free environment& $10\%-30\%$ & -  & \citet{salehi2021clonal}\citet{lukow2021chromosomal} \\ 
+    % TODO citing Nowak 2004 for the rates is strange because it is a theory paper - cite the original empirical papers Ok
     \hline
@@
 The * symbol in the Name column means that for those parameters the values have not been selected from a paper.
-Aneuploid birth rate $\lambda_a$ is set to the same value as the wildtype and mutant birth rates, $\lambda_w$ and $\lambda_m$.
-Aneuploid death rate $\mu_a$ is set to an intermediate value between the wildtype and mutant death rates, $\mu_w$ and $\mu_m$.}
+Aneuploid death rate $\mu_a$ is set to the same value as the mutant death rates, $\mu_m$, given that aneuploidy increases resistance to chemotherapeutics, such as cisplatin, by antagonizing cell division \citep{replogle2020aneuploidy}.
+Aneuploid birth rate $\lambda_a$ is set such that the aneuploid growth rate $\Delta_w\ll\Delta_a\ll\Delta_m$.
+For the missegregation rate in the drug free environment $\tilde{u}$ we choose the lower-end value of $u$ as the chemotherapeutic drugs we want to model increase the rate of aneuploidy~\citep{wang2019molecular,mason2017functional}.
+We note that we have modified the parameters from \citet{bozic2013evolutionary} by starting their analysis from the fact that wildtype/mutant birth rate is $\lambda_{w,m}=\log2/T\approx0.1$ instead of their value of $0.14$ where $T$ is the doubling time in the absence of cellular death obtained from \citet{rew2000cell}.
+We suspect the value of 0.1 was obtained as $1/T$ because $T$ can also be defined as the average time between cell divisions in the absence of cell death~\citet{avanzini2019cancer}.
+We derive the fitness cost $s$ by using the formula $s=\tilde{u}\lambda_w/f$ and obtaining the fraction of aneuploid cancer cells $f$ from \citep{lukow2021chromosomal} who mixed together wildtype and aneuploid cell at $50:50$ ratio, cultured them in drug-free environment and observed the ratio evolve as a function of time.}
   \label{table1}
@@
 \textbf{(A)} A population of cancer cells is composed of wildtype, aneuploid, and mutant cells, which divide with rates $\lambda_w$, $\lambda_a$, and $\lambda_m$ and die at rates $\mu_w$, $\mu_a$, and $\mu_m$, respectively.
-Wildtype cells can become aneuploid at rate $u$.
-Both aneuploid and wildtype cells can acquire a beneficial mutation with rate $v$.
+Wildtype cells can divide and become aneuploid at rate $u\lambda_w$.
+Both aneuploid and wildtype cells can divide and acquire a beneficial mutation with rate $v\lambda_a$ and $v\lambda_w$, respectively.
 Color denotes the relative growth rates of the three genotypes such that $\lambda_w - \mu_w < \lambda_a - \mu_a < \lambda_m - \mu_m$.
 \textbf{(B)} The wildtype and the mutant are susceptible and resistant, respectively, to the drug.
-The aneuploid may be tolerant $(4)$, non-growing $(3)$, partially resistant $(2)$ or fully resistant $(1)$.
+The aneuploid may be tolerant $(4)$, stationary $(3)$, partially resistant $(2)$ or fully resistant $(1)$.
 }
 \label{figureAneuploidy}
+\end{figure}
+
+%%%%%%%%%%%
+\begin{figure}
+\begin{subfigure}{0.5\textwidth}
+A\\
+\includegraphics[width=1\textwidth]{Figures/TauLeapMeanTimeDiagramNoAneuploidy.pdf}
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
+B\\
+\includegraphics[width=1\textwidth]{Figures/TauLeapMeanTimeDiagramSmallda.pdf}
+\end{subfigure}
+\\
+\begin{subfigure}{0.5\textwidth}
+C\\
+\includegraphics[width=1\textwidth]{Figures/TauLeapMeanTimeDiagramdazero.pdf}
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
+D\\
+\includegraphics[width=1\textwidth]{Figures/TauLeapMeanTimeDiagramlargeda.pdf}
+\end{subfigure}
+\caption{
+\textbf{Sample trajectories of the different genotypes as a function of time.}
+\textbf{(A)} Whe the missegregation rate $u=0$ evolutionary rescue is only possible through direct mutation and in most cases the tumour will be cured by the drug.
+\textbf{(B)} When the aneuploidy growth rate $\Delta_a\ll0$ we observe similar dynamic to case \textbf{(A)} as direct mutation is the only viable route toward evolutionary rescue for the tumor.
+\textbf{(C)} Intermediate aneuploid growth rates $\Delta_a\approx0$ generates the appearence of aneuploid lineages even after the wildtype population has gone extinct thus increasing the chance of evolutionary rescue.
+\textbf{(D)} As the growth rate of the aneuploid becomes positive we observe that the tumour is rescued by the aneuploid cancer cell population.
+Each plot features 10 trajectories of $\left(w_t,a_t,m_t\right)$ as a function of time $t$ for the following parameter values: $\lambda_w=0.1,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, v=10^{-7},N=10^7$.
+For \textbf{(A)} we set $u=0$, for \textbf{(B)} $\lambda_a=0.065,u=10^{-2}$, for \textbf{(C)} $\lambda_a=0.0899,u=10^{-2}$ and for \textbf{(D)}   $\lambda_a=0.095,u=10^{-2}$.
+}
+\label{sampleTrajectories}
 \end{figure}
@@
 \begin{figure}
+\includegraphics[width=1\textwidth]{Figures/ProbvNPlot.pdf}
+\caption{\textbf{Aneuploidy facilitates evolutionary rescue of cancer under drug treatment.}
+The probability of evolutionary rescue (i.e.
+the probability that the population does not go to extinction), $\presc$, as a function of the initial tumor size, $N$.
+Dashed vertical line shows the threshold tumor size, $N_a^*$, above which the probability is very high.
+Blue dashed line represents the probability of evolutionary rescue as a function of $N$ without aneuploidy ($u=0$).
+The black line represents the case with tolerant aneuploidy ($u=10^{-2}, \lambda_a=0.0899$), the red line represents the case with stationary aneuploidy ($u=10^{-2}, \lambda_a=0.08999$) and the green line represents the case with partially resistant aneuploidy ($u=10^{-2}, \lambda_a=0.095$).
+The dots represent numerical simulations and the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the probability of that the tumor has adapted to the stress and $n=100$ is the number of simulations.
+Parameters: $\lambda_w=0.1,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, v=10^{-7}$.}
+\label{rescue_prob}
+\end{figure}
+
+\begin{figure}
 \begin{subfigure}{0.5\textwidth}
 A\\
-\includegraphics[width=1\textwidth]{Figures/ProbvNPlot.pdf}
-\end{subfigure}
-\\
-\begin{subfigure}{0.5\textwidth}
-B\\
 \includegraphics[width=1\textwidth]{Figures/ThresholdPopulationSizePlot.pdf}
@@
 \begin{subfigure}{0.5\textwidth}
-C\\
+B\\
 \includegraphics[width=1\textwidth]{Figures/ThresholdPopulationSizeVersusRatioPlot.pdf}
@@
 \caption{
-\textbf{Aneuploidy facilitates evolutionary rescue of cancer under drug treatment.}
-\textbf{(A)} The probability of evolutionary rescue (i.e.
-the probability that the population does not go to extinction), $\presc$, as a function of the initial tumor size, $N$.
-Dashed vertical line shows the threshold tumor size, $N_a^*$, above which the probability is very high.
-Blue dashed line represents the probability of evolutionary rescue as a function of $N$ without aneuploidy ($u=0$).
-The blue and and red dots represent numerical simulations for the case with aneuploidy and without aneuploidy, respectively.
-The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the  probability of that a succesful mutant has not ben generated and $n=100$ is the number of simulations.
-\textbf{(B)} The threshold tumor size $N_a^*$ as a function of the aneuploid growth rate $\Delta_a$.
+\textbf{Aneuploidy growth rate and missegregation rate impact the evolutionary rescue of cancer under drug treatment.}
+\textbf{(A)} The threshold tumor size $N_a^*$ as a function of the aneuploid growth rate $\Delta_a$.
 The dashed horizontal line shows $N^*_m$, the threshold tumor size without aneuploidy ($u=0$).
@@
 When aneuploid growth rate is close to or higher than zero, aneuploidy decreases the threshold tumor size, thereby facilitating evolutionary rescue.
-\textbf{(C)} The threshold tumor size $N_a^*$ as a function of the ratio of aneuploidy and mutation rates, $u/v$.
+Parameters: $\lambda_w=0.1,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.
+\textbf{(B)} The threshold tumor size $N_a^*$ as a function of the ratio of aneuploidy and mutation rates, $u/v$.
 The dashed horizontal line shows $N^*_m$, the threshold tumor size without aneuploidy ($u=0$).
-The blue line represents the exact formula for threshold tumor size $N_a^*$.
+The blue line represents the exact formula for threshold tumor size $N_a^*$ while the solid black line represents the approximation \cref{eq:N_a}.
 The red dots represent numerical simulations.
 When the aneuploidy rate is much higher than the mutation rate, aneuploidy decreases the threshold tumor size, thereby facilitating evolutionary rescue.
+Parameters: $\lambda_w=0.1,\lambda_m=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, v=10^{-7}$.
+For the threshold tumor sizes the error bars represent the $95\%$ confidence interval obtained through bootstrapping in the following steps: \textbf{(1)} we simulate $p_{rescue}$ 100 times; \textbf{(2)} we sample with replacement which we store in $S$; \textbf{(4)} for each element of this sample we obtain $N_a^*=1/p_w$ using $p_w=-1/N_s\log\left(1-\bar{S}\right)$ where $\bar{S}$ is the mean of $S$ and $N_s$ is an arbitrary value of the initial population size we selected in order to calculate $p_{rescue}$; \textbf{(5)} we repeat steps \textbf{(2)}-\textbf{(4)} 100 times to obtain $N_a^*$ and we select the upper and lower limits such that $95\%$ of the values of $N_a^*$ lie in the interval given by the bounds.
 }
-\label{rescue_prob}
+\label{rescue_threshold}
+\end{figure}
+
+%%%%%%%%%%%
+% Fig 3A: N*/N* vs Delta_w 
+% Fig 3B: N*/N* vs ut/s
+
+\begin{figure}
+\begin{subfigure}{0.5\textwidth}
+A\\
+\includegraphics[width=1\textwidth]{Figures/RatiodwPlot.pdf}
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
+B\\
+\includegraphics[width=1\textwidth]{Figures/ratio_uPlot.pdf}
+\end{subfigure}
+\caption{
+\textbf{Standing genetic variation facilitates evolutionary rescue of cancer under drug treatment.}
+\textbf{(A)}  The ratio of threshold tumor size $\tilde{N}_a^*$ when a fraction $\frac{\tilde{u}\lambda_w}{s}$ is aneuploid at the start of treatment and $N_a^*$ as a function of the wildtype growth rate $\Delta_w$.
+The red dots represents numerical simulations and the error bars represent the $95\%$ confidence intervals obtained with bootstrapping explained below.
+Standing genetic variation will drive adaptation to the drug if $\Delta_w$ is very negative due to a stronger effect of the drug on sensitive cells.
+Parameters: $\lambda_w=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_a=0.09,\mu_m=0.09,\tilde{u}=10^{-3},u=10^{-2}, v=10^{-7}$.
+\textbf{(B)} The ratio of threshold tumor size $\tilde{N}_a^*$ when a fraction $\frac{\tilde{u}\lambda_w}{s}$ is aneuploid at the start of treatment and $N_a^*$ as a function of the the ratio of aneuploidy rates $\tilde{u}/u$.
+The red dots represents numerical simulations and the error bars represent the $95\%$ confidence intervals obtained with bootstrapping explained below.
+De-novo aneuploids will have larger contribution to the appearance of drug resistance if the drug induces the appearance of aneuploid cells ($u \gg \tilde u$).
+Parameters: $\lambda_w=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09,\tilde{u}=10^{-3}, v=10^{-7}$.
+For the ratio of the threshold tumor sizes the error bars represent the $95\%$ confidence interval obtained through bootstrapping in the following steps: \textbf{(1)} we simulate $p_{rescue}$ 100 times for both the case when $f=\tilde{u}\lambda_w/s$ and $f=0$; \textbf{(2)} we sample with replacement which we store in $S_0$ and  $S_f$; \textbf{(4)} for each element of $S_0$ we obtain $N_a^*=1/p_w$ using $p_w=-1/N_s\log\left(1-\bar{S}\right)$ where $\bar{S}$ is the mean of $S_0$ and $N_s$ is an arbitrary value of the initial population size we selected in order to calculate $p_{rescue}$; \textbf{(5)} for each element of $S_f$ we obtain $\tilde{N}_a^*=1/p_a$ using $p_a=-f/N_s\log\left(1-\bar{S}\right)$ where $\bar{S}_f$ is the mean of $S_f$ and $N_s$ is an arbitrary value of the initial population size we selected in order to calculate $p_{rescue}$; \textbf{(6)} we repeat steps \textbf{(2)}-\textbf{(5)} 100 times to obtain $\tilde{N}_a^*/N_a^*$ and we select the upper and lower limits such that $95\%$ of the values of $\tilde{N}_a^*/N_a^*$ lie in the interval given by the bounds.
+}
+\label{rescue_denovo}
 \end{figure}
@@
 \includegraphics[width=1\textwidth]{Figures/ReboundProbability.pdf}
-\caption{Plot of the probability that a succesful mutant has not appeared by time $t$.
-The blue line represents the case with aneuploidy ($u>0$), the black line represents the case without aneuploidy ($u=0$).
-The green and and red dots represent numerical simulations for the case with aneuploidy and without aneuploidy, respectively.
-For the simulations we have chosen the following parameters: $\lambda_w=0.14, \lambda_a=0.14,\lambda_m=0.14,\mu_w=0.17,\mu_a=0.1401,\mu_m=0.13, u=0.14\times10^{-2}, v=0.14\times10^{-7}$.
-The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the  probability of that a succesful mutant has not ben generated and $n=100$ is the number of simulations.
+\caption{\textbf{Aneuploidy is a key diver of cancer adaptation.}
+Plot of the probability that a successful mutant has not appeared by time $t$.
+The green line represents the case with tolerant aneuploidy ($u>0, \lambda_a=0.0899$), the blue line represents the case with non-growing aneuploidy ($u>0, \lambda_a=0.089999$), the cyan line represents the case with partially resistant aneuploidy ($u>0, \lambda_a=0.095$) and the black line represents the case without aneuploidy ($u=0$).
+The  dots represent numerical simulations.
+For the simulations we have chosen the following parameters: $\lambda_w=0.1,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7},N=10^7$.
+The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the  probability of that a successful mutant has not been generated and $n=100$ is the number of simulations.
 As time increases, aneuploidy plays an important role in helping the cancer cell population escape extinction.}
@@
 
-%%%%%%%%%%%%
-%
-%\begin{figure}
-%\begin{subfigure}{0.5\textwidth}
-%\includegraphics[width=1\textwidth]{Figures/FILENAME.pdf}
-%\end{subfigure}
-%\begin{subfigure}{0.5\textwidth}
-%\includegraphics[width=1\textwidth]{Figures/FILENAME.pdf}
-%\end{subfigure}
-%\caption{
-%\textbf{TITLE}
-%CAPTION
-%}
-%\label{fig:LABEL}
-%\end{figure}
-%%%%%%%%%%%%
-
-%%%%%%%%%%%%
-%
-%\begin{figure}
-%\begin{subfigure}{0.5\textwidth}
-%\includegraphics[width=1\textwidth]{Figures/CombinedSubplot.pdf}
-%\end{subfigure}
-%\begin{subfigure}{0.5\textwidth}
-%\includegraphics[width=1\textwidth]{Figures/CombinedSubplotLargePop.pdf}
-%\end{subfigure}
-%\caption{
-%\textbf{Evolutionary rescue probability with partially resistant or tolerant aneuploid cells.}
-%Rescue probability is very high when aneuploidy provides partial resistance ($\lambda_a=0.01$), in an initially small tumor (\textbf{A}, $N=10^4$) and even more so in an initially large tumor (\textbf{B}, $N=10^8$).
-%When aneuploidy provides tolerance (\textbf{C}, $N=10^4$; \textbf{D}, $N=10^8$), the rescue probability is much lower. 
-%In both scenarios, rescue probability increase with both the wildtype growth rate (x-axis) and the aneuploidy growth rate (colors).
-%Markers represent simulation results with 95\% CI; solid and dashed lines for the exact formula (\cref{eq:survival_prob} in \cref{eq:rescue_prob}); dashed lines for the approximate formula (\cref{rescue_prob_approx}), demonstrating that they all agree.
-%Parameters: division rate $\lambda_w=\lambda_a=\lambda_m=0.14$ (so that growth rate changes due to variable death rate); mutant death rate $\mu_m=0.13$ (so that mutant growth rate $\Delta_m=0.01$); aneuploidy rate $u=10^{-2}$; mutation rate $v=10^{-7}$.
-%}
-%\label{rescue_prob_wt_growth}
-%\end{figure}
-%%%%%%%%%%%%
-
-%\begin{figure}
-%% small population size
-%% switch from tolerance to non-growing
-%% TODO why no switch in panel B?
-%% TODO more simulations for 0 > Delta_a > Delta^*_a
-%% TODO simulations in A?
-%\begin{subfigure}{0.5\textwidth}
-%\includegraphics[width=1\textwidth]{Figures/P_est_divergence.pdf}
-%\end{subfigure}
-%\begin{subfigure}{0.5\textwidth}
-%\includegraphics[width=1\textwidth]{Figures/P_est_divergenceLargePopulation.pdf}
-%\end{subfigure}
-%\caption{\textbf{Evolutionary rescue probability with tolerant or non-growing aneuploid cells.}
-%Rescue probability grows with the aneuploid growth rate $\Delta_a$ (x-axis), and is much higher in an initially large tumor than in a small one (\textbf{A}: $N=10^4$; \textbf{B}: $N=10^8$). 
-%Markers for simulation results with 95\% CI; 
-%solid lines for the exact formula (\cref{eq:survival_prob} in \cref{eq:rescue_prob});
-%dashed lines for the approximate formula (\cref{rescue_prob_approx}).
-%The approximation agrees with the simulation and exact solution when the initial tumor size is large (panel B).
-%When the tumor size is small (panel A), we switch between the approximation for tolerant and for non-growing aneuploid cells; the switch occurs at $1/T^*$. % TODO ref to eq
-%Parameters: $\lambda_w=\lambda_a=\lambda_m=0.14$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
-%}
-%\label{rescue_prob_an_growth}
-%\end{figure}
-
-%%%%%%%%%%%%
-
-%%%%%%%%%%%%
-%
-%\begin{figure}
-%% slightly decreasing (need to check)
-% \vspace*{1\baselineskip}
-%\includegraphics[width=1\textwidth]{Figures/FractionPlot.pdf}
-%\caption{\textbf{Effect of standing variation on evolutionary rescue.}
-%In aneuploid cells already exist in the population at the onset of drug therapy as standing genetic variation, then evolutionary rescue is more likely... 
-%Parameters: $\lambda_w=\lambda_a=\lambda_m=0.14$; $\mu_w=0.17$; $\mu_a=0.145$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
-%% TODO think about this figure...
-%% TODO maybe better to show p_total / p_denovo as function of u or N.
-%}
-%\label{FractionPlot}
-%\end{figure}
-%
-%%%%%%%%%%%%
-%
-%\begin{figure}
-%\begin{subfigure}{0.5\textwidth}
-%\includegraphics[width=1\textwidth]{Figures/RatioEvolRescue.pdf}
-%\end{subfigure}
-%\begin{subfigure}{0.5\textwidth}
-%\includegraphics[width=1\textwidth]{Figures/RatioEvolRescuePopulationSize.pdf}
-%\end{subfigure}
-%\caption{\textbf{Effect of aneuploidy on evolutionary rescue.}
-%The ratio of rescue probability with and without aneuploid ($H$, \cref{ratiorescue}) increases with the aneuploid growth rate (colors) and decreases with the wildtype growth rates and initial tumor size (x-axes), except for large tumors where where the ratio converges to unity.
-%\textbf{(A, B)} Aneuploidy provides partial resistance.
-%\textbf{(C, D)} Aneuploidy provides tolerance.  
-%Solid and dashed lines apply $p_{rescue}$ from the exact formula of  (\cref{eq:survival_prob} in \cref{eq:rescue_prob}); dashed lines apply $p_{rescue}$ from the approximate formula (\cref{rescue_prob_approx}), with good agreement.
-%Parameters: $N=10^4$; $\lambda_w=\lambda_a=\lambda_m=0.14$; (B) $\mu_w=0.17$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
-%% TODO start y at 1
-%}
-%\label{rescue_ratio}
-%\end{figure}
-%
-%%%%%%%%%%%%
-
-%%%%%%%%%%%%
-%
-%\begin{figure}
-% \vspace*{1\baselineskip}
-%\includegraphics[width=1\textwidth]{Figures/MeanTimeRatioInitialPopulationSize.pdf}
-%\caption{\textbf{Ratio of evolutionary rescue time with and without aneuploidy.}
-%The ratio of the mean time to appearance of a resistance mutation that leads to evolutionary rescue with ($u>0$) and without ($u=0$) aneuploidy for variable initial tumor sizes (\cref{eq:eqMeanTimeRatioInitialPopulationSize}) when aneuploidy provides tolerance to the drug ($\Delta_a < 0$).
-When the initial tumor size is not large ($<10^8$), aneuploidy can decrease the rescue time by 10-100-fold.
-%% TODO parameter values 
-%}
-%\label{MeanTimeRatioInitialPopulationSize}
-%\end{figure}
-%
-%%%%%%%%%%%%
-
-%%%%%%%%%%
-%
-%\begin{figure}
-%% diffusion approx instead of branching process
-%% tolerance
-% \vspace*{1\baselineskip}
-%\includegraphics[width=1\textwidth]{Figures/P_est_diffusion.pdf}
-%\caption{Plot of the survival probability of an initial population consisting of $w_0=10^{4}$ wildtype cells as a function of $\Delta_a=\lambda_a-\mu_a$ for various values of $\Delta_w=\lambda_a-\mu_a$.
-The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the Feller diffusion approximation \eqref{FellerApprox}.
-The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/w_0}$ where $p$ is the mean probability of evolutionary rescue.}
-%\label{P_est_diffusion}
-%\end{figure}
-%
-%%%%%%%%%%%
-%
-%\begin{figure}
-%% partial resistance, compare 1st and 2nd order approximation and exact
-%\vspace*{1\baselineskip}
-%\includegraphics[width=1\textwidth]{Figures/SurvPlotNData.pdf}
-%\caption{Plot of the probability of survival of a population as a function of the initial population size of wildtype cells.
-The blue line represents the exact solution \eqref{survprobw}, the orange line line represents the approximation \eqref{aneuploidyresistentfirstapprox}, the green line represents the first order correction \eqref{survprobwapproxcorrected} and the red dots represents stochastic simulations.
-For the simulations we have chosen the following parameters: $\lambda_w=0.14, \lambda_a=0.14,\lambda_m=0.14,\mu_w=0.17,\mu_a=0.135,\mu_m=0.13$.
-The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n=100$ is the number of simulations.}
-%\label{compare_1st_2nd_order_approx}
-%\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{figure}
+\vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/SurvPlotNDataLogisticK.pdf}
+\caption{\textbf{Density dependent growth does not affect the accuracy of our model.} Comparison of results of simulations  with density-dependent growth (red markers with with 95\% CI) and the approximation formula (black line, \cref{eq:N_a} in \cref{eq:rescue_prob}) with maximum carrying capacity $K=10^8$ and effective carrying capacity $K_e=K\Delta_a/\lambda_a\approx10^6$.
+The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the  probability that the tumor has adapted to the stress and $n=100$ is the number of simulations.
+Parameters: $\lambda_w=0.1,\lambda_a=0.0901,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}, K=10^8$.}
+\label{LogisticPlot}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{figure}
+\vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/ProliferationTime.pdf}
+\caption{\textbf{Mean recurrence time is a decreasing function of the initial tumor size.}
+Shown is the mean time for the mutant cell population to reach size $N$, where $N$ is the initial number of cancer cells.
+Our inhomogeneous Poisson-process approximation (solid black line, \cref{meanproliferationtime}) is in agreement with simulation results (red markers with 95\% quantile intervals) for intermediary $N$.
+The numerical simulations converge to the solution of \cref{eq:t2det} (blue dashed line) for large values of $N$.
+Parameters: $\lambda_w=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}
+\label{proliferationFigure}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{figure}
+\vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/ProliferationTimeCDFN.pdf}
+\caption{Plot of the probability that a mutant cancer cell population has not reached size $N$ at time $t$.
+The green line represents the case where $N=10^6$, the red line represents $N=10^7$ and the blue line represents the case where $N=10^{10}$.
+Increasing the initial tumor size guarantees that the tumor recurrence.
+The markers represent stochastic simulations and the error bars represents $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the probability of that a mutant population size has not reached $N$ and $n=100$ is the number of simulations.
+Parameters: $\lambda_w=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}
+\label{KaplanMeierfig}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 Let $p_w$, $p_a$, and $p_m$ be the survival probabilities of a population consisting initially of single wildtype cell, aneuploid cell, or mutant cell, respectively.
-The complements $1-p_w$, $1-p_a$, and $1-p_m$ are the extinction probabilities, which satisfy each its respective equation,
+The complements $1-p_w$, $1-p_a$, and $1-p_m$ are the extinction probabilities, which satisfy each its respective equation~\citep{harris1963theory},
 \begin{equation} \label{eq:extinction_prob}
 \begin{aligned}
-1-p_w = &\frac{\mu_w}{\lambda_w+\mu_w+u+v} + 
-		  \frac{u}{\lambda_w+\mu_w+u+v}\left(1-p_a\right) + \\
-		  & \frac{\lambda_w}{\lambda_w+\mu_w+u+v}\left(1-p_w\right)^2 +
-		  \frac{v}{\lambda_w+\mu_w+u+v}\left(1-p_m\right) ,\\
-1-p_a = &\frac{\mu_a}{\lambda_a+\mu_a+v}+\frac{v}{\lambda_a+\mu_a+v}\left(1-p_m\right)+\frac{\lambda_a}{\lambda_a+\mu_a+v}\left(1-p_a\right)^2 ,\\
+1-p_w = &\frac{\mu_w}{\lambda_w+\mu_w+u\lambda_w+v\lambda_w} + 
+		  \frac{u\lambda_w}{\lambda_w+\mu_w+u\lambda_w+v\lambda_w}\left(1-p_a\right)\left(1-p_w\right) + \\
+		  & \frac{\lambda_w}{\lambda_w+\mu_w+u\lambda_w+v\lambda_w}\left(1-p_w\right)^2 +
+		  \frac{v\lambda_w}{\lambda_w+\mu_w+u\lambda_w+v\lambda_w}\left(1-p_m\right)\left(1-p_w\right) ,\\
+1-p_a = &\frac{\mu_a}{\lambda_a+\mu_a+v\lambda_a}+\frac{v\lambda_a}{\lambda_a+\mu_a+v\lambda_a}\left(1-p_m\right)\left(1-p_a\right)+\frac{\lambda_a}{\lambda_a+\mu_a+v\lambda_a}\left(1-p_a\right)^2 ,\\
 1-p_m = &\frac{\mu_m}{\lambda_m+\mu_m}+\frac{\lambda_m}{\lambda_m+\mu_m}\left(1-p_m\right)^2 .	 
@@
 \begin{aligned}
-p_w &= \frac{\lambda_w-\mu_w-u-v+\sqrt{\left(\lambda_w-\mu_w-u-v\right)^2+4\lambda_w\left(up_a+vp_m\right)}}{2\lambda_w} ,\\
-p_a &= \frac{\lambda_a-\mu_a-v+\sqrt{\left(\lambda_a-\mu_a-v\right)^2+4\lambda_avp_m}}{2\lambda_a}, \\
+p_w &= \frac{\lambda_w-\mu_w-u\lambda_wp_a-v\lambda_wp_m+\sqrt{\left(\lambda_w-\mu_w-u\lambda_wp_a-v\lambda_wp_m\right)^2+4\lambda_w^2\left(up_a+vp_m\right)}}{2\lambda_w} ,\\
+p_a &= \frac{\lambda_a-\mu_a-v\lambda_ap_m+\sqrt{\left(\lambda_a-\mu_a-v\lambda_ap_m\right)^2+4\lambda_a^2vp_m}}{2\lambda_a}, \\
 p_m &= \frac{\lambda_m-\mu_m}{\lambda_m} .
@@
 
-We first assume that aneuploidy provides partial resistance to drug therapy, $\lambda_a>\mu_a$, and that this resistance is significant, $\left(\lambda_a-\mu_a-v\right)^2 > 4\lambda_a v p_m$.
+We first assume that aneuploidy provides partial resistance to drug therapy, $\lambda_a>\mu_a$, and that this resistance is significant, $\left(\lambda_a-\mu_a-v\lambda_ap_m\right)^2 > 4\lambda_a^2 v p_m$.
 We thus rewrite \cref{eq:survival_prob} as
 \begin{align*}
-p_w&=\frac{\lambda_w-\mu_w-u-v}{2\lambda_w}\left(1-\sqrt{1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}}\right) ,
+p_w&=\frac{\lambda_w-\mu_w-u\lambda_wp_a-v\lambda_wp_m}{2\lambda_w}\left(1-\sqrt{1+\frac{4\lambda_w^2\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u\lambda_wp_a-v\lambda_wp_m\right)^2}}\right) ,
 \text{and} \\
-p_a&=\frac{\lambda_a-\mu_a-v}{2\lambda_a}\left(1+\sqrt{1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}}\right) . 
+p_a&=\frac{\lambda_a-\mu_a-v\lambda_ap_m}{2\lambda_a}\left(1+\sqrt{1+\frac{4\lambda_a^2vp_m}{\left(\lambda_a-\mu_a-v\lambda_ap_m\right)^2}}\right) . 
 \end{align*}
@@
 p_w 
-&\approx -\frac{vp_m+up_a}{\lambda_w-\mu_w-u-v}\\
+&\approx -\frac{v\lambda_wp_m+u\lambda_wp_a}{\lambda_w-\mu_w-u\lambda_wp_a-v\lambda_wp_m}\\
 \nonumber
-%&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{v\left(\lambda_a-\mu_a-u\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-u\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\ \label{survprobw2}
-&\approx-\frac{1}{\lambda_w-\mu_w}\left[\frac{u\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\
+&\approx-\frac{1}{\lambda_w-\mu_w}\left[\frac{u\lambda_w\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\lambda_w\lambda_a\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\lambda_w\left(\lambda_m-\mu_m\right)}{\lambda_m}\right].
 \end{align}
-Now $u v$ is very small, and if we assume $v \ll u$, we have
-\begin{equation}
-p_w \approx \frac{u}{\abs{\Delta_w}} \cdot \frac{\Delta_a}{\lambda_a} .
-\end{equation}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\paragraph{Second-order approximation.} 
-To improve our approximation, we can consider the second term of the Taylor series expansion,
+Now $u v$ is very small, and if we use the fact that $v \ll u$, we have:
+\begin{equation}\label{eq:pw_parttolerant}
+p_w \approx \frac{u\lambda_w}{\abs{\Delta_w}} \cdot \frac{\Delta_a}{\lambda_a} .
+\end{equation}
+However, if aneuploidy is very rare such that
 \begin{align*}
-\left(1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}\right)^{\frac{1}{2}}=1+\frac{2\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}-\frac{\left(\lambda_avp_m\right)^2}{4\left(\lambda_a-\mu_a-v\right)^4}+\cdots,
+\frac{u\lambda_w\Delta_a}{\lambda_a}<\frac{v\lambda_w\Delta_m}{\lambda_m}\Rightarrow u\lambda_a<\frac{v\lambda_a^2\Delta_m}{\lambda_m}\cdot\frac{1}{\Delta_a}<\frac{v\lambda_a^2\Delta_m}{\lambda_m}\cdot\frac{1}{\sqrt{4\lambda_a^2 v p_m}}\Rightarrow u\lambda_a<T^*,
 \end{align*}
-which gives us the following approximation,
-\begin{align}
-p_a \approx \frac{\lambda_a-\mu_a-v}{\lambda_a}+\frac{vp_m}{\lambda_a-\mu_a-v}-\frac{\lambda_a\left(vp_m\right)^2}{8\left(\lambda_a-\mu_a-v\right)^3} .
-\end{align}
-We therefore have
-\begin{align}\nonumber
-p_w&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{u\left(\lambda_a-\mu_a-v\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-v\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}-\frac{uv^2\lambda_a\left(\lambda_m-\mu_m\right)^2}{8\lambda_m^2\left(\lambda_a-\mu_a-v\right)^3}\right]\\ \label{survprobw3}
-&\approx-\frac{1}{\lambda_w-\mu_w}\left[\frac{u\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}-\frac{uv^2\lambda_a\left(\lambda_m-\mu_m\right)^2}{8\lambda_m^2\left(\lambda_a-\mu_a\right)^3}\right],
-\end{align}
-and using $\Delta_k=\lambda_k-\mu_k$, we can write the above equation as
-\begin{equation}\label{survprobwapproxcorrected}
-p_w \approx -\frac{1}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}-\frac{uv^2\lambda_a\Delta_m^2}{8\lambda_m^2\Delta_a^3}\right).
-\end{equation}
-
+where $T^* = (4 v \lambda_a^2 \Delta_m/\lambda_m)^{-1/2}$ and in the second inequality we used the fact that $\Delta_a^2 > 4\lambda_a^2 v p_m$. In this case adaptation is through direct mutation and:
+\begin{equation*}
+p_w \approx \frac{v\lambda_w}{\abs{\Delta_w}} \cdot \frac{\Delta_m}{\lambda_m} .
+\end{equation*}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 We now assume that aneuploidy provides tolerance to drug therapy, that is, the number of aneuploid cells significantly declines over time, but at a lower rate than the number of wildtype cells, $\lambda_w - \mu_w < \lambda_a - \mu_a < 0$.
-We also assume that the decline are significant, $\left(\lambda_a-\mu_a-v\right)^2 > 4\lambda_a v p_m$.
+We also assume that the decline are significant, $\left(\lambda_a-\mu_a-v\lambda_ap_m\right)^2 > 4\lambda_a^2 v p_m$.
 We rewrite \cref{eq:survival_prob} as
@@
 \begin{aligned}
-p_w&=\frac{\lambda_w-\mu_w-u-v}{2\lambda_w}\left(1-\sqrt{1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}}\right) ,
-\text{and} \\
-p_a&=\frac{\lambda_a-\mu_a-v}{2\lambda_a}\left(1-\sqrt{1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}}\right) .
+p_w&=\frac{\lambda_w-\mu_w-u\lambda_wp_a-v\lambda_wp_m}{2\lambda_w}\left(1-\sqrt{1+\frac{4\lambda_w^2\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u\lambda_wp_a-v\lambda_wp_m\right)^2}}\right), \\
+p_a&=\frac{\lambda_a-\mu_a-v\lambda_ap_m}{2\lambda_a}\left(1-\sqrt{1+\frac{4\lambda_a^2vp_m}{\left(\lambda_a-\mu_a-v\lambda_ap_m\right)^2}}\right) .
 \end{aligned}
@@
 \begin{aligned}
-p_w&\approx-\frac{vp_m+up_a}{\lambda_w-\mu_w-u-v}\\
-&\approx\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-v\right)}-\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\ %\label{survprobw2}
-&\approx\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_w-\mu_w\right)}\left[\frac{u}{\left(\lambda_a-\mu_a\right)}-1\right] \\
-&=\frac{v\Delta_m}{\lambda_m \abs{\Delta_w}}\left(\frac{u}{\abs{\Delta_a}}+1\right) .
+p_w&\approx-\frac{v\lambda_wp_m+u\lambda_wp_a}{\lambda_w-\mu_w-u\lambda_wp_a-v\lambda_wp_m}\\
+&\approx\frac{1}{\lambda_w-\mu_w-u\lambda_wp_a-v\lambda_wp_m}\left[\frac{uv\lambda_w\lambda_a\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-v\lambda_a\right)}-\frac{v\lambda_w\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\ %\label{survprobw2}
+&\approx\frac{v\lambda_w\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_w-\mu_w\right)}\left[\frac{u\lambda_a}{\left(\lambda_a-\mu_a\right)}-1\right] \\
+&=\frac{v\lambda_w\Delta_m}{\lambda_m \abs{\Delta_w}}\left(\frac{u\lambda_a}{\abs{\Delta_a}}+1\right) .
 \end{aligned}
 \end{equation}
-If we assume that $u>\abs{\Delta_a}$ then we have:
-\begin{equation*}
-p_w\approx\frac{u}{\abs{\Delta_w}}\cdot\frac{v\Delta_m}{\lambda_m \abs{\Delta_a}}.
-\end{equation*}
+If we assume that $u\lambda_a>\abs{\Delta_a}$ then we have:
+\begin{equation}\label{eq:pw_tolerant}
+p_w\approx\frac{u\lambda_w}{\abs{\Delta_w}}\cdot\frac{v\lambda_a}{\abs{\Delta_a}}\cdot\frac{\Delta_m}{\lambda_m}.
+\end{equation}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{Scenario 3: Aneuploid cells are non-growing} % TODO consider this name ok
-
-We now assume that the growth rate of aneuploid cells is close to zero (either positive or negative), such that  $\left(\lambda_a-\mu_a-v\right)^2 < 4\lambda_avp_m$.
+\subsubsection*{Scenario 3: Aneuploid cells are stationary} % TODO consider this name ok
+
+We now assume that the growth rate of aneuploid cells is close to zero (either positive or negative), such that  $\left(\Delta_a-v\lambda_ap_m\right)^2 \ll 4\lambda_a^2vp_m$.
 We rewrite \cref{eq:survival_prob} as
 \begin{equation}
-p_a = \frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}\left(1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\lambda_avp_m}\right)^{\frac12}}{2\lambda_a} .
-\end{equation}
-Using a following Taylor series expansion for small $\left(\lambda_a-\mu_a-v\right)^2 / 4\lambda_avp_m$,
+p_a = \frac{\lambda_a-\mu_a-v\lambda_ap_m+2\sqrt{\lambda_a^2 vp_m}\left(1+\frac{\left(\lambda_a-\mu_a-v\lambda_ap_m\right)^2}{4\lambda_a^2vp_m}\right)^{\frac12}}{2\lambda_a} .
+\end{equation}
+Using a following Taylor series expansion for small $\left(\lambda_a-\mu_a-v\lambda_ap_m\right)^2 / 4\lambda_a^2vp_m$,
 \begin{equation*}
-\left(1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\lambda_avp_m}\right)^{\frac{1}{2}}=1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{8\lambda_avp_m}+\cdots,
+\left(1+\frac{\left(\lambda_a-\mu_a-v\lambda_ap_m\right)^2}{4\lambda_a^2vp_m}\right)^{\frac{1}{2}}=1+\frac{\left(\lambda_a-\mu_a-v\lambda_ap_m\right)^2}{8\lambda_a^2vp_m}+\cdots,
 \end{equation*}
@@
 \begin{aligned}
-p_a&\approx\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}\left[1+\frac{\left(\lambda_a-\mu_a-v\right)^2}{8\lambda_avp_m}\right]}{2\lambda_a}\\
-&=\frac{\lambda_a-\mu_a-v+2\sqrt{\lambda_a vp_m}+\frac{\left(\lambda_a-\mu_a-v\right)^2}{4\sqrt{\lambda_avp_m}}}{2\lambda_a}\\
-&=\frac{\left(\lambda_a-\mu_a-v+2\sqrt{\lambda_avp_m}\right)^2+4\lambda_avp_m}{8\lambda_a\sqrt{\lambda_avp_m}}\\
-&=\frac{4\lambda_avp_m+4\lambda_avp_m\left(1+\frac{\lambda_a-\mu_a-v}{2\sqrt{\lambda_avp_m}}\right)^2}{8\lambda_a\sqrt{\lambda_avp_m}}\\
-&=\frac{1}{2\lambda_a}\left(\lambda_a-\mu_a-v+2\sqrt{\lambda_avp_m}\right).
+p_a&\approx\frac{\lambda_a-\mu_a-v\lambda_ap_m+2\sqrt{\lambda_a^2 vp_m}\left[1+\frac{\left(\lambda_a-\mu_a-v\lambda_ap_m\right)^2}{8\lambda_a^2vp_m}\right]}{2\lambda_a}\\
+&=\frac{\lambda_a-\mu_a-v\lambda_ap_m+2\sqrt{\lambda_a^2 vp_m}+\frac{\left(\lambda_a-\mu_a-v\lambda_ap_m\right)^2}{4\sqrt{\lambda_a^2vp_m}}}{2\lambda_a}\\
+&=\frac{\left(\lambda_a-\mu_a-v\lambda_ap_m+2\sqrt{\lambda_a^2vp_m}\right)^2+4\lambda_a^2vp_m}{8\lambda_a\sqrt{\lambda_a^2vp_m}}\\
+&=\frac{4\lambda_a^2vp_m+4\lambda_a^2vp_m\left(1+\frac{\lambda_a-\mu_a-v\lambda_ap_m}{2\sqrt{\lambda_a^2vp_m}}\right)^2}{8\lambda_a\sqrt{\lambda_a^2vp_m}}\\
+&=\frac{1}{2\lambda_a}\left(\lambda_a-\mu_a-v\lambda_ap_m+2\sqrt{\lambda_a^2vp_m}\right).
 \end{aligned}
 \end{equation}
-Plugging this in \cref{eq:survprobwinitial}, the survival probability of a population starting from one wildtype individual is
+Plugging this in \cref{eq:survprobwapprox1}, the survival probability of a population starting from one wildtype individual is
 \begin{equation}\label{eq:scenario3}
 \begin{aligned}
-p_w&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[v\frac{\lambda_m-\mu_m}{\lambda_m}+\frac{u}{2\lambda_a}\left(\lambda_a-\mu_a-v+2\sqrt{\lambda_avp_m}\right)\right]\\
-&=-\frac{1}{\lambda_w-\mu_w-u-v}\left[v\frac{\lambda_m-\mu_m}{\lambda_m}+\frac{u}{2\lambda_a}\left(\lambda_a-\mu_a-v\right)+u\sqrt{\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_a\lambda_m}}\right]\\
-&\approx-\frac{1}{\Delta_w}\left[v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right].
+p_w&\approx-\frac{1}{\lambda_w-\mu_w-u\lambda_wp_a-v\lambda_wp_m}\left[v\lambda_w\frac{\lambda_m-\mu_m}{\lambda_m}+\frac{u\lambda_w}{2\lambda_a}\left(\lambda_a-\mu_a-v\lambda_ap_m+2\sqrt{\lambda_a^2vp_m}\right)\right]\\
+&=-\frac{1}{\lambda_w-\mu_w-u\lambda_w-v\lambda_w}\left[v\lambda_w\frac{\lambda_m-\mu_m}{\lambda_m}+\frac{u\lambda_w}{2\lambda_a}\left(\lambda_a-\mu_a-v\lambda_ap_m\right)+u\lambda_w\sqrt{\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}}\right]\\
+&\approx-\frac{1}{\Delta_w}\left[v\lambda_w\frac{\Delta_m}{\lambda_m}+\frac{u\lambda_w\left(\Delta_a-v\lambda_a\right)}{2\lambda_a}+u\lambda_w\sqrt{\frac{v\Delta_m}{\lambda_m}}\right].
 \end{aligned}
@@
 \begin{equation*}
-\left(\Delta_a-v\right)^2 < 4\lambda_avp_m\Rightarrow\frac{\Delta_a-v}{2\lambda_a} < \sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}},
+\left(\Delta_a-v\lambda_ap_m\right)^2 \ll 4\lambda_a^2vp_m\Rightarrow\frac{\Delta_a-v\lambda_ap_m}{2\lambda_a} \ll \sqrt{\frac{v\lambda_a\Delta_m}{\lambda_m}},
 \end{equation*}
 and $v\ll u$ we obtain:
-\begin{equation}
-p_w\approx\frac{u}{\abs{\Delta_w}}\cdot\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}.
+\begin{equation}\label{eq:pw_partrest}
+p_w\approx\frac{u\lambda_w}{\abs{\Delta_w}}\cdot\sqrt{\frac{v\lambda_a\Delta_m}{\lambda_m}}.
 \end{equation}
@@
 \section*{Appendix B: Evolutionary rescue probability}\label{sec:appendix-rescue-prob}
-
-Substituting \cref{eq:survprobwapprox1,eq:survprobwinitial,eq:scenario3} into \cref{eq:rescue_prob}, the evolutionary rescue probability can be approximated by
+Using the fact that $\Delta_a-v\lambda_ap_m\approx\Delta_a$ we write the condition $\left(\Delta_a-v\lambda_ap_m\right)^2 \ll 4\lambda_a^2vp_m$ as:
+\begin{equation*}
+\Delta_a^2 \ll 4\lambda_a^2vp_m\Rightarrow -1\ll\Delta_aT^*\ll1,
+\end{equation*}
+where $T^* = (4 v \lambda_a^2 \Delta_m/\lambda_m)^{-1/2}$.
+Substituting \cref{eq:pw_parttolerant,eq:pw_partrest,eq:pw_tolerant} into \cref{eq:rescue_prob}, the evolutionary rescue probability can be approximated by
 \begin{equation}\label{rescue_prob_approx}
@@
   &\begin{cases}
-  1-\exp\left[\frac{N}{\Delta_w-u-v}\left(v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right)\right] ,&
-  4\lambda_avp_m>\left(\Delta_a-v\right)^2 ,\\
-   1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right] ,&
-   \Delta_a<0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2 ,\\
-   1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right] ,&
-   \Delta_a>0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2 .
+   1-\exp\left[-\frac{u\lambda_a}{\abs{\Delta_w}}\cdot\frac{v\lambda_w}{\abs{\Delta_a}}\cdot\frac{\Delta_m}{\lambda_m}\cdot N\right] ,&
+   \Delta_aT^*\ll-1 ,\\
+   1-\exp\left[-\frac{u\lambda_w}{\abs{\Delta_w}}\cdot\sqrt{\frac{v\lambda_a\Delta_m}{\lambda_m}}\cdot N\right] ,&
+  -1\ll\Delta_aT^*\ll1 ,\\
+   1-\exp\left[-\frac{u\lambda_w}{\abs{\Delta_w}} \cdot \frac{\Delta_a}{\lambda_a}\cdot N\right] ,&
+   1\ll\Delta_aT^*.
   \end{cases}
@@
 \end{equation}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-The number of successful mutants generated until time $t$ can be approximated by an inhomogeneous Poisson process with rate $R\left(t\right) = v p_m w_t$,
+The number of successful mutants generated until time $t$ can be approximated by an inhomogeneous Poisson process with rate $R_1\left(t\right) = v\lambda_w p_m w_t$,
 where $w_t=N\e^{\Delta_w t}$ is the number of wildtype cells at time $t$.
@@
 \begin{equation}\label{eq:integralR}
-\int_0^{t}{R(z)\d z} = 
-v p_m N \frac{\exp[{\Delta_w t}]-1}{\Delta_w} \approx 
-v p_m N t,
+\int_0^{t}{R_1(z)\d z} = 
+v\lambda_w p_m N \frac{\exp[{\Delta_w t}]-1}{\Delta_w} \approx 
+v\lambda_w p_m N t,
 \end{equation}
@@
 The probability density function of $T_1$ is thus
-$R\left(t\right)\exp\left(-\int_0^{t}{R(z)\d z}\right)$. % TODO ref
+$R_1\left(t\right)\exp\left(-\int_0^{t}{R_1(z)\d z}\right)$ \citep{allen2010introduction}. % TODO ref ok
 Therefore, the probability density function of the conditional random variable $(T_1 \mid T_1 < \infty)$ is
-$f_1(t) = \frac{R\left(t\right)\exp\left(-\int_0^{t}{R(z)\d z}\right)}{\presc}$. 
+$f_1(t) = \frac{R_1\left(t\right)\exp\left(-\int_0^{t}{R_1(z)\d z}\right)}{\presc}$. 
 \\
@@
 \int_{0}^{\infty}{t f_1(t) \d t} = 
-\frac{\int_{0}^{\infty}{tR(t)\exp\left(-\int_0^{t}{R(z)\d z}\right) \d t}}{p_{rescue}} = 
-\frac{\int_{0}^{\infty}{\exp\left(-\int_0^{t}{R(z)\d z}\right) \d t}}{p_{rescue}}
+\frac{\int_{0}^{\infty}{tR_1(t)\exp\left(-\int_0^{t}{R_1(z)\d z}\right) \d t}}{p_{rescue}},% = \int_{0}^{\infty}\left[1-\frac{1-\exp\left(-\int_0^{t}{R(z)\d z}\right) }{p_{rescue}}\right]\d t
 \end{aligned}
 \end{equation}
-after applying integration by parts.
+%after applying integration by parts and using the fact that $f_1(t)=-\frac{d}{dt}\left[1-F_1\left(t\right)\right]$, where $F_1\left(t\right)$ is the cumulative distribution function of $T_1$.
 Therefore, plugging \cref{eq:integralR,eq:rescue_prob} in \cref{eq:meantime1}, 
@@
 \tau_1 = 
-%\frac{\int_0^\infty\exp\left(-\int_0^\tau R(z)\d z\right)\d t}{1-\left(1-p_w\right)^N} = 
-\frac{\int_0^\infty\e^{-v N p_m\frac{\e^{\Delta_w t}-1}{\Delta_w}} \d t}{1-\left(1-p_w\right)^N} \approx
-\frac{\int_{0}^{\infty}{\exp\left(-v p_m N t\right) \d t}}{1-\e^{-N p_w}} \approx \\
-\left(1+\e^{-N p_w}\right)\int_0^\infty\e^{-v p_m N t} \d t =
-\frac{1+\e^{-N p_w}}{v p_m N} ,
+\int_0^\infty tv\lambda_w N\e^{\Delta_wt}\frac{\e^{-v\lambda_w N p_m\frac{\e^{\Delta_w t}-1}{\Delta_w}} }{1-\left(1-p_w\right)^N} \d t\approx
+\int_{0}^{\infty} tv\lambda_w N\e^{\Delta_wt}\frac{\e^{-v\lambda_w N p_mt} }{1-\e^{-Np_w}}\d t. 
 \end{align}
-where we use the approximations 
-$\frac{\e^{\Delta_w\tau}-1}{\Delta_w}=\frac{1+\Delta_w\tau+O(\tau^2)-1}{\Delta_w}=\tau+O(\tau^2)$ and $(1-\e^{-N p_w})^{-1} \approx 1+\e^{-N p_w}$ and integrate the exponent.
-\Cref{MeanTimeGrowthAneuploidyPlot}B show the agreement between this approximating and simulation results for intermediate and large tumor sizes.
+\Cref{MeanTimeGrowthAneuploidyPlot}B show the agreement between this approximating and simulation results.
 \\
 
-When $Nu\gg1$ the aneuploid frequency dynamics is roughly deterministic and therefore can be approximated by 
-\begin{equation}
-a_t \approx \frac{Nu\e^{\Delta_wt}}{\Delta_w-\Delta_a}\left[1-\e^{\left(\Delta_w-\Delta_a\right)t}\right].
-\end{equation}
-As a result, when $N\gg1$ % TODO N>>1 ? or Nu >> 1?
-the number of successful mutants created by direct mutation and via aneuploidy can be approximated by inhomogeneous Poisson processes with the rates
-\begin{align}
-r_1\left(t\right)&=vp_m\int_0^ta_{z} \d z = \frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_wt}-1}{\Delta_w}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right),\\
-r_2\left(t\right)&=vp_m\int_0^tw_{z} \d z = vNp_m\frac{\e^{\Delta_w t}-1}{\Delta_w}.
+When $Nu\lambda_w/\abs{\Delta_w}\gg1$ the aneuploid frequency dynamics is roughly deterministic and therefore can be approximated by 
+\begin{equation}\label{aneuploidpopeq}
+a_t \approx \frac{Nu\lambda_w\e^{\Delta_wt}}{\Delta_w-\Delta_a}\left[1-\e^{-\left(\Delta_w-\Delta_a\right)t}\right].
+\end{equation}
+As a result, the number of successful mutants created by direct mutation and via aneuploidy can be approximated by inhomogeneous Poisson processes with the rates
+\begin{align}\label{eq:twosteplineage}
+r_1\left(t\right)&=v\lambda_ap_m\int_0^ta_{z} \d z = \frac{uv\lambda_w\lambda_aNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_wt}-1}{\Delta_w}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right),\\ \label{eq:twosteplineagedirect}
+r_2\left(t\right)&=v\lambda_wp_m\int_0^tw_{z} \d z = v\lambda_wNp_m\frac{\e^{\Delta_w t}-1}{\Delta_w}.
 \end{align}
-For large initial population sizes we assume that the two processes are independent and as a result, they can be merged into a single Poisson process with rate $\left(r_1+r_2\right)\left(t\right)$.
+For large initial population sizes we assume that the two processes are independent and as a result, they can be merged into a single Poisson process with rate $R_2(t)=\left(r_1+r_2\right)\left(t\right)$.
 Consequently, the mean time to the appearance of the first rescue mutant is
-\begin{align}\label{meantimet2}
-\tau_2 = 
-\frac{\int_0^\infty\e^{-\left(r_1(t)+r_2(t)\right)} \d t}{1-\left(1-p_w\right)^N} = 
-\frac{\int_0^\infty\exp\left[-\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_w t}-1}{\Delta_w}-\frac{\e^{\Delta_a t}-1}{\Delta_a}\right)-vNp_m\frac{\e^{\Delta_w t}-1}{\Delta_w}\right] \d t}{1-\left(1-p_w\right)^N},
+\begin{align}\nonumber
+\tau_2 &= \frac{\int_{0}^{\infty}{tR_2(t)\exp\left(-\int_0^{t}{R_2(z)\d z}\right) \d t}}{p_{rescue}}\\ \label{meantimet2}
+&=
+\int_0^\infty t\left(v\lambda_ap_ma_t+v\lambda_wp_mw_t\right)\frac{\exp\left[-\frac{uv\lambda_w\lambda_aNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_w t}-1}{\Delta_w}-\frac{\e^{\Delta_a t}-1}{\Delta_a}\right)-v\lambda_wNp_m\frac{\e^{\Delta_w t}-1}{\Delta_w}\right] }{1-\e^{-Np_w}}\d t,
 \end{align}
 which we plot in \Cref{MeanTimeGrowthAneuploidyPlot}A as a function of the initial population size, $N$.
-
-%We wish to obtain a simpler formula for $\tau_2$, similar to \cref{eq:limitapprox}.
-%We thus have the following expansions,
-%\begin{align*}
-%\frac{\e^{\Delta_w\tau}-1}{\Delta_w}&=\frac{1+\Delta_w\tau+\frac{\Delta_w^2\tau^2}{2}+O(\tau^3)-1}{\Delta_w}=\tau+\frac{\Delta_w}{2}\tau^2+O(\tau^3) ,\\
-%\frac{\e^{\Delta_a\tau}-1}{\Delta_a}&=\frac{1+\Delta_a\tau+\frac{\Delta_a^2\tau^2}{2}+O(\tau^3)-1}{\Delta_a}=\tau+\frac{\Delta_w}{2}\tau^2+O(\tau^3) ,
-%\end{align*}
-%which allow us to write
-%\begin{equation*}
-%\frac{\e^{\Delta_w\tau}-1}{\Delta_w}-\frac{\e^{\Delta_a\tau}-1}{\Delta_a} \approx \frac{\left(\Delta_w-\Delta_a\right)\tau^2}{2}.
-%\end{equation*}
-%As a result, the integrand in \cref{meantimet2} can be written as:
-%\begin{align*}
-%&\exp\left[-\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_w\tau}-1}{\Delta_w}-\frac{\e^{\Delta_a\tau}-1}{\Delta_a}\right)-vNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}\right]\approx \\
-%& \exp\left(-uvNp_m\tau^2-vNp_m\tau\right) =
-%\exp\left(\frac{vNp_m}{2}\right)\exp\left[-\frac{uvNp_m}{2}\left(\tau+\frac{1}{u}\right)\right].
-%\end{align*}
-%Consequently, the mean conditional time $\tau_2$ is approximated by
-%\begin{align}\label{limitapprox2}
-%\tau_2 \approx
-%\frac{\left(1+\exp\left(-Np_w\right)\right) \exp\left(\frac{vNp_m}{2u}\right) \erfc\left(\sqrt{\frac{vNp_m}{2u}}\right)}{\sqrt{\frac{2uvNp_m}{\pi}}},
-%\end{align}
-%where $\erfc(z)$ is the complementary error function. 
-%
-%If we apply linear approximations,
-%\begin{align*}
-%\frac{\e^{\Delta_w t}-1}{\Delta_w}&=\frac{1+\Delta_w t+\mathcal{O}(t^2)-1}{\Delta_w} = t+O(t^2),\\
-%\frac{\e^{\Delta_a t}-1}{\Delta_a}&=\frac{1+\Delta_a t+\mathcal{O}(t^2)-1}{\Delta_a} = t+O(t^2),
-%\end{align*}
-%which we use to derive a first-order approximation for $\tau_2$,
-%\begin{align}\label{limitapprox3_appendix}
-%\tau_2 \approx
-%\left(1+\e^{-Np_w}\right)\int_0^\infty\e^{-u N p_m t} \d t = 
-%\frac{\left(1+\e^{-Np_w}\right)}{uNp_m} ,
-%\end{align}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Appendix D: Proliferation time}
-We define the proliferation time to be the time it takes the population of mutant cancer cells to reach the initial tumor size $N$.
-We distinguish between two cases.
-Firstly, when the time it takes for the mutant population to increase by a factor of $e$ is significantly smaller then the time it takes for a mutant cells, which rescues the population, to be generated.
-This is given by the condition:
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section*{Appendix D: Recurrence time}
+We define the proliferation time $\tau_2^r$  to be the time it takes the population of mutant cancer cells to reach the initial tumor size $N$.
+The number of rescue lineages generated by the wildtype population is given by  \cref{eq:twosteplineage} (see \Cref{ExpectedNumberRescueLineages}):
+\begin{equation*}
+r_1\left(\infty\right)=\frac{uv\lambda_w\lambda_aNp_m}{\abs{\Delta_w}\abs{\Delta_a}},
+\end{equation*}
+where we ignore lineages created by direct mutation because we assumed $u\lambda_a > \max{(-\Delta_a, 1/T^*)}$.
+
+This helps us distinguish between two cases for the proliferation time. Firstly, when we have at most one lineages which rescues the cancer cell population:
+\begin{equation*}
+N\ll\frac{\abs{\Delta_w}\abs{\Delta_a}}{uv\lambda_w\lambda_ap_m}.
+\end{equation*}
+As a result, the recurrence time is given by~\citep{avanzini2019cancer}:
+\begin{align}\label{meanproliferationtime}
+\tau_2^r&\approx\tau_2+\frac{\log p_mN}{\Delta_m}.
+\end{align}
+The factor of $p_m$ in the second term of \cref{meanproliferationtime} is due to the fact that the lineage is conditioned to survive genetic drift and the time to reach $N$ is shorter then the case without this property.
+
+The second case is when the wildtype population produces a large number of rescue lineages in a short period of time. This is given by the condition:
 \begin{align*}
-\frac{1}{\left(r_1+r_2\right)\left(\tau_2\right)}\gg\frac{1}{\Delta_m}.
+N\gg\frac{\abs{\Delta_w}\abs{\Delta_a}}{uv\lambda_w\lambda_ap_m}.
 \end{align*}
-As a result, the prolifereation time is given by~\citep{pompei2023fitness}:
-\begin{align}
-\tau_2'=\tau_2+\frac{\log \Delta_mN}{\Delta_m}.
+As a result, the recurrence time is obtained by solving the following system of ODEs:
+\begin{equation}\label{detODE}
+\begin{aligned}
+\frac{dw}{dt}&=\Delta_ww,\\
+\frac{da}{dt}&=\Delta_aa+u\lambda_ww,\\
+\frac{dm}{dt}&=\Delta_mm+v\lambda_aa+v\lambda_ww.
+\end{aligned}
+\end{equation}
+Solving the system of ODEs for initial condition $\left(w(0), a(0), m(0)\right)=\left(N,0,0\right)$ we obtain:
+\begin{align*}
+m\left(t\right)=\frac{Nuv\lambda_a\lambda_w}{\Delta_w-\Delta_a}\left[\frac{\e^{\Delta_wt}-\e^{\Delta_mt}}{\Delta_w-\Delta_m}-\frac{\e^{\Delta_at}-\e^{\Delta_mt}}{\Delta_a-\Delta_m}\right]+Nv\lambda_w\frac{\e^{\Delta_wt}-\e^{\Delta_mt}}{\Delta_w-\Delta_m}.
+\end{align*}
+We obtain $\tau_2^r$ such that $m\left(\tau_2^r\right)=N$ by solving:
+\begin{equation}\label{eq:t2det}
+1=\frac{uv\lambda_a\lambda_w}{\Delta_w-\Delta_a}\left[\frac{\e^{\Delta_w\tau_2^r}-\e^{\Delta_m\tau_2^r}}{\Delta_w-\Delta_m}-\frac{\e^{\Delta_a\tau_2^r}-\e^{\Delta_m\tau_2^r}}{\Delta_a-\Delta_m}\right]+v\lambda_w\frac{\e^{\Delta_w\tau_2^r}-\e^{\Delta_m\tau_2^r}}{\Delta_w-\Delta_m}.
+\end{equation}
+This is a transcendental equation which cannot be solved analytically but can solved numerically using one of the root solver algorithm in $\bold{Python}$. 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section*{Appendix E: Distribution of evolutionary rescue time}
+The probability that a successful mutant has been generated by time $t$ is given by:
+\begin{align*}
+P\left(rescue,t\right)&=P\left(T_2<t\right)\\
+&=1-\exp\left\{-\left[r_1\left(t\right)+r_2\left(t\right)\right]\right\}\\
+&=1-\exp\left\{-\left[\frac{uv\lambda_w\lambda_aNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_wt}-1}{\Delta_w}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right)+ v\lambda_wNp_m\frac{\e^{\Delta_w t}-1}{\Delta_w}\right]\right\},
+\end{align*}
+where $T_2$ is the time at which the first mutant cell appears that will avoid extinction and which was defined in Appendix C.
+
+As a result, the probability that a successful mutant has not been generated by time $t$ is:
+\begin{equation}
+1-P\left(rescue,t\right)=\exp\left\{-\left[\frac{uv\lambda_w\lambda_aNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_wt}-1}{\Delta_w}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right)+ v\lambda_wNp_m\frac{\e^{\Delta_w t}-1}{\Delta_w}\right]\right\}.
+\end{equation}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section*{Appendix F: Distribution of recurrence time}
+The probability distribution of the time that a lineage, consisting initially of a single cell, will reach size $N$ as time $t$ is given by the Gumbel distribution $\text{Gumb}_{max}\left(\frac{\log Np_m}{\Delta_m},\frac{1}{\Delta_m}\right)$~\citep{avanzini2019cancer} with probability density function:
+\begin{align*}
+G\left(t\right)=\e^{-p_mN\e^{-\Delta_mt}}.
+\end{align*}
+A mutant lineage initiated at time $s$, through aneuploidy, at rate $v\lambda_ap_ma_s$ reaches size $N$ before time $t$ with probability $G\left(t-s\right)$ where $s\leq t$.
+As a result, the number of successful mutant lineages which reach size $N$ by time $t$ can be approximated by inhomogeneous Poisson random variable with rate:
+\begin{align*}
+r\left(t\right)=v\lambda_ap_m\int_0^t a_sG\left(t-s\right)\,\d s
+\end{align*}
+where $a_t$ is aneuploid population size at time $t$ defined in \cref{aneuploidpopeq}.
+The proliferation time is defined as the first time the size of all lineages reaches $N$.
+When $N\ll\abs{\Delta_w}\abs{\Delta_a}/uv\lambda_w\lambda_ap_m$ there is at most a single mutant lineage that will survive and reach size $N$ (\Cref{ExpectedNumberRescueLineages}) and the probability that the size of that lineage has not reached $N$ by time $t$ is given by:
+\begin{align}\nonumber
+P\left(m_t\leq N\right)&=1-P\left(\tau>t\right)=\exp\left[-r\left(t\right)\right]\\ \label{eqmNs}
+&=\frac{Nuv\lambda_w\lambda_ap_m}{\Delta_w-\Delta_a}\int_0^t\left[\e^{\Delta_wt}-\e^{\Delta_at}\right]\e^{-p_mN\e^{-\Delta_m\left(t-s\right)}}\,\d s.
 \end{align}
-The second case is when the mutant population to increase by a factor of $e$ much faster then new mutants which rescue the population are generated. This is given by the condition:
-\begin{align*}
-\frac{1}{\left(r_1+r_2\right)\left(\tau_2\right)}\ll\frac{1}{\Delta_m}.
-\end{align*}
-As a result, the proliferation time is obtained by solving the following system of ODEs:
-\begin{align*}
-\frac{dw}{dt}&=\Delta_ww,\\
-\frac{da}{dt}&=\Delta_aa+uw,\\
-\frac{dm}{dt}&=\Delta_mm+v\left(w+a\right),
-\end{align*}
-for initial condition $\left(w(0), a(0), m(0)\right)=\left(N,0,0\right)$ and obtaining $\tau_2'$ from $m\left(\tau_2'\right)=N$. Consequently, we obtain the following approximation:
-\begin{equation}
-\tau_2'=\frac{\Delta_w-\Delta_m}{v\Delta_w}\left(1-\frac{u}{\Delta_a-\Delta_m}\right)^{-1},
-\end{equation}
-using the fact that $\e^{x}=1+x+\mathcal{O}\left(x^2\right)$.
-%
-%\begin{align}
-%\tau_2'=
-%   \begin{cases}
-%  \tau_2+\frac{\log \Delta_mN}{\Delta_m} ,&
-% \Delta_m\ll\left(r_1+r_2\right)\left(\tau_2\right) ,\\
-%   \frac{1}{\Delta_a}\log\frac{\left(\Delta_a-\Delta_m\right)\left(\Delta_a\Delta_m-uv\right)}{uv\Delta_w},&
-%    \Delta_m\gg\left(r_1+r_2\right)\left(\tau_2\right)
-%  \end{cases}
-%\end{align}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section*{Appendix E: Distribution of recurrence time}
-The probability that a succesful mutant has been generated by time $t$ is given by:
-\begin{align*}
-P\left(rescue,t\right)&=P\left(T_1<t\right)\\
-&=1-\exp\left\{-\left[r_1\left(t\right)+r_2\left(t\right)\right]\right\}\\
-&=1-\exp\left\{-\left[\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_wt}-1}{\Delta_w}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right)+ vNp_m\frac{\e^{\Delta_w t}-1}{\Delta_w}\right]\right\},
-\end{align*}
-where $T_1$ is the time at which the first mutant cell appears that will avoid extinction and which was defined in Appendix C.
-
-As a result, the probability that a succesful mutant has not been generated by time $t$ is:
-\begin{equation}
-1-P\left(rescue,t\right)=\exp\left\{-\left[\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_wt}-1}{\Delta_w}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right)+ vNp_m\frac{\e^{\Delta_w t}-1}{\Delta_w}\right]\right\}.
-\end{equation}
+When $N\gg\abs{\Delta_w}\abs{\Delta_a}/uv\lambda_w\lambda_ap_m$ the dynamics of the cancer cell populations is deterministic and approximated by the system of ODEs shown in \cref{detODE}.
+As a result, the size of the mutant cell population will always be below $N$ until time $\tau_2^r$ and will always be greater after:
+\begin{align}\label{eqmNd}
+P\left(m_t\leq N\right)=1-H\left(t-\tau_2^r\right),
+\end{align}
+where $H(x)$ is the Heaviside function:
+\begin{equation*}
+H\left(x\right) = \begin{cases}
+    0 ,&
+  x<0 ,\\ 
+  1 ,&
+  x\geq0 .
+  \end{cases}
+\end{equation*}
+We plot \cref{eqmNs} and \cref{eqmNd} in Figure and compare with stochastic simulations and observe that our approximation are in agreement.
+
+We observe that for $N=10^7$ our formula overestimates the probability that the mutant population will be smaller then $N$ at time $t$.
+This can be explained by the fact that $N=10^7$ is an intermediary case where the wildtype population produces a number of rescue lineages that is greater then one but still sufficiently small such that stochasticity plays an important role in the population dynamics.
+As a result, the number of mutant cancer cells will reach $N$ faster then the case with a single mutant lineage.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \newpage
+
 \section*{Appendix F: Figures}
@@
 A\\
-\includegraphics[width=1\textwidth]{Figures/MeanTimeDeleteriousPlot.pdf}
+\includegraphics[width=1\textwidth]{Figures/EvolutionaryRescueTime.pdf}
 \end{subfigure}
@@
 Shown is the mean time for appearance of a resistance mutation the leads to evolutionary rescue \textbf{(left)} with aneuploidy ($u>0$) and \textbf{(right)} without aneuploidy ($u=0$).
-Our inhomogeneous Poisson-process approximations (solid blue lines, right: \cref{eq:meantime1}, left: \cref{meantimet2}) is in agreement with simulation results (orange markers with 95\% CI). 
-%Our 1st-order (dashed red lines, right: \cref{eq:limitapprox}, left: \cref{limitapprox2}) and 2nd-order (green line, left: \cref{limitapprox3}) approximations work well when the initial tumor size is large (here $>10^8$ cells).
-Parameters: $\lambda_w=\lambda_a=\lambda_m=0.14$; $\mu_w=0.17$; (A) $\mu_a=0.145$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
+Our inhomogeneous Poisson-process approximations (solid black lines, right: \cref{eq:meantime1}, left: \cref{meantimet2}) is in agreement with simulation results (red markers with 95\% quantile intervals).
+Parameters: $\lambda_w=0.1,\lambda_m=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.
 }
@@
 \end{figure}
-
-\begin{figure}[!htb]
+%%%
+\begin{figure}
 \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/ThresholdPopulationSize.pdf}
-\caption{Minimum size that a tumor, which is exposed to chemotherapeutic drugs, reaches before adaptation sets in and tumor begins growing.
-}
-\label{MinTumorSize}
+\includegraphics[width=1\textwidth]{Figures/ExpectedNumber.pdf}
+\caption{\textbf{Aneuploidy increases the number of mutations which rescue the tumor.} Shown is the expected number of mutation, which will rescue the cancer cell population, produced through the evolutionary trajectory $wildtype \rightarrow mutant$ (blue line, \cref{eq:twosteplineagedirect}) or through the trajectory $wildtype \rightarrow aneuploid \rightarrow mutant$ (red line, \cref{eq:twosteplineage}).
+Parameters: $\lambda_w=0.1,\lambda_m=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}
+\label{ExpectedNumberRescueLineages}
 \end{figure}
 
-%\begin{figure}[!htb]
-%\begin{subfigure}{0.5\textwidth}
-%% non-growing slightly declining, compare 1st order approximation to exact and simulation
-%\includegraphics[width=1\textwidth]{Figures/DeleteriousTauLeapPlot.pdf}
-%\end{subfigure}
-%\begin{subfigure}{0.5\textwidth}
-%% density-dependent simulations, barely growing (or maybe partial resistance), with exact formula
-%\includegraphics[width=1\textwidth]{Figures/SurvPlotNDataLogisticK.pdf}
-%\end{subfigure}
-%\caption{\textbf{Evolutionary rescue probability for variable initial tumor size.}
-%\textbf{(A)} Comparison of simulation results (markers with 95\% CI, too small to appear with $10^5$ simulations per marker), the exact formula (blue line, \cref{eq:survival_prob} in \cref{eq:rescue_prob}) and the approximate formula (orange line, \cref{rescue_prob_approx}).
-%\textbf{(B)} Comparison of results of simulations  with density-dependent growth (markers with with 95\% CI) and the exact formula (blue line, \cref{eq:survival_prob} in \cref{eq:rescue_prob}) with maximum carrying capacity $K=10^9$.
-%Parameters: $\lambda_w=\lambda_a=\lambda_m=0.14$; $\mu_w=0.17$; (A) $\mu_a=0.15$, (B) $\mu_a=0.135$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
-%}
-%\label{rescue_prob_N}
-%\end{figure}
-
 \end{appendices}

### 2023-11-25 EvolutionaryRescue1.tex
Add files via upload

@@
 It is mathematically equivalent to the problem of crossing of fitness valley \citep{weissman2009rate,weissman2010rate}.
-There are three potential ways for a population to survive environmental change: migration to a new habitat similar to the one before the onset of environmental change \citep{harsch2014keeping,cobbold2020should,zhou2022range}; adaptation by phenotypic plasticity without genetic modification \citep{carja2019evolutionary,carja2017evolutionary,levien2021non}; and adaptation through genetic modifications, e.g., mutation \citep{uecker2014evolutionary,uecker2016role,uecker2011fixation}.
+There are three potential ways for a population to survive environmental change: migration to a new habitat similar to the one before the onset of environmental change \citep{harsch2014keeping,cobbold2020should,zhou2022range}; adaptation by phenotypic plasticity without genetic modification \citep{carja2019evolutionary,carja2017evolutionary,levien2021non}; and adaptation through genetic modifications, e.g., mutation \citep{uecker2014evolutionary,uecker2016role,uecker2011fixation,orr2014population}.
 
@@
 An exception was given by \citet{marrec2020adapt}, who modeled the fitness of the wildtype and mutant as time dependent.
+Evolutionary rescue in a quasi-periodic environment has been investigated by \citet{marrec2023evolutionary} with the observation that stochasticity increases the chance of extinction in comparison to the case with deterministic environmental fluctuations.
 Additionally, \citet{uecker2011fixation} investigated the probability of fixation of a beneficial mutation in a variable environment with arbitrary time-dependent selection coefficient and population size.
@@
 How multiple mutations contribute to the survival of the population is less explored, but \citet{wilson2017soft} have shown that evolutionary rescue is significantly enhanced by soft selective sweeps when multiple mutations contribute.
+Additionally, including a range of possible alleles with different fitness which could avert extinction has been coupled with Fischer's geometric model to analyze the probability of evolutionary rescue \citep{wahl2023evolutionary,martin2007distributions,osmond2020genetic}.
 Evolutionary rescue that requires two successive mutations has been investigated using diffusion approximation by \citet{martin2013probability}.

### 2023-12-04 EvolutionaryRescue1.tex
edited abstract

@@
 A prime example of evolutionary rescue is the ability of cancer to survive being exposed to various treatments.
-We are interested in the mechanisms through which a population of cancer cells are able to adapt to chemotherapy, and in particular, the role played by chromosomal instability (aneuploidy).
-Cancer cells which have aneuploidy are hypothesized to have a higher fitness in an environment altered by anti-cancer drugs as they have incomplete pathways which drugs activate in order to kill the cells.
-Aneuploidy is highly prevalent in tumors and certain drugs which attempt to combat cancers through increasing chromosomal instability.
-As a result, the question we wish to answer is how aneuploidy impacts the fate of the population of cancer cells.
-We propose to model evolutionary rescue with the help of multi-type branching processes to obtain the probability that cancer will survive.
-We obtain analytic expressions for the probability of evolutionary rescue as a function of the initial tumor size, the growth rates of the different cancer genotypes and observe that aneuploidy plays an important role in the survival of the cancer cell population when the initial tumor size is not too large and the effect of the anti-cancer drug on the aneuploid cells is not too severe.
+Chromosomal instability (aneuploidy) occurs frequently in cancer cells and introduces large genetics, suggesting that it may be an important mechanism by which populations of cancer cells to adapt to chemotherapy.
+To determine when this should be the case, we model a population of cancer cells exposed to chemotherapy that can achieve evolutionary rescue with or without aneuploidy.
+We use branching process approximations to obtain simple expressions for the probability of evolutionary rescue and the effect of aneuploidy.
+We find that even when aneuploidy by itself only confers partial drug resistance, it can still be an important stepping stone on the way to evolutionary rescue, allowing even small tumors to produce enough genetic diversity to avoid extinction.
 \end{abstract}

### 2023-12-04 EvolutionaryRescue1.tex
made comments throughout

@@
 We use branching process approximations to obtain simple expressions for the probability of evolutionary rescue and the effect of aneuploidy.
-We find that even when aneuploidy by itself only confers partial drug resistance, it can still be an important stepping stone on the way to evolutionary rescue, allowing even small tumors to produce enough genetic diversity to avoid extinction.
+We find that even when aneuploidy by itself only confers tolerance to the drug, i.e., is insufficient for rescue, it can still be an important stepping stone on the way to evolutionary rescue, allowing even small tumors to produce enough genetic diversity to avoid extinction.
 \end{abstract}
@@
 
+\daniel{RULE OF THUMB: EVERYTHING IN INTRO SHOULD LEAD TOWARD OUR QUESTION/WORK.}
+
 \paragraph{Aneuploidy in cancer.} Chromosomal instability (CIN) is the mitotic process in which cells suffer from chromosome mis-segregation that leads to aneuploidy, where cells are characterized by structural changes of the chromosomes and copy number alterations \citep{schukken2018cin}.
@@
 The role of chromosomal instability (CIN) in the emergence of cancer has been studied extensively in the past decades \citep{michor2005can,christine2018understanding,nowak2002role,pavelka2010dr,komarova2003mutation,zhu2018cellular}.
-One hypothesis is that CIN facilitates tumor genesis by accelerating the removal of tumor suppression genes (TSG) and subsequent appearance of cancer.
+\daniel{SO WHAT, WHO CARES?
+SHOULD BE ABOUT WHETHER CIN IS IMPORTANT IN CANCER.
+(AND ANSWER BETTER BE YES, OR IF ANSWER IS NO OUR HOOK SHOULD BE THAT MAYBE EVEN IF IT IS IMPORTANT IN GENERAL ITS IMPORTANT IN RESISTANCE.)}
+One hypothesis is that CIN facilitates tumorigenesis by accelerating the removal of tumor suppression genes (TSG) and subsequent appearance of cancer.
 The deletion of tumor suppression genes can happen in two ways: two point mutations deleting both alleles of the TSG (assuming a diploid genotype), or one point mutation and one chromosomal loss event.
 Initial theoretical studies have shown that aneuploidy can have a significant role in the deletion of the the tumor suppressing genes when compared to two consecutive point mutations \citep{nowak2002role,komarova2003mutation,michor2005can,komarova2008selective}.
-However, when taking into account that the appearance of aneuploidy requires a mutation to trigger CIN, the probability that CIN precedes tumor genesis is highly unlikely~\citep{komarova2004optimal}.
+However, when taking into account that the appearance of aneuploidy requires a mutation to trigger CIN, the probability that CIN precedes tumorigenesis is highly unlikely~\citep{komarova2004optimal}.
 
 \paragraph{Evolutionary rescue.} Populations adapted to a certain environment are vulnerable to environmental changes, which might cause extinction of the population.
-Examples of such environmental changes include climate change, invasive species or the onset of drug therapies.
+Examples of such environmental changes include climate change, invasive species, or the onset of drug therapies.
 Adaptation is a race against time as the population size decreases in the new environment~\citep{tanaka2022surviving}.
 \emph{Evolutionary rescue} is the process where the population acquires a trait that increases fitness in the new environment such that extinction is averted.
-It is mathematically equivalent to the problem of crossing of fitness valley \citep{weissman2009rate,weissman2010rate}.
+\daniel{It is mathematically equivalent to the problem of crossing fitness valleys \citep{weissman2009rate,weissman2010rate}.
+CUT?}
 There are three potential ways for a population to survive environmental change: migration to a new habitat similar to the one before the onset of environmental change \citep{harsch2014keeping,cobbold2020should,zhou2022range}; adaptation by phenotypic plasticity without genetic modification \citep{carja2019evolutionary,carja2017evolutionary,levien2021non}; and adaptation through genetic modifications, e.g., mutation \citep{uecker2014evolutionary,uecker2016role,uecker2011fixation,orr2014population}.
+\daniel{WHICH OF THESE ARE RELEVANT FOR CANCER? WHAT'S BEEN DONE IN THE CONTEXT OF CANCER? OR IS THERE EVEN ANY REASON WHY CANCER IS SPECIAL WHEN IT COMES TO EVOLUTIONARY RESCUE?}
 
@@
 Additionally, \citet{uecker2011fixation} investigated the probability of fixation of a beneficial mutation in a variable environment with arbitrary time-dependent selection coefficient and population size.
+\daniel{CUT ALL OF THE ABOVE IN THIS PAR?
+WE AREN'T REALLY DEALING WITH CHANGE IN TIME EITHER.}
 Most models focus on the probability that at least one mutation rescues the population.
 How multiple mutations contribute to the survival of the population is less explored, but \citet{wilson2017soft} have shown that evolutionary rescue is significantly enhanced by soft selective sweeps when multiple mutations contribute.
-Additionally, including a range of possible alleles with different fitness which could avert extinction has been coupled with Fischer's geometric model to analyze the probability of evolutionary rescue \citep{wahl2023evolutionary,martin2007distributions,osmond2020genetic}.
+Additionally, including a range of possible alleles with different fitness which could avert extinction has been coupled with Fisher's geometric model to analyze the probability of evolutionary rescue \citep{wahl2023evolutionary,martin2007distributions,osmond2020genetic}.
 Evolutionary rescue that requires two successive mutations has been investigated using diffusion approximation by \citet{martin2013probability}.
@@
 We assume the population of cells is under a strong stress, such as drug therapy, to which the wildtype genotype is susceptible or sensitive and therefore $\Delta_w<0$, whereas the mutant is resistant to the stress, $\Delta_m>0$.
-We analyze three scenarios: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_w$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing, stationary or "barely growing", that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$.
+We analyze three scenarios: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_w$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing, stationary or "barely growing", that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$, in a sense that we will make more precise below.
 We assume that both chromosomal missegregation and mutation occur during the process of mitosis.
@@
 Both aneuploid and wildtype cells may divide and mutate to become mutants at rates $v\lambda_{a}$ and $v\lambda_{w}$, respectively, which we assume is lower than the division rates, $v\cdot\max{(\lambda_w, \lambda_a)}<\min{(\lambda_w, \lambda_a, \lambda_m)}$.
-See \Cref{figureAneuploidy} for a schematic representation of the model and \Cref{sampleTrajectories} for sample trajectories of the different genotypes for $u=0 \,\textbf{(A)}, \Delta_a\ll0\, \textbf{(B)},\Delta_a\approx0\, \textbf{(C)},\Delta_a\gg0\, \textbf{(D)}$.
+See \Cref{figureAneuploidy} for a schematic representation of the model and \Cref{sampleTrajectories} for sample trajectories of the different genotypes for \textbf{(A)} $u=0$ (no aneuploidy), \textbf{(B)} $\Delta_a < 0$, \textbf{(C)} $\Delta_a\approx0$  and \textbf{(D)} $\Delta_a > 0$.
 
@@
 
+\daniel{SAY MORE ABOUT NATURAL VALUES OF THE PARAMETERS, TABLE}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-In our model, \emph{evolutionary rescue} occurs when resistant cells appear and fixate ($m_t \gg 1$) in the population before the population becomes extinct ($w_t=a_t=m_t=0$).
-Aneuploidy may contribute to evolutionary rescue by either preventing (when $\Delta_a>0$) or delaying (when $0>\Delta_a>\Delta_w$) the extinction of the population before mutant cells appear and fixate.
-We assume independence between clonal lineages starting from an initial population of $N$ wildtype cells (we check the effect of density-dependent growth on our results below).
+Aneuploidy may contribute to evolutionary rescue by either preventing population extinction directly (when aneuploidy confers resistance, $\Delta_a>0$) or by delaying extinction until resistant mutants can be establish (when $0>\Delta_a>\Delta_w$).
+\daniel{We assume independence between clonal lineages starting from an initial population of $N$ wildtype cells.
+NEED TO EXPLAIN WHY THIS IS GOOD APPROXIMATION, NOT ASSUMPTION.} (We check the effect of density-dependent growth on our results below).
 We therefore define $p_w$ as the probability that a lineage starting from a single wildtype cell avoids extinction by acquiring drug resistance.
-Thus, $N^*=1/p_w$ is the threshold tumor size above which evolutionary rescue is very likely, and the rescue probability is given by 
+The rescue probability of the whole population is then:
 \begin{equation} \label{eq:rescue_prob} 
@@
 1-\left(1-p_w\right)^N \approx
-1-\e^{-Np_w} = 
-1-e^{-N/N^*} .
-\end{equation}
-where the approximation $(1-p_w)\approx e^{-p_w}$ assumes that $p_w$ (but not necessarily $N p_w$) is small.
-Indeed, when $N<1/p_w$, then the probability for evolutionary rescue is $\presc \approx N p_w$  and when $N > 1/p_w$, it is $\presc \approx 1$, justifying the definition of $N^*$ as the threshold tumor size.
-\\
+1-\e^{-Np_w} \equiv 
+1-e^{-N/N^*},
+\end{equation}
+where $N^* \equiv 1/p_w$ is the threshold tumor size above which evolutionary rescue is very likely.
+Specifically, when $N \ll N^*$, the probability for evolutionary rescue is $\presc \approx N / N^* \ll 1$  and when $N > N^*$, it is $\presc \approx 1$.
+
 In the Appendix, we use the theory of multi-type branching processes to find approximate expressions \cref{eq:pw_parttolerant,eq:pw_partrest,eq:pw_tolerant} for $p_w$ in different regimes. 
@@
 For these approximations, an important quantity is $T^* = (4 v \lambda_a^2 \Delta_m/\lambda_m)^{-1/2}$, which is the critical time an aneuploid lineage needs to survive to produce a resistant mutant that avoids random extinction.
-First, if aneuploidy is very rare ($u\lambda_a T^*< 1$), or if aneuploidy is rare ($u\lambda_a < -\Delta_a$) and very sensitive to the drug ($\Delta_a T^* < -1$), then rescue will likely occur by a direct resistance mutation in a sensitive cell, such that
+First, if aneuploidy is very rare ($u\lambda_a T^*< 1$), or if aneuploidy is both rare ($u\lambda_a < -\Delta_a$) and very sensitive to the drug ($\Delta_a T^* < -1$), then rescue will likely occur by a direct resistance mutation in a sensitive cell, such that
 \begin{equation} \label{eq:N_m}
@@
 Otherwise, aneuploidy is frequent enough ($u\lambda_a > \max{(-\Delta_a, 1/T^*)}$) to affect the evolution of drug resistance. 
-The threshold tumor size, $N^*$, can then be approximated by one of the following cases, depending on $\Delta_a T^*$, the change in the log of the aneuploid population size during the critical time,
+The threshold tumor size, $N^*$, can then be approximated by one of the following cases, depending on $\Delta_a T^*$, the change in the log of the aneuploid population size during the critical time:
 \begin{equation}  \label{eq:N_a}
@@
 Comparing these approximations to results of stochastic evolutionary simulations, we find that the approximations perform very well (\Cref{rescue_denovo}).
-We derive $s$ by using the formula $s=\tilde{u}\lambda_w/f$ and obtaining the fraction of aneuploid cancer cells $f$ from \citep{lukow2021chromosomal} who mixed together wildtype and aneuploid cell at $50:50$ ratio, cultured them in drug-free environment and observed the ratio evolve as a function of time.
+\daniel{We derive $s$ by using the formula $s=\tilde{u}\lambda_w/f$ and obtaining the fraction of aneuploid cancer cells $f$ from \citep{lukow2021chromosomal} who mixed together wildtype and aneuploid cell at $50:50$ ratio, cultured them in drug-free environment and observed the ratio evolve as a function of time.
+SEEMS OUT OF PLACE, BELONGS IN MODEL SECTION.}
 
@@
 % TODO PLUG IN REAL VALUES ok
+\daniel{WHAT ABOUT ANEUPLOIDY VS NO-ANEUPLOIDY TAKING STANDING GENETIC VARIATION INTO ACCOUNT? THIS IS MOST IMPORTANT QUANTITY, RIGHT?}
 
@@
 However, when the expected number of resistant lineages is large the population of mutant cells behaves deterministically and the mean recurrence time cannot be separated into the mean evolutionary rescue time and mean proliferation time.
+\daniel{WHAT'S THE INTERESTING REGIME HERE? ALSO NEED TO MENTION THAT TIME IS EVEN A QUESTION EARLY IN THE MANUSCRIPT, LIKE IN THE INTRO}
 
@@
 \Cref{MeanTimeGrowthAneuploidyPlot} shows the agreement between these approximations and simulation results for intermediate and large tumor sizes.
+\daniel{MAKE NOTATION CONSISTENT WITH PREVIOUS SECTION. BE CLEARER ABOUT CONDITIONING}
 
@@
 
+\daniel{THIS WHOLE TIME SECTION SEEMS LIKE A SKETCH, NOT A DRAFT. WHAT IS THE MESSAGE?}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 We have modeled a tumor--a population of cancer cells--exposed to drug treatment that causes the population to decline in size towards potential extinction.
-In this scenario, the tumor can be "evolutionary rescued", or escape extinction, via two paths.
+In this scenario, the tumor can be ``evolutionarily rescued'', or escape extinction, via one of two paths.
 In the direct path, a sensitive cell acquires a mutation that confers resistance that allows it to rapidly grow.
-In the indirect path, a sensitive cell first becomes aneuploid, which diminishes the effect of the drug, and then an aneuploid cell acquires a mutation that confers resistance (\Cref{figureAneuploidy}).
+In the indirect path, a sensitive cell first becomes aneuploid, which diminishes the effect of the drug, and then a cell from that aneuploid lineage acquires a mutation that confers resistance (\Cref{figureAneuploidy}).
 
@@
 We obtained exact and approximate expressions for the probability of evolutionary rescue (\cref{eq:rescue_prob}). 
-Our results show that the probability of evolutionary rescue increases with the initial tumor size $N$, the sensitive growth rate $\Delta_w$, the mutation rate $v$, and the aneuploidy rate $u$.
+\daniel{Our results show that the probability of evolutionary rescue increases with the initial tumor size $N$, the sensitive growth rate $\Delta_w$, the mutation rate $v$, and the aneuploidy rate $u$.
+DIDN'T NEED A MODEL FOR ANY OF THAT. ANYTHING MORE INTERESTING?}
 
@@
 We find that aneuploidy can have a significant effect on evolutionary rescue (\Cref{rescue_prob}).
-Interestingly, aneuploidy is unlikely to contribute to evolutionary rescue in primary tumors in which the number of cells is large enough ($N>10^8$) for the appearance of resistant mutation directly in sensitive cells before these cells become extinct (\Cref{rescue_prob}).
+Aneuploidy is unlikely to contribute to evolutionary rescue in primary tumors in which the number of cells is large enough ($N>10^8$) for the appearance of resistant mutation directly in sensitive cells before these cells become extinct (\Cref{rescue_prob}).
 However, aneuploidy can have a crucial role in evolutionary rescue of secondary tumors, in which the number of sensitive cells may be below the detection threshold of $\sim10^7$  \citep{bozic2013evolutionary}.
-This can have an impact on the recurrence of cancer after the resection of the primary tumor  through secondary tumors which are too small to be detected and for which chemotherapy is employed to prevent  cancer relapse and are estimated to cause the majority of cancer-related deaths~\citep{chaffer2011perspective}.
+\daniel{This can have an impact on the recurrence of cancer after the resection of the primary tumor  through secondary tumors which are too small to be detected and for which chemotherapy is employed to prevent  cancer relapse and are estimated to cause the majority of cancer-related deaths~\citep{chaffer2011perspective}.
+THIS IS A GOOD POINT!}
 
@@
 The theoretical predictions for the mean rescue time for tumors smaller then  $10^8$ is greater then 4 years which is consistent with previous estimates of the recurrence time of tumors after resection~\citep{avanzini2019cancer}.
-
-Paradoxically, we observe from \Cref{MeanTimeGrowthAneuploidyPlot} that the mean time of a rescue mutation to appear is significatly shorter for the case when $u=0$ when compared to the case $u>0$, however this can be explained by the fact this mean time is conditioned on evolutionary rescue and, as a result, aneuploidy increase the \emph{window of opportunity} in which a rescue mutation could appear thus increasing the mean time as well (\Cref{sampleTrajectories}).
-
-We hypothesized that presence of \emph{standing variation} 
+\daniel{WHAT ABOUT DISTRIBUTIONS OF RECURRENCE TIME? CAN RAISE THE QUESTION EVEN IF WE DON'T HAVE ANY RESULTS}
+
+\daniel{Paradoxically, we observe from \Cref{MeanTimeGrowthAneuploidyPlot} that the mean time of a rescue mutation to appear is significantly shorter for the case when $u=0$ when compared to the case $u>0$, however this can be explained by the fact this mean time is conditioned on evolutionary rescue and, as a result, aneuploidy increase the \emph{window of opportunity} in which a rescue mutation could appear thus increasing the mean time as well (\Cref{sampleTrajectories}).
+BETTER TO JUST BE CLEAR AND AVOID PARADOXES IN THE FIRST PLACE}
+
+\daniel{We hypothesized that presence of \emph{standing variation} 
 --the existence of a subpopulation of aneuploid cancer cells before therapy begins--can facilitate evolutionary rescue by reducing the waiting time for the appearance of aneuploid cells.
 Indeed, we observe that even when a small fraction of the initial tumor is aneuploid, evolutionary rescue is more likely to occur through this existing standing variation, rather then through \emph{de novo} aneuploid cells (\Cref{rescue_denovo}).
-
-We observe from \Cref{KaplanMeierfig} that the probability of the mutant cell population reaching size $N$ is approximately zero before time $\tau_2^r$ which is the recurrence time for the deterministic case.
+IF RESCUE PRIMARILY OCCURS THROUGH SGV, THAT SHOULD BE THE CASE WE FOCUS ON, WITH DE NOVO AS THE LITTLE EXTRA SECTION}
+
+\daniel{We observe from \Cref{KaplanMeierfig} that the probability of the mutant cell population reaching size $N$ is approximately zero before time $\tau_2^r$ which is the recurrence time for the deterministic case.
 This can be explained as follows: in the deterministic case there is a sufficient number of lineages produced such that there exists a lineage where each descendant will only reproduce and not die; the time it takes for this lineage to reach $N$ is the lower bound for the time of all other lineages to reach $N$ and this time cannot be smaller then $\tau_2^r$ by definition.
 Given that for small values of $N$ we expect that at most a single lineage will rescue the tumor, this lineage cannot reach $N$ before $\tau_2^r$ for the deterministic case \cref{eq:t2det}.
+DOESN'T BELONG IN DISCUSSION.
+MAYBE IN RESULTS?}
 
@@
 Our model's predictions may be tested by experiments \citep{martin2013probability}.
-For example, to study the effects of initial tumor size on the probability of evolutionary rescue, a large culture mass can be propagated from a single cancer cell in permissive conditions and then diluted to a  range of starting tumor sizes.
+\daniel{DO WE REALLY NEED A REFERENCE FOR THIS STATEMENT?} For example, to study the effects of initial tumor size on the probability of evolutionary rescue, a large culture mass can be propagated from a single cancer cell in permissive conditions and then diluted to a  range of starting tumor sizes.
 Afterwards, these tumors may be exposed to anti-cancer drugs that induces aneuploidy % TODO ref OK
@@
 Ok
+\daniel{WHY DO THIS KIND OF WET SIMULATION? NO POINT. BETTER TO TRY TO EXTRACT MODEL PARAMETER VALUES FROM EXPERIMENTS AND THEN SEE IF MODEL PREDICTIONS FIT REAL PATIENT DATA}
 
@@
 
-We have assumed that cancer cell lineages are independent of each other.
+%We have assumed that cancer cell lineages are independent of each other.
 However, this may not be the case, as cancer cells compete for resources (e.g., blood supply).
@@
 ok
-our approximation for the probability of evolutionary rescue agrees with results of stochastic simulations with density-dependent growth  (\Cref{LogisticPlot}).
-Future work may focus on scenarios with small carrying capacity by analyzing density-dependent branching processes~\citep{klebaner1997population,harris1963theory}. % TODO ref? OK
-
-\paragraph{Conclusions}
+%our approximation for the probability of evolutionary rescue agrees with results of stochastic simulations with density-dependent growth  (\Cref{LogisticPlot}).
+%Future work may focus on scenarios with small carrying capacity by analyzing density-dependent branching processes~\citep{klebaner1997population,harris1963theory}. % TODO ref? OK
+% I THINK WE HAVE A GOOD ARGUMENT FOR WHY THIS ISN'T IMPORTANT
+
+\paragraph{Conclusions \daniel{WOULD USUALLY GO AT BEGINNING OF DISCUSSION}}
 
@@
 \bibliographystyle{agsm}
+\daniel{MANY PROBLEMS WITH REFERENCES}
 \bibliography{evo2022}
@@
 \caption{\textbf{Model parameters.} %NEED DIFFERENT REFS---THESE ARE MOSTLY THEORY PAPERS. CHECK BIONUMBERS? OK
+\daniel{WHY ARE SO MANY REFERENCES INCOMPLETE? WHY STILL CITING BOZIC FOR VALUES? (GOOD IF REFERENCE WAS FOUND IN BOZIC, BUT UNNECESSARY IF REFERENCE CAME FROM BIONUMBERS)}
 The * symbol in the Name column means that for those parameters the values have not been selected from a paper.
@@
 %%%%%%%%%%
+
+\clearpage
+
+\daniel{FIGURES 1,2, 4, 5 : TEXT AND LABELS NEED TO BE MUCH BIGGER.}
 
@@
 The aneuploid may be tolerant $(4)$, stationary $(3)$, partially resistant $(2)$ or fully resistant $(1)$.
+\daniel{BETTER TO PUT THESE WORDS RIGHT ON THE GRAPH, INCLUDE MATHEMATICAL VERSIONS (EG, $\Delta_A > 0$)}
 }
@@
 \textbf{Sample trajectories of the different genotypes as a function of time.}
-\textbf{(A)} Whe the missegregation rate $u=0$ evolutionary rescue is only possible through direct mutation and in most cases the tumour will be cured by the drug.
+\textbf{(A)} When the missegregation rate $u=0$ evolutionary rescue is only possible through direct mutation and in most cases the tumour will be cured by the drug.
 \textbf{(B)} When the aneuploidy growth rate $\Delta_a\ll0$ we observe similar dynamic to case \textbf{(A)} as direct mutation is the only viable route toward evolutionary rescue for the tumor.
@@
 \includegraphics[width=1\textwidth]{Figures/ReboundProbability.pdf}
-\caption{\textbf{Aneuploidy is a key diver of cancer adaptation.}
+\caption{\textbf{Aneuploidy is a key driver of cancer adaptation.} \daniel{NO, THIS IS MESSAGE OF EARLIER FIG.}
 Plot of the probability that a successful mutant has not appeared by time $t$.
@@
 \includegraphics[width=1\textwidth]{Figures/SurvPlotNDataLogisticK.pdf}
-\caption{\textbf{Density dependent growth does not affect the accuracy of our model.} Comparison of results of simulations  with density-dependent growth (red markers with with 95\% CI) and the approximation formula (black line, \cref{eq:N_a} in \cref{eq:rescue_prob}) with maximum carrying capacity $K=10^8$ and effective carrying capacity $K_e=K\Delta_a/\lambda_a\approx10^6$.
+\caption{\textbf{Density dependent growth does not affect the accuracy of our model.} \daniel{SUPPLEMENTARY FIG} Comparison of results of simulations  with density-dependent growth (red markers with with 95\% CI) and the approximation formula (black line, \cref{eq:N_a} in \cref{eq:rescue_prob}) with maximum carrying capacity $K=10^8$ and effective carrying capacity $K_e=K\Delta_a/\lambda_a\approx10^6$.
 The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the  probability that the tumor has adapted to the stress and $n=100$ is the number of simulations.
@@
 \caption{Plot of the probability that a mutant cancer cell population has not reached size $N$ at time $t$.
-The green line represents the case where $N=10^6$, the red line represents $N=10^7$ and the blue line represents the case where $N=10^{10}$.
+\daniel{WHY IS THIS SEPARATE FROM FIG 6?} The green line represents the case where $N=10^6$, the red line represents $N=10^7$ and the blue line represents the case where $N=10^{10}$.
 Increasing the initial tumor size guarantees that the tumor recurrence.

### 2023-12-31 EvolutionaryRescue1.tex
Add files via upload
Remus updates 31 December

--- 2023-11-25 EvolutionaryRescue1.tex
@@
 \usepackage{xspace}
+
+\allowdisplaybreaks
 
@@
 We propose to model evolutionary rescue with the help of multi-type branching processes to obtain the probability that cancer will survive.
-We obtain analytic expressions for the probability of evolutionary rescue as a function of the initial tumor size, the growth rates of the different cancer genotypes and observe that aneuploidy plays an important role in the survival of the cancer cell population when the initial tumor size is not too large and the effect of the anti-cancer drug on the aneuploid cells is not too severe.
+We obtain analytic expressions for the probability of evolutionary rescue as a function of the initial tumor size, the growth rates of the different cancer genotypes.
+Additionally we obtain the mean recurrence time for the tumor to reach its initial size.
+We observe that aneuploidy plays an important role in the relapse of secondary tumors who have not yet been detected and thus are smaller in size.
 \end{abstract}
@@
 
-\paragraph{Aneuploidy in cancer.} Chromosomal instability (CIN) is the mitotic process in which cells suffer from chromosome mis-segregation that leads to aneuploidy, where cells are characterized by structural changes of the chromosomes and copy number alterations \citep{schukken2018cin}.
+\paragraph{Aneuploidy in cancer.}  Each year approximately 10 million people die from cancer \citep{kocarnik2022cancer}, so understanding what factors contribute to failure of interventions is of great importance.
+One hypothesized factor is aneuploidy, where cells are characterized by structural changes of the chromosomes and copy number alterations \citep{schukken2018cin}, which is caused by chromosomal instability (CIN), the mitotic process in which cells suffer from chromosome mis-segregation that leads to aneuploidy.
 Interestingly, aberrations in chromosome copy number have been shown to allow cancer cells to survive under stressful conditions such as drug therapy \citep{lukow2021chromosomal,rutledge2016selective}.
-Indeed, cancer cells are often likely to be aneuploid, and aneuploidy is associated with poor patient outcomes \citep{ben2020context}.
-
-The role of chromosomal instability (CIN) in the emergence of cancer has been studied extensively in the past decades \citep{michor2005can,christine2018understanding,nowak2002role,pavelka2010dr,komarova2003mutation,zhu2018cellular}.
-One hypothesis is that CIN facilitates tumor genesis by accelerating the removal of tumor suppression genes (TSG) and subsequent appearance of cancer.
-The deletion of tumor suppression genes can happen in two ways: two point mutations deleting both alleles of the TSG (assuming a diploid genotype), or one point mutation and one chromosomal loss event.
-Initial theoretical studies have shown that aneuploidy can have a significant role in the deletion of the the tumor suppressing genes when compared to two consecutive point mutations \citep{nowak2002role,komarova2003mutation,michor2005can,komarova2008selective}.
-However, when taking into account that the appearance of aneuploidy requires a mutation to trigger CIN, the probability that CIN precedes tumor genesis is highly unlikely~\citep{komarova2004optimal}.
+Indeed, cancer cells are often likely to be aneuploid, and aneuploidy is associated with poor patient outcomes \citep{ben2020context,smith2018systematic}.
+
+ \citet{ippolito2021gene} induced aneuploidy in cancer cell lines by exposing them to reversine, a small-molecule inhibitor of the mitotic kinase Mpsi1, and then exposed to chemotherapeutic agents such as vemurafenib.
+When compared to wildtype cancer cell the reversine treated cells had higher proliferation in the environment altered by anti-cancer drugs.
+Similarly, \citet{lukow2021chromosomal} observed that cancer cells which have been induced to be aneuploid have an advantage when compare to wildtype cell even while they had lower fitness before the onset of chemotherapy.
+A proposed mechanism through which aneuploidy is able to confer resistance to chemotherapeutics is bu antagonizing cell division  which prevents the drugs from damaging DNA and microtubules \citep{replogle2020aneuploidy}
+
+ An important aspect of aneuploidy is the rate with which cell acquire it, which is several orders of magnitude higher when compare to mutation rate \citep{bakker2023predicting}.
+Consequently, a cell exposed to a stress such as chemotherapeutic drugs can acquire aneuploidy faster when compared to acquiring a mutation, especially when several proposed anti-cancer drugs elevate the rate of missegration in order to combat cancer \citep{lee2016effects}.
 
@@
 It is mathematically equivalent to the problem of crossing of fitness valley \citep{weissman2009rate,weissman2010rate}.
-There are three potential ways for a population to survive environmental change: migration to a new habitat similar to the one before the onset of environmental change \citep{harsch2014keeping,cobbold2020should,zhou2022range}; adaptation by phenotypic plasticity without genetic modification \citep{carja2019evolutionary,carja2017evolutionary,levien2021non}; and adaptation through genetic modifications, e.g., mutation \citep{uecker2014evolutionary,uecker2016role,uecker2011fixation,orr2014population}.
-
-Models of evolutionary rescue usually assume that the fitness of the wildtype and mutant are homogeneous in time.
-An exception was given by \citet{marrec2020adapt}, who modeled the fitness of the wildtype and mutant as time dependent.
-Evolutionary rescue in a quasi-periodic environment has been investigated by \citet{marrec2023evolutionary} with the observation that stochasticity increases the chance of extinction in comparison to the case with deterministic environmental fluctuations.
-Additionally, \citet{uecker2011fixation} investigated the probability of fixation of a beneficial mutation in a variable environment with arbitrary time-dependent selection coefficient and population size.
+There are three potential ways for a population to survive environmental change: migration to a new habitat similar to the one before the onset of environmental change \citep{harsch2014keeping,cobbold2020should,zhou2022range}; adaptation by phenotypic plasticity without genetic modification \citep{carja2019evolutionary,carja2017evolutionary,levien2021non,gunnarsson2020understanding}; and adaptation through genetic modifications, e.g., mutation \citep{gomulkiewicz1995does,uecker2014evolutionary,uecker2016role,uecker2011fixation,orr2014population}.
+
+\citet{gunnarsson2020understanding} analyze a model where a tumor consisting of two populations of cancer cells, one drug resistant and the other drug sensitive, is able to evade extinction by cells switching between the two phenotypes through epigenetic mutations.
+They found that even when the drug resistant type is barely viable the epimutations have the effect of guaranteeing evolutionary rescue.
+Evolutionary rescue in one step in which an initially declining population, after a sudden environmental change, has to acquire a mutation has been studied in the context of population genetics  by \citep{orr2008population,orr2014population}.
+They analyzed a model where the mutant strain is present in small number at the onset of therapy and concluded that this can significantly enhance the chance that the population will survive.
+
 Most models focus on the probability that at least one mutation rescues the population.
 How multiple mutations contribute to the survival of the population is less explored, but \citet{wilson2017soft} have shown that evolutionary rescue is significantly enhanced by soft selective sweeps when multiple mutations contribute.
-Additionally, including a range of possible alleles with different fitness which could avert extinction has been coupled with Fischer's geometric model to analyze the probability of evolutionary rescue \citep{wahl2023evolutionary,martin2007distributions,osmond2020genetic}.
 Evolutionary rescue that requires two successive mutations has been investigated using diffusion approximation by \citet{martin2013probability}.
+
+Here we build on this previous work on evolutionary rescue after a sudden environment change caused by the initiation of chemotherapy.
+We wish to understand what effect does aneuploidy have on probability of evolutionary rescue when it acts as an intermediary between the wildtype and the mutant cancer cells.
+Besides the probability of evolutionary rescue we calculate the calculate the mean time that an initially declining tumor cell population reaches the initial population size.
+Given that aneuploidy is present in many tumors even before the onset of therapy \citep{lukow2021chromosomal} we also take into consideration the effect that standing genetic variation has on the tumor dynamics.
 
@@
 Wildtype cells may divide and then missegregate to become aneuploids at rate $u\lambda_w$.
-Both aneuploid and wildtype cells may divide and mutate to become mutants at rates $v\lambda_{a}$ and $v\lambda_{w}$, respectively, which we assume is lower than the division rates, $v\cdot\max{(\lambda_w, \lambda_a)}<\min{(\lambda_w, \lambda_a, \lambda_m)}$.
+Both aneuploid and wildtype cells may divide and mutate to become mutants at rates $v\lambda_{a}$ and $v\lambda_{w}$, respectively.
 See \Cref{figureAneuploidy} for a schematic representation of the model and \Cref{sampleTrajectories} for sample trajectories of the different genotypes for $u=0 \,\textbf{(A)}, \Delta_a\ll0\, \textbf{(B)},\Delta_a\approx0\, \textbf{(C)},\Delta_a\gg0\, \textbf{(D)}$.
-
-%Thus, the changes in the number of each cell type is described by 
-%\begin{equation}
-%\begin{aligned}
-%w_t&\rightarrow w_t+1:\quad \lambda_ww_t,\\
-%w_t&\rightarrow w_t-1:\quad \left(\mu_w+u+v\right)w_t,\\
-%a_t&\rightarrow a_t+1:\quad \lambda_aa_t+uw_t,\\
-%a_t&\rightarrow a_t-1:\quad \left(\mu_a+v\right)a_t,\\
-%m_t&\rightarrow m_t+1:\quad \lambda_am_t+va_t+vm_t,\\
-%m_t&\rightarrow m_t-1:\quad \mu_am_t.
-%\end{aligned}
-%\end{equation}
-
+We assume that wildtype cells acquire aneuploidy before the onset of therapy with rate $\tilde{u}\lambda_w$ and that aneuploid cells have a fitness cost $s$ in the drug-free environment.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 The following describes the events that may occur (right column), the rates at which they occur (middle column), and the effect these events have on the state (left column, see \Cref{figureAneuploidy}):
+
 \begin{subequations}
@@
 
+\paragraph{Parameterization.}
+To parametrize the model we assume that the cells under consideration are melanoma cells and and rely on \citep{rew2000cell} and \citep{bozic2013evolutionary} for the division and death rates, respectively.
+\citet{rew2000cell} report in vivo measurements of the potential doubling times (which is the time which the tumor cells would take to double in number if cell loss was not present) for a large set of cancer types.
+The division rate is obtained as $\lambda=\log2/T\approx0.1\text{ days}^{-1}$ We select this to be the division rate for wildtype and mutant cells.
+%We select the mutant death rate $\mu_m=0.09\text{ days}^{-1}$ such that $\Delta_m>0$.
+
+\citet{bozic2013evolutionary} report the growth rate $\Delta_w$ for the wildtype melanoma cancer cells from which they deduce the death rate $\mu_w$ to in the interval $\left[0.11,0.17\right]$ and we choose $\mu_w=0.14days^{-1}$.
+Additionally, they observe the growth rate of cancer cells prior to treatment to be 0.01 which we select as the growth rate of mutant cells as we assume that the mutation makes the cancer cells immune to the chemotherapeutic agent.
+As a result, we choose $\mu_m=0.1-0.01=0.09\text{ days}^{-1}$ as the death rate for mutant cells.
+
+ Aneuploid death rate $\mu_a$ is set to the same value as the mutant death rates, $\mu_m=0.09\text{ days}^{-1}$, given that aneuploidy increases resistance to chemotherapeutics, such as cisplatin, by antagonizing cell division \citep{replogle2020aneuploidy}.
+Aneuploid birth rate is selected such that the aneuploid growth rate $\Delta_w\ll\Delta_a\ll\Delta_m$ which means that $\lambda_a$ belong to the interval $\left[0.06,0.1\right]$.
+For most of the paper we will use $\lambda_a=0.0899\text{ days}^{-1}$ such that aneuploidy can only act as a \emph{stepping stone} for the generation of the mutant which rescues the cancer cell population.
+
+We set the mutation rate $v$ to be equal to $10^{-7}$ per gene per cell division \citep{loeb2001mutator} and since we assume that a single target gene confers immunity to chemotherapy we select $v=10^{-7}$ per cell division.
+The missegregation rate is chosen from \citet{bakker2023predicting} who determined \textit{in silico} that it must be in the interval $10^{-3}-10^{-2}$ per chromosome per cell division.
+Given that only about one to two chromosome additions are ``beneficial'' for cancer resistance (see details next paragraph) we choose $u=10^{-2}$ per cell division.
+For the missegregation rate in the drug free environment $\tilde{u}$ we choose the lower-end value of $u$ (i.e.
+$\tilde{u}=10^{-3}$ per cell division) as the chemotherapeutic drugs we want to model increase the rate of aneuploidy~\citep{wang2019molecular,mason2017functional}.
+
+The fitness cost $s$ of aneuploidy before the onset of therapy is difficult to estimate as we are interested in a specific type of aneuploidy which improved the fitness of cancer cells in an environment altered by chemotherapeutic drugs.
+We derive $s$ by using the formula $s=\tilde{u}\lambda_w/f$ and obtaining the fraction of aneuploid cancer cells $f$ from \citep{lukow2021chromosomal} who mixed together wildtype and aneuploid A375 melanoma cells at $50:50$ ratio, cultured them in drug-free environment and observed the ratio evolve as a function of time with the aneuploid cells declining to $15\%$ after 24 days.
+\citet{ippolito2021gene} observed that trisomy 2 and 6 are most likely to confer increased resistance against the chemotherapeutic agent vemurafenib for A375 cells.
+We assume that if a cancer is aneuploid then it most likely has trisomy \citep{gisselsson2010generation} in the pre-treatment environment and that cells with more then one trisomy are very unlikely to survive.
+Additionally, we assume that all trisomies are equally likely and, as a result, we obtain the percentage of cancer cells with favorable aneuploidy as $2/22\times0.15\approx 0.0136$ (i.e.
+$1.36\%$ of pre-treatment cancer cells have the ``beneficial'' aneuploidy) which we select as the value for $f$ (this gives us a fitness cost of $s= 7.3\times10^{-3}\text{ days}^{-1}$).
+We note that when we refer to wildtype cancer cells we include those cells which have any aneuploidy except trisomy 2 and 6 as those are the aneuploid cells which are hypothesized to have higher fitness in the environment altered by therapeutics such as vemurafenib.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \paragraph{Density-dependent growth.}
@@
 In our analysis we assume that lineages produced by cells from the initial population divide and die independently of each other, which may be unrealistic, as cells usually compete for resources.
+This is a good approximation as we assume that the drug cause the wildtype population to decline rapidly and, as a result, density effect can be neglected.
 A more realistic model includes competition for limited resources and spatial structure, which may play an important role in the development of cancer \citep[e.g.,][]{martens2011spatial}.
@@
 \end{align*}
-where $K$ is the maximum carrying capacity.
-The effective carrying capacity of this model is $K_e=K\Delta_a/\lambda_a\approx10^6$ for $K=10^8, \lambda_a=0.0901,\mu_a=0.09$.
-% TODO what constants?
-see Remus version.
-Ok
+where $K$ is the maximum carrying capacity. The effective carrying capacity of this model is $K_e=K\Delta_a/\lambda_a\approx10^6$ for $K=10^8, \lambda_a=0.0901,\mu_a=0.09$.
 
@@
 \section*{Results}
-
-% TODO: OVERVIEW OF RESULTS  ok
 
@@
 Thus, if aneuploids rapidly die due to the drug but compensate by rapidly dividing, further increasing  the division rate will \emph{not} facilitate adaptation.
+This is consistent with experimental findings where aneuploidy confers resistance by decreasing the division rate~\citep{replogle2020aneuploidy}.
 For the parameter values of  $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$ we are in the partially resistant aneuploid case and obtain the ratio $N^*_a/N^*_m\approx\abs{\Delta_a}/u\lambda_a=0.11$ which means that the presence of aneuploidy reduces the threshold population by approximately $90\%$.
@@
 However, aneuploid cells are likely generated even before onset of treatment at some rate $\tilde{u} \le u$ (because the treatment itself may promote generation of aneuploid cells \citep{wang2019molecular,mason2017functional}), which are likely to have a deleterious effect in the absence of the drug, $s$ \citep{replogle2020aneuploidy,giam2015aneuploidy}.
-% TODO refs OK
 But if the number of cells in the tumor $N$ is large (as expected if the tumor is to be treated with a drug), there may already be a fraction $f \approx \tilde{u}\lambda_w/s$ (we assume that the division rate of the wildtype in a drug-free environment is equal to the division rate of the wildtype in the environment affected by the anti-cancer drug) of aneuploid cells in the population.
@@
 Therefore, the threshold tumor size with standing generation variation, $\tilde{N}^*_{a}$, is similar to the ratio with de-novo variation, $N^*_a$, except that the sensitive growth rate $\abs{\Delta_w}$ is replaced with the aneuploidy cost, $s$, such that
-\begin{equation}
+\begin{equation}\label{eq:Nsgv}
 \frac{\tilde{N}^*_{a}}{N^*_{a}} = \frac{u}{\tilde{u}} \; \frac{s}{\abs{\Delta_w}}.
@@
 Comparing these approximations to results of stochastic evolutionary simulations, we find that the approximations perform very well (\Cref{rescue_denovo}).
-We derive $s$ by using the formula $s=\tilde{u}\lambda_w/f$ and obtaining the fraction of aneuploid cancer cells $f$ from \citep{lukow2021chromosomal} who mixed together wildtype and aneuploid cell at $50:50$ ratio, cultured them in drug-free environment and observed the ratio evolve as a function of time.
-
-For the parameter values of  $\lambda_w=0.1,\mu_w=0.14, u=10^{-2}, \tilde{u}=10^{-3}, s=5\cdot10^{-4}$ the ratio of the threshold population sizes is $\tilde{N}^*_a/N^*_a\approx0.125$ which means that standing genetic variation is the main driver of adaptation even when a relatively small fraction cancer cells are aneuploid in the drug-free environment (in this case $f=20\%$).
-Furthermore, most tumors display significantly higher levels of aneuploidy such that $f\approx 95\%$ even before cancer treatment.
-In this case, the population threshold ratio becomes $\tilde{N}^*_a/N^*_a\approx0.0025$ and standing genetic variation is a very significant driver of adaptation to the chemotherapeutic drug.
-% TODO PLUG IN REAL VALUES ok
+
+For the parameter values of  $\lambda_w=0.1,\mu_w=0.14, u=10^{-2}, \tilde{u}=10^{-3}, s=7.3\cdot10^{-3}$ the ratio of the threshold population sizes is $\tilde{N}^*_a/N^*_a\approx1.825$ which means that de-novo genetic variation is  the main driver of adaptation.
+
+Using \cref{eq:N_a,eq:N_m,eq:Nsgv}, we can find the ratio of threshold tumor size for rescue via standing genetic variation  to the threshold for direct mutation,
+\begin{equation} \label{eq:N_sgv_ratio}
+\frac{\tilde{N}^*_a}{N^*_m}=\frac{\tilde{N}^*_{a}}{N^*_{a}}\times\frac{N^*_a}{N^*_m} \approx \frac{s}{\abs{\Delta_w}}\begin{cases}
+    \frac{\abs{\Delta_a}}{\tilde{u}\lambda_a} ,&
+  \Delta_a T^* \ll -1 ,\\ 
+  \frac{1}{\tilde{u}}\left(v \cdot \frac{\Delta_m}{\lambda_m}\right)^{1/2} ,&
+  -1 \ll \Delta_a T^* \ll 1  ,\\ 
+  v \frac{\Delta_m}{\lambda_m} \cdot \left(\tilde{u}\frac{\Delta_a}{\lambda_a}\right)^{-1}  ,&
+   \Delta_a T^* \gg 1 .
+  \end{cases}
+\end{equation}
+We observe that evolutionary rescue through direct mutation is more likely if the cost of aneuploidy $s$ is very large or the effect of the drug $\Delta_w$ is small.
+In contrast, standing genetic variation will drive adaptation if the pre-treatment chromosome missegreagation rate $\tilde{u}$ is very large.
+The ratio does not depend on the rate of chromosome missegregation induced by the drug $u$.
+However, if the growth rate of aneuploidy $\Delta_a$ increases then evolutionary rescue is driven by standing genetic variation.
+For the parameter values of  $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, \tilde{u}=10^{-3}, v=10^{-7}$ we are in the partially resistant aneuploid case and obtain the ratio $\tilde{N}^*_a/N^*_m\approx\frac{s}{\abs{\Delta_w}}\times\frac{\abs{\Delta_a}}{\tilde{u}\lambda_a}=0.356$ which means that standing genetic variation reduces the threshold population by approximately $65\%$.
+We observe that while standing genetic variation does not drive emergence of drug resistance when compared to de-novo aneuploids it does offer an advantage when compared with direct mutation.
 
@@
 However, when the expected number of resistant lineages is large the population of mutant cells behaves deterministically and the mean recurrence time cannot be separated into the mean evolutionary rescue time and mean proliferation time.
+Of particular interest is the distribution of the evolutionary rescue time and recurrence time in the parameter regime  $\Delta_a T^*\ll1$ (partially tolerant aneuploids) for which we select  $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$ .
 
 \paragraph{Evolutionary rescue time.}
-In Appendix C we derive an approximation for $\tau_1$, the expected rescue time without aneuploidy ($u=0$), and $\tau_2$, the expected rescue time with aneuploidy ($u>0$).
+In Appendix C we derive an approximation for $\tau_m$, the expected evolutionary rescue time without aneuploidy ($u=0$), and $\tau_a$, the expected rescue time with aneuploidy ($u>0$).
 \Cref{MeanTimeGrowthAneuploidyPlot} shows the agreement between these approximations and simulation results for intermediate and large tumor sizes.
@@
 \Cref{proliferationFigure} shows the agreement between our approximations and simulations.
-For values of $N$ smaller then $10^8$ the mean recurrence time can be approximated as the sum of the mean time for the first rescue muation to appear and the mean time for the lineage generated by that mutation to reach size $N$.
+For values of $N$ smaller then $10^8$ the mean recurrence time can be approximated as the sum of the mean time for the first rescue mutation to appear and the mean time for the lineage generated by that mutation to reach size $N$.
 For $N>10^8$ the behavior of the mutant cancer cell population is deterministic and the mean proliferation time becomes constant.
@@
 In Appendix E we derive the probability that by time $t$ a successful mutant has been generated.
-\Cref{ReboundProbability} show the agreement between our formula and simulation results for the case when aneuploidy is present and when it is absent.
+\Cref{cdffig}A show the agreement between our formula and simulation results for the case when aneuploidy is present and when it is absent.
 We observe that aneuploidy start to have an effect on adaptation for timescales greater then 100 days.
-% ? Distribution of recurrence time -- with and without aneuploidy Ok
-% Kaplan-Meier curves -- x: time; y: P(# cells >= N) -- once for u=0; u>0 with the three scenarios Ok
+
 \paragraph{Distribution of  recurrence time.}
 In Appendix F we derive the probability that mutant cancer cell population reaches $N$ at time $t$.
-\Cref{KaplanMeierfig} shows agreement between our approximations and stochastic simulations for various values of $N$.
+\Cref{cdffig}B shows agreement between our approximations and stochastic simulations for various values of $N$.
+Additionally, we derived the distribution of the recurrence time and plotted our result in \Cref{KMdistribution} alongside numerical simulations for the case $N=10^6$.
+Most clinical trials report data on the distribution of recurrence time measured from a varying time of surgery~\citep{avanzini2019cancer} with chemotherapy usually following after.
+Therefore, since only undetected secondary tumors are present at the time of the administration of the anti-cancer drug we lack knowledge of the size of the tumors and we cannot compare the empirical distributions to our predictions.
+However, we expect that the variability of secondary tumors to average out across large cohorts of patients and, as a result, we can use the mean recurrence time.
+
+We observe from \Cref{cdffig}B that the probability of the mutant cell population reaching size $N$ is approximately zero before time $\tau_a^r$ which is the recurrence time for the deterministic case.
+This can be explained as follows: in the deterministic case there is a sufficient number of lineages produced such that there exists a lineage where each descendant will only reproduce and not die; the time it takes for this lineage to reach $N$ is the lower bound for the time of all other lineages to reach $N$ and this time cannot be smaller then $\tau_a^r$ by definition.
+Given that for small values of $N$ we expect that at most a single lineage will rescue the tumor, this lineage cannot reach $N$ before $\tau_a^r$ for the deterministic case \cref{eq:t2det}.
 
@@
 This can have an impact on the recurrence of cancer after the resection of the primary tumor  through secondary tumors which are too small to be detected and for which chemotherapy is employed to prevent  cancer relapse and are estimated to cause the majority of cancer-related deaths~\citep{chaffer2011perspective}.
-
-Given the fact that the mean time for such secondary tumors to overcome chemotherapy can be of the order of 1000 days (\Cref{MeanTimeGrowthAneuploidyPlot}A), % TODO do we see tau=100 here...? OK
+This is reinforced by the fact that metastases have been shown to have higher chromosome missagregation rate two to three orders of magnitude higher then primary tumors~\citep{kimmel2023intra}.
+
+Given the fact that the mean time for such secondary tumors to overcome chemotherapy can be of the order of 1000 days (\Cref{MeanTimeGrowthAneuploidyPlot}A),
 this can explain the reappearance of cancer even after initial remission.
@@
 
-Paradoxically, we observe from \Cref{MeanTimeGrowthAneuploidyPlot} that the mean time of a rescue mutation to appear is significatly shorter for the case when $u=0$ when compared to the case $u>0$, however this can be explained by the fact this mean time is conditioned on evolutionary rescue and, as a result, aneuploidy increase the \emph{window of opportunity} in which a rescue mutation could appear thus increasing the mean time as well (\Cref{sampleTrajectories}).
-
 We hypothesized that presence of \emph{standing variation} 
 --the existence of a subpopulation of aneuploid cancer cells before therapy begins--can facilitate evolutionary rescue by reducing the waiting time for the appearance of aneuploid cells.
-Indeed, we observe that even when a small fraction of the initial tumor is aneuploid, evolutionary rescue is more likely to occur through this existing standing variation, rather then through \emph{de novo} aneuploid cells (\Cref{rescue_denovo}).
-
-We observe from \Cref{KaplanMeierfig} that the probability of the mutant cell population reaching size $N$ is approximately zero before time $\tau_2^r$ which is the recurrence time for the deterministic case.
-This can be explained as follows: in the deterministic case there is a sufficient number of lineages produced such that there exists a lineage where each descendant will only reproduce and not die; the time it takes for this lineage to reach $N$ is the lower bound for the time of all other lineages to reach $N$ and this time cannot be smaller then $\tau_2^r$ by definition.
-Given that for small values of $N$ we expect that at most a single lineage will rescue the tumor, this lineage cannot reach $N$ before $\tau_2^r$ for the deterministic case \cref{eq:t2det}.
+However, we observe that for reasonable parameter values evolutionary rescue is more likely to occur through de-novo aneuploidy (\Cref{rescue_denovo}).
+If the the fraction of the tumor cells which have the ``beneficial'' aneuploidy is $f\gg\frac{u\lambda_w}{\abs{\Delta_w}}\approx2.5\%$ then evolutionary rescue is more likely to occur through this existing standing variation, rather then through \emph{de novo} aneuploid cells.
+In that case, evolutionary rescue occurs through a single step by the aneuploid cells acquiring a beneficial mutation and the probability of evolutionary rescue is given by $1-\exp\left(-N/\tilde{N}_a^*\right)$  (\Cref{rescue_prob_sgv}).
 
@@
 For example, to study the effects of initial tumor size on the probability of evolutionary rescue, a large culture mass can be propagated from a single cancer cell in permissive conditions and then diluted to a  range of starting tumor sizes.
-Afterwards, these tumors may be exposed to anti-cancer drugs that induces aneuploidy % TODO ref OK
-or to saline solution for control~\citep{ippolito2021gene}. 
-Cell density can then be measured by optical density and a population exposed to the drug is considered extinct if the optical density is lower when compared to the control case.
+Afterwards, these tumors may be exposed to anti-cancer drugs that induces aneuploidy or to saline solution for control~\citep{ippolito2021gene}.
+Cell density can then be measured by optical density and a population exposed to the drug is considered extinct if the optical density is lower when compared to the control case with no cells present.
 We compare to the predictions of our model to see if the tumors with initial size bellow the threshold \eqref{eq:N_a} are more likely to go extinct.
-% TODO not clear what is to be compared...
-what does our model predict...?
-Ok
-
-% TODO discussion of mean time? ok
+
+Additionally, we can test our model's predictions with data from patients.
+Given the type of cancer they have and the degree of ``beneficial'' aneuploidy,  we can obtain the parameters for our model and predict with what probability will the cancer relapse after treatment.
 
@@
 However, this may not be the case, as cancer cells compete for resources (e.g., blood supply).
-Nevertheless, we find that when the effective carrying capacity is not too large ($K\sim10^8, K_e\sim10^6$) % TODO HOW LARGE?
-ok
+Nevertheless, we find that even when the effective carrying capacity is not too large ($K\sim10^8, K_e\sim10^6$)
 our approximation for the probability of evolutionary rescue agrees with results of stochastic simulations with density-dependent growth  (\Cref{LogisticPlot}).
-Future work may focus on scenarios with small carrying capacity by analyzing density-dependent branching processes~\citep{klebaner1997population,harris1963theory}. % TODO ref? OK
-
+Future work may focus on scenarios with small carrying capacity by analyzing density-dependent branching processes~\citep{klebaner1997population,harris1963theory}.
+
+Solid tumors are spatially heterogeneous with different genotypes inhabiting cellular niches and immune infiltration impacting growth in affected regions~\citep{varrone2023cellcharter,galon2010immune}.
+This has the potential to impact the probability of overcoming chemotherapy~\citep{martens2011spatial}.
+Future work should take into consideration the spatial structure of the tumor and its effect on the probability of evolutionary rescue.
+
+An additional limitation of our model is the choice of parameters which are based on a specific set of assumptions which might not be true for many tumor.
+As a result, incorporating a wider distribution of parameters would be a beneficial extension for our model.
 \paragraph{Conclusions}
@@
 \section*{Acknowledgements}
-We thank Hildegard Uecker for discussions and comments. % TODO who? ok
+We thank Hildegard Uecker for discussions and comments. 
 This work was supported in part by
@@
     $u$ & Missegregation rate& $10^{-3}-10^{-2}$ & 1$\slash$cell division  & \citet{bakker2023predicting} \\ \hline
-    $v$ & Mutation rate& $10^{-7}-10^{-5}$ &  1$\slash$gene$\slash$cell division  & \citet{loeb2001mutator} \\  \hline
+    $v$ & Mutation rate& $10^{-9}-10^{-7}$ &  1$\slash$cell division  & \citet{bozic2013evolutionary,loeb2001mutator} \\  \hline
     $\tilde{u}$ & Missegregation rate in the drug free environment$^\ast$& $10^{-3}$ & 1$\slash$cell division  & - \\ \hline
-    $s$ & Selection coefficient of aneuploidy in the drug free environment& $0.3\cdot10^{-3}-10^{-3}$ &  1/days   & \citet{lukow2021chromosomal} \\  
-    %$f$ & Fraction of cancer cells which are aneuploid in the drug free environment& $10\%-30\%$ & -  & \citet{salehi2021clonal}\citet{lukow2021chromosomal} \\ 
-    % TODO citing Nowak 2004 for the rates is strange because it is a theory paper - cite the original empirical papers Ok
+    $s$ & Selection coefficient of aneuploidy in the drug free environment& $7.3\cdot10^{-3}$ &  1/days   & \citet{lukow2021chromosomal} \\  
     \hline
   \end{tabular}
-\caption{\textbf{Model parameters.} %NEED DIFFERENT REFS---THESE ARE MOSTLY THEORY PAPERS. CHECK BIONUMBERS? OK
+\caption{\textbf{Model parameters.} 
 The * symbol in the Name column means that for those parameters the values have not been selected from a paper.
-Aneuploid death rate $\mu_a$ is set to the same value as the mutant death rates, $\mu_m$, given that aneuploidy increases resistance to chemotherapeutics, such as cisplatin, by antagonizing cell division \citep{replogle2020aneuploidy}.
-Aneuploid birth rate $\lambda_a$ is set such that the aneuploid growth rate $\Delta_w\ll\Delta_a\ll\Delta_m$.
-For the missegregation rate in the drug free environment $\tilde{u}$ we choose the lower-end value of $u$ as the chemotherapeutic drugs we want to model increase the rate of aneuploidy~\citep{wang2019molecular,mason2017functional}.
-We note that we have modified the parameters from \citet{bozic2013evolutionary} by starting their analysis from the fact that wildtype/mutant birth rate is $\lambda_{w,m}=\log2/T\approx0.1$ instead of their value of $0.14$ where $T$ is the doubling time in the absence of cellular death obtained from \citet{rew2000cell}.
-We suspect the value of 0.1 was obtained as $1/T$ because $T$ can also be defined as the average time between cell divisions in the absence of cell death~\citet{avanzini2019cancer}.
-We derive the fitness cost $s$ by using the formula $s=\tilde{u}\lambda_w/f$ and obtaining the fraction of aneuploid cancer cells $f$ from \citep{lukow2021chromosomal} who mixed together wildtype and aneuploid cell at $50:50$ ratio, cultured them in drug-free environment and observed the ratio evolve as a function of time.}
+We note that we have modified the parameters from \citet{bozic2013evolutionary} by starting their analysis from the fact that wildtype/mutant division rate is $\lambda_{w,m}=\log2/T\approx0.1$ instead of their value of $0.14$ where $T$ is the doubling time in the absence of cellular death obtained from \citet{rew2000cell}.
+We suspect the value of 0.1 was obtained as $1/T$ because $T$ can also be defined as the average time between cell divisions in the absence of cell death~\citet{avanzini2019cancer}.}
   \label{table1}
@@
 \textbf{Sample trajectories of the different genotypes as a function of time.}
-\textbf{(A)} Whe the missegregation rate $u=0$ evolutionary rescue is only possible through direct mutation and in most cases the tumour will be cured by the drug.
+\textbf{(A)} When the missegregation rate $u=0$ evolutionary rescue is only possible through direct mutation and in most cases the tumor will be cured by the drug.
 \textbf{(B)} When the aneuploidy growth rate $\Delta_a\ll0$ we observe similar dynamic to case \textbf{(A)} as direct mutation is the only viable route toward evolutionary rescue for the tumor.
-\textbf{(C)} Intermediate aneuploid growth rates $\Delta_a\approx0$ generates the appearence of aneuploid lineages even after the wildtype population has gone extinct thus increasing the chance of evolutionary rescue.
+\textbf{(C)} Intermediate aneuploid growth rates $\Delta_a\approx0$ generates the appearance of aneuploid lineages even after the wildtype population has gone extinct thus increasing the chance of evolutionary rescue.
 \textbf{(D)} As the growth rate of the aneuploid becomes positive we observe that the tumour is rescued by the aneuploid cancer cell population.
@@
 %%%%%%%%%%%
+
 \begin{figure}
 \vspace*{1\baselineskip}
+\begin{subfigure}{0.5\textwidth}
+A\\
 \includegraphics[width=1\textwidth]{Figures/ReboundProbability.pdf}
-\caption{\textbf{Aneuploidy is a key diver of cancer adaptation.}
-Plot of the probability that a successful mutant has not appeared by time $t$.
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
+B\\
+\includegraphics[width=1\textwidth]{Figures/ProliferationTimeCDFN.pdf}
+\end{subfigure}
+\caption{\textbf{Aneuploidy has a impact on cancer relapse early on.}
+\textbf{(A)} Plot of the probability that a successful mutant has not appeared by time $t$.
 The green line represents the case with tolerant aneuploidy ($u>0, \lambda_a=0.0899$), the blue line represents the case with non-growing aneuploidy ($u>0, \lambda_a=0.089999$), the cyan line represents the case with partially resistant aneuploidy ($u>0, \lambda_a=0.095$) and the black line represents the case without aneuploidy ($u=0$).
@@
 The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the  probability of that a successful mutant has not been generated and $n=100$ is the number of simulations.
-As time increases, aneuploidy plays an important role in helping the cancer cell population escape extinction.}
-\label{ReboundProbability}
-\end{figure}
-%%%%%%%%%%%
-
-\begin{figure}
-\vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/SurvPlotNDataLogisticK.pdf}
-\caption{\textbf{Density dependent growth does not affect the accuracy of our model.} Comparison of results of simulations  with density-dependent growth (red markers with with 95\% CI) and the approximation formula (black line, \cref{eq:N_a} in \cref{eq:rescue_prob}) with maximum carrying capacity $K=10^8$ and effective carrying capacity $K_e=K\Delta_a/\lambda_a\approx10^6$.
-The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the  probability that the tumor has adapted to the stress and $n=100$ is the number of simulations.
-Parameters: $\lambda_w=0.1,\lambda_a=0.0901,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}, K=10^8$.}
-\label{LogisticPlot}
+As time increases, aneuploidy plays an important role in helping the cancer cell population escape extinction.
+\textbf{(B)} Plot of the probability that a mutant cancer cell population has not reached size $N$ at time $t$.
+The green line represents the case where $N=10^6$, the red line represents $N=10^7$ and the blue line represents the case where $N=10^{10}$.
+Increasing the initial tumor size guarantees that the tumor recurrence.
+The markers represent stochastic simulations and the error bars represents $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the probability of that a mutant population size has not reached $N$ and $n=100$ is the number of simulations.
+Parameters: $\lambda_w=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}
+\label{cdffig}
 \end{figure}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \begin{figure}
 \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/ProliferationTimeCDFN.pdf}
-\caption{Plot of the probability that a mutant cancer cell population has not reached size $N$ at time $t$.
-The green line represents the case where $N=10^6$, the red line represents $N=10^7$ and the blue line represents the case where $N=10^{10}$.
-Increasing the initial tumor size guarantees that the tumor recurrence.
-The markers represent stochastic simulations and the error bars represents $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the probability of that a mutant population size has not reached $N$ and $n=100$ is the number of simulations.
-Parameters: $\lambda_w=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}
-\label{KaplanMeierfig}
+\includegraphics[width=1\textwidth]{Figures/KaplanMeierDistribution.pdf}
+\caption{\textbf{Distribution of the recurrence time.}
+Shown is the distribution of the time for the mutant cell population to reach size $N$, where $N$ is the initial number of cancer cells.
+The red line is analytic result \cref{distribution} overlaid over the histogram of numerical simulations.
+Parameters: $N=10^6, \lambda_w=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}
+\label{KMdistribution}
 \end{figure}
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 p_w&\approx-\frac{v\lambda_wp_m+u\lambda_wp_a}{\lambda_w-\mu_w-u\lambda_wp_a-v\lambda_wp_m}\\
-&\approx\frac{1}{\lambda_w-\mu_w-u\lambda_wp_a-v\lambda_wp_m}\left[\frac{uv\lambda_w\lambda_a\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-v\lambda_a\right)}-\frac{v\lambda_w\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\ %\label{survprobw2}
+&\approx\frac{1}{\lambda_w-\mu_w-u\lambda_wp_a-v\lambda_wp_m}\left[\frac{uv\lambda_w\lambda_a\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-v\lambda_a\right)}-\frac{v\lambda_w\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\
 &\approx\frac{v\lambda_w\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_w-\mu_w\right)}\left[\frac{u\lambda_a}{\left(\lambda_a-\mu_a\right)}-1\right] \\
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection*{Scenario 3: Aneuploid cells are stationary} % TODO consider this name ok
-
+\subsubsection*{Scenario 3: Aneuploid cells are stationary} 
 We now assume that the growth rate of aneuploid cells is close to zero (either positive or negative), such that  $\left(\Delta_a-v\lambda_ap_m\right)^2 \ll 4\lambda_a^2vp_m$.
@@
 \section*{Appendix C: Evolutionary rescue time}
-% TODO remove the linear approximation, it is not good. OK
 
@@
 
-Assuming no aneuploidy ($u=0$), we define $T_1$ to be the time at which the first mutant cell appears that will avoid extinction and will therefore rescue the population.
-Note that if extinction occurs, that is the frequency of mutants after a very long time is zero, $m_{\infty}=0$, then it is implied that $T_1=\infty$, and vice versa if $T_1<\infty$ then $m_{\infty}>0$.
-
-The number of successful mutants generated until time $t$ can be approximated by an inhomogeneous Poisson process with rate $R_1\left(t\right) = v\lambda_w p_m w_t$,
+Assuming no aneuploidy ($u=0$), we define $T_m$ to be the time at which the first mutant cell appears that will avoid extinction and will therefore rescue the population.
+Note that if extinction occurs, that is the frequency of mutants after a very long time is zero, $m_{\infty}=0$, then it is implied that $T_m=\infty$, and vice versa if $T_m<\infty$ then $m_{\infty}>0$.
+
+The number of successful mutants generated until time $t$ can be approximated by an inhomogeneous Poisson process with rate $R_m\left(t\right) = v\lambda_w p_m w_t$,
 where $w_t=N\e^{\Delta_w t}$ is the number of wildtype cells at time $t$.
@@
 \begin{equation}\label{eq:integralR}
-\int_0^{t}{R_1(z)\d z} = 
+\int_0^{t}{R_m(z)\d z} = 
 v\lambda_w p_m N \frac{\exp[{\Delta_w t}]-1}{\Delta_w} \approx 
@@
 by integrating the exponential and because $\frac{\exp[\Delta_w t]-1}{\Delta_w}=\frac{1+\Delta_w t+\mathcal{O}(t^2)-1}{\Delta_w}=t+O(t^2)$.
-The probability density function of $T_1$ is thus
-$R_1\left(t\right)\exp\left(-\int_0^{t}{R_1(z)\d z}\right)$ \citep{allen2010introduction}. % TODO ref ok
-Therefore, the probability density function of the conditional random variable $(T_1 \mid T_1 < \infty)$ is
-$f_1(t) = \frac{R_1\left(t\right)\exp\left(-\int_0^{t}{R_1(z)\d z}\right)}{\presc}$. 
+The probability density function of $T_m$ is thus
+$R_m\left(t\right)\exp\left(-\int_0^{t}{R_1(z)\d z}\right)$ \citep{allen2010introduction}. 
+Therefore, the probability density function of the conditional random variable $(T_m \mid T_m < \infty)$ is
+$f_m(t) = \frac{R_m\left(t\right)\exp\left(-\int_0^{t}{R_m(z)\d z}\right)}{\presc}$. 
 \\
 
-We are interested in the mean conditional time, $\tau_1=\mathbb{E}\left[T_1 \mid T_1<\infty\right]$, which is given by
+We are interested in the mean conditional time, $\tau_m=\mathbb{E}\left[T_m \mid T_m<\infty\right]$, which is given by
 \begin{equation}\label{eq:meantime1}
 \begin{aligned}
-\tau_1 =
-\int_{0}^{\infty}{t f_1(t) \d t} = 
-\frac{\int_{0}^{\infty}{tR_1(t)\exp\left(-\int_0^{t}{R_1(z)\d z}\right) \d t}}{p_{rescue}},% = \int_{0}^{\infty}\left[1-\frac{1-\exp\left(-\int_0^{t}{R(z)\d z}\right) }{p_{rescue}}\right]\d t
+\tau_m =
+\int_{0}^{\infty}{t f_m(t) \d t} = 
+\frac{\int_{0}^{\infty}{tR_m(t)\exp\left(-\int_0^{t}{R_m(z)\d z}\right) \d t}}{p_{rescue}},
 \end{aligned}
 \end{equation}
-%after applying integration by parts and using the fact that $f_1(t)=-\frac{d}{dt}\left[1-F_1\left(t\right)\right]$, where $F_1\left(t\right)$ is the cumulative distribution function of $T_1$.
 Therefore, plugging \cref{eq:integralR,eq:rescue_prob} in \cref{eq:meantime1}, 
 \begin{align}\label{eq:limitapprox_appendix}
-\tau_1 = 
+\tau_m = 
 \int_0^\infty tv\lambda_w N\e^{\Delta_wt}\frac{\e^{-v\lambda_w N p_m\frac{\e^{\Delta_w t}-1}{\Delta_w}} }{1-\left(1-p_w\right)^N} \d t\approx
@@
 \\
+Assuming aneuploidy is possible ($u>0$), we define $T_a$ to be the time at which the first mutant cell appears that will rescue the population.
+We are interested in the mean conditional time, $\tau_a=\mathbb{E}\left[T_a \mid T_a<\infty\right]$.
 
@@
 \end{align}
-For large initial population sizes we assume that the two processes are independent and as a result, they can be merged into a single Poisson process with rate $R_2(t)=\left(r_1+r_2\right)\left(t\right)$.
+For large initial population sizes we assume that the two processes are independent and as a result, they can be merged into a single Poisson process with rate $R_a(t)=\left(r_1+r_2\right)\left(t\right)$.
 Consequently, the mean time to the appearance of the first rescue mutant is
 \begin{align}\nonumber
-\tau_2 &= \frac{\int_{0}^{\infty}{tR_2(t)\exp\left(-\int_0^{t}{R_2(z)\d z}\right) \d t}}{p_{rescue}}\\ \label{meantimet2}
+\tau_a &= \frac{\int_{0}^{\infty}{tR_a(t)\exp\left(-\int_0^{t}{R_a(z)\d z}\right) \d t}}{p_{rescue}}\\ \label{meantimet2}
 &=
@@
 which we plot in \Cref{MeanTimeGrowthAneuploidyPlot}A as a function of the initial population size, $N$.
+
+Paradoxically, we observe from \Cref{MeanTimeGrowthAneuploidyPlot} that the mean time of a rescue mutation to appear is significantly shorter for the case when $u=0$ when compared to the case $u>0$, however this can be explained by the fact this mean time is conditioned on evolutionary rescue and, as a result, aneuploidy increase the \emph{window of opportunity} in which a rescue mutation could appear thus increasing the mean time as well (\Cref{sampleTrajectories}).
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section*{Appendix D: Recurrence time}
-We define the proliferation time $\tau_2^r$  to be the time it takes the population of mutant cancer cells to reach the initial tumor size $N$.
+We define the proliferation time $\tau_a^r$  to be the time it takes the population of mutant cancer cells to reach the initial tumor size $N$.
 The number of rescue lineages generated by the wildtype population is given by  \cref{eq:twosteplineage} (see \Cref{ExpectedNumberRescueLineages}):
@@
 \begin{align}\label{meanproliferationtime}
-\tau_2^r&\approx\tau_2+\frac{\log p_mN}{\Delta_m}.
+\tau_a^r&\approx\tau_a+\frac{\log p_mN}{\Delta_m}.
 \end{align}
@@
 \end{align*}
-We obtain $\tau_2^r$ such that $m\left(\tau_2^r\right)=N$ by solving:
+We obtain $\tau_a^r$ such that $m\left(\tau_a^r\right)=N$ by solving:
 \begin{equation}\label{eq:t2det}
-1=\frac{uv\lambda_a\lambda_w}{\Delta_w-\Delta_a}\left[\frac{\e^{\Delta_w\tau_2^r}-\e^{\Delta_m\tau_2^r}}{\Delta_w-\Delta_m}-\frac{\e^{\Delta_a\tau_2^r}-\e^{\Delta_m\tau_2^r}}{\Delta_a-\Delta_m}\right]+v\lambda_w\frac{\e^{\Delta_w\tau_2^r}-\e^{\Delta_m\tau_2^r}}{\Delta_w-\Delta_m}.
+1=\frac{uv\lambda_a\lambda_w}{\Delta_w-\Delta_a}\left[\frac{\e^{\Delta_w\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_w-\Delta_m}-\frac{\e^{\Delta_a\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_a-\Delta_m}\right]+v\lambda_w\frac{\e^{\Delta_w\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_w-\Delta_m}.
 \end{equation}
@@
 \begin{align*}
-P\left(rescue,t\right)&=P\left(T_2<t\right)\\
+P\left(rescue,t\right)&=P\left(T_a<t\right)\\
 &=1-\exp\left\{-\left[r_1\left(t\right)+r_2\left(t\right)\right]\right\}\\
@@
 \end{align*}
-where $T_2$ is the time at which the first mutant cell appears that will avoid extinction and which was defined in Appendix C.
+where $T_a$ is the time at which the first mutant cell appears that will avoid extinction and which was defined in Appendix C.
 
@@
 \end{align*}
-where $a_t$ is aneuploid population size at time $t$ defined in \cref{aneuploidpopeq}.
+where $a_s$ is aneuploid population size at time $s$ defined in \cref{aneuploidpopeq}.
 The proliferation time is defined as the first time the size of all lineages reaches $N$.
@@
 \begin{align}\nonumber
-P\left(m_t\leq N\right)&=1-P\left(\tau>t\right)=\exp\left[-r\left(t\right)\right]\\ \label{eqmNs}
-&=\frac{Nuv\lambda_w\lambda_ap_m}{\Delta_w-\Delta_a}\int_0^t\left[\e^{\Delta_wt}-\e^{\Delta_at}\right]\e^{-p_mN\e^{-\Delta_m\left(t-s\right)}}\,\d s.
+P\left(m_t\leq N\right)&=\exp\left[-r\left(t\right)\right]\\ \label{eqmNs}
+&=\exp\left[-\frac{Nuv\lambda_w\lambda_ap_m}{\Delta_w-\Delta_a}\int_0^t\left[\e^{\Delta_ws}-\e^{\Delta_as}\right]\e^{-p_mN\e^{-\Delta_m\left(t-s\right)}}\,\d s\right].
 \end{align}
 When $N\gg\abs{\Delta_w}\abs{\Delta_a}/uv\lambda_w\lambda_ap_m$ the dynamics of the cancer cell populations is deterministic and approximated by the system of ODEs shown in \cref{detODE}.
-As a result, the size of the mutant cell population will always be below $N$ until time $\tau_2^r$ and will always be greater after:
+As a result, the size of the mutant cell population will always be below $N$ until time $\tau_a^r$ and will always be greater after:
 \begin{align}\label{eqmNd}
-P\left(m_t\leq N\right)=1-H\left(t-\tau_2^r\right),
+P\left(m_t\leq N\right)=1-H\left(t-\tau_a^r\right),
 \end{align}
@@
 As a result, the number of mutant cancer cells will reach $N$ faster then the case with a single mutant lineage.
+
+From  \cref{eqmNd} we obtain the distribution of the recurrence time conditional of evolutionary rescue:
+\begin{align}\label{distribution}
+f\left(t\right)&=\frac{d}{dt}\left[\frac{P\left(m_t\geq N\right)}{p_{rescue}}\right]=r'\left(t\right)\frac{\exp\left[-r\left(t\right)\right]}{p_{rescue}},
+\end{align}
+which we plot in Figure and compare with numerical simulations.
+We note that in the case $N\gg\abs{\Delta_w}\abs{\Delta_a}/uv\lambda_w\lambda_ap_m$ the distribution becomes the Dirac $\delta$-function~\citep{barton1989elements}.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\newpage
+\clearpage
 
@@
 
+\begin{figure}[h]
+\vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/SurvPlotNDataLogisticK.pdf}
+\caption{\textbf{Density dependent growth does not affect the accuracy of our model.} Comparison of results of simulations  with density-dependent growth (red markers with with 95\% CI) and the approximation formula (black line, \cref{eq:N_a} in \cref{eq:rescue_prob}) with maximum carrying capacity $K=10^8$ and effective carrying capacity $K_e=K\Delta_a/\lambda_a\approx10^6$.
+The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the  probability that the tumor has adapted to the stress and $n=100$ is the number of simulations.
+Parameters: $\lambda_w=0.1,\lambda_a=0.0901,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}, K=10^8$.}
+\label{LogisticPlot}
+\end{figure}
+%%%
 \begin{figure}[!htb]
@@
 \end{figure}
-
+%%%
+\begin{figure}
+\includegraphics[width=1\textwidth]{Figures/ProbvN_SGV_Plot.pdf}
+\caption{The probability of evolutionary rescue (i.e.
+the probability that the population does not go to extinction), $\presc$, as a function of the initial tumor size, $N$.
+Dashed vertical line shows the threshold tumor size, above which the probability is very high.
+Blue dashed line represents the probability of evolutionary rescue as a function of $N$ without aneuploidy ($u=0$).
+The black line represents the case with tolerant aneuploidy a fraction $f=0\%$ of the initial tumor being aneuploid, the red line represents the case with a fraction $f=5\%$ and the green line represents the case with $f=50\%$.
+The dots represent numerical simulations and the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the probability of that the tumor has adapted to the stress and $n=100$ is the number of simulations.
+Parameters: $\lambda_w=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}
+\label{rescue_prob_sgv}
+\end{figure}
 \end{appendices}

### 2024-01-20 EvolutionaryRescue1.tex
Add files via upload

@@
 
- \citet{ippolito2021gene} induced aneuploidy in cancer cell lines by exposing them to reversine, a small-molecule inhibitor of the mitotic kinase Mpsi1, and then exposed to chemotherapeutic agents such as vemurafenib.
+ \citet{ippolito2021gene} induced aneuploidy in cancer cell lines by exposing them to reversine, a small-molecule inhibitor of the mitotic kinase Mpsi1, and then to chemotherapeutic agents such as vemurafenib.
 When compared to wildtype cancer cell the reversine treated cells had higher proliferation in the environment altered by anti-cancer drugs.
-Similarly, \citet{lukow2021chromosomal} observed that cancer cells which have been induced to be aneuploid have an advantage when compare to wildtype cell even while they had lower fitness before the onset of chemotherapy.
-A proposed mechanism through which aneuploidy is able to confer resistance to chemotherapeutics is bu antagonizing cell division  which prevents the drugs from damaging DNA and microtubules \citep{replogle2020aneuploidy}
-
- An important aspect of aneuploidy is the rate with which cell acquire it, which is several orders of magnitude higher when compare to mutation rate \citep{bakker2023predicting}.
+Similarly, \citet{lukow2021chromosomal} observed that cancer cells which have been induced to be aneuploid have an advantage when compare to wildtype even while they had lower fitness before the onset of chemotherapy.
+A proposed mechanism through which aneuploidy is able to confer resistance to chemotherapeutics is by antagonizing cell division  which prevents the drugs from damaging DNA and microtubules \citep{replogle2020aneuploidy}
+
+ An important aspect of aneuploidy is the rate with which cells acquire it, which is several orders of magnitude higher when compare to the mutation rate \citep{bakker2023predicting}.
 Consequently, a cell exposed to a stress such as chemotherapeutic drugs can acquire aneuploidy faster when compared to acquiring a mutation, especially when several proposed anti-cancer drugs elevate the rate of missegration in order to combat cancer \citep{lee2016effects}.
@@
 
-Here we build on this previous work on evolutionary rescue after a sudden environment change caused by the initiation of chemotherapy.
-We wish to understand what effect does aneuploidy have on probability of evolutionary rescue when it acts as an intermediary between the wildtype and the mutant cancer cells.
-Besides the probability of evolutionary rescue we calculate the calculate the mean time that an initially declining tumor cell population reaches the initial population size.
+Here we build on previous work on evolutionary rescue after a sudden environment change caused by the initiation of chemotherapy.
+We wish to understand what effect does aneuploidy have on the probability of evolutionary rescue when it acts as an intermediary between the wildtype and the mutant cancer cells.
+We also calculate the mean time that an initially declining tumor cell population reaches its pre-treatment size.
 Given that aneuploidy is present in many tumors even before the onset of therapy \citep{lukow2021chromosomal} we also take into consideration the effect that standing genetic variation has on the tumor dynamics.
@@
 Both aneuploid and wildtype cells may divide and mutate to become mutants at rates $v\lambda_{a}$ and $v\lambda_{w}$, respectively.
-See \Cref{figureAneuploidy} for a schematic representation of the model and \Cref{sampleTrajectories} for sample trajectories of the different genotypes for $u=0 \,\textbf{(A)}, \Delta_a\ll0\, \textbf{(B)},\Delta_a\approx0\, \textbf{(C)},\Delta_a\gg0\, \textbf{(D)}$.
+See \Cref{figureAneuploidy} for a schematic representation of the model and \Cref{sampleTrajectories} for sample trajectories of the different genotypes for $u=0 \, (A), \Delta_a\ll0\, (B),\Delta_a\approx0\, (C),\Delta_a\gg0\, (D)$.
 We assume that wildtype cells acquire aneuploidy before the onset of therapy with rate $\tilde{u}\lambda_w$ and that aneuploid cells have a fitness cost $s$ in the drug-free environment.
@@
 \end{equation*}
-
+where we obtain the threshold mutant size by solving $1-(1-p_m)^{m_t}=0.999$ for $m_t$, where $p_m=\Delta_m/\lambda_m$ is the probability that a single mutant escapes stochastic extinction  (see Apendix A).
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 If the change in number of cells is negative and larger then the subpopulation size then the subpopulation size is updated to be zero.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\paragraph{Bootstrapping}
+We  use bootstrap resampling method in order to compute  confidence intervals for the mean threshold tumor sizes and mean times by resampling 100 times with replacement from the data set obtain from simulations.
+We then compute the interval which contain $95\%$ of the means of each resampled set.
 
@@
 \citet{ippolito2021gene} observed that trisomy 2 and 6 are most likely to confer increased resistance against the chemotherapeutic agent vemurafenib for A375 cells.
-We assume that if a cancer is aneuploid then it most likely has trisomy \citep{gisselsson2010generation} in the pre-treatment environment and that cells with more then one trisomy are very unlikely to survive.
+We assume that if a tumor is aneuploid then it most likely has trisomy \citep{gisselsson2010generation} in the pre-treatment environment and that cells with more then one trisomy are very unlikely to survive.
 Additionally, we assume that all trisomies are equally likely and, as a result, we obtain the percentage of cancer cells with favorable aneuploidy as $2/22\times0.15\approx 0.0136$ (i.e.
@@
 \end{align*}
-where $K$ is the maximum carrying capacity. The effective carrying capacity of this model is $K_e=K\Delta_a/\lambda_a\approx10^6$ for $K=10^8, \lambda_a=0.0901,\mu_a=0.09$.
+where $K$ is the maximum carrying capacity.
+The effective carrying capacity of this model is $K_e=K\Delta_a/\lambda_a\approx10^6$ for $K=10^8, \lambda_a=0.0901,\mu_a=0.09$, where we define the effective carrying capacity to be the population size at which the aneuploid division rate is equal to the aneuploid death rate.
 
@@
 \\
-In the Appendix, we use the theory of multi-type branching processes to find approximate expressions \cref{eq:pw_parttolerant,eq:pw_partrest,eq:pw_tolerant} for $p_w$ in different regimes. 
+In Appendix A, we use the theory of multi-type branching processes to find approximate expressions \cref{eq:pw_parttolerant,eq:pw_partrest,eq:pw_tolerant} for $p_w$ in different regimes. 
 Substituting these  into $N^*=1/p_w$, we find approximations for the threshold tumor size, $N^*$. 
@@
     \frac{\abs{\Delta_a}}{v\lambda_a} \cdot \frac{\lambda_m}{\Delta_m} ,&
-  \Delta_a T^* \ll -1 \text{ (partially tolerant aneuploids)},\\ 
+  \Delta_a T^* \ll -1 \text{ (tolerant aneuploids)},\\ 
   %\left(\frac{\lambda_a}{v} \cdot \frac{\lambda_m}{\Delta_m}\right)^{1/2} ,&
@@
 For the parameter values of  $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$ we are in the partially resistant aneuploid case and obtain the ratio $N^*_a/N^*_m\approx\abs{\Delta_a}/u\lambda_a=0.11$ which means that the presence of aneuploidy reduces the threshold population by approximately $90\%$.
-We observe that aneuploidy has a significant contribution towards the tumor overcoming chemotherapeutic drugs.
-% TODO PLUG IN REAL VALUES ok
+We note that for this parameter regime we have $N_a^*\approx4.4\times10^6$ and $N_m^*\approx4\times10^7$ and this allows us to define the following categories: small tumors have size $N<N_a^*$, intermediate sized tumors have $N_a^*<N<N_m^*$ and large tumors have size $N>N_m^*$ .
+For the remaining of this paper we use these definitions when referring to small, intermediate or large tumors.
+Interestingly, the threshold between small and intermediate tumors is similar to the detection threshold of $4.19\times10^6$ for a wide variety of tumors as reported by \citet{avanzini2019cancer}.
 
@@
 Even when evolutionary rescue occurs and leads to recurrence of the tumor, it may take a long time.
-When the expected number of resistant lineages that avoid extinction is small the overall expected recurrence time can be estimated by adding two terms: the mean waiting time for evolutionary rescue--the appearance of a resistant lineage that avoid extinction-- (which we denote the mean evolutionary rescue time) and the expected time for proliferation of that lineage back to the original tumor size, $N$ (which we denote the mean proliferation time).
+When the expected number of resistant lineages that avoid extinction is small the overall expected recurrence time can be estimated by adding two terms: the mean waiting time for evolutionary rescue--the appearance of a resistant lineage that avoid extinction (which we denote the mean evolutionary rescue time) and the expected time for proliferation of that lineage back to the original tumor size, $N$ (which we denote the mean proliferation time).
 However, when the expected number of resistant lineages is large the population of mutant cells behaves deterministically and the mean recurrence time cannot be separated into the mean evolutionary rescue time and mean proliferation time.
@@
 In Appendix C we derive an approximation for $\tau_m$, the expected evolutionary rescue time without aneuploidy ($u=0$), and $\tau_a$, the expected rescue time with aneuploidy ($u>0$).
-\Cref{MeanTimeGrowthAneuploidyPlot} shows the agreement between these approximations and simulation results for intermediate and large tumor sizes.
-
+\Cref{MeanTimeGrowthAneuploidyPlot,EvolutionaryRescueTimeComplete} shows the agreement between these approximations and simulation results for small, intermediate and large tumor sizes.
+Additionally, we note that $\tau_a$ has the following asymptotic behaviour:
+\begin{equation}  \label{eq:AsymptoticTimeRules}
+\tau_a \approx \begin{cases}
+    \frac{1}{\abs{\Delta_w}}+\frac{1}{\abs{\Delta_a}} ,&
+ N \ll N_a^* ,\\ 
+  \left(v\lambda_wNp_m\right)^{-1}  ,&
+  N \gg N_m^* .
+  \end{cases}
+\end{equation}
+Interestingly, for small tumors the mean evolutionary rescue time is a function of the wildtype and aneuploid growth rates and independent of all of other model parameters including tumor size $N$ (\Cref{EvolutionaryRescueTimeComplete}).
+Increasing either the wildtype or aneuploid growth rates lead to an increase in the mean time.
+This can be interpreted that by increasing either growth rates the wildtype or aneuploid cells will last longer and will be able to produce a rescue mutation at latter times.
+Using the parameters that we have selected we observe that $\abs{\Delta_w}^{-1}=0.04$ and $\abs{\Delta_a}^{-1}=10^{-4}$ such that the mean evolutionary rescue time is determined by the aneuploid growth rate, i.e.
+$\tau_a\approx1/\abs{\Delta_a}$.
+For large tumors, we observe that the evolutionary rescue time is exponentially distributed with rate $v\lambda_wNp_m$ which is independent of parameters characterizing aneuploids or their production ($u$, $\lambda_a$, and $\Delta_a$).
+Increasing the mutation rate per mitosis $v\lambda_w$ leads to a faster appearance of a rescue mutation.
+Note that in the second case increasing the tumor size leads to smaller mean evolutionary rescue time as the wildtype is able to produce a rescue lineage faster.
+
+When a fraction $f\approx 1.36\%$ of the initial cancer cell population is aneuploid we calculate the mean evolutionary rescue time in \Cref{meantimet2SGV} which we plot in \Cref{SGVEvolutionaryRescueTimeComplete}.
+We note that standing genetic variation reduces the mean evolutionary rescue time only for intermediary sized tumors but not for large or small tumors.
 \paragraph{Recurrence time.}
@@
 For values of $N$ smaller then $10^8$ the mean recurrence time can be approximated as the sum of the mean time for the first rescue mutation to appear and the mean time for the lineage generated by that mutation to reach size $N$.
-For $N>10^8$ the behavior of the mutant cancer cell population is deterministic and the mean proliferation time becomes constant.
-
-\paragraph{Distribution of evolutionary rescue time -- with and without aneuploidy.}
+For $N>10^8$ the behavior of the mutant cancer cell population is deterministic and the mean recurrence time becomes constant.
+We note that $\tau_a^r$ has the following asymptotic behavior:
+\begin{equation} \label{eq:AsymptoticTimeRulesRecurrence}
+\tau_a^r \approx \begin{cases}
+   \frac{1}{ \abs{\Delta_w}}+\frac{1}{ \abs{\Delta_a}}+\frac{\log p_mN}{\Delta_m} ,&
+ N \ll N_a^* ,\\ 
+  \frac{1}{\Delta_m}\log\frac{\Delta_m-\Delta_w}{v\lambda_w}  ,&
+  N \gg N_m^* .
+  \end{cases}
+\end{equation}
+In the first case, when tumor is small, the mean recurrence time grows logarithmically with tumor size $N$ and is the same order of magnitude as the mean evolutionary rescue time (\Cref{ProliferationTimeLarge}).
+Increasing the mutant growth rate $\Delta_m$ leads to shorter recurrence times while increasing the wildtype and aneuploid growth rates $\Delta_w$ and $\Delta_a$, retrospectively, increases the recurrence time.
+In the second case, when tumor is large, we observe that increasing either the mutant growth rare $\Delta_m$ or the mutation rate $v$ leads to a decrease in the time needed for the tumor to rebound to its initial size.
+Additionally, drugs which significantly increase the wildtype death rate $\mu_w$ and do not affect the division rate $\lambda_w$ delays cancer relapse.
+Consequently, patients which take such drugs require a longer period of monitoring to guarantee the effectiveness of the treatment.
+\paragraph{Distribution of evolutionary rescue time with and without aneuploidy.}
 
@@
 \Cref{cdffig}B shows agreement between our approximations and stochastic simulations for various values of $N$.
-Additionally, we derived the distribution of the recurrence time and plotted our result in \Cref{KMdistribution} alongside numerical simulations for the case $N=10^6$.
+Additionally, we derived the distribution of the recurrence time and plotted our result in \Cref{KMdistribution} alongside simulations for the case $N=10^6$ (small tumor) and we note that the distribution is wide and rights-skewed.
+It is highly unlikely to observe the recurrence of tumors at times smaller then $\frac{1}{\Delta_m}\log\frac{\Delta_m-\Delta_w}{v\lambda_w}\approx1542$ days for the parameter values $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$ and independent of initial tumor size $N$ (\Cref{cdffig}B).
+
+The recurrence can also be defined as the time necessary for the tumor size to reach a detection threshold $M$.
+We derive the mean recurrence time to detection size $M=10^7$ in Appendix D and plot in \Cref{RecurrencePlot} with simulations.
+We observe that  for small and intermediate sized tumors the effect of the detection size is negligible when compared to $\tau_a^r$.
+However, for large tumors the mean recurrence time to detection size $M$ decreases logarithmically with tumor size $N$.
+
 Most clinical trials report data on the distribution of recurrence time measured from a varying time of surgery~\citep{avanzini2019cancer} with chemotherapy usually following after.
@@
 However, we expect that the variability of secondary tumors to average out across large cohorts of patients and, as a result, we can use the mean recurrence time.
-
-We observe from \Cref{cdffig}B that the probability of the mutant cell population reaching size $N$ is approximately zero before time $\tau_a^r$ which is the recurrence time for the deterministic case.
-This can be explained as follows: in the deterministic case there is a sufficient number of lineages produced such that there exists a lineage where each descendant will only reproduce and not die; the time it takes for this lineage to reach $N$ is the lower bound for the time of all other lineages to reach $N$ and this time cannot be smaller then $\tau_a^r$ by definition.
-Given that for small values of $N$ we expect that at most a single lineage will rescue the tumor, this lineage cannot reach $N$ before $\tau_a^r$ for the deterministic case \cref{eq:t2det}.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-We find that aneuploidy can have a significant effect on evolutionary rescue (\Cref{rescue_prob}).
-Interestingly, aneuploidy is unlikely to contribute to evolutionary rescue in primary tumors in which the number of cells is large enough ($N>10^8$) for the appearance of resistant mutation directly in sensitive cells before these cells become extinct (\Cref{rescue_prob}).
+We find that aneuploidy can have a significant effect on evolutionary rescue as it reduces the threshold population size by at least one order of magnitude even when aneuploidy is tolerant (\Cref{rescue_prob}).
+Interestingly, aneuploidy is unlikely to contribute to evolutionary rescue in primary tumors in which the number of cells is large enough ($N\gg N_m^*\approx 4\times10^7$) for the appearance of resistant mutation directly in sensitive cells before these cells become extinct (\Cref{rescue_prob}).
 However, aneuploidy can have a crucial role in evolutionary rescue of secondary tumors, in which the number of sensitive cells may be below the detection threshold of $\sim10^7$  \citep{bozic2013evolutionary}.
@@
 However, this may not be the case, as cancer cells compete for resources (e.g., blood supply).
-Nevertheless, we find that even when the effective carrying capacity is not too large ($K\sim10^8, K_e\sim10^6$)
+Nevertheless, we find that when the effective carrying capacity is not too large ($K\sim10^8, K_e\sim10^6$)
 our approximation for the probability of evolutionary rescue agrees with results of stochastic simulations with density-dependent growth  (\Cref{LogisticPlot}).
@@
 \caption{\textbf{Model parameters.} 
-The * symbol in the Name column means that for those parameters the values have not been selected from a paper.
-We note that we have modified the parameters from \citet{bozic2013evolutionary} by starting their analysis from the fact that wildtype/mutant division rate is $\lambda_{w,m}=\log2/T\approx0.1$ instead of their value of $0.14$ where $T$ is the doubling time in the absence of cellular death obtained from \citet{rew2000cell}.
-We suspect the value of 0.1 was obtained as $1/T$ because $T$ can also be defined as the average time between cell divisions in the absence of cell death~\citet{avanzini2019cancer}.}
+We have modified the parameters from \citet{bozic2013evolutionary} such that wildtype/mutant division rate is $\lambda_{w,m}=\log2/T\approx0.1$ instead of their value of $0.14$ where $T$ is the doubling time in the absence of cellular death obtained from \citet{rew2000cell}.}
   \label{table1}
@@
 \textbf{(B)} The wildtype and the mutant are susceptible and resistant, respectively, to the drug.
-The aneuploid may be tolerant $(4)$, stationary $(3)$, partially resistant $(2)$ or fully resistant $(1)$.
+The aneuploid may be tolerant, stationary and partially resistant.
 }
@@
 \caption{
-\textbf{Sample trajectories of the different genotypes as a function of time.}
-\textbf{(A)} When the missegregation rate $u=0$ evolutionary rescue is only possible through direct mutation and in most cases the tumor will be cured by the drug.
-\textbf{(B)} When the aneuploidy growth rate $\Delta_a\ll0$ we observe similar dynamic to case \textbf{(A)} as direct mutation is the only viable route toward evolutionary rescue for the tumor.
-\textbf{(C)} Intermediate aneuploid growth rates $\Delta_a\approx0$ generates the appearance of aneuploid lineages even after the wildtype population has gone extinct thus increasing the chance of evolutionary rescue.
-\textbf{(D)} As the growth rate of the aneuploid becomes positive we observe that the tumour is rescued by the aneuploid cancer cell population.
+\textbf{Sample trajectories of the different genotype frequencies.}
+(A) When the missegregation rate $u=0$ evolutionary rescue is only possible through direct mutation and in most cases the tumor will be killed by the drug.
+(B) When the aneuploidy growth rate $\Delta_a\ll0$ we observe similar dynamic to case (A) as direct mutation is the only viable route toward evolutionary rescue for the tumor.
+(C) Intermediate aneuploid growth rates $\Delta_a\approx0$ leads to the appearance of aneuploid lineages even after the wildtype population has gone extinct thus increasing the chance of evolutionary rescue.
+(D) As the growth rate of the aneuploid becomes positive we observe that the tumour is rescued by the aneuploid cancer cell population.
 Each plot features 10 trajectories of $\left(w_t,a_t,m_t\right)$ as a function of time $t$ for the following parameter values: $\lambda_w=0.1,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, v=10^{-7},N=10^7$.
-For \textbf{(A)} we set $u=0$, for \textbf{(B)} $\lambda_a=0.065,u=10^{-2}$, for \textbf{(C)} $\lambda_a=0.0899,u=10^{-2}$ and for \textbf{(D)}   $\lambda_a=0.095,u=10^{-2}$.
+For (A) we set $u=0$, for (B) $\lambda_a=0.065,u=10^{-2}$, for (C) $\lambda_a=0.0899,u=10^{-2}$ and for (D) $\lambda_a=0.095,u=10^{-2}$.
 }
@@
 The probability of evolutionary rescue (i.e.
-the probability that the population does not go to extinction), $\presc$, as a function of the initial tumor size, $N$.
-Dashed vertical line shows the threshold tumor size, $N_a^*$, above which the probability is very high.
+the probability that the population does not go to extinction), $\presc$, as a function of the initial tumor size, $N$ (see \cref{eq:rescue_prob}).
+Dashed vertical line shows the threshold tumor size, $N_a^*$, above which the probability is very high (see \cref{eq:N_a}).
 Blue dashed line represents the probability of evolutionary rescue as a function of $N$ without aneuploidy ($u=0$).
 The black line represents the case with tolerant aneuploidy ($u=10^{-2}, \lambda_a=0.0899$), the red line represents the case with stationary aneuploidy ($u=10^{-2}, \lambda_a=0.08999$) and the green line represents the case with partially resistant aneuploidy ($u=10^{-2}, \lambda_a=0.095$).
-The dots represent numerical simulations and the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the probability of that the tumor has adapted to the stress and $n=100$ is the number of simulations.
+The dots represent simulations and the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the fraction of simulations in which the tumor has adapted to the stress and $n=100$ is the number of simulations.
 Parameters: $\lambda_w=0.1,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, v=10^{-7}$.}
@@
 \caption{
-\textbf{Aneuploidy growth rate and missegregation rate impact the evolutionary rescue of cancer under drug treatment.}
-\textbf{(A)} The threshold tumor size $N_a^*$ as a function of the aneuploid growth rate $\Delta_a$.
+\textbf{Aneuploidy facilitates the evolutionary rescue of cancer under drug treatment.}
+(A) The threshold tumor size $N_a^*$ as a function of the aneuploid growth rate $\Delta_a$.
 The dashed horizontal line shows $N^*_m$, the threshold tumor size without aneuploidy ($u=0$).
-The red dots represent numerical simulations.
+When aneuploid growth rate is close to or higher than zero, aneuploidy decreases the threshold tumor size, thereby facilitating evolutionary rescue.
 The inset highlights the case when aneuploidy cancer cells are non-growing.
-When aneuploid growth rate is close to or higher than zero, aneuploidy decreases the threshold tumor size, thereby facilitating evolutionary rescue.
+The red dots represents simulations and the error bars represent the $95\%$ confidence intervals obtained with bootstrapping, see Appendix G.
 Parameters: $\lambda_w=0.1,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.
-\textbf{(B)} The threshold tumor size $N_a^*$ as a function of the ratio of aneuploidy and mutation rates, $u/v$.
+(B) The threshold tumor size $N_a^*$ as a function of the ratio of aneuploidy and mutation rates, $u/v$.
 The dashed horizontal line shows $N^*_m$, the threshold tumor size without aneuploidy ($u=0$).
+When the aneuploidy rate is much higher than the mutation rate, aneuploidy decreases the threshold tumor size, thereby facilitating evolutionary rescue.
 The blue line represents the exact formula for threshold tumor size $N_a^*$ while the solid black line represents the approximation \cref{eq:N_a}.
-The red dots represent numerical simulations.
-When the aneuploidy rate is much higher than the mutation rate, aneuploidy decreases the threshold tumor size, thereby facilitating evolutionary rescue.
+The red dots represents simulations and the error bars represent the $95\%$ confidence intervals obtained with bootstrapping, see Appendix G.
 Parameters: $\lambda_w=0.1,\lambda_m=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, v=10^{-7}$.
-For the threshold tumor sizes the error bars represent the $95\%$ confidence interval obtained through bootstrapping in the following steps: \textbf{(1)} we simulate $p_{rescue}$ 100 times; \textbf{(2)} we sample with replacement which we store in $S$; \textbf{(4)} for each element of this sample we obtain $N_a^*=1/p_w$ using $p_w=-1/N_s\log\left(1-\bar{S}\right)$ where $\bar{S}$ is the mean of $S$ and $N_s$ is an arbitrary value of the initial population size we selected in order to calculate $p_{rescue}$; \textbf{(5)} we repeat steps \textbf{(2)}-\textbf{(4)} 100 times to obtain $N_a^*$ and we select the upper and lower limits such that $95\%$ of the values of $N_a^*$ lie in the interval given by the bounds.
 }
@@
 \textbf{Standing genetic variation facilitates evolutionary rescue of cancer under drug treatment.}
-\textbf{(A)}  The ratio of threshold tumor size $\tilde{N}_a^*$ when a fraction $\frac{\tilde{u}\lambda_w}{s}$ is aneuploid at the start of treatment and $N_a^*$ as a function of the wildtype growth rate $\Delta_w$.
-The red dots represents numerical simulations and the error bars represent the $95\%$ confidence intervals obtained with bootstrapping explained below.
+(A)  The ratio of threshold tumor size $\tilde{N}_a^*$ when a fraction $\frac{\tilde{u}\lambda_w}{s}$ is aneuploid at the start of treatment and $N_a^*$ as a function of the wildtype growth rate $\Delta_w$.
 Standing genetic variation will drive adaptation to the drug if $\Delta_w$ is very negative due to a stronger effect of the drug on sensitive cells.
+The red dots represents simulations and the error bars represent the $95\%$ confidence intervals obtained with bootstrapping, see Appendix G.
 Parameters: $\lambda_w=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_a=0.09,\mu_m=0.09,\tilde{u}=10^{-3},u=10^{-2}, v=10^{-7}$.
-\textbf{(B)} The ratio of threshold tumor size $\tilde{N}_a^*$ when a fraction $\frac{\tilde{u}\lambda_w}{s}$ is aneuploid at the start of treatment and $N_a^*$ as a function of the the ratio of aneuploidy rates $\tilde{u}/u$.
-The red dots represents numerical simulations and the error bars represent the $95\%$ confidence intervals obtained with bootstrapping explained below.
+(B) The ratio of threshold tumor size $\tilde{N}_a^*$ when a fraction $\frac{\tilde{u}\lambda_w}{s}$ is aneuploid at the start of treatment and $N_a^*$ as a function of the the ratio of aneuploidy rates $\tilde{u}/u$.
 De-novo aneuploids will have larger contribution to the appearance of drug resistance if the drug induces the appearance of aneuploid cells ($u \gg \tilde u$).
+The red dots represents  simulations and the error bars represent the $95\%$ confidence intervals obtained with bootstrapping, see Appendix G.
 Parameters: $\lambda_w=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09,\tilde{u}=10^{-3}, v=10^{-7}$.
-For the ratio of the threshold tumor sizes the error bars represent the $95\%$ confidence interval obtained through bootstrapping in the following steps: \textbf{(1)} we simulate $p_{rescue}$ 100 times for both the case when $f=\tilde{u}\lambda_w/s$ and $f=0$; \textbf{(2)} we sample with replacement which we store in $S_0$ and  $S_f$; \textbf{(4)} for each element of $S_0$ we obtain $N_a^*=1/p_w$ using $p_w=-1/N_s\log\left(1-\bar{S}\right)$ where $\bar{S}$ is the mean of $S_0$ and $N_s$ is an arbitrary value of the initial population size we selected in order to calculate $p_{rescue}$; \textbf{(5)} for each element of $S_f$ we obtain $\tilde{N}_a^*=1/p_a$ using $p_a=-f/N_s\log\left(1-\bar{S}\right)$ where $\bar{S}_f$ is the mean of $S_f$ and $N_s$ is an arbitrary value of the initial population size we selected in order to calculate $p_{rescue}$; \textbf{(6)} we repeat steps \textbf{(2)}-\textbf{(5)} 100 times to obtain $\tilde{N}_a^*/N_a^*$ and we select the upper and lower limits such that $95\%$ of the values of $\tilde{N}_a^*/N_a^*$ lie in the interval given by the bounds.
 }
@@
 \caption{\textbf{Aneuploidy has a impact on cancer relapse early on.}
-\textbf{(A)} Plot of the probability that a successful mutant has not appeared by time $t$.
+(A) The probability that a successful mutant has not appeared by time $t$.
 The green line represents the case with tolerant aneuploidy ($u>0, \lambda_a=0.0899$), the blue line represents the case with non-growing aneuploidy ($u>0, \lambda_a=0.089999$), the cyan line represents the case with partially resistant aneuploidy ($u>0, \lambda_a=0.095$) and the black line represents the case without aneuploidy ($u=0$).
-The  dots represent numerical simulations.
-For the simulations we have chosen the following parameters: $\lambda_w=0.1,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7},N=10^7$.
-The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the  probability of that a successful mutant has not been generated and $n=100$ is the number of simulations.
 As time increases, aneuploidy plays an important role in helping the cancer cell population escape extinction.
-\textbf{(B)} Plot of the probability that a mutant cancer cell population has not reached size $N$ at time $t$.
-The green line represents the case where $N=10^6$, the red line represents $N=10^7$ and the blue line represents the case where $N=10^{10}$.
-Increasing the initial tumor size guarantees that the tumor recurrence.
-The markers represent stochastic simulations and the error bars represents $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the probability of that a mutant population size has not reached $N$ and $n=100$ is the number of simulations.
+The markers represent simulations and the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the fraction of simulations in which a successful mutant has not been generated and $n=100$ is the number of simulations.
+Parameters: $\lambda_w=0.1,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7},N=10^7$.
+(B) The probability that a mutant cancer cell population has not reached size $N$ at time $t$.
+The green line represents the case where $N=10^6$ (small tumor), the red line represents $N=10^7$ (intermediate sized tumor) and the blue line represents the case where $N=10^{10}$ (large tumor).
+Increasing the initial tumor size guarantees that the tumor will regrow.
+The markers represent simulations and the error bars represents $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the fraction of the simulations in which the mutant population size has not reached $N$ and $n=100$ is the number of simulations.
 Parameters: $\lambda_w=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}
@@
 \includegraphics[width=1\textwidth]{Figures/ProliferationTime.pdf}
-\caption{\textbf{Mean recurrence time is a decreasing function of the initial tumor size.}
-Shown is the mean time for the mutant cell population to reach size $N$, where $N$ is the initial number of cancer cells.
-Our inhomogeneous Poisson-process approximation (solid black line, \cref{meanproliferationtime}) is in agreement with simulation results (red markers with 95\% quantile intervals) for intermediary $N$.
-The numerical simulations converge to the solution of \cref{eq:t2det} (blue dashed line) for large values of $N$.
+\caption{\textbf{Tumor size decreases the mean recurrence time.}
+The mean time for the mutant cell population to reach size $N$, the initial number of cancer cells.
+Our inhomogeneous Poisson-process approximation (solid black line, \cref{meanproliferationtime}) is in agreement with simulation results (red markers with 95\% confidence interval obtained with bootstrapping, see Appendix G) for intermediary $N$.
+The simulations converge to \cref{eq:t2det} (blue dashed line) for large values of $N$.
 Parameters: $\lambda_w=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}
@@
 
-\begin{figure}
-\vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/KaplanMeierDistribution.pdf}
-\caption{\textbf{Distribution of the recurrence time.}
-Shown is the distribution of the time for the mutant cell population to reach size $N$, where $N$ is the initial number of cancer cells.
-The red line is analytic result \cref{distribution} overlaid over the histogram of numerical simulations.
-Parameters: $N=10^6, \lambda_w=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}
-\label{KMdistribution}
-\end{figure}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 The probability density function of $T_m$ is thus
-$R_m\left(t\right)\exp\left(-\int_0^{t}{R_1(z)\d z}\right)$ \citep{allen2010introduction}. 
+$R_m\left(t\right)\exp\left(-\int_0^{t}{R_m(z)\d z}\right)$ \citep{allen2010introduction}. 
 Therefore, the probability density function of the conditional random variable $(T_m \mid T_m < \infty)$ is
@@
 
+%%%
+If $N\gg N_m^*$ then the mean time $\tau_a$ can be written as:
+\begin{align*}
+\tau_a=\int_0^\infty\e^{-R_a\left(\tau\right)}\,\d\tau=\int_0^\infty\exp\left[-\frac{uv\lambda_w\lambda_aNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_w\tau}-1}{\Delta_w}-\frac{\e^{\Delta_a\tau}-1}{\Delta_a}\right)-v\lambda_wNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}\right]\,\d\tau,
+\end{align*}
+and we use the following Taylor series expansions:
+\begin{align*}
+\frac{\e^{\Delta_w\tau}-1}{\Delta_w}&=\frac{1+\Delta_w\tau+O(\tau^2)-1}{\Delta_w}=\tau+O(\tau^2).\\
+\frac{\e^{\Delta_a\tau}-1}{\Delta_a}&=\frac{1+\Delta_a\tau+O(\tau^2)-1}{\Delta_a}=\tau+O(\tau^2),
+\end{align*}
+to obtain a simpler approximation for $\tau_a$:
+\begin{align}\label{limitapprox3}
+\tau_a\approx\int_0^\infty\e^{-v\lambda_wNp_m\tau}\,\d\tau=\frac{1}{v\lambda_wNp_m}.
+\end{align}
+If $N\ll N_a^*$ then we can write \Cref{meantimet2} as:
+\begin{align}\nonumber
+\tau_a&\approx\frac{\int_0^\infty tv\lambda_ap_ma_\tau\,\d\tau}{1-\e^{-Np_w}}\approx\frac{uv\lambda_a\lambda_wp_m\abs{\Delta_w+\Delta_a}}{p_w\Delta_a^2\Delta_w^2}\\ \label{limitapprox4}
+&=\frac{1}{\abs{\Delta_w}}+\frac{1}{\abs{\Delta_a}},
+\end{align}
+where in the last line we used the fact that $1/p_w=N_a^*$ and \Cref{eq:N_a}.
+
+If a fraction $f$ of the cancer cells are aneuploid when the drug is administered then the rates at which the rescue mutations are generated can be written as:
+\begin{align*}
+r_1^f\left(t\right)&=v\lambda_ap_m\int_0^ta_{z} \d z = \left(1-f\right)\frac{uv\lambda_w\lambda_aNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_wt}-1}{\Delta_w}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right)+fv\lambda_aNp_m\frac{\e^{\Delta_a t}-1}{\Delta_a},\\
+r_2^f\left(t\right)&=v\lambda_wp_m\int_0^tw_{z} \d z = \left(1-f\right)v\lambda_wNp_m\frac{\e^{\Delta_w t}-1}{\Delta_w},
+\end{align*} 
+and the mean evolutionary rescue time is given by:
+\begin{align}\label{meantimet2SGV}
+\tau_a &= \frac{\int_{0}^{\infty}{tR_a^f(t)\exp\left(-\int_0^{t}{R_a^f(z)\d z}\right) \d t}}{p_{rescue}},
+\end{align}
+where $R_a^f(t)=r_1^f\left(t\right)+r_2^f\left(t\right)$ and $p_{rescue}=1-\exp\left[-\left(1-f\right)p_wN-fp_aN\right]$.
+We plot our approximation in  \Cref{SGVEvolutionaryRescueTimeComplete} together with simulated data.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \begin{equation*}
-r_1\left(\infty\right)=\frac{uv\lambda_w\lambda_aNp_m}{\abs{\Delta_w}\abs{\Delta_a}},
+r_1\left(\infty\right)=\frac{uv\lambda_w\lambda_aNp_m}{\abs{\Delta_w}\abs{\Delta_a}}=\frac{N}{N_a^*},
 \end{equation*}
-where we ignore lineages created by direct mutation because we assumed $u\lambda_a > \max{(-\Delta_a, 1/T^*)}$.
+where we ignore lineages created by direct mutation because we assumed $u\lambda_a > \max{(-\Delta_a, 1/T^*)}$, $N\ll N_m^*$ and used \Cref{eq:N_a}.
 
@@
 \begin{equation*}
-N\ll\frac{\abs{\Delta_w}\abs{\Delta_a}}{uv\lambda_w\lambda_ap_m}.
+N\ll N_a^*.
 \end{equation*}
@@
 \begin{align*}
-N\gg\frac{\abs{\Delta_w}\abs{\Delta_a}}{uv\lambda_w\lambda_ap_m}.
+N\gg N_a^*.
 \end{align*}
@@
 We obtain $\tau_a^r$ such that $m\left(\tau_a^r\right)=N$ by solving:
+\begin{equation}%\label{eq:t2det}
+1=\frac{uv\lambda_a\lambda_w}{\Delta_w-\Delta_a}\left[\frac{\e^{\Delta_w\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_w-\Delta_m}-\frac{\e^{\Delta_a\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_a-\Delta_m}\right]+v\lambda_w\frac{\e^{\Delta_w\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_w-\Delta_m}.
+\end{equation}
+This is a transcendental equation which cannot be solved exactly but we can obtain an approximation by noting that for large $\tau_a^r$ the above equation can be written as: 
+\begin{equation*}
+1=v\lambda_w\frac{\e^{\Delta_m\tau_a^r}}{\abs{\Delta_w-\Delta_m}},
+\end{equation*}
+which has solution:
 \begin{equation}\label{eq:t2det}
-1=\frac{uv\lambda_a\lambda_w}{\Delta_w-\Delta_a}\left[\frac{\e^{\Delta_w\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_w-\Delta_m}-\frac{\e^{\Delta_a\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_a-\Delta_m}\right]+v\lambda_w\frac{\e^{\Delta_w\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_w-\Delta_m}.
-\end{equation}
-This is a transcendental equation which cannot be solved analytically but can solved numerically using one of the root solver algorithm in $\bold{Python}$. 
+\tau_a^r\approx\frac{1}{\Delta_m}\log\frac{\Delta_m-\Delta_w}{v\lambda_w}.
+\end{equation}
+We observe that the terms given by the evolutionary trajectory $wildtype \rightarrow aneuploid \rightarrow mutant$ do not contribute to the above approximation and, as a result, we deduce that it accurate only for $N\gg N_m^*>N_a^*$.
+
+Additionally, we note that if we are interested in the time until the tumor reaches a detectable size $M$ then our above analysis is valid but in \Cref{meanproliferationtime} we change:
+\begin{align}\label{meanproliferationtime2}
+\tau_a^{r,M}&\approx\tau_a+\frac{\log p_mM}{\Delta_m},
+\end{align}
+and \Cref{eq:t2det} becomes:
+\begin{equation}\label{eq:t2det2}
+\tau_a^{r,M}\approx\frac{1}{\Delta_m}\log\frac{M\left(\Delta_m-\Delta_w\right)}{v\lambda_wN},
+\end{equation}
+which we plot in \Cref{RecurrencePlot} and observe that our approximations are in agreement with simulations.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 As a result, the number of mutant cancer cells will reach $N$ faster then the case with a single mutant lineage.
+Additionally, we observe from \Cref{cdffig}B that the probability of the mutant cell population reaching size $N$ is approximately zero before time $\tau_a^r$ which is the recurrence time for the deterministic case.
+This can be explained as follows: in the deterministic case there is a sufficient number of lineages produced such that there exists a lineage where each descendant will only reproduce and not die; the time it takes for this lineage to reach $N$ is the lower bound for the time of all other lineages to reach $N$ and this time cannot be smaller then $\tau_a^r$ by definition.
+Given that for small values of $N$ we expect that at most a single lineage will rescue the tumor, this lineage cannot reach $N$ before $\tau_a^r$ for the deterministic case \cref{eq:t2det}.
 
@@
 \end{align}
-which we plot in Figure and compare with numerical simulations.
+which we plot in Figure and compare with simulations.
 We note that in the case $N\gg\abs{\Delta_w}\abs{\Delta_a}/uv\lambda_w\lambda_ap_m$ the distribution becomes the Dirac $\delta$-function~\citep{barton1989elements}.
+%%%
+\section*{Appendix G: Bootstrapping}
+For the mean times the $95\%$ confidence interval is obtained through bootstrapping in the following steps: (1) we simulate $T$ 100 times; (2) we sample with replacement which we store in $T'$; (3) for each element of this sample we obtain $\tau=\mathbb{E}\left[T'\right]$; (4) we repeat steps (2)-(3) 100 times to obtain $\tau$ and we select the upper and lower limits such that $95\%$ of the values of $\tau$ lie in the interval given by the bounds.
+
+For the threshold tumor sizes the $95\%$ confidence interval is obtained through bootstrapping in the following steps: (1) we simulate $p_{rescue}$ 100 times; (2) we sample with replacement which we store in $S$; (3) for each element of this sample we obtain $N_a^*=1/p_w$ using $p_w=-1/N_s\log\left(1-\bar{S}\right)$ where $\bar{S}$ is the mean of $S$ and $N_s$ is an arbitrary value of the initial population size we selected in order to calculate $p_{rescue}$; (4) we repeat steps (2)-(3) 100 times to obtain $N_a^*$ and we select the upper and lower limits such that $95\%$ of the values of $N_a^*$ lie in the interval given by the bounds.
+
+For the ratio of the threshold tumor sizes the $95\%$ confidence interval is obtained through bootstrapping in the following steps: (1) we simulate $p_{rescue}$ 100 times for both the case when $f=\tilde{u}\lambda_w/s$ and $f=0$; (2) we sample with replacement which we store in $S_0$ and  $S_f$; (4) for each element of $S_0$ we obtain $N_a^*=1/p_w$ using $p_w=-1/N_s\log\left(1-\bar{S}\right)$ where $\bar{S}$ is the mean of $S_0$ and $N_s$ is an arbitrary value of the initial population size we selected in order to calculate $p_{rescue}$; (5) for each element of $S_f$ we obtain $\tilde{N}_a^*=1/p_a$ using $p_a=-f/N_s\log\left(1-\bar{S}\right)$ where $\bar{S}_f$ is the mean of $S_f$ and $N_s$ is an arbitrary value of the initial population size we selected in order to calculate $p_{rescue}$; (6) we repeat steps (2)-(5) 100 times to obtain $\tilde{N}_a^*/N_a^*$ and we select the upper and lower limits such that $95\%$ of the values of $\tilde{N}_a^*/N_a^*$ lie in the interval given by the bounds.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \caption{\textbf{Density dependent growth does not affect the accuracy of our model.} Comparison of results of simulations  with density-dependent growth (red markers with with 95\% CI) and the approximation formula (black line, \cref{eq:N_a} in \cref{eq:rescue_prob}) with maximum carrying capacity $K=10^8$ and effective carrying capacity $K_e=K\Delta_a/\lambda_a\approx10^6$.
-The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the  probability that the tumor has adapted to the stress and $n=100$ is the number of simulations.
+The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the  fraction of simulations in which the tumor has adapted to the stress and $n=100$ is the number of simulations.
 Parameters: $\lambda_w=0.1,\lambda_a=0.0901,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}, K=10^8$.}
@@
 \caption{\textbf{Evolutionary rescue time.}
-Shown is the mean time for appearance of a resistance mutation the leads to evolutionary rescue \textbf{(left)} with aneuploidy ($u>0$) and \textbf{(right)} without aneuploidy ($u=0$).
-Our inhomogeneous Poisson-process approximations (solid black lines, right: \cref{eq:meantime1}, left: \cref{meantimet2}) is in agreement with simulation results (red markers with 95\% quantile intervals).
+Shown is the mean time for appearance of a resistance mutation the leads to evolutionary rescue (A) with aneuploidy ($u>0$) and (B) without aneuploidy ($u=0$).
+Our inhomogeneous Poisson-process approximations (solid black lines, right: \cref{eq:meantime1}, left: \cref{meantimet2}) is in agreement with simulation results (red markers with 95\% quantile intervals obtained with bootstrapping, see see Appendix G).
 Parameters: $\lambda_w=0.1,\lambda_m=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.
@@
 \label{ExpectedNumberRescueLineages}
+\end{figure}
+%%%
+\begin{figure}
+\vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/KaplanMeierDistribution.pdf}
+\caption{\textbf{Distribution of the recurrence time.}
+Shown is the distribution of the time for the mutant cell population to reach size $N$, where $N$ is the initial number of cancer cells.
+The red line is analytic result \cref{distribution} overlaid over the histogram of simulations.
+Parameters: $N=10^6, \lambda_w=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}
+\label{KMdistribution}
 \end{figure}
@@
 The black line represents the case with tolerant aneuploidy a fraction $f=0\%$ of the initial tumor being aneuploid, the red line represents the case with a fraction $f=5\%$ and the green line represents the case with $f=50\%$.
-The dots represent numerical simulations and the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the probability of that the tumor has adapted to the stress and $n=100$ is the number of simulations.
+The dots represent simulations and the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the probability of that the tumor has adapted to the stress and $n=100$ is the number of simulations.
 Parameters: $\lambda_w=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}
@@
 \end{appendices}
-
+%%%
+\begin{figure}
+\vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/EvolutionaryRescueTimeComplete.pdf}
+\caption{Shown is the mean time for appearance of a resistance mutation the leads to evolutionary rescue with aneuploidy ($u>0$).
+Our inhomogeneous Poisson-process approximations (solid black lines, right: \cref{meantimet2}) is in agreement with simulation results (red markers with 95\% confidence intervals obtained with bootstrapping, see Appendix G).
+Dashed vertical blue line represents the threshold tumor size above which evolutionary rescue is very likely through aneuploidy \cref{eq:N_a} and the dashed vertical green line represents the threshold tumor size above which evolutionary rescue is very likely through direct mutation \cref{eq:N_m}.
+Solid lines represents the approximations \cref{eq:AsymptoticTimeRules} ($N<N_a^*$ blue line and $N>N_m^*$ green line).
+Parameters: $\lambda_w=0.1,\lambda_m=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}
+\label{EvolutionaryRescueTimeComplete}
+\end{figure}
+%%%
+\begin{figure}
+\vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/ProliferationTimeLarge.pdf}
+\caption{The mean time for the mutant cell population to reach size $N$, where $N$ is the initial number of cancer cells.
+Our inhomogeneous Poisson-process approximation (solid black line, \cref{meanproliferationtime}) is in agreement with simulation results (red markers with 95\% confidence intervals obtained with bootstrapping, see Appendix G) for small and intermediate values of $N$.
+Dashed vertical blue line represents the threshold tumor size above which evolutionary rescue is very likely through aneuploidy \cref{eq:N_a} and the dashed vertical green line represents the threshold tumor size above which evolutionary rescue is very likely through direct mutation \cref{eq:N_m}.
+Solid lines represents the approximations \cref{eq:AsymptoticTimeRulesRecurrence} ($N<N_a^*$ blue line and $N>N_m^*$ green line).
+The simulations converge to the solution of \cref{eq:t2det} (green line) for large values of $N\gg N_m^*$.
+Parameters: $\lambda_w=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}
+\label{ProliferationTimeLarge}
+\end{figure}
+%%%
+\begin{figure}
+\vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/RecurrencePlot.pdf}
+\caption{The mean time for the mutant cell population to reach size $N$, where $N$ is the initial number of cancer cells.
+Our inhomogeneous Poisson-process approximation (solid black line, \cref{meanproliferationtime}) is in agreement with simulation results (red markers with 95\% confidence intervals obtained with bootstrapping, see Appendix G) for small and intermediate values of $N$.
+Dashed vertical blue line represents the threshold tumor size above which evolutionary rescue is very likely through aneuploidy \cref{eq:N_a} and the dashed vertical green line represents the threshold tumor size above which evolutionary rescue is very likely through direct mutation \cref{eq:N_m}.
+Solid lines represents the approximations \cref{eq:AsymptoticTimeRulesRecurrence} ($N<N_a^*$ blue line and $N>N_m^*$ green line).
+The simulations converge to the solution of \cref{eq:t2det} (green line) for large values of $N\gg N_m^*$.
+Parameters: $\lambda_w=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}
+\label{RecurrencePlot}
+\end{figure}
+%%%
+\begin{figure}
+\vspace*{1\baselineskip}
+\includegraphics[width=1\textwidth]{Figures/SGVEvolutionaryRescueTimeComplete.pdf}
+\caption{Shown is the mean time for appearance of a resistance mutation the leads to evolutionary rescue with aneuploidy ($u>0$).
+Our inhomogeneous Poisson-process approximations (solid black lines, right: \cref{meantimet2}).
+Dashed black line is the  inhomogeneous Poisson-process approximation \cref{meantimet2SGV} where a fraction $f$ of tumor is aneuploid at the onset of drug therapy which is in agreement with simulation results (red markers with 95\% confidence intervals obtained with bootstrapping, see Appendix G).
+Dashed vertical blue line represents the threshold tumor size above which evolutionary rescue is very likely through aneuploidy \cref{eq:N_a} and the dashed vertical green line represents the threshold tumor size above which evolutionary rescue is very likely through direct mutation \cref{eq:N_m}.
+Solid lines represents the approximations \cref{eq:AsymptoticTimeRules} ($N<N_a^*$ blue line and $N>N_m^*$ green line).
+Parameters: $\lambda_w=0.1,\lambda_m=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7},f=1.36\%$.}
+\label{SGVEvolutionaryRescueTimeComplete}
+\end{figure}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2024-01-23 EvolutionaryRescue1.tex
revised abstract and part of intro

@@
 \begin{abstract}
-Evolutionary rescue is the process by which a population is able to survive a sudden environmental change which initially causes the population to decline towards extinction.
-A prime example of evolutionary rescue is the ability of cancer to survive being exposed to various treatments.
-We are interested in the mechanisms through which a population of cancer cells are able to adapt to chemotherapy, and in particular, the role played by chromosomal instability (aneuploidy).
-Cancer cells which have aneuploidy are hypothesized to have a higher fitness in an environment altered by anti-cancer drugs as they have incomplete pathways which drugs activate in order to kill the cells.
-Aneuploidy is highly prevalent in tumors and certain drugs which attempt to combat cancers through increasing chromosomal instability.
-As a result, the question we wish to answer is how aneuploidy impacts the fate of the population of cancer cells.
-We propose to model evolutionary rescue with the help of multi-type branching processes to obtain the probability that cancer will survive.
-We obtain analytic expressions for the probability of evolutionary rescue as a function of the initial tumor size, the growth rates of the different cancer genotypes.
-Additionally we obtain the mean recurrence time for the tumor to reach its initial size.
-We observe that aneuploidy plays an important role in the relapse of secondary tumors who have not yet been detected and thus are smaller in size.
+Evolutionary rescue is the process by which a population survives a sudden environmental change that initially causes the population to decline towards extinction.
+A prime example of evolutionary rescue is the ability of cancer to survive exposure to various treatments. 
+One evolutionary mechanism by which a population of cancer cells can to adapt to chemotherapy is aneuploidy. 
+Aneuploid cancer cells can have higher fitness in an environment altered by anti-cancer drugs, e.g., because of incomplete pathways targeted by the drugs. 
+Indeed, aneuploidy is highly prevalent in tumors, and moreover, some anti-cancer drugs fight cancer by increasing chromosomal instability.
+Here, we examine how aneuploidy impacts the fate of a population of cancer cells.
+We use multi-type branching processes to approximate the probability that a tumor survives drug treatment as a function of the initial tumor size, and rates at which aneuploidy and other beneficial mutations occur, and the growth rates of the sensitive and resistant cells.
+We also approximate the mean recurrence time for the tumor to revert back to its initial size.
+We find that aneuploidy can play an important role in the relapse of secondary tumors who have not yet been detected and thus are smaller in size.
+% TODO missing something here...
 \end{abstract}
+
+Keywords: whole-chromosome duplication, evolutionary model, adaptive evolution, cancer, drug resistance, chromosome instability
 
@@
 
-% OVERVIEW OF INTRODUCTION:
-% - background on CIN in cancer
-% - what hasn't been done? (aneuploidy + drug resistance)
-% - why is that important?
-% - how we tackle this background
-% - summary of our analysis
-% history of evolutionary rescue literature (that is not specific to cancer) should move to literature, and we should have a clear statement on what we add to that literature.
-
-% TODO two papers that show aneuploidy provides resistance Ok
-% Lukow, Devon A., Erin L.
-Sausville, Pavit Suri, Narendra Kumar Chunduri, Angela Wieland, Justin Leu, Joan C.
-Smith, et al.
-2021.
-“Chromosomal Instability Accelerates the Evolution of Resistance to Anti-Cancer Therapies.” Developmental Cell 56 (17): 2427-2439.e4.
-https://doi.org/10.1016/j.devcel.2021.07.009.
-% Rutledge, Samuel D., Temple A.
-Douglas, Joshua M.
-Nicholson, Maria Vila-Casadesús, Courtney L.
-Kantzler, Darawalee Wangsa, Monika Barroso-Vilares, Shiv D.
-Kale, Elsa Logarinho, and Daniela Cimini.
-2016.
-“Selective Advantage of Trisomic Human Cells Cultured in Non-Standard Conditions.” Scientific Reports 6 (June 2015): 1–12.
-https://doi.org/10.1038/srep22828.
-
-\paragraph{Aneuploidy in cancer.}  Each year approximately 10 million people die from cancer \citep{kocarnik2022cancer}, so understanding what factors contribute to failure of interventions is of great importance.
-One hypothesized factor is aneuploidy, where cells are characterized by structural changes of the chromosomes and copy number alterations \citep{schukken2018cin}, which is caused by chromosomal instability (CIN), the mitotic process in which cells suffer from chromosome mis-segregation that leads to aneuploidy.
-Interestingly, aberrations in chromosome copy number have been shown to allow cancer cells to survive under stressful conditions such as drug therapy \citep{lukow2021chromosomal,rutledge2016selective}.
+\paragraph{Aneuploidy in cancer.}  Each year approximately 10 million people die from cancer \citep{kocarnik2022cancer}, so understanding that factors that contribute to failure of interventions is of great importance.
+One hypothesized factor is aneuploidy, where cells are characterized by an imbalanced karyotype and chromosome copy number alterations \citep{schukken2018cin}, caused by chromosomal instability (CIN), the mitotic process in which cells suffer from chromosome mis-segregation that leads to aneuploidy.
+Importantly, aberrations in chromosome copy number have been shown to allow cancer cells to survive under stressful conditions such as drug therapy \citep{lukow2021chromosomal,rutledge2016selective}.
 Indeed, cancer cells are often likely to be aneuploid, and aneuploidy is associated with poor patient outcomes \citep{ben2020context,smith2018systematic}.
@@
  \citet{ippolito2021gene} induced aneuploidy in cancer cell lines by exposing them to reversine, a small-molecule inhibitor of the mitotic kinase Mpsi1, and then to chemotherapeutic agents such as vemurafenib.
-When compared to wildtype cancer cell the reversine treated cells had higher proliferation in the environment altered by anti-cancer drugs.
-Similarly, \citet{lukow2021chromosomal} observed that cancer cells which have been induced to be aneuploid have an advantage when compare to wildtype even while they had lower fitness before the onset of chemotherapy.
-A proposed mechanism through which aneuploidy is able to confer resistance to chemotherapeutics is by antagonizing cell division  which prevents the drugs from damaging DNA and microtubules \citep{replogle2020aneuploidy}
-
- An important aspect of aneuploidy is the rate with which cells acquire it, which is several orders of magnitude higher when compare to the mutation rate \citep{bakker2023predicting}.
-Consequently, a cell exposed to a stress such as chemotherapeutic drugs can acquire aneuploidy faster when compared to acquiring a mutation, especially when several proposed anti-cancer drugs elevate the rate of missegration in order to combat cancer \citep{lee2016effects}.
+Reversine-treated cells had higher proliferation rate in the environment altered by anti-cancer drugs compared to wildtype cancer cells.
+ Similarly, \citet{lukow2021chromosomal} induced aneuploidy in cancer cells and observed that such cells have an advantage compared to wildtype cells during chemotherapy, despite having lower fitness before the onset of chemotherapy.
+One proposed mechanism through which aneuploidy is able to confer resistance to chemotherapeutics is by antagonizing cell division, which prevents the drugs from damaging DNA and microtubules \citep{replogle2020aneuploidy}.
+
+An important aspect of aneuploidy is the rate with which cells become aneuploid, which is several orders of magnitude higher than the beneficial mutation rate \citep{bakker2023predicting}.
+Consequently, a cell exposed to a stress such as chemotherapeutic drugs can acquire aneuploidy faster when compared to acquiring a mutation, especially when several proposed anti-cancer drugs elevate the rate of mis-segregation in order to fight cancer \citep{lee2016effects}.
 

### 2024-01-23 EvolutionaryRescue1.tex
add comments to manuscript

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Josh's comments
+
+% To me the big picture of the paper isn't clear.
+I appreciate that the model can replicate what one expects, but what are the clinical implications, or surprising results?
+Which of the scenarios are most realistic given the available clinical and experimental data?
+What are the biological interpretations of the derived formulate
+
+% it's also not clear to me from the introduction that this is a novel model, which it is, no?
+
+% Big picture:
+% (1) Any time you say "we observe" or "interestingly" it sets up the expectation for biological interpretation - going through the paper and ensuring that such interpretation is present will help a lot
+% (2) What's the overall story of the paper beyond we made a model and it does what we expect.
+I'm not saying that there isn't more to the story here, I think there is, but it's not immediately clear to me what that more is.
+% (3) adding interpretation to figure discussion/caption.
+For example, in figure 3  (to me, but perhaps I'm wrong) it is unsurprising that the probability of rescue increases with population size, and that partially resistant saturates faster than stationary which is faster than partially tolerant.
+It seems to me the main message of the figure right now is that the presence of aneuploidy in turmor cells, even if only partially tolerant to treatment, facilitates rescue.
+However, it seems like the more crucial message in the figure is the number of cells at which the curves saturates.
+Perhaps one could plot the saturation level in terms of number of cancer cells as a function of the parameter value(s) which define each scenario.
+or perhaps some combination of the N*s and the saturation points
+
+% other comments:
+% figures 2, 3 shouldn't have the same bolded caption 
+%
+
+% some line item comments are below
 \begin{abstract}
@@
 Given that aneuploidy is present in many tumors even before the onset of therapy \citep{lukow2021chromosomal} we also take into consideration the effect that standing genetic variation has on the tumor dynamics.
-
+% It feels like conclusions are missing from this paragraph, we did these things and found what?...
+% from the abstract "aneuploidy can play an important role in the relapse of secondary tumors who have not yet been detected and thus are smaller in size." is this not already suggested by the sources in paragraph two above?
+% This paragraph should make clear the contribution of this paper, to me at the moment it's not...
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 We set the mutation rate $v$ to be equal to $10^{-7}$ per gene per cell division \citep{loeb2001mutator} and since we assume that a single target gene confers immunity to chemotherapy we select $v=10^{-7}$ per cell division.
+% the above sentence feels repetitive
 The missegregation rate is chosen from \citet{bakker2023predicting} who determined \textit{in silico} that it must be in the interval $10^{-3}-10^{-2}$ per chromosome per cell division.
@@
 \end{equation}
+% the bracket here is meant to denote multiplication, approximation, what?
+% Can you interpret this more directly in a biological sense?
 In all cases, the effect of aneuploidy increases (i.e., the threshold size ratio decreases) when the aneuploidy rate $u$ increases (\Cref{rescue_threshold}B).
@@
 Interestingly, the threshold between small and intermediate tumors is similar to the detection threshold of $4.19\times10^6$ for a wide variety of tumors as reported by \citet{avanzini2019cancer}.
-
+% might just be because I'm not familiar with this area of biology, but can this be interpreted further?
+i.e.
+what are the consequences, if any, of this?
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 Interestingly, for small tumors the mean evolutionary rescue time is a function of the wildtype and aneuploid growth rates and independent of all of other model parameters including tumor size $N$ (\Cref{EvolutionaryRescueTimeComplete}).
+% might just be because I'm not familiar with this area of biology, but why is this interesting, does it defy a priori expectations?
 Increasing either the wildtype or aneuploid growth rates lead to an increase in the mean time.
@@
 We note that standing genetic variation reduces the mean evolutionary rescue time only for intermediary sized tumors but not for large or small tumors.
-\paragraph{Recurrence time.}
+\paragraph{Recurrence time.} % and? whats the further interpretation of this?
 
@@
 Consequently, patients which take such drugs require a longer period of monitoring to guarantee the effectiveness of the treatment.
+% this is good, more things like this!
 \paragraph{Distribution of evolutionary rescue time with and without aneuploidy.}
@@
 We observe that aneuploidy start to have an effect on adaptation for timescales greater then 100 days.
+% and?
+so what...
 
@@
 In that case, evolutionary rescue occurs through a single step by the aneuploid cells acquiring a beneficial mutation and the probability of evolutionary rescue is given by $1-\exp\left(-N/\tilde{N}_a^*\right)$  (\Cref{rescue_prob_sgv}).
+% maybe pull some of this out for the last paragraph of the introduction?
 
@@
 We compare to the predictions of our model to see if the tumors with initial size bellow the threshold \eqref{eq:N_a} are more likely to go extinct.
+% I may need to reread, but what are the specific predictions in this paper that can be tested by experiments?
 

### 2024-02-01 EvolutionaryRescue1.tex
Add files via upload

--- 2024-01-23 EvolutionaryRescue1.tex
@@
 
-\paragraph{Aneuploidy in cancer.}  Each year approximately 10 million people die from cancer \citep{kocarnik2022cancer}, so understanding that factors that contribute to failure of interventions is of great importance.
+\paragraph{Aneuploidy in cancer.}  Each year approximately 10 million people die from cancer \citep{kocarnik2022cancer}, so understanding the factors that contribute to failure of interventions is of great importance.
 One hypothesized factor is aneuploidy, where cells are characterized by an imbalanced karyotype and chromosome copy number alterations \citep{schukken2018cin}, caused by chromosomal instability (CIN), the mitotic process in which cells suffer from chromosome mis-segregation that leads to aneuploidy.
@@
 \paragraph{Bootstrapping}
-We  use bootstrap resampling method in order to compute  confidence intervals for the mean threshold tumor sizes and mean times by resampling 100 times with replacement from the data set obtain from simulations.
+We  use bootstrap resampling method in order to compute  confidence intervals for the mean threshold tumor sizes and mean times by resampling 100 times with replacement from the data set obtained from simulations.
 We then compute the interval which contain $95\%$ of the means of each resampled set.
@@
 \paragraph{Directions of future research}
-Even though our model has been designed for cancer it can be extended to understand to understand evolutionary in different biological contexts, for example, how yeast subject to stress can overcome extinction with the help of aneuploidy~\citep{pompei2023fitness}.
+Even though our model has been designed for cancer it can be extended to understand evolutionary rescue in different biological contexts, for example, how yeast subject to stress can overcome extinction with the help of aneuploidy~\citep{pompei2023fitness}.
 Additionally, the heterogeneity of aneuploidy is a fact the we did not account for, as not all the aneuploidy lineages generated have the same growth rate $\Delta_a$.
@@
 Blue dashed line represents the probability of evolutionary rescue as a function of $N$ without aneuploidy ($u=0$).
-The black line represents the case with tolerant aneuploidy a fraction $f=0\%$ of the initial tumor being aneuploid, the red line represents the case with a fraction $f=5\%$ and the green line represents the case with $f=50\%$.
-The dots represent simulations and the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the probability of that the tumor has adapted to the stress and $n=100$ is the number of simulations.
+The black line represents the case where a fraction $f=0\%$ of the initial tumor is aneuploid, the red line represents the case with $f=5\%$ and the green line represents the case with $f=50\%$.
+The dots represent simulations and the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the fraction of simulations in which the tumor has adapted to the stress and $n=100$ is the number of simulations.
 Parameters: $\lambda_w=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}
@@
 Solid lines represents the approximations \cref{eq:AsymptoticTimeRulesRecurrence} ($N<N_a^*$ blue line and $N>N_m^*$ green line).
-The simulations converge to the solution of \cref{eq:t2det} (green line) for large values of $N\gg N_m^*$.
+The simulations converge to \cref{eq:t2det} (green line) for large values of $N\gg N_m^*$.
 Parameters: $\lambda_w=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}
@@
 \includegraphics[width=1\textwidth]{Figures/RecurrencePlot.pdf}
-\caption{The mean time for the mutant cell population to reach size $N$, where $N$ is the initial number of cancer cells.
-Our inhomogeneous Poisson-process approximation (solid black line, \cref{meanproliferationtime}) is in agreement with simulation results (red markers with 95\% confidence intervals obtained with bootstrapping, see Appendix G) for small and intermediate values of $N$.
+\caption{The mean time for the mutant cell population to reach size $M$, where $M$ is the tumor detection size.
+Our inhomogeneous Poisson-process approximation (solid black line, \cref{meanproliferationtime2}) is in agreement with simulation results (red markers with 95\% confidence intervals obtained with bootstrapping, see Appendix G) for small and intermediate values of $N$.
 Dashed vertical blue line represents the threshold tumor size above which evolutionary rescue is very likely through aneuploidy \cref{eq:N_a} and the dashed vertical green line represents the threshold tumor size above which evolutionary rescue is very likely through direct mutation \cref{eq:N_m}.
-Solid lines represents the approximations \cref{eq:AsymptoticTimeRulesRecurrence} ($N<N_a^*$ blue line and $N>N_m^*$ green line).
-The simulations converge to the solution of \cref{eq:t2det} (green line) for large values of $N\gg N_m^*$.
+Solid blue line represents the approximation \cref{meanproliferationtime2} with $\tau_a$ from \cref{eq:AsymptoticTimeRules} for $N<N_a^*$ and the solid green line  represents the approximation \cref{eq:t2det2} for $N>N_m^*$.
+The simulations converge to \cref{eq:t2det2} (green line) for large values of $N\gg N_m^*$.
 Parameters: $\lambda_w=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}
@@
 \caption{Shown is the mean time for appearance of a resistance mutation the leads to evolutionary rescue with aneuploidy ($u>0$).
-Our inhomogeneous Poisson-process approximations (solid black lines, right: \cref{meantimet2}).
-Dashed black line is the  inhomogeneous Poisson-process approximation \cref{meantimet2SGV} where a fraction $f$ of tumor is aneuploid at the onset of drug therapy which is in agreement with simulation results (red markers with 95\% confidence intervals obtained with bootstrapping, see Appendix G).
+Black lines represent our inhomogeneous Poisson-process approximations (solid black line, \cref{meantimet2}; dashed black line \cref{meantimet2SGV}).
+Dashed black line is the  inhomogeneous Poisson-process approximation where a fraction $f$ of tumor is aneuploid at the onset of drug therapy which is in agreement with simulation results (red markers with 95\% confidence intervals obtained with bootstrapping, see Appendix G).
 Dashed vertical blue line represents the threshold tumor size above which evolutionary rescue is very likely through aneuploidy \cref{eq:N_a} and the dashed vertical green line represents the threshold tumor size above which evolutionary rescue is very likely through direct mutation \cref{eq:N_m}.

### 2024-02-04 EvolutionaryRescue1.tex
edit methods

@@
 \section*{Methods}
-\subsection*{Evolutionary model}
-
+
+\paragraph{Evolutionary model.}
 We follow the number of cancer cells that have one of three different genotypes at time $t$: wildtype, $w_t$; aneuploid, $a_t$; and mutant, $m_t$. 
@@
 We assume the population of cells is under a strong stress, such as drug therapy, to which the wildtype genotype is susceptible or sensitive and therefore $\Delta_w<0$, whereas the mutant is resistant to the stress, $\Delta_m>0$.
-We analyze three scenarios: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_w$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing, stationary or "barely growing", that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$.
+We analyze three scenarios: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_w$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing, stationary or ``barely growing'', that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$.
 We assume that both chromosomal missegregation and mutation occur during the process of mitosis. 
@@
 Both aneuploid and wildtype cells may divide and mutate to become mutants at rates $v\lambda_{a}$ and $v\lambda_{w}$, respectively.
-See \Cref{figureAneuploidy} for a schematic representation of the model and \Cref{sampleTrajectories} for sample trajectories of the different genotypes for $u=0 \, (A), \Delta_a\ll0\, (B),\Delta_a\approx0\, (C),\Delta_a\gg0\, (D)$.
-We assume that wildtype cells acquire aneuploidy before the onset of therapy with rate $\tilde{u}\lambda_w$ and that aneuploid cells have a fitness cost $s$ in the drug-free environment.
+To model standing genetic variation, we assume that before the onset of therapy, wildtype cells become aneuploid with rate $\tilde{u}\lambda_w$ (which may differ from $u \lambda_w$) and that aneuploidy confers a fitness cost $s$ in the drug-free environment.
+% TODO unclear how s affects \lambda_w
+See \Cref{figureAneuploidy} for a schematic representation of the model and \Cref{sampleTrajectories} for sample trajectories of the different genotypes.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Stochastic simulations} 
-Simulations are performed using a \emph{Gillespie algorithm} \citep{gillespie1976general,gillespie1977exact} implemented in Python \citep{python}.
+
+\paragraph{Stochastic simulations.} 
+Simulations are performed using the \emph{Gillespie stochastic simulation algorithm} \citep{gillespie1976general,gillespie1977exact} implemented in Python \citep{python}.
 The simulation monitors the number of cells of each type: wildtype, aneuploid, and mutant. 
@@
 For the remaining of this paper we assume that the division rates for wildtype and aneuploid cells can be written as $\lambda_ww_t\left(1-u-v\right)\approx \lambda_ww_t$ and $\lambda_aa_t\left(1-v\right)\approx\lambda_aa_t$ because $u,v\ll1$ (see Table \ref{table1}).
-Each iteration of the simulation loop starts with computing the rates $\nu_j$ of each event $j$.
+Each iteration of the simulation loop starts with computing the rates $\nu_k$ of each event $k$.
 We then draw the time until the next event, $\Delta t$, from an exponential distribution whose rate parameter is the sum of the rates of all events, such that $\Delta t \sim \textit{Exp}(\sum_j \nu_j)$.
-Then, we randomly determine which event occurred, where the probability for event $j$ is $p_j=\nu_j/\sum_i \nu_i$.
+Then, we randomly determine which event occurred, where the probability for event $k$ is $p_k = \nu_k/\sum_j \nu_j$.
 Finally, we update the number of cells of each type according to the event that occurred and update the time from $t$ to $t+\Delta t$.
@@
 \begin{equation*}
-m_t > \left\lfloor\frac{3\log10}{\log\left(\lambda_m / \mu_m\right)}\right\rfloor + 1.
+m_t > \left\lfloor\frac{3\log{10}}{\log{\left(\lambda_m / \mu_m\right)}}\right\rfloor + 1 ,
 \end{equation*}
-where we obtain the threshold mutant size by solving $1-(1-p_m)^{m_t}=0.999$ for $m_t$, where $p_m=\Delta_m/\lambda_m$ is the probability that a single mutant escapes stochastic extinction  (see Apendix A).
+which we obtain by solving $1-(1-p_m)^{m_t}=0.999$ for $m_t$ with $p_m=\Delta_m/\lambda_m$ as the probability that a single mutant escapes stochastic extinction  (see Appendix A).
+
+When simulations are slow (e.g., due to large population size), we use $\tau$-leaping \citep{gillespie2001approximate}, where we assume that the change in the number of cells of genotype $k$ in a fixed time interval $\Delta t$ is Poisson distributed with mean $\nu_k \Delta t$.
+If the change in the number of cells is negative and larger than the subpopulation size then the subpopulation size is updated to be zero.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\paragraph{$\tau$-leaping.}
-When simulations are slow (e.g.
-due to large population size), we utilize $\tau$-leaping \citep{gillespie2001approximate}, where change in number of cells of genotype $i$ in a fixed time interval $\Delta t$ is Poisson distributed with mean $\nu_i\Delta t$.
-If the change in number of cells is negative and larger then the subpopulation size then the subpopulation size is updated to be zero.
+% YR:  I commented this out because it is entirely standard
+%\paragraph{Bootstrapping}
+%We  use bootstrap resampling method in order to compute  confidence intervals for the mean threshold tumor sizes and mean times by resampling 100 times with replacement from the data set obtained from simulations.
+We then compute the interval which contain $95\%$ of the means of each resampled set.
 
@@
 
-\paragraph{Bootstrapping}
-We  use bootstrap resampling method in order to compute  confidence intervals for the mean threshold tumor sizes and mean times by resampling 100 times with replacement from the data set obtained from simulations.
-We then compute the interval which contain $95\%$ of the means of each resampled set.
-
+\paragraph{Parameterization.}
+To parametrize the simulations, we assume that the cells under consideration are melanoma cells and rely on \citet{rew2000cell} and \citet{bozic2013evolutionary} for the division and death rates, respectively.
+\citet{rew2000cell} report \emph{in vivo} measurements of the potential doubling times (the waiting time for the number of cells in the tumor to double disregarding cell death) for a large set of cancer types.
+The division rate is obtained as $\lambda=\log{2} / T \approx 0.1$ per day.
+We select this to be the division rate for wildtype and mutant cells.
+%We select the mutant death rate $\mu_m=0.09\text{ days}^{-1}$ such that $\Delta_m>0$.
+
+\citet{bozic2013evolutionary} report the growth rate $\Delta_w$ for wildtype melanoma cancer cells from which they deduce the death rate $0.11 \le \mu_w \le 0.17$.
+We use  $\mu_w=0.14$ per day.
+Additionally, they observe the growth rate of cancer cells prior to treatment to be 0.01, which we use as the growth rate of mutant cells, which are resistant to the drug.
+Thus, we use $\mu_m=0.1-0.01=0.09$ per day as the death rate for mutant cells.
+
+ Aneuploid death rate $\mu_a$ is set to the same value as the mutant death rates, $\mu_m=0.09$ per day, given that aneuploidy increases resistance to the drug, such as cisplatin, by antagonizing cell division \citep{replogle2020aneuploidy}.
+Aneuploid birth rate is selected such that the aneuploid growth rate $\Delta_w\ll\Delta_a\ll\Delta_m$ which means that $0.06 \le \lambda_a \le 0.1$.
+For most of our simulations we use $\lambda_a=0.0899$ per day, such that aneuploidy can only act as a \emph{stepping stone} for the generation of the mutant that rescues the cancer cell population.
+
+We assume the mutation rate is $10^{-7}$ per gene per cell division \citep{loeb2001mutator} and since we assume that a single target gene confers resistance to the drug, we use $v=10^{-7}$ per cell division.
+\citet{bakker2023predicting} determined that the missegregation rate must be between $10^{-3}-10^{-2}$ per chromosome per cell division.
+Given that a duplication is is beneficial for cancer resistance only in one or two chromosomes (see details next paragraph) we use $u=10^{-2}$ per cell division.
+For the missegregation rate in the drug-free environment, $\tilde{u}$, we use the lower-end value of $\tilde{u}=10^{-3}$ per cell division, as some drugs increase the rate of aneuploidy~\citep{wang2019molecular,mason2017functional}.
+
+The fitness cost $s$ of aneuploidy before the onset of therapy is difficult to estimate as we are interested in a specific type of aneuploidy which improves the fitness of cancer cells in an environment altered by drugs.
+We derive $s$ by using the formula $s=\tilde{u}\lambda_w / f$, where $f$ is the fraction of aneuploid cancer cells.
+To estimate $f$, we note that \citet{lukow2021chromosomal} mixed together wildtype and aneuploid A375 melanoma cells at $50:50$ ratio, cultured them in drug-free environment and observed the ratio evolve as a function of time with the aneuploid cells declining to $15\%$ after 24 days.
+\citet{ippolito2021gene} observed that trisomy in Chr II and VI are most likely to confer increased resistance against the chemotherapeutic agent vemurafenib for A375 cells.
+We assume that if a tumor is aneuploid then it most likely has trisomy \citep{gisselsson2010generation} in the pre-treatment environment and that cells with more than one trisomy are very unlikely to survive.
+Additionally, we assume that all trisomies are equally likely and, as a result, we obtain the fraction of cancer cells with favorable aneuploidy as $f=2/22\times0.15\approx 0.0136$ (i.e.
+$1.36\%$ of pre-treatment cancer cells have the beneficial aneuploidy).
+This gives us a fitness cost of $s= 7.3\times10^{-3}$ per day.
+We note that when we refer to wildtype cancer cells we include those cells that have any aneuploidy except trisomy in Chr II and VI as those are the aneuploid cells which are hypothesized to have higher fitness in the environment altered by drugs such as vemurafenib.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\paragraph{Parameterization.}
-To parametrize the model we assume that the cells under consideration are melanoma cells and and rely on \citep{rew2000cell} and \citep{bozic2013evolutionary} for the division and death rates, respectively.
-\citet{rew2000cell} report in vivo measurements of the potential doubling times (which is the time which the tumor cells would take to double in number if cell loss was not present) for a large set of cancer types.
-The division rate is obtained as $\lambda=\log2/T\approx0.1\text{ days}^{-1}$ We select this to be the division rate for wildtype and mutant cells.
-%We select the mutant death rate $\mu_m=0.09\text{ days}^{-1}$ such that $\Delta_m>0$.
-
-\citet{bozic2013evolutionary} report the growth rate $\Delta_w$ for the wildtype melanoma cancer cells from which they deduce the death rate $\mu_w$ to in the interval $\left[0.11,0.17\right]$ and we choose $\mu_w=0.14days^{-1}$.
-Additionally, they observe the growth rate of cancer cells prior to treatment to be 0.01 which we select as the growth rate of mutant cells as we assume that the mutation makes the cancer cells immune to the chemotherapeutic agent.
-As a result, we choose $\mu_m=0.1-0.01=0.09\text{ days}^{-1}$ as the death rate for mutant cells.
-
- Aneuploid death rate $\mu_a$ is set to the same value as the mutant death rates, $\mu_m=0.09\text{ days}^{-1}$, given that aneuploidy increases resistance to chemotherapeutics, such as cisplatin, by antagonizing cell division \citep{replogle2020aneuploidy}.
-Aneuploid birth rate is selected such that the aneuploid growth rate $\Delta_w\ll\Delta_a\ll\Delta_m$ which means that $\lambda_a$ belong to the interval $\left[0.06,0.1\right]$.
-For most of the paper we will use $\lambda_a=0.0899\text{ days}^{-1}$ such that aneuploidy can only act as a \emph{stepping stone} for the generation of the mutant which rescues the cancer cell population.
-
-We set the mutation rate $v$ to be equal to $10^{-7}$ per gene per cell division \citep{loeb2001mutator} and since we assume that a single target gene confers immunity to chemotherapy we select $v=10^{-7}$ per cell division.
-The missegregation rate is chosen from \citet{bakker2023predicting} who determined \textit{in silico} that it must be in the interval $10^{-3}-10^{-2}$ per chromosome per cell division.
-Given that only about one to two chromosome additions are ``beneficial'' for cancer resistance (see details next paragraph) we choose $u=10^{-2}$ per cell division.
-For the missegregation rate in the drug free environment $\tilde{u}$ we choose the lower-end value of $u$ (i.e.
-$\tilde{u}=10^{-3}$ per cell division) as the chemotherapeutic drugs we want to model increase the rate of aneuploidy~\citep{wang2019molecular,mason2017functional}.
-
-The fitness cost $s$ of aneuploidy before the onset of therapy is difficult to estimate as we are interested in a specific type of aneuploidy which improved the fitness of cancer cells in an environment altered by chemotherapeutic drugs.
-We derive $s$ by using the formula $s=\tilde{u}\lambda_w/f$ and obtaining the fraction of aneuploid cancer cells $f$ from \citep{lukow2021chromosomal} who mixed together wildtype and aneuploid A375 melanoma cells at $50:50$ ratio, cultured them in drug-free environment and observed the ratio evolve as a function of time with the aneuploid cells declining to $15\%$ after 24 days.
-\citet{ippolito2021gene} observed that trisomy 2 and 6 are most likely to confer increased resistance against the chemotherapeutic agent vemurafenib for A375 cells.
-We assume that if a tumor is aneuploid then it most likely has trisomy \citep{gisselsson2010generation} in the pre-treatment environment and that cells with more then one trisomy are very unlikely to survive.
-Additionally, we assume that all trisomies are equally likely and, as a result, we obtain the percentage of cancer cells with favorable aneuploidy as $2/22\times0.15\approx 0.0136$ (i.e.
-$1.36\%$ of pre-treatment cancer cells have the ``beneficial'' aneuploidy) which we select as the value for $f$ (this gives us a fitness cost of $s= 7.3\times10^{-3}\text{ days}^{-1}$).
-We note that when we refer to wildtype cancer cells we include those cells which have any aneuploidy except trisomy 2 and 6 as those are the aneuploid cells which are hypothesized to have higher fitness in the environment altered by therapeutics such as vemurafenib.
+\paragraph{Density-dependent growth.}
+
+In our analysis we assume that lineages produced by cells from the initial population divide and die independently of each other, which may be unrealistic, as cells usually compete for resources.
+This is a good approximation as we assume that the drugs cause the wildtype population to decline rapidly and, as a result, density-dependent effects can be neglected.
+A more realistic model includes competition for limited resources and spatial structure, which may play an important role in the development of cancer \citep[e.g.,][]{martens2011spatial}.
+To simulate birth and death rates that depend on the number of cells in the population, we assume the rates of division and death to be
+\begin{align*}
+\lambda_w' &= \lambda_w , \\
+\mu_w' &= \mu_w ,\\
+\lambda_a' &= \lambda_a ,\\ 
+\mu_a' &= \mu_a + \lambda_a\frac{w+a+m}{K} ,\\
+\lambda_m' &= \lambda_m ,\\ 
+\mu_m' &= \mu_m + \lambda_m\frac{w+a+m}{K} ,
+\end{align*}
+where $K$ is the tumor carrying capacity. 
+The effective carrying capacity of this model is $K_e=K\Delta_a/\lambda_a\approx10^6$ for $K=10^8, \lambda_a=0.0901,\mu_a=0.09$, where we define the effective carrying capacity to be the population size at which the aneuploid division rate is equal to the aneuploid death rate.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\paragraph{Density-dependent growth.}
-
-In our analysis we assume that lineages produced by cells from the initial population divide and die independently of each other, which may be unrealistic, as cells usually compete for resources.
-This is a good approximation as we assume that the drug cause the wildtype population to decline rapidly and, as a result, density effect can be neglected.
-A more realistic model includes competition for limited resources and spatial structure, which may play an important role in the development of cancer \citep[e.g.,][]{martens2011spatial}.
-To simulate birth and death rates that depend on the number of cells in the population, we transform the rates of division and death to the following:
-\begin{align*}
-\lambda_w' &= \lambda_w, \\
-\mu_w' &= \mu_w,\\
-\lambda_a' &= \lambda_a,\\ 
-\mu_a' &= \mu_a + \lambda_a\frac{w+a+m}{K},\\
-\lambda_m' &= \lambda_m,\\ 
-\mu_m' &= \mu_m + \lambda_m\frac{w+a+m}{K}.
-\end{align*}
-where $K$ is the maximum carrying capacity.
-The effective carrying capacity of this model is $K_e=K\Delta_a/\lambda_a\approx10^6$ for $K=10^8, \lambda_a=0.0901,\mu_a=0.09$, where we define the effective carrying capacity to be the population size at which the aneuploid division rate is equal to the aneuploid death rate.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection*{Code and data availability.} All source code is available online at \url{https://github.com/yoavram-lab/EvolutionaryRescue}.
+\paragraph{Code and data availability.} All source code is available online at~\url{https://github.com/yoavram-lab/EvolutionaryRescue}.
 

### 2024-02-04 EvolutionaryRescue1.tex
reformat appendices

@@
 
-In our model, \emph{evolutionary rescue} occurs when resistant cells appear and fixate ($m_t \gg 1$) in the population before the population becomes extinct ($w_t=a_t=m_t=0$).
-Aneuploidy may contribute to evolutionary rescue by either preventing (when $\Delta_a>0$) or delaying (when $0>\Delta_a>\Delta_w$) the extinction of the population before mutant cells appear and fixate.
+In our model, \emph{evolutionary rescue} occurs when resistant cells appear and establish (avoid random extinction) in the population  ($m_t \gg 1$) before the population becomes extinct ($w_t=a_t=m_t=0$).
+Aneuploidy may contribute to evolutionary rescue by either preventing (when $\Delta_a>0$) or delaying (when $0>\Delta_a>\Delta_w$) the extinction of the population before mutant cells appear and establish.
 We assume independence between clonal lineages starting from an initial population of $N$ wildtype cells (we check the effect of density-dependent growth on our results below).
@@
 1-\e^{-Np_w} = 
-1-e^{-N/N^*} .
+1-e^{-N/N^*} ,
 \end{equation}
 where the approximation $(1-p_w)\approx e^{-p_w}$ assumes that $p_w$ (but not necessarily $N p_w$) is small.
-Indeed, when $N<1/p_w$, then the probability for evolutionary rescue is $\presc \approx N p_w$  and when $N > 1/p_w$, it is $\presc \approx 1$, justifying the definition of $N^*$ as the threshold tumor size.
-\\
+Indeed, when $N<1/p_w$, then the probability for evolutionary rescue is $\presc \approx N p_w$  and when $N > 1/p_w$, it is $\presc \approx 1$, justifying the definition of $N^*$ as the threshold tumor size for evolutionary rescue.
+
 In Appendix A, we use the theory of multi-type branching processes to find approximate expressions \cref{eq:pw_parttolerant,eq:pw_partrest,eq:pw_tolerant} for $p_w$ in different regimes. 
@@
 the NSF grant ($\#2146260$, DBW),
-{\color{red} Uri Ben-David acknowledgements}
-
-% TODO add funding for Uri and Daniel ok
+% TODO add funding for Uri
 }
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\section*{References}
-
-\nolinenumbers
-%\bibliographystyle{unsrtnat}
-\bibliographystyle{agsm}
-\bibliography{evo2022}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\section*{References}
+
+\nolinenumbers
+%\bibliographystyle{unsrtnat}
+\bibliographystyle{agsm}
+\bibliography{evo2022}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
 \newpage
-\clearpage
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Appendix A: Survival probability of a single lineage}\label{sec:appendix-surv-prob}
+\section{Survival probability of a single lineage}\label{sec:appendix-surv-prob}
 
@@
 
-\section*{Appendix B: Evolutionary rescue probability}\label{sec:appendix-rescue-prob}
+\section{Evolutionary rescue probability}\label{sec:appendix-rescue-prob}
 Using the fact that $\Delta_a-v\lambda_ap_m\approx\Delta_a$ we write the condition $\left(\Delta_a-v\lambda_ap_m\right)^2 \ll 4\lambda_a^2vp_m$ as:
@@
 
-\section*{Appendix C: Evolutionary rescue time}
+\section{Evolutionary rescue time}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Appendix D: Recurrence time}
+
+\section{Recurrence time}
 We define the proliferation time $\tau_a^r$  to be the time it takes the population of mutant cancer cells to reach the initial tumor size $N$.
@@
 
-\section*{Appendix E: Distribution of evolutionary rescue time}
+\section{Distribution of evolutionary rescue time}
 The probability that a successful mutant has been generated by time $t$ is given by:
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\clearpage
-
-\section*{Appendix F: Figures}
+
+\newpage
+\section*{Supplementary Figures}
+\beginsupplement % https://support.authorea.com/en-us/article/how-to-create-an-appendix-section-or-supplementary-information-1g25i5a/
 
@@
 \end{appendices}
+
 %%%

### 2024-02-04 EvolutionaryRescue1.tex
start editing results

@@
 
-In Appendix A, we use the theory of multi-type branching processes to find approximate expressions \cref{eq:pw_parttolerant,eq:pw_partrest,eq:pw_tolerant} for $p_w$ in different regimes. 
+We use the theory of multi-type branching processes to find approximate expressions \cref{eq:pw_parttolerant,eq:pw_partrest,eq:pw_tolerant} for $p_w$ in different regimes (see appendix~\ref{sec:appendix-surv-prob}).
 Substituting these into $N^*=1/p_w$, we find approximations for the threshold tumor size, $N^*$. 
 For these approximations, an important quantity is $T^* = (4 v \lambda_a^2 \Delta_m/\lambda_m)^{-1/2}$, which is the critical time an aneuploid lineage needs to survive to produce a resistant mutant that avoids random extinction.
-First, if aneuploidy is very rare ($u\lambda_a T^*< 1$), or if aneuploidy is rare ($u\lambda_a < -\Delta_a$) and very sensitive to the drug ($\Delta_a T^* < -1$), then rescue will likely occur by a direct resistance mutation in a sensitive cell, such that
+First, if aneuploidy is very rare ($u\lambda_a T^*< 1$), or if aneuploidy is rare ($u\lambda_a < -\Delta_a$) and very sensitive to the drug ($\Delta_a T^* < -1$), then evolutionary rescue will likely occur by a direct resistance mutation in a sensitive wildtype cell without the involvement of aneuploidy, such that
 \begin{equation} \label{eq:N_m}
@@
 \end{equation}
-Here, $\abs{\Delta_w}/\left(v\lambda_w\right)$ is the ratio of the rate at which wild-type cells are decreasing in number and the rate at which they are mutating.
-
-Otherwise, aneuploidy is frequent enough ($u\lambda_a > \max{(-\Delta_a, 1/T^*)}$) to affect the evolution of drug resistance. 
-The threshold tumor size, $N^*$, can then be approximated by one of the following cases, depending on $\Delta_a T^*$, the change in the log of the aneuploid population size during the critical time,
+Here, $\abs{\Delta_w}/\left(v\lambda_w\right)$ is the ratio of the rate at which wildtype cells are decreasing in number and the rate at which they are mutating.
+Notably, the aneuploidy parameters ($u$, $\lambda_a$, $\mu_a$) do not affect $N_m^*$.
+
+Otherwise, aneuploidy is frequent enough ($u\lambda_a > \max{\left(-\Delta_a, 1/T^*\right)}$) to affect the evolution of drug resistance. 
+The threshold tumor size can then be approximated by one of the following cases, depending on $\Delta_a T^*$, the change in the aneuploid log-population size during the critical time,
 \begin{equation}  \label{eq:N_a}
@@
 \end{equation}
-The first line describes the case in which aneuploids are still effectively killed by the treatment, but not as quickly as the wild type. 
-In the second case, the aneuploids are sufficiently resistant that the size of each aneuploid lineage is expected to remain roughly constant. 
-In both of these first two cases, aneuploidy increases the probability of rescue by slowing or halting the decrease of the cancer population, allowing more opportunities for producing resistant mutants.
-In the third case, aneuploidy provides sufficient resistance for the aneuploid population to re-grow the tumor even without additional resistance mutations.
-Note that in this case there is no dependence on the parameters characterizing mutants or their production ($v$, $\lambda_m$, and $\Delta_m$).
-Comparing these approximations to results of stochastic evolutionary simulations, we find that the approximations perform very well (\Cref{rescue_prob,rescue_threshold}).
-Increasing the the missegregation rate $u$ and aneuploid growth rate $\Delta_a$ reduces the threshold tumor size and facilitates the evolution of drug resistance (\Cref{rescue_threshold}).
+These approximations perform very well when compared to results of stochastic evolutionary simulations (\Cref{rescue_prob,rescue_threshold}).
+The first line describes the case in which aneuploid cells are still effectively killed by the treatment, but not as quickly as the wild type. 
+In the second case, aneuploid cells are sufficiently resistant that the size of each aneuploid lineage is expected to remain roughly constant.
+% TODO that's a weird sentence; a lineage starts with a single cell, if the lineage size stays constant that it stays one.
+so what are we talking about?
+try to rephrase.
+In both of these cases, aneuploidy increases the probability of rescue by slowing or halting the decrease of the cancer population, allowing more opportunities for producing resistant mutants. 
+In the third case, aneuploid cells are sufficiently resistance for the cancer population to re-grow the tumor even without additional resistance mutations.
+Notably, in this case there the mutant parameters ($v$, $\lambda_m$, and $\Delta_m$) do not affect $N_a^*$ beyond their effect on $T^*$.
+In all cases, $N_a^*$ is proportional to $1/u$ such that increasing the missegregation rate $u$ will decrease the threshold tumor size (\Cref{rescue_threshold}B).
+Furthermore, increasing the aneuploid growth rate $\Delta_a$ (which appears both in the terms and in the conditions), also reduces the threshold tumor size, with a sharp decrease around $\Delta_a=0$, but the effect is minor when $|\Delta_a|$ is small compared to $T^*$ (\Cref{rescue_threshold}A).
+% TODO not sure what the last sentence "but ..." means and how do you see it
+% TODO these are pretty straightforward - can you glean something more interesting from eq 4?
 
@@
 \end{equation}
-In all cases, the effect of aneuploidy increases (i.e., the threshold size ratio decreases) when the aneuploidy rate $u$ increases (\Cref{rescue_threshold}B).
-Increasing the aneuploid growth rate $\Delta_a$ also leads to an increased role of aneuploidy, although the effect is minor when $|\Delta_a|$ is small compared to $T^*$ (\Cref{rescue_threshold}A).
-Increasing the mutation rate $v$ decreases the effect of aneuploidy and increases the probability of evolutionary rescue.
-
+As expected, this ratio increases with the mutation rate $v$ and decreases with the aneuploidy rate $u$.
 In the first case, $\abs{\Delta_a}/\left(u\lambda_a\right)$ is  the ratio of the expected time for an aneuploid lineage to appear, $1/\left(u\lambda_a\right)$, and the expected time until that lineage disappears, $1/\abs{\Delta_a}$.
@@
 In this case, increasing the division rate $\lambda_a$ should also increase the mutation rate $v\lambda_a$ in aneuploid cells, as mutations mostly occur during division, so overall the threshold tumor size $N_a^*$ is unaffected by the division rate $\lambda_a$ (i.e., $d \lambda_a T^*/d\lambda_a = 0$).
-Thus, if aneuploids rapidly die due to the drug but compensate by rapidly dividing, further increasing  the division rate will \emph{not} facilitate adaptation.
+Thus, if aneuploid cells rapidly die due to the drug but compensate by rapidly dividing, further increasing  the division rate will \emph{not} facilitate adaptation.
+% NOTE this is the kind of conclusions we want to drive home
 This is consistent with experimental findings where aneuploidy confers resistance by decreasing the division rate~\citep{replogle2020aneuploidy}.
-For the parameter values of  $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$ we are in the partially resistant aneuploid case and obtain the ratio $N^*_a/N^*_m\approx\abs{\Delta_a}/u\lambda_a=0.11$ which means that the presence of aneuploidy reduces the threshold population by approximately $90\%$.
-We note that for this parameter regime we have $N_a^*\approx4.4\times10^6$ and $N_m^*\approx4\times10^7$ and this allows us to define the following categories: small tumors have size $N<N_a^*$, intermediate sized tumors have $N_a^*<N<N_m^*$ and large tumors have size $N>N_m^*$ .
-For the remaining of this paper we use these definitions when referring to small, intermediate or large tumors.
-Interestingly, the threshold between small and intermediate tumors is similar to the detection threshold of $4.19\times10^6$ for a wide variety of tumors as reported by \citet{avanzini2019cancer}.
+
+We can categorize tumors by their size: small tumors with size $N<N_a^*$, intermediate tumors with size $N_a^*<N<N_m^*$, and large tumors with size $N>N_m^*$ .
+For the parameter values in \Cref{table1} with $\lambda_a=0.0899,\mu_w=0.14, u=10^{-2}, v=10^{-7}$, we are in the partially resistant aneuploid case, and substituting in \cref{eq:N_a,eq:N_m}, we have $N_a^* \approx 4.4 \cdot 10^6$ and $N_m^* \approx 4 \cdot 10^7$.
+Hence, we obtain the ratio $N^*_a/N^*_m \approx 0.11$ (\cref{eq:N_ratio}), that is, aneuploidy reduces the threshold population size by approximately 90\% 
+Interestingly, the threshold between small and intermediate tumors, $N_a^*$, is similar to the tumor detection threshold of $4.19 \cdot 10^6$ cells for a wide variety of tumors as reported by \citet{avanzini2019cancer}.
 
@@
 However, growth may be limited by resources (oxygen, nutrients, etc.) and therefore depend on cell density.
-We therefore performed stochastic simulations of a logistic growth model with carrying capacity $K$ (see Methods). 
-We find that our approximations agree with results of simulations with density-dependent growth for biologically relevant parameter values (\Cref{LogisticPlot}).
+We therefore performed stochastic simulations of a logistic growth model with a carrying capacity (see Methods). 
+We find that our density-independent approximations agree with results of simulations with density-dependent growth for biologically relevant parameter values (\Cref{LogisticPlot}).
 
@@
 In the above we assumed that at the onset of drug treatment, the initial tumor consisted entirely of wildtype cells that are drug sensitive.
-However, aneuploid cells are likely generated even before onset of treatment at some rate $\tilde{u} \le u$ (because the treatment itself may promote generation of aneuploid cells \citep{wang2019molecular,mason2017functional}), which are likely to have a deleterious effect in the absence of the drug, $s$ \citep{replogle2020aneuploidy,giam2015aneuploidy}.
+However, aneuploidy is likely produced even before onset of treatment at some rate $\tilde{u} \le u$ and confer a deleterious fitness effect $s$ in the absence of the drug \citep{replogle2020aneuploidy,giam2015aneuploidy}, and the aneuploidy rate in the presence of drugs, $u$, is likely higher than in their absence, $\tilde{u}$ \citep{wang2019molecular,mason2017functional}.
 But if the number of cells in the tumor $N$ is large (as expected if the tumor is to be treated with a drug), there may already be a fraction $f \approx \tilde{u}\lambda_w/s$ (we assume that the division rate of the wildtype in a drug-free environment is equal to the division rate of the wildtype in the environment affected by the anti-cancer drug) of aneuploid cells in the population.
 
-Therefore, the threshold tumor size with standing generation variation, $\tilde{N}^*_{a}$, is similar to the ratio with de-novo variation, $N^*_a$, except that the sensitive growth rate $\abs{\Delta_w}$ is replaced with the aneuploidy cost, $s$, such that
+Therefore, the ratio of threshold tumor size with standing generation variation, $\tilde{N}^*_{a}$, is similar to the ratio with de-novo variation, $N^*_a$, except that the sensitive growth rate $\abs{\Delta_w}$ is replaced with the aneuploidy cost, $s$, such that
 \begin{equation}\label{eq:Nsgv}
@@
 \end{equation}
-
-Therefore, standing genetic variation will drive adaptation to the drug if $\Delta_w$ is very negative due to a stronger effect of the drug on sensitive cells, or if $s$ is very small due to a low cost of aneuploidy in the pre-drug conditions.
-In contrast, de-novo aneuploids will have a stronger effect on adaptation if the aneuploidy cost $s$ is large, the effect of the drug is weak ($\Delta_w$ is small), or if the drug induces the appearance of aneuploid cells ($u > \tilde u$).
-Comparing these approximations to results of stochastic evolutionary simulations, we find that the approximations perform very well (\Cref{rescue_denovo}).
-
-For the parameter values of  $\lambda_w=0.1,\mu_w=0.14, u=10^{-2}, \tilde{u}=10^{-3}, s=7.3\cdot10^{-3}$ the ratio of the threshold population sizes is $\tilde{N}^*_a/N^*_a\approx1.825$ which means that de-novo genetic variation is  the main driver of adaptation.
+Comparing these approximations to results of stochastic evolutionary simulations, we find that the approximations perform very well (\Cref{rescue_denovo}).  % TODO which approximations? unclear
+Standing genetic variation will drive evolutionary rescue to the drug if wildtype growth rate $\Delta_w$ is very negative due to a stronger effect of the drug on sensitive cells, or if the aneuploidy cost $s$ is very small due to a low cost of aneuploidy in the pre-drug conditions.
+In contrast, de-novo aneuploid cells will have a greater contribution to rescue if the aneuploidy cost, $s$, is large, the effect of the drug is weak ($\Delta_w$ is small), or if the drug induces the appearance of aneuploid cells ($u > \tilde u$).
+For example, with  $\lambda_w=0.1,\mu_w=0.14, u=10^{-2}, \tilde{u}=10^{-3}, s=7.3\cdot10^{-3}$, the ratio of the threshold population sizes for standing vs de-novo variation is $\tilde{N}^*_a/N^*_a \approx 1.825$, which means that de-novo genetic variation is  the main driver of evolutionary rescue.
 
@@
 \begin{equation} \label{eq:N_sgv_ratio}
-\frac{\tilde{N}^*_a}{N^*_m}=\frac{\tilde{N}^*_{a}}{N^*_{a}}\times\frac{N^*_a}{N^*_m} \approx \frac{s}{\abs{\Delta_w}}\begin{cases}
+\frac{\tilde{N}^*_a}{N^*_m}=\frac{\tilde{N}^*_{a}}{N^*_{a}} \cdot \frac{N^*_a}{N^*_m} \approx \frac{s}{\abs{\Delta_w}}\begin{cases}
     \frac{\abs{\Delta_a}}{\tilde{u}\lambda_a} ,&

### 2024-02-04 EvolutionaryRescue1.tex
continue editing results

@@
 For the parameter values in \Cref{table1} with $\lambda_a=0.0899,\mu_w=0.14, u=10^{-2}, v=10^{-7}$, we are in the partially resistant aneuploid case, and substituting in \cref{eq:N_a,eq:N_m}, we have $N_a^* \approx 4.4 \cdot 10^6$ and $N_m^* \approx 4 \cdot 10^7$.
-Hence, we obtain the ratio $N^*_a/N^*_m \approx 0.11$ (\cref{eq:N_ratio}), that is, aneuploidy reduces the threshold population size by approximately 90\% 
+Hence, we obtain the ratio $N^*_a/N^*_m \approx 0.11$ (\cref{eq:N_ratio}), that is, aneuploidy reduces the threshold tumor size by approximately 90\% 
 Interestingly, the threshold between small and intermediate tumors, $N_a^*$, is similar to the tumor detection threshold of $4.19 \cdot 10^6$ cells for a wide variety of tumors as reported by \citet{avanzini2019cancer}.
@@
 In contrast, de-novo aneuploid cells will have a greater contribution to rescue if the aneuploidy cost, $s$, is large, the effect of the drug is weak ($\Delta_w$ is small), or if the drug induces the appearance of aneuploid cells ($u > \tilde u$).
-For example, with  $\lambda_w=0.1,\mu_w=0.14, u=10^{-2}, \tilde{u}=10^{-3}, s=7.3\cdot10^{-3}$, the ratio of the threshold population sizes for standing vs de-novo variation is $\tilde{N}^*_a/N^*_a \approx 1.825$, which means that de-novo genetic variation is  the main driver of evolutionary rescue.
+For example, with  $\lambda_w=0.1,\mu_w=0.14, u=10^{-2}, \tilde{u}=10^{-3}, s=7.3\cdot10^{-3}$, the ratio of the threshold tumor sizes for standing vs de-novo variation is $\tilde{N}^*_a/N^*_a \approx 1.825$, which means that de-novo genetic variation is  the main driver of evolutionary rescue.
 
@@
 The ratio does not depend on the rate of chromosome missegregation induced by the drug $u$.
-However, if the growth rate of aneuploidy $\Delta_a$ increases then evolutionary rescue is driven by standing genetic variation.
-For the parameter values of  $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, \tilde{u}=10^{-3}, v=10^{-7}$ we are in the partially resistant aneuploid case and obtain the ratio $\tilde{N}^*_a/N^*_m\approx\frac{s}{\abs{\Delta_w}}\times\frac{\abs{\Delta_a}}{\tilde{u}\lambda_a}=0.356$ which means that standing genetic variation reduces the threshold population by approximately $65\%$.
-We observe that while standing genetic variation does not drive emergence of drug resistance when compared to de-novo aneuploids it does offer an advantage when compared with direct mutation.
+However, if the aneuploid growth rate $\Delta_a$ increases, then evolutionary rescue is driven by standing genetic variation.
+For the parameter values of  $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, \tilde{u}=10^{-3}, v=10^{-7}$, we are in the partially resistant aneuploid case and obtain the ratio $\tilde{N}^*_a/N^*_m \approx 0.356$, which means that standing genetic variation reduces the threshold tumor size by approximately $65\%$.
+We observe that while standing genetic variation does not drive evolution of drug resistance when compared to de-novo aneuploidy, it does offer an advantage when compared with direct mutation.
 
@@
 
-Even when evolutionary rescue occurs and leads to recurrence of the tumor, it may take a long time.
-When the expected number of resistant lineages that avoid extinction is small the overall expected recurrence time can be estimated by adding two terms: the mean waiting time for evolutionary rescue--the appearance of a resistant lineage that avoid extinction (which we denote the mean evolutionary rescue time) and the expected time for proliferation of that lineage back to the original tumor size, $N$ (which we denote the mean proliferation time).
-However, when the expected number of resistant lineages is large the population of mutant cells behaves deterministically and the mean recurrence time cannot be separated into the mean evolutionary rescue time and mean proliferation time.
-Of particular interest is the distribution of the evolutionary rescue time and recurrence time in the parameter regime  $\Delta_a T^*\ll1$ (partially tolerant aneuploids) for which we select  $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$ .
+When evolutionary rescue occurs and leads to recurrence of the tumor, it may still take a long time.
+When the expected number of resistant lineages that avoid extinction is small, the overall expected recurrence time can be estimated by adding two terms: the \emph{mean evolutionary rescue time}, which is the waiting time for appearance of a resistant lineage that avoids extinction (conditioned on such an even occurring in the first place), and the \emph{mean proliferation time}, which is the expected time for that lineage to reach the original tumor size, $N$.
+However, when the expected number of resistant lineages is large, the population of mutant cells behaves deterministically and the mean recurrence time cannot be separated into the mean evolutionary rescue time and mean proliferation time.
+% TODO I dont like the word "behaves", please be specific.
+- what exactly is deterministic and why does it mean you cannot do the separation.
+Of particular interest is the distribution of the evolutionary rescue time and recurrence time with partially tolerant aneuploid cells ($\Delta_a T^*\ll1$), for which we focus on the parameter values  $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$ .
 
 \paragraph{Evolutionary rescue time.}
-In Appendix C we derive an approximation for $\tau_m$, the expected evolutionary rescue time without aneuploidy ($u=0$), and $\tau_a$, the expected rescue time with aneuploidy ($u>0$).
-\Cref{MeanTimeGrowthAneuploidyPlot,EvolutionaryRescueTimeComplete} shows the agreement between these approximations and simulation results for small, intermediate and large tumor sizes.
-Additionally, we note that $\tau_a$ has the following asymptotic behaviour:
+In appendix \ref{sec:appendix_rescue_time} we derive an approximation for $\tau_m$, the expected evolutionary rescue time without aneuploidy ($u=0$), and $\tau_a$, the expected rescue time with aneuploidy ($u>0$), both conditioned on evolutionay rescue occuring.
+\Cref{EvolutionaryRescueTimeComplete,MeanTimeGrowthAneuploidyPlot} shows the agreement between these approximations and simulation results for small, intermediate and large tumor sizes.
+Notably, the expected rescue time with aneuploidy, $\tau_a$, follows these expressions for very small and very large tumors,
 \begin{equation}  \label{eq:AsymptoticTimeRules}
@@
 \end{equation}
-Interestingly, for small tumors the mean evolutionary rescue time is a function of the wildtype and aneuploid growth rates and independent of all of other model parameters including tumor size $N$ (\Cref{EvolutionaryRescueTimeComplete}).
-Increasing either the wildtype or aneuploid growth rates lead to an increase in the mean time.
-This can be interpreted that by increasing either growth rates the wildtype or aneuploid cells will last longer and will be able to produce a rescue mutation at latter times.
-Using the parameters that we have selected we observe that $\abs{\Delta_w}^{-1}=0.04$ and $\abs{\Delta_a}^{-1}=10^{-4}$ such that the mean evolutionary rescue time is determined by the aneuploid growth rate, i.e.
-$\tau_a\approx1/\abs{\Delta_a}$.
-For large tumors, we observe that the evolutionary rescue time is exponentially distributed with rate $v\lambda_wNp_m$ which is independent of parameters characterizing aneuploids or their production ($u$, $\lambda_a$, and $\Delta_a$).
-Increasing the mutation rate per mitosis $v\lambda_w$ leads to a faster appearance of a rescue mutation.
-Note that in the second case increasing the tumor size leads to smaller mean evolutionary rescue time as the wildtype is able to produce a rescue lineage faster.
-
-When a fraction $f\approx 1.36\%$ of the initial cancer cell population is aneuploid we calculate the mean evolutionary rescue time in \Cref{meantimet2SGV} which we plot in \Cref{SGVEvolutionaryRescueTimeComplete}.
+
+For small tumors, the mean rescue time is a function of the wildtype and aneuploid growth rates and independent the other model parameters, including tumor size (blue line in \Cref{EvolutionaryRescueTimeComplete}).
+Increasing the wildtype or aneuploid growth rates leads to an increase in the mean rescue time, because the corresponding cells will survive for longer and will produce additional rescue mutations at latter times.
+In our focus parameter regime, we have $\abs{\Delta_w}^{-1}=0.04$ and $\abs{\Delta_a}^{-1}=10^{-4}$, such that the mean rescue time is mainly determined by the aneuploid growth rate: $\tau_a \approx 10^4$.
+
+For large tumors, the mean evolutionary rescue time is $v\lambda_wNp_m$ (\cref{eq:AsymptoticTimeRules}), % I removed the mention of exponential distribution as it was not explained and I dont see why you need it here
+ which is independent of parameters characterizing aneuploid cells or their production ($u$, $\lambda_a$, and $\Delta_a$). 
+ Increasing the per division mutation rate, $v\lambda_w$, leads to faster appearance of a rescue mutation and hence reduced mean rescue time. 
+Finally, increasing the tumor size leads to shorter mean rescue time, as there are more wildtype cells that can mutate to become resistant.
+
+When a fraction $f\approx 1.36\%$ of the initial cancer cell population is aneuploid we calculate the mean evolutionary rescue time in \cref{meantimet2SGV} which we plot in \Cref{SGVEvolutionaryRescueTimeComplete}.
 We note that standing genetic variation reduces the mean evolutionary rescue time only for intermediary sized tumors but not for large or small tumors.
@@
 
-In Appendix D we approximate the mean time for the population of mutant cancer cells to reach the initial population size $N$ which we denote as the recurrence time.
+In appendix \ref{sec:appendix_recurrence_time} we approximate the mean time for the population of mutant cancer cells to reach the initial population size $N$ which we denote as the recurrence time.
 \Cref{proliferationFigure} shows the agreement between our approximations and simulations.
@@
 The recurrence can also be defined as the time necessary for the tumor size to reach a detection threshold $M$.
-We derive the mean recurrence time to detection size $M=10^7$ in Appendix D and plot in \Cref{RecurrencePlot} with simulations.
+We derive the mean recurrence time to detection size $M=10^7$ in appendix \ref{sec:appendix_recurrence_time} and plot in \Cref{RecurrencePlot} with simulations.
 We observe that  for small and intermediate sized tumors the effect of the detection size is negligible when compared to $\tau_a^r$.
@@
 
-We find that aneuploidy can have a significant effect on evolutionary rescue as it reduces the threshold population size by at least one order of magnitude even when aneuploidy is tolerant (\Cref{rescue_prob}).
+We find that aneuploidy can have a significant effect on evolutionary rescue as it reduces the threshold tumor size by at least one order of magnitude even when aneuploidy is tolerant (\Cref{rescue_prob}).
 Interestingly, aneuploidy is unlikely to contribute to evolutionary rescue in primary tumors in which the number of cells is large enough ($N\gg N_m^*\approx 4\times10^7$) for the appearance of resistant mutation directly in sensitive cells before these cells become extinct (\Cref{rescue_prob}).
@@
 
-\section{Evolutionary rescue time}
+\section{Evolutionary rescue time}\label{sec:appendix_rescue_time}
 
@@
 
-\section{Recurrence time}
+\section{Recurrence time}\label{sec:appendix_recurrence_time}
 We define the proliferation time $\tau_a^r$  to be the time it takes the population of mutant cancer cells to reach the initial tumor size $N$.
@@
 \end{align*}
-where $T_a$ is the time at which the first mutant cell appears that will avoid extinction and which was defined in Appendix C.
+where $T_a$ is the time at which the first mutant cell appears that will avoid extinction and which was defined in appendix \ref{sec:appendix_rescue_time}.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Appendix F: Distribution of recurrence time}
+\section{Distribution of recurrence time}
 The probability distribution of the time that a lineage, consisting initially of a single cell, will reach size $N$ as time $t$ is given by the Gumbel distribution $\text{Gumb}_{max}\left(\frac{\log Np_m}{\Delta_m},\frac{1}{\Delta_m}\right)$~\citep{avanzini2019cancer} with probability density function:

### 2024-02-08 EvolutionaryRescue1.tex
edits

@@
 
-When evolutionary rescue occurs and leads to recurrence of the tumor, it may still take a long time.
-When the expected number of resistant lineages that avoid extinction is small, the overall expected recurrence time can be estimated by adding two terms: the \emph{mean evolutionary rescue time}, which is the waiting time for appearance of a resistant lineage that avoids extinction (conditioned on such an even occurring in the first place), and the \emph{mean proliferation time}, which is the expected time for that lineage to reach the original tumor size, $N$.
+When evolutionary rescue occurs, the time until recurrence of the tumor may still be very long.
+We therefore explored the time until recurrence of the tumor, that is, the time until the tumor reaches its original size, $N$.
+When the expected number of resistant lineages that avoid extinction is small, the expected recurrence time can be estimated by adding two terms: the \emph{mean evolutionary rescue time}, which is the waiting time for appearance of a resistant lineage that avoids extinction (conditioned on such an even occurring in the first place), and the \emph{mean proliferation time}, which is the expected time for that lineage to grow to $N$ cells.
 However, when the expected number of resistant lineages is large, the population of mutant cells behaves deterministically and the mean recurrence time cannot be separated into the mean evolutionary rescue time and mean proliferation time.
@@
 \paragraph{Evolutionary rescue time.}
-In appendix \ref{sec:appendix_rescue_time} we derive an approximation for $\tau_m$, the expected evolutionary rescue time without aneuploidy ($u=0$), and $\tau_a$, the expected rescue time with aneuploidy ($u>0$), both conditioned on evolutionay rescue occuring.
-\Cref{EvolutionaryRescueTimeComplete,MeanTimeGrowthAneuploidyPlot} shows the agreement between these approximations and simulation results for small, intermediate and large tumor sizes.
-Notably, the expected rescue time with aneuploidy, $\tau_a$, follows these expressions for very small and very large tumors,
+We have derived approximations for $\tau_m$, the mean evolutionary rescue time without aneuploidy ($u=0$), and $\tau_a$, the mean rescue time with aneuploidy ($u>0$), both conditioned on evolutionary rescue occurring (see appendix \ref{sec:appendix_rescue_time}).
+These approximations are in good agreement with simulation results for small, intermediate, and large tumor sizes (\Cref{EvolutionaryRescueTimeComplete,MeanTimeGrowthAneuploidyPlot}).
+The mean rescue time with aneuploidy for small and large tumors follows these expressions:
 \begin{equation}  \label{eq:AsymptoticTimeRules}
@@
  N \ll N_a^* ,\\ 
-  \left(v\lambda_wNp_m\right)^{-1}  ,&
+  \frac{1}{v\lambda_w N} \cdot  \frac{\lambda_m}{\Delta_m} ,& % I replaced pm with 𝜆𝑚 − 𝜇𝑚 / 𝜆𝑚
   N \gg N_m^* .
@@
 
-For small tumors, the mean rescue time is a function of the wildtype and aneuploid growth rates and independent the other model parameters, including tumor size (blue line in \Cref{EvolutionaryRescueTimeComplete}).
+For small tumors ($ N \ll N_a^*$), the mean rescue time is a function of the wildtype and aneuploid growth rates and independent of the other model parameters, including tumor size (blue line in \Cref{EvolutionaryRescueTimeComplete}).
 Increasing the wildtype or aneuploid growth rates leads to an increase in the mean rescue time, because the corresponding cells will survive for longer and will produce additional rescue mutations at latter times.
-In our focus parameter regime, we have $\abs{\Delta_w}^{-1}=0.04$ and $\abs{\Delta_a}^{-1}=10^{-4}$, such that the mean rescue time is mainly determined by the aneuploid growth rate: $\tau_a \approx 10^4$.
-
-For large tumors, the mean evolutionary rescue time is $v\lambda_wNp_m$ (\cref{eq:AsymptoticTimeRules}), % I removed the mention of exponential distribution as it was not explained and I dont see why you need it here
- which is independent of parameters characterizing aneuploid cells or their production ($u$, $\lambda_a$, and $\Delta_a$). 
+In our focus parameter regime, we have $\Delta_w=-0.04$ and $\Delta_a=-10^{-4}$, such that the mean rescue time is mainly determined by the aneuploid growth rate, $\tau_a \approx 10^4$.
+
+For large tumors ($N \gg N_m^*$), the mean evolutionary rescue time (\cref{eq:AsymptoticTimeRules}) is independent of parameters characterizing aneuploid cells or their production ($u$, $\lambda_a$, and $\Delta_a$).
+% I removed the mention of exponential distribution as it was not explained and I dont see why you need it here
  Increasing the per division mutation rate, $v\lambda_w$, leads to faster appearance of a rescue mutation and hence reduced mean rescue time. 

### 2024-02-12 EvolutionaryRescue1.tex
Add files via upload
edits Remus

@@
 We use multi-type branching processes to approximate the probability that a tumor survives drug treatment as a function of the initial tumor size, and rates at which aneuploidy and other beneficial mutations occur, and the growth rates of the sensitive and resistant cells.
+Additionally, we investigate what effect having a fraction of tumor cells being aneuploid before the onset of therapy have on the probability of evolutionary rescue.
 We also approximate the mean recurrence time for the tumor to revert back to its initial size.
 We find that aneuploidy can play an important role in the relapse of secondary tumors who have not yet been detected and thus are smaller in size.
-% TODO missing something here...
+% TODO missing something here... ok
 \end{abstract}
@@
 Both aneuploid and wildtype cells may divide and mutate to become mutants at rates $v\lambda_{a}$ and $v\lambda_{w}$, respectively.
-To model standing genetic variation, we assume that before the onset of therapy, wildtype cells become aneuploid with rate $\tilde{u}\lambda_w$ (which may differ from $u \lambda_w$) and that aneuploidy confers a fitness cost $s$ in the drug-free environment.
-% TODO unclear how s affects \lambda_w
+To model standing genetic variation, we assume that before the onset of therapy, wildtype cells become aneuploid with rate $\tilde{u}\lambda_w$ (which may differ from $u \lambda_w$) and that aneuploidy confers a fitness cost $s$ in the drug-free environment (we assume that aneuploidy significantly increases the death rate of wildtype cancer cells in an environment which is not affected by chemotherapeutic drugs).
+% TODO unclear how s affects \lambda_w; I do not think that s has any effect on \lambda_w as s is the fitness of the aneuploid in the drug free environment
 See \Cref{figureAneuploidy} for a schematic representation of the model and \Cref{sampleTrajectories} for sample trajectories of the different genotypes.
@@
  Aneuploid death rate $\mu_a$ is set to the same value as the mutant death rates, $\mu_m=0.09$ per day, given that aneuploidy increases resistance to the drug, such as cisplatin, by antagonizing cell division \citep{replogle2020aneuploidy}.
-Aneuploid birth rate is selected such that the aneuploid growth rate $\Delta_w\ll\Delta_a\ll\Delta_m$ which means that $0.06 \le \lambda_a \le 0.1$.
+Aneuploid division rate is selected such that the aneuploid growth rate $\Delta_w\ll\Delta_a\ll\Delta_m$ which means that $0.06 \le \lambda_a \le 0.1$.
 For most of our simulations we use $\lambda_a=0.0899$ per day, such that aneuploidy can only act as a \emph{stepping stone} for the generation of the mutant that rescues the cancer cell population.
@@
 \citet{bakker2023predicting} determined that the missegregation rate must be between $10^{-3}-10^{-2}$ per chromosome per cell division.
-Given that a duplication is is beneficial for cancer resistance only in one or two chromosomes (see details next paragraph) we use $u=10^{-2}$ per cell division.
+Given that a duplication is beneficial for cancer resistance only in one or two chromosomes (see details next paragraph) we use $u=10^{-2}$ per cell division.
 For the missegregation rate in the drug-free environment, $\tilde{u}$, we use the lower-end value of $\tilde{u}=10^{-3}$ per cell division, as some drugs increase the rate of aneuploidy~\citep{wang2019molecular,mason2017functional}.
@@
 A more realistic model includes competition for limited resources and spatial structure, which may play an important role in the development of cancer \citep[e.g.,][]{martens2011spatial}.
-To simulate birth and death rates that depend on the number of cells in the population, we assume the rates of division and death to be
+To simulate division and death rates that depend on the number of cells in the population, we assume the rates of division and death to be
 \begin{align*}
@@
 The first line describes the case in which aneuploid cells are still effectively killed by the treatment, but not as quickly as the wild type. 
-In the second case, aneuploid cells are sufficiently resistant that the size of each aneuploid lineage is expected to remain roughly constant.
+In the second case, aneuploid cells are sufficiently resistant that the expected size of each aneuploid lineage is roughly 1.
 % TODO that's a weird sentence; a lineage starts with a single cell, if the lineage size stays constant that it stays one.
@@
 try to rephrase.
+ok
 In both of these cases, aneuploidy increases the probability of rescue by slowing or halting the decrease of the cancer population, allowing more opportunities for producing resistant mutants. 
 In the third case, aneuploid cells are sufficiently resistance for the cancer population to re-grow the tumor even without additional resistance mutations.
-Notably, in this case there the mutant parameters ($v$, $\lambda_m$, and $\Delta_m$) do not affect $N_a^*$ beyond their effect on $T^*$.
+Notably, in this case the mutant parameters ($v$, $\lambda_m$, and $\Delta_m$) do not affect $N_a^*$ beyond their effect on $T^*$.
 In all cases, $N_a^*$ is proportional to $1/u$ such that increasing the missegregation rate $u$ will decrease the threshold tumor size (\Cref{rescue_threshold}B).
-Furthermore, increasing the aneuploid growth rate $\Delta_a$ (which appears both in the terms and in the conditions), also reduces the threshold tumor size, with a sharp decrease around $\Delta_a=0$, but the effect is minor when $|\Delta_a|$ is small compared to $T^*$ (\Cref{rescue_threshold}A).
-% TODO not sure what the last sentence "but ..." means and how do you see it
-% TODO these are pretty straightforward - can you glean something more interesting from eq 4?
+Furthermore, increasing the aneuploid growth rate $\Delta_a$ (which appears both in the terms and in the conditions), also reduces the threshold tumor size, with a sharp decrease around $\Delta_a=0$, but the effect is minor when $|\Delta_a|$ is small compared to $T^*$ as this would result in the second case where  $dN_a^*/d\Delta_a=0$ (\Cref{rescue_threshold}A).
+% TODO not sure what the last sentence "but ..." means and how do you see it ok
+% TODO these are pretty straightforward - can you glean something more interesting from eq 4? I am still looking for more interesting results from eq4
 
@@
 We can categorize tumors by their size: small tumors with size $N<N_a^*$, intermediate tumors with size $N_a^*<N<N_m^*$, and large tumors with size $N>N_m^*$ .
-For the parameter values in \Cref{table1} with $\lambda_a=0.0899,\mu_w=0.14, u=10^{-2}, v=10^{-7}$, we are in the partially resistant aneuploid case, and substituting in \cref{eq:N_a,eq:N_m}, we have $N_a^* \approx 4.4 \cdot 10^6$ and $N_m^* \approx 4 \cdot 10^7$.
+For the parameter values in \Cref{table1} with $\lambda_a=0.0899,\mu_w=0.14, u=10^{-2}, v=10^{-7}$, we are in the tolerant aneuploid case, and substituting in \cref{eq:N_a,eq:N_m}, we have $N_a^* \approx 4.4 \cdot 10^6$ and $N_m^* \approx 4 \cdot 10^7$.
 Hence, we obtain the ratio $N^*_a/N^*_m \approx 0.11$ (\cref{eq:N_ratio}), that is, aneuploidy reduces the threshold tumor size by approximately 90\% 
@@
 \end{equation}
-Comparing these approximations to results of stochastic evolutionary simulations, we find that the approximations perform very well (\Cref{rescue_denovo}).  % TODO which approximations? unclear
+Comparing the above approximation to results of stochastic evolutionary simulations, we find that the approximations perform very well (\Cref{rescue_denovo}).  % TODO which approximations? unclear ok
 Standing genetic variation will drive evolutionary rescue to the drug if wildtype growth rate $\Delta_w$ is very negative due to a stronger effect of the drug on sensitive cells, or if the aneuploidy cost $s$ is very small due to a low cost of aneuploidy in the pre-drug conditions.
@@
 However, if the aneuploid growth rate $\Delta_a$ increases, then evolutionary rescue is driven by standing genetic variation.
-For the parameter values of  $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, \tilde{u}=10^{-3}, v=10^{-7}$, we are in the partially resistant aneuploid case and obtain the ratio $\tilde{N}^*_a/N^*_m \approx 0.356$, which means that standing genetic variation reduces the threshold tumor size by approximately $65\%$.
+For the parameter values of  $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, \tilde{u}=10^{-3}, v=10^{-7}$, we are in the tolerant aneuploid case and obtain the ratio $\tilde{N}^*_a/N^*_m \approx 0.356$, which means that standing genetic variation reduces the threshold tumor size by approximately $65\%$.
 We observe that while standing genetic variation does not drive evolution of drug resistance when compared to de-novo aneuploidy, it does offer an advantage when compared with direct mutation.
@@
 When the expected number of resistant lineages that avoid extinction is small, the expected recurrence time can be estimated by adding two terms: the \emph{mean evolutionary rescue time}, which is the waiting time for appearance of a resistant lineage that avoids extinction (conditioned on such an even occurring in the first place), and the \emph{mean proliferation time}, which is the expected time for that lineage to grow to $N$ cells.
-However, when the expected number of resistant lineages is large, the population of mutant cells behaves deterministically and the mean recurrence time cannot be separated into the mean evolutionary rescue time and mean proliferation time.
+However, when the expected number of resistant lineages is large, the behavior of the size of the population of mutant cells is deterministic (i.e.
+it is modeled by a system of ODEs, see \Cref{detODE}) and the mean recurrence time cannot be separated into the mean evolutionary rescue time and mean proliferation time because multiple mutant lineages contribute towards the mutant population size reaching the initial tumor size.
 % TODO I dont like the word "behaves", please be specific.
 - what exactly is deterministic and why does it mean you cannot do the separation.
-Of particular interest is the distribution of the evolutionary rescue time and recurrence time with partially tolerant aneuploid cells ($\Delta_a T^*\ll1$), for which we focus on the parameter values  $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$ .
+ok
+Of particular interest is the distribution of the evolutionary rescue time and recurrence time with tolerant aneuploid cells ($\Delta_a T^*\ll1$), for which we focus on the parameter values  $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$ .
 
@@
 % I removed the mention of exponential distribution as it was not explained and I dont see why you need it here
- Increasing the per division mutation rate, $v\lambda_w$, leads to faster appearance of a rescue mutation and hence reduced mean rescue time. 
+ Increasing the per division mutation rate, $v$, leads to faster appearance of a rescue mutation and hence reduced mean rescue time. 
 Finally, increasing the tumor size leads to shorter mean rescue time, as there are more wildtype cells that can mutate to become resistant.

### 2024-02-15 EvolutionaryRescue1.tex
minor edits

@@
 \paragraph{Aneuploidy in cancer.}  Each year approximately 10 million people die from cancer \citep{kocarnik2022cancer}, so understanding the factors that contribute to failure of interventions is of great importance.
-One hypothesized factor is aneuploidy, where cells are characterized by an imbalanced karyotype and chromosome copy number alterations \citep{schukken2018cin}, caused by chromosomal instability (CIN), the mitotic process in which cells suffer from chromosome mis-segregation that leads to aneuploidy.
+One hypothesized factor is aneuploidy, where cells are characterized by an imbalanced karyotype and chromosome copy number alterations \citep{schukken2018cin}, caused by chromosomal instability, the mitotic process in which cells suffer from chromosome mis-segregation that leads to aneuploidy.
 Importantly, aberrations in chromosome copy number have been shown to allow cancer cells to survive under stressful conditions such as drug therapy \citep{lukow2021chromosomal,rutledge2016selective}.
@@
 We assume the population of cells is under a strong stress, such as drug therapy, to which the wildtype genotype is susceptible or sensitive and therefore $\Delta_w<0$, whereas the mutant is resistant to the stress, $\Delta_m>0$.
-We analyze three scenarios: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_w$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing, stationary or ``barely growing'', that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$.
+We analyze three scenarios: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_w$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing, stationary or growing or dying only very slowly, that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$, in a sense that we will make precise below.
 We assume that both chromosomal missegregation and mutation occur during the process of mitosis. 
@@
 Each iteration of the simulation loop starts with computing the rates $\nu_k$ of each event $k$.
-We then draw the time until the next event, $\Delta t$, from an exponential distribution whose rate parameter is the sum of the rates of all events, such that $\Delta t \sim \textit{Exp}(\sum_j \nu_j)$.
+We then draw the time until the next event, $\Delta t$, from an exponential distribution whose rate parameter is the sum of the rates of all events, such that $\Delta t \sim \text{Exp}(\sum_j \nu_j)$.
 Then, we randomly determine which event occurred, where the probability for event $k$ is $p_k = \nu_k/\sum_j \nu_j$.
@@
 This gives us a fitness cost of $s= 7.3\times10^{-3}$ per day.
+\daniel{THIS DERIVATION DOESN'T MAKE SENSE.
+FROM THE LUKOW EXPERIMENT WE CAN DIRECTLY ESTIMATE THE FITNESS COST AS $s \approx \ln(0.5/0.15)/24 \approx .05$ PER DAY.
+DOING IT MORE CAREFULLY (LOGISTIC RATHER THAN EXPONENTIAL) GIVES ABOUT THE SAME ANSWER.}
 We note that when we refer to wildtype cancer cells we include those cells that have any aneuploidy except trisomy in Chr II and VI as those are the aneuploid cells which are hypothesized to have higher fitness in the environment altered by drugs such as vemurafenib.
@@
 
-In our analysis we assume that lineages produced by cells from the initial population divide and die independently of each other, which may be unrealistic, as cells usually compete for resources.
-This is a good approximation as we assume that the drugs cause the wildtype population to decline rapidly and, as a result, density-dependent effects can be neglected.
-A more realistic model includes competition for limited resources and spatial structure, which may play an important role in the development of cancer \citep[e.g.,][]{martens2011spatial}.
-To simulate division and death rates that depend on the number of cells in the population, we assume the rates of division and death to be
+In our analytical calculations, we assume that lineages produced by cells from the initial population divide and die independently of each other.
+Of course, the cells will actually compete for resources, but we expect that this can be neglected because the drug will
+cause the cell density to rapidly drop far below the carrying capacity at which these interactions are important.
+To test this, we simulate a logistic growth model, with division and death rates given by:
 \begin{align*}
@@
 \begin{equation} \label{eq:N_m}
-N_m^* \approx \frac{\abs{\Delta_w}}{v\lambda_w} \cdot \frac{\lambda_m}{\Delta_m} .
+N_m^* \approx \frac{\abs{\Delta_w}}{v\lambda_w}  \frac{\lambda_m}{\Delta_m} .
 \end{equation}
@@
   \frac{\abs{\Delta_w}}{u\lambda_w} \cdot \begin{cases}
-    \frac{\abs{\Delta_a}}{v\lambda_a} \cdot \frac{\lambda_m}{\Delta_m} ,&
+    \frac{\abs{\Delta_a}}{v\lambda_a}  \frac{\lambda_m}{\Delta_m} ,&
   \Delta_a T^* \ll -1 \text{ (tolerant aneuploids)},\\ 
-  %\left(\frac{\lambda_a}{v} \cdot \frac{\lambda_m}{\Delta_m}\right)^{1/2} ,&
+  %\left(\frac{\lambda_a}{v}  \frac{\lambda_m}{\Delta_m}\right)^{1/2} ,&
   2\lambda_a T^* ,&
@@
   \Delta_a T^* \ll -1 ,\\ 
-  \frac{1}{u}\left(v \cdot \frac{\Delta_m}{\lambda_m}\right)^{1/2} ,&
+  \frac{1}{u}\left(v  \frac{\Delta_m}{\lambda_m}\right)^{1/2} ,&
   -1 \ll \Delta_a T^* \ll 1  ,\\ 
-  v \frac{\Delta_m}{\lambda_m} \cdot \left(u\frac{\Delta_a}{\lambda_a}\right)^{-1}  ,&
+  v \frac{\Delta_m}{\lambda_m}  \left(u\frac{\Delta_a}{\lambda_a}\right)^{-1}  ,&
    \Delta_a T^* \gg 1 .
@@
 In the third case, $\left(v \frac{\Delta_m}{\lambda_m}\right) / \left(u \frac{\Delta_a}{\lambda_a}\right)$ is the ratio of the rates of formation of resistant mutants that avoid extinction and partially resistant aneuploids that avoid extinction.
-In the second case, $\frac{1}{u}\left(v \cdot \frac{\Delta_m}{\lambda_m}\right)^{1/2}=\sqrt{\frac{\Delta_a}{u\lambda_a} \cdot v \frac{\Delta_m}{\lambda_m} \cdot \left(u\frac{\Delta_a}{\lambda_a}\right)^{-1}}$, which is the geometric mean of the first and third cases.
-
-Interestingly, increasing both the aneuploid division rate, $\lambda_a$, and the aneuploid death rate, $\mu_a$, such that the growth rate $\Delta_a$ remains constant, leads to decreases in $T^*$, and therefore to the second case.
+In the second case, $\frac{1}{u}\left(v  \frac{\Delta_m}{\lambda_m}\right)^{1/2}=\sqrt{\frac{\Delta_a}{u\lambda_a}  v \frac{\Delta_m}{\lambda_m}  \left(u\frac{\Delta_a}{\lambda_a}\right)^{-1}}$, which is the geometric mean of the first and third cases.
+
+Interestingly, increasing both the aneuploid division rate, $\lambda_a$, and the aneuploid death rate, $\mu_a$, such that the growth rate $\Delta_a$ remains constant, leads to decreases in $T^*$, pushing the system to the second case.
 In this case, increasing the division rate $\lambda_a$ should also increase the mutation rate $v\lambda_a$ in aneuploid cells, as mutations mostly occur during division, so overall the threshold tumor size $N_a^*$ is unaffected by the division rate $\lambda_a$ (i.e., $d \lambda_a T^*/d\lambda_a = 0$).
@@
 
-We can categorize tumors by their size: small tumors with size $N<N_a^*$, intermediate tumors with size $N_a^*<N<N_m^*$, and large tumors with size $N>N_m^*$ .
-For the parameter values in \Cref{table1} with $\lambda_a=0.0899,\mu_w=0.14, u=10^{-2}, v=10^{-7}$, we are in the tolerant aneuploid case, and substituting in \cref{eq:N_a,eq:N_m}, we have $N_a^* \approx 4.4 \cdot 10^6$ and $N_m^* \approx 4 \cdot 10^7$.
-Hence, we obtain the ratio $N^*_a/N^*_m \approx 0.11$ (\cref{eq:N_ratio}), that is, aneuploidy reduces the threshold tumor size by approximately 90\% 
-Interestingly, the threshold between small and intermediate tumors, $N_a^*$, is similar to the tumor detection threshold of $4.19 \cdot 10^6$ cells for a wide variety of tumors as reported by \citet{avanzini2019cancer}.
+We can categorize tumors by their size: small tumors with size $N<N_a^*$ that are unlikely to survive treatment, intermediate tumors with size $N_a^* < N < N_m^*$ that rely on aneuploidy for evolutionary rescue, and large tumors with size $N > N_m^*$ that could be evolutionarily rescued even without aneuploidy.
+For the parameter values in \Cref{table1} with $\lambda_a=0.0899,\mu_w=0.14, u=10^{-2}, v=10^{-7}$, we are in the tolerant aneuploid case, and substituting in \cref{eq:N_a,eq:N_m}, we have $N_a^* \approx 4 \times 10^6$ and $N_m^* \approx 4 \times 10^7$.
+Hence, we obtain the ratio $N^*_a/N^*_m \approx 0.11$ (\cref{eq:N_ratio}), that is, aneuploidy reduces the threshold tumor size by approximately 90\%.
+Interestingly, the threshold between small and intermediate tumors, $N_a^*$, is similar to the tumor detection threshold of $4 \times 10^6$ cells for a wide variety of tumors as reported by \citet{avanzini2019cancer}.
 
@@
 In contrast, de-novo aneuploid cells will have a greater contribution to rescue if the aneuploidy cost, $s$, is large, the effect of the drug is weak ($\Delta_w$ is small), or if the drug induces the appearance of aneuploid cells ($u > \tilde u$).
-For example, with  $\lambda_w=0.1,\mu_w=0.14, u=10^{-2}, \tilde{u}=10^{-3}, s=7.3\cdot10^{-3}$, the ratio of the threshold tumor sizes for standing vs de-novo variation is $\tilde{N}^*_a/N^*_a \approx 1.825$, which means that de-novo genetic variation is  the main driver of evolutionary rescue.
+For example, with  $\lambda_w=0.1,\mu_w=0.14, u=10^{-2}, \tilde{u}=10^{-3}, s=7.3\times10^{-3}$, the ratio of the threshold tumor sizes for standing vs de-novo variation is $\tilde{N}^*_a/N^*_a \approx 1.825$, which means that de-novo genetic variation is  the main driver of evolutionary rescue.
 
@@
 \begin{equation} \label{eq:N_sgv_ratio}
-\frac{\tilde{N}^*_a}{N^*_m}=\frac{\tilde{N}^*_{a}}{N^*_{a}} \cdot \frac{N^*_a}{N^*_m} \approx \frac{s}{\abs{\Delta_w}}\begin{cases}
+\frac{\tilde{N}^*_a}{N^*_m}=\frac{\tilde{N}^*_{a}}{N^*_{a}} \frac{N^*_a}{N^*_m} \approx \frac{s}{\abs{\Delta_w}}\begin{cases}
     \frac{\abs{\Delta_a}}{\tilde{u}\lambda_a} ,&
   \Delta_a T^* \ll -1 ,\\ 
-  \frac{1}{\tilde{u}}\left(v \cdot \frac{\Delta_m}{\lambda_m}\right)^{1/2} ,&
+  \frac{1}{\tilde{u}}\left(v \frac{\Delta_m}{\lambda_m}\right)^{1/2} ,&
   -1 \ll \Delta_a T^* \ll 1  ,\\ 
-  v \frac{\Delta_m}{\lambda_m} \cdot \left(\tilde{u}\frac{\Delta_a}{\lambda_a}\right)^{-1}  ,&
+  v \frac{\Delta_m}{\lambda_m} \left(\tilde{u}\frac{\Delta_a}{\lambda_a}\right)^{-1}  ,&
    \Delta_a T^* \gg 1 .
@@
  N \ll N_a^* ,\\ 
-  \frac{1}{v\lambda_w N} \cdot  \frac{\lambda_m}{\Delta_m} ,& % I replaced pm with 𝜆𝑚 − 𝜇𝑚 / 𝜆𝑚
+  \frac{1}{v\lambda_w N}   \frac{\lambda_m}{\Delta_m} ,& % I replaced pm with 𝜆𝑚 − 𝜇𝑚 / 𝜆𝑚
   N \gg N_m^* .
@@
 
-We have assumed that cancer cell lineages are independent of each other.
-However, this may not be the case, as cancer cells compete for resources (e.g., blood supply).
-Nevertheless, we find that when the effective carrying capacity is not too large ($K\sim10^8, K_e\sim10^6$)
-our approximation for the probability of evolutionary rescue agrees with results of stochastic simulations with density-dependent growth  (\Cref{LogisticPlot}).
-Future work may focus on scenarios with small carrying capacity by analyzing density-dependent branching processes~\citep{klebaner1997population,harris1963theory}.
-
-Solid tumors are spatially heterogeneous with different genotypes inhabiting cellular niches and immune infiltration impacting growth in affected regions~\citep{varrone2023cellcharter,galon2010immune}.
-This has the potential to impact the probability of overcoming chemotherapy~\citep{martens2011spatial}.
+We have assumed that cancer cell lineages are independent of each other,
+and have verified that this approximation is accurate under simple logistic growth. 
+However, this neglects potential effects of spatial structure and local interactions, which may be very important in solid tumors.
+Such tumors can be spatially heterogeneous with different genotypes inhabiting cellular niches and immune infiltration impacting growth in affected regions \citep{varrone2023cellcharter,galon2010immune}.
+This has the potential to impact the probability of overcoming chemotherapy \citep{martens2011spatial}.
 Future work should take into consideration the spatial structure of the tumor and its effect on the probability of evolutionary rescue.
@@
 As a result, incorporating a wider distribution of parameters would be a beneficial extension for our model.
+
 \paragraph{Conclusions}
@@
 the Simons Foundation (Investigator in Mathematical Modeling of Living Systems $\#508600$, DBW),
-the Sloan Foundation Research Fellowship (FG-2021-16667, DBW),
-the NSF grant ($\#2146260$, DBW),
+the Sloan Foundation (Research Fellowship FG-2021-16667, DBW),
+the National Science Foundation (grant $\#2146260$, DBW),
 % TODO add funding for Uri
 }
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\section*{References}
+
+\nolinenumbers
+%\bibliographystyle{unsrtnat}
+\bibliographystyle{agsm}
+\bibliography{evo2022}
+
+\clearpage
 
@@
     $\tilde{u}$ & Missegregation rate in the drug free environment$^\ast$& $10^{-3}$ & 1$\slash$cell division  & - \\ \hline
-    $s$ & Selection coefficient of aneuploidy in the drug free environment& $7.3\cdot10^{-3}$ &  1/days   & \citet{lukow2021chromosomal} \\  
+    $s$ & Selection coefficient of aneuploidy in the drug free environment& $7.3 \times 10^{-3}$ &  1/days   & \citet{lukow2021chromosomal} \\  
     \hline
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\section*{References}
-
-\nolinenumbers
-%\bibliographystyle{unsrtnat}
-\bibliographystyle{agsm}
-\bibliography{evo2022}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \begin{equation}\label{eq:pw_parttolerant}
-p_w \approx \frac{u\lambda_w}{\abs{\Delta_w}} \cdot \frac{\Delta_a}{\lambda_a} .
+p_w \approx \frac{u\lambda_w}{\abs{\Delta_w}}  \frac{\Delta_a}{\lambda_a} .
 \end{equation}
@@
 \begin{align*}
-\frac{u\lambda_w\Delta_a}{\lambda_a}<\frac{v\lambda_w\Delta_m}{\lambda_m}\Rightarrow u\lambda_a<\frac{v\lambda_a^2\Delta_m}{\lambda_m}\cdot\frac{1}{\Delta_a}<\frac{v\lambda_a^2\Delta_m}{\lambda_m}\cdot\frac{1}{\sqrt{4\lambda_a^2 v p_m}}\Rightarrow u\lambda_a<T^*,
+\frac{u\lambda_w\Delta_a}{\lambda_a}<\frac{v\lambda_w\Delta_m}{\lambda_m}\Rightarrow u\lambda_a<\frac{v\lambda_a^2\Delta_m}{\lambda_m} \frac{1}{\Delta_a}<\frac{v\lambda_a^2\Delta_m}{\lambda_m} \frac{1}{\sqrt{4\lambda_a^2 v p_m}}\Rightarrow u\lambda_a<T^*,
 \end{align*}
@@
 \begin{equation*}
-p_w \approx \frac{v\lambda_w}{\abs{\Delta_w}} \cdot \frac{\Delta_m}{\lambda_m} .
+p_w \approx \frac{v\lambda_w}{\abs{\Delta_w}}  \frac{\Delta_m}{\lambda_m} .
 \end{equation*}
@@
 \begin{equation}\label{eq:pw_tolerant}
-p_w\approx\frac{u\lambda_w}{\abs{\Delta_w}}\cdot\frac{v\lambda_a}{\abs{\Delta_a}}\cdot\frac{\Delta_m}{\lambda_m}.
+p_w\approx\frac{u\lambda_w}{\abs{\Delta_w}} \frac{v\lambda_a}{\abs{\Delta_a}} \frac{\Delta_m}{\lambda_m}.
 \end{equation}
@@
 \begin{equation}\label{eq:pw_partrest}
-p_w\approx\frac{u\lambda_w}{\abs{\Delta_w}}\cdot\sqrt{\frac{v\lambda_a\Delta_m}{\lambda_m}}.
+p_w\approx\frac{u\lambda_w}{\abs{\Delta_w}} \sqrt{\frac{v\lambda_a\Delta_m}{\lambda_m}}.
 \end{equation}
@@
   &\begin{cases}
-   1-\exp\left[-\frac{u\lambda_a}{\abs{\Delta_w}}\cdot\frac{v\lambda_w}{\abs{\Delta_a}}\cdot\frac{\Delta_m}{\lambda_m}\cdot N\right] ,&
+   1-\exp\left[-\frac{u\lambda_a}{\abs{\Delta_w}} \frac{v\lambda_w}{\abs{\Delta_a}} \frac{\Delta_m}{\lambda_m}  N\right] ,&
    \Delta_aT^*\ll-1 ,\\
-   1-\exp\left[-\frac{u\lambda_w}{\abs{\Delta_w}}\cdot\sqrt{\frac{v\lambda_a\Delta_m}{\lambda_m}}\cdot N\right] ,&
+   1-\exp\left[-\frac{u\lambda_w}{\abs{\Delta_w}} \sqrt{\frac{v\lambda_a\Delta_m}{\lambda_m}} N\right] ,&
   -1\ll\Delta_aT^*\ll1 ,\\
-   1-\exp\left[-\frac{u\lambda_w}{\abs{\Delta_w}} \cdot \frac{\Delta_a}{\lambda_a}\cdot N\right] ,&
+   1-\exp\left[-\frac{u\lambda_w}{\abs{\Delta_w}}  \frac{\Delta_a}{\lambda_a}  N\right] ,&
    1\ll\Delta_aT^*.

### 2024-02-19 EvolutionaryRescue1.tex
edits and comments

@@
 Both aneuploid and wildtype cells may divide and mutate to become mutants at rates $v\lambda_{a}$ and $v\lambda_{w}$, respectively.
-To model standing genetic variation, we assume that before the onset of therapy, wildtype cells become aneuploid with rate $\tilde{u}\lambda_w$ (which may differ from $u \lambda_w$) and that aneuploidy confers a fitness cost $s$ in the drug-free environment (we assume that aneuploidy significantly increases the death rate of wildtype cancer cells in an environment which is not affected by chemotherapeutic drugs).
-% TODO unclear how s affects \lambda_w; I do not think that s has any effect on \lambda_w as s is the fitness of the aneuploid in the drug free environment
+To model standing genetic variation, we assume that before the onset of therapy, wildtype cells become aneuploid with rate $\tilde{u}\lambda_w$ (which may differ from $u \lambda_w$) and that aneuploidy confers a fitness cost $s$ in the drug-free environment, that is, we assume that aneuploid cells have an increased death rate compared to wildtype cells in a drug-free environment.
 See \Cref{figureAneuploidy} for a schematic representation of the model and \Cref{sampleTrajectories} for sample trajectories of the different genotypes.
@@
 In the second case, aneuploid cells are sufficiently resistant that the expected size of each aneuploid lineage is roughly 1.
-% TODO that's a weird sentence; a lineage starts with a single cell, if the lineage size stays constant that it stays one.
-so what are we talking about?
-try to rephrase.
-ok
 In both of these cases, aneuploidy increases the probability of rescue by slowing or halting the decrease of the cancer population, allowing more opportunities for producing resistant mutants. 
@@
 Furthermore, increasing the aneuploid growth rate $\Delta_a$ (which appears both in the terms and in the conditions), also reduces the threshold tumor size, with a sharp decrease around $\Delta_a=0$, but the effect is minor when $|\Delta_a|$ is small compared to $T^*$ as this would result in the second case where  $dN_a^*/d\Delta_a=0$ (\Cref{rescue_threshold}A).
-% TODO not sure what the last sentence "but ..." means and how do you see it ok
-% TODO these are pretty straightforward - can you glean something more interesting from eq 4? I am still looking for more interesting results from eq4
+% TODO these are pretty straightforward - can you glean something more interesting from eq 4?
+I am still looking for more interesting results from eq4
 
@@
 Thus, if aneuploid cells rapidly die due to the drug but compensate by rapidly dividing, further increasing  the division rate will \emph{not} facilitate adaptation.
-% NOTE this is the kind of conclusions we want to drive home
 This is consistent with experimental findings where aneuploidy confers resistance by decreasing the division rate~\citep{replogle2020aneuploidy}.
@@
 In the above we assumed that at the onset of drug treatment, the initial tumor consisted entirely of wildtype cells that are drug sensitive.
-However, aneuploidy is likely produced even before onset of treatment at some rate $\tilde{u} \le u$ and confer a deleterious fitness effect $s$ in the absence of the drug \citep{replogle2020aneuploidy,giam2015aneuploidy}, and the aneuploidy rate in the presence of drugs, $u$, is likely higher than in their absence, $\tilde{u}$ \citep{wang2019molecular,mason2017functional}.
-But if the number of cells in the tumor $N$ is large (as expected if the tumor is to be treated with a drug), there may already be a fraction $f \approx \tilde{u}\lambda_w/s$ (we assume that the division rate of the wildtype in a drug-free environment is equal to the division rate of the wildtype in the environment affected by the anti-cancer drug) of aneuploid cells in the population.
-
-Therefore, the ratio of threshold tumor size with standing generation variation, $\tilde{N}^*_{a}$, is similar to the ratio with de-novo variation, $N^*_a$, except that the sensitive growth rate $\abs{\Delta_w}$ is replaced with the aneuploidy cost, $s$, such that
+However, aneuploidy is likely produced even before onset of treatment at some rate $\tilde{u}$ and confers a deleterious fitness effect $s$ in the absence of the drug \citep{replogle2020aneuploidy,giam2015aneuploidy}.
+Furthermore, the aneuploidy rate in the presence of drugs is likely higher than in their absence, $\tilde{u} < u$ \citep{wang2019molecular,mason2017functional}.
+But if the number of cells in the tumor $N$ is large (as expected if the tumor is treated with a drug), there may already be a fraction $f \approx \tilde{u}\lambda_w/s$ of aneuploid cells in the population (here we assume that the drug affects the wildtype death rate but no the division rate and therefre use $\lambda_w$ for the wildtype division rate in the drug-free enviornment).
+
+Therefore, the ratio of threshold tumor size with standing generation variation, $\tilde{N}^*_{a}$, is similar to the ratio with de-novo variation, $N^*_a$, except that the sensitive growth rate $\abs{\Delta_w}$ is replaced with the aneuploidy cost $s$, such that
 \begin{equation}\label{eq:Nsgv}
@@
 \end{equation}
-Comparing the above approximation to results of stochastic evolutionary simulations, we find that the approximations perform very well (\Cref{rescue_denovo}).  % TODO which approximations? unclear ok
-Standing genetic variation will drive evolutionary rescue to the drug if wildtype growth rate $\Delta_w$ is very negative due to a stronger effect of the drug on sensitive cells, or if the aneuploidy cost $s$ is very small due to a low cost of aneuploidy in the pre-drug conditions.
-In contrast, de-novo aneuploid cells will have a greater contribution to rescue if the aneuploidy cost, $s$, is large, the effect of the drug is weak ($\Delta_w$ is small), or if the drug induces the appearance of aneuploid cells ($u > \tilde u$).
+Comparing this approximation of $\tilde{N}^*_{a}/N^*_{a}$ to results of stochastic simulations, we find that the approximations perform very well (\Cref{rescue_denovo}). 
+Standing genetic variation will drive evolutionary rescue to the drug if wildtype growth rate $\Delta_w$ is very negative due to a strong effect of the drug on sensitive cells, or if the cost aneuploidy cost in the drug-free environment, $s$, is small.
+In contrast, de-novo aneuploid cells will have a greater contribution to rescue if the aneuploidy cost $s$ is large, the effect of the drug on sensitive cells is weak ($\Delta_w$ is close to zero), or if the drug induces the appearance of aneuploid cells ($u > \tilde u$).
 For example, with  $\lambda_w=0.1,\mu_w=0.14, u=10^{-2}, \tilde{u}=10^{-3}, s=7.3\times10^{-3}$, the ratio of the threshold tumor sizes for standing vs de-novo variation is $\tilde{N}^*_a/N^*_a \approx 1.825$, which means that de-novo genetic variation is  the main driver of evolutionary rescue.
@@
 \end{equation}
-We observe that evolutionary rescue through direct mutation is more likely if the cost of aneuploidy $s$ is very large or the effect of the drug $\Delta_w$ is small.
+Evolutionary rescue through direct mutation is more likely if the cost of aneuploidy $s$ is very large or the effect of the drug $\Delta_w$ is small.
 In contrast, standing genetic variation will drive adaptation if the pre-treatment chromosome missegreagation rate $\tilde{u}$ is very large.
@@
 For the parameter values of  $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, \tilde{u}=10^{-3}, v=10^{-7}$, we are in the tolerant aneuploid case and obtain the ratio $\tilde{N}^*_a/N^*_m \approx 0.356$, which means that standing genetic variation reduces the threshold tumor size by approximately $65\%$.
-We observe that while standing genetic variation does not drive evolution of drug resistance when compared to de-novo aneuploidy, it does offer an advantage when compared with direct mutation.
+Therefore, standing genetic variation does not drive evolution of drug resistance when compared to de-novo aneuploidy, but it does offer an advantage when compared with direct mutation.
 
@@
 When the expected number of resistant lineages that avoid extinction is small, the expected recurrence time can be estimated by adding two terms: the \emph{mean evolutionary rescue time}, which is the waiting time for appearance of a resistant lineage that avoids extinction (conditioned on such an even occurring in the first place), and the \emph{mean proliferation time}, which is the expected time for that lineage to grow to $N$ cells.
-However, when the expected number of resistant lineages is large, the behavior of the size of the population of mutant cells is deterministic (i.e.
-it is modeled by a system of ODEs, see \Cref{detODE}) and the mean recurrence time cannot be separated into the mean evolutionary rescue time and mean proliferation time because multiple mutant lineages contribute towards the mutant population size reaching the initial tumor size.
+However, when the expected number of resistant lineages is large, the dynamics of number of mutant cells is deterministic (i.e., it can be modeled by a system of ODEs, \cref{detODE}) and the mean recurrence time cannot be separated into the mean evolutionary rescue time and mean proliferation time because multiple mutant lineages contribute towards the mutant population size reaching the initial tumor size.
 % TODO I dont like the word "behaves", please be specific.
@@
 \paragraph{Evolutionary rescue time.}
-We have derived approximations for $\tau_m$, the mean evolutionary rescue time without aneuploidy ($u=0$), and $\tau_a$, the mean rescue time with aneuploidy ($u>0$), both conditioned on evolutionary rescue occurring (see appendix \ref{sec:appendix_rescue_time}).
+We have derived approximations for $\tau_m$, the mean evolutionary rescue time without aneuploidy ($u=0$), and $\tau_a$, the mean rescue time with aneuploidy ($u>0$), both conditioned on evolutionary rescue occurring.
 These approximations are in good agreement with simulation results for small, intermediate, and large tumor sizes (\Cref{EvolutionaryRescueTimeComplete,MeanTimeGrowthAneuploidyPlot}).
-The mean rescue time with aneuploidy for small and large tumors follows these expressions:
+The mean rescue time with aneuploidy for small and large tumors follows these expressions (Appendix \ref{sec:appendix_rescue_time}),
 \begin{equation}  \label{eq:AsymptoticTimeRules}
@@
 
-When a fraction $f\approx 1.36\%$ of the initial cancer cell population is aneuploid we calculate the mean evolutionary rescue time in \cref{meantimet2SGV} which we plot in \Cref{SGVEvolutionaryRescueTimeComplete}.
+When a fraction $f\approx 1.36\%$ of the initial cancer cell population is aneuploid, the mean evolutionary rescue time (\cref{meantimet2SGV}; \Cref{SGVEvolutionaryRescueTimeComplete}.
 We note that standing genetic variation reduces the mean evolutionary rescue time only for intermediary sized tumors but not for large or small tumors.
+% WHAT is the purpose of this sentence?
+
+In Appendix E we derive the probability that by time $t$ a successful mutant has been generated.
+\Cref{cdffig}A show the agreement between our formula and simulation results for the case when aneuploidy is present and when it is absent.
+We observe that aneuploidy start to have an effect on adaptation for timescales greater then 100 days.
+% WHAT is the purpose of this sentence?
+
 \paragraph{Recurrence time.}
 
-In appendix \ref{sec:appendix_recurrence_time} we approximate the mean time for the population of mutant cancer cells to reach the initial population size $N$ which we denote as the recurrence time.
-\Cref{proliferationFigure} shows the agreement between our approximations and simulations.
-For values of $N$ smaller then $10^8$ the mean recurrence time can be approximated as the sum of the mean time for the first rescue mutation to appear and the mean time for the lineage generated by that mutation to reach size $N$.
-For $N>10^8$ the behavior of the mutant cancer cell population is deterministic and the mean recurrence time becomes constant.
-We note that $\tau_a^r$ has the following asymptotic behavior:
+We next approximate the mean time for the population of mutant cancer cells to reach the initial, pre-treatment population size $N$, which we denote the recurrence time $\tau_a^r$ (Appendix \ref{sec:appendix_recurrence_time}),
 \begin{equation} \label{eq:AsymptoticTimeRulesRecurrence}
@@
 \end{equation}
-In the first case, when tumor is small, the mean recurrence time grows logarithmically with tumor size $N$ and is the same order of magnitude as the mean evolutionary rescue time (\Cref{ProliferationTimeLarge}).
-Increasing the mutant growth rate $\Delta_m$ leads to shorter recurrence times while increasing the wildtype and aneuploid growth rates $\Delta_w$ and $\Delta_a$, retrospectively, increases the recurrence time.
-In the second case, when tumor is large, we observe that increasing either the mutant growth rare $\Delta_m$ or the mutation rate $v$ leads to a decrease in the time needed for the tumor to rebound to its initial size.
-Additionally, drugs which significantly increase the wildtype death rate $\mu_w$ and do not affect the division rate $\lambda_w$ delays cancer relapse.
-Consequently, patients which take such drugs require a longer period of monitoring to guarantee the effectiveness of the treatment.
-\paragraph{Distribution of evolutionary rescue time with and without aneuploidy.}
-
-In Appendix E we derive the probability that by time $t$ a successful mutant has been generated.
-\Cref{cdffig}A show the agreement between our formula and simulation results for the case when aneuploidy is present and when it is absent.
-We observe that aneuploidy start to have an effect on adaptation for timescales greater then 100 days.
-
-\paragraph{Distribution of  recurrence time.}
+\Cref{proliferationFigure} shows the agreement between our approximations and simulations.
+For small tumors ($Nt \ll 10^8$), the mean recurrence time can be approximated as the sum of the mean time for the first rescue mutation to appear and the mean time for the lineage generated by that mutation to reach size $N$.
+The mean recurrence time grows logarithmically with tumor size $N$ and is the same order of magnitude as the mean evolutionary rescue time.
+Increasing the mutant growth rate $\Delta_m$ leads to shorter recurrence times while increasing the wildtype and aneuploid growth rates, $\Delta_w$ and $\Delta_a$ respectively, increases the recurrence time.
+For large tumors ($N \gg 10^8$), the dynamics of the number of mutant cells is deterministic and the mean recurrence time becomes constant.
+Increasing either the mutant growth rate $\Delta_m$ or the mutation rate $v$ leads to a decrease in the time for the tumor to rebound to its initial size.
+In addition, drugs that significantly increase the wildtype death rate $\mu_w$ and do not affect the division rate $\lambda_w$ delay cancer relapse.
+Consequently, patients treated with such drugs require a longer period of monitoring to guarantee the effectiveness of the treatment.
+% Remus the next paragraphs look like you wanted to find approximations so you did, you reference them here, but the results are buried in the technical text: WHAT IS IT THAT YOU WANT US TO LEARN from this?
 In Appendix F we derive the probability that mutant cancer cell population reaches $N$ at time $t$.
@@
 We observe that  for small and intermediate sized tumors the effect of the detection size is negligible when compared to $\tau_a^r$.
+% WHAT DOES THAT MEAN?
 However, for large tumors the mean recurrence time to detection size $M$ decreases logarithmically with tumor size $N$.
@@
 Most clinical trials report data on the distribution of recurrence time measured from a varying time of surgery~\citep{avanzini2019cancer} with chemotherapy usually following after.
+% WHAT DOES "varying" mean here?
 Therefore, since only undetected secondary tumors are present at the time of the administration of the anti-cancer drug we lack knowledge of the size of the tumors and we cannot compare the empirical distributions to our predictions.

### 2024-03-11 EvolutionaryRescue1.tex
Add files via upload

@@
 We assume the mutation rate is $10^{-7}$ per gene per cell division \citep{loeb2001mutator} and since we assume that a single target gene confers resistance to the drug, we use $v=10^{-7}$ per cell division.
-\citet{bakker2023predicting} determined that the missegregation rate must be between $10^{-3}-10^{-2}$ per chromosome per cell division.
-Given that a duplication is beneficial for cancer resistance only in one or two chromosomes (see details next paragraph) we use $u=10^{-2}$ per cell division.
-For the missegregation rate in the drug-free environment, $\tilde{u}$, we use the lower-end value of $\tilde{u}=10^{-3}$ per cell division, as some drugs increase the rate of aneuploidy~\citep{wang2019molecular,mason2017functional}.
+\citet{bakker2023predicting} determined that the missegregation rate must be between $10^{-3}-10^{-2}$ per chromosome per cell division with the optimal value being $6.21\times10^{-3}$ per chromosome per cell division.
+\citet{ippolito2021gene} observed that trisomy in Chr II and VI are most likely to confer increased resistance against the chemotherapeutic agent vemurafenib for A375 cells.
+We assume that if a tumor is aneuploid then it most likely has trisomy \citep{gisselsson2010generation} in the pre-treatment environment and that cells with more than one trisomy are very unlikely to survive.
+Additionally, we assume that all trisomies are equally likely and, as a result, we select $u=10^{-2}$ per cell division.
+For the missegregation rate in the drug-free environment, $\tilde{u}$, we use the lower-end value of $\tilde{u}=2\times10^{-3}$ per cell division, as some drugs increase the rate of aneuploidy~\citep{wang2019molecular,mason2017functional}.
 
@@
 To estimate $f$, we note that \citet{lukow2021chromosomal} mixed together wildtype and aneuploid A375 melanoma cells at $50:50$ ratio, cultured them in drug-free environment and observed the ratio evolve as a function of time with the aneuploid cells declining to $15\%$ after 24 days.
-\citet{ippolito2021gene} observed that trisomy in Chr II and VI are most likely to confer increased resistance against the chemotherapeutic agent vemurafenib for A375 cells.
-We assume that if a tumor is aneuploid then it most likely has trisomy \citep{gisselsson2010generation} in the pre-treatment environment and that cells with more than one trisomy are very unlikely to survive.
-Additionally, we assume that all trisomies are equally likely and, as a result, we obtain the fraction of cancer cells with favorable aneuploidy as $f=2/22\times0.15\approx 0.0136$ (i.e.
-$1.36\%$ of pre-treatment cancer cells have the beneficial aneuploidy).
-This gives us a fitness cost of $s= 7.3\times10^{-3}$ per day.
-\daniel{THIS DERIVATION DOESN'T MAKE SENSE.
-FROM THE LUKOW EXPERIMENT WE CAN DIRECTLY ESTIMATE THE FITNESS COST AS $s \approx \ln(0.5/0.15)/24 \approx .05$ PER DAY.
-DOING IT MORE CAREFULLY (LOGISTIC RATHER THAN EXPONENTIAL) GIVES ABOUT THE SAME ANSWER.}
+We obtain the fitness cost $s$ using the formula $s=\abs{\log\left[0.15/(1-0.15)\right]/24}\approx0.07$ per day \citep{chevin2011measuring}.
+As a result, the fraction of cancer cells with ``beneficial'' aneuploidy is $f=2\times10^{-3}\times10^{-1}/0.07=0.284\%$ (i.e.
+$0.284\%$ of pre-treatment cancer cells have the ``beneficial'' aneuploidy).
+
 We note that when we refer to wildtype cancer cells we include those cells that have any aneuploidy except trisomy in Chr II and VI as those are the aneuploid cells which are hypothesized to have higher fitness in the environment altered by drugs such as vemurafenib.
@@
 For the parameter values in \Cref{table1} with $\lambda_a=0.0899,\mu_w=0.14, u=10^{-2}, v=10^{-7}$, we are in the tolerant aneuploid case, and substituting in \cref{eq:N_a,eq:N_m}, we have $N_a^* \approx 4 \times 10^6$ and $N_m^* \approx 4 \times 10^7$.
-Hence, we obtain the ratio $N^*_a/N^*_m \approx 0.11$ (\cref{eq:N_ratio}), that is, aneuploidy reduces the threshold tumor size by approximately 90\%.
+Hence, we obtain the ratio $N^*_a/N^*_m \approx 0.11$ (\cref{eq:N_ratio}), that is, aneuploidy reduces the threshold tumor size by approximately 89\%.
 Interestingly, the threshold between small and intermediate tumors, $N_a^*$, is similar to the tumor detection threshold of $4 \times 10^6$ cells for a wide variety of tumors as reported by \citet{avanzini2019cancer}.
@@
 
-Therefore, the ratio of threshold tumor size with standing generation variation, $\tilde{N}^*_{a}$, is similar to the ratio with de-novo variation, $N^*_a$, except that the sensitive growth rate $\abs{\Delta_w}$ is replaced with the aneuploidy cost $s$, such that
+Therefore, the threshold tumor size with standing generation variation, $\tilde{N}^*_{a}$, is similar to the threshold with de-novo variation, $N^*_a$, except that the sensitive growth rate $\abs{\Delta_w}$ is replaced with the aneuploidy cost $s$, such that
 \begin{equation}\label{eq:Nsgv}
@@
 In contrast, de-novo aneuploid cells will have a greater contribution to rescue if the aneuploidy cost $s$ is large, the effect of the drug on sensitive cells is weak ($\Delta_w$ is close to zero), or if the drug induces the appearance of aneuploid cells ($u > \tilde u$).
-For example, with  $\lambda_w=0.1,\mu_w=0.14, u=10^{-2}, \tilde{u}=10^{-3}, s=7.3\times10^{-3}$, the ratio of the threshold tumor sizes for standing vs de-novo variation is $\tilde{N}^*_a/N^*_a \approx 1.825$, which means that de-novo genetic variation is  the main driver of evolutionary rescue.
+For example, with  $\lambda_w=0.1,\mu_w=0.14, u=10^{-2}, \tilde{u}=2\times10^{-3}, s=0.07$, the ratio of the threshold tumor sizes for standing vs de-novo variation is $\tilde{N}^*_a/N^*_a \approx 8.75$, which means that de-novo genetic variation is  the main driver of evolutionary rescue.
 
@@
 However, if the aneuploid growth rate $\Delta_a$ increases, then evolutionary rescue is driven by standing genetic variation.
-For the parameter values of  $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, \tilde{u}=10^{-3}, v=10^{-7}$, we are in the tolerant aneuploid case and obtain the ratio $\tilde{N}^*_a/N^*_m \approx 0.356$, which means that standing genetic variation reduces the threshold tumor size by approximately $65\%$.
-Therefore, standing genetic variation does not drive evolution of drug resistance when compared to de-novo aneuploidy, but it does offer an advantage when compared with direct mutation.
+For the parameter values of  $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, \tilde{u}=10^{-3}, v=10^{-7}$, we are in the tolerant aneuploid case and obtain the ratio $\tilde{N}^*_a/N^*_m \approx 0.9625$, which means that standing genetic variation reduces the threshold tumor size by approximately $4\%$.
+Therefore, standing genetic variation does not drive evolution of drug resistance when compared to de-novo aneuploidy, but it does offer a slight advantage when compared with direct mutation.
 
@@
 
-When a fraction $f\approx 1.36\%$ of the initial cancer cell population is aneuploid, the mean evolutionary rescue time (\cref{meantimet2SGV}; \Cref{SGVEvolutionaryRescueTimeComplete}.
-We note that standing genetic variation reduces the mean evolutionary rescue time only for intermediary sized tumors but not for large or small tumors.
+When a fraction $f\approx 0.284\%$ of the initial cancer cell population is aneuploid we want to know whether the mean evolutionary rescue time is affected by the standing genetic variation.
+For this purpose we calculate  $\tau_a$ (see \cref{meantimet2SGV}) and compare our result with simulations in  \Cref{SGVEvolutionaryRescueTimeComplete}.
+We note that standing genetic variation does not have a significant effect on the mean evolutionary rescue time.
 % WHAT is the purpose of this sentence?
+ok
 
@@
 \Cref{cdffig}A show the agreement between our formula and simulation results for the case when aneuploidy is present and when it is absent.
-We observe that aneuploidy start to have an effect on adaptation for timescales greater then 100 days.
+We observe that aneuploidy starts to have an effect on adaptation for timescales greater then $1/\Delta_a\approx100$ days after which no more rescue mutations are generated through direct mutation.
+This shows that rescue mutations are generated through aneuploidy at latter timescales then direct mutation and, as a consequence, aneuploidy increases the \emph{window of opportunity} for evolutionary rescue.
 % WHAT is the purpose of this sentence?
+ok
 
@@
 \end{equation}
-\Cref{proliferationFigure} shows the agreement between our approximations and simulations.
-For small tumors ($Nt \ll 10^8$), the mean recurrence time can be approximated as the sum of the mean time for the first rescue mutation to appear and the mean time for the lineage generated by that mutation to reach size $N$.
+\Cref{proliferationFigure,ProliferationTimeLarge} shows the agreement between our approximations and simulations.
+For small tumors ($N \ll N_a^*$), the mean recurrence time can be approximated as the sum of the mean time for the first rescue mutation to appear and the mean time for the lineage generated by that mutation to reach size $N$.
 The mean recurrence time grows logarithmically with tumor size $N$ and is the same order of magnitude as the mean evolutionary rescue time.
 Increasing the mutant growth rate $\Delta_m$ leads to shorter recurrence times while increasing the wildtype and aneuploid growth rates, $\Delta_w$ and $\Delta_a$ respectively, increases the recurrence time.
-For large tumors ($N \gg 10^8$), the dynamics of the number of mutant cells is deterministic and the mean recurrence time becomes constant.
+For large tumors ($N \gg N_m^*$), the dynamics of the number of mutant cells is deterministic and the mean recurrence time becomes constant (and independent of the initial tumor size $N$).
 Increasing either the mutant growth rate $\Delta_m$ or the mutation rate $v$ leads to a decrease in the time for the tumor to rebound to its initial size.
@@
 Consequently, patients treated with such drugs require a longer period of monitoring to guarantee the effectiveness of the treatment.
+
+We notice that, for small and large tumors, the asymptotic expressions for the mean recurrence time are independent of the chromosome missagregation rate $u$ and, as a result, the rate at which the chemotherapeutic drug induces aneuploidy has no effect on the time necessary for the tumor to rebound to its initial size $N$ when $N \ll N_a^*$ or $N \gg N_m^*$.
 % Remus the next paragraphs look like you wanted to find approximations so you did, you reference them here, but the results are buried in the technical text: WHAT IS IT THAT YOU WANT US TO LEARN from this?
-In Appendix F we derive the probability that mutant cancer cell population reaches $N$ at time $t$.
+ok
+
+In Appendix F we derive the probability that mutant cancer cell population has not reached size $N$ at time $t$.
 \Cref{cdffig}B shows agreement between our approximations and stochastic simulations for various values of $N$.
@@
 We derive the mean recurrence time to detection size $M=10^7$ in appendix \ref{sec:appendix_recurrence_time} and plot in \Cref{RecurrencePlot} with simulations.
-We observe that  for small and intermediate sized tumors the effect of the detection size is negligible when compared to $\tau_a^r$.
+We observe that  for small and intermediate sized tumors the effect of the detection size $M$ on $\tau_a^{r,M}$ is negligible (i.e.
+$\tau_a^r\approx\tau_a^{r,M}$ for $N<N_m^*$).
 % WHAT DOES THAT MEAN?
+ok
 However, for large tumors the mean recurrence time to detection size $M$ decreases logarithmically with tumor size $N$.
 
-Most clinical trials report data on the distribution of recurrence time measured from a varying time of surgery~\citep{avanzini2019cancer} with chemotherapy usually following after.
+Most clinical trials report data on the distribution of recurrence time measured from the time of surgery~\citep{avanzini2019cancer} with chemotherapy usually following after.
 % WHAT DOES "varying" mean here?
+ok
 Therefore, since only undetected secondary tumors are present at the time of the administration of the anti-cancer drug we lack knowledge of the size of the tumors and we cannot compare the empirical distributions to our predictions.
@@
     $v$ & Mutation rate& $10^{-9}-10^{-7}$ &  1$\slash$cell division  & \citet{bozic2013evolutionary,loeb2001mutator} \\  \hline
-    $\tilde{u}$ & Missegregation rate in the drug free environment$^\ast$& $10^{-3}$ & 1$\slash$cell division  & - \\ \hline
-    $s$ & Selection coefficient of aneuploidy in the drug free environment& $7.3 \times 10^{-3}$ &  1/days   & \citet{lukow2021chromosomal} \\  
+    $\tilde{u}$ & Missegregation rate in the drug free environment$^\ast$& $2\times10^{-3}$ & 1$\slash$cell division  & - \\ \hline
+    $s$ & Selection coefficient of aneuploidy in the drug free environment& $0.07$ &  1/days   & \citet{lukow2021chromosomal} \\  
     \hline
@@
 \end{align}
-which we plot in Figure and compare with simulations.
+which we plot in \Cref{KMdistribution} and compare with simulations.
 We note that in the case $N\gg\abs{\Delta_w}\abs{\Delta_a}/uv\lambda_w\lambda_ap_m$ the distribution becomes the Dirac $\delta$-function~\citep{barton1989elements}.
@@
 \end{figure}
-\end{appendices}
-
 %%%
@@
 The simulations converge to \cref{eq:t2det2} (green line) for large values of $N\gg N_m^*$.
-Parameters: $\lambda_w=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}
+Parameters: $\lambda_w=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}, M=10^7$.}
 \label{RecurrencePlot}
@@
 Solid lines represents the approximations \cref{eq:AsymptoticTimeRules} ($N<N_a^*$ blue line and $N>N_m^*$ green line).
-Parameters: $\lambda_w=0.1,\lambda_m=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7},f=1.36\%$.}
+Parameters: $\lambda_w=0.1,\lambda_m=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7},f=0.284\%$.}
 \label{SGVEvolutionaryRescueTimeComplete}
 \end{figure}
+\end{appendices}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2024-04-03 EvolutionaryRescue1.tex
Add files via upload

@@
 Given that aneuploidy is present in many tumors even before the onset of therapy \citep{lukow2021chromosomal} we also take into consideration the effect that standing genetic variation has on the tumor dynamics.
+Additionally, we are interested in the timescale of evolutionary rescue and the effect that aneuploidy has on the time necessary for the tumor to overcome drug therapy.
 
@@
 
-We can categorize tumors by their size: small tumors with size $N<N_a^*$ that are unlikely to survive treatment, intermediate tumors with size $N_a^* < N < N_m^*$ that rely on aneuploidy for evolutionary rescue, and large tumors with size $N > N_m^*$ that could be evolutionarily rescued even without aneuploidy.
+We can categorize tumors by their size: small tumors with size $N<N_a^*$ that are unlikely to survive treatment, intermediate tumors with size $N_a^* < N < N_m^*$ that rely on aneuploidy for evolutionary rescue, and large tumors with size $N > N_m^*$ that could overcome the effect of drug treatment even without aneuploidy.
 For the parameter values in \Cref{table1} with $\lambda_a=0.0899,\mu_w=0.14, u=10^{-2}, v=10^{-7}$, we are in the tolerant aneuploid case, and substituting in \cref{eq:N_a,eq:N_m}, we have $N_a^* \approx 4 \times 10^6$ and $N_m^* \approx 4 \times 10^7$.
@@
 Furthermore, the aneuploidy rate in the presence of drugs is likely higher than in their absence, $\tilde{u} < u$ \citep{wang2019molecular,mason2017functional}.
-But if the number of cells in the tumor $N$ is large (as expected if the tumor is treated with a drug), there may already be a fraction $f \approx \tilde{u}\lambda_w/s$ of aneuploid cells in the population (here we assume that the drug affects the wildtype death rate but no the division rate and therefre use $\lambda_w$ for the wildtype division rate in the drug-free enviornment).
+But if the number of cells in the tumor $N$ is large (as expected if the tumor is treated with a drug), there may already be a fraction $f \approx \tilde{u}\lambda_w/s$ of aneuploid cells in the population (here we assume that the drug affects the wildtype death rate but no the division rate and therefore use $\lambda_w$ for the wildtype division rate in the drug-free environment).
 
@@
 Comparing this approximation of $\tilde{N}^*_{a}/N^*_{a}$ to results of stochastic simulations, we find that the approximations perform very well (\Cref{rescue_denovo}). 
-Standing genetic variation will drive evolutionary rescue to the drug if wildtype growth rate $\Delta_w$ is very negative due to a strong effect of the drug on sensitive cells, or if the cost aneuploidy cost in the drug-free environment, $s$, is small.
+Standing genetic variation will drive evolutionary rescue if wildtype growth rate $\Delta_w$ is very negative due to a strong effect of the drug on sensitive cells, or if the aneuploidy cost in the drug-free environment, $s$, is small.
 In contrast, de-novo aneuploid cells will have a greater contribution to rescue if the aneuploidy cost $s$ is large, the effect of the drug on sensitive cells is weak ($\Delta_w$ is close to zero), or if the drug induces the appearance of aneuploid cells ($u > \tilde u$).
@@
 
-Using \cref{eq:N_a,eq:N_m,eq:Nsgv}, we can find the ratio of threshold tumor size for rescue via standing genetic variation  to the threshold for direct mutation,
+Using \cref{eq:N_a,eq:N_m,eq:Nsgv}, we can find the ratio of threshold tumor size for rescue via standing genetic variation  to the threshold for rescue via direct mutation,
 \begin{equation} \label{eq:N_sgv_ratio}
@@
 \paragraph{Evolutionary rescue time.}
-We have derived approximations for $\tau_m$, the mean evolutionary rescue time without aneuploidy ($u=0$), and $\tau_a$, the mean rescue time with aneuploidy ($u>0$), both conditioned on evolutionary rescue occurring.
+In Appendix \ref{sec:appendix_rescue_time} we have derived approximations for $\tau_m$, the mean evolutionary rescue time without aneuploidy ($u=0$), and $\tau_a$, the mean rescue time with aneuploidy ($u>0$), both conditioned on evolutionary rescue occurring.
 These approximations are in good agreement with simulation results for small, intermediate, and large tumor sizes (\Cref{EvolutionaryRescueTimeComplete,MeanTimeGrowthAneuploidyPlot}).
@@
 
-When a fraction $f\approx 0.284\%$ of the initial cancer cell population is aneuploid we want to know whether the mean evolutionary rescue time is affected by the standing genetic variation.
-For this purpose we calculate  $\tau_a$ (see \cref{meantimet2SGV}) and compare our result with simulations in  \Cref{SGVEvolutionaryRescueTimeComplete}.
+Given that a fraction $f\approx 0.284\%$ of the initial cancer cell population is expected to be aneuploid even before the drug administration we want to know whether the mean evolutionary rescue time is affected by the standing genetic variation.
+For this purpose we calculate the mean evolutionary rescue time with standing genetic variation $\tau_a^f$ (see \cref{meantimet2SGV}) and compare our result with simulations in  \Cref{SGVEvolutionaryRescueTimeComplete}.
 We note that standing genetic variation does not have a significant effect on the mean evolutionary rescue time.
@@
 
-In Appendix E we derive the probability that by time $t$ a successful mutant has been generated.
-\Cref{cdffig}A show the agreement between our formula and simulation results for the case when aneuploidy is present and when it is absent.
+In Appendix \ref{sec:appendix_distribution_time} we calculate the probability that a successful mutation, which will rescue the population, has been generated by time $t$.
+This allows us to observe whether aneuploidy accelerates or delays adaptation.
+We plot our results in  \Cref{cdffig}A alongside simulations for different aneuploid growth rates ($u>0$) and the case when aneuploidy is absent ($u=0$).
 We observe that aneuploidy starts to have an effect on adaptation for timescales greater then $1/\Delta_a\approx100$ days after which no more rescue mutations are generated through direct mutation.
@@
 \end{equation}
-\Cref{proliferationFigure,ProliferationTimeLarge} shows the agreement between our approximations and simulations.
-For small tumors ($N \ll N_a^*$), the mean recurrence time can be approximated as the sum of the mean time for the first rescue mutation to appear and the mean time for the lineage generated by that mutation to reach size $N$.
+\Cref{proliferationFigure,ProliferationTimeLarge} show the agreement between our approximations and simulations.
+For small tumors ($N \ll N_a^*$), the mean recurrence time can be approximated as the sum of the mean time for the first rescue mutation to appear and the mean time for its lineage to reach size $N$.
 The mean recurrence time grows logarithmically with tumor size $N$ and is the same order of magnitude as the mean evolutionary rescue time.
-Increasing the mutant growth rate $\Delta_m$ leads to shorter recurrence times while increasing the wildtype and aneuploid growth rates, $\Delta_w$ and $\Delta_a$ respectively, increases the recurrence time.
+Increasing the mutant growth rate $\Delta_m$ decreases recurrence times while increasing the wildtype and aneuploid growth rates, $\Delta_w$ and $\Delta_a$ respectively, increases the recurrence time.
 For large tumors ($N \gg N_m^*$), the dynamics of the number of mutant cells is deterministic and the mean recurrence time becomes constant (and independent of the initial tumor size $N$).
 Increasing either the mutant growth rate $\Delta_m$ or the mutation rate $v$ leads to a decrease in the time for the tumor to rebound to its initial size.
-In addition, drugs that significantly increase the wildtype death rate $\mu_w$ and do not affect the division rate $\lambda_w$ delay cancer relapse.
-Consequently, patients treated with such drugs require a longer period of monitoring to guarantee the effectiveness of the treatment.
-
-We notice that, for small and large tumors, the asymptotic expressions for the mean recurrence time are independent of the chromosome missagregation rate $u$ and, as a result, the rate at which the chemotherapeutic drug induces aneuploidy has no effect on the time necessary for the tumor to rebound to its initial size $N$ when $N \ll N_a^*$ or $N \gg N_m^*$.
+In addition, drugs that significantly increase the wildtype death rate $\mu_w$ and do not affect the division rate $\lambda_w$ delay cancer recurrence.
+Consequently, patients treated with such drugs may require a longer period of monitoring to guarantee the effectiveness of the treatment.
+
+We note that, for small and large tumors, when $N \ll N_a^*$ or $N \gg N_m^*$, the asymptotic expressions for the mean recurrence time are independent of the chromosome missagregation rate $u$, therefore the rate at which the drug induces aneuploidy has no effect on the time necessary for the tumor to rebound to its initial size $N$.
 % Remus the next paragraphs look like you wanted to find approximations so you did, you reference them here, but the results are buried in the technical text: WHAT IS IT THAT YOU WANT US TO LEARN from this?
@@
 
-In Appendix F we derive the probability that mutant cancer cell population has not reached size $N$ at time $t$.
+In Appendix F we derive the probability that a mutant cancer cell population has not reached size $N$ by time $t$.
 \Cref{cdffig}B shows agreement between our approximations and stochastic simulations for various values of $N$.
-Additionally, we derived the distribution of the recurrence time and plotted our result in \Cref{KMdistribution} alongside simulations for the case $N=10^6$ (small tumor) and we note that the distribution is wide and rights-skewed.
-It is highly unlikely to observe the recurrence of tumors at times smaller then $\frac{1}{\Delta_m}\log\frac{\Delta_m-\Delta_w}{v\lambda_w}\approx1542$ days for the parameter values $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$ and independent of initial tumor size $N$ (\Cref{cdffig}B).
-
-The recurrence can also be defined as the time necessary for the tumor size to reach a detection threshold $M$.
-We derive the mean recurrence time to detection size $M=10^7$ in appendix \ref{sec:appendix_recurrence_time} and plot in \Cref{RecurrencePlot} with simulations.
-We observe that  for small and intermediate sized tumors the effect of the detection size $M$ on $\tau_a^{r,M}$ is negligible (i.e.
+Additionally, we derive the distribution of the recurrence time for the case $N=10^6$ (small tumor), noting that the distribution is wide and right-skewed (\Cref{KMdistribution}).
+It is highly unlikely to observe the recurrence of tumors at times smaller then $\frac{1}{\Delta_m}\log\frac{\Delta_m-\Delta_w}{v\lambda_w}\approx1542$ days for the parameter values $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$ and independent of initial tumor size $N$~(\Cref{cdffig}B).
+
+The detection time $\tau_a^{r,M}$ can be defined as the time necessary for the tumor size to reach detection threshold $M$.
+We derive the mean recurrence time to detection size $M=10^7$ in appendix \ref{sec:appendix_recurrence_time}.
+We observe that  for small and intermediate sized tumors the effect of the detection size $M$ on $\tau_a^{r,M}$ is negligible when compared to the case where the detection size is $N$ (i.e.
 $\tau_a^r\approx\tau_a^{r,M}$ for $N<N_m^*$).
@@
 ok
-However, for large tumors the mean recurrence time to detection size $M$ decreases logarithmically with tumor size $N$.
-
-Most clinical trials report data on the distribution of recurrence time measured from the time of surgery~\citep{avanzini2019cancer} with chemotherapy usually following after.
+However, for large tumors the mean recurrence time to detection size $M$ decreases logarithmically with tumor size $N$ while $\tau_a^{r}$ is constant (\Cref{RecurrencePlot}).
+Additionally, for large tumors we have $M<N_m^*<N$ which tells us that the mean detection time is smaller for such tumors when compared to the mean recurrence time.
+
+Most clinical trials report data on the distribution of recurrence time measured from the time of surgery~\citep{avanzini2019cancer} with drug therapy usually following after.
 % WHAT DOES "varying" mean here?
@@
 Therefore, since only undetected secondary tumors are present at the time of the administration of the anti-cancer drug we lack knowledge of the size of the tumors and we cannot compare the empirical distributions to our predictions.
-However, we expect that the variability of secondary tumors to average out across large cohorts of patients and, as a result, we can use the mean recurrence time.
+However, we expect that the variability of secondary tumors to average out across large cohorts of patients, thus we can use the mean recurrence time to compare with clinical data.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-We have modeled a tumor--a population of cancer cells--exposed to drug treatment that causes the population to decline in size towards potential extinction.
+We have modeled a tumor--a population of cancer cells--exposed to drug treatment that causes it to decline in size towards potential extinction.
 In this scenario, the tumor can be "evolutionary rescued", or escape extinction, via two paths.
@@
 
-Using multitype branching processes, we derived the probability of evolutionary rescue of the population of cancer cells under different scenarios for the effect of aneuploidy, ranging from tolerance to partial resistance.
+\paragraph{Evolutionary rescue}
+Using multitype branching processes, we derived the probability of evolutionary rescue of the tumor under different scenarios for the effect of aneuploidy, ranging from tolerance to partial resistance.
 We obtained exact and approximate expressions for the probability of evolutionary rescue (\cref{eq:rescue_prob}). 
@@
 
-\paragraph{Evolutionary rescue}
-When aneuploid cells are partially resistant to the drug ($\Delta_w\ll0\ll\Delta_a\ll\Delta_m$), evolutionary rescue can be approximated by a one-step process in which aneuploidy itself rescues the population (\Cref{rescue_threshold}A).
+When aneuploid cells are partially resistant to the drug ($\Delta_w\ll0\ll\Delta_a\ll\Delta_m$), aneuploidy itself rescues the population (\Cref{rescue_threshold}A). 
 When aneuploidy only provides tolerance to the drug ($\Delta_w\ll\Delta_a\ll0\ll\Delta_m$), it cannot rescue the population.
-Instead, it acts as a \emph{stepping stone} through which the resistant mutant can appear more rapidly, given that the aneuploid cell population declines slower then the sensitive cell population (\Cref{sampleTrajectories}).
+Instead, it acts as a \emph{stepping stone} through which the resistant mutant can appear more rapidly, given that the aneuploid cell population size declines slower then that of the sensitive cell population (\Cref{sampleTrajectories}).
 In this case, aneuploidy provides two benefits.
-First, it delays the extinction of the population--providing more time for appearance of the resistance mutation.
-Second, it increases the population size relative to a sensitive population--providing more cells in which mutations can occur, i.e., it increases the mutation supply, $Nuv\lambda_w\lambda_a/\abs{\Delta_w\Delta_a}$.
+First, it delays the extinction of the population, providing more time for appearance of the resistance mutation.
+Second, it increases the population size relative to a sensitive population, providing more cells in which mutations can occur, i.e., it increases the mutation supply (i.e.
+$Nuv\lambda_w\lambda_a/\abs{\Delta_w\Delta_a}$).
 
 We find that aneuploidy can have a significant effect on evolutionary rescue as it reduces the threshold tumor size by at least one order of magnitude even when aneuploidy is tolerant (\Cref{rescue_prob}).
-Interestingly, aneuploidy is unlikely to contribute to evolutionary rescue in primary tumors in which the number of cells is large enough ($N\gg N_m^*\approx 4\times10^7$) for the appearance of resistant mutation directly in sensitive cells before these cells become extinct (\Cref{rescue_prob}).
-However, aneuploidy can have a crucial role in evolutionary rescue of secondary tumors, in which the number of sensitive cells may be below the detection threshold of $\sim10^7$  \citep{bozic2013evolutionary}.
-This can have an impact on the recurrence of cancer after the resection of the primary tumor  through secondary tumors which are too small to be detected and for which chemotherapy is employed to prevent  cancer relapse and are estimated to cause the majority of cancer-related deaths~\citep{chaffer2011perspective}.
-This is reinforced by the fact that metastases have been shown to have higher chromosome missagregation rate two to three orders of magnitude higher then primary tumors~\citep{kimmel2023intra}.
+Interestingly, aneuploidy is unlikely to contribute to evolutionary rescue in primary tumors in which the number of cells is large enough (i.e.
+$N\gg N_m^*\approx 4\times10^7$) for the appearance of resistant mutation directly in sensitive cells before these cells become extinct (\Cref{rescue_prob}).
+However, aneuploidy can have a crucial role in evolutionary rescue of secondary tumors, in which the number of sensitive cells may be below the detection threshold of $\sim10^7$  \citep{bozic2013evolutionary}, and this can have an impact on the recurrence of cancer after the resection of the primary tumor  through secondary tumors which are too small to be detected and for which chemotherapy is employed to prevent  cancer relapse and are estimated to cause the majority of cancer-related deaths~\citep{chaffer2011perspective}.
+The importance of aneuploidy in the evolutionary rescue of secondary tumors is reinforced by the fact that metastases have been shown to have a chromosome missagregation rate two to three orders of magnitude higher then that of primary tumors~\citep{kimmel2023intra}.
 
 Given the fact that the mean time for such secondary tumors to overcome chemotherapy can be of the order of 1000 days (\Cref{MeanTimeGrowthAneuploidyPlot}A),
-this can explain the reappearance of cancer even after initial remission.
-The theoretical predictions for the mean rescue time for tumors smaller then  $10^8$ is greater then 4 years which is consistent with previous estimates of the recurrence time of tumors after resection~\citep{avanzini2019cancer}.
-
-We hypothesized that presence of \emph{standing variation} 
---the existence of a subpopulation of aneuploid cancer cells before therapy begins--can facilitate evolutionary rescue by reducing the waiting time for the appearance of aneuploid cells.
-However, we observe that for reasonable parameter values evolutionary rescue is more likely to occur through de-novo aneuploidy (\Cref{rescue_denovo}).
-If the the fraction of the tumor cells which have the ``beneficial'' aneuploidy is $f\gg\frac{u\lambda_w}{\abs{\Delta_w}}\approx2.5\%$ then evolutionary rescue is more likely to occur through this existing standing variation, rather then through \emph{de novo} aneuploid cells.
-In that case, evolutionary rescue occurs through a single step by the aneuploid cells acquiring a beneficial mutation and the probability of evolutionary rescue is given by $1-\exp\left(-N/\tilde{N}_a^*\right)$  (\Cref{rescue_prob_sgv}).
+aneuploidy can explain the reappearance of cancer even after initial remission.
+The theoretical prediction for the mean rescue time for tumors smaller then  $10^8$ cells is greater then 4 years, consistent with previous estimates of the recurrence time of tumors after resection~\citep{avanzini2019cancer}.
+We observe from \Cref{cdffig}A that aneuploidy complements evolutionary rescue through direct mutation given that aneuploidy is most likely to generate rescue mutations at time scales larger then the timescales at which rescue mutations are generated through direct mutation.
+
+We hypothesized that \emph{standing variation} (the existence of a subpopulation of aneuploid cancer cells before the onset of therapy) can facilitate evolutionary rescue by reducing the waiting time for the appearance of aneuploid cells.
+From \Cref{eq:Nsgv} we observe that a drug or combination of drugs that reduces the wildtype growth rate and does not significantly increase the chromosome missegregation rate is more likely to cause evolutionary rescue to occur through direct mutation.
+Furthermore, we find that for reasonable parameter values evolutionary rescue is more likely to occur through \emph{de-novo} aneuploidy (\Cref{rescue_denovo}).
+If the fraction of tumor cells that have the ``advantageous'' aneuploidy is $f\gg\frac{u\lambda_w}{\abs{\Delta_w}}\approx2.5\%$, then evolutionary rescue is more likely to occur via standing variation, rather then through \emph{de novo} aneuploid cells.
+In this case, evolutionary rescue likely occurs via aneuploid cells that acquire a resistance mutation and the probability of evolutionary rescue declines exponentially with initial tumor size  (the probability of evolutionary rescue is given by $1-\exp\left(-N/\tilde{N}_a^*\right)$, see \Cref{rescue_prob_sgv}).
 
 \paragraph{Experimental future direction}
-Our model's predictions may be tested by experiments \citep{martin2013probability}.
-For example, to study the effects of initial tumor size on the probability of evolutionary rescue, a large culture mass can be propagated from a single cancer cell in permissive conditions and then diluted to a  range of starting tumor sizes.
-Afterwards, these tumors may be exposed to anti-cancer drugs that induces aneuploidy or to saline solution for control~\citep{ippolito2021gene}.
-Cell density can then be measured by optical density and a population exposed to the drug is considered extinct if the optical density is lower when compared to the control case with no cells present.
-We compare to the predictions of our model to see if the tumors with initial size bellow the threshold \eqref{eq:N_a} are more likely to go extinct.
-
-Additionally, we can test our model's predictions with data from patients.
-Given the type of cancer they have and the degree of ``beneficial'' aneuploidy,  we can obtain the parameters for our model and predict with what probability will the cancer relapse after treatment.
+Our model prediction could be tested by experiments \citep{martin2013probability}.
+For example, to assess the effect of initial tumor size on the probability of evolutionary rescue, a large culture mass can be propagated from a single cancer cell in permissive conditions and then diluted to a  range of starting tumor sizes.
+Then, these tumors may be exposed to anti-cancer drugs that induce aneuploidy or to saline solution for control~\citep{ippolito2021gene}.
+Cell density can be measured by optical density and a population exposed to the drug is considered extinct if the optical density is lower when compared to the control case with no cells present.
+We can then compare the results of the experiments to the predictions of our model to see if the tumors with initial size bellow the threshold \eqref{eq:N_a} are more likely to go killed by the drug.
+
+Additionally, our model predictions can be tested with data from patients.
+Given the type of cancer and how ``advantageous'' the aneuploidy is,  our model can predict the probability for cancer relapse after treatment.
 
 \paragraph{Directions of future research}
-Even though our model has been designed for cancer it can be extended to understand evolutionary rescue in different biological contexts, for example, how yeast subject to stress can overcome extinction with the help of aneuploidy~\citep{pompei2023fitness}.
-Additionally, the heterogeneity of aneuploidy is a fact the we did not account for, as not all the aneuploidy lineages generated have the same growth rate $\Delta_a$.
-This can be corrected by  fixing the aneuploidy death rates and sampling the aneuploidy growth rates from a distribution $f\left(\Delta_a|\mu_a\right)$~\citet{martin2013probability}.
-Furthermore, tumor heterogeneity is also a factor for pre-existing genetic variation where the fitness of the aneuploid cells can be drawn from a distribution upon which selection acts when the tumor is exposed to a strong stress.
-
-We have assumed that cancer cell lineages are independent of each other,
+ Our model can be extended to understand evolutionary rescue in different biological contexts, for example, how yeast subject to stress can overcome extinction via aneuploidy~\citep{pompei2023fitness,kohanovski2024aneuploidy}.
+Additionally, we did not account for the heterogeneity of aneuploidy as not all the aneuploidy lineages generated have the same growth rate $\Delta_a$ as we have assumed in our model~\citep{avecilla2023copy,yang2021fitness}.
+Such heterogeneity can be accounted for  by  sampling the aneuploidy growth or death rates from a distribution~\citet{martin2013probability}.
+Furthermore, tumor heterogeneity is also important for pre-existing genetic variation where the fitness of the aneuploid cells can be drawn from a distribution and the fittest genotype is selected when the tumor is exposed to the drug treatment.
+
+We have assumed that cancer cell lineages are independent,
 and have verified that this approximation is accurate under simple logistic growth. 
-However, this neglects potential effects of spatial structure and local interactions, which may be very important in solid tumors.
+However, this neglects potential effects of spatial structure and local interactions, which may be important in solid tumors.
 Such tumors can be spatially heterogeneous with different genotypes inhabiting cellular niches and immune infiltration impacting growth in affected regions \citep{varrone2023cellcharter,galon2010immune}.
@@
 An additional limitation of our model is the choice of parameters which are based on a specific set of assumptions which might not be true for many tumor.
-As a result, incorporating a wider distribution of parameters would be a beneficial extension for our model.
+As a result, incorporating a wider distribution of parameters would be a beneficial extension for our model, however parameter values for different types of tumors are difficult to obtain.
 
@@
 
-Our paper analyzed the role that aneuploidy plays in tumor adaptation to the effects of anti-cancer drugs.
-Our study quantitatively confirms that aneuploidy plays an important role in tumors overcoming exposure to chemotherapeutic drugs when the tumor size is small or intermediate.
-Very large tumors can escape anti-cancer drugs through direct mutation while smaller ones are able to obtain the beneficial mutation through an aneuploid intermediary (\Cref{rescue_prob}).
-As a result, therapies which increase the rate of aneuploidy in tumors in order to combat cancer might have an adverse effect on patient outcomes.
+Our results quantitatively show that aneuploidy plays an important role in tumors overcoming exposure to chemotherapeutic drugs when tumor size is small or intermediate.
+Large tumors can escape anti-cancer drugs through direct mutation while smaller ones are able to obtain the beneficial mutation through an aneuploid ``\emph{stepping stone}'' (\Cref{rescue_prob}).
+As a result, therapies that increase the rate of aneuploidy in tumors in order to combat cancer may have an adverse effect on patient outcomes.
 % TODO summary ok
@@
 \begin{align}\label{meantimet2SGV}
-\tau_a &= \frac{\int_{0}^{\infty}{tR_a^f(t)\exp\left(-\int_0^{t}{R_a^f(z)\d z}\right) \d t}}{p_{rescue}},
+\tau_a^f&= \frac{\int_{0}^{\infty}{tR_a^f(t)\exp\left(-\int_0^{t}{R_a^f(z)\d z}\right) \d t}}{p_{rescue}},
 \end{align}
@@
 
-\section{Distribution of evolutionary rescue time}
+\section{Distribution of evolutionary rescue time}\label{sec:appendix_distribution_time}
 The probability that a successful mutant has been generated by time $t$ is given by:

### 2024-04-06 EvolutionaryRescue1.tex
Add files via upload

@@
 Given that aneuploidy is present in many tumors even before the onset of therapy \citep{lukow2021chromosomal} we also take into consideration the effect that standing genetic variation has on the tumor dynamics.
-Additionally, we are interested in the timescale of evolutionary rescue and the effect that aneuploidy has on the time necessary for the tumor to overcome drug therapy.
+\textcolor{blue}{Additionally, we are interested in the timescale of evolutionary rescue and the effect that aneuploidy has on the time necessary for the tumor to overcome drug therapy.}
 
@@
 
-We can categorize tumors by their size: small tumors with size $N<N_a^*$ that are unlikely to survive treatment, intermediate tumors with size $N_a^* < N < N_m^*$ that rely on aneuploidy for evolutionary rescue, and large tumors with size $N > N_m^*$ that could overcome the effect of drug treatment even without aneuploidy.
+We can categorize tumors by their size: small tumors with size $N<N_a^*$ that are unlikely to survive treatment, intermediate tumors with size $N_a^* < N < N_m^*$ that rely on aneuploidy for evolutionary rescue, and large tumors with size $N > N_m^*$ that could \textcolor{blue}{overcome the effect of drug treatment} even without aneuploidy.
 For the parameter values in \Cref{table1} with $\lambda_a=0.0899,\mu_w=0.14, u=10^{-2}, v=10^{-7}$, we are in the tolerant aneuploid case, and substituting in \cref{eq:N_a,eq:N_m}, we have $N_a^* \approx 4 \times 10^6$ and $N_m^* \approx 4 \times 10^7$.
@@
 
-Given that a fraction $f\approx 0.284\%$ of the initial cancer cell population is expected to be aneuploid even before the drug administration we want to know whether the mean evolutionary rescue time is affected by the standing genetic variation.
+\textcolor{blue}{Given that a fraction $f\approx 0.284\%$ of the initial cancer cell population is expected to be aneuploid even before the drug administration we want to know whether the mean evolutionary rescue time is affected by the standing genetic variation.
 For this purpose we calculate the mean evolutionary rescue time with standing genetic variation $\tau_a^f$ (see \cref{meantimet2SGV}) and compare our result with simulations in  \Cref{SGVEvolutionaryRescueTimeComplete}.
-We note that standing genetic variation does not have a significant effect on the mean evolutionary rescue time.
-% WHAT is the purpose of this sentence?
+We note that standing genetic variation does not have a significant effect on the mean evolutionary rescue time.} % WHAT is the purpose of this sentence?
 ok
 
-In Appendix \ref{sec:appendix_distribution_time} we calculate the probability that a successful mutation, which will rescue the population, has been generated by time $t$.
+\textcolor{blue}{In Appendix \ref{sec:appendix_distribution_time} we calculate the probability that a successful mutation, which will rescue the population, has been generated by time $t$.
 This allows us to observe whether aneuploidy accelerates or delays adaptation.
-We plot our results in  \Cref{cdffig}A alongside simulations for different aneuploid growth rates ($u>0$) and the case when aneuploidy is absent ($u=0$).
-We observe that aneuploidy starts to have an effect on adaptation for timescales greater then $1/\Delta_a\approx100$ days after which no more rescue mutations are generated through direct mutation.
+We plot our results in  \Cref{cdffig}A alongside simulations for different aneuploid growth rates ($u>0$) and the case when aneuploidy is absent ($u=0$).} We observe that aneuploidy starts to have an effect on adaptation for timescales greater then $1/\Delta_a\approx100$ days after which no more rescue mutations are generated through direct mutation.
 This shows that rescue mutations are generated through aneuploidy at latter timescales then direct mutation and, as a consequence, aneuploidy increases the \emph{window of opportunity} for evolutionary rescue.
@@
 \Cref{proliferationFigure,ProliferationTimeLarge} show the agreement between our approximations and simulations.
-For small tumors ($N \ll N_a^*$), the mean recurrence time can be approximated as the sum of the mean time for the first rescue mutation to appear and the mean time for its lineage to reach size $N$.
+For small tumors ($N \ll N_a^*$), the mean recurrence time can be approximated as the sum of the mean time for the first rescue mutation to appear and the mean time for \textcolor{blue}{its lineage to reach size $N$}.
 The mean recurrence time grows logarithmically with tumor size $N$ and is the same order of magnitude as the mean evolutionary rescue time.
-Increasing the mutant growth rate $\Delta_m$ decreases recurrence times while increasing the wildtype and aneuploid growth rates, $\Delta_w$ and $\Delta_a$ respectively, increases the recurrence time.
+Increasing the mutant growth rate $\Delta_m$ \textcolor{blue}{decreases} recurrence times while increasing the wildtype and aneuploid growth rates, $\Delta_w$ and $\Delta_a$ respectively, increases the recurrence time.
 For large tumors ($N \gg N_m^*$), the dynamics of the number of mutant cells is deterministic and the mean recurrence time becomes constant (and independent of the initial tumor size $N$).
@@
 
-We note that, for small and large tumors, when $N \ll N_a^*$ or $N \gg N_m^*$, the asymptotic expressions for the mean recurrence time are independent of the chromosome missagregation rate $u$, therefore the rate at which the drug induces aneuploidy has no effect on the time necessary for the tumor to rebound to its initial size $N$.
+We \textcolor{blue}{note} that, for small and large tumors, when $N \ll N_a^*$ or $N \gg N_m^*$, the asymptotic expressions for the mean recurrence time are independent of the chromosome missagregation rate $u$, \textcolor{blue}{therefore} the rate at which the drug induces aneuploidy has no effect on the time necessary for the tumor to rebound to its initial size $N$.
 % Remus the next paragraphs look like you wanted to find approximations so you did, you reference them here, but the results are buried in the technical text: WHAT IS IT THAT YOU WANT US TO LEARN from this?
@@
 
-In Appendix F we derive the probability that a mutant cancer cell population has not reached size $N$ by time $t$.
+In Appendix F we derive the probability that a mutant cancer cell population has not reached size $N$ \textcolor{blue}{by} time $t$.
 \Cref{cdffig}B shows agreement between our approximations and stochastic simulations for various values of $N$.
@@
 
-The detection time $\tau_a^{r,M}$ can be defined as the time necessary for the tumor size to reach detection threshold $M$.
-We derive the mean recurrence time to detection size $M=10^7$ in appendix \ref{sec:appendix_recurrence_time}.
+\textcolor{blue}{The detection time $\tau_a^{r,M}$ can be defined as the time necessary for the tumor size to reach detection threshold $M$.
+We derive the mean recurrence time to detection size $M=10^7$ in Appendix \ref{sec:appendix_recurrence_time}.
 We observe that  for small and intermediate sized tumors the effect of the detection size $M$ on $\tau_a^{r,M}$ is negligible when compared to the case where the detection size is $N$ (i.e.
@@
 However, for large tumors the mean recurrence time to detection size $M$ decreases logarithmically with tumor size $N$ while $\tau_a^{r}$ is constant (\Cref{RecurrencePlot}).
-Additionally, for large tumors we have $M<N_m^*<N$ which tells us that the mean detection time is smaller for such tumors when compared to the mean recurrence time.
+Additionally, for large tumors we have $M<N_m^*<N$ which tells us that the mean detection time is smaller for such tumors when compared to the mean recurrence time.}
 
@@
 Therefore, since only undetected secondary tumors are present at the time of the administration of the anti-cancer drug we lack knowledge of the size of the tumors and we cannot compare the empirical distributions to our predictions.
-However, we expect that the variability of secondary tumors to average out across large cohorts of patients, thus we can use the mean recurrence time to compare with clinical data.
+\textcolor{blue}{However, we expect that the variability of secondary tumors to average out across large cohorts of patients, thus we can use the mean recurrence time to compare with clinical data.}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \paragraph{Evolutionary rescue}
-Using multitype branching processes, we derived the probability of evolutionary rescue of the tumor under different scenarios for the effect of aneuploidy, ranging from tolerance to partial resistance.
+\textcolor{blue}{Using multitype branching processes, we derived the probability of evolutionary rescue of the tumor under different scenarios for the effect of aneuploidy, ranging from tolerance to partial resistance.
 We obtained exact and approximate expressions for the probability of evolutionary rescue (\cref{eq:rescue_prob}). 
-Our results show that the probability of evolutionary rescue increases with the initial tumor size $N$, the sensitive growth rate $\Delta_w$, the mutation rate $v$, and the aneuploidy rate $u$.
+Our results show that the probability of evolutionary rescue increases with the initial tumor size $N$, the sensitive growth rate $\Delta_w$, the mutation rate $v$, and the aneuploidy rate $u$.}
 
@@
 When aneuploidy only provides tolerance to the drug ($\Delta_w\ll\Delta_a\ll0\ll\Delta_m$), it cannot rescue the population.
-Instead, it acts as a \emph{stepping stone} through which the resistant mutant can appear more rapidly, given that the aneuploid cell population size declines slower then that of the sensitive cell population (\Cref{sampleTrajectories}).
+Instead, it acts as a \emph{stepping stone} through which the resistant mutant can appear more rapidly, given that the aneuploid cell population \textcolor{blue}{size} declines slower then \textcolor{blue}{that} of the sensitive cell population (\Cref{sampleTrajectories}).
 In this case, aneuploidy provides two benefits.
@@
 $N\gg N_m^*\approx 4\times10^7$) for the appearance of resistant mutation directly in sensitive cells before these cells become extinct (\Cref{rescue_prob}).
-However, aneuploidy can have a crucial role in evolutionary rescue of secondary tumors, in which the number of sensitive cells may be below the detection threshold of $\sim10^7$  \citep{bozic2013evolutionary}, and this can have an impact on the recurrence of cancer after the resection of the primary tumor  through secondary tumors which are too small to be detected and for which chemotherapy is employed to prevent  cancer relapse and are estimated to cause the majority of cancer-related deaths~\citep{chaffer2011perspective}.
-The importance of aneuploidy in the evolutionary rescue of secondary tumors is reinforced by the fact that metastases have been shown to have a chromosome missagregation rate two to three orders of magnitude higher then that of primary tumors~\citep{kimmel2023intra}.
+However, aneuploidy can have a crucial role in evolutionary rescue of secondary tumors, in which the number of sensitive cells may be below the detection threshold of $\sim10^7$  \citep{bozic2013evolutionary}, \textcolor{blue}{and this} can have an impact on the recurrence of cancer after the resection of the primary tumor  through secondary tumors which are too small to be detected and for which chemotherapy is employed to prevent  cancer relapse and are estimated to cause the majority of cancer-related deaths~\citep{chaffer2011perspective}.
+\textcolor{blue}{The importance of aneuploidy in the evolutionary rescue of secondary tumors} is reinforced by the fact that metastases have been shown to have a chromosome missagregation rate two to three orders of magnitude higher then \textcolor{blue}{that of} primary tumors~\citep{kimmel2023intra}.
 
@@
 The theoretical prediction for the mean rescue time for tumors smaller then  $10^8$ cells is greater then 4 years, consistent with previous estimates of the recurrence time of tumors after resection~\citep{avanzini2019cancer}.
-We observe from \Cref{cdffig}A that aneuploidy complements evolutionary rescue through direct mutation given that aneuploidy is most likely to generate rescue mutations at time scales larger then the timescales at which rescue mutations are generated through direct mutation.
-
-We hypothesized that \emph{standing variation} (the existence of a subpopulation of aneuploid cancer cells before the onset of therapy) can facilitate evolutionary rescue by reducing the waiting time for the appearance of aneuploid cells.
-From \Cref{eq:Nsgv} we observe that a drug or combination of drugs that reduces the wildtype growth rate and does not significantly increase the chromosome missegregation rate is more likely to cause evolutionary rescue to occur through direct mutation.
-Furthermore, we find that for reasonable parameter values evolutionary rescue is more likely to occur through \emph{de-novo} aneuploidy (\Cref{rescue_denovo}).
-If the fraction of tumor cells that have the ``advantageous'' aneuploidy is $f\gg\frac{u\lambda_w}{\abs{\Delta_w}}\approx2.5\%$, then evolutionary rescue is more likely to occur via standing variation, rather then through \emph{de novo} aneuploid cells.
-In this case, evolutionary rescue likely occurs via aneuploid cells that acquire a resistance mutation and the probability of evolutionary rescue declines exponentially with initial tumor size  (the probability of evolutionary rescue is given by $1-\exp\left(-N/\tilde{N}_a^*\right)$, see \Cref{rescue_prob_sgv}).
+\textcolor{blue}{We observe from \Cref{cdffig}A that aneuploidy complements evolutionary rescue through direct mutation given that aneuploidy is most likely to generate rescue mutations at time scales larger then the timescales at which rescue mutations are generated through direct mutation.}
+
+We hypothesized that \emph{standing variation} (the existence of a subpopulation of aneuploid cancer cells before the \textcolor{blue}{onset of} therapy) can facilitate evolutionary rescue by reducing the waiting time for the appearance of aneuploid cells.
+\textcolor{blue}{From \Cref{eq:Nsgv} we observe that a drug or combination of drugs that reduces the wildtype growth rate and does not significantly increase the chromosome missegregation rate is more likely to cause evolutionary rescue to occur through direct mutation.} Furthermore, we find that for reasonable parameter values evolutionary rescue is more likely to occur through \emph{de-novo} aneuploidy (\Cref{rescue_denovo}).
+If the fraction of tumor cells \textcolor{blue}{that} have the \textcolor{blue}{``advantageous''} aneuploidy is $f\gg\frac{u\lambda_w}{\abs{\Delta_w}}\approx2.5\%$, then evolutionary rescue is more likely to occur \textcolor{blue}{via} standing variation, rather then through \emph{de novo} aneuploid cells.
+In \textcolor{blue}{this} case, evolutionary rescue \textcolor{blue}{likely} occurs \textcolor{blue}{via} aneuploid cells that acquire a \textcolor{blue}{resistance} mutation and the \textcolor{blue}{probability of evolutionary rescue declines exponentially with initial tumor size  (the probability of evolutionary rescue is given by $1-\exp\left(-N/\tilde{N}_a^*\right)$, see \Cref{rescue_prob_sgv}).}
 
 \paragraph{Experimental future direction}
-Our model prediction could be tested by experiments \citep{martin2013probability}.
-For example, to assess the effect of initial tumor size on the probability of evolutionary rescue, a large culture mass can be propagated from a single cancer cell in permissive conditions and then diluted to a  range of starting tumor sizes.
-Then, these tumors may be exposed to anti-cancer drugs that induce aneuploidy or to saline solution for control~\citep{ippolito2021gene}.
+Our model prediction \textcolor{blue}{could} be tested by experiments \citep{martin2013probability}.
+For example, to \textcolor{blue}{assess} the effect of initial tumor size on the probability of evolutionary rescue, a large culture mass can be propagated from a single cancer cell in permissive conditions and then diluted to a  range of starting tumor sizes.
+\textcolor{blue}{Then}, these tumors may be exposed to anti-cancer drugs that induce aneuploidy or to saline solution for control~\citep{ippolito2021gene}.
 Cell density can be measured by optical density and a population exposed to the drug is considered extinct if the optical density is lower when compared to the control case with no cells present.
-We can then compare the results of the experiments to the predictions of our model to see if the tumors with initial size bellow the threshold \eqref{eq:N_a} are more likely to go killed by the drug.
-
-Additionally, our model predictions can be tested with data from patients.
-Given the type of cancer and how ``advantageous'' the aneuploidy is,  our model can predict the probability for cancer relapse after treatment.
+\textcolor{blue}{We can then compare the results of the experiments to the predictions of our model to see if the tumors with initial size bellow the threshold \eqref{eq:N_a} are more likely to go killed by the drug.}
+
+Additionally, our model predictions \textcolor{blue}{can be tested} with data from patients.
+Given the type of cancer and how \textcolor{blue}{``advantageous''} the aneuploidy is,  our model can predict \textcolor{blue}{the} probability \textcolor{blue}{for} cancer relapse after treatment.
 
 \paragraph{Directions of future research}
- Our model can be extended to understand evolutionary rescue in different biological contexts, for example, how yeast subject to stress can overcome extinction via aneuploidy~\citep{pompei2023fitness,kohanovski2024aneuploidy}.
-Additionally, we did not account for the heterogeneity of aneuploidy as not all the aneuploidy lineages generated have the same growth rate $\Delta_a$ as we have assumed in our model~\citep{avecilla2023copy,yang2021fitness}.
-Such heterogeneity can be accounted for  by  sampling the aneuploidy growth or death rates from a distribution~\citet{martin2013probability}.
-Furthermore, tumor heterogeneity is also important for pre-existing genetic variation where the fitness of the aneuploid cells can be drawn from a distribution and the fittest genotype is selected when the tumor is exposed to the drug treatment.
+ Our model can be extended to understand evolutionary rescue in different biological contexts, for example, how yeast subject to stress can overcome extinction \textcolor{blue}{via} aneuploidy~\citep{pompei2023fitness,kohanovski2024aneuploidy}.
+Additionally, \textcolor{blue}{we did not account for} the heterogeneity of aneuploidy as not all the aneuploid lineages generated have the same growth rate $\Delta_a$ \textcolor{blue}{as we have assumed in our model~\citep{avecilla2023copy,yang2021fitness}}.
+Such heterogeneity can be accounted for  by  sampling the aneuploidy growth or \textcolor{blue}{death} rates from a distribution~\citet{martin2013probability}.
+\textcolor{blue}{Furthermore, tumor heterogeneity is also important for pre-existing genetic variation where the fitness of the aneuploid cells can be drawn from a distribution and the fittest genotype is selected when the tumor is exposed to the drug treatment.}
 
@@
 
-An additional limitation of our model is the choice of parameters which are based on a specific set of assumptions which might not be true for many tumor.
-As a result, incorporating a wider distribution of parameters would be a beneficial extension for our model, however parameter values for different types of tumors are difficult to obtain.
+An additional limitation of our model is the choice of parameters, which are based on a specific set of assumptions \textcolor{blue}{that may not hold for some tumors.} \textcolor{blue}{Thus}, incorporating a wider distribution of parameters would be a beneficial extension for our model, \textcolor{blue}{however parameter values for different types of tumors are difficult to obtain.}
 

### 2024-05-06 EvolutionaryRescue1.tex
Add files via upload

--- 2024-04-03 EvolutionaryRescue1.tex
@@
 \affil[3]{Department of Physics, Emory University, Atlanta, GA}
-\affil[*]{Corresponding author: yoav@yoavram.com}
+\affil[*]{Corresponding author: Yoav Ram \textcolor{green}{(e-mail: yoavram@tauex.tau.ac.il)}}
 
@@
 Evolutionary rescue is the process by which a population survives a sudden environmental change that initially causes the population to decline towards extinction.
-A prime example of evolutionary rescue is the ability of cancer to survive exposure to various treatments. 
-One evolutionary mechanism by which a population of cancer cells can to adapt to chemotherapy is aneuploidy. 
+A prime example of evolutionary rescue is the ability of cancer to survive exposure to treatment. 
+One evolutionary mechanism by which a population of cancer cells can adapt to chemotherapy is aneuploidy. 
 Aneuploid cancer cells can have higher fitness in an environment altered by anti-cancer drugs, e.g., because of incomplete pathways targeted by the drugs. 
@@
 Here, we examine how aneuploidy impacts the fate of a population of cancer cells.
-We use multi-type branching processes to approximate the probability that a tumor survives drug treatment as a function of the initial tumor size, and rates at which aneuploidy and other beneficial mutations occur, and the growth rates of the sensitive and resistant cells.
-Additionally, we investigate what effect having a fraction of tumor cells being aneuploid before the onset of therapy have on the probability of evolutionary rescue.
-We also approximate the mean recurrence time for the tumor to revert back to its initial size.
-We find that aneuploidy can play an important role in the relapse of secondary tumors who have not yet been detected and thus are smaller in size.
+We use multi-type branching processes to approximate the probability that a tumor survives drug treatment as a function of the initial tumor size, the rates at which aneuploidy and other beneficial mutations occur, and the growth rates of the sensitive and resistant cells.
+Additionally, we investigate the effect on the probability of evolutionary rescue of having a fraction of aneuploid tumor cells before the onset of therapy.
+We also approximate the mean recurrence time for the tumor to revert back to its initial size following treatment.
+We find that aneuploidy can play an important role in the relapse of smaller secondary tumors who have not yet been detected.
 % TODO missing something here... ok
@@
 \paragraph{Aneuploidy in cancer.}  Each year approximately 10 million people die from cancer \citep{kocarnik2022cancer}, so understanding the factors that contribute to failure of interventions is of great importance.
-One hypothesized factor is aneuploidy, where cells are characterized by an imbalanced karyotype and chromosome copy number alterations \citep{schukken2018cin}, caused by chromosomal instability, the mitotic process in which cells suffer from chromosome mis-segregation that leads to aneuploidy.
-Importantly, aberrations in chromosome copy number have been shown to allow cancer cells to survive under stressful conditions such as drug therapy \citep{lukow2021chromosomal,rutledge2016selective}.
-Indeed, cancer cells are often likely to be aneuploid, and aneuploidy is associated with poor patient outcomes \citep{ben2020context,smith2018systematic}.
-
- \citet{ippolito2021gene} induced aneuploidy in cancer cell lines by exposing them to reversine, a small-molecule inhibitor of the mitotic kinase Mpsi1, and then to chemotherapeutic agents such as vemurafenib.
+One hypothesized factor is aneuploidy, where cells are characterized by an imbalanced karyotype and chromosome copy number alterations \citep{schukken2018cin}.
+Aneuploidy is caused by chromosomal instability, the mitotic process in which cells suffer from chromosome mis-segregation.
+Importantly, aberrations in chromosome copy number allow cancer cells to survive under stressful conditions such as drug therapy \citep{lukow2021chromosomal,rutledge2016selective}.
+Indeed, cancer cells are often aneuploid and aneuploidy is associated with poor patient outcomes \citep{ben2020context,smith2018systematic}.
+
+ \citet{ippolito2021gene} induced aneuploidy in cancer cell lines by exposing them to reversine, a small-molecule inhibitor of the mitotic kinase Mpsi1 and then to chemotherapeutic agents such as vemurafenib.
 Reversine-treated cells had higher proliferation rate in the environment altered by anti-cancer drugs compared to wildtype cancer cells.
@@
 An important aspect of aneuploidy is the rate with which cells become aneuploid, which is several orders of magnitude higher than the beneficial mutation rate \citep{bakker2023predicting}.
-Consequently, a cell exposed to a stress such as chemotherapeutic drugs can acquire aneuploidy faster when compared to acquiring a mutation, especially when several proposed anti-cancer drugs elevate the rate of mis-segregation in order to fight cancer \citep{lee2016effects}.
+Consequently, a cell exposed to a stress such as chemotherapeutic drugs can acquire aneuploidy faster compared to acquiring a mutation.
+Morover several proposed anti-cancer drugs elevate the rate of mis-segregation in order to fight cancer cells~\citep{lee2016effects}.
 
 \paragraph{Evolutionary rescue.} Populations adapted to a certain environment are vulnerable to environmental changes, which might cause extinction of the population.
-Examples of such environmental changes include climate change, invasive species or the onset of drug therapies.
+Examples of such environmental changes include climate change, invasive species, and the onset of drug therapies.
 Adaptation is a race against time as the population size decreases in the new environment~\citep{tanaka2022surviving}.
@@
 
-\citet{gunnarsson2020understanding} analyze a model where a tumor consisting of two populations of cancer cells, one drug resistant and the other drug sensitive, is able to evade extinction by cells switching between the two phenotypes through epigenetic mutations.
+\citet{gunnarsson2020understanding} analyzed a model where a tumor consisting of two populations of cancer cells, one drug resistant and the other drug sensitive, is able to evade extinction by cells switching between the two phenotypes through epigenetic mutations.
 They found that even when the drug resistant type is barely viable the epimutations have the effect of guaranteeing evolutionary rescue.
-Evolutionary rescue in one step in which an initially declining population, after a sudden environmental change, has to acquire a mutation has been studied in the context of population genetics  by \citep{orr2008population,orr2014population}.
-They analyzed a model where the mutant strain is present in small number at the onset of therapy and concluded that this can significantly enhance the chance that the population will survive.
-
-Most models focus on the probability that at least one mutation rescues the population.
-How multiple mutations contribute to the survival of the population is less explored, but \citet{wilson2017soft} have shown that evolutionary rescue is significantly enhanced by soft selective sweeps when multiple mutations contribute.
-Evolutionary rescue that requires two successive mutations has been investigated using diffusion approximation by \citet{martin2013probability}.
+Evolutionary rescue in one step where an initially declining population, after a sudden environmental change, has to acquire a mutation in order to survive extinction has been studied by \citet{orr2008population,orr2014population}.
+They analyzed a model where the mutant strain is present in small numbers at the onset of therapy and concluded that this can significantly enhance the probability that the population will survive.
+
+Most models focus on the probability that at most one mutation rescues the population.
+How multiple mutations contribute to the survival of the population is less explored, but \citet{wilson2017soft} have shown that evolutionary rescue is significantly enhanced by soft selective sweeps when multiple mutations contribute \textcolor{green}{towards evolutionary rescue}.
+Evolutionary rescue that requires two successive mutations (i.e., two steps) has been investigated using diffusion approximation by \citet{martin2013probability} \textcolor{green}{and tested the predictions of their model with data from experiments with yeast and bacterium}.
 
 Here we build on previous work on evolutionary rescue after a sudden environment change caused by the initiation of chemotherapy.
-We wish to understand what effect does aneuploidy have on the probability of evolutionary rescue when it acts as an intermediary between the wildtype and the mutant cancer cells.
-We also calculate the mean time that an initially declining tumor cell population reaches its pre-treatment size.
-Given that aneuploidy is present in many tumors even before the onset of therapy \citep{lukow2021chromosomal} we also take into consideration the effect that standing genetic variation has on the tumor dynamics.
+We \textcolor{green}{estimate the} effect of aneuploidy on the probability of evolutionary rescue when it acts as an \textcolor{green}{evolutionary stepping stone} between the \textcolor{green}{sensitive} and the \textcolor{green}{resistant} cancer cells.
+We also estimate the mean time \textcolor{green}{until the} tumor cell population reaches its pre-treatment size.
+Given that aneuploidy is present in many tumors even before the onset of therapy \citep{lukow2021chromosomal}, we also consider the effect of \textcolor{green}{pre-treatment} standing genetic variation on the \textcolor{green}{evolutionary} dynamics.
 Additionally, we are interested in the timescale of evolutionary rescue and the effect that aneuploidy has on the time necessary for the tumor to overcome drug therapy.
@@
 We assume the population of cells is under a strong stress, such as drug therapy, to which the wildtype genotype is susceptible or sensitive and therefore $\Delta_w<0$, whereas the mutant is resistant to the stress, $\Delta_m>0$.
-We analyze three scenarios: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_w$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing, stationary or growing or dying only very slowly, that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$, in a sense that we will make precise below.
+We analyze three scenarios: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_w$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing, stationary, or growing or dying only very slowly, that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$, in a sense that we will make precise below.
 We assume that both chromosomal missegregation and mutation occur during the process of mitosis. 
@@
 
-The state of the stochastic system at time $t$ is represented by the triplet $\left(w_t,a_t,m_t\right)$.
-The following describes the events that may occur (right column), the rates at which they occur (middle column), and the effect these events have on the state (left column, see \Cref{figureAneuploidy}):
+The \textcolor{green}{cell population} at time $t$ is represented by the triplet $\left(w_t,a_t,m_t\right)$.
+The following describes the events that may occur (right column), the rates at which they occur (middle column), and the effect these events have on the \textcolor{green}{population} (left column, see \Cref{figureAneuploidy}):
 
@@
 \end{subequations}
-For the remaining of this paper we assume that the division rates for wildtype and aneuploid cells can be written as $\lambda_ww_t\left(1-u-v\right)\approx \lambda_ww_t$ and $\lambda_aa_t\left(1-v\right)\approx\lambda_aa_t$ because $u,v\ll1$ (see Table \ref{table1}).
+For the remaining of this paper we assume that the division rates for wildtype and aneuploid cells can be written as $\lambda_ww_t\left(1-u-v\right)\approx \lambda_ww_t$ and $\lambda_aa_t\left(1-v\right)\approx\lambda_aa_t$ because $u,v\ll1$ (Table \ref{table1}).
 Each iteration of the simulation loop starts with computing the rates $\nu_k$ of each event $k$.
@@
 We assume the mutation rate is $10^{-7}$ per gene per cell division \citep{loeb2001mutator} and since we assume that a single target gene confers resistance to the drug, we use $v=10^{-7}$ per cell division.
-\citet{bakker2023predicting} determined that the missegregation rate must be between $10^{-3}-10^{-2}$ per chromosome per cell division with the optimal value being $6.21\times10^{-3}$ per chromosome per cell division.
+\citet{bakker2023predicting} determined that the missegregation rate must be between $10^{-3}-10^{-2}$ per chromosome per cell division with the optimal value being $6.2\times10^{-3}$ per chromosome per cell division \textcolor{green}{(for $u\ll10^{-3}$ no adaptation happens and for $u\gg10^{-2}$ karyotypes will acquire lethal nullisomies)}.
 \citet{ippolito2021gene} observed that trisomy in Chr II and VI are most likely to confer increased resistance against the chemotherapeutic agent vemurafenib for A375 cells.
@@
 
-The fitness cost $s$ of aneuploidy before the onset of therapy is difficult to estimate as we are interested in a specific type of aneuploidy which improves the fitness of cancer cells in an environment altered by drugs.
+The fitness cost $s$ of aneuploidy before the onset of therapy is difficult to estimate as we are interested in a specific type of aneuploidy that improves the fitness of cancer cells in an environment altered by drugs.
 We derive $s$ by using the formula $s=\tilde{u}\lambda_w / f$, where $f$ is the fraction of aneuploid cancer cells.
-To estimate $f$, we note that \citet{lukow2021chromosomal} mixed together wildtype and aneuploid A375 melanoma cells at $50:50$ ratio, cultured them in drug-free environment and observed the ratio evolve as a function of time with the aneuploid cells declining to $15\%$ after 24 days.
+To estimate $f$, we note that \citet{lukow2021chromosomal} mixed wildtype and aneuploid A375 melanoma cells at $50:50$ ratio, cultured them in drug-free environment and observed the ratio evolve as a function of time with the aneuploid cells declining to $15\%$ after 24 days.
 We obtain the fitness cost $s$ using the formula $s=\abs{\log\left[0.15/(1-0.15)\right]/24}\approx0.07$ per day \citep{chevin2011measuring}.
-As a result, the fraction of cancer cells with ``beneficial'' aneuploidy is $f=2\times10^{-3}\times10^{-1}/0.07=0.284\%$ (i.e.
-$0.284\%$ of pre-treatment cancer cells have the ``beneficial'' aneuploidy).
-
-We note that when we refer to wildtype cancer cells we include those cells that have any aneuploidy except trisomy in Chr II and VI as those are the aneuploid cells which are hypothesized to have higher fitness in the environment altered by drugs such as vemurafenib.
+As a result, the fraction of cancer cells with aneuploidy \textcolor{green}{of interest} is $f=2\times10^{-3}\times10^{-1}/0.07=0.28\%$ (i.e., $0.28\%$ of pre-treatment cancer cells have the aneuploidy \textcolor{green}{of interest}).
+
+We note that when we refer to wildtype cancer cells we include those cells that have any aneuploidy \textcolor{green}{other than} trisomy in Chr II and VI as those are the aneuploid cells that are hypothesized to have higher fitness in the environment altered by drugs such as vemurafenib.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-In our analytical calculations, we assume that lineages produced by cells from the initial population divide and die independently of each other.
-Of course, the cells will actually compete for resources, but we expect that this can be neglected because the drug will
-cause the cell density to rapidly drop far below the carrying capacity at which these interactions are important.
+In our \textcolor{green}{analysis}, we assume that lineages produced by cells from the initial population divide and die independently of each other.
+\textcolor{green}{However}, the cells will compete for resources, but we expect that this \textcolor{green}{competition} can be neglected because the drug will cause the cell density to rapidly drop below the carrying capacity at which \textcolor{green}{competition is most} important.
 To test this, we simulate a logistic growth model, with division and death rates given by:
@@
 where $K$ is the tumor carrying capacity. 
-The effective carrying capacity of this model is $K_e=K\Delta_a/\lambda_a\approx10^6$ for $K=10^8, \lambda_a=0.0901,\mu_a=0.09$, where we define the effective carrying capacity to be the population size at which the aneuploid division rate is equal to the aneuploid death rate.
+The effective carrying capacity \textcolor{green}{in} this model is $K_e=K\Delta_a/\lambda_a\approx10^6$ for $K=10^8, \lambda_a=0.0901,\mu_a=0.09$, where we define the effective carrying capacity to be the population size at which the aneuploid division rate is equal to the aneuploid death rate.
 
@@
 These approximations perform very well when compared to results of stochastic evolutionary simulations (\Cref{rescue_prob,rescue_threshold}).
-The first line describes the case in which aneuploid cells are still effectively killed by the treatment, but not as quickly as the wild type. 
+The first line describes the case in which aneuploid cells are still effectively killed by the treatment, but not as quickly as the wildtype. 
 In the second case, aneuploid cells are sufficiently resistant that the expected size of each aneuploid lineage is roughly 1.
-In both of these cases, aneuploidy increases the probability of rescue by slowing or halting the decrease of the cancer population, allowing more opportunities for producing resistant mutants. 
-In the third case, aneuploid cells are sufficiently resistance for the cancer population to re-grow the tumor even without additional resistance mutations.
+In both of these cases, aneuploidy increases the probability of rescue by slowing or halting the decrease of the \textcolor{green}{tumor} population \textcolor{green}{size}, allowing more opportunities for producing resistant mutants.
+In the third case, aneuploid cells are sufficiently resistant for the population to re-grow the tumor even without additional resistance mutations.
 Notably, in this case the mutant parameters ($v$, $\lambda_m$, and $\Delta_m$) do not affect $N_a^*$ beyond their effect on $T^*$.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\paragraph*{Standing vs. de-novo genetic variation.}
+\paragraph*{Standing vs. \emph{de-novo} genetic variation.}
 
 In the above we assumed that at the onset of drug treatment, the initial tumor consisted entirely of wildtype cells that are drug sensitive.
-However, aneuploidy is likely produced even before onset of treatment at some rate $\tilde{u}$ and confers a deleterious fitness effect $s$ in the absence of the drug \citep{replogle2020aneuploidy,giam2015aneuploidy}.
+However, aneuploidy is likely produced even before \textcolor{green}{the} onset of treatment at some rate $\tilde{u}$ and confers a deleterious fitness effect $s$ in the absence of the drug \citep{replogle2020aneuploidy,giam2015aneuploidy}.
 Furthermore, the aneuploidy rate in the presence of drugs is likely higher than in their absence, $\tilde{u} < u$ \citep{wang2019molecular,mason2017functional}.
-But if the number of cells in the tumor $N$ is large (as expected if the tumor is treated with a drug), there may already be a fraction $f \approx \tilde{u}\lambda_w/s$ of aneuploid cells in the population (here we assume that the drug affects the wildtype death rate but no the division rate and therefore use $\lambda_w$ for the wildtype division rate in the drug-free environment).
+But if the number of cells in the tumor $N$ is large (as expected if the tumor is treated with a drug), there may already be a fraction $f \approx \tilde{u}\lambda_w/s$ of aneuploid cells in the population (here we assume that the drug affects the wildtype death rate but not the division rate and therefore \textcolor{green}{we} use $\lambda_w$ for the wildtype division rate in the drug-free environment).
 
@@
 Comparing this approximation of $\tilde{N}^*_{a}/N^*_{a}$ to results of stochastic simulations, we find that the approximations perform very well (\Cref{rescue_denovo}). 
-Standing genetic variation will drive evolutionary rescue if wildtype growth rate $\Delta_w$ is very negative due to a strong effect of the drug on sensitive cells, or if the aneuploidy cost in the drug-free environment, $s$, is small.
-In contrast, de-novo aneuploid cells will have a greater contribution to rescue if the aneuploidy cost $s$ is large, the effect of the drug on sensitive cells is weak ($\Delta_w$ is close to zero), or if the drug induces the appearance of aneuploid cells ($u > \tilde u$).
-For example, with  $\lambda_w=0.1,\mu_w=0.14, u=10^{-2}, \tilde{u}=2\times10^{-3}, s=0.07$, the ratio of the threshold tumor sizes for standing vs de-novo variation is $\tilde{N}^*_a/N^*_a \approx 8.75$, which means that de-novo genetic variation is  the main driver of evolutionary rescue.
+Standing genetic variation will drive evolutionary rescue if wildtype cells \textcolor{green}{die rapidly} (growth rate $\Delta_w$ is very negative) due to a strong effect of the drug on sensitive cells, or if the aneuploidy cost in the drug-free environment, $s$, is small.
+In contrast, \emph{de-novo} aneuploid cells will have a greater contribution to rescue if the aneuploidy cost $s$ is large, the effect of the drug on sensitive cells is weak ($\Delta_w$ is close to zero), or if the drug induces the appearance of aneuploid cells ($u > \tilde u$).
+For example, with  $\lambda_w=0.1,\mu_w=0.14, u=10^{-2}, \tilde{u}=2\times10^{-3}, s=0.07$, the ratio of the threshold tumor sizes for standing vs.
+\emph{de-novo} variation is $\tilde{N}^*_a/N^*_a \approx 8.75$, which means that \emph{de-novo} genetic variation is  the main driver of evolutionary rescue.
 
@@
 For the parameter values of  $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, \tilde{u}=10^{-3}, v=10^{-7}$, we are in the tolerant aneuploid case and obtain the ratio $\tilde{N}^*_a/N^*_m \approx 0.9625$, which means that standing genetic variation reduces the threshold tumor size by approximately $4\%$.
-Therefore, standing genetic variation does not drive evolution of drug resistance when compared to de-novo aneuploidy, but it does offer a slight advantage when compared with direct mutation.
+Therefore, standing genetic variation does not drive evolution of drug resistance when compared to \emph{de-novo} aneuploidy, but it does offer a slight advantage when compared with direct mutation.
 
@@
 We therefore explored the time until recurrence of the tumor, that is, the time until the tumor reaches its original size, $N$.
-When the expected number of resistant lineages that avoid extinction is small, the expected recurrence time can be estimated by adding two terms: the \emph{mean evolutionary rescue time}, which is the waiting time for appearance of a resistant lineage that avoids extinction (conditioned on such an even occurring in the first place), and the \emph{mean proliferation time}, which is the expected time for that lineage to grow to $N$ cells.
-However, when the expected number of resistant lineages is large, the dynamics of number of mutant cells is deterministic (i.e., it can be modeled by a system of ODEs, \cref{detODE}) and the mean recurrence time cannot be separated into the mean evolutionary rescue time and mean proliferation time because multiple mutant lineages contribute towards the mutant population size reaching the initial tumor size.
+When the expected number of resistant lineages that avoid extinction is small, the expected recurrence time can be estimated by adding two terms: the \emph{mean evolutionary rescue time}, which is the waiting time for appearance of a resistant lineage that avoids extinction (conditioned on such an event occurring in the first place), and the \emph{mean proliferation time}, which is the expected time for that lineage to grow to $N$ cells.
+However, when the expected number of resistant lineages is large, the dynamics of \textcolor{green}{the} number of mutant cells is deterministic (i.e., it can be modeled by a system of ODEs, \cref{detODE}) and the mean recurrence time cannot be separated into the mean evolutionary rescue time and mean proliferation time because multiple mutant lineages contribute towards the mutant population size reaching the initial tumor size.
 % TODO I dont like the word "behaves", please be specific.
@@
 \tau_a \approx \begin{cases}
-    \frac{1}{\abs{\Delta_w}}+\frac{1}{\abs{\Delta_a}} ,&
+    -\frac{1}{\Delta_w}-\frac{1}{\Delta_a} ,&
  N \ll N_a^* ,\\ 
@@
 Increasing the wildtype or aneuploid growth rates leads to an increase in the mean rescue time, because the corresponding cells will survive for longer and will produce additional rescue mutations at latter times.
-In our focus parameter regime, we have $\Delta_w=-0.04$ and $\Delta_a=-10^{-4}$, such that the mean rescue time is mainly determined by the aneuploid growth rate, $\tau_a \approx 10^4$.
+In our focus parameter regime, we have $\Delta_w=-0.04$ and $\Delta_a=-10^{-4}$, such that the mean rescue time is mainly determined by the aneuploid growth rate, $\tau_a \approx 10^4$ days.
 
@@
 
-Given that a fraction $f\approx 0.284\%$ of the initial cancer cell population is expected to be aneuploid even before the drug administration we want to know whether the mean evolutionary rescue time is affected by the standing genetic variation.
-For this purpose we calculate the mean evolutionary rescue time with standing genetic variation $\tau_a^f$ (see \cref{meantimet2SGV}) and compare our result with simulations in  \Cref{SGVEvolutionaryRescueTimeComplete}.
+Given that a fraction $f\approx 0.28\%$ of the initial cancer cell population is expected to be aneuploid even before \textcolor{green}{onset} of the drug \textcolor{green}{treatment}, we want to know whether the mean evolutionary rescue time is affected by the standing genetic variation.
+We calculated the mean evolutionary rescue time with standing genetic variation $\tilde{\tau}_a$ (\cref{meantimet2SGV}) and compared our result with simulations in  \Cref{SGVEvolutionaryRescueTimeComplete}.
 We note that standing genetic variation does not have a significant effect on the mean evolutionary rescue time.
@@
 
-In Appendix \ref{sec:appendix_distribution_time} we calculate the probability that a successful mutation, which will rescue the population, has been generated by time $t$.
-This allows us to observe whether aneuploidy accelerates or delays adaptation.
-We plot our results in  \Cref{cdffig}A alongside simulations for different aneuploid growth rates ($u>0$) and the case when aneuploidy is absent ($u=0$).
-We observe that aneuploidy starts to have an effect on adaptation for timescales greater then $1/\Delta_a\approx100$ days after which no more rescue mutations are generated through direct mutation.
-This shows that rescue mutations are generated through aneuploidy at latter timescales then direct mutation and, as a consequence, aneuploidy increases the \emph{window of opportunity} for evolutionary rescue.
+In Appendix \ref{sec:appendix_distribution_time} we calculated the probability that a \textcolor{green}{rescue} mutation has been generated by time $t$.
+This allows us to \textcolor{green}{examine} whether aneuploidy accelerates or delays evolutionary rescue.
+\textcolor{green}{We find that anenploidy significantly affects evolutionary rescue after} $1/\Delta_a\approx100$, days after which no more rescue mutations are generated through direct mutation (\Cref{cdffig}A).
+This shows that rescue mutations are generated \textcolor{green}{in aneuploid cells} at latter timescales then \textcolor{green}{in wildtype cells} and, \textcolor{green}{thus}, aneuploidy increases the \emph{window of opportunity} for evolutionary rescue.
 % WHAT is the purpose of this sentence?
@@
 
-We next approximate the mean time for the population of mutant cancer cells to reach the initial, pre-treatment population size $N$, which we denote the recurrence time $\tau_a^r$ (Appendix \ref{sec:appendix_recurrence_time}),
+We next approximated the mean time for the population of mutant cancer cells to reach the initial, pre-treatment population size $N$, which we denote the recurrence time $\tau_a^r$ (Appendix \ref{sec:appendix_recurrence_time}),
 \begin{equation} \label{eq:AsymptoticTimeRulesRecurrence}
 \tau_a^r \approx \begin{cases}
-   \frac{1}{ \abs{\Delta_w}}+\frac{1}{ \abs{\Delta_a}}+\frac{\log p_mN}{\Delta_m} ,&
+   -\frac{1}{\Delta_w}-\frac{1}{\Delta_a}+\frac{\log p_mN}{\Delta_m} ,&
  N \ll N_a^* ,\\ 
@@
 Increasing the mutant growth rate $\Delta_m$ decreases recurrence times while increasing the wildtype and aneuploid growth rates, $\Delta_w$ and $\Delta_a$ respectively, increases the recurrence time.
-For large tumors ($N \gg N_m^*$), the dynamics of the number of mutant cells is deterministic and the mean recurrence time becomes constant (and independent of the initial tumor size $N$).
+For large tumors ($N \gg N_m^*$), the dynamics of the number of mutant cells is deterministic and the mean recurrence time becomes independent of the initial tumor size $N$.
 Increasing either the mutant growth rate $\Delta_m$ or the mutation rate $v$ leads to a decrease in the time for the tumor to rebound to its initial size.
@@
 
-We note that, for small and large tumors, when $N \ll N_a^*$ or $N \gg N_m^*$, the asymptotic expressions for the mean recurrence time are independent of the chromosome missagregation rate $u$, therefore the rate at which the drug induces aneuploidy has no effect on the time necessary for the tumor to rebound to its initial size $N$.
+We note that, for small and large tumors, when $N \ll N_a^*$ or $N \gg N_m^*$, the asymptotic expressions for the mean recurrence time are independent of the chromosome missegregation rate $u$, \textcolor{green}{and} therefore the rate at which the drug induces aneuploidy has no effect on the time necessary for the tumor to rebound to its initial size $N$.
 % Remus the next paragraphs look like you wanted to find approximations so you did, you reference them here, but the results are buried in the technical text: WHAT IS IT THAT YOU WANT US TO LEARN from this?
@@
 
-The detection time $\tau_a^{r,M}$ can be defined as the time necessary for the tumor size to reach detection threshold $M$.
-We derive the mean recurrence time to detection size $M=10^7$ in appendix \ref{sec:appendix_recurrence_time}.
-We observe that  for small and intermediate sized tumors the effect of the detection size $M$ on $\tau_a^{r,M}$ is negligible when compared to the case where the detection size is $N$ (i.e.
-$\tau_a^r\approx\tau_a^{r,M}$ for $N<N_m^*$).
+The detection time $\tau_a^{M}$ can be defined as the time for the tumor size to reach detection threshold $M$.
+We derive the mean \textcolor{green}{detection} time for $M=10^7$ in Appendix \ref{sec:appendix_recurrence_time}.
+We observe that  for small and intermediate sized tumors the effect of the detection size $M$ on $\tau_a^{M}$ is negligible when compared to the case where the detection size is \textcolor{green}{equal to the initial population size} $N$ (i.e., $\tau_a^r\approx\tau_a^{M}$ for $N<N_m^*$).
 % WHAT DOES THAT MEAN?
@@
 However, for large tumors the mean recurrence time to detection size $M$ decreases logarithmically with tumor size $N$ while $\tau_a^{r}$ is constant (\Cref{RecurrencePlot}).
-Additionally, for large tumors we have $M<N_m^*<N$ which tells us that the mean detection time is smaller for such tumors when compared to the mean recurrence time.
+Additionally, for large tumors we have $M<N_m^*<N$ so the mean detection time is shorter for such tumors when compared to the mean recurrence time, \textcolor{green}{that is the resistant tumor can be detected before reaching its initial size}.
 
@@
 ok
-Therefore, since only undetected secondary tumors are present at the time of the administration of the anti-cancer drug we lack knowledge of the size of the tumors and we cannot compare the empirical distributions to our predictions.
-However, we expect that the variability of secondary tumors to average out across large cohorts of patients, thus we can use the mean recurrence time to compare with clinical data.
+Therefore, since only undetected secondary tumors are present at the time of the \textcolor{green}{treatment with} anti-cancer drug, we lack data on the size of the tumors and cannot compare the empirical distributions to our predictions.
+However, we expect  the variability of secondary tumors to average out across large cohorts of patients, thus we can use the mean recurrence time to compare with clinical data.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \paragraph{Evolutionary rescue}
-Using multitype branching processes, we derived the probability of evolutionary rescue of the tumor under different scenarios for the effect of aneuploidy, ranging from tolerance to partial resistance.
+Using multitype branching processes, we derived the probability of evolutionary rescue of the tumor under effects of aneuploidy, ranging from tolerance to partial resistance.
 We obtained exact and approximate expressions for the probability of evolutionary rescue (\cref{eq:rescue_prob}). 
@@
 First, it delays the extinction of the population, providing more time for appearance of the resistance mutation.
-Second, it increases the population size relative to a sensitive population, providing more cells in which mutations can occur, i.e., it increases the mutation supply (i.e.
-$Nuv\lambda_w\lambda_a/\abs{\Delta_w\Delta_a}$).
-
-We find that aneuploidy can have a significant effect on evolutionary rescue as it reduces the threshold tumor size by at least one order of magnitude even when aneuploidy is tolerant (\Cref{rescue_prob}).
-Interestingly, aneuploidy is unlikely to contribute to evolutionary rescue in primary tumors in which the number of cells is large enough (i.e.
-$N\gg N_m^*\approx 4\times10^7$) for the appearance of resistant mutation directly in sensitive cells before these cells become extinct (\Cref{rescue_prob}).
-However, aneuploidy can have a crucial role in evolutionary rescue of secondary tumors, in which the number of sensitive cells may be below the detection threshold of $\sim10^7$  \citep{bozic2013evolutionary}, and this can have an impact on the recurrence of cancer after the resection of the primary tumor  through secondary tumors which are too small to be detected and for which chemotherapy is employed to prevent  cancer relapse and are estimated to cause the majority of cancer-related deaths~\citep{chaffer2011perspective}.
-The importance of aneuploidy in the evolutionary rescue of secondary tumors is reinforced by the fact that metastases have been shown to have a chromosome missagregation rate two to three orders of magnitude higher then that of primary tumors~\citep{kimmel2023intra}.
-
-Given the fact that the mean time for such secondary tumors to overcome chemotherapy can be of the order of 1000 days (\Cref{MeanTimeGrowthAneuploidyPlot}A),
-aneuploidy can explain the reappearance of cancer even after initial remission.
+Second, it increases the population size relative to a sensitive population, providing more cells in which mutations can occur, i.e., it increases the mutation supply (i.e., $Nuv\lambda_w\lambda_a/\abs{\Delta_w\Delta_a}$).
+
+We find that aneuploidy can have a significant effect on evolutionary rescue as it reduces the threshold tumor size by at least an order-of-magnitude even when aneuploidy only provides tolerance (\Cref{rescue_prob}).
+Interestingly, aneuploidy is unlikely to contribute to evolutionary rescue in primary tumors in which the number of cells is large enough (i.e., $N\gg N_m^*\approx 4\times10^7$) for the appearance of resistant mutation in sensitive cells before these cells become extinct (\Cref{rescue_prob}).
+However, aneuploidy can have a crucial role in the evolutionary rescue of secondary tumors, in which the number of sensitive cells may be below the detection threshold of $\sim10^7$  \citep{bozic2013evolutionary}, and this can have an impact on the recurrence of cancer after the resection of the primary tumor  through secondary tumors, which are too small to be detected and for which chemotherapy is employed to prevent  cancer relapse.
+\textcolor{green}{These secondary tumors} are estimated to cause the majority of cancer-related deaths~\citep{chaffer2011perspective}.
+The importance of aneuploidy in the evolutionary rescue of secondary tumors is reinforced by the fact that metastases have been shown to have a chromosome missegregation rate two to three orders of magnitude higher then that of primary tumors~\citep{kimmel2023intra}.
+
+Given the fact that the mean time for secondary tumors to overcome chemotherapy can be of the order of 1,000 days (\Cref{MeanTimeGrowthAneuploidyPlot}A), aneuploidy can explain the reappearance of cancer even after initial remission.
 The theoretical prediction for the mean rescue time for tumors smaller then  $10^8$ cells is greater then 4 years, consistent with previous estimates of the recurrence time of tumors after resection~\citep{avanzini2019cancer}.
-We observe from \Cref{cdffig}A that aneuploidy complements evolutionary rescue through direct mutation given that aneuploidy is most likely to generate rescue mutations at time scales larger then the timescales at which rescue mutations are generated through direct mutation.
-
-We hypothesized that \emph{standing variation} (the existence of a subpopulation of aneuploid cancer cells before the onset of therapy) can facilitate evolutionary rescue by reducing the waiting time for the appearance of aneuploid cells.
-From \Cref{eq:Nsgv} we observe that a drug or combination of drugs that reduces the wildtype growth rate and does not significantly increase the chromosome missegregation rate is more likely to cause evolutionary rescue to occur through direct mutation.
-Furthermore, we find that for reasonable parameter values evolutionary rescue is more likely to occur through \emph{de-novo} aneuploidy (\Cref{rescue_denovo}).
-If the fraction of tumor cells that have the ``advantageous'' aneuploidy is $f\gg\frac{u\lambda_w}{\abs{\Delta_w}}\approx2.5\%$, then evolutionary rescue is more likely to occur via standing variation, rather then through \emph{de novo} aneuploid cells.
+\textcolor{green}{We found that aneuploidy complements evolutionary rescue through direct mutation because it generate rescue mutations only after the wildtype population had died out and direct mutation is no longer a viable option for evolutionary rescue (\Cref{cdffig}A).}
+
+We hypothesized that \emph{standing \textcolor{green}{genetic} variation} (the existence of a subpopulation of aneuploid cancer cells before the onset of therapy) can facilitate evolutionary rescue by reducing the waiting time for the appearance of aneuploid cells.
+We found that a drug or combination of drugs that reduces the wildtype growth rate and does not significantly increase the chromosome missegregation rate is more likely to cause evolutionary rescue to occur through direct mutation (\cref{eq:Nsgv}).
+Furthermore, we found that for reasonable parameter values evolutionary rescue is more likely to occur through \emph{de-novo} aneuploidy (\Cref{rescue_denovo}).
+If the fraction of tumor cells that have the \textcolor{green}{beneficial} aneuploidy is $f\gg\frac{u\lambda_w}{\abs{\Delta_w}}\approx2.5\%$, then evolutionary rescue is more likely to occur via standing variation, rather then through \emph{de-novo} aneuploidy.
 In this case, evolutionary rescue likely occurs via aneuploid cells that acquire a resistance mutation and the probability of evolutionary rescue declines exponentially with initial tumor size  (the probability of evolutionary rescue is given by $1-\exp\left(-N/\tilde{N}_a^*\right)$, see \Cref{rescue_prob_sgv}).
@@
 \paragraph{Experimental future direction}
-Our model prediction could be tested by experiments \citep{martin2013probability}.
+Our model predictions could be tested by experiments \citep{martin2013probability}.
 For example, to assess the effect of initial tumor size on the probability of evolutionary rescue, a large culture mass can be propagated from a single cancer cell in permissive conditions and then diluted to a  range of starting tumor sizes.
@@
 Cell density can be measured by optical density and a population exposed to the drug is considered extinct if the optical density is lower when compared to the control case with no cells present.
-We can then compare the results of the experiments to the predictions of our model to see if the tumors with initial size bellow the threshold \eqref{eq:N_a} are more likely to go killed by the drug.
+We can then compare the results of the experiments to the predictions of our model to see if tumors with initial size bellow the threshold \cref{eq:N_a} are more likely to become extinct due to the drug.
 
 Additionally, our model predictions can be tested with data from patients.
-Given the type of cancer and how ``advantageous'' the aneuploidy is,  our model can predict the probability for cancer relapse after treatment.
+Given the type of cancer and how \textcolor{green}{beneficial} aneuploidy is,  our model can predict the probability for cancer relapse after treatment.
 
 \paragraph{Directions of future research}
- Our model can be extended to understand evolutionary rescue in different biological contexts, for example, how yeast subject to stress can overcome extinction via aneuploidy~\citep{pompei2023fitness,kohanovski2024aneuploidy}.
-Additionally, we did not account for the heterogeneity of aneuploidy as not all the aneuploidy lineages generated have the same growth rate $\Delta_a$ as we have assumed in our model~\citep{avecilla2023copy,yang2021fitness}.
-Such heterogeneity can be accounted for  by  sampling the aneuploidy growth or death rates from a distribution~\citet{martin2013probability}.
-Furthermore, tumor heterogeneity is also important for pre-existing genetic variation where the fitness of the aneuploid cells can be drawn from a distribution and the fittest genotype is selected when the tumor is exposed to the drug treatment.
-
-We have assumed that cancer cell lineages are independent,
-and have verified that this approximation is accurate under simple logistic growth. 
-However, this neglects potential effects of spatial structure and local interactions, which may be important in solid tumors.
+ Our model can be extended to understand evolutionary rescue in different biological contexts, for example, how yeast \textcolor{green}{populations under} stress overcome extinction via aneuploidy~\citep{pompei2023fitness,kohanovski2024aneuploidy}.
+Additionally, we did not account for the heterogeneity of aneuploidy as not all aneuploid lineages have the same growth rate $\Delta_a$ as assumed in our model~\citep{avecilla2023copy,yang2021fitness}.
+Such heterogeneity can be accounted for  by  sampling the aneuploidy growth or death rates from a distribution~\citep{martin2013probability}.
+\textcolor{green}{Furthermore, tumor heterogeneity may be important for genetic variation.}
+
+We have assumed that cancer cell lineages are independent and have verified that this is accurate under simple logistic growth. 
+However, this assumption neglects potential effects of spatial structure and local interactions, which may be important in solid tumors.
 Such tumors can be spatially heterogeneous with different genotypes inhabiting cellular niches and immune infiltration impacting growth in affected regions \citep{varrone2023cellcharter,galon2010immune}.
-This has the potential to impact the probability of overcoming chemotherapy \citep{martens2011spatial}.
-Future work should take into consideration the spatial structure of the tumor and its effect on the probability of evolutionary rescue.
-
-An additional limitation of our model is the choice of parameters which are based on a specific set of assumptions which might not be true for many tumor.
-As a result, incorporating a wider distribution of parameters would be a beneficial extension for our model, however parameter values for different types of tumors are difficult to obtain.
+This has the potential to impact the probability of evolutionary rescue \citep{martens2011spatial}.
+
+An additional \textcolor{green}{extension} of our model \textcolor{green}{could incorporate} a wider range of parameters, however parameter values for different types of tumors are difficult to obtain.
 
@@
 Our results quantitatively show that aneuploidy plays an important role in tumors overcoming exposure to chemotherapeutic drugs when tumor size is small or intermediate.
-Large tumors can escape anti-cancer drugs through direct mutation while smaller ones are able to obtain the beneficial mutation through an aneuploid ``\emph{stepping stone}'' (\Cref{rescue_prob}).
-As a result, therapies that increase the rate of aneuploidy in tumors in order to combat cancer may have an adverse effect on patient outcomes.
+Large tumors can escape anti-cancer drugs through direct mutation while smaller ones are able to obtain a resistance mutation through an aneuploid \emph{stepping stone} (\Cref{rescue_prob}).
+As a result, therapies that increase the rate of aneuploidy in tumors to combat cancer may have an adverse effect on patient outcomes.
 % TODO summary ok
@@
 Wildtype cells can divide and become aneuploid at rate $u\lambda_w$.
-Both aneuploid and wildtype cells can divide and acquire a beneficial mutation with rate $v\lambda_a$ and $v\lambda_w$, respectively.
+Both aneuploid and wildtype cells can divide and acquire a mutation with rate $v\lambda_a$ and $v\lambda_w$, respectively.
 Color denotes the relative growth rates of the three genotypes such that $\lambda_w - \mu_w < \lambda_a - \mu_a < \lambda_m - \mu_m$.
@@
 \textbf{Sample trajectories of the different genotype frequencies.}
-(A) When the missegregation rate $u=0$ evolutionary rescue is only possible through direct mutation and in most cases the tumor will be killed by the drug.
-(B) When the aneuploidy growth rate $\Delta_a\ll0$ we observe similar dynamic to case (A) as direct mutation is the only viable route toward evolutionary rescue for the tumor.
-(C) Intermediate aneuploid growth rates $\Delta_a\approx0$ leads to the appearance of aneuploid lineages even after the wildtype population has gone extinct thus increasing the chance of evolutionary rescue.
-(D) As the growth rate of the aneuploid becomes positive we observe that the tumour is rescued by the aneuploid cancer cell population.
-Each plot features 10 trajectories of $\left(w_t,a_t,m_t\right)$ as a function of time $t$ for the following parameter values: $\lambda_w=0.1,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, v=10^{-7},N=10^7$.
+(A) \textcolor{green}{Without aneuploidy} ($u=0$0 evolutionary rescue is possible through direct mutation and in most cases the tumor will \textcolor{green}{become extinct due to} the drug.
+(B) When the aneuploidy \textcolor{green}{cells are tolerant} ($\Delta_a<0$) we observe that, similar to case (A), direct mutation is the only \textcolor{green}{path for} evolutionary rescue.
+(C) \textcolor{green}{When aneuploidy cells are stationary} ($\Delta_a\approx0$), we observe the appearance of aneuploid lineages even after the wildtype population has gone extinct \textcolor{green}{thus showing that stationary aneuploidy increases the probability of evolutionary rescue}.
+(D) \textcolor{green}{When the aneuploid cells are partially resistant} ($\Delta_a>0$) the tumor is rescued by the aneuploid cell population.
+Each plot shows 10 simulations of the number of \textcolor{green}{wildtype, aneuploid and mutant cells} $\left(w_t,a_t,m_t\right)$ \textcolor{green}{over} time $t$ \textcolor{green}{with} the following parameter values: $\lambda_w=0.1,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, v=10^{-7},N=10^7$.
 For (A) we set $u=0$, for (B) $\lambda_a=0.065,u=10^{-2}$, for (C) $\lambda_a=0.0899,u=10^{-2}$ and for (D) $\lambda_a=0.095,u=10^{-2}$.
@@
 \caption{\textbf{Aneuploidy facilitates evolutionary rescue of cancer under drug treatment.}
-The probability of evolutionary rescue (i.e.
-the probability that the population does not go to extinction), $\presc$, as a function of the initial tumor size, $N$ (see \cref{eq:rescue_prob}).
-Dashed vertical line shows the threshold tumor size, $N_a^*$, above which the probability is very high (see \cref{eq:N_a}).
-Blue dashed line represents the probability of evolutionary rescue as a function of $N$ without aneuploidy ($u=0$).
-The black line represents the case with tolerant aneuploidy ($u=10^{-2}, \lambda_a=0.0899$), the red line represents the case with stationary aneuploidy ($u=10^{-2}, \lambda_a=0.08999$) and the green line represents the case with partially resistant aneuploidy ($u=10^{-2}, \lambda_a=0.095$).
-The dots represent simulations and the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the fraction of simulations in which the tumor has adapted to the stress and $n=100$ is the number of simulations.
+The probability of evolutionary rescue (i.e., the probability that the population does not become extinct), $\presc$, as a function of the initial tumor size, $N$ (\cref{eq:rescue_prob}).
+Dashed vertical line shows the threshold tumor size, $N_a^*$, above which the probability is very high (\cref{eq:N_a}).
+Blue dashed line: without aneuploidy ($u=0$).
+Black line: tolerant aneuploidy ($u=10^{-2}, \lambda_a=0.0899$).
+Red line: stationary aneuploidy ($u=10^{-2}, \lambda_a=0.08999$).
+Green line represents the case with partially resistant aneuploidy ($u=10^{-2}, \lambda_a=0.095$).
+Dots for simulations and the error bars for $95\%$ confidence interval ($p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the fraction of simulations in which the tumor has been rescued and $n=100$ is the number of simulations).
 Parameters: $\lambda_w=0.1,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, v=10^{-7}$.}
@@
 \caption{
-\textbf{Aneuploidy facilitates the evolutionary rescue of cancer under drug treatment.}
+\textbf{\textcolor{green}{The effect of aneuploidy on tumor threshold size.}}
 (A) The threshold tumor size $N_a^*$ as a function of the aneuploid growth rate $\Delta_a$.
@@
 When aneuploid growth rate is close to or higher than zero, aneuploidy decreases the threshold tumor size, thereby facilitating evolutionary rescue.
-The inset highlights the case when aneuploidy cancer cells are non-growing.
-The red dots represents simulations and the error bars represent the $95\%$ confidence intervals obtained with bootstrapping, see Appendix G.
+The inset highlights the case when aneuploidy cells are \textcolor{green}{stationary}.
+Red dots for simulations and error bars for the $95\%$ confidence intervals obtained with bootstrap (Appendix G).
 Parameters: $\lambda_w=0.1,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.
-(B) The threshold tumor size $N_a^*$ as a function of the ratio of aneuploidy and mutation rates, $u/v$.
-The dashed horizontal line shows $N^*_m$, the threshold tumor size without aneuploidy ($u=0$).
+(B) Threshold tumor size $N_a^*$ as a function of the ratio of aneuploidy and mutation rates, $u/v$.
+Dashed horizontal line shows $N^*_m$, the threshold tumor size without aneuploidy ($u=0$).
 When the aneuploidy rate is much higher than the mutation rate, aneuploidy decreases the threshold tumor size, thereby facilitating evolutionary rescue.
-The blue line represents the exact formula for threshold tumor size $N_a^*$ while the solid black line represents the approximation \cref{eq:N_a}.
-The red dots represents simulations and the error bars represent the $95\%$ confidence intervals obtained with bootstrapping, see Appendix G.
+Blue line represents the exact formula for threshold tumor size $N_a^*$ while the solid black line represents the approximation (\cref{eq:N_a}).
+Red dots represents simulation results and the error bars represent the $95\%$ confidence intervals obtained with bootstrap (Appendix G).
 Parameters: $\lambda_w=0.1,\lambda_m=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, v=10^{-7}$.
@@
 \caption{
-\textbf{Standing genetic variation facilitates evolutionary rescue of cancer under drug treatment.}
-(A)  The ratio of threshold tumor size $\tilde{N}_a^*$ when a fraction $\frac{\tilde{u}\lambda_w}{s}$ is aneuploid at the start of treatment and $N_a^*$ as a function of the wildtype growth rate $\Delta_w$.
-Standing genetic variation will drive adaptation to the drug if $\Delta_w$ is very negative due to a stronger effect of the drug on sensitive cells.
-The red dots represents simulations and the error bars represent the $95\%$ confidence intervals obtained with bootstrapping, see Appendix G.
+\textbf{\textcolor{green}{De-novo} standing genetic variation facilitates evolutionary rescue of cancer under drug treatment.}
+(A)  Ratio of threshold tumor size $\tilde{N}_a^*$ when a fraction $\frac{\tilde{u}\lambda_w}{s}$ is aneuploid at the start of treatment and $N_a^*$ as a function of the wildtype growth rate $\Delta_w$.
+Standing genetic variation will drive adaptation to the drug if \textcolor{green}{the wildtype population is rapidly declining} ($\Delta_w\ll0$) due to a stronger effect of the drug on sensitive cells.
+Red dots represents simulation \textcolor{green}{results} and the error bars represent the $95\%$ confidence intervals obtained with bootstrap (Appendix G).
 Parameters: $\lambda_w=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_a=0.09,\mu_m=0.09,\tilde{u}=10^{-3},u=10^{-2}, v=10^{-7}$.
-(B) The ratio of threshold tumor size $\tilde{N}_a^*$ when a fraction $\frac{\tilde{u}\lambda_w}{s}$ is aneuploid at the start of treatment and $N_a^*$ as a function of the the ratio of aneuploidy rates $\tilde{u}/u$.
-De-novo aneuploids will have larger contribution to the appearance of drug resistance if the drug induces the appearance of aneuploid cells ($u \gg \tilde u$).
-The red dots represents  simulations and the error bars represent the $95\%$ confidence intervals obtained with bootstrapping, see Appendix G.
+(B) Ratio of threshold tumor size $\tilde{N}_a^*$ when a fraction $\frac{\tilde{u}\lambda_w}{s}$ is aneuploid at the start of treatment and $N_a^*$ as a function of the the ratio of aneuploidy rates $\tilde{u}/u$.
+\emph{De-novo} aneuploids will have larger contribution to the appearance of drug resistance if the drug induces the appearance of aneuploid cells ($u \gg \tilde u$).
+Red dots represents  simulation \textcolor{green}{results} and the error bars represent the $95\%$ confidence intervals obtained with bootstrap (Appendix G).
 Parameters: $\lambda_w=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09,\tilde{u}=10^{-3}, v=10^{-7}$.
@@
 \end{subfigure}
-\caption{\textbf{Aneuploidy has a impact on cancer relapse early on.}
+\caption{\textbf{Aneuploidy \textcolor{green}{extends the window of opportunity for evolutionary rescue.}}
 (A) The probability that a successful mutant has not appeared by time $t$.
-The green line represents the case with tolerant aneuploidy ($u>0, \lambda_a=0.0899$), the blue line represents the case with non-growing aneuploidy ($u>0, \lambda_a=0.089999$), the cyan line represents the case with partially resistant aneuploidy ($u>0, \lambda_a=0.095$) and the black line represents the case without aneuploidy ($u=0$).
-As time increases, aneuploidy plays an important role in helping the cancer cell population escape extinction.
-The markers represent simulations and the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the fraction of simulations in which a successful mutant has not been generated and $n=100$ is the number of simulations.
+Green line: tolerant aneuploidy ($u>0, \lambda_a=0.0899$).
+Blue line: stationary aneuploidy ($u>0, \lambda_a=0.089999$).
+Cyan line: partially resistant aneuploidy ($u>0, \lambda_a=0.095$).
+Black line: no aneuploidy ($u=0$).
+Aneuploidy starts to play an important role in \textcolor{green}{rescuing} the cancer cell population \textcolor{green}{as the wildtype population becomes extinct}.
+Markers represent simulation results and the error bars represent $95\%$ confidence interval ($p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the fraction of simulations in which a successful mutant has not been generated and $n=100$ is the number of simulations).
 Parameters: $\lambda_w=0.1,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7},N=10^7$.
-(B) The probability that a mutant cancer cell population has not reached size $N$ at time $t$.
-The green line represents the case where $N=10^6$ (small tumor), the red line represents $N=10^7$ (intermediate sized tumor) and the blue line represents the case where $N=10^{10}$ (large tumor).
-Increasing the initial tumor size guarantees that the tumor will regrow.
-The markers represent simulations and the error bars represents $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the fraction of the simulations in which the mutant population size has not reached $N$ and $n=100$ is the number of simulations.
+(B) Probability that a mutant cancer cell population has not reached size $N$ at time $t$.
+Green line: $N=10^6$ (small tumor).
+Red line: $N=10^7$ (intermediate sized tumor).
+Blue line: $N=10^{10}$ (large tumor).
+Increasing the initial tumor size guarantees that the cancer will relapse.
+Markers represent simulations and the error bars represents $95\%$ confidence interval ($p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the fraction of the simulations in which the mutant population size has not reached $N$ and $n=100$ is the number of simulations).
 Parameters: $\lambda_w=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}
@@
 
-To analyze evolutionary rescue in this model, we use the framework of \emph{multitype branching processes} \citep{harris1963theory, weissman2009rate}. 
+To analyze evolutionary rescue in our model, we use the framework of \emph{multitype branching processes} \citep{harris1963theory, weissman2009rate}. 
 This allows us to find explicit expressions for the \emph{survival probability}: the probability that a lineage descended from a single cell does not become extinct.
@@
 \end{align*}
-Using the quadratic Taylor expansion $\sqrt{1+x}=1+x/2+\mathcal{O}(x^2)$ and assuming $u,v \ll 1$,
+Using the Taylor expansion $\sqrt{1+x}=1+x/2+\mathcal{O}(x^2)$ and assuming $u,v \ll 1$,
 we obtain the following approximation for the survival probability of a population initially consisting of a single wildtype cell,
@@
 \end{equation}
-Since $u,v\ll1$, the term in the root can be approximated using a 1st-order Taylor expansion. So, substituting the expressions for $p_a$ and $p_m$, we have
+Since $u,v\ll1$, the term in the root can be approximated using a Taylor expansion. So, substituting the expressions for $p_a$ and $p_m$, we have
 \begin{equation} \label{eq:survprobwinitial}
@@
 \end{equation}
-If we assume that $u\lambda_a>\abs{\Delta_a}$ then we have:
+If we assume that aneuploidy is not rare ($u\lambda_a>\abs{\Delta_a}$) then we have:
 \begin{equation}\label{eq:pw_tolerant}
@@
 \end{equation}
-Plugging this in \cref{eq:survprobwapprox1}, the survival probability of a population starting from one wildtype individual is
+Plugging this in \cref{eq:survprobwapprox1}, the survival probability of a population starting from one wildtype cell is
 \begin{equation}\label{eq:scenario3}
@@
 \begin{equation}\label{aneuploidpopeq}
-a_t \approx \frac{Nu\lambda_w\e^{\Delta_wt}}{\Delta_w-\Delta_a}\left[1-\e^{-\left(\Delta_w-\Delta_a\right)t}\right].
+a_t \approx \frac{Nu\lambda_w}{\Delta_w-\Delta_a}\left(\e^{\Delta_wt}-\e^{\Delta_a t}\right).
 \end{equation}
@@
 \begin{align}\label{meantimet2SGV}
-\tau_a^f&= \frac{\int_{0}^{\infty}{tR_a^f(t)\exp\left(-\int_0^{t}{R_a^f(z)\d z}\right) \d t}}{p_{rescue}},
+\tilde{\tau}_a&= \frac{\int_{0}^{\infty}{tR_a^f(t)\exp\left(-\int_0^{t}{R_a^f(z)\d z}\right) \d t}}{p_{rescue}},
 \end{align}
@@
 \section{Recurrence time}\label{sec:appendix_recurrence_time}
-We define the proliferation time $\tau_a^r$  to be the time it takes the population of mutant cancer cells to reach the initial tumor size $N$.
+We define the proliferation time $\tau_a^p$  to be the time it takes the population of mutant cancer cells to reach the initial tumor size $N$.
 The number of rescue lineages generated by the wildtype population is given by  \cref{eq:twosteplineage} (see \Cref{ExpectedNumberRescueLineages}):
@@
 \end{equation*}
-We plot \cref{eqmNs} and \cref{eqmNd} in Figure and compare with stochastic simulations and observe that our approximation are in agreement.
+We plot \cref{eqmNs} and \cref{eqmNd} in Figure \Cref{cdffig}B and compare with stochastic simulations and observe that our approximation are in agreement.
 
@@
 
-For the ratio of the threshold tumor sizes the $95\%$ confidence interval is obtained through bootstrapping in the following steps: (1) we simulate $p_{rescue}$ 100 times for both the case when $f=\tilde{u}\lambda_w/s$ and $f=0$; (2) we sample with replacement which we store in $S_0$ and  $S_f$; (4) for each element of $S_0$ we obtain $N_a^*=1/p_w$ using $p_w=-1/N_s\log\left(1-\bar{S}\right)$ where $\bar{S}$ is the mean of $S_0$ and $N_s$ is an arbitrary value of the initial population size we selected in order to calculate $p_{rescue}$; (5) for each element of $S_f$ we obtain $\tilde{N}_a^*=1/p_a$ using $p_a=-f/N_s\log\left(1-\bar{S}\right)$ where $\bar{S}_f$ is the mean of $S_f$ and $N_s$ is an arbitrary value of the initial population size we selected in order to calculate $p_{rescue}$; (6) we repeat steps (2)-(5) 100 times to obtain $\tilde{N}_a^*/N_a^*$ and we select the upper and lower limits such that $95\%$ of the values of $\tilde{N}_a^*/N_a^*$ lie in the interval given by the bounds.
+For the ratio of the threshold tumor sizes the $95\%$ confidence interval is obtained through bootstrapping in the following steps: (1) we simulate $p_{rescue}$ 100 times for both the case when $f=\tilde{u}\lambda_w/s$ and $f=0$; (2) we sample with replacement which we store in $S_f$ and  $S_0$; (4) for each element of $S_0$ we obtain $N_a^*=1/p_w$ using $p_w=-1/N_s\log\left(1-\bar{S}\right)$ where $\bar{S}$ is the mean of $S_0$ and $N_s$ is an arbitrary value of the initial population size we selected in order to calculate $p_{rescue}$; (5) for each element of $S_f$ we obtain $\tilde{N}_a^*=1/p_a$ using $p_a=-f/N_s\log\left(1-\bar{S}\right)$ where $\bar{S}_f$ is the mean of $S_f$ and $N_s$ is an arbitrary value of the initial population size we selected in order to calculate $p_{rescue}$; (6) we repeat steps (2)-(5) 100 times to obtain $\tilde{N}_a^*/N_a^*$ and we select the upper and lower limits such that $95\%$ of the values of $\tilde{N}_a^*/N_a^*$ lie in the interval given by the bounds.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \caption{\textbf{Aneuploidy increases the number of mutations which rescue the tumor.} Shown is the expected number of mutation, which will rescue the cancer cell population, produced through the evolutionary trajectory $wildtype \rightarrow mutant$ (blue line, \cref{eq:twosteplineagedirect}) or through the trajectory $wildtype \rightarrow aneuploid \rightarrow mutant$ (red line, \cref{eq:twosteplineage}).
+Dashed vertical red line represents the threshold tumor size above which evolutionary rescue is very likely through aneuploidy \cref{eq:N_a} and the dashed vertical blue line represents the threshold tumor size above which evolutionary rescue is very likely through direct mutation \cref{eq:N_m}.
 Parameters: $\lambda_w=0.1,\lambda_m=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}
@@
 \includegraphics[width=1\textwidth]{Figures/ProbvN_SGV_Plot.pdf}
-\caption{The probability of evolutionary rescue (i.e.
-the probability that the population does not go to extinction), $\presc$, as a function of the initial tumor size, $N$.
+\caption{The probability of evolutionary rescue (i.e., the probability that the population does not go to extinction), $\presc$, as a function of the initial tumor size, $N$.
 Dashed vertical line shows the threshold tumor size, above which the probability is very high.
@@
 The black line represents the case where a fraction $f=0\%$ of the initial tumor is aneuploid, the red line represents the case with $f=5\%$ and the green line represents the case with $f=50\%$.
-The dots represent simulations and the error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the fraction of simulations in which the tumor has adapted to the stress and $n=100$ is the number of simulations.
+The dots represent simulation results and the error bars represent $95\%$ confidence intervals ($p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the fraction of simulations in which the tumor has adapted to the stress and $n=100$ is the number of simulations).
 Parameters: $\lambda_w=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}

### 2024-05-15 EvolutionaryRescue1.tex
Add files via upload

@@
 \affil[3]{Department of Physics, Emory University, Atlanta, GA}
-\affil[*]{Corresponding author: Yoav Ram \textcolor{green}{(e-mail: yoavram@tauex.tau.ac.il)}}
+\affil[*]{Corresponding author: Yoav Ram (e-mail: yoavram@tauex.tau.ac.il)}
 
@@
 Most models focus on the probability that at most one mutation rescues the population.
-How multiple mutations contribute to the survival of the population is less explored, but \citet{wilson2017soft} have shown that evolutionary rescue is significantly enhanced by soft selective sweeps when multiple mutations contribute \textcolor{green}{towards evolutionary rescue}.
-Evolutionary rescue that requires two successive mutations (i.e., two steps) has been investigated using diffusion approximation by \citet{martin2013probability} \textcolor{green}{and tested the predictions of their model with data from experiments with yeast and bacterium}.
-
-Here we build on previous work on evolutionary rescue after a sudden environment change caused by the initiation of chemotherapy.
-We \textcolor{green}{estimate the} effect of aneuploidy on the probability of evolutionary rescue when it acts as an \textcolor{green}{evolutionary stepping stone} between the \textcolor{green}{sensitive} and the \textcolor{green}{resistant} cancer cells.
-We also estimate the mean time \textcolor{green}{until the} tumor cell population reaches its pre-treatment size.
-Given that aneuploidy is present in many tumors even before the onset of therapy \citep{lukow2021chromosomal}, we also consider the effect of \textcolor{green}{pre-treatment} standing genetic variation on the \textcolor{green}{evolutionary} dynamics.
+How multiple mutations contribute to the survival of the population is less explored, but \citet{wilson2017soft} have shown that evolutionary rescue is significantly enhanced by soft selective sweeps when multiple mutations contribute towards evolutionary rescue.
+Evolutionary rescue that requires two successive mutations (i.e., two steps) has been investigated using diffusion approximation by \citet{martin2013probability} who tested the predictions of their model with data from experiments with yeast and bacteria.
+
+Here we study evolutionary rescue after a sudden environment change caused by the initiation of chemotherapy.
+We estimate the effect of aneuploidy on the probability of evolutionary rescue when it acts as an evolutionary stepping stone towards resistant cancer cells.
+We also estimate the mean time until a tumor cell population reaches its pre-treatment size following drug therapy.
+Given that aneuploidy is present in many tumors even before the onset of therapy \citep{lukow2021chromosomal}, we also consider the effect of pre-treatment standing genetic variation on the evolutionary dynamics.
 Additionally, we are interested in the timescale of evolutionary rescue and the effect that aneuploidy has on the time necessary for the tumor to overcome drug therapy.
@@
 
-The \textcolor{green}{cell population} at time $t$ is represented by the triplet $\left(w_t,a_t,m_t\right)$.
-The following describes the events that may occur (right column), the rates at which they occur (middle column), and the effect these events have on the \textcolor{green}{population} (left column, see \Cref{figureAneuploidy}):
+The cell population at time $t$ is represented by the triplet $\left(w_t,a_t,m_t\right)$.
+The following describes the events that may occur (right column), the rates at which they occur (middle column), and the effect these events have on the population (left column, see \Cref{figureAneuploidy}):
 
@@
 We assume the mutation rate is $10^{-7}$ per gene per cell division \citep{loeb2001mutator} and since we assume that a single target gene confers resistance to the drug, we use $v=10^{-7}$ per cell division.
-\citet{bakker2023predicting} determined that the missegregation rate must be between $10^{-3}-10^{-2}$ per chromosome per cell division with the optimal value being $6.2\times10^{-3}$ per chromosome per cell division \textcolor{green}{(for $u\ll10^{-3}$ no adaptation happens and for $u\gg10^{-2}$ karyotypes will acquire lethal nullisomies)}.
+\citet{bakker2023predicting} determined that the missegregation rate must be between $10^{-3}-10^{-2}$ per chromosome per cell division with the optimal value being $6.2\times10^{-3}$ per chromosome per cell division (for $u\ll10^{-3}$ no adaptation occurs and for $u\gg10^{-2}$ cells will acquire lethal nullisomies where both pairs of chromosomes are lost).
 \citet{ippolito2021gene} observed that trisomy in Chr II and VI are most likely to confer increased resistance against the chemotherapeutic agent vemurafenib for A375 cells.
@@
 We obtain the fitness cost $s$ using the formula $s=\abs{\log\left[0.15/(1-0.15)\right]/24}\approx0.07$ per day \citep{chevin2011measuring}.
-As a result, the fraction of cancer cells with aneuploidy \textcolor{green}{of interest} is $f=2\times10^{-3}\times10^{-1}/0.07=0.28\%$ (i.e., $0.28\%$ of pre-treatment cancer cells have the aneuploidy \textcolor{green}{of interest}).
-
-We note that when we refer to wildtype cancer cells we include those cells that have any aneuploidy \textcolor{green}{other than} trisomy in Chr II and VI as those are the aneuploid cells that are hypothesized to have higher fitness in the environment altered by drugs such as vemurafenib.
+As a result, the fraction of cancer cells with aneuploidy of interest is $f=2\times10^{-3}\times10^{-1}/0.07=0.28\%$ (i.e., $0.28\%$ of pre-treatment cancer cells have the aneuploidy of interest).
+
+We note that when we refer to wildtype cancer cells we include those cells that have any aneuploidy other than trisomy in Chr II and VI as those are the aneuploid cells that are hypothesized to have higher fitness in the environment altered by drugs such as vemurafenib.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-In our \textcolor{green}{analysis}, we assume that lineages produced by cells from the initial population divide and die independently of each other.
-\textcolor{green}{However}, the cells will compete for resources, but we expect that this \textcolor{green}{competition} can be neglected because the drug will cause the cell density to rapidly drop below the carrying capacity at which \textcolor{green}{competition is most} important.
+In our analysis, we assume that lineages produced by cells from the initial population divide and die independently of each other.
+However, the cells will compete for resources, but we expect that this competition can be neglected because the drug will cause the cell density to rapidly drop below the carrying capacity at which competition is most important.
 To test this, we simulate a logistic growth model, with division and death rates given by:
@@
 where $K$ is the tumor carrying capacity. 
-The effective carrying capacity \textcolor{green}{in} this model is $K_e=K\Delta_a/\lambda_a\approx10^6$ for $K=10^8, \lambda_a=0.0901,\mu_a=0.09$, where we define the effective carrying capacity to be the population size at which the aneuploid division rate is equal to the aneuploid death rate.
+The effective carrying capacity in this model is $K_e=K\Delta_a/\lambda_a\approx10^6$ for $K=10^8, \lambda_a=0.0901,\mu_a=0.09$, where we define the effective carrying capacity to be the population size at which the aneuploid division rate is equal to the aneuploid death rate.
 
@@
 Otherwise, aneuploidy is frequent enough ($u\lambda_a > \max{\left(-\Delta_a, 1/T^*\right)}$) to affect the evolution of drug resistance. 
-The threshold tumor size can then be approximated by one of the following cases, depending on $\Delta_a T^*$, the change in the aneuploid log-population size during the critical time,
+The threshold tumor size can then be approximated by one of the following scenarios, depending on $\Delta_a T^*$, the change in the aneuploid log-population size during the critical time,
 \begin{equation}  \label{eq:N_a}
@@
 These approximations perform very well when compared to results of stochastic evolutionary simulations (\Cref{rescue_prob,rescue_threshold}).
-The first line describes the case in which aneuploid cells are still effectively killed by the treatment, but not as quickly as the wildtype. 
-In the second case, aneuploid cells are sufficiently resistant that the expected size of each aneuploid lineage is roughly 1.
-In both of these cases, aneuploidy increases the probability of rescue by slowing or halting the decrease of the \textcolor{green}{tumor} population \textcolor{green}{size}, allowing more opportunities for producing resistant mutants.
-In the third case, aneuploid cells are sufficiently resistant for the population to re-grow the tumor even without additional resistance mutations.
-Notably, in this case the mutant parameters ($v$, $\lambda_m$, and $\Delta_m$) do not affect $N_a^*$ beyond their effect on $T^*$.
-In all cases, $N_a^*$ is proportional to $1/u$ such that increasing the missegregation rate $u$ will decrease the threshold tumor size (\Cref{rescue_threshold}B).
-Furthermore, increasing the aneuploid growth rate $\Delta_a$ (which appears both in the terms and in the conditions), also reduces the threshold tumor size, with a sharp decrease around $\Delta_a=0$, but the effect is minor when $|\Delta_a|$ is small compared to $T^*$ as this would result in the second case where  $dN_a^*/d\Delta_a=0$ (\Cref{rescue_threshold}A).
+The first line describes the scenario in which aneuploid cells are still effectively killed by the treatment, but not as quickly as the wildtype. 
+In the second scenario, aneuploid cells are sufficiently resistant that the expected size of each aneuploid lineage is roughly 1.
+In both of these scenarios, aneuploidy increases the probability of rescue by slowing or halting the decrease of the tumor population size, allowing more opportunities for producing resistant mutants.
+In the third scenario, aneuploid cells are sufficiently resistant for the population to re-grow the tumor even without additional resistance mutations.
+Notably, in this scenario the mutant parameters ($v$, $\lambda_m$, and $\Delta_m$) do not affect $N_a^*$ beyond their effect on $T^*$.
+In all scenarios, $N_a^*$ is proportional to $1/u$ such that increasing the missegregation rate $u$ will decrease the threshold tumor size (\Cref{rescue_threshold}B).
+Furthermore, increasing the aneuploid growth rate $\Delta_a$ (which appears both in the terms and in the conditions), also reduces the threshold tumor size, with a sharp decrease around $\Delta_a=0$, but the effect is minor when $|\Delta_a|$ is small compared to $T^*$ as this would result in the second scenario where  $dN_a^*/d\Delta_a=0$ (\Cref{rescue_threshold}A).
 % TODO these are pretty straightforward - can you glean something more interesting from eq 4?
@@
 As expected, this ratio increases with the mutation rate $v$ and decreases with the aneuploidy rate $u$.
-In the first case, $\abs{\Delta_a}/\left(u\lambda_a\right)$ is  the ratio of the expected time for an aneuploid lineage to appear, $1/\left(u\lambda_a\right)$, and the expected time until that lineage disappears, $1/\abs{\Delta_a}$.
-In the third case, $\left(v \frac{\Delta_m}{\lambda_m}\right) / \left(u \frac{\Delta_a}{\lambda_a}\right)$ is the ratio of the rates of formation of resistant mutants that avoid extinction and partially resistant aneuploids that avoid extinction.
-In the second case, $\frac{1}{u}\left(v  \frac{\Delta_m}{\lambda_m}\right)^{1/2}=\sqrt{\frac{\Delta_a}{u\lambda_a}  v \frac{\Delta_m}{\lambda_m}  \left(u\frac{\Delta_a}{\lambda_a}\right)^{-1}}$, which is the geometric mean of the first and third cases.
-
-Interestingly, increasing both the aneuploid division rate, $\lambda_a$, and the aneuploid death rate, $\mu_a$, such that the growth rate $\Delta_a$ remains constant, leads to decreases in $T^*$, pushing the system to the second case.
-In this case, increasing the division rate $\lambda_a$ should also increase the mutation rate $v\lambda_a$ in aneuploid cells, as mutations mostly occur during division, so overall the threshold tumor size $N_a^*$ is unaffected by the division rate $\lambda_a$ (i.e., $d \lambda_a T^*/d\lambda_a = 0$).
+In the first scenario, $\abs{\Delta_a}/\left(u\lambda_a\right)$ is  the ratio of the expected time for an aneuploid lineage to appear, $1/\left(u\lambda_a\right)$, and the expected time until that lineage disappears, $1/\abs{\Delta_a}$.
+In the third scenario, $\left(v \frac{\Delta_m}{\lambda_m}\right) / \left(u \frac{\Delta_a}{\lambda_a}\right)$ is the ratio of the rates of formation of resistant mutants that avoid extinction and partially resistant aneuploids that avoid extinction.
+In the second scenario, $\frac{1}{u}\left(v  \frac{\Delta_m}{\lambda_m}\right)^{1/2}=\sqrt{\frac{\Delta_a}{u\lambda_a}  v \frac{\Delta_m}{\lambda_m}  \left(u\frac{\Delta_a}{\lambda_a}\right)^{-1}}$, which is the geometric mean of the first and third scenarios.
+
+Interestingly, increasing both the aneuploid division rate, $\lambda_a$, and the aneuploid death rate, $\mu_a$, such that the growth rate $\Delta_a$ remains constant, leads to decreases in $T^*$, pushing the system to the second scenario.
+In this scenario, increasing the division rate $\lambda_a$ should also increase the mutation rate $v\lambda_a$ in aneuploid cells, as mutations mostly occur during division, so overall the threshold tumor size $N_a^*$ is unaffected by the division rate $\lambda_a$ (i.e., $d \lambda_a T^*/d\lambda_a = 0$).
 Thus, if aneuploid cells rapidly die due to the drug but compensate by rapidly dividing, further increasing  the division rate will \emph{not} facilitate adaptation.
@@
 We can categorize tumors by their size: small tumors with size $N<N_a^*$ that are unlikely to survive treatment, intermediate tumors with size $N_a^* < N < N_m^*$ that rely on aneuploidy for evolutionary rescue, and large tumors with size $N > N_m^*$ that could overcome the effect of drug treatment even without aneuploidy.
-For the parameter values in \Cref{table1} with $\lambda_a=0.0899,\mu_w=0.14, u=10^{-2}, v=10^{-7}$, we are in the tolerant aneuploid case, and substituting in \cref{eq:N_a,eq:N_m}, we have $N_a^* \approx 4 \times 10^6$ and $N_m^* \approx 4 \times 10^7$.
+For the parameter values in \Cref{table1} with $\lambda_a=0.0899,\mu_w=0.14, u=10^{-2}, v=10^{-7}$, we are in the tolerant aneuploid scenario, and substituting in \cref{eq:N_a,eq:N_m}, we have $N_a^* \approx 4 \times 10^6$ and $N_m^* \approx 4 \times 10^7$.
 Hence, we obtain the ratio $N^*_a/N^*_m \approx 0.11$ (\cref{eq:N_ratio}), that is, aneuploidy reduces the threshold tumor size by approximately 89\%.
@@
 In the above we assumed that at the onset of drug treatment, the initial tumor consisted entirely of wildtype cells that are drug sensitive.
-However, aneuploidy is likely produced even before \textcolor{green}{the} onset of treatment at some rate $\tilde{u}$ and confers a deleterious fitness effect $s$ in the absence of the drug \citep{replogle2020aneuploidy,giam2015aneuploidy}.
+However, aneuploidy is likely produced even before the onset of treatment at some rate $\tilde{u}$ and confers a deleterious fitness effect $s$ in the absence of the drug \citep{replogle2020aneuploidy,giam2015aneuploidy}.
 Furthermore, the aneuploidy rate in the presence of drugs is likely higher than in their absence, $\tilde{u} < u$ \citep{wang2019molecular,mason2017functional}.
-But if the number of cells in the tumor $N$ is large (as expected if the tumor is treated with a drug), there may already be a fraction $f \approx \tilde{u}\lambda_w/s$ of aneuploid cells in the population (here we assume that the drug affects the wildtype death rate but not the division rate and therefore \textcolor{green}{we} use $\lambda_w$ for the wildtype division rate in the drug-free environment).
+But if the number of cells in the tumor $N$ is large (as expected if the tumor is treated with a drug), there may already be a fraction $f \approx \tilde{u}\lambda_w/s$ of aneuploid cells in the population (here we assume that the drug affects the wildtype death rate but not the division rate and therefore we use $\lambda_w$ for the wildtype division rate in the drug-free environment).
 
@@
 Comparing this approximation of $\tilde{N}^*_{a}/N^*_{a}$ to results of stochastic simulations, we find that the approximations perform very well (\Cref{rescue_denovo}). 
-Standing genetic variation will drive evolutionary rescue if wildtype cells \textcolor{green}{die rapidly} (growth rate $\Delta_w$ is very negative) due to a strong effect of the drug on sensitive cells, or if the aneuploidy cost in the drug-free environment, $s$, is small.
+Standing genetic variation will drive evolutionary rescue if wildtype cells die rapidly (growth rate $\Delta_w$ is very negative) due to a strong effect of the drug on sensitive cells, or if the aneuploidy cost in the drug-free environment, $s$, is small.
 In contrast, \emph{de-novo} aneuploid cells will have a greater contribution to rescue if the aneuploidy cost $s$ is large, the effect of the drug on sensitive cells is weak ($\Delta_w$ is close to zero), or if the drug induces the appearance of aneuploid cells ($u > \tilde u$).
@@
 However, if the aneuploid growth rate $\Delta_a$ increases, then evolutionary rescue is driven by standing genetic variation.
-For the parameter values of  $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, \tilde{u}=10^{-3}, v=10^{-7}$, we are in the tolerant aneuploid case and obtain the ratio $\tilde{N}^*_a/N^*_m \approx 0.9625$, which means that standing genetic variation reduces the threshold tumor size by approximately $4\%$.
+For the parameter values of  $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, \tilde{u}=10^{-3}, v=10^{-7}$, we are in the tolerant aneuploid scenario and obtain the ratio $\tilde{N}^*_a/N^*_m \approx 0.9625$, which means that standing genetic variation reduces the threshold tumor size by approximately $4\%$.
 Therefore, standing genetic variation does not drive evolution of drug resistance when compared to \emph{de-novo} aneuploidy, but it does offer a slight advantage when compared with direct mutation.
@@
 When the expected number of resistant lineages that avoid extinction is small, the expected recurrence time can be estimated by adding two terms: the \emph{mean evolutionary rescue time}, which is the waiting time for appearance of a resistant lineage that avoids extinction (conditioned on such an event occurring in the first place), and the \emph{mean proliferation time}, which is the expected time for that lineage to grow to $N$ cells.
-However, when the expected number of resistant lineages is large, the dynamics of \textcolor{green}{the} number of mutant cells is deterministic (i.e., it can be modeled by a system of ODEs, \cref{detODE}) and the mean recurrence time cannot be separated into the mean evolutionary rescue time and mean proliferation time because multiple mutant lineages contribute towards the mutant population size reaching the initial tumor size.
+However, when the expected number of resistant lineages is large, the dynamics of the number of mutant cells is deterministic (i.e., it can be modeled by a system of ODEs, \cref{detODE}) and the mean recurrence time cannot be separated into the mean evolutionary rescue time and mean proliferation time because multiple mutant lineages contribute towards the mutant population size reaching the initial tumor size.
 % TODO I dont like the word "behaves", please be specific.
@@
 
-Given that a fraction $f\approx 0.28\%$ of the initial cancer cell population is expected to be aneuploid even before \textcolor{green}{onset} of the drug \textcolor{green}{treatment}, we want to know whether the mean evolutionary rescue time is affected by the standing genetic variation.
+Given that a fraction $f\approx 0.28\%$ of the initial cancer cell population is expected to be aneuploid even before the onset of drug treatment, we want to know whether the mean evolutionary rescue time is affected by the standing genetic variation.
 We calculated the mean evolutionary rescue time with standing genetic variation $\tilde{\tau}_a$ (\cref{meantimet2SGV}) and compared our result with simulations in  \Cref{SGVEvolutionaryRescueTimeComplete}.
@@
 
-In Appendix \ref{sec:appendix_distribution_time} we calculated the probability that a \textcolor{green}{rescue} mutation has been generated by time $t$.
-This allows us to \textcolor{green}{examine} whether aneuploidy accelerates or delays evolutionary rescue.
-\textcolor{green}{We find that anenploidy significantly affects evolutionary rescue after} $1/\Delta_a\approx100$, days after which no more rescue mutations are generated through direct mutation (\Cref{cdffig}A).
-This shows that rescue mutations are generated \textcolor{green}{in aneuploid cells} at latter timescales then \textcolor{green}{in wildtype cells} and, \textcolor{green}{thus}, aneuploidy increases the \emph{window of opportunity} for evolutionary rescue.
+In Appendix \ref{sec:appendix_distribution_time} we calculated the probability that a rescue mutation has been generated by time $t$.
+This allows us to examine whether aneuploidy accelerates or delays evolutionary rescue.
+We find that anenploidy  accelerates evolutionary rescue after $1/\Delta_a\approx100$, days after which no more rescue mutations are generated through direct mutation (\Cref{cdffig}A).
+This shows that rescue mutations are generated in aneuploid cells at latter timescales then in sensitive cells and, thus, aneuploidy increases the \emph{window of opportunity} for evolutionary rescue.
 % WHAT is the purpose of this sentence?
@@
 
-We note that, for small and large tumors, when $N \ll N_a^*$ or $N \gg N_m^*$, the asymptotic expressions for the mean recurrence time are independent of the chromosome missegregation rate $u$, \textcolor{green}{and} therefore the rate at which the drug induces aneuploidy has no effect on the time necessary for the tumor to rebound to its initial size $N$.
+We note that, for small and large tumors, when $N \ll N_a^*$ or $N \gg N_m^*$, the asymptotic expressions for the mean recurrence time are independent of the chromosome missegregation rate $u$, and therefore the rate at which the drug induces aneuploidy has no effect on the time necessary for the tumor to rebound to its initial size $N$.
 % Remus the next paragraphs look like you wanted to find approximations so you did, you reference them here, but the results are buried in the technical text: WHAT IS IT THAT YOU WANT US TO LEARN from this?
@@
 \Cref{cdffig}B shows agreement between our approximations and stochastic simulations for various values of $N$.
-Additionally, we derive the distribution of the recurrence time for the case $N=10^6$ (small tumor), noting that the distribution is wide and right-skewed (\Cref{KMdistribution}).
+Additionally, we derive the distribution of the recurrence time for the scenario $N=10^6$ (small tumor), noting that the distribution is wide and right-skewed (\Cref{KMdistribution}).
 It is highly unlikely to observe the recurrence of tumors at times smaller then $\frac{1}{\Delta_m}\log\frac{\Delta_m-\Delta_w}{v\lambda_w}\approx1542$ days for the parameter values $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$ and independent of initial tumor size $N$~(\Cref{cdffig}B).
@@
 The detection time $\tau_a^{M}$ can be defined as the time for the tumor size to reach detection threshold $M$.
-We derive the mean \textcolor{green}{detection} time for $M=10^7$ in Appendix \ref{sec:appendix_recurrence_time}.
-We observe that  for small and intermediate sized tumors the effect of the detection size $M$ on $\tau_a^{M}$ is negligible when compared to the case where the detection size is \textcolor{green}{equal to the initial population size} $N$ (i.e., $\tau_a^r\approx\tau_a^{M}$ for $N<N_m^*$).
+We derive the mean detection time for $M=10^7$ in Appendix \ref{sec:appendix_recurrence_time}.
+We observe that  for small and intermediate sized tumors the effect of the detection size $M$ on $\tau_a^{M}$ is negligible when compared to the scenario where the detection size is equal to the initial population size $N$ (i.e., $\tau_a^r\approx\tau_a^{M}$ for $N<N_m^*$).
 % WHAT DOES THAT MEAN?
@@
 However, for large tumors the mean recurrence time to detection size $M$ decreases logarithmically with tumor size $N$ while $\tau_a^{r}$ is constant (\Cref{RecurrencePlot}).
-Additionally, for large tumors we have $M<N_m^*<N$ so the mean detection time is shorter for such tumors when compared to the mean recurrence time, \textcolor{green}{that is the resistant tumor can be detected before reaching its initial size}.
+Additionally, for large tumors we have $M<N_m^*<N$ so the mean detection time is shorter for such tumors compared to the mean recurrence time, that is, the resistant tumor can be detected before reaching its initial size.
 
@@
 ok
-Therefore, since only undetected secondary tumors are present at the time of the \textcolor{green}{treatment with} anti-cancer drug, we lack data on the size of the tumors and cannot compare the empirical distributions to our predictions.
+Therefore, since only undetected secondary tumors are present at the time of treatment with anti-cancer drug, we lack data on the size of the tumors and cannot compare the empirical distributions to our predictions.
 However, we expect  the variability of secondary tumors to average out across large cohorts of patients, thus we can use the mean recurrence time to compare with clinical data.
@@
 Instead, it acts as a \emph{stepping stone} through which the resistant mutant can appear more rapidly, given that the aneuploid cell population size declines slower then that of the sensitive cell population (\Cref{sampleTrajectories}).
-In this case, aneuploidy provides two benefits.
+In this scenario, aneuploidy provides two benefits.
 First, it delays the extinction of the population, providing more time for appearance of the resistance mutation.
@@
 However, aneuploidy can have a crucial role in the evolutionary rescue of secondary tumors, in which the number of sensitive cells may be below the detection threshold of $\sim10^7$  \citep{bozic2013evolutionary}, and this can have an impact on the recurrence of cancer after the resection of the primary tumor  through secondary tumors, which are too small to be detected and for which chemotherapy is employed to prevent  cancer relapse.
-\textcolor{green}{These secondary tumors} are estimated to cause the majority of cancer-related deaths~\citep{chaffer2011perspective}.
+These secondary tumors are estimated to cause the majority of cancer-related deaths~\citep{chaffer2011perspective}.
 The importance of aneuploidy in the evolutionary rescue of secondary tumors is reinforced by the fact that metastases have been shown to have a chromosome missegregation rate two to three orders of magnitude higher then that of primary tumors~\citep{kimmel2023intra}.
@@
 Given the fact that the mean time for secondary tumors to overcome chemotherapy can be of the order of 1,000 days (\Cref{MeanTimeGrowthAneuploidyPlot}A), aneuploidy can explain the reappearance of cancer even after initial remission.
-The theoretical prediction for the mean rescue time for tumors smaller then  $10^8$ cells is greater then 4 years, consistent with previous estimates of the recurrence time of tumors after resection~\citep{avanzini2019cancer}.
-\textcolor{green}{We found that aneuploidy complements evolutionary rescue through direct mutation because it generate rescue mutations only after the wildtype population had died out and direct mutation is no longer a viable option for evolutionary rescue (\Cref{cdffig}A).}
-
-We hypothesized that \emph{standing \textcolor{green}{genetic} variation} (the existence of a subpopulation of aneuploid cancer cells before the onset of therapy) can facilitate evolutionary rescue by reducing the waiting time for the appearance of aneuploid cells.
+The theoretical prediction for the mean rescue time for tumors smaller then  $10^8$ cells is greater than 4 years, consistent with previous estimates of the recurrence time of tumors after resection~\citep{avanzini2019cancer}.
+We found that aneuploidy complements evolutionary rescue through direct mutation because it generates rescue mutations mostly after the wildtype population had died out and direct mutation is no longer a viable option for evolutionary rescue (\Cref{cdffig}A).
+
+We hypothesized that \emph{standing genetic variation} (the existence of a subpopulation of aneuploid cancer cells before the onset of therapy) can facilitate evolutionary rescue by reducing the waiting time for the appearance of aneuploid cells.
 We found that a drug or combination of drugs that reduces the wildtype growth rate and does not significantly increase the chromosome missegregation rate is more likely to cause evolutionary rescue to occur through direct mutation (\cref{eq:Nsgv}).
 Furthermore, we found that for reasonable parameter values evolutionary rescue is more likely to occur through \emph{de-novo} aneuploidy (\Cref{rescue_denovo}).
-If the fraction of tumor cells that have the \textcolor{green}{beneficial} aneuploidy is $f\gg\frac{u\lambda_w}{\abs{\Delta_w}}\approx2.5\%$, then evolutionary rescue is more likely to occur via standing variation, rather then through \emph{de-novo} aneuploidy.
-In this case, evolutionary rescue likely occurs via aneuploid cells that acquire a resistance mutation and the probability of evolutionary rescue declines exponentially with initial tumor size  (the probability of evolutionary rescue is given by $1-\exp\left(-N/\tilde{N}_a^*\right)$, see \Cref{rescue_prob_sgv}).
+If the fraction of tumor cells that have the beneficial aneuploidy is $f\gg\frac{u\lambda_w}{\abs{\Delta_w}}\approx2.5\%$, then evolutionary rescue is more likely to occur via standing variation, rather then through \emph{de-novo} aneuploidy.
+In this scenario, evolutionary rescue likely occurs via aneuploid cells that acquire a resistance mutation and the probability of evolutionary rescue declines exponentially with initial tumor size  (the probability of evolutionary rescue is given by $1-\exp\left(-N/\tilde{N}_a^*\right)$, see \Cref{rescue_prob_sgv}).
 
@@
 Additionally, our model predictions can be tested with data from patients.
-Given the type of cancer and how \textcolor{green}{beneficial} aneuploidy is,  our model can predict the probability for cancer relapse after treatment.
+Given the type of cancer and how beneficial aneuploidy is,  our model can predict the probability for cancer relapse after treatment.
 
 \paragraph{Directions of future research}
- Our model can be extended to understand evolutionary rescue in different biological contexts, for example, how yeast \textcolor{green}{populations under} stress overcome extinction via aneuploidy~\citep{pompei2023fitness,kohanovski2024aneuploidy}.
+ Our model can be extended to understand evolutionary rescue in different biological contexts, for example, how yeast populations under stress overcome extinction via aneuploidy~\citep{pompei2023fitness,kohanovski2024aneuploidy}.
 Additionally, we did not account for the heterogeneity of aneuploidy as not all aneuploid lineages have the same growth rate $\Delta_a$ as assumed in our model~\citep{avecilla2023copy,yang2021fitness}.
 Such heterogeneity can be accounted for  by  sampling the aneuploidy growth or death rates from a distribution~\citep{martin2013probability}.
-\textcolor{green}{Furthermore, tumor heterogeneity may be important for genetic variation.}
 
@@
 
-An additional \textcolor{green}{extension} of our model \textcolor{green}{could incorporate} a wider range of parameters, however parameter values for different types of tumors are difficult to obtain.
+An additional extension of our model could incorporate a wider range of parameters, however parameter values for different types of tumors are difficult to obtain.
 
@@
 \textbf{Sample trajectories of the different genotype frequencies.}
-(A) \textcolor{green}{Without aneuploidy} ($u=0$0 evolutionary rescue is possible through direct mutation and in most cases the tumor will \textcolor{green}{become extinct due to} the drug.
-(B) When the aneuploidy \textcolor{green}{cells are tolerant} ($\Delta_a<0$) we observe that, similar to case (A), direct mutation is the only \textcolor{green}{path for} evolutionary rescue.
-(C) \textcolor{green}{When aneuploidy cells are stationary} ($\Delta_a\approx0$), we observe the appearance of aneuploid lineages even after the wildtype population has gone extinct \textcolor{green}{thus showing that stationary aneuploidy increases the probability of evolutionary rescue}.
-(D) \textcolor{green}{When the aneuploid cells are partially resistant} ($\Delta_a>0$) the tumor is rescued by the aneuploid cell population.
-Each plot shows 10 simulations of the number of \textcolor{green}{wildtype, aneuploid and mutant cells} $\left(w_t,a_t,m_t\right)$ \textcolor{green}{over} time $t$ \textcolor{green}{with} the following parameter values: $\lambda_w=0.1,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, v=10^{-7},N=10^7$.
+(A) Without aneuploidy ($u=0$) evolutionary rescue is possible through direct mutation and in most scenarios the tumor will become extinct due to the drug.
+(B) When aneuploid cells are tolerant ($\Delta_a<0$), we observe that, similar to A, direct mutation is the only path for evolutionary rescue.
+(C) When aneuploid cells are stationary ($\Delta_a\approx0$), we observe the appearance of aneuploid lineages even after the wildtype population has gone extinct, thus showing that stationary aneuploidy increases the probability of evolutionary rescue.
+(D) When aneuploid cells are partially resistant ($\Delta_a>0$), the tumor is rescued by the aneuploid cell population.
+Each plot shows 10 simulations of the number of wildtype, aneuploid and mutant cells $\left(w_t,a_t,m_t\right)$ over time $t$ with the following parameter values: $\lambda_w=0.1,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, v=10^{-7},N=10^7$.
 For (A) we set $u=0$, for (B) $\lambda_a=0.065,u=10^{-2}$, for (C) $\lambda_a=0.0899,u=10^{-2}$ and for (D) $\lambda_a=0.095,u=10^{-2}$.
@@
 Red line: stationary aneuploidy ($u=10^{-2}, \lambda_a=0.08999$).
-Green line represents the case with partially resistant aneuploidy ($u=10^{-2}, \lambda_a=0.095$).
+Green line represents the scenario with partially resistant aneuploidy ($u=10^{-2}, \lambda_a=0.095$).
 Dots for simulations and the error bars for $95\%$ confidence interval ($p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the fraction of simulations in which the tumor has been rescued and $n=100$ is the number of simulations).
@@
 \caption{
-\textbf{\textcolor{green}{The effect of aneuploidy on tumor threshold size.}}
+\textbf{The effect of aneuploidy on tumor threshold size.}
 (A) The threshold tumor size $N_a^*$ as a function of the aneuploid growth rate $\Delta_a$.
@@
 When aneuploid growth rate is close to or higher than zero, aneuploidy decreases the threshold tumor size, thereby facilitating evolutionary rescue.
-The inset highlights the case when aneuploidy cells are \textcolor{green}{stationary}.
+The inset highlights the scenario when aneuploid cells are stationary.
 Red dots for simulations and error bars for the $95\%$ confidence intervals obtained with bootstrap (Appendix G).
@@
 \caption{
-\textbf{\textcolor{green}{De-novo} standing genetic variation facilitates evolutionary rescue of cancer under drug treatment.}
+\textbf{Standing genetic variation facilitates evolutionary rescue of cancer.}
 (A)  Ratio of threshold tumor size $\tilde{N}_a^*$ when a fraction $\frac{\tilde{u}\lambda_w}{s}$ is aneuploid at the start of treatment and $N_a^*$ as a function of the wildtype growth rate $\Delta_w$.
-Standing genetic variation will drive adaptation to the drug if \textcolor{green}{the wildtype population is rapidly declining} ($\Delta_w\ll0$) due to a stronger effect of the drug on sensitive cells.
-Red dots represents simulation \textcolor{green}{results} and the error bars represent the $95\%$ confidence intervals obtained with bootstrap (Appendix G).
+Standing genetic variation will drive adaptation to the drug if the wildtype population is rapidly declining ($\Delta_w\ll0$) due to a stronger effect of the drug on sensitive cells.
+Red dots represents simulation results and the error bars represent the $95\%$ confidence intervals obtained with bootstrap (Appendix G).
 Parameters: $\lambda_w=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_a=0.09,\mu_m=0.09,\tilde{u}=10^{-3},u=10^{-2}, v=10^{-7}$.
@@
 \emph{De-novo} aneuploids will have larger contribution to the appearance of drug resistance if the drug induces the appearance of aneuploid cells ($u \gg \tilde u$).
-Red dots represents  simulation \textcolor{green}{results} and the error bars represent the $95\%$ confidence intervals obtained with bootstrap (Appendix G).
+Red dots represents  simulation results and the error bars represent the $95\%$ confidence intervals obtained with bootstrap (Appendix G).
 Parameters: $\lambda_w=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09,\tilde{u}=10^{-3}, v=10^{-7}$.
@@
 \end{subfigure}
-\caption{\textbf{Aneuploidy \textcolor{green}{extends the window of opportunity for evolutionary rescue.}}
+\caption{\textbf{Aneuploidy extends the window of opportunity for evolutionary rescue.}
 (A) The probability that a successful mutant has not appeared by time $t$.
@@
 Black line: no aneuploidy ($u=0$).
-Aneuploidy starts to play an important role in \textcolor{green}{rescuing} the cancer cell population \textcolor{green}{as the wildtype population becomes extinct}.
+Aneuploidy starts to play an important role in rescuing the tumor cell population as the wildtype population becomes extinct.
 Markers represent simulation results and the error bars represent $95\%$ confidence interval ($p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the fraction of simulations in which a successful mutant has not been generated and $n=100$ is the number of simulations).
@@
 \end{align*}
-where $T^* = (4 v \lambda_a^2 \Delta_m/\lambda_m)^{-1/2}$ and in the second inequality we used the fact that $\Delta_a^2 > 4\lambda_a^2 v p_m$. In this case adaptation is through direct mutation and:
+where $T^* = (4 v \lambda_a^2 \Delta_m/\lambda_m)^{-1/2}$ and in the second inequality we used the fact that $\Delta_a^2 > 4\lambda_a^2 v p_m$.
+In this scenario adaptation is through direct mutation and:
 \begin{equation*}
@@
 
-Paradoxically, we observe from \Cref{MeanTimeGrowthAneuploidyPlot} that the mean time of a rescue mutation to appear is significantly shorter for the case when $u=0$ when compared to the case $u>0$, however this can be explained by the fact this mean time is conditioned on evolutionary rescue and, as a result, aneuploidy increase the \emph{window of opportunity} in which a rescue mutation could appear thus increasing the mean time as well (\Cref{sampleTrajectories}).
+Paradoxically, we observe from \Cref{MeanTimeGrowthAneuploidyPlot} that the mean time of a rescue mutation to appear is significantly shorter for the scenario when $u=0$ when compared to the scenario $u>0$, however this can be explained by the fact this mean time is conditioned on evolutionary rescue and, as a result, aneuploidy increase the \emph{window of opportunity} in which a rescue mutation could appear thus increasing the mean time as well (\Cref{sampleTrajectories}).
 
@@
 
-This helps us distinguish between two cases for the proliferation time. Firstly, when we have at most one lineages which rescues the cancer cell population:
+This helps us distinguish between two scenarios for the proliferation time. Firstly, when we have at most one lineages which rescues the cancer cell population:
 \begin{equation*}
@@
 \end{align}
-The factor of $p_m$ in the second term of \cref{meanproliferationtime} is due to the fact that the lineage is conditioned to survive genetic drift and the time to reach $N$ is shorter then the case without this property.
-
-The second case is when the wildtype population produces a large number of rescue lineages in a short period of time. This is given by the condition:
+The factor of $p_m$ in the second term of \cref{meanproliferationtime} is due to the fact that the lineage is conditioned to survive genetic drift and the time to reach $N$ is shorter then the scenario without this property.
+
+The second scenario is when the wildtype population produces a large number of rescue lineages in a short period of time. This is given by the condition:
 \begin{align*}
@@
 We observe that for $N=10^7$ our formula overestimates the probability that the mutant population will be smaller then $N$ at time $t$.
-This can be explained by the fact that $N=10^7$ is an intermediary case where the wildtype population produces a number of rescue lineages that is greater then one but still sufficiently small such that stochasticity plays an important role in the population dynamics.
-As a result, the number of mutant cancer cells will reach $N$ faster then the case with a single mutant lineage.
-Additionally, we observe from \Cref{cdffig}B that the probability of the mutant cell population reaching size $N$ is approximately zero before time $\tau_a^r$ which is the recurrence time for the deterministic case.
-This can be explained as follows: in the deterministic case there is a sufficient number of lineages produced such that there exists a lineage where each descendant will only reproduce and not die; the time it takes for this lineage to reach $N$ is the lower bound for the time of all other lineages to reach $N$ and this time cannot be smaller then $\tau_a^r$ by definition.
-Given that for small values of $N$ we expect that at most a single lineage will rescue the tumor, this lineage cannot reach $N$ before $\tau_a^r$ for the deterministic case \cref{eq:t2det}.
+This can be explained by the fact that $N=10^7$ is an intermediary scenario where the wildtype population produces a number of rescue lineages that is greater then one but still sufficiently small such that stochasticity plays an important role in the population dynamics.
+As a result, the number of mutant cancer cells will reach $N$ faster then the scenario with a single mutant lineage.
+Additionally, we observe from \Cref{cdffig}B that the probability of the mutant cell population reaching size $N$ is approximately zero before time $\tau_a^r$ which is the recurrence time for the deterministic scenario.
+This can be explained as follows: in the deterministic scenario there is a sufficient number of lineages produced such that there exists a lineage where each descendant will only reproduce and not die; the time it takes for this lineage to reach $N$ is the lower bound for the time of all other lineages to reach $N$ and this time cannot be smaller then $\tau_a^r$ by definition.
+Given that for small values of $N$ we expect that at most a single lineage will rescue the tumor, this lineage cannot reach $N$ before $\tau_a^r$ for the deterministic scenario \cref{eq:t2det}.
 
@@
 which we plot in \Cref{KMdistribution} and compare with simulations.
-We note that in the case $N\gg\abs{\Delta_w}\abs{\Delta_a}/uv\lambda_w\lambda_ap_m$ the distribution becomes the Dirac $\delta$-function~\citep{barton1989elements}.
+We note that in the scenario $N\gg\abs{\Delta_w}\abs{\Delta_a}/uv\lambda_w\lambda_ap_m$ the distribution becomes the Dirac $\delta$-function~\citep{barton1989elements}.
 %%%
@@
 
-For the ratio of the threshold tumor sizes the $95\%$ confidence interval is obtained through bootstrapping in the following steps: (1) we simulate $p_{rescue}$ 100 times for both the case when $f=\tilde{u}\lambda_w/s$ and $f=0$; (2) we sample with replacement which we store in $S_f$ and  $S_0$; (4) for each element of $S_0$ we obtain $N_a^*=1/p_w$ using $p_w=-1/N_s\log\left(1-\bar{S}\right)$ where $\bar{S}$ is the mean of $S_0$ and $N_s$ is an arbitrary value of the initial population size we selected in order to calculate $p_{rescue}$; (5) for each element of $S_f$ we obtain $\tilde{N}_a^*=1/p_a$ using $p_a=-f/N_s\log\left(1-\bar{S}\right)$ where $\bar{S}_f$ is the mean of $S_f$ and $N_s$ is an arbitrary value of the initial population size we selected in order to calculate $p_{rescue}$; (6) we repeat steps (2)-(5) 100 times to obtain $\tilde{N}_a^*/N_a^*$ and we select the upper and lower limits such that $95\%$ of the values of $\tilde{N}_a^*/N_a^*$ lie in the interval given by the bounds.
+For the ratio of the threshold tumor sizes the $95\%$ confidence interval is obtained through bootstrapping in the following steps: (1) we simulate $p_{rescue}$ 100 times for both the scenario when $f=\tilde{u}\lambda_w/s$ and $f=0$; (2) we sample with replacement which we store in $S_f$ and  $S_0$; (4) for each element of $S_0$ we obtain $N_a^*=1/p_w$ using $p_w=-1/N_s\log\left(1-\bar{S}\right)$ where $\bar{S}$ is the mean of $S_0$ and $N_s$ is an arbitrary value of the initial population size we selected in order to calculate $p_{rescue}$; (5) for each element of $S_f$ we obtain $\tilde{N}_a^*=1/p_a$ using $p_a=-f/N_s\log\left(1-\bar{S}\right)$ where $\bar{S}_f$ is the mean of $S_f$ and $N_s$ is an arbitrary value of the initial population size we selected in order to calculate $p_{rescue}$; (6) we repeat steps (2)-(5) 100 times to obtain $\tilde{N}_a^*/N_a^*$ and we select the upper and lower limits such that $95\%$ of the values of $\tilde{N}_a^*/N_a^*$ lie in the interval given by the bounds.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 Blue dashed line represents the probability of evolutionary rescue as a function of $N$ without aneuploidy ($u=0$).
-The black line represents the case where a fraction $f=0\%$ of the initial tumor is aneuploid, the red line represents the case with $f=5\%$ and the green line represents the case with $f=50\%$.
+The black line represents the scenario where a fraction $f=0\%$ of the initial tumor is aneuploid, the red line represents the scenario with $f=5\%$ and the green line represents the scenario with $f=50\%$.
 The dots represent simulation results and the error bars represent $95\%$ confidence intervals ($p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the fraction of simulations in which the tumor has adapted to the stress and $n=100$ is the number of simulations).

### 2024-06-10 EvolutionaryRescue1.tex
Add files via upload

@@
 % Title page
-\title{The role of aneuploidy in the evolution of cancer drug resistance}
+\title{Modeling the contribution of aneuplouidy to evolutionary rescue following anticancer drug exposure}
 % Authors
@@
 One evolutionary mechanism by which a population of cancer cells can adapt to chemotherapy is aneuploidy. 
-Aneuploid cancer cells can have higher fitness in an environment altered by anti-cancer drugs, e.g., because of incomplete pathways targeted by the drugs. 
+Aneuploid cancer cells can have higher fitness in an environment altered by anti-cancer drugs, e.g., because aneuploidy disrupts the pathways normally targeted by the drugs. 
 Indeed, aneuploidy is highly prevalent in tumors, and moreover, some anti-cancer drugs fight cancer by increasing chromosomal instability.
-Here, we examine how aneuploidy impacts the fate of a population of cancer cells.
-We use multi-type branching processes to approximate the probability that a tumor survives drug treatment as a function of the initial tumor size, the rates at which aneuploidy and other beneficial mutations occur, and the growth rates of the sensitive and resistant cells.
-Additionally, we investigate the effect on the probability of evolutionary rescue of having a fraction of aneuploid tumor cells before the onset of therapy.
-We also approximate the mean recurrence time for the tumor to revert back to its initial size following treatment.
-We find that aneuploidy can play an important role in the relapse of smaller secondary tumors who have not yet been detected.
+Here, we model how aneuploidy impacts the fate of a population of cancer cells.
+We use multi-type branching processes to approximate the probability of a tumor to survive drug treatment as a function of the initial tumor size, the rates at which aneuploidy and other beneficial mutations occur, and the growth rates of the sensitive and resistant cells.
+Additionally, we investigate the effect of the fraction of pre-existent aneuploid cells on the probability of evolutionary rescue  following therapy.
+We also approximate the mean recurrence time for the tumor to revert back to its initial size following treatment, if such evolutionary rescue has occurred.
+We propose that aneuploidy can play an important role in the relapse of smaller secondary tumors.
 % TODO missing something here... ok
@@
 
-Keywords: whole-chromosome duplication, evolutionary model, adaptive evolution, cancer, drug resistance, chromosome instability
+Keywords: aneuploidy, evolutionary model, adaptive evolution, cancer, drug resistance, chromosome instability
 
@@
 One hypothesized factor is aneuploidy, where cells are characterized by an imbalanced karyotype and chromosome copy number alterations \citep{schukken2018cin}.
-Aneuploidy is caused by chromosomal instability, the mitotic process in which cells suffer from chromosome mis-segregation.
-Importantly, aberrations in chromosome copy number allow cancer cells to survive under stressful conditions such as drug therapy \citep{lukow2021chromosomal,rutledge2016selective}.
+Aneuploidy is caused by chromosomal instability, a high rate of chromosome mis-segregation during mitosis.
+Importantly, altered  copy number of chromosomes and chromosome-arms allow cancer cells to survive under stressful conditions such as drug therapy \citep{lukow2021chromosomal,rutledge2016selective,ippolito2021gene}.
 Indeed, cancer cells are often aneuploid and aneuploidy is associated with poor patient outcomes \citep{ben2020context,smith2018systematic}.
 
- \citet{ippolito2021gene} induced aneuploidy in cancer cell lines by exposing them to reversine, a small-molecule inhibitor of the mitotic kinase Mpsi1 and then to chemotherapeutic agents such as vemurafenib.
-Reversine-treated cells had higher proliferation rate in the environment altered by anti-cancer drugs compared to wildtype cancer cells.
- Similarly, \citet{lukow2021chromosomal} induced aneuploidy in cancer cells and observed that such cells have an advantage compared to wildtype cells during chemotherapy, despite having lower fitness before the onset of chemotherapy.
+ \citet{ippolito2021gene} induced aneuploidy in cancer cell lines by exposing them to reversine, a small-molecule inhibitor of the mitotic kinase Mpsi1 and then to anticancer agents such as vemurafenib.
+Reversine-treated cells had a higher proliferation rate following drug exposure compared to wildtype cancer cells, due to selection of specific beneficial karyotypes.
+ Similarly, \citet{lukow2021chromosomal} induced aneuploidy in cancer cells and observed that such cells have an advantage compared to wildtype cells during chemotherapy treatment, despite having lower fitness before the onset of chemotherapy.
 One proposed mechanism through which aneuploidy is able to confer resistance to chemotherapeutics is by antagonizing cell division, which prevents the drugs from damaging DNA and microtubules \citep{replogle2020aneuploidy}.
-
-An important aspect of aneuploidy is the rate with which cells become aneuploid, which is several orders of magnitude higher than the beneficial mutation rate \citep{bakker2023predicting}.
+Other mechanisms are the selection for specific karyotypes that lead to reduced drug metabolism \citep{ippolito2021gene}, and elevated levels of DNA damage repair due to higher basal level of DNA damage \citep{zerbib2023human}.
+
+An important aspect of aneuploidy is that the rate with which cells become aneuploid, that is the rate of chromosome mis-segregation, is several orders of magnitude higher than mutation rates \citep{bakker2023predicting}.
 Consequently, a cell exposed to a stress such as chemotherapeutic drugs can acquire aneuploidy faster compared to acquiring a mutation.
-Morover several proposed anti-cancer drugs elevate the rate of mis-segregation in order to fight cancer cells~\citep{lee2016effects}.
+Moreover several proposed anti-cancer drugs elevate the rate of mis-segregation in order to fight cancer cells~\citep{lee2016effects}, as an extremely high chromosome mis-segregation rate is not compatible with cell survival and proliferation.
 
@@
 Adaptation is a race against time as the population size decreases in the new environment~\citep{tanaka2022surviving}.
-\emph{Evolutionary rescue} is the process where the population acquires a trait that increases fitness in the new environment such that extinction is averted.
+\emph{Evolutionary rescue} is the process by which the population acquires a trait that increases fitness in the new environment such that extinction is averted.
 It is mathematically equivalent to the problem of crossing of fitness valley \citep{weissman2009rate,weissman2010rate}.
@@
 They found that even when the drug resistant type is barely viable the epimutations have the effect of guaranteeing evolutionary rescue.
-Evolutionary rescue in one step where an initially declining population, after a sudden environmental change, has to acquire a mutation in order to survive extinction has been studied by \citet{orr2008population,orr2014population}.
-They analyzed a model where the mutant strain is present in small numbers at the onset of therapy and concluded that this can significantly enhance the probability that the population will survive.
+Evolutionary rescue in a single step, where an initially declining population, has to acquire a mutation in order to survive extinction after a sudden environmental change, has been studied by \citet{orr2008population,orr2014population}.
+They analyzed a model where the mutant strain is present in small numbers at the onset of therapy and concluded that this can significantly enhance the probability that the population would survive.
 
@@
 How multiple mutations contribute to the survival of the population is less explored, but \citet{wilson2017soft} have shown that evolutionary rescue is significantly enhanced by soft selective sweeps when multiple mutations contribute towards evolutionary rescue.
-Evolutionary rescue that requires two successive mutations (i.e., two steps) has been investigated using diffusion approximation by \citet{martin2013probability} who tested the predictions of their model with data from experiments with yeast and bacteria.
+Evolutionary rescue that requires two successive mutations (i.e., two steps) has been investigated using diffusion approximation by \citet{martin2013probability} who tested the predictions of their model with data from experiments in yeast and bacteria.
 
 Here we study evolutionary rescue after a sudden environment change caused by the initiation of chemotherapy.
-We estimate the effect of aneuploidy on the probability of evolutionary rescue when it acts as an evolutionary stepping stone towards resistant cancer cells.
+We assume that the affected cancer is melanoma, the chemotherapeutic agent is vemurfenib and that resistance is conferered by either trisomy in Chromosome 2 or trisomy in Chromosome 6.
+We estimate the effect of aneuploidy on the probability of evolutionary rescue when it acts as an evolutionary stepping stone (i.e.
+intermediary) towards resistant cancer cells.
 We also estimate the mean time until a tumor cell population reaches its pre-treatment size following drug therapy.
-Given that aneuploidy is present in many tumors even before the onset of therapy \citep{lukow2021chromosomal}, we also consider the effect of pre-treatment standing genetic variation on the evolutionary dynamics.
+Given that aneuploidy is present in many tumors already before the onset of therapy \citep{lukow2021chromosomal,ben2020context}, we also consider the effect of pre-treatment standing genetic variation on the evolutionary dynamics.
 Additionally, we are interested in the timescale of evolutionary rescue and the effect that aneuploidy has on the time necessary for the tumor to overcome drug therapy.
+We find that aneuploidy increases the \emph{window of opportunity} when evolutionary rescue is able to prevent extinction.
 
@@
 
-When simulations are slow (e.g., due to large population size), we use $\tau$-leaping \citep{gillespie2001approximate}, where we assume that the change in the number of cells of genotype $k$ in a fixed time interval $\Delta t$ is Poisson distributed with mean $\nu_k \Delta t$.
+When simulations are slow (e.g., due to large population size) with runtimes in the order of days, we use $\tau$-leaping \citep{gillespie2001approximate}, where we assume that the change in the number of cells of genotype $k$ in a fixed time interval $\Delta t$ is Poisson distributed with mean $\nu_k \Delta t$.
 If the change in the number of cells is negative and larger than the subpopulation size then the subpopulation size is updated to be zero.
@@
 
-We note that when we refer to wildtype cancer cells we include those cells that have any aneuploidy other than trisomy in Chr II and VI as those are the aneuploid cells that are hypothesized to have higher fitness in the environment altered by drugs such as vemurafenib.
+We note that when we refer to wildtype cancer cells we include those cells that have any aneuploidy other than trisomy in Chr 2 and 6 as those are the aneuploid cells that are hypothesized to have higher fitness in the environment altered by drugs such as vemurafenib.
+
+All of the parameters discussed above are shown in \Cref{table1}.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-Interestingly, increasing both the aneuploid division rate, $\lambda_a$, and the aneuploid death rate, $\mu_a$, such that the growth rate $\Delta_a$ remains constant, leads to decreases in $T^*$, pushing the system to the second scenario.
+Interestingly, increasing both the aneuploid division rate, $\lambda_a$, and the aneuploid death rate, $\mu_a$, such that the growth rate $\Delta_a$ remains constant, leads to a decrease in $T^*$, pushing the system to the second scenario.
 In this scenario, increasing the division rate $\lambda_a$ should also increase the mutation rate $v\lambda_a$ in aneuploid cells, as mutations mostly occur during division, so overall the threshold tumor size $N_a^*$ is unaffected by the division rate $\lambda_a$ (i.e., $d \lambda_a T^*/d\lambda_a = 0$).
@@
 
-Given that a fraction $f\approx 0.28\%$ of the initial cancer cell population is expected to be aneuploid even before the onset of drug treatment, we want to know whether the mean evolutionary rescue time is affected by the standing genetic variation.
+Given that a fraction $f\approx 0.28\%$ of the initial cancer cell population is expected to have the beneficial aneuploidy even before the onset of drug treatment, we want to know whether the mean evolutionary rescue time is affected by the standing genetic variation.
 We calculated the mean evolutionary rescue time with standing genetic variation $\tilde{\tau}_a$ (\cref{meantimet2SGV}) and compared our result with simulations in  \Cref{SGVEvolutionaryRescueTimeComplete}.
@@
 
-Most clinical trials report data on the distribution of recurrence time measured from the time of surgery~\citep{avanzini2019cancer} with drug therapy usually following after.
+Most clinical trials report data on the distribution of recurrence time measured from the time of surgery~\citep{avanzini2019cancer} with drug therapy usually following after (i.e., adjuvant therapy).
 % WHAT DOES "varying" mean here?
@@
 Therefore, since only undetected secondary tumors are present at the time of treatment with anti-cancer drug, we lack data on the size of the tumors and cannot compare the empirical distributions to our predictions.
-However, we expect  the variability of secondary tumors to average out across large cohorts of patients, thus we can use the mean recurrence time to compare with clinical data.
+However, we expect  the variability of secondary tumors to average out across large cohorts of patients, thus we can use the mean recurrence time to compare with clinical data on cancer relapse.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 We find that aneuploidy can have a significant effect on evolutionary rescue as it reduces the threshold tumor size by at least an order-of-magnitude even when aneuploidy only provides tolerance (\Cref{rescue_prob}).
-Interestingly, aneuploidy is unlikely to contribute to evolutionary rescue in primary tumors in which the number of cells is large enough (i.e., $N\gg N_m^*\approx 4\times10^7$) for the appearance of resistant mutation in sensitive cells before these cells become extinct (\Cref{rescue_prob}).
-However, aneuploidy can have a crucial role in the evolutionary rescue of secondary tumors, in which the number of sensitive cells may be below the detection threshold of $\sim10^7$  \citep{bozic2013evolutionary}, and this can have an impact on the recurrence of cancer after the resection of the primary tumor  through secondary tumors, which are too small to be detected and for which chemotherapy is employed to prevent  cancer relapse.
+Interestingly, aneuploidy is unlikely to contribute to evolutionary rescue in large tumors in which the number of cells is large enough (i.e., $N\gg N_m^*\approx 4\times10^7$) for the appearance of resistant mutation in sensitive cells before these cells become extinct (\Cref{rescue_prob}).
+However, aneuploidy can have a crucial role in the evolutionary rescue of small tumors, in which the number of sensitive cells may be below the detection threshold of $\sim10^7$  \citep{bozic2013evolutionary}, and this can have an impact on the recurrence of cancer after the resection of the primary tumor  through secondary tumors, which are too small to be detected and for which adjuvant chemotherapy is employed to prevent  cancer relapse (since we assume that the tumor is detected as soon as possible, large tumors are more likely to be primary tumors and small (undetected) tumors are secondary tumors).
 These secondary tumors are estimated to cause the majority of cancer-related deaths~\citep{chaffer2011perspective}.
@@
 The theoretical prediction for the mean rescue time for tumors smaller then  $10^8$ cells is greater than 4 years, consistent with previous estimates of the recurrence time of tumors after resection~\citep{avanzini2019cancer}.
-We found that aneuploidy complements evolutionary rescue through direct mutation because it generates rescue mutations mostly after the wildtype population had died out and direct mutation is no longer a viable option for evolutionary rescue (\Cref{cdffig}A).
-
-We hypothesized that \emph{standing genetic variation} (the existence of a subpopulation of aneuploid cancer cells before the onset of therapy) can facilitate evolutionary rescue by reducing the waiting time for the appearance of aneuploid cells.
-We found that a drug or combination of drugs that reduces the wildtype growth rate and does not significantly increase the chromosome missegregation rate is more likely to cause evolutionary rescue to occur through direct mutation (\cref{eq:Nsgv}).
+We found that aneuploidy complements evolutionary rescue through direct mutation because it generates rescue mutations mostly after the wildtype population has died out and direct mutation is no longer a viable option for evolutionary rescue (\Cref{cdffig}A).
+
+We hypothesized that \emph{standing genetic variation} (the existence of a subpopulation of aneuploid cancer cells before the onset of therapy) could facilitate evolutionary rescue by reducing the waiting time for the appearance of aneuploid cells.
+We found that a drug or combination of drugs that reduce the wildtype growth rate and do not significantly increase the chromosome missegregation rate are more likely to cause evolutionary rescue through direct mutation (\cref{eq:Nsgv}).
 Furthermore, we found that for reasonable parameter values evolutionary rescue is more likely to occur through \emph{de-novo} aneuploidy (\Cref{rescue_denovo}).
@@
 In this scenario, evolutionary rescue likely occurs via aneuploid cells that acquire a resistance mutation and the probability of evolutionary rescue declines exponentially with initial tumor size  (the probability of evolutionary rescue is given by $1-\exp\left(-N/\tilde{N}_a^*\right)$, see \Cref{rescue_prob_sgv}).
+
+Finally, we observe from \Cref{rescue_threshold} that only for a restricted region of the parameter space will aneuploidy act as a \emph{stepping stone} for evolutionary rescue.
+If the aneuploid division rate is smaller than $\frac{\mu_a}{1+u}\approx\mu_a(1-u)$ then evolutionary rescue will occur through direct mutation.
+As a result, for most parameter values, aneuploidy will either not play any role in evolutionary rescue or will be the main driver of adaptation without requiring any mutation (i.e.
+evolutionary rescue in one step).
 
@@
 
-Additionally, our model predictions can be tested with data from patients.
-Given the type of cancer and how beneficial aneuploidy is,  our model can predict the probability for cancer relapse after treatment.
+%Additionally, our model predictions can be tested with data from patients (e.g.
+tumor karyotype).
+Given the type of cancer and how beneficial aneuploidy is,  our model can offer predictions about the probability for cancer relapse after treatment which can be tested against patient relapse data.
 
@@
 
-Our results quantitatively show that aneuploidy plays an important role in tumors overcoming exposure to chemotherapeutic drugs when tumor size is small or intermediate.
-Large tumors can escape anti-cancer drugs through direct mutation while smaller ones are able to obtain a resistance mutation through an aneuploid \emph{stepping stone} (\Cref{rescue_prob}).
-As a result, therapies that increase the rate of aneuploidy in tumors to combat cancer may have an adverse effect on patient outcomes.
+Our results quantitatively suggests that aneuploidy may play an important role in tumors overcoming exposure to anti-cancer drugs when tumor size is small or intermediate.
+Large tumors are predicted to be more likely to escape anti-cancer drugs through direct mutation, whereas smaller tumors are predicted to obtain a resistance mutation through an aneuploid \emph{stepping stone} (\Cref{rescue_prob}) or through direct effect of aneuploidy.
+As a result, therapies that increase the rate of aneuploidy in tumors to combat cancer may also promote drug resistance.
 % TODO summary ok
@@
 the Sloan Foundation (Research Fellowship FG-2021-16667, DBW),
-the National Science Foundation (grant $\#2146260$, DBW),
-% TODO add funding for Uri
+the National Science Foundation (grant $\#2146260$, DBW), the ERC Starting Grant ($\#945674$, UBD).
+% TODO add funding for Uri ok
 }

### 2024-06-23 EvolutionaryRescue1.tex
revised after UBD comments

@@
   \species{\yeast}{Saccharomyces cerevisiae}{S.~cerevisiae}
+
+% highlight
+\newcommand{\hl}[1]{\colorbox{yellow!50}{$\displaystyle#1$}}
 
@@
 % Title page
-\title{Modeling the contribution of aneuplouidy to evolutionary rescue following anticancer drug exposure}
+\title{
+	Evolutionary rescue by aneuploidy in tumors exposed to anti-cancer drugs 
+}
 % Authors
@@
 Here, we model how aneuploidy impacts the fate of a population of cancer cells.
-We use multi-type branching processes to approximate the probability of a tumor to survive drug treatment as a function of the initial tumor size, the rates at which aneuploidy and other beneficial mutations occur, and the growth rates of the sensitive and resistant cells.
-Additionally, we investigate the effect of the fraction of pre-existent aneuploid cells on the probability of evolutionary rescue  following therapy.
+We use multi-type branching processes to approximate the probability that a tumor survives drug treatment as a function of the initial tumor size, the rates at which aneuploidy and other beneficial mutations occur, and the growth rates of the sensitive and resistant cells.
+Additionally, we investigate the effect of the fraction of pre-existent aneuploid cells on the probability of evolutionary rescue.
 We also approximate the mean recurrence time for the tumor to revert back to its initial size following treatment, if such evolutionary rescue has occurred.
@@
 
- \citet{ippolito2021gene} induced aneuploidy in cancer cell lines by exposing them to reversine, a small-molecule inhibitor of the mitotic kinase Mpsi1 and then to anticancer agents such as vemurafenib.
+ \citet{ippolito2021gene} induced aneuploidy in cancer cell lines by exposing them to reversine, a small-molecule inhibitor of the mitotic kinase Mpsi1 and then to anti-cancer drugs such as vemurafenib.
 Reversine-treated cells had a higher proliferation rate following drug exposure compared to wildtype cancer cells, due to selection of specific beneficial karyotypes.
- Similarly, \citet{lukow2021chromosomal} induced aneuploidy in cancer cells and observed that such cells have an advantage compared to wildtype cells during chemotherapy treatment, despite having lower fitness before the onset of chemotherapy.
-One proposed mechanism through which aneuploidy is able to confer resistance to chemotherapeutics is by antagonizing cell division, which prevents the drugs from damaging DNA and microtubules \citep{replogle2020aneuploidy}.
+Similarly, \citet{lukow2021chromosomal} induced aneuploidy in cancer cells and observed that such cells have an advantage compared to wildtype cells during drug treatment, despite having lower fitness before the onset of treatment.
+One proposed mechanism through which aneuploidy is able to confer resistance to anti-cancer drugs is by antagonizing cell division, which prevents the drugs from damaging DNA and microtubules \citep{replogle2020aneuploidy}.
 Other mechanisms are the selection for specific karyotypes that lead to reduced drug metabolism \citep{ippolito2021gene}, and elevated levels of DNA damage repair due to higher basal level of DNA damage \citep{zerbib2023human}.
 
-An important aspect of aneuploidy is that the rate with which cells become aneuploid, that is the rate of chromosome mis-segregation, is several orders of magnitude higher than mutation rates \citep{bakker2023predicting}.
+An important aspect of aneuploidy is that the rate with which cells become aneuploid, that is the rate of chromosome mis-segregation, is several orders of magnitude higher than mutation rate \citep{bakker2023predicting}.
 Consequently, a cell exposed to a stress such as chemotherapeutic drugs can acquire aneuploidy faster compared to acquiring a mutation.
@@
 How multiple mutations contribute to the survival of the population is less explored, but \citet{wilson2017soft} have shown that evolutionary rescue is significantly enhanced by soft selective sweeps when multiple mutations contribute towards evolutionary rescue.
-Evolutionary rescue that requires two successive mutations (i.e., two steps) has been investigated using diffusion approximation by \citet{martin2013probability} who tested the predictions of their model with data from experiments in yeast and bacteria.
-
-Here we study evolutionary rescue after a sudden environment change caused by the initiation of chemotherapy.
-We assume that the affected cancer is melanoma, the chemotherapeutic agent is vemurfenib and that resistance is conferered by either trisomy in Chromosome 2 or trisomy in Chromosome 6.
-We estimate the effect of aneuploidy on the probability of evolutionary rescue when it acts as an evolutionary stepping stone (i.e.
-intermediary) towards resistant cancer cells.
+Evolutionary rescue that requires two successive mutations (i.e., two steps) has been investigated by \citet{martin2013probability} who tested the predictions of their model with data from experiments in yeast and bacteria.
+
+Here we study evolutionary rescue after a sudden environment change caused by the initiation of anti-cancer drug treatment. 
+We consider a range of effects of aneuploidy, from tolerance to (partial) resistance to the drug \citep{brauner2016distinguishing}.
+% TODO explain distinction in a sentence
+We estimate the effect of aneuploidy on the probability of evolutionary rescue of the tumor.
+When aneuploidy provides resistance to the drug, it can directly rescue the tumor.
+However, when anueploidy provides tolerance to the drug, it may act as an evolutionary ``stepping stone'' or ``springboard'' \citep{Yona2015,martin2013probability,osmond_genetic_2020}, delaying extinction, and thereby allowing the tumor more time to acquire a resistance mutation on top of the aneuploid background.
+As mentioned above, evidence suggests that aneuploidy may be a common strategy for tumor adaptation to durg therapy, but it is unknown how often aneuploidy provides tolerance and thereby acts as a stepping stone.
 We also estimate the mean time until a tumor cell population reaches its pre-treatment size following drug therapy. 
@@
 \paragraph{Evolutionary model.}
-We follow the number of cancer cells that have one of three different genotypes at time $t$: wildtype, $w_t$; aneuploid, $a_t$; and mutant, $m_t$. 
+We follow the number of cancer cells that have one of three different genotypes at time $t$: sensitive, $w_t$; tolerant/resistant aneuploid, $a_t$; and resistant mutant, $m_t$. 
 These cells divide and die with rates $\lambda_k$ and $\mu_k$ (for $k=w, a, m$).
 The difference between the division and death rate is $\Delta_k = \lambda_k-\mu_k$.
-We assume the population of cells is under a strong stress, such as drug therapy, to which the wildtype genotype is susceptible or sensitive and therefore $\Delta_w<0$, whereas the mutant is resistant to the stress, $\Delta_m>0$.
-We analyze three scenarios: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_w$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing, stationary, or growing or dying only very slowly, that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$, in a sense that we will make precise below.
+We assume the population of cells is under a strong stress, such as drug therapy, and therefore $\Delta_w<0$, whereas the mutant is resistant to the stress, $\Delta_m>0$.
+We consider a range of possible values for $\Delta_a$, finding three distinct cases: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_w$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing, stationary, or growing or dying only very slowly, that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$, in a sense that we will make precise below.
+
 We assume that both chromosomal missegregation and mutation occur during the process of mitosis. 
-Wildtype cells may divide and then missegregate to become aneuploids at rate $u\lambda_w$.
-Both aneuploid and wildtype cells may divide and mutate to become mutants at rates $v\lambda_{a}$ and $v\lambda_{w}$, respectively.
-To model standing genetic variation, we assume that before the onset of therapy, wildtype cells become aneuploid with rate $\tilde{u}\lambda_w$ (which may differ from $u \lambda_w$) and that aneuploidy confers a fitness cost $s$ in the drug-free environment, that is, we assume that aneuploid cells have an increased death rate compared to wildtype cells in a drug-free environment.
+Sensitive cells may divide and then missegregate to become aneuploids at rate $u\lambda_w$.
+Both aneuploid and sensitive cells may divide and mutate to become mutants at rates $v\lambda_{a}$ and $v\lambda_{w}$, respectively.
+To model standing genetic variation, we assume that before the onset of therapy, sensitive cells become aneuploid with rate $\tilde{u}\lambda_w$ (which may differ from $u \lambda_w$) and that aneuploidy confers a fitness cost $s$ in the drug-free environment, that is, we assume that aneuploid cells have an increased death rate compared to sensitive cells in a drug-free environment.
+
 See \Cref{figureAneuploidy} for a schematic representation of the model and \Cref{sampleTrajectories} for sample trajectories of the different genotypes.
@@
 Simulations are performed using the \emph{Gillespie stochastic simulation algorithm} \citep{gillespie1976general,gillespie1977exact} implemented in Python \citep{python}.
-The simulation monitors the number of cells of each type: wildtype, aneuploid, and mutant. 
-The wildtype population initially consists of $w_0=N$ cells, whereas the other cell types are initially absent.
+The simulation monitors the number of cells of each type: sensitive, aneuploid, and mutant. 
+The sensitive population initially consists of $w_0=N$ cells, whereas the other cell types are initially absent.
 
@@
 Aneuploid division rate is selected such that the aneuploid growth rate $\Delta_w\ll\Delta_a\ll\Delta_m$ which means that $0.06 \le \lambda_a \le 0.1$. 
-For most of our simulations we use $\lambda_a=0.0899$ per day, such that aneuploidy can only act as a \emph{stepping stone} for the generation of the mutant that rescues the cancer cell population.
-
+For most of our simulations we use $\lambda_a=0.0899$ per day, such that aneuploidy can only act as an evolutionary ``stepping stone'' for the generation of the resistant mutant that rescues the cancer cell population (note that this mutant will occur on the background of an aneuploid genotype).
+
+{\color{red} The missegregation rate according to \citet{bakker2023predicting} is between $10^{-3}-10^{-2}$ per chromosome per cell division, with the most likely value being $6.2\times10^{-3}$ per chromosome per cell division.}
+\citet{ippolito2021gene} observed that trisomy in Chr 2 and Chr 6 are most likely to confer increased resistance against the anti-cancer drug vemurafenib for A375 cells. 
+We assume each of these trisomies are formed at the most likely rate, and as a result, we use $u=10^{-2}$ per cell division.
+For the missegregation rate in the drug-free environment, $\tilde{u}$, we use the lower-end value of $\tilde{u}=2\times10^{-3}$ per cell division, as some drugs increase the rate of aneuploidy~\citep{wang2019molecular,mason2017functional}.
 We assume the mutation rate is $10^{-7}$ per gene per cell division \citep{loeb2001mutator} and since we assume that a single target gene confers resistance to the drug, we use $v=10^{-7}$ per cell division.
-\citet{bakker2023predicting} determined that the missegregation rate must be between $10^{-3}-10^{-2}$ per chromosome per cell division with the optimal value being $6.2\times10^{-3}$ per chromosome per cell division (for $u\ll10^{-3}$ no adaptation occurs and for $u\gg10^{-2}$ cells will acquire lethal nullisomies where both pairs of chromosomes are lost).
-\citet{ippolito2021gene} observed that trisomy in Chr II and VI are most likely to confer increased resistance against the chemotherapeutic agent vemurafenib for A375 cells.
-We assume that if a tumor is aneuploid then it most likely has trisomy \citep{gisselsson2010generation} in the pre-treatment environment and that cells with more than one trisomy are very unlikely to survive.
-Additionally, we assume that all trisomies are equally likely and, as a result, we select $u=10^{-2}$ per cell division.
-For the missegregation rate in the drug-free environment, $\tilde{u}$, we use the lower-end value of $\tilde{u}=2\times10^{-3}$ per cell division, as some drugs increase the rate of aneuploidy~\citep{wang2019molecular,mason2017functional}.
 
 The fitness cost $s$ of aneuploidy before the onset of therapy is difficult to estimate as we are interested in a specific type of aneuploidy that improves the fitness of cancer cells in an environment altered by drugs.
-We derive $s$ by using the formula $s=\tilde{u}\lambda_w / f$, where $f$ is the fraction of aneuploid cancer cells.
+We estimate $s$ by using the formula $s=\tilde{u}\lambda_w / f$, where $f$ is the fraction of aneuploid cancer cells.
 To estimate $f$, we note that \citet{lukow2021chromosomal} mixed wildtype and aneuploid A375 melanoma cells at $50:50$ ratio, cultured them in drug-free environment and observed the ratio evolve as a function of time with the aneuploid cells declining to $15\%$ after 24 days.
-We obtain the fitness cost $s$ using the formula $s=\abs{\log\left[0.15/(1-0.15)\right]/24}\approx0.07$ per day \citep{chevin2011measuring}.
-As a result, the fraction of cancer cells with aneuploidy of interest is $f=2\times10^{-3}\times10^{-1}/0.07=0.28\%$ (i.e., $0.28\%$ of pre-treatment cancer cells have the aneuploidy of interest).
+Thus, our estimate for the fitness cost $s$ is $s=\abs{\log\left[0.15/(1-0.15)\right]/24}\approx0.07$ per day \citep{chevin2011measuring}.
+Therefore, the estimated fraction of cancer cells with the beneficial aneuploidy is $f=2\times10^{-3}\times10^{-1}/0.07=0.28\%$, that is, $0.28\%$ of pre-treatment cancer cells have the beneficial aneuploidy of interest.
 
@@
 
-All of the parameters discussed above are shown in \Cref{table1}.
+All the parameters discussed above are shown in \Cref{table1}.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-Therefore, the threshold tumor size with standing generation variation, $\tilde{N}^*_{a}$, is similar to the threshold with de-novo variation, $N^*_a$, except that the sensitive growth rate $\abs{\Delta_w}$ is replaced with the aneuploidy cost $s$, such that
+Therefore, the threshold tumor size for rescue by standing generation variation, $\tilde{N}^*_{a}$, is similar to the threshold for rescue by de-novo variation, $N^*_a$, except that the sensitive growth rate $\abs{\Delta_w}$ is replaced by the aneuploidy cost $s$, such that
 \begin{equation}\label{eq:Nsgv}
@@
 For large tumors ($N \gg N_m^*$), the mean evolutionary rescue time (\cref{eq:AsymptoticTimeRules}) is independent of parameters characterizing aneuploid cells or their production ($u$, $\lambda_a$, and $\Delta_a$).
-% I removed the mention of exponential distribution as it was not explained and I dont see why you need it here
 Increasing the per division mutation rate, $v$, leads to faster appearance of a rescue mutation and hence reduced mean rescue time. 
@@
 We note that standing genetic variation does not have a significant effect on the mean evolutionary rescue time.
-% WHAT is the purpose of this sentence?
-ok
-
-In Appendix \ref{sec:appendix_distribution_time} we calculated the probability that a rescue mutation has been generated by time $t$.
-This allows us to examine whether aneuploidy accelerates or delays evolutionary rescue.
-We find that anenploidy  accelerates evolutionary rescue after $1/\Delta_a\approx100$, days after which no more rescue mutations are generated through direct mutation (\Cref{cdffig}A).
-This shows that rescue mutations are generated in aneuploid cells at latter timescales then in sensitive cells and, thus, aneuploidy increases the \emph{window of opportunity} for evolutionary rescue.
-% WHAT is the purpose of this sentence?
-ok
+
+In Appendix \ref{sec:appendix_distribution_time} we calculate the probability that a rescue mutation has been generated by time $t$.
+This allows us to examine whether aneuploidy promotes or delays evolutionary rescue.
+We find that anenploidy promotes evolutionary rescue after $1/\Delta_a\approx100$ days, at a time where no more rescue mutations are generated through mutations in sensitive cells (\Cref{cdffig}A).
+Thus, aneuploidy increases the \emph{window of opportunity} for evolutionary rescue.
+This can have a counter-intuitive outcome: conditioned on rescue of the tumor, tumors rescued by aneuploid cells may acquire rescue mutations later than those rescued by sensitive cells.
 
@@
 We note that, for small and large tumors, when $N \ll N_a^*$ or $N \gg N_m^*$, the asymptotic expressions for the mean recurrence time are independent of the chromosome missegregation rate $u$, and therefore the rate at which the drug induces aneuploidy has no effect on the time necessary for the tumor to rebound to its initial size $N$.
-% Remus the next paragraphs look like you wanted to find approximations so you did, you reference them here, but the results are buried in the technical text: WHAT IS IT THAT YOU WANT US TO LEARN from this?
-ok
 
@@
 We observe that  for small and intermediate sized tumors the effect of the detection size $M$ on $\tau_a^{M}$ is negligible when compared to the scenario where the detection size is equal to the initial population size $N$ (i.e., $\tau_a^r\approx\tau_a^{M}$ for $N<N_m^*$).
-% WHAT DOES THAT MEAN?
-ok
 However, for large tumors the mean recurrence time to detection size $M$ decreases logarithmically with tumor size $N$ while $\tau_a^{r}$ is constant (\Cref{RecurrencePlot}).
-Additionally, for large tumors we have $M<N_m^*<N$ so the mean detection time is shorter for such tumors compared to the mean recurrence time, that is, the resistant tumor can be detected before reaching its initial size.
-
-Most clinical trials report data on the distribution of recurrence time measured from the time of surgery~\citep{avanzini2019cancer} with drug therapy usually following after (i.e., adjuvant therapy).
-% WHAT DOES "varying" mean here?
-ok
-Therefore, since only undetected secondary tumors are present at the time of treatment with anti-cancer drug, we lack data on the size of the tumors and cannot compare the empirical distributions to our predictions.
-However, we expect  the variability of secondary tumors to average out across large cohorts of patients, thus we can use the mean recurrence time to compare with clinical data on cancer relapse.
+Additionally, for large tumors we have $M<N_m^*<N$ so the mean detection time is shorter for such tumors compared to the mean recurrence time, that is, the resistant tumor can be detected before recovering back to its initial size.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \section*{Discussion}
@@
 We have modeled a tumor--a population of cancer cells--exposed to drug treatment that causes it to decline in size towards potential extinction.
-In this scenario, the tumor can be "evolutionary rescued", or escape extinction, via two paths.
-In the direct path, a sensitive cell acquires a mutation that confers resistance that allows it to rapidly grow.
+In this scenario, the tumor can be ``evolutionary rescued'', or escape extinction, via two paths.
+In the direct path, a sensitive cell acquires a mutation or aneuploidy that confers resistance that allows it to rapidly grow.
 In the indirect path, a sensitive cell first becomes aneuploid, which diminishes the effect of the drug, and then an aneuploid cell acquires a mutation that confers resistance (\Cref{figureAneuploidy}).
@@
 We find that aneuploidy can have a significant effect on evolutionary rescue as it reduces the threshold tumor size by at least an order-of-magnitude even when aneuploidy only provides tolerance (\Cref{rescue_prob}).
-Interestingly, aneuploidy is unlikely to contribute to evolutionary rescue in large tumors in which the number of cells is large enough (i.e., $N\gg N_m^*\approx 4\times10^7$) for the appearance of resistant mutation in sensitive cells before these cells become extinct (\Cref{rescue_prob}).
-However, aneuploidy can have a crucial role in the evolutionary rescue of small tumors, in which the number of sensitive cells may be below the detection threshold of $\sim10^7$  \citep{bozic2013evolutionary}, and this can have an impact on the recurrence of cancer after the resection of the primary tumor  through secondary tumors, which are too small to be detected and for which adjuvant chemotherapy is employed to prevent  cancer relapse (since we assume that the tumor is detected as soon as possible, large tumors are more likely to be primary tumors and small (undetected) tumors are secondary tumors).
-These secondary tumors are estimated to cause the majority of cancer-related deaths~\citep{chaffer2011perspective}.
-The importance of aneuploidy in the evolutionary rescue of secondary tumors is reinforced by the fact that metastases have been shown to have a chromosome missegregation rate two to three orders of magnitude higher then that of primary tumors~\citep{kimmel2023intra}.
+When the number of cells in the tumor is large enough (i.e., $N\gg N_m^*\approx 4\times10^7$), a resistance mutation will occur in sensitive cells before these cells become extinct (\Cref{rescue_prob}).
+Therefore, large tumors are certain to be rescued with or without aneuploidy.
+However, anti-cancer drugs are often used as adjuvant therapy after resection, in which case the number of cells in the tumor may be below the detection threshold of $\sim10^7$  \citep{bozic2013evolutionary}.
+In these cases, aneuploidy can have a crucial role in the evolutionary rescue of the tumor and therefore in cancer recurrence.
+{\color{red} Indeed, secondary tumors are estimated to cause the majority of cancer-related deaths~\citep{chaffer2011perspective}.
+The importance of aneuploidy in the evolutionary rescue of secondary tumors is reinforced by the fact that metastases have been shown to have a chromosome missegregation rate two to three orders of magnitude higher then that of primary tumors~\citep{kimmel2023intra}.}
 
@@
 
-We hypothesized that \emph{standing genetic variation} (the existence of a subpopulation of aneuploid cancer cells before the onset of therapy) could facilitate evolutionary rescue by reducing the waiting time for the appearance of aneuploid cells.
-We found that a drug or combination of drugs that reduce the wildtype growth rate and do not significantly increase the chromosome missegregation rate are more likely to cause evolutionary rescue through direct mutation (\cref{eq:Nsgv}).
-Furthermore, we found that for reasonable parameter values evolutionary rescue is more likely to occur through \emph{de-novo} aneuploidy (\Cref{rescue_denovo}).
-If the fraction of tumor cells that have the beneficial aneuploidy is $f\gg\frac{u\lambda_w}{\abs{\Delta_w}}\approx2.5\%$, then evolutionary rescue is more likely to occur via standing variation, rather then through \emph{de-novo} aneuploidy.
-In this scenario, evolutionary rescue likely occurs via aneuploid cells that acquire a resistance mutation and the probability of evolutionary rescue declines exponentially with initial tumor size  (the probability of evolutionary rescue is given by $1-\exp\left(-N/\tilde{N}_a^*\right)$, see \Cref{rescue_prob_sgv}).
+We hypothesized that standing genetic variation (the existence of a subpopulation of aneuploid cancer cells before the onset of therapy) could facilitate evolutionary rescue by reducing the waiting time for the appearance of aneuploid cells.
+We found that a drug that reduces the wildtype growth rate and does not significantly increase the chromosome missegregation rate is likely to lead to evolutionary rescue through standing genetic variation (\Cref{rescue_denovo,eq:Nsgv}).
+Furthermore, if the fraction of tumor cells that have the beneficial aneuploidy is $f\gg\frac{u\lambda_w}{\abs{\Delta_w}}\approx2.5\%$, then evolutionary rescue is more likely to occur via standing variation rather then through \emph{de-novo} aneuploidy.
+However, for the parameter values we focus on in our examples (\Cref{table1}), this fraction is an order of magnitude lower, and therefore we expect evolutionary rescue to occur by \emph{de-novo} aneuploidy.
 
@@
 
-\paragraph{Experimental future direction}
-Our model predictions could be tested by experiments \citep{martin2013probability}.
+\paragraph{Experimental validation}
+Our model predictions could be tested by experiments.
 For example, to assess the effect of initial tumor size on the probability of evolutionary rescue, a large culture mass can be propagated from a single cancer cell in permissive conditions and then diluted to a  range of starting tumor sizes.
-Then, these tumors may be exposed to anti-cancer drugs that induce aneuploidy or to saline solution for control~\citep{ippolito2021gene}.
-Cell density can be measured by optical density and a population exposed to the drug is considered extinct if the optical density is lower when compared to the control case with no cells present.
-We can then compare the results of the experiments to the predictions of our model to see if tumors with initial size bellow the threshold \cref{eq:N_a} are more likely to become extinct due to the drug.
-
-%Additionally, our model predictions can be tested with data from patients (e.g.
-tumor karyotype).
-Given the type of cancer and how beneficial aneuploidy is,  our model can offer predictions about the probability for cancer relapse after treatment which can be tested against patient relapse data.
-
-\paragraph{Directions of future research}
- Our model can be extended to understand evolutionary rescue in different biological contexts, for example, how yeast populations under stress overcome extinction via aneuploidy~\citep{pompei2023fitness,kohanovski2024aneuploidy}.
-Additionally, we did not account for the heterogeneity of aneuploidy as not all aneuploid lineages have the same growth rate $\Delta_a$ as assumed in our model~\citep{avecilla2023copy,yang2021fitness}.
-Such heterogeneity can be accounted for  by  sampling the aneuploidy growth or death rates from a distribution~\citep{martin2013probability}.
-
+Then, the extinction or survival of these tumors can be monitored during exposure to anti-cancer drugs that induce aneuploidy or to saline solution for control~\citep{ippolito2021gene}.
+We can then compare the results of these experiments to predictions of our model to see if tumors with initial size below the threshold \cref{eq:N_a} are more likely to become extinct due to drug exposure.
+
+\paragraph{Directions for future research}
 We have assumed that cancer cell lineages are independent and have verified that this is accurate under simple logistic growth.
@@
 This has the potential to impact the probability of evolutionary rescue \citep{martens2011spatial}.
-
-An additional extension of our model could incorporate a wider range of parameters, however parameter values for different types of tumors are difficult to obtain.
+In addition, our model can be extended to understand evolutionary rescue in different biological contexts, for example, how yeast populations under stress overcome extinction via aneuploidy~\citep{pompei2023fitness,kohanovski2024aneuploidy}.
 
@@
 
-Our results quantitatively suggests that aneuploidy may play an important role in tumors overcoming exposure to anti-cancer drugs when tumor size is small or intermediate.
-Large tumors are predicted to be more likely to escape anti-cancer drugs through direct mutation, whereas smaller tumors are predicted to obtain a resistance mutation through an aneuploid \emph{stepping stone} (\Cref{rescue_prob}) or through direct effect of aneuploidy.
-As a result, therapies that increase the rate of aneuploidy in tumors to combat cancer may also promote drug resistance.
-% TODO summary ok
+Our results quantitatively suggest that aneuploidy may play an important role in tumor adaptation to anti-cancer drugs when tumor size is small or intermediate.
+Large tumors are predicted to be more likely to adapt to anti-cancer drugs through direct mutation, whereas smaller tumors are predicted to become resistant either directly by aneuploidy or by a resistance mutation occuring in aneuploid cells that serve as an evolutionary ``stepping stone'' (\Cref{rescue_prob}).
+Thus, therapies that increase the rate of aneuploidy in tumors to combat cancer may also promote drug resistance.
 
@@
 the National Science Foundation (grant $\#2146260$, DBW), the ERC Starting Grant ($\#945674$, UBD).
-% TODO add funding for Uri ok
 }
@@
 \caption{\textbf{Model parameters.} 
-We have modified the parameters from \citet{bozic2013evolutionary} such that wildtype/mutant division rate is $\lambda_{w,m}=\log2/T\approx0.1$ instead of their value of $0.14$ where $T$ is the doubling time in the absence of cellular death obtained from \citet{rew2000cell}.}
+Parameters from \citet{bozic2013evolutionary} consider patiens with melanoma treated with the anti-cancer drug vemurfenib, in which resistance is conferered by trisomy in either Chr 2 or Chr 6.
+We have modified the parameters from \citet{bozic2013evolutionary} such that wildtype and mutant division rates are $\lambda_{w} = \lambda_{m} = \log2/T\approx0.1$ instead of their value of $0.14$ where $T$ is the doubling time in the absence of cellular death obtained from \citet{rew2000cell}.}
   \label{table1}
@@
 \textbf{Standing genetic variation facilitates evolutionary rescue of cancer.}
-(A)  Ratio of threshold tumor size $\tilde{N}_a^*$ when a fraction $\frac{\tilde{u}\lambda_w}{s}$ is aneuploid at the start of treatment and $N_a^*$ as a function of the wildtype growth rate $\Delta_w$.
+(A)  Ratio of threshold tumor sizes for rescue by standing genetic variation and by de-novo variation, $\tilde{N}_a^*/N_a^*$, when a fraction $\frac{\tilde{u}\lambda_w}{s}$ is aneuploid at the start of treatment, as a function of the wildtype growth rate $\Delta_w$.
 Standing genetic variation will drive adaptation to the drug if the wildtype population is rapidly declining ($\Delta_w\ll0$) due to a stronger effect of the drug on sensitive cells.

### 2024-06-27 EvolutionaryRescue1.tex
remove red in discussion as UBD approved

@@
 In these cases, aneuploidy can have a crucial role in the evolutionary rescue of the tumor and therefore in cancer recurrence.
-{\color{red} Indeed, secondary tumors are estimated to cause the majority of cancer-related deaths~\citep{chaffer2011perspective}.
-The importance of aneuploidy in the evolutionary rescue of secondary tumors is reinforced by the fact that metastases have been shown to have a chromosome missegregation rate two to three orders of magnitude higher then that of primary tumors~\citep{kimmel2023intra}.}
+Indeed, secondary tumors are estimated to cause the majority of cancer-related deaths~\citep{chaffer2011perspective}.
+The importance of aneuploidy in the evolutionary rescue of secondary tumors is reinforced by the fact that metastases have been shown to have a chromosome missegregation rate two to three orders of magnitude higher then that of primary tumors~\citep{kimmel2023intra}.
 

### 2024-07-23 EvolutionaryRescue1.tex
Add files via upload

@@
 \affil[2]{Department of Human Molecular Genetics and Biochemistry, Faculty of Medicine, Tel Aviv University, Tel Aviv, Israel}
-\affil[3]{Department of Physics, Emory University, Atlanta, GA}
+\affil[3]{Department of Physics, Emory University, Atlanta, GA, USA}
 \affil[*]{Corresponding author: Yoav Ram (e-mail: yoavram@tauex.tau.ac.il)}
@@
  \citet{ippolito2021gene} induced aneuploidy in cancer cell lines by exposing them to reversine, a small-molecule inhibitor of the mitotic kinase Mpsi1 and then to anti-cancer drugs such as vemurafenib.
-Reversine-treated cells had a higher proliferation rate following drug exposure compared to wildtype cancer cells, due to selection of specific beneficial karyotypes.
-Similarly, \citet{lukow2021chromosomal} induced aneuploidy in cancer cells and observed that such cells have an advantage compared to wildtype cells during drug treatment, despite having lower fitness before the onset of treatment.
+Reversine-treated cells had a higher proliferation rate following drug exposure compared to sensitive cancer cells, due to selection of specific beneficial karyotypes.
+Similarly, \citet{lukow2021chromosomal} induced aneuploidy in cancer cells and observed that such cells have an advantage compared to sensitive cells during drug treatment, despite having lower fitness before the onset of treatment.
 One proposed mechanism through which aneuploidy is able to confer resistance to anti-cancer drugs is by antagonizing cell division, which prevents the drugs from damaging DNA and microtubules \citep{replogle2020aneuploidy}.
@@
 \paragraph{Evolutionary model.}
-We follow the number of cancer cells that have one of three different genotypes at time $t$: sensitive, $w_t$; tolerant/resistant aneuploid, $a_t$; and resistant mutant, $m_t$. 
-These cells divide and die with rates $\lambda_k$ and $\mu_k$ (for $k=w, a, m$).
+We follow the number of cancer cells that have one of three different genotypes at time $t$: sensitive, $s_t$; tolerant/resistant aneuploid, $a_t$; and resistant mutant, $m_t$. 
+These cells divide and die with rates $\lambda_k$ and $\mu_k$ (for $k=s, a, m$).
 The difference between the division and death rate is $\Delta_k = \lambda_k-\mu_k$.
-We assume the population of cells is under a strong stress, such as drug therapy, and therefore $\Delta_w<0$, whereas the mutant is resistant to the stress, $\Delta_m>0$.
-We consider a range of possible values for $\Delta_a$, finding three distinct cases: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_w$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing, stationary, or growing or dying only very slowly, that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$, in a sense that we will make precise below.
+We assume the population of cells is under a strong stress, such as drug therapy, and therefore $\Delta_s<0$, whereas the mutant is resistant to the stress, $\Delta_m>0$.
+We consider a range of possible values for $\Delta_a$, finding three distinct cases: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_s$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing, stationary, or growing or dying only very slowly, that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$, in a sense that we will make precise below.
 
 We assume that both chromosomal missegregation and mutation occur during the process of mitosis. 
-Sensitive cells may divide and then missegregate to become aneuploids at rate $u\lambda_w$.
-Both aneuploid and sensitive cells may divide and mutate to become mutants at rates $v\lambda_{a}$ and $v\lambda_{w}$, respectively.
-To model standing genetic variation, we assume that before the onset of therapy, sensitive cells become aneuploid with rate $\tilde{u}\lambda_w$ (which may differ from $u \lambda_w$) and that aneuploidy confers a fitness cost $s$ in the drug-free environment, that is, we assume that aneuploid cells have an increased death rate compared to sensitive cells in a drug-free environment.
+Sensitive cells may divide and then missegregate to become aneuploids at rate $u\lambda_s$.
+Both aneuploid and sensitive cells may divide and mutate to become mutants at rates $v\lambda_{a}$ and $v\lambda_{s}$, respectively.
+To model standing genetic variation, we assume that before the onset of therapy, sensitive cells become aneuploid with rate $\tilde{u}\lambda_s$ (which may differ from $u \lambda_s$) and that aneuploidy confers a fitness cost $c$ in the drug-free environment, that is, we assume that aneuploid cells have an increased death rate compared to sensitive cells in a drug-free environment.
 
@@
 
-The cell population at time $t$ is represented by the triplet $\left(w_t,a_t,m_t\right)$.
+The cell population at time $t$ is represented by the triplet $\left(s_t,a_t,m_t\right)$.
 The following describes the events that may occur (right column), the rates at which they occur (middle column), and the effect these events have on the population (left column, see \Cref{figureAneuploidy}):
@@
 \begin{flalign*}
-(+1,0,0)&:\quad \lambda_ww_t\left(1-u-v\right)\quad\left(\text{birth of wildtype cell}\right),\\
-(-1,0,0)&:\quad \mu_ww_t\quad\left(\text{death of wildtype cell}\right),\\
-(0,+1,0)&:\quad u\lambda_ww_t\quad\left(\text{wildtype cell divides and becomes aneuploid}\right),\\
-(0,0,+1)&:\quad v\lambda_ww_t\quad\left(\text{wildtype cell divides and becomes mutant}\right),\\
+(+1,0,0)&:\quad \lambda_ss_t\left(1-u-v\right)\quad\left(\text{birth of sensitive cell}\right),\\
+(-1,0,0)&:\quad \mu_ss_t\quad\left(\text{death of sensitive cell}\right),\\
+(0,+1,0)&:\quad u\lambda_ss_t\quad\left(\text{sensitive cell divides and becomes aneuploid}\right),\\
+(0,0,+1)&:\quad v\lambda_ss_t\quad\left(\text{sensitive cell divides and becomes mutant}\right),\\
 (0,+1,0)&:\quad \lambda_aa_t\left(1-v\right)\quad\left(\text{birth of aneuploid cell}\right),\\
@@
 \end{subequations}
-For the remaining of this paper we assume that the division rates for wildtype and aneuploid cells can be written as $\lambda_ww_t\left(1-u-v\right)\approx \lambda_ww_t$ and $\lambda_aa_t\left(1-v\right)\approx\lambda_aa_t$ because $u,v\ll1$ (Table \ref{table1}).
+For the remaining of this paper we assume that the division rates for sensitive and aneuploid cells can be written as $\lambda_ss_t\left(1-u-v\right)\approx \lambda_ss_t$ and $\lambda_aa_t\left(1-v\right)\approx\lambda_aa_t$ because $u,v\ll1$ (Table \ref{table1}).
 Each iteration of the simulation loop starts with computing the rates $\nu_k$ of each event $k$.
@@
 The division rate is obtained as $\lambda=\log{2} / T \approx 0.1$ per day.
-We select this to be the division rate for wildtype and mutant cells.
+We select this to be the division rate for sensitive and mutant cells.
 %We select the mutant death rate $\mu_m=0.09\text{ days}^{-1}$ such that $\Delta_m>0$.
 
-\citet{bozic2013evolutionary} report the growth rate $\Delta_w$ for wildtype melanoma cancer cells from which they deduce the death rate $0.11 \le \mu_w \le 0.17$.
-We use  $\mu_w=0.14$ per day.
+\citet{bozic2013evolutionary} report the growth rate $\Delta_s$ for sensitive melanoma cancer cells from which they deduce the death rate $0.11 \le \mu_s \le 0.17$.
+We use  $\mu_s=0.14$ per day.
 Additionally, they observe the growth rate of cancer cells prior to treatment to be 0.01, which we use as the growth rate of mutant cells, which are resistant to the drug.
@@
 Aneuploid death rate $\mu_a$ is set to the same value as the mutant death rates, $\mu_m=0.09$ per day, given that aneuploidy increases resistance to the drug, such as cisplatin, by antagonizing cell division \citep{replogle2020aneuploidy}.
-Aneuploid division rate is selected such that the aneuploid growth rate $\Delta_w\ll\Delta_a\ll\Delta_m$ which means that $0.06 \le \lambda_a \le 0.1$. 
+Aneuploid division rate is selected such that the aneuploid growth rate $\Delta_s\ll\Delta_a\ll\Delta_m$ which means that $0.06 \le \lambda_a \le 0.1$. 
 For most of our simulations we use $\lambda_a=0.0899$ per day, such that aneuploidy can only act as an evolutionary ``stepping stone'' for the generation of the resistant mutant that rescues the cancer cell population (note that this mutant will occur on the background of an aneuploid genotype).
 
-{\color{red} The missegregation rate according to \citet{bakker2023predicting} is between $10^{-3}-10^{-2}$ per chromosome per cell division, with the most likely value being $6.2\times10^{-3}$ per chromosome per cell division.}
+The missegregation rate in cancer cells is estimated to be between $2.5\times10^{-4}-10^{-2}$ per chromosome per cell divison \citep{shi2005chromosome,thompson2008examining}.
 \citet{ippolito2021gene} observed that trisomy in Chr 2 and Chr 6 are most likely to confer increased resistance against the anti-cancer drug vemurafenib for A375 cells.
-We assume each of these trisomies are formed at the most likely rate, and as a result, we use $u=10^{-2}$ per cell division.
-For the missegregation rate in the drug-free environment, $\tilde{u}$, we use the lower-end value of $\tilde{u}=2\times10^{-3}$ per cell division, as some drugs increase the rate of aneuploidy~\citep{wang2019molecular,mason2017functional}.
+We assume each of these trisomies are formed at the most likely rate, and as a result, we use $\tilde{u}=10^{-3}$ per cell division as the chromosome missegregation rate in the drug-free environment.
+\citet{lee2016effects} estimated the effect of different anti-cancer drugs on the missegregation rate and found a $3- 50$ fold increase.
+We assume that our anti-cancer drug causes a 10-fold increase in the chomosome misegregation rate which gives us $u=10^{-2}$ per cell division, as some drugs increase the rate of aneuploidy~\citep{wang2019molecular,mason2017functional}.
 We assume the mutation rate is $10^{-7}$ per gene per cell division \citep{loeb2001mutator} and since we assume that a single target gene confers resistance to the drug, we use $v=10^{-7}$ per cell division.
 
-The fitness cost $s$ of aneuploidy before the onset of therapy is difficult to estimate as we are interested in a specific type of aneuploidy that improves the fitness of cancer cells in an environment altered by drugs.
-We estimate $s$ by using the formula $s=\tilde{u}\lambda_w / f$, where $f$ is the fraction of aneuploid cancer cells.
-To estimate $f$, we note that \citet{lukow2021chromosomal} mixed wildtype and aneuploid A375 melanoma cells at $50:50$ ratio, cultured them in drug-free environment and observed the ratio evolve as a function of time with the aneuploid cells declining to $15\%$ after 24 days.
+The fitness cost $c$ of aneuploidy before the onset of therapy is difficult to estimate as we are interested in a specific type of aneuploidy that improves the fitness of cancer cells in an environment altered by drugs.
+We estimate $c$ by using the formula $c=\tilde{u}\lambda_s / f$, where $f$ is the fraction of aneuploid cancer cells.
+To estimate $f$, we note that \citet{lukow2021chromosomal} mixed sensitive and aneuploid A375 melanoma cells at $50:50$ ratio, cultured them in drug-free environment and observed the ratio evolve as a function of time with the aneuploid cells declining to $15\%$ after 24 days.
 Thus, our estimate for the fitness cost $s$ is $s=\abs{\log\left[0.15/(1-0.15)\right]/24}\approx0.07$ per day \citep{chevin2011measuring}.
-Therefore, the estimated fraction of cancer cells with the beneficial aneuploidy is $f=2\times10^{-3}\times10^{-1}/0.07=0.28\%$, that is, $0.28\%$ of pre-treatment cancer cells have the beneficial aneuploidy of interest.
-
-We note that when we refer to wildtype cancer cells we include those cells that have any aneuploidy other than trisomy in Chr 2 and 6 as those are the aneuploid cells that are hypothesized to have higher fitness in the environment altered by drugs such as vemurafenib.
+Therefore, the estimated fraction of cancer cells with the beneficial aneuploidy is $f=10^{-3}\times10^{-1}/0.07=0.14\%$, that is, $0.14\%$ of pre-treatment cancer cells have the beneficial aneuploidy of interest.
+
+We note that when we refer to sensitive cancer cells we include those cells that have any aneuploidy other than trisomy in Chr 2 and 6 as those are the aneuploid cells that are hypothesized to have higher fitness in the environment altered by drugs such as vemurafenib.
 
@@
 \begin{align*}
-\lambda_w' &= \lambda_w , \\
-\mu_w' &= \mu_w ,\\
+\lambda_s' &= \lambda_s , \\
+\mu_s' &= \mu_s ,\\
 \lambda_a' &= \lambda_a ,\\ 
@@
 In our model, \emph{evolutionary rescue} occurs when resistant cells appear and establish (avoid random extinction) in the population  ($m_t \gg 1$) before the population becomes extinct ($w_t=a_t=m_t=0$).
-Aneuploidy may contribute to evolutionary rescue by either preventing (when $\Delta_a>0$) or delaying (when $0>\Delta_a>\Delta_w$) the extinction of the population before mutant cells appear and establish.
-We assume independence between clonal lineages starting from an initial population of $N$ wildtype cells (we check the effect of density-dependent growth on our results below).
-We therefore define $p_w$ as the probability that a lineage starting from a single wildtype cell avoids extinction by acquiring drug resistance.
-Thus, $N^*=1/p_w$ is the threshold tumor size above which evolutionary rescue is very likely, and the rescue probability is given by 
+Aneuploidy may contribute to evolutionary rescue by either preventing (when $\Delta_a>0$) or delaying (when $0>\Delta_a>\Delta_s$) the extinction of the population before mutant cells appear and establish.
+We assume independence between clonal lineages starting from an initial population of $N$ sensitive cells (we check the effect of density-dependent growth on our results below).
+We therefore define $p_s$ as the probability that a lineage starting from a single sensitive cell avoids extinction by acquiring drug resistance.
+Thus, $N^*=1/p_s$ is the threshold tumor size above which evolutionary rescue is very likely, and the rescue probability is given by 
 \begin{equation} \label{eq:rescue_prob} 
 \presc = 
-1-\left(1-p_w\right)^N \approx
-1-\e^{-Np_w} = 
+1-\left(1-p_s\right)^N \approx
+1-\e^{-Np_s} = 
 1-e^{-N/N^*} ,
 \end{equation}
-where the approximation $(1-p_w)\approx e^{-p_w}$ assumes that $p_w$ (but not necessarily $N p_w$) is small.
-Indeed, when $N<1/p_w$, then the probability for evolutionary rescue is $\presc \approx N p_w$  and when $N > 1/p_w$, it is $\presc \approx 1$, justifying the definition of $N^*$ as the threshold tumor size for evolutionary rescue.
-
-We use the theory of multi-type branching processes to find approximate expressions \cref{eq:pw_parttolerant,eq:pw_partrest,eq:pw_tolerant} for $p_w$ in different regimes (see appendix~\ref{sec:appendix-surv-prob}).
-Substituting these into $N^*=1/p_w$, we find approximations for the threshold tumor size, $N^*$. 
+where the approximation $(1-p_s)\approx e^{-p_s}$ assumes that $p_s$ (but not necessarily $N p_s$) is small.
+Indeed, when $N<1/p_s$, then the probability for evolutionary rescue is $\presc \approx N p_s$  and when $N > 1/p_s$, it is $\presc \approx 1$, justifying the definition of $N^*$ as the threshold tumor size for evolutionary rescue.
+
+We use the theory of multi-type branching processes to find approximate expressions \cref{eq:pw_parttolerant,eq:pw_partrest,eq:pw_tolerant} for $p_s$ in different regimes (see appendix~\ref{sec:appendix-surv-prob}).
+Substituting these into $N^*=1/p_s$, we find approximations for the threshold tumor size, $N^*$. 
 For these approximations, an important quantity is $T^* = (4 v \lambda_a^2 \Delta_m/\lambda_m)^{-1/2}$, which is the critical time an aneuploid lineage needs to survive to produce a resistant mutant that avoids random extinction.
-First, if aneuploidy is very rare ($u\lambda_a T^*< 1$), or if aneuploidy is rare ($u\lambda_a < -\Delta_a$) and very sensitive to the drug ($\Delta_a T^* < -1$), then evolutionary rescue will likely occur by a direct resistance mutation in a sensitive wildtype cell without the involvement of aneuploidy, such that
+First, if aneuploidy is very rare ($u\lambda_a T^*< 1$), or if aneuploidy is rare ($u\lambda_a < -\Delta_a$) and very sensitive to the drug ($\Delta_a T^* < -1$), then evolutionary rescue will likely occur by a direct resistance mutation in a sensitive cell without the involvement of aneuploidy, such that
 \begin{equation} \label{eq:N_m}
-N_m^* \approx \frac{\abs{\Delta_w}}{v\lambda_w}  \frac{\lambda_m}{\Delta_m} .
-\end{equation}
-Here, $\abs{\Delta_w}/\left(v\lambda_w\right)$ is the ratio of the rate at which wildtype cells are decreasing in number and the rate at which they are mutating.
+N_m^* \approx \frac{\abs{\Delta_s}}{v\lambda_s}  \frac{\lambda_m}{\Delta_m} .
+\end{equation}
+Here, $\abs{\Delta_s}/\left(v\lambda_s\right)$ is the ratio of the rate at which sensitive cells are decreasing in number and the rate at which they are mutating.
 Notably, the aneuploidy parameters ($u$, $\lambda_a$, $\mu_a$) do not affect $N_m^*$.
@@
 N_a^* \approx 
-  \frac{\abs{\Delta_w}}{u\lambda_w} \cdot \begin{cases}
+  \frac{\abs{\Delta_s}}{u\lambda_s} \cdot \begin{cases}
     \frac{\abs{\Delta_a}}{v\lambda_a}  \frac{\lambda_m}{\Delta_m} ,&
@@
 These approximations perform very well when compared to results of stochastic evolutionary simulations (\Cref{rescue_prob,rescue_threshold}).
-The first line describes the scenario in which aneuploid cells are still effectively killed by the treatment, but not as quickly as the wildtype. 
+The first line describes the scenario in which aneuploid cells are still effectively killed by the treatment, but not as quickly as the sensitive cells. 
 In the second scenario, aneuploid cells are sufficiently resistant that the expected size of each aneuploid lineage is roughly 1.
@@
 We can categorize tumors by their size: small tumors with size $N<N_a^*$ that are unlikely to survive treatment, intermediate tumors with size $N_a^* < N < N_m^*$ that rely on aneuploidy for evolutionary rescue, and large tumors with size $N > N_m^*$ that could overcome the effect of drug treatment even without aneuploidy.
-For the parameter values in \Cref{table1} with $\lambda_a=0.0899,\mu_w=0.14, u=10^{-2}, v=10^{-7}$, we are in the tolerant aneuploid scenario, and substituting in \cref{eq:N_a,eq:N_m}, we have $N_a^* \approx 4 \times 10^6$ and $N_m^* \approx 4 \times 10^7$.
+For the parameter values in \Cref{table1} with $\lambda_a=0.0899,\mu_s=0.14, u=10^{-2}, v=10^{-7}$, we are in the tolerant aneuploid scenario, and substituting in \cref{eq:N_a,eq:N_m}, we have $N_a^* \approx 4 \times 10^6$ and $N_m^* \approx 4 \times 10^7$.
 Hence, we obtain the ratio $N^*_a/N^*_m \approx 0.11$ (\cref{eq:N_ratio}), that is, aneuploidy reduces the threshold tumor size by approximately 89\%.
-Interestingly, the threshold between small and intermediate tumors, $N_a^*$, is similar to the tumor detection threshold of $4 \times 10^6$ cells for a wide variety of tumors as reported by \citet{avanzini2019cancer}.
+Interestingly, the threshold between small and intermediate tumors, $N_a^*$, is similar to the tumor detection threshold of $4.19 \times 10^6$ cells for a wide variety of tumors as reported by \citet{avanzini2019cancer}.
+
+Aneuploidy can increase the point mutation rate in cancer cells by increasing the number of genes that can mutate or by causing DNA damage, which can in turn elevates the mutation rate \citep{garribba2023short,passerini2016presence,janssen2011chromosome}.
+Thus, we re-write our model to take this into account in Appendix \ref{sec:appendix-diff-mut}.
+We observe from \cref{eq:N_a2} and \Cref{rescue_prob_ane} that increasing the mutation rate in aneuploid cells by one order of magnitude leads to a decrease in the threshold tumor size of approximately one order of magnitude.
+We note that for elevated mutation rates the aneuploids transition from tolerant to stationary without change in their growth rate $\Delta_a$.
 
@@
 
-In the above we assumed that at the onset of drug treatment, the initial tumor consisted entirely of wildtype cells that are drug sensitive.
-However, aneuploidy is likely produced even before the onset of treatment at some rate $\tilde{u}$ and confers a deleterious fitness effect $s$ in the absence of the drug \citep{replogle2020aneuploidy,giam2015aneuploidy}.
+In the above we assumed that at the onset of drug treatment, the initial tumor consisted entirely of sensitive cells that are drug sensitive.
+However, aneuploidy is likely produced even before the onset of treatment at some rate $\tilde{u}$ and confers a deleterious fitness effect $c$ in the absence of the drug \citep{replogle2020aneuploidy,giam2015aneuploidy}.
 Furthermore, the aneuploidy rate in the presence of drugs is likely higher than in their absence, $\tilde{u} < u$ \citep{wang2019molecular,mason2017functional}.
-But if the number of cells in the tumor $N$ is large (as expected if the tumor is treated with a drug), there may already be a fraction $f \approx \tilde{u}\lambda_w/s$ of aneuploid cells in the population (here we assume that the drug affects the wildtype death rate but not the division rate and therefore we use $\lambda_w$ for the wildtype division rate in the drug-free environment).
-
-Therefore, the threshold tumor size for rescue by standing generation variation, $\tilde{N}^*_{a}$, is similar to the threshold for rescue by de-novo variation, $N^*_a$, except that the sensitive growth rate $\abs{\Delta_w}$ is replaced by the aneuploidy cost $s$, such that
+But if the number of cells in the tumor $N$ is large (as expected if the tumor is treated with a drug), there may already be a fraction $f \approx \tilde{u}\lambda_s/c$ of aneuploid cells in the population (here we assume that the drug affects the sensitive death rate but not the division rate and therefore we use $\lambda_s$ for the sensitive division rate in the drug-free environment).
+
+Therefore, the threshold tumor size for rescue by standing generation variation, $\tilde{N}^*_{a}$, is similar to the threshold for rescue by de-novo variation, $N^*_a$, except that the sensitive growth rate $\abs{\Delta_s}$ is replaced by the aneuploidy cost $c$, such that
 \begin{equation}\label{eq:Nsgv}
-\frac{\tilde{N}^*_{a}}{N^*_{a}} = \frac{u}{\tilde{u}} \; \frac{s}{\abs{\Delta_w}}.
+\frac{\tilde{N}^*_{a}}{N^*_{a}} = \frac{u}{\tilde{u}} \; \frac{c}{\abs{\Delta_s}}.
 \end{equation}
 Comparing this approximation of $\tilde{N}^*_{a}/N^*_{a}$ to results of stochastic simulations, we find that the approximations perform very well (\Cref{rescue_denovo}). 
-Standing genetic variation will drive evolutionary rescue if wildtype cells die rapidly (growth rate $\Delta_w$ is very negative) due to a strong effect of the drug on sensitive cells, or if the aneuploidy cost in the drug-free environment, $s$, is small.
-In contrast, \emph{de-novo} aneuploid cells will have a greater contribution to rescue if the aneuploidy cost $s$ is large, the effect of the drug on sensitive cells is weak ($\Delta_w$ is close to zero), or if the drug induces the appearance of aneuploid cells ($u > \tilde u$).
-For example, with  $\lambda_w=0.1,\mu_w=0.14, u=10^{-2}, \tilde{u}=2\times10^{-3}, s=0.07$, the ratio of the threshold tumor sizes for standing vs.
-\emph{de-novo} variation is $\tilde{N}^*_a/N^*_a \approx 8.75$, which means that \emph{de-novo} genetic variation is  the main driver of evolutionary rescue.
+Standing genetic variation will drive evolutionary rescue if sensitive cells die rapidly (growth rate $\Delta_s$ is very negative) due to a strong effect of the drug on sensitive cells, or if the aneuploidy cost in the drug-free environment, $c$, is small.
+In contrast, \emph{de-novo} aneuploid cells will have a greater contribution to rescue if the aneuploidy cost $c$ is large, the effect of the drug on sensitive cells is weak ($\Delta_s$ is close to zero), or if the drug induces the appearance of aneuploid cells ($u > \tilde u$).
+For example, with  $\lambda_s=0.1,\mu_s=0.14, u=10^{-2}, \tilde{u}=10^{-3}, c=0.07$, the ratio of the threshold tumor sizes for standing vs.
+\emph{de-novo} variation is $\tilde{N}^*_a/N^*_a \approx 17.5$, which means that \emph{de-novo} genetic variation is  the main driver of evolutionary rescue.
 
@@
 \begin{equation} \label{eq:N_sgv_ratio}
-\frac{\tilde{N}^*_a}{N^*_m}=\frac{\tilde{N}^*_{a}}{N^*_{a}} \frac{N^*_a}{N^*_m} \approx \frac{s}{\abs{\Delta_w}}\begin{cases}
+\frac{\tilde{N}^*_a}{N^*_m}=\frac{\tilde{N}^*_{a}}{N^*_{a}} \frac{N^*_a}{N^*_m} \approx \frac{c}{\abs{\Delta_s}}\begin{cases}
     \frac{\abs{\Delta_a}}{\tilde{u}\lambda_a} ,&
@@
 \end{equation}
-Evolutionary rescue through direct mutation is more likely if the cost of aneuploidy $s$ is very large or the effect of the drug $\Delta_w$ is small.
+Evolutionary rescue through direct mutation is more likely if the cost of aneuploidy $c$ is very large or the effect of the drug $\Delta_s$ is small.
 In contrast, standing genetic variation will drive adaptation if the pre-treatment chromosome missegreagation rate $\tilde{u}$ is very large.
@@
 However, if the aneuploid growth rate $\Delta_a$ increases, then evolutionary rescue is driven by standing genetic variation.
-For the parameter values of  $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, \tilde{u}=10^{-3}, v=10^{-7}$, we are in the tolerant aneuploid scenario and obtain the ratio $\tilde{N}^*_a/N^*_m \approx 0.9625$, which means that standing genetic variation reduces the threshold tumor size by approximately $4\%$.
-Therefore, standing genetic variation does not drive evolution of drug resistance when compared to \emph{de-novo} aneuploidy, but it does offer a slight advantage when compared with direct mutation.
+For the parameter values of  $\lambda_s=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, \tilde{u}=10^{-3}, v=10^{-7}$, we are in the tolerant aneuploid scenario and obtain the ratio $\tilde{N}^*_a/N^*_m \approx 1.94$, which means that standing genetic variation does not drive evolution of drug resistance when compared to direct mutation.
+We note that for larger values of the pre-treatment chromosome missegregation rate, which are consistent with empirical studies (see \Cref{table1}), standing genetic variation can drive adaptation when compared to direct mutaion.
 
@@
 ok
-Of particular interest is the distribution of the evolutionary rescue time and recurrence time with tolerant aneuploid cells ($\Delta_a T^*\ll1$), for which we focus on the parameter values  $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$ .
+Of particular interest is the distribution of the evolutionary rescue time and recurrence time with tolerant aneuploid cells ($\Delta_a T^*\ll1$), for which we focus on the parameter values  $\lambda_s=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$ .
 
@@
 \tau_a \approx \begin{cases}
-    -\frac{1}{\Delta_w}-\frac{1}{\Delta_a} ,&
+    -\frac{1}{\Delta_s}-\frac{1}{\Delta_a} ,&
  N \ll N_a^* ,\\ 
-  \frac{1}{v\lambda_w N}   \frac{\lambda_m}{\Delta_m} ,& % I replaced pm with 𝜆𝑚 − 𝜇𝑚 / 𝜆𝑚
+  \frac{1}{v\lambda_s N}   \frac{\lambda_m}{\Delta_m} ,& % I replaced pm with 𝜆𝑚 − 𝜇𝑚 / 𝜆𝑚
   N \gg N_m^* .
@@
 
-For small tumors ($ N \ll N_a^*$), the mean rescue time is a function of the wildtype and aneuploid growth rates and independent of the other model parameters, including tumor size (blue line in \Cref{EvolutionaryRescueTimeComplete}).
-Increasing the wildtype or aneuploid growth rates leads to an increase in the mean rescue time, because the corresponding cells will survive for longer and will produce additional rescue mutations at latter times.
-In our focus parameter regime, we have $\Delta_w=-0.04$ and $\Delta_a=-10^{-4}$, such that the mean rescue time is mainly determined by the aneuploid growth rate, $\tau_a \approx 10^4$ days.
+For small tumors ($ N \ll N_a^*$), the mean rescue time is a function of the sensitive and aneuploid growth rates and independent of the other model parameters, including tumor size (blue line in \Cref{EvolutionaryRescueTimeComplete}).
+Increasing the sensitive or aneuploid growth rates leads to an increase in the mean rescue time, because the corresponding cells will survive for longer and will produce additional rescue mutations at latter times.
+In our focus parameter regime, we have $\Delta_s=-0.04$ and $\Delta_a=-10^{-4}$, such that the mean rescue time is mainly determined by the aneuploid growth rate, $\tau_a \approx 10^4$ days.
 
@@
 Increasing the per division mutation rate, $v$, leads to faster appearance of a rescue mutation and hence reduced mean rescue time. 
-Finally, increasing the tumor size leads to shorter mean rescue time, as there are more wildtype cells that can mutate to become resistant.
-
-Given that a fraction $f\approx 0.28\%$ of the initial cancer cell population is expected to have the beneficial aneuploidy even before the onset of drug treatment, we want to know whether the mean evolutionary rescue time is affected by the standing genetic variation.
+Finally, increasing the tumor size leads to shorter mean rescue time, as there are more sensitive cells that can mutate to become resistant.
+
+Given that a fraction $f\approx 0.14\%$ of the initial cancer cell population is expected to have the beneficial aneuploidy even before the onset of drug treatment, we want to know whether the mean evolutionary rescue time is affected by the standing genetic variation.
 We calculated the mean evolutionary rescue time with standing genetic variation $\tilde{\tau}_a$ (\cref{meantimet2SGV}) and compared our result with simulations in  \Cref{SGVEvolutionaryRescueTimeComplete}.
@@
 \tau_a^r \approx \begin{cases}
-   -\frac{1}{\Delta_w}-\frac{1}{\Delta_a}+\frac{\log p_mN}{\Delta_m} ,&
+   -\frac{1}{\Delta_s}-\frac{1}{\Delta_a}+\frac{\log p_mN}{\Delta_m} ,&
  N \ll N_a^* ,\\ 
-  \frac{1}{\Delta_m}\log\frac{\Delta_m-\Delta_w}{v\lambda_w}  ,&
+  \frac{1}{\Delta_m}\log\frac{\Delta_m-\Delta_s}{v\lambda_s}  ,&
   N \gg N_m^* .
@@
 The mean recurrence time grows logarithmically with tumor size $N$ and is the same order of magnitude as the mean evolutionary rescue time.
-Increasing the mutant growth rate $\Delta_m$ decreases recurrence times while increasing the wildtype and aneuploid growth rates, $\Delta_w$ and $\Delta_a$ respectively, increases the recurrence time.
+Increasing the mutant growth rate $\Delta_m$ decreases recurrence times while increasing the sensitive and aneuploid growth rates, $\Delta_s$ and $\Delta_a$ respectively, increases the recurrence time.
 For large tumors ($N \gg N_m^*$), the dynamics of the number of mutant cells is deterministic and the mean recurrence time becomes independent of the initial tumor size $N$.
 Increasing either the mutant growth rate $\Delta_m$ or the mutation rate $v$ leads to a decrease in the time for the tumor to rebound to its initial size.
-In addition, drugs that significantly increase the wildtype death rate $\mu_w$ and do not affect the division rate $\lambda_w$ delay cancer recurrence.
+In addition, drugs that significantly increase the sensitive death rate $\mu_s$ and do not affect the division rate $\lambda_s$ delay cancer recurrence.
 Consequently, patients treated with such drugs may require a longer period of monitoring to guarantee the effectiveness of the treatment.
@@
 Additionally, we derive the distribution of the recurrence time for the scenario $N=10^6$ (small tumor), noting that the distribution is wide and right-skewed (\Cref{KMdistribution}).
-It is highly unlikely to observe the recurrence of tumors at times smaller then $\frac{1}{\Delta_m}\log\frac{\Delta_m-\Delta_w}{v\lambda_w}\approx1542$ days for the parameter values $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$ and independent of initial tumor size $N$~(\Cref{cdffig}B).
+It is highly unlikely to observe the recurrence of tumors at times smaller then $\frac{1}{\Delta_m}\log\frac{\Delta_m-\Delta_s}{v\lambda_s}\approx1542$ days for the parameter values $\lambda_s=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$ and independent of initial tumor size $N$~(\Cref{cdffig}B).
 
@@
 We obtained exact and approximate expressions for the probability of evolutionary rescue (\cref{eq:rescue_prob}). 
-Our results show that the probability of evolutionary rescue increases with the initial tumor size $N$, the sensitive growth rate $\Delta_w$, the mutation rate $v$, and the aneuploidy rate $u$.
-
-When aneuploid cells are partially resistant to the drug ($\Delta_w\ll0\ll\Delta_a\ll\Delta_m$), aneuploidy itself rescues the population (\Cref{rescue_threshold}A). 
-When aneuploidy only provides tolerance to the drug ($\Delta_w\ll\Delta_a\ll0\ll\Delta_m$), it cannot rescue the population.
+Our results show that the probability of evolutionary rescue increases with the initial tumor size $N$, the sensitive growth rate $\Delta_s$, the mutation rate $v$, and the aneuploidy rate $u$.
+
+When aneuploid cells are partially resistant to the drug ($\Delta_s\ll0\ll\Delta_a\ll\Delta_m$), aneuploidy itself rescues the population (\Cref{rescue_threshold}A). 
+When aneuploidy only provides tolerance to the drug ($\Delta_s\ll\Delta_a\ll0\ll\Delta_m$), it cannot rescue the population.
 Instead, it acts as a \emph{stepping stone} through which the resistant mutant can appear more rapidly, given that the aneuploid cell population size declines slower then that of the sensitive cell population (\Cref{sampleTrajectories}).
@@
 First, it delays the extinction of the population, providing more time for appearance of the resistance mutation.
-Second, it increases the population size relative to a sensitive population, providing more cells in which mutations can occur, i.e., it increases the mutation supply (i.e., $Nuv\lambda_w\lambda_a/\abs{\Delta_w\Delta_a}$).
+Second, it increases the population size relative to a sensitive population, providing more cells in which mutations can occur, i.e., it increases the mutation supply (i.e., $Nuv\lambda_s\lambda_a/\abs{\Delta_s\Delta_a}$).
 
@@
 The theoretical prediction for the mean rescue time for tumors smaller then  $10^8$ cells is greater than 4 years, consistent with previous estimates of the recurrence time of tumors after resection~\citep{avanzini2019cancer}.
-We found that aneuploidy complements evolutionary rescue through direct mutation because it generates rescue mutations mostly after the wildtype population has died out and direct mutation is no longer a viable option for evolutionary rescue (\Cref{cdffig}A).
+We found that aneuploidy complements evolutionary rescue through direct mutation because it generates rescue mutations mostly after the sensitive population has died out and direct mutation is no longer a viable option for evolutionary rescue (\Cref{cdffig}A).
 
 We hypothesized that standing genetic variation (the existence of a subpopulation of aneuploid cancer cells before the onset of therapy) could facilitate evolutionary rescue by reducing the waiting time for the appearance of aneuploid cells.
-We found that a drug that reduces the wildtype growth rate and does not significantly increase the chromosome missegregation rate is likely to lead to evolutionary rescue through standing genetic variation (\Cref{rescue_denovo,eq:Nsgv}).
-Furthermore, if the fraction of tumor cells that have the beneficial aneuploidy is $f\gg\frac{u\lambda_w}{\abs{\Delta_w}}\approx2.5\%$, then evolutionary rescue is more likely to occur via standing variation rather then through \emph{de-novo} aneuploidy.
+We found that a drug that reduces the sensitive growth rate and does not significantly increase the chromosome missegregation rate is likely to lead to evolutionary rescue through standing genetic variation (\Cref{rescue_denovo,eq:Nsgv}).
+Furthermore, if the fraction of tumor cells that have the beneficial aneuploidy is $f\gg\frac{u\lambda_s}{\abs{\Delta_s}}\approx2.5\%$, then evolutionary rescue is more likely to occur via standing variation rather then through \emph{de-novo} aneuploidy.
 However, for the parameter values we focus on in our examples (\Cref{table1}), this fraction is an order of magnitude lower, and therefore we expect evolutionary rescue to occur by \emph{de-novo} aneuploidy.
@@
     $N$ & Initial tumor size & $10^7-10^9$ & cells  & \citet{del2009does} \\ \hline
-    $\lambda_w$ & Wildtype division rate& 0.1 & 1/days  & \citet{bozic2013evolutionary,rew2000cell} \\ \hline
-    $\mu_w$ & Wildtype death rate& $0.11-0.17$ & 1/days  & \citet{bozic2013evolutionary} \\ \hline
+    $\lambda_s$ & Sensitive division rate& 0.1 & 1/days  & \citet{bozic2013evolutionary,rew2000cell} \\ \hline
+    $\mu_s$ & Sensitive death rate& $0.11-0.17$ & 1/days  & \citet{bozic2013evolutionary} \\ \hline
     $\lambda_a$  & Aneuploid division rate$^\ast$ & $0.06-0.1$ & 1/days  & - \\ \hline
@@
     $\mu_m$ & Mutant death rate& 0.09 & 1/days  & \citet{bozic2013evolutionary,carlson2003tumor} \\ \hline
-    $u$ & Missegregation rate& $10^{-3}-10^{-2}$ & 1$\slash$cell division  & \citet{bakker2023predicting} \\ \hline
+    $u$ & Missegregation rate& $10^{-2}$ & 1$\slash$cell division  & \citet{lee2016effects} \\ \hline
     $v$ & Mutation rate& $10^{-9}-10^{-7}$ &  1$\slash$cell division  & \citet{bozic2013evolutionary,loeb2001mutator} \\  \hline
-    $\tilde{u}$ & Missegregation rate in the drug free environment$^\ast$& $2\times10^{-3}$ & 1$\slash$cell division  & - \\ \hline
+    $\tilde{u}$ & Missegregation rate in the drug free environment$^\ast$& $5\times10^{-4}-2\times10^{-2}$ & 1$\slash$cell division  & \citet{shi2005chromosome,thompson2008examining} \\ \hline
     $s$ & Selection coefficient of aneuploidy in the drug free environment& $0.07$ &  1/days   & \citet{lukow2021chromosomal} \\  
@@
 \caption{\textbf{Model parameters.} 
-Parameters from \citet{bozic2013evolutionary} consider patiens with melanoma treated with the anti-cancer drug vemurfenib, in which resistance is conferered by trisomy in either Chr 2 or Chr 6.
-We have modified the parameters from \citet{bozic2013evolutionary} such that wildtype and mutant division rates are $\lambda_{w} = \lambda_{m} = \log2/T\approx0.1$ instead of their value of $0.14$ where $T$ is the doubling time in the absence of cellular death obtained from \citet{rew2000cell}.}
+Parameters from \citet{bozic2013evolutionary} consider patients with melanoma treated with the anti-cancer drug vemurfenib, in which resistance is conferered by trisomy in either Chr 2 or Chr 6.
+We have modified the parameters from \citet{bozic2013evolutionary} such that sensitive and mutant division rates are $\lambda_{s} = \lambda_{m} = \log2/T\approx0.1$ instead of their value of $0.14$ where $T$ is the doubling time in the absence of cellular death obtained from \citet{rew2000cell}.
+For a discussion of the different interpretations of the tumor doubling times see \citet{avanzini2019cancer}.}
   \label{table1}
@@
 \textbf{Model illustration.}
-\textbf{(A)} A population of cancer cells is composed of wildtype, aneuploid, and mutant cells, which divide with rates $\lambda_w$, $\lambda_a$, and $\lambda_m$ and die at rates $\mu_w$, $\mu_a$, and $\mu_m$, respectively.
-Wildtype cells can divide and become aneuploid at rate $u\lambda_w$.
-Both aneuploid and wildtype cells can divide and acquire a mutation with rate $v\lambda_a$ and $v\lambda_w$, respectively.
-Color denotes the relative growth rates of the three genotypes such that $\lambda_w - \mu_w < \lambda_a - \mu_a < \lambda_m - \mu_m$.
-\textbf{(B)} The wildtype and the mutant are susceptible and resistant, respectively, to the drug.
+\textbf{(A)} A population of cancer cells is composed of sensitive, aneuploid, and mutant cells, which divide with rates $\lambda_s$, $\lambda_a$, and $\lambda_m$ and die at rates $\mu_s$, $\mu_a$, and $\mu_m$, respectively.
+Sensitive cells can divide and become aneuploid at rate $u\lambda_s$.
+Both aneuploid and sensitive cells can divide and acquire a mutation with rate $v\lambda_a$ and $v\lambda_s$, respectively.
+Color denotes the relative growth rates of the three genotypes such that $\lambda_s - \mu_s < \lambda_a - \mu_a < \lambda_m - \mu_m$.
+\textbf{(B)} Sensitive cells are susceptible to the drug, while mutant cells are resistant.
 The aneuploid may be tolerant, stationary and partially resistant.
@@
 (A) Without aneuploidy ($u=0$) evolutionary rescue is possible through direct mutation and in most scenarios the tumor will become extinct due to the drug.
-(B) When aneuploid cells are tolerant ($\Delta_a<0$), we observe that, similar to A, direct mutation is the only path for evolutionary rescue.
-(C) When aneuploid cells are stationary ($\Delta_a\approx0$), we observe the appearance of aneuploid lineages even after the wildtype population has gone extinct, thus showing that stationary aneuploidy increases the probability of evolutionary rescue.
+(B) When aneuploid cells are tolerant ($\Delta_a<0$), we observe that mutant lineages are generated at later times then in A.
+(C) When aneuploid cells are stationary ($\Delta_a\approx0$), we observe the appearance of mutant lineages even after the sensitive population has gone extinct, thus showing that stationary aneuploidy increases the probability of evolutionary rescue.
 (D) When aneuploid cells are partially resistant ($\Delta_a>0$), the tumor is rescued by the aneuploid cell population.
-Each plot shows 10 simulations of the number of wildtype, aneuploid and mutant cells $\left(w_t,a_t,m_t\right)$ over time $t$ with the following parameter values: $\lambda_w=0.1,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, v=10^{-7},N=10^7$.
-For (A) we set $u=0$, for (B) $\lambda_a=0.065,u=10^{-2}$, for (C) $\lambda_a=0.0899,u=10^{-2}$ and for (D) $\lambda_a=0.095,u=10^{-2}$.
+Each plot shows 10 simulations of the number of sensitive, aneuploid and mutant cells $\left(s_t,a_t,m_t\right)$ over time $t$ with the following parameter values: $\lambda_s=0.1,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, v=10^{-7},N=10^7$.
+For (A) we set $u=0$, for (B) $\lambda_a=0.085,u=10^{-2}$, for (C) $\lambda_a=0.0899,u=10^{-2}$ and for (D) $\lambda_a=0.095,u=10^{-2}$.
 }
@@
 Dots for simulations and the error bars for $95\%$ confidence interval ($p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the fraction of simulations in which the tumor has been rescued and $n=100$ is the number of simulations).
-Parameters: $\lambda_w=0.1,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, v=10^{-7}$.}
+Parameters: $\lambda_s=0.1,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, v=10^{-7}$.}
 \label{rescue_prob}
@@
 Red dots for simulations and error bars for the $95\%$ confidence intervals obtained with bootstrap (Appendix G).
-Parameters: $\lambda_w=0.1,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.
+Parameters: $\lambda_s=0.1,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.
 (B) Threshold tumor size $N_a^*$ as a function of the ratio of aneuploidy and mutation rates, $u/v$.
@@
 Red dots represents simulation results and the error bars represent the $95\%$ confidence intervals obtained with bootstrap (Appendix G).
-Parameters: $\lambda_w=0.1,\lambda_m=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, v=10^{-7}$.
+Parameters: $\lambda_s=0.1,\lambda_m=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, v=10^{-7}$.
 }
@@
 A\\
-\includegraphics[width=1\textwidth]{Figures/RatiodwPlot.pdf}
+\includegraphics[width=1\textwidth]{Figures/RatiodsPlot.pdf}
 \end{subfigure}
@@
 \textbf{Standing genetic variation facilitates evolutionary rescue of cancer.}
-(A)  Ratio of threshold tumor sizes for rescue by standing genetic variation and by de-novo variation, $\tilde{N}_a^*/N_a^*$, when a fraction $\frac{\tilde{u}\lambda_w}{s}$ is aneuploid at the start of treatment, as a function of the wildtype growth rate $\Delta_w$.
-Standing genetic variation will drive adaptation to the drug if the wildtype population is rapidly declining ($\Delta_w\ll0$) due to a stronger effect of the drug on sensitive cells.
+(A)  Ratio of threshold tumor sizes for rescue by standing genetic variation and by de-novo variation, $\tilde{N}_a^*/N_a^*$, when a fraction $\frac{\tilde{u}\lambda_s}{c}$ is aneuploid at the start of treatment, as a function of the sensitive growth rate $\Delta_s$.
+Standing genetic variation will drive adaptation to the drug if the sensitive population is rapidly declining ($\Delta_s\ll0$) due to a stronger effect of the drug on sensitive cells.
 Red dots represents simulation results and the error bars represent the $95\%$ confidence intervals obtained with bootstrap (Appendix G).
-Parameters: $\lambda_w=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_a=0.09,\mu_m=0.09,\tilde{u}=10^{-3},u=10^{-2}, v=10^{-7}$.
-(B) Ratio of threshold tumor size $\tilde{N}_a^*$ when a fraction $\frac{\tilde{u}\lambda_w}{s}$ is aneuploid at the start of treatment and $N_a^*$ as a function of the the ratio of aneuploidy rates $\tilde{u}/u$.
+Parameters: $\lambda_s=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_a=0.09,\mu_m=0.09,\tilde{u}=10^{-3},u=10^{-2}, v=10^{-7}$.
+(B) Ratio of threshold tumor size $\tilde{N}_a^*$, when a fraction $\frac{\tilde{u}\lambda_s}{c}$ is aneuploid at the start of treatment, and $N_a^*$ as a function of the the ratio of aneuploidy rates $\tilde{u}/u$.
 \emph{De-novo} aneuploids will have larger contribution to the appearance of drug resistance if the drug induces the appearance of aneuploid cells ($u \gg \tilde u$).
 Red dots represents  simulation results and the error bars represent the $95\%$ confidence intervals obtained with bootstrap (Appendix G).
-Parameters: $\lambda_w=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09,\tilde{u}=10^{-3}, v=10^{-7}$.
+Parameters: $\lambda_s=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09,\tilde{u}=10^{-3}, v=10^{-7}$.
 }
@@
 Black line: no aneuploidy ($u=0$).
-Aneuploidy starts to play an important role in rescuing the tumor cell population as the wildtype population becomes extinct.
+Aneuploidy starts to play an important role in rescuing the tumor cell population as the sensitive population becomes extinct.
 Markers represent simulation results and the error bars represent $95\%$ confidence interval ($p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the fraction of simulations in which a successful mutant has not been generated and $n=100$ is the number of simulations).
-Parameters: $\lambda_w=0.1,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7},N=10^7$.
+Parameters: $\lambda_s=0.1,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7},N=10^7$.
 (B) Probability that a mutant cancer cell population has not reached size $N$ at time $t$.
@@
 Markers represent simulations and the error bars represents $95\%$ confidence interval ($p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the fraction of the simulations in which the mutant population size has not reached $N$ and $n=100$ is the number of simulations).
-Parameters: $\lambda_w=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}
+Parameters: $\lambda_s=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}
 \label{cdffig}
@@
 The simulations converge to \cref{eq:t2det} (blue dashed line) for large values of $N$.
-Parameters: $\lambda_w=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}
+Parameters: $\lambda_s=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}
 \label{proliferationFigure}
@@
 
-Let $p_w$, $p_a$, and $p_m$ be the survival probabilities of a population consisting initially of single wildtype cell, aneuploid cell, or mutant cell, respectively.
-The complements $1-p_w$, $1-p_a$, and $1-p_m$ are the extinction probabilities, which satisfy each its respective equation~\citep{harris1963theory},
+Let $p_s$, $p_a$, and $p_m$ be the survival probabilities of a population consisting initially of single sensitive cell, aneuploid cell, or mutant cell, respectively.
+The complements $1-p_s$, $1-p_a$, and $1-p_m$ are the extinction probabilities, which satisfy each its respective equation~\citep{harris1963theory},
 \begin{equation} \label{eq:extinction_prob}
 \begin{aligned}
-1-p_w = &\frac{\mu_w}{\lambda_w+\mu_w+u\lambda_w+v\lambda_w} + 
-		  \frac{u\lambda_w}{\lambda_w+\mu_w+u\lambda_w+v\lambda_w}\left(1-p_a\right)\left(1-p_w\right) + \\
-		  & \frac{\lambda_w}{\lambda_w+\mu_w+u\lambda_w+v\lambda_w}\left(1-p_w\right)^2 +
-		  \frac{v\lambda_w}{\lambda_w+\mu_w+u\lambda_w+v\lambda_w}\left(1-p_m\right)\left(1-p_w\right) ,\\
+1-p_s = &\frac{\mu_s}{\lambda_s+\mu_s+u\lambda_s+v\lambda_s} + 
+		  \frac{u\lambda_s}{\lambda_s+\mu_s+u\lambda_s+v\lambda_s}\left(1-p_a\right)\left(1-p_s\right) + \\
+		  & \frac{\lambda_s}{\lambda_s+\mu_s+u\lambda_s+v\lambda_s}\left(1-p_s\right)^2 +
+		  \frac{v\lambda_s}{\lambda_s+\mu_s+u\lambda_s+v\lambda_s}\left(1-p_m\right)\left(1-p_s\right) ,\\
 1-p_a = &\frac{\mu_a}{\lambda_a+\mu_a+v\lambda_a}+\frac{v\lambda_a}{\lambda_a+\mu_a+v\lambda_a}\left(1-p_m\right)\left(1-p_a\right)+\frac{\lambda_a}{\lambda_a+\mu_a+v\lambda_a}\left(1-p_a\right)^2 ,\\
@@
 \begin{aligned}
-p_w &= \frac{\lambda_w-\mu_w-u\lambda_wp_a-v\lambda_wp_m+\sqrt{\left(\lambda_w-\mu_w-u\lambda_wp_a-v\lambda_wp_m\right)^2+4\lambda_w^2\left(up_a+vp_m\right)}}{2\lambda_w} ,\\
+p_s &= \frac{\lambda_s-\mu_s-u\lambda_sp_a-v\lambda_sp_m+\sqrt{\left(\lambda_s-\mu_s-u\lambda_sp_a-v\lambda_sp_m\right)^2+4\lambda_s^2\left(up_a+vp_m\right)}}{2\lambda_s} ,\\
 p_a &= \frac{\lambda_a-\mu_a-v\lambda_ap_m+\sqrt{\left(\lambda_a-\mu_a-v\lambda_ap_m\right)^2+4\lambda_a^2vp_m}}{2\lambda_a}, \\
@@
 \end{equation}
-Note that the equation for $p_w$ depends on both $p_a$ and $p_m$, and the equation for $p_a$ depends on $p_m$.
-To proceed, we can plug the solution for $p_m$ and $p_a$ into the solution for $p_w$. We perform this for three different scenarios.
+Note that the equation for $p_s$ depends on both $p_a$ and $p_m$, and the equation for $p_a$ depends on $p_m$.
+To proceed, we can plug the solution for $p_m$ and $p_a$ into the solution for $p_s$. We perform this for three different scenarios.
 
@@
 \begin{align*}
-p_w&=\frac{\lambda_w-\mu_w-u\lambda_wp_a-v\lambda_wp_m}{2\lambda_w}\left(1-\sqrt{1+\frac{4\lambda_w^2\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u\lambda_wp_a-v\lambda_wp_m\right)^2}}\right) ,
+p_s&=\frac{\lambda_s-\mu_s-u\lambda_sp_a-v\lambda_sp_m}{2\lambda_s}\left(1-\sqrt{1+\frac{4\lambda_s^2\left(vp_m+up_a\right)}{\left(\lambda_s-\mu_s-u\lambda_sp_a-v\lambda_sp_m\right)^2}}\right) ,
 \text{and} \\
@@
 Using the Taylor expansion $\sqrt{1+x}=1+x/2+\mathcal{O}(x^2)$ and assuming $u,v \ll 1$,
-we obtain the following approximation for the survival probability of a population initially consisting of a single wildtype cell,
+we obtain the following approximation for the survival probability of a population initially consisting of a single sensitive cell,
 \begin{align} \label{eq:survprobwapprox1}
-p_w 
-&\approx -\frac{v\lambda_wp_m+u\lambda_wp_a}{\lambda_w-\mu_w-u\lambda_wp_a-v\lambda_wp_m}\\
+p_s
+&\approx -\frac{v\lambda_sp_m+u\lambda_sp_a}{\lambda_s-\mu_s-u\lambda_sp_a-v\lambda_sp_m}\\
 \nonumber
-&\approx-\frac{1}{\lambda_w-\mu_w}\left[\frac{u\lambda_w\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\lambda_w\lambda_a\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\lambda_w\left(\lambda_m-\mu_m\right)}{\lambda_m}\right].
+&\approx-\frac{1}{\lambda_s-\mu_s}\left[\frac{u\lambda_s\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\lambda_s\lambda_a\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\lambda_s\left(\lambda_m-\mu_m\right)}{\lambda_m}\right].
 \end{align}
@@
 \begin{equation}\label{eq:pw_parttolerant}
-p_w \approx \frac{u\lambda_w}{\abs{\Delta_w}}  \frac{\Delta_a}{\lambda_a} .
+p_s \approx \frac{u\lambda_s}{\abs{\Delta_s}}  \frac{\Delta_a}{\lambda_a} .
 \end{equation}
@@
 \begin{align*}
-\frac{u\lambda_w\Delta_a}{\lambda_a}<\frac{v\lambda_w\Delta_m}{\lambda_m}\Rightarrow u\lambda_a<\frac{v\lambda_a^2\Delta_m}{\lambda_m} \frac{1}{\Delta_a}<\frac{v\lambda_a^2\Delta_m}{\lambda_m} \frac{1}{\sqrt{4\lambda_a^2 v p_m}}\Rightarrow u\lambda_a<T^*,
+\frac{u\lambda_s\Delta_a}{\lambda_a}<\frac{v\lambda_s\Delta_m}{\lambda_m}\Rightarrow u\lambda_a<\frac{v\lambda_a^2\Delta_m}{\lambda_m} \frac{1}{\Delta_a}<\frac{v\lambda_a^2\Delta_m}{\lambda_m} \frac{1}{\sqrt{4\lambda_a^2 v p_m}}\Rightarrow u\lambda_a<T^*,
 \end{align*}
@@
 \begin{equation*}
-p_w \approx \frac{v\lambda_w}{\abs{\Delta_w}}  \frac{\Delta_m}{\lambda_m} .
+p_s \approx \frac{v\lambda_s}{\abs{\Delta_s}}  \frac{\Delta_m}{\lambda_m} .
 \end{equation*}
@@
 
-We now assume that aneuploidy provides tolerance to drug therapy, that is, the number of aneuploid cells significantly declines over time, but at a lower rate than the number of wildtype cells, $\lambda_w - \mu_w < \lambda_a - \mu_a < 0$.
+We now assume that aneuploidy provides tolerance to drug therapy, that is, the number of aneuploid cells significantly declines over time, but at a lower rate than the number of sensitive cells, $\lambda_s - \mu_s < \lambda_a - \mu_a < 0$.
 We also assume that the decline are significant, $\left(\lambda_a-\mu_a-v\lambda_ap_m\right)^2 > 4\lambda_a^2 v p_m$.
@@
 \begin{aligned}
-p_w&=\frac{\lambda_w-\mu_w-u\lambda_wp_a-v\lambda_wp_m}{2\lambda_w}\left(1-\sqrt{1+\frac{4\lambda_w^2\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u\lambda_wp_a-v\lambda_wp_m\right)^2}}\right), \\
+p_s&=\frac{\lambda_s-\mu_s-u\lambda_sp_a-v\lambda_sp_m}{2\lambda_s}\left(1-\sqrt{1+\frac{4\lambda_s^2\left(vp_m+up_a\right)}{\left(\lambda_s-\mu_s-u\lambda_sp_a-v\lambda_sp_m\right)^2}}\right), \\
 p_a&=\frac{\lambda_a-\mu_a-v\lambda_ap_m}{2\lambda_a}\left(1-\sqrt{1+\frac{4\lambda_a^2vp_m}{\left(\lambda_a-\mu_a-v\lambda_ap_m\right)^2}}\right) .
@@
 \begin{aligned}
-p_w&\approx-\frac{v\lambda_wp_m+u\lambda_wp_a}{\lambda_w-\mu_w-u\lambda_wp_a-v\lambda_wp_m}\\
-&\approx\frac{1}{\lambda_w-\mu_w-u\lambda_wp_a-v\lambda_wp_m}\left[\frac{uv\lambda_w\lambda_a\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-v\lambda_a\right)}-\frac{v\lambda_w\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\
-&\approx\frac{v\lambda_w\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_w-\mu_w\right)}\left[\frac{u\lambda_a}{\left(\lambda_a-\mu_a\right)}-1\right] \\
-&=\frac{v\lambda_w\Delta_m}{\lambda_m \abs{\Delta_w}}\left(\frac{u\lambda_a}{\abs{\Delta_a}}+1\right) .
+p_s&\approx-\frac{v\lambda_sp_m+u\lambda_sp_a}{\lambda_s-\mu_s-u\lambda_sp_a-v\lambda_sp_m}\\
+&\approx\frac{1}{\lambda_s-\mu_s-u\lambda_sp_a-v\lambda_sp_m}\left[\frac{uv\lambda_s\lambda_a\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-v\lambda_a\right)}-\frac{v\lambda_s\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\
+&\approx\frac{v\lambda_s\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_s-\mu_s\right)}\left[\frac{u\lambda_a}{\left(\lambda_a-\mu_a\right)}-1\right] \\
+&=\frac{v\lambda_s\Delta_m}{\lambda_m \abs{\Delta_s}}\left(\frac{u\lambda_a}{\abs{\Delta_a}}+1\right) .
 \end{aligned}
@@
 \begin{equation}\label{eq:pw_tolerant}
-p_w\approx\frac{u\lambda_w}{\abs{\Delta_w}} \frac{v\lambda_a}{\abs{\Delta_a}} \frac{\Delta_m}{\lambda_m}.
+p_s\approx\frac{u\lambda_s}{\abs{\Delta_s}} \frac{v\lambda_a}{\abs{\Delta_a}} \frac{\Delta_m}{\lambda_m}.
 \end{equation}
@@
 \end{equation}
-Plugging this in \cref{eq:survprobwapprox1}, the survival probability of a population starting from one wildtype cell is
+Plugging this in \cref{eq:survprobwapprox1}, the survival probability of a population starting from one sensitive cell is
 \begin{equation}\label{eq:scenario3}
 \begin{aligned}
-p_w&\approx-\frac{1}{\lambda_w-\mu_w-u\lambda_wp_a-v\lambda_wp_m}\left[v\lambda_w\frac{\lambda_m-\mu_m}{\lambda_m}+\frac{u\lambda_w}{2\lambda_a}\left(\lambda_a-\mu_a-v\lambda_ap_m+2\sqrt{\lambda_a^2vp_m}\right)\right]\\
-&=-\frac{1}{\lambda_w-\mu_w-u\lambda_w-v\lambda_w}\left[v\lambda_w\frac{\lambda_m-\mu_m}{\lambda_m}+\frac{u\lambda_w}{2\lambda_a}\left(\lambda_a-\mu_a-v\lambda_ap_m\right)+u\lambda_w\sqrt{\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}}\right]\\
-&\approx-\frac{1}{\Delta_w}\left[v\lambda_w\frac{\Delta_m}{\lambda_m}+\frac{u\lambda_w\left(\Delta_a-v\lambda_a\right)}{2\lambda_a}+u\lambda_w\sqrt{\frac{v\Delta_m}{\lambda_m}}\right].
+p_s&\approx-\frac{1}{\lambda_s-\mu_s-u\lambda_sp_a-v\lambda_sp_m}\left[v\lambda_s\frac{\lambda_m-\mu_m}{\lambda_m}+\frac{u\lambda_s}{2\lambda_a}\left(\lambda_a-\mu_a-v\lambda_ap_m+2\sqrt{\lambda_a^2vp_m}\right)\right]\\
+&=-\frac{1}{\lambda_s-\mu_s-u\lambda_s-v\lambda_s}\left[v\lambda_s\frac{\lambda_m-\mu_m}{\lambda_m}+\frac{u\lambda_s}{2\lambda_a}\left(\lambda_a-\mu_a-v\lambda_ap_m\right)+u\lambda_s\sqrt{\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}}\right]\\
+&\approx-\frac{1}{\Delta_s}\left[v\lambda_s\frac{\Delta_m}{\lambda_m}+\frac{u\lambda_s\left(\Delta_a-v\lambda_a\right)}{2\lambda_a}+u\lambda_s\sqrt{\frac{v\Delta_m}{\lambda_m}}\right].
 \end{aligned}
@@
 \begin{equation}\label{eq:pw_partrest}
-p_w\approx\frac{u\lambda_w}{\abs{\Delta_w}} \sqrt{\frac{v\lambda_a\Delta_m}{\lambda_m}}.
+p_s\approx\frac{u\lambda_s}{\abs{\Delta_s}} \sqrt{\frac{v\lambda_a\Delta_m}{\lambda_m}}.
 \end{equation}
@@
   &\begin{cases}
-   1-\exp\left[-\frac{u\lambda_a}{\abs{\Delta_w}} \frac{v\lambda_w}{\abs{\Delta_a}} \frac{\Delta_m}{\lambda_m}  N\right] ,&
+   1-\exp\left[-\frac{u\lambda_a}{\abs{\Delta_s}} \frac{v\lambda_s}{\abs{\Delta_a}} \frac{\Delta_m}{\lambda_m}  N\right] ,&
    \Delta_aT^*\ll-1 ,\\
-   1-\exp\left[-\frac{u\lambda_w}{\abs{\Delta_w}} \sqrt{\frac{v\lambda_a\Delta_m}{\lambda_m}} N\right] ,&
+   1-\exp\left[-\frac{u\lambda_s}{\abs{\Delta_s}} \sqrt{\frac{v\lambda_a\Delta_m}{\lambda_m}} N\right] ,&
   -1\ll\Delta_aT^*\ll1 ,\\
-   1-\exp\left[-\frac{u\lambda_w}{\abs{\Delta_w}}  \frac{\Delta_a}{\lambda_a}  N\right] ,&
+   1-\exp\left[-\frac{u\lambda_s}{\abs{\Delta_s}}  \frac{\Delta_a}{\lambda_a}  N\right] ,&
    1\ll\Delta_aT^*.
@@
 We first calculate the expected time for the appearance of the first mutant that rescues the cell population.
-This can occur either through the evolutionary trajectory $wildtype \rightarrow mutant$ or through the trajectory $wildtype \rightarrow aneuploid \rightarrow mutant$.
+This can occur either through the evolutionary trajectory $sensitive \rightarrow mutant$ or through the trajectory $sensitive \rightarrow aneuploid \rightarrow mutant$.
 We start with the former.
@@
 
-The number of successful mutants generated until time $t$ can be approximated by an inhomogeneous Poisson process with rate $R_m\left(t\right) = v\lambda_w p_m w_t$,
-where $w_t=N\e^{\Delta_w t}$ is the number of wildtype cells at time $t$.
+The number of successful mutants generated until time $t$ can be approximated by an inhomogeneous Poisson process with rate $R_m\left(t\right) = v\lambda_s p_m w_t$,
+where $s_t=N\e^{\Delta_s t}$ is the number of sensitive cells at time $t$.
 Note that 
@@
 \int_0^{t}{R_m(z)\d z} = 
-v\lambda_w p_m N \frac{\exp[{\Delta_w t}]-1}{\Delta_w} \approx 
-v\lambda_w p_m N t,
-\end{equation}
-by integrating the exponential and because $\frac{\exp[\Delta_w t]-1}{\Delta_w}=\frac{1+\Delta_w t+\mathcal{O}(t^2)-1}{\Delta_w}=t+O(t^2)$.
+v\lambda_s p_m N \frac{\exp[{\Delta_s t}]-1}{\Delta_s} \approx 
+v\lambda_s p_m N t,
+\end{equation}
+by integrating the exponential and because $\frac{\exp[\Delta_s t]-1}{\Delta_s}=\frac{1+\Delta_s t+\mathcal{O}(t^2)-1}{\Delta_s}=t+O(t^2)$.
 The probability density function of $T_m$ is thus
@@
 \tau_m = 
-\int_0^\infty tv\lambda_w N\e^{\Delta_wt}\frac{\e^{-v\lambda_w N p_m\frac{\e^{\Delta_w t}-1}{\Delta_w}} }{1-\left(1-p_w\right)^N} \d t\approx
-\int_{0}^{\infty} tv\lambda_w N\e^{\Delta_wt}\frac{\e^{-v\lambda_w N p_mt} }{1-\e^{-Np_w}}\d t. 
+\int_0^\infty tv\lambda_s N\e^{\Delta_st}\frac{\e^{-v\lambda_s N p_m\frac{\e^{\Delta_s t}-1}{\Delta_s}} }{1-\left(1-p_s\right)^N} \d t\approx
+\int_{0}^{\infty} tv\lambda_s N\e^{\Delta_st}\frac{\e^{-v\lambda_s N p_mt} }{1-\e^{-Np_s}}\d t. 
 \end{align}
@@
 
-When $Nu\lambda_w/\abs{\Delta_w}\gg1$ the aneuploid frequency dynamics is roughly deterministic and therefore can be approximated by 
+When $Nu\lambda_s/\abs{\Delta_s}\gg1$ the aneuploid frequency dynamics is roughly deterministic and therefore can be approximated by 
 \begin{equation}\label{aneuploidpopeq}
-a_t \approx \frac{Nu\lambda_w}{\Delta_w-\Delta_a}\left(\e^{\Delta_wt}-\e^{\Delta_a t}\right).
+a_t \approx \frac{Nu\lambda_s}{\Delta_s-\Delta_a}\left(\e^{\Delta_st}-\e^{\Delta_a t}\right).
 \end{equation}
@@
 \begin{align}\label{eq:twosteplineage}
-r_1\left(t\right)&=v\lambda_ap_m\int_0^ta_{z} \d z = \frac{uv\lambda_w\lambda_aNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_wt}-1}{\Delta_w}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right),\\ \label{eq:twosteplineagedirect}
-r_2\left(t\right)&=v\lambda_wp_m\int_0^tw_{z} \d z = v\lambda_wNp_m\frac{\e^{\Delta_w t}-1}{\Delta_w}.
+r_1\left(t\right)&=v\lambda_ap_m\int_0^ta_{z} \d z = \frac{uv\lambda_s\lambda_aNp_m}{\Delta_s-\Delta_a}\left(\frac{\e^{\Delta_st}-1}{\Delta_s}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right),\\ \label{eq:twosteplineagedirect}
+r_2\left(t\right)&=v\lambda_sp_m\int_0^ts_{z} \d z = v\lambda_sNp_m\frac{\e^{\Delta_s t}-1}{\Delta_s}.
 \end{align}
@@
 &=
-\int_0^\infty t\left(v\lambda_ap_ma_t+v\lambda_wp_mw_t\right)\frac{\exp\left[-\frac{uv\lambda_w\lambda_aNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_w t}-1}{\Delta_w}-\frac{\e^{\Delta_a t}-1}{\Delta_a}\right)-v\lambda_wNp_m\frac{\e^{\Delta_w t}-1}{\Delta_w}\right] }{1-\e^{-Np_w}}\d t,
+\int_0^\infty t\left(v\lambda_ap_ma_t+v\lambda_sp_ms_t\right)\frac{\exp\left[-\frac{uv\lambda_s\lambda_aNp_m}{\Delta_s-\Delta_a}\left(\frac{\e^{\Delta_s t}-1}{\Delta_s}-\frac{\e^{\Delta_a t}-1}{\Delta_a}\right)-v\lambda_sNp_m\frac{\e^{\Delta_s t}-1}{\Delta_s}\right] }{1-\e^{-Np_s}}\d t,
 \end{align}
@@
 \begin{align*}
-\tau_a=\int_0^\infty\e^{-R_a\left(\tau\right)}\,\d\tau=\int_0^\infty\exp\left[-\frac{uv\lambda_w\lambda_aNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_w\tau}-1}{\Delta_w}-\frac{\e^{\Delta_a\tau}-1}{\Delta_a}\right)-v\lambda_wNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}\right]\,\d\tau,
+\tau_a=\int_0^\infty\e^{-R_a\left(\tau\right)}\,\d\tau=\int_0^\infty\exp\left[-\frac{uv\lambda_s\lambda_aNp_m}{\Delta_s-\Delta_a}\left(\frac{\e^{\Delta_s\tau}-1}{\Delta_s}-\frac{\e^{\Delta_a\tau}-1}{\Delta_a}\right)-v\lambda_sNp_m\frac{\e^{\Delta_s\tau}-1}{\Delta_s}\right]\,\d\tau,
 \end{align*}
@@
 \begin{align*}
-\frac{\e^{\Delta_w\tau}-1}{\Delta_w}&=\frac{1+\Delta_w\tau+O(\tau^2)-1}{\Delta_w}=\tau+O(\tau^2).\\
+\frac{\e^{\Delta_s\tau}-1}{\Delta_s}&=\frac{1+\Delta_s\tau+O(\tau^2)-1}{\Delta_s}=\tau+O(\tau^2).\\
 \frac{\e^{\Delta_a\tau}-1}{\Delta_a}&=\frac{1+\Delta_a\tau+O(\tau^2)-1}{\Delta_a}=\tau+O(\tau^2),
@@
 \begin{align}\label{limitapprox3}
-\tau_a\approx\int_0^\infty\e^{-v\lambda_wNp_m\tau}\,\d\tau=\frac{1}{v\lambda_wNp_m}.
+\tau_a\approx\int_0^\infty\e^{-v\lambda_sNp_m\tau}\,\d\tau=\frac{1}{v\lambda_sNp_m}.
 \end{align}
@@
 \begin{align}\nonumber
-\tau_a&\approx\frac{\int_0^\infty tv\lambda_ap_ma_\tau\,\d\tau}{1-\e^{-Np_w}}\approx\frac{uv\lambda_a\lambda_wp_m\abs{\Delta_w+\Delta_a}}{p_w\Delta_a^2\Delta_w^2}\\ \label{limitapprox4}
-&=\frac{1}{\abs{\Delta_w}}+\frac{1}{\abs{\Delta_a}},
+\tau_a&\approx\frac{\int_0^\infty tv\lambda_ap_ma_\tau\,\d\tau}{1-\e^{-Np_s}}\approx\frac{uv\lambda_a\lambda_sp_m\abs{\Delta_s+\Delta_a}}{p_s\Delta_a^2\Delta_s^2}\\ \label{limitapprox4}
+&=\frac{1}{\abs{\Delta_s}}+\frac{1}{\abs{\Delta_a}},
 \end{align}
-where in the last line we used the fact that $1/p_w=N_a^*$ and \Cref{eq:N_a}.
+where in the last line we used the fact that $1/p_s=N_a^*$ and \Cref{eq:N_a}.
 
@@
 \begin{align*}
-r_1^f\left(t\right)&=v\lambda_ap_m\int_0^ta_{z} \d z = \left(1-f\right)\frac{uv\lambda_w\lambda_aNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_wt}-1}{\Delta_w}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right)+fv\lambda_aNp_m\frac{\e^{\Delta_a t}-1}{\Delta_a},\\
-r_2^f\left(t\right)&=v\lambda_wp_m\int_0^tw_{z} \d z = \left(1-f\right)v\lambda_wNp_m\frac{\e^{\Delta_w t}-1}{\Delta_w},
+r_1^f\left(t\right)&=v\lambda_ap_m\int_0^ta_{z} \d z = \left(1-f\right)\frac{uv\lambda_s\lambda_aNp_m}{\Delta_s-\Delta_a}\left(\frac{\e^{\Delta_st}-1}{\Delta_s}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right)+fv\lambda_aNp_m\frac{\e^{\Delta_a t}-1}{\Delta_a},\\
+r_2^f\left(t\right)&=v\lambda_sp_m\int_0^ts_{z} \d z = \left(1-f\right)v\lambda_sNp_m\frac{\e^{\Delta_s t}-1}{\Delta_s},
 \end{align*} 
@@
 \end{align}
-where $R_a^f(t)=r_1^f\left(t\right)+r_2^f\left(t\right)$ and $p_{rescue}=1-\exp\left[-\left(1-f\right)p_wN-fp_aN\right]$.
+where $R_a^f(t)=r_1^f\left(t\right)+r_2^f\left(t\right)$ and $p_{rescue}=1-\exp\left[-\left(1-f\right)p_sN-fp_aN\right]$.
 We plot our approximation in  \Cref{SGVEvolutionaryRescueTimeComplete} together with simulated data.
@@
 We define the proliferation time $\tau_a^p$  to be the time it takes the population of mutant cancer cells to reach the initial tumor size $N$.
-The number of rescue lineages generated by the wildtype population is given by  \cref{eq:twosteplineage} (see \Cref{ExpectedNumberRescueLineages}):
+The number of rescue lineages generated by the sensitive population is given by  \cref{eq:twosteplineage} (see \Cref{ExpectedNumberRescueLineages}):
 \begin{equation*}
-r_1\left(\infty\right)=\frac{uv\lambda_w\lambda_aNp_m}{\abs{\Delta_w}\abs{\Delta_a}}=\frac{N}{N_a^*},
+r_1\left(\infty\right)=\frac{uv\lambda_s\lambda_aNp_m}{\abs{\Delta_s}\abs{\Delta_a}}=\frac{N}{N_a^*},
 \end{equation*}
@@
 
-The second scenario is when the wildtype population produces a large number of rescue lineages in a short period of time. This is given by the condition:
+The second scenario is when the sensitive population produces a large number of rescue lineages in a short period of time. This is given by the condition:
 \begin{align*}
@@
 \begin{aligned}
-\frac{dw}{dt}&=\Delta_ww,\\
-\frac{da}{dt}&=\Delta_aa+u\lambda_ww,\\
-\frac{dm}{dt}&=\Delta_mm+v\lambda_aa+v\lambda_ww.
+\frac{ds}{dt}&=\Delta_ss,\\
+\frac{da}{dt}&=\Delta_aa+u\lambda_ss,\\
+\frac{dm}{dt}&=\Delta_mm+v\lambda_aa+v\lambda_ss.
 \end{aligned}
 \end{equation}
-Solving the system of ODEs for initial condition $\left(w(0), a(0), m(0)\right)=\left(N,0,0\right)$ we obtain:
+Solving the system of ODEs for initial condition $\left(s(0), a(0), m(0)\right)=\left(N,0,0\right)$ we obtain:
 \begin{align*}
-m\left(t\right)=\frac{Nuv\lambda_a\lambda_w}{\Delta_w-\Delta_a}\left[\frac{\e^{\Delta_wt}-\e^{\Delta_mt}}{\Delta_w-\Delta_m}-\frac{\e^{\Delta_at}-\e^{\Delta_mt}}{\Delta_a-\Delta_m}\right]+Nv\lambda_w\frac{\e^{\Delta_wt}-\e^{\Delta_mt}}{\Delta_w-\Delta_m}.
+m\left(t\right)=\frac{Nuv\lambda_a\lambda_s}{\Delta_s-\Delta_a}\left[\frac{\e^{\Delta_st}-\e^{\Delta_mt}}{\Delta_s-\Delta_m}-\frac{\e^{\Delta_at}-\e^{\Delta_mt}}{\Delta_a-\Delta_m}\right]+Nv\lambda_s\frac{\e^{\Delta_st}-\e^{\Delta_mt}}{\Delta_s-\Delta_m}.
 \end{align*}
@@
 \begin{equation}%\label{eq:t2det}
-1=\frac{uv\lambda_a\lambda_w}{\Delta_w-\Delta_a}\left[\frac{\e^{\Delta_w\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_w-\Delta_m}-\frac{\e^{\Delta_a\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_a-\Delta_m}\right]+v\lambda_w\frac{\e^{\Delta_w\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_w-\Delta_m}.
+1=\frac{uv\lambda_a\lambda_s}{\Delta_s-\Delta_a}\left[\frac{\e^{\Delta_s\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_s-\Delta_m}-\frac{\e^{\Delta_a\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_a-\Delta_m}\right]+v\lambda_s\frac{\e^{\Delta_s\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_s-\Delta_m}.
 \end{equation}
@@
 \begin{equation*}
-1=v\lambda_w\frac{\e^{\Delta_m\tau_a^r}}{\abs{\Delta_w-\Delta_m}},
+1=v\lambda_s\frac{\e^{\Delta_m\tau_a^r}}{\abs{\Delta_s-\Delta_m}},
 \end{equation*}
@@
 \begin{equation}\label{eq:t2det}
-\tau_a^r\approx\frac{1}{\Delta_m}\log\frac{\Delta_m-\Delta_w}{v\lambda_w}.
-\end{equation}
-We observe that the terms given by the evolutionary trajectory $wildtype \rightarrow aneuploid \rightarrow mutant$ do not contribute to the above approximation and, as a result, we deduce that it accurate only for $N\gg N_m^*>N_a^*$.
+\tau_a^r\approx\frac{1}{\Delta_m}\log\frac{\Delta_m-\Delta_s}{v\lambda_s}.
+\end{equation}
+We observe that the terms given by the evolutionary trajectory $sensitive \rightarrow aneuploid \rightarrow mutant$ do not contribute to the above approximation and, as a result, we deduce that it accurate only for $N\gg N_m^*>N_a^*$.
 
@@
 \begin{equation}\label{eq:t2det2}
-\tau_a^{r,M}\approx\frac{1}{\Delta_m}\log\frac{M\left(\Delta_m-\Delta_w\right)}{v\lambda_wN},
+\tau_a^{r,M}\approx\frac{1}{\Delta_m}\log\frac{M\left(\Delta_m-\Delta_s\right)}{v\lambda_sN},
 \end{equation}
@@
 &=1-\exp\left\{-\left[r_1\left(t\right)+r_2\left(t\right)\right]\right\}\\
-&=1-\exp\left\{-\left[\frac{uv\lambda_w\lambda_aNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_wt}-1}{\Delta_w}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right)+ v\lambda_wNp_m\frac{\e^{\Delta_w t}-1}{\Delta_w}\right]\right\},
+&=1-\exp\left\{-\left[\frac{uv\lambda_s\lambda_aNp_m}{\Delta_s-\Delta_a}\left(\frac{\e^{\Delta_st}-1}{\Delta_s}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right)+ v\lambda_sNp_m\frac{\e^{\Delta_s t}-1}{\Delta_s}\right]\right\},
 \end{align*}
@@
 \begin{equation}
-1-P\left(rescue,t\right)=\exp\left\{-\left[\frac{uv\lambda_w\lambda_aNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_wt}-1}{\Delta_w}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right)+ v\lambda_wNp_m\frac{\e^{\Delta_w t}-1}{\Delta_w}\right]\right\}.
+1-P\left(rescue,t\right)=\exp\left\{-\left[\frac{uv\lambda_s\lambda_aNp_m}{\Delta_s-\Delta_a}\left(\frac{\e^{\Delta_st}-1}{\Delta_s}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right)+ v\lambda_sNp_m\frac{\e^{\Delta_s t}-1}{\Delta_s}\right]\right\}.
 \end{equation}
@@
 The proliferation time is defined as the first time the size of all lineages reaches $N$.
-When $N\ll\abs{\Delta_w}\abs{\Delta_a}/uv\lambda_w\lambda_ap_m$ there is at most a single mutant lineage that will survive and reach size $N$ (\Cref{ExpectedNumberRescueLineages}) and the probability that the size of that lineage has not reached $N$ by time $t$ is given by:
+When $N\ll\abs{\Delta_s}\abs{\Delta_a}/uv\lambda_s\lambda_ap_m$ there is at most a single mutant lineage that will survive and reach size $N$ (\Cref{ExpectedNumberRescueLineages}) and the probability that the size of that lineage has not reached $N$ by time $t$ is given by:
 \begin{align}\nonumber
 P\left(m_t\leq N\right)&=\exp\left[-r\left(t\right)\right]\\ \label{eqmNs}
-&=\exp\left[-\frac{Nuv\lambda_w\lambda_ap_m}{\Delta_w-\Delta_a}\int_0^t\left[\e^{\Delta_ws}-\e^{\Delta_as}\right]\e^{-p_mN\e^{-\Delta_m\left(t-s\right)}}\,\d s\right].
+&=\exp\left[-\frac{Nuv\lambda_s\lambda_ap_m}{\Delta_s-\Delta_a}\int_0^t\left[\e^{\Delta_sx}-\e^{\Delta_ax}\right]\e^{-p_mN\e^{-\Delta_m\left(t-x\right)}}\,\d x\right].
 \end{align}
-When $N\gg\abs{\Delta_w}\abs{\Delta_a}/uv\lambda_w\lambda_ap_m$ the dynamics of the cancer cell populations is deterministic and approximated by the system of ODEs shown in \cref{detODE}.
+When $N\gg\abs{\Delta_s}\abs{\Delta_a}/uv\lambda_s\lambda_ap_m$ the dynamics of the cancer cell populations is deterministic and approximated by the system of ODEs shown in \cref{detODE}.
 As a result, the size of the mutant cell population will always be below $N$ until time $\tau_a^r$ and will always be greater after:
@@
 We observe that for $N=10^7$ our formula overestimates the probability that the mutant population will be smaller then $N$ at time $t$.
-This can be explained by the fact that $N=10^7$ is an intermediary scenario where the wildtype population produces a number of rescue lineages that is greater then one but still sufficiently small such that stochasticity plays an important role in the population dynamics.
+This can be explained by the fact that $N=10^7$ is an intermediary scenario where the sensitive population produces a number of rescue lineages that is greater then one but still sufficiently small such that stochasticity plays an important role in the population dynamics.
 As a result, the number of mutant cancer cells will reach $N$ faster then the scenario with a single mutant lineage.
@@
 which we plot in \Cref{KMdistribution} and compare with simulations.
-We note that in the scenario $N\gg\abs{\Delta_w}\abs{\Delta_a}/uv\lambda_w\lambda_ap_m$ the distribution becomes the Dirac $\delta$-function~\citep{barton1989elements}.
+We note that in the scenario $N\gg\abs{\Delta_s}\abs{\Delta_a}/uv\lambda_s\lambda_ap_m$ the distribution becomes the Dirac $\delta$-function~\citep{barton1989elements}.
 %%%
-\section*{Appendix G: Bootstrapping}
+\section{Bootstrapping}
 For the mean times the $95\%$ confidence interval is obtained through bootstrapping in the following steps: (1) we simulate $T$ 100 times; (2) we sample with replacement which we store in $T'$; (3) for each element of this sample we obtain $\tau=\mathbb{E}\left[T'\right]$; (4) we repeat steps (2)-(3) 100 times to obtain $\tau$ and we select the upper and lower limits such that $95\%$ of the values of $\tau$ lie in the interval given by the bounds.
 
-For the threshold tumor sizes the $95\%$ confidence interval is obtained through bootstrapping in the following steps: (1) we simulate $p_{rescue}$ 100 times; (2) we sample with replacement which we store in $S$; (3) for each element of this sample we obtain $N_a^*=1/p_w$ using $p_w=-1/N_s\log\left(1-\bar{S}\right)$ where $\bar{S}$ is the mean of $S$ and $N_s$ is an arbitrary value of the initial population size we selected in order to calculate $p_{rescue}$; (4) we repeat steps (2)-(3) 100 times to obtain $N_a^*$ and we select the upper and lower limits such that $95\%$ of the values of $N_a^*$ lie in the interval given by the bounds.
-
-For the ratio of the threshold tumor sizes the $95\%$ confidence interval is obtained through bootstrapping in the following steps: (1) we simulate $p_{rescue}$ 100 times for both the scenario when $f=\tilde{u}\lambda_w/s$ and $f=0$; (2) we sample with replacement which we store in $S_f$ and  $S_0$; (4) for each element of $S_0$ we obtain $N_a^*=1/p_w$ using $p_w=-1/N_s\log\left(1-\bar{S}\right)$ where $\bar{S}$ is the mean of $S_0$ and $N_s$ is an arbitrary value of the initial population size we selected in order to calculate $p_{rescue}$; (5) for each element of $S_f$ we obtain $\tilde{N}_a^*=1/p_a$ using $p_a=-f/N_s\log\left(1-\bar{S}\right)$ where $\bar{S}_f$ is the mean of $S_f$ and $N_s$ is an arbitrary value of the initial population size we selected in order to calculate $p_{rescue}$; (6) we repeat steps (2)-(5) 100 times to obtain $\tilde{N}_a^*/N_a^*$ and we select the upper and lower limits such that $95\%$ of the values of $\tilde{N}_a^*/N_a^*$ lie in the interval given by the bounds.
+For the threshold tumor sizes the $95\%$ confidence interval is obtained through bootstrapping in the following steps: (1) we simulate $p_{rescue}$ 100 times; (2) we sample with replacement which we store in $S$; (3) for each element of this sample we obtain $N_a^*=1/p_s$ using $p_s=-1/N_e\log\left(1-\bar{S}\right)$ where $\bar{S}$ is the mean of $S$ and $N_e$ is an arbitrary value of the initial population size we selected in order to calculate $p_{rescue}$; (4) we repeat steps (2)-(3) 100 times to obtain $N_a^*$ and we select the upper and lower limits such that $95\%$ of the values of $N_a^*$ lie in the interval given by the bounds.
+
+For the ratio of the threshold tumor sizes the $95\%$ confidence interval is obtained through bootstrapping in the following steps: (1) we simulate $p_{rescue}$ 100 times for both the scenario when $f=\tilde{u}\lambda_s/c$ and $f=0$; (2) we sample with replacement which we store in $S_f$ and  $S_0$; (4) for each element of $S_0$ we obtain $N_a^*=1/p_s$ using $p_s=-1/N_e\log\left(1-\bar{S}\right)$ where $\bar{S}$ is the mean of $S_0$ and $N_e$ is an arbitrary value of the initial population size we selected in order to calculate $p_{rescue}$; (5) for each element of $S_f$ we obtain $\tilde{N}_a^*=1/p_a$ using $p_a=-f/N_e\log\left(1-\bar{S}\right)$ where $\bar{S}_f$ is the mean of $S_f$ and $N_e$ is an arbitrary value of the initial population size we selected in order to calculate $p_{rescue}$; (6) we repeat steps (2)-(5) 100 times to obtain $\tilde{N}_a^*/N_a^*$ and we select the upper and lower limits such that $95\%$ of the values of $\tilde{N}_a^*/N_a^*$ lie in the interval given by the bounds.
+%%%
+\section{Aneuploidy inhances mutation rates}\label{sec:appendix-diff-mut}
+Aneuploidy has the potential to increase the mutation rate in cells which are in this state.
+We change our model such that sensitive cells divide and mutate with rate $v_s\lambda_s$ and aneuploid cells divide and mutate with rate $v_a\lambda_a$.
+As a result, we re-calculate the survival probabilities $p_s,p_a$ and $p_s$ as we did in Appendix \ref{sec:appendix-surv-prob}:
+\begin{equation} \label{eq:extinction_prob_ane}
+\begin{aligned}
+1-p_s = &\frac{\mu_s}{\lambda_s+\mu_s+u\lambda_s+v_s\lambda_s} + 
+		  \frac{u\lambda_s}{\lambda_s+\mu_s+u\lambda_s+v_s\lambda_s}\left(1-p_a\right)\left(1-p_s\right) + \\
+		  & \frac{\lambda_s}{\lambda_s+\mu_s+u\lambda_s+v_s\lambda_s}\left(1-p_s\right)^2 +
+		  \frac{v_s\lambda_s}{\lambda_s+\mu_s+u\lambda_s+v_s\lambda_s}\left(1-p_m\right)\left(1-p_s\right) ,\\
+1-p_a = &\frac{\mu_a}{\lambda_a+\mu_a+v_a\lambda_a}+\frac{v_a\lambda_a}{\lambda_a+\mu_a+v_a\lambda_a}\left(1-p_m\right)\left(1-p_a\right)+\frac{\lambda_a}{\lambda_a+\mu_a+v_a\lambda_a}\left(1-p_a\right)^2 ,\\
+1-p_m = &\frac{\mu_m}{\lambda_m+\mu_m}+\frac{\lambda_m}{\lambda_m+\mu_m}\left(1-p_m\right)^2 .	 
+\end{aligned}
+\end{equation}
+Solving the above equations we obtain:
+\begin{equation}\label{eq:survival_prob_ane}
+\begin{aligned}
+p_s &= \frac{\lambda_s-\mu_s-u\lambda_sp_a-v_s\lambda_sp_m+\sqrt{\left(\lambda_s-\mu_s-u\lambda_sp_a-v_s\lambda_sp_m\right)^2+4\lambda_s^2\left(up_a+v_sp_m\right)}}{2\lambda_s} ,\\
+p_a &= \frac{\lambda_a-\mu_a-v_a\lambda_ap_m+\sqrt{\left(\lambda_a-\mu_a-v_a\lambda_ap_m\right)^2+4\lambda_a^2v_ap_m}}{2\lambda_a}, \\
+p_m &= \frac{\lambda_m-\mu_m}{\lambda_m} .
+\end{aligned} 
+\end{equation}
+Consequently, the threshold population size can be written as:
+\begin{equation}\label{eq:N_a2}
+\begin{aligned}
+N_a^* \approx 
+  \frac{\abs{\Delta_s}}{u\lambda_s} \cdot \begin{cases}
+    \frac{\abs{\Delta_a}}{v_a\lambda_a}  \frac{\lambda_m}{\Delta_m} ,&
+  \Delta_a T^* \ll -1 \text{ (tolerant aneuploids)},\\ 
+  %\left(\frac{\lambda_a}{v}  \frac{\lambda_m}{\Delta_m}\right)^{1/2} ,&
+  2\lambda_a T^* ,&
+  -1 \ll \Delta_a T^* \ll 1  \text{ (stationary aneuploids)},\\ 
+  \frac{\lambda_a}{\Delta_a} ,&
+   \Delta_a T^* \gg 1 \text{ (resistant aneuploids)},
+  \end{cases}
+\end{aligned}
+\end{equation}
+where $T^*=\left(4v_a\lambda_a^2p_m\right)^{-\frac{1}{2}}$. The probability of evolutionary rescue is:
+\begin{equation} \label{eq:rescue_prob_ane} 
+\presc = 
+1-\left(1-p_s\right)^N \approx
+1-\e^{-Np_s} = 
+1-e^{-N/N_a^*} ,
+\end{equation}
+which we plot in Figure \ref{rescue_prob_ane} for multiple values of $v_a$. We note that for $v_a=10^{-5}$ we are in the case of stationary aneuploidy (i.e. $\Delta_a T^*\approx -0.55$ ). 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the  fraction of simulations in which the tumor has adapted to the stress and $n=100$ is the number of simulations.
-Parameters: $\lambda_w=0.1,\lambda_a=0.0901,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}, K=10^8$.}
+Parameters: $\lambda_s=0.1,\lambda_a=0.0901,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}, K=10^8$.}
 \label{LogisticPlot}
@@
 Our inhomogeneous Poisson-process approximations (solid black lines, right: \cref{eq:meantime1}, left: \cref{meantimet2}) is in agreement with simulation results (red markers with 95\% quantile intervals obtained with bootstrapping, see see Appendix G).
-Parameters: $\lambda_w=0.1,\lambda_m=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.
+Parameters: $\lambda_s=0.1,\lambda_m=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.
 }
@@
 \includegraphics[width=1\textwidth]{Figures/ExpectedNumber.pdf}
-\caption{\textbf{Aneuploidy increases the number of mutations which rescue the tumor.} Shown is the expected number of mutation, which will rescue the cancer cell population, produced through the evolutionary trajectory $wildtype \rightarrow mutant$ (blue line, \cref{eq:twosteplineagedirect}) or through the trajectory $wildtype \rightarrow aneuploid \rightarrow mutant$ (red line, \cref{eq:twosteplineage}).
+\caption{\textbf{Aneuploidy increases the number of mutations which rescue the tumor.} Shown is the expected number of mutation, which will rescue the cancer cell population, produced through the evolutionary trajectory $sensitive \rightarrow mutant$ (blue line, \cref{eq:twosteplineagedirect}) or through the trajectory $sensitive \rightarrow aneuploid \rightarrow mutant$ (red line, \cref{eq:twosteplineage}).
 Dashed vertical red line represents the threshold tumor size above which evolutionary rescue is very likely through aneuploidy \cref{eq:N_a} and the dashed vertical blue line represents the threshold tumor size above which evolutionary rescue is very likely through direct mutation \cref{eq:N_m}.
-Parameters: $\lambda_w=0.1,\lambda_m=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}
+Parameters: $\lambda_s=0.1,\lambda_m=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}
 \label{ExpectedNumberRescueLineages}
@@
 The red line is analytic result \cref{distribution} overlaid over the histogram of simulations.
-Parameters: $N=10^6, \lambda_w=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}
+Parameters: $N=10^6, \lambda_s=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}
 \label{KMdistribution}
@@
 The dots represent simulation results and the error bars represent $95\%$ confidence intervals ($p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the fraction of simulations in which the tumor has adapted to the stress and $n=100$ is the number of simulations).
-Parameters: $\lambda_w=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}
+Parameters: $\lambda_s=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}
 \label{rescue_prob_sgv}
@@
 Solid lines represents the approximations \cref{eq:AsymptoticTimeRules} ($N<N_a^*$ blue line and $N>N_m^*$ green line).
-Parameters: $\lambda_w=0.1,\lambda_m=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}
+Parameters: $\lambda_s=0.1,\lambda_m=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}
 \label{EvolutionaryRescueTimeComplete}
@@
 The simulations converge to \cref{eq:t2det} (green line) for large values of $N\gg N_m^*$.
-Parameters: $\lambda_w=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}
+Parameters: $\lambda_s=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}
 \label{ProliferationTimeLarge}
@@
 The simulations converge to \cref{eq:t2det2} (green line) for large values of $N\gg N_m^*$.
-Parameters: $\lambda_w=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}, M=10^7$.}
+Parameters: $\lambda_s=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}, M=10^7$.}
 \label{RecurrencePlot}
@@
 \includegraphics[width=1\textwidth]{Figures/SGVEvolutionaryRescueTimeComplete.pdf}
-\caption{Shown is the mean time for appearance of a resistance mutation the leads to evolutionary rescue with aneuploidy ($u>0$).
+\caption{Shown is the mean time for appearance of a resistance mutation the leads to evolutionary rescue with aneuploidy ($u>0$) when a fraction $f$ of cancer cells are aneuploid at the start of therapy.
 Black lines represent our inhomogeneous Poisson-process approximations (solid black line, \cref{meantimet2}; dashed black line \cref{meantimet2SGV}).
@@
 Solid lines represents the approximations \cref{eq:AsymptoticTimeRules} ($N<N_a^*$ blue line and $N>N_m^*$ green line).
-Parameters: $\lambda_w=0.1,\lambda_m=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7},f=0.284\%$.}
+Parameters: $\lambda_s=0.1,\lambda_m=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7},f=0.14\%$.}
 \label{SGVEvolutionaryRescueTimeComplete}
+\end{figure}
+%%%
+\begin{figure}
+\includegraphics[width=1\textwidth]{Figures/ProbvNPlot_ane.pdf}
+\caption{The probability of evolutionary rescue (i.e., the probability that the population does not become extinct), $\presc$, as a function of the initial tumor size, $N$ (\cref{eq:rescue_prob}).
+Dashed vertical line shows the threshold tumor size, $N_a^*$, above which the probability is very high (\cref{eq:N_a2}).
+Red dashed line: $v_a=10^{-7}$.
+blue line: $v_a=10^{-6}$.
+Green line: $v_a=10^{-5}$.
+Dots for simulations and the error bars for $95\%$ confidence interval ($p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the fraction of simulations in which the tumor has been rescued and $n=100$ is the number of simulations).
+Parameters: $\lambda_s=0.1,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}$.}
+\label{rescue_prob_ane}
 \end{figure}

### 2024-07-24 EvolutionaryRescue1.tex
edited and revised text

@@
 
-%\usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent
+\usepackage{parskip} % Activate to begin paragraphs with an empty line rather than an indent
+\setlength{\parindent}{25pt} % This sets the indentation length for each paragraph
 \usepackage{graphicx} % Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
@@
 \begin{abstract}
-Evolutionary rescue is the process by which a population survives a sudden environmental change that initially causes the population to decline towards extinction.
+Evolutionary rescue occurs when a population survives a sudden environmental change that initially causes the population to decline toward extinction.
 A prime example of evolutionary rescue is the ability of cancer to survive exposure to treatment. 
 One evolutionary mechanism by which a population of cancer cells can adapt to chemotherapy is aneuploidy. 
-Aneuploid cancer cells can have higher fitness in an environment altered by anti-cancer drugs, e.g., because aneuploidy disrupts the pathways normally targeted by the drugs. 
-Indeed, aneuploidy is highly prevalent in tumors, and moreover, some anti-cancer drugs fight cancer by increasing chromosomal instability.
-Here, we model how aneuploidy impacts the fate of a population of cancer cells.
-We use multi-type branching processes to approximate the probability that a tumor survives drug treatment as a function of the initial tumor size, the rates at which aneuploidy and other beneficial mutations occur, and the growth rates of the sensitive and resistant cells.
-Additionally, we investigate the effect of the fraction of pre-existent aneuploid cells on the probability of evolutionary rescue.
-We also approximate the mean recurrence time for the tumor to revert back to its initial size following treatment, if such evolutionary rescue has occurred.
-We propose that aneuploidy can play an important role in the relapse of smaller secondary tumors.
-% TODO missing something here... ok
+Aneuploid cancer cells can be fitter in an environment altered by anti-cancer drugs, e.g., because aneuploidy disrupts the pathways usually targeted by the drugs. 
+Indeed, aneuploidy is highly prevalent in tumors, and some anti-cancer drugs fight cancer by increasing chromosomal instability.
+Here, we model the impact of aneuploidy on the fate of a population of cancer cells.
+We use multi-type branching processes to approximate the probability that a tumor survives drug treatment as a function of the initial tumor size, the rates at which aneuploidy and other beneficial mutations occur, and the growth rates of the drug-sensitive and drug-resistant cells.
+Additionally, we investigate the effect of the pre-existent aneuploid cells on the probability of evolutionary rescue.
+Finally, we estimate the tumor's mean recurrence time to revert to its initial size following treatment and evolutionary rescue.
+We propose that aneuploidy can play an essential role in the relapse of smaller secondary tumors.
 \end{abstract}
@@
 
-\paragraph{Aneuploidy in cancer.}  Each year approximately 10 million people die from cancer \citep{kocarnik2022cancer}, so understanding the factors that contribute to failure of interventions is of great importance.
-One hypothesized factor is aneuploidy, where cells are characterized by an imbalanced karyotype and chromosome copy number alterations \citep{schukken2018cin}.
-Aneuploidy is caused by chromosomal instability, a high rate of chromosome mis-segregation during mitosis.
-Importantly, altered  copy number of chromosomes and chromosome-arms allow cancer cells to survive under stressful conditions such as drug therapy \citep{lukow2021chromosomal,rutledge2016selective,ippolito2021gene}.
-Indeed, cancer cells are often aneuploid and aneuploidy is associated with poor patient outcomes \citep{ben2020context,smith2018systematic}.
-
- \citet{ippolito2021gene} induced aneuploidy in cancer cell lines by exposing them to reversine, a small-molecule inhibitor of the mitotic kinase Mpsi1 and then to anti-cancer drugs such as vemurafenib.
-Reversine-treated cells had a higher proliferation rate following drug exposure compared to sensitive cancer cells, due to selection of specific beneficial karyotypes.
-Similarly, \citet{lukow2021chromosomal} induced aneuploidy in cancer cells and observed that such cells have an advantage compared to sensitive cells during drug treatment, despite having lower fitness before the onset of treatment.
-One proposed mechanism through which aneuploidy is able to confer resistance to anti-cancer drugs is by antagonizing cell division, which prevents the drugs from damaging DNA and microtubules \citep{replogle2020aneuploidy}.
-Other mechanisms are the selection for specific karyotypes that lead to reduced drug metabolism \citep{ippolito2021gene}, and elevated levels of DNA damage repair due to higher basal level of DNA damage \citep{zerbib2023human}.
-
-An important aspect of aneuploidy is that the rate with which cells become aneuploid, that is the rate of chromosome mis-segregation, is several orders of magnitude higher than mutation rate \citep{bakker2023predicting}.
-Consequently, a cell exposed to a stress such as chemotherapeutic drugs can acquire aneuploidy faster compared to acquiring a mutation.
-Moreover several proposed anti-cancer drugs elevate the rate of mis-segregation in order to fight cancer cells~\citep{lee2016effects}, as an extremely high chromosome mis-segregation rate is not compatible with cell survival and proliferation.
-
-\paragraph{Evolutionary rescue.} Populations adapted to a certain environment are vulnerable to environmental changes, which might cause extinction of the population.
+\paragraph{Aneuploidy in cancer.}  
+Each year, approximately 10 million people die from cancer \citep{kocarnik2022cancer}. Understanding the factors that contribute to the failure of interventions is of great importance.
+One suggested factor is aneuploidy, in which cells are characterized by an imbalanced karyotype and alterations in the number of chromosomes \citep{schukken2018cin}. 
+Aneuploidy is caused by chromosomal instability and mis-segregation of chromosomes during mitosis.
+Importantly, changes in the number of chromosomes and chromosome arms copies allow cancer cells to survive under stressful conditions such as drug therapy \citep{lukow2021chromosomal,rutledge2016selective,ippolito2021gene}.
+Indeed, cancer cells are often aneuploid, and aneuploidy is associated with poor patient outcomes \citep{ben2020context,smith2018systematic}.
+
+ \citet{ippolito2021gene} induced aneuploidy in cancer cell lines by exposing them to reversine, a small-molecule inhibitor of the mitotic kinase Mpsi1, and then to anti-cancer drugs such as vemurafenib.
+Reversine-treated cells had a higher proliferation rate following drug exposure than sensitive cancer cells, due to selection of specific beneficial karyotypes.
+Similarly, \citet{lukow2021chromosomal} induced aneuploidy in cancer cells and observed that such cells have an advantage compared to sensitive cells during drug treatment despite having lower fitness before the onset of treatment.
+One proposed mechanism through which aneuploidy can confer resistance to anti-cancer drugs is by antagonizing cell division, which prevents the drugs from damaging DNA and microtubules \citep{replogle2020aneuploidy}.
+Other mechanisms are the selection for specific karyotypes that lead to reduced drug metabolism \citep{ippolito2021gene} and elevated levels of DNA damage repair due to higher basal levels of DNA damage \citep{zerbib2023human}.
+
+An essential aspect of aneuploidy is that the rate with which cells become aneuploid, that is, the rate of chromosome missegregation, is several orders of magnitude higher than mutation rate \citep{bakker2023predicting}.
+Consequently, a cell exposed to stress, such as chemotherapeutic drugs, can acquire aneuploidy faster than a mutation.
+Moreover, several proposed anti-cancer drugs elevate the missegregation rate to fight cancer cells~\citep{lee2016effects}, as an extremely high chromosome missegregation rate is incompatible with cell survival and proliferation.
+
+\paragraph{Evolutionary rescue.} Populations adapted to a specific environment are vulnerable to environmental changes, which might cause the population's extinction.
 Examples of such environmental changes include climate change, invasive species, and the onset of drug therapies.
@@
 
-\citet{gunnarsson2020understanding} analyzed a model where a tumor consisting of two populations of cancer cells, one drug resistant and the other drug sensitive, is able to evade extinction by cells switching between the two phenotypes through epigenetic mutations.
-They found that even when the drug resistant type is barely viable the epimutations have the effect of guaranteeing evolutionary rescue.
-Evolutionary rescue in a single step, where an initially declining population, has to acquire a mutation in order to survive extinction after a sudden environmental change, has been studied by \citet{orr2008population,orr2014population}.
-They analyzed a model where the mutant strain is present in small numbers at the onset of therapy and concluded that this can significantly enhance the probability that the population would survive.
+\citet{gunnarsson2020understanding} analyzed a model where a tumor consisting of two populations of cancer cells, one drug-resistant and the other drug-sensitive, can evade extinction by cells switching between the two phenotypes through epigenetic mutations.
+They found that even when drug-resistant cells are barely viable, the epimutations guarantee evolutionary rescue.
+Evolutionary rescue in a single step, where an initially declining population has to acquire a mutation to survive extinction after a sudden environmental change, has been studied by \citet{orr2008population,orr2014population}.
+They analyzed a model where the mutant strain is present in small numbers at the onset of therapy.
+They concluded that this can significantly enhance the population's probability of survival.
 
@@
 How multiple mutations contribute to the survival of the population is less explored, but \citet{wilson2017soft} have shown that evolutionary rescue is significantly enhanced by soft selective sweeps when multiple mutations contribute towards evolutionary rescue.
-Evolutionary rescue that requires two successive mutations (i.e., two steps) has been investigated by \citet{martin2013probability} who tested the predictions of their model with data from experiments in yeast and bacteria.
-
-Here we study evolutionary rescue after a sudden environment change caused by the initiation of anti-cancer drug treatment. 
+Evolutionary rescue that requires two successive mutations (i.e., two steps) has been investigated by \citet{martin2013probability}, who tested their model's predictions with data from yeast and bacteria experiments.
+
+Here, we study evolutionary rescue after a sudden environmental change caused by initiating anti-cancer drug treatment. 
 We consider a range of effects of aneuploidy, from tolerance to (partial) resistance to the drug \citep{brauner2016distinguishing}.
-% TODO explain distinction in a sentence
-We estimate the effect of aneuploidy on the probability of evolutionary rescue of the tumor.
-When aneuploidy provides resistance to the drug, it can directly rescue the tumor.
-However, when anueploidy provides tolerance to the drug, it may act as an evolutionary ``stepping stone'' or ``springboard'' \citep{Yona2015,martin2013probability,osmond_genetic_2020}, delaying extinction, and thereby allowing the tumor more time to acquire a resistance mutation on top of the aneuploid background.
-As mentioned above, evidence suggests that aneuploidy may be a common strategy for tumor adaptation to durg therapy, but it is unknown how often aneuploidy provides tolerance and thereby acts as a stepping stone.
+We estimate the effect of aneuploidy on the tumor's evolutionary rescue probability.
+When aneuploidy provides drug resistance, it can directly rescue the tumor.
+However, when aneuploidy provides tolerance to the drug, it may act as an evolutionary ``stepping stone'' or ``springboard'' \citep{Yona2015,martin2013probability,osmond_genetic_2020}, delaying extinction, and thereby allowing the tumor more time to acquire a resistance mutation on top of the aneuploid background.
+As mentioned above, evidence suggests that aneuploidy may be a common strategy for tumor adaptation to drug therapy.
+Still, it is unknown how often aneuploidy provides tolerance and acts as a stepping stone.
 We also estimate the mean time until a tumor cell population reaches its pre-treatment size following drug therapy. 
-Given that aneuploidy is present in many tumors already before the onset of therapy \citep{lukow2021chromosomal,ben2020context}, we also consider the effect of pre-treatment standing genetic variation on the evolutionary dynamics.
-Additionally, we are interested in the timescale of evolutionary rescue and the effect that aneuploidy has on the time necessary for the tumor to overcome drug therapy.
-We find that aneuploidy increases the \emph{window of opportunity} when evolutionary rescue is able to prevent extinction.
+Given that aneuploidy is present in many tumors before the onset of therapy \citep{lukow2021chromosomal,ben2020context}, we also consider the effect of pre-treatment standing genetic variation on the evolutionary dynamics.
+Additionally, we are interested in the timescale of evolutionary rescue and the impact that aneuploidy has on the time necessary for the tumor to overcome drug therapy.
 
@@
 \paragraph{Evolutionary model.}
-We follow the number of cancer cells that have one of three different genotypes at time $t$: sensitive, $s_t$; tolerant/resistant aneuploid, $a_t$; and resistant mutant, $m_t$. 
+We follow the number of cancer cells with one of three different genotypes at time $t$: sensitive, $s_t$; tolerant/resistant aneuploid, $a_t$; and resistant mutant, $m_t$. 
 These cells divide and die with rates $\lambda_k$ and $\mu_k$ (for $k=s, a, m$).
-The difference between the division and death rate is $\Delta_k = \lambda_k-\mu_k$.
+The division and death rate difference is $\Delta_k = \lambda_k-\mu_k$.
 We assume the population of cells is under a strong stress, such as drug therapy, and therefore $\Delta_s<0$, whereas the mutant is resistant to the stress, $\Delta_m>0$.
-We consider a range of possible values for $\Delta_a$, finding three distinct cases: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_s$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing, stationary, or growing or dying only very slowly, that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$, in a sense that we will make precise below.
-
-We assume that both chromosomal missegregation and mutation occur during the process of mitosis. 
+We consider a range of possible values for $\Delta_a$, finding three distinct scenarios: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_s$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing, stationary, or growing or dying only very slowly, that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$, in a sense that we will make precise below.
+
+We assume that both chromosomal missegregation and mutations occur during the process of mitosis. 
 Sensitive cells may divide and then missegregate to become aneuploids at rate $u\lambda_s$.
@@
 The simulation monitors the number of cells of each type: sensitive, aneuploid, and mutant. 
-The sensitive population initially consists of $w_0=N$ cells, whereas the other cell types are initially absent.
+Initially, the population starts with only sensitive cells, $s_0=N$, and the other genotypes are initially absent.
 
@@
 \end{subequations}
-For the remaining of this paper we assume that the division rates for sensitive and aneuploid cells can be written as $\lambda_ss_t\left(1-u-v\right)\approx \lambda_ss_t$ and $\lambda_aa_t\left(1-v\right)\approx\lambda_aa_t$ because $u,v\ll1$ (Table \ref{table1}).
-Each iteration of the simulation loop starts with computing the rates $\nu_k$ of each event $k$.
+For the remaining of this paper, we assume that the division rates of sensitive and aneuploid cells can be written as $\lambda_ss_t\left(1-u-v\right)\approx \lambda_ss_t$ and $\lambda_aa_t\left(1-v\right)\approx\lambda_aa_t$ because $u,v\ll1$ (\Cref{table1}).
+Each iteration of the simulation loop starts by computing the rates $\nu_k$ of each event $k$.
 We then draw the time until the next event, $\Delta t$, from an exponential distribution whose rate parameter is the sum of the rates of all events, such that $\Delta t \sim \text{Exp}(\sum_j \nu_j)$.
 Then, we randomly determine which event occurred, where the probability for event $k$ is $p_k = \nu_k/\sum_j \nu_j$.
-Finally, we update the number of cells of each type according to the event that occurred and update the time from $t$ to $t+\Delta t$.
-We repeat these iterations until either the population becomes extinct (the number of cells of all types is zero) or the number of mutant cells is high enough so that its extinction probability is $<0.1\%$, that is until
+Finally, we update the number of cells of each genotype according to the event that occurred and update the time from $t$ to $t+\Delta t$.
+We repeat these iterations until either the population becomes extinct (the number of cells of all genotypes is zero) or the number of mutant cells is high enough so that their extinction probability is $<0.1\%$, that is, until
 \begin{equation*}
@@
 \end{equation*}
-which we obtain by solving $1-(1-p_m)^{m_t}=0.999$ for $m_t$ with $p_m=\Delta_m/\lambda_m$ as the probability that a single mutant escapes stochastic extinction  (see Appendix A).
+which we obtain by solving $1-(1-p_m)^{m_t}=0.999$ for $m_t$ with $p_m=\Delta_m/\lambda_m$ as the probability that a single mutant escapes stochastic extinction (Appendix A).
 
 When simulations are slow (e.g., due to large population size) with runtimes in the order of days, we use $\tau$-leaping \citep{gillespie2001approximate}, where we assume that the change in the number of cells of genotype $k$ in a fixed time interval $\Delta t$ is Poisson distributed with mean $\nu_k \Delta t$.
-If the change in the number of cells is negative and larger than the subpopulation size then the subpopulation size is updated to be zero.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% YR:  I commented this out because it is entirely standard
-%\paragraph{Bootstrapping}
-%We  use bootstrap resampling method in order to compute  confidence intervals for the mean threshold tumor sizes and mean times by resampling 100 times with replacement from the data set obtained from simulations.
-We then compute the interval which contain $95\%$ of the means of each resampled set.
+If the number of cells of genotype $k$ becomes negative, we change it to zero.
 
@@
 To parametrize the simulations, we assume that the cells under consideration are melanoma cells and rely on \citet{rew2000cell} and \citet{bozic2013evolutionary} for the division and death rates, respectively.
-\citet{rew2000cell} report \emph{in vivo} measurements of the potential doubling times (the waiting time for the number of cells in the tumor to double disregarding cell death) for a large set of cancer types.
+\citet{rew2000cell} report \emph{in vivo} measurements of the potential doubling times (the waiting time for the number of cells in the tumor to double, disregarding cell death) for a large set of cancer types.
 The division rate is obtained as $\lambda=\log{2} / T \approx 0.1$ per day.
@@
 We use  $\mu_s=0.14$ per day.
-Additionally, they observe the growth rate of cancer cells prior to treatment to be 0.01, which we use as the growth rate of mutant cells, which are resistant to the drug.
+Additionally, they observed the growth rate of cancer cells before treatment to be 0.01, which we use as the growth rate of mutant cells, which are resistant to the drug.
 Thus, we use $\mu_m=0.1-0.01=0.09$ per day as the death rate for mutant cells.
@@
 Aneuploid death rate $\mu_a$ is set to the same value as the mutant death rates, $\mu_m=0.09$ per day, given that aneuploidy increases resistance to the drug, such as cisplatin, by antagonizing cell division \citep{replogle2020aneuploidy}.
-Aneuploid division rate is selected such that the aneuploid growth rate $\Delta_s\ll\Delta_a\ll\Delta_m$ which means that $0.06 \le \lambda_a \le 0.1$. 
-For most of our simulations we use $\lambda_a=0.0899$ per day, such that aneuploidy can only act as an evolutionary ``stepping stone'' for the generation of the resistant mutant that rescues the cancer cell population (note that this mutant will occur on the background of an aneuploid genotype).
-
-The missegregation rate in cancer cells is estimated to be between $2.5\times10^{-4}-10^{-2}$ per chromosome per cell divison \citep{shi2005chromosome,thompson2008examining}.
+The aneuploid division rate is selected such that the aneuploid growth rate $\Delta_s\ll\Delta_a\ll\Delta_m$ means that $0.06 \le \lambda_a \le 0.1$. 
+For most of our simulations, we use $\lambda_a=0.0899$ per day, so that aneuploid cells are tolerant and aneuploidy can only act as an evolutionary ``stepping stone'' for the generation of the resistant mutant that rescues the tumor (note that this mutant will occur on the background of an aneuploid genotype).
+
+The missegregation rate in cancer cells is estimated to be between $2.5\times10^{-4}-10^{-2}$ per chromosome per cell division \citep{shi2005chromosome,thompson2008examining}.
 \citet{ippolito2021gene} observed that trisomy in Chr 2 and Chr 6 are most likely to confer increased resistance against the anti-cancer drug vemurafenib for A375 cells.
-We assume each of these trisomies are formed at the most likely rate, and as a result, we use $\tilde{u}=10^{-3}$ per cell division as the chromosome missegregation rate in the drug-free environment.
-\citet{lee2016effects} estimated the effect of different anti-cancer drugs on the missegregation rate and found a $3- 50$ fold increase.
-We assume that our anti-cancer drug causes a 10-fold increase in the chomosome misegregation rate which gives us $u=10^{-2}$ per cell division, as some drugs increase the rate of aneuploidy~\citep{wang2019molecular,mason2017functional}.
-We assume the mutation rate is $10^{-7}$ per gene per cell division \citep{loeb2001mutator} and since we assume that a single target gene confers resistance to the drug, we use $v=10^{-7}$ per cell division.
+We assume each of these trisomies is formed at the most likely rate, and as a result, we use $\tilde{u}=10^{-3}$ per cell division as the chromosome missegregation rate in the drug-free environment.
+Some drugs are known to increase chromosome instability~\citep{wang2019molecular,mason2017functional}. 
+Specifically, \citet{lee2016effects} estimated the effect of different anti-cancer drugs on the missegregation rate and found a 3-50-fold increase.
+We thus assume an anti-cancer drug that causes a 10-fold increase in the chromosome missegregation rate, which gives us $u=10^{-2}$ per cell division.
+We assume the mutation rate is $10^{-7}$ per gene per cell division \citep{loeb2001mutator}, and since we assume that a single target gene confers resistance to the drug, we use $v=10^{-7}$ per cell division.
 
 The fitness cost $c$ of aneuploidy before the onset of therapy is difficult to estimate as we are interested in a specific type of aneuploidy that improves the fitness of cancer cells in an environment altered by drugs.
-We estimate $c$ by using the formula $c=\tilde{u}\lambda_s / f$, where $f$ is the fraction of aneuploid cancer cells.
-To estimate $f$, we note that \citet{lukow2021chromosomal} mixed sensitive and aneuploid A375 melanoma cells at $50:50$ ratio, cultured them in drug-free environment and observed the ratio evolve as a function of time with the aneuploid cells declining to $15\%$ after 24 days.
-Thus, our estimate for the fitness cost $s$ is $s=\abs{\log\left[0.15/(1-0.15)\right]/24}\approx0.07$ per day \citep{chevin2011measuring}.
-Therefore, the estimated fraction of cancer cells with the beneficial aneuploidy is $f=10^{-3}\times10^{-1}/0.07=0.14\%$, that is, $0.14\%$ of pre-treatment cancer cells have the beneficial aneuploidy of interest.
-
-We note that when we refer to sensitive cancer cells we include those cells that have any aneuploidy other than trisomy in Chr 2 and 6 as those are the aneuploid cells that are hypothesized to have higher fitness in the environment altered by drugs such as vemurafenib.
+We estimate $c=\tilde{u}\lambda_s / f$, where $f$ is the fraction of aneuploid cancer cells.
+To estimate $f$, we note that \citet{lukow2021chromosomal} mixed sensitive and aneuploid A375 melanoma cells at $1:1$ ratio, cultured them in a drug-free environment, and observed the ratio evolve as a function of time with the aneuploid cells declining to $15\%$ after 24 days.
+Thus, our estimate for the fitness cost is $c=\abs{\log\left[0.15/(1-0.15)\right]/24}\approx0.07$ per day \citep{chevin2011measuring}, and the estimated fraction of cancer cells with the beneficial aneuploidy is $f=10^{-3}\times10^{-1}/0.07=0.14\%$, that is, $0.14\%$ of pre-treatment cancer cells have the beneficial aneuploidy of interest.
+
+We note that when we refer to drug-sensitive cells, we include those cells that have any aneuploidy other than trisomy in Chr 2 and 6, as those are the aneuploid cells that are hypothesized to have higher fitness in the environment altered by drugs such as vemurafenib.
 
@@
 
-In our analysis, we assume that lineages produced by cells from the initial population divide and die independently of each other.
-However, the cells will compete for resources, but we expect that this competition can be neglected because the drug will cause the cell density to rapidly drop below the carrying capacity at which competition is most important.
-To test this, we simulate a logistic growth model, with division and death rates given by:
+In our analysis, we assume that cells from the initial population divide and die independently of each other.
+However, these cells will compete for resources.
+We assume this competition can be ignored because the drug will cause the cell density to rapidly drop below the carrying capacity where competition is important.
+To test this assumption, we simulate a logistic growth model, with division and death rates given by
 \begin{align*}
@@
 
-In our model, \emph{evolutionary rescue} occurs when resistant cells appear and establish (avoid random extinction) in the population  ($m_t \gg 1$) before the population becomes extinct ($w_t=a_t=m_t=0$).
+In our model, \emph{evolutionary rescue} occurs when drug-resistant cells appear and establish (avoid random extinction) in the population  ($m_t \gg 1$) before the population becomes extinct ($w_t=a_t=m_t=0$).
 Aneuploidy may contribute to evolutionary rescue by either preventing (when $\Delta_a>0$) or delaying (when $0>\Delta_a>\Delta_s$) the extinction of the population before mutant cells appear and establish.
 We assume independence between clonal lineages starting from an initial population of $N$ sensitive cells (we check the effect of density-dependent growth on our results below).
-We therefore define $p_s$ as the probability that a lineage starting from a single sensitive cell avoids extinction by acquiring drug resistance.
+Define $p_s$ as the probability that a lineage starting from a single drug-sensitive cell avoids extinction by acquiring drug resistance.
 Thus, $N^*=1/p_s$ is the threshold tumor size above which evolutionary rescue is very likely, and the rescue probability is given by 
@@
 where the approximation $(1-p_s)\approx e^{-p_s}$ assumes that $p_s$ (but not necessarily $N p_s$) is small.
-Indeed, when $N<1/p_s$, then the probability for evolutionary rescue is $\presc \approx N p_s$  and when $N > 1/p_s$, it is $\presc \approx 1$, justifying the definition of $N^*$ as the threshold tumor size for evolutionary rescue.
-
-We use the theory of multi-type branching processes to find approximate expressions \cref{eq:pw_parttolerant,eq:pw_partrest,eq:pw_tolerant} for $p_s$ in different regimes (see appendix~\ref{sec:appendix-surv-prob}).
+Indeed, when $N<1/p_s$, then the probability for evolutionary rescue is $\presc \approx N p_s$, and when $N > 1/p_s$, it is $\presc \approx 1$, justifying the definition of $N^*$ as the threshold tumor size for evolutionary rescue.
+
+We use multi-type branching-process theory to find approximate expressions \cref{eq:pw_parttolerant,eq:pw_partrest,eq:pw_tolerant} for $p_s$ in three distinct scenarios (Appendix~\ref{sec:appendix-surv-prob}).
 Substituting these into $N^*=1/p_s$, we find approximations for the threshold tumor size, $N^*$. 
-For these approximations, an important quantity is $T^* = (4 v \lambda_a^2 \Delta_m/\lambda_m)^{-1/2}$, which is the critical time an aneuploid lineage needs to survive to produce a resistant mutant that avoids random extinction.
-First, if aneuploidy is very rare ($u\lambda_a T^*< 1$), or if aneuploidy is rare ($u\lambda_a < -\Delta_a$) and very sensitive to the drug ($\Delta_a T^* < -1$), then evolutionary rescue will likely occur by a direct resistance mutation in a sensitive cell without the involvement of aneuploidy, such that
+In these approximations, an important quantity is $T^* = (4 v \lambda_a^2 \Delta_m/\lambda_m)^{-1/2}$, which is the critical time an aneuploid lineage needs to survive to produce a resistant mutant that avoids random extinction.
+First, if aneuploidy is very rare ($u\lambda_a T^*< 1$), or if aneuploidy is rare ($u\lambda_a < -\Delta_a$) and very sensitive to the drug ($\Delta_a T^* < -1$), then it is likely that evolutionary rescue will occur through a direct resistance mutation in a sensitive cell without aneuploidy playing a role in the adaptive dynamics, such that
 \begin{equation} \label{eq:N_m}
@@
 \end{equation}
-Here, $\abs{\Delta_s}/\left(v\lambda_s\right)$ is the ratio of the rate at which sensitive cells are decreasing in number and the rate at which they are mutating.
+Here, $\abs{\Delta_s}/\left(v\lambda_s\right)$ is the ratio of the rate at which sensitive cells decrease in number and the rate at which they are mutating.
 Notably, the aneuploidy parameters ($u$, $\lambda_a$, $\mu_a$) do not affect $N_m^*$.
@@
 Otherwise, aneuploidy is frequent enough ($u\lambda_a > \max{\left(-\Delta_a, 1/T^*\right)}$) to affect the evolution of drug resistance. 
-The threshold tumor size can then be approximated by one of the following scenarios, depending on $\Delta_a T^*$, the change in the aneuploid log-population size during the critical time,
+The threshold tumor size can be approximated by one of the following scenarios, depending on $\Delta_a T^*$, which represents the change in the aneuploid log-population size during the critical time,
 \begin{equation}  \label{eq:N_a}
@@
 \end{equation}
-These approximations perform very well when compared to results of stochastic evolutionary simulations (\Cref{rescue_prob,rescue_threshold}).
-The first line describes the scenario in which aneuploid cells are still effectively killed by the treatment, but not as quickly as the sensitive cells. 
-In the second scenario, aneuploid cells are sufficiently resistant that the expected size of each aneuploid lineage is roughly 1.
-In both of these scenarios, aneuploidy increases the probability of rescue by slowing or halting the decrease of the tumor population size, allowing more opportunities for producing resistant mutants.
-In the third scenario, aneuploid cells are sufficiently resistant for the population to re-grow the tumor even without additional resistance mutations.
+These approximations perform very well when compared to the results of stochastic evolutionary simulations (\Cref{rescue_prob,rescue_threshold}).
+The first line describes the scenario in which the treatment still effectively kills aneuploid cells but not as quickly as the sensitive cells. 
+In the second scenario, aneuploid cells are sufficiently resistant, and the expected size of each aneuploid lineage is roughly 1.
+In both of these scenarios, aneuploidy increases the probability of rescue by slowing or halting the decrease in the tumor population size, allowing more opportunities to produce resistant mutants. 
+In the third scenario, aneuploid cells are sufficiently resistant for the population to re-grow the tumor without additional resistance mutations.
 Notably, in this scenario the mutant parameters ($v$, $\lambda_m$, and $\Delta_m$) do not affect $N_a^*$ beyond their effect on $T^*$.
@@
 As expected, this ratio increases with the mutation rate $v$ and decreases with the aneuploidy rate $u$.
-In the first scenario, $\abs{\Delta_a}/\left(u\lambda_a\right)$ is  the ratio of the expected time for an aneuploid lineage to appear, $1/\left(u\lambda_a\right)$, and the expected time until that lineage disappears, $1/\abs{\Delta_a}$.
-In the third scenario, $\left(v \frac{\Delta_m}{\lambda_m}\right) / \left(u \frac{\Delta_a}{\lambda_a}\right)$ is the ratio of the rates of formation of resistant mutants that avoid extinction and partially resistant aneuploids that avoid extinction.
+In the first scenario, $\abs{\Delta_a}/u\lambda_a$ is the ratio of the expected time for an aneuploid lineage to appear, $1/u\lambda_a$, and the expected time until that lineage disappears, $1/\abs{\Delta_a}$.
+In the third scenario, $\left(v \frac{\Delta_m}{\lambda_m}\right) / \left(u \frac{\Delta_a}{\lambda_a}\right)$ is the ratio of the rates of appearance of resistant mutants that avoid extinction and partially resistant aneuploids that avoid extinction.
 In the second scenario, $\frac{1}{u}\left(v  \frac{\Delta_m}{\lambda_m}\right)^{1/2}=\sqrt{\frac{\Delta_a}{u\lambda_a}  v \frac{\Delta_m}{\lambda_m}  \left(u\frac{\Delta_a}{\lambda_a}\right)^{-1}}$, which is the geometric mean of the first and third scenarios.
@@
 Interestingly, increasing both the aneuploid division rate, $\lambda_a$, and the aneuploid death rate, $\mu_a$, such that the growth rate $\Delta_a$ remains constant, leads to a decrease in $T^*$, pushing the system to the second scenario.
-In this scenario, increasing the division rate $\lambda_a$ should also increase the mutation rate $v\lambda_a$ in aneuploid cells, as mutations mostly occur during division, so overall the threshold tumor size $N_a^*$ is unaffected by the division rate $\lambda_a$ (i.e., $d \lambda_a T^*/d\lambda_a = 0$).
-Thus, if aneuploid cells rapidly die due to the drug but compensate by rapidly dividing, further increasing  the division rate will \emph{not} facilitate adaptation.
+In this scenario, increasing the division rate $\lambda_a$ should also increase the mutation rate $v\lambda_a$ in aneuploid cells, as mutations mostly occur during division, so overall, the threshold tumor size $N_a^*$ is unaffected by the division rate $\lambda_a$ (i.e., $d \lambda_a T^*/d\lambda_a = 0$).
+Thus, if aneuploid cells rapidly die due to the drug but compensate by rapidly dividing, increasing the division rate will \emph{not} facilitate adaptation.
 This is consistent with experimental findings where aneuploidy confers resistance by decreasing the division rate~\citep{replogle2020aneuploidy}.
 
-We can categorize tumors by their size: small tumors with size $N<N_a^*$ that are unlikely to survive treatment, intermediate tumors with size $N_a^* < N < N_m^*$ that rely on aneuploidy for evolutionary rescue, and large tumors with size $N > N_m^*$ that could overcome the effect of drug treatment even without aneuploidy.
+We can categorize tumors by their size: small tumors with size $N<N_a^*$ that are unlikely to survive treatment, intermediate tumors with size $N_a^* < N < N_m^*$ that rely on aneuploidy for evolutionary rescue, and large tumors with size $N > N_m^*$ that could overcome the effect of drug treatment without aneuploidy.
 For the parameter values in \Cref{table1} with $\lambda_a=0.0899,\mu_s=0.14, u=10^{-2}, v=10^{-7}$, we are in the tolerant aneuploid scenario, and substituting in \cref{eq:N_a,eq:N_m}, we have $N_a^* \approx 4 \times 10^6$ and $N_m^* \approx 4 \times 10^7$.
 Hence, we obtain the ratio $N^*_a/N^*_m \approx 0.11$ (\cref{eq:N_ratio}), that is, aneuploidy reduces the threshold tumor size by approximately 89\%.
-Interestingly, the threshold between small and intermediate tumors, $N_a^*$, is similar to the tumor detection threshold of $4.19 \times 10^6$ cells for a wide variety of tumors as reported by \citet{avanzini2019cancer}.
-
-Aneuploidy can increase the point mutation rate in cancer cells by increasing the number of genes that can mutate or by causing DNA damage, which can in turn elevates the mutation rate \citep{garribba2023short,passerini2016presence,janssen2011chromosome}.
-Thus, we re-write our model to take this into account in Appendix \ref{sec:appendix-diff-mut}.
-We observe from \cref{eq:N_a2} and \Cref{rescue_prob_ane} that increasing the mutation rate in aneuploid cells by one order of magnitude leads to a decrease in the threshold tumor size of approximately one order of magnitude.
-We note that for elevated mutation rates the aneuploids transition from tolerant to stationary without change in their growth rate $\Delta_a$.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\paragraph*{Density-dependent growth.}
-
-In our analysis we used branching processes, which assume that growth (division and death) is density-independent.
+Interestingly, the threshold between small and intermediate tumors, $N_a^*$, is similar to the tumor detection threshold of $4.19 \times 10^6$ cells for a wide variety of tumors~\citep{avanzini2019cancer}.
+
+Aneuploidy may lead to an increased mutation rate in cancer cells\citep{garribba2023short,passerini2016presence,janssen2011chromosome}.
+Thus, we extended our model to account for this in Appendix \ref{sec:appendix-diff-mut}.
+We find that increasing the mutation rate in aneuploid cells by one order of magnitude leads to a decrease in the threshold tumor size of approximately one order of magnitude.
+Also, it transitions the system from the first scenario (tolerant aneuploids) to the second scenario (stationary aneuploids) without changing the aneuploid growth rate, $\Delta_a$.
+
+In our analysis, we used branching processes, which assume that growth (division and death) is density-independent.
 However, growth may be limited by resources (oxygen, nutrients, etc.) and therefore depend on cell density.
-We therefore performed stochastic simulations of a logistic growth model with a carrying capacity (see Methods). 
-We find that our density-independent approximations agree with results of simulations with density-dependent growth for biologically relevant parameter values (\Cref{LogisticPlot}).
+Therefore, we performed stochastic simulations of a logistic growth model with a carrying capacity. 
+We find that our density-independent approximations agree with the results of simulations with density-dependent growth for biologically relevant parameter values (\Cref{LogisticPlot}).
 
@@
 
-In the above we assumed that at the onset of drug treatment, the initial tumor consisted entirely of sensitive cells that are drug sensitive.
-However, aneuploidy is likely produced even before the onset of treatment at some rate $\tilde{u}$ and confers a deleterious fitness effect $c$ in the absence of the drug \citep{replogle2020aneuploidy,giam2015aneuploidy}.
-Furthermore, the aneuploidy rate in the presence of drugs is likely higher than in their absence, $\tilde{u} < u$ \citep{wang2019molecular,mason2017functional}.
-But if the number of cells in the tumor $N$ is large (as expected if the tumor is treated with a drug), there may already be a fraction $f \approx \tilde{u}\lambda_s/c$ of aneuploid cells in the population (here we assume that the drug affects the sensitive death rate but not the division rate and therefore we use $\lambda_s$ for the sensitive division rate in the drug-free environment).
-
-Therefore, the threshold tumor size for rescue by standing generation variation, $\tilde{N}^*_{a}$, is similar to the threshold for rescue by de-novo variation, $N^*_a$, except that the sensitive growth rate $\abs{\Delta_s}$ is replaced by the aneuploidy cost $c$, such that
+In the above, we assumed that at the onset of drug treatment, the initial tumor consisted entirely of drug-sensitive cells.
+However, aneuploidy is likely produced even before the onset of treatment at some rate $\tilde{u}$, which may be lower in the absence of drugs, $\tilde{u} < u$ \citep{wang2019molecular,mason2017functional}.
+Moreover, aneuploidy likely confers a fitness cost $c$ in the absence of drugs \citep{replogle2020aneuploidy,giam2015aneuploidy}.
+Hence, if the number of cells in the tumor $N$ is large (as expected if the tumor is treated with a drug), there may already be a fraction $f \approx \tilde{u}\lambda_s/c$ of aneuploid cells in the population (here we assume that the drug affects the sensitive death rate but not the division rate and therefore we use $\lambda_s$ for the sensitive division rate in the drug-free environment).
+
+Therefore, the threshold tumor size for rescue by standing generation variation, $\tilde{N}^*_{a}$, is similar to the threshold for rescue by de-novo variation, $N^*_a$, except that the sensitive growth rate $\abs{\Delta_s}$ is replaced by the cost of aneuploidy $c$, such that
 \begin{equation}\label{eq:Nsgv}
@@
 Comparing this approximation of $\tilde{N}^*_{a}/N^*_{a}$ to results of stochastic simulations, we find that the approximations perform very well (\Cref{rescue_denovo}). 
-Standing genetic variation will drive evolutionary rescue if sensitive cells die rapidly (growth rate $\Delta_s$ is very negative) due to a strong effect of the drug on sensitive cells, or if the aneuploidy cost in the drug-free environment, $c$, is small.
-In contrast, \emph{de-novo} aneuploid cells will have a greater contribution to rescue if the aneuploidy cost $c$ is large, the effect of the drug on sensitive cells is weak ($\Delta_s$ is close to zero), or if the drug induces the appearance of aneuploid cells ($u > \tilde u$).
-For example, with  $\lambda_s=0.1,\mu_s=0.14, u=10^{-2}, \tilde{u}=10^{-3}, c=0.07$, the ratio of the threshold tumor sizes for standing vs.
+Standing genetic variation will drive evolutionary rescue if sensitive cells die rapidly (growth rate $\Delta_s$ is very negative) due to a strong effect of the drug on sensitive cells or if the cost of aneuploidy in the drug-free environment, $c$, is small.
+In contrast, \emph{de-novo} aneuploid cells will have a greater contribution to rescue if the cost of aneuploidy, $c$, is large, the effect of the drug on sensitive cells is weak ($\Delta_s$ is close to zero), or if the drug induces the appearance of aneuploid cells ($u > \tilde u$).
+For example, with  $\lambda_s=0.1$, $\mu_s=0.14$, $u=10^{-2}$, $\tilde{u}=10^{-3}$, and $c=0.07$, the ratio of the threshold tumor sizes for standing vs.
 \emph{de-novo} variation is $\tilde{N}^*_a/N^*_a \approx 17.5$, which means that \emph{de-novo} genetic variation is  the main driver of evolutionary rescue.
@@
 \end{equation}
-Evolutionary rescue through direct mutation is more likely if the cost of aneuploidy $c$ is very large or the effect of the drug $\Delta_s$ is small.
-In contrast, standing genetic variation will drive adaptation if the pre-treatment chromosome missegreagation rate $\tilde{u}$ is very large.
-The ratio does not depend on the rate of chromosome missegregation induced by the drug $u$.
-However, if the aneuploid growth rate $\Delta_a$ increases, then evolutionary rescue is driven by standing genetic variation.
-For the parameter values of  $\lambda_s=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, \tilde{u}=10^{-3}, v=10^{-7}$, we are in the tolerant aneuploid scenario and obtain the ratio $\tilde{N}^*_a/N^*_m \approx 1.94$, which means that standing genetic variation does not drive evolution of drug resistance when compared to direct mutation.
-We note that for larger values of the pre-treatment chromosome missegregation rate, which are consistent with empirical studies (see \Cref{table1}), standing genetic variation can drive adaptation when compared to direct mutaion.
+Evolutionary rescue through direct mutation is more likely if the cost of aneuploidy, $c$, is very large or the effect of the drug $\Delta_s$ is small.
+In contrast, standing genetic variation will drive adaptation if the pre-treatment chromosome missegregation rate, $\tilde{u}$, is very large.
+The ratio does not depend on the rate of chromosome missegregation induced by the drug, $u$.
+However, if the aneuploid growth rate, $\Delta_a$, increases, evolutionary rescue is driven by standing genetic variation.
+For the parameter values of  $\lambda_s=0.1$, $\lambda_a=0.0899$, $\lambda_m=0.1$, $\mu_s=0.14$, $\mu_a=0.09$, $\mu_m=0.09$, $\tilde{u}=10^{-3}$, and $v=10^{-7}$, we are in the first scenario (tolerant aneuploids) and obtain the ratio $\tilde{N}^*_a/N^*_m \approx 1.94$, which means that standing genetic variation does not drive evolution of drug resistance when compared to direct mutation.
+We note that for larger values of the pre-treatment chromosome missegregation rate, $\tilde{u}$, which are consistent with empirical studies (\Cref{table1}), standing genetic variation can drive adaptation when compared to direct mutation.
 
@@
 
-When evolutionary rescue occurs, the time until recurrence of the tumor may still be very long.
-We therefore explored the time until recurrence of the tumor, that is, the time until the tumor reaches its original size, $N$.
-When the expected number of resistant lineages that avoid extinction is small, the expected recurrence time can be estimated by adding two terms: the \emph{mean evolutionary rescue time}, which is the waiting time for appearance of a resistant lineage that avoids extinction (conditioned on such an event occurring in the first place), and the \emph{mean proliferation time}, which is the expected time for that lineage to grow to $N$ cells.
-However, when the expected number of resistant lineages is large, the dynamics of the number of mutant cells is deterministic (i.e., it can be modeled by a system of ODEs, \cref{detODE}) and the mean recurrence time cannot be separated into the mean evolutionary rescue time and mean proliferation time because multiple mutant lineages contribute towards the mutant population size reaching the initial tumor size.
-% TODO I dont like the word "behaves", please be specific.
-- what exactly is deterministic and why does it mean you cannot do the separation.
-ok
-Of particular interest is the distribution of the evolutionary rescue time and recurrence time with tolerant aneuploid cells ($\Delta_a T^*\ll1$), for which we focus on the parameter values  $\lambda_s=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$ .
+When evolutionary rescue occurs, the time until the tumor recurs may still be very long.
+We therefore explored the time until the tumor recurs, that is, the time until the tumor reaches its original size, $N$.
+When the expected number of resistant lineages that avoid extinction is small, the expected recurrence time can be estimated by adding two terms: the \emph{mean evolutionary rescue time}, which is the waiting time for the appearance of a resistant lineage that avoids extinction (conditioned on such an event occurring in the first place), and the \emph{mean proliferation time}, which is the expected time for that lineage to grow to $N$ cells.
+However, when the expected number of resistant lineages is large, the dynamics of the number of mutant cells is deterministic (i.e., it can be modeled by a system of ODEs, \cref{detODE}), and the mean recurrence time cannot be separated into the mean evolutionary rescue time and mean proliferation time because multiple mutant lineages contribute towards the mutant population size reaching the initial tumor size.
+Of particular interest is the distribution of the evolutionary rescue time and recurrence time with tolerant aneuploid cells ($\Delta_a T^*\ll1$), for which we focus on the parameter values $\lambda_s=0.1$, $\lambda_a=0.0899$, $\lambda_m=0.1$, $\mu_s=0.14$, $\mu_a=0.09$, $\mu_m=0.09$, $u=10^{-2}$, and $v=10^{-7}$.
 
 \paragraph{Evolutionary rescue time.}
-In Appendix \ref{sec:appendix_rescue_time} we have derived approximations for $\tau_m$, the mean evolutionary rescue time without aneuploidy ($u=0$), and $\tau_a$, the mean rescue time with aneuploidy ($u>0$), both conditioned on evolutionary rescue occurring.
-These approximations are in good agreement with simulation results for small, intermediate, and large tumor sizes (\Cref{EvolutionaryRescueTimeComplete,MeanTimeGrowthAneuploidyPlot}).
-The mean rescue time with aneuploidy for small and large tumors follows these expressions (Appendix \ref{sec:appendix_rescue_time}),
+We have derived approximations for $\tau_m$, the mean evolutionary rescue time without aneuploidy ($u=0$), and $\tau_a$, the mean rescue time with aneuploidy ($u>0$), both conditioned on evolutionary rescue occurring (Appendix~\ref{sec:appendix_rescue_time}).
+These approximations agree with simulation results for small, intermediate, and large tumor sizes (\Cref{EvolutionaryRescueTimeComplete,MeanTimeGrowthAneuploidyPlot}).
+The mean rescue time with aneuploidy for small and large tumors follows
 \begin{equation}  \label{eq:AsymptoticTimeRules}
@@
 \end{equation} 
-
 For small tumors ($ N \ll N_a^*$), the mean rescue time is a function of the sensitive and aneuploid growth rates and independent of the other model parameters, including tumor size (blue line in \Cref{EvolutionaryRescueTimeComplete}).
@@
 For large tumors ($N \gg N_m^*$), the mean evolutionary rescue time (\cref{eq:AsymptoticTimeRules}) is independent of parameters characterizing aneuploid cells or their production ($u$, $\lambda_a$, and $\Delta_a$).
-Increasing the per division mutation rate, $v$, leads to faster appearance of a rescue mutation and hence reduced mean rescue time. 
-Finally, increasing the tumor size leads to shorter mean rescue time, as there are more sensitive cells that can mutate to become resistant.
-
-Given that a fraction $f\approx 0.14\%$ of the initial cancer cell population is expected to have the beneficial aneuploidy even before the onset of drug treatment, we want to know whether the mean evolutionary rescue time is affected by the standing genetic variation.
-We calculated the mean evolutionary rescue time with standing genetic variation $\tilde{\tau}_a$ (\cref{meantimet2SGV}) and compared our result with simulations in  \Cref{SGVEvolutionaryRescueTimeComplete}.
-We note that standing genetic variation does not have a significant effect on the mean evolutionary rescue time.
-
-In Appendix \ref{sec:appendix_distribution_time} we calculate the probability that a rescue mutation has been generated by time $t$.
+Increasing the per division mutation rate, $v$ leads to the faster appearance of a rescue mutation and hence reduced mean rescue time. 
+Finally, increasing the tumor size leads to shorter mean rescue time, as more sensitive cells can mutate to become resistant.
+
+Given that a fraction $f\approx 0.14\%$ of the initial cancer cell population is expected to have beneficial aneuploidy even before the onset of drug treatment, we want to know whether the mean evolutionary rescue time is affected by the standing genetic variation.
+We calculated the mean evolutionary rescue time with standing genetic variation, $\tilde{\tau}_a$ (\cref{meantimet2SGV}), and compared our result with simulations (\Cref{SGVEvolutionaryRescueTimeComplete}).
+We note that standing genetic variation does not significantly affect the mean evolutionary rescue time.
+
+We calculate the probability that a rescue mutation has been generated by time $t$ in Appendix~\ref{sec:appendix_distribution_time}.
 This allows us to examine whether aneuploidy promotes or delays evolutionary rescue.
-We find that anenploidy promotes evolutionary rescue after $1/\Delta_a\approx100$ days, at a time where no more rescue mutations are generated through mutations in sensitive cells (\Cref{cdffig}A).
+We find that aneuploidy promotes evolutionary rescue after $1/\Delta_a\approx100$ days, at a time when no more rescue mutations are generated through mutations in sensitive cells (\Cref{cdffig}A).
 Thus, aneuploidy increases the \emph{window of opportunity} for evolutionary rescue.
-This can have a counter-intuitive outcome: conditioned on rescue of the tumor, tumors rescued by aneuploid cells may acquire rescue mutations later than those rescued by sensitive cells.
+This can have a counter-intuitive outcome: conditioned on the rescue of the tumor, tumors rescued by aneuploid cells may acquire rescue mutations later than those rescued by sensitive cells.
 
@@
 
-We next approximated the mean time for the population of mutant cancer cells to reach the initial, pre-treatment population size $N$, which we denote the recurrence time $\tau_a^r$ (Appendix \ref{sec:appendix_recurrence_time}),
+We next approximated the mean time for the population of mutant cancer cells to reach the initial, pre-treatment population size $N$, which we denote the recurrence time $\tau_a^r$ (Appendix~\ref{sec:appendix_recurrence_time}),
 \begin{equation} \label{eq:AsymptoticTimeRulesRecurrence}
@@
 \end{equation}
-\Cref{proliferationFigure,ProliferationTimeLarge} show the agreement between our approximations and simulations.
+\Cref{proliferationFigure,ProliferationTimeLarge} show the agreement between our approximations and simulation results.
 For small tumors ($N \ll N_a^*$), the mean recurrence time can be approximated as the sum of the mean time for the first rescue mutation to appear and the mean time for its lineage to reach size $N$.
 The mean recurrence time grows logarithmically with tumor size $N$ and is the same order of magnitude as the mean evolutionary rescue time.
-Increasing the mutant growth rate $\Delta_m$ decreases recurrence times while increasing the sensitive and aneuploid growth rates, $\Delta_s$ and $\Delta_a$ respectively, increases the recurrence time.
-For large tumors ($N \gg N_m^*$), the dynamics of the number of mutant cells is deterministic and the mean recurrence time becomes independent of the initial tumor size $N$.
-Increasing either the mutant growth rate $\Delta_m$ or the mutation rate $v$ leads to a decrease in the time for the tumor to rebound to its initial size.
-In addition, drugs that significantly increase the sensitive death rate $\mu_s$ and do not affect the division rate $\lambda_s$ delay cancer recurrence.
+Increasing the mutant growth rate, $\Delta_m$, decreases the recurrence time, while increasing the sensitive and aneuploid growth rates, $\Delta_s$ and $\Delta_a$, respectively, increases the recurrence time.
+For large tumors ($N \gg N_m^*$), the dynamics of the number of mutant cells is deterministic, and the mean recurrence time becomes independent of the initial tumor size $N$.
+Increasing either the mutant growth rate, $\Delta_m$, or the mutation rate, $v$, decreases the time for the tumor to rebound to its initial size.
+In addition, drugs that significantly increase the death rate of drug-sensitive cells, $\mu_s$, but do not affect their division rate, $\lambda_s$, delay cancer recurrence (conditioned on evolutionary rescue).
 Consequently, patients treated with such drugs may require a longer period of monitoring to guarantee the effectiveness of the treatment.
 
-We note that, for small and large tumors, when $N \ll N_a^*$ or $N \gg N_m^*$, the asymptotic expressions for the mean recurrence time are independent of the chromosome missegregation rate $u$, and therefore the rate at which the drug induces aneuploidy has no effect on the time necessary for the tumor to rebound to its initial size $N$.
-
-In Appendix F we derive the probability that a mutant cancer cell population has not reached size $N$ by time $t$.
-\Cref{cdffig}B shows agreement between our approximations and stochastic simulations for various values of $N$.
-Additionally, we derive the distribution of the recurrence time for the scenario $N=10^6$ (small tumor), noting that the distribution is wide and right-skewed (\Cref{KMdistribution}).
-It is highly unlikely to observe the recurrence of tumors at times smaller then $\frac{1}{\Delta_m}\log\frac{\Delta_m-\Delta_s}{v\lambda_s}\approx1542$ days for the parameter values $\lambda_s=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$ and independent of initial tumor size $N$~(\Cref{cdffig}B).
-
-The detection time $\tau_a^{M}$ can be defined as the time for the tumor size to reach detection threshold $M$.
+We note that, for small and large tumors, when $N \ll N_a^*$ or $N \gg N_m^*$, the asymptotic expressions for the mean recurrence time are independent of the chromosome missegregation rate $u$, and therefore, the rate at which the drug induces aneuploidy has no effect on the time for the tumor to rebound to its initial size $N$.
+
+Appendix F gives us the probability that a mutant cancer cell population has not reached size $N$ by time $t$.
+\Cref{cdffig}B shows agreement between our approximations and simulation results for various values of $N$.
+Additionally, we derive the distribution of the recurrence time for a small tumor with $N=10^6$ cells, noting that the distribution is wide and right-skewed (\Cref{KMdistribution}).
+It is highly unlikely to observe the recurrence of cancer at times smaller than $\frac{1}{\Delta_m}\log\frac{\Delta_m-\Delta_s}{v\lambda_s}\approx1542$ days for the parameter values in \Cref{table1} with $\lambda_a=0.0899$, $\mu_s=0.14$, and $v=10^{-7}$ and independent of initial tumor size $N$~(\Cref{cdffig}B).
+
+The detection time $\tau_a^{M}$ is defined as the time for the tumor size to reach detection threshold $M$.
 We derive the mean detection time for $M=10^7$ in Appendix \ref{sec:appendix_recurrence_time}.
-We observe that  for small and intermediate sized tumors the effect of the detection size $M$ on $\tau_a^{M}$ is negligible when compared to the scenario where the detection size is equal to the initial population size $N$ (i.e., $\tau_a^r\approx\tau_a^{M}$ for $N<N_m^*$).
-However, for large tumors the mean recurrence time to detection size $M$ decreases logarithmically with tumor size $N$ while $\tau_a^{r}$ is constant (\Cref{RecurrencePlot}).
-Additionally, for large tumors we have $M<N_m^*<N$ so the mean detection time is shorter for such tumors compared to the mean recurrence time, that is, the resistant tumor can be detected before recovering back to its initial size.
+We find for small and intermediate-sized tumors, the detection size $M$ has a negligible effect on the mean detection time $\tau_a^{M}$ compared to when the detection size equals the initial population size, $N$ (i.e., $\tau_a^r\approx\tau_a^{M}$ for $N<N_m^*$).
+However, for large tumors, the mean detection time $\tau_a^{M}$ decreases logarithmically with tumor size $N$, while the recurrence time $\tau_a^{r}$ is constant (\Cref{RecurrencePlot}).
+Additionally, for large tumors, we have $M<N_m^*<N$, so the mean detection time is shorter compared to the mean recurrence time, that is, the resistant tumor may be detected before recovering back to its initial size.
 
@@
 
-We have modeled a tumor--a population of cancer cells--exposed to drug treatment that causes it to decline in size towards potential extinction.
-In this scenario, the tumor can be ``evolutionary rescued'', or escape extinction, via two paths.
-In the direct path, a sensitive cell acquires a mutation or aneuploidy that confers resistance that allows it to rapidly grow.
-In the indirect path, a sensitive cell first becomes aneuploid, which diminishes the effect of the drug, and then an aneuploid cell acquires a mutation that confers resistance (\Cref{figureAneuploidy}).
-
-\paragraph{Evolutionary rescue}
-Using multitype branching processes, we derived the probability of evolutionary rescue of the tumor under effects of aneuploidy, ranging from tolerance to partial resistance.
+We have modeled a tumor--a population of cancer cells--exposed to drug treatment that causes it to decline in size toward potential extinction.
+In this scenario, the tumor can be ``evolutionary rescued'' or escape extinction via two paths.
+In the direct path, a drug-sensitive cell acquires a mutation or aneuploidy that confers resistance and allows it to grow rapidly.
+In the indirect path, a sensitive cell first becomes aneuploid, which diminishes the drug's effect, and then an aneuploid cell acquires a mutation that confers resistance (\Cref{figureAneuploidy}).
+
+Using multi-type branching processes, we derived the probability of evolutionary rescue of the tumor under the effects of aneuploidy, ranging from tolerance to partial resistance.
 We obtained exact and approximate expressions for the probability of evolutionary rescue (\cref{eq:rescue_prob}). 
-Our results show that the probability of evolutionary rescue increases with the initial tumor size $N$, the sensitive growth rate $\Delta_s$, the mutation rate $v$, and the aneuploidy rate $u$.
+Our results show that the probability of evolutionary rescue increases with the initial tumor size $N$, the drug-sensitive growth rate $\Delta_s$, the mutation rate $v$, and the aneuploidy rate $u$.
 
@@
 When aneuploidy only provides tolerance to the drug ($\Delta_s\ll\Delta_a\ll0\ll\Delta_m$), it cannot rescue the population.
-Instead, it acts as a \emph{stepping stone} through which the resistant mutant can appear more rapidly, given that the aneuploid cell population size declines slower then that of the sensitive cell population (\Cref{sampleTrajectories}).
-In this scenario, aneuploidy provides two benefits.
-First, it delays the extinction of the population, providing more time for appearance of the resistance mutation.
-Second, it increases the population size relative to a sensitive population, providing more cells in which mutations can occur, i.e., it increases the mutation supply (i.e., $Nuv\lambda_s\lambda_a/\abs{\Delta_s\Delta_a}$).
-
-We find that aneuploidy can have a significant effect on evolutionary rescue as it reduces the threshold tumor size by at least an order-of-magnitude even when aneuploidy only provides tolerance (\Cref{rescue_prob}).
-When the number of cells in the tumor is large enough (i.e., $N\gg N_m^*\approx 4\times10^7$), a resistance mutation will occur in sensitive cells before these cells become extinct (\Cref{rescue_prob}).
-Therefore, large tumors are certain to be rescued with or without aneuploidy.
+Instead, it acts as a ``stepping stone'' through which the resistant mutant can appear more rapidly, given that the number of aneuploid cells declines slower than the number of drug-sensitive cells (\Cref{sampleTrajectories}).
+In this scenario, aneuploidy provides two advantages.
+First, it delays the extinction of the population, providing more time for the appearance of the resistance mutation.
+Second, it increases the population size relative to a drug-sensitive population, providing more cells in which mutations can occur, i.e., it increases the mutation supply (i.e., $Nuv\lambda_s\lambda_a/\abs{\Delta_s\Delta_a}$).
+
+We find that aneuploidy can significantly affect evolutionary rescue as it reduces the threshold tumor size by at least an order of magnitude even when aneuploidy only provides tolerance (\Cref{rescue_prob}).
+When the number of cells in the tumor is large enough (i.e., $N\gg N_m^*\approx 4\times10^7$), a resistance mutation will occur in drug-sensitive cells before these cells become extinct (\Cref{rescue_prob}).
+Therefore, large tumors are likely to be rescued with or without aneuploidy.
 However, anti-cancer drugs are often used as adjuvant therapy after resection, in which case the number of cells in the tumor may be below the detection threshold of $\sim10^7$  \citep{bozic2013evolutionary}.
-In these cases, aneuploidy can have a crucial role in the evolutionary rescue of the tumor and therefore in cancer recurrence.
+In these cases, aneuploidy can have a crucial role in the evolutionary rescue of the tumor and, therefore, in cancer recurrence.
 Indeed, secondary tumors are estimated to cause the majority of cancer-related deaths~\citep{chaffer2011perspective}.
-The importance of aneuploidy in the evolutionary rescue of secondary tumors is reinforced by the fact that metastases have been shown to have a chromosome missegregation rate two to three orders of magnitude higher then that of primary tumors~\citep{kimmel2023intra}.
-
-Given the fact that the mean time for secondary tumors to overcome chemotherapy can be of the order of 1,000 days (\Cref{MeanTimeGrowthAneuploidyPlot}A), aneuploidy can explain the reappearance of cancer even after initial remission.
-The theoretical prediction for the mean rescue time for tumors smaller then  $10^8$ cells is greater than 4 years, consistent with previous estimates of the recurrence time of tumors after resection~\citep{avanzini2019cancer}.
-We found that aneuploidy complements evolutionary rescue through direct mutation because it generates rescue mutations mostly after the sensitive population has died out and direct mutation is no longer a viable option for evolutionary rescue (\Cref{cdffig}A).
-
-We hypothesized that standing genetic variation (the existence of a subpopulation of aneuploid cancer cells before the onset of therapy) could facilitate evolutionary rescue by reducing the waiting time for the appearance of aneuploid cells.
-We found that a drug that reduces the sensitive growth rate and does not significantly increase the chromosome missegregation rate is likely to lead to evolutionary rescue through standing genetic variation (\Cref{rescue_denovo,eq:Nsgv}).
-Furthermore, if the fraction of tumor cells that have the beneficial aneuploidy is $f\gg\frac{u\lambda_s}{\abs{\Delta_s}}\approx2.5\%$, then evolutionary rescue is more likely to occur via standing variation rather then through \emph{de-novo} aneuploidy.
-However, for the parameter values we focus on in our examples (\Cref{table1}), this fraction is an order of magnitude lower, and therefore we expect evolutionary rescue to occur by \emph{de-novo} aneuploidy.
-
-Finally, we observe from \Cref{rescue_threshold} that only for a restricted region of the parameter space will aneuploidy act as a \emph{stepping stone} for evolutionary rescue.
-If the aneuploid division rate is smaller than $\frac{\mu_a}{1+u}\approx\mu_a(1-u)$ then evolutionary rescue will occur through direct mutation.
-As a result, for most parameter values, aneuploidy will either not play any role in evolutionary rescue or will be the main driver of adaptation without requiring any mutation (i.e.
-evolutionary rescue in one step).
-
-\paragraph{Experimental validation}
-Our model predictions could be tested by experiments.
+The importance of aneuploidy in the evolutionary rescue of secondary tumors is reinforced by the fact that metastases have been shown to have a chromosome missegregation rate two to three orders of magnitude higher compared to primary tumors~\citep{kimmel2023intra}.
+
+Given that the mean time for secondary tumors to adapt to anti-cancer drugs can be of the order of 1,000 days (\Cref{MeanTimeGrowthAneuploidyPlot}A), aneuploidy can explain the reappearance of cancer even after initial remission.
+The theoretical prediction for the mean rescue time of tumors smaller than $10^8$ cells is greater than 4 years, consistent with previous estimates of the recurrence time of tumors after resection~\citep{avanzini2019cancer}.
+We found that aneuploidy complements evolutionary rescue through direct mutation because it produces rescue mutations mostly after the number of sensitive cells has decreased to a point where a direct mutation is no longer a feasible option for evolutionary rescue (\Cref{cdffig}A).
+
+We hypothesized that standing genetic variation (the existence of aneuploid cancer cells in the tumor before the onset of therapy) could facilitate evolutionary rescue by reducing the waiting time for the appearance of aneuploid cells.
+We found that a drug that reduces the sensitive growth rate and does not significantly increase the chromosome missegregation rate will likely lead to evolutionary rescue through standing genetic variation (\Cref{rescue_denovo,eq:Nsgv}).
+Furthermore, if the fraction of tumor cells that have the beneficial aneuploidy is $f\gg\frac{u\lambda_s}{\abs{\Delta_s}}\approx2.5\%$, then evolutionary rescue is more likely to occur via standing variation rather than through \emph{de-novo} aneuploidy.
+However, for the parameter values we focus on in our examples (\Cref{table1}), this fraction is an order of magnitude lower, and therefore, we expect evolutionary rescue to occur by \emph{de-novo} aneuploidy.
+
+Finally, we observe from \Cref{rescue_threshold} that only for a restricted region of the parameter space will aneuploidy act as a ``stepping stone'' for evolutionary rescue.
+If the aneuploid division rate is smaller than $\frac{\mu_a}{1+u}\approx\mu_a(1-u)$, then evolutionary rescue will occur through direct mutation.
+As a result, for most parameter values, aneuploidy will either not play any role in evolutionary rescue or will be the main driver of adaptation without requiring any mutation (i.e., evolutionary rescue in one step).
+
+\paragraph{Directions for future research}
+Experiments could test our model predictions.
 For example, to assess the effect of initial tumor size on the probability of evolutionary rescue, a large culture mass can be propagated from a single cancer cell in permissive conditions and then diluted to a range of starting tumor sizes.
@@
 
-\paragraph{Directions for future research}
 We have assumed that cancer cell lineages are independent and have verified that this is accurate under simple logistic growth.
-However, this assumption neglects potential effects of spatial structure and local interactions, which may be important in solid tumors.
-Such tumors can be spatially heterogeneous with different genotypes inhabiting cellular niches and immune infiltration impacting growth in affected regions \citep{varrone2023cellcharter,galon2010immune}.
-This has the potential to impact the probability of evolutionary rescue \citep{martens2011spatial}.
+However, this assumption neglects the potential effects of spatial structure and local interactions, which may be important in solid tumors.
+Such tumors can be spatially heterogeneous, with different genotypes inhabiting cellular niches and immune infiltration impacting growth in affected regions \citep{varrone2023cellcharter,galon2010immune}.
+This can potentially impact the probability of evolutionary rescue \citep{martens2011spatial}.
 In addition, our model can be extended to understand evolutionary rescue in different biological contexts, for example, how yeast populations under stress overcome extinction via aneuploidy~\citep{pompei2023fitness,kohanovski2024aneuploidy}.
@@
 
-Our results quantitatively suggest that aneuploidy may play an important role in tumor adaptation to anti-cancer drugs when tumor size is small or intermediate.
-Large tumors are predicted to be more likely to adapt to anti-cancer drugs through direct mutation, whereas smaller tumors are predicted to become resistant either directly by aneuploidy or by a resistance mutation occuring in aneuploid cells that serve as an evolutionary ``stepping stone'' (\Cref{rescue_prob}).
+Our results quantitatively suggest that aneuploidy may play an important role in tumor adaptation to anti-cancer drugs when the tumor size is small or intermediate.
+Large tumors are predicted to adapt to anti-cancer drugs through direct mutation.
+In contrast, smaller tumors are predicted to become resistant either directly by aneuploidy or by a resistance mutation occurring in aneuploid cells that serve as an evolutionary ``stepping stones'' (\Cref{rescue_prob}).
 Thus, therapies that increase the rate of aneuploidy in tumors to combat cancer may also promote drug resistance.
@@
 \caption{\textbf{Model parameters.} 
-Parameters from \citet{bozic2013evolutionary} consider patients with melanoma treated with the anti-cancer drug vemurfenib, in which resistance is conferered by trisomy in either Chr 2 or Chr 6.
+Parameters from \citet{bozic2013evolutionary} consider patients with melanoma treated with the anti-cancer drug vemurfenib, in which resistance is conferred by trisomy in either Chr 2 or Chr 6.
 We have modified the parameters from \citet{bozic2013evolutionary} such that sensitive and mutant division rates are $\lambda_{s} = \lambda_{m} = \log2/T\approx0.1$ instead of their value of $0.14$ where $T$ is the doubling time in the absence of cellular death obtained from \citet{rew2000cell}.
@@
 \textbf{Model illustration.}
-\textbf{(A)} A population of cancer cells is composed of sensitive, aneuploid, and mutant cells, which divide with rates $\lambda_s$, $\lambda_a$, and $\lambda_m$ and die at rates $\mu_s$, $\mu_a$, and $\mu_m$, respectively.
+\textbf{(A)} A population of cancer cells is composed of drug-sensitive, aneuploid, and mutant cells, which divide with rates $\lambda_s$, $\lambda_a$, and $\lambda_m$ and die at rates $\mu_s$, $\mu_a$, and $\mu_m$, respectively.
 Sensitive cells can divide and become aneuploid at rate $u\lambda_s$.
-Both aneuploid and sensitive cells can divide and acquire a mutation with rate $v\lambda_a$ and $v\lambda_s$, respectively.
+Both aneuploid and sensitive cells can divide and acquire a mutation with rates $v\lambda_a$ and $v\lambda_s$, respectively.
 Color denotes the relative growth rates of the three genotypes such that $\lambda_s - \mu_s < \lambda_a - \mu_a < \lambda_m - \mu_m$.
-\textbf{(B)} Sensitive cells are susceptible to the drug, while mutant cells are resistant.
-The aneuploid may be tolerant, stationary and partially resistant.
+\textbf{(B)} Sensitive cells are sensitive to the drug, while mutant cells are drug-resistant.
+The aneuploid may be tolerant, stationary, or partially resistant.
 }
@@
 \caption{
-\textbf{Sample trajectories of the different genotype frequencies.}
-(A) Without aneuploidy ($u=0$) evolutionary rescue is possible through direct mutation and in most scenarios the tumor will become extinct due to the drug.
-(B) When aneuploid cells are tolerant ($\Delta_a<0$), we observe that mutant lineages are generated at later times then in A.
+\textbf{Sample trajectories of the genotype frequencies.}
+(A) Without aneuploidy ($u=0$), evolutionary rescue is possible through direct mutation, and in most scenarios, the tumor will become extinct due to the drug.
+(B) When aneuploid cells are tolerant ($\Delta_a<0$), we observe that mutant lineages are generated at later times than in A.
 (C) When aneuploid cells are stationary ($\Delta_a\approx0$), we observe the appearance of mutant lineages even after the sensitive population has gone extinct, thus showing that stationary aneuploidy increases the probability of evolutionary rescue.
 (D) When aneuploid cells are partially resistant ($\Delta_a>0$), the tumor is rescued by the aneuploid cell population.
-Each plot shows 10 simulations of the number of sensitive, aneuploid and mutant cells $\left(s_t,a_t,m_t\right)$ over time $t$ with the following parameter values: $\lambda_s=0.1,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, v=10^{-7},N=10^7$.
-For (A) we set $u=0$, for (B) $\lambda_a=0.085,u=10^{-2}$, for (C) $\lambda_a=0.0899,u=10^{-2}$ and for (D) $\lambda_a=0.095,u=10^{-2}$.
+Each plot shows 10 simulations of the number of sensitive, aneuploid, and mutant cells $\left(s_t,a_t,m_t\right)$ over time $t$.
+Here, $\lambda_s=0.1$, $\lambda_m=0.1$, $\mu_s=0.14$, $\mu_a=0.09$, $\mu_m=0.09$, $v=10^{-7}$ , $N=10^7$; (A) $u=0$; (B) $\lambda_a=0.085$, $u=10^{-2}$; (C) $\lambda_a=0.0899$, $u=10^{-2}$; (D) $\lambda_a=0.095$, $u=10^{-2}$.
+% TODO panel C run simulations for \lambda_a=0.08999 instead of 0.0899 so that it would give T\Delta_a =~ -0.555 to be in the stationary case (right now it is ~-5.5 which is the tolerant case)
 }
@@
 \includegraphics[width=1\textwidth]{Figures/ProbvNPlot.pdf}
-\caption{\textbf{Aneuploidy facilitates evolutionary rescue of cancer under drug treatment.}
+\caption{\textbf{Aneuploidy facilitates the evolutionary rescue of cancer under drug treatment.}
 The probability of evolutionary rescue (i.e., the probability that the population does not become extinct), $\presc$, as a function of the initial tumor size, $N$ (\cref{eq:rescue_prob}).
@@
 Green line represents the scenario with partially resistant aneuploidy ($u=10^{-2}, \lambda_a=0.095$).
-Dots for simulations and the error bars for $95\%$ confidence interval ($p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the fraction of simulations in which the tumor has been rescued and $n=100$ is the number of simulations).
+Dots for simulations and the error bars for $95\%$ confidence interval ($p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the fraction of simulations in which the tumor has been rescued, and $n=100$ is the number of simulations).
 Parameters: $\lambda_s=0.1,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, v=10^{-7}$.}
@@
 The dashed horizontal line shows $N^*_m$, the threshold tumor size without aneuploidy ($u=0$).
-When aneuploid growth rate is close to or higher than zero, aneuploidy decreases the threshold tumor size, thereby facilitating evolutionary rescue.
+When aneuploid growth rate is close to or higher than zero, aneuploidy decreases the threshold tumor size, facilitating evolutionary rescue.
 The inset highlights the scenario when aneuploid cells are stationary.
@@
 Dashed horizontal line shows $N^*_m$, the threshold tumor size without aneuploidy ($u=0$).
-When the aneuploidy rate is much higher than the mutation rate, aneuploidy decreases the threshold tumor size, thereby facilitating evolutionary rescue.
+When the aneuploidy rate is much higher than the mutation rate, aneuploidy decreases the threshold tumor size, facilitating evolutionary rescue.
 Blue line represents the exact formula for threshold tumor size $N_a^*$ while the solid black line represents the approximation (\cref{eq:N_a}).
-Red dots represents simulation results and the error bars represent the $95\%$ confidence intervals obtained with bootstrap (Appendix G).
+Red dots represent simulation results, and the error bars represent the $95\%$ confidence intervals obtained with bootstrap (Appendix G).
 Parameters: $\lambda_s=0.1,\lambda_m=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, v=10^{-7}$.
@@
 \caption{
-\textbf{Standing genetic variation facilitates evolutionary rescue of cancer.}
+\textbf{Standing genetic variation facilitates the evolutionary rescue of cancer.}
 (A)  Ratio of threshold tumor sizes for rescue by standing genetic variation and by de-novo variation, $\tilde{N}_a^*/N_a^*$, when a fraction $\frac{\tilde{u}\lambda_s}{c}$ is aneuploid at the start of treatment, as a function of the sensitive growth rate $\Delta_s$.
 Standing genetic variation will drive adaptation to the drug if the sensitive population is rapidly declining ($\Delta_s\ll0$) due to a stronger effect of the drug on sensitive cells.
-Red dots represents simulation results and the error bars represent the $95\%$ confidence intervals obtained with bootstrap (Appendix G).
+Red dots represent simulation results, and the error bars represent the $95\%$ confidence intervals obtained with bootstrap (Appendix G).
 Parameters: $\lambda_s=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_a=0.09,\mu_m=0.09,\tilde{u}=10^{-3},u=10^{-2}, v=10^{-7}$.
-(B) Ratio of threshold tumor size $\tilde{N}_a^*$, when a fraction $\frac{\tilde{u}\lambda_s}{c}$ is aneuploid at the start of treatment, and $N_a^*$ as a function of the the ratio of aneuploidy rates $\tilde{u}/u$.
-\emph{De-novo} aneuploids will have larger contribution to the appearance of drug resistance if the drug induces the appearance of aneuploid cells ($u \gg \tilde u$).
-Red dots represents  simulation results and the error bars represent the $95\%$ confidence intervals obtained with bootstrap (Appendix G).
+(B) Ratio of threshold tumor size $\tilde{N}_a^*$, when a fraction $\frac{\tilde{u}\lambda_s}{c}$ is aneuploid at the start of treatment, and $N_a^*$ as a function of the ratio of aneuploidy rates $\tilde{u}/u$.
+\emph{De-novo} aneuploids will have a larger contribution to the appearance of drug resistance if the drug induces the appearance of aneuploid cells ($u \gg \tilde u$).
+Red dots represent simulation results, and the error bars represent the $95\%$ confidence intervals obtained with bootstrap (Appendix G).
 Parameters: $\lambda_s=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09,\tilde{u}=10^{-3}, v=10^{-7}$.
@@
 Black line: no aneuploidy ($u=0$).
-Aneuploidy starts to play an important role in rescuing the tumor cell population as the sensitive population becomes extinct.
-Markers represent simulation results and the error bars represent $95\%$ confidence interval ($p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the fraction of simulations in which a successful mutant has not been generated and $n=100$ is the number of simulations).
+Aneuploidy plays an important role in rescuing the tumor cell population as the sensitive population becomes extinct.
+Markers represent simulation results, and the error bars represent $95\%$ confidence interval ($p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the fraction of simulations in which a successful mutant has not been generated, and $n=100$ is the number of simulations).
 Parameters: $\lambda_s=0.1,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7},N=10^7$.
@@
 Green line: $N=10^6$ (small tumor).
-Red line: $N=10^7$ (intermediate sized tumor).
+Red line: $N=10^7$ (intermediate-sized tumor).
 Blue line: $N=10^{10}$ (large tumor).
 Increasing the initial tumor size guarantees that the cancer will relapse.
-Markers represent simulations and the error bars represents $95\%$ confidence interval ($p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the fraction of the simulations in which the mutant population size has not reached $N$ and $n=100$ is the number of simulations).
+Markers represent simulations, and the error bars represent $95\%$ confidence interval ($p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the fraction of the simulations in which the mutant population size has not reached $N$ and $n=100$ is the number of simulations).
 Parameters: $\lambda_s=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}
@@
 %%%
-\section{Aneuploidy inhances mutation rates}\label{sec:appendix-diff-mut}
-Aneuploidy has the potential to increase the mutation rate in cells which are in this state.
-We change our model such that sensitive cells divide and mutate with rate $v_s\lambda_s$ and aneuploid cells divide and mutate with rate $v_a\lambda_a$.
-As a result, we re-calculate the survival probabilities $p_s,p_a$ and $p_s$ as we did in Appendix \ref{sec:appendix-surv-prob}:
+\section{Aneuploidy-induced mutation rate}\label{sec:appendix-diff-mut}
+The mutation rate may be increased in aneuploid cells. 
+To account for an increased mutation rate in cells with the aneuploidy that provides a fitness advantage in the presence of a drug, we extend our model such that sensitive cells mutate with rate $v_s$ and aneuploid cells mutate with rate $v_a$.
+Note that sensitive cells include those cells with any other aneuploidy, including those that may cause an increased mutation rate, and therefore those cases are already covered by the model presented in the main text.
+We then calculate the survival probabilities $p_s$, $p_a$ and $p_s$ as in Appendix \ref{sec:appendix-surv-prob},
 \begin{equation} \label{eq:extinction_prob_ane}
@@
 \end{equation}
-Solving the above equations we obtain:
+Solving the above equations we obtain
 \begin{equation}\label{eq:survival_prob_ane}
@@
 \end{equation}
-Consequently, the threshold population size can be written as:
+Consequently, the threshold population size can be written as
 \begin{equation}\label{eq:N_a2}
@@
 \end{equation}
-where $T^*=\left(4v_a\lambda_a^2p_m\right)^{-\frac{1}{2}}$. The probability of evolutionary rescue is:
+where $T^*=\left(4v_a\lambda_a^2p_m\right)^{-\frac{1}{2}}$.
+This is the same as in \cref{eq:N_a} except with $v_a$ instead of $v$.
+
+The probability of evolutionary rescue is
 \begin{equation} \label{eq:rescue_prob_ane} 
@@
 \end{equation}
-which we plot in Figure \ref{rescue_prob_ane} for multiple values of $v_a$. We note that for $v_a=10^{-5}$ we are in the case of stationary aneuploidy (i.e. $\Delta_a T^*\approx -0.55$ ). 
+which we plot in Figure \ref{rescue_prob_ane} for multiple values of $v_a$. We note that when $v_a=10^{-5}$, we are in the case of stationary aneuploidy (i.e., $\Delta_a T^*\approx -0.55$ ). 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \label{rescue_prob_ane}
+% TODO what is the value of v_s?
 \end{figure}

### 2024-07-30 EvolutionaryRescue1.tex
Add files via upload

@@
 They concluded that this can significantly enhance the population's probability of survival.
+\citet{iwasa2003evolutionary,iwasa2004evolutionary} analysed the probability that a population under strong selective pressure can survive extinction   through succesive mutations with the help of multitype branching processes.
 
@@
 Furthermore, increasing the aneuploid growth rate $\Delta_a$ (which appears both in the terms and in the conditions), also reduces the threshold tumor size, with a sharp decrease around $\Delta_a=0$, but the effect is minor when $|\Delta_a|$ is small compared to $T^*$ as this would result in the second scenario where  $dN_a^*/d\Delta_a=0$ (\Cref{rescue_threshold}A).
-% TODO these are pretty straightforward - can you glean something more interesting from eq 4?
-I am still looking for more interesting results from eq4
+If the mutation rate $v$ increases the tumor threshold size for tolerant aneuploids will decrease $\sim v^{-1}$ while the threshold size for stationary aneuploids will decrease $\sim v^{-\frac{1}{2}}$.
+The growth rate for which tolerant aneuploids can offer a viable path to rescuing the cancer cell population is in the interval $\left(-u\lambda_a,-1/T^*\right)$ which will become smaller as $v$ increases.
+As a result, for any value of $\Delta_a\ll-\left(T^*\right)^{-1}$ increasing the mutation rate will improve probability of evolutionary rescue of the trajectory $sensitive\rightarrow aneuploid\rightarrow mutant$ until the threshold $\Delta_a=-\left(T^*\right)^{-1}$ is reached.
+% TODO these are pretty straightforward - can you glean something more interesting from eq 4? I am still looking for more interesting results from eq4 ok
 
@@
 (A) Without aneuploidy ($u=0$), evolutionary rescue is possible through direct mutation, and in most scenarios, the tumor will become extinct due to the drug.
-(B) When aneuploid cells are tolerant ($\Delta_a<0$), we observe that mutant lineages are generated at later times than in A.
+(B) When aneuploid cells are tolerant ($\Delta_a<0$), we observe, similar to A, direct mutation is the only path for evolutionary rescue.
 (C) When aneuploid cells are stationary ($\Delta_a\approx0$), we observe the appearance of mutant lineages even after the sensitive population has gone extinct, thus showing that stationary aneuploidy increases the probability of evolutionary rescue.
@@
 Each plot shows 10 simulations of the number of sensitive, aneuploid, and mutant cells $\left(s_t,a_t,m_t\right)$ over time $t$.
-Here, $\lambda_s=0.1$, $\lambda_m=0.1$, $\mu_s=0.14$, $\mu_a=0.09$, $\mu_m=0.09$, $v=10^{-7}$ , $N=10^7$; (A) $u=0$; (B) $\lambda_a=0.085$, $u=10^{-2}$; (C) $\lambda_a=0.0899$, $u=10^{-2}$; (D) $\lambda_a=0.095$, $u=10^{-2}$.
-% TODO panel C run simulations for \lambda_a=0.08999 instead of 0.0899 so that it would give T\Delta_a =~ -0.555 to be in the stationary case (right now it is ~-5.5 which is the tolerant case)
+Here, $\lambda_s=0.1$, $\lambda_m=0.1$, $\mu_s=0.14$, $\mu_a=0.09$, $\mu_m=0.09$, $v=10^{-7}$ , $N=10^7$; (A) $u=0$; (B) $\lambda_a=0.065$, $u=10^{-2}$; (C) $\lambda_a=0.08999$, $u=10^{-2}$; (D) $\lambda_a=0.095$, $u=10^{-2}$.
+% TODO panel C run simulations for \lambda_a=0.08999 instead of 0.0899 so that it would give T\Delta_a =~ -0.555 to be in the stationary case (right now it is ~-5.5 which is the tolerant case) OK
 }
@@
 Dots for simulations and the error bars for $95\%$ confidence interval ($p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the fraction of simulations in which the tumor has been rescued and $n=100$ is the number of simulations).
-Parameters: $\lambda_s=0.1,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}$.}
+Parameters: $\lambda_s=0.1,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v_s=10^{-7}$.}
 \label{rescue_prob_ane}
-% TODO what is the value of v_s?
+% TODO what is the value of v_s? OK
 \end{figure}

### 2024-07-31 EvolutionaryRescue1.tex
revise latest changes by RS

@@
 They concluded that this can significantly enhance the population's probability of survival.
-\citet{iwasa2003evolutionary,iwasa2004evolutionary} analysed the probability that a population under strong selective pressure can survive extinction   through succesive mutations with the help of multitype branching processes.
 
@@
 Evolutionary rescue that requires two successive mutations (i.e., two steps) has been investigated by \citet{martin2013probability}, who tested their model's predictions with data from yeast and bacteria experiments.
+\citet{iwasa2003evolutionary,iwasa2004evolutionary} used multi-type branching process theory to approximate the probability that a population under strong selective pressure can survive extinction via succesive mutations.
 
@@
 Substituting these into $N^*=1/p_s$, we find approximations for the threshold tumor size, $N^*$. 
-In these approximations, an important quantity is $T^* = (4 v \lambda_a^2 \Delta_m/\lambda_m)^{-1/2}$, which is the critical time an aneuploid lineage needs to survive to produce a resistant mutant that avoids random extinction.
+In these approximations, an important quantity is $T^* = \sqrt{\lambda_m / 4 v \lambda_a^2 \Delta_m}$, which is the critical time an aneuploid lineage needs to survive to produce a resistant mutant that avoids random extinction.
 First, if aneuploidy is very rare ($u\lambda_a T^*< 1$), or if aneuploidy is rare ($u\lambda_a < -\Delta_a$) and very sensitive to the drug ($\Delta_a T^* < -1$), then it is likely that evolutionary rescue will occur through a direct resistance mutation in a sensitive cell without aneuploidy playing a role in the adaptive dynamics, such that
@@
 Furthermore, increasing the aneuploid growth rate $\Delta_a$ (which appears both in the terms and in the conditions), also reduces the threshold tumor size, with a sharp decrease around $\Delta_a=0$, but the effect is minor when $|\Delta_a|$ is small compared to $T^*$ as this would result in the second scenario where  $dN_a^*/d\Delta_a=0$ (\Cref{rescue_threshold}A).
-If the mutation rate $v$ increases the tumor threshold size for tolerant aneuploids will decrease $\sim v^{-1}$ while the threshold size for stationary aneuploids will decrease $\sim v^{-\frac{1}{2}}$.
-The growth rate for which tolerant aneuploids can offer a viable path to rescuing the cancer cell population is in the interval $\left(-u\lambda_a,-1/T^*\right)$ which will become smaller as $v$ increases.
-As a result, for any value of $\Delta_a\ll-\left(T^*\right)^{-1}$ increasing the mutation rate will improve probability of evolutionary rescue of the trajectory $sensitive\rightarrow aneuploid\rightarrow mutant$ until the threshold $\Delta_a=-\left(T^*\right)^{-1}$ is reached.
-% TODO these are pretty straightforward - can you glean something more interesting from eq 4? I am still looking for more interesting results from eq4 ok
+The tumor threshold size decreases with the mutation rate in the first and second scenarios: $N_a^*$ is proportional to $1/v$ in the first scenario (tolerant aneuploids) and to $\sqrt{1/v}$ in the second scenario (stationary aneuploids).
+Furthermore, the growth rate $\Delta_a<0$ that allows tolerant aneuploids to rescue the tumor is between $-u\lambda_a$ and  $-1/T^*$, which is proportional to $-\sqrt{v}$.
+Thus, increasing the mutation rate $v$ will decrease the tumor threshold size $N^*_a$, making evolutionary rescue more likely, but only until $T^*$ decerases to a point where  $\Delta_a =1/T^*$.
 
@@
 Large tumors are predicted to adapt to anti-cancer drugs through direct mutation.
-In contrast, smaller tumors are predicted to become resistant either directly by aneuploidy or by a resistance mutation occurring in aneuploid cells that serve as an evolutionary ``stepping stones'' (\Cref{rescue_prob}).
+In contrast, smaller tumors are predicted to become resistant either directly by aneuploidy or by a resistance mutation occurring in aneuploid cells that serve as evolutionary ``stepping stones'' (\Cref{rescue_prob}).
 Thus, therapies that increase the rate of aneuploidy in tumors to combat cancer may also promote drug resistance.

### 2024-07-31 EvolutionaryRescue1.tex
remove todo comments
abstract minor change

@@
 
-% MESSAGES ok
-% 1. aneuploidy increases the prob of rescue -- reduces the tumor threshold size for rescue
-% 2. aneuploidy changes the survival curve in a distinct way
-
-% FIGURES ok
-% Fig 1A: model graph chart
-% Fig 1B: Delta_w, Delta_a, Delta_m illustration
-
-% Fig 2: example simulations - w_t, a_t, m_t over time 
-% 	A u=0; B Delta_a << 0; C Delta_a ~ 0; D Delta_a >>0
-
-% Fig 3A: prob rescue vs N for various Delta_a; markup N*
-% Fig 3B: N* vs Delta_a
-% Fig 3C: N* vs u/v
-
-% Fig 4: survival curves
-% 	A u=0; B Delta_a << 0; C Delta_a ~ 0; D Delta_a >>0
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{abstract}
-Evolutionary rescue occurs when a population survives a sudden environmental change that initially causes the population to decline toward extinction.
+Evolutionary rescue happens when a population survives a sudden environmental change that initially causes the population to decline toward extinction.
 A prime example of evolutionary rescue is the ability of cancer to survive exposure to treatment. 
@@
 Here, $\lambda_s=0.1$, $\lambda_m=0.1$, $\mu_s=0.14$, $\mu_a=0.09$, $\mu_m=0.09$, $v=10^{-7}$ , $N=10^7$; (A) $u=0$; (B) $\lambda_a=0.065$, $u=10^{-2}$; (C) $\lambda_a=0.08999$, $u=10^{-2}$; (D) $\lambda_a=0.095$, $u=10^{-2}$.
-% TODO panel C run simulations for \lambda_a=0.08999 instead of 0.0899 so that it would give T\Delta_a =~ -0.555 to be in the stationary case (right now it is ~-5.5 which is the tolerant case) OK
 }
@@
 \label{rescue_prob_ane}
-% TODO what is the value of v_s? OK
 \end{figure}

### 2024-12-22 ResponseLetter.tex
Remus revision

\documentclass[11pt,a4paper]{article}
\usepackage{xcolor}
\usepackage{amssymb,amsmath,amsthm}
\usepackage{natbib}

\renewcommand{\d}{{\rm d}}
\newcommand{\e}{{\rm e}}
\begin{document}
Reviewer $\#1$ :

Reviewer Summary:

In this paper, the authors consider the role of aneuploidy in creating drug-resistant tumor cell lines in cancer. Specifically, they address how aneuploidy contributes to evolutionary rescue of a tumor cell population following treatment with tumor-cell killing drugs. They explore how different types of aneuploidy (stationary, tolerant and partially resistant aneuploid cell populations), time, mutation, rates of change, growth rates of drug resistant cells, pre-treatment aneuploid cell populations, and standing genetic variation, all might contribute, and importantly be exacerbated by aneuploidy, leading to the persistence of tumor cells following chemotherapy that result in a recurrence of tumor, even well past the point of extended remission. The authors use evolutionary models to identify the probability that a tumor could survive and proliferate under different genetic scenarios, rooting their parameters within ranges defined by previous studies, and identify several ways in which aneuploidy may serve a role, directly or in stepping-stone fashion, to facilitate evolutionary rescue of tumor cell populations, especially in smaller tumors often characteristic of primary tumors following resection and secondary tumors following metastasis.

Major Edits:

I have no major edits to recommend on this manuscript and I am comfortable with the authors considering the minor edits and making adjustments as they see fit.

Minor Edits:

1. You do an incredible job of walking the reader through this manuscript, making clear your definitions of important concepts, like evolutionary rescue, along the way. Two places where you didn't do that, but that could really push this paper over the top into perfection territory, are the following: 1) offering your reader an explanation, even if brief, of the multi-type branching method central to your analyses/models, and why it was the right selection for this endeavor among other methods or evolutionary modeling frameworks used in the past or available to you, and, 2) putting something real-world up against the parameter ranges you used to start the models, or just a brief statement about how they were established from the literature you drew upon (i.e., from experimental tests, patient data, other models, etc.). The latter is just in an effort to put something real to the numerical ranges that a reader can relate to before launching into the modeling.

\textcolor{blue}{Thank you for your comment. 
\begin{enumerate}
\item We have added a comment explaining that we use multi-type branching processes because we assume independence between clonal lineages starting from an initial population of $N$ sensitive cells.   change 53 
\item For the sensitive and mutant division rate the paper already states that this parameter is obtain in vivo by Rew and Wilson (2000). I have updated the second paragraph  of "Parameterization" with the fact that Bozic et al. (2013) use patient data in order to obtain the growth rate of sensitive cell. change 54
\end{enumerate}
}

2. Do a search for multi-type branching and make sure you consistently use the hyphen or not.

\textcolor{blue}{Thank you for your comment. We have made the change to use the hyphen in all mentions of "multi-type branching" such that it is consistent with textbooks such as Athreya $\&$ Ney (1972).  change 31}

3. You identify that at least in certain scenarios aneuploidy plays an important role in the evolutionary rescue of tumor cell populations following chemotherapy, at least with one type of drug, and maybe limited to small primary or secondary tumors. But, we never really know the extent that this scenario is applicable. Is it specific to this particular chemotherapy, small tumors, only certain types of cancer? In conjunction with this, there are places in the manuscript where you say things like ". . .for most parameter values aneuploidy will either play no role in evolutionary rescue or will be the main driver of adaptation without requiring any mutation (one step)." Both parts of the quoted sentence stand in opposition to one of the more interesting scenarios found in your modeling, which is that aneuploidy can buy a tumor cell population enough time to accumulate a mutation that leads to evolutionary rescue. Like in point 1, making clear the broader implications of your findings, or the limited scenarios in which they might be applicable, pushes this paper over the top into perfection territory because it roots your reader into something relatable, manageable, or applicable to patient care. If the answer is "I don't know how applicable this finding is or what the broader implications might be," you could then add something to the future research section that talks about how this might be addressed experimentally or theoretically. It's ok to not know for sure, but you shouldn't pass the opportunity to talk about it in limits and future research.

\textcolor{blue}{Thank you for your comment.  We have made an addition to "Future research" that out model can be used on a larger set of tissue-drug combinations to establish whether aneuploidy can play an important role in the evolution of drug resistance. change 55}

4. I would revisit some of your figures to make sure groups/categories are not colored in contradictory ways from one graph to the next (i.e., like "no aneuploidy" being one color on one graph and a different color in another graph or different categories ("no aneuploidy" vs. "aneuploid" vs. "stationary") using the same color in different figures).

\textcolor{blue}{Thank you for your comment.  We have changed Figure 6A to be consistent with the other figures. I have also changed the caption for Figure 6 to be consistent with the new coloring. change 56}

5. The results on detection time associated with tumor size were not intuitive to me on first read. You might want to clarify or rewrite those sentences to make it more clear. It read to me like something about this finding/scenario changed the baseline detection size required to know the tumor cell population was still present and viable.

\textcolor{blue}{Thank you for your comment. We have rewritten the sentence about the mean detection time to explain that for the small and intermediate tumors the mean detection time is approximately  equal to the mean recurrence time. change 48}

6. Table 1 caption doesn't define what the asterisk means.

\textcolor{blue}{Thank you for your comment. We have added the definition of asterisk in the table. change 1}

7. You never tell your reader what ODEs is. . .won't be obvious to everyone. I would just right it out on page 9 where I think is the first time you use it.

\textcolor{blue}{Thank you for your comment. We have added an explanation of the meaning of ODEs to the text in page 9. change 2}

8. Are the tests for standing genetic variation essentially genetic drift? Variation is present and in a tumor cell population wildly fluctuating in size due to treatment you just hit on a "right" variant stochastically when tumor size is small (less cells) that is resistant and it's driven to fixation? If so, using genetic drift in this section is rooting language and another way of describing alternative scenarios/hypotheses for tumor rebound.

\textcolor{blue}{Thank you for your comment. In the context we are discussing, standing genetic variation is generated by the balance between "mutation" (which increases the number of aneuploid cells) and selection (which reduces the number of aneuploid cells). The aneuploid cells are selected against in the pre-tratment regime but once the tumor is treated with anti-cancer drugs they have a fitness advantage over the sensitive cells. If the aneuploids are partially resistant they will grow to fixation if they escape stochastic extinction. However, if the they are stationary of tolerant they will outlast the sensitive cells and increase the porbability of evolutionary rescue.}

9. Do we know if it's reasonable to assume the division rates between cell types (i.e., "sensitive" vs. "mutant") are the same? Maybe ground that assumption into something we know, or just say we are not sure if that is reasonable or not. This relates to point 1 and rooting the reader into these parameter values with something real or being honest about the assumption they are equal.

\textcolor{blue}{Thank you for your comment. It is reasonable to assume that the sensitive and mutant growth rates are similar given that we assume that the mutation makes the cancer cells immune to the chemoterapeutic drugs. change 40}

10. Mid page 7, "decreases" is mis-spelled and I would swap out "very well" with something more descriptive where you use it.

\textcolor{blue}{Thank you for your comment. We have made the change from "decerases" to "decreases". change 27}

Reviewer $\#$2 :

Summary

This is a paper about the likelihood and time for a tumour to evolve drug resistance, via direct mutation or indirectly through aneuploidy. The authors develop a multi-type branching process model to approximate these quantities analytically and to simulate the full behaviour. They also parameterize their approximations and simulations for melanoma, with adaptive trisomies for chromosome 2 and 6. The approximations provide intuition and match simulations well while demonstrating, for instance, that even tolerant (ie, subcritical) aneuploids -- which appears to be the case for melanoma -- can substantially facilitate drug resistance (fig 3) and extend recurrence times (fig 6).

I think this is a very nice application of evolutionary rescue theory. The presentation is clear and the neat approximations provide intuition. Parameterizing for melanoma demonstrates the utility of the theory.

My main concern (elaborated below), is that all of the theoretical results derived here have extremely close analogues in previous evolutionary rescue theory. This substantially reduces the novelty of this paper, in the sense that much of the intuition developed here has been given elsewhere. However, I do think it is useful to derive all of these approximations for a single model that can be parameterized with existing cancer data. What I would like to see, then, is the authors connect their results with existing rescue theory, which would allow them to clarify what is new here and also help develop a largely missing link between the evolutionary rescue and cancer literatures.

$\#$ Major comments

1. The paper is framed as a model about the evolution of drug resistance in tumours via direct mutation or indirectly through aneuploidy. However, the model is much more general than this, it is a model of 1- and 2-step evolutionary rescue that is then parameterized for cancer. Because of this, I would like to see the authors spend more time discussing what is already known about 1- and 2-step rescue, which should also elucidate why another model is needed rather than simply parameterizing a model that already exists. I do think the authors can justify a new model, but I would like to see this developed with a more insightful review of existing models (currently Intro mentions some models exist but doesn't explain why they aren't sufficient) and a comparison of results derived here to those derived previously (currently no rescue papers cited after Intro). In addition to clarifying what is novel here, these comparisons may help connect the currently disparate fields of evolutionary rescue and cancer (see fig 1 in Alexander et al 2014 Evol. Appl.).

In case it helps, my understanding of multi-step rescue theory is that there was an early model for the emergence of epidemics via multiple mutations (Antia et al 2003 Nature, and some follow ups like Yates et al 2006 PRSB and Alexander $\&$ Day 2010 Interface), a very general treatment of emergence and rescue via multiple mutations (Iwasa et al 2004 JTB), a 2-locus 2-allele rescue model (Uecker $\&$ Hermisson 2016 Genetics), and multi-step rescue models on a fitness landscape (Martin et al 2013, Osmond et al 2020), all of which also use branching processes. Many of the results in this paper have been derived in the above studies. For example, eqn 3 is essentially the 1-step rescue probability from Orr $\&$ Unckless 2008 (Am Nat, also their eqn 3) and eqn 4 is essentially the emergence probability from Iwasa et al 2004 (their eqn 3) multiplied by the number of first step mutations that occur during the wildtypes decline. Even more similarly, Eqn 4 is a special case of the results in Osmond et al (2020), who also used the reasoning from fitness valley crossing (Weissman et al 2009) to derive approximations for the probability of 2-step rescue, with the added complication of a distribution of mutation effects. Setting the distribution of fitness effects to a dirac and assuming fast wildtype decline reduces the first two lines of their eqn 8 to the first two lines of eqn 4 (the final line of eqn 4 being the well known case of 1-step rescue). The de novo vs standing variance result (eqn 6) was pointed out in Orr $\&$ Unckless 2008 (right side of page 163) and given in Martin et al 2013 (eqn 3.8). The rescue and recurrence times (eqns 8 and 9) are perhaps the most novel results derived here, but have some close similarities with times derived in Orr $\&$ Unckless 2014 (see eqn 18 and text S2 for rescue time and eqn 23 for recurrence time).

2. I have two general modelling-choice questions whose answers may substantially influence the results. 1) How heritable is a particular aneuploidy? 2) Are aneuploids with a resistance mutation really equivalent to diploids with a resistance mutation? Can the modeling choices made be empirically justified, or can the model (or results) be generalized to include these complications?

$\#$ Minor comments

- line numbers would have been helpful

\textcolor{blue}{Thank you for your comment. We have added line numbers.}

- abstract: first sentence could give more precise definition of evolutionary rescue

\textcolor{blue}{Thank you for your commnet. We have changed the definition of evolutionary rescue such that it highlights the fact that the population must adapt to the environmental change. change 32}

- abstract: 'fitter'$\rightarrow$'more fit'

\textcolor{blue}{Thank you for your comment. We have made the requested change. change 3}

- intro: 'the number of chromosomes and chromosome arms copies'$\rightarrow$'the number of chromosomes and chromosome arm copies'

\textcolor{blue}{Thank you for your comment. We have made the requested change. change 4}

- intro: while there are some similarities, I do not think that evolutionary rescue is mathematically equivalent to the problem of crossing fitness valleys. If that was the case, you wouldn't have to create a new model here and could instead parameterize the model of Weissman et al 2009 for cancer. One difference, for example, is that with fitness valley crossing the population doesn't go extinct and so it is a question about when the valley will be crossed, while in rescue the wildtype is declining to extinction and so the question is largely if rescue will occur. I do agree that the case of two step rescue is similar to fitness valley crossing, but that is a much weaker statement.

\textcolor{blue}{Thank you for your comment. We have changed the text such that it states that evolutionary rescue in two steps is similar to the problem of crossing fitness valley. change 6}

- intro: there is a relatively arbitrary list of rescue papers given at the end of the paragraph, which could be replaced (or supplemented) with synthetic reviews (eg, Alexander et al 2014 Evol Appl, Bell 2017 Annual Reviews)

\textcolor{blue}{Thank you for your comment. We have added the recommend synthetic reviews to the list of rescue papers cited. change 45}

- intro: 'there are three ways for a population to survive ...' - I think this statement would be better placed at the beginning of the paragraph, after which you focus on genetic adaptation

\textcolor{blue}{Thank you for your comment. We have  placed the statement at the beginning of the paragraph. change 28}

- intro: 'most models focus on the probability that at most one mutation rescues the population' - I disagree with this statement. I think the focus is rather on the probability that at least one mutation rescues the population. That is the logic behind the $1-\exp(-Np)$ form: $\exp(-Np)$ is the probability that no mutations rescue the population. In addition to Wilson et al 2017, Osmond $\&$ Coop 2020 (Genetics) also calculate the probability rescue is soft and the expected number of rescuing mutations in a variety of rescue scenarios.

\textcolor{blue}{Thank you for your comment. We have changed the statement to just explain the concepts of soft and hard selective sweeps. Additionaly, we have added a reference to Osmond $\&$ Coop 2020 (Genetics). change 41}

- methods: 'The aneuploid division rate is selected...' - I found the grammar odd here.

\textcolor{blue}{Thank you for your comment. We have changed the statement to be 'The aneuploid division rate is chosen...'.   change 36}

- methods: 'We estimate c=...' - It sounds like you will estimate c from the observed fraction of aneuploid cells but that is not what you describe below.

\textcolor{blue}{Thank you for your comment. We have changed the paragraph to make it clear that we obtaint the fitness cost from the experimental data and then use the formula $f=\frac{\tilde{u}\lambda_s}{c}$ to obtain $f$. change 5}

- methods: is there a justification for not including density-dependence for the sensitives as well? This would hasten their initial decline, which would compound with the reduced initial establishment probability of the mutants. It seems odd they can exert competition on the mutants but don't themselves feel any competition.

\textcolor{blue}{Thank you for your comment. The justification for not including density dependence for sensitive cells is because of the fact that they are under the influence of a powerful stress which causes the population to decline very fast. As a result, density dependent effects will not have an effect on the initial establishment probability of mutants.  As shown in Figure S1, the presence of a carrying capacity does not effect the probability of evolutionary rescue.}

- methods: I got a 404 error for the source code link: https://github.com/yoavram-lab/EvolutionaryRescue

- results: there doesn't appear to be an eqn 1

\textcolor{blue}{Thank you for your comment. We have made changes such that equation 1 appears. change 29}

- results: eqn 3 is an alternate form of the classic rescue probability, eg, $1/N_m^*=u/r*2(s-r)$ in Orr $\&$ Unckless 2008. I think it would be helpful to mention this.

\textcolor{blue}{Thank you for your comment. We have mentioned that equation 2 (the numbers are changes in the updated version) is an alternative form of the formula from Orr $\&$ Unckless 2008. change 33}

- results: eqn 4 is an approximation (fast wildtype decline) of a special case (dirac distribution of mutational effects) in Osmond et al 2020, who used the exact same reasoning from Weissman et al 2009, and is also very similar to eqn 3 in Iwasa et al 2004. I think it would be helpful to mention this.

\textcolor{blue}{Thank you for your comment. We have added a paragraph which explains the similarity of equation 4 (now equation 3) to previous results. change 57}

- results: '$T^*$ decerases'$\rightarrow$'$T^*$ decreases'

\textcolor{blue}{Thank you for your comment. We have made the change from "decerases" to "decreases". change 27}

- results: I thought it was interesting that the second scenario in eqn 5 was the geometric mean of the other two scenarios. Is there some intution for why this is so?

\textcolor{blue}{Thank you for your comment. We have not been able to identify any intuition as to why the second scenario is the geometric mean of the other two scenarios.?}

- results: I thought it might be helpful to give a little more intution about the effect of simultaneously increasing the aneuploid birth and death rates (paragraph starting below eqn 5). The reason why this leads to a decrease in $T^*$ and pushes the system towards the second scenario is that the mutation rate has increased, meaning a lineage doesn't have to persist for as long to almost surely lead to a successful rescue mutant, while the expected persistence time hasn't changed, making it more likely that a rare long-lived aneuploid lineage will lead to rescue.

\textcolor{blue}{Thank you for your comment. We have added the recommended comment about the fact that the reason why $T^*$ decreases, if we increase the aneuploid division and death rates such that $\Delta_a$ remains constant, is because of the fact that more mutants are generated and a lineage doesn't have to persist as long in order to generate a succesful mutant. Change 51 }

- results: I don't get the logic at the end of this paragraph, that your results are consistent with aneuploidy conferring resistance by decreasing the division rate. My understanding is that your result says that, holding aneuploid growth rate constant, increasing aneuploid division rate helps until you reach scenario 2 (effectively stationary/critical aneuploids) but beyond that doesn't help. I don't see how reducing division rate helps rescue.

\textcolor{blue}{Thank you for your comment. In the scenario discribed decreasing the division rate also leads to a decrease in the death rate as the drugs affect the cancer cells during cellular division. We have expanded on the explanation of this scenario in the paper. Change 52}

- results: if $N_a^*$ is similar to the detection threshold, then you predict that any tumour detected is very likely (eg, $>50\%$) to evolve resistance? Does this fit with data on resistance rates?

\textcolor{blue}{Thank you for your comment.  We believe that our model does fit the data on resistance rates given that \citep{handa2022long} state that 68$\%$ of melanoma patients treated with vemurafenib experienced relapse after initial response. change 58}

- results: if aneuploidy may increase the mutation rate and this is easy to implement in the model, why not do it from the start, and reduce to the simpler version $v_a=v$ when needed?

\textcolor{blue}{Thank you for your comment. We started with the simpler model and thought that the more advanced model should be included in Appendix H. Additionally, we do not have biologically realistic values for the aneuploidy induced mutation rate and, as a result, we decided it was better to start with the model for which we have biologically realistic parameters. }

- results: 'cancer cells(' $\rightarrow$ 'cancer cells ('

\textcolor{blue}{Thank you for your comment. We have made the change from 'cancer cells(' to 'cancer cells ('. change 25}

- results: eqn 6 has been pointed out by Orr $\&$ Unckless 2008 and Martin et al 2013

\textcolor{blue}{Thank you for your comment. We have mentioned that the results has been obtained previously by Orr $\&$ Unckless 2008 and Martin et al 2013. change 42}

- results: much of the discussion of the results (eg below eqn 7) is about the probability of rescue, but the equations all give the inverse of this, critical tumour sizes. I find it a little annoying to keep converting between them, and presenting the probabilities of rescue would also help compare with previous papers. Either way is fine but perhaps something to think about.

\textcolor{blue}{Thank you for your comment. We thought about this and we think that tumor size is critical for the interpretation of our model}

- results: I think the rescue times given in the main text (eqn 8) could be explained a little more. For small tumours, this is the 2-step version of the 1-step rescue time given in Orr $\&$ Unckless 2014 (the expectation of their eqn 18), $-1/r$. I.e., the faster the population declines the earlier rescue occurs (conditioned on it occurring), because that is when most mutations happen. For larger tumours this is just a deterministic constant population size result, where the successful mutation is produced at a constant rate.

\textcolor{blue}{Thank you for your comment. We have added comments that better explain the mean rescue time in the case of small and large tumors. change 46}

%- results: why doesn't standing genetic variance affect the mean rescue time? This might have something to do with what I just mentioned (successful mutants tend to arise early) and more clues might be found in Orr $\&$ Unckless 2014, who also compare times with and without standing genetic variance.

- results: 'aneuploidy promotes evolutionary rescue after $1/\Delta_a$' $\rightarrow$ 'aneuploidy promotes evolutionary rescue after $1/\Delta_s$'

\textcolor{blue}{Thank you for your comment. We have made the change from $1/\Delta_a$ to $1/\Delta_s$. change 37}

- results: in eqn 9 the variable $p_m$ is used, but it is not defined in the main text

\textcolor{blue}{Thank you for your comment. We have added the definition of $p_m$ below equation 9. change 26}

- results: the final term of the recurrence time for small tumours is equivalent to the return time of Orr $\&$ Unckless 2014 with standing genetic variation (eqn 16), but different from their return time with new mutation (eqn 23). Why?

\textcolor{blue}{Thank you for your comment. The return time with new mutation (eqn 23) from Orr $\&$ Unckless 2014 can be written as:
\begin{align*}
t_{return}&\sim\frac{1}{s-r}\log\left[\frac{2N_0s(s-r)}{r}\right]\\
&\sim \frac{1}{s-r}\log[2N_0(s-r)]+\frac{1}{s-r}\log\left(\frac{s}{r}\right)
\end{align*}
Additionally, when $|s-r|<r$ we have:
\begin{equation*}
\log s \approx \log r +\frac{s-r}{r}
\end{equation*}
which can be rewritten as:
\begin{align*}
\frac{1}{s-r}\log\left(\frac{s}{r}\right)\approx \frac{1}{r}
\end{align*}
As a result, the return time with new mutation can be written as:
\begin{equation*}
t_{return}\sim \frac{1}{r}+\frac{1}{s-r}\log[2N_0(s-r)]
\end{equation*}
which is the one-step equivalent to our recurrence time for small tumors.
} 
- results: you mention that drugs increasing the death rate of sensitive cells delays recurrence, but given that mutations happen during division, it might be worth mentioning that drugs that decrease the division rate of sensitive cells by the same amount delay recurrence even more (holding mutant division rate constant - which is maybe why you didn't mention this for your parameterization, but seems like a worthwhile general point?).

\textcolor{blue}{Thank you for your comment. For large tumors, decreasing the division rate of sensitive cells will lead to a decrease in the cancer recurrence time when conditioned on evolutionary rescue. We have added this comment to the text. change 49}

- results: why doesn't standing genetic variance affect the mean rescue time? This might have something to do with what I just mentioned (successful mutants tend to arise early) and more clues might be found in Orr $\&$ Unckless 2014, who also compare times with and without standing genetic variance.

\textcolor{blue}{Thank you for your comment. Standing genetic variation does not have a significant impact on mean rescue time because the percentage of cell which have the beneficial aneuploidy is only $0.14\%$.}

- discussion: I find the discussion of the two advantages of aneuploidy for rescue a little confusing. Yes, aneuploidy delays population extinction and yes it increases the population size at a given time, but the important fact is that it increases the cumulative number of mutants that are created, which is the expression you give at the end of the sentence. Perhaps it would help to insert an additional sentence before the expression, like "Together, this increases the cumulative number of mutants that arise."

\textcolor{blue}{Thank you for your comment. We have added a sentence before the expression which states that aneuploidy increases the cumulant number of mutants that arise. change 39}

- discussion: the last sentence mentions that some cancer therapies are designed to increase aneuploidy, which felt like a new and important piece of information for me. Some discussion of this in the intro could be helpful as motivation for this work.

\textcolor{blue}{Thank you for your comment. In the last sentence of "Discussion" we state that certain therapies increase the rate of aneuploidy which has been shown by Lee et al. (2016) which is referenced in the "Introduction" section as a motivation for our research.}

- discussion: there are no citations of evolutionary rescue theory in the discussion, despite most of the results already existing or having very close parallels in the rescue literature. I think it would be nice to build more connection between rescue theory and cancer here.

\textcolor{blue}{Daniel ?.}

- table 1: what are the asterisks in the name column?

\textcolor{blue}{Thank you for your comment. We have added the definition of asterisk in the table. change 1}

- fig 2: when aneuploids are tolerant rescue can also happen indirectly (caption incorrect for panel B)

\textcolor{blue}{Thank you for your comment. We have made the change such that it says "When aneuploid cells are tolerant ($\Delta_a<0$) evolutionary rescue can occur indirectly, but direct mutation is the most likely path for evolutionary rescue". change 24}

- fig 4: give eqn $\#$ of exact formula (this goes for all formulas used in all figures)

\textcolor{blue}{Thank you for your comment. We added the references to the equation numbers for the formulas of $N_a^*$ and $N_m^*$. change 43}

- fig 4: $\lambda_m$ value given twice

\textcolor{blue}{Thank you for your comment. We have made the change from $\lambda_m$ to $\lambda_a$ in the first instance of $\lambda_m$. change 18}

- appendix A: 'decline are significant'$\rightarrow$'decline is significant'

\textcolor{blue}{Thank you for your comment. We have made the recommended change. change 19}

- appendix A: in eqn A10, the $p_a$ and $p_m$ terms in the denominator seem to disappear in the second line (they will disappear in the third line anyway, given small mutation, but I think they should technically be in the second line)

\textcolor{blue}{Thank you for your comment. We have made the recommended change. change 20}

- appendix A: and in the third line of eqn A10 the $p_m$ in the numerator disappears (again, this doesn't matter for the result, eqn A11, but is an error and is required for the approximation to get from A10 to A11)

\textcolor{blue}{Thank you for your comment. We have made the recommended change. change 21}

- appendix C: $w_t\rightarrow s_t$

\textcolor{blue}{Thank you for your comment. We have made the requested change from $w_t$ to $s_t$. change 7}

- appendix C: I think the approximation in eqn C1 makes more sense by taking the Taylor series of the exponential term with respect to $\Delta_st$, rather than with respect to $t$ alone. For instance, you write $t + O(t^2)$, but the higher order terms are only smaller than t when $t\ll1$.

\textcolor{blue}{Thank you for your comment. We have changed $\mathcal{O}\left(t^2\right)$ to $\mathcal{O}\left(\Delta_st^2\right)$. change 22}

- appendix C: grammar mistakes in sentence below eqn C3

\textcolor{blue}{Thank you for your comment. We have corrected the grammatical mistakes. change 8}

- appendix C: I think it would be helpful to define $N_m^*$ and $N_a^*$ again in the appendix before using them

\textcolor{blue}{Thank you for your comment. We have added the definitions of $N_m^*$ and $N_a^*$ in Appendix C. change 38}

- appendix C: I would like to hear the justification for the initial expression of $\tau_a$ (integral of exponential) when $N>N_m^*$

\textcolor{blue}{Thank you for your comment. The justification is the following: 
\begin{align*}
\tau_a&=\int_0^\infty tf(t)dt=\int_0^\infty t\frac{d}{dt}F(t)dt=-\int_0^\infty t\frac{d}{dt}S(t)dt\\
&=-\left[tS(t)\right]_0^\infty+\int_0^\infty S(t)dt\\
&=\int_0^\infty S(t)dt,
\end{align*}
where f(t), F(t) and S(t) are the probability density function, cumulative distribution function and survival function.
In our case, $S(t)=\exp\left(-\int_0^t R_a(z)dz\right)$. Additionally, for $N\gg N_m^*$ we have $1-\e^{-Np_s}\approx1$. I have corrected the following mistake:
\begin{equation*}
\tau_a=\int_0^\infty \e^{-\int_0^\tau R_a(t)\d t}\d\tau
\end{equation*}
instead of:
\begin{equation*}
\tau_a=\int_0^\infty \e^{-R_a(\tau)}\d\tau.
\end{equation*}
change 23
}

- appendix C: similarly, I think it would be nice to hear some justification for the approximations in eqn C9, ie, can ignore direct rescue and rescue via aneuploidy very unlikely

\textcolor{blue}{Thank you for your comment. We know that
\begin{align*}
\tau_a=\int_0^\infty t\left(v\lambda_ap_ma_t+v\lambda_sp_ms_t\right)\frac{\exp\left[-\frac{uv\lambda_s\lambda_aNp_m}{\Delta_s-\Delta_a}\left(\frac{\e^{\Delta_s t}-1}{\Delta_s}-\frac{\e^{\Delta_a t}-1}{\Delta_a}\right)-v\lambda_sNp_m\frac{\e^{\Delta_s t}-1}{\Delta_s}\right] }{1-\e^{-Np_s}}\d t.
\end{align*}
Additionally, if $N\ll N_a^*$ then the probability distribution of evolutionary rescue time is skewed toward small values (i.e. evolutionary rescue occurs conditioned on it happening). As a result, we have: 
\begin{align*}
\frac{\e^{\Delta_s t}-1}{\Delta_s}&=\frac{1+\Delta_s t+\mathcal{O}\left(\left(\Delta_st\right)^2\right)-1}{\Delta_s}=t+\mathcal{O}\left(\Delta_st^2\right)\\
\frac{\e^{\Delta_a t}-1}{\Delta_a}&=\frac{1+\Delta_a t+\mathcal{O}\left(\left(\Delta_at\right)^2\right)-1}{\Delta_a}=t+\mathcal{O}\left(\Delta_at^2\right)
\end{align*}
We observe that the first term in the exponential can be approximated to be zero and the second term as -$v\lambda_sNp_mt$ where $t\ll1$. As a result, we can approximate:
\begin{equation*}
\exp\left[-\frac{uv\lambda_s\lambda_aNp_m}{\Delta_s-\Delta_a}\left(\frac{\e^{\Delta_s t}-1}{\Delta_s}-\frac{\e^{\Delta_a t}-1}{\Delta_a}\right)-v\lambda_sNp_m\frac{\e^{\Delta_s t}-1}{\Delta_s}\right] \sim 1
\end{equation*}
Additionally, since $N\ll N_a^*\ll N_m^*$, evolutionary rescue is more likely to occur through the trajectory $sensitive \rightarrow aneuploid \rightarrow mutant$. As a result,
\begin{equation*}
\tau_a\sim\int_0^\infty \frac{tv\lambda_ap_ma_t}{1-\e^{-Np_s}}\d t.
\end{equation*}
}

- appendix D: the fact that lineages conditioned to survive drift reach N earlier is nicely discussed in Orr $\&$ Unckless 2014, who connect it back to classic work on selective sweeps by Maynard Smith $\&$ Haigh 1974

\textcolor{blue}{Thank you for your comment. We have added citations for Orr $\&$ Unckless 2014 and Maynard Smith $\&$ Haigh 1974. change 35}

- appendix D: 'it accurate' $\rightarrow$ it is accurate'

\textcolor{blue}{Thank you for your comment. We have made the change from 'it accurate' to 'it is accurate'. change 9}

- appendix F: 'Figure Figure' $\rightarrow$ 'Figure'

\textcolor{blue}{Thank you for your comment. We have made the recommended change. change 10}

- appendix F: 'our approximation are'$\rightarrow$'our approximations are'

\textcolor{blue}{Thank you for your comment. We have made the recommended change. change 11}

- appendix F: 'From eq. (F2) we obtain'$\rightarrow$'From eq. (F1) we obtain'

\textcolor{blue}{Thank you for your comment. We have made the recommended change. change 12}

- fig S2: 'the leads'$\rightarrow$ 'that leads', 'is in agreement'$\rightarrow$'are in agreement', $\lambda_m$ value assigned twice

\textcolor{blue}{Thank you for your comment. We have made the recommended change. change 13}

- fig S3: 'number of mutation'$\rightarrow$'number of mutations', $\lambda_m$ value assigned twice

\textcolor{blue}{Thank you for your comment. We have made the recommended change. change 14}

- fig S5: $\mu_a$ value assigned twice

\textcolor{blue}{Thank you for your comment. We have made the recommended change. change 15}

- fig S6: 'the leads'$\rightarrow$'that leads', 'lines, right'$\rightarrow$'lines,', 'is in agreement'$\rightarrow$'are in agreement', 'lines represents'$\rightarrow$'lines represent', $\lambda_m$ value assigned twice

\textcolor{blue}{Thank you for your comment. We have made the recommended change. change 16}

- fig S9: 'the leads'$\rightarrow$'that leads', $\lambda_m$ value assigned twice

\textcolor{blue}{Thank you for your comment. We have made the recommended change. change 17}

Reviewer $\#3$ :

Summary

Stana et al. study the effects of aneuploidy (mutations that give rise to atypical karyotypes) on the survival of a tumor- a population of malignant cells- in presence of an intervention, here, an anti-cancer drug. While the role of aneuploidy in development of drug-resistance has been acknowledged in the field of evolutionary oncology, the precise mechanism by which it incurs drug-resistance is not clear.

This manuscript is about understanding the evolutionary role of aneuploidy in drug-resistance in cancer. The authors model tumor evolutionary dynamics under treatment as a multi-type branching process, where the types correspond to three genotypes: sensitive to therapy, mutant (resistant), and aneuploid. The key assumption is that aneuploid cells are less-sensitive than sensitive cells to therapy. They parameterized their model by the pre/post treatment cost of aneuploidy, mutation rates and birth/death rates for each genotype.

Stana et al. derive equations for relevant evolutionary quantities under multiple parameter regimes. The evolutionary quantities include the probability of evolutionary rescue, and time to recurrence. They further investigate the effects of standing variation, and density-dependent growth. A central finding is that aneuploidy can directly or indirectly result in evolutionary rescue. In the latter, it delays tumor extinction and provides a window of opportunity for mutations to rescue the population.

Overall, the paper proposes a rigorous evolutionary approach to understand the effects of aneuploidy in development of drug-resistance in cancer. The authors verify the accuracy of their derivations via extensive stochastic simulations. The paper is well written and the ideas are presented in a cohesive and rational manner. The accompanying python implementation appears sufficient (see below) to reproduce the results. The problem is interesting and subject of active research in evolutionary oncology, specially development of drug-resistance subpopulations (PMID: 39223250) and new avenues of targeting aneuploid cells (PMID: 38992122).

Major points

- The importance of time-resolved measurements of clonal dynamics have been recognized and more such datasets are becoming available (e.g., PMID: 34163070, PMID: 38987605). The simulation studies in this manuscript use a parameter regime largely based on a melanoma cell-line (A375) and treatment with a specific BRAF-inhibitor (vemurafenib). The relevance of the paper would be improved if consistent results can be shown in other datasets, including the citations above.

Minor points

- An exciting finding of the paper is that aneuploidy can act as an evolutionary stepping stone (Figure 4). However, in the discussion, you point out that only in a small parameter regime this would be the case, and in most cases, aneuploidy either directly results in evolutionary rescue or does not play a role. If this were the case, wouldn't you expect that a majority of cancer cells in a tumor to either have clonal/truncal aneuploidies or have no clonal aneuploidies at all? Given the prevalence of aneuploidies in patient cohorts, how can we interpret the restricted parameter regime that allows it?

\textcolor{blue}{Thank you for your comment. In our model, we are interested in a specific type of aneuploidy which increases the fitness of cancer cells in an environment affected by anti-cancer drugs. As a consequence, we catalogued all other forms of aneuploids as sensitive.}

- Code Base
The python scripts used for simulation studies are provided in the supplementary package, however the github link is not accessible.
It would improve the reproducibility of the manuscript if the code to reproduce individual figures is provided.
Documenting the code as well as adding a short tutorial will improve the reusability of the scripts.

- In your model, mutations appear to only confer resistance, however, most mutations are deleterious. How easy is it to incorporate the effects of deleterious mutations in your model? This should be addressed in the discussion.

\textcolor{blue}{Thank you for your comment. We consider the fitness of the sensitive, aneuploid or mutant cells to be the average of each population. As a result, mutations that are deleterious are taken into consideration through the mean fitness of each population. I have added this comment in the "Parametrisation" section change 50}

- You report the exact parameters used in most figures. Sometimes it is difficult to follow why a specific value for a parameter was chosen. It would improve the readability if you included a table where the variation in the parameter used is noted and explained.

\textcolor{blue}{Thank you for your comment. The variation in our parameters in noted in Table 1 and is obtained from the papers which are cited there.}

Associate Editor Comments:

The paper sits in an odd place, as noted by reviewer $\#2$. Existing theory already explores evolutionary rescue by one and two step mutations, and it is hard to see the connection between the current theory and the existing theory. Which of the results are new? It is important that the reader knows what is novel. If similar results are derived in a different way, please explain the connection. When the quantity considered is different (e.g., timing of certain events), please alert the reader and maybe explain why these quantities are particularly relevant in the context of cancer. Reviewer $\#2$ has many helpful suggests along these lines.

Another oddity is that the theory developed applies much more broadly to evolutionary rescue scenarios, but the presentation makes it seems that the results only apply to cancer. Similarly, the initial facilitating mutation could be any type of mutation, not just aneuploidy.

A sufficient revision must, in my view, properly connect this work to the existing theoretical literature on evolutionary rescue (as noted by reviewer $\#2$) and to the broader empirical literature on cancer about multi-step drug resistance (as noted by reviewer $\#3$).

Minor comments
Introduction:
* 10 millions people die: add globally

\textcolor{blue}{Thank you for your comment. We have made the recommended change. change 30}

* mathematically equivalent: this is incorrect because fitness-valley crossing papers typically ignore population size dynamics, whereas evolutionary rescue is explicitly set in the context of a declining population size.

\textcolor{blue}{Thank you for your commnet. We have changed the text such that it states that evolutionary rescue in two steps is similar to the problem of crossing fitness valley. change 6}

* how multiple mutations contribute: several important papers on this subject are referenced later but not in this paragraph. Please revise, reviewing the existing theory and saying how the current theoretical work differs.

Methods:
* The methods only describe the simulations, and the results focus on analytical results that don't follow from the methods. I suggest expanding the "Methods" to discuss the general approach, model structure, and quantities derived (you can rename this as "Model") and then move to "Results".

\textcolor{blue}{Thank you for your commnet. The model is described in "Methods" under the subsection "Evolutionary model" and the analytic results follow directly from this model. Additionally, the quantities derived are stated in the paragraph before "Methods". }

* Delta would be better as "r" for the growth rate

\textcolor{blue}{Thank you for your commnet. The are about 300 mentions of $\Delta$ and we do not want to introduce any errors by changing $\Delta$ to $r$.}

* The use of the terms "tolerant" (for negative growth) and "resistant" (for positive growth) is not standard (e.g., see different for fungi used by Berman $\&$ Krysan 2020 Nature Reviews Microbiology). Maybe alert the reader that "tolerant" is used in different ways in the literature and is used here solely for notational convenience to refer to mutations that, on their own, do not allow growth? Or replace with "sub-critical", a term already used in this context?

\textcolor{blue}{Thank you for your commnet. We have added a short explanation for "tolerant" and that it has other meanings in the literature. change 47.}

* "antagonizing cell division" - The meaning of this isn't clear. Rephrase as "by extending the length of the cell cycle"?

\textcolor{blue}{Thank you for your comment. We have made the recommended change from "antagonizing cell division" to "by extending the length of the cell cycle". change 34}

Results:
* Figure 7 - Add a numerical solution to equation (D3)? This should make it clearer that the non-monotonicity of the black curve is an artefact of the simplifying assumptions in (D4) and not a problem with the approach. By the way, It wasn't clear to me what happened to the other large $\tau_a^r$ terms in the first portion of equation (D3) when approximating the equation into the form shown in (D4)? Could this be briefly clarified?

\textcolor{blue}{Thank you for your comment. We have changed Figure 7 by adding a green which represents the numerical solution of (D3).  In the first term of (D3) we have:
\begin{align*}
\frac{uv\lambda_a\lambda_s}{\Delta_s-\Delta_a}\left[\frac{\e^{\Delta_s\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_s-\Delta_m}-\frac{\e^{\Delta_a\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_a-\Delta_m}\right]\approx\frac{uv\lambda_a\lambda_s\e^{\Delta_m\tau_a^r}}{\left(\Delta_m-\Delta_s\right)\left(\Delta_m-\Delta_a\right)}
\end{align*}
which is significantly smaller then the second term of (D3):
\begin{align*}
v\lambda_s\frac{\e^{\Delta_s\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_s-\Delta_m}\approx\frac{v\lambda_s\e^{\Delta_m\tau_a^r}}{\left(\Delta_m-\Delta_a\right)}
\end{align*}
change 44}

Remus changes:

\textcolor{red}{$w_t\rightarrow s_t$. change r1}

\textcolor{red}{$t\rightarrow \tau$. change r2}

\textcolor{red}{
We have made the following changes: the rates of the inhomogenous Poisson process are:
\begin{align*}
r_1\left(t\right)&=v\lambda_ap_ma_{t},\\ 
r_2\left(t\right)&=v\lambda_sp_ms_{t},
\end{align*}
and the expected number of successful mutants created by direct mutation and via aneuploidy until time $t$ is given by:
\begin{align*}
M_a\left(t\right)&=v\lambda_ap_m\int_0^ta_{z} d z = \frac{uv\lambda_s\lambda_aNp_m}{\Delta_s-\Delta_a}\left(\frac{\e^{\Delta_st}-1}{\Delta_s}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right),\\ 
M_m\left(t\right)&=v\lambda_sp_m\int_0^ts_{z} d z = v\lambda_sNp_m\frac{\e^{\Delta_s t}-1}{\Delta_s}.
\end{align*}
change r3}

\textcolor{red}{We have made changes to Appendix C such that all the dummy variable used in integration are consistent. change r4}

\textcolor{red}{We have made changes to Figure S6, S7, S8, S9 by changing the subscripts of several variables from $w$ to $s$ . change r5}

\textcolor{red}{ We have made the following change: "If a fraction $f$ of the cancer cells are aneuploid when the drug is administered then the rates at which the rescue mutations are generated can be written as" $\rightarrow$
"If a fraction $f$ of the cancer cells are aneuploid when the drug is administered then the expected number of successful mutants generated until time $t$ is given by". change r6}

\bibliographystyle{plain}
\bibliography{evo2022}
\end{document}

### 2024-12-22 ResponseLetter.tex
went through R1 and R2 comments

@@
-\documentclass[11pt,a4paper]{article}
+\documentclass[12pt]{extarticle}
+\usepackage{geometry}
+\geometry{
+a4paper,
+total={170mm,257mm},
+left=20mm,
+top=20mm,
+headheight=12pt
+}
+
 \usepackage{xcolor}
@@
 \usepackage{natbib}
+\usepackage{commath}
+\usepackage{url}
 
@@
 \newcommand{\e}{{\rm e}}
+
+% Title page
+\title{
+	Response to reviewers\\ Evolutionary rescue by aneuploidy in tumors exposed to anti-cancer drugs 
+}
+
 \begin{document}
-Reviewer $\#1$ :
+\maketitle
 
-Reviewer Summary:
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\textbf{Reviewer $\#$1}
 
-In this paper, the authors consider the role of aneuploidy in creating drug-resistant tumor cell lines in cancer.
+%Reviewer Summary:
+%In this paper, the authors consider the role of aneuploidy in creating drug-resistant tumor cell lines in cancer.
 Specifically, they address how aneuploidy contributes to evolutionary rescue of a tumor cell population following treatment with tumor-cell killing drugs.
@@
 
-Major Edits:
+%Major Edits:
 
-I have no major edits to recommend on this manuscript and I am comfortable with the authors considering the minor edits and making adjustments as they see fit.
+%I have no major edits to recommend on this manuscript and I am comfortable with the authors considering the minor edits and making adjustments as they see fit.
 
@@
 
-1.
-You do an incredible job of walking the reader through this manuscript, making clear your definitions of important concepts, like evolutionary rescue, along the way.
-Two places where you didn't do that, but that could really push this paper over the top into perfection territory, are the following: 1) offering your reader an explanation, even if brief, of the multi-type branching method central to your analyses/models, and why it was the right selection for this endeavor among other methods or evolutionary modeling frameworks used in the past or available to you, and, 2) putting something real-world up against the parameter ranges you used to start the models, or just a brief statement about how they were established from the literature you drew upon (i.e., from experimental tests, patient data, other models, etc.).
+1. You do an incredible job of walking the reader through this manuscript, making clear your definitions of important concepts, like evolutionary rescue, along the way.
+
+\textcolor{blue}{Thank you for your kind feedback! We are glad our explanations were clear and helpful.}
+
+Two places where you didn't do that, but that could really push this paper over the top into perfection territory, are the following: 1) offering your reader an explanation, even if brief, of the multi-type branching method central to your analyses/models, and why it was the right selection for this endeavor among other methods or evolutionary modeling frameworks used in the past or available to you,
+
+\textcolor{blue}{ % change 53 
+We have added a comment in line XXX: "We assume independence between clonal lineages starting from an initial population of $N$ sensitive cells (we check the effect of density-dependent growth on our results below) and therefore, we use multi-type branching processes to model the dynamics of the cancer cell population.
+Multi-type branching process models the growth and evolution of populations with distinct types, capturing dynamics like mutation and selection.
+In evolutionary rescue, it predicts the probability of population survival by tracking adaptive mutations that emerge and spread under environmental stress."
+}
+
+2) putting something real-world up against the parameter ranges you used to start the models, or just a brief statement about how they were established from the literature you drew upon (i.e., from experimental tests, patient data, other models, etc.).
 The latter is just in an effort to put something real to the numerical ranges that a reader can relate to before launching into the modeling.
 
-\textcolor{blue}{Thank you for your comment. 
-\begin{enumerate}
-\item We have added a comment explaining that we use multi-type branching processes because we assume independence between clonal lineages starting from an initial population of $N$ sensitive cells.
-change 53
-\item For the sensitive and mutant division rate the paper already states that this parameter is obtain in vivo by Rew and Wilson (2000).
-I have updated the second paragraph  of "Parameterization" with the fact that Bozic et al.
-(2013) use patient data in order to obtain the growth rate of sensitive cell.
-change 54
-\end{enumerate}
+\textcolor{blue}{ % change 54
+We have updated the "Parameterization" section: "To parametrize the simulations, we assume that the cells under consideration are melanoma cells and rely on Rew and Wilson (2000) and Bozic et al.
+(2013) for the division and death rates, respectively.
+Rew and Wilson (2000)  report \emph{in vivo} measurements of the potential doubling times (the waiting time for the number of cells in the tumor to double, disregarding cell death) for a large set of cancer types.
+The division rate is obtained as $\lambda=\log{2} / T \approx 0.1$ per day.
+We select this to be the division rate for sensitive and mutant cells.
+This is a reasonable assumption given that we assume that the mutation makes the cancer cells immune to the chemotherapeutic drugs.
+Bozic et al.
+(2013) report the growth rate $\Delta_s$ for sensitive melanoma cancer cells, from patient data, from which they deduce the death rate $0.11 \le \mu_s \le 0.17$.
+We use  $\mu_s=0.14$ per day.
+Additionally, they observed the growth rate of cancer cells before treatment to be 0.01, which we use as the growth rate of mutant cells, which are resistant to the drug.
+Thus, we use $\mu_m=0.1-0.01=0.09$ per day as the death rate for mutant cells."
 }
@@
 
-\textcolor{blue}{Thank you for your comment.
+\textcolor{blue}{%change 31
 We have made the change to use the hyphen in all mentions of "multi-type branching" such that it is consistent with textbooks such as Athreya $\&$ Ney (1972).
-change 31}
+}
 
@@
 Is it specific to this particular chemotherapy, small tumors, only certain types of cancer?
-In conjunction with this, there are places in the manuscript where you say things like ".
-.
-.for most parameter values aneuploidy will either play no role in evolutionary rescue or will be the main driver of adaptation without requiring any mutation (one step)." Both parts of the quoted sentence stand in opposition to one of the more interesting scenarios found in your modeling, which is that aneuploidy can buy a tumor cell population enough time to accumulate a mutation that leads to evolutionary rescue.
+In conjunction with this, there are places in the manuscript where you say things like "for most parameter values aneuploidy will either play no role in evolutionary rescue or will be the main driver of adaptation without requiring any mutation (one step)." Both parts of the quoted sentence stand in opposition to one of the more interesting scenarios found in your modeling, which is that aneuploidy can buy a tumor cell population enough time to accumulate a mutation that leads to evolutionary rescue.
 Like in point 1, making clear the broader implications of your findings, or the limited scenarios in which they might be applicable, pushes this paper over the top into perfection territory because it roots your reader into something relatable, manageable, or applicable to patient care.
@@
 
-\textcolor{blue}{Thank you for your comment.
-We have made an addition to "Future research" that out model can be used on a larger set of tissue-drug combinations to establish whether aneuploidy can play an important role in the evolution of drug resistance.
-change 55}
+\textcolor{red}{
+TODO
+}
 
@@
 
-\textcolor{blue}{Thank you for your comment.
-We have changed Figure 6A to be consistent with the other figures.
-I have also changed the caption for Figure 6 to be consistent with the new coloring.
-change 56}
+\textcolor{blue}{ %change 56
+We have changed Figure 6A to be consistent with the other figures and changed the caption for Figure 6 to be consistent with the new coloring.
+}
 
@@
 
-\textcolor{blue}{Thank you for your comment.
-We have rewritten the sentence about the mean detection time to explain that for the small and intermediate tumors the mean detection time is approximately  equal to the mean recurrence time.
-change 48}
+\textcolor{blue}{%change 48
+We have rewritten the sentence about the mean detection time (line XXX): "We find that for small and intermediate-sized tumors the mean detection time is approximately equal to the mean recurrence time (i.e., $\tau_a^r\approx\tau_a^{M}$ for $N<N_m^*$).
+"
+}
 
@@
 
-\textcolor{blue}{Thank you for your comment. We have added the definition of asterisk in the table. change 1}
+\textcolor{blue}{ %change 1
+We have added the definition of asterisk in the table.
+}
 
-7. You never tell your reader what ODEs is. . .won't be obvious to everyone. I would just right it out on page 9 where I think is the first time you use it.
+7. You never tell your reader what ODEs is... won't be obvious to everyone. I would just right it out on page 9 where I think is the first time you use it.
 
-\textcolor{blue}{Thank you for your comment. We have added an explanation of the meaning of ODEs to the text in page 9. change 2}
+\textcolor{blue}{%change 2
+We have added an explanation of ODE (line XXX): "In this case the dynamics of the number of mutant cells is deterministic and can therefore be modeled by a system of ordinary differential equations (ODE), which describe how the number of mutant cells changes over time by its time derivative (eq.~D2)."
+}
 
@@
 
-\textcolor{blue}{Thank you for your comment.
-In the context we are discussing, standing genetic variation is generated by the balance between "mutation" (which increases the number of aneuploid cells) and selection (which reduces the number of aneuploid cells).
-The aneuploid cells are selected against in the pre-tratment regime but once the tumor is treated with anti-cancer drugs they have a fitness advantage over the sensitive cells.
-If the aneuploids are partially resistant they will grow to fixation if they escape stochastic extinction.
-However, if the they are stationary of tolerant they will outlast the sensitive cells and increase the porbability of evolutionary rescue.}
+\textcolor{blue}{In our model, where aneuploidy is deleterious prior to drug treatment, standing variation is generated mis-segragation and purged by selection.
+Since aneuploids are carry a considerable fitness cost, random drift is not expected to effect standing genetic variation.
+Standing variation actually increases with the population size, whereas the effect of drift decreases with the population size.
+We now explain this in line XXX: "We assume that $c$ is not too small (Table 1), so that standing variation is generated by a balance between the generation of anuploids and selection against them, without the effect of genetic drift."
+}
 
@@
 
-\textcolor{blue}{Thank you for your comment.
+\textcolor{red}{ % change 40
 It is reasonable to assume that the sensitive and mutant growth rates are similar given that we assume that the mutation makes the cancer cells immune to the chemoterapeutic drugs. 
-change 40}
+% TODO the reveiwer is asking if this assumption is reasoanbe and if you can ground it in something
+}
 
-10. Mid page 7, "decreases" is mis-spelled and I would swap out "very well" with something more descriptive where you use it.
+10. Mid page 7, "decreases" is mis-spelled
 
-\textcolor{blue}{Thank you for your comment. We have made the change from "decerases" to "decreases". change 27}
+\textcolor{blue}{%change 27
+Fixed. 
+}
 
-Reviewer $\#$2 :
+and I would swap out "very well" with something more descriptive where you use it.
+
+\textcolor{blue}{
+We changed "the approximations perform very wll" to "the approximations are accurate" in two places.
+}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\textbf{Reviewer $\#$2}
 
@@
 
-This is a paper about the likelihood and time for a tumour to evolve drug resistance, via direct mutation or indirectly through aneuploidy.
+%This is a paper about the likelihood and time for a tumour to evolve drug resistance, via direct mutation or indirectly through aneuploidy.
 The authors develop a multi-type branching process model to approximate these quantities analytically and to simulate the full behaviour.
@@
 Parameterizing for melanoma demonstrates the utility of the theory.
+
+\textcolor{blue}{
+Thank you, we are pleased you found the application clear and intuitive.
+}
 
@@
 
+\textcolor{red}{
+TODO - Daniel
+}
+
 2.
@@
 
+\textcolor{blue}{% change y1
+Aneuploidy is indeed heritable.
+The extra-chromosome is lost at a rate similar to it's gain.
+A recent paper estimated the per generation loss rate in yeast at 8e-5-6e-4, depending on the specific chromosome (Hose et al.
+Genetics 2024).
+However, this transition from aneuploidy to euploidy is rare enough that we can neglect it, because the euploid has lower fitness than the aneuploid in the presence of drug.
+As for the fitness difference between euploid and aneuploid mutants, this is a good question and the answer probably varies depending on the drug, the chromosome, and other factors.
+We now address both your questions under "Directions for future reseach" in line XXX: "Furthermore, loss of extra-chromosome may occur at a high rate (for estimates in yeast, see Hose et al.
+(2024)) and the fitness of a euploid cell with a resistance mutation may be higher than that of an aneuploid cell with the same mutation.
+Future work could test when these effects have significant effect on the results of our model."
+}
+
 $\#$ Minor comments
@@
 
-\textcolor{blue}{Thank you for your comment. We have added line numbers.}
+\textcolor{blue}{We have added line numbers.}
 
@@
 
-\textcolor{blue}{Thank you for your commnet.
-We have changed the definition of evolutionary rescue such that it highlights the fact that the population must adapt to the environmental change.
-change 32}
+\textcolor{blue}{%change 32
+We changed the first sentence in the abstract: "Evolutionary rescue occurs when a population, facing a sudden environmental change that would otherwise lead to extinction, adapts through beneficial mutations, allowing it to recover and persist."
+}
 
@@
 
-\textcolor{blue}{Thank you for your comment. We have made the requested change. change 3}
+\textcolor{blue}{Done} % change 3
 
@@
 
-\textcolor{blue}{Thank you for your comment. We have made the requested change. change 4}
+\textcolor{blue}{Done.}% change 4
 
@@
 
-\textcolor{blue}{Thank you for your comment. We have changed the text such that it states that evolutionary rescue in two steps is similar to the problem of crossing fitness valley. change 6}
+\textcolor{blue}{
+We removed the statement about the equivalence to valley crossing.
+}
 
@@
 
-\textcolor{blue}{Thank you for your comment. We have added the recommend synthetic reviews to the list of rescue papers cited. change 45}
+\textcolor{blue}{%change 45
+We have added reference to the recommended synthetic reviews in line XXX.
+}
 
@@
 
-\textcolor{blue}{Thank you for your comment. We have  placed the statement at the beginning of the paragraph. change 28}
+\textcolor{blue}{%change 28
+We have  moved the statement at the beginning of the paragraph, line XXX. 
+}
 
@@
 
-\textcolor{blue}{Thank you for your comment.
-We have changed the statement to just explain the concepts of soft and hard selective sweeps.
-Additionaly, we have added a reference to Osmond $\&$ Coop 2020 (Genetics).
-change 41}
+\textcolor{blue}{%change 41
+We have changed the statement to explain the concepts of soft and hard selective sweeps and added a reference to Osmond and Coop 2020 (line XXX): "Evolutionary rescue may occur when a beneficial mutation appears in a single copy, leading to hard selective sweeps, or when multiple mutations appear independently in different individuals, leading to soft selective sweeps.
+Wilson et al.
+(2017) have shown that evolutionary rescue is significantly enhanced by soft selective sweeps when multiple mutations contribute towards evolutionary rescue.
+Additionally, Osmond and Coop (2020) have calculated the probability that rescue is soft and the expected number of rescue mutations in different scenarios.
+Evolutionary rescue that requires two successive mutations (i.e., two steps) has been investigated by Martin et al.
+(2013), who tested their model's predictions with data from yeast and bacteria experiments.
+Iwasa et al.
+(2003, 2004) used multi-type branching process theory to approximate the probability that a population under strong selective pressure can survive extinction via succesive mutations.
+"
+}
 
@@
 
-\textcolor{blue}{Thank you for your comment. We have changed the statement to be 'The aneuploid division rate is chosen...'.   change 36}
+\textcolor{blue}{%change 36
+We repharased to (line XXX): "For the aneuploid growth rate to be intermediate between those of sensitive and mutant cells, $\Delta_s\ll\Delta_a\ll\Delta_m$, we set the aneuploid division rate to be $0.06 \le \lambda_a \le 0.1$."
+}
 
@@
 
-\textcolor{blue}{Thank you for your comment.
-We have changed the paragraph to make it clear that we obtaint the fitness cost from the experimental data and then use the formula $f=\frac{\tilde{u}\lambda_s}{c}$ to obtain $f$.
-change 5}
+\textcolor{blue}{%change 5
+We revised the text to make it clear that we estimate $c$ from an experimental result and $f$ from $c$ (line XXX): "To estimate fitness cost of aneuploidy, $c$, we note that Lukow et al.
+(2021) mixed sensitive and aneuploid A375 melanoma cells at $1:1$ ratio, cultured them in a drug-free environment, and observed the ratio evolve as a function of time with the aneuploid cells declining to $15\%$ after 24 days.
+Thus, our estimate for the fitness cost is $c=\abs{\log\left(0.15/(1-0.15)\right)/24}\approx0.07$ per day (Chevin 2011).
+We estimate the fraction of aneuploid cancer cells in the pre-treatment environment using the formula $f=\tilde{u}\lambda_s / c$, which produces an estimate of $f=10^{-3}\times10^{-1}/0.07$, that is, $0.14\%$ of pre-treatment cancer cells have the beneficial aneuploidy of interest."
+}
 
@@
 
-\textcolor{blue}{Thank you for your comment.
-The justification for not including density dependence for sensitive cells is because of the fact that they are under the influence of a powerful stress which causes the population to decline very fast.
-As a result, density dependent effects will not have an effect on the initial establishment probability of mutants.
-As shown in Figure S1, the presence of a carrying capacity does not effect the probability of evolutionary rescue.}
+\textcolor{red}{We did not include density dependence for sensitive cells because they are under strong stress that causes their number to rapidly decline.
+As a result, density will not affect the initial establishment probability of mutants.
+As shown in Figure S1, the presence of a carrying capacity does not effect the probability of evolutionary rescue.
+DOES THIS MAKE SENSE?
+WHY NOT JUST CHANGE TO INCLUDE DENSITY DEPENDENCE FOR SENSITIVES?}
 
@@
 
+\textcolor{blue}{
+The repository will become available when the paper is accepted.
+Meanwhile, the source code for our analysis can be privately accessed here: \url{https://www.dropbox.com/s/60zyc6rcjp8q859/StanaEtAlCode.zip}
+}
+
 - results: there doesn't appear to be an eqn 1
 
-\textcolor{blue}{Thank you for your comment. We have made changes such that equation 1 appears. change 29}
+\textcolor{blue}{Fixed.} %change 29
 
@@
 
-\textcolor{blue}{Thank you for your comment.
-We have mentioned that equation 2 (the numbers are changes in the updated version) is an alternative form of the formula from Orr $\&$ Unckless 2008.
-change 33}
+\textcolor{blue}{ % change 33
+We now mention that eq. 2 (previously eq.3 ) is an alternative form of an eq. from Orr and Unckless 2008 (line XXX).
+}
 
@@
 
-\textcolor{blue}{Thank you for your comment. We have added a paragraph which explains the similarity of equation 4 (now equation 3) to previous results. change 57}
+\textcolor{blue}{%change 57
+We now note the similarity of eq.~3 (previously eq.~3) to Osmond et al 2020 and Iwasa et al 2004.
+}
 
@@
 
-\textcolor{blue}{Thank you for your comment. We have made the change from "decerases" to "decreases". change 27}
+\textcolor{blue}{Fixed.} % change 27
 
@@
 
-\textcolor{blue}{Thank you for your comment. We have not been able to identify any intuition as to why the second scenario is the geometric mean of the other two scenarios.?}
+\textcolor{blue}{We agree, but we couldn't come up with such intuition.}
 
@@
 
-\textcolor{blue}{Thank you for your comment.
+\textcolor{red}{%change 51
 We have added the recommended comment about the fact that the reason why $T^*$ decreases, if we increase the aneuploid division and death rates such that $\Delta_a$ remains constant, is because of the fact that more mutants are generated and a lineage doesn't have to persist as long in order to generate a succesful mutant.
-Change 51 }
+}
 
@@
 
-\textcolor{blue}{Thank you for your comment.
+\textcolor{red}{%change 52
 In the scenario discribed decreasing the division rate also leads to a decrease in the death rate as the drugs affect the cancer cells during cellular division.
 We have expanded on the explanation of this scenario in the paper.
-Change 52}
+}
 
@@
 
-\textcolor{blue}{Thank you for your comment.
-We believe that our model does fit the data on resistance rates given that \citep{handa2022long} state that 68$\%$ of melanoma patients treated with vemurafenib experienced relapse after initial response.
-change 58}
+\textcolor{red}{ % change 58
+We believe that our model does fit with data on resistance rates: \citep{handa2022long} state that 68$\%$ of melanoma patients treated with vemurafenib experienced relapse after initial response.
+} % THE SENTENCE IN THE TEXT IS UNCLEAR.
+REVISE IT AND COPY IT HERE
 
@@
 
-\textcolor{blue}{Thank you for your comment.
-We started with the simpler model and thought that the more advanced model should be included in Appendix H.
-Additionally, we do not have biologically realistic values for the aneuploidy induced mutation rate and, as a result, we decided it was better to start with the model for which we have biologically realistic parameters.
+\textcolor{blue}{
+We start with the simpler model (same mutation rate in aneuploid and sensitive cells) and include the more complex model (different mutation rates) in Appendix H as we think the presentation is already complex enough and the literature does not have good estimates for aneuploidy induced mutation rates in cancer cells.
 }
@@
 
-\textcolor{blue}{Thank you for your comment. We have made the change from 'cancer cells(' to 'cancer cells ('. change 25}
+\textcolor{blue}{Fixed.} % change 25
 
@@
 
-\textcolor{blue}{Thank you for your comment. We have mentioned that the results has been obtained previously by Orr $\&$ Unckless 2008 and Martin et al 2013. change 42}
+\textcolor{blue}{%change 42
+We now mention that this equation has been previously shown by Orr and Unckless 2008 and by Martin et al 2013 in line XXX.
+}
 
@@
 
-\textcolor{blue}{Thank you for your comment. We thought about this and we think that tumor size is critical for the interpretation of our model}
+\textcolor{blue}{
+We agree that having both quantities can be confusing. But we think that in the cotext of cancer evolutionary rescue, the tumor size has a more intuitive interpretation than the rescue probability.
+}
 
@@
 
-\textcolor{blue}{Thank you for your comment. We have added comments that better explain the mean rescue time in the case of small and large tumors. change 46}
-
-%- results: why doesn't standing genetic variance affect the mean rescue time?
-This might have something to do with what I just mentioned (successful mutants tend to arise early) and more clues might be found in Orr $\&$ Unckless 2014, who also compare times with and without standing genetic variance.
+\textcolor{blue}{%change 46
+We have added the following text in line XXX: "For small tumors ($ N \ll N_a^*$), the mean rescue time is the two-step equivalent of the one-step result from Orr and Unckleass (2014; expectation of eq.~18).
+The mean rescue time (conditioned on rescue occuring) is a decreasing function of the sensitive and aneuploid growth rates and independent of the other model parameters, including tumor size (blue line in Figure S6).
+This is because if the population rapidly declines but is then rescued, than the resistance mutation must have appeared early; if the population slowly declines, than mutations can appear later and the mean time will be longer.
+In our focus parameter regime, we have $\Delta_s=-0.04$ and $\Delta_a=-10^{-4}$, such that the mean rescue time is mainly determined by the aneuploid growth rate, $\tau_a \approx 10^4$ days.
+\\
+For large tumors ($N \gg N_m^*$), the dynamics are equivalent to a scenario where rescue mutations appear at a constant rate, and the mean rescue time is independent of the aneuploid parameters ($u$, $\lambda_a$, and $\Delta_a$).
+Increasing the per division mutation rate, $v$, leads to the faster appearance of a rescue mutations and hence reduced mean rescue time.
+Finally, increasing the tumor size leads to shorter mean rescue time, as more sensitive cells can mutate to become resistant."
+}
 
@@
 
-\textcolor{blue}{Thank you for your comment. We have made the change from $1/\Delta_a$ to $1/\Delta_s$. change 37}
+\textcolor{blue}{Fixed.} % change 37
 
@@
 
-\textcolor{blue}{Thank you for your comment. We have added the definition of $p_m$ below equation 9. change 26}
+\textcolor{blue}{%change 26
+We have added the definition of $p_m$ in line XXX: "where $p_m$ is the probability that a lineage starting from a single mutant cell escapes stochastic extinction."
+}
 
@@
 
-\textcolor{blue}{Thank you for your comment. The return time with new mutation (eqn 23) from Orr $\&$ Unckless 2014 can be written as:
+\textcolor{blue}{The return time with new mutation (eqn 23) from Orr $\&$ Unckless 2014 can be written as:
 \begin{align*}
@@
 }
+
 - results: you mention that drugs increasing the death rate of sensitive cells delays recurrence, but given that mutations happen during division, it might be worth mentioning that drugs that decrease the division rate of sensitive cells by the same amount delay recurrence even more (holding mutant division rate constant - which is maybe why you didn't mention this for your parameterization, but seems like a worthwhile general point?).
 
-\textcolor{blue}{Thank you for your comment.
+\textcolor{red}{%change 49
 For large tumors, decreasing the division rate of sensitive cells will lead to a decrease in the cancer recurrence time when conditioned on evolutionary rescue.
 We have added this comment to the text.
-change 49}
+PLEASE CHECK AGAIN, YOU WROTE increase AND DID NOT REALLY ADDRESS THE REVIEWER COMMENT.}
 
@@
 
-\textcolor{blue}{Thank you for your comment.
-Standing genetic variation does not have a significant impact on mean rescue time because the percentage of cell which have the beneficial aneuploidy is only $0.14\%$.}
+\textcolor{blue}{Standing genetic variation does not have a significant impact on mean rescue time because the percentage of cells that have the beneficial aneuploidy, in our parameterization, is only $0.14\%$.}
 
@@
 
-\textcolor{blue}{Thank you for your comment. We have added a sentence before the expression which states that aneuploidy increases the cumulant number of mutants that arise. change 39}
+\textcolor{blue}{%change 39
+Thank you for this suggestions.
+We have revised the text (line XXX): "aneuploidy provides two advantages.
+First, it delays the extinction of the population, providing more time for the appearance of the resistance mutation.
+Second, it increases the population size relative to a drug-sensitive population, providing more cells in which mutations can occur.
+Together, this increases the cumulative number of mutants that arise (i.e., $Nuv\lambda_s\lambda_a/\abs{\Delta_s\Delta_a}$)."
+}
 
@@
 
-\textcolor{blue}{Thank you for your comment.
-In the last sentence of "Discussion" we state that certain therapies increase the rate of aneuploidy which has been shown by Lee et al.
-(2016) which is referenced in the "Introduction" section as a motivation for our research.}
+\textcolor{blue}{%change x1
+Line XXX in the Introduction reads: "Futhermore, some proposed anti-cancer drugs elevate the missegregation rate to fight cancer cells (Lee et al.
+2016), as an extremely high chromosome missegregation rate is incompatible with cell survival and proliferation."
+}
 
@@
 
-\textcolor{blue}{Daniel ?.}
+\textcolor{red}{TODO}
 
@@
 
-\textcolor{blue}{Thank you for your comment. We have added the definition of asterisk in the table. change 1}
+\textcolor{blue}{%change 1
+We have added the definition of asterisk in the table.
+}
 
@@
 
-\textcolor{blue}{Thank you for your comment.
-We have made the change such that it says "When aneuploid cells are tolerant ($\Delta_a<0$) evolutionary rescue can occur indirectly, but direct mutation is the most likely path for evolutionary rescue".
-change 24}
+\textcolor{blue}{%change 24
+We changed the caption to "When aneuploid cells are tolerant ($\Delta_a<0$) evolutionary rescue can occur indirectly, but direct mutation is the most likely path for evolutionary rescue".
+}
 
@@
 
-\textcolor{blue}{Thank you for your comment. We added the references to the equation numbers for the formulas of $N_a^*$ and $N_m^*$. change 43}
+\textcolor{blue}{%change 43
+We added equation references the formulas of $N_a^*$ and $N_m^*$.
+}
 
@@
 
-\textcolor{blue}{Thank you for your comment. We have made the change from $\lambda_m$ to $\lambda_a$ in the first instance of $\lambda_m$. change 18}
+\textcolor{blue}{%change 18
+We changed $\lambda_m$ to $\lambda_a$ in the first instance in the caption of Fig. 4 .
+}
 
@@
 
-\textcolor{blue}{Thank you for your comment. We have made the recommended change. change 19}
+\textcolor{blue}{Fixed.}  % change 19
 
@@
 
-\textcolor{blue}{Thank you for your comment. We have made the recommended change. change 20}
+\textcolor{blue}{Fixed.} % change 20
 
@@
 
-\textcolor{blue}{Thank you for your comment. We have made the recommended change. change 21}
+\textcolor{blue}{Fixed.} % change 21
 
@@
 
-\textcolor{blue}{Thank you for your comment. We have made the requested change from $w_t$ to $s_t$. change 7}
+\textcolor{blue}{Fixed.} % change 7
 
@@
 
-\textcolor{blue}{Thank you for your comment. We have changed $\mathcal{O}\left(t^2\right)$ to $\mathcal{O}\left(\Delta_st^2\right)$. change 22}
+\textcolor{blue}{We have changed $\mathcal{O}\left(t^2\right)$ to $\mathcal{O}\left(\Delta_st^2\right)$. } %change 22
 
@@
 
-\textcolor{blue}{Thank you for your comment. We have corrected the grammatical mistakes. change 8}
+\textcolor{blue}{We have corrected the grammatical mistakes. } %change 8
 
@@
 
-\textcolor{blue}{Thank you for your comment. We have added the definitions of $N_m^*$ and $N_a^*$ in Appendix C. change 38}
+\textcolor{blue}{We have added the definitions of $N_m^*$ and $N_a^*$ in Appendix C. } %change 38
 
@@
 
-\textcolor{blue}{Thank you for your comment. The justification is the following: 
+\textcolor{blue}{%change 23
+The justification is the following: 
 \begin{align*}
@@
 \end{align*}
-where f(t), F(t) and S(t) are the probability density function, cumulative distribution function and survival function.
-In our case, $S(t)=\exp\left(-\int_0^t R_a(z)dz\right)$. Additionally, for $N\gg N_m^*$ we have $1-\e^{-Np_s}\approx1$. I have corrected the following mistake:
+where $f(t)$, $F(t)$ and $S(t)$ are the probability density function, cumulative distribution function and survival function.
+In our case, $S(t)=\exp\left(-\int_0^t R_a(z)dz\right)$.
+Additionally, for $N\gg N_m^*$ we have $1-\e^{-Np_s}\approx1$. I have corrected the following mistake:
 \begin{equation*}
@@
 \end{equation*}
-change 23
 }
+\textcolor{red}{REMUS DID YOU ADD THIS TO THE APPENDIX? NOT JUST HERE}
 
@@
 
-\textcolor{blue}{Thank you for your comment. We know that
+\textcolor{blue}{We know that
 \begin{align*}
@@
 }
+\textcolor{red}{REMUS DID YOU ADD THIS TO THE APPENDIX? NOT JUST HERE}
 
@@
 
-\textcolor{blue}{Thank you for your comment. We have added citations for Orr $\&$ Unckless 2014 and Maynard Smith $\&$ Haigh 1974. change 35}
+\textcolor{blue}{We have added citations for Orr and Unckless 2014 and Maynard Smith and Haigh 1974. } %change 35
 
@@
 
-\textcolor{blue}{Thank you for your comment. We have made the change from 'it accurate' to 'it is accurate'. change 9}
+\textcolor{blue}{Fixed.} % change 9
 
@@
 
-\textcolor{blue}{Thank you for your comment. We have made the recommended change. change 10}
+\textcolor{blue}{Fixed.} %change 10
 
@@
 
-\textcolor{blue}{Thank you for your comment. We have made the recommended change. change 11}
+\textcolor{blue}{Fixed.}% change 11
 
@@
 
-\textcolor{blue}{Thank you for your comment. We have made the recommended change. change 12}
+\textcolor{blue}{Fixed.}% change 12
 
@@
 
-\textcolor{blue}{Thank you for your comment. We have made the recommended change. change 13}
+\textcolor{blue}{Fixed.}% change 13
 
@@
 
-\textcolor{blue}{Thank you for your comment. We have made the recommended change. change 14}
+\textcolor{blue}{Fixed.}% change 14
 
@@
 
-\textcolor{blue}{Thank you for your comment. We have made the recommended change. change 15}
+\textcolor{blue}{Fixed.} % change 15
 
@@
 
-\textcolor{blue}{Thank you for your comment. We have made the recommended change. change 16}
+\textcolor{blue}{Fixed.} % change 16
 
@@
 
-\textcolor{blue}{Thank you for your comment. We have made the recommended change. change 17}
+\textcolor{blue}{Fixed.} % change 17
 
-Reviewer $\#3$ :
+\textbf{Reviewer $\#3$}
 
@@
 
-\textcolor{blue}{Thank you for your comment.
-In our model, we are interested in a specific type of aneuploidy which increases the fitness of cancer cells in an environment affected by anti-cancer drugs.
+\textcolor{blue}{In our model, we are interested in a specific type of aneuploidy which increases the fitness of cancer cells in an environment affected by anti-cancer drugs.
 As a consequence, we catalogued all other forms of aneuploids as sensitive.}
@@
 
-\textcolor{blue}{Thank you for your comment.
-We consider the fitness of the sensitive, aneuploid or mutant cells to be the average of each population.
+\textcolor{blue}{We consider the fitness of the sensitive, aneuploid or mutant cells to be the average of each population.
 As a result, mutations that are deleterious are taken into consideration through the mean fitness of each population.
@@
 
-\textcolor{blue}{Thank you for your comment. The variation in our parameters in noted in Table 1 and is obtained from the papers which are cited there.}
+\textcolor{blue}{The variation in our parameters in noted in Table 1 and is obtained from the papers which are cited there.}
 
@@
 
-\textcolor{blue}{Thank you for your comment. We have made the recommended change. change 30}
+\textcolor{blue}{We have made the recommended change. change 30}
 
@@
 
-\textcolor{blue}{Thank you for your comment. We have made the recommended change from "antagonizing cell division" to "by extending the length of the cell cycle". change 34}
+\textcolor{blue}{We have made the recommended change from "antagonizing cell division" to "by extending the length of the cell cycle". change 34}
 
@@
 
-\textcolor{blue}{Thank you for your comment. We have changed Figure 7 by adding a green which represents the numerical solution of (D3).  In the first term of (D3) we have:
+\textcolor{blue}{We have changed Figure 7 by adding a green which represents the numerical solution of (D3).  In the first term of (D3) we have:
 \begin{align*}

### 2024-12-22 EvolutionaryRescue1.tex
went through R1 and R2 comments

@@
 \begin{abstract}
-Evolutionary rescue happens when a population survives a sudden environmental change that initially causes the population to decline toward extinction.
+Evolutionary rescue occurs when a population, facing a sudden environmental change that would otherwise lead to extinction, adapts through beneficial mutations, allowing it to recover and persist.
+% change 32
 A prime example of evolutionary rescue is the ability of cancer to survive exposure to treatment. 
 One evolutionary mechanism by which a population of cancer cells can adapt to chemotherapy is aneuploidy. 
-Aneuploid cancer cells can be fitter in an environment altered by anti-cancer drugs, e.g., because aneuploidy disrupts the pathways usually targeted by the drugs. 
+Aneuploid cancer cells can be more fit in an environment altered by anti-cancer drugs, e.g., because aneuploidy disrupts the pathways usually targeted by the drugs. %change 3 
 Indeed, aneuploidy is highly prevalent in tumors, and some anti-cancer drugs fight cancer by increasing chromosomal instability.
@@
 \paragraph{Aneuploidy in cancer.}  
-Each year, approximately 10 million people die from cancer \citep{kocarnik2022cancer}. Understanding the factors that contribute to the failure of interventions is of great importance.
+Each year, approximately 10 million people die from cancer globally \citep{kocarnik2022cancer}.
+Understanding the factors that contribute to the failure of interventions is of great importance.
+%change 30
 One suggested factor is aneuploidy, in which cells are characterized by an imbalanced karyotype and alterations in the number of chromosomes \citep{schukken2018cin}. 
 Aneuploidy is caused by chromosomal instability and mis-segregation of chromosomes during mitosis.
-Importantly, changes in the number of chromosomes and chromosome arms copies allow cancer cells to survive under stressful conditions such as drug therapy \citep{lukow2021chromosomal,rutledge2016selective,ippolito2021gene}.
+Importantly, changes in the number of chromosomes and chromosome arm copies allow cancer cells to survive under stressful conditions such as drug therapy \citep{lukow2021chromosomal,rutledge2016selective,ippolito2021gene}.
+%change 4
 Indeed, cancer cells are often aneuploid, and aneuploidy is associated with poor patient outcomes \citep{ben2020context,smith2018systematic}.
@@
 Similarly, \citet{lukow2021chromosomal} induced aneuploidy in cancer cells and observed that such cells have an advantage compared to sensitive cells during drug treatment despite having lower fitness before the onset of treatment.
-One proposed mechanism through which aneuploidy can confer resistance to anti-cancer drugs is by antagonizing cell division, which prevents the drugs from damaging DNA and microtubules \citep{replogle2020aneuploidy}.
+One proposed mechanism through which aneuploidy can confer resistance to anti-cancer drugs is by extending the length of the cell cycle, which prevents the drugs from damaging DNA and microtubules \citep{replogle2020aneuploidy}.
 Other mechanisms are the selection for specific karyotypes that lead to reduced drug metabolism \citep{ippolito2021gene} and elevated levels of DNA damage repair due to higher basal levels of DNA damage \citep{zerbib2023human}.
+%change 34
 
@@
 Consequently, a cell exposed to stress, such as chemotherapeutic drugs, can acquire aneuploidy faster than a mutation.
-Moreover, several proposed anti-cancer drugs elevate the missegregation rate to fight cancer cells~\citep{lee2016effects}, as an extremely high chromosome missegregation rate is incompatible with cell survival and proliferation.
+Futhermore, some proposed anti-cancer drugs elevate the missegregation rate to fight cancer cells~\citep{lee2016effects}, as an extremely high chromosome missegregation rate is incompatible with cell survival and proliferation.
+% change x1
 
@@
 Adaptation is a race against time as the population size decreases in the new environment~\citep{tanaka2022surviving}.
-\emph{Evolutionary rescue} is the process by which the population acquires a trait that increases fitness in the new environment such that extinction is averted.
-It is mathematically equivalent to the problem of crossing of fitness valley \citep{weissman2009rate,weissman2010rate}.
-There are three potential ways for a population to survive environmental change: migration to a new habitat similar to the one before the onset of environmental change \citep{harsch2014keeping,cobbold2020should,zhou2022range}; adaptation by phenotypic plasticity without genetic modification \citep{carja2019evolutionary,carja2017evolutionary,levien2021non,gunnarsson2020understanding}; and adaptation through genetic modifications, e.g., mutation \citep{gomulkiewicz1995does,uecker2014evolutionary,uecker2016role,uecker2011fixation,orr2014population}.
+\emph{Evolutionary rescue} is the process by which the population acquires an adaptation that increases fitness in the new environment such that extinction is averted.
+There are three potential ways for a population to survive environmental change \citep{alexander2014evolutionary,bell2017evolutionary}: % change 45 %change 28
+migration to a new habitat similar to the one before the onset of environmental change \citep{harsch2014keeping,cobbold2020should,zhou2022range}; 
+adaptation by phenotypic plasticity without genetic modification \citep{carja2019evolutionary,carja2017evolutionary,levien2021non,gunnarsson2020understanding}; 
+and adaptation through genetic modifications, e.g., mutation \citep{gomulkiewicz1995does,uecker2014evolutionary,uecker2016role,uecker2011fixation,orr2014population}. 
+Here, we focus on the latter.
 
@@
 
-Most models focus on the probability that at most one mutation rescues the population.
-How multiple mutations contribute to the survival of the population is less explored, but \citet{wilson2017soft} have shown that evolutionary rescue is significantly enhanced by soft selective sweeps when multiple mutations contribute towards evolutionary rescue.
+Evolutionary rescue may occur when a beneficial mutation appears in a single copy, leading to hard selective sweeps, or when multiple mutations appear independently in different individuals, leading to soft selective sweeps.
+\citet{wilson2017soft} have shown that evolutionary rescue is significantly enhanced by soft selective sweeps when multiple mutations contribute towards evolutionary rescue.
+Additionally, \citet{osmond2020genetic} have calculated the probability that rescue is soft and the expected number of rescue mutations in different scenarios.
 Evolutionary rescue that requires two successive mutations (i.e., two steps) has been investigated by \citet{martin2013probability}, who tested their model's predictions with data from yeast and bacteria experiments.
 \citet{iwasa2003evolutionary,iwasa2004evolutionary} used multi-type branching process theory to approximate the probability that a population under strong selective pressure can survive extinction via succesive mutations.
+%change 41
 
 Here, we study evolutionary rescue after a sudden environmental change caused by initiating anti-cancer drug treatment. 
-We consider a range of effects of aneuploidy, from tolerance to (partial) resistance to the drug \citep{brauner2016distinguishing}.
+We consider a range of effects of aneuploidy, from tolerance to (partial) resistance to the drug \citep{brauner2016distinguishing} (we note that we use tolerant to refer to aneuploids which have negative growth rates, which is different from other references in the literature such as \citet{berman2020drug}).
+%change 47
 We estimate the effect of aneuploidy on the tumor's evolutionary rescue probability.
@@
 To model standing genetic variation, we assume that before the onset of therapy, sensitive cells become aneuploid with rate $\tilde{u}\lambda_s$ (which may differ from $u \lambda_s$) and that aneuploidy confers a fitness cost $c$ in the drug-free environment, that is, we assume that aneuploid cells have an increased death rate compared to sensitive cells in a drug-free environment.
+We assume that $c$ is not too small (\Cref{table1}), so that standing variation is generated by a balance between the generation of anuploids and selection against them, without the effect of genetic drift.
 
@@
 
-\begin{subequations}
-\begin{flalign*}
+\begin{align*}
 (+1,0,0)&:\quad \lambda_ss_t\left(1-u-v\right)\quad\left(\text{birth of sensitive cell}\right),\\
@@
 (0,0,-1)&:\quad \mu_mm_t\quad\left(\text{death of mutant cell}\right).
-\end{flalign*}
-\end{subequations}
+\end{align*}
 For the remaining of this paper, we assume that the division rates of sensitive and aneuploid cells can be written as $\lambda_ss_t\left(1-u-v\right)\approx \lambda_ss_t$ and $\lambda_aa_t\left(1-v\right)\approx\lambda_aa_t$ because $u,v\ll1$ (\Cref{table1}).
@@
 We select this to be the division rate for sensitive and mutant cells.
-%We select the mutant death rate $\mu_m=0.09\text{ days}^{-1}$ such that $\Delta_m>0$.
-
-\citet{bozic2013evolutionary} report the growth rate $\Delta_s$ for sensitive melanoma cancer cells from which they deduce the death rate $0.11 \le \mu_s \le 0.17$.
+This is a reasonable assumption given that we assume that the mutation makes the cancer cells immune to the chemotherapeutic drugs.
+%change 40
+
+\citet{bozic2013evolutionary} report the growth rate $\Delta_s$ for sensitive melanoma cancer cells, from patient data, from which they deduce the death rate $0.11 \le \mu_s \le 0.17$.
 We use  $\mu_s=0.14$ per day.
@@
 Thus, we use $\mu_m=0.1-0.01=0.09$ per day as the death rate for mutant cells.
-
-Aneuploid death rate $\mu_a$ is set to the same value as the mutant death rates, $\mu_m=0.09$ per day, given that aneuploidy increases resistance to the drug, such as cisplatin, by antagonizing cell division \citep{replogle2020aneuploidy}.
-The aneuploid division rate is selected such that the aneuploid growth rate $\Delta_s\ll\Delta_a\ll\Delta_m$ means that $0.06 \le \lambda_a \le 0.1$. 
-For most of our simulations, we use $\lambda_a=0.0899$ per day, so that aneuploid cells are tolerant and aneuploidy can only act as an evolutionary ``stepping stone'' for the generation of the resistant mutant that rescues the tumor (note that this mutant will occur on the background of an aneuploid genotype).
+%change 54
+
+The aneuploid death rate $\mu_a$ is set to the mutant death rate, $\mu_m=0.09$ per day, assuming aneuploidy increases resistance to the drug, such as cisplatin, by antagonizing cell division \citep{replogle2020aneuploidy}.
+For the aneuploid growth rate to be intermediate between those of sensitive and mutant cells, $\Delta_s\ll\Delta_a\ll\Delta_m$, we set the aneuploid division rate to be $0.06 \le \lambda_a \le 0.1$.
+% change 36
+In most of our simulations, we use $\lambda_a=0.0899$ per day, so that aneuploid cells are tolerant and aneuploidy can only act as an evolutionary ``stepping stone'' for the generation of the resistant mutant that rescues the tumor (note that this mutant will occur on the background of an aneuploid genotype).
 
@@
 
-The fitness cost $c$ of aneuploidy before the onset of therapy is difficult to estimate as we are interested in a specific type of aneuploidy that improves the fitness of cancer cells in an environment altered by drugs.
-We estimate $c=\tilde{u}\lambda_s / f$, where $f$ is the fraction of aneuploid cancer cells.
-To estimate $f$, we note that \citet{lukow2021chromosomal} mixed sensitive and aneuploid A375 melanoma cells at $1:1$ ratio, cultured them in a drug-free environment, and observed the ratio evolve as a function of time with the aneuploid cells declining to $15\%$ after 24 days.
-Thus, our estimate for the fitness cost is $c=\abs{\log\left[0.15/(1-0.15)\right]/24}\approx0.07$ per day \citep{chevin2011measuring}, and the estimated fraction of cancer cells with the beneficial aneuploidy is $f=10^{-3}\times10^{-1}/0.07=0.14\%$, that is, $0.14\%$ of pre-treatment cancer cells have the beneficial aneuploidy of interest.
+%The fitness cost $c$ of aneuploidy before the onset of therapy is difficult to estimate as we are interested in a specific type of aneuploidy that improves the fitness of cancer cells in an environment altered by drugs.
+To estimate fitness cost of aneuploidy, $c$, we note that \citet{lukow2021chromosomal} mixed sensitive and aneuploid A375 melanoma cells at $1:1$ ratio, cultured them in a drug-free environment, and observed the ratio evolve as a function of time with the aneuploid cells declining to $15\%$ after 24 days.
+Thus, our estimate for the fitness cost is $c=\abs{\log\left(0.15/(1-0.15)\right)/24}\approx0.07$ per day \citep{chevin2011measuring}.
+We estimate the fraction of aneuploid cancer cells in the pre-treatment environment using the formula $f=\tilde{u}\lambda_s / c$, which produces an estimate of $f=10^{-3}\times10^{-1}/0.07$, that is, $0.14\%$ of pre-treatment cancer cells have the beneficial aneuploidy of interest.
+%change 5
 
 We note that when we refer to drug-sensitive cells, we include those cells that have any aneuploidy other than trisomy in Chr 2 and 6, as those are the aneuploid cells that are hypothesized to have higher fitness in the environment altered by drugs such as vemurafenib.
+Additionally, in our model, mutations only confer resistance, however, most mutations are deleterious.
+We consider the fitness of the sensitive, aneuploid or mutant cells to be the average of each population.
+As a result, mutations that are deleterious are taken into consideration through the mean fitness of each population.
+%change 50
 
@@
 
-In our model, \emph{evolutionary rescue} occurs when drug-resistant cells appear and establish (avoid random extinction) in the population  ($m_t \gg 1$) before the population becomes extinct ($w_t=a_t=m_t=0$).
+In our model, \emph{evolutionary rescue} occurs when drug-resistant cells appear and establish (avoid random extinction) in the population  ($m_t \gg 1$) before the population becomes extinct ($s_t=a_t=m_t=0$).
+%change r1
 Aneuploidy may contribute to evolutionary rescue by either preventing (when $\Delta_a>0$) or delaying (when $0>\Delta_a>\Delta_s$) the extinction of the population before mutant cells appear and establish.
-We assume independence between clonal lineages starting from an initial population of $N$ sensitive cells (we check the effect of density-dependent growth on our results below).
+We assume independence between clonal lineages starting from an initial population of $N$ sensitive cells (we check the effect of density-dependent growth on our results below) and therefore, we use multi-type branching processes to model the dynamics of the cancer cell population.
+Multi-type branching process models the growth and evolution of populations with distinct types, capturing dynamics like mutation and selection.
+In evolutionary rescue, it predicts the probability of population survival by tracking adaptive mutations that emerge and spread under environmental stress.
+% change 53
 Define $p_s$ as the probability that a lineage starting from a single drug-sensitive cell avoids extinction by acquiring drug resistance.
 Thus, $N^*=1/p_s$ is the threshold tumor size above which evolutionary rescue is very likely, and the rescue probability is given by 
-\begin{equation} \label{eq:rescue_prob} 
+\begin{equation}\label{eq:rescue_prob} 
 \presc = 
@@
 \begin{equation} \label{eq:N_m}
-N_m^* \approx \frac{\abs{\Delta_s}}{v\lambda_s}  \frac{\lambda_m}{\Delta_m} .
-\end{equation}
+N_m^* \approx \frac{\abs{\Delta_s}}{v\lambda_s}  \frac{\lambda_m}{\Delta_m} ,
+\end{equation}
+which is similar to a classical result by \citet{orr2008population}. %change 33
 Here, $\abs{\Delta_s}/\left(v\lambda_s\right)$ is the ratio of the rate at which sensitive cells decrease in number and the rate at which they are mutating.
@@
 \end{equation}
-These approximations perform very well when compared to the results of stochastic evolutionary simulations (\Cref{rescue_prob,rescue_threshold}).
+Note that that \cref{eq:N_a} is an approximation of the two step-evolutionary rescue probability derived by \citet{osmond_genetic_2020} with Dirac delta fitness distribution, which is also similar to eq.~A3 in \citep{iwasa2004evolutionary}.
+%change 57
+These approximations are accurate when compared to results of stochastic evolutionary simulations (\Cref{rescue_prob,rescue_threshold}). 
 The first line describes the scenario in which the treatment still effectively kills aneuploid cells but not as quickly as the sensitive cells. 
@@
 Furthermore, the growth rate $\Delta_a<0$ that allows tolerant aneuploids to rescue the tumor is between $-u\lambda_a$ and  $-1/T^*$, which is proportional to $-\sqrt{v}$.
-Thus, increasing the mutation rate $v$ will decrease the tumor threshold size $N^*_a$, making evolutionary rescue more likely, but only until $T^*$ decerases to a point where  $\Delta_a =1/T^*$.
+Thus, increasing the mutation rate $v$ will decrease the tumor threshold size $N^*_a$, making evolutionary rescue more likely, but only until $T^*$ decreases to a point where  $\Delta_a =1/T^*$.
+%change 27
 
@@
 
-Interestingly, increasing both the aneuploid division rate, $\lambda_a$, and the aneuploid death rate, $\mu_a$, such that the growth rate $\Delta_a$ remains constant, leads to a decrease in $T^*$, pushing the system to the second scenario.
+Interestingly, increasing both the aneuploid division rate, $\lambda_a$, and the aneuploid death rate, $\mu_a$, such that the growth rate $\Delta_a$ remains constant, leads to a decrease in $T^*$, pushing the system to the second scenario (increasing $\lambda_a$ causes a decrease in $T^*$ because  it increases the mutation rate and a lineage does not have to survive as long in order to generate a successful mutant).
 In this scenario, increasing the division rate $\lambda_a$ should also increase the mutation rate $v\lambda_a$ in aneuploid cells, as mutations mostly occur during division, so overall, the threshold tumor size $N_a^*$ is unaffected by the division rate $\lambda_a$ (i.e., $d \lambda_a T^*/d\lambda_a = 0$).
 Thus, if aneuploid cells rapidly die due to the drug but compensate by rapidly dividing, increasing the division rate will \emph{not} facilitate adaptation.
-This is consistent with experimental findings where aneuploidy confers resistance by decreasing the division rate~\citep{replogle2020aneuploidy}.
+%change 51
+This is consistent with experimental findings where aneuploidy confers resistance by decreasing the division rate which in turn decreases the death rate as certain drugs affect cancer cells during division~\citep{replogle2020aneuploidy}.
+%change 52
 
@@
 Interestingly, the threshold between small and intermediate tumors, $N_a^*$, is similar to the tumor detection threshold of $4.19 \times 10^6$ cells for a wide variety of tumors~\citep{avanzini2019cancer}.
-
-Aneuploidy may lead to an increased mutation rate in cancer cells\citep{garribba2023short,passerini2016presence,janssen2011chromosome}.
+We note that vemurafenib-treated melanomas (i.e.
+melanomas with sizes above the detection threshold) have a probability $>50\%$ to relapse \citep{piejko2023long,handa2022long}.
+%change 58
+
+Aneuploidy may lead to an increased mutation rate in cancer cells \citep{garribba2023short,passerini2016presence,janssen2011chromosome}.
 Thus, we extended our model to account for this in Appendix \ref{sec:appendix-diff-mut}.
@@
 Also, it transitions the system from the first scenario (tolerant aneuploids) to the second scenario (stationary aneuploids) without changing the aneuploid growth rate, $\Delta_a$.
+%change 25
 
@@
 \end{equation}
-Comparing this approximation of $\tilde{N}^*_{a}/N^*_{a}$ to results of stochastic simulations, we find that the approximations perform very well (\Cref{rescue_denovo}). 
+This result has been previously reported by \citet{orr2008population} for the one-step evolutionary rescue and by \citet{martin2013probability} for two-step evolutionary rescue. % change 42
+Comparing this approximation of $\tilde{N}^*_{a}/N^*_{a}$ to results of stochastic simulations, we find that the approximations are accurate (\Cref{rescue_denovo}). 
 Standing genetic variation will drive evolutionary rescue if sensitive cells die rapidly (growth rate $\Delta_s$ is very negative) due to a strong effect of the drug on sensitive cells or if the cost of aneuploidy in the drug-free environment, $c$, is small.
@@
 When the expected number of resistant lineages that avoid extinction is small, the expected recurrence time can be estimated by adding two terms: the \emph{mean evolutionary rescue time}, which is the waiting time for the appearance of a resistant lineage that avoids extinction (conditioned on such an event occurring in the first place), and the \emph{mean proliferation time}, which is the expected time for that lineage to grow to $N$ cells.
-However, when the expected number of resistant lineages is large, the dynamics of the number of mutant cells is deterministic (i.e., it can be modeled by a system of ODEs, \cref{detODE}), and the mean recurrence time cannot be separated into the mean evolutionary rescue time and mean proliferation time because multiple mutant lineages contribute towards the mutant population size reaching the initial tumor size.
+However, when the expected number of resistant lineages is large, the mean recurrence time cannot be separated into the mean evolutionary rescue time and mean proliferation time because multiple mutant lineages contribute towards the mutant population size reaching the initial tumor size.
+In this case the dynamics of the number of mutant cells is deterministic and can therefore be modeled by a system of ordinary differential equations (ODE), which describe how the number of mutant cells changes over time by its time derivative (eq.~\ref{detODE}).
+%change 2
 Of particular interest is the distribution of the evolutionary rescue time and recurrence time with tolerant aneuploid cells ($\Delta_a T^*\ll1$), for which we focus on the parameter values $\lambda_s=0.1$, $\lambda_a=0.0899$, $\lambda_m=0.1$, $\mu_s=0.14$, $\mu_a=0.09$, $\mu_m=0.09$, $u=10^{-2}$, and $v=10^{-7}$.
@@
  N \ll N_a^* ,\\ 
-  \frac{1}{v\lambda_s N}   \frac{\lambda_m}{\Delta_m} ,& % I replaced pm with 𝜆𝑚 − 𝜇𝑚 / 𝜆𝑚
+  \frac{1}{v\lambda_s N}   \frac{\lambda_m}{\Delta_m} ,& 
   N \gg N_m^* .
@@
 \end{equation} 
-For small tumors ($ N \ll N_a^*$), the mean rescue time is a function of the sensitive and aneuploid growth rates and independent of the other model parameters, including tumor size (blue line in \Cref{EvolutionaryRescueTimeComplete}).
-Increasing the sensitive or aneuploid growth rates leads to an increase in the mean rescue time, because the corresponding cells will survive for longer and will produce additional rescue mutations at latter times.
+For small tumors ($ N \ll N_a^*$), the mean rescue time is the two-step equivalent of the one-step result from \citet[expectation of eq.~18]{orr2014population}.
+The mean rescue time (conditioned on rescue occuring) is a decreasing function of the sensitive and aneuploid growth rates and independent of the other model parameters, including tumor size (blue line in \Cref{EvolutionaryRescueTimeComplete}).
+This is because if the population rapidly declines but is then rescued, than the resistance mutation must have appeared early; if the population slowly declines, than mutations can appear later and the mean time will be longer.
 In our focus parameter regime, we have $\Delta_s=-0.04$ and $\Delta_a=-10^{-4}$, such that the mean rescue time is mainly determined by the aneuploid growth rate, $\tau_a \approx 10^4$ days.
-
-For large tumors ($N \gg N_m^*$), the mean evolutionary rescue time (\cref{eq:AsymptoticTimeRules}) is independent of parameters characterizing aneuploid cells or their production ($u$, $\lambda_a$, and $\Delta_a$).
-Increasing the per division mutation rate, $v$ leads to the faster appearance of a rescue mutation and hence reduced mean rescue time. 
+% change 46
+
+For large tumors ($N \gg N_m^*$), the dynamics are equivalent to a scenario where rescue mutations appear at a constant rate, and the mean rescue time (\cref{eq:AsymptoticTimeRules}) is independent of the aneuploid parameters ($u$, $\lambda_a$, and $\Delta_a$).
+Increasing the per division mutation rate, $v$, leads to the faster appearance of a rescue mutations and hence reduced mean rescue time.
 Finally, increasing the tumor size leads to shorter mean rescue time, as more sensitive cells can mutate to become resistant.
+% change 46
 
@@
 This allows us to examine whether aneuploidy promotes or delays evolutionary rescue.
-We find that aneuploidy promotes evolutionary rescue after $1/\Delta_a\approx100$ days, at a time when no more rescue mutations are generated through mutations in sensitive cells (\Cref{cdffig}A).
+We find that aneuploidy promotes evolutionary rescue after $1/\Delta_s\approx100$ days, at a time when no more rescue mutations are generated through mutations in sensitive cells (\Cref{cdffig}A).
 Thus, aneuploidy increases the \emph{window of opportunity} for evolutionary rescue.
 This can have a counter-intuitive outcome: conditioned on the rescue of the tumor, tumors rescued by aneuploid cells may acquire rescue mutations later than those rescued by sensitive cells.
+%change 37
 
@@
   \frac{1}{\Delta_m}\log\frac{\Delta_m-\Delta_s}{v\lambda_s}  ,&
-  N \gg N_m^* .
+  N \gg N_m^* ,
   \end{cases}
 \end{equation}
-\Cref{proliferationFigure,ProliferationTimeLarge} show the agreement between our approximations and simulation results.
+where $p_m$ is the probability that a lineage starting from a single mutant cell escapes stochastic extinction. %change 26 
+  ~\Cref{proliferationFigure,ProliferationTimeLarge} show the agreement between our approximations and simulation results.
 For small tumors ($N \ll N_a^*$), the mean recurrence time can be approximated as the sum of the mean time for the first rescue mutation to appear and the mean time for its lineage to reach size $N$.
@@
 Increasing either the mutant growth rate, $\Delta_m$, or the mutation rate, $v$, decreases the time for the tumor to rebound to its initial size.
-In addition, drugs that significantly increase the death rate of drug-sensitive cells, $\mu_s$, but do not affect their division rate, $\lambda_s$, delay cancer recurrence (conditioned on evolutionary rescue).
+Drugs that significantly increase the death rate of sensitive cells, $\mu_s$, but do not affect their division rate, $\lambda_s$, delay cancer recurrence (conditioned on evolutionary rescue).
 Consequently, patients treated with such drugs may require a longer period of monitoring to guarantee the effectiveness of the treatment.
+In addition, drugs that significantly increase the division rate of sensitive cells, $\lambda_s$, but do not affect their death rate, $\mu_s$, decrease cancer recurrence times (conditioned on evolutionary rescue).
+%change 49
 
@@
 We derive the mean detection time for $M=10^7$ in Appendix \ref{sec:appendix_recurrence_time}.
-We find for small and intermediate-sized tumors, the detection size $M$ has a negligible effect on the mean detection time $\tau_a^{M}$ compared to when the detection size equals the initial population size, $N$ (i.e., $\tau_a^r\approx\tau_a^{M}$ for $N<N_m^*$).
+We find that for small and intermediate-sized tumors the mean detection time is approximately equal to the mean recurrence time (i.e., $\tau_a^r\approx\tau_a^{M}$ for $N<N_m^*$).
+%change 48
 However, for large tumors, the mean detection time $\tau_a^{M}$ decreases logarithmically with tumor size $N$, while the recurrence time $\tau_a^{r}$ is constant (\Cref{RecurrencePlot}).
@@
 First, it delays the extinction of the population, providing more time for the appearance of the resistance mutation.
-Second, it increases the population size relative to a drug-sensitive population, providing more cells in which mutations can occur, i.e., it increases the mutation supply (i.e., $Nuv\lambda_s\lambda_a/\abs{\Delta_s\Delta_a}$).
+Second, it increases the population size relative to a drug-sensitive population, providing more cells in which mutations can occur.
+Together, this increases the cumulative number of mutants that arise (i.e., $Nuv\lambda_s\lambda_a/\abs{\Delta_s\Delta_a}$).
+% change 39
 
@@
 This can potentially impact the probability of evolutionary rescue \citep{martens2011spatial}.
-In addition, our model can be extended to understand evolutionary rescue in different biological contexts, for example, how yeast populations under stress overcome extinction via aneuploidy~\citep{pompei2023fitness,kohanovski2024aneuploidy}.
+Furthermore, loss of extra-chromosome may occur at a high rate (for estimates in yeast, see \citet{hose_rate_2024}) and the fitness of a euploid cell with a resistance mutation may be higher than that of an aneuploid cell with the same mutation.
+Future work could test when these effects have significant effect on the results of our model.
+% change y1
+In addition, our model can be applied to study evolutionary rescue in other biological systems, such as evolution of yeast populations under different stress conditions~\citep{pompei2023fitness,kohanovski2024aneuploidy}.
 
@@
     $\tilde{u}$ & Missegregation rate in the drug free environment$^\ast$& $5\times10^{-4}-2\times10^{-2}$ & 1$\slash$cell division  & \citet{shi2005chromosome,thompson2008examining} \\ \hline
-    $s$ & Selection coefficient of aneuploidy in the drug free environment& $0.07$ &  1/days   & \citet{lukow2021chromosomal} \\  
+    $c$ & Selection coefficient against aneuploidy in the drug free environment& $0.07$ &  1/days   & \citet{lukow2021chromosomal} \\  
     \hline
@@
 We have modified the parameters from \citet{bozic2013evolutionary} such that sensitive and mutant division rates are $\lambda_{s} = \lambda_{m} = \log2/T\approx0.1$ instead of their value of $0.14$ where $T$ is the doubling time in the absence of cellular death obtained from \citet{rew2000cell}.
-For a discussion of the different interpretations of the tumor doubling times see \citet{avanzini2019cancer}.}
+For a discussion of the different interpretations of the tumor doubling times see \citet{avanzini2019cancer}.
+Parameters marked with $^\ast$ are not obtained from the literature.} %change 1
   \label{table1}
@@
 (A) Without aneuploidy ($u=0$), evolutionary rescue is possible through direct mutation, and in most scenarios, the tumor will become extinct due to the drug.
-(B) When aneuploid cells are tolerant ($\Delta_a<0$), we observe, similar to A, direct mutation is the only path for evolutionary rescue.
+(B) When aneuploid cells are tolerant ($\Delta_a<0$) evolutionary rescue can occur indirectly, but direct mutation is the most likely path for evolutionary rescue.
+%change 24
 (C) When aneuploid cells are stationary ($\Delta_a\approx0$), we observe the appearance of mutant lineages even after the sensitive population has gone extinct, thus showing that stationary aneuploidy increases the probability of evolutionary rescue.
@@
 \textbf{The effect of aneuploidy on tumor threshold size.}
-(A) The threshold tumor size $N_a^*$ as a function of the aneuploid growth rate $\Delta_a$.
-The dashed horizontal line shows $N^*_m$, the threshold tumor size without aneuploidy ($u=0$).
+(A) The threshold tumor size $N_a^*$ (\cref{eq:N_a}) as a function of the aneuploid growth rate $\Delta_a$.
+The dashed horizontal line shows $N^*_m$ (\cref{eq:N_m}), the threshold tumor size without aneuploidy ($u=0$).
 When aneuploid growth rate is close to or higher than zero, aneuploidy decreases the threshold tumor size, facilitating evolutionary rescue.
@@
 Parameters: $\lambda_s=0.1,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.
-(B) Threshold tumor size $N_a^*$ as a function of the ratio of aneuploidy and mutation rates, $u/v$.
-Dashed horizontal line shows $N^*_m$, the threshold tumor size without aneuploidy ($u=0$).
+%change 43
+(B) Threshold tumor size $N_a^*$ (\cref{eq:N_a}) as a function of the ratio of aneuploidy and mutation rates, $u/v$.
+Dashed horizontal line shows $N^*_m$ (\cref{eq:N_m}), the threshold tumor size without aneuploidy ($u=0$).
 When the aneuploidy rate is much higher than the mutation rate, aneuploidy decreases the threshold tumor size, facilitating evolutionary rescue.
@@
 Red dots represent simulation results, and the error bars represent the $95\%$ confidence intervals obtained with bootstrap (Appendix G).
-Parameters: $\lambda_s=0.1,\lambda_m=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, v=10^{-7}$.
+Parameters: $\lambda_s=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, v=10^{-7}$.
+%change 18, 43
 }
@@
 (A) The probability that a successful mutant has not appeared by time $t$.
-Green line: tolerant aneuploidy ($u>0, \lambda_a=0.0899$).
-Blue line: stationary aneuploidy ($u>0, \lambda_a=0.089999$).
-Cyan line: partially resistant aneuploidy ($u>0, \lambda_a=0.095$).
-Black line: no aneuploidy ($u=0$).
+Black line: tolerant aneuploidy ($u>0, \lambda_a=0.0899$).
+Red line: stationary aneuploidy ($u>0, \lambda_a=0.089999$).
+Green line: partially resistant aneuploidy ($u>0, \lambda_a=0.095$).
+Blue line: no aneuploidy ($u=0$).
 Aneuploidy plays an important role in rescuing the tumor cell population as the sensitive population becomes extinct.
@@
 Parameters: $\lambda_s=0.1,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7},N=10^7$.
+%change 56
 (B) Probability that a mutant cancer cell population has not reached size $N$ at time $t$.
@@
 The simulations converge to \cref{eq:t2det} (blue dashed line) for large values of $N$.
+The green line represents the numerical solution of \cref{eq:t2det2024}.
+% change 44
 Parameters: $\lambda_s=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}
@@
 
-To analyze evolutionary rescue in our model, we use the framework of \emph{multitype branching processes} \citep{harris1963theory, weissman2009rate}. 
+To analyze evolutionary rescue in our model, we use the framework of \emph{multi-type branching processes} \citep{harris1963theory, weissman2009rate}. % change 31
 This allows us to find explicit expressions for the \emph{survival probability}: the probability that a lineage descended from a single cell does not become extinct.
@@
 We now assume that aneuploidy provides tolerance to drug therapy, that is, the number of aneuploid cells significantly declines over time, but at a lower rate than the number of sensitive cells, $\lambda_s - \mu_s < \lambda_a - \mu_a < 0$.
-We also assume that the decline are significant, $\left(\lambda_a-\mu_a-v\lambda_ap_m\right)^2 > 4\lambda_a^2 v p_m$.
+We also assume that the decline is significant, $\left(\lambda_a-\mu_a-v\lambda_ap_m\right)^2 > 4\lambda_a^2 v p_m$.
+%change 19
 We rewrite \cref{eq:survival_prob} as
@@
 p_s&\approx-\frac{1}{\lambda_s-\mu_s-u\lambda_sp_a-v\lambda_sp_m}\left[v\lambda_s\frac{\lambda_m-\mu_m}{\lambda_m}+\frac{u\lambda_s}{2\lambda_a}\left(\lambda_a-\mu_a-v\lambda_ap_m+2\sqrt{\lambda_a^2vp_m}\right)\right]\\
-&=-\frac{1}{\lambda_s-\mu_s-u\lambda_s-v\lambda_s}\left[v\lambda_s\frac{\lambda_m-\mu_m}{\lambda_m}+\frac{u\lambda_s}{2\lambda_a}\left(\lambda_a-\mu_a-v\lambda_ap_m\right)+u\lambda_s\sqrt{\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}}\right]\\
-&\approx-\frac{1}{\Delta_s}\left[v\lambda_s\frac{\Delta_m}{\lambda_m}+\frac{u\lambda_s\left(\Delta_a-v\lambda_a\right)}{2\lambda_a}+u\lambda_s\sqrt{\frac{v\Delta_m}{\lambda_m}}\right].
+&=-\frac{1}{\lambda_s-\mu_s-u\lambda_sp_a-v\lambda_sp_m}\left[v\lambda_s\frac{\lambda_m-\mu_m} {\lambda_m}+\frac{u\lambda_s}{2\lambda_a}\left(\lambda_a-\mu_a-v\lambda_ap_m\right)+u\lambda_s\sqrt{\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}}\right]\\ %change 20
+&\approx-\frac{1}{\Delta_s}\left[v\lambda_s\frac{\Delta_m}{\lambda_m}+\frac{u\lambda_s\left(\Delta_a-v\lambda_ap_m\right)}{2\lambda_a}+u\lambda_s\sqrt{\frac{v\Delta_m}{\lambda_m}}\right]. %change 21
 \end{aligned}
@@
 
-The number of successful mutants generated until time $t$ can be approximated by an inhomogeneous Poisson process with rate $R_m\left(t\right) = v\lambda_s p_m w_t$,
+The number of successful mutants generated until time $t$ can be approximated by an inhomogeneous Poisson process with rate $R_m\left(t\right) = v\lambda_s p_m s_t$, %change 7
 where $s_t=N\e^{\Delta_s t}$ is the number of sensitive cells at time $t$.
@@
 \end{equation}
-by integrating the exponential and because $\frac{\exp[\Delta_s t]-1}{\Delta_s}=\frac{1+\Delta_s t+\mathcal{O}(t^2)-1}{\Delta_s}=t+O(t^2)$.
+by integrating the exponential and because $\frac{\exp[\Delta_s t]-1}{\Delta_s}=\frac{1+\Delta_s t+\mathcal{O}\left(\left(\Delta_st\right)^2\right)-1}{\Delta_s}=t+O(\Delta_st^2)$. %change 22
 The probability density function of $T_m$ is thus
@@
 \end{align}
-\Cref{MeanTimeGrowthAneuploidyPlot}B show the agreement between this approximating and simulation results.
+\Cref{MeanTimeGrowthAneuploidyPlot}B shows the agreement between this approximation and simulation results. %change  8
 \\
@@
 As a result, the number of successful mutants created by direct mutation and via aneuploidy can be approximated by inhomogeneous Poisson processes with the rates
+\begin{align*}
+r_1\left(t\right)&=v\lambda_ap_ma_{t},\\
+r_2\left(t\right)&=v\lambda_sp_ms_{t},
+\end{align*}
+and the expected number of successful mutants created by direct mutation and via aneuploidy until time $t$ is given by:
 \begin{align}\label{eq:twosteplineage}
-r_1\left(t\right)&=v\lambda_ap_m\int_0^ta_{z} \d z = \frac{uv\lambda_s\lambda_aNp_m}{\Delta_s-\Delta_a}\left(\frac{\e^{\Delta_st}-1}{\Delta_s}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right),\\ \label{eq:twosteplineagedirect}
-r_2\left(t\right)&=v\lambda_sp_m\int_0^ts_{z} \d z = v\lambda_sNp_m\frac{\e^{\Delta_s t}-1}{\Delta_s}.
-\end{align}
+M_a\left(t\right)&=v\lambda_ap_m\int_0^ta_{z} d z = \frac{uv\lambda_s\lambda_aNp_m}{\Delta_s-\Delta_a}\left(\frac{\e^{\Delta_st}-1}{\Delta_s}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right),\\ 
+M_m\left(t\right)&=v\lambda_sp_m\int_0^ts_{z} d z = v\lambda_sNp_m\frac{\e^{\Delta_s t}-1}{\Delta_s}.
+\end{align} %change r3
+
 For large initial population sizes we assume that the two processes are independent and as a result, they can be merged into a single Poisson process with rate $R_a(t)=\left(r_1+r_2\right)\left(t\right)$.
@@
 %%%
-If $N\gg N_m^*$ then the mean time $\tau_a$ can be written as:
+Let $N_a^*$ and $N_m^*$ be the threshold population size above which evolutionary rescue through aneuploidy or direct mutation, respectively, is very likely.
+If $N\gg N_m^*$ then $p_{rescue}\approx1$ and the mean time $\tau_a$ can be written as: % change 38
 \begin{align*}
-\tau_a=\int_0^\infty\e^{-R_a\left(\tau\right)}\,\d\tau=\int_0^\infty\exp\left[-\frac{uv\lambda_s\lambda_aNp_m}{\Delta_s-\Delta_a}\left(\frac{\e^{\Delta_s\tau}-1}{\Delta_s}-\frac{\e^{\Delta_a\tau}-1}{\Delta_a}\right)-v\lambda_sNp_m\frac{\e^{\Delta_s\tau}-1}{\Delta_s}\right]\,\d\tau,
-\end{align*}
+\tau_a=\int_0^\infty\e^{-\int_0^t R_a\left(z\right)\d z}\,\d t=\int_0^\infty\exp\left[-\frac{uv\lambda_s\lambda_aNp_m}{\Delta_s-\Delta_a}\left(\frac{\e^{\Delta_s t}-1}{\Delta_s}-\frac{\e^{\Delta_a t}-1}{\Delta_a}\right)-v\lambda_sNp_m\frac{\e^{\Delta_s t}-1}{\Delta_s}\right]\,\d t,
+\end{align*} %change 23 ,r4
 and we use the following Taylor series expansions:
 \begin{align*}
-\frac{\e^{\Delta_s\tau}-1}{\Delta_s}&=\frac{1+\Delta_s\tau+O(\tau^2)-1}{\Delta_s}=\tau+O(\tau^2).\\
-\frac{\e^{\Delta_a\tau}-1}{\Delta_a}&=\frac{1+\Delta_a\tau+O(\tau^2)-1}{\Delta_a}=\tau+O(\tau^2),
+\frac{\e^{\Delta_s t}-1}{\Delta_s}&=\frac{1+\Delta_s t+O(t^2)-1}{\Delta_s}=t+O(t^2).\\ % change r4
+\frac{\e^{\Delta_a t}-1}{\Delta_a}&=\frac{1+\Delta_a t+O(t^2)-1}{\Delta_a}=t+O(t^2), % change r4
 \end{align*}
@@
 \begin{align}\label{limitapprox3}
-\tau_a\approx\int_0^\infty\e^{-v\lambda_sNp_m\tau}\,\d\tau=\frac{1}{v\lambda_sNp_m}.
+\tau_a\approx\int_0^\infty\e^{-v\lambda_sNp_m t}\,\d t=\frac{1}{v\lambda_sNp_m}. % change r4
 \end{align}
@@
 \begin{align}\nonumber
-\tau_a&\approx\frac{\int_0^\infty tv\lambda_ap_ma_\tau\,\d\tau}{1-\e^{-Np_s}}\approx\frac{uv\lambda_a\lambda_sp_m\abs{\Delta_s+\Delta_a}}{p_s\Delta_a^2\Delta_s^2}\\ \label{limitapprox4}
+\tau_a&\approx\frac{\int_0^\infty t v\lambda_ap_ma_t \,\d t}{1-\e^{-Np_s}}\approx\frac{uv\lambda_a\lambda_sp_m\abs{\Delta_s+\Delta_a}}{p_s\Delta_a^2\Delta_s^2}\\ \label{limitapprox4} %change r2
 &=\frac{1}{\abs{\Delta_s}}+\frac{1}{\abs{\Delta_a}},
-\end{align}
+\end{align} % change r4
 where in the last line we used the fact that $1/p_s=N_a^*$ and \Cref{eq:N_a}.
 
-If a fraction $f$ of the cancer cells are aneuploid when the drug is administered then the rates at which the rescue mutations are generated can be written as:
+If a fraction $f$ of the cancer cells are aneuploid when the drug is administered then the expected number of successful mutants generated until time $t$ is given by: %change r6
 \begin{align*}
@@
 \end{align}
-The factor of $p_m$ in the second term of \cref{meanproliferationtime} is due to the fact that the lineage is conditioned to survive genetic drift and the time to reach $N$ is shorter then the scenario without this property.
+The factor of $p_m$ in the second term of \cref{meanproliferationtime} is due to the fact that the lineage is conditioned to survive genetic drift and the time to reach $N$ is shorter then the scenario without this property \citep{orr2014population,smith1974hitch}.
+% change 35
 
@@
 We obtain $\tau_a^r$ such that $m\left(\tau_a^r\right)=N$ by solving:
-\begin{equation}%\label{eq:t2det}
+\begin{equation}\label{eq:t2det2024}
 1=\frac{uv\lambda_a\lambda_s}{\Delta_s-\Delta_a}\left[\frac{\e^{\Delta_s\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_s-\Delta_m}-\frac{\e^{\Delta_a\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_a-\Delta_m}\right]+v\lambda_s\frac{\e^{\Delta_s\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_s-\Delta_m}.
@@
 \end{equation}
-We observe that the terms given by the evolutionary trajectory $sensitive \rightarrow aneuploid \rightarrow mutant$ do not contribute to the above approximation and, as a result, we deduce that it accurate only for $N\gg N_m^*>N_a^*$.
+We observe that the terms given by the evolutionary trajectory $sensitive \rightarrow aneuploid \rightarrow mutant$ do not contribute to the above approximation and, as a result, we deduce that it is accurate only for $N\gg N_m^*>N_a^*$.
+%change 9
 
@@
 \end{equation*}
-We plot \cref{eqmNs} and \cref{eqmNd} in Figure \Cref{cdffig}B and compare with stochastic simulations and observe that our approximation are in agreement.
+We plot \cref{eqmNs} and \cref{eqmNd} in \Cref{cdffig}B and compare with stochastic simulations and observe that our approximations are in agreement. %change 10, change 11
 
@@
 
-From  \cref{eqmNd} we obtain the distribution of the recurrence time conditional of evolutionary rescue:
+From  \cref{eqmNs} we obtain the distribution of the recurrence time conditional of evolutionary rescue: %change 12
 \begin{align}\label{distribution}
@@
 \caption{\textbf{Evolutionary rescue time.}
-Shown is the mean time for appearance of a resistance mutation the leads to evolutionary rescue (A) with aneuploidy ($u>0$) and (B) without aneuploidy ($u=0$).
-Our inhomogeneous Poisson-process approximations (solid black lines, right: \cref{eq:meantime1}, left: \cref{meantimet2}) is in agreement with simulation results (red markers with 95\% quantile intervals obtained with bootstrapping, see see Appendix G).
-Parameters: $\lambda_s=0.1,\lambda_m=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.
+Shown is the mean time for appearance of a resistance mutation that leads to evolutionary rescue (A) with aneuploidy ($u>0$) and (B) without aneuploidy ($u=0$).
+Our inhomogeneous Poisson-process approximations (solid black lines, right: \cref{eq:meantime1}, left: \cref{meantimet2}) are in agreement with simulation results (red markers with 95\% quantile intervals obtained with bootstrapping, see see Appendix G).
+Parameters: $\lambda_s=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$. %change 13
 }
@@
 \includegraphics[width=1\textwidth]{Figures/ExpectedNumber.pdf}
-\caption{\textbf{Aneuploidy increases the number of mutations which rescue the tumor.} Shown is the expected number of mutation, which will rescue the cancer cell population, produced through the evolutionary trajectory $sensitive \rightarrow mutant$ (blue line, \cref{eq:twosteplineagedirect}) or through the trajectory $sensitive \rightarrow aneuploid \rightarrow mutant$ (red line, \cref{eq:twosteplineage}).
+\caption{\textbf{Aneuploidy increases the number of mutations which rescue the tumor.} Shown is the expected number of mutations, which will rescue the cancer cell population, produced through the evolutionary trajectory $sensitive \rightarrow mutant$ (blue line, \cref{eq:twosteplineagedirect}) or through the trajectory $sensitive \rightarrow aneuploid \rightarrow mutant$ (red line, \cref{eq:twosteplineage}).
 Dashed vertical red line represents the threshold tumor size above which evolutionary rescue is very likely through aneuploidy \cref{eq:N_a} and the dashed vertical blue line represents the threshold tumor size above which evolutionary rescue is very likely through direct mutation \cref{eq:N_m}.
-Parameters: $\lambda_s=0.1,\lambda_m=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}
+Parameters: $\lambda_s=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.} %change 14
 \label{ExpectedNumberRescueLineages}
@@
 The dots represent simulation results and the error bars represent $95\%$ confidence intervals ($p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the fraction of simulations in which the tumor has adapted to the stress and $n=100$ is the number of simulations).
-Parameters: $\lambda_s=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}
+Parameters: $\lambda_s=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.} %change 15
 \label{rescue_prob_sgv}
@@
 \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/EvolutionaryRescueTimeComplete.pdf}
-\caption{Shown is the mean time for appearance of a resistance mutation the leads to evolutionary rescue with aneuploidy ($u>0$).
-Our inhomogeneous Poisson-process approximations (solid black lines, right: \cref{meantimet2}) is in agreement with simulation results (red markers with 95\% confidence intervals obtained with bootstrapping, see Appendix G).
+\includegraphics[width=1\textwidth]{Figures/EvolutionaryRescueTimeComplete.pdf} %change r5
+\caption{Shown is the mean time for appearance of a resistance mutation that leads to evolutionary rescue with aneuploidy ($u>0$).
+Our inhomogeneous Poisson-process approximations (solid black lines: \cref{meantimet2}) are in agreement with simulation results (red markers with 95\% confidence intervals obtained with bootstrapping, see Appendix G).
 Dashed vertical blue line represents the threshold tumor size above which evolutionary rescue is very likely through aneuploidy \cref{eq:N_a} and the dashed vertical green line represents the threshold tumor size above which evolutionary rescue is very likely through direct mutation \cref{eq:N_m}.
-Solid lines represents the approximations \cref{eq:AsymptoticTimeRules} ($N<N_a^*$ blue line and $N>N_m^*$ green line).
-Parameters: $\lambda_s=0.1,\lambda_m=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}
+Solid lines represent the approximations \cref{eq:AsymptoticTimeRules} ($N<N_a^*$ blue line and $N>N_m^*$ green line).
+Parameters: $\lambda_s=0.1,\lambda_m=0.0899,\lambda_a=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.} %change 16
 \label{EvolutionaryRescueTimeComplete}
@@
 \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/ProliferationTimeLarge.pdf}
+\includegraphics[width=1\textwidth]{Figures/ProliferationTimeLarge.pdf} %change r5
 \caption{The mean time for the mutant cell population to reach size $N$, where $N$ is the initial number of cancer cells.
@@
 \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/RecurrencePlot.pdf}
+\includegraphics[width=1\textwidth]{Figures/RecurrencePlot.pdf} %change r5
 \caption{The mean time for the mutant cell population to reach size $M$, where $M$ is the tumor detection size.
@@
 \vspace*{1\baselineskip}
-\includegraphics[width=1\textwidth]{Figures/SGVEvolutionaryRescueTimeComplete.pdf}
-\caption{Shown is the mean time for appearance of a resistance mutation the leads to evolutionary rescue with aneuploidy ($u>0$) when a fraction $f$ of cancer cells are aneuploid at the start of therapy.
+\includegraphics[width=1\textwidth]{Figures/SGVEvolutionaryRescueTimeComplete.pdf} %change r5
+\caption{Shown is the mean time for appearance of a resistance mutation that leads to evolutionary rescue with aneuploidy ($u>0$) when a fraction $f$ of cancer cells are aneuploid at the start of therapy.
 Black lines represent our inhomogeneous Poisson-process approximations (solid black line, \cref{meantimet2}; dashed black line \cref{meantimet2SGV}).
@@
 Solid lines represents the approximations \cref{eq:AsymptoticTimeRules} ($N<N_a^*$ blue line and $N>N_m^*$ green line).
-Parameters: $\lambda_s=0.1,\lambda_m=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7},f=0.14\%$.}
+Parameters: $\lambda_s=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7},f=0.14\%$.} %change 17
 \label{SGVEvolutionaryRescueTimeComplete}

### 2025-01-08 EvolutionaryRescue1.tex
changed align to equation + aligned
apparently align doesn't play well with line numbering

@@
 The following describes the events that may occur (right column), the rates at which they occur (middle column), and the effect these events have on the population (left column, see \Cref{figureAneuploidy}):
-
-\begin{align*}
+\[
+\begin{aligned}
 (+1,0,0)&:\quad \lambda_ss_t\left(1-u-v\right)\quad\left(\text{birth of sensitive cell}\right),\\
@@
 (0,0,-1)&:\quad \mu_mm_t\quad\left(\text{death of mutant cell}\right).
-\end{align*}
-For the remaining of this paper, we assume that the division rates of sensitive and aneuploid cells can be written as $\lambda_ss_t\left(1-u-v\right)\approx \lambda_ss_t$ and $\lambda_aa_t\left(1-v\right)\approx\lambda_aa_t$ because $u,v\ll1$ (\Cref{table1}).
+\end{aligned}
+\]
+
+For the remainder of this paper, we assume that the division rates of sensitive and aneuploid cells can be written as $\lambda_ss_t\left(1-u-v\right)\approx \lambda_ss_t$ and $\lambda_aa_t\left(1-v\right)\approx\lambda_aa_t$ because $u,v\ll1$ (\Cref{table1}).
 Each iteration of the simulation loop starts by computing the rates $\nu_k$ of each event $k$.
@@
 To test this assumption, we simulate a logistic growth model, with division and death rates given by
-\begin{align*}
+\[
+\begin{aligned}
 \lambda_s' &= \lambda_s , \\
@@
 \mu_m' &= \mu_m + \lambda_m\frac{w+a+m}{K} ,
-\end{align*}
+\end{aligned}
+\]
 where $K$ is the tumor carrying capacity. 

### 2025-01-08 EvolutionaryRescue1.tex
minor typography fixes

@@
 \newcommand{\psgv}{p_\text{SGV}}
+\newcommand{\bigo}[1]{\mathcal{O}\left(#1\right)}
 
@@
 We first calculate the expected time for the appearance of the first mutant that rescues the cell population.
-This can occur either through the evolutionary trajectory $sensitive \rightarrow mutant$ or through the trajectory $sensitive \rightarrow aneuploid \rightarrow mutant$.
+This can occur either through the evolutionary trajectory \textit{sensitive}$\rightarrow$\textit{mutant} or through the trajectory \textit{sensitive}$\rightarrow$\textit{aneuploid}$\rightarrow$\textit{mutant}.
 We start with the former.
@@
 \end{equation}
-by integrating the exponential and because $\frac{\exp[\Delta_s t]-1}{\Delta_s}=\frac{1+\Delta_s t+\mathcal{O}\left(\left(\Delta_st\right)^2\right)-1}{\Delta_s}=t+O(\Delta_st^2)$. %change 22
+by integrating the exponential and because $\frac{1}{\Delta_s} \left(\exp[\Delta_s t]-1\right) = t+ \bigo{\Delta_s t^2}$. %change 22
 The probability density function of $T_m$ is thus
@@
 Therefore, the probability density function of the conditional random variable $(T_m \mid T_m < \infty)$ is
-$f_m(t) = \frac{R_m\left(t\right)\exp\left(-\int_0^{t}{R_m(z)\d z}\right)}{\presc}$. 
+$f_m(t) = R_m \left(t\right)\exp\left(-\int_0^{t}{R_m(z)\d z}\right) / \presc$. 
 \\
@@
 \int_{0}^{\infty}{t f_m(t) \d t} = 
-\frac{\int_{0}^{\infty}{tR_m(t)\exp\left(-\int_0^{t}{R_m(z)\d z}\right) \d t}}{p_{rescue}},
+\frac{\int_{0}^{\infty}{tR_m(t)\exp\left(-\int_0^{t}{R_m(z)\d z}\right) \d t}}{\presc},
 \end{aligned}
@@
 \begin{align}\nonumber
-\tau_a &= \frac{\int_{0}^{\infty}{tR_a(t)\exp\left(-\int_0^{t}{R_a(z)\d z}\right) \d t}}{p_{rescue}}\\ \label{meantimet2}
+\tau_a &= \frac{\int_{0}^{\infty}{tR_a(t)\exp\left(-\int_0^{t}{R_a(z)\d z}\right) \d t}}{\presc}\\ \label{meantimet2}
 &=
@@
 Let $N_a^*$ and $N_m^*$ be the threshold population size above which evolutionary rescue through aneuploidy or direct mutation, respectively, is very likely.
-If $N\gg N_m^*$ then $p_{rescue}\approx1$ and the mean time $\tau_a$ can be written as: % change 38
+If $N\gg N_m^*$ then $\presc\approx1$ and the mean time $\tau_a$ can be written as: % change 38
 \begin{align*}
@@
 \begin{align}\label{meantimet2SGV}
-\tilde{\tau}_a&= \frac{\int_{0}^{\infty}{tR_a^f(t)\exp\left(-\int_0^{t}{R_a^f(z)\d z}\right) \d t}}{p_{rescue}},
+\tilde{\tau}_a&= \frac{\int_{0}^{\infty}{tR_a^f(t)\exp\left(-\int_0^{t}{R_a^f(z)\d z}\right) \d t}}{\presc},
 \end{align}
-where $R_a^f(t)=r_1^f\left(t\right)+r_2^f\left(t\right)$ and $p_{rescue}=1-\exp\left[-\left(1-f\right)p_sN-fp_aN\right]$.
+where $R_a^f(t)=r_1^f\left(t\right)+r_2^f\left(t\right)$ and $\presc = 1-\exp\left[-\left(1-f\right)p_sN-fp_aN\right]$.
 We plot our approximation in  \Cref{SGVEvolutionaryRescueTimeComplete} together with simulated data.
@@
 \begin{align}\label{distribution}
-f\left(t\right)&=\frac{d}{dt}\left[\frac{P\left(m_t\geq N\right)}{p_{rescue}}\right]=r'\left(t\right)\frac{\exp\left[-r\left(t\right)\right]}{p_{rescue}},
+f\left(t\right)&=\frac{d}{dt}\left[\frac{P\left(m_t\geq N\right)}{\presc}\right]=r'\left(t\right)\frac{\exp\left[-r\left(t\right)\right]}{\presc},
 \end{align}
@@
 
-For the threshold tumor sizes the $95\%$ confidence interval is obtained through bootstrapping in the following steps: (1) we simulate $p_{rescue}$ 100 times; (2) we sample with replacement which we store in $S$; (3) for each element of this sample we obtain $N_a^*=1/p_s$ using $p_s=-1/N_e\log\left(1-\bar{S}\right)$ where $\bar{S}$ is the mean of $S$ and $N_e$ is an arbitrary value of the initial population size we selected in order to calculate $p_{rescue}$; (4) we repeat steps (2)-(3) 100 times to obtain $N_a^*$ and we select the upper and lower limits such that $95\%$ of the values of $N_a^*$ lie in the interval given by the bounds.
-
-For the ratio of the threshold tumor sizes the $95\%$ confidence interval is obtained through bootstrapping in the following steps: (1) we simulate $p_{rescue}$ 100 times for both the scenario when $f=\tilde{u}\lambda_s/c$ and $f=0$; (2) we sample with replacement which we store in $S_f$ and  $S_0$; (4) for each element of $S_0$ we obtain $N_a^*=1/p_s$ using $p_s=-1/N_e\log\left(1-\bar{S}\right)$ where $\bar{S}$ is the mean of $S_0$ and $N_e$ is an arbitrary value of the initial population size we selected in order to calculate $p_{rescue}$; (5) for each element of $S_f$ we obtain $\tilde{N}_a^*=1/p_a$ using $p_a=-f/N_e\log\left(1-\bar{S}\right)$ where $\bar{S}_f$ is the mean of $S_f$ and $N_e$ is an arbitrary value of the initial population size we selected in order to calculate $p_{rescue}$; (6) we repeat steps (2)-(5) 100 times to obtain $\tilde{N}_a^*/N_a^*$ and we select the upper and lower limits such that $95\%$ of the values of $\tilde{N}_a^*/N_a^*$ lie in the interval given by the bounds.
+For the threshold tumor sizes the $95\%$ confidence interval is obtained through bootstrapping in the following steps: (1) we simulate $\presc$ 100 times; (2) we sample with replacement which we store in $S$; (3) for each element of this sample we obtain $N_a^*=1/p_s$ using $p_s=-1/N_e\log\left(1-\bar{S}\right)$ where $\bar{S}$ is the mean of $S$ and $N_e$ is an arbitrary value of the initial population size we selected in order to calculate $\presc$; (4) we repeat steps (2)-(3) 100 times to obtain $N_a^*$ and we select the upper and lower limits such that $95\%$ of the values of $N_a^*$ lie in the interval given by the bounds.
+
+For the ratio of the threshold tumor sizes the $95\%$ confidence interval is obtained through bootstrapping in the following steps: (1) we simulate $\presc$ 100 times for both the scenario when $f=\tilde{u}\lambda_s/c$ and $f=0$; (2) we sample with replacement which we store in $S_f$ and  $S_0$; (4) for each element of $S_0$ we obtain $N_a^*=1/p_s$ using $p_s=-1/N_e\log\left(1-\bar{S}\right)$ where $\bar{S}$ is the mean of $S_0$ and $N_e$ is an arbitrary value of the initial population size we selected in order to calculate $\presc$; (5) for each element of $S_f$ we obtain $\tilde{N}_a^*=1/p_a$ using $p_a=-f/N_e\log\left(1-\bar{S}\right)$ where $\bar{S}_f$ is the mean of $S_f$ and $N_e$ is an arbitrary value of the initial population size we selected in order to calculate $\presc$; (6) we repeat steps (2)-(5) 100 times to obtain $\tilde{N}_a^*/N_a^*$ and we select the upper and lower limits such that $95\%$ of the values of $\tilde{N}_a^*/N_a^*$ lie in the interval given by the bounds.
 %%%

### 2025-01-08 EvolutionaryRescue1.tex
minor formatting fixes

@@
 % line numbers
-\usepackage[displaymath, mathlines]{lineno}
+\usepackage[mathlines]{lineno}
 \renewcommand\linenumberfont{\normalfont\small\sffamily}
@@
 and the expected number of successful mutants created by direct mutation and via aneuploidy until time $t$ is given by:
-\begin{align}\label{eq:twosteplineage}
-M_a\left(t\right)&=v\lambda_ap_m\int_0^ta_{z} d z = \frac{uv\lambda_s\lambda_aNp_m}{\Delta_s-\Delta_a}\left(\frac{\e^{\Delta_st}-1}{\Delta_s}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right),\\ 
-M_m\left(t\right)&=v\lambda_sp_m\int_0^ts_{z} d z = v\lambda_sNp_m\frac{\e^{\Delta_s t}-1}{\Delta_s}.
+\begin{align}
+M_a\left(t\right)&=v\lambda_ap_m\int_0^ta_{z} d z = \frac{uv\lambda_s\lambda_aNp_m}{\Delta_s-\Delta_a}\left(\frac{\e^{\Delta_st}-1}{\Delta_s}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right),\label{eq:twosteplineage}\\
+M_m\left(t\right)&=v\lambda_sp_m\int_0^ts_{z} d z = v\lambda_sNp_m\frac{\e^{\Delta_s t}-1}{\Delta_s}.\label{eq:twosteplineagedirect}
 \end{align} %change r3
@@
 \includegraphics[width=1\textwidth]{Figures/ExpectedNumber.pdf}
-\caption{\textbf{Aneuploidy increases the number of mutations which rescue the tumor.} Shown is the expected number of mutations, which will rescue the cancer cell population, produced through the evolutionary trajectory $sensitive \rightarrow mutant$ (blue line, \cref{eq:twosteplineagedirect}) or through the trajectory $sensitive \rightarrow aneuploid \rightarrow mutant$ (red line, \cref{eq:twosteplineage}).
+\caption{\textbf{Aneuploidy increases the number of mutations which rescue the tumor.} Shown is the expected number of mutations, which will rescue the cancer cell population, produced through the evolutionary trajectory \textit{sensitive} $\rightarrow$ \textit{mutant} (blue line, \cref{eq:twosteplineagedirect}) or through the trajectory \textit{sensitive} $\rightarrow$ \textit{aneuploid} $\rightarrow$ \textit{mutant} (red line, \cref{eq:twosteplineage}).
 Dashed vertical red line represents the threshold tumor size above which evolutionary rescue is very likely through aneuploidy \cref{eq:N_a} and the dashed vertical blue line represents the threshold tumor size above which evolutionary rescue is very likely through direct mutation \cref{eq:N_m}.

### 2025-01-08 EvolutionaryRescue1.tex
added citation to orr & unckless for recurrence time

@@
 Here, we study evolutionary rescue after a sudden environmental change caused by initiating anti-cancer drug treatment. 
+Essentially, this consists of combining known results from the evolutionary rescue literature, focusing on the quantities and parameter values that are relevant for cancer.
+For completeness, however, we re-derive all results within a single model.
 We consider a range of effects of aneuploidy, from tolerance to (partial) resistance to the drug \citep{brauner2016distinguishing} (we note that we use tolerant to refer to aneuploids which have negative growth rates, which is different from other references in the literature such as \citet{berman2020drug}).
@@
   ~\Cref{proliferationFigure,ProliferationTimeLarge} show the agreement between our approximations and simulation results.
-For small tumors ($N \ll N_a^*$), the mean recurrence time can be approximated as the sum of the mean time for the first rescue mutation to appear and the mean time for its lineage to reach size $N$.
-The mean recurrence time grows logarithmically with tumor size $N$ and is the same order of magnitude as the mean evolutionary rescue time.
+For small tumors ($N \ll N_a^*$), the mean recurrence time can be approximated as the sum of the mean time for the first rescue mutation to appear ($\tau_a$) and the additional mean time for its lineage to reach size $N$.
+This additional time is the equivalent of \citet{orr2014population}'s ``$t_\text{return}$'' (their eq.~23). 
+It grows logarithmically with tumor size $N$ and may be the same order of magnitude as the mean evolutionary rescue time. 
 Increasing the mutant growth rate, $\Delta_m$, decreases the recurrence time, while increasing the sensitive and aneuploid growth rates, $\Delta_s$ and $\Delta_a$, respectively, increases the recurrence time.
+
 For large tumors ($N \gg N_m^*$), the dynamics of the number of mutant cells is deterministic, and the mean recurrence time becomes independent of the initial tumor size $N$.
@@
 
-We have modeled a tumor--a population of cancer cells--exposed to drug treatment that causes it to decline in size toward potential extinction.
+We have modeled a tumor---a population of cancer cells---exposed to drug treatment that causes it to decline in size toward potential extinction.
 In this scenario, the tumor can be ``evolutionary rescued'' or escape extinction via two paths.

### 2025-01-08 EvolutionaryRescue1.tex
cut final para before future directions

@@
 Essentially, this consists of combining known results from the evolutionary rescue literature, focusing on the quantities and parameter values that are relevant for cancer.
-For completeness, however, we re-derive all results within a single model.
+Because these results are scattered across multiple papers, and because they all follow easily from simple branching process approximations, we re-derive all results here within our single cancer model.
 We consider a range of effects of aneuploidy, from tolerance to (partial) resistance to the drug \citep{brauner2016distinguishing} (we note that we use tolerant to refer to aneuploids which have negative growth rates, which is different from other references in the literature such as \citet{berman2020drug}).
@@
 Additionally, we derive the distribution of the recurrence time for a small tumor with $N=10^6$ cells, noting that the distribution is wide and right-skewed (\Cref{KMdistribution}).
-It is highly unlikely to observe the recurrence of cancer at times smaller than $\frac{1}{\Delta_m}\log\frac{\Delta_m-\Delta_s}{v\lambda_s}\approx1542$ days for the parameter values in \Cref{table1} with $\lambda_a=0.0899$, $\mu_s=0.14$, and $v=10^{-7}$ and independent of initial tumor size $N$~(\Cref{cdffig}B).
+It is highly unlikely to observe the recurrence of cancer at times smaller than $\frac{1}{\Delta_m}\log\frac{\Delta_m-\Delta_s}{v\lambda_s}\approx 1500$ days for the parameter values in \Cref{table1} with $\lambda_a=0.0899$, $\mu_s=0.14$, and $v=10^{-7}$ and independent of initial tumor size $N$~(\Cref{cdffig}B).
 
@@
 We found that a drug that reduces the sensitive growth rate and does not significantly increase the chromosome missegregation rate will likely lead to evolutionary rescue through standing genetic variation (\Cref{rescue_denovo,eq:Nsgv}).
-Furthermore, if the fraction of tumor cells that have the beneficial aneuploidy is $f\gg\frac{u\lambda_s}{\abs{\Delta_s}}\approx2.5\%$, then evolutionary rescue is more likely to occur via standing variation rather than through \emph{de-novo} aneuploidy.
+Furthermore, if the fraction of tumor cells that have the beneficial aneuploidy is $f \gg u \lambda_s / \abs{\Delta_s} \approx2.5\%$, then evolutionary rescue is more likely to occur via standing variation rather than through \emph{de-novo} aneuploidy.
 However, for the parameter values we focus on in our examples (\Cref{table1}), this fraction is an order of magnitude lower, and therefore, we expect evolutionary rescue to occur by \emph{de-novo} aneuploidy.
 
-Finally, we observe from \Cref{rescue_threshold} that only for a restricted region of the parameter space will aneuploidy act as a ``stepping stone'' for evolutionary rescue.
+%Finally, we observe from \Cref{rescue_threshold} that only for a restricted region of the parameter space will aneuploidy act as a ``stepping stone'' for evolutionary rescue.
 If the aneuploid division rate is smaller than $\frac{\mu_a}{1+u}\approx\mu_a(1-u)$, then evolutionary rescue will occur through direct mutation.
 As a result, for most parameter values, aneuploidy will either not play any role in evolutionary rescue or will be the main driver of adaptation without requiring any mutation (i.e., evolutionary rescue in one step).
+% NOTE FROM DANIEL: I cut this paragraph because I think it's implicitly based on a uniform distribution of aneuploid growth rates.
+But the whole point of why we're doing this is because there are apparently some interesting aneuploidies that have growth rates near 0
 

### 2025-01-08 EvolutionaryRescue1.tex
rewrote evolutionary rescue intro

@@
 
+There has been extensive theoretical analysis of evolutionary rescue.
+\citet{orr2008population,orr2014population} thoroughly investigated the simplest case in which there is a specific single mutation or low-frequency variant that is sufficient to rescue the population.
+\citet{martin2013probability} extended these results to consider the effects of a range of rescue genotypes that can vary in both growth rate and variance in offspring number.
+\citep{uecker2016role} modeled a situation in which \textit{two} mutations at different loci (with recombination between them) are necessary for rescue. 
+\citet{osmond2020genetic} considered a population that must achieve rescue, potentially via multiple mutations, by adapting on a fitness landscape described by Fisher's geometric model.
+
+Resistance to cancer therapy is a natural application of evolutionary rescue theory \citep{alexander2014evolutionary}. 
+\citet{iwasa2003evolutionary} used multi-type branching process theory to approximate the probability that a population under strong selective pressure can survive extinction, focusing on the case in which the cancer must get two mutations to be rescued, with the single-mutant still dying rapidly.
+\citet{iwasa2004evolutionary} used a similar approach to find simple approximate expressions for the probability that a single lineage would evolve to survive via a wide range of possible mutational pathways.
+A very similar set of models describe adaptation to new environments (e.g., a pathogen evolving to cross a species barrier) and fitness valley crossing \citep{antia2003role, weissman2009rate}.
 \citet{gunnarsson2020understanding} analyzed a model where a tumor consisting of two populations of cancer cells, one drug-resistant and the other drug-sensitive, can evade extinction by cells switching between the two phenotypes through epigenetic mutations.
 They found that even when drug-resistant cells are barely viable, the epimutations guarantee evolutionary rescue.
-Evolutionary rescue in a single step, where an initially declining population has to acquire a mutation to survive extinction after a sudden environmental change, has been studied by \citet{orr2008population,orr2014population}.
-They analyzed a model where the mutant strain is present in small numbers at the onset of therapy.
-They concluded that this can significantly enhance the population's probability of survival.
-
-Evolutionary rescue may occur when a beneficial mutation appears in a single copy, leading to hard selective sweeps, or when multiple mutations appear independently in different individuals, leading to soft selective sweeps.
+
+%Evolutionary rescue may occur when a beneficial mutation appears in a single copy, leading to hard selective sweeps, or when multiple mutations appear independently in different individuals, leading to soft selective sweeps.
 \citet{wilson2017soft} have shown that evolutionary rescue is significantly enhanced by soft selective sweeps when multiple mutations contribute towards evolutionary rescue.
@@
 Evolutionary rescue that requires two successive mutations (i.e., two steps) has been investigated by \citet{martin2013probability}, who tested their model's predictions with data from yeast and bacteria experiments.
-\citet{iwasa2003evolutionary,iwasa2004evolutionary} used multi-type branching process theory to approximate the probability that a population under strong selective pressure can survive extinction via succesive mutations.
 %change 41
@@
 Essentially, this consists of combining known results from the evolutionary rescue literature, focusing on the quantities and parameter values that are relevant for cancer.
-Because these results are scattered across multiple papers, and because they all follow easily from simple branching process approximations, we re-derive all results here within our single cancer model.
+Because these results are scattered across multiple papers, and because they all follow easily from simple branching process approximations, we re-derive all results here within our single model.
+While our focus is on the application to cancer, our model is a general one, and in the general evolutionary rescue context, this work can be seen as a characterization of the circumstances in which two-step processes contribute to rescue even when single-step rescue mutations are available.
 We consider a range of effects of aneuploidy, from tolerance to (partial) resistance to the drug \citep{brauner2016distinguishing} (we note that we use tolerant to refer to aneuploids which have negative growth rates, which is different from other references in the literature such as \citet{berman2020drug}).
@@
 
-In our analysis, we assume that cells from the initial population divide and die independently of each other.
+In most of our analysis, we assume that cells from the initial population divide and die independently of each other. 
 However, these cells will compete for resources. 
@@
 Define $p_s$ as the probability that a lineage starting from a single drug-sensitive cell avoids extinction by acquiring drug resistance.
-Thus, $N^*=1/p_s$ is the threshold tumor size above which evolutionary rescue is very likely, and the rescue probability is given by 
+Thus, $N^*=1/p_s$ is the threshold tumor size above which evolutionary rescue is very likely \citep{iwasa2003evolutionary}, and the rescue probability is given by 
 \begin{equation}\label{eq:rescue_prob} 

### 2025-01-09 EvolutionaryRescue1.tex
minor edits to DBW changes

@@
 \citet{martin2013probability} extended these results to consider the effects of a range of rescue genotypes that can vary in both growth rate and variance in offspring number.
-\citep{uecker2016role} modeled a situation in which \textit{two} mutations at different loci (with recombination between them) are necessary for rescue. 
+\citet{uecker2016role} modeled a situation in which \textit{two} mutations at different loci (with recombination between them) are necessary for rescue. 
 \citet{osmond2020genetic} considered a population that must achieve rescue, potentially via multiple mutations, by adapting on a fitness landscape described by Fisher's geometric model.
@@
 Essentially, this consists of combining known results from the evolutionary rescue literature, focusing on the quantities and parameter values that are relevant for cancer.
-Because these results are scattered across multiple papers, and because they all follow easily from simple branching process approximations, we re-derive all results here within our single model.
-While our focus is on the application to cancer, our model is a general one, and in the general evolutionary rescue context, this work can be seen as a characterization of the circumstances in which two-step processes contribute to rescue even when single-step rescue mutations are available.
-We consider a range of effects of aneuploidy, from tolerance to (partial) resistance to the drug \citep{brauner2016distinguishing} (we note that we use tolerant to refer to aneuploids which have negative growth rates, which is different from other references in the literature such as \citet{berman2020drug}).
+Because these results are scattered across multiple papers, and because they all follow from simple branching process approximations, we re-derive all results here for our model.
+While our focus is on the application of evolutionary rescue theory to evolution of drug resistance in cancer, our model is a general one, and in the general evolutionary rescue context, this work can be seen as a characterization of the circumstances in which two-step processes contribute to rescue even when single-step rescue mutations are available.
+We consider a range of effects of aneuploidy, from tolerance to (partial) resistance to the drug \citep{brauner2016distinguishing}; note that we use tolerant to refer to aneuploids which have negative growth rates, which is different from other references in the literature such as \citet{berman2020drug}.
 %change 47

### 2025-01-09 ResponseLetter.tex
added daniel's part to response

@@
 
+\textcolor{blue}{This is a great point. We have basically incorporated this comment into our Introduction.}
+
 $\#$ Major comments
@@
 
+\textcolor{blue}{We have added line XX on this in the Introduction: for readers interested more in general evolutionary rescue than in the application to cancer, we think that the primary novelty in the paper is the focus on comparing direct, single-step evolutionary rescue and rescue via an intermediate mutant genotype.}
+
 In case it helps, my understanding of multi-step rescue theory is that there was an early model for the emergence of epidemics via multiple mutations (Antia et al 2003 Nature, and some follow ups like Yates et al 2006 PRSB and Alexander $\&$ Day 2010 Interface), a very general treatment of emergence and rescue via multiple mutations (Iwasa et al 2004 JTB), a 2-locus 2-allele rescue model (Uecker $\&$ Hermisson 2016 Genetics), and multi-step rescue models on a fitness landscape (Martin et al 2013, Osmond et al 2020), all of which also use branching processes.
@@
 
-\textcolor{red}{
-TODO - Daniel
+\textcolor{blue}{
+We owe a serious debt of gratitude to the reviewer for this comment.
+They have basically done all the work for us.
+We have incorporated these references in the Introduction and throughout the Results at the equations mentioned.
+The only slight changes we have made are omitting the follow-ups to Antia et al 2003 and instead including Iwasa et al 2003 Proc B, which we really should have mentioned before and is arguably the closest thing in the literature to this manuscript.
 }

### 2025-01-12 EvolutionaryRescue1.tex
added daniel's part to response
fixed osmond ref

@@
 \citet{uecker2016role} modeled a situation in which \textit{two} mutations at different loci (with recombination between them) are necessary for rescue. 
-\citet{osmond2020genetic} considered a population that must achieve rescue, potentially via multiple mutations, by adapting on a fitness landscape described by Fisher's geometric model.
+\citet{osmond_genetic_2020} considered a population that must achieve rescue, potentially via multiple mutations, by adapting on a fitness landscape described by Fisher's geometric model.
 
@@
 \citet{wilson2017soft} have shown that evolutionary rescue is significantly enhanced by soft selective sweeps when multiple mutations contribute towards evolutionary rescue.
-Additionally, \citet{osmond2020genetic} have calculated the probability that rescue is soft and the expected number of rescue mutations in different scenarios.
+Additionally, \citet{osmond_genetic_2020} have calculated the probability that rescue is soft and the expected number of rescue mutations in different scenarios.
 Evolutionary rescue that requires two successive mutations (i.e., two steps) has been investigated by \citet{martin2013probability}, who tested their model's predictions with data from yeast and bacteria experiments.
@@
 \end{equation}
-Note that that \cref{eq:N_a} is an approximation of the two step-evolutionary rescue probability derived by \citet{osmond_genetic_2020} with Dirac delta fitness distribution, which is also similar to eq.~A3 in \citep{iwasa2004evolutionary}.
+\cref{eq:N_a} is basically our model's equivalent of \citet{iwasa2004evolutionary}'s eq.~A3, or eq.~8 of \citet{osmond_genetic_2020} with the distribution of fitness effects set to a delta function.
 %change 57
@@
 We have modeled a tumor---a population of cancer cells---exposed to drug treatment that causes it to decline in size toward potential extinction.
-In this scenario, the tumor can be ``evolutionary rescued'' or escape extinction via two paths.
+In this scenario, the tumor can be ``evolutionarily rescued'' or escape extinction via two paths.
 In the direct path, a drug-sensitive cell acquires a mutation or aneuploidy that confers resistance and allows it to grow rapidly.

### 2025-01-12 EvolutionaryRescue1.tex
copy-editing

@@
 \citet{uecker2016role} modeled a situation in which \textit{two} mutations at different loci (with recombination between them) are necessary for rescue. 
-\citet{osmond_genetic_2020} considered a population that must achieve rescue, potentially via multiple mutations, by adapting on a fitness landscape described by Fisher's geometric model.
+\citet{osmond_genetic_2020} considered a population that must achieve rescue, potentially via multiple mutations, by adapting on a fitness landscape described by Fisher's geometric model,
+and found that for some parameter values rescue by two mutational steps is more likely than rescue by a single step.
 
@@
 Because these results are scattered across multiple papers, and because they all follow from simple branching process approximations, we re-derive all results here for our model.
-While our focus is on the application of evolutionary rescue theory to evolution of drug resistance in cancer, our model is a general one, and in the general evolutionary rescue context, this work can be seen as a characterization of the circumstances in which two-step processes contribute to rescue even when single-step rescue mutations are available.
+While our focus is on the application of evolutionary rescue theory to evolution of drug resistance in cancer, our model is a general one, and in the general evolutionary rescue context, this work can be seen as a characterization of the circumstances in which two-step processes contribute to rescue even when single-step rescue mutations are available, similar to part of \citet{osmond_genetic_2020} but focused on a different fitness landscape.
+
 We consider a range of effects of aneuploidy, from tolerance to (partial) resistance to the drug \citep{brauner2016distinguishing}; note that we use tolerant to refer to aneuploids which have negative growth rates, which is different from other references in the literature such as \citet{berman2020drug}.
@@
 To parametrize the simulations, we assume that the cells under consideration are melanoma cells and rely on \citet{rew2000cell} and \citet{bozic2013evolutionary} for the division and death rates, respectively.
-\citet{rew2000cell} report \emph{in vivo} measurements of the potential doubling times (the waiting time for the number of cells in the tumor to double, disregarding cell death) for a large set of cancer types.
+\citet{rew2000cell} report \textit{in vivo} measurements of the potential doubling times (the waiting time for the number of cells in the tumor to double, disregarding cell death) for a large set of cancer types.
 The division rate is obtained as $\lambda=\log{2} / T \approx 0.1$ per day.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\paragraph*{Standing vs. \emph{de-novo} genetic variation.}
+\paragraph*{Standing vs. \textit{de novo} genetic variation.}
 
@@
 
-Therefore, the threshold tumor size for rescue by standing generation variation, $\tilde{N}^*_{a}$, is similar to the threshold for rescue by de-novo variation, $N^*_a$, except that the sensitive growth rate $\abs{\Delta_s}$ is replaced by the cost of aneuploidy $c$, such that
+Therefore, the threshold tumor size for rescue by standing generation variation, $\tilde{N}^*_{a}$, is similar to the threshold for rescue by \textit{de novo} variation, $N^*_a$, except that the sensitive growth rate $\abs{\Delta_s}$ is replaced by the cost of aneuploidy $c$, such that
 \begin{equation}\label{eq:Nsgv}
@@
 Standing genetic variation will drive evolutionary rescue if sensitive cells die rapidly (growth rate $\Delta_s$ is very negative) due to a strong effect of the drug on sensitive cells or if the cost of aneuploidy in the drug-free environment, $c$, is small.
-In contrast, \emph{de-novo} aneuploid cells will have a greater contribution to rescue if the cost of aneuploidy, $c$, is large, the effect of the drug on sensitive cells is weak ($\Delta_s$ is close to zero), or if the drug induces the appearance of aneuploid cells ($u > \tilde u$).
+In contrast, \textit{de novo} aneuploid cells will have a greater contribution to rescue if the cost of aneuploidy, $c$, is large, the effect of the drug on sensitive cells is weak ($\Delta_s$ is close to zero), or if the drug induces the appearance of aneuploid cells ($u > \tilde u$).
 For example, with  $\lambda_s=0.1$, $\mu_s=0.14$, $u=10^{-2}$, $\tilde{u}=10^{-3}$, and $c=0.07$, the ratio of the threshold tumor sizes for standing vs.
-\emph{de-novo} variation is $\tilde{N}^*_a/N^*_a \approx 17.5$, which means that \emph{de-novo} genetic variation is  the main driver of evolutionary rescue.
+\textit{de novo} variation is $\tilde{N}^*_a/N^*_a \approx 17.5$, which means that \textit{de novo} genetic variation is  the main driver of evolutionary rescue.
 
@@
 We found that a drug that reduces the sensitive growth rate and does not significantly increase the chromosome missegregation rate will likely lead to evolutionary rescue through standing genetic variation (\Cref{rescue_denovo,eq:Nsgv}).
-Furthermore, if the fraction of tumor cells that have the beneficial aneuploidy is $f \gg u \lambda_s / \abs{\Delta_s} \approx2.5\%$, then evolutionary rescue is more likely to occur via standing variation rather than through \emph{de-novo} aneuploidy.
-However, for the parameter values we focus on in our examples (\Cref{table1}), this fraction is an order of magnitude lower, and therefore, we expect evolutionary rescue to occur by \emph{de-novo} aneuploidy.
+Furthermore, if the fraction of tumor cells that have the beneficial aneuploidy is $f \gg u \lambda_s / \abs{\Delta_s} \approx2.5\%$, then evolutionary rescue is more likely to occur via standing variation rather than through \textit{de novo} aneuploidy.
+However, for the parameter values we focus on in our examples (\Cref{table1}), this fraction is an order of magnitude lower, and therefore, we expect evolutionary rescue to occur by \textit{de novo} aneuploidy.
 
@@
 \textbf{Standing genetic variation facilitates the evolutionary rescue of cancer.}
-(A)  Ratio of threshold tumor sizes for rescue by standing genetic variation and by de-novo variation, $\tilde{N}_a^*/N_a^*$, when a fraction $\frac{\tilde{u}\lambda_s}{c}$ is aneuploid at the start of treatment, as a function of the sensitive growth rate $\Delta_s$.
+(A)  Ratio of threshold tumor sizes for rescue by standing genetic variation and by \textit{de novo} variation, $\tilde{N}_a^*/N_a^*$, when a fraction $\frac{\tilde{u}\lambda_s}{c}$ is aneuploid at the start of treatment, as a function of the sensitive growth rate $\Delta_s$.
 Standing genetic variation will drive adaptation to the drug if the sensitive population is rapidly declining ($\Delta_s\ll0$) due to a stronger effect of the drug on sensitive cells.
@@
 (B) Ratio of threshold tumor size $\tilde{N}_a^*$, when a fraction $\frac{\tilde{u}\lambda_s}{c}$ is aneuploid at the start of treatment, and $N_a^*$ as a function of the ratio of aneuploidy rates $\tilde{u}/u$.
-\emph{De-novo} aneuploids will have a larger contribution to the appearance of drug resistance if the drug induces the appearance of aneuploid cells ($u \gg \tilde u$).
+\textit{De novo} aneuploids will have a larger contribution to the appearance of drug resistance if the drug induces the appearance of aneuploid cells ($u \gg \tilde u$).
 Red dots represent simulation results, and the error bars represent the $95\%$ confidence intervals obtained with bootstrap (Appendix G).

### 2025-01-12 EvolutionaryRescue1.tex
added kendall / birth-death ref

@@
 Instead, it acts as a ``stepping stone'' through which the resistant mutant can appear more rapidly, given that the number of aneuploid cells declines slower than the number of drug-sensitive cells (\Cref{sampleTrajectories}).
-In this scenario, aneuploidy provides two advantages.
-First, it delays the extinction of the population, providing more time for the appearance of the resistance mutation.
+In this scenario, aneuploidy provides two advantages. First, it delays the extinction of the population, providing more time for the appearance of the resistance mutation. 
 Second, it increases the population size relative to a drug-sensitive population, providing more cells in which mutations can occur. 
-Together, this increases the cumulative number of mutants that arise (i.e., $Nuv\lambda_s\lambda_a/\abs{\Delta_s\Delta_a}$).
-% change 39
+Together, this increases the cumulative number of mutants that arise (i.e., $N u v \lambda_s \lambda_a / \abs{\Delta_s \Delta_a}$). % change 39
 
 We find that aneuploidy can significantly affect evolutionary rescue as it reduces the threshold tumor size by at least an order of magnitude even when aneuploidy only provides tolerance (\Cref{rescue_prob}).
-When the number of cells in the tumor is large enough (i.e., $N\gg N_m^*\approx 4\times10^7$), a resistance mutation will occur in drug-sensitive cells before these cells become extinct (\Cref{rescue_prob}).
+When the number of cells in the tumor is large enough (i.e., $N \gg N_m^* \approx 4\times10^7$), a resistance mutation will occur in drug-sensitive cells before these cells become extinct (\Cref{rescue_prob}).
 Therefore, large tumors are likely to be rescued with or without aneuploidy.
@@
 However, for the parameter values we focus on in our examples (\Cref{table1}), this fraction is an order of magnitude lower, and therefore, we expect evolutionary rescue to occur by \textit{de novo} aneuploidy.
+
+Comparing our results to the evolutionary rescue literature,
 
@@
 
-To analyze evolutionary rescue in our model, we use the framework of \emph{multi-type branching processes} \citep{harris1963theory, weissman2009rate}. % change 31
+To analyze evolutionary rescue in our model, we use the framework of \emph{multi-type branching processes} \citep{harris1963theory, weissman2009rate}. 
+In the case of cancer, where all reproduction is binary fission, these branching processes are also \emph{birth-death processes} \citep{kendall1948}% change 31
 This allows us to find explicit expressions for the \emph{survival probability}: the probability that a lineage descended from a single cell does not become extinct.

### 2025-01-12 EvolutionaryRescue1.tex
added evo rescue refs to discussion

@@
 \end{equation}
-\cref{eq:N_a} is basically our model's equivalent of \citet{iwasa2004evolutionary}'s eq.~A3, or eq.~8 of \citet{osmond_genetic_2020} with the distribution of fitness effects set to a delta function.
+\Cref{eq:N_a} is our model's equivalent of \citet{iwasa2004evolutionary}'s eq.~A3, or eq.~8 of \citet{osmond_genetic_2020} with the distribution of fitness effects set to a delta function;
+our ``tolerant'', ``stationary'', and ``resistant'' correspond to \citet{osmond_genetic_2020}'s ``sufficiently subcritical'', ``sufficiently critical'', and ``sufficiently supercritical'', respectively.
 %change 57
@@
 
-Comparing our results to the evolutionary rescue literature,
+Comparing our results to the evolutionary rescue literature, we complement \citet{iwasa2003evolutionary} and \citet{osmond_genetic_2020}'s work on multi-step rescue by deriving simple approximations for a different fitness landscape.
+Similar to \citet{osmond_genetic_2020}, we find that two-step rescues can be more likely than simple one-step rescues. 
+The situation in which the intermediate growth rate is close to zero in the presence of drug (``stationary'') is particularly interesting.
+While in a general model of evolutionary rescue this might appear to be only a tiny region of parameter space 
+(Fig.~\ref{rescue_threshold}A, inset),
+the example of some aneuploidies in cancer suggests that it may be biologically relevant.
 

### 2025-01-12 ResponseLetter.tex
Update ResponseLetter.tex

@@
 
-\textcolor{blue}{We have added line XX on this in the Introduction: for readers interested more in general evolutionary rescue than in the application to cancer, we think that the primary novelty in the paper is the focus on comparing direct, single-step evolutionary rescue and rescue via an intermediate mutant genotype.}
+\textcolor{blue}{We have added line XX on this in the Introduction: for readers interested more in general evolutionary rescue than in the application to cancer, we think that the primary novelty in the paper is the focus on comparing direct, single-step evolutionary rescue and rescue via an intermediate mutant genotype.
+This was also explicitly considered in Osmond et al, 2020, focusing on Fisher's geometric model;
+we focus on a different parametrization of the fitness landscape.}
 
@@
 
-\textcolor{red}{TODO}
+\textcolor{blue}{We have added a paragraph in the Discussion on the connections to Iwasa et al 2003 and Osmond et al 2020, which we think are the two closest parallels for the main results. 
+For the other more detailed expressions, we have just added the references in the Results as they come up.}
 

### 2025-01-16 ResponseLetter.tex
Add files via upload

@@
 \textcolor{red}{ % change 40
-It is reasonable to assume that the sensitive and mutant growth rates are similar given that we assume that the mutation makes the cancer cells immune to the chemoterapeutic drugs. 
-% TODO the reveiwer is asking if this assumption is reasoanbe and if you can ground it in something
+The doubling times for tumors after relapse is smaller than the doubling times for primary tumors for a variety of tissues \citep{tezuka2007growth,rodgers2024glioblastoma}.
+Additionally, metastatic melanoma has been shown to grow faster then primary melonoma \citep{carlson2003tumor}.
+As a result, it is reasonable to assume that the sensitive and mutant growth rates are at the very least equal.
+% TODO the reveiwer is asking if this assumption is reasoanbe and if you can ground it in something DONE
 }
@@
 
-\textcolor{red}{We did not include density dependence for sensitive cells because they are under strong stress that causes their number to rapidly decline.
-As a result, density will not affect the initial establishment probability of mutants.
-As shown in Figure S1, the presence of a carrying capacity does not effect the probability of evolutionary rescue.
-DOES THIS MAKE SENSE?
-WHY NOT JUST CHANGE TO INCLUDE DENSITY DEPENDENCE FOR SENSITIVES?}
+\textcolor{red}{We have added density-dependence for the sensitives as well.}% DOES THIS MAKE SENSE? WHY NOT JUST CHANGE TO INCLUDE DENSITY DEPENDENCE FOR SENSITIVES?}  Remus: DONE
 
@@
 \textcolor{red}{%change 49
-For large tumors, decreasing the division rate of sensitive cells will lead to a decrease in the cancer recurrence time when conditioned on evolutionary rescue.
-We have added this comment to the text.
-PLEASE CHECK AGAIN, YOU WROTE increase AND DID NOT REALLY ADDRESS THE REVIEWER COMMENT.}
+We have incorporated this comment into our Results. %PLEASE CHECK AGAIN, YOU WROTE increase AND DID NOT REALLY ADDRESS THE REVIEWER COMMENT.} Remus: DONE
+}
 
@@
 }
-\textcolor{red}{REMUS DID YOU ADD THIS TO THE APPENDIX? NOT JUST HERE}
+\textcolor{red}{REMUS DID YOU ADD THIS TO THE APPENDIX? NOT JUST HERE. Remus: DONE}
 
@@
 }
-\textcolor{red}{REMUS DID YOU ADD THIS TO THE APPENDIX? NOT JUST HERE}
+\textcolor{red}{REMUS DID YOU ADD THIS TO THE APPENDIX? NOT JUST HERE. Remus: DONE, change 59}
 

### 2025-01-16 EvolutionaryRescue1.tex
Add files via upload

@@
 \lambda_s' &= \lambda_s , \\
-\mu_s' &= \mu_s ,\\
+\mu_s' &= \mu_s + \lambda_s\frac{s+a+m}{K},\\
 \lambda_a' &= \lambda_a ,\\ 
-\mu_a' &= \mu_a + \lambda_a\frac{w+a+m}{K} ,\\
+\mu_a' &= \mu_a + \lambda_a\frac{s+a+m}{K} ,\\
 \lambda_m' &= \lambda_m ,\\ 
-\mu_m' &= \mu_m + \lambda_m\frac{w+a+m}{K} ,
+\mu_m' &= \mu_m + \lambda_m\frac{s+a+m}{K} ,
 \end{aligned}
@@
 Increasing either the mutant growth rate, $\Delta_m$, or the mutation rate, $v$, decreases the time for the tumor to rebound to its initial size.
-Drugs that significantly increase the death rate of sensitive cells, $\mu_s$, but do not affect their division rate, $\lambda_s$, delay cancer recurrence (conditioned on evolutionary rescue).
+Drugs that significantly increase the death rate of sensitive cells, $\mu_s$, but do not affect their division rate, $\lambda_s$, delay cancer recurrence.
+Additionally, decreasing the sensitive division rate will also the cancer recurrence time.
 Consequently, patients treated with such drugs may require a longer period of monitoring to guarantee the effectiveness of the treatment.
+%change 49
 In addition, drugs that significantly increase the division rate of sensitive cells, $\lambda_s$, but do not affect their death rate, $\mu_s$, decrease cancer recurrence times (conditioned on evolutionary rescue).
-%change 49
 
@@
 Let $N_a^*$ and $N_m^*$ be the threshold population size above which evolutionary rescue through aneuploidy or direct mutation, respectively, is very likely.
-If $N\gg N_m^*$ then $\presc\approx1$ and the mean time $\tau_a$ can be written as: % change 38
+The mean time $\tau_a$ can be written as: % change 38
+\begin{align*}
+\tau_a&=\int_0^\infty tf(t)dt=\int_0^\infty t\frac{d}{dt}F(t)dt=-\int_0^\infty t\frac{d}{dt}S(t)dt\\
+&=-\left[tS(t)\right]_0^\infty+\int_0^\infty S(t)dt\\
+&=\int_0^\infty S(t)dt,
+\end{align*}
+where $f(t)$, $F(t)$ and $S(t)$ are the probability density function, cumulative distribution function and survival function.
+In our case, $S(t)=\exp\left(-\int_0^t R_a(z)dz\right)$.
+Additionally, for $N\gg N_m^*$ we have $1-\e^{-Np_s}\approx1$ and, as a result, we have: %change 23
 \begin{align*}
 \tau_a=\int_0^\infty\e^{-\int_0^t R_a\left(z\right)\d z}\,\d t=\int_0^\infty\exp\left[-\frac{uv\lambda_s\lambda_aNp_m}{\Delta_s-\Delta_a}\left(\frac{\e^{\Delta_s t}-1}{\Delta_s}-\frac{\e^{\Delta_a t}-1}{\Delta_a}\right)-v\lambda_sNp_m\frac{\e^{\Delta_s t}-1}{\Delta_s}\right]\,\d t,
-\end{align*} %change 23 ,r4
+\end{align*} %change r4
 and we use the following Taylor series expansions:
@@
 \end{align}
-If $N\ll N_a^*$ then we can write \Cref{meantimet2} as:
+If $N\ll N_a^*$ then the probability distribution of evolutionary rescue time is skewed toward small values (i.e. evolutionary rescue occurs conditioned on it happening). As a result, we have: 
+\begin{align*}
+\frac{\e^{\Delta_s t}-1}{\Delta_s}&=\frac{1+\Delta_s t+\mathcal{O}\left(\left(\Delta_st\right)^2\right)-1}{\Delta_s}=t+\mathcal{O}\left(\Delta_st^2\right)\\
+\frac{\e^{\Delta_a t}-1}{\Delta_a}&=\frac{1+\Delta_a t+\mathcal{O}\left(\left(\Delta_at\right)^2\right)-1}{\Delta_a}=t+\mathcal{O}\left(\Delta_at^2\right)
+\end{align*}
+We observe that the first term in the exponential in \cref{meantimet2} can be approximated to be zero and the second term as -$v\lambda_sNp_mt$ where $t\ll1$. As a result, we can approximate:
+\begin{equation*}
+\exp\left[-\frac{uv\lambda_s\lambda_aNp_m}{\Delta_s-\Delta_a}\left(\frac{\e^{\Delta_s t}-1}{\Delta_s}-\frac{\e^{\Delta_a t}-1}{\Delta_a}\right)-v\lambda_sNp_m\frac{\e^{\Delta_s t}-1}{\Delta_s}\right] \sim 1
+\end{equation*}
+Additionally, since $N\ll N_a^*\ll N_m^*$, evolutionary rescue is more likely to occur through the trajectory $sensitive \rightarrow aneuploid \rightarrow mutant$.
+As a result, we can write \Cref{meantimet2} as:
 \begin{align}\nonumber
@@
 \end{align} % change r4
-where in the last line we used the fact that $1/p_s=N_a^*$ and \Cref{eq:N_a}.
+where in the last line we used the fact that $1/p_s=N_a^*$ and \Cref{eq:N_a}. %change 59
 

### 2025-01-16 EvolutionaryRescue1.tex
added breast cancer params, figure, discussion

@@
 \modulolinenumbers[2]
+
+% widow and orphan
+\widowpenalty=10000
+\clubpenalty=10000
 
@@
 \paragraph{Parameterization.}
-To parametrize the simulations, we assume that the cells under consideration are melanoma cells and rely on \citet{rew2000cell} and \citet{bozic2013evolutionary} for the division and death rates, respectively.
+We parametrize most of our simulations by considering melanoma cells and rely on \citet{rew2000cell} and \citet{bozic2013evolutionary} for the division and death rates, respectively. 
 \citet{rew2000cell} report \textit{in vivo} measurements of the potential doubling times (the waiting time for the number of cells in the tumor to double, disregarding cell death) for a large set of cancer types.
@@
 % change 36
-In most of our simulations, we use $\lambda_a=0.0899$ per day, so that aneuploid cells are tolerant and aneuploidy can only act as an evolutionary ``stepping stone'' for the generation of the resistant mutant that rescues the tumor (note that this mutant will occur on the background of an aneuploid genotype).
+In most of our simulations, we use $\lambda_a=0.0899$ per day, so that aneuploid cells are tolerant and aneuploidy can only act as an evolutionary ``stepping stone'' for the generation of the resistant mutant that rescues the tumor (note that this mutant will occur on the background of an aneuploid genotype) .
+
+In some of the analysis, we consider breast cancer.
+We assume that the division and death rates of the aneuploid breast cancer cells in the presence of cisplatin is the same as those of the sensitive cells in the absence of the drug.
+The growth rate of breast cancer cells in the absence of drug is $0.0085$ per day \citep{spratt1996rates}.
+The doubling time of breast cancer cells in the absence of cellular death is 8.2 days \citep{rew2000cell}.
+Thus, the division rate of aneuploid cells is $\lambda_a=\log{(2)}/8.2=0.0845$ per day, and their death rate is $\mu_a=0.076$ per day.
+For the division and death rates of sensitive cells, we turn to results described in \citet{salehi2021clonal}.
+We select TNBC-SA535 PDX clone H, which has a relative fitness of $\approx 1.011$.
+Thus, the selection coefficient is the ratio of the aneuploid and sensitive Wrightian fitness values minus 1 \citep{chevin2011measuring}.
+The Malthusian fitness is the log of the Wrightian fitness \citep{Wu2013a}.
+Thus, $\Delta_s = \log{0.9975} -0.0025$.
+We assume cisplatin affects cancer cells by increasing their death rate but does not affect the their division rate, hence $\lambda_s=0.08$ and $\mu_s=0.0825$ per day.
+A similar analysis for TNBC-SA609 clone A and TNBC-SA1035 clone H from \citet{salehi2021clonal} gives $\lambda_s=0.08, \mu_s=0.12$ and $\lambda_s=0.08, \mu=0.1$, respectively.
 
@@
 \end{equation}
-\Cref{eq:N_a} is our model's equivalent of \citet{iwasa2004evolutionary}'s eq.~A3, or eq.~8 of \citet{osmond_genetic_2020} with the distribution of fitness effects set to a delta function;
-our ``tolerant'', ``stationary'', and ``resistant'' correspond to \citet{osmond_genetic_2020}'s ``sufficiently subcritical'', ``sufficiently critical'', and ``sufficiently supercritical'', respectively.
+This equation is equivalent to eq.~A3 of \citet{iwasa2004evolutionary} and to eq.~8 of \citet{osmond_genetic_2020} with the distribution of fitness effects set to a delta function;
+our ``tolerant'', ``stationary'', and ``resistant'' scenarios correspond to \citet{osmond_genetic_2020}'s ``sufficiently subcritical'', ``sufficiently critical'', and ``sufficiently supercritical'', respectively.
 %change 57
 These approximations are accurate when compared to results of stochastic evolutionary simulations (\Cref{rescue_prob,rescue_threshold}). 
-The first line describes the scenario in which the treatment still effectively kills aneuploid cells but not as quickly as the sensitive cells. 
+In the first scenario, the treatment effectively kills aneuploid cells but not as quickly as it kills sensitive cells. 
 In the second scenario, aneuploid cells are sufficiently resistant, and the expected size of each aneuploid lineage is roughly 1.
@@
 
+We compared the probability of evolutionary rescue in melanoma cells and triple-negative breast cancer cells (TNBC) under drug threapy (black and color lines, respectively, in \Cref{rescue_prob}B).
+This comparison suggests that TNBC has a higher probability of relapsing compared to melanoma, because smaller tumors have higher resuce probability.
+Indeed, the probability of relapse in TNBC patients is higher than 50\% in the first 3-5 years after diagnosis \citep{taushanova2023synchronous}, whereas in melanoma patients it is approximately 10-20\% in the first 5 years \citep{wan2022prediction,von_schuckmann2019risk}.
+% change A
+
 Given that the mean time for secondary tumors to adapt to anti-cancer drugs can be of the order of 1,000 days (\Cref{MeanTimeGrowthAneuploidyPlot}A), aneuploidy can explain the reappearance of cancer even after initial remission.
@@
 
-Comparing our results to the evolutionary rescue literature, we complement \citet{iwasa2003evolutionary} and \citet{osmond_genetic_2020}'s work on multi-step rescue by deriving simple approximations for a different fitness landscape.
+Comparing our results to the evolutionary rescue literature, we complement the studies by \citet{iwasa2003evolutionary} and \citet{osmond_genetic_2020} on multi-step rescue by deriving simple approximations for a different fitness landscape.
 Similar to \citet{osmond_genetic_2020}, we find that two-step rescues can be more likely than simple one-step rescues. 
-The situation in which the intermediate growth rate is close to zero in the presence of drug (``stationary'') is particularly interesting.
-While in a general model of evolutionary rescue this might appear to be only a tiny region of parameter space 
-(Fig.~\ref{rescue_threshold}A, inset),
-the example of some aneuploidies in cancer suggests that it may be biologically relevant.
-
-%Finally, we observe from \Cref{rescue_threshold} that only for a restricted region of the parameter space will aneuploidy act as a ``stepping stone'' for evolutionary rescue.
-If the aneuploid division rate is smaller than $\frac{\mu_a}{1+u}\approx\mu_a(1-u)$, then evolutionary rescue will occur through direct mutation.
-As a result, for most parameter values, aneuploidy will either not play any role in evolutionary rescue or will be the main driver of adaptation without requiring any mutation (i.e., evolutionary rescue in one step).
-% NOTE FROM DANIEL: I cut this paragraph because I think it's implicitly based on a uniform distribution of aneuploid growth rates.
-But the whole point of why we're doing this is because there are apparently some interesting aneuploidies that have growth rates near 0
-
-\paragraph{Directions for future research}
+The ``stationary'' scenario, in which the aneuploid growth rate is close to zero in the presence of drug, is particularly interesting.
+Although this may seem like a small region of parameter space in a general model of evolutionary rescue (Fig.~\ref{rescue_threshold}A, inset), the example of certain aneuploidies in cancer suggests it could be biologically significant.
+
+%\paragraph{Directions for future research}
 Experiments could test our model predictions.
@@
 
-\paragraph{Conclusions}
+\paragraph{Conclusions.}
 
@@
     $\tilde{u}$ & Missegregation rate in the drug free environment$^\ast$& $5\times10^{-4}-2\times10^{-2}$ & 1$\slash$cell division  & \citet{shi2005chromosome,thompson2008examining} \\ \hline
-    $c$ & Selection coefficient against aneuploidy in the drug free environment& $0.07$ &  1/days   & \citet{lukow2021chromosomal} \\  
-    \hline
+    $c$ & Selection coefficient against aneuploidy in the drug free environment& $0.07$ &  1/days   & \citet{lukow2021chromosomal} \\ \hline
   \end{tabular}
-\caption{\textbf{Model parameters.} 
+\caption{\textbf{Model parameters: Melanoma.} 
 Parameters from \citet{bozic2013evolutionary} consider patients with melanoma treated with the anti-cancer drug vemurfenib, in which resistance is conferred by trisomy in either Chr 2 or Chr 6.
-We have modified the parameters from \citet{bozic2013evolutionary} such that sensitive and mutant division rates are $\lambda_{s} = \lambda_{m} = \log2/T\approx0.1$ instead of their value of $0.14$ where $T$ is the doubling time in the absence of cellular death obtained from \citet{rew2000cell}.
+We have modified the parameters from \citet{bozic2013evolutionary} such that sensitive and mutant division rates are $\lambda_{s} = \lambda_{m} = \log{(2)}/T\approx0.1$ instead of their value of $0.14$ where $T$ is the doubling time in the absence of cellular death obtained from \citet{rew2000cell}.
 For a discussion of the different interpretations of the tumor doubling times see \citet{avanzini2019cancer}.
@@
   \label{table1}
+\end{center}
+\end{table}
+
+\begin{table}
+\begin{center}
+  \begin{tabular}{| l |p{5cm}| c | c | p{3cm} |}
+    \hline
+     & Name & Value & Units & References \\ \hline
+    $\lambda_s$ & Sensitive division rate& 0.08 & 1/days  & \citet{salehi2021clonal} \\ \hline
+    $\mu_s$ & Sensitive death rate& $0.0825, 0.1, 0.12$ & 1/days  & \citet{salehi2021clonal} \\ \hline
+    $\lambda_a$  & Aneuploid division rate & $0.0845$ & 1/days  & \citet{spratt1996rates,rew2000cell}  \\ \hline
+    $\mu_a$ & Aneuploid death rate & $0.076$ & 1/days  & \citet{spratt1996rates,rew2000cell}  \\ \hline
+  \end{tabular}
+\caption{\textbf{Model parameters: Breast cancer.} 
+Parameters not listed here remain unchanged from the values provided in \cref{table1}.
+}
+  \label{table2}
 \end{center}
@@
 \begin{figure}
+\begin{subfigure}{0.48\textwidth}
+A\\
 \includegraphics[width=1\textwidth]{Figures/ProbvNPlot.pdf}
+\end{subfigure}
+\begin{subfigure}{0.52\textwidth}
+B\\
+\includegraphics[width=1\textwidth]{Figures/PDXModelProb.pdf}
+\end{subfigure}
 \caption{\textbf{Aneuploidy facilitates the evolutionary rescue of cancer under drug treatment.}
-The probability of evolutionary rescue (i.e., the probability that the population does not become extinct), $\presc$, as a function of the initial tumor size, $N$ (\cref{eq:rescue_prob}).
-Dashed vertical line shows the threshold tumor size, $N_a^*$, above which the probability is very high (\cref{eq:N_a}).
-Blue dashed line: without aneuploidy ($u=0$).
+The probability of evolutionary rescue, $\presc$, as a function of the initial tumor size, $N$ (\cref{eq:rescue_prob}). 
+Dashed vertical lines show the threshold tumor size, $N_a^*$, above which the probability is very high (\cref{eq:N_a}). 
+\textbf{(A)} Blue dashed line: without aneuploidy ($u=0$).
 Black line: tolerant aneuploidy ($u=10^{-2}, \lambda_a=0.0899$).
 Red line: stationary aneuploidy ($u=10^{-2}, \lambda_a=0.08999$).
-Green line represents the scenario with partially resistant aneuploidy ($u=10^{-2}, \lambda_a=0.095$).
-Dots for simulations and the error bars for $95\%$ confidence interval ($p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the fraction of simulations in which the tumor has been rescued, and $n=100$ is the number of simulations).
-Parameters: $\lambda_s=0.1,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, v=10^{-7}$.}
+Green line: partially resistant aneuploidy ($u=10^{-2}, \lambda_a=0.095$).
+Markers and error bars for averages of simulation results with $95\%$ confidence interval ($p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the fraction of simulations in which the tumor has been rescued, and $n=100$ is the number of simulations).
+Parameters: $\lambda_s=0.1,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, v=10^{-7}$.
+\textbf{(B)} Comparison for melanoma (black) and breast cancer (colors)  with parameters from the literature (\Cref{table1,table2}).
+Black line: Melanoma A375.
+Blue line: Breast cancer TNBC-SA1035 clone H.
+Red line: Breast cancer TNBC-SA609 clone A.
+Green line: Breast cancer TNBC-SA535 clone H.
+}
 \label{rescue_prob}

### 2025-01-16 ResponseLetter.tex
continue editing the revision and response

@@
 \title{
-	Response to reviewers\\ Evolutionary rescue by aneuploidy in tumors exposed to anti-cancer drugs 
+	Response to reviewers: \\ Evolutionary rescue by aneuploidy in tumors exposed to anti-cancer drugs \\ 
 }
@@
 
-\textcolor{red}{ % change 40
-The doubling times for tumors after relapse is smaller than the doubling times for primary tumors for a variety of tissues \citep{tezuka2007growth,rodgers2024glioblastoma}.
-Additionally, metastatic melanoma has been shown to grow faster then primary melonoma \citep{carlson2003tumor}.
-As a result, it is reasonable to assume that the sensitive and mutant growth rates are at the very least equal.
-% TODO the reveiwer is asking if this assumption is reasoanbe and if you can ground it in something DONE
+\textcolor{blue}{ % change 40
+We now ground this assumption in results from the literature (line XXX): "The division rate is obtained as $\lambda=\log{2} / T \approx 0.1$ per day.
+We take this to be the division rate for sensitive and assume that mutant cells, which are resistant to therapy, have the same division rate.
+Indeed, doubling times for tumors after relapse is lower than that of primary tumors for a variety of tissues \citep{tezuka2007growth,rodgers2024glioblastoma} and metastatic melanoma has been shown to grow faster then primary melonoma \citep{carlson2003tumor}."
 }
@@
 
-\textcolor{red}{We have added density-dependence for the sensitives as well.}% DOES THIS MAKE SENSE? WHY NOT JUST CHANGE TO INCLUDE DENSITY DEPENDENCE FOR SENSITIVES?}  Remus: DONE
+\textcolor{blue}{We have now added density-dependence for the sensitives. The results are unchanged.}
 
@@
 The repository will become available when the paper is accepted.
-Meanwhile, the source code for our analysis can be privately accessed here: \url{https://www.dropbox.com/s/60zyc6rcjp8q859/StanaEtAlCode.zip}
+Meanwhile, the source code for our analysis can be privately accessed here: \url{https://www.dropbox.com/scl/fi/hrvqdocsr67qur7k59wxw/StanaEtAlCode.zip?rlkey=n7bcr3usx5ym6q3s2qkotanhx}
 }
@@
 
-\textcolor{red}{ % change 58
-We believe that our model does fit with data on resistance rates: \citep{handa2022long} state that 68$\%$ of melanoma patients treated with vemurafenib experienced relapse after initial response.
-} % THE SENTENCE IN THE TEXT IS UNCLEAR.
-REVISE IT AND COPY IT HERE
+\textcolor{blue}{ % change 58
+We believe that our model does fit with data on resistance rates, and the text now reads (line XXX): "Interestingly, the threshold between small and intermediate tumors, $N_a^*$, is similar to the tumor detection threshold of $4.19 \times 10^6$ cells for a wide variety of tumors~\citep{avanzini2019cancer}.
+We note that vemurafenib-treated melanomas (i.e.
+melanomas with sizes above the detection threshold) have a probability $>50\%$ to relapse \citep{piejko2023long,handa2022long}."
+}
 
@@
 
-\textcolor{red}{%change 49
-We have incorporated this comment into our Results. %PLEASE CHECK AGAIN, YOU WROTE increase AND DID NOT REALLY ADDRESS THE REVIEWER COMMENT.} Remus: DONE
+\textcolor{blue}{% change 49
+The text now reads (line XXX): "Drugs that increase the sensitive death rate, $\mu_s$, but do not affect the sensitive division rate, $\lambda_s$, delay cancer recurrence.
+Drugs that decrease the sensitive division rate delay cancer recurrence time even more, because they effectively decrease the mutation rate (assuming mutations occur during division)."
 }
@@
 \textcolor{blue}{%change 23
-The justification is the following: 
+The justification, which is now included in Appendix C, is the following: 
 \begin{align*}
@@
 In our case, $S(t)=\exp\left(-\int_0^t R_a(z)dz\right)$.
-Additionally, for $N\gg N_m^*$ we have $1-\e^{-Np_s}\approx1$. I have corrected the following mistake:
-\begin{equation*}
-\tau_a=\int_0^\infty \e^{-\int_0^\tau R_a(t)\d t}\d\tau
-\end{equation*}
-instead of:
-\begin{equation*}
-\tau_a=\int_0^\infty \e^{-R_a(\tau)}\d\tau.
-\end{equation*}
+Additionally, for $N\gg N_m^*$ we have $1-\e^{-Np_s}\approx1$. 
+We have corrected a mistake, we now write $\tau_a=\int_0^\infty \e^{-\int_0^\tau R_a(t)\d t}\d\tau$
+instead of $\tau_a=\int_0^\infty \e^{-R_a(\tau)}\d\tau$.
 }
-\textcolor{red}{REMUS DID YOU ADD THIS TO THE APPENDIX? NOT JUST HERE. Remus: DONE}
 
@@
 
-\textcolor{blue}{We know that
+\textcolor{blue}{
+The justificaiont, also included now in Appendix C, is in the following. 
+We know that
 \begin{align*}
@@
 }
-\textcolor{red}{REMUS DID YOU ADD THIS TO THE APPENDIX? NOT JUST HERE. Remus: DONE, change 59}
 
@@
 As a result, mutations that are deleterious are taken into consideration through the mean fitness of each population.
-I have added this comment in the "Parametrisation" section change 50}
+We have added this comment in the "Parametrisation" section change 50}
 
@@
 
-\textcolor{blue}{The variation in our parameters in noted in Table 1 and is obtained from the papers which are cited there.}
+\textcolor{blue}{The variation in our parameters in noted in Table 1 and the new Table 2 and is obtained from the papers which are cited there.}
 
@@
 
-\textcolor{blue}{We have made the recommended change. change 30}
+\textcolor{blue}{We have made the recommended change. }  % change 30
 
@@
 
-\textcolor{blue}{Thank you for your commnet. We have changed the text such that it states that evolutionary rescue in two steps is similar to the problem of crossing fitness valley. change 6}
+\textcolor{blue}{We removed the statement about the mathematical equivalence of valley crossing and evolutionary rescue.}
 
@@
 
-\textcolor{blue}{Thank you for your commnet.
-The model is described in "Methods" under the subsection "Evolutionary model" and the analytic results follow directly from this model.
-Additionally, the quantities derived are stated in the paragraph before "Methods".
-}
+\textcolor{blue}{
+The Methods section has several subsections: "Evolutionary model" that describes the model, from which the analytic results are derived; "Stochastic simulations", which describes the simulations used to validate the analytic results; "Parameterization", which describes how we parameterize the model; and "Denstity-dependent growth", which described the extended model that includes density dependence.
+Note that the "Evolutionary model" subsection is not focused on the simulations but rather described the general model that we are studying.}
 
@@
 
-\textcolor{blue}{Thank you for your commnet. The are about 300 mentions of $\Delta$ and we do not want to introduce any errors by changing $\Delta$ to $r$.}
+\textcolor{blue}{The manuscript has over 300 mentions of $\Delta$, and we do not want to introduce any errors by changing it.}
 
@@
 
-\textcolor{blue}{Thank you for your commnet. We have added a short explanation for "tolerant" and that it has other meanings in the literature. change 47.}
+\textcolor{blue}{We have added a short explanation for "tolerant" and that it has other meanings in the literature.
+We also mention the equivalence to in "sub/super/critical" from Osmond et al 2020.} % change 47, 57
 
@@
 
-\textcolor{blue}{We have made the recommended change from "antagonizing cell division" to "by extending the length of the cell cycle". change 34}
+\textcolor{blue}{We have made the recommended change from "antagonizing cell division" to "by extending the length of the cell cycle".}  % change 34
 
@@
 This should make it clearer that the non-monotonicity of the black curve is an artefact of the simplifying assumptions in (D4) and not a problem with the approach.
-By the way, It wasn't clear to me what happened to the other large $\tau_a^r$ terms in the first portion of equation (D3) when approximating the equation into the form shown in (D4)?
+
+\textcolor{blue}{We have added the green line for eq. D3 to Figure 7 as suggested, it is very close to the blue line for eq. D4.}
+
+* By the way, It wasn't clear to me what happened to the other large $\tau_a^r$ terms in the first portion of equation (D3) when approximating the equation into the form shown in (D4)?
 Could this be briefly clarified?
 
-\textcolor{blue}{We have changed Figure 7 by adding a green which represents the numerical solution of (D3).  In the first term of (D3) we have:
+\textcolor{blue}{In the first term of (D3) we have: 
 \begin{align*}
@@
 \end{align*}
-change 44}
+} % change 44
 
-Remus changes:
+\textbf{Additional changes}
 
-\textcolor{red}{$w_t\rightarrow s_t$. change r1}
+We have made the following additional changes:
 
-\textcolor{red}{$t\rightarrow \tau$. change r2}
+$w_t\rightarrow s_t$. %change r1
 
-\textcolor{red}{
-We have made the following changes: the rates of the inhomogenous Poisson process are:
+$t\rightarrow \tau$. %change r2
+
+The rates of the inhomogenous Poisson process are: %change r3
 \begin{align*}
@@
 \end{align*}
-change r3}
 
-\textcolor{red}{We have made changes to Appendix C such that all the dummy variable used in integration are consistent. change r4}
+Appendix C: all dummy variable used in integration are consistent. %change r4
 
-\textcolor{red}{We have made changes to Figure S6, S7, S8, S9 by changing the subscripts of several variables from $w$ to $s$ . change r5}
+Figure S6, S7, S8, S9: changed the subscripts of several variables from $w$ to $s$ . %change r5
 
-\textcolor{red}{ We have made the following change: "If a fraction $f$ of the cancer cells are aneuploid when the drug is administered then the rates at which the rescue mutations are generated can be written as" $\rightarrow$
-"If a fraction $f$ of the cancer cells are aneuploid when the drug is administered then the expected number of successful mutants generated until time $t$ is given by". change r6}
+"If a fraction $f$ of the cancer cells are aneuploid when the drug is administered then the rates at which the rescue mutations are generated can be written as" $\rightarrow$
+"If a fraction $f$ of the cancer cells are aneuploid when the drug is administered then the expected number of successful mutants generated until time $t$ is given by". % change r6
 

### 2025-01-16 EvolutionaryRescue1.tex
continue editing the revision and response

@@
 The division rate is obtained as $\lambda=\log{2} / T \approx 0.1$ per day.
-We select this to be the division rate for sensitive and mutant cells.
-This is a reasonable assumption given that we assume that the mutation makes the cancer cells immune to the chemotherapeutic drugs.
+We take this to be the division rate for sensitive and assume that mutant cells, which are resistant to therapy, have the same division rate.
+Indeed, doubling times for tumors after relapse is lower than that of primary tumors for a variety of tissues \citep{tezuka2007growth,rodgers2024glioblastoma} and metastatic melanoma has been shown to grow faster then primary melonoma \citep{carlson2003tumor}.
 %change 40
@@
 The mean time for the mutant cell population to reach size $N$, the initial number of cancer cells.
-Our inhomogeneous Poisson-process approximation (solid black line, \cref{meanproliferationtime}) is in agreement with simulation results (red markers with 95\% confidence interval obtained with bootstrapping, see Appendix G) for intermediary $N$.
-The simulations converge to \cref{eq:t2det} (blue dashed line) for large values of $N$.
-The green line represents the numerical solution of \cref{eq:t2det2024}.
-% change 44
+Our inhomogeneous Poisson-process approximation (solid black line, \cref{meanproliferationtime}) is in agreement with simulation results (red markers with error bars for 95\% CI) for intermediate size $N$.
+Simulation results converge to \cref{eq:t2det} (blue dashed line) for large values of $N$. The green line represents the numerical solution of \cref{eq:t2det2024}. % change 44  
 Parameters: $\lambda_s=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}

### 2025-01-16 ResponseLetter.tex
more edits

@@
 \usepackage{url}
+\usepackage{graphicx} % Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
 
@@
 
-%I have no major edits to recommend on this manuscript and I am comfortable with the authors considering the minor edits and making adjustments as they see fit.
+I have no major edits to recommend on this manuscript and I am comfortable with the authors considering the minor edits and making adjustments as they see fit.
 
@@
 } 
-
+\\
+\\
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \textcolor{blue}{Fixed.} % change 17
-
+\\
+\\
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \textbf{Reviewer $\#3$}
 
-Summary
+%Summary
 
-Stana et al.
+%Stana et al.
 study the effects of aneuploidy (mutations that give rise to atypical karyotypes) on the survival of a tumor- a population of malignant cells- in presence of an intervention, here, an anti-cancer drug.
@@
 
-This manuscript is about understanding the evolutionary role of aneuploidy in drug-resistance in cancer.
+%This manuscript is about understanding the evolutionary role of aneuploidy in drug-resistance in cancer.
 The authors model tumor evolutionary dynamics under treatment as a multi-type branching process, where the types correspond to three genotypes: sensitive to therapy, mutant (resistant), and aneuploid.
@@
 
-Stana et al.
+%Stana et al.
 derive equations for relevant evolutionary quantities under multiple parameter regimes.
@@
 The relevance of the paper would be improved if consistent results can be shown in other datasets, including the citations above.
+
+\textcolor{blue}{Following the reviewer's suggestion, we have evaluated our model using parameters from breast cancer (PMID: 34163070), see Methods (line XXX), Table 2, and Figure 3B, which also appear below.
+We have added the following text to the discussion: "We compared the probability of evolutionary rescue in melanoma cells and triple-negative breast cancer cells (TNBC) under drug threapy (black and color lines, respectively, in Figure 3B).
+This comparison suggests that TNBC has a higher probability of relapsing compared to melanoma, because smaller tumors have higher resuce probability.
+Indeed, the probability of relapse in TNBC patients is higher than 50\% in the first 3-5 years after diagnosis \citep{taushanova2023synchronous}, whereas in melanoma patients it is approximately 10-20\% in the first 5 years \citep{wan2022prediction,von_schuckmann2019risk}.
+}% change A
 
@@
 } % change 44
-
+\\
+\\
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \textbf{Additional changes}

### 2025-01-16 ResponseLetter.tex
red something to discuss:

@@
 
-\textcolor{blue}{We have added line XX on this in the Introduction: for readers interested more in general evolutionary rescue than in the application to cancer, we think that the primary novelty in the paper is the focus on comparing direct, single-step evolutionary rescue and rescue via an intermediate mutant genotype.
+\textcolor{red}{We have added line XX on this in the Introduction: "for readers interested more in general evolutionary rescue than in the application to cancer, we think that the primary novelty in the paper is the focus on comparing direct, single-step evolutionary rescue and rescue via an intermediate mutant genotype.
 This was also explicitly considered in Osmond et al, 2020, focusing on Fisher's geometric model;
-we focus on a different parametrization of the fitness landscape.}
+we focus on a different parametrization of the fitness landscape."\\ IS THIS IN THE TEXT? CANT FIND IT}
 

### 2025-01-16 EvolutionaryRescue1.tex
edited intro

@@
 Here, we study evolutionary rescue after a sudden environmental change caused by initiating anti-cancer drug treatment. 
-Essentially, this consists of combining known results from the evolutionary rescue literature, focusing on the quantities and parameter values that are relevant for cancer.
-Because these results are scattered across multiple papers, and because they all follow from simple branching process approximations, we re-derive all results here for our model.
-While our focus is on the application of evolutionary rescue theory to evolution of drug resistance in cancer, our model is a general one, and in the general evolutionary rescue context, this work can be seen as a characterization of the circumstances in which two-step processes contribute to rescue even when single-step rescue mutations are available, similar to part of \citet{osmond_genetic_2020} but focused on a different fitness landscape.
-
 We consider a range of effects of aneuploidy, from tolerance to (partial) resistance to the drug \citep{brauner2016distinguishing}; note that we use tolerant to refer to aneuploids which have negative growth rates, which is different from other references in the literature such as \citet{berman2020drug}.
@@
 Additionally, we are interested in the timescale of evolutionary rescue and the impact that aneuploidy has on the time necessary for the tumor to overcome drug therapy.
+
+We will address these questions essentially by combining known results from the evolutionary rescue literature, and considering the quantities and parameter values that are relevant for cancer.
+Because these results are scattered across multiple papers, and because they all follow from simple branching process approximations, we re-derive all results here for our model.
+While our focus is on the application of evolutionary rescue theory to evolution of drug resistance in cancer, our model is a general one, and in the general evolutionary rescue context, this work can be seen as a characterization of the circumstances in which two-step processes contribute to rescue even when single-step rescue mutations are available.
+This is similar to part of \citet{osmond_genetic_2020} but focused on a different fitness landscape.
 

### 2025-01-16 EvolutionaryRescue1.tex
edited on zoom with DBW

--- 2025-01-16 EvolutionaryRescue1.tex
@@
 In the third scenario, aneuploid cells are sufficiently resistant for the population to re-grow the tumor without additional resistance mutations.
+In our parameterization for melanoma and breast cancer, the value of $\Delta_a T^*$ is XXX. Therefore, in these specific cases, we are in the XXX scenario. %TODO fill XXX % change b
+
 Notably, in this scenario the mutant parameters ($v$, $\lambda_m$, and $\Delta_m$) do not affect $N_a^*$ beyond their effect on $T^*$.
@@
 
-Interestingly, increasing both the aneuploid division rate, $\lambda_a$, and the aneuploid death rate, $\mu_a$, such that the growth rate $\Delta_a$ remains constant, leads to a decrease in $T^*$, pushing the system to the second scenario (increasing $\lambda_a$ causes a decrease in $T^*$ because  it increases the mutation rate and a lineage does not have to survive as long in order to generate a successful mutant).
-In this scenario, increasing the division rate $\lambda_a$ should also increase the mutation rate $v\lambda_a$ in aneuploid cells, as mutations mostly occur during division, so overall, the threshold tumor size $N_a^*$ is unaffected by the division rate $\lambda_a$ (i.e., $d \lambda_a T^*/d\lambda_a = 0$).
+Interestingly, increasing both the aneuploid division rate, $\lambda_a$, and the aneuploid death rate, $\mu_a$, such that the growth rate $\Delta_a$ remains constant, leads to a decrease in $T^*$, pushing the system to the second scenario.
+This is because increasing $\lambda_a$ causes a decrease in $T^*$ because  it increases the effective mutation rate $v\lambda_a$ (as mutations mostly occur during division) and a lineage does not have to survive as long in order to generate a successful mutant.
+Overall, the threshold tumor size $N_a^*$ is unaffected by the division rate $\lambda_a$ (i.e., $d \lambda_a T^*/d\lambda_a = 0$).
 Thus, if aneuploid cells rapidly die due to the drug but compensate by rapidly dividing, increasing the division rate will \emph{not} facilitate adaptation.
@@
 
-We compared the probability of evolutionary rescue in melanoma cells and triple-negative breast cancer cells (TNBC) under drug threapy (black and color lines, respectively, in \Cref{rescue_prob}B).
+As an example, we have parameterized our model using estimates for melanoma and triple-negative breast cancer cells (TNBC) under drug therapy from the literature. 
+We find that in these cases we are in the XXX scenario. %TODO fill XXX % change B
+We have also compared the probability of evolutionary rescue between melanoma and TNBC (black and color lines, respectively, in \Cref{rescue_prob}B).
 This comparison suggests that TNBC has a higher probability of relapsing compared to melanoma, because smaller tumors have higher resuce probability.
@@
 Although this may seem like a small region of parameter space in a general model of evolutionary rescue (Fig.~\ref{rescue_threshold}A, inset), the example of certain aneuploidies in cancer suggests it could be biologically significant.
+% TODO if change B gives stationary scenario, mention it here; otherwise, say that other cancer might produce this
 

### 2025-01-16 ResponseLetter.tex
Merge branch 'main' of https://github.com/yoavram-lab/EvolutionaryRescue

--- 2025-01-16 ResponseLetter.tex
@@
 \textcolor{red}{
-TODO
+TODO change B
 }
@@
 
-\textcolor{blue}{We have added line XX on this in the Introduction: for readers interested more in general evolutionary rescue than in the application to cancer, we think that the primary novelty in the paper is the focus on comparing direct, single-step evolutionary rescue and rescue via an intermediate mutant genotype.
-This was also explicitly considered in Osmond et al, 2020, focusing on Fisher's geometric model;
-we focus on a different parametrization of the fitness landscape.}
-
 In case it helps, my understanding of multi-step rescue theory is that there was an early model for the emergence of epidemics via multiple mutations (Antia et al 2003 Nature, and some follow ups like Yates et al 2006 PRSB and Alexander $\&$ Day 2010 Interface), a very general treatment of emergence and rescue via multiple mutations (Iwasa et al 2004 JTB), a 2-locus 2-allele rescue model (Uecker $\&$ Hermisson 2016 Genetics), and multi-step rescue models on a fitness landscape (Martin et al 2013, Osmond et al 2020), all of which also use branching processes.
@@
 They have basically done all the work for us.
-We have incorporated these references in the Introduction and throughout the Results at the equations mentioned.
+We have incorporated a review of the literature in the Introduction (line XXX), and refer to these papers throughout the Results at the equations mentioned.
 The only slight changes we have made are omitting the follow-ups to Antia et al 2003 and instead including Iwasa et al 2003 Proc B, which we really should have mentioned before and is arguably the closest thing in the literature to this manuscript.
+Because the results mentioned by the reviewer are scattered across multiple papers, and because they all follow from simple branching
+process approximations, we re-derive all results here for our model.
 }
@@
 
-\textcolor{red}{%change 51
-We have added the recommended comment about the fact that the reason why $T^*$ decreases, if we increase the aneuploid division and death rates such that $\Delta_a$ remains constant, is because of the fact that more mutants are generated and a lineage doesn't have to persist as long in order to generate a succesful mutant.
+\textcolor{blue}{%change 51
+We have added the recommended comment (line XXX): "Interestingly, increasing both the aneuploid division rate, $\lambda_a$, and the aneuploid death rate, $\mu_a$, such that the growth rate $\Delta_a$ remains constant, leads to a decrease in $T^*$, pushing the system to the second scenario.
+This is because increasing $\lambda_a$ causes a decrease in $T^*$ because  it increases the effective mutation rate $v\lambda_a$ (as mutations mostly occur during division) and a lineage does not have to survive as long in order to generate a successful mutant."
 }

### 2025-01-19 ResponseLetter.tex
edited breast cancer stuff; ms.bib

@@
 
-\textcolor{blue}{This is a great point. We have basically incorporated this comment into our Introduction.}
+\textcolor{blue}{This is a great point. We have revised the Introduction accordingly.}
 
@@
 We have added the recommended comment (line XXX): "Interestingly, increasing both the aneuploid division rate, $\lambda_a$, and the aneuploid death rate, $\mu_a$, such that the growth rate $\Delta_a$ remains constant, leads to a decrease in $T^*$, pushing the system to the second scenario.
-This is because increasing $\lambda_a$ causes a decrease in $T^*$ because  it increases the effective mutation rate $v\lambda_a$ (as mutations mostly occur during division) and a lineage does not have to survive as long in order to generate a successful mutant."
+This is because increasing $\lambda_a$ causes a decrease in $T^*$ as  it increases the effective mutation rate $v\lambda_a$ (as mutations mostly occur during division) and a lineage does not have to survive as long in order to generate a successful mutant."
 }
@@
 
-\textcolor{red}{%change 52
-In the scenario discribed decreasing the division rate also leads to a decrease in the death rate as the drugs affect the cancer cells during cellular division.
-We have expanded on the explanation of this scenario in the paper.
+\textcolor{blue}{%change 52
+We have removed the sentence about these results being consistent with experimental finding.
 }
@@
 \bibliographystyle{plain}
-\bibliography{evo2022}
+\bibliography{ms}
 \end{document}

### 2025-01-19 ms.tex
edited breast cancer stuff; ms.bib

@@
 Because these results are scattered across multiple papers, and because they all follow from simple branching process approximations, we re-derive all results here for our model.
-While our focus is on the application of evolutionary rescue theory to evolution of drug resistance in cancer, our model is a general one, and in the general evolutionary rescue context, this work can be seen as a characterization of the circumstances in which two-step processes contribute to rescue even when single-step rescue mutations are available.
+While our focus is on the application of evolutionary rescue theory to the evolution of drug resistance in cancer, our model is a general one, and in the general evolutionary rescue context, this work can be seen as a characterization of the circumstances in which two-step processes contribute to rescue even when single-step rescue mutations are available.
 This is similar to part of \citet{osmond_genetic_2020} but focused on a different fitness landscape.
@@
 In some of the analysis, we consider breast cancer. 
-We assume that the division and death rates of the aneuploid breast cancer cells in the presence of cisplatin is the same as those of the sensitive cells in the absence of the drug.
+\citet{salehi2021clonal} studied multiple TNBC PDX clones (triple-negative breast cancer patient-derived xenografts).
+Their results suggest that aneuploidy and other copy number alterations provid resistance to the drug cisplatin.  
+Thus, we assume that aneuploid breast cancer cells are are partially or even completely resistant, i.e., $\Delta_m \ge \Delta_a > 0$, and that the division and death rates of aneuploid cells in the presence of cisplatin is the same as those of  sensitive breast cancer cells in the absence of the drug.
 The growth rate of breast cancer cells in the absence of drug is $0.0085$ per day \citep{spratt1996rates}.
@@
 Thus, the division rate of aneuploid cells is $\lambda_a=\log{(2)}/8.2=0.0845$ per day, and their death rate is $\mu_a=0.076$ per day.
-For the division and death rates of sensitive cells, we turn to results described in \citet{salehi2021clonal}.
-We select TNBC-SA535 PDX clone H, which has a relative fitness of $\approx 1.011$.
+To determine the division and death rates of sensitive cells, we chose TNBC-SA535 clone H from \citet{salehi2021clonal}, which has a relative fitness of $\approx 1.011$.
 Thus, the selection coefficient is the ratio of the aneuploid and sensitive Wrightian fitness values minus 1 \citep{chevin2011measuring}.
@@
 A similar analysis for TNBC-SA609 clone A and TNBC-SA1035 clone H from \citet{salehi2021clonal} gives $\lambda_s=0.08, \mu_s=0.12$ and $\lambda_s=0.08, \mu=0.1$, respectively.
+% TODO it does not make sense that when you change aneuploid clones what changes is the sensitive growth rate - the sensitive growth rate should be constant and the aneuploid clone growth rate should change...
 
@@
 In the third scenario, aneuploid cells are sufficiently resistant for the population to re-grow the tumor without additional resistance mutations.
-In our parameterization for melanoma and breast cancer, the value of $\Delta_a T^*$ is XXX. Therefore, in these specific cases, we are in the XXX scenario. %TODO fill XXX % change b
+In our parameterization for melanoma and breast cancer, the value of $\Delta_a T^*$ is over 500.
+Therefore, in these cases, we are in the third scenario, in which aneuploidy provides resistance to the drug. % change b
 
@@
 Interestingly, increasing both the aneuploid division rate, $\lambda_a$, and the aneuploid death rate, $\mu_a$, such that the growth rate $\Delta_a$ remains constant, leads to a decrease in $T^*$, pushing the system to the second scenario.
-This is because increasing $\lambda_a$ causes a decrease in $T^*$ because  it increases the effective mutation rate $v\lambda_a$ (as mutations mostly occur during division) and a lineage does not have to survive as long in order to generate a successful mutant.
-Overall, the threshold tumor size $N_a^*$ is unaffected by the division rate $\lambda_a$ (i.e., $d \lambda_a T^*/d\lambda_a = 0$).
+This is because increasing $\lambda_a$ causes a decrease in $T^*$ as  it increases the effective mutation rate $v\lambda_a$ (as mutations mostly occur during division) and a lineage does not have to survive as long in order to generate a successful mutant.
+In the second scenario, the threshold tumor size $N_a^*$ is unaffected by the division rate $\lambda_a$ (i.e., $d \lambda_a T^*/d\lambda_a = 0$).
 Thus, if aneuploid cells rapidly die due to the drug but compensate by rapidly dividing, increasing the division rate will \emph{not} facilitate adaptation.
 %change 51
-This is consistent with experimental findings where aneuploidy confers resistance by decreasing the division rate which in turn decreases the death rate as certain drugs affect cancer cells during division~\citep{replogle2020aneuploidy}.
+%This is consistent with experimental findings where aneuploidy confers resistance by decreasing the division rate which in turn decreases the death rate as certain drugs affect cancer cells during division~\citep{replogle2020aneuploidy}.
 %change 52
@@
 
-As an example, we have parameterized our model using estimates for melanoma and triple-negative breast cancer cells (TNBC) under drug therapy from the literature. 
-We find that in these cases we are in the XXX scenario. %TODO fill XXX % change B
+As an example, we have parameterized our model using estimates for melanoma and triple-negative breast cancer (TNBC) cells under drug therapy from the literature. 
+We find that in these cases we are in the third scenario, in which aneuploidy provides at least partial resistance to the drug. 
+It remains to be seen which tumor type and drug combinations produce tolerant and stationary aneuploidies. % change B
 We have also compared the probability of evolutionary rescue between melanoma and TNBC (black and color lines, respectively, in \Cref{rescue_prob}B).
@@
 The ``stationary'' scenario, in which the aneuploid growth rate is close to zero in the presence of drug, is particularly interesting.
-Although this may seem like a small region of parameter space in a general model of evolutionary rescue (Fig.~\ref{rescue_threshold}A, inset), the example of certain aneuploidies in cancer suggests it could be biologically significant.
-% TODO if change B gives stationary scenario, mention it here; otherwise, say that other cancer might produce this
+Although this may seem like a small region of parameter space in a general model of evolutionary rescue (Fig.~\ref{rescue_threshold}A, inset), it could be biologically significant for some tumor under specific drug therapies.
 
@@
 \bibliographystyle{agsm}
-\bibliography{evo2022}
+\bibliography{ms}
 

### 2025-01-22 ms.tex
revise TNBC again

@@
 
-In some of the analysis, we consider breast cancer. 
-\citet{salehi2021clonal} studied multiple TNBC PDX clones (triple-negative breast cancer patient-derived xenografts).
-Their results suggest that aneuploidy and other copy number alterations provid resistance to the drug cisplatin.  
-Thus, we assume that aneuploid breast cancer cells are are partially or even completely resistant, i.e., $\Delta_m \ge \Delta_a > 0$, and that the division and death rates of aneuploid cells in the presence of cisplatin is the same as those of  sensitive breast cancer cells in the absence of the drug.
-The growth rate of breast cancer cells in the absence of drug is $0.0085$ per day \citep{spratt1996rates}.
+We also consider breast cancer cells, relying on \citet{salehi2021clonal}, who studied multiple TNBC PDX clones (triple-negative breast cancer patient-derived xenografts.)
+Their results suggest that aneuploidy and other copy number alterations provide resistance to the drug cisplatin.  
+Thus, we assume that aneuploid breast cancer cells are partially or even fully drug-resistant, i.e., $\Delta_m \ge \Delta_a > 0$.
+We chose three clones for which \cite{salehi2021clonal} report their relative Wrightian fitness, $1+s$, in the presence of the drug cisplatin: TNBC-SA609 clone A with $1+s=1.047$, TNBC-SA1035 clone H with $1+s=1.02$, and TNBC-SA535 clone H with $1+s=1.01$.
+The growth rate $\Delta$, or the Malthusian fitness, is the log of the Wrightian fitness \citep{Wu2013a}. 
+We therefore have the following set of equations: $0.047 = \exp{(\Delta_{SA609}-\Delta_s)}, 0.02=\exp{(\Delta_{SA1035}-\Delta_s)}, 0.01=\exp{(\Delta_{SA535}-\Delta_s)}$.
+The growth rate of breast cancer cells in the absence of the drug is $0.0085$ per day \citep{spratt1996rates}.
 The doubling time of breast cancer cells in the absence of cellular death is 8.2 days \citep{rew2000cell}.
-Thus, the division rate of aneuploid cells is $\lambda_a=\log{(2)}/8.2=0.0845$ per day, and their death rate is $\mu_a=0.076$ per day.
-To determine the division and death rates of sensitive cells, we chose TNBC-SA535 clone H from \citet{salehi2021clonal}, which has a relative fitness of $\approx 1.011$.
-Thus, the selection coefficient is the ratio of the aneuploid and sensitive Wrightian fitness values minus 1 \citep{chevin2011measuring}.
-The Malthusian fitness is the log of the Wrightian fitness \citep{Wu2013a}.
-Thus, $\Delta_s = \log{0.9975} -0.0025$.
-We assume cisplatin affects cancer cells by increasing their death rate but does not affect the their division rate, hence $\lambda_s=0.08$ and $\mu_s=0.0825$ per day.
-A similar analysis for TNBC-SA609 clone A and TNBC-SA1035 clone H from \citet{salehi2021clonal} gives $\lambda_s=0.08, \mu_s=0.12$ and $\lambda_s=0.08, \mu=0.1$, respectively.
-% TODO it does not make sense that when you change aneuploid clones what changes is the sensitive growth rate - the sensitive growth rate should be constant and the aneuploid clone growth rate should change...
+Thus, the division rate of breast cancer cells in the absence of the drug is $\log{(2)}/8.2=0.0845$ per day, and their death rate is $0.076$ per day.
+As all three clones grow in the presence of the drug, we set the growth rate of the least fit, TNBC-SA535 clone H, to the growth rate of breast cancer cells in the absence of the drug, $\Delta_{SA535}=0.0085$, and then solve the set of equations.
+% TODO needs to be the most fit, wait for Remus
+We further assume the drug affects death rather than division.
+We therefore have $\lambda_s=\lambda_a=0.0845$ for all clones,  $\mu_{SA609} = 0.0415$, $\mu_{SA1035}=0.0667$, $\mu_{SA535}=0.085$, and $\mu_s=0.085$.
 
@@
     \hline
-     & Name & Value & Units & References \\ \hline
-    $\lambda_s$ & Sensitive division rate& 0.08 & 1/days  & \citet{salehi2021clonal} \\ \hline
-    $\mu_s$ & Sensitive death rate& $0.0825, 0.1, 0.12$ & 1/days  & \citet{salehi2021clonal} \\ \hline
-    $\lambda_a$  & Aneuploid division rate & $0.0845$ & 1/days  & \citet{spratt1996rates,rew2000cell}  \\ \hline
-    $\mu_a$ & Aneuploid death rate & $0.076$ & 1/days  & \citet{spratt1996rates,rew2000cell}  \\ \hline
+     & Name & Value & Units \\ \hline
+    $\lambda_s$ & Sensitive division rate& 0.0848& 1/days \\ \hline
+    $\mu_s$ & Sensitive death rate& $0.085$ & 1/days   \\ \hline
+    $\lambda_a$  & Aneuploid division rate & $0.0845$ & 1/days    \\ \hline
+    $\mu_a$ & Aneuploid death rate & $0.0415$, $0.0667$, $0.085$ & 1/days   \\ \hline
   \end{tabular}
 \caption{\textbf{Model parameters: Breast cancer.} 
+The derivation of these values, described in the methods section, uses fitness estimates from \citet{salehi2021clonal}.
 Parameters not listed here remain unchanged from the values provided in \cref{table1}.

### 2025-01-22 ResponseLetter.tex
respond to last open comment

@@
 Additionally, they observed the growth rate of cancer cells before treatment to be 0.01, which we use as the growth rate of mutant cells, which are resistant to the drug.
-Thus, we use $\mu_m=0.1-0.01=0.09$ per day as the death rate for mutant cells."
+Thus, we use $\mu_m=0.1-0.01=0.09$ per day as the death rate for mutant cells." \\
+We also have another paragraph in the same section about parameterization of the model using evidence from breast cancer.
 }
@@
 
-\textcolor{red}{
-TODO change B
+\textcolor{blue}{
+Thank you for this remark, you are correct to note that this statement is in opposition to some of our results.
+We have removed this statement and instead, we now write "The “stationary” scenario, in which the aneuploid growth rate is close to zero in the presence of drug, is particularly interesting.
+Although this may seem like a small region of parameter space in a general model of evolutionary rescue (Fig.
+4A, inset), it could be biologically significant for some tumors under specific drug therapies.
+The current literature may be biased towards aneuploidies that provide drug resistance.
+Our results suggest that identifying aneuploidies that are tolerant or stationary may be worthwhile, as these are also expected to enhance the probability of rescue (Figure 3) and extend the window of opportunity for rescue (Figure 6)."\\
+We also added the following sentences to the Discussion.
+In line XXX: "Our results show that the probability of evolutionary rescue increases with the initial tumor size $N$, the drug-sensitive growth rate $\Delta_s$, the mutation rate $v$, and the aneuploidy rate $u$.
+\emph{The latter indicates that aneuploidy, even when it only provides tolerance, increases the probability that the tumor will be rescued.}" \\
+And in line XXX: "As an example, we have parameterized our model using estimates for melanoma and triple-negative breast cancer (TNBC) cells under drug therapy from the literature.
+We find that in these cases we are in the third scenario, in which aneuploidy provides at least partial resistance to the drug.
+It remains to be seen which tumor type and drug combinations produce tolerant and stationary aneuploidies."\\
+The following sentences were already in the Discussion (line XXX): "In this scenario, aneuploidy provides two advantages.
+First, it delays the extinction of the population, providing more time for the appearance of the resistance mutation.
+Second, it increases the population size relative to a drug-sensitive population, providing more cells in which mutations can occur. 
+Together, this increases the cumulative number of mutants that arise." and "We find that aneuploidy can significantly affect evolutionary rescue as it reduces the threshold tumor size by at least an order of magnitude even when aneuploidy only provides tolerance."
 }

### 2025-01-22 ms.tex
respond to last open comment

@@
     \frac{\abs{\Delta_a}}{u\lambda_a} ,&
-  \Delta_a T^* \ll -1 ,\\ 
+  \Delta_a T^* \ll -1 \text{ (tolerant aneuploids)} ,\\ 
   \frac{1}{u}\left(v  \frac{\Delta_m}{\lambda_m}\right)^{1/2} ,&
-  -1 \ll \Delta_a T^* \ll 1  ,\\ 
+  -1 \ll \Delta_a T^* \ll 1 \text{ (stationary aneuploids)} ,\\ 
   v \frac{\Delta_m}{\lambda_m}  \left(u\frac{\Delta_a}{\lambda_a}\right)^{-1}  ,&
-   \Delta_a T^* \gg 1 .
+   \Delta_a T^* \gg 1 \text{ (resistant aneuploids)} .
   \end{cases}
@@
     \frac{\abs{\Delta_a}}{\tilde{u}\lambda_a} ,&
-  \Delta_a T^* \ll -1 ,\\ 
+  \Delta_a T^* \ll -1 \text{ (tolerant aneuploids)} ,\\ 
   \frac{1}{\tilde{u}}\left(v \frac{\Delta_m}{\lambda_m}\right)^{1/2} ,&
-  -1 \ll \Delta_a T^* \ll 1  ,\\ 
+  -1 \ll \Delta_a T^* \ll 1 \text{ (stationary aneuploids)} ,\\ 
   v \frac{\Delta_m}{\lambda_m} \left(\tilde{u}\frac{\Delta_a}{\lambda_a}\right)^{-1}  ,&
-   \Delta_a T^* \gg 1 .
+   \Delta_a T^* \gg 1\text{ (resistant aneuploids)} .
   \end{cases}
@@
 Our results show that the probability of evolutionary rescue increases with the initial tumor size $N$, the drug-sensitive growth rate $\Delta_s$, the mutation rate $v$, and the aneuploidy rate $u$.
+The latter indicates that aneuploidy, even when it only provides tolerance, increases the probability that the tumor will be rescued (\Cref{figureAneuploidy}).
 
@@
 Similar to \citet{osmond_genetic_2020}, we find that two-step rescues can be more likely than simple one-step rescues. 
-The ``stationary'' scenario, in which the aneuploid growth rate is close to zero in the presence of drug, is particularly interesting.
-Although this may seem like a small region of parameter space in a general model of evolutionary rescue (Fig.~\ref{rescue_threshold}A, inset), it could be biologically significant for some tumor under specific drug therapies.
+The ``stationary'' scenario, in which the aneuploid growth rate is close to zero in the presence of the drug, is particularly interesting.
+Although this may seem like a small region of parameter space in a general model of evolutionary rescue (Fig.~\ref{rescue_threshold}A, inset), it could be biologically significant for some tumors under specific drug therapies.
+The current literature may be biased towards aneuploidies that provide drug resistance. 
+Our results suggest that identifying aneuploidies that are tolerant or stationary may be worthwhile, as these are also expected to enhance the probability of rescue (\Cref{rescue_prob}) and extend the window of opportunity for rescue (\Cref{cdffig}).
 
@@
 %change 56
-(B) Probability that a mutant cancer cell population has not reached size $N$ at time $t$.
+(B) Probability that a mutant cancer cell population has not reached size $N$ at time $t$ when aneuploidy provides tolerance.
 Green line: $N=10^6$ (small tumor).

### 2025-01-22 ResponseLetter.tex
revise responses and add line numbers

--- 2025-01-19 ResponseLetter.tex
@@
 \textcolor{blue}{ % change 53 
-We have added a comment in line XXX: "We assume independence between clonal lineages starting from an initial population of $N$ sensitive cells (we check the effect of density-dependent growth on our results below) and therefore, we use multi-type branching processes to model the dynamics of the cancer cell population.
+We have added a comment in line 239: "We assume independence between clonal lineages starting from an initial population of $N$ sensitive cells (we check the effect of density-dependent growth on our results below) and therefore, we use multi-type branching processes to model the dynamics of the cancer cell population.
 Multi-type branching process models the growth and evolution of populations with distinct types, capturing dynamics like mutation and selection.
@@
 \textcolor{blue}{ % change 54
-We have updated the "Parameterization" section: "To parametrize the simulations, we assume that the cells under consideration are melanoma cells and rely on Rew and Wilson (2000) and Bozic et al.
+We have updated the "Parameterization" section (line 159): "We parametrize most of our simulations by considering melanoma cells and rely on Rew and Wilson (2000) and Bozic et al.
 (2013) for the division and death rates, respectively.
@@
 
-\textcolor{red}{
-TODO change B
+\textcolor{blue}{
+Thank you for this remark, you are correct to note that this statement is in opposition to some of our results.
+We have removed this statement and instead, we now write (line 523): "The “stationary” scenario, in which the aneuploid growth rate is close to zero in the presence of drug, is particularly interesting.
+Although this may seem like a small region of parameter space in a general model of evolutionary rescue (Fig.
+4A, inset), it could be biologically significant for some tumors under specific drug therapies.
+Our results suggest that identifying aneuploidies that are tolerant or stationary may be worthwhile, as these are also expected to enhance the probability of rescue (Figure 3) and extend the window of opportunity for rescue (Figure 6)."\\
+We also added the following sentences to the Discussion.
+In line 468, emphasis on new text): "Our results show that the probability of evolutionary rescue increases with the initial tumor size $N$, the drug-sensitive growth rate $\Delta_s$, the mutation rate $v$, and the aneuploidy rate $u$.
+\emph{The latter indicates that aneuploidy, even when it only provides tolerance, increases the probability that the tumor will be rescued.}" \\
+And in line 493: "As an example, we have parameterized our model using estimates for melanoma and triple-negative breast cancer (TNBC) cells under drug therapy from the literature.
+We find that in these cases we are in the third scenario, in which aneuploidy provides at least partial resistance to the drug.
+It remains to be seen which tumor type and drug combinations produce tolerant and stationary aneuploidies."\\
+The following sentences were already in the Discussion (line 476): "In this scenario, aneuploidy provides two advantages.
+First, it delays the extinction of the population, providing more time for the appearance of the resistance mutation.
+Second, it increases the population size relative to a drug-sensitive population, providing more cells in which mutations can occur. 
+Together, this increases the cumulative number of mutants that arise." and in line 481: "We find that aneuploidy can significantly affect evolutionary rescue as it reduces the threshold tumor size by at least an order of magnitude even when aneuploidy only provides tolerance."
 }
@@
 \textcolor{blue}{%change 48
-We have rewritten the sentence about the mean detection time (line XXX): "We find that for small and intermediate-sized tumors the mean detection time is approximately equal to the mean recurrence time (i.e., $\tau_a^r\approx\tau_a^{M}$ for $N<N_m^*$).
+We have rewritten the sentence about the mean detection time (line 453): "We find that for small and intermediate-sized tumors the mean detection time is approximately equal to the mean recurrence time (i.e., $\tau_a^r\approx\tau_a^{M}$ for $N<N_m^*$).
 "
@@
 \textcolor{blue}{%change 2
-We have added an explanation of ODE (line XXX): "In this case the dynamics of the number of mutant cells is deterministic and can therefore be modeled by a system of ordinary differential equations (ODE), which describe how the number of mutant cells changes over time by its time derivative (eq.~D2)."
+We have added an explanation of ODE (line 381): "In this case the dynamics of the number of mutant cells is deterministic and can therefore be modeled by a system of ordinary differential equations (ODE), which describe how the number of mutant cells changes over time by its time derivative (eq.~D2)."
 }
@@
 Standing variation actually increases with the population size, whereas the effect of drift decreases with the population size.
-We now explain this in line XXX: "We assume that $c$ is not too small (Table 1), so that standing variation is generated by a balance between the generation of anuploids and selection against them, without the effect of genetic drift."
+We now explain this in line 130: "We assume that $c$ is not too small (Table 1), so that standing variation is generated by a balance between the generation of anuploids and selection against them, without the effect of genetic drift."
 }
@@
 \textcolor{blue}{ % change 40
-We now ground this assumption in results from the literature (line XXX): "The division rate is obtained as $\lambda=\log{2} / T \approx 0.1$ per day.
+We now ground this assumption in results from the literature (line 163): "The division rate is obtained as $\lambda=\log{2} / T \approx 0.1$ per day.
 We take this to be the division rate for sensitive and assume that mutant cells, which are resistant to therapy, have the same division rate.
@@
 They have basically done all the work for us.
-We have incorporated a review of the literature in the Introduction (line XXX), and refer to these papers throughout the Results at the equations mentioned.
+We have incorporated a review of the literature in the Introduction (lines 55-87), and refer to these papers throughout the Results at the equations mentioned.
 The only slight changes we have made are omitting the follow-ups to Antia et al 2003 and instead including Iwasa et al 2003 Proc B, which we really should have mentioned before and is arguably the closest thing in the literature to this manuscript.
@@
 As for the fitness difference between euploid and aneuploid mutants, this is a good question and the answer probably varies depending on the drug, the chromosome, and other factors.
-We now address both your questions under "Directions for future reseach" in line XXX: "Furthermore, loss of extra-chromosome may occur at a high rate (for estimates in yeast, see Hose et al.
+We now address both your questions in the Discussion, in line 542: "Furthermore, loss of extra-chromosome may occur at a high rate (for estimates in yeast, see Hose et al.
 (2024)) and the fitness of a euploid cell with a resistance mutation may be higher than that of an aneuploid cell with the same mutation.
@@
 \textcolor{blue}{%change 45
-We have added reference to the recommended synthetic reviews in line XXX.
+We have added reference to the recommended synthetic reviews in line 61.
 }
@@
 \textcolor{blue}{%change 28
-We have  moved the statement at the beginning of the paragraph, line XXX. 
+We have  moved the statement at the beginning of the paragraph, line 60. 
 }
@@
 \textcolor{blue}{%change 41
-We have changed the statement to explain the concepts of soft and hard selective sweeps and added a reference to Osmond and Coop 2020 (line XXX): "Evolutionary rescue may occur when a beneficial mutation appears in a single copy, leading to hard selective sweeps, or when multiple mutations appear independently in different individuals, leading to soft selective sweeps.
-Wilson et al.
-(2017) have shown that evolutionary rescue is significantly enhanced by soft selective sweeps when multiple mutations contribute towards evolutionary rescue.
-Additionally, Osmond and Coop (2020) have calculated the probability that rescue is soft and the expected number of rescue mutations in different scenarios.
-Evolutionary rescue that requires two successive mutations (i.e., two steps) has been investigated by Martin et al.
-(2013), who tested their model's predictions with data from yeast and bacteria experiments.
-Iwasa et al.
-(2003, 2004) used multi-type branching process theory to approximate the probability that a population under strong selective pressure can survive extinction via succesive mutations.
-"
+We have removed the statement "most models focus...". Instead, we include a review of the evolutionary rescue literature (lines 55-87).
 }
@@
 \textcolor{blue}{%change 36
-We repharased to (line XXX): "For the aneuploid growth rate to be intermediate between those of sensitive and mutant cells, $\Delta_s\ll\Delta_a\ll\Delta_m$, we set the aneuploid division rate to be $0.06 \le \lambda_a \le 0.1$."
+We repharased to (line line 175): "For the aneuploid growth rate to be intermediate between those of sensitive and mutant cells, $\Delta_s\ll\Delta_a\ll\Delta_m$, we set the aneuploid division rate to be $0.06 \le \lambda_a \le 0.1$."
 }
@@
 \textcolor{blue}{%change 5
-We revised the text to make it clear that we estimate $c$ from an experimental result and $f$ from $c$ (line XXX): "To estimate fitness cost of aneuploidy, $c$, we note that Lukow et al.
+We revised the text to make it clear that we estimate $c$ from an experimental result and $f$ from $c$ (line 208): "To estimate fitness cost of aneuploidy, $c$, we note that Lukow et al.
 (2021) mixed sensitive and aneuploid A375 melanoma cells at $1:1$ ratio, cultured them in a drug-free environment, and observed the ratio evolve as a function of time with the aneuploid cells declining to $15\%$ after 24 days.
@@
 
-\textcolor{blue}{We have now added density-dependence for the sensitives. The results are unchanged.}
+\textcolor{blue}{We have now added density-dependence for the sensitives (Eq. at line 226). The results are unchanged.}
 
@@
 \textcolor{blue}{ % change 33
-We now mention that eq. 2 (previously eq.3 ) is an alternative form of an eq. from Orr and Unckless 2008 (line XXX).
+We now mention that eq. 2 (previously eq.3 ) is an alternative form of an eq. from Orr and Unckless 2008 (line 261).
 }
@@
 \textcolor{blue}{%change 57
-We now note the similarity of eq.~3 (previously eq.~3) to Osmond et al 2020 and Iwasa et al 2004.
+We now note the similarity of eq.~3 (previously eq.~3) to Osmond et al 2020 and Iwasa et al 2004 (line 269).
 }
@@
 \textcolor{blue}{%change 51
-We have added the recommended comment (line XXX): "Interestingly, increasing both the aneuploid division rate, $\lambda_a$, and the aneuploid death rate, $\mu_a$, such that the growth rate $\Delta_a$ remains constant, leads to a decrease in $T^*$, pushing the system to the second scenario.
+We have added the recommended comment (line 303): "Interestingly, increasing both the aneuploid division rate, $\lambda_a$, and the aneuploid death rate, $\mu_a$, such that the growth rate $\Delta_a$ remains constant, leads to a decrease in $T^*$, pushing the system to the second scenario.
 This is because increasing $\lambda_a$ causes a decrease in $T^*$ as  it increases the effective mutation rate $v\lambda_a$ (as mutations mostly occur during division) and a lineage does not have to survive as long in order to generate a successful mutant."
@@
 \textcolor{blue}{ % change 58
-We believe that our model does fit with data on resistance rates, and the text now reads (line XXX): "Interestingly, the threshold between small and intermediate tumors, $N_a^*$, is similar to the tumor detection threshold of $4.19 \times 10^6$ cells for a wide variety of tumors~\citep{avanzini2019cancer}.
+We believe that our model does fit with data on resistance rates, and the text now reads (line 316): "Interestingly, the threshold between small and intermediate tumors, $N_a^*$, is similar to the tumor detection threshold of $4.19 \times 10^6$ cells for a wide variety of tumors~\citep{avanzini2019cancer}.
 We note that vemurafenib-treated melanomas (i.e.
@@
 \textcolor{blue}{%change 42
-We now mention that this equation has been previously shown by Orr and Unckless 2008 and by Martin et al 2013 in line XXX.
+We now mention that this equation (eq.~5 in the revision) has been previously shown by Orr and Unckless 2008 and by Martin et al 2013 in line 346.
 }
@@
 \textcolor{blue}{%change 46
-We have added the following text in line XXX: "For small tumors ($ N \ll N_a^*$), the mean rescue time is the two-step equivalent of the one-step result from Orr and Unckleass (2014; expectation of eq.~18).
+We have added the following text in line 393: "For small tumors ($ N \ll N_a^*$), the mean rescue time is the two-step equivalent of the one-step result from Orr and Unckleass (2014; expectation of eq.~18).
 The mean rescue time (conditioned on rescue occuring) is a decreasing function of the sensitive and aneuploid growth rates and independent of the other model parameters, including tumor size (blue line in Figure S6).
@@
 \textcolor{blue}{%change 26
-We have added the definition of $p_m$ in line XXX: "where $p_m$ is the probability that a lineage starting from a single mutant cell escapes stochastic extinction."
+We have added the definition of $p_m$ in line 421: "where $p_m$ is the probability that a lineage starting from a single mutant cell escapes stochastic extinction."
 }
@@
 
-\textcolor{blue}{The return time with new mutation (eqn 23) from Orr $\&$ Unckless 2014 can be written as:
+\textcolor{blue}{
+The return time with new mutation (eqn 23) from Orr $\&$ Unckless 2014 can be written as:
 \begin{align*}
@@
 \textcolor{blue}{% change 49
-The text now reads (line XXX): "Drugs that increase the sensitive death rate, $\mu_s$, but do not affect the sensitive division rate, $\lambda_s$, delay cancer recurrence.
+The text now reads (line 435): "Drugs that significantly increase the sensitive death rate, $\mu_s$, but do not affect the sensitive division rate, $\lambda_s$, delay cancer recurrence.
 Drugs that decrease the sensitive division rate delay cancer recurrence time even more, because they effectively decrease the mutation rate (assuming mutations occur during division)."
@@
 Thank you for this suggestions.
-We have revised the text (line XXX): "aneuploidy provides two advantages.
+We have revised the text (line 476): "...
+aneuploidy provides two advantages.
 First, it delays the extinction of the population, providing more time for the appearance of the resistance mutation.
@@
 \textcolor{blue}{%change x1
-Line XXX in the Introduction reads: "Futhermore, some proposed anti-cancer drugs elevate the missegregation rate to fight cancer cells (Lee et al.
+Line 52 in the Introduction reads: "Futhermore, some proposed anti-cancer drugs elevate the missegregation rate to fight cancer cells (Lee et al.
 2016), as an extremely high chromosome missegregation rate is incompatible with cell survival and proliferation."
@@
 
-\textcolor{blue}{We have added a paragraph in the Discussion on the connections to Iwasa et al 2003 and Osmond et al 2020, which we think are the two closest parallels for the main results. 
-For the other more detailed expressions, we have just added the references in the Results as they come up.}
+\textcolor{blue}{We have added a paragraph in the Discussion (line 520) on the connections to Iwasa et al 2003 and Osmond et al 2020, which we think are the two closest parallels for the main results.
+For the other more detailed expressions, we have added the references in the Results as they come up.}
 
@@
 
+\textcolor{blue}{Thank you for the positive review of our manuscript.}
+
 Major points
@@
 
-\textcolor{blue}{Following the reviewer's suggestion, we have evaluated our model using parameters from breast cancer (PMID: 34163070), see Methods (line XXX), Table 2, and Figure 3B, which also appear below.
-We have added the following text to the discussion: "We compared the probability of evolutionary rescue in melanoma cells and triple-negative breast cancer cells (TNBC) under drug threapy (black and color lines, respectively, in Figure 3B).
+\textcolor{blue}{Following the reviewer's suggestion, we have evaluated our model using parameters from breast cancer (PMID: 34163070), see Methods (line 180), Table 2, and Figure 3B, which also appear below.\\
+We have aldo added the following text to the Discussion (line 493): "As an example, we have parameterized our model using estimates for melanoma and triple-negative breast cancer (TNBC) cells under drug therapy from the literature.
+We find that in these cases we are in the third scenario, in which aneuploidy provides at least partial resistance to the drug. 
+It remains to be seen which tumor type and drug combinations produce tolerant and stationary aneuploidies. % change B
+We have also compared the probability of evolutionary rescue between melanoma and TNBC (black and color lines, respectively, in Figure 3B).
 This comparison suggests that TNBC has a higher probability of relapsing compared to melanoma, because smaller tumors have higher resuce probability.
 Indeed, the probability of relapse in TNBC patients is higher than 50\% in the first 3-5 years after diagnosis \citep{taushanova2023synchronous}, whereas in melanoma patients it is approximately 10-20\% in the first 5 years \citep{wan2022prediction,von_schuckmann2019risk}.
+% change A
 }% change A
@@
 
-\textcolor{blue}{In our model, we are interested in a specific type of aneuploidy which increases the fitness of cancer cells in an environment affected by anti-cancer drugs.
-As a consequence, we catalogued all other forms of aneuploids as sensitive.}
+\textcolor{blue}{
+We have removed the misleading sentence "aneuploidy will either not play any role in evolutionary rescue or will be the main driver of adaptation", which is incorrect.
+Instead, we write (line 526): "Our results suggest that identifying aneuploidies that are tolerant or stationary may be worthwhile, as these are also expected to enhance the probability of rescue (Figure 3) and extend the window of opportunity for rescue (Figure 6)."\\
+Furthermore, In our model, we are interested in a specific type of aneuploidy which increases the fitness of cancer cells in an environment affected by anti-cancer drugs.
+As a consequence, we catalogued all other forms of aneuploids as sensitive.
+These aneuploidies can appear in patient cohorts because, for example, they drive cancer itself or increase genome instability, which leads to mutations and abberations that promote cancer.
+}
 
@@
 
-\textcolor{blue}{We consider the fitness of the sensitive, aneuploid or mutant cells to be the average of each population.
-As a result, mutations that are deleterious are taken into consideration through the mean fitness of each population.
-We have added this comment in the "Parametrisation" section change 50}
+\textcolor{blue}{Our focus is on mutations that confer resistance.
+We assume deleterious mutations can occur with equal rates in all genotypes and therefore would not change the dynamics.
+We now explain these assumptions in line 215: "Note that when we refer to drug-sensitive cells, we include those cells that have any aneuploidy that does not affect fitness in the presence of the drug.
+Additionally, we focus on mutations that confer resistance, neglecting deleterious mutations (which are more common).
+We assume deleterious mutations and other aneuplidies occur at similar rates across genotypes and therefore neglect their effect on the dynamics."
+}
 
@@
 \textcolor{blue}{The variation in our parameters in noted in Table 1 and the new Table 2 and is obtained from the papers which are cited there.} 
-
-Associate Editor Comments:
+\\\\
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\textbf{Associate Editor Comments}
 
@@
 A sufficient revision must, in my view, properly connect this work to the existing theoretical literature on evolutionary rescue (as noted by reviewer $\#2$) and to the broader empirical literature on cancer about multi-step drug resistance (as noted by reviewer $\#3$).
+
+\textcolor{blue}{
+We have have revised the introduction and discussion accordingly, see lines 55-87, 520-529.
+}
 
@@
 
-\textcolor{blue}{The manuscript has over 300 mentions of $\Delta$, and we do not want to introduce any errors by changing it.}
+\textcolor{blue}{The manuscript has over 300 mentions of $\Delta$, plus mentions in the figures, and we do not want to introduce any errors by changing it.}
 

### 2025-01-22 ms.tex
revise responses and add line numbers

@@
 
-We note that when we refer to drug-sensitive cells, we include those cells that have any aneuploidy other than trisomy in Chr 2 and 6, as those are the aneuploid cells that are hypothesized to have higher fitness in the environment altered by drugs such as vemurafenib.
-Additionally, in our model, mutations only confer resistance, however, most mutations are deleterious.
-We consider the fitness of the sensitive, aneuploid or mutant cells to be the average of each population.
-As a result, mutations that are deleterious are taken into consideration through the mean fitness of each population.
-%change 50
-
-All the parameters discussed above are shown in \Cref{table1}.
+Note that when we refer to drug-sensitive cells, we include those cells that have any aneuploidy that does not affect fitness in the presence of the drug.
+Additionally, we focus on mutations that confer resistance, neglecting deleterious mutations (which are more common).
+We assume deleterious mutations and other aneuplidies occur at similar rates across genotypes and therefore neglect their effect on the dynamics.
+
+All the parameters discussed above are shown in \Cref{table1}.\\
 
@@
 Our results show that the probability of evolutionary rescue increases with the initial tumor size $N$, the drug-sensitive growth rate $\Delta_s$, the mutation rate $v$, and the aneuploidy rate $u$.
-The latter indicates that aneuploidy, even when it only provides tolerance, increases the probability that the tumor will be rescued (\Cref{figureAneuploidy}).
+The latter indicates that aneuploidy, even when it only provides tolerance, increases the probability that the tumor will be rescued (\Cref{rescue_prob}).
 
@@
 Although this may seem like a small region of parameter space in a general model of evolutionary rescue (Fig.~\ref{rescue_threshold}A, inset), it could be biologically significant for some tumors under specific drug therapies.
-The current literature may be biased towards aneuploidies that provide drug resistance. 
 Our results suggest that identifying aneuploidies that are tolerant or stationary may be worthwhile, as these are also expected to enhance the probability of rescue (\Cref{rescue_prob}) and extend the window of opportunity for rescue (\Cref{cdffig}).

### 2025-01-22 ResponseLetter.tex
fixed quote marks

@@
 \textcolor{blue}{ % change 53 
-We have added a comment in line 239: "We assume independence between clonal lineages starting from an initial population of $N$ sensitive cells (we check the effect of density-dependent growth on our results below) and therefore, we use multi-type branching processes to model the dynamics of the cancer cell population.
+We have added a comment in line 239: ``We assume independence between clonal lineages starting from an initial population of $N$ sensitive cells (we check the effect of density-dependent growth on our results below) and therefore, we use multi-type branching processes to model the dynamics of the cancer cell population.
 Multi-type branching process models the growth and evolution of populations with distinct types, capturing dynamics like mutation and selection.
-In evolutionary rescue, it predicts the probability of population survival by tracking adaptive mutations that emerge and spread under environmental stress."
+In evolutionary rescue, it predicts the probability of population survival by tracking adaptive mutations that emerge and spread under environmental stress.''
 }
@@
 \textcolor{blue}{ % change 54
-We have updated the "Parameterization" section (line 159): "We parametrize most of our simulations by considering melanoma cells and rely on Rew and Wilson (2000) and Bozic et al.
+We have updated the ``Parameterization'' section (line 159): ``We parametrize most of our simulations by considering melanoma cells and rely on Rew and Wilson (2000) and Bozic et al.
 (2013) for the division and death rates, respectively.
@@
 Additionally, they observed the growth rate of cancer cells before treatment to be 0.01, which we use as the growth rate of mutant cells, which are resistant to the drug.
-Thus, we use $\mu_m=0.1-0.01=0.09$ per day as the death rate for mutant cells." \\
+Thus, we use $\mu_m=0.1-0.01=0.09$ per day as the death rate for mutant cells.'' \\
 We also have another paragraph in the same section about parameterization of the model using evidence from breast cancer.
@@
 \textcolor{blue}{%change 31
-We have made the change to use the hyphen in all mentions of "multi-type branching" such that it is consistent with textbooks such as Athreya $\&$ Ney (1972).
+We have made the change to use the hyphen in all mentions of ``multi-type branching'' such that it is consistent with textbooks such as Athreya $\&$ Ney (1972).
 }
@@
 Is it specific to this particular chemotherapy, small tumors, only certain types of cancer?
-In conjunction with this, there are places in the manuscript where you say things like "for most parameter values aneuploidy will either play no role in evolutionary rescue or will be the main driver of adaptation without requiring any mutation (one step)." Both parts of the quoted sentence stand in opposition to one of the more interesting scenarios found in your modeling, which is that aneuploidy can buy a tumor cell population enough time to accumulate a mutation that leads to evolutionary rescue.
+In conjunction with this, there are places in the manuscript where you say things like ``for most parameter values aneuploidy will either play no role in evolutionary rescue or will be the main driver of adaptation without requiring any mutation (one step).'' Both parts of the quoted sentence stand in opposition to one of the more interesting scenarios found in your modeling, which is that aneuploidy can buy a tumor cell population enough time to accumulate a mutation that leads to evolutionary rescue.
 Like in point 1, making clear the broader implications of your findings, or the limited scenarios in which they might be applicable, pushes this paper over the top into perfection territory because it roots your reader into something relatable, manageable, or applicable to patient care.
-If the answer is "I don't know how applicable this finding is or what the broader implications might be," you could then add something to the future research section that talks about how this might be addressed experimentally or theoretically.
+If the answer is ``I don't know how applicable this finding is or what the broader implications might be,'' you could then add something to the future research section that talks about how this might be addressed experimentally or theoretically.
 It's ok to not know for sure, but you shouldn't pass the opportunity to talk about it in limits and future research.
@@
 Thank you for this remark, you are correct to note that this statement is in opposition to some of our results.
-We have removed this statement and instead, we now write (line 523): "The “stationary” scenario, in which the aneuploid growth rate is close to zero in the presence of drug, is particularly interesting.
+We have removed this statement and instead, we now write (line 523): ``The `stationary' scenario, in which the aneuploid growth rate is close to zero in the presence of drug, is particularly interesting.
 Although this may seem like a small region of parameter space in a general model of evolutionary rescue (Fig.
 4A, inset), it could be biologically significant for some tumors under specific drug therapies.
-Our results suggest that identifying aneuploidies that are tolerant or stationary may be worthwhile, as these are also expected to enhance the probability of rescue (Figure 3) and extend the window of opportunity for rescue (Figure 6)."\\
+Our results suggest that identifying aneuploidies that are tolerant or stationary may be worthwhile, as these are also expected to enhance the probability of rescue (Figure 3) and extend the window of opportunity for rescue (Figure 6).''\\
 We also added the following sentences to the Discussion.
-In line 468, emphasis on new text): "Our results show that the probability of evolutionary rescue increases with the initial tumor size $N$, the drug-sensitive growth rate $\Delta_s$, the mutation rate $v$, and the aneuploidy rate $u$.
-\emph{The latter indicates that aneuploidy, even when it only provides tolerance, increases the probability that the tumor will be rescued.}" \\
-And in line 493: "As an example, we have parameterized our model using estimates for melanoma and triple-negative breast cancer (TNBC) cells under drug therapy from the literature.
+In line 468, emphasis on new text): ``Our results show that the probability of evolutionary rescue increases with the initial tumor size $N$, the drug-sensitive growth rate $\Delta_s$, the mutation rate $v$, and the aneuploidy rate $u$.
+\emph{The latter indicates that aneuploidy, even when it only provides tolerance, increases the probability that the tumor will be rescued.}'' \\
+And in line 493: ``As an example, we have parameterized our model using estimates for melanoma and triple-negative breast cancer (TNBC) cells under drug therapy from the literature.
 We find that in these cases we are in the third scenario, in which aneuploidy provides at least partial resistance to the drug.
-It remains to be seen which tumor type and drug combinations produce tolerant and stationary aneuploidies."\\
-The following sentences were already in the Discussion (line 476): "In this scenario, aneuploidy provides two advantages.
+It remains to be seen which tumor type and drug combinations produce tolerant and stationary aneuploidies.''\\
+The following sentences were already in the Discussion (line 476): ``In this scenario, aneuploidy provides two advantages.
 First, it delays the extinction of the population, providing more time for the appearance of the resistance mutation.
 Second, it increases the population size relative to a drug-sensitive population, providing more cells in which mutations can occur. 
-Together, this increases the cumulative number of mutants that arise." and in line 481: "We find that aneuploidy can significantly affect evolutionary rescue as it reduces the threshold tumor size by at least an order of magnitude even when aneuploidy only provides tolerance."
+Together, this increases the cumulative number of mutants that arise.'' and in line 481: ``We find that aneuploidy can significantly affect evolutionary rescue as it reduces the threshold tumor size by at least an order of magnitude even when aneuploidy only provides tolerance.''
 }
@@
 4.
-I would revisit some of your figures to make sure groups/categories are not colored in contradictory ways from one graph to the next (i.e., like "no aneuploidy" being one color on one graph and a different color in another graph or different categories ("no aneuploidy" vs.
-"aneuploid" vs.
-"stationary") using the same color in different figures).
+I would revisit some of your figures to make sure groups/categories are not colored in contradictory ways from one graph to the next (i.e., like ``no aneuploidy'' being one color on one graph and a different color in another graph or different categories (``no aneuploidy'' vs.
+``aneuploid'' vs.
+``stationary'') using the same color in different figures).
 
@@
 \textcolor{blue}{%change 48
-We have rewritten the sentence about the mean detection time (line 453): "We find that for small and intermediate-sized tumors the mean detection time is approximately equal to the mean recurrence time (i.e., $\tau_a^r\approx\tau_a^{M}$ for $N<N_m^*$).
-"
+We have rewritten the sentence about the mean detection time (line 453): ``We find that for small and intermediate-sized tumors the mean detection time is approximately equal to the mean recurrence time (i.e., $\tau_a^r\approx\tau_a^{M}$ for $N<N_m^*$).''
 }
@@
 \textcolor{blue}{%change 2
-We have added an explanation of ODE (line 381): "In this case the dynamics of the number of mutant cells is deterministic and can therefore be modeled by a system of ordinary differential equations (ODE), which describe how the number of mutant cells changes over time by its time derivative (eq.~D2)."
+We have added an explanation of ODE (line 381): ``In this case the dynamics of the number of mutant cells is deterministic and can therefore be modeled by a system of ordinary differential equations (ODE), which describe how the number of mutant cells changes over time by its time derivative (eq.~D2).''
 }
@@
 Are the tests for standing genetic variation essentially genetic drift?
-Variation is present and in a tumor cell population wildly fluctuating in size due to treatment you just hit on a "right" variant stochastically when tumor size is small (less cells) that is resistant and it's driven to fixation?
+Variation is present and in a tumor cell population wildly fluctuating in size due to treatment you just hit on a ``right'' variant stochastically when tumor size is small (less cells) that is resistant and it's driven to fixation?
 If so, using genetic drift in this section is rooting language and another way of describing alternative scenarios/hypotheses for tumor rebound.
@@
 Standing variation actually increases with the population size, whereas the effect of drift decreases with the population size.
-We now explain this in line 130: "We assume that $c$ is not too small (Table 1), so that standing variation is generated by a balance between the generation of anuploids and selection against them, without the effect of genetic drift."
+We now explain this in line 130: ``We assume that $c$ is not too small (Table 1), so that standing variation is generated by a balance between the generation of anuploids and selection against them, without the effect of genetic drift.''
 }
@@
 9.
-Do we know if it's reasonable to assume the division rates between cell types (i.e., "sensitive" vs.
-"mutant") are the same?
+Do we know if it's reasonable to assume the division rates between cell types (i.e., ``sensitive'' vs.
+``mutant'') are the same?
 Maybe ground that assumption into something we know, or just say we are not sure if that is reasonable or not.
@@
 \textcolor{blue}{ % change 40
-We now ground this assumption in results from the literature (line 163): "The division rate is obtained as $\lambda=\log{2} / T \approx 0.1$ per day.
+We now ground this assumption in results from the literature (line 163): ``The division rate is obtained as $\lambda=\log{2} / T \approx 0.1$ per day.
 We take this to be the division rate for sensitive and assume that mutant cells, which are resistant to therapy, have the same division rate.
-Indeed, doubling times for tumors after relapse is lower than that of primary tumors for a variety of tissues \citep{tezuka2007growth,rodgers2024glioblastoma} and metastatic melanoma has been shown to grow faster then primary melonoma \citep{carlson2003tumor}."
+Indeed, doubling times for tumors after relapse is lower than that of primary tumors for a variety of tissues \citep{tezuka2007growth,rodgers2024glioblastoma} and metastatic melanoma has been shown to grow faster then primary melonoma \citep{carlson2003tumor}.''
 }
 
-10. Mid page 7, "decreases" is mis-spelled
+10. Mid page 7, ``decreases'' is mis-spelled
 
@@
 
-and I would swap out "very well" with something more descriptive where you use it.
+and I would swap out ``very well'' with something more descriptive where you use it.
 
 \textcolor{blue}{
-We changed "the approximations perform very wll" to "the approximations are accurate" in two places.
+We changed ``the approximations perform very well'' to ``the approximations are accurate'' in two places.
 } 
@@
 As for the fitness difference between euploid and aneuploid mutants, this is a good question and the answer probably varies depending on the drug, the chromosome, and other factors.
-We now address both your questions in the Discussion, in line 542: "Furthermore, loss of extra-chromosome may occur at a high rate (for estimates in yeast, see Hose et al.
+We now address both your questions in the Discussion, in line 542: ``Furthermore, loss of extra-chromosome may occur at a high rate (for estimates in yeast, see Hose et al.
 (2024)) and the fitness of a euploid cell with a resistance mutation may be higher than that of an aneuploid cell with the same mutation.
-Future work could test when these effects have significant effect on the results of our model."
+Future work could test when these effects have significant effect on the results of our model.''
 }
@@
 \textcolor{blue}{%change 32
-We changed the first sentence in the abstract: "Evolutionary rescue occurs when a population, facing a sudden environmental change that would otherwise lead to extinction, adapts through beneficial mutations, allowing it to recover and persist."
+We changed the first sentence in the abstract: ``Evolutionary rescue occurs when a population, facing a sudden environmental change that would otherwise lead to extinction, adapts through beneficial mutations, allowing it to recover and persist.''
 }
@@
 
-- intro: 'most models focus on the probability that at most one mutation rescues the population' - I disagree with this statement.
+- intro: `most models focus on the probability that at most one mutation rescues the population' - I disagree with this statement.
 I think the focus is rather on the probability that at least one mutation rescues the population.
@@
 \textcolor{blue}{%change 41
-We have removed the statement "most models focus...". Instead, we include a review of the evolutionary rescue literature (lines 55-87).
+We have removed the statement ``most models focus...''. Instead, we include a review of the evolutionary rescue literature (lines 55-87).
 }
@@
 \textcolor{blue}{%change 36
-We repharased to (line line 175): "For the aneuploid growth rate to be intermediate between those of sensitive and mutant cells, $\Delta_s\ll\Delta_a\ll\Delta_m$, we set the aneuploid division rate to be $0.06 \le \lambda_a \le 0.1$."
+We repharased to (line line 175): ``For the aneuploid growth rate to be intermediate between those of sensitive and mutant cells, $\Delta_s\ll\Delta_a\ll\Delta_m$, we set the aneuploid division rate to be $0.06 \le \lambda_a \le 0.1$.''
 }
 
-- methods: 'We estimate c=...' - It sounds like you will estimate c from the observed fraction of aneuploid cells but that is not what you describe below.
+- methods: `We estimate c=...' - It sounds like you will estimate c from the observed fraction of aneuploid cells but that is not what you describe below.
 
 \textcolor{blue}{%change 5
-We revised the text to make it clear that we estimate $c$ from an experimental result and $f$ from $c$ (line 208): "To estimate fitness cost of aneuploidy, $c$, we note that Lukow et al.
+We revised the text to make it clear that we estimate $c$ from an experimental result and $f$ from $c$ (line 208): ``To estimate fitness cost of aneuploidy, $c$, we note that Lukow et al.
 (2021) mixed sensitive and aneuploid A375 melanoma cells at $1:1$ ratio, cultured them in a drug-free environment, and observed the ratio evolve as a function of time with the aneuploid cells declining to $15\%$ after 24 days.
 Thus, our estimate for the fitness cost is $c=\abs{\log\left(0.15/(1-0.15)\right)/24}\approx0.07$ per day (Chevin 2011).
-We estimate the fraction of aneuploid cancer cells in the pre-treatment environment using the formula $f=\tilde{u}\lambda_s / c$, which produces an estimate of $f=10^{-3}\times10^{-1}/0.07$, that is, $0.14\%$ of pre-treatment cancer cells have the beneficial aneuploidy of interest."
+We estimate the fraction of aneuploid cancer cells in the pre-treatment environment using the formula $f=\tilde{u}\lambda_s / c$, which produces an estimate of $f=10^{-3}\times10^{-1}/0.07$, that is, $0.14\%$ of pre-treatment cancer cells have the beneficial aneuploidy of interest.''
 }
@@
 \textcolor{blue}{%change 51
-We have added the recommended comment (line 303): "Interestingly, increasing both the aneuploid division rate, $\lambda_a$, and the aneuploid death rate, $\mu_a$, such that the growth rate $\Delta_a$ remains constant, leads to a decrease in $T^*$, pushing the system to the second scenario.
-This is because increasing $\lambda_a$ causes a decrease in $T^*$ as  it increases the effective mutation rate $v\lambda_a$ (as mutations mostly occur during division) and a lineage does not have to survive as long in order to generate a successful mutant."
+We have added the recommended comment (line 303): ``Interestingly, increasing both the aneuploid division rate, $\lambda_a$, and the aneuploid death rate, $\mu_a$, such that the growth rate $\Delta_a$ remains constant, leads to a decrease in $T^*$, pushing the system to the second scenario.
+This is because increasing $\lambda_a$ causes a decrease in $T^*$ as  it increases the effective mutation rate $v\lambda_a$ (as mutations mostly occur during division) and a lineage does not have to survive as long in order to generate a successful mutant.''
 }
@@
 \textcolor{blue}{ % change 58
-We believe that our model does fit with data on resistance rates, and the text now reads (line 316): "Interestingly, the threshold between small and intermediate tumors, $N_a^*$, is similar to the tumor detection threshold of $4.19 \times 10^6$ cells for a wide variety of tumors~\citep{avanzini2019cancer}.
+We believe that our model does fit with data on resistance rates, and the text now reads (line 316): ``Interestingly, the threshold between small and intermediate tumors, $N_a^*$, is similar to the tumor detection threshold of $4.19 \times 10^6$ cells for a wide variety of tumors~\citep{avanzini2019cancer}.
 We note that vemurafenib-treated melanomas (i.e.
-melanomas with sizes above the detection threshold) have a probability $>50\%$ to relapse \citep{piejko2023long,handa2022long}."
+melanomas with sizes above the detection threshold) have a probability $>50\%$ to relapse \citep{piejko2023long,handa2022long}.''
 }
@@
 \textcolor{blue}{%change 46
-We have added the following text in line 393: "For small tumors ($ N \ll N_a^*$), the mean rescue time is the two-step equivalent of the one-step result from Orr and Unckleass (2014; expectation of eq.~18).
+We have added the following text in line 393: ``For small tumors ($ N \ll N_a^*$), the mean rescue time is the two-step equivalent of the one-step result from Orr and Unckleass (2014; expectation of eq.~18).
 The mean rescue time (conditioned on rescue occuring) is a decreasing function of the sensitive and aneuploid growth rates and independent of the other model parameters, including tumor size (blue line in Figure S6).
@@
 Increasing the per division mutation rate, $v$, leads to the faster appearance of a rescue mutations and hence reduced mean rescue time.
-Finally, increasing the tumor size leads to shorter mean rescue time, as more sensitive cells can mutate to become resistant."
+Finally, increasing the tumor size leads to shorter mean rescue time, as more sensitive cells can mutate to become resistant.''
 }
 
-- results: 'aneuploidy promotes evolutionary rescue after $1/\Delta_a$' $\rightarrow$ 'aneuploidy promotes evolutionary rescue after $1/\Delta_s$'
+- results: `aneuploidy promotes evolutionary rescue after $1/\Delta_a$' $\rightarrow$ `aneuploidy promotes evolutionary rescue after $1/\Delta_s$'
 
@@
 \textcolor{blue}{%change 26
-We have added the definition of $p_m$ in line 421: "where $p_m$ is the probability that a lineage starting from a single mutant cell escapes stochastic extinction."
+We have added the definition of $p_m$ in line 421: ``where $p_m$ is the probability that a lineage starting from a single mutant cell escapes stochastic extinction.''
 }
@@
 \textcolor{blue}{% change 49
-The text now reads (line 435): "Drugs that significantly increase the sensitive death rate, $\mu_s$, but do not affect the sensitive division rate, $\lambda_s$, delay cancer recurrence.
-Drugs that decrease the sensitive division rate delay cancer recurrence time even more, because they effectively decrease the mutation rate (assuming mutations occur during division)."
+The text now reads (line 435): ``Drugs that significantly increase the sensitive death rate, $\mu_s$, but do not affect the sensitive division rate, $\lambda_s$, delay cancer recurrence.
+Drugs that decrease the sensitive division rate delay cancer recurrence time even more, because they effectively decrease the mutation rate (assuming mutations occur during division).''
 }
@@
 Yes, aneuploidy delays population extinction and yes it increases the population size at a given time, but the important fact is that it increases the cumulative number of mutants that are created, which is the expression you give at the end of the sentence.
-Perhaps it would help to insert an additional sentence before the expression, like "Together, this increases the cumulative number of mutants that arise."
+Perhaps it would help to insert an additional sentence before the expression, like ``Together, this increases the cumulative number of mutants that arise.''
 
@@
 Thank you for this suggestions.
-We have revised the text (line 476): "...
+We have revised the text (line 476): ``...
 aneuploidy provides two advantages.
@@
 Second, it increases the population size relative to a drug-sensitive population, providing more cells in which mutations can occur.
-Together, this increases the cumulative number of mutants that arise (i.e., $Nuv\lambda_s\lambda_a/\abs{\Delta_s\Delta_a}$)."
+Together, this increases the cumulative number of mutants that arise (i.e., $Nuv\lambda_s\lambda_a/\abs{\Delta_s\Delta_a}$).''
 }
@@
 \textcolor{blue}{%change x1
-Line 52 in the Introduction reads: "Futhermore, some proposed anti-cancer drugs elevate the missegregation rate to fight cancer cells (Lee et al.
-2016), as an extremely high chromosome missegregation rate is incompatible with cell survival and proliferation."
+Line 52 in the Introduction reads: ``Furthermore, some proposed anti-cancer drugs elevate the missegregation rate to fight cancer cells (Lee et al.
+2016), as an extremely high chromosome missegregation rate is incompatible with cell survival and proliferation.''
 }
@@
 \textcolor{blue}{%change 24
-We changed the caption to "When aneuploid cells are tolerant ($\Delta_a<0$) evolutionary rescue can occur indirectly, but direct mutation is the most likely path for evolutionary rescue".
+We changed the caption to ``When aneuploid cells are tolerant ($\Delta_a<0$) evolutionary rescue can occur indirectly, but direct mutation is the most likely path for evolutionary rescue''.
 }
@@
 
-- appendix A: 'decline are significant'$\rightarrow$'decline is significant'
+- appendix A: `decline are significant'$\rightarrow$`decline is significant'
 
@@
 \textcolor{blue}{
-The justificaiont, also included now in Appendix C, is in the following. 
+The justificaion, also included now in Appendix C, is in the following. 
 We know that
@@
 
-- appendix D: 'it accurate' $\rightarrow$ it is accurate'
+- appendix D: `it accurate' $\rightarrow$ `it is accurate'
 
@@
 
-- appendix F: 'Figure Figure' $\rightarrow$ 'Figure'
+- appendix F: `Figure Figure' $\rightarrow$ `Figure'
 
@@
 
-- appendix F: 'our approximation are'$\rightarrow$'our approximations are'
+- appendix F: `our approximation are' $\rightarrow$ `our approximations are'
 
@@
 
-- appendix F: 'From eq. (F2) we obtain'$\rightarrow$'From eq. (F1) we obtain'
+- appendix F: `From eq. (F2) we obtain' $\rightarrow$ `From eq. (F1) we obtain'
 
@@
 
-- fig S2: 'the leads'$\rightarrow$ 'that leads', 'is in agreement'$\rightarrow$'are in agreement', $\lambda_m$ value assigned twice
+- fig S2: `the leads'$\rightarrow$ `that leads', `is in agreement' $\rightarrow$ `are in agreement', $\lambda_m$ value assigned twice
 
@@
 
-- fig S3: 'number of mutation'$\rightarrow$'number of mutations', $\lambda_m$ value assigned twice
+- fig S3: `number of mutation'$\rightarrow$ `number of mutations', $\lambda_m$ value assigned twice
 
@@
 
-- fig S6: 'the leads'$\rightarrow$'that leads', 'lines, right'$\rightarrow$'lines,', 'is in agreement'$\rightarrow$'are in agreement', 'lines represents'$\rightarrow$'lines represent', $\lambda_m$ value assigned twice
+- fig S6: `the leads' $\rightarrow$ `that leads', `lines, right' $\rightarrow$ `lines,', `is in agreement' $\rightarrow$ `are in agreement', `lines represents' $\rightarrow$ `lines represent', $\lambda_m$ value assigned twice
 
@@
 
-- fig S9: 'the leads'$\rightarrow$'that leads', $\lambda_m$ value assigned twice
+- fig S9: `the leads'$\rightarrow$ `that leads', $\lambda_m$ value assigned twice
 
@@
 \textcolor{blue}{Following the reviewer's suggestion, we have evaluated our model using parameters from breast cancer (PMID: 34163070), see Methods (line 180), Table 2, and Figure 3B, which also appear below.\\
-We have aldo added the following text to the Discussion (line 493): "As an example, we have parameterized our model using estimates for melanoma and triple-negative breast cancer (TNBC) cells under drug therapy from the literature.
+We have aldo added the following text to the Discussion (line 493): ``As an example, we have parameterized our model using estimates for melanoma and triple-negative breast cancer (TNBC) cells under drug therapy from the literature.
 We find that in these cases we are in the third scenario, in which aneuploidy provides at least partial resistance to the drug. 
@@
 This comparison suggests that TNBC has a higher probability of relapsing compared to melanoma, because smaller tumors have higher resuce probability.
-Indeed, the probability of relapse in TNBC patients is higher than 50\% in the first 3-5 years after diagnosis \citep{taushanova2023synchronous}, whereas in melanoma patients it is approximately 10-20\% in the first 5 years \citep{wan2022prediction,von_schuckmann2019risk}.
-% change A
+Indeed, the probability of relapse in TNBC patients is higher than 50\% in the first 3-5 years after diagnosis \citep{taushanova2023synchronous}, whereas in melanoma patients it is approximately 10-20\% in the first 5 years \citep{wan2022prediction,von_schuckmann2019risk}.'' % change A
 }% change A
@@
 \textcolor{blue}{
-We have removed the misleading sentence "aneuploidy will either not play any role in evolutionary rescue or will be the main driver of adaptation", which is incorrect.
-Instead, we write (line 526): "Our results suggest that identifying aneuploidies that are tolerant or stationary may be worthwhile, as these are also expected to enhance the probability of rescue (Figure 3) and extend the window of opportunity for rescue (Figure 6)."\\
+We have removed the misleading sentence ``aneuploidy will either not play any role in evolutionary rescue or will be the main driver of adaptation'', which is incorrect.
+Instead, we write (line 526): ``Our results suggest that identifying aneuploidies that are tolerant or stationary may be worthwhile, as these are also expected to enhance the probability of rescue (Figure 3) and extend the window of opportunity for rescue (Figure 6).''\\
 Furthermore, In our model, we are interested in a specific type of aneuploidy which increases the fitness of cancer cells in an environment affected by anti-cancer drugs.
@@
 We assume deleterious mutations can occur with equal rates in all genotypes and therefore would not change the dynamics.
-We now explain these assumptions in line 215: "Note that when we refer to drug-sensitive cells, we include those cells that have any aneuploidy that does not affect fitness in the presence of the drug.
+We now explain these assumptions in line 215: ``Note that when we refer to drug-sensitive cells, we include those cells that have any aneuploidy that does not affect fitness in the presence of the drug.
 Additionally, we focus on mutations that confer resistance, neglecting deleterious mutations (which are more common).
-We assume deleterious mutations and other aneuplidies occur at similar rates across genotypes and therefore neglect their effect on the dynamics."
+We assume deleterious mutations and other aneuplidies occur at similar rates across genotypes and therefore neglect their effect on the dynamics.''
 }
@@
 * The methods only describe the simulations, and the results focus on analytical results that don't follow from the methods.
-I suggest expanding the "Methods" to discuss the general approach, model structure, and quantities derived (you can rename this as "Model") and then move to "Results".
+I suggest expanding the ``Methods'' to discuss the general approach, model structure, and quantities derived (you can rename this as ``Model'') and then move to ``Results''.
 
 \textcolor{blue}{
-The Methods section has several subsections: "Evolutionary model" that describes the model, from which the analytic results are derived; "Stochastic simulations", which describes the simulations used to validate the analytic results; "Parameterization", which describes how we parameterize the model; and "Denstity-dependent growth", which described the extended model that includes density dependence.
-Note that the "Evolutionary model" subsection is not focused on the simulations but rather described the general model that we are studying.}
+The Methods section has several subsections: ``Evolutionary model'' that describes the model, from which the analytic results are derived; ``Stochastic simulations'', which describes the simulations used to validate the analytic results; ``Parameterization'', which describes how we parameterize the model; and ``Denstity-dependent growth'', which described the extended model that includes density dependence.
+Note that the ``Evolutionary model'' subsection is not focused on the simulations but rather described the general model that we are studying.}
 
-* Delta would be better as "r" for the growth rate
+* Delta would be better as ``r'' for the growth rate
 
@@
 
-* The use of the terms "tolerant" (for negative growth) and "resistant" (for positive growth) is not standard (e.g., see different for fungi used by Berman $\&$ Krysan 2020 Nature Reviews Microbiology).
-Maybe alert the reader that "tolerant" is used in different ways in the literature and is used here solely for notational convenience to refer to mutations that, on their own, do not allow growth?
-Or replace with "sub-critical", a term already used in this context?
+* The use of the terms ``tolerant'' (for negative growth) and ``resistant'' (for positive growth) is not standard (e.g., see different for fungi used by Berman $\&$ Krysan 2020 Nature Reviews Microbiology).
+Maybe alert the reader that ``tolerant'' is used in different ways in the literature and is used here solely for notational convenience to refer to mutations that, on their own, do not allow growth?
+Or replace with ``sub-critical'', a term already used in this context?
 
-\textcolor{blue}{We have added a short explanation for "tolerant" and that it has other meanings in the literature.
-We also mention the equivalence to in "sub/super/critical" from Osmond et al 2020.} % change 47, 57
+\textcolor{blue}{We have added a short explanation for ``tolerant'' and that it has other meanings in the literature.
+We also mention the equivalence to in ``sub/super/critical'' from Osmond et al 2020.} % change 47, 57
 
-* "antagonizing cell division" - The meaning of this isn't clear. Rephrase as "by extending the length of the cell cycle"?
+* ``antagonizing cell division'' - The meaning of this isn't clear. Rephrase as ``by extending the length of the cell cycle''?
 
-\textcolor{blue}{We have made the recommended change from "antagonizing cell division" to "by extending the length of the cell cycle".}  % change 34
+\textcolor{blue}{We have made the recommended change from ``antagonizing cell division'' to ``by extending the length of the cell cycle''.}  % change 34
 
@@
 
-"If a fraction $f$ of the cancer cells are aneuploid when the drug is administered then the rates at which the rescue mutations are generated can be written as" $\rightarrow$
-"If a fraction $f$ of the cancer cells are aneuploid when the drug is administered then the expected number of successful mutants generated until time $t$ is given by". % change r6
+``If a fraction $f$ of the cancer cells are aneuploid when the drug is administered then the rates at which the rescue mutations are generated can be written as'' $\rightarrow$
+``If a fraction $f$ of the cancer cells are aneuploid when the drug is administered then the expected number of successful mutants generated until time $t$ is given by''. % change r6
 

### 2025-01-22 ResponseLetter.tex
edited response to editor, fixed bib

@@
 \usepackage{amssymb,amsmath,amsthm}
-\usepackage{natbib}
+\usepackage[comma,sort]{natbib}
 \usepackage{commath}
@@
 \usepackage{graphicx} % Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
+\usepackage[T1]{fontenc}
 
@@
 Instead, we write (line 526): ``Our results suggest that identifying aneuploidies that are tolerant or stationary may be worthwhile, as these are also expected to enhance the probability of rescue (Figure 3) and extend the window of opportunity for rescue (Figure 6).''\\
-Furthermore, In our model, we are interested in a specific type of aneuploidy which increases the fitness of cancer cells in an environment affected by anti-cancer drugs.
+Furthermore, in our model, we are interested in a specific type of aneuploidy which increases the fitness of cancer cells in an environment affected by anti-cancer drugs.
 As a consequence, we catalogued all other forms of aneuploids as sensitive.
-These aneuploidies can appear in patient cohorts because, for example, they drive cancer itself or increase genome instability, which leads to mutations and abberations that promote cancer.
+These aneuploidies can appear in patient cohorts because, for example, they drive cancer itself or increase genome instability, which leads to mutations and aberrations that promote cancer.
 }
@@
 \textcolor{blue}{
-We have have revised the introduction and discussion accordingly, see lines 55-87, 520-529.
+We agree with all of these points. We have revised the introduction and discussion accordingly, see lines 55-87, 520-529.
 }
@@
 
+\textcolor{blue}{We have added the material, as described in our response to reviewer 2.}
+
 Methods:
@@
 \textcolor{blue}{The manuscript has over 300 mentions of $\Delta$, plus mentions in the figures, and we do not want to introduce any errors by changing it.}
+\textcolor{red}{I don't think this is a good argument.
+It's literally one command in \LaTeX to change all the ``$\Delta$''s in the text to ``$r$''s.
+So it would just be changing the figures, which isn't a big deal.
+I think we should do it.}
 
@@
 \textcolor{blue}{We have added a short explanation for ``tolerant'' and that it has other meanings in the literature.
-We also mention the equivalence to in ``sub/super/critical'' from Osmond et al 2020.} % change 47, 57
+We also mention the equivalence to in ``sub/super/critical'' from Osmond et al 2020.
+We use these terms because we are trying to translate concepts familiar to the evolutionary rescue community into ones more familiar to cancer researchers.} % change 47, 57
 
@@
 
-\textcolor{blue}{We have added the green line for eq. D3 to Figure 7 as suggested, it is very close to the blue line for eq. D4.}
+\textcolor{blue}{We have added the green line for eq. D3 to Figure 7 as suggested. It is very close to the blue line for eq. D4.}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \textbf{Additional changes}
+
+{\color{blue}
 
@@
 
-\bibliographystyle{plain}
+}
+
+\bibliographystyle{agsm}
 \bibliography{ms}

### 2025-01-22 ms.tex
reformatting appendix D

@@
 
-This helps us distinguish between two scenarios for the proliferation time. Firstly, when we have at most one lineages which rescues the cancer cell population:
+This helps us distinguish between two scenarios for the proliferation time. Firstly, when we have at most one lineage which rescues the cancer cell population:
 \begin{equation*}
@@
 Solving the system of ODEs for initial condition $\left(s(0), a(0), m(0)\right)=\left(N,0,0\right)$ we obtain:
-\begin{align*}
-m\left(t\right)=\frac{Nuv\lambda_a\lambda_s}{\Delta_s-\Delta_a}\left[\frac{\e^{\Delta_st}-\e^{\Delta_mt}}{\Delta_s-\Delta_m}-\frac{\e^{\Delta_at}-\e^{\Delta_mt}}{\Delta_a-\Delta_m}\right]+Nv\lambda_s\frac{\e^{\Delta_st}-\e^{\Delta_mt}}{\Delta_s-\Delta_m}.
-\end{align*}
-We obtain $\tau_a^r$ such that $m\left(\tau_a^r\right)=N$ by solving:
+\begin{equation} \label{m_t}
+m(t) = \frac{N u v \lambda_a \lambda_s}{\Delta_a - \Delta_s}\left[ \frac{\e^{\Delta_m t} - \e^{\Delta_a t}}{\Delta_m - \Delta_a} - \frac{\e^{\Delta_m t} - \e^{\Delta_s t}}{\Delta_m - \Delta_s} \right] + N v \lambda_s \frac{\e^{\Delta_m t} - \e^{\Delta_s t}}{\Delta_m - \Delta_s}.
+\end{equation}
+We obtain $\tau_a^r$ by solving \eqref{m_t} for the time at which the number of mutant cells reaches $N$, i.e., $m\left(\tau_a^r\right)=N$:
 \begin{equation}\label{eq:t2det2024}
-1=\frac{uv\lambda_a\lambda_s}{\Delta_s-\Delta_a}\left[\frac{\e^{\Delta_s\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_s-\Delta_m}-\frac{\e^{\Delta_a\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_a-\Delta_m}\right]+v\lambda_s\frac{\e^{\Delta_s\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_s-\Delta_m}.
+1 = \frac{u v \lambda_a \lambda_s}{\Delta_a - \Delta_s}\left[ \frac{\e^{\Delta_m \tau_a^r} - \e^{\Delta_a \tau_a^r}}{\Delta_m - \Delta_a} - \frac{\e^{\Delta_m \tau_a^r} - \e^{\Delta_s \tau_a^r}}{\Delta_m - \Delta_s} \right] + v \lambda_s \frac{\e^{\Delta_m \tau_a^r} - \e^{\Delta_s \tau_a^r}}{\Delta_m - \Delta_s}.
 \end{equation}
@@
 \begin{equation*}
-1=v\lambda_s\frac{\e^{\Delta_m\tau_a^r}}{\abs{\Delta_s-\Delta_m}},
+1=v\lambda_s\frac{\e^{\Delta_m\tau_a^r}}{\Delta_m - \Delta_s},
 \end{equation*}
@@
 \end{equation}
-We observe that the terms given by the evolutionary trajectory $sensitive \rightarrow aneuploid \rightarrow mutant$ do not contribute to the above approximation and, as a result, we deduce that it is accurate only for $N\gg N_m^*>N_a^*$.
+We observe that the terms given by the evolutionary trajectory \textit{sensitive} $\rightarrow$ \textit{aneuploid} $\rightarrow$ \textit{mutant} do not contribute to the above approximation and, as a result, we deduce that it is accurate only for $N \gg N_m^* > N_a^*$.
 %change 9

### 2025-01-23 ms.tex
revise TNBC again

@@
 We also consider breast cancer cells, relying on \citet{salehi2021clonal}, who studied multiple TNBC PDX clones (triple-negative breast cancer patient-derived xenografts.)
-Their results suggest that aneuploidy and other copy number alterations provide resistance to the drug cisplatin.  
-Thus, we assume that aneuploid breast cancer cells are partially or even fully drug-resistant, i.e., $\Delta_m \ge \Delta_a > 0$.
 We chose three clones for which \cite{salehi2021clonal} report their relative Wrightian fitness, $1+s$, in the presence of the drug cisplatin: TNBC-SA609 clone A with $1+s=1.047$, TNBC-SA1035 clone H with $1+s=1.02$, and TNBC-SA535 clone H with $1+s=1.01$.
@@
 Thus, the division rate of breast cancer cells in the absence of the drug is $\log{(2)}/8.2=0.0845$ per day, and their death rate is $0.076$ per day.
-As all three clones grow in the presence of the drug, we set the growth rate of the least fit, TNBC-SA535 clone H, to the growth rate of breast cancer cells in the absence of the drug, $\Delta_{SA535}=0.0085$, and then solve the set of equations.
-% TODO needs to be the most fit, wait for Remus
-We further assume the drug affects death rather than division.
-We therefore have $\lambda_s=\lambda_a=0.0845$ for all clones,  $\mu_{SA609} = 0.0415$, $\mu_{SA1035}=0.0667$, $\mu_{SA535}=0.085$, and $\mu_s=0.085$.
+\cite{salehi2021clonal} report relative fitness for the aneuploid clones rather than division and growth rate.
+We, therefore, use the above estimates from breast cancer cells in the absence of drugs for the fittest clone, TNBC-SA609 clone A, and solve the set of equations assuming that the drug affects death rather than division and that mutant cells have the same division and death rates as the fittest aneuploid cells TNBC-SA609 clone A.
+We therefore have $\lambda_s=\lambda_a=\lambda_m=0.0845$ for all clones, $\mu_s=0.1215$, $\mu_{SA609} = 0.076$, $\mu_{SA1035}=0.1015$, $\mu_{SA535}=0.1115$, and $\mu_m=0.076$. 
+These estimates are approximate at best, and we hope that future experimental research could provide accurate estimates of these quantities.
 
@@
 These approximations are accurate when compared to results of stochastic evolutionary simulations (\Cref{rescue_prob,rescue_threshold}). 
+Our parameterization for two types of cancers suggests that melanoma and for TNBC-SA609 clone A are resistant to the drug ($\Delta_a T^*>500$), whereas TNBC-SA1035 clone H and TNBC-SA535 clone H are tolerant ($\Delta_a T^*<-1000$).
 In the first scenario, the treatment effectively kills aneuploid cells but not as quickly as it kills sensitive cells. 
@@
 In the third scenario, aneuploid cells are sufficiently resistant for the population to re-grow the tumor without additional resistance mutations.
-In our parameterization for melanoma and breast cancer, the value of $\Delta_a T^*$ is over 500.
-Therefore, in these cases, we are in the third scenario, in which aneuploidy provides resistance to the drug. % change b
 
@@
 
-As an example, we have parameterized our model using estimates for melanoma and triple-negative breast cancer (TNBC) cells under drug therapy from the literature. 
-We find that in these cases we are in the third scenario, in which aneuploidy provides at least partial resistance to the drug. 
-It remains to be seen which tumor type and drug combinations produce tolerant and stationary aneuploidies. % change B
-We have also compared the probability of evolutionary rescue between melanoma and TNBC (black and color lines, respectively, in \Cref{rescue_prob}B).
-This comparison suggests that TNBC has a higher probability of relapsing compared to melanoma, because smaller tumors have higher resuce probability.
-Indeed, the probability of relapse in TNBC patients is higher than 50\% in the first 3-5 years after diagnosis \citep{taushanova2023synchronous}, whereas in melanoma patients it is approximately 10-20\% in the first 5 years \citep{wan2022prediction,von_schuckmann2019risk}.
+As an example, we have parameterized our model using estimates for melanoma and three triple-negative breast cancer (TNBC) cells under drug therapy from the literature. 
+We find that in these cases we are either in the first scenario (melanoma and one of the TNBC clones) in which aneuploidy provides at partial or full resistance to the drug, or the third scenario (two TNBC), in which aneuploidy provides tolerance to the drug.
+It remains to be seen which tumor type and drug combinations produce stationary aneuploidies. % change B
+Comparing the probability of evolutionary rescue between different tumors (\Cref{rescue_prob}B) suggests that some TNBC clones have a much higher probability of relapsing compared to melanoma and other TNBC clones, because smaller tumors have higher resuce probability.
+Notably, the probability of relapse in TNBC patients is higher than 50\% in the first 3-5 years after diagnosis \citep{taushanova2023synchronous}, whereas in melanoma patients it is approximately 10-20\% in the first 5 years \citep{wan2022prediction,von_schuckmann2019risk}.
 % change A
@@
      & Name & Value & Units \\ \hline
-    $\lambda_s$ & Sensitive division rate& 0.0848& 1/days \\ \hline
-    $\mu_s$ & Sensitive death rate& $0.085$ & 1/days   \\ \hline
+    $\lambda_s$ & Sensitive division rate& 0.0845& 1/days \\ \hline
+    $\mu_s$ & Sensitive death rate& $0.1215$ & 1/days   \\ \hline
     $\lambda_a$  & Aneuploid division rate & $0.0845$ & 1/days    \\ \hline
-    $\mu_a$ & Aneuploid death rate & $0.0415$, $0.0667$, $0.085$ & 1/days   \\ \hline
+    $\mu_a$ & Aneuploid death rate & $0.076$, $0.1015$, $0.1115$ & 1/days   \\ \hline
+    $\lambda_m$  & Mutant division rate & $0.0845$ & 1/days    \\ \hline
+    $\mu_m$ & Mutant death rate & $0.076$ & 1/days   \\ \hline
   \end{tabular}
@@
 Parameters: $\lambda_s=0.1,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, v=10^{-7}$.
-\textbf{(B)} Comparison for melanoma (black) and breast cancer (colors)  with parameters from the literature (\Cref{table1,table2}).
+\textbf{(B)} Comparison for melanoma and breast cancer with parameters from the literature (\Cref{table1,table2}).
 Black line: Melanoma A375.

### 2025-01-23 ms.tex
Delta to r
minor formatting

@@
 
+% Change Delta to r
+\renewcommand{\Delta}{r}
+
 % Supplementary
@@
 
-All the parameters discussed above are shown in \Cref{table1}.\\
+All the parameters discussed above are shown in \Cref{table1}.
 
@@
 As an example, we have parameterized our model using estimates for melanoma and three triple-negative breast cancer (TNBC) cells under drug therapy from the literature. 
-We find that in these cases we are either in the first scenario (melanoma and one of the TNBC clones) in which aneuploidy provides at partial or full resistance to the drug, or the third scenario (two TNBC), in which aneuploidy provides tolerance to the drug.
+We find that in these cases we are either in the first scenario (melanoma and one of the TNBC clones), in which aneuploidy provides at partial or full resistance to the drug, or the third scenario (two TNBC), in which aneuploidy provides tolerance to the drug.
 It remains to be seen which tumor type and drug combinations produce stationary aneuploidies. % change B

### 2025-01-23 ResponseLetter.tex
commented on response
fix TNBC

@@
 \emph{The latter indicates that aneuploidy, even when it only provides tolerance, increases the probability that the tumor will be rescued.}'' \\
-And in line 493: ``As an example, we have parameterized our model using estimates for melanoma and triple-negative breast cancer (TNBC) cells under drug therapy from the literature.
-We find that in these cases we are in the third scenario, in which aneuploidy provides at least partial resistance to the drug.
-It remains to be seen which tumor type and drug combinations produce tolerant and stationary aneuploidies.''\\
+And in line 493: ``AAs an example, we have parameterized our model using estimates for melanoma and three triple-negative breast cancer (TNBC) cells under drug therapy from the literature. 
+We find that in these cases we are either in the first scenario (melanoma and one of the TNBC clones) in which aneuploidy provides at partial or full resistance to the drug, or the third scenario (two TNBC), in which aneuploidy provides tolerance to the drug.
+It remains to be seen which tumor type and drug combinations produce stationary aneuploidies.''\\
 The following sentences were already in the Discussion (line 476): ``In this scenario, aneuploidy provides two advantages.
@@
 
-\textcolor{blue}{Following the reviewer's suggestion, we have evaluated our model using parameters from breast cancer (PMID: 34163070), see Methods (line 180), Table 2, and Figure 3B, which also appear below.\\
-We have aldo added the following text to the Discussion (line 493): ``As an example, we have parameterized our model using estimates for melanoma and triple-negative breast cancer (TNBC) cells under drug therapy from the literature.
-We find that in these cases we are in the third scenario, in which aneuploidy provides at least partial resistance to the drug. 
-It remains to be seen which tumor type and drug combinations produce tolerant and stationary aneuploidies. % change B
-We have also compared the probability of evolutionary rescue between melanoma and TNBC (black and color lines, respectively, in Figure 3B).
-This comparison suggests that TNBC has a higher probability of relapsing compared to melanoma, because smaller tumors have higher resuce probability.
-Indeed, the probability of relapse in TNBC patients is higher than 50\% in the first 3-5 years after diagnosis \citep{taushanova2023synchronous}, whereas in melanoma patients it is approximately 10-20\% in the first 5 years \citep{wan2022prediction,von_schuckmann2019risk}.'' % change A
+\textcolor{blue}{Following the reviewer's suggestion, we have evaluated our model using parameters from breast cancer (PMID: 34163070), see Methods (line 180), Table 2, and Figure 3B.\\
+We have aldo added the following text to the Discussion (line 493): ``As an example, we have parameterized our model using estimates for melanoma and three triple-negative breast cancer (TNBC) cells under drug therapy from the literature.
+We find that in these cases we are either in the first scenario (melanoma and one of the TNBC clones), in which aneuploidy provides at partial or full resistance to the drug, or the third scenario (two TNBC), in which aneuploidy provides tolerance to the drug.
+It remains to be seen which tumor type and drug combinations produce stationary aneuploidies. % change B
+Comparing the probability of evolutionary rescue between different tumors (Figure 3B) suggests that some TNBC clones have a much higher probability of relapsing compared to melanoma and other TNBC clones, because smaller tumors have higher resuce probability.
+Notably, the probability of relapse in TNBC patients is higher than 50\% in the first 3-5 years after diagnosis \citep{taushanova2023synchronous}, whereas in melanoma patients it is approximately 10-20\% in the first 5 years \citep{wan2022prediction,von_schuckmann2019risk}.'' % change A
 }% change A
@@
 
-\textcolor{blue}{The manuscript has over 300 mentions of $\Delta$, plus mentions in the figures, and we do not want to introduce any errors by changing it.}
-\textcolor{red}{I don't think this is a good argument.
-It's literally one command in \LaTeX to change all the ``$\Delta$''s in the text to ``$r$''s.
-So it would just be changing the figures, which isn't a big deal.
-I think we should do it.}
+\textcolor{blue}{We changed Delta to r.}
 
@@
 \begin{align*}
-v\lambda_s\frac{\e^{\Delta_s\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_s-\Delta_m}\approx\frac{v\lambda_s\e^{\Delta_m\tau_a^r}}{\left(\Delta_m-\Delta_a\right)}
+v\lambda_s\frac{\e^{\Delta_s\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_s-\Delta_m}\approx\frac{v\lambda_s\e^{\Delta_m\tau_a^r}}{\Delta_m-\Delta_s}
 \end{align*}
@@
 \\
-\\
+\textcolor{red}{I don't understand our answer here.
+If aneuploids are common and pretty good, can't they still matter?
+And I don't understand what we're saying about ``large $\tau_a^r$''---large compared to what?
+Aren't we describing the regime of relatively small $\tau_a^r$ here?
+}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2025-01-23 ms.tex
minor formatting

@@
 These approximations are accurate when compared to results of stochastic evolutionary simulations (\Cref{rescue_prob,rescue_threshold}). 
-Our parameterization for two types of cancers suggests that melanoma and for TNBC-SA609 clone A are resistant to the drug ($\Delta_a T^*>500$), whereas TNBC-SA1035 clone H and TNBC-SA535 clone H are tolerant ($\Delta_a T^*<-1000$).
+Our parameterization for two types of cancers suggests that melanoma and TNBC-SA609 clone A are resistant to the drug ($\Delta_a T^*>500$), whereas TNBC-SA1035 clone H and TNBC-SA535 clone H are tolerant ($\Delta_a T^*<-1000$).
 In the first scenario, the treatment effectively kills aneuploid cells but not as quickly as it kills sensitive cells. 
@@
 \end{equation*}
-where we ignore lineages created by direct mutation because we assumed $u\lambda_a > \max{(-\Delta_a, 1/T^*)}$, $N\ll N_m^*$ and used \Cref{eq:N_a}.
+where we ignore lineages created by direct mutation because we assumed $u\lambda_a > \max{(-\Delta_a, 1/T^*)}$, $N\ll N_m^*$ and used \cref{eq:N_a}.
 
@@
 Solving the system of ODEs for initial condition $\left(s(0), a(0), m(0)\right)=\left(N,0,0\right)$ we obtain:
-\begin{equation} \label{m_t}
+\begin{equation*} \label{m_t}
 m(t) = \frac{N u v \lambda_a \lambda_s}{\Delta_a - \Delta_s}\left[ \frac{\e^{\Delta_m t} - \e^{\Delta_a t}}{\Delta_m - \Delta_a} - \frac{\e^{\Delta_m t} - \e^{\Delta_s t}}{\Delta_m - \Delta_s} \right] + N v \lambda_s \frac{\e^{\Delta_m t} - \e^{\Delta_s t}}{\Delta_m - \Delta_s}.
-\end{equation}
-We obtain $\tau_a^r$ by solving \eqref{m_t} for the time at which the number of mutant cells reaches $N$, i.e., $m\left(\tau_a^r\right)=N$:
+\end{equation*}
+We obtain $\tau_a^r$ by solving the above equation for time $t=\tau_a^r$ at which the number of mutant cells reaches $N$, i.e., $m\left(\tau_a^r\right)=N$:
 \begin{equation}\label{eq:t2det2024}
@@
 
-Additionally, we note that if we are interested in the time until the tumor reaches a detectable size $M$ then our above analysis is valid but in \Cref{meanproliferationtime} we change:
+Additionally, we note that if we are interested in the time until the tumor reaches a detectable size $M$ then our above analysis is valid but in \cref{meanproliferationtime} we change:
 \begin{align}\label{meanproliferationtime2}

### 2025-01-26 ms.tex
small fixes, TODOs for RS

@@
 The growth rate $\Delta$, or the Malthusian fitness, is the log of the Wrightian fitness \citep{Wu2013a}. 
-We therefore have the following set of equations: $0.047 = \exp{(\Delta_{SA609}-\Delta_s)}, 0.02=\exp{(\Delta_{SA1035}-\Delta_s)}, 0.01=\exp{(\Delta_{SA535}-\Delta_s)}$.
+We therefore have the following set of equations: $1.047 = \exp{(\Delta_{SA609}-\Delta_s)}, 1.02=\exp{(\Delta_{SA1035}-\Delta_s)}, 1.01=\exp{(\Delta_{SA535}-\Delta_s)}$.
 The growth rate of breast cancer cells in the absence of the drug is $0.0085$ per day \citep{spratt1996rates}.
@@
 \cite{salehi2021clonal} report relative fitness for the aneuploid clones rather than division and growth rate.
-We, therefore, use the above estimates from breast cancer cells in the absence of drugs for the fittest clone, TNBC-SA609 clone A, and solve the set of equations assuming that the drug affects death rather than division and that mutant cells have the same division and death rates as the fittest aneuploid cells TNBC-SA609 clone A.
+We, therefore, use the above estimates from breast cancer cells in the absence of drugs for the fittest clone, TNBC-SA609 clone A, and solve the set of equations assuming that the drug affects death rather than division and that mutant cells have the same division and death rates as the fittest aneuploid cells of TNBC-SA609 clone A.
 We therefore have $\lambda_s=\lambda_a=\lambda_m=0.0845$ for all clones, $\mu_s=0.1215$, $\mu_{SA609} = 0.076$, $\mu_{SA1035}=0.1015$, $\mu_{SA535}=0.1115$, and $\mu_m=0.076$. 
@@
 %change 2
-Of particular interest is the distribution of the evolutionary rescue time and recurrence time with tolerant aneuploid cells ($\Delta_a T^*\ll1$), for which we focus on the parameter values $\lambda_s=0.1$, $\lambda_a=0.0899$, $\lambda_m=0.1$, $\mu_s=0.14$, $\mu_a=0.09$, $\mu_m=0.09$, $u=10^{-2}$, and $v=10^{-7}$.
+Of particular interest is the distribution of the evolutionary rescue time and recurrence time with tolerant aneuploid cells ($\Delta_a T^*\ll1$), for which we focus on the parameter values in \Cref{table1} with $\lambda_a=0.0899$ (tolerant aneuploids), $\mu_s=0.14$ (intermediate sensitive death rate), and $v=10^{-7}$ (high end of the mutation rate).
 
@@
 This is because if the population rapidly declines but is then rescued, than the resistance mutation must have appeared early; if the population slowly declines, than mutations can appear later and the mean time will be longer.
-In our focus parameter regime, we have $\Delta_s=-0.04$ and $\Delta_a=-10^{-4}$, such that the mean rescue time is mainly determined by the aneuploid growth rate, $\tau_a \approx 10^4$ days.
+In our focus parameter regime, we have $\Delta_s=-0.04$ and $\Delta_a=-10^{-4}$, such that the mean rescue time is mainly determined by the aneuploid growth rate, and $\tau_a \approx 10^4$ days (\cref{eq:AsymptoticTimeRules}).
 % change 46
 
-For large tumors ($N \gg N_m^*$), the dynamics are equivalent to a scenario where rescue mutations appear at a constant rate, and the mean rescue time (\cref{eq:AsymptoticTimeRules}) is independent of the aneuploid parameters ($u$, $\lambda_a$, and $\Delta_a$).
+For large tumors ($N \gg N_m^*$), the dynamics are equivalent to a scenario where rescue mutations appear at a constant rate, and the mean rescue time is independent of the aneuploid parameters ($u$, $\lambda_a$, and $\Delta_a$).
 Increasing the per division mutation rate, $v$, leads to the faster appearance of a rescue mutations and hence reduced mean rescue time.
@@
 We calculated the mean evolutionary rescue time with standing genetic variation, $\tilde{\tau}_a$ (\cref{meantimet2SGV}), and compared our result with simulations (\Cref{SGVEvolutionaryRescueTimeComplete}).
-We note that standing genetic variation does not significantly affect the mean evolutionary rescue time.
+We find that standing genetic variation does not significantly affect the mean evolutionary rescue time.
 
@@
 
-We next approximated the mean time for the population of mutant cancer cells to reach the initial, pre-treatment population size $N$, which we denote the recurrence time $\tau_a^r$ (Appendix~\ref{sec:appendix_recurrence_time}),
+We next approximated the mean time for the population of mutant cancer cells to reach the initial, pre-treatment population size $N$, which we denote the recurrence time $\tau_a^r$ (\cref{eq:AsymptoticTimeRules,meanproliferationtime,eq:t2det}),
 \begin{equation} \label{eq:AsymptoticTimeRulesRecurrence}
@@
 \begin{align*}
-r_1\left(t\right)&=v\lambda_ap_ma_{t},\\
+n_1\left(t\right)&=v\lambda_ap_ma_{t},\\
 r_2\left(t\right)&=v\lambda_sp_ms_{t},
@@
 
-For large initial population sizes we assume that the two processes are independent and as a result, they can be merged into a single Poisson process with rate $R_a(t)=\left(r_1+r_2\right)\left(t\right)$.
+For large initial population sizes we assume that the two processes are independent and as a result, they can be merged into a single Poisson process with rate $R_a(t)=\left(n_1+r_2\right)\left(t\right)$.
 Consequently, the mean time to the appearance of the first rescue mutant is
@@
 \begin{align*}
-r_1^f\left(t\right)&=v\lambda_ap_m\int_0^ta_{z} \d z = \left(1-f\right)\frac{uv\lambda_s\lambda_aNp_m}{\Delta_s-\Delta_a}\left(\frac{\e^{\Delta_st}-1}{\Delta_s}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right)+fv\lambda_aNp_m\frac{\e^{\Delta_a t}-1}{\Delta_a},\\
+n_1^f\left(t\right)&=v\lambda_ap_m\int_0^ta_{z} \d z = \left(1-f\right)\frac{uv\lambda_s\lambda_aNp_m}{\Delta_s-\Delta_a}\left(\frac{\e^{\Delta_st}-1}{\Delta_s}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right)+fv\lambda_aNp_m\frac{\e^{\Delta_a t}-1}{\Delta_a},\\
 r_2^f\left(t\right)&=v\lambda_sp_m\int_0^ts_{z} \d z = \left(1-f\right)v\lambda_sNp_m\frac{\e^{\Delta_s t}-1}{\Delta_s},
@@
 \end{align}
-where $R_a^f(t)=r_1^f\left(t\right)+r_2^f\left(t\right)$ and $\presc = 1-\exp\left[-\left(1-f\right)p_sN-fp_aN\right]$.
+where $R_a^f(t)=n_1^f\left(t\right)+r_2^f\left(t\right)$ and $\presc = 1-\exp\left[-\left(1-f\right)p_sN-fp_aN\right]$.
 We plot our approximation in  \Cref{SGVEvolutionaryRescueTimeComplete} together with simulated data.
@@
 
-\section{Recurrence time}\label{sec:appendix_recurrence_time}
-We define the proliferation time $\tau_a^p$  to be the time it takes the population of mutant cancer cells to reach the initial tumor size $N$.
+\section{Recurrence time}\label{sec:appendix_recurrence_time} % TODO it is not clear what are the assumptions here; reading the main text and then coming here I think that you assume tolerant aneuploidy - is that the case?
+you need to be EXPLICIT
+We define the proliferation time $\tau_a^p$  to be the expected time for a resistant lineage to grow to the initial tumor size $N$.
 The number of rescue lineages generated by the sensitive population is given by  \cref{eq:twosteplineage} (see \Cref{ExpectedNumberRescueLineages}):
 \begin{equation*}
-r_1\left(\infty\right)=\frac{uv\lambda_s\lambda_aNp_m}{\abs{\Delta_s}\abs{\Delta_a}}=\frac{N}{N_a^*},
+n_1\left(\infty\right)=\frac{uv\lambda_s\lambda_aNp_m}{\abs{\Delta_s}\abs{\Delta_a}}=\frac{N}{N_a^*},
 \end{equation*}
 where we ignore lineages created by direct mutation because we assumed $u\lambda_a > \max{(-\Delta_a, 1/T^*)}$, $N\ll N_m^*$ and used \cref{eq:N_a}.
-
-This helps us distinguish between two scenarios for the proliferation time. Firstly, when we have at most one lineage which rescues the cancer cell population:
-\begin{equation*}
-N\ll N_a^*.
-\end{equation*}
+% TODO Remus why did you assume these assumptions?
+say the biological assumption, not just the math.
+and how did you use eq 3?
+write it EXPLICITLY.
+
+This helps us distinguish between two scenarios for the proliferation time. 
+In the first scenario, $N\ll N_a^*$, and we have at most one lineage which rescues the cancer cell population.
 As a result, the recurrence time is given by~\citep{avanzini2019cancer}:
 \begin{align}\label{meanproliferationtime}
-\tau_a^r&\approx\tau_a+\frac{\log p_mN}{\Delta_m}.
+\tau_a^r&\approx\tau_a+\frac{\log p_mN}{\Delta_m}. % TODO is it log(pm) N or log(pm N) ? add parentheses.
 \end{align}
 The factor of $p_m$ in the second term of \cref{meanproliferationtime} is due to the fact that the lineage is conditioned to survive genetic drift and the time to reach $N$ is shorter then the scenario without this property \citep{orr2014population,smith1974hitch}.
-% change 35
-
-The second scenario is when the sensitive population produces a large number of rescue lineages in a short period of time. This is given by the condition:
-\begin{align*}
-N\gg N_a^*.
-\end{align*}
-As a result, the recurrence time is obtained by solving the following system of ODEs:
+% change 35 % TODO without what property?
+
+In the second scenario, $N\gg N_a^*$, and the sensitive population produces a large number of rescue lineages in a short period of time.
+As a result, the recurrence time is obtained by solving the following system of oridnary differential equations (ODEs),
 \begin{equation}\label{detODE}
@@
 \end{equation}
-Solving the system of ODEs for initial condition $\left(s(0), a(0), m(0)\right)=\left(N,0,0\right)$ we obtain:
+Solving the system of ODEs for initial condition $\left(s(0), a(0), m(0)\right)=\left(N,0,0\right)$ we obtain
 \begin{equation*} \label{m_t}
@@
 \end{equation*}
-We obtain $\tau_a^r$ by solving the above equation for time $t=\tau_a^r$ at which the number of mutant cells reaches $N$, i.e., $m\left(\tau_a^r\right)=N$:
+We obtain $\tau_a^r$ by solving the above equation for time $t=\tau_a^r$ at which the number of mutant cells reaches $N$, that is, we solve $m\left(\tau_a^r\right)=N$,
 \begin{equation}\label{eq:t2det2024}
@@
 \end{equation}
-This is a transcendental equation which cannot be solved exactly but we can obtain an approximation by noting that for large $\tau_a^r$ the above equation can be written as: 
+This is a transcendental equation which cannot be solved exactly but we can obtain an approximation by noting that for large $\tau_a^r$ the above equation can be written as % TODO Remus what do you mean large tau?
+large compared to what?
+why do you assume it is large?
+what is the biological assumption?
+It is not clear why you neglect two out of three terms, and you need to be explicit about this -- say both what the math is but also what the biological assumption is.
+the editor  has asked for as much.
 \begin{equation*}
@@
 \end{equation*}
-which has solution:
+which has an approximate solution
 \begin{equation}\label{eq:t2det}
@@
 
-Additionally, we note that if we are interested in the time until the tumor reaches a detectable size $M$ then our above analysis is valid but in \cref{meanproliferationtime} we change:
+Additionally, we note that if we are interested in the time until the tumor reaches a detectable size $M$ then our above analysis is valid but in \cref{meanproliferationtime} we change
 \begin{align}\label{meanproliferationtime2}
@@
 \end{align}
-and \Cref{eq:t2det} becomes:
+and \cref{eq:t2det} becomes
 \begin{equation}\label{eq:t2det2}
@@
 P\left(rescue,t\right)&=P\left(T_a<t\right)\\
-&=1-\exp\left\{-\left[r_1\left(t\right)+r_2\left(t\right)\right]\right\}\\
+&=1-\exp\left\{-\left[n_1\left(t\right)+r_2\left(t\right)\right]\right\}\\
 &=1-\exp\left\{-\left[\frac{uv\lambda_s\lambda_aNp_m}{\Delta_s-\Delta_a}\left(\frac{\e^{\Delta_st}-1}{\Delta_s}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right)+ v\lambda_sNp_m\frac{\e^{\Delta_s t}-1}{\Delta_s}\right]\right\},

### 2025-01-26 ResponseLetter.tex
format D3 as in text
revised to emph eq 4 result (rarity); fix some phrasing

@@
 \newcommand{\e}{{\rm e}}
+
+% Change Delta to r
+\renewcommand{\Delta}{r}
 
@@
 In line 468, emphasis on new text): ``Our results show that the probability of evolutionary rescue increases with the initial tumor size $N$, the drug-sensitive growth rate $\Delta_s$, the mutation rate $v$, and the aneuploidy rate $u$.
-\emph{The latter indicates that aneuploidy, even when it only provides tolerance, increases the probability that the tumor will be rescued.}'' \\
-And in line 493: ``AAs an example, we have parameterized our model using estimates for melanoma and three triple-negative breast cancer (TNBC) cells under drug therapy from the literature. 
+\emph{Notably, the latter indicates that aneuploidy, even when it only provides tolerance, increases the probability that the tumor will be rescued, as long as it is produced fast enough.}'' \\
+And in line 493: ``As an example, we have parameterized our model using estimates for melanoma and three triple-negative breast cancer (TNBC) cells under drug therapy from the literature. 
 We find that in these cases we are either in the first scenario (melanoma and one of the TNBC clones) in which aneuploidy provides at partial or full resistance to the drug, or the third scenario (two TNBC), in which aneuploidy provides tolerance to the drug.
@@
 Second, it increases the population size relative to a drug-sensitive population, providing more cells in which mutations can occur. 
-Together, this increases the cumulative number of mutants that arise.'' and in line 481: ``We find that aneuploidy can significantly affect evolutionary rescue as it reduces the threshold tumor size by at least an order of magnitude even when aneuploidy only provides tolerance.''
+Together, this increases the cumulative number of mutants that arise.'' and in line 481: ``We find that aneuploidy can significantly affect evolutionary rescue by reducing the threshold tumor size by several orders of magnitude, even when aneuploidy only provides tolerance, provided that the aneuploidy rate is high enough, $\lambda_a u > |\Delta_a|$.''
 }
@@
 
-\textcolor{blue}{We changed Delta to r.}
+\textcolor{blue}{We changed $\Delta$ to $r$. We therefore also changed $r_1$ to $n_1$ ($r_1$ only appeared in the appendices).}
 
@@
 \begin{align*}
-\frac{uv\lambda_a\lambda_s}{\Delta_s-\Delta_a}\left[\frac{\e^{\Delta_s\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_s-\Delta_m}-\frac{\e^{\Delta_a\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_a-\Delta_m}\right]\approx\frac{uv\lambda_a\lambda_s\e^{\Delta_m\tau_a^r}}{\left(\Delta_m-\Delta_s\right)\left(\Delta_m-\Delta_a\right)}
+\frac{u v \lambda_a \lambda_s}{\Delta_a - \Delta_s}\left[ \frac{\e^{\Delta_m \tau_a^r} - \e^{\Delta_a \tau_a^r}}{\Delta_m - \Delta_a} - \frac{\e^{\Delta_m \tau_a^r} - \e^{\Delta_s \tau_a^r}}{\Delta_m - \Delta_s} \right]
+\approx \frac{uv\lambda_a\lambda_s\e^{\Delta_m\tau_a^r}}{\left(\Delta_m-\Delta_s\right)\left(\Delta_m-\Delta_a\right)}
 \end{align*}
@@
 \begin{align*}
-v\lambda_s\frac{\e^{\Delta_s\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_s-\Delta_m}\approx\frac{v\lambda_s\e^{\Delta_m\tau_a^r}}{\Delta_m-\Delta_s}
+v \lambda_s \frac{\e^{\Delta_m \tau_a^r} - \e^{\Delta_s \tau_a^r}}{\Delta_m - \Delta_s} \approx \frac{v\lambda_s\e^{\Delta_m\tau_a^r}}{\Delta_m-\Delta_s}
 \end{align*}
@@
 We have made the following additional changes:
+
+Added just below eq 4: "Imporantly, when this threshold is smaller than 1, aneuploidy is expected to play a role in evolutionary rescue, as it resuced the threshold tumor size for rescue."
 

### 2025-01-26 ms.tex
revised to emph eq 4 result (rarity); fix some phrasing

@@
 In the first scenario, the treatment effectively kills aneuploid cells but not as quickly as it kills sensitive cells. 
-In the second scenario, aneuploid cells are sufficiently resistant, and the expected size of each aneuploid lineage is roughly 1.
+In the second scenario, aneuploid cells are sufficiently resistant, and the expected size of each aneuploid lineage is roughly one.
 In both of these scenarios, aneuploidy increases the probability of rescue by slowing or halting the decrease in the tumor population size, allowing more opportunities to produce resistant mutants.
@@
 \end{equation}
+Imporantly, when this threshold is smaller than 1, aneuploidy is expected to play a role in evolutionary rescue, as it resuced the threshold tumor size for rescue. 
 As expected, this ratio increases with the mutation rate $v$ and decreases with the aneuploidy rate $u$.
@@
 Our results show that the probability of evolutionary rescue increases with the initial tumor size $N$, the drug-sensitive growth rate $\Delta_s$, the mutation rate $v$, and the aneuploidy rate $u$.
-The latter indicates that aneuploidy, even when it only provides tolerance, increases the probability that the tumor will be rescued (\Cref{rescue_prob}).
+Notably, the latter indicates that aneuploidy, even when it only provides tolerance, increases the probability that the tumor will be rescued, as long as it is produced fast enough (\Cref{rescue_prob}A, \cref{eq:N_ratio}).
 
@@
 When aneuploidy only provides tolerance to the drug ($\Delta_s\ll\Delta_a\ll0\ll\Delta_m$), it cannot rescue the population.
-Instead, it acts as a ``stepping stone'' through which the resistant mutant can appear more rapidly, given that the number of aneuploid cells declines slower than the number of drug-sensitive cells (\Cref{sampleTrajectories}).
+Instead, if the aneuploidy rate is fast enough (\cref{eq:N_ratio}) then it may act as a ``stepping stone'' through which the resistant mutant can appear more rapidly, given that the number of aneuploid cells declines slower than the number of drug-sensitive cells (\Cref{sampleTrajectories}).
 In this scenario, aneuploidy provides two advantages. First, it delays the extinction of the population, providing more time for the appearance of the resistance mutation. 
@@
 
-We find that aneuploidy can significantly affect evolutionary rescue as it reduces the threshold tumor size by at least an order of magnitude even when aneuploidy only provides tolerance (\Cref{rescue_prob}).
-When the number of cells in the tumor is large enough (i.e., $N \gg N_m^* \approx 4\times10^7$), a resistance mutation will occur in drug-sensitive cells before these cells become extinct (\Cref{rescue_prob}).
+We find that aneuploidy can significantly affect evolutionary rescue by reducing the threshold tumor size by several orders of magnitude, even when aneuploidy only provides tolerance, provided that the aneuploidy rate is high enough, $\lambda_a u > |\Delta_a|$ (\Cref{rescue_prob}A, \cref{eq:N_ratio}).
+When the number of cells in the tumor is large enough (i.e., $N \gg N_m^* \approx 4\times10^7$), a resistance mutation will occur in drug-sensitive cells before these cells become extinct.
 Therefore, large tumors are likely to be rescued with or without aneuploidy.
@@
 As an example, we have parameterized our model using estimates for melanoma and three triple-negative breast cancer (TNBC) cells under drug therapy from the literature. 
-We find that in these cases we are either in the first scenario (melanoma and one of the TNBC clones), in which aneuploidy provides at partial or full resistance to the drug, or the third scenario (two TNBC), in which aneuploidy provides tolerance to the drug.
+We find that in these cases we are either in the first scenario (melanoma and one of the TNBC clones), in which aneuploidy provides at partial or full resistance to the drug, or the third scenario (two TNBC clones), in which aneuploidy provides tolerance to the drug.
 It remains to be seen which tumor type and drug combinations produce stationary aneuploidies. % change B
-Comparing the probability of evolutionary rescue between different tumors (\Cref{rescue_prob}B) suggests that some TNBC clones have a much higher probability of relapsing compared to melanoma and other TNBC clones, because smaller tumors have higher resuce probability.
+Comparing the probability of evolutionary rescue between different tumors (\Cref{rescue_prob}B) suggests that some TNBC clones have a much higher probability of relapsing compared to melanoma and other TNBC clones.
 Notably, the probability of relapse in TNBC patients is higher than 50\% in the first 3-5 years after diagnosis \citep{taushanova2023synchronous}, whereas in melanoma patients it is approximately 10-20\% in the first 5 years \citep{wan2022prediction,von_schuckmann2019risk}.
@@
 Although this may seem like a small region of parameter space in a general model of evolutionary rescue (Fig.~\ref{rescue_threshold}A, inset), it could be biologically significant for some tumors under specific drug therapies.
-Our results suggest that identifying aneuploidies that are tolerant or stationary may be worthwhile, as these are also expected to enhance the probability of rescue (\Cref{rescue_prob}) and extend the window of opportunity for rescue (\Cref{cdffig}).
+Our results suggest that identifying aneuploidies that are tolerant or stationary may be worthwhile, as they may enhance the probability of rescue (\Cref{rescue_prob}A) and extend the window of opportunity for rescue (\Cref{cdffig}).
 
@@
 
-Our results quantitatively suggest that aneuploidy may play an important role in tumor adaptation to anti-cancer drugs when the tumor size is small or intermediate.
+Our results quantitatively suggest that aneuploidy can play an important role in tumor adaptation to anti-cancer drugs when the tumor size is small or intermediate.
 Large tumors are predicted to adapt to anti-cancer drugs through direct mutation.
-In contrast, smaller tumors are predicted to become resistant either directly by aneuploidy or by a resistance mutation occurring in aneuploid cells that serve as evolutionary ``stepping stones'' (\Cref{rescue_prob}).
+In contrast, smaller tumors are predicted to become resistant either directly by aneuploidy or by a resistance mutation occurring in aneuploid cells that serve as evolutionary ``stepping stones''.
 Thus, therapies that increase the rate of aneuploidy in tumors to combat cancer may also promote drug resistance.

### 2025-01-29 ms.tex
Add files via upload

@@
 \section{Recurrence time}\label{sec:appendix_recurrence_time} % TODO it is not clear what are the assumptions here; reading the main text and then coming here I think that you assume tolerant aneuploidy - is that the case?
-you need to be EXPLICIT
-We define the proliferation time $\tau_a^p$  to be the expected time for a resistant lineage to grow to the initial tumor size $N$.
+you need to be EXPLICIT OK
+For this section we assume that aneuploids are tolerant and we define the proliferation time $\tau_a^p$  to be the expected time for the number of resistant cells to grow to the initial tumor size $N$.
 The number of rescue lineages generated by the sensitive population is given by  \cref{eq:twosteplineage} (see \Cref{ExpectedNumberRescueLineages}):
 \begin{equation*}
-n_1\left(\infty\right)=\frac{uv\lambda_s\lambda_aNp_m}{\abs{\Delta_s}\abs{\Delta_a}}=\frac{N}{N_a^*},
+n_1\left(\infty\right)=\frac{uv\lambda_s\lambda_aNp_m}{\abs{\Delta_s}\abs{\Delta_a}}+\frac{v\lambda_sNp_m}{\abs{\Delta_s}}=\frac{N}{N_a^*}+\frac{N}{N_m^*},
 \end{equation*}
-where we ignore lineages created by direct mutation because we assumed $u\lambda_a > \max{(-\Delta_a, 1/T^*)}$, $N\ll N_m^*$ and used \cref{eq:N_a}.
+where we assumed that aneuploidy is frequent enough to affect the evolution of drug resistance (i.e.
+$u\lambda_a > \max{(-\Delta_a, 1/T^*)}$).
+Additionally, we have used \cref{eq:N_a,eq:N_m} to write $N_a^*=\frac{\abs{\Delta_s}\abs{\Delta_a}}{uv\lambda_s\lambda_ap_m}$ and $N_m^*=\frac{\abs{\Delta_s}}{v\lambda_sp_m}$.
 % TODO Remus why did you assume these assumptions?
@@
 write it EXPLICITLY.
+OK
 
 This helps us distinguish between two scenarios for the proliferation time. 
-In the first scenario, $N\ll N_a^*$, and we have at most one lineage which rescues the cancer cell population.
+In the first scenario, $N\ll N_a^*\ll N_m^*$, and we have at most one lineage which rescues the cancer cell population.
 As a result, the recurrence time is given by~\citep{avanzini2019cancer}:
 \begin{align}\label{meanproliferationtime}
-\tau_a^r&\approx\tau_a+\frac{\log p_mN}{\Delta_m}. % TODO is it log(pm) N or log(pm N) ? add parentheses.
+\tau_a^r&\approx\tau_a+\frac{\log \left(p_mN\right)}{\Delta_m}. % TODO is it log(pm) N or log(pm N) ? add parentheses. OK
 \end{align}
-The factor of $p_m$ in the second term of \cref{meanproliferationtime} is due to the fact that the lineage is conditioned to survive genetic drift and the time to reach $N$ is shorter then the scenario without this property \citep{orr2014population,smith1974hitch}.
+The factor of $p_m$ in the second term of \cref{meanproliferationtime} is due to the fact that the lineage is conditioned to survive genetic drift and the time to reach $N$ is shorter then the scenario when the lineage is not conditioned to survive stochastic extinction~\citep{orr2014population,smith1974hitch}.
 % change 35 % TODO without what property?
-
-In the second scenario, $N\gg N_a^*$, and the sensitive population produces a large number of rescue lineages in a short period of time.
+OK
+
+In the second scenario, $N\gg N_m^*$, and the sensitive population produces a large number of rescue lineages in a short period of time.
 As a result, the recurrence time is obtained by solving the following system of oridnary differential equations (ODEs),
@@
 \end{equation}
-This is a transcendental equation which cannot be solved exactly but we can obtain an approximation by noting that for large $\tau_a^r$ the above equation can be written as % TODO Remus what do you mean large tau?
+This is a transcendental equation which cannot be solved exactly but we can obtain an approximation by noting that $\e^{\Delta_m \tau_a^r} \gg\e^{\Delta_s \tau_a^r},\e^{\Delta_a \tau_a^r}$ for even small values of $\tau_a^r$ because $\Delta_m>0$ and $\Delta_s,\Delta_a<0$.
+Given we assumed that the tumor is large (i.e.
+$N\gg N_a^*$) then the time $\tau_a^r$ necessary for the resistant cell population to reach $N$ is large enough such that the first term of \cref{eq:t2det2024} can be written as: % TODO Remus what do you mean large tau?
 large compared to what?
@@
 the editor  has asked for as much.
+OK
+\begin{align*}
+\frac{uv\lambda_a\lambda_s}{\Delta_s-\Delta_a}\left[\frac{\e^{\Delta_s\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_s-\Delta_m}-\frac{\e^{\Delta_a\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_a-\Delta_m}\right]\approx\frac{uv\lambda_a\lambda_s\e^{\Delta_m\tau_a^r}}{\left(\Delta_m-\Delta_s\right)\left(\Delta_m-\Delta_a\right)}
+\end{align*}
+which is significantly smaller then the second term of \cref{eq:t2det2024}:
+\begin{align*}
+v\lambda_s\frac{\e^{\Delta_s\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_s-\Delta_m}\approx\frac{v\lambda_s\e^{\Delta_m\tau_a^r}}{\left(\Delta_m-\Delta_a\right)}
+\end{align*}
+Consequently, \cref{eq:t2det2024} can be approximated as:
 \begin{equation*}
@@
 \begin{align}\label{meanproliferationtime2}
-\tau_a^{r,M}&\approx\tau_a+\frac{\log p_mM}{\Delta_m},
+\tau_a^{r,M}&\approx\tau_a+\frac{\log \left(p_mM\right)}{\Delta_m},
 \end{align}

### 2025-01-29 src/Figure3/Figure3A.py
Add files via upload

@@ -1,0 +1,223 @@
+import numpy as np
+import matplotlib.pyplot as plt
+import seaborn as sns
+sns.set_context('talk')
+
+# s = sensitive
+# a = aneuploid
+# m = mutant
+ 
+def NaS(λs, λa, λm,  μs,  μa,  μm, u, v):
+    
+    ds = λs - μs
+    da = λa - μa
+    dm = λm - μm
+    
+    Ts = 1/np.sqrt(4*v*λa*λa*dm/λm)
+    
+    Ret = 0
+    
+    if da*Ts<-1:
+        
+        Ret = abs(da)/(v*λa)*λm/dm 
+        
+    if abs(da*Ts)<1:
+        
+        Ret = 2*λa*Ts
+        
+    if da*Ts>1:
+        
+        Ret = λa/da
+        
+    return abs(ds)/(u*λs)*Ret
+
+def prob(N, λs, λa, λm,  μs,  μa,  μm, u, v):
+    
+    return 1 - np.exp(-N/NaS(λs, λa, λm,  μs,  μa,  μm, u, v))
+        
+#Model parameters
+λs, λa, λm, μs, μa, μm, u, v = 0.1, 0.0899, 0.1, 0.14, 0.09, 0.09, 10**(-2),10**(-7)
+
+ds = λs - μs
+dm = λm - μm
+
+Ts = 1/np.sqrt(4*v*λa*λa*dm/λm)
+
+fig = plt.figure(figsize=(7,5.5),frameon=False)
+ax = plt.subplot(111)
+
+ax.spines['top'].set_visible(False)
+ax.spines['right'].set_visible(False)
+
+ax.yaxis.set_ticks_position('left')
+ax.xaxis.set_ticks_position('bottom')
+
+Col = ['k','b','g','m']
+
+Col1 = ['r--','b--','g--']
+
+PopSize = np.geomspace(10**0,10**9,10000)
+Nms = abs(ds)/(v*λs)*λm/dm
+
+#Partially resistant aneuploids
+Prob = [prob(x, λs, 0.091, λm,  μs,  μa,  μm, u, v) for x in PopSize]
+
+plt.plot(PopSize, Prob, color=Col[2], linewidth=1.5, label=r'partially resistant') 
+
+#Stationary aneuploids
+Prob = [prob(x, λs, 0.09-10**(-5), λm,  μs,  μa,  μm, u, v) for x in PopSize]
+
+plt.plot(PopSize, Prob, color='r', linewidth=1.5, label=r'stationary') 
[…]

### 2025-01-29 src/Figure3/Figure3.py
Delete src/Figure3/Figure3.py

@@ -1,223 +1,0 @@
-import numpy as np
-import matplotlib.pyplot as plt
-import seaborn as sns
-sns.set_context('talk')
-
-# s = sensitive
-# a = aneuploid
-# m = mutant
- 
-def NaS(λs, λa, λm,  μs,  μa,  μm, u, v):
-    
-    ds = λs - μs
-    da = λa - μa
-    dm = λm - μm
-    
-    Ts = 1/np.sqrt(4*v*λa*λa*dm/λm)
-    
-    Ret = 0
-    
-    if da*Ts<-1:
-        
-        Ret = abs(da)/(v*λa)*λm/dm 
-        
-    if abs(da*Ts)<1:
-        
-        Ret = 2*λa*Ts
-        
-    if da*Ts>1:
-        
-        Ret = λa/da
-        
-    return abs(ds)/(u*λs)*Ret
-
-def prob(N, λs, λa, λm,  μs,  μa,  μm, u, v):
-    
-    return 1 - np.exp(-N/NaS(λs, λa, λm,  μs,  μa,  μm, u, v))
-        
-#Model parameters
-λs, λa, λm, μs, μa, μm, u, v = 0.1, 0.0899, 0.1, 0.14, 0.09, 0.09, 10**(-2),10**(-7)
-
-ds = λs - μs
-dm = λm - μm
-
-Ts = 1/np.sqrt(4*v*λa*λa*dm/λm)
-
-fig = plt.figure(figsize=(7,5.5),frameon=False)
-ax = plt.subplot(111)
-
-ax.spines['top'].set_visible(False)
-ax.spines['right'].set_visible(False)
-
-ax.yaxis.set_ticks_position('left')
-ax.xaxis.set_ticks_position('bottom')
-
-Col = ['k','b','g','m']
-
-Col1 = ['r--','b--','g--']
-
-PopSize = np.geomspace(10**0,10**9,10000)
-Nms = abs(ds)/(v*λs)*λm/dm
-
-#Partially resistant aneuploids
-Prob = [prob(x, λs, 0.091, λm,  μs,  μa,  μm, u, v) for x in PopSize]
-
-plt.plot(PopSize, Prob, color=Col[2], linewidth=1.5, label=r'partially resistant') 
-
-#Stationary aneuploids
-Prob = [prob(x, λs, 0.09-10**(-5), λm,  μs,  μa,  μm, u, v) for x in PopSize]
-
-plt.plot(PopSize, Prob, color='r', linewidth=1.5, label=r'stationary') 
[…]

### 2025-01-29 src/Figure3/Figure3B.py
Add files via upload

@@ -1,0 +1,112 @@
+import numpy as np
+import matplotlib.pyplot as plt
+
+fig = plt.figure(figsize=(7.4,5.5),frameon=False)
+ax = plt.subplot(111)
+
+ax.spines['top'].set_visible(False)
+ax.spines['right'].set_visible(False)
+
+ax.yaxis.set_ticks_position('left')
+ax.xaxis.set_ticks_position('bottom')
+
+def prob1cellMelanoma(λs, λa, λm,  μs,  μa,  μm, u, v):
+    
+    ds = λs - μs
+    da = λa - μa
+    dm = λm - μm
+    
+    Ts = 1/np.sqrt(4*v*λa*λa*dm/λm)
+    
+    Ret = 0
+    
+    if da*Ts<-1:
+        
+        Ret = abs(da)/(v*λa)*λm/dm 
+        
+    if abs(da*Ts)<1:
+        
+        Ret = 2*λa*Ts
+        
+    if da*Ts>1:
+        
+        Ret = λa/da
+        
+    return (abs(ds)/(u*λs)*Ret)**(-1)
+
+
+def probNcellMelanoma(N, λs, λa, λm,  μs,  μa,  μm, u, v):
+    
+    return 1 - np.exp(-N*prob1cellMelanoma(λs, λa, λm,  μs,  μa,  μm, u, v))
+    
+u = 10**(-2)
+v = 10**(-7)
+λa1, μa1 = 0.0845, 0.076
+λa2, μa2 = 0.0845, 0.1015
+λa3, μa3 = 0.0845, 0.1115
+
+λs, μs = 0.0845, 0.1215
+
+λm, μm = 0.0845, 0.076
+dm = λm - μm
+
+Ts1 = 1/np.sqrt(4*v*λa1*λa1*dm/λm)
+Ts2 = 1/np.sqrt(4*v*λa2*λa2*dm/λm)
+Ts3 = 1/np.sqrt(4*v*λa3*λa3*dm/λm)
+
+da1 = λa1 - μa1
+da2 = λa2 - μa2
+da3 = λa3 - μa3
+
+N = np.geomspace(10**0, 10**11, 1000)
+
+plt.plot(N,probNcellMelanoma(N, λs, λa1, λm,  μs,  μa1,  μm, u, v),color='r', clip_on=False, label='TNBC-SA609 \nclone A')
[…]

### 2025-01-29 src/Figure5/Figure5A.py
Add files via upload

@@ -1,76 +1,17 @@
-#import matplotlib.inline
 import numpy as np
 import matplotlib.pyplot as plt
-import numba
-import scipy.stats
-from scipy.integrate import solve_ivp
-from scipy.integrate import odeint
 import seaborn as sns
 sns.set_context('talk')
 
-from functools import partial
 
-# n = naive
+# s = sensitive
 # a = aneuploidy
 # m = mutant
- 
-def NaS(λw, λa, λm,  μw,  μa,  μm, u, v):
-    
-    dw = λw - μw
-    da = λa - μa
-    dm = λm - μm
-    
-    Ts = 1/np.sqrt(4*v*λa*λa*dm/λm)
-    
-    Ret = 0
-    
-    if da*Ts<-1:
         
-        Ret = abs(da)/(v*λa)*λm/dm # abs(dw)*(1-u/abs(da))/v*λm/dm
-        
-    if abs(da*Ts)<1:
-        
-        Ret = 2*λa*Ts
-        
-    if da*Ts>1:
-        
-        Ret = λa/da
-        
-    return abs(dw)/(u*λw)*Ret
-
-def NaS2(λw, λa, λm,  μw,  μa,  μm, u, v):
-    
-    dw = λw - μw
-    da = λa - μa
-    dm = λm - μm
-    
-    Ts = 1/np.sqrt(4*v*λa*λa*dm/λm)
-    
-    Ret = 0
-    
-    if da*Ts<-1:
-        
-        Ret = abs(da)/(v*λa)*λm/dm # abs(dw)*(1-u/abs(da))/v*λm/dm
-        
-    if abs(da*Ts)<1:
-        
-        Ret = 2*λa*Ts
-        
-    if da*Ts>1:
-        
-        Ret = λa/da
-        
-    return Ret
-        
-λw, λa, λm, μa, μm, u, v = 0.1, 0.0899, 0.1, 0.09, 0.09, 10**(-2), 10**(-7)
+λs, λa, λm, μa, μm, u, v = 0.1, 0.0899, 0.1, 0.09, 0.09, 10**(-2), 10**(-7)
 s, us = 7.3*10**(-3), 10**(-3)
-print(u/us*(abs(s)/abs(λw - 0.14)),'ratio')
 
-#dw = λw - μw
 dm = λm - μm
-
-#Ts = np.sqrt(4*v*λa*λa*dm/λm)
-#print(Ts,r'Ts')
 
 fig = plt.figure(frameon=False)
[…]

### 2025-01-29 src/Figure5/Figure5B.py

@@ -10,40 +10,12 @@ sns.set_context('talk')
 
 from functools import partial
 
-# n = naive
+# s = sensitive
 # a = aneuploidy
 # m = mutant
- 
-def NaS(λw, λa, λm,  μw,  μa,  μm, u, v):
-    
-    dw = λw - μw
-    da = λa - μa
-    dm = λm - μm
-
-    
-    Ts = 1/np.sqrt(4*v*λa*dm/λm)
-    
-    Ret = 0
-    
-    if da*Ts<-1:
-        
-        Ret = abs(da)/v*λm/dm
-        
-    if abs(da*Ts)<1:
-        
-        Ret = 2*λa*Ts
-        
-    if da*Ts>1:
-        
-        Ret = λa/da
-        
-    return abs(dw)/u*Ret
-        
-λw, λa, λm, μw, μa, μm, v = 0.1, 0.0899, 0.1, 0.14, 0.09, 0.09, 10**(-7)
+       
+λs, λa, λm, μs, μa, μm, v = 0.1, 0.0899, 0.1, 0.14, 0.09, 0.09, 10**(-7)
 s, us = 7.3*10**(-3), 10**(-3)
-
-#Ts = 1/np.sqrt(4*v*λa*dm/λm)
-#print(Ts,r'Ts')
 
 fig = plt.figure(frameon=False)
 ax = plt.subplot(111)
@@ -54,7 +26,6 @@ ax.spines['right'].set_visible(False)
 ax.yaxis.set_ticks_position('left')
 ax.xaxis.set_ticks_position('bottom')
 
-#Nm = 10**(4)
 
 Col = ['r','b','g']
 
@@ -63,16 +34,15 @@ Col1 = ['r--','b--','g--']
 Y = [-0.01,-0.005,-0.001]
 
 u = np.geomspace(10**(-3),10**(-1),1000)
-N = [U/us*(abs(s)/abs(λw - μw)) for U in u]
+N = [U/us*(abs(s)/abs(λs - μs)) for U in u]
 x = [us/U for U in u]
 plt.plot(x, N, color='k', linewidth=1.5)   
-
 
 #Stochastic simulations plot
     
 color = ['ro','bo','go']
 
-txt=str("DataGenerationFractionAneuploidyDanielratiomutations.txt")
+txt=str("DataGenerationFractionAneuploidyratiomutations.txt")
 text_file = open(txt, "r")
[…]

### 2025-01-30 ms.tex
edited A4 and added q for RS

--- 2025-01-26 ms.tex
@@
 
-\section{Recurrence time}\label{sec:appendix_recurrence_time} % TODO it is not clear what are the assumptions here; reading the main text and then coming here I think that you assume tolerant aneuploidy - is that the case?
-you need to be EXPLICIT
-We define the proliferation time $\tau_a^p$  to be the expected time for a resistant lineage to grow to the initial tumor size $N$.
-The number of rescue lineages generated by the sensitive population is given by  \cref{eq:twosteplineage} (see \Cref{ExpectedNumberRescueLineages}):
+\section{Recurrence time}\label{sec:appendix_recurrence_time} 
+In the following, we assume aneuploid cells are tolerant ($\Delta_a T^* \ll 1$) and frequent enough to affect the evolution of drug resistance (i.e. $u\lambda_a > \max{(-\Delta_a, 1/T^*)}$). 
+We define the proliferation time $\tau_a^p$  to be the expected time for the number of resistant cells to grow to the initial tumor size $N$.
+The number of rescue lineages generated by the sensitive population is given by  \cref{eq:twosteplineage} (see \Cref{ExpectedNumberRescueLineages}),
 \begin{equation*}
-n_1\left(\infty\right)=\frac{uv\lambda_s\lambda_aNp_m}{\abs{\Delta_s}\abs{\Delta_a}}=\frac{N}{N_a^*},
+n_1\left(\infty\right)=\frac{uv\lambda_s\lambda_aNp_m}{\abs{\Delta_s}\abs{\Delta_a}}+\frac{v\lambda_sNp_m}{\abs{\Delta_s}}=\frac{N}{N_a^*}+\frac{N}{N_m^*} .
 \end{equation*}
-where we ignore lineages created by direct mutation because we assumed $u\lambda_a > \max{(-\Delta_a, 1/T^*)}$, $N\ll N_m^*$ and used \cref{eq:N_a}.
-% TODO Remus why did you assume these assumptions?
-say the biological assumption, not just the math.
-and how did you use eq 3?
-write it EXPLICITLY.
-
-This helps us distinguish between two scenarios for the proliferation time. 
-In the first scenario, $N\ll N_a^*$, and we have at most one lineage which rescues the cancer cell population.
-As a result, the recurrence time is given by~\citep{avanzini2019cancer}:
+We use \cref{eq:N_a,eq:N_m} to write $N_a^*=\frac{\abs{\Delta_s}\; \abs{\Delta_a}}{uv\lambda_s\lambda_ap_m}$ and $N_m^*=\frac{\abs{\Delta_s}}{v\lambda_sp_m}$.
+
+We distinguish between two scenarios for the proliferation time. 
+In the first scenario, $N\ll N_a^*\ll N_m^*$, and we have at most one lineage that rescues the cancer cell population.
+As a result, the recurrence time is given by~\citep{avanzini2019cancer}
 \begin{align}\label{meanproliferationtime}
-\tau_a^r&\approx\tau_a+\frac{\log p_mN}{\Delta_m}. % TODO is it log(pm) N or log(pm N) ? add parentheses.
+\tau_a^r&\approx\tau_a+\frac{\log \left(p_mN\right)}{\Delta_m}.
 \end{align}
-The factor of $p_m$ in the second term of \cref{meanproliferationtime} is due to the fact that the lineage is conditioned to survive genetic drift and the time to reach $N$ is shorter then the scenario without this property \citep{orr2014population,smith1974hitch}.
-% change 35 % TODO without what property?
-
-In the second scenario, $N\gg N_a^*$, and the sensitive population produces a large number of rescue lineages in a short period of time.
+The factor $p_m$ in the second term of \cref{meanproliferationtime} occurs because the lineage is conditioned to survive stochastic extinction and the time to reach $N$ is shorter compared to the scenario where the lineage is not conditioned to survive stochastic extinction~\citep{orr2014population,smith1974hitch}.
+% change 35
+
+In the second scenario, $N\gg N_m^*$, and the sensitive population produces a large number of rescue lineages in a short period of time.
 As a result, the recurrence time is obtained by solving the following system of oridnary differential equations (ODEs),
@@
 \end{equation}
-This is a transcendental equation which cannot be solved exactly but we can obtain an approximation by noting that for large $\tau_a^r$ the above equation can be written as % TODO Remus what do you mean large tau?
-large compared to what?
-why do you assume it is large?
-what is the biological assumption?
-It is not clear why you neglect two out of three terms, and you need to be explicit about this -- say both what the math is but also what the biological assumption is.
-the editor  has asked for as much.
+This is a transcendental equation which cannot be solved exactly but we can obtain an approximation by noting that $\e^{\Delta_m \tau_a^r} \gg\e^{\Delta_s \tau_a^r},\e^{\Delta_a \tau_a^r}$ for even small values of $\tau_a^r$ because $\Delta_m>0$ and $\Delta_s,\Delta_a<0$.
+We assume that the tumor is large ($N\gg N_a^*$) % REMUS you assumed N\gg N_m^* not N\gg N_a^*
+so the time $\tau_a^r$ for the resistant cell population to reach $N$ is large enough such that the first term of \cref{eq:t2det2024} can be written as
+\begin{align*}
+\frac{uv\lambda_a\lambda_s}{\Delta_s-\Delta_a}\left[\frac{\e^{\Delta_s\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_s-\Delta_m}-\frac{\e^{\Delta_a\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_a-\Delta_m}\right]\approx\frac{uv\lambda_a\lambda_s\e^{\Delta_m\tau_a^r}}{\left(\Delta_m-\Delta_s\right)\left(\Delta_m-\Delta_a\right)} ,
+\end{align*}
+which is significantly smaller then the second term of \cref{eq:t2det2024},
+\begin{align*}
+v\lambda_s\frac{\e^{\Delta_s\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_s-\Delta_m}\approx\frac{v\lambda_s\e^{\Delta_m\tau_a^r}}{\left(\Delta_m-\Delta_a\right)} .
+\end{align*}
+Consequently, \cref{eq:t2det2024} can be approximated by
 \begin{equation*}
@@
 \begin{equation}\label{eq:t2det}
-\tau_a^r\approx\frac{1}{\Delta_m}\log\frac{\Delta_m-\Delta_s}{v\lambda_s}.
+\tau_a^r\approx\frac{1}{\Delta_m}\log{\left(\frac{\Delta_m-\Delta_s}{v\lambda_s}\right)}.
 \end{equation}
@@
 \begin{equation}\label{eq:t2det2}
-\tau_a^{r,M}\approx\frac{1}{\Delta_m}\log\frac{M\left(\Delta_m-\Delta_s\right)}{v\lambda_sN},
+\tau_a^{r,M}\approx\frac{1}{\Delta_m}\log{\left(\frac{M\left(\Delta_m-\Delta_s\right)}{v\lambda_sN}\right)},
 \end{equation}

### 2025-01-30 ms.tex
Add files via upload

@@
 This is a transcendental equation which cannot be solved exactly but we can obtain an approximation by noting that $\e^{\Delta_m \tau_a^r} \gg\e^{\Delta_s \tau_a^r},\e^{\Delta_a \tau_a^r}$ for even small values of $\tau_a^r$ because $\Delta_m>0$ and $\Delta_s,\Delta_a<0$.
-We assume that the tumor is large ($N\gg N_a^*$) % REMUS you assumed N\gg N_m^* not N\gg N_a^*
+We assume that the tumor is large ($N\gg N_m^*$) % REMUS you assumed N\gg N_m^* not N\gg N_a^* FIXED
 so the time $\tau_a^r$ for the resistant cell population to reach $N$ is large enough such that the first term of \cref{eq:t2det2024} can be written as

### 2025-02-01 ms.tex
edited appendix D

--- 2025-01-30 ms.tex
@@
 \renewcommand{\d}{{\rm d}}
-\newcommand{\e}{{\rm e}}
+\newcommand{\e}{e}
 \newcommand{\erfc}{{\rm erfc}}
@@
 \section{Recurrence time}\label{sec:appendix_recurrence_time}
-In the following, we assume aneuploid cells are tolerant ($\Delta_a T^* \ll 1$) and frequent enough to affect the evolution of drug resistance (i.e. $u\lambda_a > \max{(-\Delta_a, 1/T^*)}$). 
+
+In the following, we assume aneuploid cells are tolerant ($\Delta_a T^* \ll -1$) and frequent enough to affect the evolution of drug resistance (i.e. $u\lambda_a \gg \max{(-\Delta_a, 1/T^*)}$). 
 We define the proliferation time $\tau_a^p$  to be the expected time for the number of resistant cells to grow to the initial tumor size $N$.
@@
 In the second scenario, $N\gg N_m^*$, and the sensitive population produces a large number of rescue lineages in a short period of time.
-As a result, the recurrence time is obtained by solving the following system of oridnary differential equations (ODEs),
+As a result, the recurrence time is obtained by solving the following system of ordinary differential equations (ODEs),
 \begin{equation}\label{detODE}
@@
 \end{equation}
-Solving the system of ODEs for initial condition $\left(s(0), a(0), m(0)\right)=\left(N,0,0\right)$ we obtain
+Solving the system of ODEs for initial condition $\left(s(0), a(0), m(0)\right)=\left(N,0,0\right)$ we obtain:
 \begin{equation*} \label{m_t}
@@
 \end{equation}
-This is a transcendental equation which cannot be solved exactly but we can obtain an approximation by noting that $\e^{\Delta_m \tau_a^r} \gg\e^{\Delta_s \tau_a^r},\e^{\Delta_a \tau_a^r}$ for even small values of $\tau_a^r$ because $\Delta_m>0$ and $\Delta_s,\Delta_a<0$.
-We assume that the tumor is large ($N\gg N_a^*$) % REMUS you assumed N\gg N_m^* not N\gg N_a^*
-so the time $\tau_a^r$ for the resistant cell population to reach $N$ is large enough such that the first term of \cref{eq:t2det2024} can be written as
-\begin{align*}
-\frac{uv\lambda_a\lambda_s}{\Delta_s-\Delta_a}\left[\frac{\e^{\Delta_s\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_s-\Delta_m}-\frac{\e^{\Delta_a\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_a-\Delta_m}\right]\approx\frac{uv\lambda_a\lambda_s\e^{\Delta_m\tau_a^r}}{\left(\Delta_m-\Delta_s\right)\left(\Delta_m-\Delta_a\right)} ,
-\end{align*}
-which is significantly smaller then the second term of \cref{eq:t2det2024},
-\begin{align*}
-v\lambda_s\frac{\e^{\Delta_s\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_s-\Delta_m}\approx\frac{v\lambda_s\e^{\Delta_m\tau_a^r}}{\left(\Delta_m-\Delta_a\right)} .
-\end{align*}
-Consequently, \cref{eq:t2det2024} can be approximated by
-\begin{equation*}
-1=v\lambda_s\frac{\e^{\Delta_m\tau_a^r}}{\Delta_m - \Delta_s},
-\end{equation*}
-which has an approximate solution
+
+\eqref{eq:t2det2024} is a transcendental equation which cannot be solved exactly but we can obtain an approximation by noting that the final tumor size $N$ is very large, and that since the sensitive and aneuploid cells have negative growth rates, almost all of the growth needed to rebound to size $N$ will be by mutant cells.
+Mathematically, this corresponds to $\e^{\Delta_m \tau_a^r}$ being much larger than the other exponential terms,  $\e^{\Delta_s \tau_a^r}$ and $\e^{\Delta_a \tau_a^r}$. 
+\eqref{eq:t2det2024} can then be rewritten as:
+\[
+	1 \approx \frac{v \lambda_s e^{\Delta_m \tau_a^r}}{\Delta_m - \Delta_s} 
+	\left(\frac{u \lambda_a}{\Delta_m - \Delta_a} + 1\right).
+\]
+Assuming that rate of aneuploidy is not so high that it overwhelms the aneuploids' growth disadvantage relative to the mutants (i.e., that $u \lambda_a \ll \Delta_m - \Delta_a$), we can neglect the first term in parentheses above, yielding:
 \begin{equation}\label{eq:t2det}
@@
 \end{equation}
-We observe that the terms given by the evolutionary trajectory \textit{sensitive} $\rightarrow$ \textit{aneuploid} $\rightarrow$ \textit{mutant} do not contribute to the above approximation and, as a result, we deduce that it is accurate only for $N \gg N_m^* > N_a^*$.
+We emphasize that in this case the population is large enough to produce mutants without needing to pass through aneuploidy ($N \gg N_m^*$), and so  the terms arising from the evolutionary trajectory \textit{sensitive} $\rightarrow$ \textit{aneuploid} $\rightarrow$ \textit{mutant} do not contribute to the above approximation.
 %change 9
@@
 \section{Distribution of evolutionary rescue time}\label{sec:appendix_distribution_time}
+
 The probability that a successful mutant has been generated by time $t$ is given by:

### 2025-02-02 ms.tex
edited app D

@@
 
-In the following, we assume aneuploid cells are tolerant ($\Delta_a T^* \ll -1$) and frequent enough to affect the evolution of drug resistance (i.e. $u\lambda_a \gg \max{(-\Delta_a, 1/T^*)}$). 
+In the following, we assume aneuploid cells are tolerant ($\Delta_a T^* \ll -1$) and frequent enough to affect the evolution of drug resistance ($u\lambda_a \gg \max{(-\Delta_a, 1/T^*)}$). 
 We define the proliferation time $\tau_a^p$  to be the expected time for the number of resistant cells to grow to the initial tumor size $N$.
@@
 We distinguish between two scenarios for the proliferation time. 
-In the first scenario, $N\ll N_a^*\ll N_m^*$, and we have at most one lineage that rescues the cancer cell population.
+In the first scenario, the tumor is small, $N\ll N_a^*\ll N_m^*$, and we have at most one lineage that rescues the cancer cell population.
 As a result, the recurrence time is given by~\citep{avanzini2019cancer}
@@
 
-In the second scenario, $N\gg N_m^*$, and the sensitive population produces a large number of rescue lineages in a short period of time.
+In the second scenario, the tumor is large, $N\gg N_m^*$, and the sensitive population produces a large number of rescue lineages in a short period of time.
 As a result, the recurrence time is obtained by solving the following system of ordinary differential equations (ODEs),
@@
 
-\eqref{eq:t2det2024} is a transcendental equation which cannot be solved exactly but we can obtain an approximation by noting that the final tumor size $N$ is very large, and that since the sensitive and aneuploid cells have negative growth rates, almost all of the growth needed to rebound to size $N$ will be by mutant cells.
+\Cref{eq:t2det2024} is a transcendental equation which cannot be solved exactly but we can obtain an approximation by noting that the tumor size $N$ is large, and because the sensitive and aneuploid cells have negative growth rates, almost all of the growth needed to rebound to size $N$ will be by mutant cells.
 Mathematically, this corresponds to $\e^{\Delta_m \tau_a^r}$ being much larger than the other exponential terms,  $\e^{\Delta_s \tau_a^r}$ and $\e^{\Delta_a \tau_a^r}$. 
-\eqref{eq:t2det2024} can then be rewritten as:
+\Cref{eq:t2det2024} can then be rewritten as
 \[
@@
 \]
-Assuming that rate of aneuploidy is not so high that it overwhelms the aneuploids' growth disadvantage relative to the mutants (i.e., that $u \lambda_a \ll \Delta_m - \Delta_a$), we can neglect the first term in parentheses above, yielding:
+Assuming that rate of aneuploidy is not so high that it overwhelms the aneuploids' growth disadvantage relative to the mutants ($u \lambda_a \ll \Delta_m - \Delta_a$), we can neglect the first term in parentheses above, yielding
 \begin{equation}\label{eq:t2det}

### 2025-02-03 ms.tex
Add files via upload

@@
 % change 35
+We note that using \cref{meantimet2} for $\tau_a$ in \cref{meanproliferationtime} offers a good approximation for the recurrence time for both $N\ll N_a^*$ and $N_a^*\ll N\ll N_m^*$ as seen in \Cref{ProliferationTimeLarge} when compared to stochastic simulations.
 
@@
 \caption{The mean time for the mutant cell population to reach size $N$, where $N$ is the initial number of cancer cells.
-Our inhomogeneous Poisson-process approximation (solid black line, \cref{meanproliferationtime}) is in agreement with simulation results (red markers with 95\% confidence intervals obtained with bootstrapping, see Appendix G) for small and intermediate values of $N$.
+Our inhomogeneous Poisson-process approximation (solid black line, \cref{meanproliferationtime} with \cref{meantimet2} for $\tau_a$) is in agreement with simulation results (red markers with 95\% confidence intervals obtained with bootstrapping, see Appendix G) for small and intermediate values of $N$.
 Dashed vertical blue line represents the threshold tumor size above which evolutionary rescue is very likely through aneuploidy \cref{eq:N_a} and the dashed vertical green line represents the threshold tumor size above which evolutionary rescue is very likely through direct mutation \cref{eq:N_m}.

### 2025-02-03 ms.tex
revise fig S6-S9 captions and AppD

--- 2025-02-02 ms.tex
@@
 
-We distinguish between two scenarios for the proliferation time. 
-In the first scenario, the tumor is small, $N\ll N_a^*\ll N_m^*$, and we have at most one lineage that rescues the cancer cell population.
+We distinguish between small, intermediate, or large tumors.
+In small ($N\ll N_a^*\ll N_m^*$) and intermediate ($N_a^* \ll N \ll N_m^*$) tumors, we have at most one lineage that rescues the cancer cell population.
 As a result, the recurrence time is given by~\citep{avanzini2019cancer}
@@
 % change 35
-
-In the second scenario, the tumor is large, $N\gg N_m^*$, and the sensitive population produces a large number of rescue lineages in a short period of time.
+In both scenarios, plugging \cref{meantimet2} for $\tau_a$ in \cref{meanproliferationtime} provides a good approximation for the recurrence time (black line in \Cref{ProliferationTimeLarge}).
+For small tumors, we have a simpler approximation by plugging \cref{limitapprox4} for $\tau_a$ to get $\tau_a^r\approx1/|\Delta_a| + \log{\left(p_mN\right)}/\Delta_m$ (blue line in \Cref{ProliferationTimeLarge}).
+
+In a large tumor, $N_m^* \ll N$, the sensitive population produces a large number of rescue lineages in a short period of time.
 As a result, the recurrence time is obtained by solving the following system of ordinary differential equations (ODEs),
@@
 \]
-Assuming that rate of aneuploidy is not so high that it overwhelms the aneuploids' growth disadvantage relative to the mutants ($u \lambda_a \ll \Delta_m - \Delta_a$), we can neglect the first term in parentheses above, yielding
+Assuming that rate of aneuploidy is not so high that it overwhelms the aneuploids' growth disadvantage relative to the mutants ($u \lambda_a \ll \Delta_m - \Delta_a$), we can neglect the first term in parentheses above, yielding (green line in \Cref{ProliferationTimeLarge})
 \begin{equation}\label{eq:t2det}
@@
 %%%
-\section{Bootstrapping}
+\section{Bootstrap}
 For the mean times the $95\%$ confidence interval is obtained through bootstrapping in the following steps: (1) we simulate $T$ 100 times; (2) we sample with replacement which we store in $T'$; (3) for each element of this sample we obtain $\tau=\mathbb{E}\left[T'\right]$; (4) we repeat steps (2)-(3) 100 times to obtain $\tau$ and we select the upper and lower limits such that $95\%$ of the values of $\tau$ lie in the interval given by the bounds.
@@
 \includegraphics[width=1\textwidth]{Figures/EvolutionaryRescueTimeComplete.pdf} %change r5
-\caption{Shown is the mean time for appearance of a resistance mutation that leads to evolutionary rescue with aneuploidy ($u>0$).
-Our inhomogeneous Poisson-process approximations (solid black lines: \cref{meantimet2}) are in agreement with simulation results (red markers with 95\% confidence intervals obtained with bootstrapping, see Appendix G).
-Dashed vertical blue line represents the threshold tumor size above which evolutionary rescue is very likely through aneuploidy \cref{eq:N_a} and the dashed vertical green line represents the threshold tumor size above which evolutionary rescue is very likely through direct mutation \cref{eq:N_m}.
-Solid lines represent the approximations \cref{eq:AsymptoticTimeRules} ($N<N_a^*$ blue line and $N>N_m^*$ green line).
+\caption{The mean time for appearance of a resistance mutation that leads to evolutionary rescue with aneuploidy ($u>0$). 
+Solid lines show our approximations (\cref{eq:AsymptoticTimeRules}) compared to red markers that show mean of simulations results (with error bars for 95\% confidence intervalues obtained with bootstrap, see Appendix G).
+Blue dashed line, $N_a^*$.
+Green dashed line, $N_m^*$.
 Parameters: $\lambda_s=0.1,\lambda_m=0.0899,\lambda_a=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.} %change 16
@@
 \caption{The mean time for the mutant cell population to reach size $N$, where $N$ is the initial number of cancer cells.
-Our inhomogeneous Poisson-process approximation (solid black line, \cref{meanproliferationtime}) is in agreement with simulation results (red markers with 95\% confidence intervals obtained with bootstrapping, see Appendix G) for small and intermediate values of $N$.
-Dashed vertical blue line represents the threshold tumor size above which evolutionary rescue is very likely through aneuploidy \cref{eq:N_a} and the dashed vertical green line represents the threshold tumor size above which evolutionary rescue is very likely through direct mutation \cref{eq:N_m}.
-Solid lines represents the approximations \cref{eq:AsymptoticTimeRulesRecurrence} ($N<N_a^*$ blue line and $N>N_m^*$ green line).
-The simulations converge to \cref{eq:t2det} (green line) for large values of $N\gg N_m^*$.
+Solid lines show our approximations (green: \cref{eq:t2det}, blue: \cref{meanproliferationtime} with \cref{limitapprox4} for $\tau_a$; black: \cref{meanproliferationtime} with \cref{meantimet2} for $\tau_a$) compared to red markers that show mean of simulations results (with error bars for 95\% confidence intervals obtained with bootstrap, see Appendix G).
+Blue dashed line, $N_a^*$.
+Green dashed line, $N_m^*$.
 Parameters: $\lambda_s=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}
@@
 \caption{The mean time for the mutant cell population to reach size $M$, where $M$ is the tumor detection size.
-Our inhomogeneous Poisson-process approximation (solid black line, \cref{meanproliferationtime2}) is in agreement with simulation results (red markers with 95\% confidence intervals obtained with bootstrapping, see Appendix G) for small and intermediate values of $N$.
-Dashed vertical blue line represents the threshold tumor size above which evolutionary rescue is very likely through aneuploidy \cref{eq:N_a} and the dashed vertical green line represents the threshold tumor size above which evolutionary rescue is very likely through direct mutation \cref{eq:N_m}.
-Solid blue line represents the approximation \cref{meanproliferationtime2} with $\tau_a$ from \cref{eq:AsymptoticTimeRules} for $N<N_a^*$ and the solid green line  represents the approximation \cref{eq:t2det2} for $N>N_m^*$.
-The simulations converge to \cref{eq:t2det2} (green line) for large values of $N\gg N_m^*$.
+Solid lines show our approximations (green: \cref{eq:t2det2} for $N>N_m^*$, blue:  \cref{meanproliferationtime2} with $\tau_a$ from \cref{eq:AsymptoticTimeRules} for $N<N_a^*$; black: \cref{meanproliferationtime2}) compared to red markers that show mean of simulations results (with error bars for 95\% confidence intervals obtained with bootstrap, see Appendix G).
+Blue dashed line, $N_a^*$.
+Green dashed line, $N_m^*$.
 Parameters: $\lambda_s=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}, M=10^7$.}
@@
 \includegraphics[width=1\textwidth]{Figures/SGVEvolutionaryRescueTimeComplete.pdf} %change r5
-\caption{Shown is the mean time for appearance of a resistance mutation that leads to evolutionary rescue with aneuploidy ($u>0$) when a fraction $f$ of cancer cells are aneuploid at the start of therapy.
-Black lines represent our inhomogeneous Poisson-process approximations (solid black line, \cref{meantimet2}; dashed black line \cref{meantimet2SGV}).
-Dashed black line is the  inhomogeneous Poisson-process approximation where a fraction $f$ of tumor is aneuploid at the onset of drug therapy which is in agreement with simulation results (red markers with 95\% confidence intervals obtained with bootstrapping, see Appendix G).
-Dashed vertical blue line represents the threshold tumor size above which evolutionary rescue is very likely through aneuploidy \cref{eq:N_a} and the dashed vertical green line represents the threshold tumor size above which evolutionary rescue is very likely through direct mutation \cref{eq:N_m}.
-Solid lines represents the approximations \cref{eq:AsymptoticTimeRules} ($N<N_a^*$ blue line and $N>N_m^*$ green line).
+\caption{The mean time for appearance of a resistance mutation that leads to evolutionary rescue with aneuploidy ($u>0$) when a fraction $f$ of cancer cells are aneuploid at the start of drug treatment.
+Lines show our approximations (solid green: \cref{eq:AsymptoticTimeRules} for $N>N_m^*$; solid blue:  \cref{eq:AsymptoticTimeRules} for $N<N_a^*$; solid black: \cref{meantimet2}; dashed black: \cref{meantimet2SGV}) compared to red markers that show mean of simulations results (with error bars for 95\% confidence intervals obtained with bootstrap, see Appendix G).
+Blue dashed line, $N_a^*$.
+Green dashed line, $N_m^*$.
 Parameters: $\lambda_s=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7},f=0.14\%$.} %change 17

### 2025-02-03 ResponseLetter.tex
revised response to editor on appendix D

@@
 
-\textcolor{blue}{In the first term of (D3) we have: 
-\begin{align*}
-\frac{u v \lambda_a \lambda_s}{\Delta_a - \Delta_s}\left[ \frac{\e^{\Delta_m \tau_a^r} - \e^{\Delta_a \tau_a^r}}{\Delta_m - \Delta_a} - \frac{\e^{\Delta_m \tau_a^r} - \e^{\Delta_s \tau_a^r}}{\Delta_m - \Delta_s} \right]
-\approx \frac{uv\lambda_a\lambda_s\e^{\Delta_m\tau_a^r}}{\left(\Delta_m-\Delta_s\right)\left(\Delta_m-\Delta_a\right)}
-\end{align*}
-which is significantly smaller then the second term of (D3):
-\begin{align*}
-v \lambda_s \frac{\e^{\Delta_m \tau_a^r} - \e^{\Delta_s \tau_a^r}}{\Delta_m - \Delta_s} \approx \frac{v\lambda_s\e^{\Delta_m\tau_a^r}}{\Delta_m-\Delta_s}
-\end{align*}
-} % change 44
+\textcolor{blue}{
+We have thoroughly revised Appendix D and the corresponding Figure S7 to make the derivations clearer.
+}
 \\
-\textcolor{red}{I don't understand our answer here.
-If aneuploids are common and pretty good, can't they still matter?
-And I don't understand what we're saying about ``large $\tau_a^r$''---large compared to what?
-Aren't we describing the regime of relatively small $\tau_a^r$ here?
-}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+Revised the captions of supporting Figures S6-S9 to make them shorter and clearer.
+
 }

### 2025-02-04 src/Figure3/Figure3ASimulation.py
Add files via upload

@@ -1,0 +1,111 @@
+import numpy as np
+import matplotlib.pyplot as plt
+import numba
+import scipy.stats as stats
+import seaborn as sns
+sns.set_context('talk')
+
+@numba.jit # 2-fold faster
+def get_rates(s, a, m, λs, λa, λm, μs, μa, μm, u, v):
+    return np.array([
+        λs * s,    
+        μs * s,     
+        u * λs * s,
+        v * λs * s,
+        λa * a, 
+        μa * a,
+        v * λa * a,
+        λm * m,    
+        μm * m,
+        0,
+    ])
+    
+updates = np.array([
+    [1, 0, 0],  # growth sensitive
+    [-1, 0, 0], # death sensitive
+    [0, 1, 0], #mutation s->a
+    [0, 0, 1], #mutation s->m
+    [0, 1, 0],  # growth aneuploidy type
+    [0, -1, 0],  # death aneuploidy type
+    [0, 0, 1], # #mutation a->m
+    [0, 0, 1],  # growth mutant type
+    [0, 0, -1],   # death mutant type
+    [0, 0, 0]
+])
+    
+def τ_leap(s, a, m, λs, λa, λm, μs, μa, μm, u, v, τ):
+
+    rates = get_rates(s, a, m, λs, λa, λm, μs, μa, μm, u, v)
+    
+    try:
+        adj_rates = np.random.poisson(rates * τ)
+    except ValueError:
+        print(rates, τ, s, a, m)
+        raise
+        
+    adj_rates = np.random.poisson(rates * τ)
+        
+    Δs, Δa, Δm = updates.T @ adj_rates
+   
+    return τ, Δs, Δa, Δm
+
+
+def gillespie_ssa(λs, λa, λm, μs, μa, μm, u, v, s0, tmax, t0=0,  a0=0, m0=0, t_steps=1000):
+    
+    ΔT = (tmax-t0)/t_steps
+        
+    next_t = t0
+    
+    t = t0
+    
+    s, a, m = s0, a0, m0
+    
+    Δs, Δa, Δm = 0, 0, 0
+    
+    T = []
+    S = []
+    A = []
+    M = []
+    
+    Lim = int(-3*np.log(10)/np.log(μm/λm)) + 1
+    # loop over recording times
+    while s + a + m > 0:
+        while t < next_t:
+            Δt, Δs, Δa, Δm = τ_leap(s, a, m, λs, λa, λm, μs, μa, μm, u, v, τ)
[…]

### 2025-02-04 src/Figure4/Figure4ASimulation.py
Add files via upload

@@ -1,0 +1,151 @@
+import mpmath as mp
+import numpy as np
+import numba
+import seaborn as sns
+from random import choices
+
+sns.set_context('talk')
+
+@numba.jit # 2-fold faster
+def get_rates(w, a, m, λw, λa, λm, μw, μa, μm, u, v):
+    return np.array([
+        λw * w,    
+        μw * w,     
+        u * λw * w,
+        v * λw * w,
+        λa * a, 
+        μa * a,
+        v * λa * a,
+        λm * m,    
+        μm * m,
+        0,
+    ])
+        
+updates = np.array([
+    [1, 0, 0],  # growth sensitive
+    [-1, 0, 0], # death sensitive
+    [0, 1, 0], #mutation s->a
+    [0, 0, 1], #mutation s->m
+    [0, 1, 0],  # growth aneuploidy type
+    [0, -1, 0],  # death aneuploidy type
+    [0, 0, 1], # #mutation a->m
+    [0, 0, 1],  # growth mutant type
+    [0, 0, -1],   # death mutant type
+    [0, 0, 0]
+])
+    
+def τ_leap(s, a, m, λs, λa, λm, μs, μa, μm, u, v, τ):
+
+    rates = get_rates(s, a, m, λs, λa, λm, μs, μa, μm, u, v)
+    
+    try:
+        adj_rates = np.random.poisson(rates * τ)
+    except ValueError:
+        print(rates, τ, s, a, m)
+        raise
+        
+    adj_rates = np.random.poisson(rates * τ) 
+    
+    Δs, Δa, Δm = updates.T @ adj_rates
+   
+    return τ, Δs, Δa, Δm
+
+
+def gillespie_ssa(λs, λa, λm, μs, μa, μm, u, v, s0, tmax, t0=0,  a0=0, m0=0, t_steps=1000):
+    
+    ΔT = (tmax-t0)/t_steps
+        
+    next_t = t0
+    
+    t = t0
+    
+    s, a, m = s0, a0, m0
+    
+    Δs, Δa, Δm = 0, 0, 0
+    
+    T = []
+    S = []
+    A = []
+    M = []
+    
+    Lim = int(-3*np.log(10)/np.log(μm/λm)) + 1
+    # loop over recording times
+    while s + a + m > 0:
+        while t < next_t:
+            Δt, Δs, Δa, Δm = τ_leap(s, a, m, λs, λa, λm, μs, μa, μm, u, v, τ)
[…]

### 2025-02-04 src/Figure4/Figure4BSimulation.py

@@ -1,0 +1,153 @@
+import mpmath as mp
+import numpy as np
+import numba
+import seaborn as sns
+from random import choices
+
+sns.set_context('talk')
+
+@numba.jit # 2-fold faster
+def get_rates(s, a, m, λs, λa, λm, μs, μa, μm, u, v):
+    return np.array([
+        λs * s,    
+        μs * s,     
+        u * λs * s,
+        v * λs * s,
+        λa * a, 
+        μa * a,
+        v * λa * a,
+        λm * m,    
+        μm * m,
+        0,
+    ])
+    
+updates = np.array([
+    [1, 0, 0],  # growth sensitive
+    [-1, 0, 0], # death sensitive
+    [0, 1, 0], #mutation s->a
+    [0, 0, 1], #mutation s->m
+    [0, 1, 0],  # growth aneuploidy type
+    [0, -1, 0],  # death aneuploidy type
+    [0, 0, 1], # #mutation a->m
+    [0, 0, 1],  # growth mutant type
+    [0, 0, -1],   # death mutant type
+    [0, 0, 0]
+])
+    
+def τ_leap(s, a, m, λs, λa, λm, μs, μa, μm, u, v, τ):
+
+    rates = get_rates(s, a, m, λs, λa, λm, μs, μa, μm, u, v)
+    
+    try:
+        adj_rates = np.random.poisson(rates * τ)
+    except ValueError:
+        print(rates, τ, s, a, m)
+        raise
+       
+    adj_rates = np.random.poisson(rates * τ) 
+    Δs, Δa, Δm = updates.T @ adj_rates
+   
+    return τ, Δs, Δa, Δm
+
+
+def gillespie_ssa(λs, λa, λm, μs, μa, μm, u, v, s0, tmax, t0=0,  a0=0, m0=0, t_steps=1000):
+    
+    ΔT = (tmax-t0)/t_steps
+        
+    next_t = t0
+    
+    t = t0
+    
+    s, a, m = s0, a0, m0
+    
+    Δs, Δa, Δm = 0, 0, 0
+    
+    T = []
+    S = []
+    A = []
+    M = []
+    
+    Lim = int(-3*np.log(10)/np.log(μm/λm)) + 1
+    # loop over recording times
+    while s + a + m > 0:
+        while t < next_t:
+            Δt, Δs, Δa, Δm = τ_leap(s, a, m, λs, λa, λm, μs, μa, μm, u, v, τ)
[…]

### 2025-02-04 src/Figure5/Figure5ASimulation.py
Add files via upload

@@ -1,0 +1,194 @@
+import mpmath as mp
+import numpy as np
+import numba
+import seaborn as sns
+sns.set_context('talk')
+from random import choices
+
+@numba.jit # 2-fold faster
+def get_rates(s, a, m, λs, λa, λm, μs, μa, μm, u, v):
+    return np.array([
+        λs * s,    
+        μs * s,     
+        u * λs * s,
+        v * λs * s,
+        λa * a, 
+        μa * a,
+        v * λa * a,
+        λm * m,    
+        μm * m,
+        0,
+    ])
+
+updates = np.array([
+    [1, 0, 0],  # growth sensitive
+    [-1, 0, 0], # death sensitive
+    [0, 1, 0], #mutation s->a
+    [0, 0, 1], #mutation s->m
+    [0, 1, 0],  # growth aneuploidy type
+    [0, -1, 0],  # death aneuploidy type
+    [0, 0, 1], # #mutation a->m
+    [0, 0, 1],  # growth mutant type
+    [0, 0, -1],   # death mutant type
+    [0, 0, 0]
+])
+
+def τ_leap(s, a, m, λs, λa, λm, μs, μa, μm, u, v, τ):
+
+    rates = get_rates(s, a, m, λs, λa, λm, μs, μa, μm, u, v)
+    try:
+        adj_rates = np.random.poisson(rates * τ)
+    except ValueError:
+        print(rates, τ, s, a, m)
+        raise
+        
+    adj_rates = np.random.poisson(rates * τ)
+        
+    Δs, Δa, Δm = updates.T @ adj_rates
+   
+    return τ, Δs, Δa, Δm
+
+
+def gillespie_ssa(λs, λa, λm, μs, μa, μm, u, v, s0, a0, tmax, t0=0, m0=0, t_steps=1000):
+    
+    ΔT = (tmax-t0)/t_steps
+        
+    next_t = t0
+    
+    t = t0
+    
+    s, a, m = s0, a0, m0
+    
+    
+    Δs, Δa, Δm = 0, 0, 0
+    
+    T = []
+    S = []
+    A = []
+    M = []
+    
+    Lim = int(-3*np.log(10)/np.log(μm/λm)) + 1
+    # loop over recording times
+    while s + a + m > 0:
+        while t < next_t:
[…]

### 2025-02-04 src/Figure5/Figure5BSimulation.py

@@ -1,0 +1,192 @@
+import mpmath as mp
+import numpy as np
+import numba
+import seaborn as sns
+sns.set_context('talk')
+from random import choices
+
+@numba.jit # 2-fold faster
+def get_rates(s, a, m, λs, λa, λm, μs, μa, μm, u, v):
+    return np.array([
+        λs * s,    
+        μs * s,     
+        u * λs * s,
+        v * λs * s,
+        λa * a, 
+        μa * a,
+        v * λa * a,
+        λm * m,    
+        μm * m,
+        0,
+    ])
+    
+updates = np.array([
+    [1, 0, 0],  # growth sensitive
+    [-1, 0, 0], # death sensitive
+    [0, 1, 0], #mutation s->a
+    [0, 0, 1], #mutation s->m
+    [0, 1, 0],  # growth aneuploidy type
+    [0, -1, 0],  # death aneuploidy type
+    [0, 0, 1], # #mutation a->m
+    [0, 0, 1],  # growth mutant type
+    [0, 0, -1],   # death mutant type
+    [0, 0, 0]
+])
+
+def τ_leap(s, a, m, λs, λa, λm, μs, μa, μm, u, v, τ):
+
+    rates = get_rates(s, a, m, λs, λa, λm, μs, μa, μm, u, v)
+    try:
+        adj_rates = np.random.poisson(rates * τ)
+    except ValueError:
+        print(rates, τ, s, a, m)
+        raise
+        
+    adj_rates = np.random.poisson(rates * τ)
+        
+    Δs, Δa, Δm = updates.T @ adj_rates
+   
+    return τ, Δs, Δa, Δm
+
+
+def gillespie_ssa(λs, λa, λm, μs, μa, μm, u, v, s0, a0, tmax, t0=0, m0=0, t_steps=1000):
+    
+    ΔT = (tmax-t0)/t_steps
+        
+    next_t = t0
+    
+    t = t0
+    
+    s, a, m = s0, a0, m0
+        
+    Δs, Δa, Δm = 0, 0, 0
+    
+    T = []
+    S = []
+    A = []
+    M = []
+    
+    Lim = int(-3*np.log(10)/np.log(μm/λm)) + 1
+    # loop over recording times
+    while s + a + m > 0:
+        while t < next_t:
[…]

### 2025-02-04 src/Figure6/Figure6ASimulation.py
Add files via upload

@@ -1,0 +1,143 @@
+import numpy as np
+import matplotlib.pyplot as plt
+import numba
+import seaborn as sns
+
+sns.set_context('talk')
+
+
+@numba.jit # 2-fold faster
+def get_rates(s, a, m, λs, λa, λm, μs, μa, μm, u, v):
+    return np.array([
+        λs * s,    
+        μs * s,     
+        u * λs * s,
+        v * λs * s,
+        λa * a, 
+        μa * a,
+        v * λa * a,
+        λm * m,    
+        μm * m,
+        0,
+    ])
+    
+@numba.jit # 2-fold faster
+def draw_time(rates):
+    total_rate = rates.sum()
+    
+    if total_rate > 0:
+        return np.random.exponential(1/total_rate)
+    if total_rate == 0:
+        return 1
+
+def draw_reaction(rates):
+    if rates.sum() > 0:
+        rates /= rates.sum()
+        return np.random.multinomial(1, rates).argmax()
+    if rates.sum() == 0:
+        return 9
+
+    
+updates = np.array([
+    [1, 0, 0],  # growth sensitive
+    [-1, 0, 0], # death sensitive
+    [0, 1, 0], #mutation s->a
+    [0, 0, 1], #mutation s->m
+    [0, 1, 0],  # growth aneuploidy type
+    [0, -1, 0],  # death aneuploidy type
+    [0, 0, 1], # #mutation a->m
+    [0, 0, 1],  # growth mutant type
+    [0, 0, -1],   # death mutant type
+    [0, 0, 0]
+])
+    
+def gillespie_step(s, a, m, λs, λa, λm, μs, μa, μm, u, v):
+    rates = get_rates(s, a, m, λs, λa, λm, μs, μa, μm, u, v)
+    Δt = draw_time(rates)
+    ri = draw_reaction(rates)
+    Δw, Δa, Δm= updates[ri]
+    return Δt, Δw, Δa, Δm
+
+def τ_leap(s, a, m, λs, λa, λm, μs, μa, μm, u, v, τ):
+
+    rates = get_rates(s, a, m, λs, λa, λm, μs, μa, μm, u, v)
+    try:
+        adj_rates = np.random.poisson(rates * τ)
+    except ValueError:
+        print(rates, τ, s, a, m)
+        raise
+
+    adj_rates = np.random.poisson(rates * τ)
+    Δs, Δa, Δm = updates.T @ adj_rates
+   
+    return τ, Δs, Δa, Δm
+
[…]

### 2025-02-04 src/Figure6/Figure6BSimulation.py

@@ -1,0 +1,100 @@
+import numpy as np
+import matplotlib.pyplot as plt
+import numba
+import seaborn as sns
+
+sns.set_context('talk')
+
+@numba.jit # 2-fold faster
+def get_rates(s, a, m, λs, λa, λm, μs, μa, μm, u, v):
+    return np.array([
+        λs * s,    
+        μs * s,     
+        u * λs * s,
+        v * λs * s,
+        λa * a, 
+        μa * a,
+        v * λa * a,
+        λm * m,    
+        μm * m,
+        0,
+    ])
+    
+updates = np.array([
+    [1, 0, 0],  # growth sensitive
+    [-1, 0, 0], # death sensitive
+    [0, 1, 0], #mutation s->a
+    [0, 0, 1], #mutation s->m
+    [0, 1, 0],  # growth aneuploidy type
+    [0, -1, 0],  # death aneuploidy type
+    [0, 0, 1], # #mutation a->m
+    [0, 0, 1],  # growth mutant type
+    [0, 0, -1],   # death mutant type
+    [0, 0, 0]
+])
+    
+def τ_leap(s, a, m, λs, λa, λm, μs, μa, μm, u, v, τ):
+
+    rates = get_rates(s, a, m, λs, λa, λm, μs, μa, μm, u, v)
+    try:
+        adj_rates = np.random.poisson(rates * τ)
+    except ValueError:
+        print(rates, τ, s, a, m)
+        raise
+    
+    Δs, Δa, Δm = updates.T @ adj_rates
+   
+    return τ, Δs, Δa, Δm
+
+def gillespie_ssa(λs, λa, λm, μs, μa, μm, u, v, s0, tmax, t0=0,  a0=0, m0=0, t_steps=1000):
+    
+    t = t0
+    s, a, m = s0, a0, m0
+    Δs, Δa, Δm = 0, 0, 0
+    
+    MT = []
+    # loop over recording times
+    while s + a + m > 0: 
+                    
+        Δt, Δs, Δa, Δm = τ_leap(s, a, m, λs, λa, λm, μs, μa, μm, u, v, τ)
+        t, s, a, m = t + Δt, max(s + Δs, 0), max(a + Δa, 0),  max(m + Δm, 0)  
+                    
+        if m > s0: 
+            MT.append(t)
+            break          
+                            
+    if len(MT) == 0: return 0
[…]

### 2025-02-04 src/Figure7/Figure7Simulation.py
Add files via upload

@@ -1,0 +1,104 @@
+import numpy as np
+import matplotlib.pyplot as plt
+import numba
+import seaborn as sns
+
+sns.set_context('talk')
+
+@numba.jit # 2-fold faster
+def get_rates(s, a, m, λs, λa, λm, μs, μa, μm, u, v):
+    return np.array([
+        λs * s,    
+        μs * s,     
+        u * s * λs,
+        v * s * λs,
+        λa * a, 
+        μa * a,
+        v * a * λa,
+        λm * m,    
+        μm * m,
+        0,
+    ])
+    
+updates = np.array([
+    [1, 0, 0],  # growth sensitive
+    [-1, 0, 0], # death sensitive
+    [-1, 1, 0], #mutation s->a
+    [-1, 0, 1], #mutation s->m
+    [0, 1, 0],  # growth aneuploidy type
+    [0, -1, 0],  # death aneuploidy type
+    [0, 0, 1], # #mutation a->m
+    [0, 0, 1],  # growth mutant type
+    [0, 0, -1],   # death mutant type
+    [0, 0, 0]
+])
+
+
+def τ_leap(s, a, m, λs, λa, λm, μs, μa, μm, u, v, τ):
+  
+    rates = get_rates(s, a, m, λs, λa, λm, μs, μa, μm, u, v)
+    try:
+        adj_rates = np.random.poisson(rates * τ)
+    except ValueError:
+        print(rates, τ, s, a, m)
+        raise
+
+    Δs, Δa, Δm = updates.T @ adj_rates
+   
+    return τ, Δs, Δa, Δm
+
+def gillespie_ssa(λs, λa, λm, μs, μa, μm, u, v, s0, tmax, t0=0,  a0=0, m0=0, t_steps=1000):
+    
+    t = t0
+    s, a, m = s0, a0, m0
+    Δs, Δa, Δm = 0, 0, 0
+    
+    while s + a + m > 0: 
+              
+        #print(n,a,m)      
+        Δt, Δs, Δa, Δm = τ_leap(s, a, m, λs, λa, λm, μs, μa, μm, u, v, τ)
+        t, s, a, m = t + Δt, max(s + Δs, 0), max(a + Δa, 0), max(m + Δm, 0)
+            
+        if m>=s0: break          
+                            
+    if m == 0: return 0
+    if m >= s0: return t
+
+
+λs, λa, λm, μs, μa, μm, u, v = 0.1, 0.0899, 0.1, 0.14, 0.09, 0.09, 10**(-2), 10**(-7)
+
[…]

### 2025-02-04 ms.tex
Add files via upload
Fixed Figure S6

@@
 \caption{The mean time for appearance of a resistance mutation that leads to evolutionary rescue with aneuploidy ($u>0$). 
-Solid lines show our approximations (\cref{eq:AsymptoticTimeRules}) compared to red markers that show mean of simulations results (with error bars for 95\% confidence intervalues obtained with bootstrap, see Appendix G).
+Solid lines show our approximations (green: \cref{limitapprox3}, blue: \cref{limitapprox4}, black: \cref{meantimet2}) compared to red markers that show mean of simulations results (with error bars for 95\% confidence intervalues obtained with bootstrap, see Appendix G).
 Blue dashed line, $N_a^*$.
@@
 \caption{The mean time for the mutant cell population to reach size $M$, where $M$ is the tumor detection size.
-Solid lines show our approximations (green: \cref{eq:t2det2} for $N>N_m^*$, blue:  \cref{meanproliferationtime2} with $\tau_a$ from \cref{eq:AsymptoticTimeRules} for $N<N_a^*$; black: \cref{meanproliferationtime2}) compared to red markers that show mean of simulations results (with error bars for 95\% confidence intervals obtained with bootstrap, see Appendix G).
+Solid lines show our approximations (green: \cref{eq:t2det2} for $N>N_m^*$, blue:  \cref{meanproliferationtime2} with $\tau_a$ from \cref{eq:AsymptoticTimeRules} for $N<N_a^*$; black: \cref{meanproliferationtime2} with $\tau_a$ from \cref{meantimet2}) compared to red markers that show mean of simulations results (with error bars for 95\% confidence intervals obtained with bootstrap, see Appendix G).
 Blue dashed line, $N_a^*$.

### 2025-02-05 ms.tex
final fix to app D

@@
 We distinguish between small, intermediate, or large tumors.
-In small ($N\ll N_a^*\ll N_m^*$) and intermediate ($N_a^* \ll N \ll N_m^*$) tumors, we have at most one lineage that rescues the cancer cell population.
+In small ($N\ll N_a^*\ll N_m^*$) tumors, we have at most one lineage that rescues the cancer cell population.
 As a result, the recurrence time is given by~\citep{avanzini2019cancer}
@@
 % change 35
-In both scenarios, plugging \cref{meantimet2} for $\tau_a$ in \cref{meanproliferationtime} provides a good approximation for the recurrence time (black line in \Cref{ProliferationTimeLarge}).
-For small tumors, we have a simpler approximation by plugging \cref{limitapprox4} for $\tau_a$ to get $\tau_a^r\approx1/|\Delta_a| + \log{\left(p_mN\right)}/\Delta_m$ (blue line in \Cref{ProliferationTimeLarge}).
-
-In a large tumor, $N_m^* \ll N$, the sensitive population produces a large number of rescue lineages in a short period of time.
+Therefore, in small tumors, we can substitue \cref{limitapprox4} for $\tau_a$ in \cref{meanproliferationtime} to get (blue line in \Cref{ProliferationTimeLarge})
+\begin{equation*}
+\tau_a^r\approx1/|\Delta_a| + \log{\left(p_mN\right)}/\Delta_m .
+\end{equation*}
+
+In intermediate tumors, $N_a^* << N << N_m^*$, the sensitive cell population generates a number of rescue lineages in a short period of time after $\tau_a$.
+The time it takes the rescue lineages to reach the initial population size $N$ is small compared to the time it took the lineages to appear.
+Therefore, the mean recurrence time can be approximated by substituting \cref{meantimet2} for $\tau_a$ in \cref{meanproliferationtime} (black line in \Cref{ProliferationTimeLarge}).
+
+In large tumors, $N_m^* \ll N$, the sensitive population produces a large number of rescue lineages in a short period of time.
 As a result, the recurrence time is obtained by solving the following system of ordinary differential equations (ODEs),

### 2025-02-05 ResponseLetter.tex
correct lines, typos

@@
 Thank you for this remark, you are correct to note that this statement is in opposition to some of our results.
-We have removed this statement and instead, we now write (line 523): ``The `stationary' scenario, in which the aneuploid growth rate is close to zero in the presence of drug, is particularly interesting.
+We have removed this statement and instead, we now write (line 527): ``The `stationary' scenario, in which the aneuploid growth rate is close to zero in the presence of drug, is particularly interesting.
 Although this may seem like a small region of parameter space in a general model of evolutionary rescue (Fig.
@@
 \emph{Notably, the latter indicates that aneuploidy, even when it only provides tolerance, increases the probability that the tumor will be rescued, as long as it is produced fast enough.}'' \\
-And in line 493: ``As an example, we have parameterized our model using estimates for melanoma and three triple-negative breast cancer (TNBC) cells under drug therapy from the literature. 
-We find that in these cases we are either in the first scenario (melanoma and one of the TNBC clones) in which aneuploidy provides at partial or full resistance to the drug, or the third scenario (two TNBC), in which aneuploidy provides tolerance to the drug.
+And in line 497: ``As an example, we have parameterized our model using estimates three triple-negative breast cancer (TNBC) clones under drug therapy from the literature. 
+We find that TNBC clones are either in the first scenario, in which aneuploidy provides at partial or full resistance to the drug, or the third scenario, in which aneuploidy provides tolerance to the drug.
 It remains to be seen which tumor type and drug combinations produce stationary aneuploidies.''\\
-The following sentences were already in the Discussion (line 476): ``In this scenario, aneuploidy provides two advantages.
+The following sentences were already in the Discussion (line 480): ``In this scenario, aneuploidy provides two advantages.
 First, it delays the extinction of the population, providing more time for the appearance of the resistance mutation.
@@
 \textcolor{blue}{%change 48
-We have rewritten the sentence about the mean detection time (line 453): ``We find that for small and intermediate-sized tumors the mean detection time is approximately equal to the mean recurrence time (i.e., $\tau_a^r\approx\tau_a^{M}$ for $N<N_m^*$).''
+We have rewritten the sentence about the mean detection time (line 455): ``We find that for small and intermediate-sized tumors the mean detection time is approximately equal to the mean recurrence time (i.e., $\tau_a^r\approx\tau_a^{M}$ for $N<N_m^*$).''
 }
@@
 \textcolor{blue}{%change 2
-We have added an explanation of ODE (line 381): ``In this case the dynamics of the number of mutant cells is deterministic and can therefore be modeled by a system of ordinary differential equations (ODE), which describe how the number of mutant cells changes over time by its time derivative (eq.~D2).''
+We have added an explanation of ODE (line 382): ``In this case the dynamics of the number of mutant cells is deterministic and can therefore be modeled by a system of ordinary differential equations (ODE), which describe how the number of mutant cells changes over time by its time derivative (eq.~D2).''
 }
@@
 Aneuploidy is indeed heritable.
-The extra-chromosome is lost at a rate similar to it's gain.
-A recent paper estimated the per generation loss rate in yeast at 8e-5-6e-4, depending on the specific chromosome (Hose et al.
+The extra-chromosome is lost at a rate similar to its gain.
+A recent paper estimated the per generation loss rate in yeast at $8\cdot10^{-5} - 6\cdot 10^{-4}$, depending on the specific chromosome (Hose et al.
 Genetics 2024).
@@
 As for the fitness difference between euploid and aneuploid mutants, this is a good question and the answer probably varies depending on the drug, the chromosome, and other factors.
-We now address both your questions in the Discussion, in line 542: ``Furthermore, loss of extra-chromosome may occur at a high rate (for estimates in yeast, see Hose et al.
+We now address both your questions in the Discussion, in line~545: ``Furthermore, loss of extra-chromosome may occur at a high rate (for estimates in yeast, see Hose et al.
 (2024)) and the fitness of a euploid cell with a resistance mutation may be higher than that of an aneuploid cell with the same mutation.
@@
 \textcolor{blue}{%change 5
-We revised the text to make it clear that we estimate $c$ from an experimental result and $f$ from $c$ (line 208): ``To estimate fitness cost of aneuploidy, $c$, we note that Lukow et al.
+We revised the text to make it clear that we estimate $c$ from an experimental result and $f$ from $c$ (line 209): ``To estimate the fitness cost of aneuploidy, $c$, we note that Lukow et al.
 (2021) mixed sensitive and aneuploid A375 melanoma cells at $1:1$ ratio, cultured them in a drug-free environment, and observed the ratio evolve as a function of time with the aneuploid cells declining to $15\%$ after 24 days.
@@
 
-\textcolor{blue}{We have now added density-dependence for the sensitives (Eq. at line 226). The results are unchanged.}
+\textcolor{blue}{We have now added density-dependence for the sensitives (line 226). The results are unchanged.}
 
-- methods: I got a 404 error for the source code link: https://github.com/yoavram-lab/EvolutionaryRescue
+- methods: I got a 404 error for the source code link: \url{https://github.com/yoavram-lab/EvolutionaryRescue}
 
@@
 \textcolor{blue}{ % change 33
-We now mention that eq. 2 (previously eq.3 ) is an alternative form of an eq. from Orr and Unckless 2008 (line 261).
+We now mention that eq. 2 (previously eq. 3) is an alternative form of an equation from Orr and Unckless 2008 (line 261).
 }
@@
 \textcolor{blue}{%change 51
-We have added the recommended comment (line 303): ``Interestingly, increasing both the aneuploid division rate, $\lambda_a$, and the aneuploid death rate, $\mu_a$, such that the growth rate $\Delta_a$ remains constant, leads to a decrease in $T^*$, pushing the system to the second scenario.
+We have added the recommended comment (line 304): ``Interestingly, increasing both the aneuploid division rate, $\lambda_a$, and the aneuploid death rate, $\mu_a$, such that the growth rate $\Delta_a$ remains constant, leads to a decrease in $T^*$, pushing the system to the second scenario.
 This is because increasing $\lambda_a$ causes a decrease in $T^*$ as  it increases the effective mutation rate $v\lambda_a$ (as mutations mostly occur during division) and a lineage does not have to survive as long in order to generate a successful mutant.''
@@
 \textcolor{blue}{ % change 58
-We believe that our model does fit with data on resistance rates, and the text now reads (line 316): ``Interestingly, the threshold between small and intermediate tumors, $N_a^*$, is similar to the tumor detection threshold of $4.19 \times 10^6$ cells for a wide variety of tumors~\citep{avanzini2019cancer}.
+We believe that our model does fit with data on resistance rates, and the text now reads (line 317): ``Interestingly, the threshold between small and intermediate tumors, $N_a^*$, is similar to the tumor detection threshold of $4.19 \times 10^6$ cells for a wide variety of tumors~\citep{avanzini2019cancer}.
 We note that vemurafenib-treated melanomas (i.e.
@@
 \textcolor{blue}{%change 46
-We have added the following text in line 393: ``For small tumors ($ N \ll N_a^*$), the mean rescue time is the two-step equivalent of the one-step result from Orr and Unckleass (2014; expectation of eq.~18).
+We have added the following text in line 395: ``For small tumors ($ N \ll N_a^*$), the mean rescue time is the two-step equivalent of the one-step result from Orr and Unckleass (2014; expectation of eq.~18).
 The mean rescue time (conditioned on rescue occuring) is a decreasing function of the sensitive and aneuploid growth rates and independent of the other model parameters, including tumor size (blue line in Figure S6).
@@
 \textcolor{blue}{%change 26
-We have added the definition of $p_m$ in line 421: ``where $p_m$ is the probability that a lineage starting from a single mutant cell escapes stochastic extinction.''
+We have added the definition of $p_m$ in line 425: ``where $p_m$ is the probability that a lineage starting from a single mutant cell escapes stochastic extinction.''
 }
@@
 \textcolor{blue}{
-The return time with new mutation (eqn 23) from Orr $\&$ Unckless 2014 can be written as:
+The return time with new mutation from Orr $\&$ Unckless 2014 (eq. 23) can be written as:
 \begin{align*}
@@
 \textcolor{blue}{% change 49
-The text now reads (line 435): ``Drugs that significantly increase the sensitive death rate, $\mu_s$, but do not affect the sensitive division rate, $\lambda_s$, delay cancer recurrence.
+The text now reads (line 437): ``Drugs that significantly increase the sensitive death rate, $\mu_s$, but do not affect the sensitive division rate, $\lambda_s$, delay cancer recurrence.
 Drugs that decrease the sensitive division rate delay cancer recurrence time even more, because they effectively decrease the mutation rate (assuming mutations occur during division).''
@@
 Thank you for this suggestions.
-We have revised the text (line 476): ``...
+We have revised the text (line 480): ``...
 aneuploidy provides two advantages.
@@
 
-\textcolor{blue}{We have added a paragraph in the Discussion (line 520) on the connections to Iwasa et al 2003 and Osmond et al 2020, which we think are the two closest parallels for the main results.
+\textcolor{blue}{We have added a paragraph in the Discussion (line 524) on the connections to Iwasa et al 2003 and Osmond et al 2020, which we think are the two closest parallels for the main results.
 For the other more detailed expressions, we have added the references in the Results as they come up.}
@@
 \textcolor{blue}{Following the reviewer's suggestion, we have evaluated our model using parameters from breast cancer (PMID: 34163070), see Methods (line 180), Table 2, and Figure 3B.\\
-We have aldo added the following text to the Discussion (line 493): ``As an example, we have parameterized our model using estimates for melanoma and three triple-negative breast cancer (TNBC) cells under drug therapy from the literature.
-We find that in these cases we are either in the first scenario (melanoma and one of the TNBC clones), in which aneuploidy provides at partial or full resistance to the drug, or the third scenario (two TNBC), in which aneuploidy provides tolerance to the drug.
-It remains to be seen which tumor type and drug combinations produce stationary aneuploidies. % change B
-Comparing the probability of evolutionary rescue between different tumors (Figure 3B) suggests that some TNBC clones have a much higher probability of relapsing compared to melanoma and other TNBC clones, because smaller tumors have higher resuce probability.
-Notably, the probability of relapse in TNBC patients is higher than 50\% in the first 3-5 years after diagnosis \citep{taushanova2023synchronous}, whereas in melanoma patients it is approximately 10-20\% in the first 5 years \citep{wan2022prediction,von_schuckmann2019risk}.'' % change A
+We have aldo added the following text to the Discussion (line 497): ``As an example, we have parameterized our model using estimates from three triple-negative breast cancer (TNBC) clones under drug therapy.
+We find that TNBC clones are either in the first scenario, in which aneuploidy provides at partial or full resistance to the drug, or the third scenario, in which aneuploidy provides tolerance to the drug.
+Comparing the probability of evolutionary rescue between different tumors (Figure 3B) suggests that some TNBC clones have a much higher probability of relapsing compared to melanoma and other TNBC clones.
+Notably, the probability of relapse in TNBC patients is higher than 50\% in the first 3-5 years after diagnosis \citep{taushanova2023synchronous}, whereas in melanoma patients it is approximately 10-20\% in the first 5 years \citep{wan2022prediction,von_schuckmann2019risk}."
 }% change A
@@
 We have removed the misleading sentence ``aneuploidy will either not play any role in evolutionary rescue or will be the main driver of adaptation'', which is incorrect.
-Instead, we write (line 526): ``Our results suggest that identifying aneuploidies that are tolerant or stationary may be worthwhile, as these are also expected to enhance the probability of rescue (Figure 3) and extend the window of opportunity for rescue (Figure 6).''\\
+Instead, we write (line 530): ``Our results suggest that identifying aneuploidies that are tolerant or stationary may be worthwhile, as these are also expected to enhance the probability of rescue (Figure 3) and extend the window of opportunity for rescue (Figure 6).''\\
 Furthermore, in our model, we are interested in a specific type of aneuploidy which increases the fitness of cancer cells in an environment affected by anti-cancer drugs.
@@
 
+\textcolor{blue}{The GitHub repository will become available when the paper is accepted and will include a README file that explains the repository structure.
+Meanwhile, the source code for our analysis can be privately accessed here: \url{https://www.dropbox.com/scl/fi/hrvqdocsr67qur7k59wxw/StanaEtAlCode.zip?rlkey=n7bcr3usx5ym6q3s2qkotanhx}
+}
+
 - In your model, mutations appear to only confer resistance, however, most mutations are deleterious.
@@
 We assume deleterious mutations can occur with equal rates in all genotypes and therefore would not change the dynamics.
-We now explain these assumptions in line 215: ``Note that when we refer to drug-sensitive cells, we include those cells that have any aneuploidy that does not affect fitness in the presence of the drug.
+We now explain these assumptions in line 216: ``Note that when we refer to drug-sensitive cells, we include those cells that have any aneuploidy that does not affect fitness in the presence of the drug.
 Additionally, we focus on mutations that confer resistance, neglecting deleterious mutations (which are more common).
@@
 \textcolor{blue}{
-We agree with all of these points. We have revised the introduction and discussion accordingly, see lines 55-87, 520-529.
+We agree with all of these points. We have revised the introduction and discussion accordingly, see lines 55-87, 524-532.
 }
@@
 
-\textcolor{blue}{We changed $\Delta$ to $r$. We therefore also changed $r_1$ to $n_1$ ($r_1$ only appeared in the appendices).}
+\textcolor{blue}{We changed Delta to r. We therefore also changed $r_1$ to $n_1$ ($r_1$ only appeared in the appendices).}
 
@@
 
-\textcolor{blue}{We have added a short explanation for ``tolerant'' and that it has other meanings in the literature.
-We also mention the equivalence to in ``sub/super/critical'' from Osmond et al 2020.
+\textcolor{blue}{We have added a short explanation for ``tolerant'' and that it has other meanings in the literature (line 90).
+We also mention the equivalence to in ``sub/super/critical'' from Osmond et al 2020 (line 270).
 We use these terms because we are trying to translate concepts familiar to the evolutionary rescue community into ones more familiar to cancer researchers.} % change 47, 57

### 2025-02-05 ms.tex
correct lines, typos

@@
 %The fitness cost $c$ of aneuploidy before the onset of therapy is difficult to estimate as we are interested in a specific type of aneuploidy that improves the fitness of cancer cells in an environment altered by drugs.
-To estimate fitness cost of aneuploidy, $c$, we note that \citet{lukow2021chromosomal} mixed sensitive and aneuploid A375 melanoma cells at $1:1$ ratio, cultured them in a drug-free environment, and observed the ratio evolve as a function of time with the aneuploid cells declining to $15\%$ after 24 days.
+To estimate the fitness cost of aneuploidy, $c$, we note that \citet{lukow2021chromosomal} mixed sensitive and aneuploid A375 melanoma cells at $1:1$ ratio, cultured them in a drug-free environment, and observed the ratio evolve as a function of time with the aneuploid cells declining to $15\%$ after 24 days.
 Thus, our estimate for the fitness cost is $c=\abs{\log\left(0.15/(1-0.15)\right)/24}\approx0.07$ per day \citep{chevin2011measuring}.
@@
 These approximations are accurate when compared to results of stochastic evolutionary simulations (\Cref{rescue_prob,rescue_threshold}). 
-Our parameterization for two types of cancers suggests that melanoma and TNBC-SA609 clone A are resistant to the drug ($\Delta_a T^*>500$), whereas TNBC-SA1035 clone H and TNBC-SA535 clone H are tolerant ($\Delta_a T^*<-1000$).
+Our parameterization for triple-negative breast cancer clones suggests that TNBC-SA609 clone A is resistant to the drug ($\Delta_a T^*>500$), whereas TNBC-SA1035 clone H and TNBC-SA535 clone H are tolerant ($\Delta_a T^*<-1000$).
+
 In the first scenario, the treatment effectively kills aneuploid cells but not as quickly as it kills sensitive cells. 
@@
 In the third scenario, aneuploid cells are sufficiently resistant for the population to re-grow the tumor without additional resistance mutations.
-
 Notably, in this scenario the mutant parameters ($v$, $\lambda_m$, and $\Delta_m$) do not affect $N_a^*$ beyond their effect on $T^*$.
@@
 Furthermore, the growth rate $\Delta_a<0$ that allows tolerant aneuploids to rescue the tumor is between $-u\lambda_a$ and  $-1/T^*$, which is proportional to $-\sqrt{v}$.
-Thus, increasing the mutation rate $v$ will decrease the tumor threshold size $N^*_a$, making evolutionary rescue more likely, but only until $T^*$ decreases to a point where  $\Delta_a =1/T^*$.
+Thus, increasing the mutation rate $v$ will decrease the tumor threshold size $N^*_a$, making evolutionary rescue more likely.
 %change 27
@@
 
-As an example, we have parameterized our model using estimates for melanoma and three triple-negative breast cancer (TNBC) cells under drug therapy from the literature. 
-We find that in these cases we are either in the first scenario (melanoma and one of the TNBC clones), in which aneuploidy provides at partial or full resistance to the drug, or the third scenario (two TNBC clones), in which aneuploidy provides tolerance to the drug.
+As an example, we have parameterized our model using estimates from three triple-negative breast cancer (TNBC) clones under drug therapy.
+We find that TNBC clones are either in the first scenario, in which aneuploidy provides at partial or full resistance to the drug, or the third scenario, in which aneuploidy provides tolerance to the drug.
 It remains to be seen which tumor type and drug combinations produce stationary aneuploidies. % change B

### 2025-02-27 ms.tex
ack
edited evo rescue refs

@@
 \citet{martin2013probability} extended these results to consider the effects of a range of rescue genotypes that can vary in both growth rate and variance in offspring number.
-\citet{uecker2016role} modeled a situation in which \textit{two} mutations at different loci (with recombination between them) are necessary for rescue. 
+Importantly for the present work, they also argued that two-step processes involving a ``transient'' intermediate genotype with growth rate close to zero may be an important form of rescue;
+however, they only calculated that the rate of such rescues should scale with the total number of such transient genotypes produced.
+%\citet{uecker2016role} modeled a situation in which \textit{two} mutations at different loci (with recombination between them) are necessary for rescue. 
 \citet{osmond_genetic_2020} considered a population that must achieve rescue, potentially via multiple mutations, by adapting on a fitness landscape described by Fisher's geometric model,
@@
 \section*{Acknowledgements}
-We thank Hildegard Uecker for discussions and comments. 
+We thank Hildegard Uecker for discussions and the editor and three reviewers for comments on the manuscript.
 This work was supported in part by

### 2025-03-09 ms.tex
Add files via upload

@@
 One evolutionary mechanism by which a population of cancer cells can adapt to chemotherapy is aneuploidy. 
-Aneuploid cancer cells can be more fit in an environment altered by anti-cancer drugs, e.g., because aneuploidy disrupts the pathways usually targeted by the drugs. %change 3 
+Aneuploid cancer cells can be more fit in an environment altered by anti-cancer drugs, because aneuploidy disrupts the pathways usually targeted by the drugs. %change 3 %change S2
 Indeed, aneuploidy is highly prevalent in tumors, and some anti-cancer drugs fight cancer by increasing chromosomal instability.
@@
 We use multi-type branching processes to approximate the probability that a tumor survives drug treatment as a function of the initial tumor size, the rates at which aneuploidy and other beneficial mutations occur, and the growth rates of the drug-sensitive and drug-resistant cells.
-Additionally, we investigate the effect of the pre-existent aneuploid cells on the probability of evolutionary rescue.
+We also investigate the effect of the pre-existent aneuploid cells on the probability of evolutionary rescue.
+%change S2
  Finally, we estimate the tumor's mean recurrence time to revert to its initial size following treatment and evolutionary rescue.
@@
 One suggested factor is aneuploidy, in which cells are characterized by an imbalanced karyotype and alterations in the number of chromosomes \citep{schukken2018cin}. 
-Aneuploidy is caused by chromosomal instability and mis-segregation of chromosomes during mitosis.
+Aneuploidy is caused by chromosomal instability and missegregation of chromosomes during mitosis. % change S4
 Importantly, changes in the number of chromosomes and chromosome arm copies allow cancer cells to survive under stressful conditions such as drug therapy \citep{lukow2021chromosomal,rutledge2016selective,ippolito2021gene}.
@@
 An essential aspect of aneuploidy is that the rate with which cells become aneuploid, that is, the rate of chromosome missegregation, is several orders of magnitude higher than mutation rate \citep{bakker2023predicting}.
-Consequently, a cell exposed to stress, such as chemotherapeutic drugs, can acquire aneuploidy faster than a mutation.
-Futhermore, some proposed anti-cancer drugs elevate the missegregation rate to fight cancer cells~\citep{lee2016effects}, as an extremely high chromosome missegregation rate is incompatible with cell survival and proliferation.
-% change x1
+A cell exposed to stress, such as chemotherapeutic drugs, can acquire aneuploidy faster than a mutation.
+%change S16
+Some proposed anti-cancer drugs elevate the missegregation rate to fight cancer cells~\citep{lee2016effects}, as an extremely high chromosome missegregation rate is incompatible with cell survival and proliferation.
+% change x1 %change S16
 
@@
 adaptation by phenotypic plasticity without genetic modification \citep{carja2019evolutionary,carja2017evolutionary,levien2021non,gunnarsson2020understanding}; 
-and adaptation through genetic modifications, e.g., mutation \citep{gomulkiewicz1995does,uecker2014evolutionary,uecker2016role,uecker2011fixation,orr2014population}. 
+and adaptation through genetic modifications, like mutation \citep{gomulkiewicz1995does,uecker2014evolutionary,uecker2016role,uecker2011fixation,orr2014population}. %change S5
 Here, we focus on the latter.
@@
 \citet{martin2013probability} extended these results to consider the effects of a range of rescue genotypes that can vary in both growth rate and variance in offspring number. 
-Importantly for the present work, they also argued that two-step processes involving a ``transient'' intermediate genotype with growth rate close to zero may be an important form of rescue;
-however, they only calculated that the rate of such rescues should scale with the total number of such transient genotypes produced.
+Importantly for the present work, they also argued that two-step processes involving a ``transient'' intermediate genotype with growth rate close to zero may be an important form of rescue; however, they only calculated that the rate of such rescues should scale with the total number of such transient genotypes produced.
+%change S19
 %\citet{uecker2016role} modeled a situation in which \textit{two} mutations at different loci (with recombination between them) are necessary for rescue. 
@@
 \citet{iwasa2004evolutionary} used a similar approach to find simple approximate expressions for the probability that a single lineage would evolve to survive via a wide range of possible mutational pathways.
-A very similar set of models describe adaptation to new environments (e.g., a pathogen evolving to cross a species barrier) and fitness valley crossing \citep{antia2003role, weissman2009rate}.
+A similar set of models describe adaptation to new environments (e.g., a pathogen evolving to cross a species barrier) and fitness valley crossing \citep{antia2003role, weissman2009rate}. % change S1
 \citet{gunnarsson2020understanding} analyzed a model where a tumor consisting of two populations of cancer cells, one drug-resistant and the other drug-sensitive, can evade extinction by cells switching between the two phenotypes through epigenetic mutations.
@@
 We assume the population of cells is under a strong stress, such as drug therapy, and therefore $\Delta_s<0$, whereas the mutant is resistant to the stress, $\Delta_m>0$.
-We consider a range of possible values for $\Delta_a$, finding three distinct scenarios: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_s$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing, stationary, or growing or dying only very slowly, that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$, in a sense that we will make precise below.
+We consider a range of possible values for $\Delta_a$, finding three distinct scenarios: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_s$ \citep[tolerant cells are able to survive the effects of the drug for an extended period of time compared to susceptible cells while resistant cells require a higher drug concentration to be affected, see][for a detailed explanation of the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing, stationary, or growing or dying slowly, that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$, in a sense that we will make precise below.
+% change S1 %change S17
 
@@
 To model standing genetic variation, we assume that before the onset of therapy, sensitive cells become aneuploid with rate $\tilde{u}\lambda_s$ (which may differ from $u \lambda_s$) and that aneuploidy confers a fitness cost $c$ in the drug-free environment, that is, we assume that aneuploid cells have an increased death rate compared to sensitive cells in a drug-free environment.
-We assume that $c$ is not too small (\Cref{table1}), so that standing variation is generated by a balance between the generation of anuploids and selection against them, without the effect of genetic drift.
+We assume that $c$ is not too small (\Cref{table1}), so that standing variation is generated by a balance between the generation of aneuploids and selection against them, without the effect of genetic drift.
+%change RS3
 
@@
 \begin{aligned}
-(+1,0,0)&:\quad \lambda_ss_t\left(1-u-v\right)\quad\left(\text{birth of sensitive cell}\right),\\
-(-1,0,0)&:\quad \mu_ss_t\quad\left(\text{death of sensitive cell}\right),\\
-(0,+1,0)&:\quad u\lambda_ss_t\quad\left(\text{sensitive cell divides and becomes aneuploid}\right),\\
-(0,0,+1)&:\quad v\lambda_ss_t\quad\left(\text{sensitive cell divides and becomes mutant}\right),\\
-(0,+1,0)&:\quad \lambda_aa_t\left(1-v\right)\quad\left(\text{birth of aneuploid cell}\right),\\
-(0,-1,0)&:\quad \mu_aa_t\quad\left(\text{death of aneuploid cell}\right),\\
-(0,0,+1)&:\quad v\lambda_aa_t\quad\left(\text{aneuploid cell divides and becomes mutant}\right),\\
-(0,0,+1)&:\quad \lambda_mm_t\quad\left(\text{birth of mutant cell}\right),\\
-(0,0,-1)&:\quad \mu_mm_t\quad\left(\text{death of mutant cell}\right).
+(+1,0,0)&:\quad \lambda_ss_t\left(1-u-v\right)&&\left(\text{birth of sensitive cell}\right),\\
+(-1,0,0)&:\quad \mu_ss_t&&\left(\text{death of sensitive cell}\right),\\
+(0,+1,0)&:\quad u\lambda_ss_t&&\left(\text{sensitive cell divides and becomes aneuploid}\right),\\
+(0,0,+1)&:\quad v\lambda_ss_t&&\left(\text{sensitive cell divides and becomes mutant}\right),\\
+(0,+1,0)&:\quad \lambda_aa_t\left(1-v\right)&&\left(\text{birth of aneuploid cell}\right),\\
+(0,-1,0)&:\quad \mu_aa_t&&\left(\text{death of aneuploid cell}\right),\\
+(0,0,+1)&:\quad v\lambda_aa_t&&\left(\text{aneuploid cell divides and becomes mutant}\right),\\
+(0,0,+1)&:\quad \lambda_mm_t&&\left(\text{birth of mutant cell}\right),\\
+(0,0,-1)&:\quad \mu_mm_t&&\left(\text{death of mutant cell}\right).
 \end{aligned}
-\]
+\]% change S6
 
@@
 
-Note that when we refer to drug-sensitive cells, we include those cells that have any aneuploidy that does not affect fitness in the presence of the drug.
+Note that when we refer to drug-sensitive cells, we include those cells that have any aneuploidy that does not affect fitness in the presence of the drug (see \Cref{table1}). %change S7
 Additionally, we focus on mutations that confer resistance, neglecting deleterious mutations (which are more common).
-We assume deleterious mutations and other aneuplidies occur at similar rates across genotypes and therefore neglect their effect on the dynamics.
-
-All the parameters discussed above are shown in \Cref{table1}.
+We assume deleterious mutations and other aneuploidies occur at similar rates across genotypes and therefore neglect their effect on the dynamics.
+%change RS4
 
@@
 Define $p_s$ as the probability that a lineage starting from a single drug-sensitive cell avoids extinction by acquiring drug resistance.
-Thus, $N^*=1/p_s$ is the threshold tumor size above which evolutionary rescue is very likely \citep{iwasa2003evolutionary}, and the rescue probability is given by 
+Thus, $N^*=1/p_s$ is the threshold tumor size above which evolutionary rescue is likely \citep{iwasa2003evolutionary}, and the rescue probability is given by % change S1
 \begin{equation}\label{eq:rescue_prob} 
@@
 In these approximations, an important quantity is $T^* = \sqrt{\lambda_m / 4 v \lambda_a^2 \Delta_m}$, which is the critical time an aneuploid lineage needs to survive to produce a resistant mutant that avoids random extinction.
-First, if aneuploidy is very rare ($u\lambda_a T^*< 1$), or if aneuploidy is rare ($u\lambda_a < -\Delta_a$) and very sensitive to the drug ($\Delta_a T^* < -1$), then it is likely that evolutionary rescue will occur through a direct resistance mutation in a sensitive cell without aneuploidy playing a role in the adaptive dynamics, such that
+First, if aneuploidy is sufficiently rare ($u\lambda_a T^*< 1$), or if aneuploidy is rare ($u\lambda_a < -\Delta_a$) and sensitive to the drug ($\Delta_a T^* < -1$), then it is likely that evolutionary rescue will occur through a direct resistance mutation in a sensitive cell without aneuploidy playing a role in the adaptive dynamics, such that % change S1
 \begin{equation} \label{eq:N_m}
@@
 \end{equation}
-Imporantly, when this threshold is smaller than 1, aneuploidy is expected to play a role in evolutionary rescue, as it resuced the threshold tumor size for rescue. 
+Imporantly, when this threshold is smaller than 1, aneuploidy is expected to play a role in evolutionary rescue, as it reduces the threshold tumor size neeede for rescue. %change S8
 As expected, this ratio increases with the mutation rate $v$ and decreases with the aneuploidy rate $u$.
@@
 However, growth may be limited by resources (oxygen, nutrients, etc.) and therefore depend on cell density.
-Therefore, we performed stochastic simulations of a logistic growth model with a carrying capacity. 
+ We performed stochastic simulations of a logistic growth model with a carrying capacity. %change S16
 We find that our density-independent approximations agree with the results of simulations with density-dependent growth for biologically relevant parameter values (\Cref{LogisticPlot}).
@@
 
-In the above, we assumed that at the onset of drug treatment, the initial tumor consisted entirely of drug-sensitive cells.
-However, aneuploidy is likely produced even before the onset of treatment at some rate $\tilde{u}$, which may be lower in the absence of drugs, $\tilde{u} < u$ \citep{wang2019molecular,mason2017functional}.
-Moreover, aneuploidy likely confers a fitness cost $c$ in the absence of drugs \citep{replogle2020aneuploidy,giam2015aneuploidy}.
-Hence, if the number of cells in the tumor $N$ is large (as expected if the tumor is treated with a drug), there may already be a fraction $f \approx \tilde{u}\lambda_s/c$ of aneuploid cells in the population (here we assume that the drug affects the sensitive death rate but not the division rate and therefore we use $\lambda_s$ for the sensitive division rate in the drug-free environment).
-
-Therefore, the threshold tumor size for rescue by standing generation variation, $\tilde{N}^*_{a}$, is similar to the threshold for rescue by \textit{de novo} variation, $N^*_a$, except that the sensitive growth rate $\abs{\Delta_s}$ is replaced by the cost of aneuploidy $c$, such that
+We assumed that at the onset of drug treatment the initial tumor consisted entirely of drug-sensitive cells but aneuploidy is likely produced even before the onset of treatment at some rate $\tilde{u}$, which may be lower in the absence of drugs, $\tilde{u} < u$ \citep{wang2019molecular,mason2017functional}.
+%change S9, S19
+ Aneuploidy likely confers a fitness cost $c$ in the absence of drugs \citep{replogle2020aneuploidy,giam2015aneuploidy}. %change S16, S19 
+If the number of cells in the tumor $N$ is large (as expected if the tumor is treated with a drug), there may already be a fraction $f \approx \tilde{u}\lambda_s/c$ of aneuploid cells in the population (here we assume that the drug affects the sensitive death rate but not the division rate and therefore we use $\lambda_s$ for the sensitive division rate in the drug-free environment).
+%change S16, S19
+
+Therefore, the threshold tumor size for rescue by standing genetic variation, $\tilde{N}^*_{a}$, is similar to the threshold for rescue by \textit{de novo} variation, $N^*_a$, except that the sensitive growth rate $\abs{\Delta_s}$ is replaced by the cost of aneuploidy $c$, such that %change S10
 \begin{equation}\label{eq:Nsgv}
@@
 Comparing this approximation of $\tilde{N}^*_{a}/N^*_{a}$ to results of stochastic simulations, we find that the approximations are accurate (\Cref{rescue_denovo}). 
-Standing genetic variation will drive evolutionary rescue if sensitive cells die rapidly (growth rate $\Delta_s$ is very negative) due to a strong effect of the drug on sensitive cells or if the cost of aneuploidy in the drug-free environment, $c$, is small.
+Standing genetic variation will drive evolutionary rescue if sensitive cells die rapidly (growth rate $\Delta_s$ is negative) due to a strong effect of the drug on sensitive cells or if the cost of aneuploidy in the drug-free environment, $c$, is small.
 In contrast, \textit{de novo} aneuploid cells will have a greater contribution to rescue if the cost of aneuploidy, $c$, is large, the effect of the drug on sensitive cells is weak ($\Delta_s$ is close to zero), or if the drug induces the appearance of aneuploid cells ($u > \tilde u$).
+% change S1
 For example, with  $\lambda_s=0.1$, $\mu_s=0.14$, $u=10^{-2}$, $\tilde{u}=10^{-3}$, and $c=0.07$, the ratio of the threshold tumor sizes for standing vs.
@@
 \end{equation}
-Evolutionary rescue through direct mutation is more likely if the cost of aneuploidy, $c$, is very large or the effect of the drug $\Delta_s$ is small.
-In contrast, standing genetic variation will drive adaptation if the pre-treatment chromosome missegregation rate, $\tilde{u}$, is very large.
+Evolutionary rescue through direct mutation is more likely if the cost of aneuploidy, $c$, is large or the effect of the drug $\Delta_s$ is small.
+In contrast, standing genetic variation will drive adaptation if the pre-treatment chromosome missegregation rate, $\tilde{u}$, is large.
+% change S1
 The ratio does not depend on the rate of chromosome missegregation induced by the drug, $u$.
@@
 
-When evolutionary rescue occurs, the time until the tumor recurs may still be very long.
+When evolutionary rescue occurs, the time until the tumor recurs may still be long.
 We therefore explored the time until the tumor recurs, that is, the time until the tumor reaches its original size, $N$.
+% change S1
 When the expected number of resistant lineages that avoid extinction is small, the expected recurrence time can be estimated by adding two terms: the \emph{mean evolutionary rescue time}, which is the waiting time for the appearance of a resistant lineage that avoids extinction (conditioned on such an event occurring in the first place), and the \emph{mean proliferation time}, which is the expected time for that lineage to grow to $N$ cells.
@@
 %change 2
-Of particular interest is the distribution of the evolutionary rescue time and recurrence time with tolerant aneuploid cells ($\Delta_a T^*\ll1$), for which we focus on the parameter values in \Cref{table1} with $\lambda_a=0.0899$ (tolerant aneuploids), $\mu_s=0.14$ (intermediate sensitive death rate), and $v=10^{-7}$ (high end of the mutation rate).
+Of particular interest is the distribution of the evolutionary rescue time and recurrence time with tolerant aneuploid cells ($\Delta_a T^*\ll1$) given that this represents the two step evolutionary rescue scenario.
+We focus on the parameter values in \Cref{table1} with $\lambda_a=0.0899$ (tolerant aneuploids), $\mu_s=0.14$ (intermediate sensitive death rate), and $v=10^{-7}$ (high end of the mutation rate).
+%change S11
 
@@
 Increasing the per division mutation rate, $v$, leads to the faster appearance of a rescue mutations and hence reduced mean rescue time.
-Finally, increasing the tumor size leads to shorter mean rescue time, as more sensitive cells can mutate to become resistant.
-% change 46
+Increasing the tumor size leads to shorter mean rescue time, as more sensitive cells can mutate to become resistant.
+% change 46 change S16
 
@@
 It grows logarithmically with tumor size $N$ and may be the same order of magnitude as the mean evolutionary rescue time. 
-Increasing the mutant growth rate, $\Delta_m$, decreases the recurrence time, while increasing the sensitive and aneuploid growth rates, $\Delta_s$ and $\Delta_a$, respectively, increases the recurrence time.
+Increasing the sensitive $\Delta_s$, aneuploid $\Delta_a$ and mutant $\Delta_m$ growth rates decreases the recurrence time.%, while increasing the sensitive and aneuploid growth rates, $\Delta_s$ and $\Delta_a$, respectively, increases the recurrence time.
+%change RS2
 
@@
 %change 49
-In addition, drugs that significantly increase the division rate of sensitive cells, $\lambda_s$, but do not affect their death rate, $\mu_s$, decrease cancer recurrence times (conditioned on evolutionary rescue).
+Drugs that significantly increase the division rate of sensitive cells, $\lambda_s$, but do not affect their death rate, $\mu_s$, decrease cancer recurrence times (conditioned on evolutionary rescue).
+%change S12
 
@@
 %change 48
-However, for large tumors, the mean detection time $\tau_a^{M}$ decreases logarithmically with tumor size $N$, while the recurrence time $\tau_a^{r}$ is constant (\Cref{RecurrencePlot}).
-Additionally, for large tumors, we have $M<N_m^*<N$, so the mean detection time is shorter compared to the mean recurrence time, that is, the resistant tumor may be detected before recovering back to its initial size.
+For large tumors the mean detection time $\tau_a^{M}$ decreases logarithmically with tumor size $N$, while the recurrence time $\tau_a^{r}$ is constant (\Cref{RecurrencePlot}). %change S16
+ For large tumors we also have $M<N_m^*<N$, so the mean detection time is shorter compared to the mean recurrence time, that is, the resistant tumor may be detected before recovering back to its initial size.
+%change S16
 
@@
 When aneuploidy only provides tolerance to the drug ($\Delta_s\ll\Delta_a\ll0\ll\Delta_m$), it cannot rescue the population.
-Instead, if the aneuploidy rate is fast enough (\cref{eq:N_ratio}) then it may act as a ``stepping stone'' through which the resistant mutant can appear more rapidly, given that the number of aneuploid cells declines slower than the number of drug-sensitive cells (\Cref{sampleTrajectories}).
-In this scenario, aneuploidy provides two advantages. First, it delays the extinction of the population, providing more time for the appearance of the resistance mutation. 
+Instead, if the aneuploidy rate is fast enough (\cref{eq:N_ratio}) then it may act as a ``stepping stone'' through which the resistant mutant can appear more easily, given that the number of aneuploid cells declines slower than the number of drug-sensitive cells (\Cref{sampleTrajectories}).
+%change RS1
+In this scenario, aneuploidy provides two advantages. First, it delays the extinction of the population, providing more time for the appearance of a resistance mutation. %change S13
 Second, it increases the population size relative to a drug-sensitive population, providing more cells in which mutations can occur. 
@@
 Therefore, large tumors are likely to be rescued with or without aneuploidy.
-However, anti-cancer drugs are often used as adjuvant therapy after resection, in which case the number of cells in the tumor may be below the detection threshold of $\sim10^7$  \citep{bozic2013evolutionary}.
+ Anti-cancer drugs are often used as adjuvant therapy after resection, in which case the number of cells in the tumor may be below the detection threshold of $\sim10^7$  \citep{bozic2013evolutionary}.
+%change S16
 In these cases, aneuploidy can have a crucial role in the evolutionary rescue of the tumor and, therefore, in cancer recurrence.
@@
 As an example, we have parameterized our model using estimates from three triple-negative breast cancer (TNBC) clones under drug therapy.
-We find that TNBC clones are either in the first scenario, in which aneuploidy provides at partial or full resistance to the drug, or the third scenario, in which aneuploidy provides tolerance to the drug.
+We find that TNBC clones are either in the first scenario, in which aneuploidy provides a partial or full resistance to the drug, or the third scenario, in which aneuploidy provides tolerance to the drug.
+%change S14
 It remains to be seen which tumor type and drug combinations produce stationary aneuploidies. % change B
@@
 We found that a drug that reduces the sensitive growth rate and does not significantly increase the chromosome missegregation rate will likely lead to evolutionary rescue through standing genetic variation (\Cref{rescue_denovo,eq:Nsgv}).
-Furthermore, if the fraction of tumor cells that have the beneficial aneuploidy is $f \gg u \lambda_s / \abs{\Delta_s} \approx2.5\%$, then evolutionary rescue is more likely to occur via standing variation rather than through \textit{de novo} aneuploidy.
-However, for the parameter values we focus on in our examples (\Cref{table1}), this fraction is an order of magnitude lower, and therefore, we expect evolutionary rescue to occur by \textit{de novo} aneuploidy.
+ If the fraction of tumor cells that have the beneficial aneuploidy is $f \gg u \lambda_s / \abs{\Delta_s} \approx2.5\%$, then evolutionary rescue is more likely to occur via standing variation rather than through \textit{de novo} aneuploidy.
+%change S16
+ For the parameter values we focus on in our examples (\Cref{table1}), this fraction is an order of magnitude lower, and therefore, we expect evolutionary rescue to occur primarily by \textit{de novo} aneuploidy.
+%change S15 change S16
 
@@
 We have assumed that cancer cell lineages are independent and have verified that this is accurate under simple logistic growth.
-However, this assumption neglects the potential effects of spatial structure and local interactions, which may be important in solid tumors.
+This assumption neglects the potential effects of spatial structure and local interactions, which may be important in solid tumors.
+%change S16
 Such tumors can be spatially heterogeneous, with different genotypes inhabiting cellular niches and immune infiltration impacting growth in affected regions \citep{varrone2023cellcharter,galon2010immune}.
 This can potentially impact the probability of evolutionary rescue \citep{martens2011spatial}.
-Furthermore, loss of extra-chromosome may occur at a high rate (for estimates in yeast, see \citet{hose_rate_2024}) and the fitness of a euploid cell with a resistance mutation may be higher than that of an aneuploid cell with the same mutation.
-Future work could test when these effects have significant effect on the results of our model.
-% change y1
+Loss of extra-chromosome may occur at a high rate (for estimates in yeast, see \citet{hose_rate_2024}) and the fitness of a euploid cell with a resistance mutation may be higher than that of an aneuploid cell with the same mutation.
+%change S16
+Future work could test when these effects have significant effect on the results of our model. % change y1
 In addition, our model can be applied to study evolutionary rescue in other biological systems, such as evolution of yeast populations under different stress conditions~\citep{pompei2023fitness,kohanovski2024aneuploidy}.
@@
 The probability of evolutionary rescue, $\presc$, as a function of the initial tumor size, $N$ (\cref{eq:rescue_prob}). 
-Dashed vertical lines show the threshold tumor size, $N_a^*$, above which the probability is very high (\cref{eq:N_a}). 
+Dashed vertical lines show the threshold tumor size, $N_a^*$, above which the probability is high (\cref{eq:N_a}). % change S1
 \textbf{(A)} Blue dashed line: without aneuploidy ($u=0$).
@@
 \caption{
-\textbf{The effect of aneuploidy on tumor threshold size.}
+\textbf{Aneuploidy reduces tumor threshold size.} %change S18
 (A) The threshold tumor size $N_a^*$ (\cref{eq:N_a}) as a function of the aneuploid growth rate $\Delta_a$.
@@
 \end{align}
-Now $u v$ is very small, and if we use the fact that $v \ll u$, we have:
+Now $u v$ is small, and if we use the fact that $v \ll u$, we have: % change S1
 \begin{equation}\label{eq:pw_parttolerant}
@@
 \end{equation}
-However, if aneuploidy is very rare such that
+However, if aneuploidy is sufficiently rare such that % change S1
 \begin{align*}
@@
 Assuming no aneuploidy ($u=0$), we define $T_m$ to be the time at which the first mutant cell appears that will avoid extinction and will therefore rescue the population.
-Note that if extinction occurs, that is the frequency of mutants after a very long time is zero, $m_{\infty}=0$, then it is implied that $T_m=\infty$, and vice versa if $T_m<\infty$ then $m_{\infty}>0$.
+Note that if extinction occurs, that is the frequency of mutants after a long time is zero, $m_{\infty}=0$, then it is implied that $T_m=\infty$, and vice versa if $T_m<\infty$ then $m_{\infty}>0$.
+% change S1
 
@@
 For large initial population sizes we assume that the two processes are independent and as a result, they can be merged into a single Poisson process with rate $R_a(t)=\left(n_1+r_2\right)\left(t\right)$.
-Consequently, the mean time to the appearance of the first rescue mutant is
+ The mean time to the appearance of the first rescue mutant is %change S16
 \begin{align}\nonumber
@@
 %%%
-Let $N_a^*$ and $N_m^*$ be the threshold population size above which evolutionary rescue through aneuploidy or direct mutation, respectively, is very likely.
-The mean time $\tau_a$ can be written as: % change 38
+Let $N_a^*$ and $N_m^*$ be the threshold population size above which evolutionary rescue through aneuploidy or direct mutation, respectively, is likely.
+The mean time $\tau_a$ can be written as: % change 38 % change S1
 \begin{align*}
@@
 \end{equation}
-Consequently, the threshold population size can be written as
+The threshold population size can be written as %change S16
 \begin{equation}\label{eq:N_a2}
@@
 \caption{\textbf{Aneuploidy increases the number of mutations which rescue the tumor.} Shown is the expected number of mutations, which will rescue the cancer cell population, produced through the evolutionary trajectory \textit{sensitive} $\rightarrow$ \textit{mutant} (blue line, \cref{eq:twosteplineagedirect}) or through the trajectory \textit{sensitive} $\rightarrow$ \textit{aneuploid} $\rightarrow$ \textit{mutant} (red line, \cref{eq:twosteplineage}).
-Dashed vertical red line represents the threshold tumor size above which evolutionary rescue is very likely through aneuploidy \cref{eq:N_a} and the dashed vertical blue line represents the threshold tumor size above which evolutionary rescue is very likely through direct mutation \cref{eq:N_m}.
+Dashed vertical red line represents the threshold tumor size above which evolutionary rescue is likely through aneuploidy \cref{eq:N_a} and the dashed vertical blue line represents the threshold tumor size above which evolutionary rescue is very likely through direct mutation \cref{eq:N_m}.
+% change S1
 Parameters: $\lambda_s=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.} %change 14
@@
 \includegraphics[width=1\textwidth]{Figures/ProbvN_SGV_Plot.pdf}
-\caption{The probability of evolutionary rescue (i.e., the probability that the population does not go to extinction), $\presc$, as a function of the initial tumor size, $N$.
-Dashed vertical line shows the threshold tumor size, above which the probability is very high.
+\caption{\textbf{Standing genetic variation reduces tumor threshold size.} %change S18
+The probability of evolutionary rescue (i.e., the probability that the population does not go to extinction), $\presc$, as a function of the initial tumor size, $N$.
+Dashed vertical line shows the threshold tumor size, above which the probability is high.
 Blue dashed line represents the probability of evolutionary rescue as a function of $N$ without aneuploidy ($u=0$).
@@
 The dots represent simulation results and the error bars represent $95\%$ confidence intervals ($p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the fraction of simulations in which the tumor has adapted to the stress and $n=100$ is the number of simulations).
-Parameters: $\lambda_s=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.} %change 15
+Parameters: $\lambda_s=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.} %change 15 % change S1
 \label{rescue_prob_sgv}
@@
 \caption{The probability of evolutionary rescue (i.e., the probability that the population does not become extinct), $\presc$, as a function of the initial tumor size, $N$ (\cref{eq:rescue_prob}).
-Dashed vertical line shows the threshold tumor size, $N_a^*$, above which the probability is very high (\cref{eq:N_a2}).
+Dashed vertical line shows the threshold tumor size, $N_a^*$, above which the probability is high (\cref{eq:N_a2}).
 Red dashed line: $v_a=10^{-7}$.
@@
 Dots for simulations and the error bars for $95\%$ confidence interval ($p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the fraction of simulations in which the tumor has been rescued and $n=100$ is the number of simulations).
-Parameters: $\lambda_s=0.1,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v_s=10^{-7}$.}
+Parameters: $\lambda_s=0.1,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v_s=10^{-7}$.} % change S1
 \label{rescue_prob_ane}

### 2025-03-09 ResponseLetter2.tex
Add files via upload

\documentclass[12pt]{extarticle}
\usepackage{geometry}
\geometry{
a4paper,
total={170mm,257mm},
left=20mm,
top=20mm,
headheight=12pt
}

\usepackage{xcolor}
\usepackage{amssymb,amsmath,amsthm}
\usepackage{natbib}
\usepackage{commath}
\usepackage{url}

\renewcommand{\d}{{\rm d}}
\newcommand{\e}{{\rm e}}

% Title page
\title{
	Response to reviewers\\ Evolutionary rescue by aneuploidy in tumors exposed to anti-cancer drugs 
}

\begin{document}
\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\textbf{Reviewer $\#$1}

Minor Edits:

1. Review all Figure titles (main/supplement) and turn the Figure Titles into impact statements of the main result, like you did for Figures 3, 5, and 6, which give your reader an immediate connection to the purpose of the figure and the take-away. Change Figure 3 Title to ". . .evolutionary rescue of tumors under drug treatment. . ." In Table 1, ". . . 0.01, instead. . ." In Appendix A, ". . .(Kendall, 1948). This allows. . ." In Figure 5, "Standing genetic variation will drive drug resistance when the sensitive cell population is rapidly declining." In Figure 5, ". . .Na* is a function. . ."

\textcolor{blue}{We have changed the title of Figure 4 and added a title to Figure S5. Change S18}

2. Please do a search for "very" across the whole manuscript and remove. The world holds no meaning here. If there is a need to describe an overwhelming amount of anything, then quantify it.

\textcolor{blue}{Change S1}

3. There is a tendency in the manuscript to use some kind of connection word to start sentences, and to do it back-to-back. For example, in Lines 22-25 in the Abstract, "Additionally. . .Finally," and lines 51-54 in the Introduction, "Consequently. . .Furthermore." It is a bit distracting and unnecessary and detracts from the overall quality of the paper, which is excellent. Go through the manuscript and remove as much of this is possible.

\textcolor{blue}{Change S16}

4. Line 16, remove ", e.g.,"

\textcolor{blue}{Change S2}

5. Line 25, ". . .small to intermediate size primary and secondary tumors." You use small and intermediate in the discussion. Not sure if you need primary and secondary here.

\textcolor{blue}{We decided to keep "small secondary tumors" as this is an important insight from our model.}

6. Global search on "missegregation" to make sure you use it with/without hyphen consistently.

\textcolor{blue}{Change S4}

7. Line 65, ". . .like mutation. . ."

\textcolor{blue}{Change S5}

8. Lines 120/121, describe susceptible, resistant, and tolerant here.

\textcolor{blue}{We have aded a description of susceptible, resistant, and tolerant in the main text. Change S17}

9. Table at 142, tab the right column so that each column lines up from row to row.

\textcolor{blue}{Change S6}

10. Line 221, put "Table 1" in parentheses in a previous sentence.

\textcolor{blue}{Change S7}

11. Sometimes you set variable names between commas, sometimes in parenthesis, sometimes without a leading comma. I would recommend a consistent strategy.

\textcolor{blue}{We believe that both paranthesis and commas are appropiate depending on the context.}

12. Line 297, ". . .as it reduced the threshold tumor sized needed for rescue."

\textcolor{blue}{Change S8}

13. Lines 334-343: "In the above. . .However. . .Moreover. . .Hence. . .Therefore. . ."

\textcolor{blue}{We have removed all of the connecting words highlighted. Change S19}

14. Remove comma end of line 334

\textcolor{blue}{Change S9}

15. Line 343, ". . .standing genetic variation. . ."

\textcolor{blue}{Change S10}

16. Line 356, ". . .de novo genetic variation. . ."

\textcolor{blue}{The correct terminology is "standing genetic variation" as we have $\tilde{N}_a^*$ as the numerator in equation 6.}

17. Lines 384-386, offer a reason for the statement that starts "Of particular interest. . ."

\textcolor{blue}{We are intersted in the case of tolerant aneuploidy given that this represents a two step evolutionary rescue scenario. Change S11}

18. Line 438, "Decreasing the sensitive cell division rate will also decrease the cancer recurrence time." Note: This is a perfect example of how the leading word "Additionally" is unnecessary bc you use "also" in the sentence.

\textcolor{blue}{Change S12}

19. There a couple of places in the Discussion where you use "fast enough" to communicate the idea of timing with regard to aneuploidy. "Fast enough" indicates something about rate (and maybe timing), but "early enough" is about timing. Consider whether you want to use one or both to get this idea across. (Lines 474 and 477).

\textcolor{blue}{We decided to keep using "fast enough" as in both contexts we talk about the rate at which aneuploids are produced.}

20. Line 481, swap "the resistance mutation" with "a resistance mutation"

\textcolor{blue}{Change S13}

21. Line 499, ". . .aneuploidy provides a partial. . ."

\textcolor{blue}{Change S14}

22. Line 523, ". . .primarily by de novo. . ."

\textcolor{blue}{Change S15}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\textbf{Reviewer $\#$2}

Thanks to the authors for improving the connection with previous rescue theory. A few minor comments on the responses to my comments below. Line numbers in tracked changes version.

lines 69-93: The review of multi-step rescue in the intro makes some helpful connections but the logical flow could be improved. Yes, Martin et al 2013 extended Orr $\&$ Unckless 2008 by allowing for multiple different rescue mutations to occur during wildtype decline, but for the purposes of this paper the more important (and yet unmentioned) extension is to allow rescue by multiple (and in particular, two) mutational steps. I think it is then logical to skip to Osmond et al 2020 and instead of focusing on the fitness landscape, which is irrelevant here, focus on the fact that they approximate the 2-step result of Martin et al 2013 in two regimes: sufficiently subcritical and sufficiently critical single mutants. Ignoring the fitness landscape, this is essentially the same result as in Iwasa et al 2003 and 2004, who were specifically thinking about drug resistance and host shifts. I know I mentioned the 2-locus Uecker and Hermisson (2016) result, as it is 2-step rescue, but without more explanation its not clear how it is relevant to the current paper and, in its current placement, breaks up the flow.

\textcolor{blue}{We mentioned that Martin et al. 2013 considered two-step rescue processes involving a ``transient'' intermediate genotype with growth rate close to zero. Change S19}

lines 126-127: I think the last sentence of the intro, "This is similar to part of Osmond et al. (2020) but focused on a different fitness landscape.", could cause a little confusion. In Osmond et al 2020 there is very explicitly a fitness landscape, with mutational kernels, such that the fitnesses of mutations that arise from a genotype depend on that genotype's fitness. In constrast, here there are three genotypes with independently chosen fitnesses, $\lambda_i - \mu_i$. To avoid confusion I would perhaps just attach a small statement to the previous sentence (about studying 2-step rescue), e.g., ", as in Osmond et al 2020 and Iwasa et al 2003, 2004." and avoid mention of landscapes.

lines 582-585: The same issue about a landscape arises here. Again, I don't think it is helpful to think of the current paper as Osmond et al 2020 with a different fitness lanscape. Relatedly, I find the sentence "Comparing our results to the evolutionary rescue literature, we complement the studies by Iwasa et al. (2003) and Osmond et al. (2020) on multi-step rescue by deriving simple approximations for a different fitness landscape" confusing, in part also because Iwasa et al 2003 say that their results hold for any fitness landscape. Why not just drop that sentence and move on to the 1 vs 2 step rescue result?

lines 585-586: "Similar to Osmond et al. (2020), we find that two-step rescues can be more likely than simple one-step rescues." I think it would be worth adding the reason, because of a higher mutation rate to stepping-stone vs rescue mutations.

lines 606-607: "Future work could test when these effects have significant effect on the results of our model." This sentence could be replaced with something more informative. I.e., one could hypothesize that rates of aneuploid loss are small enough relative to the fitness gain over the wildtype to ignore and that allowing a resistant euploid to have a higher fitness than a resistant aneuploid would decrease the importance of 2-step rescue (by some relatively small constant proportion?).

lines 392-393: The equivalent result in Martin et al 2013 is also for 1-step rescue, not 2-step rescue. See their eqn 3.8, which is derived from eqns 3.3 (de novo 1-step) and 3.6 (standing variance 1-step).

appendix D: Smith $\&$ Haigh 1974 should be Maynard Smith $\&$ Haigh 1974.

\textcolor{red}{Remus comments:}

\textcolor{red}{We change "rapidly" to "easily" because $\tau_m<\tau_a$ conditioned on evolutionary rescue and the word "rapidly" suggests that aneuploid speeds up the process of rescue which is not true. Change RS1}

\textcolor{red}{Recurrence time is a decreasing function of sensitive and aneuploid growth rates not an increasing function as stated in the manuscript. Change RS2}

\textcolor{red}{"anuploids"$\rightarrow$"aneuploids". Change RS3}

\textcolor{red}{"aneuplidies"$\rightarrow$"aneuploids". Change RS4}

\bibliographystyle{plain}
\bibliography{evo2022}
\end{document}

### 2025-03-11 ResponseLetter2.tex
response to reviewer 2

@@
 \usepackage{url}
+\usepackage{csquotes}
 
@@
 
+\textcolor{blue}{We thank the reviewer for their patience and thorough suggestions. At this point, they have basically taught us the evolutionary rescue literature. 
+This was going above and beyond their responsibilities. 
+We apologize that we didn't do this ourselves, and really appreciate the reviewer's efforts.}
+
 lines 69-93: The review of multi-step rescue in the intro makes some helpful connections but the logical flow could be improved.
@@
 
-\textcolor{blue}{We mentioned that Martin et al. 2013 considered two-step rescue processes involving a ``transient'' intermediate genotype with growth rate close to zero. Change S19}
+\textcolor{blue}{ We have tried to make these changes. The relevant paragraphs now read:
+\begin{displayquote}
+There has been extensive theoretical analysis of evolutionary rescue. Orr and Unckless (2008,
+2014) thoroughly investigated the simplest case in which there is a specific single mutation or low-
+frequency variant that i ssufficient to rescue the population. 
+Martin et al. (2013) extended these results
+to consider the effects of a range of rescue genotypes that can vary in both growth rate and variance in
+offspring number. 
+Importantly for the present work, they also argued that two-step processes involving
+a ``transient'' intermediate genotype with growth rate close to zero may be an important form of rescue;
+however, they only calculated that the rate of such rescues should scale with the total number of such
+transient genotypes produced. Osmond et al. (2020) derived approximate expressions for the rate of
+these two-step rescues in limiting cases and found that for some parameter values two-step rescue is
+more likely than rescue by a single step.
+\end{displayquote}
+}
 

### 2025-03-11 ms.tex
response to reviewer 2

@@
 Importantly for the present work, they also argued that two-step processes involving a ``transient'' intermediate genotype with growth rate close to zero may be an important form of rescue; however, they only calculated that the rate of such rescues should scale with the total number of such transient genotypes produced.
-%change S19
-%\citet{uecker2016role} modeled a situation in which \textit{two} mutations at different loci (with recombination between them) are necessary for rescue. 
-\citet{osmond_genetic_2020} considered a population that must achieve rescue, potentially via multiple mutations, by adapting on a fitness landscape described by Fisher's geometric model,
-and found that for some parameter values rescue by two mutational steps is more likely than rescue by a single step.
+%\citet{uecker2016role} modeled a situation in which \textit{two} mutations at different loci (with recombination between them) are necessary for rescue. %change S19
+\citet{osmond_genetic_2020} derived approximate expressions for the rate of these two-step rescues in limiting cases
+and found that for some parameter values two-step rescue is more likely than rescue by a single step.
 
@@
 \citet{iwasa2004evolutionary} used a similar approach to find simple approximate expressions for the probability that a single lineage would evolve to survive via a wide range of possible mutational pathways.
+These results are essentially the same as those described later in the general context by \citet{osmond_genetic_2020} for the probability of evolutionary rescue by multiple mutations,
+which we mentioned in the previous paragraph. 
 A similar set of models describe adaptation to new environments (e.g., a pathogen evolving to cross a species barrier) and fitness valley crossing \citep{antia2003role, weissman2009rate}. % change S1
@@
 Because these results are scattered across multiple papers, and because they all follow from simple branching process approximations, we re-derive all results here for our model.
-While our focus is on the application of evolutionary rescue theory to the evolution of drug resistance in cancer, our model is a general one, and in the general evolutionary rescue context, this work can be seen as a characterization of the circumstances in which two-step processes contribute to rescue even when single-step rescue mutations are available.
-This is similar to part of \citet{osmond_genetic_2020} but focused on a different fitness landscape.
+While our focus is on the application of evolutionary rescue theory to the evolution of drug resistance in cancer, our model is a general one, and in the general evolutionary rescue context, this work can be seen as a characterization of the circumstances in which two-step processes contribute to rescue even when single-step rescue mutations are available, as in \citet{iwasa2003evolutionary, iwasa2004evolutionary, osmond_genetic_2020}.
 
@@
 \end{equation}
-This result has been previously reported by \citet{orr2008population} for the one-step evolutionary rescue and by \citet{martin2013probability} for two-step evolutionary rescue. % change 42
+This result has been previously reported by \citet{orr2008population} and \citet{martin2013probability} for one-step evolutionary rescue. % change 42
 Comparing this approximation of $\tilde{N}^*_{a}/N^*_{a}$ to results of stochastic simulations, we find that the approximations are accurate (\Cref{rescue_denovo}). 
@@
 
-Comparing our results to the evolutionary rescue literature, we complement the studies by \citet{iwasa2003evolutionary} and \citet{osmond_genetic_2020} on multi-step rescue by deriving simple approximations for a different fitness landscape.
-Similar to \citet{osmond_genetic_2020}, we find that two-step rescues can be more likely than simple one-step rescues. 
+Comparing our results to the evolutionary rescue literature, we complement the studies by \citet{iwasa2003evolutionary} and \citet{osmond_genetic_2020}, re-deriving some of their results and extending them by including intermediate (aneuploid) genotypes with growth rates substantially above one.
+Similar to \citet{osmond_genetic_2020}, we find that two-step rescues can be more likely than simple one-step rescues, because the mutation rate to stepping-stone (aneuploid) genotypes can be much higher than that to fully resistant genotypes.
 The ``stationary'' scenario, in which the aneuploid growth rate is close to zero in the presence of the drug, is particularly interesting.
@@
 We can then compare the results of these experiments to predictions of our model to see if tumors with initial size below the threshold \cref{eq:N_a} are more likely to become extinct due to drug exposure.
+It may also be interesting to look for the involvement of aneuploidy in evolutionary rescue in other biological systems, such as evolution of yeast populations under different stress conditions~\citep{pompei2023fitness,kohanovski2024aneuploidy}.
+
+Our model neglects the possibility of back-mutations from aneuploidy to euploidy, and also the possibility that the fitness of a euploid cell with a resistance mutation may be higher than that of an aneuploid cell with the same mutation.
+Both of these would be expected to reduce the importance of two-step rescues via aneuploidy.
+But for the former, unless the rate of loss of extra chromosomes is extremely high (for estimates in yeast, see \citet{hose_rate_2024}), comparable in magnitude to the growth rate $\Delta_a$, we expect that its effect on the dynamics will be negligible.
+%change S16
+The latter possibility, in which aneuploidy substantially reduces the mutant growth rate, seems more likely
+to have an effect. 
+Including it would be a straightforward extension to our model.
+In this case, it could actually be more important to consider the possibility of the loss of aneuploidy, 
+as one would need to check the relative rates of the simple \textit{sensitive} $\rightarrow$ \textit{euploid mutant} path and the three-step \textit{sensitive} $\rightarrow$ \textit{aneuploid} $\rightarrow$ \textit{aneuploid mutant} $\rightarrow$ \textit{euploid mutant} path \citep{kohanovski2024aneuploidy}.
 
@@
 This can potentially impact the probability of evolutionary rescue \citep{martens2011spatial}.
-Loss of extra-chromosome may occur at a high rate (for estimates in yeast, see \citet{hose_rate_2024}) and the fitness of a euploid cell with a resistance mutation may be higher than that of an aneuploid cell with the same mutation.
-%change S16
-Future work could test when these effects have significant effect on the results of our model. % change y1
-In addition, our model can be applied to study evolutionary rescue in other biological systems, such as evolution of yeast populations under different stress conditions~\citep{pompei2023fitness,kohanovski2024aneuploidy}.
 
@@
 \end{align}
-The factor $p_m$ in the second term of \cref{meanproliferationtime} occurs because the lineage is conditioned to survive stochastic extinction and the time to reach $N$ is shorter compared to the scenario where the lineage is not conditioned to survive stochastic extinction~\citep{orr2014population,smith1974hitch}.
+The factor $p_m$ in the second term of \cref{meanproliferationtime} occurs because the lineage is conditioned to survive stochastic extinction and the time to reach $N$ is shorter compared to the scenario where the lineage is not conditioned to survive stochastic extinction~\citep{orr2014population,maynardsmith1974hitch}.
 % change 35

### 2025-03-11 ResponseLetter2.tex
updated response to reviewer 2

@@
 
-\textcolor{blue}{ We have tried to make these changes. The relevant paragraphs now read:
+\textcolor{blue}{ We have tried to make these changes. The relevant paragraphs now read:}
 \begin{displayquote}
+	\textcolor{blue}{
     There has been extensive theoretical analysis of evolutionary rescue. Orr and Unckless (2008,
@@
     these two-step rescues in limiting cases and found that for some parameter values two-step rescue is
-more likely than rescue by a single step.
+    more likely than rescue by a single step.}
+
+    \textcolor{blue}{
+    Resistance to cancer therapy is a natural application of evolutionary rescue theory (Alexander et al, 2014). 
+    Iwasa et al (2003) used multi-type branching process theory to approximate the probability that a population under strong selective pressure can survive extinction, focusing on the case in which the cancer must get two mutations to be rescued, with the single-mutant still dying rapidly.
+    Iwasa et al (2004) used a similar approach to find simple approximate expressions for the probability that a single lineage would evolve to survive via a wide range of possible mutational pathways.
+    These results are essentially the same as those described later in the general context by Osmond et al (2020) for the probability of evolutionary rescue by multiple mutations,
+    which we mentioned in the previous paragraph\ldots}
 \end{displayquote}
-}
 
@@
 To avoid confusion I would perhaps just attach a small statement to the previous sentence (about studying 2-step rescue), e.g., ", as in Osmond et al 2020 and Iwasa et al 2003, 2004." and avoid mention of landscapes.
+
+\textcolor{blue}{We have made this change.}
 
@@
 
+\textcolor{blue}{We have made both these changes.
+The paragraph now begins, ``Similar to Osmond et al (2020), we find that two-step rescues can be more likely than simple one-step rescues, because the mutation rate to stepping-stone (aneuploid) genotypes can be much higher than that to fully resistant genotypes.''}
+
 lines 606-607: "Future work could test when these effects have significant effect on the results of our model." This sentence could be replaced with something more informative.
 I.e., one could hypothesize that rates of aneuploid loss are small enough relative to the fitness gain over the wildtype to ignore and that allowing a resistant euploid to have a higher fitness than a resistant aneuploid would decrease the importance of 2-step rescue (by some relatively small constant proportion?).
+
+\textcolor{blue}{Agreed. We have rewritten this paragraph as follows:
+\begin{displayquote}
+    Our model neglects the possibility of back-mutations from aneuploidy to euploidy, and also the possibility that the fitness of a euploid cell with a resistance mutation may be higher than that of an aneuploid cell with the same mutation.
+    Both of these would be expected to reduce the importance of two-step rescues via aneuploidy.
+    But for the former, unless the rate of loss of extra chromosomes is extremely high (for estimates in yeast, see Hose et al (2024)), comparable in magnitude to the growth rate $r_a$, we expect that its effect on the dynamics will be negligible.
+    The latter possibility, in which aneuploidy substantially reduces the mutant growth rate, seems more likely
+    to have an effect. 
+    Including it would be a straightforward extension to our model.
+    In this case, it could actually be more important to consider the possibility of the loss of aneuploidy, 
+    as one would need to check the relative rates of the simple \textit{sensitive} $\rightarrow$ \textit{euploid mutant} path and the three-step \textit{sensitive} $\rightarrow$ \textit{aneuploid} $\rightarrow$ \textit{aneuploid mutant} $\rightarrow$ \textit{euploid mutant} path (Kohanovski et al, 2024).
+\end{displayquote}
+}
 
@@
 
+\textcolor{blue}{Fixed.}
+
 appendix D: Smith $\&$ Haigh 1974 should be Maynard Smith $\&$ Haigh 1974.
+
+\textcolor{blue}{Fixed.}
 

### 2025-03-11 ms.tex
updated response to reviewer 2

@@
 Because these results are scattered across multiple papers, and because they all follow from simple branching process approximations, we re-derive all results here for our model.
-While our focus is on the application of evolutionary rescue theory to the evolution of drug resistance in cancer, our model is a general one, and in the general evolutionary rescue context, this work can be seen as a characterization of the circumstances in which two-step processes contribute to rescue even when single-step rescue mutations are available, as in \citet{iwasa2003evolutionary, iwasa2004evolutionary, osmond_genetic_2020}.
+While our focus is on the application of evolutionary rescue theory to the evolution of drug resistance in cancer, our model is a general one, and in the general evolutionary rescue context, this work can be seen as a characterization of the circumstances in which two-step processes contribute to rescue even when single-step rescue mutations are available, as in \citet{iwasa2003evolutionary, iwasa2004evolutionary} and \citet{osmond_genetic_2020}.
 
@@
 
-Comparing our results to the evolutionary rescue literature, we complement the studies by \citet{iwasa2003evolutionary} and \citet{osmond_genetic_2020}, re-deriving some of their results and extending them by including intermediate (aneuploid) genotypes with growth rates substantially above one.
 Similar to \citet{osmond_genetic_2020}, we find that two-step rescues can be more likely than simple one-step rescues, because the mutation rate to stepping-stone (aneuploid) genotypes can be much higher than that to fully resistant genotypes.

### 2025-03-11 ResponseLetter2.tex
edited response to reviewer 1

@@
 1.
-Review all Figure titles (main/supplement) and turn the Figure Titles into impact statements of the main result, like you did for Figures 3, 5, and 6, which give your reader an immediate connection to the purpose of the figure and the take-away.
-Change Figure 3 Title to ".
-.
-.evolutionary rescue of tumors under drug treatment.
-.
-." In Table 1, ".
-.
-.
-0.01, instead.
-.
-." In Appendix A, ".
-.
-.(Kendall, 1948).
-This allows.
-.
-." In Figure 5, "Standing genetic variation will drive drug resistance when the sensitive cell population is rapidly declining." In Figure 5, ".
-.
-.Na* is a function.
-.
-."
+Review all Figure titles (main/supplement) and turn the Figure titles into impact statements of the main result, like you did for Figures 3, 5, and 6, which give your reader an immediate connection to the purpose of the figure and the take-away.
+Change Figure 3 title to ``\ldots evolutionary rescue of tumors under drug treatment \ldots''.
+In Table 1, ``\ldots 0.01, instead \dots''.
+In Appendix A, ``\ldots (Kendall, 1948).
+This allows\ldots''.
+In Figure 5, ``Standing genetic variation will drive drug resistance when the sensitive cell population is rapidly declining.'' In Figure 5, ``\ldots $N_a^*$ is a function\ldots''
 
@@
 
-2. Please do a search for "very" across the whole manuscript and remove. The world holds no meaning here. If there is a need to describe an overwhelming amount of anything, then quantify it.
+2. Please do a search for ``very'' across the whole manuscript and remove. The world holds no meaning here. If there is a need to describe an overwhelming amount of anything, then quantify it.
 
@@
 There is a tendency in the manuscript to use some kind of connection word to start sentences, and to do it back-to-back.
-For example, in Lines 22-25 in the Abstract, "Additionally.
-.
-.Finally," and lines 51-54 in the Introduction, "Consequently.
-.
-.Furthermore." It is a bit distracting and unnecessary and detracts from the overall quality of the paper, which is excellent.
+For example, in Lines 22-25 in the Abstract, ``Additionally\ldots Finally,'' and lines 51-54 in the Introduction, ``Consequently\ldots Furthermore.'' It is a bit distracting and unnecessary and detracts from the overall quality of the paper, which is excellent.
 Go through the manuscript and remove as much of this is possible.
@@
 
-4. Line 16, remove ", e.g.,"
+4. Line 16, remove ``, e.g.,''
 
@@
 
-5. Line 25, ". . .small to intermediate size primary and secondary tumors." You use small and intermediate in the discussion. Not sure if you need primary and secondary here.
+5. Line 25, ``\ldots small to intermediate size primary and secondary tumors.'' You use small and intermediate in the discussion. Not sure if you need primary and secondary here.
 
-\textcolor{blue}{We decided to keep "small secondary tumors" as this is an important insight from our model.}
+\textcolor{blue}{We decided to keep ``small secondary tumors'' as this is an important insight from our model.}
 
-6. Global search on "missegregation" to make sure you use it with/without hyphen consistently.
+6. Global search on ``missegregation'' to make sure you use it with/without hyphen consistently.
 
@@
 
-7. Line 65, ". . .like mutation. . ."
+7. Line 65, ``\dots like mutation\ldots''
 
@@
 
-\textcolor{blue}{We believe that both paranthesis and commas are appropiate depending on the context.}
+\textcolor{blue}{We believe that both parenthesis and commas are appropriate depending on the context.}
 
-12. Line 297, ". . .as it reduced the threshold tumor sized needed for rescue."
+12. Line 297, ``\ldots as it reduced the threshold tumor sized needed for rescue.''
 
@@
 
-13. Lines 334-343: "In the above. . .However. . .Moreover. . .Hence. . .Therefore. . ."
+13. Lines 334-343: ``In the above\ldots However\ldots Moreover\ldots Hence\dots Therefore\ldots ''
 
@@
 19.
-There a couple of places in the Discussion where you use "fast enough" to communicate the idea of timing with regard to aneuploidy.
-"Fast enough" indicates something about rate (and maybe timing), but "early enough" is about timing.
+There a couple of places in the Discussion where you use ``fast enough'' to communicate the idea of timing with regard to aneuploidy.
+``Fast enough'' indicates something about rate (and maybe timing), but ``early enough'' is about timing.
 Consider whether you want to use one or both to get this idea across.
@@
 
-\textcolor{blue}{We decided to keep using "fast enough" as in both contexts we talk about the rate at which aneuploids are produced.}
+\textcolor{blue}{We have changed both usages to make it clear that we are talking specifically about the rate of occurrence of aneuploidy, not the timing.
+They now read ``as long as it occurs frequently enough'' and ``if the rate of occurrence of aneuploidy is high enough''.}
 
@@
 
-\textcolor{red}{Remus comments:}
-
-\textcolor{red}{We change "rapidly" to "easily" because $\tau_m<\tau_a$ conditioned on evolutionary rescue and the word "rapidly" suggests that aneuploid speeds up the process of rescue which is not true.
+%\textcolor{red}{Remus comments:}
+%
+%\textcolor{red}{We change "rapidly" to "easily" because $\tau_m<\tau_a$ conditioned on evolutionary rescue and the word "rapidly" suggests that aneuploid speeds up the process of rescue which is not true.
 Change RS1}
-
-\textcolor{red}{Recurrence time is a decreasing function of sensitive and aneuploid growth rates not an increasing function as stated in the manuscript. Change RS2}
-
-\textcolor{red}{"anuploids"$\rightarrow$"aneuploids". Change RS3}
-
-\textcolor{red}{"aneuplidies"$\rightarrow$"aneuploids". Change RS4}
+%
+%\textcolor{red}{Recurrence time is a decreasing function of sensitive and aneuploid growth rates not an increasing function as stated in the manuscript. Change RS2}
+%
+%\textcolor{red}{"anuploids"$\rightarrow$"aneuploids". Change RS3}
+%
+%\textcolor{red}{"aneuplidies"$\rightarrow$"aneuploids". Change RS4}
+%
+%
+%
 

### 2025-03-11 ms.tex
edited response to reviewer 1

@@
 Our results show that the probability of evolutionary rescue increases with the initial tumor size $N$, the drug-sensitive growth rate $\Delta_s$, the mutation rate $v$, and the aneuploidy rate $u$.
-Notably, the latter indicates that aneuploidy, even when it only provides tolerance, increases the probability that the tumor will be rescued, as long as it is produced fast enough (\Cref{rescue_prob}A, \cref{eq:N_ratio}).
+Notably, the latter indicates that aneuploidy, even when it only provides tolerance, increases the probability that the tumor will be rescued, as long as it occurs frequently enough (\Cref{rescue_prob}A, \cref{eq:N_ratio}).
 
@@
 When aneuploidy only provides tolerance to the drug ($\Delta_s\ll\Delta_a\ll0\ll\Delta_m$), it cannot rescue the population.
-Instead, if the aneuploidy rate is fast enough (\cref{eq:N_ratio}) then it may act as a ``stepping stone'' through which the resistant mutant can appear more easily, given that the number of aneuploid cells declines slower than the number of drug-sensitive cells (\Cref{sampleTrajectories}).
+Instead, if the rate of occurrence of aneuploidy is high enough (\cref{eq:N_ratio}) then aneuploidy may act as a ``stepping stone'' through which the resistant mutant can appear more easily, given that the number of aneuploid cells declines slower than the number of drug-sensitive cells (\Cref{sampleTrajectories}).
 %change RS1

### 2025-03-14 ResponseLetter2.tex
Remus updates

@@
 
-\textcolor{blue}{Change S1}
+\textcolor{blue}{We have made these changes. Change S1}
 
@@
 
-\textcolor{blue}{Change S16}
+\textcolor{blue}{Fixed. Change S16}
 
@@
 
-\textcolor{blue}{Change S2}
+\textcolor{blue}{Fixed. Change S2}
 
@@
 
-\textcolor{blue}{Change S4}
+\textcolor{blue}{Fixed. Change S4}
 
@@
 
-\textcolor{blue}{Change S5}
+\textcolor{blue}{Fixed. Change S5}
 
@@
 
-\textcolor{blue}{Change S6}
+\textcolor{blue}{Fixed. Change S6}
 
@@
 
-\textcolor{blue}{Change S7}
+\textcolor{blue}{Fixed. Change S7}
 
@@
 
-\textcolor{blue}{Change S8}
+\textcolor{blue}{Fixed. Change S8}
 
@@
 
-\textcolor{blue}{Change S9}
+\textcolor{blue}{Fixed. Change S9}
 
@@
 
-\textcolor{blue}{Change S10}
+\textcolor{blue}{Fixed. Change S10}
 
@@
 
-\textcolor{blue}{We are intersted in the case of tolerant aneuploidy given that this represents a two step evolutionary rescue scenario. Change S11}
+\textcolor{blue}{We are interested in the case of tolerant aneuploidy given that this represents a two step evolutionary rescue scenario. Change S11}
 
@@
 
-\textcolor{blue}{Change S12}
+\textcolor{blue}{Decreasing the sensitive cell division will increase the cancer recurrence time. Change S12}
 
@@
 \textcolor{blue}{We have changed both usages to make it clear that we are talking specifically about the rate of occurrence of aneuploidy, not the timing.
-They now read ``as long as it occurs frequently enough'' and ``if the rate of occurrence of aneuploidy is high enough''.}
+They now read ``as long as it occurs frequently enough'' and ``if the rate of occurrence of aneuploidy is high enough''.
+Change S20}
 
@@
 
-\textcolor{blue}{Change S13}
+\textcolor{blue}{We have made this change. Change S13}
 
@@
 
-\textcolor{blue}{Change S14}
+\textcolor{blue}{Fixed. Change S14}
 
@@
 
-\textcolor{blue}{Change S15}
+\textcolor{blue}{Fixed. Change S15}
 

### 2025-03-14 ms.tex
Remus update

@@
 Drugs that significantly increase the death rate of sensitive cells, $\mu_s$, but do not affect their division rate, $\lambda_s$, delay cancer recurrence.
-Additionally, decreasing the sensitive division rate will also the cancer recurrence time.
-Consequently, patients treated with such drugs may require a longer period of monitoring to guarantee the effectiveness of the treatment.
-%change 49
+Additionally, decreasing the sensitive division rate will also increase the cancer recurrence time.
+Patients treated with such drugs may require a longer period of monitoring to guarantee the effectiveness of the treatment.
+%change 49 change S12
 Drugs that significantly increase the division rate of sensitive cells, $\lambda_s$, but do not affect their death rate, $\mu_s$, decrease cancer recurrence times (conditioned on evolutionary rescue).
-%change S12
 
@@
 Notably, the latter indicates that aneuploidy, even when it only provides tolerance, increases the probability that the tumor will be rescued, as long as it occurs frequently enough (\Cref{rescue_prob}A, \cref{eq:N_ratio}).
+%change S20
 
@@
 Instead, if the rate of occurrence of aneuploidy is high enough (\cref{eq:N_ratio}) then aneuploidy may act as a ``stepping stone'' through which the resistant mutant can appear more easily, given that the number of aneuploid cells declines slower than the number of drug-sensitive cells (\Cref{sampleTrajectories}).
-%change RS1
+%change RS1, S20
 In this scenario, aneuploidy provides two advantages. First, it delays the extinction of the population, providing more time for the appearance of a resistance mutation. %change S13
@@
 \caption{\textbf{Aneuploidy increases the number of mutations which rescue the tumor.} Shown is the expected number of mutations, which will rescue the cancer cell population, produced through the evolutionary trajectory \textit{sensitive} $\rightarrow$ \textit{mutant} (blue line, \cref{eq:twosteplineagedirect}) or through the trajectory \textit{sensitive} $\rightarrow$ \textit{aneuploid} $\rightarrow$ \textit{mutant} (red line, \cref{eq:twosteplineage}).
-Dashed vertical red line represents the threshold tumor size above which evolutionary rescue is likely through aneuploidy \cref{eq:N_a} and the dashed vertical blue line represents the threshold tumor size above which evolutionary rescue is very likely through direct mutation \cref{eq:N_m}.
+Dashed vertical red line represents the threshold tumor size above which evolutionary rescue is likely through aneuploidy \cref{eq:N_a} and the dashed vertical blue line represents the threshold tumor size above which evolutionary rescue is likely to occur through direct mutation \cref{eq:N_m}.
 % change S1

### 2025-03-27 ms.tex
Add files via upload

@@
 For small tumors ($ N \ll N_a^*$), the mean rescue time is the two-step equivalent of the one-step result from \citet[expectation of eq.~18]{orr2014population}.
-The mean rescue time (conditioned on rescue occuring) is a decreasing function of the sensitive and aneuploid growth rates and independent of the other model parameters, including tumor size (blue line in \Cref{EvolutionaryRescueTimeComplete}).
+The mean rescue time (conditioned on rescue occuring) is an increasing function of the sensitive and aneuploid growth rates and independent of the other model parameters, including tumor size (blue line in \Cref{EvolutionaryRescueTimeComplete}).
+%change RS2
 This is because if the population rapidly declines but is then rescued, than the resistance mutation must have appeared early; if the population slowly declines, than mutations can appear later and the mean time will be longer.
@@
 It grows logarithmically with tumor size $N$ and may be the same order of magnitude as the mean evolutionary rescue time. 
-Increasing the sensitive $\Delta_s$, aneuploid $\Delta_a$ and mutant $\Delta_m$ growth rates decreases the recurrence time.%, while increasing the sensitive and aneuploid growth rates, $\Delta_s$ and $\Delta_a$, respectively, increases the recurrence time.
-%change RS2
+Increasing mutant $\Delta_m$ growth rates decreases the recurrence time, while increasing the sensitive and aneuploid growth rates, $\Delta_s$ and $\Delta_a$, respectively, increases the recurrence time.
 

### 2025-03-27 ResponseLetter2.tex
Add files via upload

@@
 
-%\textcolor{red}{Remus comments:}
-%
-%\textcolor{red}{We change "rapidly" to "easily" because $\tau_m<\tau_a$ conditioned on evolutionary rescue and the word "rapidly" suggests that aneuploid speeds up the process of rescue which is not true.
+\textcolor{red}{Remus comments:}
+
+\textcolor{red}{We change "rapidly" to "easily" because $\tau_m<\tau_a$ conditioned on evolutionary rescue and the word "rapidly" suggests that aneuploid speeds up the process of rescue which is not true.
 Change RS1}
-%
-%\textcolor{red}{Recurrence time is a decreasing function of sensitive and aneuploid growth rates not an increasing function as stated in the manuscript. Change RS2}
-%
-%\textcolor{red}{"anuploids"$\rightarrow$"aneuploids". Change RS3}
-%
-%\textcolor{red}{"aneuplidies"$\rightarrow$"aneuploids". Change RS4}
-%
-%
-%
+
+\textcolor{red}{Recurrence time is a increasing function of sensitive and aneuploid growth rates not a decreasing function as stated in the manuscript. Change RS2}
+
+\textcolor{red}{"anuploids"$\rightarrow$"aneuploids". Change RS3}
+
+\textcolor{red}{"aneuplidies"$\rightarrow$"aneuploids". Change RS4}
 

### 2025-03-30 ms.tex
Add files via upload

@@
 As expected, this ratio increases with the mutation rate $v$ and decreases with the aneuploidy rate $u$.
-In the first scenario, $\abs{\Delta_a}/u\lambda_a$ is the ratio of the expected time for an aneuploid lineage to appear, $1/u\lambda_a$, and the expected time until that lineage disappears, $1/\abs{\Delta_a}$.
+In the first scenario, $\abs{\Delta_a}/u\lambda_a$ is the ratio of the number of sensitive divisions required to produce an aneuploid, $1/u$, to the number of divisions in the resulting aneuploid lineage before it goes extinct, $\lambda_a / |\Delta_a|$.
+% RS5
+%is the ratio of the expected time for an aneuploid lineage to appear, $1/u\lambda_a$, and the expected time until that lineage disappears, $1/\abs{\Delta_a}$.
 In the third scenario, $\left(v \frac{\Delta_m}{\lambda_m}\right) / \left(u \frac{\Delta_a}{\lambda_a}\right)$ is the ratio of the rates of appearance of resistant mutants that avoid extinction and partially resistant aneuploids that avoid extinction.
@@
 Drugs that significantly increase the death rate of sensitive cells, $\mu_s$, but do not affect their division rate, $\lambda_s$, delay cancer recurrence.
-Additionally, decreasing the sensitive division rate will also increase the cancer recurrence time.
+Drugs that decrease the sensitive division rate delay cancer recurrence time even more, because they effectively decrease the mutation rate (assuming mutations occur during division).
 Patients treated with such drugs may require a longer period of monitoring to guarantee the effectiveness of the treatment.
 %change 49 change S12
-Drugs that significantly increase the division rate of sensitive cells, $\lambda_s$, but do not affect their death rate, $\mu_s$, decrease cancer recurrence times (conditioned on evolutionary rescue).
+%Drugs that significantly increase the division rate of sensitive cells, $\lambda_s$, but do not affect their death rate, $\mu_s$, decrease cancer recurrence times (conditioned on evolutionary rescue).
+%RS6
 

### 2025-03-30 ResponseLetter2.tex
Add files via upload

@@
 
+\textcolor{red}{$1/u\lambda_a$ is not the expected time for an aneuploid lineage to appear and we have corrected the interpretation of $\abs{\Delta_a}/u\lambda_a$ as it is the ratio of the number of sensitive births required to produce an aneuploid, $1/u$, to the number of births in the resulting aneuploid lineage before it goes extinct, $\lambda_a / |\Delta_a|$.
+Change RS5}
+
+\textcolor{red}{We deleted the sentence "Drugs that significantly increase the division rate of sensitive cells, $\lambda_s$, but do not affect their death rate, $\mu_s$, decrease cancer recurrence times (conditioned on evolutionary rescue)" because while it is mathematically true it not make sense for the drug to increase the fitness of sensitive cells.
+Change RS6}
+
 \bibliographystyle{plain}

### 2025-04-20 ResponseLetter2.tex
revised responses

--- 2025-03-27 ResponseLetter2.tex
@@
 \title{
-	Response to reviewers\\ Evolutionary rescue by aneuploidy in tumors exposed to anti-cancer drugs 
+	Response to reviewers Evolutionary rescue by aneuploidy in tumors exposed to anti-cancer drugs 
 }
@@
 
+I would like to congratulate the authors on the continued progress of this work.
+I am an evolutionary biologist, but not generally a reader or reviewer for this kind of work.
+I would like to reiterate my pleasure in reviewing this manuscript, both on first and second submission.
+I feel strongly that you have an aptitude for making this kind of work digestible for a more general genetics/evolutionary biology audience, and I am happy to see the manuscript improved on second submission, given how solid and interesting it was on first submission.
+
+\textcolor{blue}{We thank the reviewer for his kind words and his helpful comments that helped to make the manuscript clearer and better!}
+
 Minor Edits:
@@
 Review all Figure titles (main/supplement) and turn the Figure titles into impact statements of the main result, like you did for Figures 3, 5, and 6, which give your reader an immediate connection to the purpose of the figure and the take-away.
+\textcolor{blue}{Done.}
+
 Change Figure 3 title to ``\ldots evolutionary rescue of tumors under drug treatment \ldots''. 
+\textcolor{blue}{Changed.}
+
 In Table 1, ``\ldots 0.01, instead \dots''. 
-In Appendix A, ``\ldots (Kendall, 1948).
-This allows\ldots''.
-In Figure 5, ``Standing genetic variation will drive drug resistance when the sensitive cell population is rapidly declining.'' In Figure 5, ``\ldots $N_a^*$ is a function\ldots''
+\textcolor{blue}{Fixed.}
 
-\textcolor{blue}{We have changed the title of Figure 4 and added a title to Figure S5. Change S18}
+In Appendix A, ``\ldots (Kendall, 1948). This allows\ldots''. 
+\textcolor{blue}{Fixed.}
 
-2. Please do a search for ``very'' across the whole manuscript and remove. The world holds no meaning here. If there is a need to describe an overwhelming amount of anything, then quantify it.
+In Figure 5, ``Standing genetic variation will drive drug resistance when the sensitive cell population is rapidly declining.'' 
+\textcolor{blue}{Changed.}
 
-\textcolor{blue}{We have made these changes. Change S1}
+In Figure 5, ``\ldots $N_a^*$ is a function\ldots''
+
+\textcolor{blue}{We kept "as", but revised the sentence to be clearer: ``Ratio of threshold tumor size $\tilde{N}_a^*$ (when a fraction $\frac{\tilde{u}\lambda_s}{c}$ is aneuploid at the start of treatment) and $N_a^*$ as a function of the ratio of aneuploidy rates $\tilde{u}/u$''.}
+
+2. Please do a search for ``very'' across the whole manuscript and remove. The word holds no meaning here. If there is a need to describe an overwhelming amount of anything, then quantify it. 
+\textcolor{blue}{Done.}
 
@@
 Go through the manuscript and remove as much of this is possible.
-
-\textcolor{blue}{Fixed. Change S16}
+\textcolor{blue}{Done.} % s16
 
 4. Line 16, remove ``, e.g.,''
-
-\textcolor{blue}{Fixed. Change S2}
+\textcolor{blue}{Changed to ``Aneuploid cancer cells can be more fit in an environment altered by anti-cancer drugs, in part because aneuploidy may disrupt the pathways  targeted by the drugs.''.}
 
@@
 6. Global search on ``missegregation'' to make sure you use it with/without hyphen consistently.
-
-\textcolor{blue}{Fixed. Change S4}
+\textcolor{blue}{Done.}
 
 7. Line 65, ``\dots like mutation\ldots''
-
-\textcolor{blue}{Fixed. Change S5}
+\textcolor{blue}{Fixed.}
 
@@
 
-\textcolor{blue}{We have aded a description of susceptible, resistant, and tolerant in the main text. Change S17}
+\textcolor{blue}{We added (line 124): ``We define tolerant cells as those that survive drug exposure longer than susceptible cells, while resistant cells require higher drug concentrations to be affected.
+For details on these distinctions, see Brauner et al.
+(2016).''} %Change S17
 
 9. Table at 142, tab the right column so that each column lines up from row to row.
-
-\textcolor{blue}{Fixed. Change S6}
+\textcolor{blue}{Done.}
 
 10. Line 221, put "Table 1" in parentheses in a previous sentence.
-
-\textcolor{blue}{Fixed. Change S7}
+\textcolor{blue}{Fixed.}
 
@@
 
-\textcolor{blue}{We believe that both parenthesis and commas are appropriate depending on the context.}
+\textcolor{blue}{We went through all occurences and ensured that parenthesis and commas are used appropriately depending on the context.}
 
 12. Line 297, ``\ldots as it reduced the threshold tumor sized needed for rescue.''
-
-\textcolor{blue}{Fixed. Change S8}
+\textcolor{blue}{Fixed.} % Change S8
 
@@
 
-\textcolor{blue}{We have removed all of the connecting words highlighted. Change S19}
+\textcolor{blue}{We revised the paragraph at line X to \\
+``We initially assumed that at the onset of drug treatment, the tumor consisted entirely of drug-sensitive cells.
+In reality, aneuploidy is likely generated even before treatment begins, at some rate $\tilde{u}$, which may be lower in the absence of drugs, $\tilde{u} < u$ (Mason et al., 2016, Wang et al., 2019).
+Aneuploidy is also thought to carry a fitness cost $c$ when drugs are absent (Giam and Rancati, 2015, Replogle et al., 2020). 
+Therefore, if the tumor size $N$ is large--as expected at the time of treatment--a fraction of cells, approximately $f \approx \tilde{u}\lambda_s/c$, may already be aneuploid.
+Here we assume that the drug affects the sensitive death rate but not the division rate and therefore we use $\lambda_s$ for the sensitive division rate in the drug-free environment.''}
+% Change S19
 
 14. Remove comma end of line 334
-
-\textcolor{blue}{Fixed. Change S9}
+\textcolor{blue}{Fixed. }%Change S9
 
 15. Line 343, ". . .standing genetic variation. . ."
-
-\textcolor{blue}{Fixed. Change S10}
+\textcolor{blue}{Fixed. }% Change S10
 
 16. Line 356, ". . .de novo genetic variation. . ."
-
-\textcolor{blue}{The correct terminology is "standing genetic variation" as we have $\tilde{N}_a^*$ as the numerator in equation 6.}
+\textcolor{blue}{Fixed.}
 
@@
 
-\textcolor{blue}{We are interested in the case of tolerant aneuploidy given that this represents a two step evolutionary rescue scenario. Change S11}
+\textcolor{blue}{We revised the text (line 384): ``Of particular interest is the distribution of the evolutionary rescue time and recurrence time with tolerant aneuploid cells ($\Delta_a T^*\ll1$) \emph{as it represents the two-step evolutionary rescue scenario.}}
+%Change S11
 
@@
 
-\textcolor{blue}{Decreasing the sensitive cell division will increase the cancer recurrence time. Change S12}
+\textcolor{blue}{Revised to (line 437): ``Drugs that significantly increase the death rate of sensitive cells, $\mu_s$, but do not affect their division rate, $\lambda_s$, delay cancer recurrence.
+Drugs that decrease the sensitive division rate delay cancer recurrence time even more, because they effectively decrease the mutation rate (assuming mutations occur during division).
+Patients treated with such drugs may require a longer period of monitoring to guarantee the effectiveness of the treatment.''}
+% Change S12}
 
@@
 They now read ``as long as it occurs frequently enough'' and ``if the rate of occurrence of aneuploidy is high enough''.
-Change S20}
+}%Change S20}
 
 20. Line 481, swap "the resistance mutation" with "a resistance mutation"
+\textcolor{blue}{Done. }%Change S13}
 
-\textcolor{blue}{We have made this change. Change S13}
-
-21. Line 499, ". . .aneuploidy provides a partial. . ."
-
-\textcolor{blue}{Fixed. Change S14}
+21. Line 499, ". . .aneuploidy provides a partial. . .
+\textcolor{blue}{Fixed.}%Change S14}
 
 22. Line 523, ". . .primarily by de novo. . ."
-
-\textcolor{blue}{Fixed. Change S15}
+\textcolor{blue}{Fixed. }% S15}
+\\
 
@@
 
-\textcolor{blue}{ We have tried to make these changes. The relevant paragraphs now read:}
+\textcolor{blue}{ We have tried to make these changes. The relevant paragraphs now read (lines 67-85):}
 \begin{displayquote}
 	\textcolor{blue}{
-    There has been extensive theoretical analysis of evolutionary rescue. Orr and Unckless (2008,
+    ``There has been extensive theoretical analysis of evolutionary rescue. Orr and Unckless (2008,
     2014) thoroughly investigated the simplest case in which there is a specific single mutation or low-
@@
     these two-step rescues in limiting cases and found that for some parameter values two-step rescue is
-    more likely than rescue by a single step.}
-
-    \textcolor{blue}{
+    more likely than rescue by a single step.
+    \\
     Resistance to cancer therapy is a natural application of evolutionary rescue theory (Alexander et al, 2014). 
@@
     Iwasa et al (2004) used a similar approach to find simple approximate expressions for the probability that a single lineage would evolve to survive via a wide range of possible mutational pathways.
-    These results are essentially the same as those described later in the general context by Osmond et al (2020) for the probability of evolutionary rescue by multiple mutations,
-    which we mentioned in the previous paragraph\ldots}
+    These results are essentially the same as those described later in the general context by Osmond et al (2020) for the probability of evolutionary rescue by multiple mutations, which we mentioned in the previous paragraph.''}
 \end{displayquote}
@@
 To avoid confusion I would perhaps just attach a small statement to the previous sentence (about studying 2-step rescue), e.g., ", as in Osmond et al 2020 and Iwasa et al 2003, 2004." and avoid mention of landscapes.
-
-\textcolor{blue}{We have made this change.}
+\textcolor{blue}{Done, line 114.}
 
@@
 \textcolor{blue}{We have made both these changes.
-The paragraph now begins, ``Similar to Osmond et al (2020), we find that two-step rescues can be more likely than simple one-step rescues, because the mutation rate to stepping-stone (aneuploid) genotypes can be much higher than that to fully resistant genotypes.''}
+The paragraph now begins (line 523), ``Similar to Osmond et al.
+(2020), we find that two-step rescues can be more likely than simple one-step rescues, because the mutation rate to stepping-stone (aneuploid) genotypes can be much higher than that to fully resistant genotypes.''}
 
@@
 
-\textcolor{blue}{Agreed. We have rewritten this paragraph as follows:
+\textcolor{blue}{Agreed. We have rewritten this paragraph as follows (line 541):
 \begin{displayquote}
@@
 See their eqn 3.8, which is derived from eqns 3.3 (de novo 1-step) and 3.6 (standing variance 1-step).
-
 \textcolor{blue}{Fixed.}
@@
 appendix D: Smith $\&$ Haigh 1974 should be Maynard Smith $\&$ Haigh 1974.
+\textcolor{blue}{Fixed.}
+\\
 
-\textcolor{blue}{Fixed.}
+\textbf{Additional changes by the authors}
 
-\textcolor{red}{Remus comments:}
+\textcolor{blue}{Line 477: We changed "rapidly" to "easily" because $\tau_m<\tau_a$ is conditioned on evolutionary rescue and the word "rapidly" may suggest that aneuploidy speeds up the process of rescue, which is incorrect.
+}%Change RS1}
 
-\textcolor{red}{We change "rapidly" to "easily" because $\tau_m<\tau_a$ conditioned on evolutionary rescue and the word "rapidly" suggests that aneuploid speeds up the process of rescue which is not true.
-Change RS1}
+\textcolor{blue}{Line 397: Recurrence time is an increasing function of sensitive and aneuploid growth rates, not a decreasing function}%. Change RS2}
 
-\textcolor{red}{Recurrence time is a increasing function of sensitive and aneuploid growth rates not a decreasing function as stated in the manuscript. Change RS2}
+\textcolor{blue}{Line 134 typo: "anuploids"$\rightarrow$"aneuploids".} % Change RS3}
 
-\textcolor{red}{"anuploids"$\rightarrow$"aneuploids". Change RS3}
+\textcolor{blue}{Line 111 typo: "aneuplidies"$\rightarrow$"aneuploids".}% Change RS4}
 
-\textcolor{red}{"aneuplidies"$\rightarrow$"aneuploids". Change RS4}
+\textcolor{blue}{Line 300: We have corrected the interpretation of $\abs{\Delta_a}/u\lambda_a$.
+It is the ratio of the number of sensitive births required to produce an aneuploid, $1/u$, to the number of births in the resulting aneuploid lineage before it goes extinct, $\lambda_a / |\Delta_a|$.}% Change RS5}
 
-\bibliographystyle{plain}
-\bibliography{evo2022}
+\textcolor{blue}{We deleted the sentence "Drugs that significantly increase the division rate of sensitive cells, $\lambda_s$, but do not affect their death rate, $\mu_s$, decrease cancer recurrence times (conditioned on evolutionary rescue)" because, while it is mathematically true, it does not make sense for the drug to increase the fitness of sensitive cells.}% Change RS6}
+
+%\bibliographystyle{plain}
+%\bibliography{evo2022}
 \end{document}

### 2025-04-20 ms.tex
revised responses

@@
 One evolutionary mechanism by which a population of cancer cells can adapt to chemotherapy is aneuploidy. 
-Aneuploid cancer cells can be more fit in an environment altered by anti-cancer drugs, because aneuploidy disrupts the pathways usually targeted by the drugs. %change 3 %change S2
+Aneuploid cancer cells can be more fit in an environment altered by anti-cancer drugs, in part because aneuploidy may disrupt the pathways targeted by the drugs.
 Indeed, aneuploidy is highly prevalent in tumors, and some anti-cancer drugs fight cancer by increasing chromosomal instability.
@@
 adaptation by phenotypic plasticity without genetic modification \citep{carja2019evolutionary,carja2017evolutionary,levien2021non,gunnarsson2020understanding}; 
-and adaptation through genetic modifications, like mutation \citep{gomulkiewicz1995does,uecker2014evolutionary,uecker2016role,uecker2011fixation,orr2014population}. %change S5
+and adaptation through genetic modifications like mutation \citep{gomulkiewicz1995does,uecker2014evolutionary,uecker2016role,uecker2011fixation,orr2014population}. %change S5
 Here, we focus on the latter.
@@
 \citet{iwasa2004evolutionary} used a similar approach to find simple approximate expressions for the probability that a single lineage would evolve to survive via a wide range of possible mutational pathways.
-These results are essentially the same as those described later in the general context by \citet{osmond_genetic_2020} for the probability of evolutionary rescue by multiple mutations,
-which we mentioned in the previous paragraph. 
+These results are essentially the same as those described later in the general context by \citet{osmond_genetic_2020} for the probability of evolutionary rescue by multiple mutations, which we mentioned in the previous paragraph.
 A similar set of models describe adaptation to new environments (e.g., a pathogen evolving to cross a species barrier) and fitness valley crossing \citep{antia2003role, weissman2009rate}. % change S1
@@
 Because these results are scattered across multiple papers, and because they all follow from simple branching process approximations, we re-derive all results here for our model.
-While our focus is on the application of evolutionary rescue theory to the evolution of drug resistance in cancer, our model is a general one, and in the general evolutionary rescue context, this work can be seen as a characterization of the circumstances in which two-step processes contribute to rescue even when single-step rescue mutations are available, as in \citet{iwasa2003evolutionary, iwasa2004evolutionary} and \citet{osmond_genetic_2020}.
+While our focus is on the application of evolutionary rescue theory to the evolution of drug resistance in cancer, our model is a general one, and in the general evolutionary rescue context, this work can be seen as a characterization of the circumstances in which two-step processes contribute to rescue even when single-step rescue mutations are available, as in \citet{osmond_genetic_2020} and  \citet{iwasa2003evolutionary, iwasa2004evolutionary}.
 
@@
 We assume the population of cells is under a strong stress, such as drug therapy, and therefore $\Delta_s<0$, whereas the mutant is resistant to the stress, $\Delta_m>0$.
-We consider a range of possible values for $\Delta_a$, finding three distinct scenarios: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_s$ \citep[tolerant cells are able to survive the effects of the drug for an extended period of time compared to susceptible cells while resistant cells require a higher drug concentration to be affected, see][for a detailed explanation of the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing, stationary, or growing or dying slowly, that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$, in a sense that we will make precise below.
+We consider a range of possible values for $\Delta_a$, finding three distinct scenarios: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_s$; in the third, aneuploid cells are non-growing, stationary, or growing or dying slowly, that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$, in a sense that we will make precise below.
 % change S1 %change S17
+We define tolerant cells as those that survive drug exposure longer than susceptible cells, while resistant cells require higher drug concentrations to be affected.
+For details on these distinctions, see \citet{brauner2016distinguishing}.
 
@@
 
-Note that when we refer to drug-sensitive cells, we include those cells that have any aneuploidy that does not affect fitness in the presence of the drug (see \Cref{table1}). %change S7
+Note that when we refer to drug-sensitive cells, we include those cells that have any aneuploidy that does not affect fitness in the presence of the drug (\Cref{table1}). %change S7
 Additionally, we focus on mutations that confer resistance, neglecting deleterious mutations (which are more common).
@@
 \end{equation}
-Imporantly, when this threshold is smaller than 1, aneuploidy is expected to play a role in evolutionary rescue, as it reduces the threshold tumor size neeede for rescue. %change S8
+Imporantly, when this threshold is smaller than 1, aneuploidy is expected to play a role in evolutionary rescue, as it reduces the threshold tumor size needed for rescue. %change S8
 As expected, this ratio increases with the mutation rate $v$ and decreases with the aneuploidy rate $u$.
@@
 \paragraph*{Standing vs. \textit{de novo} genetic variation.}
-
-We assumed that at the onset of drug treatment the initial tumor consisted entirely of drug-sensitive cells but aneuploidy is likely produced even before the onset of treatment at some rate $\tilde{u}$, which may be lower in the absence of drugs, $\tilde{u} < u$ \citep{wang2019molecular,mason2017functional}.
+We initially assumed that at the onset of drug treatment, the tumor consisted entirely of drug-sensitive cells.
+In reality, aneuploidy is likely generated even before treatment begins, at some rate $\tilde{u}$, which may be lower in the absence of drugs, $\tilde{u} < u$ \citep{wang2019molecular,mason2017functional}.
 %change S9, S19
- Aneuploidy likely confers a fitness cost $c$ in the absence of drugs \citep{replogle2020aneuploidy,giam2015aneuploidy}. %change S16, S19 
-If the number of cells in the tumor $N$ is large (as expected if the tumor is treated with a drug), there may already be a fraction $f \approx \tilde{u}\lambda_s/c$ of aneuploid cells in the population (here we assume that the drug affects the sensitive death rate but not the division rate and therefore we use $\lambda_s$ for the sensitive division rate in the drug-free environment).
-%change S16, S19
+Aneuploidy is also thought to carry a fitness cost $c$ when drugs are absent \citep{replogle2020aneuploidy,giam2015aneuploidy}. %change S16, S19 
+Therefore, if the tumor size $N$ is large--as expected at the time of treatment--a fraction of cells, approximately $f \approx \tilde{u}\lambda_s/c$, may already be aneuploid.
+Here we assume that the drug affects the sensitive death rate but not the division rate and therefore we use $\lambda_s$ for the sensitive division rate in the drug-free environment. %change S16, S19
 
@@
 
-Using \cref{eq:N_a,eq:N_m,eq:Nsgv}, we can find the ratio of threshold tumor size for rescue via standing genetic variation  to the threshold for rescue via direct mutation,
+Using \cref{eq:N_a,eq:N_m,eq:Nsgv}, we can find the ratio of threshold tumor size for rescue via standing genetic variation  to the threshold for rescue via \textit{de novo} genetic variation,
 \begin{equation} \label{eq:N_sgv_ratio}
@@
 %change 2
-Of particular interest is the distribution of the evolutionary rescue time and recurrence time with tolerant aneuploid cells ($\Delta_a T^*\ll1$) given that this represents the two step evolutionary rescue scenario.
+Of particular interest is the distribution of the evolutionary rescue time and recurrence time with tolerant aneuploid cells ($\Delta_a T^*\ll1$) as it represents the two-step evolutionary rescue scenario.
 We focus on the parameter values in \Cref{table1} with $\lambda_a=0.0899$ (tolerant aneuploids), $\mu_s=0.14$ (intermediate sensitive death rate), and $v=10^{-7}$ (high end of the mutation rate).
@@
 Parameters from \citet{bozic2013evolutionary} consider patients with melanoma treated with the anti-cancer drug vemurfenib, in which resistance is conferred by trisomy in either Chr 2 or Chr 6.
-We have modified the parameters from \citet{bozic2013evolutionary} such that sensitive and mutant division rates are $\lambda_{s} = \lambda_{m} = \log{(2)}/T\approx0.1$ instead of their value of $0.14$ where $T$ is the doubling time in the absence of cellular death obtained from \citet{rew2000cell}.
+We have modified the parameters from \citet{bozic2013evolutionary} such that sensitive and mutant division rates are $\lambda_{s} = \lambda_{m} = \log{(2)}/T\approx0.1$ instead of their value of $0.14$, where $T$ is the doubling time in the absence of cellular death obtained from \citet{rew2000cell}.
 For a discussion of the different interpretations of the tumor doubling times see \citet{avanzini2019cancer}.
@@
 \end{subfigure}
-\caption{\textbf{Aneuploidy facilitates the evolutionary rescue of cancer under drug treatment.}
+\caption{\textbf{Aneuploidy facilitates the evolutionary rescue of tumors under drug treatment.}
 The probability of evolutionary rescue, $\presc$, as a function of the initial tumor size, $N$ (\cref{eq:rescue_prob}). 
@@
 \caption{
-\textbf{Standing genetic variation facilitates the evolutionary rescue of cancer.}
+\textbf{Standing genetic variation facilitates the evolutionary rescue of tumors under drug treatment.}
 (A)  Ratio of threshold tumor sizes for rescue by standing genetic variation and by \textit{de novo} variation, $\tilde{N}_a^*/N_a^*$, when a fraction $\frac{\tilde{u}\lambda_s}{c}$ is aneuploid at the start of treatment, as a function of the sensitive growth rate $\Delta_s$.
-Standing genetic variation will drive adaptation to the drug if the sensitive population is rapidly declining ($\Delta_s\ll0$) due to a stronger effect of the drug on sensitive cells.
+Standing genetic variation will drive drug resistance when the sensitive cell
+population is rapidly declining ($\Delta_s\ll0$) due to a stronger effect of the drug on sensitive cells.
 Red dots represent simulation results, and the error bars represent the $95\%$ confidence intervals obtained with bootstrap (Appendix G).
 Parameters: $\lambda_s=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_a=0.09,\mu_m=0.09,\tilde{u}=10^{-3},u=10^{-2}, v=10^{-7}$.
-(B) Ratio of threshold tumor size $\tilde{N}_a^*$, when a fraction $\frac{\tilde{u}\lambda_s}{c}$ is aneuploid at the start of treatment, and $N_a^*$ as a function of the ratio of aneuploidy rates $\tilde{u}/u$.
+(B) Ratio of threshold tumor size $\tilde{N}_a^*$ (when a fraction $\frac{\tilde{u}\lambda_s}{c}$ is aneuploid at the start of treatment) and $N_a^*$ as a function of the ratio of aneuploidy rates $\tilde{u}/u$.
 \textit{De novo} aneuploids will have a larger contribution to the appearance of drug resistance if the drug induces the appearance of aneuploid cells ($u \gg \tilde u$).
@@
 \end{subfigure}
-\caption{\textbf{Aneuploidy extends the window of opportunity for evolutionary rescue.}
+\caption{\textbf{Aneuploidy extends the window of opportunity for evolutionary rescue of tumors under drug treatment.}
 (A) The probability that a successful mutant has not appeared by time $t$.
@@
 To analyze evolutionary rescue in our model, we use the framework of \emph{multi-type branching processes} \citep{harris1963theory, weissman2009rate}. 
-In the case of cancer, where all reproduction is binary fission, these branching processes are also \emph{birth-death processes} \citep{kendall1948}% change 31
+In the case of cancer, where all reproduction is binary fission, these branching processes are also \emph{birth-death processes} \citep{kendall1948}. % change 31
 This allows us to find explicit expressions for the \emph{survival probability}: the probability that a lineage descended from a single cell does not become extinct.
@@
 \includegraphics[width=1\textwidth]{Figures/ExpectedNumber.pdf}
-\caption{\textbf{Aneuploidy increases the number of mutations which rescue the tumor.} Shown is the expected number of mutations, which will rescue the cancer cell population, produced through the evolutionary trajectory \textit{sensitive} $\rightarrow$ \textit{mutant} (blue line, \cref{eq:twosteplineagedirect}) or through the trajectory \textit{sensitive} $\rightarrow$ \textit{aneuploid} $\rightarrow$ \textit{mutant} (red line, \cref{eq:twosteplineage}).
+\caption{\textbf{Aneuploidy increases the number of rescue mutations.} Shown is the expected number of mutations, which will rescue the cancer cell population, produced through the evolutionary trajectory \textit{sensitive} $\rightarrow$ \textit{mutant} (blue line, \cref{eq:twosteplineagedirect}) or through the trajectory \textit{sensitive} $\rightarrow$ \textit{aneuploid} $\rightarrow$ \textit{mutant} (red line, \cref{eq:twosteplineage}).
 Dashed vertical red line represents the threshold tumor size above which evolutionary rescue is likely through aneuploidy \cref{eq:N_a} and the dashed vertical blue line represents the threshold tumor size above which evolutionary rescue is likely to occur through direct mutation \cref{eq:N_m}.
@@
 \includegraphics[width=1\textwidth]{Figures/EvolutionaryRescueTimeComplete.pdf} %change r5
-\caption{The mean time for appearance of a resistance mutation that leads to evolutionary rescue with aneuploidy ($u>0$). 
+\caption{\textbf{Tumor size reduces rescue time in a non-linear fashion.}
+The mean time for appearance of a resistance mutation that leads to evolutionary rescue with aneuploidy ($u>0$). 
 Solid lines show our approximations (green: \cref{limitapprox3}, blue: \cref{limitapprox4}, black: \cref{meantimet2}) compared to red markers that show mean of simulations results (with error bars for 95\% confidence intervalues obtained with bootstrap, see Appendix G).
@@
 \includegraphics[width=1\textwidth]{Figures/ProliferationTimeLarge.pdf} %change r5
-\caption{The mean time for the mutant cell population to reach size $N$, where $N$ is the initial number of cancer cells.
+\caption{\textbf{Tumor size reduces resucrrence time in a non-linear fashion.}
+The mean time for the mutant cell population to reach size $N$, where $N$ is the initial number of cancer cells.
 Solid lines show our approximations (green: \cref{eq:t2det}, blue: \cref{meanproliferationtime} with \cref{limitapprox4} for $\tau_a$; black: \cref{meanproliferationtime} with \cref{meantimet2} for $\tau_a$) compared to red markers that show mean of simulations results (with error bars for 95\% confidence intervals obtained with bootstrap, see Appendix G).
@@
 \includegraphics[width=1\textwidth]{Figures/RecurrencePlot.pdf} %change r5
-\caption{The mean time for the mutant cell population to reach size $M$, where $M$ is the tumor detection size.
+\caption{\textbf{Tumor size reduces time to detection in a non-linear fashion.}
+The mean time for the mutant cell population to reach size $M$, where $M$ is the tumor detection size.
 Solid lines show our approximations (green: \cref{eq:t2det2} for $N>N_m^*$, blue:  \cref{meanproliferationtime2} with $\tau_a$ from \cref{eq:AsymptoticTimeRules} for $N<N_a^*$; black: \cref{meanproliferationtime2} with $\tau_a$ from \cref{meantimet2}) compared to red markers that show mean of simulations results (with error bars for 95\% confidence intervals obtained with bootstrap, see Appendix G).
@@
 \includegraphics[width=1\textwidth]{Figures/SGVEvolutionaryRescueTimeComplete.pdf} %change r5
-\caption{The mean time for appearance of a resistance mutation that leads to evolutionary rescue with aneuploidy ($u>0$) when a fraction $f$ of cancer cells are aneuploid at the start of drug treatment.
+\caption{\textbf{Tumor size reduces rescue time in a non-linear fashion: with standing genetic variation.}
+The mean time for appearance of a resistance mutation that leads to evolutionary rescue with aneuploidy ($u>0$) when a fraction $f$ of cancer cells are aneuploid at the start of drug treatment. 
 Lines show our approximations (solid green: \cref{eq:AsymptoticTimeRules} for $N>N_m^*$; solid blue:  \cref{eq:AsymptoticTimeRules} for $N<N_a^*$; solid black: \cref{meantimet2}; dashed black: \cref{meantimet2SGV}) compared to red markers that show mean of simulations results (with error bars for 95\% confidence intervals obtained with bootstrap, see Appendix G).
@@
 \includegraphics[width=1\textwidth]{Figures/ProbvNPlot_ane.pdf}
-\caption{The probability of evolutionary rescue (i.e., the probability that the population does not become extinct), $\presc$, as a function of the initial tumor size, $N$ (\cref{eq:rescue_prob}).
+\caption{\textbf{Effect of tumor size on probability of evolutionary rescue is sensitive to the mutation rate in aneuploid cells.}
+The probability of evolutionary rescue (i.e., the probability that the population does not become extinct), $\presc$, as a function of the initial tumor size, $N$ (\cref{eq:rescue_prob}).
 Dashed vertical line shows the threshold tumor size, $N_a^*$, above which the probability is high (\cref{eq:N_a2}).
-Red dashed line: $v_a=10^{-7}$.
-blue line: $v_a=10^{-6}$.
-Green line: $v_a=10^{-5}$.
-Dots for simulations and the error bars for $95\%$ confidence interval ($p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the fraction of simulations in which the tumor has been rescued and $n=100$ is the number of simulations).
+Red dashed line: low mutation rate in aneuploid cells, $v_a=10^{-7}$.
+Blue line: interemediate mutation rate in aneuploid cellsm $v_a=10^{-6}$.
+Green line: high mutation rate in aneuploid cells, $v_a=10^{-5}$.
+Markers for simulation results, with error bars for $95\%$ confidence intervals ($p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the fraction of simulations in which the tumor has been rescued and $n=100$ is the number of simulations).
 Parameters: $\lambda_s=0.1,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v_s=10^{-7}$.} % change S1

### 2025-04-21 ResponseLetter2.tex
typos

@@
 
-\textcolor{blue}{We added (line 124): ``We define tolerant cells as those that survive drug exposure longer than susceptible cells, while resistant cells require higher drug concentrations to be affected.
+\textcolor{blue}{We added (line 124): ``We define tolerant cells as those that survive drug exposure longer than sensitive cells, while resistant cells require higher drug concentrations to be affected.
 For details on these distinctions, see Brauner et al.
@@
 
-15. Line 343, ". . .standing genetic variation. . ."
+15. Line 356, ". . .standing genetic variation. . ."
 \textcolor{blue}{Fixed. }% Change S10

### 2025-04-21 ms.tex
typos

@@
 Aneuploidy is caused by chromosomal instability and missegregation of chromosomes during mitosis. % change S4
-Importantly, changes in the number of chromosomes and chromosome arm copies allow cancer cells to survive under stressful conditions such as drug therapy \citep{lukow2021chromosomal,rutledge2016selective,ippolito2021gene}.
+Changes in the number of chromosomes and chromosome arm copies allow cancer cells to survive under stressful conditions such as drug therapy \citep{lukow2021chromosomal,rutledge2016selective,ippolito2021gene}.
 %change 4
@@
  \citet{ippolito2021gene} induced aneuploidy in cancer cell lines by exposing them to reversine, a small-molecule inhibitor of the mitotic kinase Mpsi1, and then to anti-cancer drugs such as vemurafenib.
-Reversine-treated cells had a higher proliferation rate following drug exposure than sensitive cancer cells, due to selection of specific beneficial karyotypes.
+Reversine-treated cells had a higher proliferation rate following drug exposure compared to sensitive cancer cells, due to selection of specific beneficial karyotypes.
 Similarly, \citet{lukow2021chromosomal} induced aneuploidy in cancer cells and observed that such cells have an advantage compared to sensitive cells during drug treatment despite having lower fitness before the onset of treatment.
 One proposed mechanism through which aneuploidy can confer resistance to anti-cancer drugs is by extending the length of the cell cycle, which prevents the drugs from damaging DNA and microtubules \citep{replogle2020aneuploidy}.
-Other mechanisms are the selection for specific karyotypes that lead to reduced drug metabolism \citep{ippolito2021gene} and elevated levels of DNA damage repair due to higher basal levels of DNA damage \citep{zerbib2023human}.
+Other mechanisms include altering the expression of drug-metabolizing genes \citep{ippolito2021gene} and enhancing DNA repair due to chronically elevated baseline levels of DNA damage in aneuploid cells \citep{zerbib2023human}.
 %change 34
 
-An essential aspect of aneuploidy is that the rate with which cells become aneuploid, that is, the rate of chromosome missegregation, is several orders of magnitude higher than mutation rate \citep{bakker2023predicting}.
+An essential aspect of aneuploidy is that the rate with which cells become aneuploid, that is, the rate of chromosome missegregation, is several orders of magnitude higher than the mutation rate \citep{bakker2023predicting}.
 A cell exposed to stress, such as chemotherapeutic drugs, can acquire aneuploidy faster than a mutation.
@@
 A similar set of models describe adaptation to new environments (e.g., a pathogen evolving to cross a species barrier) and fitness valley crossing \citep{antia2003role, weissman2009rate}. % change S1
-\citet{gunnarsson2020understanding} analyzed a model where a tumor consisting of two populations of cancer cells, one drug-resistant and the other drug-sensitive, can evade extinction by cells switching between the two phenotypes through epigenetic mutations.
+\citet{gunnarsson2020understanding} analyzed a model where a tumor consisting of two populations of cancer cells, one drug resistant and the other drug sensitive, can evade extinction by cells switching between the two phenotypes through epigenetic mutations.
 They found that even when drug-resistant cells are barely viable, the epimutations guarantee evolutionary rescue.
@@
 Here, we study evolutionary rescue after a sudden environmental change caused by initiating anti-cancer drug treatment. 
-We consider a range of effects of aneuploidy, from tolerance to (partial) resistance to the drug \citep{brauner2016distinguishing}; note that we use tolerant to refer to aneuploids which have negative growth rates, which is different from other references in the literature such as \citet{berman2020drug}.
+We consider a range of effects of aneuploidy, from tolerance to (partial) resistance to the drug \citep{brauner2016distinguishing}.
+Note that we use ``tolerance'' to refer to aneuploids that have negative growth rates, which is different from other references in the literature such as \citet{berman2020drug}.
 %change 47
@@
 When aneuploidy provides drug resistance, it can directly rescue the tumor.
-However, when aneuploidy provides tolerance to the drug, it may act as an evolutionary ``stepping stone'' or ``springboard'' \citep{Yona2015,martin2013probability,osmond_genetic_2020}, delaying extinction, and thereby allowing the tumor more time to acquire a resistance mutation on top of the aneuploid background.
+However, when aneuploidy provides tolerance to the drug, it may act as an evolutionary ``stepping stone'' or ``springboard'' \citep{Yona2015,martin2013probability,osmond_genetic_2020}, delaying extinction and thereby allowing the tumor more time to acquire a resistance mutation on top of the aneuploid background.
 As mentioned above, evidence suggests that aneuploidy may be a common strategy for tumor adaptation to drug therapy.
@@
 
-We will address these questions essentially by combining known results from the evolutionary rescue literature, and considering the quantities and parameter values that are relevant for cancer.
+We address these questions by combining known results from the evolutionary rescue literature and considering the quantities and parameter values that are relevant for cancer.
 Because these results are scattered across multiple papers, and because they all follow from simple branching process approximations, we re-derive all results here for our model.
@@
 % change S1 %change S17
-We define tolerant cells as those that survive drug exposure longer than susceptible cells, while resistant cells require higher drug concentrations to be affected.
+We define tolerant cells as those that survive drug exposure longer than sensitive cells, while resistant cells require higher drug concentrations to be affected.
 For details on these distinctions, see \citet{brauner2016distinguishing}.
@@
 Both aneuploid and sensitive cells may divide and mutate to become mutants at rates $v\lambda_{a}$ and $v\lambda_{s}$, respectively.
-To model standing genetic variation, we assume that before the onset of therapy, sensitive cells become aneuploid with rate $\tilde{u}\lambda_s$ (which may differ from $u \lambda_s$) and that aneuploidy confers a fitness cost $c$ in the drug-free environment, that is, we assume that aneuploid cells have an increased death rate compared to sensitive cells in a drug-free environment.
-We assume that $c$ is not too small (\Cref{table1}), so that standing variation is generated by a balance between the generation of aneuploids and selection against them, without the effect of genetic drift.
+To model standing genetic variation, we assume that before the onset of therapy, sensitive cells become aneuploid with rate $\tilde{u}\lambda_s$ (which may differ from $u \lambda_s$) and that aneuploidy confers a fitness cost $c$ in a drug-free environment, that is, we assume that aneuploid cells have an increased death rate compared to sensitive cells in a drug-free environment.
+We assume that $c$ is not too small, so that standing variation is generated by a balance between the generation of aneuploids and selection against them, without the effect of genetic drift.
 %change RS3
 
-See \Cref{figureAneuploidy} for a schematic representation of the model and \Cref{sampleTrajectories} for sample trajectories of the different genotypes.
+See \Cref{figureAneuploidy} for a schematic representation of the model, \Cref{sampleTrajectories} for sample trajectories of the different genotypes, and \Cref{table1} for a summary of model parameters.
 
@@
 For the remainder of this paper, we assume that the division rates of sensitive and aneuploid cells can be written as $\lambda_ss_t\left(1-u-v\right)\approx \lambda_ss_t$ and $\lambda_aa_t\left(1-v\right)\approx\lambda_aa_t$ because $u,v\ll1$ (\Cref{table1}).
-Each iteration of the simulation loop starts by computing the rates $\nu_k$ of each event $k$.
-We then draw the time until the next event, $\Delta t$, from an exponential distribution whose rate parameter is the sum of the rates of all events, such that $\Delta t \sim \text{Exp}(\sum_j \nu_j)$.
-Then, we randomly determine which event occurred, where the probability for event $k$ is $p_k = \nu_k/\sum_j \nu_j$.
-Finally, we update the number of cells of each genotype according to the event that occurred and update the time from $t$ to $t+\Delta t$.
+Each iteration of the simulation loop starts by computing the rates $\rho_k$ of each event $k$.
+We then draw the time until the next event, $\delta$, from an exponential distribution whose rate parameter is the sum of the rates of all events, such that $\delta \sim \text{Exp}(\sum_j \rho_j)$.
+Then, we randomly determine which event occurred, where the probability for event $k$ is $p_k = \rho_k/\sum_j \rho_j$.
+Finally, we update the number of cells of each genotype according to the event that occurred and update the time from $t$ to $t+\delta$.
 We repeat these iterations until either the population becomes extinct (the number of cells of all genotypes is zero) or the number of mutant cells is high enough so that their extinction probability is $<0.1\%$, that is, until
@@
 
-When simulations are slow (e.g., due to large population size) with runtimes in the order of days, we use $\tau$-leaping \citep{gillespie2001approximate}, where we assume that the change in the number of cells of genotype $k$ in a fixed time interval $\Delta t$ is Poisson distributed with mean $\nu_k \Delta t$.
+When simulations are slow (e.g., due to large population size) with runtimes in the order of days, we use $\tau$-leaping \citep{gillespie2001approximate}, where we assume that the change in the number of cells of genotype $k$ in a fixed time interval $\delta$ is Poisson distributed with mean $\rho_k \delta$.
 If the number of cells of genotype $k$ becomes negative, we change it to zero.
@@
 
-Note that when we refer to drug-sensitive cells, we include those cells that have any aneuploidy that does not affect fitness in the presence of the drug (\Cref{table1}). %change S7
+Note that when we refer to drug-sensitive cells, we include those cells that have any aneuploidy that does not increase fitness in the presence of the drug (\Cref{table1}). %change S7
 Additionally, we focus on mutations that confer resistance, neglecting deleterious mutations (which are more common).
@@
 
-We can categorize tumors by their size: small tumors with size $N<N_a^*$ that are unlikely to survive treatment, intermediate tumors with size $N_a^* < N < N_m^*$ that rely on aneuploidy for evolutionary rescue, and large tumors with size $N > N_m^*$ that could overcome the effect of drug treatment without aneuploidy.
+We can categorize tumors by their size: small tumors with size $N<N_a^*$ that are unlikely to survive treatment, intermediate tumors with size $N_a^* < N < N_m^*$ that rely on aneuploidy for evolutionary rescue, and large tumors with size $N > N_m^*$ that could overcome drug treatment without aneuploidy.
 For the parameter values in \Cref{table1} with $\lambda_a=0.0899,\mu_s=0.14, u=10^{-2}, v=10^{-7}$, we are in the tolerant aneuploid scenario, and substituting in \cref{eq:N_a,eq:N_m}, we have $N_a^* \approx 4 \times 10^6$ and $N_m^* \approx 4 \times 10^7$.
@@
 Interestingly, the threshold between small and intermediate tumors, $N_a^*$, is similar to the tumor detection threshold of $4.19 \times 10^6$ cells for a wide variety of tumors~\citep{avanzini2019cancer}.
-We note that vemurafenib-treated melanomas (i.e.
-melanomas with sizes above the detection threshold) have a probability $>50\%$ to relapse \citep{piejko2023long,handa2022long}.
+We note that vemurafenib-treated melanomas (i.e., melanomas with sizes above the detection threshold) have a probability $>50\%$ to relapse \citep{piejko2023long,handa2022long}.
 %change 58
@@
 Aneuploidy may lead to an increased mutation rate in cancer cells \citep{garribba2023short,passerini2016presence,janssen2011chromosome}.
-Thus, we extended our model to account for this in Appendix \ref{sec:appendix-diff-mut}.
+Thus, we extended our model to account for this in Appendix~\ref{sec:appendix-diff-mut}.
 We find that increasing the mutation rate in aneuploid cells by one order of magnitude leads to a decrease in the threshold tumor size of approximately one order of magnitude.
@@
 
-Using \cref{eq:N_a,eq:N_m,eq:Nsgv}, we can find the ratio of threshold tumor size for rescue via standing genetic variation  to the threshold for rescue via \textit{de novo} genetic variation,
+Using \cref{eq:N_a,eq:N_m,eq:Nsgv}, we can find the ratio of threshold tumor size for rescue via standing genetic variation  to the threshold for rescue via mutation,
 \begin{equation} \label{eq:N_sgv_ratio}
@@
 \end{equation}
-Evolutionary rescue through direct mutation is more likely if the cost of aneuploidy, $c$, is large or the effect of the drug $\Delta_s$ is small.
+Evolutionary rescue through direct mutation is more likely if the cost of aneuploidy, $c$, is large or the effect of the drug, $\Delta_s$, is small.
 In contrast, standing genetic variation will drive adaptation if the pre-treatment chromosome missegregation rate, $\tilde{u}$, is large.
@@
 It grows logarithmically with tumor size $N$ and may be the same order of magnitude as the mean evolutionary rescue time. 
-Increasing mutant $\Delta_m$ growth rates decreases the recurrence time, while increasing the sensitive and aneuploid growth rates, $\Delta_s$ and $\Delta_a$, respectively, increases the recurrence time.
+Increasing the mutant growth rate, $\Delta_m$, decreases the recurrence time, while increasing the sensitive and aneuploid growth rates, $\Delta_s$ and $\Delta_a$, respectively, increases the recurrence time.
 
